### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/02-Gasentladungen/xpdp1/doc/xpdp1_manual.tex
Suttrop - Plasmaphysics

\documentstyle [12pt, epsf] {article}

\topmargin 0.00in
\oddsidemargin 0.25in
\textheight 9.00in
\textwidth 6.00in
\pagestyle{plain}
\renewcommand{\baselinestretch}{1.4}

\begin{document}
\title{
PDP1\\ Plasma Device 1 Dimensional Bounded Electrostatic Code
}
\author{
Plasma Theory and Simulation Group\\
EECS Department,
University of California,\\ Berkeley, CA 94720}

\date{}

%
\begin{figure}
\begin{center}
\leavevmode
\hbox{%
\epsfysize 4.0in
\epsfxsize 4.0in
\epsffile{pdp1.epsi}}
\end{center}
\end{figure}
%
\maketitle
\newpage
\tableofcontents
\newpage

\begin{section}
{\bf INTRODUCTION}

XPDP1 is a bounded electrostatic code for simulating 1 dimensional Plasma
Devices running on Unix workstations with X-Windows, and PC's with an X-Windows
emulator.  The code simulates a bounded plasma plus external circuit system
whose characteristics, including particles and electrostatic fields, are
specified by the user at run time using an input file.  The external circuit
includes R, L, and C elements, as well as AC, DC, and ramped current and
voltage sources.  The simulation proceeds in real-time, with the user viewing
output as the code runs in the form of various user specified diagnostics which
are updated each time step (animation).
\vspace{0.1in}

\noindent
The code compiles with standard C compilers and requires X-Windows libraries
(X10 or higher).  This version of PDP1 differs from the Cray FORTRAN version
(PDW1 - Plasma Device Workshop) originated by W. S. Lawson on several physics
issues summarized below.  The X-Windows version provides interactive animation
whereas the FORTRAN version produces output files requiring post processing.

\begin{subsection}
{\bf Scope}

   This document describes the XPDP1 programs running on the workstations and
   UNICOS Cray environment.  The general physics issues involved in a bounded
   plasma simulation are discussed briefly.  Program installation, operation,
   and modification are discussed.  In addition, the library of input files
   accompanying XPDP1 is described, and the guidelines to generate new input
   files are provided.
\vspace{.2in}

\noindent
   This manual makes no attempt to explain exhaustively the physics and
   computational issues of particle simulation.  Many comprehensive texts on
   particle simulation are available \cite{kn:bird}\cite{kn:hock}.
   For information on the Cray FORTRAN
   version of PDW1, see W. S. Lawson \cite{kn:law}.
\vspace{.2in}

\noindent
   Some familiarity with plasma physics is required to understand the results
   of the simulations and generate new simulations.  Knowledge of numerical
   analysis and/or particle simulation is useful for modification of the code
   and understanding of numerical errors which can occur in any computer
   simulation.

\end{subsection}

\begin{subsection}
{\bf Objectives}
   The objectives of this package include distribution of the $XPDP1$ codes in
   a
   form which makes it accessible to large and small research facilities,
   universities, and individuals interested in plasma simulation.  The Unix and
   X-Windows environments provide a vehicle for distribution since nearly every
   researcher and student has access to this type of machine.  The trend in
   recent years has been one of decreasing cost and increasing performance and
   capability.
\end{subsection}

\begin{subsection}
{\bf History}
   \mbox{PDW1} was originally written by W. S. Lawson for a plasma device seminar on
   axially bounded plasma systems at the University of California at Berkeley
   in 1983. We will discuss
   differences with the Cray \mbox{FORTRAN} version below as appropriate.
\vspace{.2in}

\noindent
   The \mbox{FORTRAN} code was translated into \mbox{Cray-C} by T. L.
   Crystal in 1987.
   Crystal made simplifications in the external circuit as well as a number of
   programming-related changes.  \mbox{PDP1} has significant differences 
   from these
   previous versions of \mbox{PDW1}, containing more accurate and 
   versatile circuit
   solvers, simultaneous solution of the circuit and potential equation coupled
   through boundary conditions at the plasma wall, a basic ionization model,
   MKS units, and Fourier transform diagnostics as well as many additional
   diagnostics and tools.
\end{subsection}

\begin{subsection}
{\bf Enhancements}
   This section summarizes the enhancements made to the \mbox{PC}
   version of \mbox{PDP1}
   since the original version (\mbox{PDW1}) written in \mbox{FORTRAN}
   by Lawson.  The
   enhancements are presented in the order they occurred, referenced by version
   number.  Note that the original version has no version number, and the
   \mbox{Cray-C} version due to Crystal has been given version number 
   \mbox{PDW1-1.0}
   although it had no explicit version number.

\begin{subsubsection}
{\bf Version \mbox{PDW1-2.0}}
      The code was translated into Microsoft C 5.~1 running under
      \mbox{MS-DOS} on \mbox{IBM}
      \mbox{PC} compatible microcomputers.  The diagnostics code was replaced with
      real time graphics animation in \mbox{EGA} resolution (640x350 pixels, 16
      colors).  The code had limitations on particles, diagnostics, etc., and
      the user interface consisted of mnemonic keys to display snapshots of
      diagnostics.  The code could only run while displaying phase space, and
      had a maximum time-step limitation.
\vspace{.2in}

\noindent
      Version \mbox{PDW1-2.0} was never officially released due to the rapid
      development of \mbox{PDP1} described below.
\end{subsubsection}

\begin{subsubsection}
{\bf Version 1.0}

      Version 1.0 is a major overhaul providing PDP1 with an improved user
      interface and output capabilities.  The core of the user interface,
      designed and written by J. P. Verboncoeur and V. Vahedi\cite{kn:verb}, 
      handles the
      programming issues of the simulation.  These include keyboard handling,
      screen graphics, and printer output for Postscript and IBM Graphics
      printers (including the Epson FX compatible family of printers) as well
      as CGM (Computer Graphics Metafile) output for editing by many drawing
      programs.  The code is implemented in an object-oriented style in
      standard C to facilitate enhancements, including new menu items and
      diagnostics.
%
\begin{figure}
\begin{center}
\leavevmode
\hbox{%
\epsfysize 4.0in
\epsfxsize 4.0in
\epsffile{f1.epsi}}
\end{center}
\caption{Schematic representation of the interaction between WinGraphics and the physics kernel.}
\label{PICFLOW}
\end{figure}
%
\noindent
      The code is separated into a physics application and the windowing core
      as shown in Figure 1.  New physics and diagnostics can be added without
      altering the windowing code, with the only restriction that any new
      diagnostic must be a linear, semi-log, or scatter plot.  A text plot is
      currently under consideration which would display parameters from the
      input file during the simulation.
\vspace{.2in}

\noindent
      Using the windowing core, all diagnostics are updated dynamically in
      time.  The core can also update in individual time-steps, pausing for a
      keystroke before continuing the simulation.
\vspace{.2in}

\noindent
      The code runs significantly faster when fewer (or zero) plots are
      displayed on screen.  Therefore, when output is not being viewed (long
      runs overnight, coffee break, etc.), it is more efficient to remove all
      plots from the screen.
\vspace{.2in}

\noindent
      PDP1 no longer has a time step limitation; it can run indefinitely.  All
      time histories are combed periodically such that there are never more
      than HISTMAX (a programmable constant) values stored.  Note that after
      long runs this can result in a loss of high frequency resolution on the
      time history plots.  This has no effect on the physics of the simulation
      since the diagnostics are simply a view of the physical results.
\vspace{.2in}

\noindent
      PDP1 can now generate hard copy output on an IBM Graphics Printer or
      compatible, including the Epson FX series of dot matrix printers.  PDP1
      can generate a CGM format file for further editing by many drawing
      packages which support ANSI standard CGM and for inclusion in many word
      processors and desktop publishing programs supporting CGM.  In addition
      PDP1 can generate a Postscript file which can be sent to any Postscript
      printer or included in a Postscript document.  The resolution and
      grey-shading of the CGM and Postscript output is dependent only on the
      output device, while the dot matrix output is at the screen resolution in
      monochrome.
\vspace{.2in}

\noindent
      PDP1 now employs four circuit solvers to handle the full range of
      external circuit parameters.  The general voltage-driven series RLC case
      is solved using a second order backward Euler method.  For the open
      circuit case, $C \rightarrow 0$, the external circuit no longer needs to be
      solved; the boundary surface charge influences the potential as always,
      but it cannot exchange charge (via current) with the other boundary.  For
      $C \rightarrow \infty$ and $R = L = 0$, the external circuit becomes a
      short circuit, so the applied potentials are applied directly to the
      plasma boundaries.  The final case is an ideal current source, which
      imposes the specified current independent of the external circuit
      parameters or boundary potentials.  For a complete discussion of the
      simultaneous solution of the external circuit and the spatial plasma
      potential, refer to the attached paper\cite{kn:alves}.
\vspace{.2in}

\noindent
      FFT routines are used to transform time histories into frequency domain
      plots.  Both amplitude and phase of the transformed quantity are
      available.  This feature is useful when looking for response of a bounded
      plasma at one or more frequencies, as well as determining the impedance
      of the plasma to external currents/voltages.
\end{subsubsection}

\begin{subsubsection}
{\bf Version 2.0}

      Version 2.0 supports a number of enhancements, including a Monte Carlo
      collisional model for electron- and ion-neutral collisions.  Dump files
      are now supported.   The current simulation state can be saved to a file
      for later restart.  This is useful for saving long runs toward
      equilibrium, and changing a few parameters and restarting close to the
      new equilibrium.

\vspace{.2in}

\noindent
      In addition, the right boundary of the simulation is now assumed to be
      the zero reference potential in PDP1.  In PDC1 and PDS1, the left
      boundary represents the inner (driven) electrode, while the right
      boundary is the zero reference potential as in PDP1.  A flag has been
      added for refluxing at the right boundary as an alternative to the
      reference potential; refluxing is explained in detail in 4.1.3 Flags.
\end{subsubsection}

\begin{subsubsection}
{\bf Version 2.1}

      An error was discovered in the Monte Carlo collisions (MCC) between
      charged particles and neutrals in calculating the scattered velocity from
      the incident velocity.  In this calculation, one must rotate the velocity
      vector and adjust its magnitude accordingly.  In the rotation part, a set
      of orthonormal vectors is needed to set up a coordinate system to rotate
      the incident velocity vector by theta and phi to the scattered velocity
      vector.  This set of orthonormal vectors was NOT calculated properly
      (they were orthogonal to each other but NOT normal). The vector set that
      we used was 
[
$\vec{v}, (\vec{v} \, {\bf \times} \, \vec{y})$, 
$\vec{v} {\bf \times} (\vec{v} \, {\bf \times} \, \vec{y})$
]  
which should have been 
[
$\vec{v} / |\vec {v}|$, 
$(\vec{v} \, {\bf \times} \, \vec{y}) / |\vec{v} \, {\bf \times} \, \vec{y}|$,
$(\vec{v} \, {\bf \times} \, (\vec{v} \, {\bf \times} \, \vec{y})) / 
|\vec{v} \, {\bf \times} \, (\vec{v} \, {\bf \times} \, \vec{y})|$  
].
	If the vector set is not normalized, the magnitude of the
	scattered velocity
      is also affected through the rotation.  In fact the scattered velocity
      will be smaller than the incident velocity.  This would act as a cooling
      mechanism for the particles, and the system would loose energy.
\vspace{.2in}

\noindent
      In addition a digital smoothing filter (1, 2, 1) is added to the planar
      case (PDP1) which requires an flag in the input file (see the list of
      input parameters).
\end{subsubsection}

\begin{subsubsection}
{\bf Version 3.0 (XPDP1)}

      XPDP1 is the X-Windows version of XPDP1 running under X11.  This version
      makes extensive use of a mouse for selecting menus, moving windows,
      resizing, etc.  In addition, XPDP1 can be run on a remote host while the
      output displays on a local graphics workstation.  XPDP1 uses XGrafix for
      the graphics display, which requires X11 libraries to compile.
\end{subsubsection}

\begin{subsubsection}
{\bf Version 3.1 (XPDP1)}

      In the previous versions, the code would allocate position and velocity
      arrays for each species based on a parameter called MAXLEN (defined
      previously in the header file) which was the maximum number of particles
      for all species.  In that case, the user would have had to recompile the
      whole code if more particles were desired.
\vspace{.2in}

\noindent
      The input files in this version have an extra parameter per species
      (called max-np) for the maximum number of particles of that species.
      This avoids the complications of recompiling and at the same time allows
      the user to define variable array sizes per species.
\end{subsubsection}

\begin{subsubsection}
{\bf Version 4.0 (XPDP1)}

      Xgrafix has been updated to use Tcl/Tk libraries for dialog boxes. 
      The installation of Xgrafix now requires that these libraries be
      also installed.
\vspace{.2in}

\noindent
      The input files in this version has been many new parameters. The 
      major one being {\bf GAS}. This parameter picks out one of the four
      gases built in to PDP1 along with the respective collisions. As a 
      result of this, the collision part of the input file is obsolete
      and no longer required. This subsection is explained in later sections
      for continuity and for help in understanding collision implementation.
      The gases are enumerated below.
\begin {itemize}
\item	GAS = 1: Helium
\item	GAS = 2: Argon
\item	GAS = 3: Neon 
\item	GAS = 4: Oxygen \footnote{Gas = Oxygen requires that {\em nsp} = 3}
\end {itemize}
	In addition to this, an implicit mover {\em imp}, volume source 
{\em psource} and sub cycling {\em k} have also been implemented.

\end{subsubsection}

\begin{subsubsection}
{\bf Version 4.1 (XPDP1)}

\noindent
	A second order accurate injection push has been implemented.  Also all
	components of velocity can be specified independently.  The inversion
	of the distribution functions is now second order accurate.
	The gas package now includes a fifth option (GAS = 5) which allows
	the use of the old cross section parameters (which are explained
	in section 4.1.5-6).  These parameters need to be added to mcc.cpp 
	before compilation.

\end{subsubsection}

\end{subsection}
\end{section}
\newpage
\begin{section}
{\bf INSTALLATION}

This section describes the contents of the \mbox{XPDP1} distribution 
and its installation procedure.

\begin{subsection}
{\bf Distribution Format}

   XPDP1 is no longer distributed on floppy diskettes. Instead the preferred
distribution procedure is through anonymous FTP. The codes are available
in the tar gzipped form from the {\bf PTSG} web site at 
{\em http://ptsg.eecs.berkeley.edu}. {\bf xgrafix} distribution is available
separately at the same web site and has to be installed for any of the PTSG
family of codes to work. The contents of the XPDP1 distribution 
contains:

\begin{tabbing}
	XPDP1.TARaaaaa  \= Text file containing the direction for installing 
			\mbox{XPDP1} on a Unix \= \kill
        README	\> Text file containing the direction for installing 
			\mbox{XPDP1} on a Unix \> \\
	        \> platform. \> \\
				\\
	XPDP1.TAR  \> This tar file contains the files required 
			for \mbox{XPDP1}. \>
\end{tabbing}

\noindent
\verb1\1{\bf xpdp1} directory:

\begin{tabbing}
        READMEaaaaa\=     All files with the .c extension are the C language
                       source files for \= \kill
        README   \>    Text file containing information about licensing. \> \\
								 	    \\
	*.c      \>    All files with the .c extension are the C language
                       source files for \> \\
		 \>    \mbox{XPDP1}.  These files should be
                       placed in the \mbox{xpdp1} directory. \> \\
								\\
        *.h      \>    All files with the .h extension are the 
			C language header files for \> \\
		 \>    \mbox{XPDP1}.  These files should be
                       placed in the \mbox{xpdp1} directory. \> \\
								\\
        makefile \>    The make file for automatically performing
                       conditional compilation \> \\
		 \>    /linking of only those files
                       which have been changed.  This file \> \\
		 \>    should be placed in the \mbox{xpdp1} directory. \> 
\end{tabbing}

\noindent
\verb1\1{\bf xpdp1}\verb1\1{\bf inp} directory:

\begin{tabbing}
        *.inpaaaaa\=      All files with the inp extension are input files.
                       For detailed information on \= \kill
        *.inp   \>      All files with the inp extension are input files.
                       For detailed information on \> \\
	        \>      each input file refer
                       to Section 4.2 Input File Library.  \= \\
	        \>      A directory called inp is set up
                       under the xpdp1 directory to include all the input
                       files. \> 
\end{tabbing}

\noindent
\verb1\1{\bf xpdp1}\verb1\1{\bf doc} directory:

\begin{tabbing}
        xpdp1.txtaaaaa\=   \mbox{ASCII} file containing the documentation. \= \kill
        xpdp1.ps  \>   \mbox{PostScript} file containing the documentation. \> \\
									\\
        xpdp1.txt \>   \mbox{ASCII} file containing the documentation. \>
\end{tabbing}

\noindent
The {\bf xgrafix} distribution  contains the required files 
for displaying graphics   in the X Windows version.  These are:

\begin{tabbing}
        xgrafix.sd*aaaaa\= The \mbox{XGrafix} icon (bitmap).  This file should be
                       placed in the \mbox{xgrafix} \= \kill

        README     \>  Text file containing some directions and
                       information on compiling \> \\
		   \>  the \mbox{XGrafix} libraries.
                       When the information in this manual conflicts \> \\
		   \>  with
                       the \mbox{README} file, assume the file is correct. \> \\
									\\
        xgrafix.c  \>  The source file for the \mbox{XGrafix} 
				graphics display
                       library.  This file \> \\
		   \>  should be placed in the 
			\mbox{xgrafix} directory. \> \\
							\\
        xgrafix.h  \>  Header file for \mbox{XGrafix}.  This file should be
                       placed in the \mbox{xgrafix} \> \\
		   \>  directory. \> \\
						\\
        xgrafix.ico \> The \mbox{XGrafix} icon (bitmap).  This file should be
                       placed in the \mbox{xgrafix} \> \\
		    \> directory. \> \\
							\\
        xgrafix.str \> Another header file containing string definitions
                       for XGrafix.  This \> \\
		    \> file should be placed in the xgrafix directory. \> \\
								\\
        makefile    \> The make file for \mbox{XGrafix}.  This file should be
                       placed in the \mbox{xgrafix} directory. \> 
\end{tabbing}

\end{subsection}

\begin{subsection}
{\bf Setup and Installation Procedure (X Windows version)}

   The installation procedure to a workstation must be done manually.  Take the
   tar file along with the README file to your workstation or Unix platform.
   Follow the directions in the README file for installation and compiling.
   Read the README file in the xgrafix directory as well as the makefile before
   compiling XGrafix, then compile xpdp1.
\end{subsection}
\end{section}
\newpage
\begin{section}		% begin 3
{\bf X-WINDOWS PROGRAM OPERATION}

The X-Windows version of PDP1, XPDP1, is operated in the same manner as
discussed in the manual for the PC version, with the exceptions noted below.

\begin{subsection}
{\bf Syntax}

   xpdp1 -i filename[.inp] -d [dumpfile.dmp]
 \vspace{.2in}

\noindent
   where $<filename.inp>$ is the name of the input file.  Although we have used
   *.inp for the input files in the library, the .INP extension is not
   required.  If no filename is provided on the command line, XPDP1 displays an
   error message.  The dumpfile parameter is optional; it must be an existing
   file created by the same version of the code.  If the input files are not in
   the same directory or are located in a sub-directory, the path must also be
   specified.  For instance, the syntax for starting XPDP1 with the input file
   vc.inp which is in a sub-directory of xpdp1 called INP is:
\vspace{.2in}

\noindent
   xpdp1 -1 inp/vc[.inp]
\vspace{.2in}

\noindent
   The input file is required since XPDP1 determines the parameters of the
   simulation at run time.
\end{subsection}

\begin{subsection}
{\bf GUI Support}

   XPDP1 fully supports a mouse for selection of items, buttons etc.  Moving,
   resizing, and iconifying of windows is supported indirectly via the X window
   manager (Motif, Open~Look, etc.).  Keystrokes are not supported for these
   actions, so a mouse is required.  The move, resize, and iconifying buttons
   and operations are governed by the window manager; consult the window
   manager manual or guru for details of these procedures.
\end{subsection}

\begin{subsection}
{\bf Main Menu}

   The buttons on the main menu can be selected using the mouse.  The functions
   available include RUN, STOP, STEP, SAVE, and QUIT, which all perform the
   same function described previously in Section 3.  Note that the SAVE
   function is equivalent to the DUMP function in the MS-DOS version which is
   also NOT implemented in this version.
\end{subsection}

\begin{subsection}
{\bf Diagnostic Window Buttons}

Every diagnostic window in XPDP1 contains four buttons: Rescale, 
Trace, Print, and Cross-hair.

\begin{subsubsection}	% begin 3.4.1
{\bf Rescale}

      The rescale button pauses the simulation and opens a dialog box
      containing editable fields for the minimum and maximum labels on the x
      and y axes.  In addition, the dialog box contains buttons for automatic
      rescaling of the x and y axis.  These buttons toggle auto rescaling of the
      respective axis on and off.  When all axes are scaled as desired, select
      OK to accept the changes or CANCEL to return to the previous status.
      Note that while rescaling the simulation is paused.
\end{subsubsection}	% end 3.4.1

\begin{subsubsection}	% begin 3.4.2
{\bf Trace}

      The trace button turns toggles the plot tracing feature on and off.  The
      previous plots are accumulated, generating a series of lines or dots as
      described above.
\end{subsubsection}	% end 3.4.2

\begin{subsubsection}	% begin 3.4.3
{\bf Print}

      The Print button generates a PostScript plot file of the current window.
      Pressing the button opens a dialog box containing the file name for the
      plot and a plot title.  Selecting OK generates the plot, CANCEL returns
      to the simulation.  Note that the simulation is paused while the dialog
      box is open.
\end{subsubsection}	% end 3.4.3

\begin{subsubsection}	% begin 3.4.4
{\bf Crosshair}

      The crosshair button activates the crosshair pointer and opens a dialog
      box displaying the coordinates of the pointer.  To display the
      coordinates of a point move the crosshair pointer to the desired location
      and click.  The simulation is paused until the crosshair is deactivated
      by selecting the Crosshair button again.
\end{subsubsection}	% end 3.4.4

\end{subsection}	% end 3.4

\begin{subsection}	% begin 3.5
{\bf Diagnostics}

   In XPDP1 diagnostics there is no diagnostic menu list.  Instead, all
   diagnostics appear as icons at the bottom of the display or in the
   designated icon area depending on the window manager.  To open a diagnostic,
   simply click on its icon.  In addition, some window managers will display a
   list of the available diagnostics with all other open windows in a window
   list.  Note that this is not a feature of XPDP1, but rather a feature of the
   X-Windows Manager used on the system.

\end{subsection}	% end 3.5
\end{section}		% end 3

\newpage
\begin{section}		% begin 4
{\bf INPUT FILES}

XPDP1 obtains its versatility through the use of input files.  The input file
contains the parameters for the simulation, specifying number of each species,
grid spacing, charge to mass ratios, etc.  This section describes the contents,
use, and modification of input files for XPDP1.

\begin{subsection}	% begin 4.1
{\bf Input File Parameters}

   The codes use input files to describe the simulation, including the physical
   bounded device parameters, external circuit, RF drive, etc. (global
   parameters), as well as the parameters describing each species of
   particles.  Units, if any, are shown in [ ].

 \begin{subsubsection}
{\bf Global Parameters}
\begin{tabbing}
      epsilonraaaaa\=    The number of particle species to simulate (0= no species
                    present, may \= \kill
     nsp    \>       The number of particle species to simulate (0= no species
                    present, may \> \\
	    \>      use this option to check the system, 1= one
                    species in the whole system, \> \\
	    \>      etc.).  If modifying an input
                    file that has, say, 2 species, to add more \> \\
	    \>      species, just
                    copy one of the blocks of parameters corresponding to
                    species \> \\
	    \> 	    1 or 2, and change the parameters to the desired
                    values.  Note that each \> \\
	    \>      species added seeks another 100
                    kByte of memory. \> \\
					\\
      nc    \>      The number of spatial cells.  The cell width is calculated using \> \\
       	    \>      {\em$\Delta x = length / nc$}. \> \\
							\\
      nc2p  \>      The number of physical particles per computer particle.
                    The number of \> \\
	    \>      super particles in the simulation is found
                    using \> \\
	    \> 	    {\em$N = initn \, \cdot \, area \, \cdot \, length\, /nc2p $}
			, where
                    {\em initn} is the uniform number density. \> \\
							\\
      dt    \>      The time step [sec]. \> \\
						\\
      length \>     The length of the system (distance between electrodes) [m]. \> \\
									\\
      area   \>     Electrode area [{\rm m$^{2}$}]. Allows application of real currents
                    and real external \> \\
	     \>     circuit parameters. \> \\
						\\
      epsilonr \>   Background relative dielectric constant of system. \> \\
							\\
      B   \>        Applied homogeneous magnetic field [Tesla]. \> \\
							\\
      PSI \>        Angle the B-field makes with the normal from the electrode,
                    x-axis \> \\
	  \>        (B-field is in the x-z plane) [deg]. \> \\
							\\
      rhoback \>    Fixed background charge density (non-accelerating) [C/{\rm m$^{3}$}]. \> \\
							\\
      backj   \>    Background current density (non-accelerating) [Amps/{\rm m$^{2}$}]. \> \\

      dde  \>       Sinusoidal perturbation of charge density
                    {\em$(\delta x / l)$} 
			at 
		   	{\em$t = 0$}; \> \\
	   \>       {\em$\delta x (x) = $l$ \, \cdot \, dde \, \cdot \, \sin(2 \pi x/l) $} \> \\
										\\
      extR \>       External circuit resistance [Ohms]. \> \\
								\\
      extL  \>      External circuit inductance [Henries]. \> \\
								\\
      extC  \>      External circuit capacitance [Farads]. \> \\
									\\
      q0   \>       Initial capacitor charge [C]. \> 
\end{tabbing}

\end{subsubsection}

\begin{subsubsection}
{\bf Applied Voltage Or Current Sources}

      When the flag dcramped is off, the general form of the applied source is:

\begin{center}
{\em$S(t) = DC + Ramp \, \cdot \, t + AC \, \cdot \, \sin(2\pi f_{0}t+ \theta_{0})$} 
\end{center}

\noindent
where {\em$S(t)$}, the applied source, is either a current or a voltage source.
\vspace{.2in}

\noindent
      The flag dcramped should be turned on (set to 1) when a step function is
      desired.  The step function can have a zero rise time, $Ramp \gg 1$,
      or can be ramped to its final DC value either with a constant slope or
      sinusoidally as shown in Figure 2. The rise time of the applied signal is
      equal to 1/2f0, in the sinusoidal case, or DC/Ramp, in the constant
      linear slope case.
%
\begin{figure}
\begin{center}
\leavevmode
\hbox{%
\epsfysize 4.0in
\epsfxsize 4.0in
\epsffile{f2.epsi}}
\end{center}
\caption{ When the flag dcramped is set, the signal is ramped to its
final DC value sinusoidally (left), or with a constant slope (right).}
\label{SINRAMP}
\end{figure}
%
\noindent
\begin{tabbing}
      dcrampedaaaaa\=  Flag for ramping external voltage/current source to a final
                    DC value \= \kill
      source    \>  Voltage or current source indicator (v=voltage, i=current). \> \\
						\\
      dcramped  \>  Flag for ramping external voltage/current source to a final
                    DC value \> \\
	        \>  (1=yes, 0=no). \> \\
						\\
      DC        \>  DC voltage or current source [V or Amps].  Zero value
                    indicates zero dc \> \\
		\>  voltage. \> \\
						\\
      Ramp      \>  Rate of ramping for voltage or current source [V/sec or
                    Amps/sec].  Zero \> \\
	        \>  value indicates zero ramping for voltage. \> \\
					\\
      AC        \>  AC voltage or current source [V or Amps]. Zero value
                    indicates zero ac \> \\
		\>  voltage and the values of {\em $f0$} and {\em $theta0$}
                    are ignored. \> \\
					\\
      f0        \>  AC source driving frequency [Hz]. \> \\
					\\
      theta0     \>  Initial phase angle of AC source [deg]. \>
\end{tabbing}
\end{subsubsection}

\begin{subsubsection}
{\bf Flags}

\begin{tabbing}
      e\_collisionalaaaaa\= The flag for ionization, elastic, and excitation
                    electron-neutral collisions \= \kill
      secondary \>  Secondary electron emission flag (0=off, 1=species 1
                    emitted, etc.).  \> \\
		\>    The emitted electron species give the
                    emitted velocity distribution at the \> \\
		\>    surface specified for
                    the species (see SPECIES PARAMETERS). \> \\
							\\
      e\_collisional \> The flag for ionization, elastic, and excitation
                    electron-neutral collisions \> \\
		\>  (0 = off, 1 = species 1 is the
                    colliding electron species, etc.). \> \\
                \>  Note: Only ONE species can be the colliding electron
                    species. \> \\
								\\
      i\_collisional \> The flag for scattering and charge exchange ion-neutral
                    collisions (0 = off, \> \\
		\>  2 = species 2 is the colliding ion
                    species, etc.). \> \\
                \>  Note: Only ONE species can be the colliding ion species. \> \\
						\\
      reflux    \>  The flag for refluxing the particles at the right wall
                    (0=off, 1=on).  In this \> \\
		\>  case, the particles hitting the
                    right wall are not absorbed but reflected back  \> \\
		\>  into the
                    system.  Since the right wall in this case does not charge
                    up, it  \> \\
		\>  serves only as a symmetry plane allowing for a
                    semi-infinite plasma at \> \\
		\>  the right wall.  The particles of
                    each species are refluxed at the temperature \> \\
		\>  specified for
                    the species. \> \\
							\\
      nfft      \>  Number of samples for the Fast Fourier Transform analyzer
                    (must be a \> \\
		\>  power of 2).  When this parameter is set to
                    zero, no FFT analysis is done, \> \\
		\>  and the diagnostics in the
                    frequency-domain are NOT shown. \> \\
									\\
		n\_ave \> Number of samples for the average diagnostics.  When this parameter is set \>\\
		\> to zero, no averages are not done and NOT shown. \> \\
						\\
      nsmoothing \> Number of time that a (1, 2, 1) digital smoothing filter is
                    applied to the \> \\
		\>  charge density arrays prior to the
                    field-solve.  This filter is only present \> \\
		\>  in PDP1. \> \\
\end{tabbing}
\end{subsubsection}

\begin{subsubsection}
{\bf Wall Emission Coefficients and Neutral Gas Parameters}

\begin{tabbing}

      seec(elect.)aaaaa\=  The coefficient of secondary electron emission due to the
                    first species \= \kill

      seec(elect.) \>  The coefficient of secondary electron emission due to the
                    first species \> \\
		   \> striking the two electrodes.  If this
                    parameter is set to say 0.1, on average \> \\
		   \> one electron is
                    injected for every 10 incident particles of this species. \> \\
								\\
      seec(ions)   \>  The coefficient of secondary electron emission due to the
                    second species \> \\	
		   \> striking the two electrodes. \> \\
							\\
      ion species  \>  indicates the ion species created by electron-neutral
                    ionization collisions \> \\
		   \> (2=the created ions are of type
                    species 2, etc.). \> \\
                   \>  Note: this also specifies the type of the background
                    neutral gas particles \> \\
		   \> colliding with electrons. \> \\
						\\
      Gpressure    \>  Background neutral gas pressure [Torr]. \> \\
						\\
      Gtemp        \>  Background neutral gas thermal temperature [eV]. \> 
\end{tabbing}
\end{subsubsection}

\begin{subsubsection} {\bf Electron-Neutral Collisional Parameters}

      The cross-sections \footnote {These details are built-in to XPDP1 in 
      version 4.0} used in the code are close fits to the values measured
      experimentally.  The general expression for all electron-neutral collision
      cross-sections, as seen in Figure 3, is:
\begin{eqnarray*}      
\sigma (E) =  0 & E  <  E_{0}&    \\ 	
\sigma (E)  \propto  E &   E_{0} \leq E  \leq E_{1}&  \\ 	
\sigma (E)  =  \sigma_{max} &  E_{1} \leq E   \leq E_{2}& \\ 	
\sigma (E)  \propto   {\ln(E)  \over E}&  E_{2}\leq E&	     
\end{eqnarray*}

\vspace{.2in}
%
\begin{figure}
\begin{center}
\leavevmode
\hbox{%
\epsfysize 4.0in
\epsfxsize 4.0in
\epsffile{f4.epsi}}
\end{center}
\caption{
General profile for electron-neutral c versus incident particle energy E.
}
\label{Elect}
\end{figure}
%
\begin{tabbing}
   elsengy0aaaaa\=Maximum electron-neutral elastic cross section
		[$m^{2}$]. \= \kill 					
   selsmax\> 	 Maximum electron-neutral elastic cross section
		 [m$^{2}$]. \> \\ 					
				\\ 
   elsengy0\> 	
Elastic collision threshold energy [eV]. \> \\ 			\\
   elsengy1\>	 Low energy of plateau for elastic collisions [eV]. \> \\
				\\ 
   elsengy2\>	 High energy of plateau for elastic collisions [eV]. \> \\ 
					\\ 
   sextmax\>	 Maximum excitation cross section [m$^{2}$]. \> \\
\\ extengy0 \>	Excitation threshold energy [eV]. \> \\
\\ extengy1 \>	Low energy of plateau for excitation [eV]. \> \\
\\ extengy2 \>	High energy of plateau for excitation [eV]. \> \\
\\ sionmax \>	Maximum ionization cross section [m$^{2}$]. \> \\
\\ ionengy0 \>	Ionization threshold energy [eV]. \> \\
					\\
      ionengy1 \> Low energy of plateau for ionization [eV]. \> \\
\\ ionengy2 \> High energy of plateau for ionization [eV]. \>
\end{tabbing}
\end{subsubsection}

\begin{subsubsection} {\bf Ion-Neutral Collisional Parameters}

      The cross-sections \footnote{These details are built-in to XPDP1 
in version 4.0} for ion-neutral collisions are assumed to be
in the form $\sigma = a + b / \sqrt{E}$, as seen in Figure 4, where a
and b are determined by finding the best fit to the experimental
values.  This empirical fit seems to closely resemble the experimental
cross-sections measured for helium, argon and neon.
%
\begin{figure}
\begin{center}
\leavevmode
\hbox{%
\epsfysize 4.0in
\epsfxsize 4.0in
\epsffile{f3.epsi}}
\end{center}
\caption{
General profile for ion-neutral collision cross-section versus incident particle energy E.
}
\label{IXSECXN}
\end{figure}
%
\noindent
\begin{tabbing}
      achrgxaaaaa\= Charge exchange cross section [m$^{2}$]. \= \kill
      achrgx \> Charge exchange cross section [m$^{2}$]. \> \\
\\ bchrgx \> Charge exchange cross section [m$^{2} \sqrt{e Volt}$].
\> \\ 							\\ ascat \>
Scattering cross section [m$^{2}$]. \> \\
\\ bscat \> Scattering cross section [m$^{2} \sqrt{e Volt}$]. \>
\end{tabbing}
\end{subsubsection}

\begin{subsubsection}
{\bf Species Parameters}

      One set for each species should be specified.

\begin{tabbing}
      max-npaaaaa\= The maximum number of particles per species. \= \kill
      max-np \> The maximum number of particles per species. \> \\
\\ q \> Charge per physical particle [C]. \> \\
\\ m \> Mass per physical particle [kg]. \> \\
\\ j0L \> Magnitude of injected current density from the left
electrode [Amps/m$^{2}$]. \> \\
\\ j0R \> Magnitude of injected current density from the right
electrode [Amps/m$^{2}$]. \> \\
\\ initn \> Initial species physical density in the system [m$^{-3}$].
\> \\
\\ max-np \> The maximum number of particles per species. \>

\end{tabbing}

\noindent
{\bf 4.1.7.1 Velocity Distribution} \\
%
\begin{figure}
\begin{center}
\leavevmode
\hbox{%
\epsfysize 4.0in
\epsfxsize 4.0in
\epsffile{f5.epsi}}
\end{center}
\caption{
Velocity distribution function in x-direction.  The
distribution function in the perpendicular direction does not have a
cutoff, but may have a drift.
}
\label{FOFVX}
\end{figure}
%
\begin{tabbing}
         vyt=vztaaaaa\= Perpendicular thermal velocity Cartesian
components \= \kill

         vx0L \> Drift velocity for v $>$ 0 particles [m/sec]. \> \\
\\ vx0R \> Drift velocity for v $<$ 0 particles [m/sec].  \> \\
\\ vxtL \> Thermal velocity for v $>$ 0 particles [m/sec]. \> \\
\\ vxtR \> Thermal velocity for v $<$ 0 particles [m/sec]. \> \\
\\ vxcL \> Cutoff velocity for v $>$ 0 thermal distribution [m/sec]. \>
\\ 								\\
vxcR \> Cutoff velocity for v $<$ 0 thermal distribution [m/sec]. \> \\
\\ v0y \> Drift velocity in the y directions for particles [m/sec]. \> \\
\\ vty \> Thermal velocity in the y direction for particles [m/sec].  \> \\
\\ v0z \> Drift velocity in the z direction for particles [m/sec]. \> \\
\\ vtz \> Thermal velocity in the z direction for particles [m/sec]. \> \\
\\ 	
\end{tabbing}
Note that the thermal velocity components are defined for a
Maxwellian distribution as 
$v_{thermal} = \left( \frac{k_{B} T}{m} \right)^{1/2}$, 
where $k_{B}$ is Boltzmann's constant, $T$ is the temperature, and
$m$ is the mass of the species.

\noindent
{\bf 4.1.7.2 Energy Distribution Diagnostics} \\

\begin{tabbing}
         Eminaaaaa\= The minimum energy seen in the energy distribution
diagnostic of the \= \kill

        nbin \> Number of bins for the energy distribution diagnostic
of the \> \\ 	 \> species at left wall. \> \\
\\ Emin \> The minimum energy seen in the energy distribution
diagnostic of the \> \\ 	 \> species at left wall [eV]. \> \\
\\ Emax \> The maximum energy seen in the energy distribution
diagnostic of the \> \\ 	 \> species at left wall [eV]. \>
\end{tabbing}

\noindent
Parameters for an energy distribution function in the system.
The parameters XStart and XFinish designate a region (a window) in the
space over which the energy distribution is calculated.

\begin{tabbing}
       XFinishaaaaa\= The right boundary of the region over which the
distribution \= \kill
         nbin \> Number of bins for the energy distribution diagnostic
of the species \> \\ 		\> 	in the system. \> \\
\\	 Emin \> The minimum energy seen in the energy distribution
diagnostic of the \> \\ 	 \>	species in the system [eV]. \>
\\ 				\\ Emax \> The maximum energy seen in
the energy distribution diagnostic of the \> \\
\>	species in the system [eV].	\> \\
\\ XStart \> The left boundary of the region over which the
distribution is calculated. \> \\ 				\\
XFinish \> The right boundary of the region over which the
distribution is calculated. \> \\
\end{tabbing}

\noindent
{\bf 4.1.7.3 Velocity Distribution Diagnostics} \\
\begin{tabbing}
         vyt=vztaaaaa\= Perpendicular thermal velocity Cartesian
components \= \kill

 vx\_lower \> Lower velocity for velocity distribution diagnostics [m/sec]. \> \\
\\ vx\_upper  \> Upper velocity for velocity distribution diagnostics [m/sec].  \> \\
\\ nxbin \> Number of bins used. (if 0 diagnostics is turned off.) \> \\
\\ vy\_lower \> Lower velocity for velocity distribution diagnostics [m/sec]. \> \\
\\ vy\_upper  \> Upper velocity for velocity distribution diagnostics [m/sec].  \> \\
\\ nybin \> Number of bins used. (if 0 diagnostics is turned off.) \> \\
\\ vz\_lower \> Lower velocity for velocity distribution diagnostics [m/sec]. \> \\
\\ vz\_upper  \> Upper velocity for velocity distribution diagnostics [m/sec].  \> \\
\\ nzbin \> Number of bins used. (if 0 diagnostics is turned off.) \> \\
\end{tabbing}

\end{subsubsection}
\end{subsection}

\begin{subsection}
{\bf Input File Library}

   Clearly the number of data sets is virtually unlimited.  PDP1 is
accompanied by a number of prepared simulations discussed below.  The
user is encouraged to make a working copy of the data sets to edit
using any ASCII word processor or editor.  Note that the relative
position of the numeric data is important for useful results, but the
number of spaces separating each number is unimportant as long as the
numbers remain on the same line.  Comments may be added on the lines
containing the descriptive text (so long as it remains on a single
line) and at the end of the input file an unlimited number of text
lines are supported.

\begin{subsubsection}
{\bf PIIIA.INP (PIIIH.INP)}

      This input file simulates plasma immersion ion implantation
(sometimes referred to as plasma source ion implantation) in an argon
(hydrogen) plasma.  A single short step potential is applied to the
target.  The electrons are pushed away from the target before the ions
have responded, generating an ion matrix (uniform) sheath.  On a
longer time scale, the ions are accelerated by the resultant electric
field across the sheath and implanted into the target.  As the ions
are implanted, the ion density in the sheath drops.  This causes the
sheath-plasma edge to recede and uncover more ions to increase the ion
density in the sheath and sustain the potential drop across the
sheath.  The velocity of the moving sheath edge depends upon, among
other factors, the pressure of the background neutral gas.  One can
vary the pressure of the neutral gas and observe the difference in the
profiles of ion energy and angular

      distribution at the target.  Electrons and ions are injected and
absorbed at the right wall.  The rate of injection for electrons is
higher than that for ions by the square root of the mass ratio (mi/me)
to reduce the source sheath produced at the right wall.
\end{subsubsection}

\begin{subsubsection}
{\bf QMACH.INP}

      This input file simulates a Q-machine, in which Maxwellian
distributions of electrons and ions are injected from the right
electrode.  The electrons reach the left electrode before the more
inertial ions, charging the (floating) electrode negatively.  After a
few ion transit times, a sheath drop region near the wall can be seen
in the spatial potential, followed by a flat region of essentially
charge-neutral plasma.  Depending on the difference in the electron
and ion injected current densities, a source sheath may be generated
at the injection electrode.
\end{subsubsection}

\begin{subsubsection}
{\bf RFDA.INP (RFDH.INP)}

      This input file simulates a capacitively coupled RF discharge
started with a few ionized particles in an argon (hydrogen) plasma.
Electrons responding to the RF field applied to the electrodes gain
energy and ionize neutrals, building the plasma initially, until some
equilibrium is reached when the electron-ion pairs generated just
balances the loss to the electrodes.  Note the symmetric (but out of
phase) sheath near the wall in the spatial plot of potential.  The
characteristics of the discharge can be changed by varying the neutral
gas pressure.  Also note that there is no physical difference between
the driven electrode and the reference electrode at zero potential in
PDP1.
\end{subsubsection}

\begin{subsubsection}
{\bf VC.INP}

      This input file simulates a virtual cathode oscillations due to
injection of a cold electron beam between shorted electrodes.  The
oscillations are periodic after an initial transient, and can be seen
in any of the time history plots.
\end{subsubsection}
\end{subsection}

\begin{subsection}
{\bf Format}

   The input file is currently a fixed format ASCII file.  The input
files may be edited using any ASCII text editor or word processor
which does not insert formatting characters into the file.  The text
lines can contain any descriptive comments, etc., and the text may
continue for as many lines as desired.  The lines containing numbers
must also remain on a single line and each number can only be
separated by white space, including any amount of spaces and tabs.
The numbers may be in floating point or exponential format (decimal
point and sign are optional).

   The preferred method of trying new input parameters is to COPY the
original data set to a new file name and edit the working copy.  This
will leaves original data set intact for future use and reference.

\end{subsection}
\end{section}

\newpage
\noindent
\begin{appendix}
\begin{section}
{\bf SYSTEM REQUIREMENTS}

\noindent
The X-Windows version requires X11 libraries (X11R4 or any superset of
X11 such as Motif), a C compiler, and an X display or X-terminal. In
addition xgrafix requires Tcl/Tk also be installed. The exact version of
the (public domain) required graphics libraries can be ascertained by 
reading the README file presented in the distribution.

To edit the input files, the user must have an ASCII text editor or
word processor that can edit and save plain ASCII files without
formatting or control characters.
\end{section}

\newpage
\noindent
\begin{section}
{\bf APPENDIX TECHNICAL SUPPORT}

The Plasma Theory amd Simulation Group maintains mailing lists for discussing
the codes developed by it. In order to subscribe to a mailing list\\
send email to\\
{\em  majordomo@langmuir.eecs.berkeley.edu} \\
with the phrase \\
{\bf subscribe $\langle {\rm {\bf list}} \rangle$}\\
in the body of the message.  This will cause you
to be automatically added to the mailing list.\\
{\bf $\langle {\rm {\bf list}} \rangle$} is the name of the mailing list 
you want to subscribe to and can be one of the following 
\begin {itemize}
\item {\bf oopic-users}: mailing list for oopic and xoopic users.
\item {\bf pdp2-users}: mailing list for xpdp2 users.
\item {\bf pdx1-users}: mailing list for xpdp1 and xpdc1 users.
\item {\bf es1-users}: mailing list for xes1 users.
\end{itemize}

The group also maintains a web site containing home pages of group
members and the anonymous FTP directory for accessing codes and can
be  accessed at\\
 {\em http://ptsg.eecs.berkeley.edu} in the {\em pub/codes} directory\\

For technical support, modifications and custom models, contact
{\bf CPS} (Computer Physics Software) via email at 
{\em cps@langmuir.eecs.berkeley.edu}
\end{section}
\end{appendix}

\newpage
\center\begin{thebibliography}{99}

\bibitem{kn:bird}
C. K. Birdsall and A. B. Langdon, {\em Plasma Physics Via Computer Simulation}, (McGraw-Hill 1985, Adam-Hilger 1991 
which has ES1 disk).

\bibitem{kn:hock}
R. W. Hockney and J. W. Eastwood, {\em Computer Simulation Using Particles}, Adam Hilger (1988).

\bibitem{kn:law}
W. S. Lawson, {\em PDW1 User's Manual}, Electronics Research Laboratory 
Report M84/37 (1984).

\bibitem{kn:verb}
J. P. Verboncoeur and V. Vahedi, {\em WinGraphics: An Optimized Windowing 
Environment for Plasma Simulation}, Proceedings of $13^{th}$ 
Numerical Simulation Conference, Sante Fe, NM (1989).

\bibitem{kn:alves}
John P. Verboncoeur, M. Virginia Alves, V. Vahedi, and
C. K. Birdsall, {\em J. Comp. Phys.} {\bf 104}, 321 (1993).
\end{thebibliography}
\end{document}

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/02-Gasentladungen/xpdp1/src/fft.c

@@ -1,0 +1,146 @@
+#include <math.h>
+#include "pdp1.h"
+
+#define SWAP(a, b) tempr=(a); (a)=(b); (b)=tempr
+
+#ifndef M_PI
+#define M_PI    3.14159265358979323846
+#endif
+
+
+
+/***************************************************************/
+/* Replace `data' by its discrete Fourier transform, if `isign'is
+	input as 1, or by its inverse discrete Fourier transform, if 
+	`isign' is input as -1. `data' is a complex array of length `nn',
+	input as a real array data[1..2*nn]. `nn' MUST be an integer
+	power of 2 (this is not checked for!?)  */
+
+void four1(SCALAR data[], int nn, int isign)
+{
+  register int i, j, m, n;
+  int  mmax, istep;
+  SCALAR wtemp, wr, wpr, wpi, wi, theta;
+  SCALAR tempr, tempi;
+  
+  n= nn<<1;
+  j=1;
+  
+  for(i=1; i<n; i+=2)
+  {
+    if(j >i)
+    {
+      SWAP(data[j], data[i]);
+      SWAP(data[j+1], data[i+1]);
+    }
+    m= n>>1;
+    while(m >=2 &&j >m)
+    {
+      j -=m;
+      m >>=1;
+    }
+    j +=m;
+  }
+  mmax =2;
+  while(n> mmax)
+  {
+    istep= 2*mmax;
+    theta= 6.28318530717959/(isign*mmax);
+    wtemp= sin(.5*theta);
+    wpr= -2.0*wtemp*wtemp;
+    wpi= sin(theta);
+    wr= 1.0;
+    wi= 0.0;
+    for(m=1; m<mmax; m+=2)
+    {
+      for(i=m; i<n; i+=istep)
+      {
+	j=i+mmax;
+	tempr= wr*data[j]-wi*data[j+1];
+	tempi= wr*data[j+1]+wi*data[j];
+	data[j]= data[i]- tempr;
+	data[j+1]= data[i+1]- tempi;
+	data[i] +=tempr;
+	data[i+1] +=tempi;
+      }
+      wr= (wtemp=wr)*wpr - wi*wpi+wr;
+      wi= wi*wpr + wtemp*wpi + wi;
+    }
+    mmax= istep;
+  }
+}
+
+/***************************************************************/
+/* Calculates the Fourier transform of a set of 2n real-valued
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/02-Gasentladungen/xpdp1/src/maxwellv.c

@@ -1,0 +1,61 @@
+#include<math.h>
+#include <stdlib.h> 
+#include <stdio.h>
+#include "pdp1.h"
+
+#define  NVTS    3.3
+
+#ifndef M_PI
+#define M_PI    3.14159265358979323846
+#endif
+
+/**************************************************************/
+SCALAR F(SCALAR);
+SCALAR frand(void);
+
+void maxwellv(SCALAR *vx_local, SCALAR *vy_local, SCALAR *vz_local, SCALAR vth)
+{
+  static int nvel, init_flag= 1;
+  static SCALAR *vsave_x, *vsave_y, *vsave_z;
+  
+  int i, n;
+  SCALAR vmag, aphi, dv, rr, sintheta, costheta;
+  
+  if (init_flag) {
+    nvel= 1./(1-F((SCALAR)NVTS));
+    dv = sqrt(M_PI)/(4.0*nvel);
+    if(nvel > 100000)
+      puts("Warning: Your choice of NVTS has made nvel > 1e5"); 
+    
+    vsave_x= (SCALAR *) malloc(nvel*sizeof(SCALAR));
+    vsave_y= (SCALAR *) malloc(nvel*sizeof(SCALAR));
+    vsave_z= (SCALAR *) malloc(nvel*sizeof(SCALAR));
+    
+    init_flag= 0;
+    i=n=0;
+    for (n=0; n<nvel; n++) {
+      rr=(1.0*n)/nvel;
+      while (F(i*dv)< rr) i++;
+      vmag=i*dv;
+      aphi=2*M_PI*frand();
+      costheta = 1-2*frand();
+      sintheta = sqrt(1-costheta*costheta);
+      vsave_x[n] = vmag*sintheta*cos(aphi);
+      vsave_y[n] = vmag*sintheta*sin(aphi);
+      vsave_z[n] = vmag*costheta;
+    }
+  }
+  n = (nvel-1)*frand();
+  *vx_local = vth*vsave_x[n];
+  *vy_local = vth*vsave_y[n];
+  *vz_local = vth*vsave_z[n];
+}
+
+/**************************************************************/
+
+SCALAR F(SCALAR v)
+{
+  return(-2*v*exp(-v*v)/sqrt(M_PI) +erf(v));
+}
+
+/**************************************************************/

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/02-Gasentladungen/xpdp1/src/mcc.c

@@ -1,0 +1,551 @@
+#include "pdp1.h"
+#include "xgrafix.h"
+
+#define   NEMAX    500
+SCALAR sigma1(SCALAR), sigma2(SCALAR), sigma3(SCALAR), sigma4(SCALAR), sigma5(SCALAR);
+SCALAR gden, vgth;
+int ecolsp, icolsp, ionsp;
+void newvel(SCALAR, SCALAR, SCALAR *, SCALAR *, SCALAR *, int);
+
+/***************************************************************/
+/* Monte Carlo scheme for electron and ion-neutrals collisions */
+/* easly convert old mcc cross sections */
+
+#define selsmax 1.0e-19 
+#define elsengy0 0.0
+#define elsengy1 0.0
+#define elsengy2 10.0
+#define sextmax 1.0e-20 
+#define extengy0 12.0    
+#define extengy1 50.0
+#define extengy2 100.0 
+#define sionmax 1.0e-20
+#define ionengy0 13.6 
+#define ionengy1 60.0
+#define ionengy2 110.0
+#define achrgx 3.0e-19
+#define bchrgx 0.0  
+#define ascat 2.0e-19     
+#define bscat 0.0
+
+void mcc(int isp)
+{
+  static int init_flag=1;
+  static SCALAR ecol_extra, icol_extra;
+  static SCALAR max_sigmav_e, max_sigmav_i;
+  static SCALAR col_prob_e, col_prob_i;
+  
+  int N, i, k, nnp, j, s, index;
+  SCALAR random, dum, vel,sigma_total;
+  SCALAR engy, rengy, del, phi1;
+  SCALAR cosphi, sinphi, coschi, sinchi;
+  SCALAR temp, vneutx, vneuty, vneutz;
+  SCALAR tempsigma1, tempsigma2;
+  SCALAR tempsigma5;
+  
+  if (init_flag)
+		{
+			gden = NperTORR*pressure/(gtemp+DBL_MIN);  /* Calculating the gas density */
+
+			ecolsp= ecollisional-1;
+			icolsp= icollisional-1;
+			ionsp= ionspecies-1;          /* Fixing the indices into the array of species */
+			
+			if(ionsp>=0) vgth= sqrt(gtemp/Escale[ionsp]);
+
+			/*****************************************/
+			/* Calculating the null collision prob.  */
+			
+			if(ecollisional)
+				{
+					max_sigmav_e = 0.0;
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/02-Gasentladungen/xpdp1/src/mccdiaginit.c

@@ -1,0 +1,100 @@
+#include "pdp1.h"
+
+void mccdiag_init(void)
+{
+  int i, j;
+  
+  if(ecollisional==0 && icollisional==0) {
+    ndiag = 0;
+    return;
+  }
+  
+	/* default gas = argon */
+
+	if ( (gas < HELIUM) || (gas > OXYGEN) ) gas = ARGON;
+
+	switch (gas) {
+
+	case HELIUM:
+		ndiag = 5;
+		
+		rate_title = (char **)malloc(ndiag*sizeof(char *));
+		rate_title[0]= "MCC Elastic";
+		rate_title[1]= "MCC Excitation";
+		rate_title[2]= "MCC Ionization";
+		rate_title[3]= "MCC Ion Scattering";
+		rate_title[4]= "MCC Charge X";
+		break;
+		
+	case ARGON:
+		ndiag = 5;
+		
+		rate_title = (char **)malloc(ndiag*sizeof(char *));
+		rate_title[0]= "MCC Elastic";
+		rate_title[1]= "MCC Excitation";
+		rate_title[2]= "MCC Ionization";
+		rate_title[3]= "MCC Ion Scattering";
+		rate_title[4]= "MCC Charge X";
+		break;
+
+	case NEON:
+		ndiag = 7;
+		
+		rate_title = (char **)malloc(ndiag*sizeof(char *));
+		rate_title[0]= "MCC Elastic";
+		rate_title[1]= "MCC Excitation1";
+		rate_title[2]= "MCC Excitation2";
+		rate_title[3]= "MCC Excitation3";
+		rate_title[4]= "MCC Excitation4";
+		rate_title[5]= "MCC Ionization";
+		rate_title[6]= "MCC Charge X";
+		break;
+		
+	case OXYGEN:
+		ndiag = 21;
+		
+		rate_title = (char **)malloc(ndiag*sizeof(char *));
+		rate_title[0]=  "MCC Elastic";
+		rate_title[1]=  "MCC Rotational";
+		rate_title[2]=  "MCC Vibrational1";
+		rate_title[3]=  "MCC Vibrational2";
+		rate_title[4]=  "MCC Vibrational3";
+		rate_title[5]=  "MCC Vibrational4";
+		rate_title[6]=  "MCC O2 Sing Delta";
+		rate_title[7]=  "MCC O2 Sing Sigma";
+		rate_title[8]=  "MCC Diss Attach";
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/02-Gasentladungen/xpdp1/src/move.c

@@ -1,0 +1,199 @@
+#include "pdp1.h"
+#include "xgrafix.h"
+
+/***************************************************************/
+
+void imp_move(int isp, const int EnergyFlag)
+{
+  register int i, j, k;
+  register SCALAR  s;
+  SCALAR atemp, ke_temp; 
+  
+  k = sp_k[isp];
+  for (j=0; j<ng; j++) {
+    a[j] = e[j]*a_scale[isp];
+    if(EnergyFlag) sp_j_x[isp][j] = sp_j_y[isp][j] = sp_j_z[isp][j] = sp_ke_x[isp][j] =  sp_ke_y[isp][j] = sp_ke_z[isp][j] = 0.0;
+  }
+  
+  for(i= 0; i<np[isp]; i++) { 
+    /****** Post-push to time n *****/
+    j = x[isp][i];
+    s = x[isp][i] - j;
+    atemp= a[j] + s*(a[j+1] - a[j]);
+    
+    x[isp][i] += 0.5*k*atemp;
+    vx[isp][i]+= atemp;
+
+    /* NOTE: Algorithmic bug: Particle may go out of bounds in prepush. */
+    /****** Get diagnostics at time n ****/
+    if (x[isp][i]< xnc) { /* exclude outsiders */
+    j = x[isp][i];
+    s = x[isp][i] - j;
+
+    if(EnergyFlag) {
+      ke_temp= vx[isp][i]*vx[isp][i] +vy[isp][i]*vy[isp][i] +vz[isp][i]*vz[isp][i];
+      sp_j_x[isp][j]   += (1-s)*vx[isp][i];
+      sp_j_x[isp][j+1] += s*vx[isp][i];
+      sp_ke_x[isp][j]  += (1-s)*ke_temp;
+      sp_ke_x[isp][j+1]+= s*ke_temp;
+    }
+    
+    /****** Pre-push to time n+1  *******/
+    atemp= a[j] + s*(a[j+1] - a[j]);
+    x[isp][i] += k*(vx[isp][i] +0.5*atemp);
+    vx[isp][i]+= atemp;
+   }
+  }
+  
+  /****** Fix diagnostics and end points and rescale ****/
+  if(EnergyFlag) {
+    sp_j_x[isp][0]  *= 2.;
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/02-Gasentladungen/xpdp1/src/xsect.h

@@ -1,0 +1,22 @@
+/**********************************************
+  Cross section tables header
+**********************************************/
+
+#ifndef __xsect_h
+#define __xsect_h
+#include "pdp1.h"
+
+struct crossSectStruct
+{
+  int n;/* number of data pairs */
+  SCALAR thresholdE; /* threshold energy */
+  SCALAR* E; /* energy array */
+  SCALAR* sigma; /* cross section array */
+};
+
+typedef struct crossSectStruct CrossSect;
+
+int readXSectionTables(CrossSect** xSect, CrossSect* totalMom, char* fileName);
+SCALAR interpolateTableLinear(CrossSect *xSect, SCALAR E);
+
+#endif

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/04-Culombstoesse/coulomb.py

@@ -1,0 +1,177 @@
+#!/usr/bin/env python3
+
+"""
+  Electron undergoing a Coulomb collision
+"""
+
+import numpy as np
+import scipy.integrate as integrate
+
+# natural constants
+e = 1.602e-19   # [As] elementary charge
+epsilon0 = 8.854e-12   # [As/Vm]  dielectric constant
+m_e = 9.11e-31   # [kg] electron mass
+
+# global parameters
+E0 = 2*e   # [eV]   incident kinetic energy
+vz0 = np.sqrt(2*E0/m_e)   # [m/s] incident velocity
+
+q1 = e          # [As] of scattering centre
+q2 = -e         # [As] of scattered particle
+r0perp = np.abs(q1*q2)/(8*np.pi*epsilon0*E0)  # [m]  impact parameter for 90 degrees deflection
+
+
+# ---------------------------------------------------------------------
+class Particle:
+    '''
+      follow one particle in electrostatic potential
+      :dt:  time step [s]   (no default value)
+      :m:   particle mass [kg]    (default: electron mass)
+      :lambda_D:  Debye shielding length [m] (None if no shielding)
+    '''
+    def __init__(self, dt, m=m_e, lambda_D=None):
+        self.m = m
+        self.lambda_D = lambda_D
+        self.dt = dt
+        self.CoulombConst = q1*q2 / (4*np.pi*epsilon0)
+        self.tot_weight = 0.0   # accumulated weight
+        self.tot_dvz = 0.0      # accumulated
+        self.r = np.array([])   # empty arrays to start with
+        self.z = np.array([])
+
+    def init(self, init_state):
+        '''
+        initialise particle position and velocity
+        :init_state:  [r0, z0, vr0, vz0]
+        '''
+        self.init_state = init_state
+        self.state = self.init_state
+        self.r = np.array([init_state[0]])
+        self.z = np.array([init_state[1]])
+
+    def dvz_acc(self, r0perp):
+        '''
+         evaluate slowing down: accumulated dvz-dvz0
+        '''
+        weight = (self.init_state[0]/r0perp)**2 # cyl-sym around z -> weight = (r0/r0perp)^2 
+        dvz = self.state[3] - self.init_state[3]  # vz - vz0
+        self.tot_weight += weight
+        self.tot_dvz += (dvz*weight)
+        return self.tot_dvz  # accumulated dvz
+
+    def energy(self):
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/05-Einzelteilchen/gyro_anim.py

@@ -1,0 +1,84 @@
+#!/usr/bin/env python3
+# animate gyromotion
+
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import matplotlib.animation as animation
+
+import particle
+
+# constants
+e = 1.602E-19   # As   elementary charge
+m_e = 9.11E-31   # kg   electron mass
+m_p = 1.67E-27   # kg   proton mass
+
+# homogeneous magnetic field, no electrical field
+B0    = [0.0, 0.0, 1.0]  # [T]
+gradB = [0.0, 0.0, 0.0]  # [T/m]
+E0 = np.array([0.0, 0.0, 0.0])
+f = particle.SimpleFields(E0, B0, 0.0)  # the EM field
+
+# particle
+Ekin0 = 200*e   # kinetic energy [J]
+if False:   # electron
+    m = m_e
+    q = -e
+    col = 'green'
+else:      # proton
+    m = m_p
+    q = e
+    col = 'red'
+
+v_perp = np.sqrt(2*Ekin0/m) # initial (perpendicular) velocity
+    
+# analytical solution
+omega_c = q*np.sqrt(np.dot(B0,B0))/m
+r_L = v_perp/np.abs(omega_c)
+    
+# simulation parameters
+ngyrations = 10
+nframes = 100 * ngyrations
+tmax = ngyrations * (2*np.pi)/np.abs(omega_c)
+tstep = tmax/nframes    # time step for each frame
+
+# the simulation
+p = particle.FullOrbit(f, tstep, m, q)
+x0 = [0.0, r_L, 0.0]
+v0 = [q/e*v_perp, 0.0, 0.0]
+
+# plot
+matplotlib.rcParams['figure.figsize'] = [8.0,8.0]
+matplotlib.rcParams['figure.subplot.left'] = 0.09
+matplotlib.rcParams['figure.subplot.right'] = 0.975
+matplotlib.rcParams['figure.subplot.bottom'] = 0.05
+matplotlib.rcParams['figure.subplot.top'] = 0.98
+
+fig = plt.figure()
+ab = 1.1e3*r_L  # plot boundary
+ax = fig.add_subplot(111, aspect='equal', autoscale_on=False,
+                     xlim=(-ab, ab), ylim=(-ab, ab))
+plt.xlabel('x [mm]')
+plt.ylabel('y [mm]')
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/05-Einzelteilchen/particle.py

@@ -1,0 +1,252 @@
+#!/usr/bin/env python3
+import numpy as np
+import scipy.integrate
+
+
+# --------------------- field definition ---------------------------
+class SimpleFields():
+  '''
+   Base class for electromagnetic field definitions
+   Implements a homogeneous E field and 
+   B field || z with gradient in x-direction
+  '''
+  def __init__(self, E, B, gradB):
+    '''
+      initialise simple static E and B fields with
+      B = [Bx, By, Bz + gradB*x]  (field in z-direction, gradient B in x-direction)
+
+      :E: = [Ex, Ey, Ez]
+      :B: = [Bx, By, Bz] homogeneous part of B field
+      :gradB: (T/m) gradient of Bz in x-direction
+    '''
+    self.E = np.array(E)
+    self.B = np.array(B)
+    self.gradB = gradB
+
+  def Evec(self, t, r):
+    '''
+      return electrical field 
+      at time t and position vector r.
+      r can be a vector, r=[x,y,z], 
+      or an array r=[xyz, caseindex]
+    '''
+    if r.ndim==1:
+      return self.E
+    else:   # same at all positions
+      return np.outer(self.E, np.ones(r.shape[1]))
+
+  def Bvec(self, t, r):
+    '''
+     return magnetic induction vector
+     at time t and position(s) r.
+     r can be a vector, r=[x,y,z], 
+     or an array r=[xyz, caseindex]
+    '''
+    if r.ndim==1:
+      return np.array([self.B[0], self.B[1], self.B[2] + self.gradB*r[0]])
+    else:
+      B = np.outer(self.B, np.ones(r.shape[1]))
+      dBz = self.gradB*r[0]
+      B[2,:] += dBz
+      return B
+
+  def grad_B(self, t, r):
+    '''
+     return grad B at time t and position vector r 
+    '''
+    if r.ndim==1:
+      return [self.gradB, 0.0, 0.0]
+    else:   # same at all positions
+      return np.outer([self.gradB, 0.0, 0.0], np.ones(r.shape[1]))
+
+    
+# --------------------- particle integrators -----------------------
+class FullOrbit():
+
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/06-HFPlasmen/RF_particle_anim.py

@@ -1,0 +1,240 @@
+#!/usr/bin/env python
+# animate a (possibly gyrating, possibly damped) particle in an oscillating electrical field
+
+import numpy as np
+import scipy.integrate
+import matplotlib
+import matplotlib.pyplot as plt
+import matplotlib.animation as animation
+
+
+epsilon0 = 8.854e-12   # [As/Vm]  free-space permissivity
+mu0 = 4*np.pi*1e-7     # [Vs/Am] free-space permeability
+c0 = 1.0/np.sqrt(epsilon0 * mu0)   # [m/s] free-space speed of light
+
+# ------------------ plane wave RF field --------------
+class RF_PlaneWave ():
+
+  def __init__(self, Ex1, Bz0, omega):
+    '''
+      plane wave - oscillating E_x field and static homogeneous B_z field 
+      E1 = [Ex1, 0, 0] * cos(omega*t)
+      B0 = [0, 0, Bz0] 
+    '''
+    self.E1 = np.array([Ex1, 0.0, 0.0])
+    self.B0 = np.array([0.0, 0.0, Bz0])
+    self.B1 = np.array([0.0, 0.0, -Ex1/c0])  # from Faraday's law, integrated
+    self.omega = omega
+    self.ky = omega / c0   # wave vector in y direction
+    print('RF wavelength: %f mm' % (1e3/self.ky))
+    print('Wave magnetic field amplitude: %f T' % (np.abs(Ex1)/c0))
+    if Bz0!=0:
+      print('  B1 / B0 = %f' % (np.abs(Ex1/c0/Bz0)))
+
+  def Evec(self, t, r):
+    '''
+      return electrical field 
+      at time t and position vector r.
+      r can be a vector, r=[x,y,z], 
+      or an array r=[xyz, caseindex]
+    '''
+    E = self.E1 * np.cos(self.omega * t)
+    if r.ndim==1:
+      return E
+    else:   # same at all positions
+      return np.outer(E, np.ones(r.shape[1]))
+
+  def Bvec(self, t, r):
+    '''
+     return magnetic induction vector
+     at time t and position(s) r.
+     r can be a vector, r=[x,y,z], 
+     or an array r=[xyz, caseindex]
+    '''
+    B = self.B0 + self.B1*np.sin(self.omega*t)
+    if r.ndim==1:
+      return self.B0
+    else:
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/08-LandauDaempfung/landaudamp.py

@@ -1,0 +1,279 @@
+#!/usr/bin/env python3
+
+'''
+  Simulation of Landau damping
+  using the most simple 1D electrostatic PIC model (one spatial, one velocity dimension)
+  after pic.py by Philip Mocs
+  https://medium.com/swlh/create-your-own-plasma-pic-simulation-with-python-39145c66578b
+'''
+
+import numpy as np
+import scipy.sparse as sp
+from scipy.sparse.linalg import spsolve
+
+class Electrostatic_1d():
+
+    def __init__(self, Nx, boxsize):
+        '''
+          electrostatic field only
+          :Nx:       number of grid points, preferrably a power of 2
+          :boxsize:  length of computational domain
+        '''
+        self.Nx = Nx
+        self.boxsize = boxsize
+        self.grid = np.linspace(0,boxsize,Nx)
+
+	# Construct matrix G to compute Gradient  (1st derivative)
+        dx = boxsize/Nx
+        e = np.ones(Nx)
+        diags = np.array([-1,1])
+        vals  = np.vstack((-e,e))
+        Gmtx = sp.spdiags(vals, diags, Nx, Nx);
+        Gmtx = sp.lil_matrix(Gmtx)
+        Gmtx[0,Nx-1] = -1
+        Gmtx[Nx-1,0] = 1
+        Gmtx /= (2*dx)
+        self.Gmtx = sp.csr_matrix(Gmtx)
+
+        # Construct matrix L to compute Laplacian (2nd derivative)
+        diags = np.array([-1,0,1])
+        vals  = np.vstack((e,-2*e,e))
+        Lmtx = sp.spdiags(vals, diags, Nx, Nx);
+        Lmtx = sp.lil_matrix(Lmtx)
+        Lmtx[0,Nx-1] = 1
+        Lmtx[Nx-1,0] = 1
+        Lmtx /= dx**2
+        self.Lmtx = sp.csr_matrix(Lmtx)
+
+    def weights(self, pos):
+        '''
+         weights for particles at positions 'pos'
+        '''
+        dx         = self.boxsize / self.Nx
+        j          = np.floor(pos/dx).astype(int)
+        jp1        = j+1
+        weight_j   = ( jp1*dx - pos  )/dx
+        weight_jp1 = ( pos    - j*dx )/dx
+        j          = np.mod(j, self.Nx)
+        jp1        = np.mod(jp1, self.Nx)   # periodic BC
+        return j, jp1, weight_j, weight_jp1
+        
[…]

### 2022-04-05 literatur/Suttrop - Einfuehrung in Plasmaphysik/10-Wellen/cold_plasma_waves.py

@@ -1,0 +1,191 @@
+'''
+  cold_plasma_waves.py  - dispersion relation for cold waves (P=0, T=0) in a plasma
+'''
+
+import numpy as np
+
+# ---------------------------------------------------------------------------------------
+
+class ColdElectronWave(object):
+    '''
+      Wave in a cold electron fluid
+    '''
+
+    def __init__(self, ompe2, omce):
+        '''
+           Initialize wave in a cold electron fluid
+           
+           ompe2: (omega_p,e / omega)^2   square of (plasma frequency normalised to frequency)
+           omcs:  (omega_c,e / omega)     cyclotron frequency normalised to frequency
+        '''
+        self.ompe2 = ompe2
+        self.omce  = omce
+        self.P = 1 - ompe2
+        self.R = 1 - ompe2/(1.0 - omce)
+        self.L = 1 - ompe2/(1.0 + omce)
+        self.S = 0.5*(self.R + self.L)
+        self.D = 0.5*(self.R - self.L)
+
+    def calc_n2(self, theta):
+        '''
+           set progagation direction of wave
+           theta: propagation angle with respect to static 
+                  background magnetic field \vec{B} (in radians)
+               theta = 0: wave vector parallel \vec{B}
+               theta = pi/s: wave vector perpendicular \vec{B}
+        '''
+        s2t = np.square(np.sin(theta))
+        c2t = np.square(np.cos(theta))
+        a = s2t*self.S + c2t*self.P
+        b = -(s2t*self.R*self.L + (1+c2t)*self.P*self.S)
+        c = self.R *self.L *self.P
+        n2a = (-b + np.sqrt(np.square(b) - 4*a*c)) / 2/a
+        n2b = (-b - np.sqrt(np.square(b) - 4*a*c)) / 2/a
+        return np.array([n2a, n2b])
+
+
+# --------------------------- main --------------------------------------------------
+
+
+import matplotlib.pyplot as plt
+
+
+# k || B, frequency dependence
+if True:
+    omfac = 0.9   # omega_p^2 / omega_c^2
+
+    logmin = -2
+    logmax = 2    
[…]

### 2022-05-03 gkw/doc/chease/gkwandchease.tex
(1) Added gkw Clone gwk  #3

\documentclass[a4paper,12pt]{article}

\newcommand{\ind}[1]{\ensuremath{_{\rm{#1}}}}
\newcommand{\expo}[1]{\ensuremath{^{\rm{#1}}}}

\usepackage{amsmath}
\usepackage{amssymb}

\setlength\parindent{0in}
\oddsidemargin 0.0in
\evensidemargin 1.0in
\textwidth 6.0in
\textheight 9.0in
\headheight 0.0in
\topmargin 0.0in

% Title Page
\title{CHEASE/GKW Interface}
\author{Y. Camenen}
\date{September 11, 2008}

\begin{document}
\maketitle

%\begin{abstract}
%
%\end{abstract}

This rather technical report aims at documenting the addition of a small module in CHEASE to compute the metric elements required for GKW and of another module in GKW to read the CHEASE outputs.

%*********************************************************************
\part{Background}
%*********************************************************************
\section{CHEASE coordinates}
In CHEASE \cite{Lutjens:Comput1992, Lutjens:Comput1996}, the Grad-Shafranov equation is solved on a $\left(\sigma,\theta,\phi\right)$ grid, where $\theta$ and $\phi$ are the poloidal and toroidal
angle, respectively. The
non-orthogonal coordinate system $\left(\sigma,\theta,\phi\right)$ is related to the cylindrical coordinates system $\left(R,Z,\phi\right)$ by
\begin{equation}
\begin{aligned}
R&=\sigma\rho\ind{s}(\theta)\cos{\theta} + R\ind{c} \nonumber \\
Z&=\sigma\rho\ind{s}(\theta)\sin{\theta} + Z\ind{c} \nonumber
\end{aligned}
\end{equation}
where $0 \leq \sigma \leq 1$, $0\leq \theta \leq 2\pi$, $\left(R\ind{c},Z\ind{c}\right)$ is the center of the $\left(\sigma,\theta\right)$ grid and $\rho\ind{s}(\theta)$ is the distance from $\left(R\ind{c},Z\ind{c}\right)$ to the plasma boundary (assumed to be known). Note that $\sigma$ is not a flux label.

After the Grad-Shafranov equation has been solved, the equilibrium is mapped into flux coordinates $\left(\Psi,\chi,\phi\right)$ where
$\Psi=\frac{1}{2\pi}\int\ind{S_{pol}}\textbf{n}\cdot\textbf{B}\,{\rm d}S$ is the poloidal magnetic flux and $\chi$ is a generalized poloidal angle specified by the choice of the jacobian. Note that in
the Grad-Shafranov solver part (i.e. before the mapping), $\Psi$ is negative inside the plasma (the plasma current flows in the direction of $\nabla\phi$) and minimum on the magnetic axis. However,
for the flux coordinates, the poloidal flux is renormalized to be positive in the plasma and to increase towards the edge. This allows to have a positive radial coordinate, as usual.\\
In CHEASE, the jacobian $J_{\Psi\chi\phi}=\left[\nabla\Psi\times\nabla\chi\cdot\nabla\phi\right]^{-1}$ can have the following forms:
$$J_{\Psi\chi\phi}=C(\Psi)R\expo{NER}\left|\nabla\Psi\right|\expo{NEGP}$$
where NER and NEGP are integers specified in the input file and $C(\Psi)$ is determined by imposing that $\chi$ increases by $2\pi$ per poloidal turn. It can be shown (see Appendix A in \cite{Lutjens:Comput1992}) that:  
$$\chi(\theta)=\int_0^\theta\frac{R\sigma\rho\ind{s}^2(\tilde{\theta})}{J_{\Psi\chi\phi}\partial\Psi/\partial{\sigma}}\,{\rm d}\tilde{\theta}$$
Defining $s\ind{B}=\textrm{sign}(\bf{B}\cdot\nabla\phi)$ and $s\ind{j}=\textrm{sign}(\bf{j}\cdot\nabla\phi)$, where $\bf{j}$ is the current density, the magnetic field (axisymmetric device) is
represented as 
$$\textbf{B}=s\ind{B}F(\Psi)\nabla{\phi}+s\ind{j}\nabla\phi\times\nabla\Psi$$
with $F>0$ and $\nabla\Psi$ is directed from the magnetic axis to the edge. It leads to:
\begin{equation}
\begin{aligned}
B^\Psi&=\textbf{B}\cdot\nabla\Psi=0 \nonumber \\
B^\chi&=\textbf{B}\cdot\nabla\chi=\frac{s\ind{j}}{J_{\Psi\chi\phi}} \nonumber \\
B^\phi&=\textbf{B}\cdot\nabla\phi=s\ind{B}F\left|\nabla\phi\right|^2=\frac{s\ind{B}F}{R^2} \nonumber
\end{aligned}
\end{equation}
By convention, in CHEASE, $s\ind{B}=-1$ and $s\ind{j}=-1$ for $\nabla \phi$ oriented as in GKW, i.e. $(R,Z,\phi)$ direct. In fact, in CHEASE $\nabla \phi\expo{ch}=-\nabla \phi$ is oriented
such as to have $(R,\phi\expo{ch},Z)$ direct. However, in the code the jacobian is always defined positive and taken to be $|J_{\Psi\chi\phi\expo{ch}}|$ .

\section{GKW coordinates}
GKW uses a field aligned flux coordinate system $\left(\psi,\zeta,s\right)$ with constant jacobian on a flux surface (Hamada coordinates with third coordinate transformed). Here, $\psi$ is a flux surface label (not necessarily equal to the $\Psi$ used in CHEASE) and $-1/2\leq s\leq1/2$ for one poloidal turn ($s=0$ on the LFS midplane).

In section 2.1 of the GKW manual, it is shown that provided a toroidal coordinate system $\left(\psi,s,\phi\right)$ with $B^s$ being a flux label, it is possible to build the required field aligned flux coordinate system $\left(\psi,\zeta,s\right)$ with constant jacobian.

\section{Construction of the GKW coordinates from the CHEASE coordinates}
In CHEASE, a constant jacobian on a flux surface can be specified for the $\left(\Psi,\chi,\phi\right)$ coordinate system by setting $NER=0$ and $NEGP=0$ in the input file. With this choice, the
$\left(\Psi,\chi,\phi\right)$ coordinate system is neither straight field line, nor field aligned, however, $B^\chi=\frac{s\ind{j}}{J_{\Psi\chi\phi}}=\frac{s\ind{j}}{C(\Psi)}$ is a flux label.

From now on, $\psi=\Psi$ is chosen for the radial coordinate of the $\left(\psi,\zeta,s\right)$ GKW coordinate system.

\subsection{Transformation to straight field line coordinates}
The following transformations are applied:
\begin{equation}
\begin{aligned}
s&= \frac{\chi}{2\pi}\nonumber \\
\gamma&= \frac{\phi}{2\pi}+s\ind{B}\frac{F}{2\pi} \int_0^\chi\frac{{\rm d}\tilde{\chi}}{\textbf{B}\cdot\nabla\tilde{\chi}}\left[\left<\frac{1}{R^2} \right>
-\frac{1}{R^2(\Psi,\tilde{\chi})}\right]\nonumber
\end{aligned}
\end{equation}
leading to
\begin{equation}
\begin{aligned}
\nabla s&= \frac{\nabla\chi}{2\pi} \nonumber \\
\nabla \gamma&= \frac{\nabla\phi}{2\pi} +
s\ind{B}\frac{FJ_{\Psi\chi\phi}}{2\pi}\left[\left<\frac{1}{R^2} \right> -\frac{1}{R^2}\right]\nabla\chi + \frac{\partial \gamma}{\partial \Psi}\nabla\Psi \nonumber 
\end{aligned}
\end{equation}
and
\begin{equation}
\begin{aligned}
B^s&=\textbf{B}\cdot\nabla s=\frac{s\ind{j}}{2\pi J_{\Psi\chi\phi}} \nonumber \\
B^\gamma&=\textbf{B}\cdot\nabla\gamma=\frac{s\ind{B}F}{2\pi}\left<\frac{1}{R^2} \right> \nonumber
\end{aligned}
\end{equation}
Therefore, $B^s/B^\gamma$ is constant on a flux surface and $\left(\Psi,s,\gamma\right)$ is a straight field line coordinate system with the jacobian 
$$J_{\Psi s \gamma}=\left[\nabla\Psi\times\nabla s\cdot\nabla\gamma\right]^{-1}=4\pi^2J_{\Psi\chi\phi}>0$$ being a flux label.
In addition, the local field line pitch $B^\gamma/B^s$ is equal to its average value, the safety factor: $$q=s\ind{B}s\ind{j}FJ_{\Psi\chi\phi}\left<\frac{1}{R^2} \right>$$
The sign of the safety factor depends on the relative direction of the magnetic field and plasma current.

\subsection{Transformation to field aligned coordinates}
The following transformation is then applied
$$\zeta=qs-\gamma$$
for which 
$$B^\zeta=\textbf{B}\cdot\nabla\zeta=qB^s-B^\gamma=0$$
%and
%$$\nabla\zeta=s\frac{\partial q}{\partial %\Psi}\nabla\Psi+\frac{q}{2\pi}\nabla\chi-\frac{\nabla\phi}{2\pi}-\frac{F}{2\pi}\frac{\nabla\chi}{\textbf{B}\cdot\nabla\chi}\left[\left<\frac{1}{R^2} \right> -\frac{1}{R^2}\right]
%- \frac{\partial \gamma}{\partial \Psi}\nabla\Psi$$
%- \frac{\nabla \Psi }{2\pi}\frac{\partial F}{\partial \Psi}\int_0^\chi\frac{{\rm d}\chi}{\textbf{B}\cdot\nabla\chi}\left[\left<\frac{1}{R^2} \right> -\frac{1}{R^2}\right]$$
The right handed coordinate system $\left(\Psi,\zeta,s\right)$ is field aligned and its jacobian is $$J_{\Psi\zeta s}=J_{\Psi s \gamma}=4\pi^2J_{\Psi\chi\phi}$$

\section{Calculation of $(\Psi,\zeta,s)$ and of the associated metric}
\subsection{Construction of $s$ and $\nabla s$}
The construction of $s$ from $\chi$ is straightforward, as well as the calculation of $\nabla s$:
\begin{align*}
s&=\frac{\chi}{2\pi}\\
\nabla s &= \frac{1}{2\pi}\nabla\chi\\
\end{align*}
By definition, $s$ increases in the same direction as the poloidal angle $\theta$.

\subsection{Construction of $\zeta$ and $\nabla \zeta$}
Remembering $q=s\ind{B}s\ind{j}FJ_{\Psi\chi\phi}\left<\frac{1}{R^2}\right>$, it can be shown that $\zeta$ simplifies to:
$$\zeta=-\frac{\phi}{2\pi}+s\ind{B}s\ind{j}\frac{F(\Psi)J_{\Psi\chi\phi}(\Psi)}{2\pi}\int_0^\chi \frac{1}{R^2(\Psi,\tilde{\chi})}\,{\rm d}\tilde{\chi}$$
The construction of $\zeta$ therefore only requires to know:
$$\,F(\Psi),\,J_{\Psi\chi\phi}(\Psi) \textrm{ and } \mathcal{A}(\Psi,\chi)=\int_0^\chi \frac{1}{R^2(\Psi,\tilde{\chi})}\,{\rm d}\tilde{\chi}$$
The last term can be calculated from the expression of $g^{33}=|\nabla\phi|^2=\frac{1}{R^2}$ given on the $(\Psi,\chi,\phi)$ grid.

For the calculation of $\nabla\zeta$, we have:
\begin{align*}
\frac{\partial\zeta}{\partial\Psi}&=s\ind{B}s\ind{j}\frac{1}{2\pi}\left[ \frac{\partial F}{\partial \Psi}J_{\Psi\chi\phi}\mathcal{A}  + F\frac{\partial J_{\Psi\chi\phi}}{\partial \Psi}\mathcal{A}
+FJ_{\Psi\chi\phi}\frac{\partial \mathcal{A}}{\partial\Psi}\right]\\ 
\frac{\partial\zeta}{\partial\chi}&=s\ind{B}s\ind{j}\frac{FJ_{\Psi\chi\phi}}{2\pi}\frac{1}{R^2}\\ 
\frac{\partial\zeta}{\partial\phi}&=-\frac{1}{2\pi}\\ 
\end{align*}
The calculation of $\nabla \gamma$ therefore requires three additional terms: 
$$\frac{\partial F}{\partial\Psi},\, \frac{\partial J_{\Psi\chi\phi}}{\partial\Psi} \textrm{ and } \frac{\partial \mathcal{A}}{\partial\Psi}$$

\subsection{Metric tensor}
For any axisymetric function $h(\Psi,\chi)$, we have:
\begin{align*}
\frac{\partial h}{\partial \zeta}&=\frac{\partial h}{\partial \Psi}\frac{\partial \Psi}{\partial \zeta}+\frac{\partial h}{\partial \chi}\frac{\partial \chi}{\partial \zeta}=0 \\
\frac{\partial h}{\partial s}&=2\pi\frac{\partial h}{\partial \chi}
\end{align*}
The metric elements are calculated in CHEASE (in the \texttt{hamada.f90} module) using:
\begin{align*}
g^{\Psi\Psi}&=g^{\Psi\Psi}\\
g^{\Psi\zeta}&=g^{\Psi\Psi}\frac{\partial\zeta}{\partial \Psi}+g^{\Psi\chi}\frac{\partial\zeta}{\partial \chi}\\
g^{\Psi s}&=\frac{g^{\Psi\chi}}{2\pi}\\
g^{\zeta\zeta}&=g^{\Psi\Psi}\left(\frac{\partial\zeta}{\partial \Psi}\right)^2+g^{\chi\chi}\left(\frac{\partial\zeta}{\partial \chi}\right)^2
	+\frac{g^{\phi\phi}}{4\pi^2}+2g^{\Psi\chi}\frac{\partial\zeta}{\partial \Psi}\frac{\partial\zeta}{\partial \chi}\\
g^{\zeta s}&=\frac{g^{\Psi\chi}}{2\pi}\frac{\partial\zeta}{\partial \Psi}+\frac{g^{\chi\chi}}{2\pi}\frac{\partial\zeta}{\partial \chi}\\
g^{s s}&=\frac{g^{\chi\chi}}{4\pi^2}\\
\end{align*}
These metric elements are an output of CHEASE with $s\ind{B}>0$ and $s\ind{j}>0$. In GKW, the radial coordinate is not $\Psi$, but $\varepsilon=\frac{1}{2R\ind{ref}}(R\ind{max}-R\ind{min)}$
and the signs of the magnetic field and the current are taken into account, leading to the following modification (done in the \texttt{geom.F90} module of GKW) of the metric elements:
\begin{align*}
g^{\varepsilon\varepsilon}_{GKW}&=\left(\frac{\partial\varepsilon}{\partial\Psi}\right)^2g^{\Psi\Psi}\\
g^{\varepsilon\zeta}_{GKW}&=s\ind{B}s\ind{j}\frac{\partial\varepsilon}{\partial\Psi}g^{\Psi\zeta}\\
g^{\varepsilon s}_{GKW}&=\frac{\partial\varepsilon}{\partial\Psi}g^{\Psi s}\\
g^{\zeta\zeta}_{GKW}&=g^{\zeta\zeta}\\
g^{\zeta s}_{GKW}&=s\ind{B}s\ind{j}g^{\zeta s}\\
g^{s s}_{GKW}&=g^{s s}\\
\end{align*}

\subsection{Other metric elements}
\subsubsection{$\mathcal{F}$}
$$\mathcal{F}=\frac{R\ind{ref}B^s}{B}=\frac{s\ind{j}R\ind{ref}}{2\pi BJ_{\Psi\chi\phi}}=\frac{2\pi s\ind{j} R\ind{ref}}{ BJ_{\Psi\zeta s}}$$
\subsubsection{$\mathcal{D}^\alpha$}
$$\mathcal{D}^\alpha=-\frac{2}{B}\mathcal{E}^{\alpha\beta}\frac{\partial B}{\partial x^\beta}$$
\subsubsection{$\mathcal{E}^{\alpha\beta}$}
Noting that $\textbf{B}$ can be written:
$$\textbf{B}=J_{\Psi\zeta s}B^s\nabla\Psi\times\nabla\zeta=4\pi^2J_{\Psi\chi\phi}B^s\nabla\Psi\times\nabla\zeta=2\pi s\ind{j}\nabla\Psi\times\nabla\zeta$$
we have:
$$\mathcal{E}^{\alpha\beta}=\frac{R\ind{ref}^2}{2BB\ind{N}}(\nabla x^\alpha\times\nabla x^\beta) \cdot \textbf{B}=s\ind{j}\frac{2\pi R\ind{ref}^2}{2BB\ind{N}}(\nabla x^\alpha\times\nabla
x^\beta)\cdot(\nabla\Psi\times\nabla\zeta)$$
$$\mathcal{E}^{\alpha\beta}=s\ind{j}\frac{\pi R\ind{ref}^2}{BB\ind{N}}(g^{\alpha\Psi}g^{\beta\zeta}-g^{\alpha\zeta}g^{\beta\Psi})$$
\subsubsection{$\mathcal{G}$}
As $\textbf{B}\cdot\nabla B=\frac{s\ind{j}}{2\pi J_{\Psi\chi\phi}}\frac{\partial B}{\partial s}$, we have:
$$\mathcal{G}=\frac{R\ind{ref}\textbf{B}\cdot\nabla B}{B^2}=\frac{1}{B}\frac{\partial B}{\partial s}\mathcal{F}$$
\subsubsection{$\mathcal{H}^\alpha$}
To calculate $\mathcal{H}^\alpha$, we need:
$$\nabla Z\cdot \nabla x^\alpha=(\nabla Z)^\alpha=\frac{\partial Z}{\partial \Psi}g^{\Psi\alpha} + \frac{\partial Z}{\partial s}g^{s\alpha} $$ 
%and
%$$|\nabla Z|^2=g_{\alpha\beta}(\nabla Z)_\alpha(\nabla Z)^\beta=\frac{1}{g^{\alpha\beta}}(\nabla Z)^\alpha(\nabla Z)^\beta$$
to obtain:
\begin{align*}
\mathcal{H}^\alpha&=\frac{R\ind{ref}^2}{B\ind{N}v\ind{thref}}\mathbf{\Omega_\perp}\cdot\nabla x^\alpha \\
\mathcal{H}^\alpha&=-s\ind{B}\frac{R\ind{ref}^2}{B\ind{N}v\ind{thref}}\frac{\Omega}{|\nabla Z|}\left( \nabla Z \cdot \nabla x^\alpha - \delta_{\alpha
s}\frac{\mathcal{F}^2}{R\ind{ref}^2}\frac{\partial Z}{\partial s}\right)\\
\end{align*}
\subsubsection{$\mathcal{I}^\alpha$}
Using again $\textbf{B}=2\pi s\ind{j}\nabla\Psi\times\nabla\zeta$, we have:
\begin{align*}
\mathcal{I}^\alpha&=\frac{R\ind{ref}^2\Omega^2}{v\ind{thref}^2B\ind{N}}\frac{R}{B}(\nabla x^\alpha\times\nabla R)\cdot\mathbf{B}\\
\mathcal{I}^\alpha&=s\ind{j}\frac{R\ind{ref}^2\Omega^2}{v\ind{thref}^2B\ind{N}}\frac{2\pi R}{B}\left( g^{\Psi\alpha}\nabla R\cdot\nabla\zeta - g^{\zeta\alpha}\nabla R\cdot\nabla\psi \right)\\
\end{align*}
which can be written as a function of $\mathcal{E}^{\alpha\beta}$:
$$\mathcal{I}^\alpha=\frac{2R\Omega^2}{v\ind{thref}^2}\mathcal{E}^{\alpha\beta}\frac{\partial R}{\partial x^\beta}$$

%*********************************************************************
\part{CHEASE modification}
%*********************************************************************
\section{New quantities}
\subsection{$\beta_{\Psi \chi}\left(\theta=0\right)$}
In CHEASE, the non orthogonality $\beta_{\Psi \chi}$ is defined as 
$$\beta_{\Psi \chi}=\frac{\nabla\Psi\cdot\nabla\chi}{\left|\nabla \Psi \right|^2}$$
The quantity BCHIN, calculated in \texttt{surface.f90} on the CHIN mesh, corresponds to $$\beta_{\Psi \chi}-\beta_{\Psi \chi}\left(\theta=0\right)$$
Even if usually small, $\beta_{\Psi \chi}\left(\theta=0\right)$ is not necessarily zero and has to be calculated. \\
As the line $\theta=0$ is a horizontal line, we have $\nabla\chi\cdot\nabla R|_{\theta=0}=0$, and hence:
$$\beta_{\Psi \chi}\left(\theta=0\right)=\frac{1}{|\nabla \Psi|^2}|\nabla \chi|\frac{\partial{\Psi}}{\partial{Z}}$$
Moreover, we have:
$$\frac{1}{J_{\Psi\chi\phi}^2}=\det\left(g^{ij}\right)=\frac{|\nabla \Psi|^2}{R^2}\left(|\nabla \chi|^2 - \beta_{\Psi \chi}^2 |\nabla \Psi|^2\right)$$
From which we can deduce:
$$\beta_{\Psi \chi}\left(\theta=0\right)=\frac{R}{J_{\Psi\chi\phi}|\nabla\Psi|^3}\frac{\partial{\Psi}}{\partial{Z}}\frac{1}{\sqrt{1-\frac{1}{|\nabla\Psi|^2}\left(\frac{\partial{\Psi}}{\partial Z}\right)^2}}$$
The calculation of $\beta_{\Psi \chi}\left(\theta=0\right)$, variable ZBETCHI0, has been added in \texttt{chipsimetrics.f90}.

\subsection{Partial derivatives of $R$ and $Z$}
From $\nabla R\cdot \nabla \Psi=\frac{\partial \Psi}{\partial R}$ and $\nabla R\cdot \nabla \Psi= \frac{\partial R}{\partial \Psi}|\nabla \Psi|^2 + \frac{\partial R}{\partial \chi}|\nabla \Psi|^2 \beta_{\Psi \chi}$, we can write:
$$\frac{\partial R}{\partial \Psi}= \frac{1}{|\nabla \Psi|^2}\left( \frac{\partial \Psi}{\partial R} - \frac{\partial R}{\partial \chi}|\nabla \Psi|^2 \beta_{\Psi \chi}\right)$$
Then, $\frac{\partial R}{\partial \chi}$ can be calculated from:
$$\frac{\partial R}{\partial \chi}=\frac{\partial \sigma}{\partial \chi}\rho_s\cos{\theta}-\sigma\rho_s\frac{\partial \theta}{\partial \chi}\sin{\theta} +\sigma  \frac{\partial \rho_s}{\partial\theta} \frac{\partial\theta}{\partial\chi}\cos{\theta}$$
and
\begin{align*}
\frac{\partial\sigma}{\partial\chi}&=-\frac{1}{|A|}\frac{\partial\Psi}{\partial\theta}\\
\frac{\partial\theta}{\partial\chi}&=\frac{1}{|A|}\frac{\partial\Psi}{\partial\sigma}\\
\end{align*}
with 
$$|A|=\frac{\partial \Psi}{\partial \sigma}\frac{\partial\chi}{\partial\theta} - \frac{\partial\Psi}{\partial\theta}\frac{\partial\chi}{\partial\sigma}= \frac{R\sigma\rho_s^2}{J_{\Psi\chi\phi}}$$
So finally,
$$\frac{\partial R}{\partial \Psi}=\frac{1}{|\nabla \Psi|^2}\frac{\partial \Psi}{\partial R}-\frac{1}{|A|}\beta_{\Psi \chi}\left( - \frac{\partial \Psi}{\partial \theta}\rho_s\cos{\theta}-\sigma\rho_s\frac{\partial \Psi}{\partial \sigma}\sin{\theta} +\sigma  \frac{\partial \rho_s}{\partial\theta} \frac{\partial\Psi}{\partial\sigma}\cos{\theta}\right)$$
A similar calculation leads to $\frac{\partial Z}{\partial \Psi}$ and $\frac{\partial Z}{\partial \chi}$. All these quantities are computed in \texttt{chipsimetrics.f90} and named ZDRDPSI, ZDRDCHI, ZDZDPSI, ZDZDCHI, respectively.

\subsection{$\mathcal{A}$ and $\partial\mathcal{A}/\partial\Psi$}
The integrals
$$\mathcal{A}(\Psi,\chi)=\int_0^\chi \frac{1}{R^2}\,{\rm d}\tilde{\chi}$$
and 
$$\frac{\partial\mathcal{A}}{\partial \Psi}(\Psi,\chi)=-2\int_0^\chi \frac{1}{R^3}\frac{\partial R}{\partial \Psi}\,{\rm d}\tilde{\chi}$$
are calculated in \texttt{surface.f90} (variables AH and AHPR) using the expression of $\frac{\partial R}{\partial \Psi}$ derived above and performing gaussian integration on the $\Psi=C\expo{te}$ surfaces. A correction (variable AHPRCOR) is applied to $\frac{\partial\mathcal{A}}{\partial \Psi}$ in \texttt{chipsimetrics.f90} to take into account equilibria with non-zero $\beta_{\Psi \chi}\left(\theta=0\right)$.

\section{Output file}
The output file containing the metric elements for the $(\Psi,\zeta,s)$ coordinate system is named \texttt{hamada.dat} and is created in \texttt{hamada.f90}. All quantities are given in MKSA units. The output file content is described below.

\textbf{Scalars}\\
NPSI: number of points for the radial grid\\
NS: number of points for the poloidal grid (specified with NCHI in the CHEASE namelist)\\
R0EXP: reference major radius used in CHEASE, generally taken to be close to the magnetic axis radius for faster convergence (but the actual position is not necessarily known beforehand)\\
B0EXP: reference magnetic field in CHEASE defined as the magnetic field at R0EXP\\
Raxis: magnetic axis radius as calculated by CHEASE
\vspace{0.5cm}

\textbf{Grids}\\
PSI: $\Psi$, the poloidal flux, zero at the magnetic axis, increasing towards the edge. The $\Psi$ grid given in output is equidistant in $\sqrt{\Psi}$ (except if mesh densification is used). The
first point is NOT on the magnetic axis (it is at $\sqrt{\Psi}=0.5*\sqrt{\Psi\ind{\max}}/\mathrm{NPSI}$, while on the magnetic axis $\sqrt{\Psi}=0$). The steps of the grid are
$\sqrt{\Psi\ind{\max}}/\mathrm{NPSI}$, except for the last one wich is half of it to have the last point on the last closed flux surface.\\
S: $s$, as defined in GKW. The grid is equidistant and for a given flux surface $s=0$ is on the low field side, at the height of the magnetic axis. 
\vspace{0.5cm}

\textbf{1D arrays}\\
Rgeom: major radius of the flux surface centre, defined as $(R\ind{max}+R\ind{min})/2$\\
amin: minor radius of the flux surface, defined as $(R\ind{max}-R\ind{min})/2$\\
damindpsi: $\partial \mathrm{amin}/\partial \Psi$\\
d2amindpsi2: $\partial^2 \mathrm{amin}/\partial \Psi^2$\\
bmax: maximum of the magnetic field strength on the flux surface\\
bmin: minimum of the magnetic field strength on the flux surface\\
q: safety factor\\
dqdpsi: $\partial q /\partial \Psi$\\
dqdpsi\_chk: value of $\partial \zeta / \partial \Psi$ at $s=1$ (equal to $\partial q /\partial \Psi$)\\
p: plasma pressure\\
dpdpsi: $\partial p /\partial \Psi$\\
jac: $J_{\Psi\zeta s}$ the jacobian of the $(\Psi,\zeta,s)$ coordinates\\
djacdpsi: $\partial J_{\Psi\zeta s} /\partial \Psi$\\
f: $F=RB\ind{t} = R^2|\mathbf{B}\cdot\nabla\phi|$ (always positive by convention)\\
dfdpsi: $\partial F /\partial \Psi$\\

\textbf{2D arrays}\\
g11: $\nabla \Psi \cdot \nabla \Psi$\\
g12: $\nabla \Psi \cdot \nabla \zeta$\\
g13: $\nabla \Psi \cdot \nabla s$\\
g22: $\nabla \zeta \cdot \nabla \zeta$\\
g23: $\nabla \zeta \cdot \nabla s$\\
g33: $\nabla s \cdot \nabla s$\\
B: the norm of the magnetic field\\
dBdpsi: $\partial B/\partial \Psi$\\
dBds: $\partial B/\partial s$\\
R: major radius of the $(\Psi,s)$ grid points (which describe flux surfaces) \\
Z: height of the $(\Psi,s)$ grid points (which describe flux surfaces) \\
dRdpsi: $\partial R/\partial \Psi$\\
dRds: $\partial R/\partial s$
dZdpsi: $\partial Z/\partial \Psi$\\
dZds: $\partial Z/\partial s$\\
theta: the poloidal angle, zero at $s=0$, increasing anticlockwise and measured from the magnetic axis\\
Ah: $\mathcal{A}$ \\
dAhdpsi: $\partial \mathcal{A}/\partial \Psi$\\ 
dzetadpsi: $\partial \zeta/\partial \Psi$\\ 
dzetadchi: $\partial \zeta/\partial \chi$\\

\section{Tests}
The matlab file used to perform the checks can be found in \texttt{/home/camenen/chease/tests/} on \texttt{hal}.
\subsection{Consistency checks}
The following checks have been performed using the outputs stored in \texttt{hamada.dat}
\begin{itemize}
 \item The output LCFS is the same as the input LCFS
 \item The output $FF'$ is the same as the input $FF'$
 \item The partial derivatives computed in matlab on the output grid are quasi-identical (excpet close to the magnetic axis) to the one computed in CHEASE for the following quantities: $q$, $p$, $F$, $J_{\Psi\zeta s}$, $R$, $Z$, $B$ and $\mathcal{A}$.\\
 There is a significant disagreement ($10$ to $20\%$) for the derivatives of $R$, $Z$ and $B$ at $s$ values where the derivatives are almost zero.
 \item $$\frac{1}{J_{\Psi\zeta s}^2}=det\left( g^{ij}\right)_{i,j=\Psi,\zeta,s}$$
 \item $|\nabla R|^2=1$ and $|\nabla Z|^2=1$ 
 \item $B^2=4\pi^2\left[ g^{\Psi\Psi}g^{\zeta\zeta} - (g^{\Psi\zeta})^2\right]$ (validation of $\frac{\partial \zeta}{\partial \chi}$)
 \item $q=F J_{\Psi \chi \phi}\left< \frac{1}{R^2}\right>$, using $\left< \frac{1}{R^2}\right>=\frac{1}{2\pi}\mathcal{A}_{\chi=2\pi}$
 \item The cartesian coordinates $(x,y,z)$ are related to the cylindrical coordinates $(R,Z,\phi)$ by:
\begin{align*}
x&=R\cos\phi\\
y&=Z\sin\phi\\
z&=Z\\
\end{align*}
with $$\phi = 2\pi\left( - \zeta + FJ_{\psi\chi\phi}\int_0^s\frac{1}{R^2}\,\rm{d}s\right)$$
Using the definition $g_{ij}=\frac{\partial x^k}{\partial\xi ^i}\frac{\partial x^k}{\partial\xi ^j}$ with $(x^1,x^2,x^3)=(x,y,z)$ and $(\xi^1,\xi^2,\xi^3)=(\Psi,\zeta,s)$ all the relationships $g^{ik}g_{kj}=\delta_{ij}$ have been checked (validation of $\frac{\partial \zeta}{\partial \psi}$, $\frac{\partial \zeta}{\partial \chi}$ and strong consistency check). Note that:
\begin{align*}
&\left[\frac{\partial \phi}{\partial\psi}\right]_{\zeta, s}=2\pi\left[\frac{\partial \zeta}{\partial\psi}\right]_{\chi, \phi}\\
&\left[\frac{\partial \phi}{\partial s}\right]_{\Psi, \zeta}=4\pi^2\left[\frac{\partial \zeta}{\partial\chi}\right]_{\psi, \phi}\\
&\left[\frac{\partial \phi}{\partial\zeta}\right]_{\psi, s}=-2\pi\\
\end{align*}

\end{itemize}

\subsection{Comparison with LIUQE and Psi-toolbox}
CHEASE is run with the $FF'$ and $p'$ values from LIUQE. The Psi-toolbox is used to compute the metric of the LIUQE equilibrium.
\begin{itemize}
 \item The flux surface description $(R,Z)$ is similar for CHEASE and LIUQE (small difference for the position of the magnetic axis).
 \item For the two codes, $\left| \nabla \Psi\right|$, $\left<\left| \nabla \Psi\right|\right>$, $\left<\frac{1}{R^2}\right>$ and $\mathcal{A}$ are similar (less than $2\%$ difference).
 \item $J_{\Psi \zeta s}^{CHEASE}=\frac{\partial \mathcal{V}}{\partial \Psi}^{LIUQE}$
\end{itemize}

\subsection{Comparison with the low aspect-ratio shifted circular flux surfaces approximation}
CHEASE is run with a circular LCFS and the metric elements are compared to the analytical formulae obtained in the low aspect-ratio approximation for circular flux surfaces. The agreement is relatively good, especially in the plasma center, as expected.

\subsection{Tilted plasmas and $\beta_{\Psi \chi}\left(\theta=0\right)$}
To check the calculation of $\beta_{\Psi \chi}\left(\theta=0\right)$, the input LCFS used for CHEASE has been rotated by $\pi/8$ to have a finite value of $\beta_{\Psi \chi}\left(\theta=0\right)$. The value of $\frac{\partial R}{\partial {\Psi}}$ obtained in CHEASE is then compared to the one directly calculated in matlab using the values of $R$ 
on the $(\Psi, \chi)$ grid. The comparison shows that if the finite $\beta_{\Psi \chi}\left(\theta=0\right)$ correction is not included, there is a significant discrepancy between the values of $\frac{\partial R}{\partial {\Psi}}$ calculated in CHEASE and matlab, whereas the agreement is perfect when the correction is included.
Note that $\beta_{\Psi \chi}\left(\theta=0\right)$ is also involved in the calculation of $g^{\Psi\chi}$ and has a strong impact on this parameter.

\section{List of modified files}
The following files have been modified: \texttt{cotrol.f90}, \texttt{chipsimetrics.f90}, \texttt{g\_2.f90}, \texttt{globals.f90}, \texttt{Makefile}, \texttt{mappin.f90}, \texttt{ogyropsi.f90}, \texttt{stepon.f90}, \texttt{surface.f90}, \texttt{write\_ogyropsi.f90} and \texttt{write\_ogyropsi\_hdf5.f90} \\
The following file has been created:
\texttt{hamada.f90}

%*********************************************************************
\part{GKW modification}
%*********************************************************************
Most of the modifications have been done in the subroutine \texttt{geom\_init\_grids} of the module \texttt{geom.f90} where the calculation of the metric elements and related quantities from the chease output file has been implemented. 
\section{Namelist GEOM}
The namelist GEOM has been extended to include the following variables:
\begin{itemize}
 \item GEOM\_TYPE\\
It is used to specify whether the metric elements and related quantities are computed using the s-alpha model or using the output of an MHD equilibrium solver. At present, two options are implemented: 's-alpha' and 'chease'.
 \item EQFILE\\
It contains the path and name of the chease output file from which the MHD equilibrium related quantities are read when GEOM\_TYPE has the value 'chease'
 \item SIGNB\\
It contains the sign of $\bf{B}\cdot\nabla\phi$
 \item SIGNJ\\
It contains the sign of $\bf{j}\cdot\nabla\phi$, where $\bf{j}$ is the plasma current density.

\end{itemize}
With the 'chease' option, the value of EPS specified in the namelist GEOM is used to find the corresponding $q$ and $\hat{s}$ from the chease output file (i.e. the Q and SHAT given in the namelist are not used). Note that EPS is defined as follow:
$$\mathrm{EPS}=\frac{r}{R\ind{ref}}=\frac{R\ind{max}-R\ind{min}}{2R\ind{axis}}$$
To be consistent, this definition of EPS must be used to calculate the normalised gradients given in the SPECIES namelist
Note that in chease, the radial coordinate is $\Psi$, not $\epsilon$, which means that a factor $\frac{\partial \epsilon}{\partial \Psi}$ has to be added in various places to the quantities loaded from the chease output file.

\section{Grids}
The chease outputs are given on a $(\Psi,s)$ grid. To ensure that every point of the GKW $s$-grid corresponds to a point of the CHEASE $s$-grid, the following conditions are needed:
\begin{itemize}
 \item NS must be odd
 \item $\frac{\mathrm{NS}}{2\mathrm{NPERIOD}-1}$ has to be an integer
 \item $\mathrm{NCHI}=2k\frac{\mathrm{NS}}{2\mathrm{NPERIOD}-1}$ with $k$ being an non-zero integer (NCHI is used in the chease input file to specify the number of points for the $s$-grid)
\end{itemize}
These conditions are checked in the subroutine \texttt{geom\_init\_grids} of the module \texttt{geom.f90}

\section{$\beta$ and $\beta'$}
The BETA parameter specified in the namelist SPCGENERAL is used in the Amp\`ere law and has the following definition:
$$\mathrm{BETA}=\frac{2\mu\ind{0}n\ind{ref}T\ind{ref}}{B\ind{ref}^2}$$
For later use, the variable BETA\_REAL has been created and is loaded from the chease output file. It is defined as:
$$\mathrm{BETA\_REAL}=\frac{2\mu\ind{0}p}{B\ind{ref}^2}$$
where $p$ is the total ``MHD'' plasma pressure (i.e. the pressure solution of the Grad-Shafranov equation) and $B\ind{ref}$ is taken to be $B\ind{axis}$, the magnetic field strength at the magnetic axis position. \\
The value of $\beta'$ is needed to calculate part of the curvature drift, with
$$\beta'=\frac{2\mu\ind{0}}{B^2}\frac{\partial p}{\partial \epsilon}$$
From the chease output, the variable 
$$\mathrm{BETA\_PRIME\_REAL}=\frac{2\mu\ind{0}}{B\ind{axis}^2}\frac{\partial p}{\partial \epsilon}$$
is loaded and $\mathrm{BETA\_PRIME}=\mathrm{BETA\_PRIME\_REAL}$ is then used in \texttt{components.f90}. So we have:
$$\beta'=\mathrm{BETA\_PRIME}\frac{B\ind{axis}^2}{B^2}$$

\section{Output file}
The subroutine \texttt{write\_geom} has been added in the \texttt{diagnostics.f90} module. It writes an output file named \texttt{geom.dat} and containing the quantities related to the MHD equilibrium.

\section{Tests}
\begin{itemize}
 \item Check that $\mathcal{E}^{\epsilon\zeta}=\frac{R\ind{ref}^2B\ind{ref}}{4\pi}\frac{\partial\epsilon}{\partial \Psi}$.
\end{itemize}

\bibliographystyle{plain}
\bibliography{gkwandchease}

\end{document}

### 2022-05-03 gkw/doc/chease/runchease.tex
(1) Added gkw Clone gwk  #3

\documentclass[a4paper,12pt]{article}

\newcommand{\ind}[1]{\ensuremath{_{\rm{#1}}}}
\newcommand{\expo}[1]{\ensuremath{^{\rm{#1}}}}

\usepackage{amsmath}
\usepackage{amssymb}

\setlength\parindent{0in}
\oddsidemargin 0.0in
\evensidemargin 1.0in
\textwidth 6.0in
\textheight 9.0in
\headheight 0.0in
\topmargin 0.0in

% Title Page
\title{Running GKW for actual tokamak geometry}
\author{Y. Camenen}
\date{March 6, 2009\\ Updated on June 9, 2011}

\begin{document}
\maketitle
This document describes how to to run GKW for actual tokamak geometry getting the metric elements from the CHEASE code for toroidal MHD equilibria \cite{Lutjens:Comput1992,Lutjens:Comput1996}.

%*********************************************************************
\part*{Fast Track}
%*********************************************************************
Get the source, compile, run a test case and check the results. It is assumed that you already have the latest GKW version in the \texttt{./gkw} folder.

\section{Get CHEASE}
The CHEASE code has been developed and is maintained at CRPP, Lausanne. A copy of the latest version can always be obtained on the public svn server via:\\
\texttt{\small svn export https://crppsvn.epfl.ch/repos/CHEASE/chease/trunk chease}\\
It will copy  the files on your machine in the \texttt{./chease/} directory. You can use \texttt{co} instead of \texttt{export} to also get the svn links.

Alternatively, a tagged CHEASE version that has been tested with GKW can be found under:\\
\texttt{\small svn export https://crppsvn.epfl.ch/repos/CHEASE/chease/tags/CHEASE\_90\_9\_3 chease}\\
This tag, and the latest version of CHEASE will work with GKW r1459 or later.

The latest CHEASE version includes the most recent developments and bug fixes, but unlike the tagged version, it has not necessarily been extensively tested with GKW (but test cases are now available to ensure that the latest CHEASE version produces the expected results).

Note that CHEASE versions after tag \texttt{CHEASE\_90\_7\_5} and before tag \texttt{CHEASE\_90\_9\_3} had a bug that prevents their use to compute metric elements for GKW. Using these versions will produce wrong results!

Note also that the GKW version in the CPC journal repository, the GKW /branch/cpc, and versions prior to r1459 only work with the CHEASE version tagged \texttt{CHEASE\_90\_7\_3}.

\section{Compile CHEASE}

\subsection{CHEASE\_90\_7\_3, CHEASE\_90\_7\_1}

For reference, use with CPC version of GKW.

Once the sources on your local machines, the compilation should be straightforward. The \texttt{Makefile} is very simple and the target machine is selected by commenting/uncommenting the appropriate block (you may need to add a block for your machine).
Note that provided the HDF5 and FUTILS libraries are installed, it is possible to use HDF5 to output the results of the CHEASE run. This option is not required (and not used) for the interface with GKW but could be implemented relatively easily.

\subsection{CHEASE\_90\_7\_5}

Use only if you have a good reason. It is recommended to use instead the latest version or tag \texttt{CHEASE\_90\_9\_3} which benefits from several improvements. Works with newer versions of GKW (from r1459).

Unfortunately this tagged version does not compile exactly as checked out.  To compile you first need to:
\texttt{ln -s euitm\_schemas\_407b.f90 euitm\_schemas.f90} and uncomment the lines in the build rules at the bottom relating to \texttt{euitm\_schemas}.

Use the following for CSC warwick desktop:
\begin{verbatim}
 # CSC desktop Warwick
F90 = mpif90
F90FLAGS = -r8 -O3 -I$(DIR_interpos)
LDFLAGS =
LIBS=-L$(DIR_interpos) -linterpos
F90FLAGS_parser = -O1
\end{verbatim}
An appropriately modified makefile is provided with GKW in \texttt{/doc/chease/chease\_90\_7\_5.makefile}

\subsection{CHEASE\_90\_9\_3 or latest version in the trunk}

These are the recommended CHEASE versions.\\
Works with newer versions of GKW (from r1459).  Should compile easily if you add a few lines for your machine and compiler in the \texttt{Makefile.define\_*} files. The HDF5 and FUTILS libraries are not required for the interface with GKW.

\section{Run the test cases}

To run the test cases desribed in this section, you will need \texttt{CHEASE\_90\_9\_3} or a later version.

\subsection{Create working directory}
To start with, create a working directory, for instance:\\
\texttt{\small mkdir ./chease/runs/}

\subsection{Input files}
CHEASE solves the Grad-Shafranov equation in toroidal geometry taking as an input the last closed flux surface description, the radial derivative of the pressure profile and the current profile. These quantities can be specified in various ways, the most common being the use of an EXPEQ or an EQDSK file. CHEASE also needs an input file containing the values of the various switches (fortran namelist).\\

For the first run, we will use the input files provided in\\
\texttt{./chease/WK/TESTCASES/NIDEAL10/input/}.\\
There are two test cases available: \texttt{splus} and \texttt{tcv.snd1}. 
To run the test case \texttt{splus}, you will need to copy the corresponding EXPEQ and namelist file in the working directory:\\
\texttt{\small cp ./chease/WK/TESTCASES/NIDEAL10/input/EXPEQ.splus ./chease/runs/EXPEQ}\\
\texttt{\small cp ./chease/WK/TESTCASES/NIDEAL10/input/i.splus ./chease/runs/chease\_namelist}\\
Note that the EXPEQ file and the namelist file HAVE TO be named \texttt{EXPEQ} and \texttt{chease\_namelist}, respectively.

\subsection{Run CHEASE}
Before running CHEASE, You will need to copy the executable compiled in the previous section in the working directory:\\
\texttt{\small cp ./chease/src-f90/chease ./chease/runs/}\\
Then, launch the executable (you must be in the working directory):\\
\texttt{\small cd ./chease/runs/}
\texttt{\small ./chease | tee o.chease}\\
A lot of outputs are displayed on the screen and stored in the file \texttt{o.chease}. The file \texttt{hamada.dat} is also created, it contains all the elements required for the metric coefficients calculation in GKW. All the quantities in \texttt{hamada.dat} are in SI units.

\subsection{Check the results}
The expected outputs are stored in\\
 \texttt{./chease/WK/TESTCASES/NIDEAL10/output\_ref/}.\\
If you are using matlab, you can use the script\\
\texttt{./chease/WK/TESTCASES/NIDEAL10/matlab/chease\_diff.m}\\
to check that you obtained the correct result (as CHEASE is quite sensitive to compiler and compilation options, a standard 'diff' will not do the job...). Just type:
\texttt{\small >> chease\_diff(1e-8,'splus',' ./chease/WK/TESTCASES/NIDEAL10/output\_ref/','','./chease/runs/');}\\
If you get no warning, all is fine.

\section{Matlab scripts}
A couple of matlab scripts intended to be used in combination with CHEASE can be found in the directory \texttt{./gkw/matlab/chease/} that you naturally have downloaded from the GKW svn repository. These scripts are:
\begin{itemize}
 \item \texttt{read\_eqdsk.m} to read the content of an EQDSK file
 \item \texttt{read\_expeq.m}  to read the content of an EXPEQ file
 \item \texttt{write\_expeq.m} to write the content of a matlab structure (with the appropriate format) into an EXPEQ file.
 \item \texttt{read\_hamada.m} to read the content of the hamada.dat file (CHEASE output when NIDEAL=10)
 \item \texttt{read\_neoart.m} to read the content of the neoart.dat file (another CHEASE output when NIDEAL=10)
 \item \texttt{chease\_diff.m} to compare the outputs hamada.dat and neoart.dat to reference files
\end{itemize}
The three last scripts can also be obtained from the CHEASE svn repository.\\

Note that I wrote the matlab routines to examine the input/output files in a very non-general way as they were not intially intended to be distributed. They may still be useful, take what you want...\\

You can for instance load the content of \texttt{hamada.dat} using the matlab routine \texttt{read\_hamada.dat} and plot the magnetic shear profile as a function of the normalised radius $\rho_\psi$ (where $\psi$ is the poloidal flux):\\
\texttt{\small >> H1=read\_hamada('hamada.dat','./chease/runs/');}\\
\texttt{\small >> plot(sqrt(H1.psi./max(H1.psi)),H1.amin./H1.q.*H1.dqdpsi./H1.damindpsi);}\\
It is of course crucial to check that the output is consistent with what you expect, in particular if you want to make comparisons with the experiments and/or use rescaling of the equilibrium in CHEASE.

\section{Run GKW}
Before running GKW, you need to specify in the \texttt{GEOM} namelist that 1) you want to use CHEASE, 2) where is the output file containing the metric (I assume here that GKW is launched from the working directory \texttt{./chease/runs/}):\\
\texttt{\small GEOM\_TYPE = 'chease'}\\
\texttt{\small EQFILE = '../WK/TESTCASES/NIDEAL10/output\_ref/hamada.splus.dat '}\\
The flux surface of interest is specified using its \texttt{EPS} parameter. Depending on the value (1 or 2) of the \texttt{EPS\_TYPE} parameter, \texttt{EPS} is either the local inverse aspect ratio (i.e. the $\psi$ coordinate of GKW) or $\rho_\Psi$ the square root of the normalised poloidal flux ($\rho_\Psi=0$ on the axis and $\rho_\Psi=1$ on the last closed flux surface).
Note that when 
$$\texttt{EPS}=\psi=\frac{R\ind{max}-R\ind{min}}{2R\ind{ref}}$$
is used, $R\ind{ref}$ is taken to be the parameter \texttt{R0EXP} used for the normalisations in the CHEASE inputs. The values given in the GKW input file for \texttt{Q} and \texttt{SHAT} are not used and taken directly from CHEASE.\\

Now let's try to run GKW using the reference input file in the \texttt{./gkw/doc/chease} folder.\\
\texttt{\small cp ./gkw/doc/chease/input.gkw.chease ./chease/runs/input.dat}\\
Then copy your GKW executable in the working directory and run it. If all goes well, you should be thrown out of the code with the message\\
\texttt{Radial grid too coarse in CHEASE: increase NPSI}.\\
This is because GKW looks in the file \texttt{hamada.splus.dat} for a flux surface with the specified \texttt{EPS} value. If a flux surface is not found close enough to the specified value, the code stops. To avoid that, you can either change the value of \texttt{EPS} in the GKW input, or rerun CHEASE taking care to have a radial grid point where you want (i.e. you increase \texttt{NPSI} or you use packing with $\texttt{NMESHA}=4$). To look where are the CHEASE grid points:\\
\texttt{\small >> H1=read\_hamada('hamada.dat','./chease/runs/');}\\
\texttt{\small >> plot(sqrt(H1.psi./max(H1.psi)),H1.amin./H1.r0exp,'+');}\\
You can then choose $\texttt{EPS}=0.75$ (keeping \texttt{EPS\_TYPE=2}) in the file \texttt{input.dat} and try again. This time, it should work. The output file \texttt{geom.dat} gives the metric elements used in GKW on the $s$ grid. This file can be read in matlab using \texttt{read\_geom.m} available in \texttt{./gkw/matlab/}. The actual value used for \texttt{EPS} is given in \texttt{geom.dat}.\\

A last paragraph about the parallel grid. In CHEASE the number of points (for one poloidal turn) used for the parallel/poloidal grid is specified through the variable \texttt{NCHI} in the input namelist. To avoid interpolation in GKW, not all values of \texttt{N\_s\_grid} are allowed for a given \texttt{NCHI}. You must have: $$\texttt{NCHI}=2k \frac{\texttt{N\_s\_grid}}{2\texttt{NPERIOD}-1}$$
with $k$ a positive integer. This is a bit of a constraint, but to rerun CHEASE with another \texttt{NCHI} value takes less than a minute. It is only annoying when you want to scan \texttt{N\_s\_grid} or \texttt{NPERIOD} which does not happen that often...
Finally, \texttt{N\_s\_grid} must be odd to ensure that a point is at the outboard midplane (this constraint could be relaxed...).

%*********************************************************************
\part*{Playing with CHEASE}
%*********************************************************************
Different ways to run CHEASE have been developped over the years. Most of the available options are described in Ref.~\cite{Lutjens:Comput1996}, the others are to my knowledge not documented. Some options, useful to compute the metric for GKW, are briefly described here.

\section{CHEASE namelist}

A minimal namelist input file for CHEASE, with some comments, can be found in \texttt{./gkw/doc/chease/short.namelist.chease}.

\subsection{Normalisation}
The input parameter \texttt{R0EXP} in [m] is the approximate radial position of the magnetic axis (you do not necessarily know it precisely before running chease) and \texttt{B0EXP} in [T] is the value of the magnetic field at this position. These two parameters are used for the normalisations in CHEASE:
\begin{eqnarray*}
 R &=& R\ind{CHEASE} * \texttt{R0EXP}\\
 Z &=& Z\ind{CHEASE} * \texttt{R0EXP}\\
 \Psi &=& \Psi\ind{CHEASE} * \texttt{R0EXP}^2 * \texttt{B0EXP}\\
 T &=& T\ind{CHEASE} * \texttt{R0EXP} * \texttt{B0EXP}\\
 TT' &=& TT'\ind{CHEASE} * \texttt{B0EXP}\\
 p' &=& p'\ind{CHEASE} * \texttt{B0EXP}/(\mu_0*\texttt{R0EXP}^2)\\
 \mu_0 &=& 4\pi 10^{-7}
\end{eqnarray*}

\textbf{The normalising radius $R\ind{ref}$ in GKW is taken to be the parameter $\texttt{R0EXP}$.  When running with an EQDSK file, CHEASE uses the 5th value in the EQDSK file RCENTR for R0EXP, which may correspond to the limiter centre and not the magnetic axis.  This value can be overridden by setting a negative R0EXP in the chease input file.  You can always check the value of R0EXP used by chease in the hamada.dat file.  Since R0EXP=$R_{\rm Ref}$ in GKW, you need to to know its value in order to input eps and all the gradients correctly in GKW.}

\subsection{NIDEAL}
The parameter \texttt{NIDEAL} is used to specify the output of CHEASE. Two options are of interest here:\\
$\texttt{NIDEAL}=6$ to generate an EQDSK, file \texttt{EQDSK.OUT}\\
$\texttt{NIDEAL}=10$ to generate the metric for GKW, file \texttt{hamada.dat}\\

\subsection{Smoothing the inputs}
Very often the description of the last closed flux surface and of the pressure/current profiles has been obtained using cubic splines. Even if the fit looks smooth, standard splines can give rise to big jumps in the second derivative. Such jumps are a nightmare for the metric computation but fortunately, they can be avoided by using a spline fit with tension. This had previously to be done with the routines \texttt{bndfit} and \texttt{proffit} but is now done automatically in CHEASE.\\
If the parameter \texttt{TENSBND} is specified in the input namelist, its value is used for the tension in the fit of the plasma boundary.
The parameter \texttt{TENSPROF} can also be specified for the tension used in the fit of the profiles, but is not used at the time of writting. 
If \texttt{TENSBND=0.}, no tension is used. If the parameter is not specified in the namelist, default tension is used.\\

If you decide to specify the tension in the namelist, you NEED to check that the last closed flux surface in the output is reasonnably close to the input one. If you give too much tension, you will get too much smoothing. Too little tension and you will not remove the potential wiggles.

\subsection{Scaling of the equilibrium}
The current profile can be rescaled using the \texttt{NCSCAL} parameter (see \cite{Lutjens:Comput1996} p.231):\\
$\texttt{NCSCAL}=1$: matches the safety factor $q(\texttt{CSSPEC})$ to \texttt{QSPEC} where \texttt{CSSPEC} is in $\rho_\psi$.\\
$\texttt{NCSCAL}=2$: matches the total current value to $\texttt{CURRT}$ (CHEASE units, works only if $\texttt{NSTTP}\neq3$)\\
$\texttt{NCSCAL}=4$: no rescaling\\
The flux function $T=F=R B\ind{T}$ is also rescaled to have $T\ind{axis}=1$ if $\texttt{NTMF0}=1$ or $T\ind{edge}=1$ if $\texttt{NTMF0}=0$ (in CHEASE units). If you choose $\texttt{R0EXP}=R\ind{axis}$
and $\texttt{B0EXP}=B\ind{axis}$, the natural choice would be to have $\texttt{NTMF0}=1$.

\subsection{Packing}
In CHEASE, you can locally increase the resolution of the grids used for the mapping. For instance, you can increase the radial resolution around a rational flux surface or the poloidal resolution
around the X-point. I have added a new option for the packing in CHEASE to specify the exact position in $\rho_\psi$ of a few radial grid points, this to avoid guessing the
correct \texttt{EPS}
value in GKW. This option can be used with $\texttt{NMESHA}=4$. The number of flux surfaces around which the packing is done is then given by \texttt{NPOIDA} and their position ($\rho_\psi$: square
root of the poloidal flux, 0 in the center, 1 in the edge) by \texttt{APLACE}. The "width" of the densification is specified by \texttt{AWIDTH} and the total fraction of the radial grid which remains
undensified is given by \texttt{SOLPDA}. Hence, to densify around $\rho_\psi=0.4$ and $\rho_\psi=0.65$, you can use:\\
\texttt{\small NMESHA=4, NPOIDA=4, SOLPDA=.50,}\\
\texttt{\small APLACE=.4,.4,.65,.65, AWIDTH=.07,.03,.07,.03,}\\
where the double specification of each point with a progressive decrease of the densification width is a "standard" way to get a "good" densification (from O. Sauter).\\
The flux surface is then selected in GKW by using \texttt{EPS\_TYPE=2} and specifying the value of $\rho_\psi$ in \texttt{EPS}.

\section{EXPEQ file}
The EXPEQ file is used as an input in  CHEASE if $\texttt{NSURF}=6$ and $\texttt{NEQDSK}=0$. All quantities in the EXPEQ file are in CHEASE units (see \cite{Lutjens:Comput1996} p.236), i.e.
they are all normalised using \texttt{R0EXP} and \texttt{B0EXP}. The content of the EXPEQ file is:\\
Line 1: $(R\ind{max}-R\ind{min})/(R\ind{max}+R\ind{min})$ of the LCFS\\
Line 2: $(Z\ind{max}+Z\ind{min})/(R\ind{max}+R\ind{min})$ of the LCFS\\
Line 3: pressure at the LCFS\\
Line 4: NP, number of points used in the discretization of the LCFS\\
Line 5 to 5+NP-1: first column R, second column Z of the LCFS points (do not forget that these quantities are normalised to \texttt{R0EXP})\\
Line 5+NP: N, number of points used in the discretization of the profiles\\
Line 6+NP: \texttt{NSTTP}, form used for the current profile specification. For $\texttt{NSTTP}=1$, you give $TT'$, see \cite{Lutjens:Comput1996} p.247 for other options. Note that the value of
\texttt{NSTTP} given in the CHEASE namelist is not used if you choose to read the EXPEQ file.\\
Line 6+NP+1 to 6+NP+N: radial grid in $\rho_\psi$\\
Line 7+NP+N to 6+NP+2N: $p'=\partial p/\partial \psi$ radial derivative of the pressure versus $\psi$, not $\rho_\psi$ (and in CHEASE units, of course)\\
Line 7+NP+2N to 6+NP+3N: $TT'$ if $\texttt{NSTTP}=1$. Here again the radial derivative is versus $\psi$ and the input is in CHEASE units.\\
The exact definition of the radial grid that has to be used is:
$$\rho_\psi=\sqrt{\frac{|\psi-\psi\ind{axis}|}{|\psi\ind{edge}-\psi\ind{axis}|}}$$
which means that 
$$p'=\frac{\partial p}{\partial \psi} = \frac{1}{2\rho_\psi|\psi\ind{edge}-\psi\ind{axis}|}\frac{\partial p}{\partial \rho_\psi}$$ 
Depending from where you get your value of $|\psi\ind{edge}-\psi\ind{axis}|$, you might need to divide it by $2\pi$ to be consistent with the CHEASE definition (which has the $2\pi$). Anyway, you will
easily spot the problem when comparing the input and output pressure profiles: they should be quasi-identical if the input value of $|\psi\ind{edge}-\psi\ind{axis}|$ is correct.\\
Did I say that all the quantities in the EXPEQ file must be in CHEASE units?

\section{EQDSK file}
The EQDSK file is used as an input in CHEASE if $\texttt{NSURF}=6$ and $\texttt{NEQDSK}=1$. All quantities in the EQDSK are in SI units (ouf...). The content of the EQDSK is relatively well described
in \cite{Lutjens:Comput1996} p.236. You can run CHEASE directly from the EQDSK obtained from a reconstruction code like EFIT. Just make sure that the EQDSK is of type 3 (first number after the date
in the first line) and that you have run BNDFIT to have a good descritption of the LCFS. Important (and a bit confusing), the EQDSK file has to be named \texttt{EXPEQ} in the working directory to be
found by CHEASE.

\section{Output file \texttt{hamada.dat}}
The CHEASE output file containing the metric elements for the $(\Psi,\zeta,s)$ coordinate system is named \texttt{hamada.dat} and is created in \texttt{hamada.f90}. All quantities are
given in MKSA units. The output file content is described below (see \texttt{gkwandchease.tex} for further information).\\

\textbf{Scalars}\\
NPSI: number of points for the radial grid\\
NS: number of points for the poloidal grid (specified with NCHI in the CHEASE namelist)\\
R0EXP: reference major radius used in CHEASE, should be taken to be close to the magnetic axis radius for faster convergence (but the actual position is not necessarily known beforehand).\\
B0EXP: reference magnetic field in CHEASE defined as the magnetic field at R0EXP\\
Raxis: magnetic axis radius as calculated by CHEASE
\vspace{0.5cm}

\textbf{Grids}\\
PSI: $\Psi$, the poloidal flux, zero at the magnetic axis, increasing towards the edge. The $\Psi$ grid given in output is equidistant in $\sqrt{\Psi}$ (except if mesh densification is used). The
first point is NOT on the magnetic axis (it is at $\sqrt{\Psi}=\sqrt{\Psi\ind{\max}}/\mathrm{NPSI}$, while on the magnetic axis $\sqrt{\Psi}=0$). The steps of the grid are $\sqrt{\Psi\ind{\max}}/\mathrm{NPSI}$ and the last point is on the last closed flux surface.\\
S: $s$, as defined in GKW. The grid is equidistant and for a given flux surface $s=0$ is on the low field side, at the height of the magnetic axis. 
\vspace{0.5cm}

\textbf{1D arrays}\\
Rgeom: major radius of the flux surface centre, defined as $(R\ind{max}+R\ind{min})/2$\\
amin: minor radius of the flux surface, defined as $(R\ind{max}-R\ind{min})/2$\\
damindpsi: $\partial \mathrm{amin}/\partial \Psi$\\
d2amindpsi2: $\partial^2 \mathrm{amin}/\partial \Psi^2$\\
bmax: maximum of the magnetic field strength on the flux surface\\
bmin: minimum of the magnetic field strength on the flux surface\\
q: safety factor\\
dqdpsi: $\partial q /\partial \Psi$\\
dqdpsi\_chk: value of $\partial \zeta / \partial \Psi$ at $s=1$ (equal to $\partial q /\partial \Psi$)\\
p: plasma pressure\\
dpdpsi: $\partial p /\partial \Psi$\\
jac: $J_{\Psi\zeta s}$ the jacobian of the $(\Psi,\zeta,s)$ coordinates\\
djacdpsi: $\partial J_{\Psi\zeta s} /\partial \Psi$\\
f: $F=RB\ind{t} = R^2|\mathbf{B}\cdot\nabla\phi|$ (always positive by convention)\\
dfdpsi: $\partial F /\partial \Psi$\\

\textbf{2D arrays}\\
g11: $\nabla \Psi \cdot \nabla \Psi$\\
g12: $\nabla \Psi \cdot \nabla \zeta$\\
g13: $\nabla \Psi \cdot \nabla s$\\
g22: $\nabla \zeta \cdot \nabla \zeta$\\
g23: $\nabla \zeta \cdot \nabla s$\\
g33: $\nabla s \cdot \nabla s$\\
B: the norm of the magnetic field\\
dBdpsi: $\partial B/\partial \Psi$\\
dBds: $\partial B/\partial s$\\
R: major radius of the $(\Psi,s)$ grid points (which describe flux surfaces) \\
Z: height of the $(\Psi,s)$ grid points (which describe flux surfaces) \\
dRdpsi: $\partial R/\partial \Psi$\\
dRds: $\partial R/\partial s$
dZdpsi: $\partial Z/\partial \Psi$\\
dZds: $\partial Z/\partial s$\\
theta: the poloidal angle, zero at $s=0$, increasing anticlockwise and measured from the magnetic axis\\
Ah: $\mathcal{A}$ \\
dAhdpsi: $\partial \mathcal{A}/\partial \Psi$\\ 
dzetadpsi: $\partial \zeta/\partial \Psi$\\ 
dzetadchi: $\partial \zeta/\partial \chi$\\

%*********************************************************************
\part*{List of reference cases}
%*********************************************************************
A couple of reference input files are included in the folder \texttt{./gkw/doc/chease/ref\_cases/}. They include the standard \texttt{chease\_namelist} file called \texttt{i.case\_name} and the corresponding EXPEQ or EQDSK. Packing is used such that the radial mesh is densified around $\rho_\psi=0.2$, $0.4$, $0.6$ and $0.8$. Furthermore, the value of \texttt{NCHI} is 126, which means that the number of points per poloidal turn in GKW can be 7, 9, 21 or 63 ($k=9$, 7, 3 or 1). You can of course change the value of NCHI in the input namelist at will to run GKW with a different number of points with poloidal turn. The input last closed flux surface and profiles are already smooth, so no tension is needed for the fits (\texttt{TENSBND=0.} and \texttt{TENSPROF=0.} in the input namelist).

\section{Circular case: \texttt{circ}}
Circular LCFS, $\texttt{R0EXP}=1$ and $\texttt{B0EXP}=1$. No rescaling of the current profile.\\
In that example, the input file is an EQDSK. Just to show it works...

\section{MAST DND: \texttt{mast.dnd1}}
MAST double-null plasma, based on \#18696 at $t=0.25\,\textrm{s}$, $\texttt{R0EXP}=1.002$ and $\texttt{B0EXP}=0.436$, no rescaling of the current profile.

\section{TCV SND: \texttt{tcv.snd1}}
TCV single-null plasma, based on \#31837 at $t=1.0\,\textrm{s}$, $\texttt{R0EXP}=0.891$ and $\texttt{B0EXP}=1.44$, no rescaling of the current profile.

\section{Waltz and Cyclone}
These two cases do not need introduction. The CHEASE input file have been tweaked so that the $q$ and shear value matches the one of the Waltz or Cyclone cases at the required radial position. The
flux surfaces are concentric. The \texttt{hamada.dat} are available for several values of $N_s$ (parallel grid). You can generate your own by changing the parameter $\texttt{NCHI}$ in the CHEASE input namelist.

\bibliographystyle{iopart-num}
\bibliography{gkwandchease}

\end{document}

### 2022-05-03 gkw/doc/etc/Ballooning.tex
(1) Added gkw Clone gwk  #3

\documentclass[a4paper,11pt]{article}
\usepackage{fullpage}
\setlength{\parskip}{8pt}
\setlength{\parindent}{0pt}

%opening
\title{Ballooning representation of the Vlasov equation}
\author{F. J. Casson}
\date{May 2008}

\begin{document}

\maketitle

\begin{abstract}

The solutions to magnetic differential equations have rapid spatial variation perpendicular to the magnetic field and slow variation along magnetic field lines. The eigensolutions may be represented as a rapidly varying eikonal in the perpendicular plane modified by a slowly varying envelope function along the field lines. In a toroidal geometry, periodicity constraints on the eigenmodes lead to the ballooning representation.  Using the gyrokinetic Vlasov equation as an example, we here introduce the concepts of the ballooning representation with certain simplifying assumptions.  This discussion is illustrative with many technical details omitted.

\end{abstract}

\section{Motivation with respect to GKW}
The ballooning representation is not exactly the method used to solve the gyrokinetic Vlasov equation in GKW, but it is closely aligned to it.  By studying the ballooning representation we hope to reveal something about the structure of the solutions that GKW finds.  In the process this additional perspective should also illuminate a new way to think about periodicity and mode connections as implemented in GKW. This understanding of the periodicity constraints is vital to make appropriate choices for the mode namelist and in interpretation of results.  We shall return to these points later.

\section{Eikonal representation}

A magnetic differential equation (\cite{H+M} \S 3.3)  is of the inhomogeneous form
\begin{equation}
 {\bf B}\cdot \nabla f = U({\bf x})
\end{equation}

The eikonal representation here outlined is applicable to the solution of many equations of this form, but here we restrict ourselves to the example of the gyrokinetic Vlasov equation for a perturbation distribution function $f({\bf X},\mu,v_\parallel, t)$:
\begin{equation}
{\partial f \over \partial t} + v_\parallel {\bf b}\cdot \nabla f + {\bf v}_{D} \cdot \nabla f -{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial f \over \partial v_\parallel} = S
%-{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial f \over \partial v_\parallel} = 0. 
\label{eq.gyro}
\end{equation}
in which S is the source term determined by the background distribution function, and ${\bf v}_{D}$ contains all curvature and ${\bf E} \times {\bf B}$ drifts.

% \begin{equation}
% {\partial f \over \partial t} + v_\parallel {\bf b}\cdot \nabla f + {\bf v}_{D} \cdot \nabla f  = 0
% %-{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial f \over \partial v_\parallel} = 0. 
% \end{equation}
%in which the mirror and source terms have been dropped.

The perpendicular and parallel dynamics are strongly separated since $v_\parallel \approx v_{thermal}$ and $v_D \approx \rho_* v_{thermal}$ and $\rho_* = {\rho \over R} \ll 1$ for most toroidal devices.  We therefore look to construct an \textit{eikonal solution} that determines the rapidly varying perpendicular mode structure.  We can also view the eikonal representation as a perturbation series solution approach (c.f. Quantum theory) with ordering parameter $\rho_*$.
%Furthermore we see that the mirror term is also of $O(\rho_*)$  since ??.
To lowest order then, we take $\rho_* = 0$ and keeping only spatial dependence our eikonal $f^{(0)}$ must satisfy:
\begin{equation}
{\partial f^{(0)} \over \partial t} + v_\parallel {\bf b}\cdot \nabla f^{(0)} = 0
\end{equation}
We would also like our base solution $f^{(0)}$ to be time independent hence we have:
\begin{equation}
{\bf b}\cdot \nabla f^{(0)}  = 0
\label{eq.mag}
\end{equation}
Expressing the magnetic field unit vector ${\bf b}$ in terms of its toroidal and polodial components we have
\begin{equation}
 {\bf b}= {{\bf B} \over B} = {1 \over B} (B_t {\bf {\hat e}}_t + B_p {\bf {\hat e}}_p)
\end{equation}
where the ${\bf {\hat e}}$ represent \textit{normalised} unit vectors.

%Perhaps not quite so illuminating?
At this point it is necessary to be more specific about the coordinates of our presentation.  A general toroidal coordinate system with $(\rho,\theta,\phi)$ is not in general orthogonal.  Assuming $\rho$ as a flux surface label coordinate, we prefer to express the above in the covariant basis ${\bf e}_i={\partial {\bf x} \over \partial x^i}$ since ${\bf e}_\theta, {\bf e}_\phi$ lie in the flux surface (whereas $\nabla \theta$ in the contravariant basis may not). In this basis:
\begin{equation}
{\bf {\hat e}}_p = {{\bf e}_\theta \over |{\bf e}_\theta|},\quad {\bf {\hat e}}_t = {{\bf e}_\phi \over |{\bf e}_\phi|},\quad B_p = {B^\theta \over |\nabla \theta|},\quad B_t = {B^\phi \over |\nabla \phi|}
\end{equation}
So we may write:
 \begin{equation}
  {\bf B} = (B^\phi {\bf {e}}_\phi + B^\theta {\bf {e}}_\theta)
 \end{equation}
Recalling that 
\begin{equation}
 {\bf a} \cdot {\bf b} = a^i b_i {\bf e}_i \cdot \nabla x_i, \quad
\nabla f = \nabla x^i {\partial f \over \partial x^i}
\label{eq.bases}
\end{equation}
we have from Eq. \ref{eq.mag}:
\begin{equation}
B^\phi\left({\partial f^{(0)} \over \partial \phi} + {B^\theta \over B^\phi}{\partial f^{(0)} \over \partial \theta}\right) = 0
\end{equation}
%which we have factorized to make it clear that the ratio of the terms depends on ${B^\phi \over B^\theta}$.
For simplicity in the solution of this equation we shall use the low aspect ratio approximation $\epsilon = r/R \ll 1$ which allows us to assume constant pitch over the poloidal angle with $q = {B^\phi \over B^\theta}$.  (Dropping this assumption would lead us to an integral over the local pitch in the eikonal.)
%This definition of the local pitch is wrong....
Hence:
\begin{equation}
 {\partial f^{(0)} \over \partial \phi} + {1 \over q}{\partial f^{(0)} \over \partial \theta} = 0
\end{equation}
To satisfy toroidal periodicity we look for solutions of the form $f_n^{(0)} = l(\theta)\exp [in\phi]$ characterised by toroidal wavenumber $n$. Solving leads us to the eikonal:
\begin{equation}
 f_n^{(0)}=exp[in(\phi-q(\theta+\theta_0))]
\end{equation}
What we have done is to construct, to lowest order in $\rho$, time independent solutions to Eq. \ref{eq.gyro} with rapid spatial variation perpendicular to the field (we shall assume $n \gg 1$) and with no phase variation along the field line.  The wavevectors are perpendicular to the field lines and parallel derivatives of the eikonal are zero by construction.

\vspace*{5 pc}

\section{Envelope function}
To approach the first order solution in $\rho_*$ to Eq. \ref{eq.gyro} we assume the form of a slowly varying envelope function $g$ around our eikonal solution $f_n^{(0)}$:
\begin{equation}
 f_n^{(1)}=g_n(\theta,v_\parallel,\mu,t,)f_n^{(0)}(\rho,\theta,\phi)
\end{equation}
Note this is NOT the same meaning for $g$ as in the GKW manual. Substitution into Eq. \ref{eq.gyro} gives
\begin{equation}
g_n \left(\underbrace{{\partial f_n^{(0)} \over \partial t} + v_\parallel {\bf b}\cdot \nabla f_n^{(0)}}_A\right) + \underbrace{g{\bf v}_D \cdot \nabla f_n^{(0)}}_I + \left({\partial g_n \over \partial t} + v_\parallel{\bf b}\cdot\nabla g_n + \underbrace{{\bf v}_D \cdot \nabla g_n}_{O(\rho_*^2)} -{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial g_n \over \partial v_\parallel}\right)f_n^{(0)} = S
\label{eq.substitute}
\end{equation}
The term labelled A is zero by our construction of the eikonal.  The term labelled $I$ is the only one of first order in $\rho_*$ for the drifts ${\bf v}_D)$ and we neglect the last of $O(\rho_*^2)$.  In a linear treatment of the drift term ${\bf v}_D$ we neglect the ${\bf E} \times {\bf B}$ drifts and using the approximation ${\bf e}_\perp \approx {\bf e}_\theta$ we can say that all other curvature drifts are in the vertical direction.  This approximation is the same as:
\begin{equation}
{B_p \over B_t} = {\epsilon \over q} \ll 1 \quad \Rightarrow q \approx 1
\end{equation}
Resolving this vertical drift into our toroidal coordinates ($|{\bf e}_\theta|=r$) we have:
\begin{equation}
 {\bf v}_D = |{\bf v}_D|\left(\cos(\theta) {{\bf e}_\theta \over r} + \sin(\theta) {{\bf e}_\rho \over |{\bf e}_\rho|}\right)
\end{equation}
and again using Eq. \ref{eq.bases} we have:
\begin{equation}
 I = g_n|{\bf v}_D|\left[\cos(\theta){1 \over r}(-inq)+{1 \over |{\bf e}_\rho|}\sin(\theta)(-{\partial q \over \partial \rho} in(\theta+\theta_0))\right]f_n^{(0)} 
\end{equation}
We now define a wavevector $k_{\theta,n}$ by analogy with a Fourier expansion such that: 
\begin{equation}
 {\bf e}_\theta \cdot \nabla f_n^{(1)} = ik_\theta f_n^{(1)} \Rightarrow k_{\theta,n} = - {nq \over r}
\end{equation}
Which allows us to write
\begin{equation}
 I = g_n|{\bf v}_D|k_{\theta,n} \left[\cos(\theta)+{\hat s}(\theta+\theta_0) \sin(\theta)\right]f_n^{(0)} 
\end{equation}
where ${\hat s}={r \over q |{\bf e}_\rho|}{\partial q \over \partial \rho}$ is the magnetic shear.  Hence we see that a consequence of toroidal geometry is that the radial wavevector is coupled to the perpendicular by
\begin{equation}
 k_{\rho,n}={\hat s}\theta k_{\theta,n}+k_{\rho_0,n}
\label{eq.connect}
\end{equation}
where $k_{\rho_0,n}={nq \over r}{\hat s}\theta_0$ can be chosen arbitrarily.  Our envelope function can now be written in terms of the wavevectors and From Eq. \ref{eq.substitute} this envelope $g_n(k_\theta,k_\rho,\mu,v_\parallel,t)$ must satisfy:
\begin{equation}
{\partial g_n \over \partial t} + v_\parallel{\bf b}\cdot\nabla g_n + g_n|{\bf v}_D|k_\theta \left[\cos(\theta)+{\hat s}\theta \sin(\theta)\right]-{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial g \over \partial v_\parallel}=S
\end{equation}

Hence each discrete mode in $k_\theta$ must have an associated continuous spectrum in $k_\rho$. We see that the on the periodic domain $-\pi < \theta < \pi$ that if we have multiple discrete $k_{\rho}$ modes they must connect after one poloidal turn according to Eq. \ref{eq.connect}. An alternative way to interpret this is that in the infinite domian $-\infty < \theta < \infty$ there is only one $k_{\rho,n_p}$ mode for each $k_\theta$.

\section{Periodicity constraints}
Earlier we used the condition of toroidal periodicity to construct an eikonal with toroidal mode number $n$. Until this point we have ignored the additional constraint of poloidal periodicity. On any rational flux surface there will be an infinite set of mode numbers $n_q$ such that $n_q q$ is also an integer.  For each eikonal $f_{n_q}^{(0)}$, poloidal periodicity will be satisfied on this rational flux surface.

%I think this bit from page 329 of H + M only makes sense using the \eta coordinate
% Furthermore \textbf{it can be shown} (check this argument before typing it up...) that the envelope functions for these eikonals are all related by
% \begin{equation}
%  g_{n_p}(\theta) = g_{n_0}(\theta+2\pi p)
% \label{eq.envelope}
% \end{equation}
% % in which $g_{n_0}$ is one arbitrarily chosen periodic solution which determines all others.  Whilst we neglected poloidal periodicity, each $f_n^{(1)}$ was one member of a linearly independent basis set for the first order solutions to Eq. \ref{eq.gyro}.  
% Applying the constraint of poloidal peridiocity to the eikonal we find that the $k_{\theta,n}$ modes are now coupled and any solution must be a sum of all of them:

Note, however, that since the radial wavevector is increasing with $\theta$ (Eq. \ref{eq.connect}, a periodic envelope function cannot be physical as the energy in the mode would diverge).  We construct a periodic envelope function
\begin{equation}
  {\hat g}_n(\theta)=\sum_{p=-\infty}^\infty g_{n}(\theta + 2\pi p)
\end{equation}
where we also require that the infinite sum converges.  This last restriction results in a $g_{n}$ that must decay away as $\theta \rightarrow \pm \infty$.  Our new periodic envelope function ${\hat g}_n(\theta)$ then allows us to find periodic solutions $f_n^{(1)}={\hat g}_n f_n^{(0)}$:
\begin{equation}
 f_n^{(1)}(\theta)=\sum_{p=-\infty}^\infty g_{n}(\theta+2\pi p)\exp(-in(\phi-q(\theta+\theta_0+2\pi p))) 
\end{equation}
This is the idea that is the key to the ballooning representation.

%I Think this is also incorrect after conversation with Fulvio Zonca
% Taking the rational surfaces to be closely packed and $n_p \gg 1$, a solution on an irrational surface can used as an accurate approximation for a nearby irrational surface.

\vspace*{13 pc}

%$f^{(1)}=\sum_n c_n f^{(1)}$ we can only admit solutions with $f^{(1)}(\theta)=f^{(1)}(\theta+2\pi)$
%some detail missing here: To prove need to use \theta_0
%If one of the solution branches with $n=n_0$ has a periodic poloidal dependence then the eikonal tells us
% \begin{equation}
% \sin(n_0 q\theta) = \sin(n_0 q(\theta+2\pi j )) = \sin(n_0q\theta + 2\pi p) 
% \end{equation}
%for all integers $j$ where $p = n_0qj$ is also an integer.  The envelope functions must then satisfy %$g_p(\theta) = g_{n_0}(\theta +2\pi p)$
%But this defines another solution branch with$n = n_0 x$

\section{The ballooning transform and its relation to GKW}
\begin{small}

% \addtolength{\textwidth}{1in}
% % \addtolength{\hoffset}{-0.5in}
% \addtolength{\textheight}{2in}
% \addtolength{\voffset}{-2in}

Here we have conducted our analysis in generalized toroidal coordinates with the assumptions $\epsilon = r/R \ll 1$ and ${\bf e}_\perp \approx {\bf e}_\theta$.  The ballooning representation can also be constructed in field aligned coordinates (see \cite{H+M} \S 7.10), rendering these simplifying assumptions unnecessary, since in the field aligned coordinates the local pitch is constant on a surface and $q(\rho)$ is flux label.

We have already seen that our wavevectors were defined analogously to those under a Fourier transform.  Extending the analogy, the Ballooning representation can be viewed as a generalized \textit{transform} with the Fourier transform as one substep \cite{H+M} \cite{CHT}.  It has been shown \cite{HazeltineNewcomb} that this `Ballooning transformation' has an inverse.

In GKW we use a spectral method in the perpendicular direction and disregard terms of $O(\rho_*^2)$.  In the field aligned coordinates, this becomes equivalent to the ballooning transform with the coupling of the modes after one poloidal rotation.  This is achieved with the routine \texttt{connect\_parallel} playing the role of Eq. \ref{eq.connect}. Note, the mode connection will only take place if $mode\_box=true$ and $nperiod=1$. In a linear run with $nperiod = a > 1$ and a single $k_\rho$ mode ($NX = 1$) then we should see a decaying parallel envelope. With identical parameters but $nperiod=1$ and with $NX =2a - 1$ modes in $k_\rho$ the use of the parallel connection should give an identical result.

Of course we only have a discrete number of radial modes for each $k_\theta$ mode. Since the spacing of the radial modes is up to us, we choose them such that we can connect discrete modes after one poloidal turn according to Eq \ref{eq.connect}. This is achieved by making the spacing of the discrete radial modes
\begin{equation}
k_{\rho,n,i+1}-k_{\rho,n,i}={k_{\theta,n}{\hat s}\over j}
\end{equation}
for some integer $j$ (called \texttt{ikxspace} in the code) related to $k_{\rho_0}$.  For more details see the section `local limit' of the GKW manual.

The effect of including the nonlinear $v_{{\bf E} \times {\bf B}} \cdot \nabla f$ term is to couple the $k_\theta$ modes in which case we must keep more than one to see this interaction.  A nonlinear run can also be viewed as the single $k_\rho$ mode nonlinearly interacting with itself in the infinite $\theta$ domain.  In practice of course we cannot extend to the infinite $\theta$ domain.

Finally, our treatment here has been a local one, with the assumption that our envelope does not have a radial dependence.  This is equivalent to the use of the periodic radial boundary in GKW which makes it a local `flux tube' code.  Ballooning theory can be extended to allow for a radial variation in the form of a secondary, radial, envelope function but this is well beyond the scope of this work. See \cite{Dewar83}.

\end{small}

%Nperiod = 1 - infinite
%S Boundaries set to zero.
%Should see a decaying envelope in par.dat%

\setlength{\parskip}{4pt}
\begin{tiny}\section{Further Reading}

For a readable presentation of the Ballooning representation and Ballooning transform in field aligned coordinates, see \cite{H+M} \S 7.10.  For the original paper by Connor describing the Ballooning transformation see \cite{CHT}.  For the proof of the existence of an inverse transformation and identification of an error in \cite{CHT} and other previous papers see \cite{HazeltineNewcomb}.  For a more complete discussion of the Ballooning representation and its relation to WKB theory see \cite{Dewar83}.  For a comparison of the representation and the transformation methods see \cite{Thy}.

\begin{thebibliography}{5}

\bibitem{H+M}
R.~D. Hazeltine and J.~D. Meiss.
\newblock {\em Plasma Confinement}.
\newblock Dover, 2003.

\bibitem{CHT}
J.~W. Connor, R~.~J. Hastie, and J.~B. Taylor.
\newblock Shear, periodicity, and plasma balooning modes.
\newblock {\em Physcial Review Letters}, {\bf 40}, 396, 1978.

\bibitem{HazeltineNewcomb}
R.~D. Hazeltine and W.~A. Newcomb.
\newblock Inversion of the ballooning transformation.
\newblock {\em Physics of Fluids B: Plasma Physics}, {\bf 2}, 7, 1990.

\bibitem{Dewar83}
R.~L. Dewar and A.~H. Glasser.
\newblock Ballooning mode spectrum in general toroidal systems.
\newblock {\em Physics of Fluids}, {\bf 26}, 3038, 1983.

\bibitem{Thy}
A.~Thyagaraja.
\newblock A spectral approach to ballooning theory. part 1.
\newblock {\em J. Plasma Physics}, {\bf 59}, 367, 1998.

\end{thebibliography}

\end{tiny}
\end{document}

### 2022-05-03 gkw/doc/etc/GKW_for_beginners.tex
(1) Added gkw Clone gwk  #3

\documentclass[a4paper,10pt]{article}
\usepackage{amssymb}
\usepackage{graphicx}
\graphicspath{{../fig/}}
\usepackage{fullpage}
\usepackage{hyperref}

\hypersetup{
bookmarks,
plainpages=false,
colorlinks=true,
pdfborder={0 0 0 0},
linkcolor=blue,
citecolor=blue,
urlcolor=blue,
pdfstartview={FitH},
bookmarksopen=false,
unicode
}

\newcommand{\src}[1]{\href{http://bitbucket.org/gkw/gkw/src/HEAD/src/#1}{#1}}
\newcommand{\doc}[1]{\href{http://bitbucket.org/gkw/gkw/src/HEAD/doc/#1}{#1}}
\newcommand{\raw}[1]{\href{http://bitbucket.org/gkw/gkw/raw/HEAD/#1}{#1}}
\newcommand{\scripts}[1]{\href{http://bitbucket.org/gkw/gkw/src/HEAD/scripts/#1}{#1}}
\newcommand{\matlab}[1]{\href{http://bitbucket.org/gkw/gkw/src/HEAD/matlab/#1}{#1}}
\newcommand{\trunk}[1]{\href{http://bitbucket.org/gkw/gkw/src/HEAD/#1}{#1}}

\setlength{\textheight}{24cm}

\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}} % for partial derivatives

%opening
\title{GKW for Beginners}
\author{F.J. Casson}
\date{11 April 2008.  Last Revised 18 Jan 2016.}
\setlength{\parindent}{0pt}

\begin{document}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}

  This introductory document aims to describe the very basic things a
  newcomer needs to know in trying to understand or use GKW. These are
  the things that are mostly obvious to anyone who has worked with the
  code on any serious level.  To anyone who has worked with another
  gyrokinetic code many of these things will also be obvious, but
  there are practicalities specific to GKW.  Therefore this document
  has to be written and improved by newcomers still in the early
  stages of learning GKW.  This is still a work in progress, as such
  you are encouraged to add to and improve it.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\tableofcontents
\setlength{\parskip}{8pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What does the code do?}

The code GKW solves the gyrokinetic Vlasov equation for the evolution
of plasma particle distribution functions self consistently with the
Maxwell equations in a toroidally symmetric geometry. More precisely,
it solves it only for the perturbation $f$ of an equilibrium
distribution $F_M$ (also called the background), similar to an Ansatz
in linear perturbation theory.
\begin{equation}
  f_{tot} = F_M + f
\end{equation}

One important idea of the gyrokinetic model is that one velocity
dimension of the general Vlasov equation is removed by averaging over
the small scale cyclotron (gyro) motion of a charged particle in a
magnetic field. Thus the equations are 5 dimensional (3 space, 2
velocity) with the two directions of velocity $\mu$ (the magnetic
moment, being connected to $v_\perp$) and $v_\parallel$.  This
simplification by one dimension makes a numerical simulation of
microscale plasma turbulence tractable.

For the purpose of simulation a tokamak core plasma, the field aligned Hamada coordinate system is
introduced, which separates the directions parallel and perpendicular to the
magnetic field. This allows to apply the following assumptions, called the
'gyrokinetic ordering': One assumes certain length
scales for parallel and perpendicular derivatives of distribution and
field perturbations.
\begin{equation}
\nabla_\parallel  \sim \frac{1}{R} \quad \mathrm{and}\quad
\nabla_\perp \sim \frac{1}{\rho_{ref}} =  \frac{1}{R}\frac{R}{\rho_{ref}} = \frac{1}{R\rho_*}
\end{equation}
where $R$ is a typical length scale of the background (like the
tokamak major radius) and $\rho_{ref}$ is a typical gyroradius. The
ratio $\rho_*=\rho_i/R$ is a small number, serves as a smallness
parameter in this model and appears also in an assumption called the
$\delta f$-approximation:
\begin{equation}
  f \sim \rho_*F_M\quad\mathrm{and}\quad \pd{f}{v_\parallel}\sim\rho_*
\end{equation}

When constructing a gyrokinetic equation including terms of at most (say) first
order in $\rho_*$, these assumption about the structures of interest
allow to neglect certain terms, e.g. such containing parallel derivatives
of perturbed quantities.

Hence, these orderings mean that the model will resolve and include structures
perpendicular to the magnetic field on the scale of the Larmor radius
$\rho_i$, whilst parallel structures will have a length scale of the
order of the system (tokamak) size $R$. This ordering is chosen to
describe drift wave structures.

The GKW code has a number of useful features, such as the ability
to model many plasma species, including non linear terms, kinetic
electrons, particle collisions, a rotating plasma and arbitrary flux
surface geometry (through interface with the CHEASE code),
tearingmodes and others.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Where is the code?}

In the git repository at bitbucket.  Non-members may clone a read-only working copy anonymously over HTTP.

\texttt{git clone git@bitbucket.org:gkw/gkw.git}.

Some additional git workflow guidance for GKW is on the \href{https://bitbucket.org/gkw/gkw/wiki/Git_Bitbucket}{wiki}.
If you have not used git before then briefly read: Chapters 1 and 2 of the git book (\href{http://git-scm.com/book/en/v2}{http://git-scm.com/book/en/v2}).  Bitbucket has a graphical user interface which makes it easy to use many of the features of git.

When you check out the code the following folders will be created:
\begin{itemize}
 \item \texttt{/src} contains all the source code for GKW.
 \item \texttt{/doc} contains documentation files. \texttt{/doc/input} contains some sample input files.
 \item \texttt{/config} contains build configuration makefiles for some hosts
 \item \texttt{/tests} contains a set of test cases we use for testing changes made to the code.
 \item \texttt{/matlab} contains matlab scripts for preparing input files and interpreting output data.
 \item \texttt{/octave} contains GNU Octave scripts for preparing input files and interpreting output data.
 \item \texttt{/python} contains Python scripts for handling output data.
 \item \texttt{/scripts} contains various scripts which (amongst other
   things) can compile GKW from any working directory
   (\texttt{gkwmake}), can check if the code is working as it should
   (\texttt{gkw\_run\_tests}), delete all output files of a GKW run
   (\texttt{gkw\_clean\_run}), start a simulation locally with the
   newest executable (\texttt{gkwrun}) or submit multiple jobs
   (\texttt{gkwnlin}). See their help messages or instruction at the
   beginnig of these scripts.
 \item \texttt{/libs} contains additional libraries that are needed
   and also provided with gkw for easier compilation.
\end{itemize}

It is required by various scripts that you define a \texttt{GKW\_HOME}
environment variable which is the path to your GKW folder, as this
variable is used in some scripts. Adding the following to your
\texttt{.bashrc} file will make these things automatic:

\begin{verbatim}
   export GKW_HOME="/path/to/gkw/"
   export PATH="$PATH:${GKW_HOME}/scripts"
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{How do I compile from source?}

The code should compile with default options on your machine by running \texttt{gkwmake -j} (or \texttt{gmake} in the top level GKW directory). This will create:
\begin{itemize}
\item \texttt{/obj} folder for all the object files (these can be deleted later)
\item \texttt{/run} folder which will contain the compiled executable. 
\end{itemize}

On many shared machines, GKW makefiles are already configured.  
When you build for the first time on a new machine, a template configuration file will be created for your machine in \textit{./config/machine}.  
The basic configuration will not contain MPI or FFTW (needed for parallel or nonlinear runs).
You can configure FFTW and MPI and other compile options specific to your machine by editing the file \textit{default.mk} in this folder. 
\texttt{gkwmake clean} will remove all the object files if you want to do a completely fresh compile.

While developing new code it is strongly recommended to compile with
\begin{verbatim}
   gkwmake -j DEBUG=on OPTFLAGS=off
\end{verbatim}
whereas executables for production runs will become optimised for speed with
\begin{verbatim}
   gkwmake -j DEBUG=off OPTFLAGS=on
\end{verbatim}

\section{How do I do a run?}

You need to create a file named \texttt{input.dat} in the same
directory as the executable.  This file is the input deck containing
all the switches which control how the code runs.  There are some
sample input files in \texttt{/doc/input}. Also in the \texttt{/doc}
folder is an \doc{input.dat.sample} file which should briefly describe
all the input variables. Many of the input switches are optional and
default values will be used if they are not provided.  The
\doc{input.dat.minimum} file shows the variables that \textit{must} be
provided.

The code is quite sensitive to bad formatting or typos in
\texttt{input.dat}, in which case you will hit a read error as soon as
you run the code.  Most of the time if you input a set of incompatible
switches the code will tell you and throw you out.

Two key output files created are \texttt{time.dat} which records the
(linear) mode growth rate against time, and \texttt{fluxes.dat} which
records particle fluxes against time.  More details on the other
output files are in the main manual.

A few salient points about the input file (many more in the full
manual):

\subsection{Output File Format}
The setting of \texttt{io\_format} controls if diagnostic data is
output as HDF5 file or as many ASCII or binary files. The setting
\texttt{io\_legacy} switches between historically grown data ordering
and a more cleaned-up ordering (e.g. data in form of a long flat
column vs. a n-dimensional matrix)\\
In the long run, it is recommended to use
\texttt{io\_format=hdf5+ascii} and \texttt{io\_legacy=F}.

\subsection{Species}
There is no default species data, hence you must provide some. The species you input must satisfy quasi-neutrality. This requires you to set ratios of densities and charge appropriately such that the sum over all species $s$ is zero:
\begin{equation}
 \sum_s z_s \cdot dens_s = 0
\end{equation}
\begin{equation}
 \sum_s z_s \cdot dens_s \cdot rln_s= 0
\end{equation}
You can input multiple ion species but only one negatively charged species (electrons) is allowed.  Note that the number of species read in is controlled by the number of species parameter in the 'gridsize' namelist and hence it is possible that later species data you specify will not be read in if this is set incorrectly.  The \texttt{number\_of\_species} input variable does not include adiabatic electrons but the species data for these will still be read if this the case.

\subsection{Timestep and resolution}
The timestep required for stability can vary greatly depending on input parameters or choice of scheme.  In particular kinetic electrons require a timestep $10^{-2}$ smaller.  The question of necessary resolution needed is even harder to answer, in fact it is one of the controversial questions in gyrokinetics.  To a large extent it depends on which physics problem you want to study.  Use the sample provided \doc{input} files to get you started.

\subsection{Parallelisation}
To run parallel (MPI) jobs you need to be in an environment which
provides the an MPI library and with an mpi controller configured and
running. It may help to execute \texttt{type mpiexec} or \texttt{type
  mpirun}, or \texttt{gkw\_run\_tests --parallel} to see if this is the
case.

If the parallelisation scheme is not explicitely given in the input
file via the \texttt{n\_procs\_*} parameters the code will select what
it considers to be the most appropriate parallelisation given your
computational grid.  The code will parallelise first
over the $\mu$ grid and (non-adiabatic) species, and then over the
$v_{\parallel}$ and $s$ grid points.  Nevertheless it is sensible to
have in mind a parallelisation setup when you set the gridsize, as the
user controls the number of processors the code will be run on and
should choose a number that fits with the gridsize.

\subsection{Modes}
The spacing and location of the modes is determined by the `modes' namelist.   For linear runs, set \texttt{mode\_box=.false.}, \texttt{NX=1} and \texttt{NMOD=1}.  For nonlinear runs, there is a spreadsheet \doc{mode\_box\_calculator.ods} in the documentation folder which should help you determine the mode spacing.

\subsection{Solver Methods}
Safest option:
\begin{verbatim}
METHOD='EXP'
METH=2
\end{verbatim}

\subsection{Batch jobs}
There are also scripts for doing multiple runs. The script \scripts{gkwnlin}  uses a directory structure with batch input files all put in an input folder (with different names), and output files will all be put in relevant folders.  The scripts submit jobs to a queue in the portable batch system (PBS).  If you haven't used PBS before, read a little about how it works.

Typical workflow for a batch job is the following:
\begin{itemize}
\item Compile the GKW executable for the machine you are on. 
\item Run \texttt{gkwnlin -p project\_name} to create a new project directory.
\item Run matlab script \matlab{create\_gkwscan} to make a scan of the given input file and put these in the input folder.  Or if you only need to make a few input files, you can do this manually.
\item Run \texttt{gkwnlin} from the input folder to submit multiple input files to the PBS queue.  Use something like \texttt{gkwnlin -np 8 -h 3 *}.  gkwnlin will take the most recent executable from \texttt{\$GKW\_HOME/run} to run with all the input files you give. 
\item Run \texttt{qstat} to the status of your jobs in the queue and see if they are running yet.
\item After the first loop has completed you should look in the output files ./runs/*/out, to check the job is running ok.
To stop an individual job use \texttt{qdel}, or do \texttt{touch gkw.stop} in its run folder for a clean stop.
\item Go get coffee.

\end{itemize}

These scripts are self documenting on the -options syntax they require.  Type \texttt{gkwnlin} for help, or in matlab \texttt{help create\_gkwscan}.  Note that \texttt{create\_gkwscan} works best if you keep your batch projects runs in \texttt{~/runs/gkw/}; or else reconfigure the matlab script \texttt{gkwpath}.

\section{Geometry and Domain}

\begin{figure}[hb!]
 \begin{center}
 \includegraphics[scale=0.3]{flux_tube.eps}
\caption{Sketch of the flux tube domain with $q=2$, $\epsilon=r/R=0.29$ and $\hat{s}=1$ showing the GKW coordinate directions.} 
 \label{fig.domain}
\end{center}
\end{figure}

\begin{figure}[ht!]
 \begin{center}
 \includegraphics[scale=0.3]{shear.eps}
\caption{Superposition of two perpendicular cross sections of the flux tube separated by one polodial circuit.  The flux tube is sheared as it follows the magnetic shear.  This picture hopes to make clear that each $k_\zeta$ mode must connect to a higher one for a periodic boundary in $s$.  Namely, the difference between connected modes is $\Delta k_\psi = k_\zeta {\partial q \over \partial \psi}$. }
 \label{fig.shear}
\end{center}
\end{figure}

The coordinates used are field aligned Hamada coordinates $(s, \psi, \zeta)$,  in which magnetic field lines are straightened. In these coordinates, the components of the magnetic field are flux functions and $B^\psi = B^\zeta = 0$ so that ${\bf B} \cdot \nabla = B^s {\partial \over \partial s}$.  Whilst $s(\theta,\phi)$ is a coordinate along the field line, $\psi(r)$ is a radial coordinate which describes the flux surfaces and $\zeta(\theta,\phi)$ is a (mostly poloidal) coordinate which lies within a flux surface.  The purpose of these coordinates is to allow separation of the different length scales of the parallel and perpendicular plasma dynamics.

GKW is written with general tensors so that in principle the code can deal with any toroidally symmetric 
equilibrium. This means that all the equations are expressed with the use of general tensors, and 
the metric elements. Currently  the geometry dependent quantities are calculated in the code only for the $\hat s - \alpha$ equilibrium. In addition, GKW can read CHEASE output for arbitrary flux surface configuration. CHEASE takes an input of the shape of the last closed flux surface, a boundary pressure, and a current profile. From these inputs, CHEASE solves the Grad-Shafranov equation to find an equilibria flux surface configurations which it outputs as the poloidal flux. See the documentation file \texttt{gkwandchease.tex} for more information.

GKW is a local gyrokinetic code modelling a flux tube domain, and is spectral (with periodic boundaries) in both directions perpendicular to a field line.  The flux tube domain is bounded by four magnetic field lines, two within each of two adjacent flux surfaces (Fig \ref{fig.domain}). As the domain progresses helically round the torus it is sheared with the magnetic shear.  When one poloidal circuit is completed the flux tube can be connected to itself, but the shear requires the $k_\psi$ modes to be be connected (Fig. \ref{fig.shear}) at the $s$ boundary.

In GKW, computation in the perpendicular directions is done in the wavevector domain for computational efficiency, whilst a finite difference scheme is used in the parallel directions.  In the general formulation, this pseudo-spectral method is more general than assuming an eikonal form for the solution. In our ordering of terms we keep only first order terms in $\rho_*=\rho/R \ll 1$, which makes the two approaches equivalent.  The transformation to field aligned coordinates is mathematically equivalent to the ballooning transform of the eikonal form in this case. The periodicity constraints equate to the coupling of the $k_\psi$ modes described above.  Since the flux tube is sheared, the consequences of magnetic shear is to couple the modes.  For further details see the section \textit{local limit} in \textit{GKW How and Why}.

\section{What is under the bonnet?}

The code is written in Fortran 95 and uses MPI for (optional) parallel implementation.  The code is split into separate Fortran modules in the \texttt{src} folder.  The program runs from \texttt{gkw.f90} and then uses the modules as called from there, each of which can recursively call other modules. 
The automatically generated reference dictionary for the code is at \href{http://www.gkw.org.uk/doxygen}{http://www.gkw.org.uk/doxygen} and is a useful way to explore the code.   The file \texttt{program\_flow} in the documentation folder shows the code progression (though it is a little out of date).
If you want to familiarise yourself with inner details of the code, we suggest that after reading the manual, you look at the source files 
\src{gkw.f90}, \src{init.f90}, \src{mode.f90}, \src{linear\_terms.F90} and \src{exp\_integration.F90}, in that order.

\begin{itemize}
\item Modules *.F90 are put through the preprocessor mainly for processing \#if blocks for MPI.
\item Modules *.f90 are not put through the preprocessor.
\end{itemize}

\section{Philosophy behind the coding}

\begin{itemize}
 \item The code is designed to make initialisation procedures easy to read.  As such they may not be the most computationally efficient ways of doing things. Only the computationally expensive parts of the code are optimised, other parts are designed to be readable.

 \item The code has a lot built in error reporting with the idea that if you break something, or if something is currently not working, it will tell you.  In addition, the script \scripts{gkw\_run\_tests} will run the code against a set of test cases and report if any changes made have altered the physics output of the code.

 \item The code uses global variables for most things.  This is a compromise with simplicity and readability by reducing the need for functions with many arguments passed to them. 
To avoid confusing problems often associated with global variables the \texttt{use control,only : number of processors} and data hiding within modules limits the use of variables declared in other modules to those explicitly named.

\end{itemize}

\section{Checking in changes to the code}

If you are going to contribute changes to the code, you will need a bitbucket.org account and your username will need to be added to the list of authorised users.  More details can be found on the \href{https://bitbucket.org/gkw/gkw/wiki/Git_Bitbucket}{GKW wiki}.

Think carefully before you push something as it will be visible for all the world to see, forever.  Do not commit any code that is not open source.  The repository should only be used for plain text files (and possibly some small figures).  Before you push any changes to the source, you should run the test script \texttt{gkw\_run\_tests} to check you have not broken the code.  This script runs the code under a number of scenarios and checks if the physics output has changed relative to a previously generated reference output.

When committing (\texttt{git commit}) you should specify a log message, for which a text editor will open.  This message is stored in the Git log. The first line should be a brief summary, followed by a blank line, followed by the details.   The default text editor can be changed by putting (for example) \texttt{export EDITOR=kwrite} in your \texttt{.bashrc}.

\section{Where can I get more help?}

The following are other useful GKW resources.

\begin{itemize}
\item The GKW manual at \href{http://www.gkw.org.uk/tikiwiki/Manual}{http://www.gkw.org.uk/tikiwiki/Manual} has a detailed description of the gyrokinetic framework as is implemented in GKW covering the equations, co-ordinate system, and normalisations of the code.  It also includes further details on running the code, input and output files, and the benchmarks.

\item A full listing of all the input variables is provided in \doc{input.dat.sample}

\item An automatically generated reference dictionary for the code is at 
\href{http://www.gkw.org.uk/doxygen}{http://www.gkw.org.uk/doxygen}

\item Browse some of the papers in which GKW has been used to get an idea of what it does.

\href{http://www.gkw.org.uk/tikiwiki/publications}{http://www.gkw.org.uk/tikiwiki/publications}

\item The wiki at \href{https://bitbucket.org/gkw/gkw/wiki}{https://bitbucket.org/gkw/gkw/wiki}.

\item If you find a problem with code, please submit it to the issue tracker:

\href{https://bitbucket.org/gkw/gkw/issues}{https://bitbucket.org/gkw/gkw/issues}

\item Arthur Peeters obviously knows the answer to most GKW questions, but other people working with the code may know the answer to simple questions, just ask.

\end{itemize}

\end{document}

### 2022-05-03 gkw/doc/etc/derivation.tex
(1) Added gkw Clone gwk  #3

\documentclass[a4paper,10pt]{article}

\usepackage{amsfonts}
\usepackage{amsmath}
% \usepackage{appendix}
\usepackage{graphicx}
\usepackage{hyperref}

\hypersetup{
pdftitle={Derivation of the gyrokinetic equations},
pdfauthor={Gabor Szepesi},
pdfkeywords={gyrokinetics, fully electro-magnetic, guiding-centre, gyrocentre, Lie transform, differential, fundamental one-form},
% bookmarks,
plainpages=false,
pdfborder={0 0 0},
colorlinks=true,
linkcolor=blue,
citecolor=red,
pdfstartview={FitH},
bookmarksopen=false,
unicode
}

%opening
\title{Derivation of the fully electro-magnetic, non-linear, gyrokinetic Vlasov--Maxwell equations in a rotating frame of reference for GKW with Lie transform perturbation method}
\author{G\'abor Szepesi}

\newcommand{\dpt}[1]{\frac{\partial #1}{\partial t}}
\newcommand{\dt}[1]{\frac{\mathrm{d} #1}{\mathrm{d} t}}
\newcommand{\st}[1]{\mathrm{#1}} %subtitle text
\renewcommand{\vec}[1]{\mathbf{#1}}

\graphicspath{{../fig/}}

\begin{document}

\maketitle

% \begin{abstract}
% The new feature of parallel perturbation in the magnetic field has been under development for the gyrokinetic simulation code GKW. Compression of the magnetic field leads to modes which are not present in electrostatic and partially electromagnetic cases, e.g. the compressional Alfv\'en wave, and it is likely to change significantly the growth rates of other modes in the high beta regime. In this work a detailed derivation of the fully electromagnetic gyrokinetic equations in Lagrangian formalism is shown. The transformation of the Lagrangian is performed with the Lie transform perturbation method. A straightforward way of implementation in GKW is discussed.
% \end{abstract}

\tableofcontents

\newpage

%-----------------------------------------------------------------------
\section{Introduction}
In this document the derivation of the gyrokinetic Vlasov--Maxwell system of equations is presented. First, in section \ref{sec:gk_concept}, the main motivation and the basic mathematical concept of the gyrokinetic transformation based on the Lie-transform perturbation method is outlined. A constructive derivation of the gyrokinetic equations is shown in section \ref{sec:gk_derivation}. This calculation is based on the work by Dannert presented in his thesis \cite{dannert}. The main difference is that this derivation includes plasma rotation and is formulated in a co-rotating frame of reference. The gyrokinetic Maxwell-equations are shown in detail, in particular the parallel component of Amp\`ere's law where a minor correction of the equation is suggested.

The equations derived here are solved by the gyrokinetic code GKW \cite{gkw}. The parallel Amp\`ere's law has recently been implemented in the code and is required for an accurate modelling of high beta discharges.

%-----------------------------------------------------------------------
\section{Motivation and Basic Mathematical Concept of Gyrokinetics} \label{sec:gk_concept}
Despite the fact that the forces acting on individual particles in a plasma can be analytically expressed, the solution of the equation of motion for every single particle is presently impossible. In order to characterize the macroscopic behaviour of the plasma, such detailed knowledge is not even required. A statistical method, based on describing the evolution of the distribution function of the particles both in real and velocity space, is sufficient. The underlying theory is called the kinetic theory and the equation governing the dynamics of the distribution function is the Vlasov-equation (see any textbook on plasma physics, for example \cite{bellan}):
\begin{equation}
 \dpt{f} + \vec{v} \cdot \frac{\partial f}{\partial \vec{x}} + \vec{a} \cdot \frac{\partial f}{\partial \vec{v}} = 0
 \label{eq:vlasov_basic}
\end{equation}
where $f=f(\vec{x},\vec{v},t)$ is the distribution of the particles in the six-dimensional phase space, $\vec{x}$ is the real spatial coordinate, $\vec{v}$ is the velocity space coordinate, $t$ is time and $\vec{a}$ is the acceleration of particles determined by the force acting on them. Binary interactions between particles can be included in this equation with a collision operator on the right hand side. The study of collisions is in itself a complex area, and for the purpose of this introduction a collisionless case is considered.

Without collisions the particles are accelerated only by electromagnetic forces. The electromagnetic fields are determined by the density and current of the plasma through Maxwell's equations, which are expressed as moments of the distribution function. The acceleration is therefore a non-trivial function of the distribution function and the third term on the left hand side becomes non-linear in $f$. This means that the Vlasov-equation is a complicated integro-differential equation. And the fact that it is six-dimensional (not counting the different species of the plasma), makes its numerical treatment challenging.

Gyrokinetic theory is basically a method that enables the numerical solution of the Vlasov-equation. The idea is based on the fact that one component of the single particle motion in a magnetized plasma is always a gyration around the magnetic field lines. 
Although gyro-motion leads to fundamental physical phenomena, such as drifts, and therefore strongly influences plasma turbulence, the exact knowledge of where the particles reside on their respective gyro-orbits is not required for estimating macroscopic plasma confinement and transport. The gyro-motion can be averaged out and the orbiting particles replaced by an associated gyro-centre that moves according to the particle drifts. The advantage of gyro-averaging is twofold: first, in an appropriately chosen coordinate system (where the gyro-angle is one of the coordinates), it reduces the dimensionality of the problem from six to five. Second, the gyro-motion typically takes place on a much faster time scale than the turbulent processes of interest. Therefore, once the gyro-averaging has been performed, there is no need to resolve the rapid gyromotion of the particles, and a significantly larger time-step can be applied in the numerical scheme.

In a stationary plasma with uniform magnetic field the projection of the particle orbits onto the plane perpendicular to the magnetic field is a circle. Integrating the particle motion along the gyro-orbit in this case is straightforward since none of the quantities depend on the gyro-phase. 
However, turbulence in plasmas is characterized by small scale and small amplitude fluctuations superimposed on the quasi-stationary equilibrium. These fluctuations vary on the length scale of the Larmor-radius and therefore they reintroduce the gyro-phase dependence to the system. Although a direct averaging over the gyro-angle is still possible, in modern gyrokinetic theory the averaging process is regarded as a phase space transformation. The basic idea is that the six-dimensional phase space manifold is mapped onto itself in a way that the gyro-phase dependence of the fluctuations are asymptotically removed from the equations of motion up to a certain order \cite{brizard}. A rigorous mathematical treatment of the problem is based on the Lie-transform perturbation method \cite{littlejohn_gc}.

%-----------------------------------------------------------------------
\subsection{The Lie-transform Perturbation Method} \label{sec:lie}
The Lie transform perturbation method is a general way of treating small-amplitude perturbations to arbitrary orders with near-identity, continuous mappings of manifolds. The underlying idea is that the points of the manifold are infinitesimally transported along a pre-defined vector field. The vector field can be pictured as a breeze blowing away the points of the manifold, like grains of sand. This defines a mapping $\Phi$ of the manifold $M$ onto itself generated by a vector field $G$: $\Phi: M \to M$.

%-----------------------------------------------------------------------
\subsubsection{Pull-back, push-forward}
As a result of this mapping not only the points of the manifold are transported, the functions (mapping of the manifold to the real numbers) and curves (mapping of the real numbers to the manifold) change, as well. Figure \ref{fig:pull-back} introduces two important concepts arising from this: the pull-back and push-forward operators. For the sake of generality we define a non-invertible mapping $\Phi$ between two manifolds $M$ and $N$. We assume that there exists a curve on the original manifold $M$ represented by the mapping $\gamma: \mathbb{R} \to M$. One possible definition of a vector space upon a manifold is through finding equivalence classes of curves (for details see \cite{fecko} or any textbook on differential geometry). A vector space upon manifold $M$ will be denoted by $\mathcal{T}^1(M)$ indicating that the vectors lie in the local tangent plane to the manifold. The upper index shows that this is a special, one times contravariant, case of a general $\mathcal{T}_q^p$ tensor space. Note, that a vector space $\mathcal{T}^1(M)$ is not the same as one particular vector field $G$ generating the transform: $G \in \mathcal{T}^1(M)$. The curve $\gamma$ is therefore related to a vector at point $x$ denoted with $\dot{\gamma} \in \mathcal{T}^1(x)$. If we now apply the mapping $\Phi$, the points of curve $\gamma \subset M$ are simply transported to $N$ and thus a curve $\Phi \circ \gamma$ and an associated vector in $\Phi(x)$ naturally arise. The two-step process of moving along the arrows $\gamma$ and $\Phi$ can be substituted by a single mapping from $\mathbb{R}$ directly to $N$. With other words, the curve $\gamma$ and the vector $\dot{\gamma}$ are \emph{pushed forward} from manifold $M$ to $N$ by the mapping $\Phi$. The push-forward operator acting on the vector space is denoted by $\Phi_*$ and the new vector by $\Phi_* \dot{\gamma}$. Note, that this process does not work in the opposite direction: if a curve originally existed on $N$ it could not be pushed forward to $M$ unless $\Phi$ was invertible. However, if a function $\Psi: N \to \mathbb{R}$ is defined on the target manifold $N$ it can be \emph{pulled back} to $M$. Any point on $N$ that is mapped to a certain real value by $\Psi$ has an origin in $M$. Thus, following the two-step process of moving along $\Phi$ and then $\Psi$ can be replaced by a single effective arrow from $M$ to $\mathbb{R}$ $\Psi \circ \Phi$.

\begin{figure}
 \begin{center}
	\includegraphics[scale=0.6]{pullback-pushforward-mod.eps}
 \end{center}
 \caption{The pull-back and push-forward operators induced by a general continuous mapping $\Phi$ between two differentiable manifolds $M$ and $N$. Since $\Phi$ is generally not invertible, the existence of a mapping $\mathbb{R} \to M$ induces a mapping $\mathbb{R} \to N$ but not the other way around. Similarly $\Psi: N \to \mathbb{R}$ naturally leads to $\Psi \circ \Phi: M \to \mathbb{R}$.}
 \label{fig:pull-back}
\end{figure}

For our present purpose, the application of Lie-transform method in gyrokinetics, we have to consider an other mathematical object: functions mapping the vectors on a manifold to the real numbers. These can also be thought of as elements of the dual space of the original vector space and denoted as $\mathcal{T}_1(M)$. If we replace $M$ and $N$ on figure \ref{fig:pull-back} with $\mathcal{T}^1(M)$ and $\mathcal{T}^1(N)$, such a mapping will be analogous to $\Psi: N \to \mathbb{R}$ in the previous example. They are therefore pulled back from the vector space upon the target manifold $N$ to that upon $M$. The pull-back operator acting on the dual space $\mathcal{T}_1(N)$ will be denoted by $\Phi^*$. The reason why we have to consider this object is because in the Lagrangian formalism of the gyrokinetic transformation the Lagrangian itself is represented by a differential one-form on the phase space. A complete introduction to differential forms is beyond the scope of this document. We simply state that differential one-forms are elements of the dual vector space upon a manifold. We can therefore conclude, that when the gyrokinetic transformation is applied, the Lagrangian will be transformed by the pull-back operator.

%-----------------------------------------------------------------------
\subsubsection{Transformation of a one-form}
Let us assume that the infinitesimal transport $\Phi$ on manifold $M$ generated by the vector field $G$ can be expressed as a function of a small parameter $\varepsilon$. If we want to calculate how a general covariant tensor field $A \in \mathcal{T}_q(M)$ changes due to the transport, we have to compare it with its original state \emph{at the same point}. In order to do this, we have to pull the resulting tensor space back to the original manifold. This operation is expressed by the Lie-derivative:
\begin{eqnarray}
	\mathcal{L}_G A &=& \left. \frac{\mathrm{d}}{\mathrm{d} \varepsilon} \right|_0 \Phi^*_{\varepsilon} A  \qquad \Longrightarrow \\
	\Phi^*_{\varepsilon} A &=& A + \varepsilon \mathcal{L}_G A + \mathcal{O}(\varepsilon^2) \qquad \varepsilon \ll 1.
\end{eqnarray}
Note, that the above formula is generally true if $\Phi^*_{\varepsilon}$, and therefore $\Phi$, is differentiable. $\Phi$ is usually assumed to be an exponential function of $\varepsilon$: $\Phi = \st{e}^{\varepsilon G}$ \cite{littlejohn}. This means that the pull-back operator can also be written as an exponential containing the Lie-derivative. If the manifold $M$ is such that Taylor-series converge ($M$ is $C^{\omega}$ class \cite{fecko}) then it can be approximated as
\begin{equation}
	\Phi^*_{\varepsilon} = e^{\varepsilon \mathcal{L}_G} = 1 + \varepsilon \mathcal{L}_G + \frac{\varepsilon^2}{2!} \mathcal{L}_G \mathcal{L}_G + \mathcal{O}(\varepsilon^3).
	\label{eq:pullback_taylor}
\end{equation}

Let $\Gamma \in \mathcal{T}_1(M)$ an arbitrary one-form on $M$. If it was defined on $N$ then it could simply be pulled back to $M$ by $\Phi^*$. However, if we want to evaluate it on the target manifold, we have to apply the inverse of the pull-back operator. The inverse generally exists if $\Phi$ itself is invertible. This, together with the condition on differentiability, requires $\Phi$ to be a diffeomorphism. In this particular case the exponential form of $\Phi$ satisfies this condition and one can write
\begin{equation}
	\left(\Phi^{*}_{\varepsilon}\right)^{-1} = e^{- \varepsilon \mathcal{L}_G} = 1 - \varepsilon \mathcal{L}_G + \frac{\varepsilon^2}{2!} \mathcal{L}_G \mathcal{L}_G + \mathcal{O}(\varepsilon^3)
	\label{eq:pullback_taylor_inv}
\end{equation}
and the one-form on the target manifold as
\begin{equation}
	\bar{\Gamma} = (\Phi_{\varepsilon}^*)^{-1} \Gamma + \mathrm{d}S
	\label{eq:gyro_lagrangian_general}
\end{equation}
where $\bar{\Gamma} \in \mathcal{T}_1(N)$. $\mathrm{d}S$ expresses gauge freedom, that is, we allow for a constant additional term in the one-form. This is related to  the motion being independent up to an additive constant in the Lagrangian. Finally, the Lie-derivative of the one-form in terms of the generating vector field can be calculated using the homotopy formula \cite{littlejohn}:
\begin{equation}
	(\mathcal{L}_G \Gamma)_i = G^j \left( \frac{\partial \Gamma_i}{\partial x^j} - \frac{\partial \Gamma_j}{\partial x^i} \right)
	\label{eq:homotopy}
\end{equation}
where $x^i$ are contravariant coordinates and $G^i$ are the components of the generating vector field.

%-----------------------------------------------------------------------
\subsubsection{Perturbations} \label{sec:perturbations}
So far we have expressed how an arbitrary one-form changes under an infinitesimal transformation of the manifold. The reason why this is interesting from a practical point of view is perturbations. Perturbations can be modelled by writing the one-form as a sum of contributions of different orders:
\begin{equation}
\Gamma = \Gamma_0 + \varepsilon \Gamma_1 + \varepsilon^2 \Gamma_2 + \dots.
\end{equation}
A series of transformations associated to each of these terms can be introduced and the overall inverse pull-back operator written as
\begin{eqnarray*}
	(\Phi_{\varepsilon}^*)^{-1} &=& \dots e^{-\varepsilon^2 \mathcal{L}_{G_2}} e^{-\varepsilon \mathcal{L}_{G_1}} \\
	&=& 1 - \varepsilon \mathcal{L}_{G_1} + \varepsilon^2 \left( \frac{1}{2} \mathcal{L}_{G_1}^2 - \mathcal{L}_{G_2} \right) + \mathcal{O}(\varepsilon^3)
\end{eqnarray*}
where the second line is obtained by using equation \ref{eq:pullback_taylor_inv} for each of the factors. Finally, substituting to equation \ref{eq:gyro_lagrangian_general} and separating the orders lead to:
\begin{eqnarray*}
	\bar{\Gamma}_0 &=& \Gamma_0 + \mathrm{d}S_0 \\
	\bar{\Gamma}_1 &=& \Gamma_1 - \mathcal{L}_G \Gamma_0 + \mathrm{d}S_1 \\
	\bar{\Gamma}_2 &=& \Gamma_2 - \mathcal{L}_{G_1} \Gamma_1 + \left( \frac{1}{2} \mathcal{L}_{G_1}^2 - \mathcal{L}_{G_2} \right) \Gamma_0 + \mathrm{d}S_2 \\
	&\vdots&
\end{eqnarray*}
This introduction shows how the perturbation method can be applied up to arbitrary orders. In the present work perturbations only up to first order in an appropriate small parameter will be considered. Choosing the zeroth order gauge function $S_0$ to be zero and using the homotopy formula \ref{eq:homotopy}, the first order perturbation can be finally expressed as
\begin{eqnarray}
	\bar{\Gamma}_0 &=& \Gamma_0 \\
	\bar{\Gamma}_{1,i} &=& \Gamma_{1,i} - G_1^{j} \left( \frac{\partial \Gamma_{0,i}}{\partial x^{j}} - \frac{\partial \Gamma_{0,j}}{\partial x^{i}} \right) + \frac{\partial S_1}{\partial x^{i}}. 
	\label{eq:gy_final}
\end{eqnarray}

%-----------------------------------------------------------------------
\subsection{Lagrangian Formalism and Lie-transform in Gyrokinetics} \label{sec:lie_transform_in_gk}
In this work the derivation of the gyrokinetic equations in Lagrangian formalism is presented. In general kinetic theory the Lagrangian is used to derive the equations of motion through the Euler--Lagrange equations (\ref{eq:euler-lagrange}). The time derivatives of the coordinates enter the Vlasov equation which is solved for the distribution function. As mentioned in the introduction of the chapter, the Vlasov-equation is coupled with Maxwell's equations through the dependence of the electro-magnetic fields on the distribution function. An analytical solution of the Vlasov--Maxwell system is rarely possible, typically an iterative method outlined on figure \ref{fig:lagrangian_formalism} is used in numerical schemes. In the first iteration an initial estimate of the electro-magnetic fields is used for the calculations of the distribution function. Then, the fields are updated by solving the Maxwell-equations are solved with the first approximation of the distribution function. The process is continued until convergence is reached. 
\begin{figure}
 \begin{center}
	\includegraphics[scale=0.5]{lagrangian-formalism.eps}
 \end{center}
 \caption{Iterative solution of the Vlasov--Maxwell system in Lagrangian formalism.}
 \label{fig:lagrangian_formalism}
\end{figure}

In the previous section (\ref{sec:lie}) it was shown what happens to an arbitrary differential one-form under an infinitesimal transformation of the manifold described by a generating vector field. In gyrokinetics the aim is to find an appropriate transformation of the six-dimensional phase space so that the gyro-angle dependence of the perturbations are asymptotically removed from the Lagrangian. This is now an \emph{inverse} problem: the vector field generating the transport is not specified, its components are derived based on prescribed conditions on the gyrokinetic Lagrangian.

If the new Lagrangian, the starting point of the graph on figure \ref{fig:lagrangian_formalism} is known, why is it still needed to derive the generating vector field? The gyrokinetic Lagrangian leads to the equations of motion in the transformed coordinates and a Vlasov-equation describing the evolution of the transformed gyrokinetic distribution function. However, the electro-magnetic fields depend on the actual, untransformed positions and velocities of the plasma particles. The transformed distribution function has to be pulled back to the original manifold in order to solve Maxwell's equations. The transformation rule, i.e. the components of the generating vector field, must be derived in order to carry out this pull-back operation.

%-----------------------------------------------------------------------
\section{Derivation of the Gyrokinetic Equations}  \label{sec:gk_derivation}
The derivation outlined here closely follows the structure of the calculation described by Dannert in his thesis \cite{dannert}. The major difference is that in this work the derivation is performed in presence of finite plasma rotation in a co-rotating frame of reference.

The derivation of the gyrokinetic equations is carried out through a two-step coordinate transformation method. The first step is a change of coordinates performed in equilibrium without perturbations. The new coordinates are the so called \emph{guiding-centre} coordinates (and the manifold the guiding-centre phase space) which are more suitable to describe the gyrating motion of the charged particles in the stationary magnetic field. The derivation of the guiding-centre Lagrangian is detailed in section \ref{sec:guiding}.

The second step is performed when the perturbations are introduced. It is based on the Lie-transform perturbation method and decouples the gyro-angle dependence of the Lagrangian in the presence of small scale fluctuations. The transformation takes us from the guiding-centre phase space to the so called \emph{gyro-centre} phase space. This process is explained in section \ref{sec:gyro}.

Before proceeding to the derivation of the gyro-kinetic Lagrangian, the ordering assumptions are clarified in section \ref{sec:ordering} and the notation required for the rotating frame of reference is introduced in section \ref{sec:rotation}.

%-----------------------------------------------------------------------
\subsection{Notes on Ordering} \label{sec:ordering}
It has been emphasized in the previous section that the Lie-transform method is applicable for small perturbations. In the general theory this has been expressed by the smallness of the parameter $\varepsilon$ and the linear approximation of the exponential transformation formulae. In gyrokinetic theory the aim is to decouple the effect of small-scale, small amplitude fluctuations of the plasma in the Lagrangian. It is therefore natural to choose either of these properties of the fluctuations as the small parameter in this problem. These statements can be formalized by the usual ordering assumptions applied in gyrokinetic theory \cite{brizard}
\begin{eqnarray}
	\frac{|\mathbf{A}_1|}{|\mathbf{A}_0|} &\sim& \frac{\Phi_1}{\Phi_0} \sim \varepsilon_{\delta} \ll 1 \\ \label{eq:eps_delta}
	\rho \frac{\nabla B_0}{B_0} &\sim& \rho \frac{\nabla E_0}{E_0} \sim \frac{\rho}{L_B} \sim \varepsilon_B  \ll 1\\ \label{eq:eps_B}
	k_{\perp} \rho &\sim& \varepsilon_{\perp} \sim 1 \\ \label{eq:eps_perp}
	\frac{\omega}{\omega_{\st{L}}} &\sim& \varepsilon_{\omega} \ll 1 \label{eq:eps_omega}
\end{eqnarray}
where $\vec{A}$ and $\Phi$ are the vector and scalar potentials, $\vec{B}$ and $\vec{E}$ are the magnetic and electric fields, $\omega$ and $k_{\perp}$ are the typical mode frequency and perpendicular wavenumber, $\rho$ and $\omega_{\st{L}}$ are the Larmor-radius and Larmor-frequency. The equilibrium quantities are denoted with 0 and the fluctuations with 1 subscript. The above equations express that the fluctuations have a much smaller magnitude than the corresponding equilibrium values, their typical time scale is much slower than the Larmor-frequency, their characteristic length scale is of the order of the Larmor-radius and it is typically much shorter than the equilibrium spatial variation scale. The gyrokinetic equation as derived in this document is valid if these ordering assumptions are true. The three different small parameters described here arise for different physical reasons. However, the general practice is to assume that they are of similar order and substitute them with one parameter. In this work the ratio of the reference thermal Larmor-radius and the equilibrium magnetic length scale $\rho_* = \frac{\rho_{\st{ref}}}{L_B} = \frac{m_{\st{ref}} v_{th,ref}}{e B_{\st{ref}}} \sim \varepsilon_B \sim \varepsilon_{\delta} \sim \varepsilon_{\omega}$ is chosen for this purpose, and the equations are derived up to first order in this quantity.

%-----------------------------------------------------------------------
\subsection{Plasma Rotation} \label{sec:rotation}
Turbulence in a rotating plasma can be conveniently described in a co-rotating
frame of reference. Although a detailed discussion of plasma rotation is outside
the main scope of this document, for the sake of generality the derivation of
the gyrokinetic equations is shown in a reference frame rigidly rotating with
velocity $\vec{u}_0$. The plasma rotation is assumed to be toroidal, the
poloidal component is typically much smaller (the order of the diamagnetic velocity) and neglected here. The reference frame is therefore chosen to rotate in the toroidal direction, and its velocity can be expressed with a constant angular frequency in the form
\begin{equation}
\vec{u}_0 = \vec{\Omega} \times \vec{x} = R^2 \Omega \nabla \varphi
\label{eq:frame_frequency}
\end{equation}
where $\varphi$ is the toroidal angle and $R \nabla \varphi$ is the unit vector in the toroidal direction \cite{peeters_rotation}.

The rotation of the plasma in the laboratory frame is typically not a rigid body rotation, it is characterized by a radial profile of angular velocity $\hat{\Omega}(\psi)$. The rotating frame is chosen in a way that its angular frequency $\Omega$ matches the plasma rotation at a certain radial point: $\Omega = \hat{\Omega}(\psi_r)$. This method is suitable for local gyrokinetic studies, since in the rotating frame on the surface labelled by $\psi_r$ the plasma rotation vanishes. However, a finite gradient of the rotation profile has to be taken into account. The plasma rotation in the co-rotating frame of reference will be denoted as $\omega_{\varphi}(\psi) = \hat{\Omega}(\psi) - \Omega$. The associated plasma rotation speed along the magnetic field line in the co-rotating frame can be expressed as
\begin{equation}
 u_{\parallel} = \frac{R B_{\st{t}}}{B} \omega_{\varphi}(\psi)
 \label{eq:parallel_rot_speed}
\end{equation}
where $B_{\st{t}}$ is the toroidal component of the magnetic field \cite{peeters_rotation}.

%-----------------------------------------------------------------------
\subsection{Lagrangian in Guiding-centre Coordinates} \label{sec:guiding}
% In this subsection the Lagrangian, or fundamental one-form, of a plasma particle in guiding-centre coordinates is derived. 
The particle phase space consists of three real and three velocity space components: $(\mathbf{x},\mathbf{v})$. A point in this space describes a particle at a certain position travelling according to a certain velocity vector. The Lagrangian, or fundamental one-form, of a particle with mass $m$ and charge number $Z$ in an electro-magnetic field is written as
\begin{equation}
	\gamma = \gamma_{\nu} \mathrm{d} z^{\nu} = \left( m \mathbf{v} + Z e \mathbf{A}(\mathbf{x}) \right) \cdot \mathrm{d} \mathbf{x} - \left( \frac{1}{2} m v^2 + Z e \Phi(\mathbf{x}) \right) \mathrm{d}t,
\end{equation}
where $\nu$ indexes the six coordinates, and summation over repeated indices is meant \cite{scott_gotit}. $\vec{A}$ and $\Phi$ are the  vector and scalar potentials, respectively. The first term, multiplied by the differential of the spatial coordinates, is called the symplectic part and the second is the Hamiltonian: $H(\mathbf{x},\mathbf{v})$.

In order to write the Lagrangian in a rotating frame of reference, both the velocity coordinates and the fields have to be modified according to the Lorentz-transformation:
\begin{equation}
\vec{v} \to \vec{v} + \vec{u}_0 \qquad \vec{E} \to \vec{E} + \vec{u}_0 \times \vec{B} \qquad \Phi \to \Phi + \vec{A} \cdot \vec{u}_0.
\end{equation}
Following the calculation outlined in \cite{peeters_rotation} the Lagrangian becomes
\begin{equation}
	\gamma = \left( m \mathbf{v} + m \vec{u}_0 + Z e \mathbf{A}(\mathbf{x}) \right) \cdot \mathrm{d} \mathbf{x} - \left( \frac{1}{2} m v^2 + Z e \Phi(\mathbf{x}) - \frac{1}{2} m u_0^2 \right) \mathrm{d}t.
\end{equation}

Let us introduce the guiding-centre phase space with the following coordinate transformation:
\begin{eqnarray}
	\mathbf{X}(\mathbf{x},\mathbf{v}) &=& \mathbf{x} - \mathbf{r} = \mathbf{x} - \rho(\mathbf{x},\vec{v}) \vec{a}(\vec{x},\vec{v}) \label{eq:X}\\
	v_{\parallel}(\mathbf{x},\mathbf{v}) &=& \mathbf{v} \cdot \mathbf{b}(\mathbf{x}) \label{eq:v_par}\\
	\mu(\mathbf{x},\mathbf{v}) &=& \frac{m v_{\st{L}}^2(\mathbf{x})}{2 B(\mathbf{x})} \label{eq:mu}\\
	\theta(\mathbf{x},\mathbf{v}) &=& \cos^{-1} \left( \frac{1}{v_{\st{L}}(\mathbf{x})} (\mathbf{b}(\mathbf{x}) \times \mathbf{v}) \cdot \mathbf{e}_1 \right). \label{eq:theta}
\end{eqnarray}
The new coordinates are the position of the centre of the particle's gyro-orbit, or guiding-centre $\mathbf{X}$, the particle velocity parallel to the equilibrium magnetic field $v_{\parallel}$, the magnetic moment $\mu$ and the gyrophase $\theta$. $\mathbf{b}(\mathbf{x})$ is the unit vector in the direction of the equilibrium magnetic field and $\rho(\mathbf{x},\vec{v}) \vec{a}(\vec{x},\vec{v})$ is the vector pointing from the guiding-centre to the particle's position. Its direction is determined by the unit vector $\vec{a}$ and its length is the Larmor-radius
\begin{equation}
	\rho(\mathbf{x},\vec{v}) = \frac{m v_{\st{L}}(\mathbf{x})}{|Z| e B(\mathbf{x})}
	\label{eq:rho}
\end{equation}
where $v_{\st{L}}$ is the velocity of the gyro-motion, or Larmor-velocity. 
The absolute value of the charge number is needed so the ion and electron Larmor-radii are both positive. The fact that their gyration is in opposite direction is expressed by the vectorial factor $\vec{a}$.
The unit vector $\vec{a}$ can be expressed in a local orthonormal basis as the function of the gyro-angle:
\begin{equation}
	\mathbf{a}(\theta) = \mathbf{e}_1 \cos \theta + \mathbf{e}_2 \sin \theta.
	\label{eq:a}
\end{equation}
The vectors $\mathbf{b}$, $\mathbf{e}_1$ and $\mathbf{e}_2$ form a local righthanded Cartesian coordinate system at the guiding-centre position.

The transformation of the one-form due to the change of coordinates can be expressed as
\begin{equation}
	\Gamma_{\eta} = \gamma_{\nu} \frac{\mathrm{d} z^{\nu}}{\mathrm{d} Z^{\eta}}
	\label{eq:gc_transform}
\end{equation}
where $\Gamma_{\eta}$ is a component of the guiding-centre fundamental one-form \cite{scott_gotit}. In order to calculate these new components the transformation equations \ref{eq:X}, \ref{eq:v_par}, \ref{eq:mu} and \ref{eq:theta} have to be inverted in order to provide the old coordinates as functions of the new ones: $z(Z)$. It is clear that the direct transformation is uniquely determined if the magnetic field is known at the particle's position. However, when the inverse is taken and particle coordinates are calculated, the particle position can not be explicitly expressed due to the dependence of the Larmor-radius on $\vec{x}$ through the magnetic field. The coordinates are therefore Taylor-expanded in space around the guiding-centre location $\vec{X}$:
\begin{eqnarray*}
	\mathbf{x} &=& \mathbf{X} + \rho(\mathbf{x}) \mathbf{a}(\theta) \\
	\rho(\mathbf{x}) &\approx& \rho(\mathbf{X}) + \left( \frac{\partial \rho(\mathbf{x})}{\partial \mathbf{x}} \right)_{\mathbf{X}} \rho(\mathbf{x}) \mathbf{a}(\theta) + \mathcal{O}((\rho(\mathbf{x}) \mathbf{a}(\theta))^2).
\end{eqnarray*} 
It can be shown that the first order correction in the Taylor-expansion containing $\rho \frac{\partial \rho}{\partial \vec{x}} \sim \rho^2$ leads to second order terms in $\rho_*$. It is thus sufficient to keep only $\rho(\vec{x}) \approx \rho(\vec{X})$ which gives
\begin{equation}
	\vec{x}(\vec{X},\theta) \approx \vec{X} + \rho(\vec{X}) \vec{a}(\theta). \label{eq:x(X)}
\end{equation}
Note that the Larmor radius $\rho$ also depends on the velocity space coordinates $\vec{v}$ in particle phase space, or the magnetic moment $\mu$ in guiding-centre phase space through the formula $\rho(\vec{X},\mu) = \frac{1}{Z e} \sqrt{\frac{2 \mu m}{B(\vec{X})}}$. This dependence will not be explicitly indicated unless greater clarity is called for.

The particle velocity is the sum of three contributions: the velocity along the magnetic field, the gyration velocity and the drift velocities. It will be shown later that the particle drifts are described by the motion of the guiding centre. Hence the velocity in the guiding-centre frame can be written as
\[\vec{v} = v_{\parallel} \mathbf{b}(\mathbf{x}) + \mathbf{v}_{\st{L}} = v_{\parallel} \mathbf{b}(\mathbf{x}) + \rho(\mathbf{x}) \dot{\mathbf{a}}(\theta)\] 
Applying Taylor expansion again around $\vec{X}$ we obtain
\begin{equation}
	\vec{v}(\vec{X},v_{\parallel},\mu,\theta) \approx v_{\parallel} \left[ \mathbf{b}(\mathbf{X}) + \frac{\partial \mathbf{b}(\mathbf{X})}{\partial \mathbf{X}} \cdot \mathbf{a}(\theta) \rho(\mathbf{X}, \mu) \right] + \rho(\mathbf{X}, \mu) \dot{\mathbf{a}}(\theta)
	\label{eq:v(X)}
\end{equation}

The transformation formula (\ref{eq:gc_transform}) can now be applied to express the fundamental one-form in the new coordinates. The $\vec{X}$ component takes the form
\[ \Gamma_{X^i} = \gamma_{x^j} \frac{\mathrm{d} x^j}{\mathrm{d} X^i} + \underbrace{\gamma_{v^j}}_{0} \frac{\mathrm{d} v^j}{\mathrm{d} X^i} + \gamma_t \underbrace{\frac{\mathrm{d} t}{\mathrm{d} X^i}}_{0}. \]
The required derivative is
\[\frac{\mathrm{d} x^j}{\mathrm{d} X^i} = \delta^j_i + \frac{\mathrm{d} \rho(\mathbf{X})}{\mathrm{d}X^i} a^j(\theta) \]
which substituted into (\ref{eq:gc_transform}) gives
\begin{eqnarray*}
	\Gamma_{X^i} &=& \left( m v_j + m u_{0j} + Z e A_j(\mathbf{x}) \right) \frac{\mathrm{d} x^j}{\mathrm{d} X^i} \\
	&=& \left( m v_{\parallel} b_j(\mathbf{x}) + m \rho(\mathbf{x}) \dot{a}_j(\theta) + m u_{0j} + Z e A_j(\mathbf{x})\right) \left( \delta^j_i + \frac{\mathrm{d} \rho(\mathbf{X})}{\mathrm{d}X^i} a^j(\theta) \right). %\\
\end{eqnarray*}
Expanding the quantities into Taylor-series around $\vec{X}$ and keeping terms up to first order in $\rho$ results
\begin{eqnarray*}
	\Gamma_{X^i} &=& m v_{\parallel} \left[ b_i(\mathbf{X}) + \frac{\partial b_i(\mathbf{X})}{\partial X^k} a^k(\theta) \rho(\mathbf{X}) \right] + m \rho(\mathbf{X}) \dot{a}_i + m u_{0i}\\
	&& + Z e \left[ A_i(\mathbf{X}) + \frac{\partial A_i(\mathbf{X})}{\partial X^k} a^k(\theta) \rho(\mathbf{X}) \right] + (Z e A_j(\mathbf{X}) + m u_{0j}) \frac{\mathrm{d} \rho(\mathbf{X})}{\mathrm{d} X^i} a^j(\theta).
\end{eqnarray*}
Note that $\mathbf{a}$ is perpendicular to both $\mathbf{b}$ and $\dot{\mathbf{a}}$ and the term containing $\rho \frac{\partial \rho}{\partial \vec{X}}$ can be neglected.

The gyro-averaging operator in guiding-centre phase space is simply an integral over the gyro-phase $\theta$:
\begin{equation}
	\langle \dots \rangle = \frac{1}{2 \pi} \int \limits_0^{2 \pi} \dots \, \mathrm{d} \theta.
\end{equation}
It follows from the definition of the vector $\mathbf{a}$ (equation \ref{eq:a}) that first order terms in $\mathbf{a}$ or $\dot{\mathbf{a}}$ disappear under gyro-averaging. Hence the $\theta$-integral yields
\begin{equation}
	\langle \Gamma_{X^i} \rangle = m v_{\parallel} b_i(\mathbf{X}) + m u_{0i} + Z e A_i(\mathbf{X}).
	\label{eq:gc_X}
\end{equation}

The remaining components are expressed in an analogous way. The parallel velocity component according to equation \ref{eq:gc_transform} becomes
\begin{equation}
	\Gamma_{v_{\parallel}} = \gamma_{\nu} \frac{\mathrm{d}z^{\nu}}{\mathrm{d}v_{\parallel}} = \gamma_{x^i} \underbrace{\frac{\mathrm{d}x^i}{\mathrm{d}v_{\parallel}}}_{0} + \underbrace{\gamma_{v^i}}_{0} \frac{\mathrm{d}v^i}{\mathrm{d}v_{\parallel}} = 0
	\label{eq:gc_vpar}
\end{equation}
meaning that this component remains zero in the guiding-centre approximation.

The $\mu$-component takes the form
\begin{eqnarray*}
	\Gamma_{\mu} &=& \gamma_{\nu} \frac{\mathrm{d} z^{\nu}}{\mathrm{d} \mu} = \gamma_{x^i} \frac{\mathrm{d}x^i}{\mathrm{d}\mu} = \left( m v_i + m u_{0i} + Z e A_i(\mathbf{x}) \right) \frac{\partial \rho(\mathbf{X},\mu)}{\partial \mu} a^i(\theta) \\
	&=& \left( m v_{\parallel} b_i(\mathbf{x}) + m \rho(\mathbf{x}) \dot{a}_i(\theta) + m u_{0i} + Z e A_i(\mathbf{x}) \right) \frac{\partial \rho(\mathbf{X},\mu)}{\partial \mu} a^i(\theta) \\
	&=& 
	\left[ m u_{0i} + Z e \left( A_i(\mathbf{X}) + \frac{\partial A_i(\mathbf{X})}{\partial X^k} a^k(\theta) \rho(\mathbf{X}, \mu) \right) \right] \frac{\partial \rho(\mathbf{X},\mu)}{\partial \mu} a^i(\theta).
\end{eqnarray*}
Since $\frac{\partial \rho}{\partial \mu} \sim \frac{\rho}{\mu}$, the term containing $\rho \frac{\partial \rho}{\partial \mu}$ leads to second order terms in $\rho_*$ and therefore it can be neglected. Finally, gyro-averaging gives
\begin{eqnarray}
	\langle \Gamma_{\mu} \rangle = 0.
	\label{eq:gc_mu}
\end{eqnarray}

Using relations $\dot{\mathbf{a}}(\theta) = \frac{\partial \mathbf{a}(\theta)}{\partial \theta} \dot{\theta}$ and $\dot{\theta} = \omega_{\st{L}} = \frac{Z e B}{m}$, the gyration velocity can be written as $\rho(\mathbf{X}) \dot{\theta} = \frac{Z}{|Z|} v_{\st{L}}(\mathbf{X})$. Since $\frac{\partial a_i}{\partial \theta} \frac{\partial a^i}{\partial \theta} = 1$ (see equations \ref{eq:a} and \ref{eq:v(X)}) the $\theta$-component becomes
\begin{equation}
	\Gamma_{\theta} = \gamma_{x^i} \frac{\mathrm{d} x^i}{\mathrm{d} \theta} \ \Longrightarrow\ \langle \Gamma_{\theta} \rangle = m \rho(\mathbf{X}) \frac{Z}{|Z|} v_{\st{L}}(\mathbf{X}) = \frac{2 \mu m}{Z e}.
	\label{eq:gc_theta}
\end{equation}

Time is not transformed thus the Hamiltonian part in guiding-centre coordinates remains
\begin{equation}
	\langle \Gamma_t \rangle = -\left( \frac{1}{2} m v_{\parallel}^2 + \mu B(\mathbf{X}) - \frac{1}{2} m u_0^2 + Z e \Phi(\mathbf{X}) \right).
	\label{eq:gc_t}
\end{equation}

Using equations \ref{eq:gc_X}, \ref{eq:gc_vpar}, \ref{eq:gc_mu}, \ref{eq:gc_theta} and \ref{eq:gc_t} describing the components of the fundamental one-form in guiding-centre coordinates we finally obtain
\begin{eqnarray}
	\langle \Gamma \rangle &=& \left( m v_{\parallel} \mathbf{b}(\mathbf{X}) + m \vec{u}_0 + Z e \mathbf{A}(\mathbf{X}) \right) \cdot \mathrm{d}\mathbf{X} + \frac{2 \mu m}{Z e} \mathrm{d} \theta \nonumber \\ 
	&& - \left( \frac{1}{2} m v_{\parallel}^2 + \mu B(\mathbf{X}) - \frac{1}{2} m u_0^2 + Z e \Phi(\mathbf{X}) \right) \mathrm{d}t.
	\label{eq:gc_lagrangian}
\end{eqnarray}
The equations of motion can be derived from equation \ref{eq:gc_lagrangian} with the Euler--Lagrange equations (\ref{eq:euler-lagrange}). The results are the well known drifts of the guiding centre and are not detailed here. However, it is important to note that as a consequence of of the Lagrangian being independent of the gyro-phase $\theta$, the magnetic moment $\mu$ (the associated conjugate coordinate pair of $\theta$) becomes an invariant of the motion: $\dot{\mu}=0$.

%-----------------------------------------------------------------------
\subsection{Lagrangian in Gyro-centre Coordinates} \label{sec:gyro}
In this section the fluctuations are added to the guiding-centre Lagrangian, and a transformation is derived that removes the gyro-phase dependence of the perturbations up to first order in $\rho_*$. The derivation is based on a list of requirements on the new Lagrangian and the transformation is defined in terms of the generating vector field.

%-----------------------------------------------------------------------
\subsubsection{Perturbed guiding-centre one-form}
Let us introduce small scale perturbations of the electromagnetic fields in the form
\[ \mathbf{A} = \mathbf{A}_0 + \mathbf{A}_1 \qquad \Phi = \Phi_0 + \Phi_1.\]
The equilibrium electric field is typically assumed zero in a stationary plasma, but it has to be kept in case of finite plasma rotation. According to the gyrokinetic ordering the perturbations are first order in the typical small parameters used in gyrokinetics: $\frac{A_1}{A_0} \sim \frac{\Phi_1}{\Phi_0} \sim \rho_*$. The perturbations appear in the particle phase space Lagrangian as
\begin{eqnarray*}
	\gamma &=& \gamma_0 + \gamma_1 \\
	\gamma_0 &=& \left( m \mathbf{v} + m \vec{u}_0 + Z e \mathbf{A}_0(\mathbf{x}) \right) \cdot \mathrm{d} \mathbf{x} - \left( \frac{1}{2} m v^2 - \frac{1}{2} m u_0^2 + Z e \Phi_0(\mathbf{x}) \right) \mathrm{d}t \\
	\gamma_1 &=& Z e \mathbf{A}_1(\mathbf{x}) \cdot \mathrm{d} \mathbf{x} - Z e \Phi_1(\mathbf{x}) \mathrm{d}t.
\end{eqnarray*}
The total Lagrangian has to be transformed first to the guiding-centre and then to gyro-centre phase space. The guiding-centre transformation of the equilibrium part has been completed in section \ref{sec:guiding}. The transformation of the perturbed part of the Lagrangian $\gamma_1$ is analogous to the calculation shown there. An important difference arises from the fact that fluctuating quantities vary on a small length scale and therefore Taylor expansion around the guiding-centre is not advantageous. Their values have to be taken at the particle position which is a function of the gyro-angle in guiding-centre coordinates.

The spatial components of the perturbed guiding-centre one-form are calculated as
\begin{eqnarray*}
	\Gamma_{1,X^i} &=& \gamma_{1,\nu} \frac{\mathrm{d} z^{\nu}}{\mathrm{d} X^i}= Z e A_{1,j}(\mathbf{x}) \left( \delta^j_i + \frac{\mathrm{d} \rho(\mathbf{X})}{\mathrm{d} X^i} a^j(\theta) \right) \\ 
	&=& Z e A_{1,i}(\mathbf{x}) + Z e \frac{\mathrm{d} \rho(\mathbf{X})}{\mathrm{d} X^i} A_{1,j}(\mathbf{x}) a^j(\theta) \\
	&\approx& Z e A_{1,i}(\mathbf{x}).
\end{eqnarray*}
The final approximation can be made because the second term contains $\frac{\partial \rho}{\partial \vec{x}} \vec{A}_1 \sim \rho^2$ and therefore can be neglected. The parallel velocity component remains zero since there are no terms added in the perturbed part
\begin{equation}
	\Gamma_{1,v_{\parallel}} = 0.
\end{equation}
The perturbed $\mu$ component is
\begin{equation}
 \Gamma_{1,\mu} = Z e A_{1,j}(\mathbf{x}) \frac{\rho(\mathbf{X})}{2 \mu} a^j(\theta) = \frac{Z}{|Z|} \frac{1}{v_{\st{L}}(\mathbf{X}, \mu)} \mathbf{A}_1(\mathbf{x}) \cdot \mathbf{a}(\theta),
\end{equation}
the $\theta$ component is
\begin{equation}
	\Gamma_{1,\theta} = Z e A_{1,j}(\mathbf{x}) \rho(\mathbf{X}) \frac{\mathrm{d} a^j(\theta)}{\mathrm{d} \theta} = \frac{Z}{|Z|} \frac{2 \mu}{v_{\st{L}}(\mathbf{X},\mu)} \mathbf{A}_1(\mathbf{x}) \cdot \frac{\mathrm{d} \mathbf{a}(\theta)}{\mathrm{d} \theta},
\end{equation}
and the Hamiltonian becomes
\begin{equation}
	\Gamma_{1,t} = -Z e \Phi_1(\mathbf{x}).
\end{equation}
The perturbed part of the one-form in the guiding-centre phase space can be written as
\begin{equation}
	\Gamma_1 = Z e \mathbf{A}_1(\mathbf{x}) \cdot \mathrm{d} \mathbf{X} + \frac{Z}{|Z|} \frac{\mathbf{A}_1(\mathbf{x}) \cdot \mathbf{a}}{v_{\st{L}}} \mathrm{d} \mu + \frac{Z}{|Z|} \frac{2 \mu}{v_{\st{L}}} \mathbf{A}_1(\mathbf{x}) \cdot \frac{\mathrm{d} \mathbf{a}}{\mathrm{d} \theta} \mathrm{d} \theta - Z e \Phi_1(\mathbf{x}) \mathrm{d}t.
	\label{eq:gc_oneform_pert}
\end{equation}
The complete guiding-centre Lagrangian including perturbations is the sum of equations \ref{eq:gc_lagrangian} and \ref{eq:gc_oneform_pert}.

%-----------------------------------------------------------------------
\subsubsection{Gyro-centre transformation} \label{sec:gyro_derive}
The aim is now to find a transformation that removes the gyro-angle dependence introduced by the fluctuations from the Lagrangian. Since the perturbations are first order in $\rho_*$, this one-form can be transformed into the gyro-centre phase space according to equation \ref{eq:gy_final}. Note, that the transformation is an inverse pull-back operator, which means that the new Lagrangian in the gyro-centre phase space is expressed in guiding-centre coordinates $(\vec{X},v_{\parallel},\mu,\theta)$ in order to allow the comparison of the two one-forms. However, the transformation does give rise to a new set of gyro-centre coordinates $(\bar{\vec{X}},\bar{v}_{\parallel},\bar{\mu},\bar{\theta)}$ which are not being used in the present section. The gyro-centre Lagrangian will be distinguished from its guiding-centre counterpart with an overbar $\bar{\Gamma}$.

It is not obvious which vector field to use to generate a transformation that removes the fluctuating quantities from the Lagrangian. Therefore, we do not perform a direct transformation. The vector field is derived based on a simple set of requirements on the gyro-centre Lagrangian suggested by Dannert \cite{dannert}:
\begin{eqnarray*}
	G_{1,t} = 0 &\to& \textrm{no transformation in time} \\
	\bar{\Gamma}_{1,v_{\parallel}} = 0 &\to& \textrm{no $v_{\parallel}$ component} \\
	\bar{\Gamma}_{1,\mu} = 0 &\to& \textrm{no $\mu$ component} \\
	\bar{\Gamma}_{1,\theta} = 0 &\to& \textrm{no change in $\theta$ component} \\
	\bar{\Gamma}_{1,\mathbf{X}} = Z e \langle \mathbf{A}_1(\vec{x}) \rangle &\to& \textrm{the transformation leads to the gyro-average}
\end{eqnarray*}

Let us now apply equation \ref{eq:gy_final} and the requirements above to express the components of the generator vector field that provides the desired Lagrangian. As it was mentioned in section \ref{sec:lie_transform_in_gk}, this can be considered as an inverse problem. The equation for the $\mu$ component is written as
\begin{eqnarray}
	\bar{\Gamma}_{1,\mu} &=& \frac{Z}{|Z|} \frac{\mathbf{A}_1(\mathbf{x}) \cdot \mathbf{a}(\theta)}{v_{\st{L}}(\mathbf{X,\mu})} + G_1^{\theta} \frac{2 m}{Z e} + \frac{\partial S_1}{\partial \mu} = 0 \ \ \Longrightarrow \nonumber \\
	G_1^{\theta} &=& - \frac{Z e}{2 m} \left( \frac{Z}{|Z|} \frac{\mathbf{A}_1(\mathbf{x}) \cdot \mathbf{a}(\theta)}{v_{\st{L}}(\mathbf{X,\mu})} + \frac{\partial S_1}{\partial \mu} \right).
\end{eqnarray}
The equation for the $\theta$ component gives
\begin{eqnarray}
	\bar{\Gamma}_{1,\theta} &=& \frac{Z}{|Z|} \frac{m v_{\st{L}}(\mathbf{X},\mu)}{B_0(\mathbf{X})} \mathbf{A}_1(\mathbf{x}) \cdot \frac{\mathrm{d}\mathbf{a}(\theta)}{\mathrm{d} \theta} - G_1^{\mu} \frac{2 m}{Z e} + \frac{\partial S_1}{\partial \theta} = 0 \ \ \Longrightarrow \nonumber \\
	G_1^{\mu} &=& \frac{Z e}{2 m} \left( \frac{\partial S_1}{\partial \theta} + \frac{Z}{|Z|} \frac{m v_{\st{L}}(\mathbf{X},\mu)}{B_0(\mathbf{X})} \mathbf{A}_1(\mathbf{x}) \cdot \frac{\mathrm{d}\mathbf{a}(\theta)}{\mathrm{d} \theta} \right).
\end{eqnarray}
The $v_{\parallel}$ component leads to
\begin{eqnarray}
	\bar{\Gamma}_{1,v_{\parallel}} &=& G_1^{\mathbf{X}} \cdot m \mathbf{b}_0 + \frac{\partial S_1}{\partial v_{\parallel}} = 0 \ \ \Longrightarrow \nonumber \\
	G_1^{\mathbf{X}} \cdot \mathbf{b}_0 &=& - \frac{1}{m} \frac{\partial S_1}{\partial v_{\parallel}}. \label{eq:G1X}
\end{eqnarray}
The spatial components' transformation can be written as
\begin{eqnarray}
	\bar{\Gamma}_{1,\mathbf{X}} &=& Z e \mathbf{A}_1(\mathbf{x}) + Z e G_1^{\mathbf{X}} \times \underbrace{\nabla \times \left( \mathbf{A}_0(\mathbf{X}) + \frac{m}{Z e} (v_{\parallel} \mathbf{b}_0(\mathbf{X})+\vec{u}_0)\right)}_{\equiv \mathbf{B}_0^{*}(\mathbf{X}) \equiv \nabla \times \mathbf{A}_0^{*}(\mathbf{X})} + \nonumber\\
	&& m G_1^{v_{\parallel}} \mathbf{b}_0(\mathbf{X}) + \nabla S_1 = Z e \langle \mathbf{A}_1(\mathbf{x}) \rangle \ \ \Longrightarrow \nonumber \\
	\noalign{\vskip 0.5 truecm}
	0 &=& Z e \underbrace{(\mathbf{A}_1(\mathbf{x}) - \langle \mathbf{A}_1(\mathbf{x})\rangle)}_{\widetilde{\mathbf{A}}_1(\mathbf{x})} + Z e G_1^{\mathbf{X}} \times \mathbf{B}_0^{*}(\mathbf{X}) - m G_1^{v_{\parallel}} \mathbf{b}_0(\mathbf{X}) + \nabla S_1 \label{eq:gamma_gy_X}
\end{eqnarray}
where the notation $\mathbf{B}_0^{*}(\mathbf{X}) = \nabla \times \left( \mathbf{A}_0(\mathbf{X}) + \frac{m}{Z e} (v_{\parallel} \mathbf{b}_0(\mathbf{X})+\vec{u}_0)\right)$ has been introduced. The scalar and vector potential perturbations are formally separated into a gyro-averaged and an oscillating part:
\begin{eqnarray}
	\mathbf{A}_1 &=& \widetilde{\vec{A}}_1 + \langle \mathbf{A}_1 \rangle \\
	\Phi_1 &=& \widetilde{\Phi}_1 + \langle \Phi_1 \rangle. \label{eq:av_osc_decomp_phi}
\end{eqnarray}
In order to express the required component of the generating vector field, first we take the scalar product of equation \ref{eq:gamma_gy_X} with $B_0^{*}$ to obtain $G_1^{v_{\parallel}}$, then the vector product with $\mathbf{b}_0$ to obtain $G_1^{\mathbf{X}}$.
\begin{eqnarray}
	G_1^{v_{\parallel}} &=& \frac{1}{m B_{0 \parallel}^{*}(\mathbf{X})} \left( Z e \widetilde{\vec{A}}_1(\vec{x}) \cdot \mathbf{B}_0^{*}(\mathbf{X}) + \nabla S_1 \cdot \mathbf{B}_0^{*}(\mathbf{X})\right) \\
	G_1^{\mathbf{X}} &=& -\frac{1}{B_{0 \parallel}^{*}(\mathbf{X})} \left( \widetilde{\vec{A}}_1(\vec{x}) \times \mathbf{b}_0(\mathbf{X}) + \frac{1}{m} \frac{\partial S_1}{\partial v_{\parallel}} \mathbf{B}_0^{*}(\mathbf{X}) + Z e \nabla S_1 \times \mathbf{b}_0(\mathbf{X}) \right)
\end{eqnarray}
where equation \ref{eq:G1X} and $B_{0 \parallel}^{*} = \mathbf{B}_0^{*} \cdot \mathbf{b}_0$ have been used. Finally, the transformation of the Hamiltonian part yields
\begin{eqnarray*}
	\bar{\Gamma}_{1,t} &=& -Z e \Phi_1(\mathbf{x}) + G_1^{\mathbf{X}} \cdot \left( \mu \nabla B_0(\mathbf{X}) + Z e \nabla \Phi_0(\mathbf{X}) \right) + G_1^{v_{\parallel}} m v_{\parallel} + G_1^{\mu} B_0(\mathbf{X}) + \frac{\partial S_1}{\partial t} \\
	&=& -Z e \Phi_1(\mathbf{x}) - \frac{1}{B_{0 \parallel}^{*}} \left( \widetilde{\vec{A}}_1(\vec{x}) \times \mathbf{b}_0(\mathbf{X}) + \frac{1}{m} \frac{\partial S_1}{\partial v_{\parallel}} \mathbf{B}_0^{*}(\mathbf{X}) + Z e \nabla S_1 \times \mathbf{b}_0(\mathbf{X}) \right) \cdot \\
	&& \left( \mu \nabla \mathbf{B}_0(\mathbf{X}) + Z e \nabla \Phi_0(\mathbf{X}) \right) + \frac{v_{\parallel}}{B_{0 \parallel^{*}}} \left( Z e \widetilde{\vec{A}}_1(\vec{x}) \cdot \mathbf{B}_0^{*}(\mathbf{X}) + \nabla S_1 \cdot \mathbf{B}_0^{*}(\mathbf{X})\right) + \\
	&& \frac{Z e}{2 m} \left( \frac{\partial S_1}{\partial \theta} + \frac{Z}{|Z|} \frac{m v_{\st{L}}(\mathbf{X},\mu)}{B_0(\mathbf{X})} \mathbf{A}_1(\mathbf{x}) \cdot \frac{\mathrm{d}\mathbf{a}(\theta)}{\mathrm{d} \theta} \right) B_0(\mathbf{X}) + \frac{\partial S_1}{\partial t}.
\end{eqnarray*}

Let us now take a closer look on the gauge transformation function $S_1$ and its derivatives. According to equations \ref{eq:eps_B}-\ref{eq:eps_perp} the ordering of these derivatives is as follows:
\begin{eqnarray*}
	\partial_t S_1 &\sim& \omega S_1 \sim \varepsilon_{\omega} \omega_{\st{L}} S_1 \sim \rho_* \omega_{\st{L}} S_1 \\
	\nabla_{\parallel} S_1 &\sim& \frac{1}{L_B} S_1 \sim \frac{\rho_*}{\rho} S_1 \\
	\nabla_{\perp} S_1 &\sim& \frac{1}{\rho} S_1 \\
	\partial _{v_{\parallel}} S_1&\sim& \frac{1}{v_{\st{th}}} S_1 \\
	\partial_{\mu} S_1 &\sim& \frac{B_0}{T} S_1 \\
	\partial_{\theta} S_1 &\sim& 1.
\end{eqnarray*}
Note that the characteristic length scale of the perturbations along the magnetic field lines is of order $L_B$ and that the characteristic parallel velocity and Larmor-speed is of order of the thermal velocity. Using the above assumptions and $|\mathbf{B}_0^{*}| = \left| \mathbf{B}_0 + \frac{m v_{\parallel}}{e} \nabla \times \mathbf{b}_0 \right| \sim B_0 + \rho_* B_0$, the ordering of the terms containing $S_1$ in the gyro-centre Hamiltonian can be written as
\begin{eqnarray*}
	\frac{\mathbf{B}_0^{*}}{B_{0 \parallel}^{*}} \frac{1}{m} \frac{\partial S_1}{\partial v_{\parallel}} \cdot (\mu \nabla B_0 + Z e \nabla \Phi_0) &\sim& \rho_* \omega_{\st{L}} \left( 1 + \frac{Z e \Phi_0}{\mu B_0} \right) S_1 \\
	\frac{1}{B_{0 \parallel}^{*}} Z e \nabla S_1 \times \mathbf{b}_0 \cdot (\mu \nabla B_0 + Z e \nabla \Phi_0) &\sim& \rho_* \omega_{\st{L}} \left( 1 + \frac{Z e \Phi_0}{\mu B_0} \right) S_1 \\
	\frac{\mathbf{B}_0^{*}}{B_{0 \parallel}^{*}} \cdot \nabla S_1 v_{\parallel} \sim \frac{v_{\parallel}}{\rho} \varepsilon_B S_1 &\sim& \rho_* \omega_{\st{L}} S_1 \\
	\frac{Z e}{2 m} \frac{\partial S_1}{\partial \theta} B_0 &\sim& \omega_{\st{L}} S_1 \\
	\frac{\partial S_1}{\partial t} &\sim& \rho_* \omega_{\st{L}} S_1
\end{eqnarray*}
Note that $\nabla_{\perp} S_1 \cdot \mathbf{B}_0 = 0$ and $\nabla_{\parallel} S_1 \times \mathbf{b}_0 = 0$. Let us assume that the order of $\frac{Z e \Phi_0}{\mu B_0}$ is no larger than $1$ ($\Phi_0$ is zero in a non-rotating plasma). Terms that are explicitly second order in $\rho_*$ have been neglected. Since $S_1$ is the first order gauge function, $\rho_* S_1$ gives rise to second order terms and therefore can be neglected, as well. The only term that remains is the one containing the $\theta$ derivative of $S_1$. Using the above orderings and applying the decomposition in equation \ref{eq:av_osc_decomp_phi} gives
\begin{eqnarray*}
	\bar{\Gamma}_{1,t} &=& - Z e \left( \widetilde{\Phi}_1(\mathbf{x}) + \langle \Phi_1(\mathbf{x})\rangle \right) - \frac{1}{B_{0 \parallel}^{*}} \widetilde{\vec{A}}_1(\vec{x}) \times \mathbf{b}_0(\mathbf{X}) \cdot \left( \mu \nabla B_0(\mathbf{X}) + Z e \nabla \Phi_0(\mathbf{X})\right) \\ 
	&& + \omega_{\st{L}} \frac{\partial S_1}{\partial \theta} + \frac{v_{\parallel}}{B_{0 \parallel}^{*}} Z e \widetilde{\vec{A}}_1(\vec{x}) \cdot \mathbf{B}_0^{*}(\mathbf{X}) \\
	&& + Z e \left( \langle \mathbf{A}_1(\mathbf{x}) \cdot \vec{v}_{\st{L}}(\mathbf{X},\mu,\theta) \rangle  + \left( \mathbf{A}_1(\mathbf{x}) \cdot \vec{v}_{\st{L}}(\mathbf{X},\mu,\theta) \right)^{\st{osc}} \right)
\end{eqnarray*}
where $\vec{v}_{\st{L}}(\mathbf{X},\mu,\theta) = v_{\st{L}}(\mathbf{X},\mu) \frac{\partial \mathbf{a}(\theta)}{\partial \theta}$. The word ''osc'' in superscript is the same as $\ \widetilde{}$, it denotes the oscillating part of the quantity between the brackets.

In order to remove the oscillating quantities from the Lagrangian, the term $\omega_{\st{L}} \frac{\partial S_1}{\partial \theta}$ has to cancel them all. This leads to the following equation for $S_1$:
\begin{eqnarray*}
	\frac{\partial S_1}{\partial \theta} &=& \frac{1}{\omega_{\st{L}}} \left( Z e \widetilde{\Phi}_1(\mathbf{x}) + \frac{1}{B_{0 \parallel}^{*}} \widetilde{\vec{A}}_1(\vec{x}) \times \mathbf{b}_0(\mathbf{X}) \cdot \left( \mu \nabla B_0(\mathbf{X}) + Z e \nabla \Phi_0(\mathbf{X})\right) - \right. \\
	&& \left. \frac{v_{\parallel}}{B_{0 \parallel}^{*}} Z e \widetilde{\vec{A}}_1(\vec{x}) \cdot \mathbf{B}_0^{*}(\mathbf{X}) - Z e \left( \mathbf{A}_1(\mathbf{x}) \cdot \vec{v}_{\st{L}}(\mathbf{X},\mu,\theta) \right)^{\st{osc}} \right).
\end{eqnarray*}
As a result, the total gyro-centre Lagrangian becomes
\begin{eqnarray}
	\bar{\Gamma} &=& \bar{\Gamma}_0 + \bar{\Gamma}_1 = \left( m v_{\parallel} \mathbf{b}_0(\mathbf{X}) + m \vec{u}_0 + Z e \mathbf{A}_0(\mathbf{X}) + Z e \langle \mathbf{A}_1(\mathbf{x}) \rangle \right) \cdot \mathrm{d} \mathbf{X} + \frac{2 \mu m}{Z e} \mathrm{d} \theta - \\
	&& \left( \frac{1}{2} m \left( v_{\parallel}^2 - u_0^2 \right) + \mu B_0(\mathbf{X}) + Z e \left( \Phi_0(\vec{X}) + \langle \Phi_1(\mathbf{x}) \rangle \right) - Z e \langle \mathbf{A}_1(\mathbf{x}) \cdot \vec{v}_{\st{L}}(\mathbf{X},\mu,\theta) \rangle \right) \mathrm{d}t. \nonumber
\end{eqnarray} 
The oscillating quantities have thus been systematically removed from the gyro-centre Lagrangian and added to the gauge function. The components of the gyro-centre Lagrangian are again independent of the gyro-phase, and therefore the magnetic moment in the new phase space remains invariant during the particles' motion.

%-----------------------------------------------------------------------
\subsubsection{Bessel functions}
Gyro-averaging of the fluctuations can be performed by in Fourier-space by separating the quantities' dependence on the gyro-centre position and the Larmor-radius vector:
\begin{eqnarray*}
 \langle \vec{A}_1(\vec{x}) \rangle &=& \langle \vec{A}_1 (\vec{X}+\vec{r}) \rangle = \langle \int \vec{\hat{A}}_1(\vec{k}) \st{e}^{i \vec{k} \cdot (\vec{X}+\vec{r})} \st{d}\vec{k} \rangle \\
 &=& \frac{1}{2 \pi} \int_0^{2 \pi} \int \vec{\hat{A}}_1(\vec{k}) \st{e}^{i \vec{k} \cdot \vec{X}} \st{e}^{i k_{\perp} \rho \cos \theta} \st{d} \vec{k} \st{d} \theta \\
 &=& \int \vec{\hat{A}}_1(\vec{k}) \st{e}^{i \vec{k} \cdot \vec{X}} \underbrace{\frac{1}{2 \pi} \int_0^{2 \pi} \st{e}^{i k_{\perp} \rho \cos \theta} \st{d} \theta}_{J_0(\rho k_{\perp})} \st{d} \vec{k} \\
 &=& \int J_0(\rho k_{\perp}) \vec{\hat{A}}_1(\vec{k}) \st{e}^{i \vec{k} \cdot \vec{X}} \st{d} \vec{k} = J_0(\lambda) \vec{A}_1(\vec{X})
\end{eqnarray*}
where the vector $\vec{r}$ is used in the sense of equation \ref{eq:X}, and the direction of the basis vector $\vec{e}_1$ has been aligned with the wavenumber vector so that $\vec{k}=\vec{e}_1 k_{\perp}$ leading to $\vec{k} \cdot \vec{r} = \rho k_{\perp} \cos \theta$. $J_0$ is a zeroth order Bessel function of the first kind as defined in \cite{a_and_s}. Its argument becomes $\lambda = i \rho \nabla_{\perp}$ during the inverse Fourier transformation.

Gyro-averaging of the $\langle \vec{A}_1 \cdot \vec{v}_{\st{L}} \rangle$ can be performed in a similar fashion, although through some lengthy algebra. The detailed calculation can be found in Appendix C of \cite{dannert}. The gyro-averaging process eventually gives
\begin{eqnarray*}
	\langle \Phi_1(\mathbf{x})\rangle &=& J_0(\lambda) \Phi_1(\mathbf{X}) \\
	\langle \mathbf{A}_1(\mathbf{x})\rangle &=& J_0(\lambda) \mathbf{A}_1(\mathbf{X}) \\
	Z e \langle \mathbf{A}_1(\mathbf{x}) \cdot \vec{v}_{\st{L}}(\mathbf{X},\mu,\theta) \rangle &=& - \hat{J}_1(\lambda) \mu B_{1 \parallel}(\mathbf{X})
\end{eqnarray*}
where $\hat{J}_1(z) = \frac{2}{z} J_1(z)$ is a modified first order Bessel function of the first kind.
Substituting the above expressions into the gyro-centre Lagrangian gives
\begin{eqnarray}
	\bar{\Gamma} &=& \left( m v_{\parallel} \mathbf{b}_0(\mathbf{X}) + Z e \mathbf{A}_0(\mathbf{X}) + m \vec{u}_0 + Z e J_0(\lambda) \mathbf{A}_1(\mathbf{X}) \right) \cdot \mathrm{d} \mathbf{X} + \frac{2 \mu m}{Z e} \mathrm{d} \theta - \nonumber\\
	&& \left( \frac{1}{2} m \left( v_{\parallel}^2 - u_0^2 \right) + Z e \left( \Phi_0(\vec{X}) + J_0(\lambda) \Phi_1(\mathbf{X}) \right) + \mu \left( B_0(\mathbf{X}) + \hat{J}_1(\lambda) B_{1 \parallel}(\mathbf{X}) \right) \right) \mathrm{d}t \nonumber \\
	&=& \left( Z e \mathbf{A}_0^{*}(\mathbf{X}) + Z e \bar{\mathbf{A}}_1(\mathbf{X}) \right) \cdot \mathrm{d} \mathbf{X} + \frac{2 \mu m}{Z e} \mathrm{d} \theta - \nonumber\\
	&& \left( \frac{1}{2} m \left( v_{\parallel}^2 - u_0^2 \right) + Z e \left( \Phi_0(\vec{X}) + \bar{\Phi}_1(\mathbf{X}) \right) + \mu \left( B_0(\mathbf{X}) + \bar{B}_{1 \parallel}(\mathbf{X}) \right) \right) \mathrm{d}t
	\label{eq:gyro_oneform}
\end{eqnarray} 
where we used $\mathbf{A}_0^{*} = \mathbf{A}_0 + \frac{m }{Z e} (v_{\parallel} \mathbf{b}_0 + \vec{u}_0 )$ again and introduced the shorter notations $J_0(\lambda) \Phi_1 = \bar{\Phi}_1$, $J_0(\lambda) \mathbf{A}_1 = \bar{\mathbf{A}}_1$ and $\hat{J}_1(\lambda) B_{1 \parallel} = \bar{B}_{1 \parallel}$.

%-----------------------------------------------------------------------
\subsection{Gyrokinetic Vlasov-equation} \label{sec:vlasov}
The time evolution of the distribution function in the phase space is described by the Vlasov equation. In particle phase space without collisions it can be written as
\[ \frac{\partial f}{\partial t} + \dot{\mathbf{x}} \cdot \frac{\partial f}{\partial \mathbf{x}} + \dot{\mathbf{v}} \cdot \frac{\partial f}{\partial \mathbf{v}} = 0.\]
Since in gyro-centre phase space the gyro-phase is an ignorable coordinate and the total time derivative of the gyro-centre magnetic moment is zero (see equation (\ref{eq:gyro_eqs_motion})), the Vlasov equation takes the form
\begin{equation}
	\frac{\partial f_{\st{gy}}}{\partial t} + \dot{\mathbf{X}} \cdot \frac{\partial f_{\st{gy}}}{\partial \mathbf{X}} + \dot{v}_{\parallel} \frac{\partial f_{\st{gy}}}{\partial v_{\parallel}} = 0
	\label{eq:gyro_vlasov}
\end{equation}
where $f_{\st{gy}}$ is the distribution function of they gyro-centres instead of the particles. In the remainder of this section the ''gy'' underscript will be dropped for simplicity and $f$ will denote the gyro-centre distribution function unless otherwise stated.

Two important modifications have to be performed on equation \ref{eq:gyro_vlasov}: first, the terms $\dot{\mathbf{X}}$ and $\dot{v}_{\parallel}$ have to be expressed from the gyro-centre Lagrangian through of the Euler--Lagrange equations, and second, the total distribution function $f$ will be decomposed into a sum of an equilibrium and a perturbation part: $f = F + \delta f$ where $\frac{\delta f}{F} \sim \varepsilon_{\delta}$. The latter step is the so called delta-f approximation.

%-----------------------------------------------------------------------
\subsubsection{The Euler--Lagrange equations}
According to Scott \cite{scott_gotit} the Euler--Lagrange equations can be written as
\begin{equation}
	\left( \frac{\partial \gamma_j}{\partial z^i} - \frac{\partial \gamma_i}{\partial z^j} \right) \frac{\mathrm{d} z^j}{\mathrm{d} t} = \frac{\partial H}{\partial z^i} + \frac{\partial \gamma_i}{\partial t}.
	\label{eq:euler-lagrange}
\end{equation}
After substituting equation \ref{eq:gyro_oneform} the equations of motion are directly obtained as
\begin{eqnarray}
 	&& m \dot{v}_{\parallel} \mathbf{b}_0 - Z e \dot{\mathbf{X}} \times \mathbf{B}_0^{*} - Z e \dot{\mathbf{X}} \times \left( \nabla \times \bar{\mathbf{A}}_1 \right) =  \label{eq:gyro_eqs_motion} \\
	&& \hspace{2cm} \underbrace{- Z e \nabla \left( \Phi_0 + \bar{\Phi}_1 \right) - Z e \frac{\mathrm{d}}{\mathrm{d}t} \bar{\mathbf{A}}_1}_{Z e \bar{\mathbf{E}}} - \mu \nabla \left( B_0 + \bar{B}_{1 \parallel}\right) + \frac{1}{2}m \nabla u_0^2 \nonumber \\
	&& v_{\parallel} = \mathbf{b}_0 \cdot \dot{\mathbf{X}} \qquad \dot{\mu} = 0 \qquad \dot{\theta} = \omega_{\st{L}} - \frac{Z e}{m} \frac{\partial}{\partial \mu} \left( Z e \bar{\mathbf{A}}_1 \cdot \dot{\mathbf{X}} - Z e \bar{\Phi}_1 - \mu \bar{B}_{1 \parallel} \right) \nonumber
\end{eqnarray}
where the relation $\left(\nabla \mathbf{A} - (\nabla \mathbf{A})^T \right) \cdot \dot{\mathbf{X}} = \dot{\mathbf{X}} \times (\nabla \times \mathbf{A})$ has been applied. Using equation \ref{eq:frame_frequency} the last term in the first equation of \ref{eq:gyro_eqs_motion} can be rewritten as $\frac{1}{2}m \nabla u_0^2 = m R \Omega^2 \nabla R$.

By introducing the notation $\nabla \times \bar{\mathbf{A}}_1 = \bar{\mathbf{B}}_1$ and taking the cross product with $\mathbf{b}_0$ the time evolution of the gyro-centre position $\dot{\mathbf{X}}$ can be expressed as
\begin{eqnarray}
	\dot{\mathbf{X}} &=& \mathbf{b}_0 v_{\parallel} + \frac{m v_{\parallel}^2}{Z e B_0} \left( \nabla \times \mathbf{b}_0 \right)_{\perp} + \frac{\bar{\mathbf{B}}_{1 \perp}}{B_0} v_{\parallel} - \frac{1}{B_0} \bar{\mathbf{E}} \times \mathbf{b}_0 + \nonumber \\
	&& \frac{\mu}{Z e B_0} \nabla \left( B_0 + \bar{B}_{1 \parallel} \right) \times \mathbf{b}_0 + \frac{2 m v_{\parallel}}{Z e B_0} \vec{\Omega}_{\perp} - \frac{m R \Omega^2}{Z e B_0} \nabla R \times \vec{b}_0 \nonumber \\
	&=& \mathbf{b}_0 v_{\parallel} + \underbrace{\mathbf{v}_{\bar{\vec{B}}_{1 \perp}} + \mathbf{v}_{\bar{\vec{E}} \times \vec{B}_0} + \mathbf{v}_{\nabla \bar{B}_{1\parallel}}}_{\vec{v}_{\chi}} + \underbrace{\mathbf{v}_{C} + \mathbf{v}_{\nabla \vec{B}_0} + \vec{v}_{\st{co}} + \vec{v}_{\st{cf}}}_{\mathbf{v}_{\st{D}}}.
	\label{eq:gyro_Xdot}
\end{eqnarray}
To obtain equation \ref{eq:gyro_Xdot} the Taylor expansion
\[\frac{1}{B_{0 \parallel}^{*} + \bar{B}_{1 \parallel}} = \frac{1}{B_0} \left( 1 - \frac{m}{Z e B_0} \mathbf{b}_0 \cdot \nabla \times ( v_{\parallel} \mathbf{b}_0 + \vec{u}_0) - \frac{\bar{B}_{1 \parallel}}{B_0}  + O(\rho^2) \right)\]
has been used and the terms were kept up to first order in $\rho_*$.

The subsequent terms in the first two lines of equation \ref{eq:gyro_Xdot} denote streaming along the equilibrium magnetic field ($\mathbf{b}_0 v_{\parallel}$), curvature drift ($\mathbf{v}_C$), streaming along the perpendicular perturbed magnetic field ($\mathbf{v}_{\bar{\vec{B}}_{1\perp}}$), $E \times B$ drift in the total electric field ($\mathbf{v}_{\bar{\vec{E}} \times \vec{B}_0}$), and grad-$B$ drifts in the gradients of the equilibrium as well as the parallel perturbed magnetic fields ($\mathbf{v}_{\nabla \bar{B}_{1 \parallel}}$ and $\mathbf{v}_{\nabla \vec{B}_0}$), the Coriolis ($\vec{v}_{\st{co}}$) and centrifugal ($\vec{v}_{\st{cf}}$) drifts, respectively.

Note that the motion along the parallel component of the perturbed magnetic field is missing in equation \ref{eq:gyro_Xdot}, it would only appear in higher orders. The perpendicular component of the perturbed vector potential $\bar{\mathbf{A}}_{1 \perp}$ is related to the parallel perturbation of the magnetic field and it has been kept in $\bar{\mathbf{E}}$. However, according to typical normalization assumptions in gyrokinetics (see for example \cite{gkw}) one can easily show that the time derivative of the vector potential is one order smaller than the gradient of the electrostatic potential and therefore its contribution can be neglected in the $\mathbf{v}_{\bar{\vec{E}} \times \vec{B}_0}$ velocity.

After dropping the vector potential term from $\mathbf{v}_{\bar{\vec{E}} \times \vec{B}_0}$, the velocities due to the perturbation of the fields can be written in a more compact form. Let us define the quantity $\chi$ as
\[\chi = \underbrace{\Phi_0 + \bar{\Phi}_1}_{\bar{\Phi}} - v_{\parallel} \bar{A}_{1 \parallel} + \frac{\mu}{Z e} \bar{B}_{1 \parallel} \]
with which one can write
\[\mathbf{v}_{\chi} = \frac{\mathbf{b} \times \nabla \chi}{B_0} = \mathbf{v}_{\bar{\vec{B}}_{1\perp}} + \mathbf{v}_{\bar{\vec{E}} \times \vec{B}_0} + \mathbf{v}_{\nabla \bar{B}_{1 \parallel}}.\]

The equation for $\dot{v}_{\parallel}$ is obtained by taking the scalar product of the first equation of equation \ref{eq:gyro_eqs_motion} with $\dot{\mathbf{X}}$:
\begin{equation}
	\dot{v}_{\parallel} = \frac{\dot{\mathbf{X}}}{m v_{\parallel}} \cdot \left( Z e \bar{\mathbf{E}} - \mu \nabla (B_0 + \bar{B}_{1 \parallel}) + \frac{1}{2}m \nabla u_0^2 \right).
\label{eq:gyro_vdot}
\end{equation}
By substituting the obtained drift velocities it can be shown that the acceleration of the $\bar{\vec{E}} \times \vec{B}_0$ velocity by the mirror force of the perturbed magnetic field $\mu \nabla \bar{B}_{1 \parallel}$ and the acceleration of the $\nabla \bar{B}_{1 \parallel} \times \vec{B}_0$ velocity by the electric field cancel each other. Noting that $\nabla \bar{\Phi} \perp \nabla \bar{\Phi} \times \vec{b_0}$ and $\nabla \bar{B}_{1 \parallel} \perp \nabla \bar{B}_{1 \parallel} \times \vec{b_0}$, the equation of motion for the parallel velocity becomes
\begin{equation}
	\dot{v}_{\parallel} = \frac{\dot{\mathbf{X}}}{m v_{\parallel}} \cdot \left( - \mu \nabla B_0 + \frac{1}{2}m \nabla u_0^2 \right) + \frac{\vec{v}_{D} +  \vec{v}_{\vec{B}_{1 \perp}}}{m v_{\parallel}} \cdot (Z e \bar{\vec{E}} - \mu \nabla \bar{B}_{1 \parallel} ).
\end{equation}

%-----------------------------------------------------------------------
\subsubsection{The delta-f approximation}
Substituting equations \ref{eq:gyro_vdot} and \ref{eq:gyro_Xdot} into the gyro-centre Vlasov equation \ref{eq:gyro_vlasov} and applying the delta-f approximation $f = F + \delta f$ leads to
\begin{equation}
	\frac{\partial \delta f}{\partial t} + \dot{\mathbf{X}} \cdot \nabla \delta f - \frac{\vec{b}_0}{m} \cdot ( \nabla \Phi_0 - m R \Omega^2 \nabla R + \nabla B_0 ) \frac{\partial \delta f}{\partial v_{\parallel}} = \underbrace{- \dot{\mathbf{X}} \cdot \nabla F - \dot{v}_\parallel \frac{\partial F}{\partial v_{\parallel}}}_{S}.
	\label{eq:vlasov_deltaf}
\end{equation}
Note that in equation \ref{eq:vlasov_deltaf} only the highest order term of $\dot{v}_{\parallel}$ has to be kept in $\dot{v}_{\parallel} \frac{\partial \delta f}{\partial v_{\parallel}}$, the contributions of $\bar{B}_{1 \parallel}$ and $\bar{\Phi}_1$ can be neglected.

The equilibrium distribution function is assumed to be a Maxwellian. In a toroidally rotating plasma, a finite radial equilibrium electric field arises in order to balance the centrifugal force (in the co-rotating frame). A combined energy term associated with the kinetic energy of the rotation and the energy stored in the equilibrium electric field has to be included in the Maxwellian. According to \cite{peeters_rotation}, in the rotating frame this term can be written as 
\begin{equation*}
 \mathcal{E} = Z e \langle \Phi_0 \rangle - \frac{1}{2} m \omega_{\varphi}^2 (R^2-R_0^2)
\end{equation*}
where the angled brackets denote flux-surface averaging, $\omega_{\varphi}$ is the plasma rotation frequency profile in the rotating frame as introduced in section \ref{sec:rotation}, $R$ is the local major radius and $R_0$ is an integration constant which can be chosen, for example, as the major radius of the plasma or the flux surface average of the major radius \cite{Casson2010}. The Maxwellian is written as
\begin{equation}
	F = F_{\st{M}} = \frac{n_0}{ (2 \pi T / m)^{\frac{3}{2}} } \exp\left( - \frac{\frac{1}{2} m (v_{\parallel} - u_{\parallel})^2 + \mu B_0 + \mathcal{E}}{T} \right) = F_{\st{M}}(\vec{X},v_{\parallel},\mu)
	\label{eq:maxwellian}
\end{equation}
where $n_0$ is the equilibrium particle density and $u_{\parallel}$ is rotation speed of the plasma in the rotating frame parallel to the magnetic field. In a local description, a reference frame rotating with $u_{\parallel}$ can be chosen in which case $u_{\parallel}$ vanishes but its gradient has to be taken into account. According to \cite{peeters_rotation} the derivatives of the Maxwellian can be expressed as
\begin{eqnarray}
	\nabla F_{\st{M}} &=& \left[ \frac{\nabla n_0}{n_0} + \left( \frac{\frac{1}{2} m v_{\parallel}^2 + \mu B_0 + \mathcal{E}}{T} - \frac{3}{2} \right) \frac{\nabla T}{T} - \frac{\mu B_0}{T} \frac{\nabla B_0}{B_0} + \right. \nonumber \\ 
	&& \left. \left( \frac{m v_{\parallel} R B_{\st{t}}}{B T} + m \Omega (R^2-R_0^2) \right) \nabla \omega_{\varphi} \right] F_{\st{M}} \nonumber \\
	\frac{\partial F_{\st{M}}}{\partial v_{\parallel}} &=& - \frac{m v_{\parallel}}{T} F_{\st{M}} \qquad \qquad \frac{\partial F_{\st{M}}}{\partial \mu} = - \frac{B_0}{T} F_{\st{M}} \label{eq:grad_F_M}
\end{eqnarray}
where the $\nabla \omega_{\varphi}$ terms in the gradient are coming from the derivatives of the $u_{\parallel}$ and $\mathcal{E}$ terms, respectively, and evaluated at zero rotation speed locally in the co-rotating frame. Using equations \ref{eq:gyro_Xdot}, \ref{eq:gyro_vdot} and \ref{eq:grad_F_M} it can be shown that the $\nabla B_0$ term in $- \dot{\vec{X}} \cdot \nabla F_{\st{M}}$ cancels with $\frac{\dot{\vec{X}}}{m v_{\parallel}} \mu \nabla B_0 \frac{\partial F_{\st{M}}}{\partial v_{\parallel}}$. In case of purely toroidal rotation $\nabla R \perp \vec{\Omega}$, and $\nabla R \perp \nabla R \times \vec{b}_0$. Using these expression the source term can be finally written as
\begin{eqnarray*} 
	S &=& - \left( \mathbf{v}_{\chi} + \mathbf{v}_{\st{D}} \right) \cdot \nabla_p F_{\st{M}} - \frac{Z e v_{\parallel}}{T} \frac{\partial \bar{A}_{1 \parallel}}{\partial t} F_{\st{M}} \\
	&& - \frac{F_{\st{M}}}{T} \left( v_{\parallel} \mathbf{b}_0 + \mathbf{v}_{D} + \mathbf{v}_{\vec{B}_{1 \perp}} \right) \cdot \left( Z e \nabla \bar{\Phi} + \mu \nabla \bar{B}_{1 \parallel} \right)
\end{eqnarray*}
where $\nabla_p$ means that only the $\nabla n$, $\nabla T$ and $\nabla \omega_{\varphi}$ terms of $\nabla F_{\st{M}}$ are considered \cite{gkw}.

The term containing the time derivative of the perturbed vector potential is computationally difficult to handle. Therefore a modified distribution function $g$ is defined as
\begin{equation}
	g = \delta f + \frac{Z e v_{\parallel}}{T} \bar{A}_{1 \parallel} F_{\st{M}}.
	\label{eq:g}
\end{equation}
Note that $g$ is not the non-adiabatic part of the distribution function, as it typically appears in the literature. 
The perturbed part of the distribution function $\delta f$ is replaced with $g$ in the first term of the left hand side of the Vlasov-equation \ref{eq:vlasov_deltaf}, and also in the second term where it is multiplied by $\vec{v}_{\chi}$. In the remaining part of the second term, where it is multiplied by $\vec{b}_0 v_{\parallel} + \vec{v}_{\st{D}}$ the original distribution function is kept. It can be easily shown that as a result of this substitutions the $\vec{v}_{\vec{B}_{1 \perp}}$ term in the source cancels out and the equation simplifies to
\begin{equation}
	\frac{\partial g}{\partial t} + \vec{v}_{\chi} \cdot \nabla g + \left( v_{\parallel} \mathbf{b}_0 + \mathbf{v}_{\st{D}} \right) \cdot \nabla \delta f - \frac{\vec{b}_0}{m} \cdot \left(\nabla \Phi_0 - m R \Omega^2 \nabla R + \nabla B_0 \right) \frac{\partial \delta f}{\partial v_{\parallel}} = S 
	\label{eq:vlasov}
\end{equation}
with the source term being
\begin{equation}
	S = - \left( \mathbf{v}_{\chi} + \mathbf{v}_{\st{D}} \right) \cdot \nabla_p F_{\st{M}} + \frac{F_{\st{M}}}{T} \left( v_{\parallel} \mathbf{b}_0 + \mathbf{v}_{D} \right) \cdot \left( - Z e \nabla \bar{\Phi}_1 - \mu \nabla \bar{B}_{1 \parallel} \right).
	\label{eq:vlasov_source}
\end{equation}
This is the form of the Vlasov-equation currently implemented in the code GKW. The normalized version of its terms is found in \cite{gkw}. The effect of the magnetic compression on the Vlasov-equation is that a new velocity component, $\mathbf{v}_{\nabla \bar{B}_{1 \parallel}}$ describing the grad-B drift in the parallel perturbation of the magnetic field, appears alongside with an associated mirror force in the source term.

%-----------------------------------------------------------------------
\subsection{Maxwell's Equations in Gyro-centre Coordinates} \label{sec:gk_maxwell}
In order to obtain a self-consistent system of equations, the perturbed electro-magnetic fields must be calculated with Maxwell's equations. In typical fusion plasmas the Gauss law is replaced by the quasi-neutrality condition. This expresses that any deviation from neutrality can only happen at small length scales (within the Debye-length) and on a time scale much shorter than that of the fluctuations. The displacement current in Amp\`ere's law is also neglected due to the non-relativistic time scale of the turbulence. With these assumptions the Maxwell-equations can be written as
\begin{eqnarray*}
	\sum_{\st{sp}} Z_{\st{sp}} e n_{\st{sp}} = 0 &\qquad& \nabla \times \mathbf{E}_1 = - \frac{\partial \mathbf{B}_1}{\partial t} \\
	\nabla \cdot \mathbf{B}_1 = 0 &\qquad& \nabla \times \mathbf{B}_1 = \mu_0 \sum_{\st{sp}} \mathbf{j}_{\st{sp}}.
\end{eqnarray*}

These equations contain densities and currents of particles that can be expressed by taking the moments of the particle phase space distribution function. The Vlasov equation, however, describes the evolution of the distribution function in the gyro-centre phase space. The connection between them is recovered by expressing the particle moments with the guiding-centre distribution function and by pulling back the gyro-centre distribution function to the guiding-centre phase space. That is, Maxwell's equations in gyrokinetic theory are written in the middle, in the guiding-centre phase space (see figure \ref{fig:maxwell-transform}).

\begin{figure}
 \begin{center}
	\includegraphics[scale=0.5]{maxwell-transform.eps}
 \end{center}
 \caption{Idea of gyrokinetic Maxwell's equations: The density and current of particles as well as the gyro-centre distribution function are expressed in guiding centre phase space.}
 \label{fig:maxwell-transform}
\end{figure}

As a first step, the particle densities and currents of one species are expressed with the guiding-centre distribution function:
\begin{eqnarray}
	n(\mathbf{x}) &=& \int f(\mathbf{x},\mathbf{v}) \mathrm{d} \mathbf{v} = \frac{B_0}{m} \int \delta(\mathbf{X}+\mathbf{r}-\mathbf{x}) f_{\st{gc}} \mathrm{d}\mathbf{X} \mathrm{d}v_{\parallel} \mathrm{d} \theta \mathrm{d}\mu \label{eq:gc_n} \\
	j_{\parallel} &=& Z e \int v_{\parallel} f(\mathbf{x},\mathbf{v}) \mathrm{d} \mathbf{v}  = \frac{Z e B_0}{m} \int \delta(\mathbf{X}+\mathbf{r}-\mathbf{x}) v_{\parallel} f_{\st{gc}} \mathrm{d}\mathbf{X} \mathrm{d}v_{\parallel} \mathrm{d} \theta \mathrm{d}\mu \label{eq:gc_j_par}\\
	\mathbf{j}_{\perp} &=& Z e \int \vec{v}_{\st{L}} f(\mathbf{x},\mathbf{v}) \mathrm{d} \mathbf{v} = \frac{Z e B_0}{m} \int \delta(\mathbf{X}+\mathbf{r}-\mathbf{x}) \vec{v}_{\st{L}} f_{\st{gc}} \mathrm{d}\mathbf{X} \mathrm{d}v_{\parallel} \mathrm{d} \theta \mathrm{d}\mu \label{eq:gc_j_perp}
\end{eqnarray}
where $\vec{r}$ is the vector pointing from the centre of the Larmor orbit to the particle's position. The $\frac{B_0}{m}$ factor is the Jacobian. Formally, the delta function appears since the change of coordinates affects the whole phase space while the integral goes only over the velocity subspace. It guarantees that the spatial region taken into account in the integral remains unchanged during the coordinate transformation. Physically it expresses that all the particles that have a Larmor orbit crossing a given point $\vec{x}$ in real space contribute to the particle density there (see figure \ref{fig:density-gc}).

\begin{figure}
 \begin{center}
	\includegraphics[scale=0.3]{density-gc.eps}
 \end{center}
 \caption{The connection between the density of particles and density of guiding-centres.}
 \label{fig:density-gc}
\end{figure}

The second step is to express the guiding-centre distribution function with the gyro-centre one and write it into equations \ref{eq:gc_n}-\ref{eq:gc_j_perp}. Since the distribution function is a regular function (or zero-form) on the manifold, it is transformed by the pull-back operator:
\begin{equation}
	f_{\st{gc}} = \Phi^* f_{\st{gy}} = \mathrm{e}^{\varepsilon \mathcal{L}_{G_1}} f_{\st{gy}} \approx \left( 1 + \varepsilon \mathcal{L}_{G_1} \right) f_{\st{gy}} = f_{\st{gy}} + G_1^{\nu} \frac{\partial f_{\st{gy}}}{\partial Z^{\nu}}. \label{eq:pullback}
\end{equation}
Substituting the components of the generator vector field derived in section \ref{sec:gyro_derive}, the expressions that connect the gyro-centre and guiding-centre distribution function are directly obtained. The delta-f approximation is applied once again and the equilibrium part of the gyro-centre distribution function $f_{\st{gy}}$ is assumed Maxwellian (equation \ref{eq:maxwellian}). Since the generator functions are already first order in $\rho_*$, in the second part of equation \ref{eq:pullback} only the equilibrium distribution function $F_{\st{M}}$ must be taken into account. Also, as $\nabla F_{\st{M}}$ only contains first order terms in $\rho_*$ the term $G_1^{\mathbf{X}} \cdot \nabla F_{\st{M}}$ can be dropped. The resulting transformation formula is
\begin{eqnarray}
 	G_1^{\mu} \frac{\partial F_{\st{M}}}{\partial \mu} &=& \frac{1}{B_0} \left( Z e \widetilde{\Phi}_1 - Z e v_{\parallel} \widetilde{A}_{1 \parallel} - Z e \left( \widetilde{\mathbf{A}_1 \cdot \vec{v}_{\st{L}}} \right) + Z e v_{\st{L}} \left| \mathbf{A}_{1 \perp}\right| \right) \frac{\partial F_{\st{M}}}{\partial \mu} \nonumber \\
	&=& \frac{1}{B_0} \left( Z e \widetilde{\Phi}_1 - Z e v_{\parallel} \widetilde{A}_{1 \parallel} - Z e \langle \mathbf{A}_1 \cdot \vec{v}_{\st{L}} \rangle \right) \frac{\partial F_{\st{M}}}{\partial \mu} \nonumber \\
	&=& -\frac{1}{T} \left( Z e \widetilde{\Phi}_1 - Z e v_{\parallel} \widetilde{A}_{1 \parallel} + \mu \bar{B}_{1 \parallel} \right) F_{\st{M}} \nonumber \\
	\noalign{\vskip 0.5 truecm}
	G_1^{v_{\parallel}} \frac{\partial F_{\st{M}}}{\partial v_{\parallel}} &=& - Z e \frac{v_{\parallel}}{T} \widetilde{A}_{1 \parallel} F_{\st{M}} \nonumber \\
	\noalign{\vskip 0.5 truecm}
	f_{\st{gc}} &=& f_{\st{gy}} + G_1^{\mu} \frac{\partial F_{\st{M}}}{\partial \mu} + G_1^{v_{\parallel}} \frac{\partial F_{\st{M}}}{\partial v_{\parallel}} = f_{\st{gy}} - \frac{F_{\st{M}}}{T} \left( Z e \widetilde{\Phi}_1 - \mu \bar{B}_{1 \parallel} \right).
	\label{eq:f_gy_to_gc}
\end{eqnarray}
The two correction terms appearing in equation \ref{eq:f_gy_to_gc} containing the fluctuations of the electro-magnetic fields are the results of the pull-back transformation from the gyro-centre to the guiding centre phase space. Physically, they describe the polarization and magnetization effects of the fluctuations on the gyro-orbit \cite{brizard}.

%-----------------------------------------------------------------------
\subsubsection{Normalization} \label{sec:normalization}
The field equations derived in the following sections will be normalized as detailed in \cite{gkw}. The definition of the normalized quantities is reiterated here for the sake of completeness. A reference mass $m_{\st{ref}}$, density $n_{\st{ref}}$, temperature $T_{\st{ref}}$, magnetic field $B_{\st{ref}}$ and major radius $R_{\st{ref}}$ is chosen. These are then used to define the reference thermal velocity $T_{\st{ref}} = \frac{1}{2} m_{\st{ref}} v_{\st{th,ref}}^2$ and reference thermal Larmor radius $\rho_{\st{ref}} = \frac{m_{\st{ref}} v_{\st{th,ref}}}{e B_{\st{ref}}}$. The normalized major radius and equilibrium magnetic field are simply $R = R_{\st{ref}} R_{st{N}}$ and $B = B_{\st{ref}} B_{\st{N}}$. For convenience, the small parameter $\rho_*$ is redefined with the reference values as $\rho_* = \frac{\rho_{\st{ref}}}{R_{\st{ref}}}$. The reference quantities are used to define the dimensionless relative quantities
\begin{equation}
m_{\st{R}} = \frac{m}{m_{\st{ref}}} \qquad n_{\st{R}} = \frac{n}{n_{\st{ref}}} \qquad T_{\st{R,sp}} = \frac{T}{T_{\st{ref}}} \qquad v_{\st{R}} = \frac{v_{\st{th}}}{v_{\st{th,ref}}}.
\end{equation}
The fluctuating fields are normalized as
\begin{equation}
\Phi_1 = \rho_* \frac{T_{\st{ref}}}{e} \Phi_{1 \st{N}} \qquad A_{1 \parallel} = B_{\st{ref}} R_{\st{ref}} \rho_*^2 A_{1 \parallel \st{N}} \qquad B_{1 \parallel} = \rho_* B_{\st{ref}} B_{1 \parallel \st{N}}.
\end{equation}
The time and the angular rotation frequency are normalized using the reference thermal velocity as
\begin{equation}
t = \frac{R_{\st{ref}}}{v_{\st{th,ref}}} t_{\st{N}} \qquad \Omega = \frac{v_{\st{th,ref}}}{R_{\st{ref}}} \Omega_{\st{N}}
\end{equation}
whereas the velocity space coordinates and the distribution functions are normalized with the thermal velocity to maintain their species dependence:
\begin{equation}
v_{\parallel} = v_{\st{th}} v_{\parallel \st{N}} \qquad \mu = \frac{m v_{\st{th}}^2}{e B_{\st{ref}}} \qquad \delta f = \rho_* \frac{n}{v_{\st{ref}}^3} \delta f_{\st{N}} \qquad F_{\st{M}} = \frac{n}{v_{\st{ref}}^3} F_{\st{M,N}}.
\end{equation}
Finally, the parallel and perpendicular gradients of the equilibrium and the wavenumber of the fluctuations are normalized, respectively, as
\begin{equation}
\nabla_{\perp} = \frac{1}{R_{\st{ref}}} \nabla_{\perp \st{N}} \qquad \nabla_{\parallel} = \frac{1}{R_{\st{ref}}} \nabla_{\parallel \st{N}} \qquad k = \frac{k_{\st{N}}}{\rho_*}.
\end{equation}

%-----------------------------------------------------------------------
\subsubsection{Gyrokinetic Poisson equation}
The derivation of the gyrokinetic Maxwell-equations is now straightforward: the distribution function transformed from gyro-centre to guiding-centre phase space by equation \ref{eq:f_gy_to_gc} has to be substituted to the guiding-centre particle density and current equations \ref{eq:gc_n}-\ref{eq:gc_j_perp}. Equation \ref{eq:gc_n} takes the form
\begin{eqnarray*}
	n(\vec{x}) &=& \frac{B_0}{m} \int \delta(\vec{X}+\vec{r}-\vec{x}) \left( f_{\st{gy}}(\vec{X}) - \frac{F_{\st{M}}}{T} \left( Z e \widetilde{\Phi}_1(\vec{X}+\vec{r}) - \bar{B}_{1 \parallel}(\vec{X}) \right)\right) \mathrm{d}\mathbf{X} \mathrm{d}v_{\parallel} \mathrm{d} \theta \mathrm{d}\mu \\
	&=& \bar{n}(\vec{x}) - \frac{B_0}{m} \int \delta(\vec{X}+\vec{r}-\vec{x}) \left( Z e \widetilde{\Phi}_1(\vec{X}+\vec{r}) - \mu \bar{B}_{1 \parallel}(\vec{X}) \right) \frac{F_{\st{M}}}{T} \mathrm{d}\mathbf{X} \mathrm{d}v_{\parallel} \mathrm{d} \theta \mathrm{d}\mu
\end{eqnarray*}
where the density of gyro-centres $\bar{n}$ has been introduced. Note that the distribution functions $f_{\st{gy}}$ depends on $\vec{X}$ as well as the velocity coordinates $v_{\parallel}$, $\mu$ and $\theta$. Here, however, only the spatial dependence is indicated for the gyroaveraging. Writing the oscillating part of the perturbed potential as $\widetilde{\Phi}_1(\vec{X}+\vec{r}) = \Phi_1(\vec{X}+\vec{r}) - \bar{\Phi}_1(\vec{X})$ and integrating the $\Phi_1(\vec{X}+\vec{r})$ term yields
\begin{eqnarray*}
	n(\vec{x}) &=& \bar{n}(\vec{x}) - \frac{ Z e \Phi_1(\vec{x})}{T} n_0(\vec{x}) \\
	&& + \frac{B_0}{m} \int \delta(\vec{X}+\vec{r}-\vec{x}) \left( Z e \bar{\Phi}_1(\vec{X}) + \mu \bar{B}_{1 \parallel}(\vec{X}) \right) \frac{F_{\st{M}}}{T} \mathrm{d}\mathbf{X} \mathrm{d}v_{\parallel} \mathrm{d} \theta \mathrm{d}\mu
\end{eqnarray*}
where $n_0$ is the equilibrium particle density. The factor $\frac{F_{\st{M}}}{T}$ changes on the equilibrium length scale. Therefore, its weak spatial dependence has been neglected in the previous step compared to the fast variation of the fluctuating $\Phi_1$. 
The integral over $\vec{X}$ together with the delta function changes the arguments of the fields to $\vec{x}-\vec{r}$. After integrating over $\theta$ this results
\begin{eqnarray*}
	n(\vec{x}) &=& \bar{n}(\vec{x}) - \frac{Z e \Phi_1(\vec{x})}{T} n_0(\vec{x}) \\
	&& + \frac{2 \pi B_0}{m} \int J_0(\lambda) \left( Z e \bar{\Phi}_1(\vec{x}) + \mu \bar{B}_{1 \parallel}(\vec{x}) \right) \frac{F_{\st{M}}}{T} \mathrm{d}v_{\parallel} \mathrm{d}\mu
\end{eqnarray*}
where $\lambda^2 = - \rho^2 \nabla_{\perp}^2$. Performing the $v_{\parallel}$ integral is straightforward as it is only the Maxwellian that depends on it: 
\begin{eqnarray*}
	F_{\st{M}} &=& F_{\st{M}}(\mu) \mathrm{e}^{-\frac{m (v_{\parallel} - u_{\parallel})^2}{2 T}} \st{e}^{-\frac{\mathcal{E}}{T}} \\
	n(\vec{x}) &=& \bar{n}(\vec{x}) - \frac{Z e \Phi_1(\vec{x})}{T} n_0(\vec{x}) \\
	&& + \frac{B_0}{\sqrt{T}} \left( \frac{2 \pi}{m} \right)^{\frac{3}{2}} \st{e}^{\frac{-\mathcal{E}}{T}} \int F_{\st{M}}(\mu) J_0(\lambda) \left( Z e \bar{\Phi}_1(\vec{x}) + \mu \bar{B}_{1 \parallel}(\vec{x}) \right) \mathrm{d}\mu.
\end{eqnarray*}
In order to evaluate the integrals of the Bessel functions, let us introduce the new variable $x = \frac{\mu B}{T}$, leading to $\lambda^2 = 2 x b$, $b = -\rho_{\st{th}}^2 \nabla^2_{\perp}$, where $\rho_{\st{th}}$ is the thermal Larmor radius. The two remaining integrals are of the form
\[ \int \limits_0^{\infty} J_0^2(\sqrt{2bx}) \mathrm{e}^{-x} \mathrm{d}x \qquad \int \limits_0^{\infty} J_0(\sqrt{2bx}) \hat{J}_1(\sqrt{2bx}) \sqrt{x} \mathrm{e}^{-x} \mathrm{d}x. \]
For the details of evaluating these integrals see appendix \ref{app:bessel_integral}. The results are
\[\mathrm{e}^{-b} I_0(b) \qquad \mathrm{e}^{-b} \left( I_0(b) - I_1(b) \right) ,\]
respectively, where $I_n$ is the n-th order modified Bessel function of the first kind. Using the notation $\Gamma_n(b) = I_n(b) \mathrm{e}^{-b}$, the expression for the particle density can be written as
\[n(\vec{x}) = \bar{n}(\vec{x}) - \frac{Z e \Phi_1(\vec{x})}{T} n_0(\vec{x}) + n_0 \st{e}^{\frac{-\mathcal{E}}{T}} \left( \Gamma_0(b) \frac{Z e \Phi_1(\vec{x})}{T} + (\Gamma_0(b) - \Gamma_1(b)) \frac{B_{1 \parallel}(\vec{x})}{B_0} \right). \]
The quasineutrality equation $\sum_{\st{sp}} Z_{\st{sp}} e n_{\st{sp}} = 0$ gives
\begin{eqnarray}
	&& \sum_{\st{sp}} Z_{\st{sp}} e \left( 1 - \st{e}^{\frac{-\mathcal{E}_{\st{sp}}}{T_{\st{sp}}}} \Gamma_0(b_{\st{sp}}) \right) \frac{Z_{\st{sp}}e \Phi_1(\vec{x}) n_{0,\st{sp}}}{T_{\st{sp}}} =  \nonumber \\
	&& \hspace{1cm} \sum_{\st{sp}} Z_{\st{sp}} e \left( \bar{n}_{\st{sp}} + \st{e}^{\frac{-\mathcal{E}_{\st{sp}}}{T_{\st{sp}}}} (\Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}})) \frac{B_{1 \parallel}(\vec{x}) n_{0,\st{sp}}}{B_0} \right).
	\label{eq:neutrality}
\end{eqnarray}

In order to bring this equation to the form implemented in GKW, the modified distribution function $g$ has to be applied. When substituting equation \ref{eq:g} into the definition of the gyro-centre density, it can be seen that the second term provides zero contribution since it contains the product of $F_{\st{M}}$, which is symmetric in $v_{\parallel}$, with $v_{\parallel}$, which is antisymmetric:
\begin{eqnarray*}
	\bar{n}(\vec{x}) &=& \frac{B_0}{m} \int \delta(\vec{X}+\vec{r}-\vec{x}) f_{\st{gy}}(\vec{X}) \mathrm{d} \vec{X} \mathrm{d} v_{\parallel} \mathrm{d} \mu \mathrm{d} \theta \\
	&=& \frac{2 \pi B_0}{m} \int J_0(\lambda) \left( g(\vec{x}) - \frac{F_{\st{M}}}{T} v_{\parallel} Z e \bar{A}_{1 \parallel}\right) \mathrm{d}v_{\parallel} \mathrm{d} \mu = \frac{2 \pi B_0}{m} \int J_0(\lambda) g(\vec{x}) \mathrm{d}v_{\parallel} \mathrm{d} \mu.
	\label{eq:density_gkw_nonorm}
\end{eqnarray*}

After Fourier-transforming and applying the normalizing expressions of section \ref{sec:normalization}, the gyrokinetic Poisson equation finally takes the form
\begin{eqnarray}
	\hspace{-5mm}
	&& \sum_{\st{sp}} Z_{\st{sp}} n_{\st{R,sp}} \left[ 2 \pi B_{\st{N}} \int J_0(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d} v_{\parallel N} \mathrm{d} \mu_{\st{N}} + \right. \nonumber \\ 
	\hspace{-5mm}
	&& \left. \frac{Z_{\st{sp}}}{T_{\st{R,sp}}} \hat{\Phi}_{1 \st{N}} \left( \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \Gamma_0(b_{\st{sp}}) - 1 \right) + \frac{\hat{B}_{1 \parallel \st{N}}}{B_{\st{N}}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} (\Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}})) \right] = 0.
	\label{eq:poisson}
\end{eqnarray}
Since the equation does not contain spatial derivatives of the fields and the distribution function, and they are linear in every term, the Fourier transform simply results in the Fourier components $\hat{g}_{\st{N,sp}}(\vec{k})$, $\hat{\Phi}_{1 \st{N}}(\vec{k})$ and $\hat{B}_{1 \parallel \st{N}}(\vec{k})$.

%-----------------------------------------------------------------------
\subsubsection{Equation for $A_{1 \parallel}$}
The method to calculate the equation of the parallel perturbation of the vector potential is analogous to that used for the Poisson-equation.
The starting point is the parallel component of Amp\`ere's law. It is first transformed into guiding-centre coordinates, then the guiding-centre distribution function is expressed with the pull-back of its gyro-centre counterpart.

The parallel component of Amp\`ere's law using Coulomb gauge $\nabla \cdot \mathbf{A}_1 = 0$ can be written as
\begin{equation} 
 \nabla^2 \mathbf{A}_{1 \parallel} = \mu_0 \mathbf{j}_{1 \parallel}. 
 \label{eq:ampere}
\end{equation}
The parallel perturbation of the current density is expressed with equation \ref{eq:gc_j_par}. After applying equation \ref{eq:f_gy_to_gc} it takes the form
\begin{eqnarray*}
	j_{1 \parallel} &=& \frac{Z e B_0}{m} \int v_{\parallel} \delta(\vec{X}+\vec{r}-\vec{x}) \left[ f_{\st{gy}}(\vec{X}) - \frac{F_{\st{M}}}{T} \left( Z e \widetilde{\Phi}_1(\vec{X} + \vec{r}) - \mu \bar{B}_{1 \parallel}(\vec{X}) \right) \right] \mathrm{d} \vec{X} \mathrm{d} v_{\parallel} \mathrm{d} \mu \mathrm{d} \theta \\
	&=& \frac{Z e B_0}{m} \int v_{\parallel} \delta(\vec{X}+\vec{r}-\vec{x}) f_{\st{gy}}(\vec{X}) \mathrm{d} \vec{X} \mathrm{d} v_{\parallel} \mathrm{d} \mu \mathrm{d} \theta - \frac{Z e B_0}{m} \underbrace{\int v_{\parallel} F_{\st{M}}(v_{\parallel}) \mathrm{d} v_{\parallel}}_{0} \int (\dots) \mathrm{d} \vec{X} \mathrm{d} \mu \mathrm{d} \theta \\
	&=& \frac{2 \pi Z e B_0}{m} \int v_{\parallel} J_0(\lambda) f_{\st{gy}}(\vec{x}) \mathrm{d} v_{\parallel} \mathrm{d} \mu.
\end{eqnarray*}
Substituting equation \ref{eq:g} gives
\begin{eqnarray*}
	j_{1 \parallel} &=& \frac{2 \pi Z e B_0}{m} \int v_{\parallel} J_0(\lambda) g \mathrm{d} v_{\parallel} \mathrm{d} \mu - \frac{2 \pi Z e B_0}{m} \int v_{\parallel}^2 J_0(\lambda) \frac{F_{\st{M}}}{T} Z e \bar{A}_{1 \parallel} \mathrm{d} v_{\parallel} \mathrm{d} \mu.
\end{eqnarray*}
In the second term the integral over $v_{\parallel}$ can be performed by parts leading to
\[-\frac{ 2 Z^2 e^2 B_0 n_0}{ m^2 v_{\st{th}}^{2}} A_{1 \parallel} \st{e}^{\frac{-\mathcal{E}}{T}} \underbrace{\int J_0^2(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} \mu}_{\frac{T}{B_0} \Gamma_0(b)},\]
and so the total expression for $j_{1 \parallel}$ becomes
\[j_{1 \parallel} = \frac{2 \pi B_0 Z e}{m} \int v_{\parallel} J_0(\lambda) g \mathrm{d} v_{\parallel} \mathrm{d} \mu - \frac{Z^2 e^2 n_0}{m} \st{e}^{\frac{-\mathcal{E}}{T}} \Gamma_0(b) A_{1 \parallel}. \]
Summing the above expression over the species, substituting into Amp\`ere's law, Fourier transforming and applying the normalization equations yield
\begin{eqnarray}
	&& \left[ k_{\perp N}^2 + \beta_{\st{ref}} \sum_{\st{sp}} \frac{Z_{\st{sp}}^2 n_{\st{R,sp}}}{m_{\st{R,sp}}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \Gamma_0(b_{\st{sp}}) \right] \hat{A}_{1 \parallel \st{N}} = \nonumber \\
	&& 2 \pi B_{N} \beta_{\st{ref}} \sum_{\st{sp}} Z_{\st{sp}} n_{\st{R,sp}} v_{\st{R,sp}} \int v_{\parallel N} J_0(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d} v_{\parallel N} \mathrm{d} \mu_{N}
	\label{eq:apar}
\end{eqnarray}
where the reference plasma beta has been defined as
\begin{equation}
	\beta_{\st{ref}} = \frac{2 \mu_0 n_{\st{ref}} T_{\st{ref}}}{B_{\st{ref}}^2}.
	\label{eq:beta}
\end{equation}

%-----------------------------------------------------------------------
\subsubsection{Equation for $B_{1 \parallel}$}
The perpendicular component of Amp\`ere's law can be written as
\[ (\nabla \times \mathbf{B}_1)_{\perp} = \left( \begin{array}{c} \partial_y B_{1 \parallel} - \partial_z B_{1 y} \\ \partial_z B_{1x} - \partial_x B_{1 \parallel} \end{array} \right) = \mu_0 \mathbf{j}_{1 \perp} \]
where $z$ is the direction of the equilibrium magnetic field. The parallel gradients of the perturbed quantities can be neglected since they are one order smaller than perpendicular ones, giving
\begin{equation*}
	\left( \begin{array}{c} \partial_y B_{1 \parallel} \\ - \partial_x B_{1 \parallel} \end{array} \right) = \nabla_{\perp} B_{1 \parallel} \times \mathbf{b} = \mu_0 \mathbf{j}_{1 \perp}.
\end{equation*}
Taking the cross-product of the above equation with $\vec{b}$ from the left gives
\[ \mu_0 \vec{b} \times \vec{j}_{1 \perp} = \vec{b} \times \nabla_{\perp} B_{1 \parallel} \times \vec{b} = \nabla_{\perp} B_{1 \parallel}. \]
Using equation \ref{eq:gc_j_perp}, assuming zero equilibrium current $\vec{j}_0 = 0$, summing over species and dotting with $\vec{k}_{\perp}$ leads to
\begin{eqnarray*}
	\vec{k}_{\perp} \cdot \nabla_{\perp} B_{1 \parallel} &=& \mu_0 B_0 e \sum_{\st{sp}} \frac{Z_{\st{sp}}}{m_{\st{sp}}} \int \delta(\vec{X}+\vec{r}-\vec{x}) \underbrace{\vec{k}_{\perp} \cdot \vec{b} \times \vec{v}_{\st{L}}}_{\vec{v}_{\st{L}} \cdot (\vec{k}_{\perp} \times \vec{b})} \\ 
	&& \left( f_{\st{gy,sp}} - \left( Z_{\st{sp}} e \widetilde{\Phi}_{1} - \mu \bar{B}_{1 \parallel} \right) \frac{F_{\st{M}}}{T_{\st{sp}}} \right) \mathrm{d} \vec{X} \mathrm{d} v_{\parallel} \mathrm{d} \mu \mathrm{d} \theta \\
	&=& \mu_0 B_0 e \sum_{\st{sp}} \frac{Z_{\st{sp}}}{m_{\st{sp}}} \left[ \int \vec{v}_{\st{L}} \cdot (\vec{k}_{\perp} \times \vec{b}(\vec{x})) f_{\st{gy,sp}}(\vec{x}-\vec{r}) \mathrm{d} v_{\parallel} \mathrm{d} \mu \mathrm{d} \theta \right. \\ 
	&& \left. + \int \vec{v}_{\st{L}} \cdot (\vec{k}_{\perp} \times \vec{b}(\vec{x})) \frac{F_{\st{M}}}{T_{\st{sp}}}\left( Z_{\st{sp}} e \bar{\Phi}_1(\vec{x}-\vec{r}) + \mu \bar{B}_{1 \parallel}(\vec{x}-\vec{r}) \right)  \mathrm{d} v_{\parallel} \mathrm{d} \mu \mathrm{d} \theta \right].
\end{eqnarray*}
Performing the $\theta$-integral gives (for details see \cite{dannert}, Appendix C)
\begin{eqnarray*}
	\vec{k}_{\perp} \cdot \nabla_{\perp} B_{1 \parallel} &=& - 2 \pi \mu_0 B_0 e \sum_{\st{sp}} \frac{Z_{\st{sp}}}{|Z_{\st{sp}}|} \frac{Z_{\st{sp}}}{m_{\st{sp}}} \left[ v_{\st{L}} \int \frac{\mathrm{d}^3 k}{(2 \pi)^3} \left( i k_{\perp}^2 (\vec{k}_{\perp} \times \vec{b} \cdot \vec{e}_1 \hat{f}_{\st{gy,sp}}) - \right. \right. \\ 
	&& \left. i k_{\perp}^1 (\vec{k}_{\perp} \times \vec{b} \cdot \vec{e}_2 \hat{f}_{\st{gy,sp}}) \right) \frac{\rho_{\st{sp}}}{2} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \mathrm{e}^{i \vec{k} \cdot \vec{r}} \mathrm{d} v_{\parallel} \mathrm{d} \mu + \\
	&& v_{\st{L}} \int \frac{\mathrm{d}^3 k}{(2 \pi)^3} i k_{\perp}^2 \left( \vec{k}_{\perp} \times \vec{b} \cdot \vec{e}_1 \frac{F_{\st{M}}}{T_{\st{sp}}} \left( Z_{\st{sp}} e \hat{\bar{\Phi}}_1 + \mu \hat{\bar{B}}_{1 \parallel} \right) \right) - \\ 
	&& \left. i k_{\perp}^1 \left( \vec{k}_{\perp} \times \vec{b} \cdot \vec{e}_2 \frac{F_{\st{M}}}{T_{\st{sp}}} \left( Z_{\st{sp}} e \hat{\bar{\Phi}}_1 + \mu \hat{\bar{B}}_{1 \parallel} \right) \right) \frac{\rho_{\st{sp}}}{2} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \mathrm{e}^{i \vec{k} \cdot \vec{r}} \mathrm{d} v_{\parallel} \mathrm{d} \mu \right].
\end{eqnarray*}
It can be easily shown that $\vec{k}_{\perp} \times \vec{b} \cdot \vec{e}_1 = k_{\perp}^2$ and $\vec{k}_{\perp} \times \vec{b} \cdot \vec{e}_2 = -k_{\perp}^1$, where the upper indices represent contravariant components. Fourier transforming both sides, dividing by the perpendicular wavenumber and transforming back to real space results
\begin{eqnarray*}
	i k_{\perp}^2 \hat{B}_{1 \parallel} &=& - 2 \pi B_0 e \mu_0 \sum_{\st{sp}} \frac{Z_{\st{sp}}}{|Z_{\st{sp}}|} \frac{Z_{\st{sp}}}{m_{\st{sp}}} \left[ v_{\st{L}} \int i k_{\perp}^2 \hat{f}_{\st{gy,sp}} \frac{\rho_{\st{sp}}}{2} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \mathrm{d} v_{\parallel} \mathrm{d} \mu \right. + \\ 
	&& \left. v_{\st{L}} \int i k_{\perp}^2 \left( Z_{\st{sp}} e \hat{\bar{\Phi}}_1 + \mu \hat{\bar{B}}_{1 \parallel} \right) \frac{F_{\st{M}}}{T_{\st{sp}}} \frac{\rho_{\st{sp}}}{2} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \mathrm{d} v_{\parallel} \mathrm{d} \mu \right] \\
	B_{1 \parallel} &=& - 2 \pi B_0 \mu_0 \sum_{\st{sp}} \frac{Z_{\st{sp}}}{|Z_{\st{sp}}|} \frac{1}{m_{\st{sp}}} \left[ \int \mu \frac{Z_{\st{sp}}}{|Z_{\st{sp}}|} \hat{J}_1(\lambda_{\st{sp}}) f_{\st{gy,sp}} \mathrm{d} v_{\parallel} \mathrm{d} \mu \right. \\
	&& \left. + \int \mu \left( Z_{\st{sp}} e J_0(\lambda_{\st{sp}}) \Phi_1 + \mu \hat{J}_1(\lambda_{\st{sp}}) B_{1 \parallel} \right) \frac{F_{\st{M}}}{T_{\st{sp}}} \frac{Z_{\st{sp}}}{|Z_{\st{sp}}|} \hat{J}_{1}(\lambda_{\st{sp}}) \mathrm{d} v_{\parallel} \mathrm{d} \mu \right].
\end{eqnarray*}
Here the upper index in $k_{\perp}^2$ means second power. The integral over $v_{\parallel}$ is elementary as again it only appears in the Maxwellian. Using equation \ref{eq:g} and evaluating the integrals over $\mu$ according to equations \ref{eq:bessel_integral_bpar} in appendix \ref{app:bessel_integral} one obtains
\begin{eqnarray*}
B_{1 \parallel} &=& - \sum_{\st{sp}} \frac{2 \mu_0 T_{\st{sp}} n_{0,\st{sp}}}{B_0} \left[ \frac{\pi B_0^2}{m_{\st{sp}} n_{0,\st{sp}} T_{\st{sp}}} \int \mu \hat{J}_1(\lambda_{\st{sp}}) g_{\st{sp}} \mathrm{d}v_{\parallel} \mathrm{d} \mu + \right. \nonumber \\
&& \left. \st{e}^{\frac{-\mathcal{E}}{T_{\st{sp}}}} (\Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}})) \left( \frac{Z_{\st{sp}} e \Phi_1}{2 T_{\st{sp}}} + \frac{B_{1 \parallel}}{B_0} \right) \right].
\end{eqnarray*}
Normalization and rearrangement of the terms after writing the fields in Fourier-space yield
\begin{eqnarray}
	&& \left[ 1 + \sum_{\st{sp}} \beta_{\st{sp}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \left( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) \right) \right] \hat{B}_{1 \parallel \st{N}} = \nonumber \\
	&& \hspace{2cm} - \sum_{\st{sp}} \beta_{\st{sp}} \left[ 2 \pi B_{\st{N}}^3 \int \mu_{N} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d}v_{\parallel N} \mathrm{d} \mu_{\st{N}} \right. + \nonumber \\ 
	&& \hspace{2cm} \left. \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \left( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) \right) \frac{Z_{\st{sp}} B_{\st{N}}}{2 T_{\st{R,sp}}} \hat{\Phi}_{1 \st{N}} \right].
	\label{eq:bpar1}
\end{eqnarray}
Using the reference instead of the full value of the plasma beta ($\beta = \beta_{\st{ref}} \beta_{\st{R}} = \beta_{\st{ref}} \frac{T_{\st{R,sp}} n_{\st{R,sp}}}{B_{\st{N}}^2}$), the equation for the parallel magnetic perturbation can be finally written as
\begin{eqnarray}
	&& \hspace{-1cm} \left[ 1 + \sum_{\st{sp}} \frac{T_{\st{R,sp}} n_{\st{R,sp}}}{B_{\st{N}}^2} \beta_{\st{ref}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \left( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) \right) \right] \hat{B}_{1 \parallel \st{N}} = \nonumber \\
	&& \hspace{1cm} - \sum_{\st{sp}} \beta_{\st{ref}} \left[ 2 \pi B_{\st{N}} T_{\st{R,sp}} n_{\st{R,sp}} \int \mu_{N} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d}v_{\parallel N} \mathrm{d} \mu_{\st{N}} \right. \nonumber\\
	&& \hspace{1cm} + \left. \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \left( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) \right) \frac{Z_{\st{sp}} n_{\st{R,sp}}}{2 B_{\st{N}}} \hat{\Phi}_{1 \st{N}} \right].
	\label{eq:bpar}
\end{eqnarray}

It can be seen that both the perpendicular (equation \ref{eq:apar}) and parallel (equation \ref{eq:bpar1}) components of the perturbed magnetic field depend strongly on the plasma $\beta$. Indeed, it has been commonly observed in gyrokinetic simulations that in low $\beta$ plasmas the amplitude of the magnetic perturbations is much lower than that of the fluctuating scalar potential. However, in high $\beta$ plasmas magnetic perturbations strongly impact mode stability and transport by destabilizing kinetic ballooning \cite{jenko_em} and micro-tearing \cite{applegate_mtm} modes.

The correction suggested in equation \ref{eq:bpar1} compared to the calculation by Dannert in \cite{dannert} is the missing $2 b$ factor in the second term on the left hand side, which is the direct consequence of equation \ref{eq:bessel_integral_bpar}. High $\beta$ simulations with GKW using the corrected form of this equation have been successfully benchmarked against an other gyrokinetic code, GS2 \cite{gs2}.

%-----------------------------------------------------------------------
\newpage

\begin{thebibliography}{99}
 \bibitem{dannert} T. Dannert: Gyrokinetische Simulation von Plasmaturbulenz mit gefangenen Teilchen und Elektromagnetischen Effekten (Doctoral Thesis, Technische Universt\"at M\"unchen, 2005)
 \bibitem{gkw} A.G. Peeters et al. (including G. Szepesi), Comput. Phys. Commun. {\bf 180}, 2650 (2009)
 \bibitem{bellan} P.M. Bellan, Fundamentals of Plasma Physics (Cambridge University Press, 2006)
 \bibitem{brizard} A.J. Brizard, T.S. Hahm, Reviews of Modern Physics {\bf 79}, 421 (2007)
 \bibitem{littlejohn_gc} R.G. Littlejohn, Phys. Fluids {\bf 24}, 1730 (1981)
 \bibitem{fecko} M. Fecko Differential Geometry and Lie Groups for Physicists (Cambridge University Press, 2006) 
 \bibitem{littlejohn} R.G. Littlejohn, J. Math. Phys. {\bf 23}, 742 (1982)
 \bibitem{peeters_rotation} A.G. Peeters et al., Phys. Plasmas {\bf 16}, 042310 (2009)
 \bibitem{scott_gotit} B. Scott: Introduction to Gyrokinetic Field Theory (Lecture notes, GOTiT course at IPP Garching, 2008 November)
 \bibitem{Casson2010} F.J. Casson et al. (including G. Szepesi), Phys. Plasmas {\bf 17}, 102305 (2010)
 \bibitem{a_and_s} M. Abramowitz, I.A. Stegun, eds., Handbook of Mathematical Functions with Formulas, Graphs, and Mathematical Tables, (New York: Dover Publications, 1972)
 \bibitem{jenko_em} F. Jenko, W. Dorland, Plasma Phys. Control. Fusion {\bf 43}, A141 (2001)
 \bibitem{applegate_mtm} D J Applegate et al., Plasma Phys. Control. Fusion {\bf 49}, 1113 (2007)
 \bibitem{gs2} M. Kotschenreuther, G. Rewoldt, and M. W. Tang, Comput. Phys. Commun. {\bf 88}, 128 (1995)
\bibitem{g_and_r} I.S Gradshteyn, I.M. Ryzhik: Table of integrals, series, and products (Academic Press, 1965)
 \bibitem{w_and_w} E.T. Whittaker, G.N. Watson, A Course of Modern Analysis (Cambridge University Press, 4th edition, 1927)
\end{thebibliography}

%-----------------------------------------------------------------------
\newpage

\appendix
% \appendixpage

%-----------------------------------------------------------------------
\section{Integrals involving products of Bessel functions} \label{app:bessel_integral}
The following three types of integrals involving products of Bessel functions, exponential and algebraic terms appear in the derivation of the gyrocentre Maxwell's equations:
\begin{eqnarray}
	1. && \int \limits_{0}^{\infty} J_0^2(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} \mu \\
	2. && \int \limits_{0}^{\infty} \mu J_0(\lambda) \hat{J}_1(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} \mu \\
	3. && \int \limits_{0}^{\infty} \mu^2 \hat{J}_1^2(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} \mu
\end{eqnarray}
where $\hat{J}_1(\lambda) = \frac{2}{\lambda} J_1(\lambda)$. In all three cases the new variable $x = \frac{\mu B_0}{T}$ is introduced according to Dannert \cite{dannert}, which also leads to $\lambda = \sqrt{2 x b}$ where $b = - \rho_{th}^2 \nabla_{\perp}^2$. The Jacobian is simply $\frac{T}{B_0}$ and the integrals can be written as
\begin{eqnarray*}
	1. && \frac{T}{B_0} \int \limits_{0}^{\infty} J_0^2(\sqrt{2 x b}) \mathrm{e}^{-x} \mathrm{d} x \\
	2. && \frac{T^2}{B_0^2} \frac{2}{\sqrt{2b}} \int \limits_{0}^{\infty} \sqrt{x} J_0(\sqrt{2 x b}) J_1(\sqrt{2 x b}) \mathrm{e}^{-x} \mathrm{d} x \\
	3. && \frac{T^3}{B_0^3} \frac{2}{b} \int \limits_{0}^{\infty} x J_1^2(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} x.
\end{eqnarray*}
The first integral can be solved using the formula 6.615 on page 710 of \cite{g_and_r}:
\[\int \limits_0^{\infty} \mathrm{e}^{- \alpha x} J_{\nu}(2 \beta \sqrt{x}) J_{nu}(2 \gamma \sqrt{x}) \mathrm{d} x = \frac{1}{\alpha} I_{\nu}\left( \frac{2 \beta \gamma}{\alpha} \right) \mathrm{e}^{-\frac{\beta^2 + \gamma^2}{\alpha^2}}\]
which immedately gives
\begin{eqnarray*}
	1. && \frac{T}{B_0} \int \limits_{0}^{\infty} J_0^2(\sqrt{2 x b}) \mathrm{e}^{-x} \mathrm{d} x = I_0(b) \mathrm{e}^{-b}.
\end{eqnarray*}
In the two remaining integrals let us use the new variable $y=\sqrt{2 b x}$. The Jacobian is $\frac{y}{b}$ and therefore it leads to
\begin{eqnarray*}
	2. && \frac{T^2}{B_0^2} \frac{2}{\sqrt{2b}} \int \limits_{0}^{\infty} \sqrt{x} J_0(\sqrt{2 x b}) J_1(\sqrt{2 x b}) \mathrm{e}^{-x} \mathrm{d} x = \frac{T^2}{B_0^2} \frac{1}{b^2} \int \limits_{0}^{\infty} y^2 J_0(y) J_1(y) \mathrm{e}^{-h y^2} \mathrm{d}y \\
	3. && \frac{T^3}{B_0^3} \frac{2}{b} \int \limits_{0}^{\infty} x J_1^2(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} \mu = \frac{T^3}{B_0^3} \frac{1}{b^3} \int \limits_{0}^{\infty} y^3 J_1^2(y) \mathrm{e}^{-h y^2} \mathrm{d} y
\end{eqnarray*}
where $h = \frac{1}{2b}$ was used in the exponentials.

The second integral (leaving the prefactors for now) can be further written as
\begin{eqnarray*}
	\int \limits_{0}^{\infty} y^2 J_0(y) J_1(y) \mathrm{e}^{-h y^2} \mathrm{d}y &=& \frac{\partial}{\partial h} \left( - \int \limits_0^{\infty} \mathrm{e}^{-hy^2} J_0(y) J_1(y) \right) \mathrm{d} y\\
	&=& \frac{1}{2} \frac{\partial}{\partial h} \left( \int \limits_0^{\infty} \mathrm{e}^{-hy^2} \frac{\mathrm{d}}{\mathrm{d} y} \left(J_0^2(y)\right) \right) \mathrm{d}y \\
	&=& \frac{1}{2} \frac{\partial}{\partial h} \left( \underbrace{\left[ J_0^2(y) \mathrm{e}^{-h y^2} \right]_0^{\infty}}_{1}  + 2 h \int \limits_0^{\infty} J_0^2(y) y \mathrm{e}^{-h y^2} \mathrm{d}y \right)
\end{eqnarray*}
where the relation $\frac{\mathrm{d} J_0(y)}{\mathrm{d} y} = - J_1(y)$ was applied (see \cite{a_and_s}). The integral can be found in \cite{w_and_w} on page 395 and it gives
\begin{equation}
	\int \limits_{0}^{\infty} y J_0^2(y) \mathrm{e}^{-h y^2} \mathrm{d}y = \frac{1}{2h} \mathrm{e}^{-\frac{1}{2h}} I_0\left(\frac{1}{2h}\right).
\end{equation}
Taking the derivative with respect to $h$ and using the identity $\frac{\mathrm{d}I_0(z)}{\mathrm{d} z} = I_1(z)$ results
\begin{eqnarray}
	\int \limits_{0}^{\infty} y^2 J_0(y) J_1(y) \mathrm{e}^{-h y^2} \mathrm{d}y &=& \frac{1}{4h^2} \mathrm{e}^{-\frac{1}{2h}} \left( I_0\left( \frac{1}{2h} \right) - I_1\left( \frac{1}{2h} \right) \right) \nonumber \\
	&=& b^2 \mathrm{e}^{-b} \left( I_0(b) - I_1(b) \right).
	\label{eq:K2}
\end{eqnarray}
And with the prefactors:
\begin{eqnarray}
	2. && \int \limits_{0}^{\infty} \mu J_0(\lambda) \hat{J}_1(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} \mu = \frac{T^2}{B_0^2} \mathrm{e}^{-b} \left( I_0(b) - I_1(b) \right).
\end{eqnarray}

The third integral (without the prefactors) is integrated by parts first to get
\begin{eqnarray*}
	\int \limits_{0}^{\infty} y^3 J_1^2(y) \mathrm{e}^{-h y^2} \mathrm{d} y &=& -\frac{1}{2 h} \int \limits_{0}^{\infty} (-2 h y) \mathrm{e}^{-h y^2} y^2 J_1(y) \mathrm{d} y \\
	&=& -\frac{1}{2h} \left( \underbrace{\left[ \mathrm{e}^{-h y^2} y^2 J_1^2(y) \right]_0^{\infty}}_{0} - 2 \int \limits_0^{\infty}  \mathrm{e}^{-h y^2} \left( y J_1^2(y) + y^2 J_1(y) \frac{\mathrm{d}J_1(y)}{\mathrm{d}y} \right) \right).
\end{eqnarray*}
Using the relation $\frac{\mathrm{d}J_1(y)}{\mathrm{d}y} = J_0(y) - \frac{1}{y} J_1(y)$ (see \cite{a_and_s}) two of the terms cancel out and the integral simplifies to
\begin{equation}
	\int \limits_{0}^{\infty} y^3 J_1^2(y) \mathrm{e}^{-h y^2} \mathrm{d} y = \frac{1}{h} \int \limits_{0}^{\infty} y^2 J_0(y) J_1(y) \mathrm{e}^{-h y^2} \mathrm{d} y. \nonumber
\end{equation}
Using equation (\ref{eq:K2}) the third integral can be written as
\begin{equation}
	\int \limits_{0}^{\infty} y^3 J_1^2(y) \mathrm{e}^{-h y^2} \mathrm{d} y = 2 b^3 \mathrm{e}^{-b} \left( I_0(b) - I_1(b) \right)
\end{equation}
and with prefactors:
\begin{eqnarray}
	3. && \int \limits_{0}^{\infty} \mu^2 \hat{J}_1^2(\lambda) \mathrm{e}^{-\frac{\mu B_0}{T}} \mathrm{d} \mu = \frac{T^3}{B_0^3} 2 \mathrm{e}^{-b} \left( I_0(b) - I_1(b) \right).
	\label{eq:bessel_integral_bpar}
\end{eqnarray}

\end{document}

### 2022-05-03 gkw/doc/etc/devel.tex
(1) Added gkw Clone gwk  #3

%% GKW developer notes: presently only rules and guidelines for developers. 
%
\documentclass[12pt,twoside,notitlepage,a4paper]{article}

%% caption modifications; default to footnotesize - tweak as appropriate
\usepackage[font={footnotesize,it}]{caption}
%\usepackage[small,labelfont=bf,textfont=it,width=0.9\textwidth]{caption}

%% for modifying description lists 
%\usepackage{expdlist}

%% the listings package; useful for blocks of code
\usepackage{listings}

%% fancy verbatim {Verbatim}
\usepackage{fancyvrb}

%% for using some regular formatting in verbatim {alltt}
%\usepackage{alltt}

%% underscore
\usepackage{underscore}

%% graphics
%\usepackage{graphicx}

%% look for figure input in ./figs
%%\graphicspath{{../figs/}}

%% fullpage
\usepackage{fullpage}

%% color
\usepackage{color}

%% makeindex ?
%\usepackage{makeidx}

\usepackage{bm}

%% hyperref
\usepackage[hyperindex]{hyperref}

% ---------------------------------------------------------------------------- %
%% Macros
\newcommand{\file}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

%%\newenvironment{mylisting}
%%{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\scriptsize\bfseries}
%%{\end{list}}
%%
%%\newenvironment{mytinylisting}
%%{\begin{list}{}{\setlength{\leftmargin}{1em}}\item\tiny\bfseries}
%%{\end{list}}

% ---------------------------------------------------------------------------- %

\begin{document}
\pagestyle{empty}
\pagestyle{plain}
\pagenumbering{roman}

% ============================================================================ %

\begin{titlepage}
  \begin{center}

  \large

  \vspace*{2cm}

  {\Large\sffamily\bfseries GKW:\\[2\parskip]
    Notes}

  \vspace{3ex}

  {\sffamily DeveloperGuidelines}

  \vspace{\stretch{2}}

  \vspace{\stretch{2}}

  \vspace{\stretch{1}}

  \emph{$ $Date$ $, $ $Revision$ $}\\
  \url{http://gkw.googlecode.com}

  \vspace{\stretch{3}}

\end{center}

\end{titlepage}

\clearpage
\tableofcontents
\cleardoublepage
%\pagestyle{fancy}
\pagenumbering{arabic}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\cleardoublepage
\appendix
\part{Appendix}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% ============================================================================ %

\section{Coding GKW: rules, style and recommendations}

It is intended that this appendix becomes a part of the main code documentation.
Its purpose should be to make clear what standards are expected of code or files
committed to the repository.
For now, we should use it as a discussion about what exactly these things should be.
The motivations for adopting such conventions are fairly clear (e.g. keeping the code
readable, portable, maintainable, flexible etc.). We should aim to
find a {\it reasonable} set of conventions to agree on, rather than iterate to perfection.

I suggest that we initially focus only on the Fortran source parts of the code, then
develop conventions for other files later as necessary. It may also be worth adding
a section of notes about ways of avoiding certain undesirable features (e.g.
non-standard/extended Fortran -- just a list of things we already know about that you should not
do).

So far we have adopted some conventions and style fairly consistently throughout the code
and these can be documented here already. There are some conventions which should be
clarified and perhaps more that should be adopted. If the code is tidied a bit, some
conventions will become self-evident and we may not need to document them here.

Please discuss (in this document file even). The following could be restructured a bit...

FC, YC:  I agree with all, except where I have commented otherwise.\\
{\bf YC: it looks like GKW is not in a state consistent with the guidelines. How do we implement the changes ?
Each of us takes a couple of files and does it ? APS: it is not difficult for one of use to do it. Some checks
are difficult so I will likely write a short script to do them. It might be
worth 2 people first independently look look at 1 module, then review what the other did to see
if it meets their understanding of the conventions...}

\subsection{General}

The code is written in standard Fortran 95. Extensions to the standard are not
allowed and redundant features should be avoided. The code should always
be able to compile and run on any system which has a working Fortran compiler.

External Fortran libraries may be added to the repository to provide extra     
functionality to the code. These should be no less portable than the rest
of the code and their license must be compatible with the one with which the
code is distributed. Anything copied from some other source should have its
origin made very clear, and again, do make sure that it is allowed to
be distributed together with the rest of the code (If in doubt, please
discuss with the code administrators.). 
%APS:
{\bf APS: Is this OK? If so, must the added code meet any conventions? For example, 
our FFT interface follows different style and conventions to the rest of the code.
Also, can any added code have unintended consequences? What should the policy be?}

{\bf AGP: Everything in /trunk/src should follow the style rules, even if we 
stole it from somewhere else. Is there a good suggestion on where to store library 
source code. APS: separate subdirectory at the same level as trunk, branches; 
README file making it clear it is not part
of the code; named `external', `lib', ? (I don't like any of these names very  much). }

The code may make use of calls to other 3rd-party libraries, provided they have
some sort of Fortran interface. Calls to such libraries should be hidden away
from the rest of the code as much as possible. The use of non-portable and
non-free software libraries is strongly discouraged; they should never be
enabled as default and must never constitute an essential code component;
they should only be used to enhance performance of the code on specific systems
rather than functionality. (FC: The OpenMP in exp integration currently makes use of the Intel MKL, 
but as discussed in issue 5 this is only a temporary arrangment during testing and development.)

We aim to make the code easy to read, understand, maintain and modify. 
To this end, we make many functions and subroutines as general as possible to encourage code re-use.
The following sections highlight particular aspects of this style and adopted conventions.

Suggestions regarding adding new features to the code, restructuring the existing code etc. can be discussed in the discussion list.

\subsection{Files}

Each source code file should contain either a Fortran program or a module.
The name of the file should correspond to the program or module contained within it,
followed by either a \file{.f90} or \file{.F90} suffix (the \file{.F90} suffix
denoting a file which must be preprocessed -- see {\it preprocessing} later).

Including files via the obsolete \code{include} statement should be avoided.
An exception is made for the interfaces to Fourier transforms and MPI libraries etc. -- if a file
must be included it should be included only once in a single file.

\subsection{Names}

Modules, subroutines, functions, variables etc. should be given meaningful or self-explanatory
names in general (be careful with spellings!). The names should be fairly concise, but not so short
that they are not understood and absolutely not more than 32 characters. For indices that are used in loops, the variable names
need only a single letter if desired (and should probably be kept short?). {\bf FC: I think two letters max is plenty.
APS: so we should explicitly discourage/disallow long indices names?}
{\bf AGP: to letters max is not a bad idea, but the code does violate this restriction rather strongly 
at present (imod). I think furthermore that we should encourage the use of standard indices for certain 
loops (for instance imod (or better im) for the toroidal mode number}
\textbf{YC: I do not really care about the length of the indice names as long as it is clear from their name that they are indices (which is I think is the case for imod) and are only used to that purpose. }
All names should be written in lower case in the code source files, apart from a few exceptions

\begin{itemize}
\item Optional function and subroutine Keywords.
\item External variable and subroutine names (e.g. the integer \code{MPI_OFFSET_KIND} is a
MPI variable.)
\item
%APS: 
{\bf APS: Anything else? We could add macro definitions, directives etc? (I vote for these)}
\end{itemize}

\subsection{Modules}

Each module should have a clearly defined purpose or set of tasks associated
with it as described in its doxygen compatible leading comment. Usually, all functions, subroutines and variables should be
private to that module my default and only those used elsewhere should be
exposed via the \code{public} statement.

\subsection{The use of \code{use} statements}

Every \code{use} statement should always be followed by the \code{only} option to explicitly
state which variables and functions are in scope. There is one exception for the MPI
interfaces, i.e. every module which has a name beginning `mpi' may drop this convention.
%APS:
{\bf APS: this need not be the case, only at the moment you would need to put all the MPI
things in too, e.g.
\begin{verbatim}
  use mpiinterface, only : MPI_ALLREDUCE, MPI_TYPE_CREATE_SUBARRAY
\end{verbatim}
etc.}

%Since the routines in a module typically work on the same arrays and structures etc.,
%a commonly used convention is to declare all the variables and routines used in
%other modules once globally at the module level via USE statements.
Rather than putting all the required \code{use} statements for a module at the beginning,
we have adopted a convention that \code{use} statements should only appear when required -- i.e.
they should pertain only to the subroutine in which they exist. 
%APS:
{\bf APS: this does not preclude the possibility of them appearing at the beginning of a module,
since they well may be necessary to define variables that are global to the module.}
{\bf AGP: if really necessary for the definition of global variables they may appear at the 
top, but only then. (This is not done consistently in the code at present)}
This enables one to see exactly what is within scope at the subroutine level.
To maximise the usefulness of this convention, only those variables
actually used should be imported to each routine. When starting a new routine, it
is very tempting to copy the \code{use} statements from a neighbouring code block without
checking if all variables are actually needed. When doing so, you should check that
any variable not used is removed.

\subsection{Interfaces}

%APS:
{\bf APS: There is a suggestion that we should restrict function overloading
to functions which have the same number of arguments. For example, the
\code{indx} function gets called with either 4 or 6 arguments, so perhaps
there should be two distinct functions that are called. Other overloads, such
as \code{mpibcast}, are fine as they always take the same number of arguments,
independent of datatype.}
{\bf FC:I see no benefit in overloading functions through interfaces, it only makes 
the actual bit of code take longer to find and makes the code harder to read. APS:
The purpose of interfaces is to make things easier all round. If something is not clear
we should change it. I don't believe the interfaces themselves are the problem.
Say I want to do a MPI_BCAST. Then doing it explicitly I have to (i) look up the interface
\url{http://www.mcs.anl.gov/research/projects/mpi/www/www3/MPI_Bcast.html}; (ii) include
preprocessor directives. (iii) check all the datatypes are correct. Using the 
overloaded function I get to know at compile time if I've made a mistake (rather than when the code
runs with some obscure MPI error message). (and the
wrapper can be used to handle all the mpi things in a consistent way rather than having to
repeat the same code over and over elsewhere).
FC: I also suggest generally avoiding optional arguments, they also make the code harder to read. APS:
sometimes we would be left with passing many lines of unnecessary variables to
a subroutine -- it would be nice to see which ones are actually being used. I agree with the examples
we discussed earlier, but these should be indentified in the
code and made clear.}

\subsection{Some things not allowed}

\begin{itemize}
\item No variable may have the pointer attribute. We do not use pointers in the code.
\item Derived data types should be avoided for the most numeric parts of the code (and in general?)
where potential compiler optimisation may be lost.
\end{itemize}

\subsection{Variable declaration}

\begin{itemize}
%APS:
\item {APS: \bf should we use \code{dimension} or not? I think putting the dimension after the
variable instead is fine, but we should stick to one or the other.} FC: Agree, no strong preference.
\item We do not use non-standard \code{real*8} etc. Instead we use \code{character (len=32)}, 
\code{real (kind=somekind)}, etc.
\item We use \code{::} for all declarations, not only for consistency but to avoid problems with some
compilers and also clarity.
\item Every dummy variable in a subroutine should have an \code{intent} (the default is
{\it undefined!}). This includes intent \code{inout}.
\end{itemize}

\subsection{Subroutines}

Individual subroutines should be kept short, not more than say $\sim200$ lines. A subroutine likely
tries to do too much if it is long and could be split into more subroutines.
%APS:
{\bf APS: perhaps part of the reason we are not encouraged to break subroutines is that we have
to set up all the use statements again? Subroutines can themselves contain subroutines, which
gets around this.}

All optional arguments should be capitalised.
e.g. the subroutine
\begin{verbatim}
subroutine mpibcast_real_array1(rscalar,nscalar,IPROC)

  real, intent(in) :: rscalar
  integer, intent(in) :: nscalar
  integer, optional, intent(in) :: IPROC
\end{verbatim}
with optional argument \code{IPROC} and its call
\begin{verbatim}
  call mpibcast(some_real,some_integer,IPROC=some_proc)
\end{verbatim}

\subsection{Code comments}

Comments are important! We can certainly have too many comments. We should aim
to explain what each routine does in the description. No need to go overboard;
comment should say what task is being performed; "This is a loop over x..."
is useless?

This is bad? (and hard to read?)
\begin{verbatim}
  ! something
  call do_something()
  ! something else
  call do_something_else()
  ! more still
  call do_more()
\end{verbatim}
is this better?
\begin{verbatim}
  ! something
  call do_something()

  ! something else
  call do_something_else()

  ! more still
  call do_more()
\end{verbatim}

Comments in GKW should be written in such a way as to be parsable by \href{http://www.doxygen.org}{Doxygen}.
The \href{http://www.gkw.org.uk/doxygen}{reference dictionary} for the code can then be automatically generated.

\begin{itemize}
\item All modules, functions and (module level) variable declarations have a useful Doxygen comment.
{\bf APS: should it be public variables, rather than module level?}
{\bf AGP: I would say module level as well, since a module can 5000 lines, and rather confusig to look through}
\item Make informative comments thinking of others {\bf APS: Sentences in English are very much appreciated!}
\item Notes to self should be marked with your initials e.g. \texttt{!APS: check this} {\bf APS: These should be in col 1 and considered as bugs! we should aim
to remove them. Note should be meaningful to all, not just creator; should it be dated?}
\textbf{YC: yes, I like the idea of dating it. It makes it easier to guess what is right when two comments are inconsistent.}
\item Questions to others should be marked e.g \texttt{!APS -> WAH: is this correct?} {\bf APS: I dont think it is so useful. Issue tracker is fine. initials 
considered as bugs; always start in the first col. Also all preprocessor and compiler directives. I think all other comments need a space between \code{!} and the first real
character of the comment.}
\item Input variables are labelled as such with default values listed if applicable at declaration comment
\item Allocatable arrays should have their size noted at declaration comment 
\item Indent comments inline with code
\item How do we do block comments?
\item Commented out code starts \code{!NAME: }? or what? Double \code{!}, with a
tag to explain who commented it out and why?
\textbf{YC: yes, I think it is good practice to explain why part of the code is commented out. Otherwise, why not remove it?}
%APS: \item Unused code should be commented out? Then comment rules apply...
%APS: comments are bugs?
\end{itemize}

\subsubsection{How to write comments compatible with the doxygen parser}

\textbf{Variables} can have either a leading comment, or trailing comments on same line.

\begin{itemize}
\item 
\textbf{Example 1:}\end{itemize}

\begin{verbatim}
!> the total energy flux per species : eflux_tot(number_of_species) 
!> electrostatic contribution 
real, allocatable, dimension(:) :: eflux_tot_es 
\end{verbatim}

\begin{itemize}
\item 
\textbf{Example 2:}\end{itemize}

\begin{verbatim}
logical :: coriolis             !< true if coriolis drift term is kept 
\end{verbatim}

\textbf{Functions}, \textbf{subroutines} and \textbf{modules} must have their description \textbf{preceding} the declaration. To keep it simple, use the \texttt{!>} comment format for all comment lines above a function declaration.  Lines will run together as a single sentence.

\begin{itemize}
\item 
\textbf{Example 3:}\end{itemize}

\begin{verbatim}
!-----------------------------------------------------------------------------
!> This routine calculates the correction necessary to go from the
!> distribution g (which includes the correction of the parallel
!> vector potential) to the distribution f
!-----------------------------------------------------------------------------
subroutine g2f_correct
..
..
\end{verbatim}

Alternatively, comments preceding a function can be written as a formatted block which will retain its layout through the parser (in next example).

Latex math mode can be used if enclosed within two \texttt{\\f\$} groups.  If used, it must
be valid latex without macros or packages - or the doxygen parser will fail.  See
\texttt{linear\_terms} for examples.

\begin{itemize}
\item 
\textbf{Example 4:}\end{itemize}

\begin{verbatim}
!++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
!> Add terms I and IV as one term : \f$ v_R {\cal F} \{H,g_N\} \f$
!! where
!!         \f$ H  = (1/2) v_{\parallel N}^2 + \mu_N B_N + (1/2) {\cal E}_R = H(s,\mu,v_{\parallel}) \f$,
!! and
!!         \f$ \{H,g_N\} = (d H/d s)(d g_N/d v) - (d H/d v)(d g_N/d s). \f$
!!
!! At the boundaries in the s-direction, we apply something similar to the
!! wills=1 scheme to do upwinding. Note that the routines that this calls to
!! deal with points near the boundaries do not presently work in second order.
!!
!! See jhg_interior for the main bit of these terms.
!!
!! ( Note that above we dropped the second index with respect to `j'(for \mu)
!!  in H; the second index is always j. Below we use `HH' for `H')
!<----------------------------------------------------------------------------
subroutine igh(disp_par,disp_vp)
...
... 
\end{verbatim}
{\bf APS: in the last two examples above, is the \code{-} character the one we
have decided to use to break up the code blocks in all cases? Only in the code we use at least \code{+,*}
as well...}
\subsection{Code blocks, line indentation, whitespace and continuation}

We do not allow TAB characters in the code. They may mean something in an extension of Fortran, so
could have unintended consequences. Also, they can make reading the source between different
editors difficult.

Blank lines may appear in the code (see comments section); they should contain no characters (including
whitespace characters). A line may be a blank comment line (e.g. in a comment block), but should
start with \code{!}. FC: Comments should indented inline with code.

\begin{itemize}
\item \code{module, subroutine, function, contains statements etc. begin in col 1} 
(unless they are functions inside subroutines?)
\item code is indented 2 columns from the program unit.  FC: I prefer 3. AGP: I prefer 2. YC: 2 for me 
\item \code{do,if,select case} require indent of 2 cols.  FC: I prefer 3. AGP: I prefer 2. YC: 2 for me. APS: so it is 0 or 2 and never 3?
\item Maximum line length. There is no need to use the old Fortran 77 limit. However, anyone
editing with emacs or vi would greatly appreciate it if most lines do not go beyond 78 cols
(I suspect most potential coders would use one editor or the other). What lines should go beyond the
limit? Lines that do not need to be read to understand what the code does? FC: I vote for more than 78.
APS: The original code used around the Fortran 77 limit. Since the style should be similar, we need not
usually use more than that. If we set a _hard_ limit around 78, and a suggest using less in general, that
already allows one to use several more cols. If we agree that this is too restrictive, why not set the hard
limit to 132 (as allowed by the standard). To make it clear - the main reason for 78 is that when remote
editing a file (such as when you do some tweaking to the source you are compiling on a cluster), it is usual
and convenient to run your editor in non-graphical mode. When doing so, lines longer than 78 are wrapped in
emacs and it can be awkward to work with. \textbf{YC: I would prefer the limit to be 132, but it is because I never run an editor in non-graphical mode...}
\item Long lines should be broken by placing a continuation character (\code{\&}) in the last col?
A continuation character should also appear in the next line, indented 4(?) cols in from the start
of the previous line. Lines which end part way through a character string should be terminated before
the continuation character, then concatenated with another string on the next line.
e.g. {\bf AGP: continuation characters should be placed in a readable place, for instance directly 
under the = in an assignment statement.APS: In the line below it could instead go under the opening brace?
Setting to a fixed number is easy for editors though - I don't have a macro to
automatically put it in the right place}
\begin{Verbatim}
call gkw_init('gkw_init: I do something related to reading files and'//      &
    &         ' allocating memory')
\end{Verbatim}
\end{itemize}

{\bf Need we discuss where whitespace is used in e.g. \code{if, do}?}

\subsection{Misc - unsorted ideas}

\begin{itemize}
\item \code{endif,enddo,endsubroutine} vs \code{end if, end do, end subroutine}; should we be consistent? 
AGP: Not a major point for me. APS: looking in the code, mostly the words are
split, apart from "endif" (roughly 4 times more common than "end if").
\item Should we use the operators $<, >, <=, >=, ==, /=$? They are easy to read and the
old ones are considered obsolete. AGP: Yes would be nice 
\item Should we avoid using multiple \code{return} statements? They are redundant at the end of subroutines
anyhow. Sometimes it is difficult to see which code gets reached if there are many. Any thoughts?
The only advantage is in an optimised routine when many more unnecessary tasks would otherwise be
performed (but these should be in sub, or subsub routines, right?) AGP: I have always considered multiple 
returns useful. Often there are special cases in which the job gets done with just a few lines. APS: certainly in
some optimal cases. Not needed at the end of subroutines or functions though. I'll refer other specific cases to you.
Some are really difficult to debug.
\item Should we use \code{()} for function and subroutine calls which have no variables to pass? FC: yes.
\item Numbered lines - are they needed? Formats in write statements are the only thing that use them, but
even that is not necessary. FC: neater and easier for (repeated) write statments. APS: Rather than have
numbered lines all over the place, I suggest using character strings instead. Even better, putting
all the formats in a module, so that they are done consistently throughout (and could potentially be
modified via a namelist). AGP: indeed numbered lines only for the write statements. Putting them in 
a module is not a bad idea. 
\item Do we allow \code{;} to e.g. put multiple assignments on one line or for multiple do loops etc.?
When and why?  FC: yes as it keeps routines shorter and more readable. AGP: After being initially against
it, I now have changed my mind. Indeed it does make the code a lot more readable. So yes. 
\item No hard coded numbers? - should use \code{parameter} instead?  FC: Clarify $10^-10$ not ``two''.  Does this include: \texttt{max_iterations}, \texttt{nmmx}? AGP : I prefer no hard coded numbers (although I am guilty) YC: what do we do for things like $2\pi$, etc.? We want to have it with the same precision everywhere in the code, no ?
\item Always use \code{implicit none} exactly once in each module. 
\item Array syntax. It should be kept simple in the optimal parts of the code, right? FC: Don't understand. APS: OK, when should we use (if ever) array syntax?
It is extremely concise and usually at least as efficient as doing it explicitly, but mistakes can be made if you try to be a bit tricky.
(when assigning with array syntax, use \code{A(:) = 0.}, rather than \code{A = 0.}? etc.) AGP : I agree use A(:) and preferably no array syntax a = b + c. It can be confusing
\end{itemize}

\subsection{Input files}

Input files are also Fortran files! (Only applies to files checked into the repository, such
as the test files and samples.)
(and we have had some problems with them! We use some non-standard features which could be avoided
by introducing simple guidelines for writing them.)
How about:

\begin{Verbatim}[commandchars=\\\{\}]
! This is a sample input file.
! Comments are allowed as in normal fortran while outside a namelist block.
! No non-Fortran chars allowed! (Including TABS)
       \textit{! Blank lines are allowed}

&CONTROL  \textit{! no need to be capitals; comments allowed this line; \code{\&} in col 1 only}
 some_variable = 1. \textit{! variables indented 1 space, comments only on same line}
 another_variable = 1, ! comment allowed here
       ! comment allowed here (unless the previous variable is a character, if that
	   ! is the correct interpretation of the standard) NO TABS!
 \textit{(some compilers apparently do not allow comments in the block at all;}
 \textit{ not sure the standard allows them on next line unless they start}
 \textit{col 1 and the previous line has no comma at the end..)}

/ \textit{! end of block starts in col 1}
\end{Verbatim}

in summary:
\begin{itemize}
\item Blocks start and end in col. 1
\item \code{var = val} starts in col. 2
\item No TABS
\item No comments after \code{char=val}. The comment can go on the next line or previous line.
\item If the line ends with \code{","} with no comment afterwards, the next line can not be a comment?
\item Stick to the same line length as for the fortran files (when we decide)
\end{itemize}

\subsection{Preprocessing}

We use C preprocessor directives in the code for the purpose of conditional compilation only.
They are used to enable/disable calls to optional libraries, such as MPI, FFTW, UMFPACK etc.
and also to get the right datatypes in those calls. Ideally, the actual calls to
these external libraries should limited to 1 module per library (this is already the
case for FFTW, and almost true for the MPI).

The compiler should see as much of the code as possible when testing, so we avoid
any other use. (One can use the timings module \file{PERF}, together with directives
for testing parts of the code, but putting these all over the checked in code make it messy?
Perhaps we could always call perf at the desired places, but either put a 
logical "perf" (which decides if the call actually does anything), or preprocess perf itself
with the list of perf names)

Allowed list:
\begin{verbatim}
#ifdef ONE || TWO
#elif defined(THREE)
#else
#endif
\end{verbatim}
(Do we need more?)
\subsection{Fortran best practices}

Do we need such a section? Most people know it is bad to put conditionals in the middle of 
optimised loops, that all cases should be covered by conditionals etc. {\bf APS: I have a list.
}. \textbf{YC: This would be useful to me. The little I know in Fortran is from copy-pasting bits of code. So I might very well have taken bad practices with it.
APS: Right, the list can begin with things we removed already.
Then whenever cleaning up, add more things to the list.
The old features can be summarised in another section again, although they can be found in most fortran books.}. We should always try to generalise bits of code though to encourage re-use, so
examples of generalisation could go here. Obsolete and old features that are not used in the code
could also be listed.
{\bf AGP : Optimized routines should be flagged as such. It is often very useful to program 
with readability in mind rather than performance. This should be encouraged in all the 
non-optimized routines. Only a few routines of the code need really be optimized for speed}

\subsection{Summary of main unresolved points}

\begin{description}
 \item Most things that were here are resolved or unimportant.
%  \item If \code{dimension} should be used or not.
%  \item What character(s) should be used to break up code blocks in comment lines.
%Presently, at least \code{-,*} and \code{+} are used.
%  \item The number of indentation whitespace characters used. We agreed on 2 in all cases, apart from the
%  start of a subroutine (no indent, or 2)? In the special case of continuation, is it always clear where
%  to put it on the next line? it could be omitted completely. Do we indent the code under each \code{case} in a \code{select case}?
%  \item The first continuation character should go at the end of the line, or not? (i.e. if we allow 78 cols it goes in col 78)
%  \item The number of allowed cols. 78 vs more?
  \item \code{endif} vs \code{end if}: see A.12
\end{description}

%\section{The previous section again}

%%%% ============================================================================ %
%%%% Index 
%%%% Put the indexes here for files, variables, then another for names,
%%%% defs, commands etc.
%%%%
%%%% ============================================================================ %

\end{document}

### 2022-05-03 gkw/doc/etc/doc_old.tex
(1) Added gkw Clone gwk  #3

%\documentclass[twocolumn,showpacs,preprintnumbers,amsmath,amssymb]{revtex4}
\documentclass{report}
\usepackage{fancyhdr}
\usepackage{fullpage}
%\pagestyle{fancy}
%\fancyhead{}
\fancyfoot[L]{$ $Date: 2009-11-26 16:34:54 +0000 (Thu, 26 Nov 2009) $ $}
\fancyfoot[R]{$ $Revision: 1279 $ $}
\usepackage[small,labelfont=bf,textfont=it,width=0.9\textwidth]{caption} 
\usepackage{bm}

\renewcommand{\vec}{\bm}
\usepackage{graphicx}
\graphicspath{{../fig/}}
\usepackage{amssymb}
\usepackage{makeidx}
\makeindex

\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\bee{\begin{eqnarray}}
\def\eee{\end{eqnarray}}
\def\l{{(0)}}
\def\f{{(1)}}
\def\minus{{\,\,\,\over \,\,\,}}
%
\def\black{\special     {ps: 0 0 0 setrgbcolor}}
\def\green{\special     {ps: 0 .5 0 setrgbcolor}}
\def\blue{\special      {ps: 0 0 1 setrgbcolor}}

%\def\black{}
%\def\blue{}
%\def\green{}

%\input{epsf}
%\newcommand{\epsfig}[3]{%
%    \begin{figure}[htbp]%
%        \begin{center}%
%            \setlength{\unitlength}{1.0cm}%
%            \begin{picture}(15,20)%
%                \epsfxsize=12cm%
%                \epsffile{#1}%
%                \end{picture}%
%            \end{center}%
%        \caption{#3}%
%        \{#2}%
%        \end{figure}%
% }

\title{GKW how and why}
\author{A.G. Peeters, D. Strintzi, Y. Camenen,\\
F.J. Casson, W.A. Hornsby, A.P. Snodin}
\date{Not maintained since CPC paper.  Please edit new version.}%\\
%\url{http://go.warwick.ac.uk/gkw}

\begin{document}

\def\compset#1{\vskip 0.2 truecm \noindent {\texttt #1 \hfill} \vskip0.2 truecm \noindent}

\pagenumbering{roman}
\maketitle
\cleardoublepage 
\centerline{\bf PREFACE} 
\vskip 1 truecm

This \blue documentation \black comes with the GKW code. This code is distributed with the hope that 
it is useful, but the authors take no responsibility for it what so ever.

There are several rules that apply to the use of this code 
\begin{itemize}

\item You are not allowed to change the name of the code under any circumstances, even 
if you use part of the code no matter how small. This also applies if you make large 
changes to the code or even if you completely rewrite it. Add ons to the name, like 
$GKW\_GA$ are not allowed either. If you modify the code you should refer to it as 
'a modified version of GKW'. You are encouraged to write what the modifications include 
or cite a paper where it is spelled out.

\item When GKW is used to produce results for a paper, you are obliged to cite the 
following references \hfill \break 
[1] A.G. Peeters, D. Strintzi, Phys. Plasmas, 11, 3748 (2004) \hfill \break
[2] A.G. Peeters, D. Strintzi, ...\hfill \break %PRL 2007?
[3] A.G. Peeters, Y. Camenen, F.J. Casson, W.A. Hornsby, A.P. Snodin, D. Strintzi, Computer Physics Communications, (2009) ...\hfill \break

The first is the original LINART code from which GKW was developed. In the second 
the code was used for the first time with major upgrades (kinetic electrons for instance) 
and the third is a general description of the code. We may change one of these citations
in the future. We promise to limit ourselves to 3 citations though (which should not stop 
you from citing any of our other excellent papers)

\end{itemize}

\noindent 
The restrictions above do not apply if the leading author (A.G.Peeters) gives explicit 
permission (He is unlikely to do so though).
The tone above is rather harsh, but you must realize that you can:

\begin{itemize}

\item Freely use the code for your own research purposes. You do not have to check with 
us or tell us what you are doing. (you are encouraged to do so, since we might be able 
to help you, and you might be able to help us)

\item Publish as many papers as you like without telling us.

\item Make as many changes as you want to the source without any restrictions or 
obligation to share the changes (again you are encouraged to send us the modifications 
you have made, since it will help to further develop the code.

\item Read in the manual what the code does and how it does it.

\item Read in the manual what are the restrictions on the physics model.

\end{itemize}

\cleardoublepage

\tableofcontents
\cleardoublepage

\pagenumbering{arabic}

\vfill \eject 
\hbox to 2 truecm{\hfill}
\vskip 8 truecm

\centerline{\selectfont \Huge I AM A USER} 
\addcontentsline{toc}{chapter}{I AM A USER}

\vfill  \eject

\chapter{I want to do a run}

Before you do any run it is important to realize that it is not all that hard to produce nonsense. 
The code is not under all conditions stable, and it is not always immediately obvious if a 
result is wrong. It is therefore important to diagnose your runs to make sure they are o.k.. 
It is your responsibility as a user to produce physically meaningful results. %See the section 'diagnose a run' to see what possibilities there are, and what to do if your run did not turn out quite the way you hoped it would.

\section{Controlling the code} %Added by FC March 2008, updated Nov 2008
\index{input file}
\index{namelists}
To run GKW you must provide an input file \texttt{input.dat}. Provided with the code is a \texttt{input.dat.sample} file which describes all the possible input variables.  The \texttt{input.dat.minimum} file shows the variables that must be provided.  Many of the input switches are optional and default values will be used if they are not provided.  There are also other example input files provided.  The input variables are provided in fortran namelists, for example the namelist `control' contains the main switches for the code. The namelist examples presented in this chapter provide you with a basic set needed to run the code.  The namelists are sensitive to bad formatting or typos, in which case the code should infrom you which namelist has the problem.

GKW tries to catch as many invalid input combinations as possible but there are probably many it can miss.  Most of the time if you input a set of incompatible switches the code will tell you and abort if necessary. For more minor errors the code will generate a warning that it is ignoring an input parameter and continue.  The actual input parameters the code runs with are written to input.out.  By renaming input.out to input.dat the code should run without the warnings.

\begin{footnotesize} \begin{verbatim}
&CONTROL
 NON_LINEAR = .false.      !Include the non linear terms, computationally expensive fft
 zonal_adiabatic = .false.,!If zonal flows corrections included for adiabiatic electrons
 order_of_the_scheme = 'fourth_order'	!or "second_order" in space		
 METHOD = 'EXP',        !Time integration "EXP" for explicit, "IMP" for implicit
 METH   = 2,            !Choice of algorithm for METHOD
 DTIM   = 0.02             !Time step size (normalized)
 NTIME  = 25,              !Number of iterations of NAVERAGE.
 NAVERAGE = 100,        !No of timesteps between re-normalisation, data output
 nlapar = .false.,      !true = keep the electromagnetic potential. false = electrostatic 
 collisions = .false.   !Turn on the collision operator
 disp_par = 1.0	        !Dissipation coefficient for parallel derivatives.
 disp_vp  = 0.0	        !Dissipation coefficient for parallel velocity space
 max_sec = 1000.        !Internal program kill time.
/
&GRIDSIZE
 NX = 83,        !Number of radial wave vectors - needs to be an odd number for fft
 N_s_grid = 50,  !Number of grid points along the field line 
 NPERIOD = 1,    !Integer - the field line makes 2*nperiod - 1 poloidal turns. 
 N_mu_grid   = 8,        !Number of magnetic moment grid points
 N_vpar_grid = 32,       !Number of grid points for parallel velocity 
 NMOD = 21,              !Number of binormal modes
 number_of_species = 1 	 !Number of kinetic species. 
/
\end{verbatim}\end{footnotesize}

\subsection{Timestep}
The timestep required for stability can vary greatly depending on input parameters or choice of scheme.  In particular kinetic electrons require a timestep $10^{-2}$ smaller.  Keeping the electromagnetic corrections for nonlinear runs is stabilising for low $\beta$ and can allow the timestep to be three times larger.  Clearly you should check the convergence of your results.

\subsection{Solver Methods}
The scheme can be either second order or fourth order in space.  The code has a choice of time integrator methods which can be used in the core solver as chosen by the input variable \texttt{METHOD}. The options are implicit (IMP) and explicit (EXP).  Within each method is a choice of algorithms chosen by the input variable \texttt{METH}. Currently the only safe options are:

\begin{itemize}
\item (METHOD, METH) = ('EXP', 1) Explicit - Modified mid point.
\item (METHOD, METH) = ('EXP', 2) Explicit - 4th order Runge Kutta.(RECOMMENDED)
\item (METHOD, METH) = ('EXP', 3) Explicit - 3rd order scheme.
\end{itemize}

The performance and stability of the different options varies and require careful usage.  Non linear runs can recalculate the time step size based on the Courant condition for numerical stability for the 4th order RK method.

\subsection{Parallelization}
\index{parallelization}
The code will parallelize first over the $\mu$ grid and (non-adiabatic) species, and then over the $v_{\parallel}$ and $s$ grid points.  The code will select what it considers to be the most appropriate parallelization given your input (this can be overridden by setting \texttt{N\_procs\_[sp|mu|vpar|s]} in the gridsize namelist).  Nevertheless it is sensible to have in mind a parallelization setup when you set the gridsize, as the user controls the number of processors the code will be run on and should choose a number that fits with the gridsize.

\subsection{Modes}
The spacing and location of the modes is determined by the `mode' namelist but is not immediately intuitive how this works.

For \texttt{mode\_box=.false.} there is only the zero radial mode. The toridal modes are determined by the input list \texttt{kthrho} . For {mode\_box=.false.} the toroidal modes are input as a list i.e. \texttt{kthrho=0.1,0.2,0.3} (The \texttt{krhomax} is ignored), and the radial mode is not coupled at the end of the field line.

For \texttt{mode\_box=.true.}, the field line has one poloidal turn \texttt{nperiod=1}, and the radial modes are coupled at the end of the field line according to the ballooning shift due to the magnetic shear.  The torodial modes will be equally spaced $k_\theta\rho =0$ to \texttt{krhomax} (The \texttt{kthrho} input is ignored).  See \ref{ikxspace} for further details (the integer $p$ described there is input into the code as \texttt{ikxspace}).  The spreadsheet \texttt{modebox\_calculator} enables quick testing of the relevant quantities without running the code.

\begin{footnotesize}\begin{verbatim}
&MODE
 KTHRHO  = 0.5,1.0,  !Poloidal wave vector(s) if mode_box=.false.
 mode_box = .false., !Determines if there is a 2D grid of ky,kx. if true use nperiod = 1
                     !Also determines if kx modes are coupled at end of field line
 krhomax = 1.0,      !For mode_box, the maximum k_theta rho_i (ky) used
                     !Evaluated on the low field side of the outboard midplane
 ikxspace = 7,       !Determines spacing between kx modes (radial resolution against radial box size)
\end{verbatim}\end{footnotesize}

\begin{table}[h!]
\begin{center}
\begin{tabular}{ccccc}
 NS & ikxspace & NX & NPERIOD & mode\_box \\
\hline
 80      &    N/A   &  1  &    3    & false    \\
 16      &     1    &  5  &    1    & true     \\ 
\hline
 $n_s.(2n_p-1)$ &    N/A   &  1  &    $n_p$    & false    \\
 $n_s$    &     1    & $2n_p-1$ &  1   & true   \\

\end{tabular}
\caption{Equivalent formulations example with \texttt{mode\_box} true and false}
\end{center}
\end{table}

\subsubsection{Generating a linear growth rate spectrum}
GKW currently outputs the global growth rate in \texttt{time.dat} so the easiest way to generate a growth rate spectrum is do a batch of linear runs with \texttt{NMOD=1} and take the growth rate from each. If \texttt{mode\_box=.true.} and you have mode than one raidial mode, scan over \texttt{krhomax} to set the toroidal wavevectors (to keep the kx spacing the same you also have to vary ikxspace with krhomax); otherwise with {mode\_box=.false.} and \texttt{nx=1}, scan over \texttt{kthrho}.  The matlab scripts \texttt{create\_gkwscan} and \texttt{growth\_rate\_spec} help with creating the input and plotting the output; script \texttt{gkwnlin} runs the batch jobs.

\subsection{Species}
There is no default species data, hence you must provide some. The species you input must satisfy quasi-neutrality. This requires you to set ratios of densities and charge appropriately.  You can input multiple ion species but only one negatively charged species is allowed.  Note that the number of species read in is controlled by the number of species parameter in the `gridsize' namelist and hence it is possible that later species data you specify will not be read in if this is set incorrectly.  The \texttt{number\_of\_species} input variable does not include adiabatic electrons but the species data for these will still be read if this the case.

\begin{footnotesize}
\begin{verbatim}
 &SPCGENERAL
  beta = 0.000,                !Plasma beta (not used if nlapar = .false.)
  adiabatic_electrons = .true. !kinetic electrons require a smaller timestep
 /
 &SPECIES
  MASS  = 1.0,  !Species mass in terms of reference mass
  Z     = 1.0,  !Species charge. If negative, assumed to be the electrons.
  TEMP  = 1.,   !Temperature of species scaled by reference temperature 
  DENS  = 1.,   !Density of species scaled by reference density 
  rlt   = 8,    !Temperature gradient R/LT
  rln   = 3.5,  !Density gradient R/Ln
  uprim = 0.0,  !Gradient of the toroidal rotation R^2 grad Omega / v_thref
 / 
 &SPECIES 
  MASS  = 2.72e-4,  !Electrons
  Z     = -1.0,     !Only one negatively charged species is allowed.
  TEMP  = 1.0, 
  dens  = 1.0,
  rlt   = 7.9,
  rln   = 3.5, 
  uprim = 0.0,
 /
\end{verbatim}
\end{footnotesize}

\section{Geometry} 
GKW is written in general coordinates so that in principle the code can deal with any toroidally symmetric 
equilibrium. This means that all the equations are expressed with the use of general tensors, and 
the metric elements. Currently the geometry dependent quantities are calculated in the code only for the $\hat s - \alpha$ equilibrium.  In addition, GKW can read CHEASE \cite{LUT96} output for
arbitrary flux
surface configurations. CHEASE takes an input of the shape of the last closed flux surface, a pressure gradient profile (and a boundary pressure), and a current profile. From these inputs, CHEASE
solves the Grad-Shafranov equation to find an equilibria flux surface configurations which it outputs as the poloidal flux. See the documentation file \texttt{gkwandchease.tex} for more information.

The simplest geometry is the 
$\hat s -\alpha$ geometry which consists of circular flux surfaces which are shifted with respect to 
each other by a magnitude determined by $\alpha$ which is the MHD parameter 
\be 
\alpha = -q^2R{d\beta \over dr}
\ee
\blue (As currently implemented in the code $\alpha = 0$).\black

The only other parameters that determine this equilibrium is the minor radius $r$ of the flux surface,
the safety factor $q$, and the magnetic shear $\hat s$. The minor radius is expressed in terms of the inverse 
aspect ratio $\epsilon$ 
\be 
\epsilon = {r \over R_{\rm ref}}  = {\tt eps}
\ee
where the reference major radius ($R_{\rm ref}$) is taken to be the major radius of the center of the
flux surface. 
In the equation above the name of the input variable is given in typewriter font. 
The shear follows from  
\be 
\hat s = {r \over q} {{\rm d} q \over {\rm d} r}  = {\tt shat} 
\ee
These parameters must be entered in the input file (which is always called input.dat) in the namelist 
called `geom'.

\begin{footnotesize}\begin{verbatim}
&GEOM
 shat = 1.0,
 q    = 2.0,
 eps  = 0.1
 /
&ROTATION
 VCOR = 0.0	!Rotation of the plasma vcor = Vtor / vthref = R Omega / vthref
 /
\end{verbatim} 
\end{footnotesize}

\section{Comparison with other codes} %Added Nov 2008 by FC
If you are trying to use GKW to compare to a result of another code, then you must understand the correspondance between the normalizations.  For many codes the thermal velocity is normalized differently and hence GKW's $k_{\theta}\rho_i$ may be a factor of $\sqrt{2}$ larger.  Note also that \texttt{rlt} in the input file is the value of $R/L_T$ and \texttt{rln} is $R/L_n$.

Here we outline how to compare GKW results with codes that use gyro-Bohm units.
In the case where $T_e=T_{\rm ref}$ and $m_i=m_{\rm ref}$ the sound speed $c_s=\sqrt{T_e / m_i}$
converts to the GKW reference velocity as $v_{\rm thref}=\sqrt{2} c_s$ and the gyro-Bohm Larmor
radius $\rho_s=c_s/\Omega_{ci}$ is related to the GKW reference Larmor radius by $\rho_s=\rho_{\rm
ref}/\sqrt{2}$. For full details see section \ref{normalization}.

Hence the GKW time normalization may be converted:
\begin{equation}
t_N = {v_{\rm thref} \over R} t = \sqrt{2}{c_s \over R} = \sqrt{2} \left({a \over R}\right) \left({c_s \over a}\right) t
\end{equation}
This time normalization is used in all GKW output of times and rates and this must be accounted for when comparing results with other codes.  %Also set B_N to 1 ????

The fluxes diagnostic outputs the heat fluxes ${\texttt eflux} = q_i^N$ which may be converted to a heat diffusivity
\begin{equation}
\chi_i^N={q_i^N \over R/L_T}={R \over \rho_{\rm ref}^2 v_{\rm thref}} \chi_i = {1 \over 2\sqrt{2}}\left({R \over a}\right)\left({a \over \rho_s^2 c_s}\right) \chi_i 
\end{equation}
The particle and momentum fluxes can be converted similarily.  For further details see section \ref{fluxes}.

%\section{Setting up a run and initial checks}
%
%For linear runs should have symmetric in s.  Stable should have a smooth solution in s
%Check convergence of growth rates and fluxes ratios

%For kinetic runs need kxmax 3*kymax.  Expect ion fluxes 3 x bigger than for adiabatic and 3 x bigger that electron fluxes
%For nonlinear runs check fluxes saturate, want range in fluxes <80% - use minimum dissipation that gives this.
%For kinetic runs use nlapar=.true. and beta 3e-4 with timestep 0.003
%Use an odd number for n_s_grid if you want a point in the middle of the field line (but for parallelisation in s should still have some factors!)
%Choose ikxspace to choose get balance between x resolution and box size - usually aiming for a square box and kxmax = 3x krhomax (in rho ref units)
%Check the parallel output - should be not too noisy for stable runs.
%Check the ky spectrum is well covered by the modes
%Check the linear growth rate spectrum is well covered by the modes.
%Nonlinear: check the timestep didn't collapse.
%suitable dissipation coeffcients.  
%guidline suitable (minimum and not necessirly well converged!) grid sizes for kinetic ITG / TEM cases. nmod=21, nx=167, krhomax=1.41, ikxspace=6, ns=16, nvpar=32, nmu=8, dtim=0.003, disp_par=disp_vp=1.5
%particle fluxes should be zero for a single species, should be same for two species
%time averaged momentum flux should be zero for no s-alpha, no shear, no uprime.

\cleardoublepage

%\pagenumbering{arabic}

\vfill \eject 
\hbox to 2 truecm{\hfill}
\vskip 8 truecm

\centerline{\selectfont \Huge I AM A PHYSICIST} 
\vskip 8 truecm 
\blue Read this section alongside the CPC paper which may have some corrections.\black
\addcontentsline{toc}{chapter}{I AM A PHYSICIST}

\vfill  \eject

\chapter{Geometry}

\section{Field aligned Hamada coordinates} 
\index{coordinates, flux tube}

In this chapter we will work out the coordinates attached to the magnetic field geometry. 
Several functions will be defined that are used in the gyro-kinetic equation. The 
development of the material in this section has benefited a lot from the write-ups 
of Dr. B.D. Scott \cite{Scott}

We start with an arbitrarily shaped, but toroidally symmetric geometry. We assume that 
we have a semi-orthogonal coordinate system $(\psi,\theta,\phi)$, where $\psi$ is the 'radial'
coordinate (i.e. ${\bf B}\cdot \nabla \psi = 0$), $\theta$ is the poloidal angle (upward 
on the outboard midplane), and $\phi$ is the toroidal angle (clockwise when viewed from 
above). Furthermore, we will assume that the magnetic field and quantities like the major 
radius are known in this coordinate system. Below we go through two coordinate transformations. 
The first will make the field lines straight in the new coordinates, and the second will 
align one of the coordinates with the magnetic field.

In the first step we transform the poloidal and toroidal angle 
\be 
s = s(\psi,\theta), \qquad \gamma = \gamma(\psi,\theta,\phi)
\ee 
in such a way that the contra-variant components of the magnetic field are flux functions 
in the new coordinate system ($B^s = B^s(\psi)$, $B^\gamma = B^\gamma(\psi)$). These 
coordinates are known as Hamada coordinates \cite{HAM62}\index{Geometry:Hamada coordinates}. 
Note that the new coordinate $\gamma$ is still an ignorable coordinate, since any function that is independent of $\phi$
will be independent of $\gamma$ ($f(\theta,\psi) \rightarrow f(s,\psi)$) after the coordinate 
transformation. The transformation of the poloidal angle is chosen such that $B^s$ becomes 
a flux function 
\be 
B^s = {\bf B} \cdot \nabla s = {\bf B}\cdot \nabla \theta {\partial s \over \partial \theta} 
+ {\bf B} \cdot \nabla \psi {\partial s \over \partial \psi} = {\bf B} \cdot \nabla \theta 
{\partial s \over \partial \theta}
\ee 
From which the relation between $s$ and $\theta$ can be derived 
\be
{\partial s \over \partial \theta}={B^{s} \over {\bf B \cdot \nabla \theta}}\Leftrightarrow
\int_{0}^{\theta}{\partial s \over \partial \theta}d\theta =s=B^{s}\int_{0}^{\theta}{d\theta \over 
{\bf B}\cdot \nabla \theta}
\ee
since $B^{s}$ does not depend on $\theta$ (flux function) and
\be
\oint {\partial s \over \partial \theta}d\theta =1=B^{s}\oint {d\theta \over {\bf B}\cdot \nabla 
\theta}\Leftrightarrow B^{s}={1 \over \oint {d\theta \over {\bf B}\cdot \nabla \theta} }
\ee
which gives the following equation for s:
\be 
s(\theta,\psi) = {\int_0^\theta {{\rm d} \theta^\prime \over {\bf B} \cdot \nabla \theta^\prime} 
\biggl /  \oint {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime}} \qquad 
{\rm with} \qquad 
B^s =  {1
\biggl /  \oint {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime}}
\ee  
where the normalizing constant has been chosen such that the domain [-1/2,1/2] in $s$ corresponds 
to one poloidal turn.

The new coordinate is closely related to the flux surface average. This average is defined 
as the volume average between two flux surfaces in the limiting case in which the distance between 
the flux functions goes to zero. One can derive the following relation  
\be 
\langle g \rangle = \oint {{\rm d} l_p \over B_p } g \biggl / \oint {{\rm d} l_p \over B_p }
\ee
where $l_p$ is the length in the poloidal direction (${\rm d}l_p = \vert {\bf e}_\theta \vert 
{\rm d} \theta$), $B_p$ is the poloidal component of the magnetic field strength ($B_p = 
{\bf B}\cdot \nabla \theta / \vert \nabla \theta \vert$), and the integral is to be taken at 
constant $\psi$. Inserting the expressions for the 
poloidal length and the poloidal magnetic field strength and using $\vert {\bf e}_\theta\vert 
\vert \nabla \theta \vert = 1$ one obtains 
\be 
\langle g \rangle = \oint {\rm d}s g 
\ee

For the transformation of the toroidal angle we choose 
\be 
\gamma = {\phi \over 2 \pi} + g(\theta,\psi) 
\ee
The $2\pi$ is added here such the domain [0,1] in $\gamma$ will correspond to one toroidal 
turn around the flux surface. The contra-variant component of the magnetic field is 
\be 
B^\gamma = {F \over 2\pi} {1\over R^2} + {\bf B} \cdot \nabla \theta {\partial g \over \partial 
\theta} 
\ee 
where $F = R B_t = F(\psi)$ is a flux function. (Here $B_t$ is the toroidal magnetic field 
strength $B_t = {\bf B} \cdot \nabla \phi / \vert \nabla \phi \vert$ where $\vert \nabla \phi 
\vert = 1/R$ with $R$ being the local major radius)
From the expression above one can derive 
\be 
g(\theta,\psi) = \int_0^\theta {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime } 
\biggl [ B^\gamma - {F \over 2\pi} {1 \over R^2} \biggr ] 
\ee 
Demanding that $g$ is periodic
\be
\oint {{\rm d}\theta^\prime \over {\bf B}\cdot 
\nabla \theta^\prime } 
\biggl [ B^\gamma - {F \over 2\pi} {1 \over R^2} \biggr ] =0
\ee
yields 
\be 
B^\gamma = {F\over 2 \pi } \biggl \langle {1\over R^2} \biggr \rangle 
\ee 
The coordinate $\gamma$ can then be expressed as 
\be 
\gamma = {\phi \over 2\pi} + {F \over 2\pi} \int_0^\theta {{\rm d} \theta^\prime \over {\bf B}
\cdot \nabla \theta^\prime} \biggl [ \biggl \langle {1\over R^2} \biggr \rangle - {1\over R^2} 
\biggr ] 
\ee 
The coordinates $(\psi,s,\gamma)$ are the Hamada coordinates. The expression above should 
allow these coordinates to be calculated for an arbitrary toroidally symmetric geometry.

We now go through the second coordinate transformation where one of the coordinates is 
aligned with the magnetic field, i.e. we demand 
\be 
{\bf B} \cdot \nabla = B^s {\partial \over \partial s}
\ee
Note that this does not mean that $\nabla s$ is in the direction of the magnetic field. 
In general it is not. The transformation is 
\be 
\zeta = \zeta(s,\gamma,\psi) 
\ee
In the new coordinates 
\be
{\bf B}\cdot \nabla =B^{\psi}{\partial \over \partial \psi}+B^{s}{\partial \over \partial s}+
B^{\zeta}{\partial \over \partial \zeta}
\ee
where $B^\psi$ is zero. The transformation to field aligned coordinates is the transformation 
for which $B^\zeta = 0$. The latter condition can be formulated as 
\be 
B^\gamma {\partial \zeta \over \partial \gamma} + B^s {\partial \zeta \over \partial s} = 0
\ee 
which is satisfied for the simple linear transformation
\be 
\zeta =  q s - \gamma \qquad {\rm where} \qquad q = {B^\gamma \over B^s} 
\ee
From the field line equation it follows that 
\be 
{{\rm d} \gamma \over {\rm d} s} = {B^\gamma \over B^s} \qquad \rightarrow \qquad 
\oint {{\rm d} \gamma \over {\rm d} s} {\rm d} s = \oint {B^\gamma \over B^s} {\rm d} s = q 
\ee 
i.e. $q = q(\psi)$ is indeed the safety factor.

Note that the coordinate transformation above flips the sign of the toroidal angle. 
The right handed coordinate system can therefore be defined as $(\psi,\zeta,s)$. 
The Jacobian of the new coordinate system can be expressed in terms of the original 
Jacobian through 
\be 
(\nabla \psi \times \nabla \zeta) \cdot \nabla s = {1 \over J_{\psi  \zeta s} } = 
{1\over 2 \pi}{\partial s \over \partial \theta} {1 \over J_{\psi \theta \phi}} 
\ee 
From which it follows 
\be 
J_{\psi  \zeta s} = 2\pi {\bf B} \cdot \nabla \theta \oint {{\rm d} \theta^\prime \over 
{{\bf B} \cdot \nabla \theta^\prime}} J_{\psi \theta \phi} 
\ee

We note here that the coordinates $s$ and $\zeta$ are dimensionless, but that the 
dimension of $\psi$ is not yet defined. We will always use a normalized $\psi$ that 
is dimensionless (see \ref{normalization} for a discussion on the normalization)

\section{Metric elements} 
%Some updates to this section in the CPC paper, check for consistency here.
\section{Short look ahead} 
\index{geometry tensors}
So far we have introduced the coordinates. Of course, to solve the gyro-kinetic 
equation we will have to calculate some specific quantities in the general 
coordinate system. In the next chapter these will be introduced. To proceed 
with the discussion of the geometry we define them here 
\be 
{\cal F} = {R_{\rm ref} B^s \over B} \qquad \rightarrow \qquad 
R_{\rm ref} {\bf b} \cdot \nabla  = {{\cal F}} {\partial \over \partial s} 
\ee
\be 
{\cal D}^\alpha = {R_{\rm ref}^2 {\bf B} \times \nabla B \over B^2 B_N} \cdot \nabla x_\alpha 
\ee 
\be 
{\cal E}^{\alpha \beta} = {R_{\rm ref}^2\over 2 B_N} (\nabla x_\alpha \times \nabla x_\beta) \cdot {\bf b} 
\ee 
\be 
{\cal G} = {R_{\rm ref} {\bf B } \cdot \nabla B \over B^2} %%={\cal F} {\partial ln B \over \partial s} ??
\ee
\be 
{\cal H}^\alpha = {R_{\rm ref}^2 \over v_{\rm thref}}{\bf \Omega}_\perp \cdot \nabla x_\alpha 
\ee
\be 
{\cal I}^\alpha = {R_{\rm ref}^2 \Omega^2 \over 2 v_{\rm thref}^2 B_N } (\nabla x_\alpha \times \nabla R^2) 
\cdot {\bf b} 
\ee
In these equations $R_{\rm ref}$ is the reference value of the major radius, i.e. it is 
simply a normalizing constant. $B_N$ is the magnetic field normalized to a reference value 
$B_{\rm ref}$, $v_{\rm thref}$ is a reference (thermal) velocity, and $\Omega$ is the angular 
(toroidal) rotation frequency of the plasma. In the next sections these coefficients will be 
worked out for different geometries.  Note that for benchmarking against other gyrokinetic codes the $B_{N}$ may need to be approximated to 1. \blue You should check your version of the code in the
geom module, as this  may have been done. \black
Note that the sign convention is $\Omega>0$ for a plasma flowing in the direction of $\mathbf{B}$, which  means that $\mathbf{\Omega}=-\Omega\mathbf{\nabla}Z$ for $\mathbf{B}\cdot\mathbf{\nabla \phi}>0$ and $\mathbf{\Omega}=\Omega\mathbf{\nabla}Z$ for $\mathbf{B}\cdot\mathbf{\nabla \phi}<0$.

\section{$s-\alpha$ geometry} 
\index{$s-\alpha$}
The simplest choice of the geometry is the $s-\alpha$ model. This model assumes circular 
flux surfaces which can be shifted with respect to each other. The inverse aspect rato $\epsilon = r / R$ 
is assumed small, with $R=R_{\rm ref}$, and only the lowest order in an $\epsilon$ expansion is kept. 
The poloidal magnetic field is assumed much smaller than the toroidal magnetic field
\be 
B \approx |B_t| \qquad 
q \approx s_{\rm B}s_{\rm j} {r |B_t| \over R |B_p|} 
\ee 
where the signs of the magnetic field ${\bf B}$ and current density ${\bf j}$ are defined as $s_{\rm B}=\textrm{sign}(\bf{B}\cdot\nabla\phi)$ and $s_{\rm j}=\textrm{sign}(\bf{j}\cdot\nabla\phi)$
with $\phi$ increasing clockwise when the torus is viewed from above. 
The coordinates can be approximated as 
\be 
\psi = {r \over R}, \qquad 
\zeta =  {1\over 2 \pi} [q \theta - \phi ] \qquad 
s = {\theta \over 2 \pi }  \qquad 
\ee
The gradients are approximated as  
\be 
\nabla \psi = {1\over R} {\bf e}_r \qquad 
\nabla \zeta = {s_{\rm B}s_{\rm j}|q| \over 2 \pi r} [ {\bf e}_\theta + \hat s \theta {\bf e}_r ]  \qquad 
\nabla s = {1 \over 2 \pi r} {\bf e}_\theta
\ee 
Note that the gradient of the toroidal angle is ordered small compared with the gradient of the poloidal
angle (times q) and is neglected in the gradient of $\zeta$. The contravariant components of the 
magnetic field are 
\be 
B^s = s_{\rm j}{|B_p| \over 2 \pi r} \qquad 
B^\zeta = -{s_{\rm B}|B_t| \over 2 \pi R}  
\ee 
The gradient of the magnetic field strength is assumed to be in the direction of $\nabla R$ and the 
drift term is approximated as 
\be 
{\bf B} \times \nabla B = -s_{\rm B}{B^2 \over R} [ \cos \theta {\bf e}_\theta + \sin \theta {\bf e}_r ] 
\ee

With this expression one can evaluate the quantities needed for the numerical computations 
\be 
{\cal F} = {s_{\rm j}\over 2 \pi |q|} 
\ee
\be 
{\cal D^\psi} = - {s_{\rm B}\over B_N} \sin (2 \pi s)
\ee 
\be 
{\cal D}^\zeta = - {s_{\rm j}|q| \over 2 \pi \epsilon B_N} [ \cos (2 \pi s) + 2 \pi \hat s s \sin (2 \pi s) ] 
\ee
\be 
{\cal D}^s = -{s_{\rm B}\over 2 \pi \epsilon B_N} \cos ( 2 \pi s) 
\ee
\be 
{\cal G} =  {s_{\rm j}\epsilon \over |q|} \sin ( 2 \pi s) 
\ee
\be 
{\cal E}^{\psi \zeta} = {s_{\rm j}|q| \over 4 \pi \epsilon B_N}
\ee
\be 
{\cal E}^{\psi s} =  {s_{\rm B} \over 4 \pi \epsilon B_N}
\ee
\be 
{\cal E}^{\zeta s} =  { s_{\rm j}|q| \over 4 \pi \epsilon^2 B_N } \hat s s 
\ee
Finally 
\be 
R^2g^{\zeta \zeta} = \biggl ( {q \over 2 \pi \epsilon} \biggr )^2 + \biggl ( {q \over\epsilon} \hat{s} s \biggr )^2 \qquad 
\ee
\be 
R_{\rm ref}^2 g^{\zeta \psi} = {s_{\rm B}s_{\rm j}|q| \over \epsilon} \hat s s 
\ee 
\be  
R_{\rm ref}^2 g^{\psi \psi} = 1 
\ee 
%{\cal E}^{\psi s} and {\cal D} not used in code, not included in CPC paper

\section{Shifted circular surface geometry}

The shifted circular surface geometry is defined in the ($R,Z$) plane (using cylindrical coordinates
$(R,\phi,z)$. 
\be 
R = r \cos \theta + \Delta(r) + R_0 \qquad {\rm with} \qquad \Delta(0) = 0 
\ee
\be 
Z = r \sin \theta
\ee
Surfaces of constant $r$ are assumed to be flux surfaces. One can confusingly define two major 
radii. One is the radius of the magnetic axis $R_0$, and one is the major radius of the centre 
of the surface $R_c$ 
\be 
R_c = (R(\theta =0) + R(\theta = \pi) )/2 = R_0 + \Delta(r) 
\ee
Since the first one is a constant while $R_c$ depends on $r$, we will choose to normalize all 
the radii with $R_0$, i.e. 
\be 
R_{\rm ref} = R_0
\ee
The normalized radial coordinate then is 
\be 
\psi = r / R_0 = \epsilon
\ee
and we will define a normalized major radius that gives the centre of the surface as 
\be 
R_{cN} = 1 + {\Delta (r) \over R_0} 
\ee

From the equations above one can work out
\be 
\biggl [ \matrix{ \nabla r \cr \nabla \theta} \biggr ]  = {1\over r (1 + \Delta^\prime \cos \theta)} 
\biggl [ \matrix{ r \cos \theta & r \sin \theta \cr - \sin \theta & \Delta^\prime + \cos \theta} 
\biggr ] \biggl [ \matrix{ \nabla R \cr \nabla Z } \biggr ] 
\ee
where 
\be 
\Delta^\prime = {{\rm d} \Delta \over {\rm d} r} 
\ee
The poloidal magnetic field can be written as 
\be 
{\bf B}_p = B_p {{\bf e}_\theta \over \vert {\bf e}_\theta \vert} \qquad 
{\bf e}_\theta = {\partial {\bf x} \over \partial \theta} = \biggl [ \matrix{-r \sin \theta \cr r \cos \theta}
\biggr ]  
\ee 
From which it folows that 
\be 
{\bf e}_\theta \cdot \nabla \theta = 1 \qquad {\bf B} \cdot \nabla \theta = B_p / r 
\ee 
From flux conservation 
\be 
R B_p {\rm d}l_\perp = {R B_p \over \vert \nabla r \vert } {\rm d} r = {\rm constant} 
\ee
It follows that  
\be 
{B_p \over B_p(\pi/2)} = {R_{cN}(\psi) \over (1 + \Delta^\prime(\psi) \cos \theta) (R_{cN}(\psi) + \psi \cos \theta)}
\ee
The integral to obtain $s$ and $B^s$ can be easily performed 
\be 
B^s = {B_p(\pi/2)\over 2 \pi ( 1 + \psi \Delta^\prime / 2 R_{Nc} )}
\ee
\be 
s = {\theta \over 2 \pi} + {2 \psi \sin \theta + 2 \Delta^\prime R_{cn} \sin \theta + \Delta^\prime \psi 
\cos \theta \sin \theta \over 4 \pi R_{cn} ( 1 + \Delta^\prime \psi / 2 R_{cn})}
\ee
After some manipulations also the flux surface average of $1/R^2$ can be expressed as 
\be 
\biggl \langle {1 \over R^2} \biggr \rangle = {1\over R_0^2} {1 \over R_{cN} + \psi \Delta^\prime / 2} 
\biggl [ {1 - \Delta^\prime \over \sqrt{R_{cN}^2 - \psi^2} } + {\Delta^\prime \over \psi} \biggr ] 
\ee

\section{General geometry}
The metric elements for general toroidal MHD equilibria can be obtained from the CHEASE code \cite{LUT96}. The sign of the magnetic field and plasma current are not taken into account into CHEASE
which always consider that ${\bf B}$ and ${\bf j}$ are in the direction opposite to $\nabla \phi$ (here
$\nabla \phi = \nabla \phi_{\rm GKW}$, as everywhere in this document). The sign of ${\bf B}$ and ${\bf j}$ are therefore specified in GKW, in the \texttt{GEOM} namelist
using the variables \texttt{SIGNB} and \texttt{SIGNJ} defined as in the previous section by $s_{\rm B}=\textrm{sign}(\bf{B}\cdot\nabla\phi)$ and $s_{\rm j}=\textrm{sign}(\bf{j}\cdot\nabla\phi)$. When
CHEASE is used, the absolute value of the safety factor and magnetic shear (used for the periodicity condition and to set the radial wave vector grid) is taken from the equilibrium reconstruction and
the values specified in the namelist \texttt{GEOM} are not read. The values of $\beta'$ is also calculated consistently with the equilibrium reconstruction.

\section{Summary of the sign conventions in GKW} 
\label{signs}
In GKW, the cylindrical coordinate system $(R,Z,\phi)$ is right handed, which means that $\phi$ is increasing clockwise when the torus is viewed from above.\\
The toroidal rotation is defined positive for a plasma flowing in the direction of ${\bf B}$.\\
The mode frequency is defined positive for a perturbation evolving in the direction opposite to $\nabla \zeta$. This corresponds to the ion $\nabla B$ drift direction if $s_{\rm j}=1$ and to the
electron $\nabla B$ drift direction if $s_{\rm j}=-1$.  See also Fig.~\ref{directions}\\
Coordinate system:
\begin{itemize}
 \item $\psi=\varepsilon=(R_{\rm max}-R_{\rm min})/(2R_{\rm ref})$ is always increasing from the plasma center to the plasma edge.
 \item $s$ is always increasing upwards from the low field side midplane. It is zero at the height of the magnetic axis, on the low field side midplane.
 \item $\zeta$ is always increasing in the direction opposite to $\phi$ (i.e. anticlockwise when viewed from above) at constant $\psi$ and $s$. The direction of $\nabla \zeta$ in the poloidal plane is
given by $\textrm{sign}(\nabla s \cdot \nabla \zeta)=\textrm{sign}(\nabla 
\zeta \cdot \nabla \theta)=s_{\rm B}s_{\rm j}$
\end{itemize}

\bee
\textrm{sign}({\bf B} \cdot \nabla \phi)&\equiv& s_B \\
\textrm{sign}({\bf j} \cdot \nabla \phi)&\equiv&s_j \\
\textrm{sign}({\bf B} \cdot \nabla \theta)&=& s_j \\
\textrm{sign}({\bf B} \cdot \nabla s)&=& s_j \\
\textrm{sign}(\nabla \phi \cdot \nabla \zeta)&=& -1 \\
\textrm{sign}(\nabla s \cdot \nabla \theta)&=& 1\\
\textrm{sign}(\nabla \theta \cdot \nabla \zeta)&=& s_B s_j \\
\textrm{sign}(B_\theta \nabla \theta \cdot \nabla \zeta)&=& s_B \\
\textrm{sign}(\nabla s \cdot \nabla \zeta)&=& s_B s_j \\
{\bf B} \cdot \nabla \zeta&=& 0 \\
{\bf \Omega} &=& -s_B \Omega \nabla z \\
u &=& R \Omega 
\eee
%The direction of zeta is flipped by both sign_j reversal and sign_B reversal.

\begin{thebibliography}{10}

\bibitem{Scott} B.D. Scott, private communications and web page 
www.rzg.mpg.de/~bds

\bibitem{HAM62} S. Hamada, Kakuyugo Kenkyu {\bf 1}, 542 (1958)

\bibitem{LUT96} H. L\"utjens, A. Bondeson and O. Sauter, Comput. Phys. Commun.{\bf 97}, 219 (1996)

\end{thebibliography}

\chapter{The gyro-kinetic set of equations }

This chapter deals with the basic equations that  are solved in the code.

\section{The gyro kinetic framework}

The starting point of the equation for the evolution of the distribution function is the 
gyro-kinetic theory formulated in Refs \cite{LIT83,HAH88,BRI88,SUG00}. In GKW these equations 
are solved up to first order in Larmor radius over the major radius ($\rho_*$), which 
is assume to be a small quantity. In order to determine which of the terms in the 
original formulation of the theory (which is accurate up to second order in $\rho_*$) 
must be kept, we will here briefly discuss the ordering.

A thermal velocity ($v_{\rm th}$) and a major radius ($R$) are used to normalize the length and 
time scales. For the gradients we assume 
\be 
R \nabla_\parallel  \approx 1 \qquad 
R \nabla_\perp \approx 1/ \rho_*
\ee 
i.e. the solution is assumed to have a length scale of the order of the system size along the 
field, but can have a structure of the order of the Larmor radius perpendicular to the 
magnetic field. This difference in length scales makes that we will evaluate the velocity 
along the field in lowest order only (i.e. of the order of $v_{\rm th}$), but have to go to 
first order (order $\rho_* v_{\rm th}$) for the perpendicular motion. Both the perturbations in 
the electro static potential $\phi$ as well as the vector potential are retained. For the 
latter it is, however, assumed that 
\be 
{\bf A} = A_\parallel {\bf b} 
\ee
i.e. no compressional effects are kept in the magnetic field evolution. 
Note that the vector potential above is the perturbed vector potential not to be confused 
with the one connected with the equilibrium magnetic field. 
The potential and the vector potential are taken to be of the order 
\be 
\phi \approx {T \over e} \rho_* \qquad 
A_\parallel \approx {T \over e v_{\rm th}} \rho_* , 
\ee
where $T$ is the temperature, and $e$ is the elementary charge.  
From this assumption it can be derived that the ${\bf E} \times {\bf B}$ velocity is of the order $\rho_* v_{\rm th}$, 
i.e. no higher order corrections must be kept. A similar argument applies to the tilting of 
the field line due to the perturbed vector potential.

Gyro-center coordinates (${\bf X},\mu,v_\parallel$) are used, with ${\bf X}$ being the gyro
center position, $v_\parallel$ the parallel velocity, and $\mu$ the magnetic moment 
\be 
\mu = {m v_\perp^2 \over 2 B} 
\ee 
where $v_\perp$ is the velocity component perpendicular to the magnetic field ${\bf B}$, $m$
is the particle mass and $B$ is the magnetic field strength. The magnetic moment is an 
invariant of motion, but the parallel velocity changes due to the electric field acceleration 
and the mirror force. It turns out that for consistency one needs to keep both the lowest 
as well as the first order (in $\rho_*$) contribution to the acceleration.

Finally, also the approximation known as $\delta f$ approximation will be made. This approximation
assumes that the perturbed distribution $f$ is much smaller than the background distribution 
$F$, but has a length scale of the order of the Larmor radius whereas the background varies 
on the length scale of the system size 
\be 
f \approx \rho_* F \qquad 
{\partial f \over \partial v_\parallel} \approx \rho_* {\partial F \over v_\parallel}  \qquad 
\nabla f \approx \nabla F 
\ee
These relations mean that what is known as the velocity non-linearity (combinations of the 
electromagnetic fields and the velocity space derivative of the perturbed distribution) can
neglected in the evolution equation for the perturbed distribution.

With the help of these approximations the gyro-kinetic equation of Refs. \cite{LIT83,HAH88,BRI88,SUG00}
can be rewritten. Starting point is the Lagrangian \index{Lagrangian}
\be
\gamma=\left({e\over c}{\bf A}+{e\over c}A_{\parallel}{\bf b}+m{\bf v} \right)\cdot d{\bf x}-\left({m\over 2}v^{2}+e\phi \right)dt
\ee
The coordinates are transformed into guiding center coordinates. The electrostatic guiding center
Lagrangian in known, and can be found in Hahm 1988. For the magnetic perturbations we take that
\be
A_{\parallel}({\bf X}+{\bf \rho}){\bf b} \cdot d({\bf X}+\rho)
\ee
where $\rho =\rho({\bf X},{\mu},{\theta})$. All the derivatives of $\rho$ will give vectors
perpendicular to the magnetic field. Thus if there is not a perpendicular magnetic potential
perturbation, the only term that survives is
$$A_{\parallel}{\bf b}\cdot d{\bf X}$$
Then the guiding center Lagrangian is
\be
\gamma =\left({e\over c}{\bf A}+{e\over c}A_{\parallel}{\bf b}+mv_{\parallel}{\bf b} \right)\cdot 
d{\bf X}+\mu d\theta -\left({m\over 2}v_{\parallel}^{2}+\mu B+e\phi \right)dt
\ee
This Lagrangian is transformed in the gyrocenter coordinates with the Lie transforms.
To the first order, if there are no perpendicular perturbations in the magnetic potential,
we can just substitute the perturbations of the fields with the respective gyroaveraged quantities. 
This can also be derived rigorously with the Lie transforms. The new Lagrangian will be
\be
\Gamma =\left({e\over c}{\bf A}+{e\over c}<A_{\parallel}>{\bf b}+mv_{\parallel}{\bf b} \right)\cdot 
d{\bf X}+\mu d\theta -\left({m\over 2}v_{\parallel}^{2}+\mu B+e<\phi> \right)dt
\ee
where now ${\bf X},\mu,v_{\parallel},\theta$ are the gyrocenter coordinates. From this one can derive 
the equations of motion using the Euler Lagrange equations
\be
\omega_{ij}{dz^{j}\over dt}={\partial H \over \partial z^{i}}-{\partial \gamma_{i}\over \partial t}
\ee
where $H$ is the Hamiltonian and 
\be
\omega_{ij}={\partial \gamma_{j}\over \partial z^{i}}-{\partial \gamma_{i}\over \partial z^{j}}
\ee
are the Lagrange brackets.

The Lagrange brackets that are of interest are
\be
\omega_{X_{i}X_{j}}={e\over c}\epsilon_{ijk}{\hat B}_{k}
\ee
\be
\omega_{{\bf X}U}=-m{\bf b}
\ee
where 
\be
{\hat {\bf B}}=\nabla \times {\bf A}+\nabla \times A_{\parallel}{\bf b}+{mc \over e}v_{\parallel}\nabla \times {\bf b}
\ee

Choosing $i=U$ we get that
\be
m{\bf b}\cdot {\dot {\bf X}}=mv_{\parallel}
\ee
Choosing $i={\bf X}$ we get that
\be
m{\dot v}_{\parallel}{\bf b}-{e\over c}{\dot {\bf X}}\times {\hat{\bf B}}=-e\nabla <\phi>-\mu \nabla B-{e\over c}
{\partial <A_{\parallel}>\over \partial t}{\bf b}
\ee
and taking that
\be
{\bf E}=-\nabla \phi -{1\over c}{\partial <A_{\parallel}>\over \partial t}{\bf b}
\ee
we get that
\be
m{\dot v}_{\parallel}{\bf b}-{e\over c}{\dot {\bf X}}\times {\hat{\bf B}}=e{\bf E}-\mu \nabla B
\ee

We can cross this equation with ${\bf b}$ to derive the equation for $\dot X$. To derive the equation of GKW we have
to approximate
\be
\nabla \times (<A_{\parallel}>{\bf b})\approx \nabla  (<A_{\parallel}>)\times {\bf b}
\ee
Then the equation of $\dot X$ is.
\be 
{{\rm d} {\bf X} \over {\rm d} t} = v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E + {\bf v}_{\delta B_\perp}, 
\ee
To derive the equation for $v_{\parallel}$ we dot with $\dot X$. This gives
\be
mv_{\parallel}{\dot v}_{\parallel}={\dot {\bf X}}\cdot \left(e{\bf E}-\mu \nabla B\right)
\ee

\section{Equation for the evolution of the distribution function}

In the previous section the evolution equation for the guiding center motion and the parallel velocity 
were formulated 
\be 
{{\rm d} {\bf X} \over {\rm d} t} = v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E + {\bf v}_{\delta B_\perp}, 
\ee
\be 
m v_\parallel {{\rm d} v_\parallel \over {\rm d} t} =  Z {e } 
{{\rm d} {\bf X} \over {\rm d} t} \cdot  {\bf E}  - \mu 
{{\rm d} {\bf X} \over {\rm d} t}\cdot \nabla B,
\ee
furthermore $\mu$ is a constant of motion. 
\be 
{{\rm d} \mu \over {\rm d} t} = 0
\ee
Here ${\bf b} = {\bf B}/B$ is the unit vector in the direction of the unperturbed magnetic field 
(${\bf B}$), $Z$ is the particle charge, and $\bf E$ is the gyro-averaged electric field 
\be 
{\bf E} = -\nabla \langle \phi \rangle  - {\partial \langle A_\parallel \rangle \over \partial t} 
\ee
where the angle brackets denote the gyro-average. 
The velocities in the first equation are from left to right, the parallel motion along the 
unperturbed field ($v_\parallel {\bf b}$), the drift motion due to the inhomogeneous field 
for which the low beta approximation will be used

the ${\bf E} \times {\bf B}$ velocity 
\be 
{\bf v}_E = {{\bf b} \times \nabla \langle \phi \rangle \over B},
\ee   
and the parallel motion along the perturbed field line 
\be 
{\bf v}_{\delta B} = - {{\bf b} \times \nabla v_\parallel \langle A_\parallel \rangle \over B} 
\ee 
The last two velocities can be combined 
\be 
\label{electromag}
\index{electromagnetic perturbations}
{\bf v}_\chi = {\bf v}_E + {\bf v}_{\delta B_\perp} = {{\bf b} \times \nabla \chi \over B} 
\ee
with 
\be 
\chi = \langle \phi \rangle - v_\parallel \langle A_\parallel \rangle 
\ee

From the equations above the gyro-kinetic equation in (${\bf X}, v_\parallel, 
\mu$) coordinates can be derived 
\be 
{\partial f \over \partial t} + {{\rm d} {\bf X} \over {\rm d} t} \cdot \nabla_\mu 
f + {{\rm d} v_\parallel \over {\rm d} t} {\partial f \over \partial v_\parallel} = 0
\ee 
The distribution function in this equation is split in a background $F$ and a perturbed 
distribution $f$. As discussed at the beginning of the section the $\delta f$ approximation 
will be employed. This approximation makes that the first order corrections to the term 
that contains the derivative towards the parallel velocity of $f$ can be neglected. 
The lowest order term in the equation for the parallel velocity can easily be identified 
to be the mirror term connected with the parallel motion in the magnetic field gradient 
\be 
{{\rm d} v_\parallel \over {\rm d} t} =   -{\mu B\over m } {{\bf B} \cdot \nabla B \over 
B^2 }
\ee
Consequently the equation for $f$ can be written in the form 
\be 
{\partial f \over \partial t} + (v_\parallel {\bf b} + {\bf v}_D + {\bf v}_\chi) \cdot \nabla f  
-{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial f \over \partial v_\parallel} = S. 
\ee
where $S$ is the source which is determined by the background distribution function.

Here we will use the simplest choice for the background which is the Maxwellian 
\be 
F = F_m = {n \over \pi^{3/2} v_{\rm th}^3 } \exp \biggl [ - {(v_\parallel-u )^2 + 2 \mu B / m \over v_{\rm th}^2} \biggr ] 
\ee 
where $v_{\rm th}$ is the thermal velocity. Note that we will formulate the equations in the co-moving 
system. The Maxwell defined here nevertheless contains the mean parallel velocity $u = u(\psi)$ to 
allow for a gradient in the toroidal velocity. The gradients of the Maxwellian will then simply be 
evaluated at $u = 0$. 
\be 
v_{\rm th} \equiv \sqrt{ 2 T / m} 
\ee
and $n$ is the density. (Note, in this area of research almost everybody uses a different definition of the 
thermal velocity, i.e. without the factor of 2 in front of the temperature. This can lead to some 
confusion when comparing results.) For the Maxwell distribution the gradients can be directly calculated 
\be
\nabla F_m = \biggl [ {\nabla n \over n} + \biggl ( { v_\parallel^{2} \over 
v_{\rm th}^2} + { 2 \mu B \over m v_{\rm th}^2} 
- {3\over 2} \biggr ) {\nabla T \over T}  + 2 {v_\parallel \over 
v_{\rm th}^{2}} \nabla u - {2 \mu B \over m v_{\rm th}^2} 
{\nabla B \over B} \biggr ]  F_m  
\ee
\be 
{\partial F_m \over \partial v_\parallel} = - {2 v_\parallel \over v_{\rm th}^2} F_M 
\ee

Using the relations for the Maxwellian one can derive 
\be 
{{\rm d} {\bf X} \over {\rm d} t } \cdot \nabla_b F_M + {{\rm d} v_\parallel \over {\rm d} t} \biggr \vert_b 
{\partial F_M \over \partial v_\parallel} = 0
\ee
where the index $b$ on the radial gradient indicates that only the $\nabla B$ term is considered, and 
the index $b$ on the time derivative of the parallel velocity indicates that only the change due to the 
$\mu \nabla B$ term is considered. 
The trapping term is, therefore, canceled from the equations due to the motion in the gradient of the 
magnetic field. This is indeed as one might expect since the Maxwell is isotropic. 
Furthermore since the remaining gradients are in the direction of $\nabla \psi$ 
\be 
{{\rm d} {\bf X } \over {\rm d} t } \cdot \nabla F_M \rightarrow ({\bf v}_\chi + {\bf v}_D) \cdot 
\nabla_p F_M,  
\ee
where the index p indicates that only the density, temperature, and toroidal velocity gradients of $F_M$ have 
to be considered. 
For the change of parallel velocity due to the motion in the electric field we note that the only 
term due to the electro-magnetic part of the electric field is 
\be 
- {Z e \over m v_\parallel } {{\rm d} {\bf X } \over {\rm d} t }\cdot {\bf b} {\partial \langle A_\parallel \rangle 
\over \partial t} {\partial F_M \over \partial v_\parallel} 
=  {Z e \over T} v_\parallel {\partial \langle A_\parallel \rangle \over \partial t} F_M 
\ee
For the electrostatic potential, finally,the ExB motion is perpendicular to $\nabla \langle \phi \rangle$ 
\be 
- {Z e \over m v_\parallel } {{\rm d} {\bf X } \over {\rm d} t }\cdot \nabla \langle \phi \rangle 
{\partial F_M \over \partial v_\parallel} = {Ze \over T} [ v_\parallel {\bf b} + {\bf v}_D + {\bf v}_{\delta B} ] 
\cdot \nabla \langle \phi \rangle F_M 
\ee 
In conclusion we can write the source as  
\be 
S =  - ({\bf v}_\chi + {\bf v}_d) \cdot \nabla_p F_M -  {Z e \over T} v_\parallel {\partial \langle A_\parallel \rangle \over \partial t} F_M 
-  {Ze \over T} [ v_\parallel {\bf b} + {\bf v}_D + {\bf v}_{\delta B} ] \cdot \nabla \langle \phi \rangle F_M 
\ee

The term containing the derivative of the vector potential towards time is a difficult term numerically. In the $\delta f$ 
formalism it can, however, easily be combined with the time derivative of $f$. If one defines a 'new' distribution $g$ 
\be 
g = f + {Z e \over T} v_\parallel \langle A_\parallel \rangle F_M 
\ee
and replaces the time derivative of $f$ in the gyro-kinetic equation with the time derivative of $g$, the cumbersome 
time derivative of $\langle A_\parallel$ is removed. 
Furthermore direct substitution yields 
\bee 
{\bf v}_\chi \cdot \nabla f &=& {\bf v}_\chi \cdot \nabla g + {\bf v}_\chi \cdot \nabla {Z e \over T} v_\parallel 
\langle A_\parallel \rangle F_M \approx  {\bf v}_\chi \cdot \nabla g + %{Z e \over T} v_\parallel {{\bf b} \times 
% \nabla \langle \phi \rangle \over B} \cdot \nabla \langle A_\parallel \rangle F_M 
\cr
\noalign{\vskip 0.2 truecm} 
& = & {\bf v}_\chi \cdot \nabla g
- {Z e \over T} v_\parallel {\bf v}_{\delta B} \cdot \nabla \langle \phi \rangle F_M 
\eee
i.e. rewriting the ExB nonlinearity on $f$ in terms of $g$ also cancels one source terms. Using again that the parallel 
motion in the gradient of the Maxwellian balances the trapping term one arrives at the equation for $g$
\be
\label{g-vlasov} 
\index{gyrokinetic Vlasov equation}
{\partial g \over \partial t} + (v_\parallel {\bf b} + {\bf v}_D + {\bf v}_\chi) \cdot \nabla g  
-{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial g \over \partial v_\parallel} = S. 
\ee
where the Source is given by 
\bee 
\label{sources}
S &=&  - ({\bf v}_\chi + {\bf v}_D) \cdot \nabla_p F_M
-  {Ze \over T} [ v_\parallel {\bf b} + {\bf v}_D ] \cdot \nabla \chi F_M \cr 
\noalign{\vskip 0.2 truecm} 
& & - {Z e \over T} {\mu B \over m} \biggl ( {{\bf B} \cdot \nabla {\bf B} \over B^2} 
\biggr ) \langle A_\parallel \rangle F_M 
\eee
The electro-static potential and vector potential appear in a relatively symmetric way. The only 
exception is the last term which is due to the trapping.

To proceed we have to work out the drift 
\be 
{\bf v}_D = {m v_\parallel^2 \over Z e B} {\bf b} \times ({\bf b}\cdot \nabla {\bf b}) + 
{\mu \over Z e} {{\bf B}\times \nabla B \over B^2} + { 2 m v_\parallel \over Z e B } {\bf 
\Omega}_\perp - {m \Omega^2 R \over Z e B }  {\bf b} \times {\nabla R}
\ee 
We will write this drift velocity in a more conventional form. One can derive 
\be 
 {\bf b} \times ({\bf b}\cdot \nabla {\bf b}) =  {1\over 2} ({\bf b} \times \nabla \psi) 
\beta^\prime + {{\bf B}\times \nabla B \over B^2} 
\ee
where 
\be 
\beta^\prime = {1 \over B^2 / 2 \mu_0} {\partial p \over \partial \psi} 
\ee
and $p$ is the equilibrium pressure. Note that $\beta^\prime$ is not exactly the derivative 
of the plasma beta (although it is very close in almost all cases) 
To derive the relation above we have used 
\be 
\nabla \times {\bf B} = \mu_0 {\bf J} 
\ee 
and 
\be 
{\bf J} \times {\bf B} = \nabla p = {\partial p \over \partial \psi} \nabla \psi
\ee
where the last equation is the equilibrium equation of MHD. 
Using the relations given above the drift velocity can be written in the form 
\bee
\label{drift}
\index{drifts}
{\bf v}_D &=& {1\over Ze} \biggl [ {m v_\parallel^2\over B} + \mu \biggr ] {{\bf B} \times \nabla B
\over B^2} + {m v_\parallel^2 \over 2 Z e B} \beta^\prime {\bf b} \times \nabla \psi 
 + { 2 m v_\parallel \over Z e B } {\bf \Omega}_\perp \cr 
\noalign{\vskip 0.2 truecm} 
&-& {m \Omega^2 R \over Z e B }  {\bf b} \times {\nabla R} = {\bf v}_d + {\bf v}_{db} + {\bf v}_{dc} 
+ {\bf v}_{df}, 
\eee
where ${\bf v}_d$ is the low beta drift (i.e. the term proportional to ${\bf B} \times \nabla B$), 
${\bf v}_{db}$ is the finite beta correction to the drift, ${\bf v}_{dc}$ is the Coriolis drift
and ${\bf v}_{df}$ is the drift due to the centrifugal force. Note that the centrifugal force also
leads to an acceleration along the magnetic field, which is not kept in the equations. At pressent
therefore, the centrifugal force must be put to zero for the formalism to be consistent. We keep 
it here for later use.

\section{Normalizations} 
\label{normalization}
\index{normalizations}

The normalization of the set of equations is a bit of a pain. Although in this chapter the code is 
not discussed as such, the normalization introduced here is of course designed to give an efficient
implementation. 
The velocity will be normalized by the thermal velocity of each of the species. This has the 
advantage that the grid in velocity space can always be defined relative to the thermal 
velocity of that species. Quantities like the potential, however, must be normalized by 
the reference temperature since it is species independent. A similar argument applies 
to the vector potential and the time.

We define a reference mass $m_{\rm ref}$, thermal velocity $v_{\rm thref}$, density $n_{\rm ref}$, 
Temperature $T_{\rm ref}$, magnetic field $B_{\rm ref}$, and major radius $R_{\rm ref}$. These 
are not unrelated since 
\be 
T_{\rm ref} = {1 \over 2} m_{\rm ref} v_{\rm thref}^2 \qquad 
\rho_{\rm ref} = {m_{\rm ref} v_{\rm thref} \over e B_{\rm ref}}
\ee
Furthermore we define a normalized Larmor radius to be 
\be 
\rho_* = \rho_{\rm ref} / R_{\rm ref} 
\ee
The reference values are used to define, for every species, a dimensionless mass $m_R$, thermal 
velocity $v_R$ density $n_R$, and temperature $T_R$. 
\be 
m_R = {m \over m_{\rm ref} } 
\qquad 
v_R = {v_{\rm th} \over v_{\rm thref}}
\qquad 
n_R = {n \over n_{\rm ref} }
\qquad 
T_R = {T \over T_{\rm ref}}
\ee 
The fields are made dimensionless using only reference values 
\be 
\phi = \rho_* {T_{\rm ref} \over e} \phi_N 
\qquad 
A_\parallel = B_{\rm ref} R_{\rm ref} \rho_*^2 A_{\parallel N}
\qquad 
\chi = \rho_* {T_{\rm ref} \over e} \chi_N 
\ee 
Here the index $N$ refers to a normalized quantity. 
Note that a factor $\rho_*$ has been added. This makes that these quantities will be of the 
order 1. 
Also time $t$, magnetic field $B$, angular rotation frequency $\Omega$, and the major radius $R$,
are made dimensionless using the reference values 
\be
t = { R_{\rm ref} \over v_{\rm thref}} t_N 
\qquad 
B = B_{\rm ref} B_N
\qquad 
\Omega = {v_{\rm thref} \over R_{\rm ref}} \Omega_N 
\qquad 
R = R_{\rm ref} R_N
\ee
The velocity space coordinates, however, are made dimensionless with the help of 
the thermal velocity of the species 
\be 
v_\parallel = v_{\parallel N} v_{\rm th} 
\qquad 
\mu = {m v_{\rm th}^2 \over B_{\rm ref}} \mu_N
\ee
It is then convenient to normalize the distribution functions according to 
\be 
f = \rho_* {n \over v_{\rm th}^3} f_N 
\qquad 
F_M = {n \over v_{\rm th}^3} F_{MN}
\ee 
Note that the normalized Maxwellian is the same for all species 
\be 
F_{MN} = {1 \over \pi^{3/2}} \exp [ - v_{\parallel N}^2 - \mu_N B_N ] 
\ee
The gradient of each species is made dimensionless using the density and 
Temperature of that species, the rotation gradient on the other hand is 
often the same for all species (although it does not have to be) and is
therefore normalized to $v_{\rm thref}$ 
\be 
{R\over L_N} = - {1\over n} {\partial n \over \partial \psi} 
\qquad 
{R \over L_T} = - {1 \over T} {\partial T \over \partial \psi} 
\qquad 
u^\prime = - {1\over v_{\rm thref}} {\partial u \over \partial \psi}
\label{eq:gradients}
\ee
Note that we will always use a normalized coordinate $\psi$. For circular surfaces 
for instance $\psi = r / R = \epsilon$. Hence for each co-ordinate
\be
x_{\alpha,N} = {x_{\alpha} \over R_{\rm ref}}
\ee
With the consequence that derivatives are normalised as
\be
{\partial \over \partial x_{\alpha,N}} = {1 \over R_{\rm ref}} {\partial \over \partial x_\alpha}
\ee
Although the quantity $\beta$ is
dimensionless we will define a different dimensionless beta ($\beta_N$) for the 
use in the code. This $\beta_N$ is directly related to the reference values and is
defined to be 
\be 
\beta_N = {n_{\rm ref} T_{\rm ref} \over B_{\rm ref}^2 / 2 \mu_0} 
\ee
The derivative $\beta^\prime$ is related and can be calculated from the pressure and temperature gradients
\be
\beta^\prime=\beta_N \sum_s n_{Rs} T_{Rs} (R/L_{Ns} + R/L_{Ts})
\ee
which is not necessiarily consistent with the geometry tensors for an arbitrary geometry.

Finally, a wave vector will be introduced in the next section, for which we define 
a normalization 
\be 
\label{knorm}
k = {k_N \over R_{\rm ref} \rho_*} = {k_N \over \rho_{\rm ref}}
\ee

With the normalizations defined we can now define the geometry dependent quantities
in normalized form. These will be defined as 
\index{geometry tensors, normalized}
\be 
{\cal F} = {R_{\rm ref} B^s \over B} \qquad \rightarrow \qquad 
R_{\rm ref} {\bf b} \cdot \nabla  = {{\cal F}} {\partial \over \partial s} 
\ee
\be 
{\cal D}^\alpha = {R_{\rm ref}^2 {\bf B} \times \nabla B \over B^2 B_N} \cdot \nabla x_\alpha 
\ee 
\be 
\label{efun}
{\cal E}^{\alpha \beta} = {R_{\rm ref}^2\over 2 B_N} (\nabla x_\alpha \times \nabla x_\beta) \cdot {\bf b} 
\ee 
\be 
{\cal G} = {R_{\rm ref} {\bf B } \cdot \nabla B \over B^2} 
\ee
\be 
{\cal H}^\alpha = {R_{\rm ref}^2 \over B_N v_{\rm thref}}{\bf \Omega}_\perp \cdot \nabla x_\alpha 
\ee
\be 
{\cal I}^\alpha = {R_{\rm ref}^2 \Omega^2 \over 2 v_{\rm thref}^2 B_N } (\nabla x_\alpha \times \nabla R^2) 
\cdot {\bf b} 
\ee

With the normalizations defined we can write the evolution equation in normalized form 
\be 
{\partial g \over \partial t} = {n \over v_{\rm th}^3} \rho_* {v_{\rm thref} \over R_{\rm ref}} 
{\partial g_N \over \partial t_N} \rightarrow {\partial g_N \over \partial t_N} 
\ee
i.e. we will divide each term in the equation by $n \rho_* v_{\rm thref} / (v_{\rm th}^3 R_{\rm ref}) $
and write it explicitely in normalized form. We furthermore define 
\be
E_D = v_{\parallel N}^2 + \mu_N B_N  \qquad 
E_T = v_{\parallel N}^2 + 2 \mu_N B_N - {3\over 2} 
\ee
This yields 
\be 
\label{terms}
\index{equation, terms as implemented}
{\partial g_N \over \partial t_N} = {\rm I} + {\rm II} + {\rm III} + {\rm IV} + {\rm V} + \ldots 
\ee
with (by comparision to \ref{g-vlasov} and \ref{sources} - see \ref{1} and \ref{2} for more detail)
\bee 
{\rm I} &=& - v_\parallel {\bf b} \cdot \nabla g \rightarrow - v_R v_{\parallel N} {\cal F} {\partial g_N \over 
\partial s} 
\cr \noalign{\vskip 0.4 truecm} 
%\ee
%\be 
{\rm II} &=& - {\bf v}_D \cdot \nabla g \rightarrow -{1 \over Z} \rho_* [ T_R E_D {\cal D}^\alpha + T_R v_{\parallel N}^2
\beta^\prime  {\cal E}^{\psi \alpha} + 
\cr \noalign{\vskip 0.2 truecm}
& & 2 m_R v_R v_{\parallel N}{\cal H}^\alpha + m_R {\cal I}^\alpha ] {\partial g_N \over \partial x_\alpha}
\cr \noalign{\vskip 0.4 truecm} 
%\ee
%\be 
{\rm III} &=& - {\bf v}_\chi \cdot \nabla g \rightarrow - \rho_*^2 {\partial \chi_N \over \partial x_\beta}
{\cal E}^{\beta \alpha} {\partial g_N \over \partial x_\alpha} 
\cr \noalign{\vskip 0.4 truecm} 
%\ee 
%\be 
{\rm IV} &=& + {\mu B \over m } {{\rm B} \cdot \nabla B \over B^2} {\partial g \over \partial v_\parallel} \rightarrow
v_R \mu_N B_N {\cal G} {\partial g_N \over \partial v_{\parallel N}} 
\cr \noalign{\vskip 0.4 truecm} 
%\ee
%\be 
{\rm V} &=& - {\bf v}_\chi \cdot \nabla_p F_M \rightarrow  \rho_* {\partial \chi_N \over \partial x_\alpha} 
{\cal E}^{\alpha \psi} \biggl [ {R \over L_N} + E_T {R \over L_T} + {2 v_{\parallel N} \over v_R} u^\prime 
\biggr ] F_{MN}
\cr \noalign{\vskip 0.4 truecm} 
%\ee
%\be 
{\rm VI} &=& - {\bf v}_D \cdot \nabla_p F_M \rightarrow  {1 \over Z} [ T_R E_D {\cal D}^\psi + 2 m_R v_R v_{\parallel N}
H^\psi + m_R I^\psi ] \times \cr \noalign{\vskip 0.2 truecm}
& &  \biggl [ {R \over L_N} + E_T {R \over L_T} + {2 v_{\parallel N} \over v_R} u^\prime  \biggr ] F_{MN}
\cr \noalign{\vskip 0.4 truecm} 
%\ee 
%\be 
{\rm VII} &=& - {Z e \over T} v_\parallel {\bf b} \cdot \nabla \chi F_M \rightarrow - {Z \over T_R} v_R v_{\parallel N} 
{\cal F} {\partial \chi_N \over \partial s} F_{MN}
\cr \noalign{\vskip 0.4 truecm}
%\ee
%\be 
{\rm VIII} &=& - {Z e \over T}{\bf v}_D \cdot \nabla \chi F_M  \rightarrow - \rho_* [ E_D {\cal D}^\alpha + \beta^\prime 
v_{\parallel N}^2 {\cal E}^{\psi \alpha} + 
\cr \noalign{\vskip 0.2 truecm}
& & {2 m_R v_R \over T_R}v_{\parallel N}{\cal H}^\alpha + {m_R \over T_R}{\cal I}^\alpha ] {\partial \chi_N \over \partial x_\alpha} F_{MN}
\cr \noalign{\vskip 0.4 truecm} 
%\ee
%\be
{\rm IX} &=& - {Z e \over T} {\mu B \over m} \biggl ( {{\bf B} \cdot \nabla {\bf B} \over B^2} 
\biggr ) \langle A_\parallel \rangle F_M  \rightarrow - {2 Z \over m_R } \mu_N B_N \langle A_{\parallel N} \rangle 
{\cal G} F_{MN}
\label{eqn_norm}
\eee
where 
\be 
\chi_N = \phi_N - 2 v_{R} v_{\parallel N} \langle A_{\parallel N} \rangle 
\ee 
and 
\be 
g_N = f_N + {2 Z \over T_{\rm R}} v_{\rm R} v_{\parallel N} \langle A_{\parallel N} \rangle F_{MN} 
\ee

\section{Local limit} 
\index{local limit}
The whole reason for aligning the coordinate system with the magnetic field lies in the strong
difference in wave vectors in the parallel and perpendicular direction $k_\parallel \ll k_\perp$. 
The scale lengths perpendicular to the field are of the order of the Larmor radius and therefore
much smaller than the system size. In the local approximation the geometry in the perpendicular 
plane (which varies on the scale of the system size) is assumed uniform. This means that all 
the tensors ${\cal D, E, F, G, H, I}$ can be assumed to be a function of the parallel coordinate $s$ 
only. 
Of course, a homogeneous system can be conveniently described by the sum of plane waves. 
\be 
\label{eq:local_limit}
f(\psi,\zeta,s) \rightarrow \sum_{k_\zeta k_\psi} h (k_\psi, k_\zeta,s) \exp [ {\rm i} k_\zeta  \zeta / \rho_*  +
 {\rm i} k_\psi  \psi / \rho_* ] 
\ee 
where the factor $\rho_*$ has been added to have a convenient normalization 
\be 
\label{rho_norm}
\rho_* {\partial \over \partial x_\alpha} \rightarrow {\rm i} k_\alpha 
\ee
and $k$ is of the order unity. Note that in the perpendicular plane a box size will be used which 
is several Larmor radius wide. The quantities $\zeta / \rho_*$, and $\psi / \rho_*$ are therefore 
of the order unity in the domain of the box. Note too, that although the variation of the geometry 
in the radial direction can be neglected for the tensors, it can not necessarily be neglected for 
the Fourier modes, due to the appearance of $1/\rho_*$ inside the exponents.

The use of Fourier harmonics automatically means that the boundary conditions in the perpendicular 
plane are periodic.  The periodicity of the torus is, however, not automatically satisfied. The 
toroidal periodicity means 
\be 
f(\psi,\zeta+ 1,s) = f(\psi,\zeta,s) 
\ee
For the Fourier representation above it follows that 
\be  
{k_\zeta \over 2 \pi \rho_*} = N 
\ee
where $N$ must be an integer. Since $\rho_*$ is small, this condition can in general be satisfied 
with very small changes to $k_\zeta$ or $\rho_*$. Since in the local limit we will not have to 
specify $\rho_*$ we can simply assume the above condition is satisfied. 
The poloidal periodicity translates to
\be 
f(\psi,\zeta + q/2, 1/2) = f(\psi,\zeta- q/2,-1/2)
\ee
This condition translates to 
\bee 
\sum_{\bf k} h( k_\psi, k_\zeta, {1 \over 2} ) \exp \biggl [ {{\rm i} k_\zeta \over \rho_*} + {{\rm i} k_\psi 
\psi \over \rho_*} + {{\rm i} q k_\zeta \over 2 \rho_*} \biggr ] \cr 
\noalign{\vskip 0.2 truecm} 
= \sum_{\bf k} h ( k_\psi, k_\zeta, - {1 \over 2} )
 \exp \biggl [ {{\rm i} k_\zeta \over \rho_*} + {{\rm i} k_\psi 
\psi \over \rho_*} - {{\rm i} q k_\zeta \over 2 \rho_*} \biggr ]
\eee
Expanding the safety factor around a reference value (the center of the radial domain) 
\be 
{q k_\zeta \over \rho_*} = {q_R k_\zeta \over \rho_*} + k_\zeta {\partial q \over \partial \psi}{\psi \over \rho_*} 
+ {1 \over 2} k_\zeta \rho_* {\partial^2 q \over \partial \psi^2} \biggl ( {\psi \over \rho_*} \biggr )^2 
\ee
and neglecting the second derivative correction (which gives only a $\rho_*$ variation over the size of the 
box) one can see that this condition can be satisfied if also $q_R k_\zeta / 2 \pi \rho_*$ is assumed to be an 
integer and if 
\be 
h(k_\psi, k_\zeta, {1\over 2} ) = h( k_\psi + k_\zeta {\partial q \over \partial \psi} , k_\zeta, -{1 \over 2} ) 
\ee
i.e. on the boundary of the $s$ domain one simply connects the mode with the appropriate higher $k_\psi$ mode. 
The above boundary condition for the Fourier amplitudes implies that a convenient choice for the spacing of 
the $k_\psi$ modes in any discrete Fourier representation is 
\be 
\Delta k_\psi = k_\zeta {\partial q \over \partial \psi} {1 \over p}
\ee
where $p $ is some integer (\texttt{ikxspace}). \label{ikxspace}
\index{ikxspace}

The other problem we need to address is that one would like to specify the wave vector in normalized 
form as $k_\theta \rho_i$. We will use the expression 
\be 
(k_\theta \rho_{\rm ref} )^2 = g^{\zeta \zeta} k_\zeta^2 R_{\rm ref}^2
\ee
evaluated at the low field side midplane to determine $k_\zeta$ from $k_\theta \rho$.   
The full perpendicular wave vector is given by 
\be 
(k_\perp \rho_{\rm ref} )^2 = R_{\rm ref}^2 g^{\alpha \beta} k_\alpha k_\beta 
\ee

\section{Gyro-kinetic field equation} 
The quasi-neutrality equation (also referred to as Poisson equation) and the Amp\`{e}re Law are used to calculate the perturbed electrostatic potential $\phi$ and vector potential $A_\parallel$ from
the perturbed distribution function. These equations apply in real space while the perturbed distribution function $f$ is calculated in the gyrocenter phase space. We therefore need to express the
distribution function in the real phase space $f^{*}_{tot}(\mathbf{x};\mu,v_\parallel,\gamma)$ as a function of the distribution in the gyrocenter phase space $f_{tot}(\mathbf{X};\mu,v_\parallel)$:
\index{field equations}
\begin{eqnarray}
\label{eq:f_real}
 f^{*}_{tot}(\mathbf{x};\mu,v_\parallel,\gamma) &=& f_{b}(\mathbf{X};\mu,v_\parallel) + f(\mathbf{X};\mu,v_\parallel) \\
 &+& \frac{Ze}{B}\frac{\partial f_{b}^{*}}{\partial \mu}(\mathbf{x}) \left[ \phi(\mathbf{x}) - \left< \phi\right>(\mathbf{X}) - \frac{v_\parallel}{c}\left(A_\parallel(\mathbf{x}) - \left<
A_\parallel\right>(\mathbf{X})\right) \right] \nonumber
\end{eqnarray}
In this expression, $<.>$ denotes the gyroaverage and $\mathbf{X}$ is the gyrocenter of the particle in $\mathbf{x}$ with velocity coordinates $(\mu,v_\parallel,\gamma)$, where $\gamma$ is the
gyroangle. The total distribution functions (subscript \textit{tot}) have been split as usual in a background part (subscript \textit{b}) and a perturbed part (no subscript) with the assumption that
the background distribution is approximately constant over a gyro-orbit. The three contributions to the real space distribution function in Eq.~(\ref{eq:f_real}) are therefore, in that order, the
background, the gyroaveraged perturbed part and the gyroangle dependent perturbed part (which arises from the variation of the potential and vector potential over a gyro-orbit). In the following,
the background distribution is taken to be a Maxwellian: 
$$f_{b}^{*}(\mathbf{x};\mu,v_\parallel,\gamma) = f_{b}(\mathbf{X};\mu,v_\parallel) = F_M(\mathbf{X};\mu,v_\parallel)$$
Using for the potential the local limit approximation already used in Eq.~(\ref{eq:local_limit}) for the distribution function, one can write:
\be
 \phi(\mathbf{x}) = \sum_\mathbf{k_\perp} \phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}} = \sum_\mathbf{k_\perp}
\phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}}\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{\rho}}
\ee
where $\mathbf{x} = \mathbf{X} + \mathbf{\rho}$  has been used, $\mathbf{\rho}$ being the gyroradius vector. Note that $\phi_{k_\perp}(\mathbf{X})$ is a slowly varying amplitude that can be
apporximated by $\phi_{k_\perp}(\mathbf{x})$. The gyro-average leads to 
\be
\label{eq:phi_gyro}
 \left<\phi\right>(\mathbf{X}) = \sum_\mathbf{k_\perp}
\phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}}\left<\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{\rho}}\right> = \sum_\mathbf{k_\perp}
\phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}}J_0(k_\perp\rho)
\ee
where the Bessel function $J_0$ arises from: 
\be
 \left<\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{\rho}}\right> = \frac{1}{2\pi}\oint \textrm{e}^{ik_\perp\rho\cos(\gamma-\gamma_{\rm{\rm ref}})} \,\textrm{d}\gamma= J_0(k_\perp\rho)
\ee
From Eq.~(\ref{eq:f_real}) and Eq.~(\ref{eq:phi_gyro}) one can then calculate the density in real space:
\begin{eqnarray*}
n(\mathbf{x}) &=& \int f^{*}_{tot}(\mathbf{x};\mu,v_\parallel,\gamma) B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma \\
 &=& \int f_{b}(\mathbf{X};\mu,v_\parallel) B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma + \int \sum_\mathbf{k_\perp} 
f_{k_\perp}(\mathbf{x};\mu,v_\parallel)\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}}  \textrm{e}^{-i\mathbf{k_\perp}\cdot\mathbf{\rho}} B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma \\
 &-& \frac{Ze}{T} \int F_M(\mathbf{X};\mu,v_\parallel) \sum_\mathbf{k_\perp} \left(\phi_{k_\perp}(\mathbf{X}) - \frac{v_\parallel}{c}A_{k_\perp}(\mathbf{X})\right)
\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}} \left(1-\textrm{e}^{-i\mathbf{k_\perp}\cdot\mathbf{\rho}}J_0(k_\perp\rho)\right)     B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma
\end{eqnarray*}
Note that the integral over $\gamma$ is NOT a gyroaverage but the sum of the contribution to the local density coming from all the
gyrocenters $\mathbf{X}$ around $\mathbf{x}$. The reference point is $\mathbf{x}$ and the perturbed distribution function has been expressed as
$$f(\mathbf{X}) = \sum_\mathbf{k_\perp} f_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}} = \sum_\mathbf{k_\perp}
f_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}} \textrm{e}^{-i\mathbf{k_\perp}\cdot\mathbf{\rho}}$$
Performing the velocity space integrals, one obtains:
\begin{eqnarray*}
\label{eq:density}
n(\mathbf{x}) &=& \overbrace{n_0 + 2\pi B \int \sum_{k_\perp} f_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}}J_0(k_\perp\rho)
\,\mathrm{d}\mu\,\mathrm{d}v_\parallel}^\textrm{\scriptsize guiding center density (background+perturbed)}\\* 
& & - \underbrace{n_0\frac{Ze}{T}\sum_{k_\perp} \phi_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}}\left(1-\Gamma(b)\right)}_\textrm{\scriptsize
polarisation density}
\end{eqnarray*}
where $n_0$ and $T$ are the density and temperature of the Maxwellian and where the following relationship has been used:
$$\frac{1}{v_{\rm th}^2}\int_0^\infty \mathrm{e}^{-v_\perp^2/v_{\rm th}^2}J_0^2(k_\perp v_\perp \frac{ZeB}{m}) v_\perp \,\mathrm{d}v_\perp = \Gamma(b)=\textrm{e}^{-b}I_0(b) \qquad \mathrm{with} \qquad b={1
\over 2} {k_\perp^2 m^2 v_{\rm th}^2 \over Z^2e^2 B^2}$$
with $I_0$ the modified Bessel function. The gyro-kinetic quasi-neutrality equation is then obtained by summing over the different species and assuming quasi-neutrality for the background (see also
Ref. \cite{BEER} Eq. (2.55)). Writing the result for a single wave vector to keep the expressions shorter, one obtains:
\be 
\sum_s Z_s n_s(\mathbf{x}) = 0 = \sum_s  Z_s \left[2\pi B \int  f_s(\mathbf{x})J_0(k_\perp\rho_s)\,\mathrm{d}\mu\,\mathrm{d}v_\parallel + {Z_s e n_s \over T_s} ( \Gamma(b_s) - 1)\phi \right] 
\ee 
The quasi-neutrality equation in normalized form is 
\be 
\sum_s  Z_s n_{Rs} \biggl [ 2 \pi B_N \int {\rm d} v_{\parallel N} {\rm d} \mu_N J_0(k_\perp\rho_s) g_{Ns} 
+ {Z_s \over T_{Rs}} [ \Gamma(b_s) -1]\phi_N \biggr ] = 0 
\ee
\blue From here on still needs to be checked \black 
with the argument of the Bessel function being 
\be 
k_\perp \rho_s = { k_\perp \rho_* R_{\rm ref} \over Z} \sqrt{ 2 m_R T_R \mu_N \over B_N}
\ee

The vector potential must be calculated from Ampere's law 
\be 
\nabla^2 A_\parallel = \mu_0 {\bf J} = \mu_0 \sum_s Z_s e \int {\rm d}^3 {\bf v} \, 
v_\parallel J_0 f_s 
\ee 
Since we the distribution $g$ forward in time, we have to express this equation 
in terms of $g$ 
\be 
\nabla^2 A_\parallel + \mu_0 \sum_s {Z_s^2 e^2 \over T} \int {\rm d}^3 {\bf v} \, 
v_\parallel^2 J_0^2 A_\parallel F_M 
 =  \mu_0 \sum_s Z_s e \int {\rm d}^3 {\bf v} \, v_\parallel J_0 g_s  
\ee 
The integral over the Maxwell distribution can be performed analytically with the 
help of the integrals 
\be 
{1 \over \sqrt{\pi}} \int_{-\infty}^\infty {\rm d} v_\parallel  v_\parallel^2 
\exp \biggl [ - {v_\parallel^2 \over v_{\rm th}^2} \biggl ] = {1\over 2} v_{\rm th}^3 
\ee 
\be 
\int_0^\infty {\rm d} x J_0(\sqrt{2 b x}) \exp [ -x ] = \Gamma(b) = \exp [ -b ] I_0(b) 
\ee 
where $I$ is the modified Bessel function. 
Using these expressions one obtains 
\be
\int {\rm d}^3 {\bf v}\, v_\parallel^2 J_0^2 F_M = {n v_{\rm th}^2 \over 2} \Gamma(b)
\ee 
with 
\be 
b = {1 \over 2} {k_\perp^2 m^2 v_{\rm th}^2 \over e^2 B^2} = {1 \over 2} m_R T_R (k_\perp 
\rho_* R_{\rm ref} / Z B_N^2)^2
\ee
Replacing furthermore $\nabla_\perp^2$ with $-k_\perp^2$ Ampere's law for $g_s$ can be 
written as 
\be 
-k_\perp^2 A_\parallel + \mu_0 \sum_s {Z_s^2 e^2 n \over m} \Gamma(b) A_\parallel =
\mu_0 \sum_s Z_s e \int {\rm d}^3 {\bf v} v_\parallel J_0^2 g_s
\ee
The normalization can then be applied to yield 
\bee 
\biggl [ -k_{\perp N}^2 + \beta \sum_s {Z_s^2 n_{rat,s} \over m_{rat,s}} \Gamma(b_s) \biggr ] A_{\parallel N}
= \beta \sum_s Z_s v_{thrat,s} n_{rat,s} \times \cr 
\noalign{\vskip 0.2 truecm} 
2 \pi B_N \int {\rm d} v_{\parallel N} \int {\rm d} \mu_N
v_{\parallel N} J_0 g_{N,s}
\eee
where $g_N$ is given by 
\be 
g_N = f_N + 2 Z {v_{\rm thrat} \over T_{\rm rat}} v_{\parallel N} J_0 A_{\parallel N} F_{MN}
\ee

\section{Collisions} 
\index{collision operator}
The previous sections have neglected the collision operator. This operator can formally be written 
as 
\be 
{\partial f_a \over \partial t} = C(f_a) 
\ee 
Here we have added the index $a$ to denote the species. 
Useful expressions for the collision operator have been published in the literature, usually in the 
($v,\theta$) coordinate system, where $v$ is the velocity and $\theta$ is the pitch angle. For a 
Maxwellian background the collision operator takes the following form 
\be 
C(f_a) = \sum_b {1\over v^2} {\partial \over \partial v} \biggl [ v^2 \biggl ( 
D_{vv}^{a/b} {\partial f_a \over \partial v} - F_v^{a/b}f_a \biggr ) \biggr ] + {1 \over v 
\sin \theta } {\partial \over \partial \theta} \biggl [ \sin \theta  D_{\theta \theta}^{a/b} 
{1\over v}{\partial f_a \over \partial \theta} \biggr ] 
\ee 
Here the sum is over all species $b$. 
$D_{\theta \theta}$ represents the pitch angle scattering, while $D_{vv}$ is the energy scatering
and $F_v$ is the slowing down force. For convenience (ours, not yours) we split this collision 
operator in three parts 
\be 
C(f) = C_\theta(f) + C_{vv}(f) + C_v(f)
\ee
where the first part is due to $D_{\theta \theta}$, the second due to $D_{vv}$ and the third 
due to $F_v$. The nomalization of the collision operator follows from the normalization of the 
gyro-kinetic equation 
\be 
C_N(f_a) = {R_{\rm ref} v_{tha}^3 \over n_a \rho_* v_{\rm thref}} C(f_a)
\ee
We will also make the diffusion coefficients and friction force coefficient dimensionless accorring 
to 
\be 
D_{\theta \theta}^{a/b} = {v_{tha}^3 \over R_{\rm ref}} D_{\theta \theta N}^{a/b} \qquad 
D_{vv}^{a/b} = {v_{tha}^3 \over R_{\rm ref}} D_{vv N}^{a/b} \qquad 
F_v^{a/b} =  {v_{tha}^2 \over R_{\rm ref}} F_{v N}^{a/b}  
\ee
The diffusion coefficients can be obtained for the literature \cite{KAR86} (and the reference cited therein) 
\be 
D_{\theta \theta N}^{a} = \sum_b {\Gamma^{a/b}_N \over 4 v_{Na}} \biggl [ \biggl ( 2 - {1\over v_{Nb}^2} \biggr ) 
{\rm erf}(v_{Nb}) + {1 \over v_{Nb}} {\rm erf}^\prime (v_{Nb}) \biggr ] 
\ee
\be 
D_{vvN}^{a} = \sum_b { \Gamma^{a/b}_N  \over 2 v_{Na}} \biggl [ {1 \over v_{Nb}^2} {\rm erf}(v_{Nb}) - 
{1\over v_{Nb}}{\rm erf}^\prime (v_{Nb}) \biggr ] 
\ee 
\be 
F_{vN}^{a} = - \sum_b {\Gamma^{a/b}_N \over v_{Na}^2} {m_{Ra} \over m_{Rb}} [ {\rm erf}(v_{Nb}) - v_{Nb} {\rm erf}^\prime (v_{Nb})  ]  
\ee 
The normalized velocity in these equations is defined as 
\be
v_{N\gamma} =  {v_{Ra} \over v_{R\gamma}}[ v_{\parallel N}^2 + 2 \mu_N B ]^{1/2} 
\ee
(with $\gamma = a$ or $\gamma = b$) and the error function are given through the relations  
\be 
{\rm erf}(u) = {2 \over \sqrt{\pi}} \int_0^u \exp [ -x^2] {\rm d} x 
\ee 
\be 
{\rm erf}^\prime(u) = {2 \over \sqrt{\pi}} \exp [ -u^2] 
\ee
Finally the constant $\Gamma^{a/b}_N$ is given by 
\be 
\Gamma^{a/b}_N = {R_{\rm thref} n_b Z_a^2 Z_b^2 e^4 \ln \Lambda^{a/b} \over 4 \pi \epsilon_0^2 m_a^2 v_{tha}^4} = 
6.5141\cdot 10^{-5} {R_{\rm ref} n_{\rm ref}^{19} \over T_{\rm ref}^k}  {n_{Rb} Z_a^2 Z_b^2 \ln \Lambda^{a/b} 
\over T_{Ra}^2}
\ee
where $R_{\rm ref}$ must be given in meters, $n_{\rm ref}^{19}$ is the reference density in units of $10^{19}$
$m^{-3}$, and $T_{\rm ref}^k$ is the reference temperature in units keV.

The relation between ($v,\theta$) and the normalized coodinates used in this document is 
\be 
v_{\rm th} v_{\parallel N} =  v \cos \theta \qquad 
v_{\rm th}^2 \mu_N = {v^2 \sin^2 \theta \over 2 B_N}
\ee 
From this one can derive 
\be 
{\partial \over \partial v} = {1\over v_{\rm th} }\cos \theta {\partial \over \partial v_{\parallel N} } 
+ {v \sin^2 \theta \over v_{\rm th}^2 B_N} {\partial \over \partial \mu_N} 
\ee
\be 
{\partial \over \partial \theta} = - {v \sin \theta \over v_{\rm th}} {\partial \over \partial 
v_{\parallel N}} + {v^2 \sin \theta \cos \theta \over v_{\rm th}^2 B_N } {\partial \over \partial \mu_N}
\ee
These allow for a coordinate transformation to the $(v_{\parallel N}, \mu_N)$ coordinate system. 
For the pitch angle scattering one arrives at 
\bee 
C_{\theta N} =  &\rightarrow & 
{2 v_R} {\partial \over \partial v_{\parallel N}} \biggl [ {\mu_N D_{\theta \theta} \over 
v_{\parallel N}^2 + 2 \mu_N B_N} \biggl ( B_N {\partial f\over \partial v_{\parallel N}} - v_{\parallel N} 
{\partial f\over \partial \mu_N} \biggr ) \biggr ] \cr 
\noalign{\vskip 0.2 truecm} 
&+& {2 v_R \over  B_N} {\partial \over \partial \mu_N} 
\biggl [ {v_{\parallel N} \mu_N D_{\theta \theta} \over v_{\parallel N}^2 + 2 \mu_N B_N} \biggl ( - B_N 
{\partial f \over \partial v_{\parallel N}} + v_{\parallel N} {\partial f\over \partial \mu_N} \biggr ) \biggr ] 
\eee 
The energy scattering yields 
\bee 
C_{vvN} & \rightarrow & 
v_R {\partial \over \partial v_{\parallel N} } 
\biggl [  {v_{\parallel N } D_{vv} \over v_{\parallel N}^2 + 2 \mu_N B_N} \biggl ( v_{\parallel N} {\partial 
f \over \partial v_{\parallel N}} + 2 \mu_N {\partial f \over \partial \mu_N} \biggr ) \biggr ] \cr 
\noalign{\vskip 0.2 truecm}
&+& v_R {\partial \over \partial \mu_N} \biggl [ {2 \mu_N D_{vv} \over v_{\parallel N}^2 + 
2 \mu_N B_N} \biggl ( v_{\parallel N} {\partial f \over \partial v_{\parallel N}} + 2 \mu_N {\partial f \over 
\partial \mu_N}\biggr ) \biggr ] 
\eee
The slowing down can be written as 
\bee 
C_{vN}  &\rightarrow & -
v_R {\partial \over \partial v_{\parallel N} } \biggl [  {v_{\parallel N} F_v \over [ v_{\parallel N}^2 
+ 2 \mu_N B_N ]^{1/2}} f  \biggr ] \cr
\noalign{\vskip 0.2 truecm}
& & - {1 \over v_{\rm th}} {\partial \over \partial \mu_N} \biggl [ {2 \mu_N F_v \over [v_{\parallel N}^2 + 
2 \mu_N B_N]^{1/2}} f \biggr ] 
\eee

Of course, the collision operator above does not conserve momentum. Below we describe a simplified model 
that generates momentum conservtion. Note though, that this is not the exact linearized Fokker Planck 
operator. We are interested in momentum conservation for like particle collisions. The change in parallel 
momentum due to the collision operator above is 
\bee 
&& 2 \pi B_N \int {\rm d}v_\parallel \int {\rm d} \mu_N v_{\parallel N} C_N(f) = \cr
\noalign{\vskip 0.2truecm}
&& - 4 \pi B_N v_R \int {\rm d} v_{\parallel N} {\rm d} \mu_N {\mu_N D_{\theta \theta}^{a/a} 
\over v_{\parallel N}^2 + 2 \mu_N B_N } \biggl ( B_N {\partial f_N \over \partial v_{\parallel N} } 
- v_{\parallel N} {\partial f_N \over \partial \mu_N} \biggr ) 
\eee 
This loss of momentum can be corrected for by adding an ad hoc correction term 
\be 
{\partial f_{Nc} \over \partial t_N} = C v_{\parallel N} F_{MN} 
\ee
The constant $C$ then follows from working out the momentum change due to this term 
\be 
2 \pi B_N C \int {\rm d} v_{\parallel N} {\rm d} \mu_N v_{\parallel N}^2 F_{MN} 
\ee 
and putting it equal to minus the term due to the pitch angle scattering.

\subsection{The effect of each of the terms}

In the previous section we derived the form of the three terms of the Fokker-Plank operator.  Here we will show, using data produced by GKW and our implementation of the previous to show what effect the operator has on the velocity space structure.

When running GKW, collisions are included if \texttt{collisions} $=$  \texttt{.true.} in the CONTROL namelist

\subsubsection{Pitch-angle scattering}

This effect is switched on by setting \texttt{pitch\_angle} $=$ \texttt{.true.} in the collisions input deck.  Pitch-angle scattering conserves momentum and therefore will have the effect of isotropising the distribution function along arcs of constant velocity $v = \sqrt{v_{\parallel}^{2} + 2\mu B}$.

If we initialise the code with narrow, shifted gaussian in velcity space, then as time progresses then pitch-angle scattering will have the effect of producing a ring in velocity space (As seen in figure \ref{pitch-angle}).

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{pitchangle.eps}
\caption{Snapshot of velocity space as three different time intervals showing the effect of pitch angle scattering on the velocity space.  Coordinates are $v_{\parallel}$ horizontally and $v_{\perp}$ vertically. All data shown in this chapter was run with 100 parallel velocity points and 50 $\mu$ points non-equally spaced.\label{pitch-angle}}
\end{center}
\end{figure}

\subsubsection{Energy scattering}

This effect is switched on by setting \texttt{en\_scatter} $=$ \texttt{.true.} in the collisions input deck.  The energy scattering term spreads the distribution function in energy.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.9\textwidth]{enscatter.eps}
\caption{Snapshot of velocity space as four different time intervals showing the effect of energy cattering on the velocity space. Coordinates are $v_{\parallel}$ horizontally and $v_{\perp}$ vertically. }
\end{center}
\end{figure}

\subsubsection{Collisional friction}

This effect is switched on by setting \texttt{friction\_coll} $=$ \texttt{.true.} in the collisions input deck.  The collisional friction term advects the distribution function towards the origin in velocity space.

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{collfric.eps}
\caption{Snapshot of velocity space at four different time intervals showing the effect of the collisional friction term on the velocity space.}
\end{center}
\end{figure}

When you combine all three effects we get an evolution of the distribution function towards a maxwellian centred on zero and with a width of a single normalized thermal velocity.

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{fulloper.eps}
\caption{Snapshot of velocity space as four different time intervals showing the effect of all the terms in the collison operator on the velocity space.  The final state is a maxwellian centred on zero.}
\end{center}
\end{figure}

\subsection{Velocity space boundary conditions}

There are currently two options for the boundary conditions in velocity space.  The input deck parameter \texttt{mass\_conserve} changes between open boundary conditions (when set to \texttt{.false.} and zero flux boundary conditions when set to \texttt{.true.} the former being slightly more physical, but the latter being desireable also.  Conservation is down to machine precision when the mu grid is uniform which is not the default.  When the mu grid is non-unifom (which is the usual case unless changed at compile time in velocity\_grid) the mass is conserved down to the fifth decimal place, although this can improve if a greater resolution is used.

\subsection{User defined collision-frequencies}

The collision frequency can be defined in two ways.  Firstly by specifying \texttt{nref} which is the particle density in units of $10^{19}$ particles per $m^{3}$, \texttt{tref}, the temperature in keV and \texttt{rref} which is the toroidal major radius.  These parameters are what the corresponding parameters are normalized to in the code and are never usually specified.  The normalized collision (normalized to the ratio of the thermal velocity to the major radius) frequency is given by,

\be 
\Gamma^{a/b}_N = {R_{\rm ref} n_b Z_a^2 Z_b^2 e^4 \ln \Lambda^{a/b} \over 4 \pi \epsilon_0^2 m_a^2 v_{tha}^4} = 
6.5141\cdot 10^{-5} {R_{\rm ref} n_{\rm ref}^{19} \over T_{\rm ref}^k}  {n_{Rb} Z_a^2 Z_b^2 \ln \Lambda^{a/b} 
\over T_{Ra}^2}
\ee

However if \texttt{freq\_override} $=$ \texttt{.true.} then,

\be
\Gamma^{a/b}_N =  n_b Z_a^2 Z_b^2 \nu_{\rm ref} n_b \Lambda^{a/b}/ T_{Na}^2 \Lambda^{i/i}
\ee

Where $\nu_{\rm ref}$ is the user defined frequency defined as \texttt{coll\_freq} in the input deck.  This is assumed to be the single charged ion - ion (proton/deuteron) collision frequency.  To get agreement with the first method this can be set to $6.5141\cdot 10^{-5}R_{\rm ref}n_{\rm ref}\Lambda^{i/i}/T_{\rm ref}^{2}$.  This value is then scaled to all the other species respectively.  For this to work the first species in the input deck should be singly charged ions.

\begin{footnotesize}
\begin{verbatim}
&COLLISIONS
 rref = 1   !reference major radius
 tref = 1   !reference temperature in units of kev
 nref = 1   !reference density in units 10^19 m^-3
 pitch_angle = .true.           !Switches for the pitch angle scattering
 en_scatter = .true.            !Energy scattering term 
 friction_coll = .true.         !Collisional friction term
 mom_conservation = .false.     !Use the correction to conserve momentum
 mass_conserve = .true.,        !Forces zero flux boundary conditions on velocity space
 freq_override=.false.          !Manually specify ii collision frequency to override ref values
 coll_freq=6.515E-5             !Ion-ion collision frequency used if freq_override=.true. 
\end{verbatim}\end{footnotesize}

\section{E x B background shear}%Last updated by Francis Jul 2009

%From CPC paper.\section{Background $E \times B$ shear flow}
The co-moving system employed in GKW corresponds to a frame that rotates as a rigid body with constant frequency 
$\Omega$. 
The possible radial gradient in the toroidal rotation is then treated through a radial gradient in the averaged parallel velocity 
of the background. 
The radial gradient of the perpendicular velocity has so far been ignored, but is known to play an important role in the 
saturation of the turbulence. 
The sheared $E\times B$ velocity leads to a stabilisation of turbulence through the breaking up of eddy structures. 
This $E\times B$ shearing is always present for a purely toroidally rotating plasma but can, of course, also be related 
to a sheared poloidal rotation. 
In this section it is described how the $E \times B$ shearing is treated in GKW.

The equilibrium $E \times B$ rotation 
\be
{\bf v}_s(\psi)= {{\bf b} \times \nabla \Phi \over B} 
\ee
lies inside the flux surface.  The shearing rate in the normalised units is defined to be 
\be
\gamma_{E}^N = {1 \over 2}\rho_*^2 {\partial^2 \Phi_N \over \partial \psi^2} ,
\label{gamma_E}
\ee
where the normalisation $\Phi= \rho_* {T_{\rm ref} \over e}\Phi_N$ is as for the
perturbed potential Eq.~(\ref{phi_norm}).  The factor $1/2$ is present due to the definition of the
reference thermal velocity.
In physical units the shearing rate is
\begin{equation}
\gamma_E = {v_{\rm thref} \over R_{\rm ref}} \gamma_E^N = 
{1 \over B_{\rm ref}}{\partial^2 \Phi \over \partial r^2} ,
%Contingent on \phi=r/R_{ref}
%True if r = R_{max}-R_{min}
\end{equation}
where $r=(R_{\rm max}-R_{\rm min})/2$.  The GKW shearing rate is assumed
radially constant and is defined as a flux function.  At the radial centre of the flux tube
$\bf{v_s}$ is chosen to be zero, with the result that there is no net flow over the domain.  In the
local limit with the approximations of the `$s-\alpha$' model geometry
the shearing rate is then equivalent to the familiar definition \cite{Hahm95,Burrell97}:
\begin{equation}
\gamma_E \approx {(RB_p)^2 \over B_t}{\partial^2 \Phi \over \partial \Psi^2}
\approx {d v_s \over d r} .
\end{equation}
As before the index $N$ is dropped in what follows.
The sign convention here is opposite to Eq.~(\ref{gradients}), so $\gamma_E<0$ corresponds to 
$\nabla E_0$ radially outwards.

Only the Doppler shift of the background $E \times B$ rotation is kept in the description, i.e. the
background $E\times B$ rotation is 
added as an additional convective term for the perturbed distribution (${\bf v}_s \cdot \nabla g$) in the gyrokinetic equation~(\ref{complete_set_eq}), 
and we neglect the acceleration due to the background potential.  
With zero net flow across the flux tube one obtains
\be
\label{IX}
{\rm IX} = -{\bf v}_s \cdot \nabla g \rightharpoonup - \rho_*^2 {\partial \Phi \over \partial
\psi}{\mathcal E}^{\psi \zeta}{\partial g \over \partial \zeta} =
 - 2 {\gamma_E \psi}{\mathcal E}^{\psi \zeta}  {\partial g \over \partial \zeta} .
\ee
Defining $\bar \gamma_E= 2 \gamma_E {\mathcal E}^{\psi \zeta} $ and taking the Fourier transform, the term can be written as a derivative in
Fourier space
\be 
{\mathcal T}(-{\bar \gamma_E \psi}{\partial g \over \partial \zeta}) = \bar \gamma_E
k_\zeta {\partial \hat g \over \partial k_\psi} .
\ee
The derivative represents a continuous advection (and shearing) in Fourier space in $k_\psi$.

There are a number of subtleties associated with numerical evaluation of this derivative. A finite difference approximation cannot resolve the fine 
scale structure in Fourier space.  The `harmonic derivative' \cite{Waltz94,Miller94} is a full convolution over all modes and is equivalent to a
pseudo-spectral implementation using FFTs (as was used for the nonlinear term in Eq.~(\ref{nl_term})).  
Although it correctly implements the $E \times B$ shearing term inside the computational domain, it does not represent a homogeneous shear 
flow because the flow profile does not satisfy the periodicity of the discrete Fourier transform; any turbulent structure passing through the radial
boundary will experience a discontinuity in the flow profile (which in the periodic domain has a sawtooth form).  
Since the local limit requires that the turbulence has no preference for the boundary, periodicity at the boundary must be preserved. 
The `shear-periodic' boundary condition \cite{Baron82,Schumann85,Gerz89} that moves with the mean flow accomplishes this for a finite 
difference representation in the radial direction, but does not naturally translate to the spectral representation.

In coordinates that move with the flow  \cite{Rogallo81,Zang88,Pumir96}, the `radial' wavenumbers become time dependent:
\bee
\psi^\prime = \psi , &\qquad& \qquad \zeta^\prime=\zeta-\psi \bar \gamma_E t ,\\
k_\zeta^\prime = k_\zeta , &\qquad& \qquad  k_\psi^\prime = k_\psi - k_\zeta \bar \gamma_E t .
\eee
In these coordinates, the derivative in Fourier space becomes part of the time derivative.

For a gyro-kinetic code, a time dependent wave vector requires the re-evaluation of the linear terms and Bessel functions at every time-step
and would be computationally prohibitive. 
By discretising the time dependence of the wavenumbers as a remapping of the solution vector between the original fixed wavevectors, 
this expensive re-evaluation can be avoided \cite{HammettAPS06}. 
Using only the fixed wavevector grid, the advection in Fourier space occurs in jumps only when the boundary periodicity is satisfied for a 
given mode. 
Explicitly, the remapping $\hat g (k_\psi, k_\zeta,s) \rightarrow \hat g (k_\psi-\Delta k_\psi, k_\zeta,s)$ occurs when
\be
k_\zeta \bar \gamma_E(t-t_{\rm last remap}) = \Delta k_\psi /2 .
\ee
At the high $k_\psi$ limit the solution vector is discarded.  
This means that the method is non conservative, but since the turbulence is characterised by a peaked spectrum 
(Fig.~\ref{cyclone_linear}), the losses are negligible if the range of radial wavenumbers is suitably wide.  
For numerical stability, the remapping must occur at the same time for all points along the flux tube. 
As the metric tensor ${\mathcal E}^{\psi \zeta}$ is a flux function, see Eq.~(\ref{efun_cst}), this condition is always satisfied for a 
constant shear rate along the field line.

The method in GKW differs from the standard spectral methods to model homogeneous shear flows in fluid turbulence of
Refs.~\cite{Rogallo81,Zang88,Pumir96}, where the wavenumbers must be recalculated and re-meshing occurs for all wavevectors 
at the same time. 
In GKW the method used is the one proposed by Hammett et al. in Ref.~\cite{HammettAPS06}.  
Here the `re-meshing' occurs continually (and at different times for each $k_\zeta$), and the wavenumbers stay on their original fixed grid.  
The accuracy of the method has been argued to be second order on average \cite{HammettAPS06} and
is able to capture the physics effects of a 
background shear flow whilst allowing the desirable features of the flux tube model to be retained.

Convergence of the remap method can be (and should be) checked (particularly by for the modes with lowest $k_\zeta$) by decreasing 
$\Delta k_\psi / k_{\zeta {\rm min}}$ (increasing $N_x$ whilst holding $N_{\rm mod}$ constant (defined in the next section)).

%As in CPC paper:
% For purely toroidal rotation (with no poloidal flows) the poloidal components of the parallel flow
% $u_\parallel=R\Omega B_t / B$ and the 
% $E \times B$ flow ${\bf v}_s$ must cancel:
% \be
% v_s = - {B_p \over B_t}u_\parallel = - {B_p R\Omega \over B} . 
% %YC = - {B_p u\over B}
% \ee
% For $s_B=1$, $v_s>0$ indicates a flow in increasing poloidal direction, ${\bf v}_s\cdot \nabla
% \theta>0$. 
% For $s_j=1$, ${\bf B}\cdot \nabla \theta>0$, with the poloidal component of the magnetic field
% positive, $B_p>0$. 
% For the gradients
% \be
% \gamma_E = - {\partial \over \partial \psi} \left({B_p R_{\rm ref} \over |\nabla
% \psi|}\Omega\right) \approx  {B_p u^\prime \over B} ,
% \ee
% since the definitions (\ref{gradients}) and (\ref{gamma_E}) have opposite sign.

%FC: I have sketches as jpeg to illustrate both sign and general geometry coupling.

%FC: Post CPC paper - Oct 09
\subsection{Purely toroidal sheared rotation in general geometry}

Here we follow the notation in Peeters (2009) on centrifugal effects \cite{PEE09}
and the normalisations and definitions of the CPC paper.  We also note that by definition
$\vec{\Omega}=-s_B \Omega \nabla z$.  We denote the plasma rotation with $\omega(\psi)$ and the
rigid frame rotation with $\Omega$, with both normalised by $v_{\rm thref} / {R_{\rm ref}}$.  Note
that in the CPC paper no distinction is made, but here we use both to avoid ambiguity.

In the case of purely toroidal rotation (in the laboratory frame in non-normalised units)

\begin{equation}
u_\parallel \vec{b} + {\bf v}_s  = s_B R \omega(\psi) R \nabla \phi
\label{toroidal}
\end{equation}
where $\phi$ is the toroidal angle.  From the normalisation definitions above one can show that
\begin{equation}
 {\bf v}_s = \left({1 \over 2} \rho_*^2 v_{\rm thref}\right) {\vec{b} \times \nabla_N \psi \over
B_N} {\partial \Phi^N \over \partial \psi}
\end{equation}
hence in the normalised units, taking the bi-normal component of \ref{toroidal} one finds
\begin{equation}
 u_{\parallel N} \underbrace{\vec{b} \cdot \nabla_N \zeta}_{=0} +  {1 \over 2}
\rho_*^2 \underbrace{{\vec{b} \times \nabla_N \psi \over B_N}\cdot \nabla_N
\zeta}_{=\mathcal{E}^{\psi \zeta}}{\partial \Phi_N \over \partial \psi} = s_B R_N^2 \omega_N(\psi)
\underbrace{\nabla_N \phi \cdot \nabla_N \zeta}_{=-1/2\pi R_N^2}
\end{equation}
hence
\begin{equation}
 {1 \over 2} \rho_*^2 {\partial \Phi_N \over \partial \psi} = - {s_B \over 2 \pi}{1 \over
\mathcal{E}^{\psi \zeta}}\omega_N(\psi)
\end{equation}
taking the radial derivative
\begin{equation}
 {1 \over 2} \rho_*^2 {\partial^2 \Phi_N \over \partial \psi^2} = - {s_B \over 2 \pi}\left({1 \over
\mathcal{E}^{\psi \zeta}}{\partial \omega_N \over \partial \psi} + \omega_N{\partial \over \partial
\psi} \left({1 \over \mathcal{E}^{\psi \zeta}}\right)\right)
\end{equation}
Finally we transform to the rotating (starred) frame in which
$\omega_N^*(\psi)=\omega_N(\psi)-\Omega_N$ such that $\omega_N^*(\psi)=0$ (and hence ${\bf
v_s}=0$)
at the centre of the radial domain.  Since the frame rotates rigidly, ${\partial \Omega_N \over
\partial \psi}=0$. The tensor $\mathcal{E}$ is invariant under the transformation (check??).  The
electric field in the rotating frame transforms \cite{PEE09} to 
\begin{equation}
\Phi_N^*=\Phi_N + \underbrace{s_B s_j}_{check ??}{2 \over \rho_*^2}\Psi_N \Omega_N 
\end{equation}
where $\Psi = R_{\rm ref}^2 B_{\rm ref}\Psi_N$ is the poloidal flux and the factor of $2 /
\rho_*^2$ arises from the normalisation of $\Phi$.  In the rotating frame the coupling condition
becomes
\begin{equation}
 \underbrace{{1 \over 2} \rho_*^2 {\partial^2 \Phi_N^* \over \partial \psi^2}
-\underbrace{s_B s_j}_{??}  \underbrace{\Omega_N}_{ \rm VCOR} \overbrace{\partial^2 \Psi_N \over
\partial \psi^2}^{\rm New: d2pffun}}_{\rm
SHEAR\_RATE^{LAB}} = - {s_B \over 2 \pi}{1 \over
\mathcal{E}^{\psi \zeta}}\underbrace{\partial \omega_N^* \over \partial \psi}_{\rm UPRIM} +
\Omega_N{\partial \over \partial \psi} \left({1 \over \mathcal{E}^{\psi \zeta}}\right) 
\end{equation}
Since $\Psi=f(\psi)$ only, from the relation 
\begin{equation} 
{\cal E}^{\psi \zeta} = \frac{s_{\rm j}}{4\pi}\frac{\partial \psi}{\partial \Psi_N}
\label{efun_cst}
\end{equation}
it follows that
\begin{equation}
 {\partial \over \partial \psi} \left({1 \over \mathcal{E}^{\psi \zeta}}\right) = s_j 4 \pi 
{\partial^2 \Psi_N \over \partial \psi^2}
\end{equation}
hence
\begin{equation}
 \underbrace{{1 \over 2} \rho_*^2 {\partial^2 \Phi_N^* \over \partial \psi^2}}_{\rm SHEAR\_RATE^*}
+s_B s_j \underbrace{\Omega_N}_{\rm VCOR} {\partial^2 \Psi_N \over
\partial \psi^2} = - {s_B \over 2 \pi}{1 \over
\mathcal{E}^{\psi \zeta}}\underbrace{\partial \omega_N^* \over \partial \psi}_{\rm UPRIM}
\end{equation}
All quantities in the above equation are flux functions, and the shear rate will always be a
flux function as required for the discrete remapping method.   At version 1135 this
\texttt{TOROIDAL\_SHEAR} coupling is only implemented for \texttt{VCOR}=0.

\begin{footnotesize}
\begin{verbatim}

&ROTATION	       !Namelist read in rotation.F90 -> read_rotation
 VCOR = 0.0            !Rotation of the plasma vcor = Rref Omega / vthref
                       !Always in direction of toroidal magnetic field if positive.
 shear_rate = 0.3      !Normalised shearing rate for the E x B perpendicular shear 
 shear_profile  = 'none'    !To include a shear flow, use `wavevector_remap'
                            !discrete mapping of wavevectors (use NX >> NMOD ) 
 toroidal_shear = 'none'    !Override uprim inputs to be toroidally consistent with shear_rate
                            !'use_uprim': shear_rate derived from input uprim for species 1.
                            !'use_shear_rate': uprim derived from input shear_rate.
/
\end{verbatim}
\end{footnotesize}

\begin{thebibliography}{10}

\bibitem{LIT83} R. G. Littlejohn, J. Plasma Phys. {\bf 29}, 111 (1983).
\bibitem{HAH88} T. S. Hahm, Phys. Fluids {\bf 31}, 2670 (1988)
\bibitem{BRI88} A. Brizard, Phys. Plasmas {\bf 41}, 541 (1988)
\bibitem{SUG00} H. Sugama, Phys. Plasmas {\bf 7}, 466 (2000).

\bibitem{KAR86} C.F.F. Karney, {\sl Fokker-Planck and quasi-linear codes}, Comp. Phys. Reports {\bf 4} 183 
(1986) 
% E x B shear method
\bibitem{Hahm95}
T.~S. Hahm and K.~H. Burrell.
\newblock {\em Phys. Plasmas}, {\bf 2} (1995) 1648

\bibitem{Burrell97}
K.~H. Burrell.
\newblock Phys. Plasmas, {\bf 4} (1997) 1499

\bibitem{Waltz94}
R.~E. Waltz, G.~D. Kerbel, and J.~Milovich, Phys. Plasmas {\bf 1} (1994) 2229

\bibitem{Miller94}
R.~L. Miller and R.~E. Waltz, Phys. Plasmas {\bf 1} (1994) 2835

\bibitem{Baron82}
F.~Baron. {PhD} in physics, (Univ. Pierre et Marie Curie, Paris, France, 1982)

\bibitem{Schumann85}
U.~{Schumann}, {Algorithms for direct numerical simulation of shear-periodic
  turbulence}, in {\em Lecture Notes in
  Physics {\bf 218}} (Berlin Springer Verlag 1985) pages 492

\bibitem{Gerz89}
T.~Gerz, U.~Schumann, and S.~E. Elghobashi, Journal of Fluid Mechanics Digital Archive {\bf 200} 
(1989) 563

\bibitem{Rogallo81}
R.~S. {Rogallo}, {\em Numerical experiments in homogeneous turbulence}.
\newblock {\em NASA STI/Recon Technical Report N} (1981) 81:31508

\bibitem{Zang88}
T.~A. {Zang}, Applied Numerical Mathematics {\bf 7} (1991) 27

\bibitem{Pumir96} Alain Pumir, Phys. Fluids, {\bf 8} (1996) 3112

\bibitem{HammettAPS06}
G.~W. {Hammett}, W.~{Dorland}, N.~F. {Loureiro}, and T.~{Tatsuno}.
\newblock {Implementation of Large Scale E x B Shear Flow in the GS2
  Gyrokinetic Turbulence Code}.
\newblock {\em APS Meeting Abstracts} (2006) pages 1136 % E x B shear method
\bibitem{Hahm95}
T.~S. Hahm and K.~H. Burrell.
\newblock {\em Phys. Plasmas}, {\bf 2} (1995) 1648

\bibitem{Burrell97}
K.~H. Burrell.
\newblock Phys. Plasmas, {\bf 4} (1997) 1499

\bibitem{Waltz94}
R.~E. Waltz, G.~D. Kerbel, and J.~Milovich, Phys. Plasmas {\bf 1} (1994) 2229

\bibitem{Miller94}
R.~L. Miller and R.~E. Waltz, Phys. Plasmas {\bf 1} (1994) 2835

\bibitem{Baron82}
F.~Baron. {PhD} in physics, (Univ. Pierre et Marie Curie, Paris, France, 1982)

\bibitem{Schumann85}
U.~{Schumann}, {Algorithms for direct numerical simulation of shear-periodic
  turbulence}, in {\em Lecture Notes in
  Physics {\bf 218}} (Berlin Springer Verlag 1985) pages 492

\bibitem{Gerz89}
T.~Gerz, U.~Schumann, and S.~E. Elghobashi, Journal of Fluid Mechanics Digital Archive {\bf 200} 
(1989) 563

\bibitem{Rogallo81}
R.~S. {Rogallo}, {\em Numerical experiments in homogeneous turbulence}.
\newblock {\em NASA STI/Recon Technical Report N} (1981) 81:31508

\bibitem{Zang88}
T.~A. {Zang}, Applied Numerical Mathematics {\bf 7} (1991) 27

\bibitem{Pumir96} Alain Pumir, Phys. Fluids, {\bf 8} (1996) 3112

\bibitem{HammettAPS06}
G.~W. {Hammett}, W.~{Dorland}, N.~F. {Loureiro}, and T.~{Tatsuno}.
\newblock {Implementation of Large Scale E x B Shear Flow in the GS2
  Gyrokinetic Turbulence Code}.
\newblock {\em APS Meeting Abstracts} (2006) pages 1136

\bibitem{PEE09} A.G. Peeters, D. Strintzi, C. Angioni, et al., Phys. Plasmas {\bf 16} (2009) 042310

\end{thebibliography}

\chapter{Tearing modes}

Here are some brief notes that describe how a magnetic island is included in the current version of GKW.

If we initially assume a constant poloidal flux perturbation with a given helicity 
\be 
A_\parallel = C \exp [ 2 \pi {\rm i} (m s - n \gamma ) ] 
\ee 
where s and $\gamma$ are the poloidal and toroidal angles, respectively.  If we introduce the helical coordinate $\zeta = qs - \gamma$ and expanding the safety factor up to first order
\be 
q = m / n + \psi\partial q /\partial\psi
\ee 
in the centre of the box (i.e. the mode is resonant at this position). We need to transform this
representation to the ($\psi, \zeta, s)$ coordinate system which gives,
\be 
A_\parallel = C \exp [ 2 \pi {\rm i} ( n \zeta - {\partial q \over \partial \psi} n s \psi ) ] 
\ee 
From which it directly follows that 
\be 
k_\zeta = 2 \pi n \rho_* 
\ee
From the periodicity contraint connected with the parallel boundary conditions it then follows 
that we have to keep radial modes with ($N = ikxspace$),
\be 
k_\psi = (p/N) k_\zeta {\partial q \over \partial \psi} = (p/N) 2 \pi n \rho_* {\partial q \over \partial \psi} 
\ee 
where $p$ is an integer. From $p = 1$ one directly obtains the radial size of the box 
\be 
\psi \rightarrow \biggl [ - { N \over 2 n \partial q / \partial \psi} , {N \over 2 n \partial q / \partial \psi} \biggr ]  
\ee

The Fourier amplitudes of the the different radial modes then can be calculated by integrating over 
the radial domain 
\be 
\hat{A}(k_\psi, k_\zeta, s) = C \int {\rm d} \psi \exp \biggl [ - 2 \pi {\rm i} n {\partial q \over \partial \psi}
\psi [ p + s] \biggr ] 
= {C N \over \pi n \partial q / \partial \psi} {\sin [ \pi (p + Ns) ] \over p + Ns } 
= C^*N {\sin [ \pi s^\circ ] \over s^\circ} 
\ee 
where we have use the ballooning angle $s^\circ = N*s + p$ in the last step.

One point to note at this point is that this perturbation is not periodic in the domain of GKW for all values of s.  This manifests itself as a discontinuity in the parallel vector potential at the edge of the computational domain.  A solution to the this problem is to relax the constant flux approximation and therefore write the perturbation as,
\be 
A_\parallel = C \exp{\imath k_{\zeta}\zeta}\sum_{p=0}^{\inf} A^{p}(s)\exp{\imath pk_{\psi}\psi}
\ee  
Where $A^{p}(s)$ is a damping function.  We have chosen a Gaussian profile to reduce any ringing.  This gives,
\be
\exp{\left((-(sN+p)^{2}/L^{2})\right)}\frac{\sin{\pi(sN+p)}}{\pi(sN+p)}
\ee
Where L is half width of the gaussian envelope (This is set to 2.0 by default).  This approximation of the leads to a more satisfactory shape for the vector potential, where the value of L can be tuned to that the damping of the small k modes is negligible and the of high k modes sufficuiently strongly to prevent the occurance of a jump at the boundary.

Thus the full vector potential has the form,

\be
A_{||} = C\exp{(\imath k_{\zeta}\zeta - \imath\omega t)}\sum_{p=-Nx/2}^{p=Nx/2}\exp{\left(-\frac{(sN+p)^2}{L^2}\right)}\frac{\sin{(\pi(sN+p))}}{\pi(sN+p)}\exp{(\imath p k_{\psi}\psi)}
\ee

Note that only a single binormal mode is used.  The island width, W, is converted to the vector potential by using the expression,
\be
w = 2\sqrt{\left[\frac{rqB_{r}}{mB_{\theta}(dq/d\psi)}\right]}
\ee
where $B_{\phi}$ and $B_{\theta}$ is the toroidal and poloidal components of the magnetic field respectively.  $dq/d\psi$ is the magnetic shear and n is the poloidal mode number.  Furthermore we use the fact that the radial magnetic field is prescribed as $B_{r}= m\tilde{\psi} /rR$, and that in a circular cross-section plasma equilibrium the perturbed magnetic flux is given by $\tilde{\psi} = R\tilde{A_{\parallel 0}}$.  This allows us to prescribe the correct island width within GKW which is normalised by $\rho_{ref}$ to give the parameter wstar.

\begin{figure}
\begin{center}
\includegraphics[width=0.45\textwidth]{MagIslGeom.eps}
\caption{Magnetic Island Geometry for an island with a half width of 30 ion gyro-radii}
\label{islandgeom}
\end{center}
\end{figure}

To run GKW with a magnetic island present the following options must be used.  Firstly the code must be set to run electro-magnetically by setting {\bf nlapar}$=$.true..

In the spcgeneral common block, {\bf finit} must be set to {\it island}, which adds the magnetic island as a perturbation of the equilibrium field.  A further input parameter is the rotation frequency of the island, which is set by the parameter {\it isl\_rot\_freq} which must be entered normalised, when plotting XY profile outputs, setting {\it list\_follow} in the diagnostic namelist will follow the magnetic island so that the O-point is always in the centre of the box.

\begin{verbatim}
 &SPCGENERAL
 beta = 0.0001,
 finit = 'island',
 wstar = 5.0
 adiabatic_electrons = .false.
 Ls = 2.0,
 isl_rot_freq = 0.1,
 /
\end{verbatim}

The half width of the island is determined by the input parameter {\bf wstar} which is the island width normalised to the ion gyro-radius.  A point of note is that this island width can not exceed the radial extent of the computational box due to the periodicity constraints of being a pseudo-spectral code.

\subsection{Profile flattening in GKW}

The full benchmark of the magnetic islands implementation requires the flattening of the temperature and density profiles.  The equilbrium profiles are linear and are present in the source terms.  Since we are solving for the perturbed part of the distribution function, the solution should respond to the source terms due to the background gradients and the magnetic island giving a perturbuation that cancels the background out through the islands O-point.

Runs must be non-linear for the correct level of flattening to be obtained.

\subsubsection{Reconstruction of radial density and temperature profiles}

To reconstruct the radial density profile firstly the XY diagnostics must enabled in the diagnostics namelist.

\begin{verbatim}
&DIAGNOSTIC
 lisl_follow = .false.,
 lphi_diagnostics = .true.,
 xy_phi = .true.,
 xy_apar = .true.,
 xy_fluxes = .true.,
 xy_fluxes_em = .true.,
 xy_dens = .true.
 xy_current = .true.,
 xy_temp = .true.,
 / 
\end{verbatim}

The files named Densi000*\_* hold the XY perturbed density data ($\tilde{n}$) for each large time step.  The background density profile can be reconstructed by using the date in the filed named {\it xphi} (Radial coordinate values same dimensions as XY slices) and the input parameter RLn (different for each species), thus the full density is:
\be
n({\Delta\psi}) = \tilde{n} - RLn*xphi
\ee

The temperature profiles are slightly more complicated.  The files Tempe000*\_* hold the second moments of the distribution function (Energy, $\tilde{E}$), thus to get the perturbed temperature $\tilde{T}$ 
\be
\tilde{T} = (1/n_{ref})(\frac{2}{3}\tilde{E} - T_{ref}\tilde{n})
\ee
which in the same way as the total density, can be made into the total temperature by,
\be
T({\Delta\psi}) = \tilde{T} - RLT*xphi
\ee

\begin{figure}
\begin{center}
\includegraphics[width=0.95\textwidth]{tempflat.eps}
\caption{Radial temperature profiles through the O-point (Blue) and X-point (Red) in the presence of a magnetic island.  With the background (Green) temperature.}
\end{center}
\end{figure}

\cleardoublepage

%\pagenumbering{arabic}

\vfill \eject 
\hbox to 2 truecm{\hfill}
\vskip 8 truecm

\centerline{\selectfont \Huge I AM A FREAK} 
\addcontentsline{toc}{chapter}{I AM A FREAK}

\vfill  \eject

\chapter{Diagnostics}

\section{fluxes} %This section has been consolidated with the separate one on the fluxes routine
\label{fluxes}
\index{fluxes}

The fluxes are calculated as guiding center fluxes and are given by the equations 
\be 
\Gamma_i^\psi  = \biggl \langle \int {\rm d}^3 {\bf v} \tilde {\bf v}_E\cdot \nabla \psi  \alpha_i f \biggr \rangle 
\ee
where i = (1,2,3) with $\alpha = (1, mv^2/2,v_\parallel)$, i.e. $\alpha = 1$ is the 
particle flux, $\alpha = 2$ is the heat flux and $\alpha = 3$ is the parallel momentum 
flux. The big angle brackets denote the flux surface average.

Both the potential as well as the distribution function are represented as a sum over 
Fourier modes. Since the flux contains an average over space it will be zero unless 
the wave vectors of the potential and distribution function add up to be zero. In the 
representation of modes used in this manuscript this means that a particular wave 
vector of the representation of $f$ must be combined with the complex conjugate of 
the same wave vector in the representation of $\phi$, i.e.  
\be 
\langle \tilde v_E f \rangle  = 2 \sum_m {\rm Re}  [ \langle \tilde {\bf v}_{Em}^\dagger 
\cdot \nabla \psi  f_n \rangle ]  
\ee 
where $v_{En}$ ($f_n$) is the Fourier amplitude of the ExB velocity (distribution function) 
of the mode $n$, and the dagger denotes the complex conjugate. Substituting the 
various definitions one can derive that a normalized flux can be defined as 
\be 
{\cal I}_i = \sum_m \biggl \{ 2 \pi B {\cal E}^{\psi \beta} k_{\beta m } \int {\rm d} 
\mu {\rm d} v_{\parallel} \,  \alpha_i {\rm Im} [ \langle \phi_m \rangle^\dagger f_m ] \biggr \}
\ee
where $\alpha_i = (1, v_\parallel^2 + 2 \mu B,v_\parallel)$  
This definition means that the fluxes in physical units can be obtained from 
\be 
R_{\rm ref} \Gamma^\psi_s  = n_s \rho_*^2 v_{\rm thref} {\cal I}_1 
\ee 
\be 
R_{\rm ref} Q_s^\psi = n_s T_s \rho_*^2 v_{\rm thref} {\cal I}_2 
\ee 
\be 
R_{\rm ref} \Gamma^\psi_{\phi s} = m_s n_s v_{ths} \rho_*^2 v_{\rm thref} {\cal I}_3   
\ee

\subsection{Details of the routine}
The radial component of the ${\bf E} \times {\bf B}$ velocity and the distribution function are given by 
\be 
\tilde {v}_{Ex} = \sum_k  {\rm i} {k_y \phi_k \over B} {\bf e}_x \exp [ {\rm i} {\bf k} 
\cdot {\bf x} ] + c.c  
\ee
\be 
f = \sum_k f_k\exp [ {\rm i} {\bf k} \cdot {\bf x} ] + c.c  
\ee
Multiplying these two expressions and averaging over space yields 
\be 
v_{Ex} f = 2 \sum_k {k_y \over B} {\rm Im} ( \phi_k^\dagger f_k) 
\ee 
where the dagger denotes the complex conjugate.

Using all the normalizations that are described in section \ref{normalization} one can 
write the equation in the form 
\be 
\Gamma_i = n_0 \rho_*^2 v_{\rm th} \sum_k \biggl \langle \int {\rm d}^3 v_N^3 \, \alpha_i k_{yN} 
{\rm Im}(\phi^\dagger_{kN}  f_{kN} ) \biggr \rangle  =  n_0 \rho_*^2 v_{\rm th} {\cal I}(\alpha_i)
\ee 
The fluxes in the routine are calculated for every mode and stored in {\texttt pflux, eflux}
and {\texttt vflux}. These are defined as 
\be 
{\texttt pflux} = {\cal I}(1)
\ee
\be 
{\texttt eflux} = {\cal I}(v_N^2)
\ee
\be 
{\texttt vflux} = {\cal I}(v_{\parallel N} ) 
\ee
Consequently in physical units the fluxes are 
\be 
\Gamma = n_0 \rho_*^2 v_{\rm th} {\texttt pflux} 
\ee
\be 
Q = n_0 T_0 \rho_*^2 v_{\rm th} {\texttt eflux} 
\ee
\be 
\Gamma_\phi = m n_0 \rho_*^2 v_{\rm th} {\texttt vflux} 
\ee    
So to get the real flux one still needs to multiply with the density and / or temperature of the particular species.  The normalization, however, has the advantage that the diffusivity coeffcients are
\be
D       = \rho_*^2 L_N v_{\rm thref} pflux
\ee
\be
\chi     = \rho_*^2 L_T v_{\rm thref} eflux 
\ee
\be
\chi_{\phi} = \rho_*^2 L_u v_{\rm thref} v_R vflux
\ee
Note that the momentum flux must be multiplied with the relative velocity. The fluxes contain both the anomalous as well as the neo-classical fluxes. The anomalous flux contains the contributions due to the ExB velocity and the magnetic flutter.

For example from the heat flux output we can calculate the normalized species heat diffusivity
\be
\chi_s^N = {{\texttt eflux} \over R/L_{T_s} } = {R \chi_s  \over \rho_{\rm ref}^2 v_{\rm thref}} 
\ee

For a linear run, the dominant mode growth rate is calculated as
\be
\gamma_{N}^{max}(t_N)={\ln \left[{\sum_m |\hat \phi|^2(t_N) \over \sum_m |\hat \phi|^2(t_N-\Delta t_N)}\right] \over \Delta t_N}
\ee
The mode frequency is calculated as
\be
\omega_N(t_N)=\frac{1}{\Delta t_N}\left(\frac{\sum_s |\hat \phi(t_N)| {\mathrm Arg}({\hat \phi}(t_N))}{\sum_s|\hat \phi(t_N)|} - \frac{\sum_s |\hat \phi(t_N-\Delta t_N)|{\mathrm Arg}({\hat
\phi}(t_N-\Delta t_N))}{\sum_s |\hat \phi(t_N-\Delta t_N)|}\right).
\ee
Positive values indicate propagation in the direction opposite to $\nabla \zeta$ (which for $s_j=1$ is in the ion diamagnetic drift direction).

\begin{table}[h!]
\begin{tabular}{l|l|l|l}
 Filename & Description & Quantity & Cols,(Rows)\\
\hline
time.dat & Dominant linear mode growth rate & $t_N, \gamma_{N}^{max}, \omega_N$ & 3  \\

fluxes.dat & Total fluxes $(i=1,2,3)$ for each species  & ${\cal I}_i=\sum_{k_\zeta} \sum_{k_\psi} \sum_{s} {\cal J}_i$  & $3N_{sp}$ \\

kyspec & Binormal mode potential spectrum &  $\sum_{k_\psi} \sum_{s} |\hat \phi(k_\psi,k_\zeta,s)|^2$    &  $N_{mod}$\\

kxspec & Radial mode potential spectrum &  $\sum_{k_\zeta} \sum_{s} |\hat \phi(k_\psi,k_\zeta,s)|^2$  & $N_x$  \\

{p/e/v}\_flux\_spectra & Binormal spectral flux for each species & $\sum_{k_\psi} \sum_{s} {\cal J}_i(k_\psi,k_\zeta,s) $ & $N_{mod}N_{sp}$ \\

{p/e/v}\_flux\_xspec  & Radial spectral fluxes for each species & $\sum_{k_\zeta} \sum_{s} {\cal J}_i(k_\psi,k_\zeta,s) $  & $N_{mod}N_{sp}$   \\

parallel\_phi.dat & Parallel  potential  & $ \sum_{k_\zeta} \sum_{k_\psi} |\hat \phi(k_\psi,k_\zeta,s)|^2$ & $N_s$        \\

spc\#     & 2D potential spectrum slice \# by timestep & $|\hat \phi(k_\psi,k_\zeta,s=0)|$ & $N_x, (N_{mod})$ \\

phi\#     & 2D potential slice \# by timestep &  $\phi(\psi,\zeta,s=0)$ & $M_x, (M_{mod})$\\

\hline
\end{tabular}
\caption{Time trace diagnostics output by GKW. The coordinate labels for the \texttt{par\#} output are in \texttt{xphi} and \texttt{yphi}.}
\end{table}

\begin{table}[h!]
\begin{tabular}{l|l|l|l}
 Filename & Description & Quantities & Cols,Rows \\
\hline
krho & Binormal wavevector grid & $(q / 2 \pi \psi) k_\zeta \rho_{\rm ref} = k_\perp(k_\psi=0,s=0) \rho_{\rm ref}$ & $N_x, N_{mod}$\\
kxrh & Radial wavevector grid & $k_\psi \rho_{\rm ref}$ & $N_x, N_{mod}$\\
yphi & Real space binormal grid & $ 2 \pi \psi (\zeta-\zeta_{min})/q\rho_{\rm ref}$ & $M_x, M_{mod}$ \\
xphi & Real space radial grid & $(\psi-\psi_{min})/\rho_{\rm ref} $ & $M_x, M_{mod}$ \\
parfun.dat & Perpendicular wavevector & $k_\perp\rho_{\rm ref}, $ & $2,N_s N_{mod} N_x$\\
par.dat & Curvature and Coriolis functions & $k_\perp\rho_{\rm ref}$ , ${\cal D}^\psi k_\psi + {\cal D}^\zeta k_\zeta$ , ${\cal H}^\psi k_\psi + {\cal H}^\zeta k_\zeta$, & $3,N_s N_{mod} N_x$\\
geom.dat & Geometry tensors & Labelled in file & N/A\\
kx\_connect & Parallel boundary connections & Labelled in file & N/A\\
\end{tabular}
\caption{Initial grid diagnostics output by GKW. The grid outputs are scaled so that $\psi$ and $\zeta$ coordinate scales can be compared directly on the input scale of $k_\perp \rho_{\rm ref}$ (see equation \ref{kperp}).} %where $D={\cal D}^\psi k_\psi + {\cal D}^\zeta k_\zeta$, and $H={\cal H}^\psi k_\psi + {\cal H}^\zeta k_\zeta$.
\end{table}

% \begin{table}[h!]
% \begin{tabular}{l|l|l|l}
%  Filename & Description & Quantities & number of columns \\
% \hline
% %distr1.dat &  $v_\parallel$ grid at maximum in potential & $v_\parallel$ & $N_{v\parallel}, &N_{\mu}$\\
% %distr2.dat & $\mu$ grid at maximum in potential & $\mu$ & $N_{v\parallel}, N_{\mu}$ \\
% %distr3.dat & Distribution function for & Im(f(1,1,max,vp,mu)) & $N_{v\parallel}, N_{\mu}$\\
% %distr4.dat & Distribution function for & Re(f(1,1,max,vp,mu)) & $N_{v\parallel}, N_{\mu}$\\
% %parallel.dat & & & \\
% \end{tabular}
% \caption{Final diagnostics ouput by GKW.}
% \end{table}

 \begin{figure}
 \label{directions}
 \begin{center}
 \includegraphics[width=0.5\textwidth]{directions.eps}
 \caption{Coordinate directions on the outboard plane ($s=0$) for $s_B=1$, as output by the potential slices diagnostic \texttt{phi\#} with the coordinate labels for each point as output by \texttt{xphi} and \texttt{yphi} (the yphi coordinate label inverts the plot). For $s_B=1$, the poloidal field is `upwards' ${\rm sign}(B_p \nabla \theta \cdot \nabla \zeta)= s_B$. If in doubt, see Sec.~\ref{signs}.  Note that the `centre' of the flux tube is in the bottom left corner when output from the FFTs.}.

 \end{center}
 \end{figure}
%This means as the matlab script mkmovie plots it.

\section{Output Files}%Added by FC Apr 2008 - Unfinished - Please improve
\index{output files}
GKW will generate a number of output files depending on the input options and `diagnostic' namelist.  If \texttt{mode\_box = false} the output of potentials is suppressed.

% \begin{footnotesize}
%  \begin{verbatim}
%  &DIAGNOSTIC
%   Add details here.
%  /
% \end{verbatim}
% \end{footnotesize}

\vspace{10pt}
\noindent
The following files are added to or created each data dump timestep (every NAVERAGE timesteps):
\begin{itemize}
\item For a linear run, \texttt{time.dat} contains three columns.  The first is the normalized time, the second is the dominant mode growth rate, and the third is the dominant mode frequency.  If \texttt{CONTROL: normalized=.false.} the third column instead gives the total norm of the distribution function.

\item The columns of \texttt{fluxes.dat} contains the particle flux, energy flux, and momentum flux, (in this order), repeated for each kinetic species in the order of the species input in \texttt{input.dat}.  The file therefore has 3 *(number of kinetic species) columns.

\item The columns of \texttt{\{p/e/v\}flux\_spectra.dat} contain the fluxes toroidal mode (ky) spectra for particle flux, energy flux, and momentum flux \{respectively\}, repeated for each kinetic species.  Therefore each file has NMOD*(number of kinetic species) columns: in for each species for each kinetic species in the order of the species input in \texttt{input.dat}.

\item The columns of \texttt{\{p/e/v\}flux\_xspec.dat} contain the fluxes radial mode (kx) spectra for particle flux, energy flux, and momentum flux \{respectively\}, repeated for each kinetic species.  Therefore each file has NX*(number of kinetic species) columns: in for each species for each kinetic species in the order of the species input in \texttt{input.dat}.

\item \texttt{kyspec} contains the peturbed potential `poloidal' mode spectrum against time.  Plot a slice across the columns for the spectrum at a given timestep.

\item \texttt{spcXXXXXX} Contains the 2D potential spectrum at the outboard side for each dump timestep.

\item \texttt{phiXXXXXX} Contains a 2D real cross section of the potential on the outboard side (can be altered) for each dump timestep.
\end {itemize}

\vspace{10pt}
\noindent
The following files are written at the start of a run:
\begin{itemize}
\item \texttt{input.out} echos the input variables the code is using (including defaults not supplied in the input file). By renaming it to \texttt{input.dat} it could be used for starting the code as if it were \texttt{input.dat}.

 \item \texttt{xphi} and \texttt{yphi} contain real space position information for the data points in the phiXXXXX files.

\item \texttt{krho} contains the values of $k_y\rho$ for each mode. %normalized against $k_{\rm th}^N = q  / ( 2 \pi \psi)$, the normfactor for calculating the $k_\zeta$ values in the code.

\item \texttt{kxrh} contains contains the values of $k_x\rho$ for each mode. %normalized against $k_{\rm th}^N = q  / ( 2 \pi \psi)$, the normfactor for calculating the $k_\zeta$ values in the code.
\end{itemize}

\vspace{10pt}
\noindent
The following files are written once at the end of a successful run:

\begin{itemize}

\item \texttt{parallel.dat} contains some diagnostics connected with the parallel mode structure. For every mode the following quantities are (sequentially) written to file, in this column order:
\begin{verbatim}
sgr         the length along the field line 
phi         the potential 
apar        the parallel component of the vector potential 
dens        the perturbed density 
tpar        the perturbed parallel temperature 
tperp       the perturbed perpendicular temperature 
wflow       the perturbed parallel flow velocity
\end{verbatim}
Note only \texttt{sgr} is real. The other quantities are complex and fill 2 columns in the output file. The file will have $2*N_{MOD}*N_X*N_S$ rows.

\item \texttt{parfun.dat} and \texttt{par.dat}? contain the curvature and $k_\perp$ function.

\item \texttt{FDSXXXXXX} Contain the distrbution function binaries for each processor needed for restarting a run. The newer restart version will output only 1 binary, \texttt{FDS}.
\end{itemize}

\chapter{Notes on individual routines}

This chapter discusses the individual routines and gives some details on the physics 
processes implemented. Here, not all the input and output or all the variables are 
discussed. It merely gives some details that should allow the reader to understand
the code better.

\section{\texttt{linear\_terms.F90}}%Added by Francis Jan 2008
In the treatment so far, the terms as implemented in the code were described in \ref{terms} as terms I-IX. Here we link these terms to the routines in the code, as well as to \ref{g-vlasov}, the Vlasov equation for the modified perturbation distribution function $g$.
\be 
\label{1}
{\partial g \over \partial t} + (\underbrace{v_\parallel {\bf b}}_{I} + \underbrace{{\bf v}_D}_{II} + \underbrace{{\bf v}_\chi}_{III}) \cdot \nabla g  
-\underbrace{{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial g \over \partial v_\parallel}}_{IV} = S. 
\ee

Where the source from the Maxwellian background is given by eqn \ref{sources}
\bee 
\label{2}
S &=&  - \underbrace{({\bf v}_\chi}_{V} + \underbrace{{\bf v}_D)}_{VI} \cdot \nabla_p F_M
-  {Ze \over T} \underbrace{[ v_\parallel {\bf b}}_{VII} + \underbrace{{\bf v}_D ]}_{VIII} \cdot \nabla \chi F_M \cr 
\noalign{\vskip 0.2 truecm} 
& & - \underbrace{{Z e \over T} {\mu B \over m} \biggl ( {{\bf B} \cdot \nabla {\bf B} \over B^2} 
\biggr ) \langle A_\parallel \rangle F_M }_{IX}
\eee

We also need the expansion of the drift velocity \ref{drift}
\bee
\label{3}
{\bf v}_D &=& \underbrace{{1\over Ze} \biggl [ \underbrace{{m v_\parallel^2\over B}}_{1} + \underbrace{\mu}_{2} \biggr ] {{\bf B} \times \nabla B
\over B^2}}_{A} + \underbrace{{m v_\parallel^2 \over 2 Z e B} \beta^\prime {\bf b} \times \nabla \psi}_{3} 
 + \underbrace{{ 2 m v_\parallel \over Z e B } {\bf \Omega}_\perp}_{4} \cr 
\noalign{\vskip 0.2 truecm} 
&-& \underbrace{{m \Omega^2 R \over Z e B }  {\bf b} \times {\nabla R}}_{5}, 
\eee

In \texttt {linear\_terms.f90} the various terms from these equations are input into the matrix in individual routines which will be described below.  Most of the computation in the code is done in the wave vector domain. All derivatives in the perpendicular direction are in Fourier space; but derivatives along the field line are done in real space (terms I and VII)

\par
Many of the routines in \texttt {linear\_terms.f90} have two versions (identified by a postfix to the routine name), one for each of the second order and fourth order methods.  We concentrate here on the fourth order method since the other is out of order. The routines do the same thing with regards to which terms are implemented; the implementation is different for each method.  Note that the new distribution $g$ is often called $f$ in the code comments, and its array is called \texttt{fdisi}. Also the normalizations outlined in \ref{normalization} are also used.

\par
\texttt{calc\_linear\_terms} is the master routine that calls the subroutines which put the terms I-IX (excluding term III) into the matrix.  Broadly, the first section puts the terms on the LHS of \ref{1}.  The second section calculates the source terms due to the maxwell background.\ref{2}.  The third section of \texttt{calc\_linear\_terms} calls the routines that input the field equations into the matrix. These are the equations that force the solution the be self consistent with the potentials calculated from the distribution function.  After each section is completed there is a call to \texttt {finish\_matrix\_section} in module \texttt {matdat} which stores the k index labels of the matrix for each section..

\par
Hence \texttt{calc\_linear\_terms} covers all the terms I-IX in \ref{terms} ( also labelled as the underbraced numbers in \ref{1}, \ref{2}, and \ref{3}), with the exception of term III, which is implemented in {\texttt non\_linear\_terms.f90}.

\subsection{\texttt{vpar\_grad\_df}, term I}
\texttt{vpar\_grad\_df} calculates the convection parallel to the field.  This routine also adds parallel velocity dissapation required for a numerically stable solution determined by the dissapation parameter \texttt{disp\_par}.  This term is the free streaming motion along the feild line, with derivative calculated in real space.

\subsection{\texttt{vdgradf}, term II}
\texttt{vdgradf} calculates the drift in the gradient of the eikonal term.

\subsection{\texttt{trapdf}, terms IV and IX}
\texttt{trapdf} calcuates the mirror trapping terms dependent on the switch variable \texttt{trapping}.   This routine also adds perpendicular velocity dissapation required for a a numerically stable solution determined by the dissapation parameter \texttt{disp\_vp}.

\subsection{\texttt{ve\_grad\_fm}, term V}
\texttt{ve\_grad\_fm} calculates the ${\bf E} \times {\bf B}$ drift due to the background distribution, as well as (optionally according to logical \texttt{nlapar}) the electromagnetic correction ${\bf v}_{\delta B_\perp}.$

\subsection{\texttt{neoclassical}, term VI}
\texttt{neoclassical} optionally includes the neoclassical terms according to the switch \texttt{neoclassics}.  Note many gyrokinetics codes disregard this term.

\subsection{\texttt{vpgrphi}, term VII}
\texttt{vpgrphi} calculates the landau damping term, dependent on the switch variable \texttt{landau}.  This is an acceleration along the field line with parallel derivative in real space.

\subsection{\texttt{vd\_grad\_phi\_fm}, term VIII}
\texttt{vd\_grad\_phi\_fm} calculates the  drift in the gradient of phi times the velocity derivative of the Maxwell background.  Note that Equation terms  \ref{3}.1 and \ref{3}.2 are combined as \ref{3}.A in the $E_D$.multiplier as defined in \ref{terms}.

\section{\texttt{non\_linear\_terms.F90}}%Added by Francis Feb 2008

The Fourier amplitudes of the potential (and perturbed distribution function) are defined as 
\be 
\phi_N(x) = \sum_k \phi_{Nk} \exp [ {\rm i} k x ] + c.c.  
\ee 
Note that because the inverse discrete Fourier transform ($H_l$) is defined such that 
the values in real space ($h_n$) are given by  
\be 
h_n = {1\over N} \sum_{l=0}^{N-1} H_l \exp [ {\rm i} 2\pi l n / N ] 
\ee 
where $N$ is the total number of grid points of the discrete transform. The relation between the 
amplitudes and the discrete Fourier transform, therefore, is 
\be 
H_l = N\phi_{Nk}
\ee 
The wave vector $k$ is given by the standard relation  
\be 
k = {2 \pi l \over N \Delta} 
\ee 
where $\Delta$ is the grid distance. This relation is used to determine the size of the box 
in real space. It follows that the lowest non-zero wave vector $k_1$ determines $\Delta$ and
the size of the box $L = N\Delta$ 
\be 
L = {2\pi \over k_1} 
\ee

\subsection{\texttt{add\_non\_linear\_terms}, term III} 
\index{Non linear terms}
This routine implements the full term III with electromagnetic corrections as described by \ref{electromag}.  Hence the routine finds the ${\bf E} \times {\bf B}$ drift of the perturbed distribution function, and adds the term:
\be 
\label{Edrift}
{\bf v}_E \cdot \nabla g = {{\bf b} \times \nabla \langle \phi \rangle \over B} \cdot \nabla g
\ee
Without the elctromagnetic corrections, Term III of \ref{terms} becomes
\be
\label{nl_term}
{\rm III} = - {\bf v}_\phi \cdot \nabla g \rightarrow - \rho_*^2 {\partial {\langle \phi_N \rangle} \over \partial x_\beta}
{\cal E}^{\beta \alpha} {\partial g_N \over \partial x_\alpha} = {\rho_*^2 {\cal E}^{\psi \zeta}}\left({{\partial {\langle \phi_{N} \rangle} \over \partial \zeta} {\partial g_N \over \partial \psi} - {\partial g_N \over \partial \zeta}{\partial {\langle \phi_{N} \rangle} \over \partial \psi}}\right)
\ee
where the expansion of the tensor sum uses the anti-symmetry of ${\cal E}^{\beta \alpha}$.  Note also that the diagonal elements of ${\cal E}^{\beta \alpha}$ are zero, and derivatives along $s$ are neglected.

The gyroaveraged potential $\langle \phi_N \rangle$ is first calculated in the wavevector domain by multiplying the potential by the Bessel function $J_0$ (\ref {Bessel}). In k space the derivative of the potential is obtained by multiplying by $ik_\alpha$.

Hence the implementation in the code is
\be
a = i J_{0}(k_{\perp}\rho) k_{N,\zeta} \phi_{N,k} = ik_{\zeta}\rho_{\rm ref}{\langle \phi_{N,k} \rangle} 
\ee
\be
b = i J_{0}(k_{\perp}\rho) k_{N,\psi} \phi_{N,k} = ik_{\psi}\rho_{\rm ref}{\langle \phi_{N,k} \rangle} 
\ee
Note the $\rho$ in the Bessel function is species dependent.  Also the normalization of the k vector \ref{knorm} $k = {k_N / \rho_{\rm ref}}$ introduces a multiplier of $\rho_{\rm ref}$
Hence the inverse Fourier transformation gives
\be
ar = \mathcal{F}^{-1}(a) =  \rho_{\rm ref} {\partial {\langle \phi_N \rangle} \over \partial \zeta_{N}} = {\rho_{\rm ref} \over R_{\rm ref}} {\partial {\langle \phi_N \rangle} \over \partial \zeta} =  \rho_* {\partial {\langle \phi_N \rangle} \over \partial \zeta}
\ee
\be
br = \mathcal{F}^{-1}(b) =  \rho_* {\partial {\langle \phi_N \rangle} \over \partial \psi}
\ee
%Incorrect, not all corrdinates are normalized the same.
Since the coordinates are normalized by $x_{\alpha,N} = {x_{\alpha} \slash R_{\rm ref}}$.  Here we have explicitly linked the normalizations of the k vector and coordinates to show that under the Fourier transform; $\rho_* {\partial \over \partial x_\alpha} \rightarrow {\rm i} k_\alpha$, as was intended in its definition (see \ref{rho_norm}). Put simply, a factor of $\rho_{*}$ appears after the inverse Fourier transform.

Similarly, the gradient of the distribution function in k space is found $a = i k_\psi g_{N,k}$, $b = i k_\zeta g_{N,k}$
and applying the inverse Fourier transform gives
\be
cr = \mathcal{F}^{-1}(a) = \rho_* {\partial g_N \over \partial \psi}
\ee
\be
dr = \mathcal{F}^{-1}(b) = \rho_* {\partial g_N \over \partial \zeta}
\ee

The tensor ${\cal E}^{\alpha \beta}$ (see \ref{efun}) is defined in {\texttt geom.F90} with name  {\texttt efun}. Note that although the tensor has two indices, each component varies along the field line due to the gradient in $B$. The array therfore has 3 indices with $efun{(i,2,1)}$ being the ${\cal E}(s)^{\zeta \psi}$ component. Calling this element of the array we now have all the components needed to calculate \ref{nl_term}.  The factor of $\rho_*^2$ comes from the 2 inverse Fourier transforms.  Term III then appears in the code as
\be
efun(i,2,1)*({ar*cr-br*dr})
\ee

Finally this is Fourier transformed back to the wavevector domain and $\mathcal{F}({\bf v}_E \cdot \nabla g)$ is added to the RHS of the equations -  which were passed to the subroutine as an argument along with the distribution function when it was called from \texttt{exp\_integration.F90}.

\par
The routine also calculates maximum velocities to be used in the timestep estimator.  The first time this routine is called it does some initialisations which are not repeated in later calls:  The sizes of the boxes in real space are calculated, the max k-vectors are calculated and the location of the $k_x = 0$ mode is determined.  Index arrays are also set up to relate the storage regime on the fast Fourier transform (fft) routine to the GKW storage regime. The remainder of the routine as described above is executed every time.

\section{The index function \texttt{indx()} in \texttt{dist.F90}}%By Francis Dec 07 - Please expand!
\index{index function}
The perturbation distribution function is a 6 dimensional object (three dimensions of space, two of velocity, and one over the species). In the code the perturbation distribution function is called \texttt{fdisi} which corresponds to $g$ in this manual.

The $v_\perp$ velocity dimension is expressed as the magnetic moment $\mu = \frac{mv_\perp}{2B}$), whilst the $v_\parallel$ is retained as a velocity. Two of the spatial dimensions are expressed in the wave vector domain, but the dimension along the field line is kept as a real spatial dimension.

This 6 dimensional distribution function object is stored as a one dimensional array in the code to allow for optimisation in the computationally intensive core solvers.  In the more readable initialisations and output calculations  the distribution function is indexed through the \texttt{indx} function

Hence to find the part of the distribution function you need, the code calls: \texttt{indx(imod,ix,i,j,k,is)} where:

\begin{itemize}
 \item \texttt{imod} is the toroidal mode number, $ imod \in \lbrace\mathbb{Z}\cap[1,nmod]\rbrace$
 \item \texttt{ix} is the radial mode number, $ ix \in \lbrace\mathbb{Z}\cap[1,nx]\rbrace$
 \item \texttt{i} is the point on spatial grid along field line, $ i \in \lbrace\mathbb{Z}\cap[1,ns]\rbrace$
 \item \texttt{j} is the magnetic moment $\mu$ grid number, $ j \in \lbrace\mathbb{Z}\cap[1,nmu]\rbrace$  
 \item \texttt{k} is the $v_\parallel$ grid number, $ k \in \lbrace\mathbb{Z}\cap[1,nvpar]\rbrace$
 \item \texttt{is} is the species number, $ is \in \lbrace\mathbb{Z}\cap[1,nsp]\rbrace$

\end{itemize}

So at many points in the code you will see the entire distribution function looped aver with 6 nested loops over the indices above.

%\vfill \eject 
%\printindex

\section{Advection boundary conditions}%Added by Will 29/2/2008
\index{boundary conditons}
The following section describes the implementation of the open boundary conditions in the s direction. (This will also be extended in the near future to the parallel velocity direction).  Away from the boundaries the scheme used is a simple central differencing, second and fourth order is available.

\begin{eqnarray}
\mbox{Fourth Order} &-& \frac{\partial f}{\partial x} = \frac{-f_{n+2} + 8f_{n+1} - 8f_{n-1} + f_{n-2}}{12dx} \nonumber\\
\mbox{Second Order} &-& \frac{\partial f}{\partial x} = \frac{f_{n+1} - f_{n-1}}{2dx} \nonumber
\end{eqnarray}
\noindent
Where n denotes the grid point of interest.

Close to the boundaries a staggered change in differencing order accuracy is used to allow the removal of ghost cells.  Because of the use of upwinding the boundary condition used is highly dependent on the sign of the advective velocity as if the wrong direction is chosen then the scheme is highly unstable.

The terms where this has been implemented are denoted I and VII in equation \ref{eqn_norm}.  Both these terms are advection equations with a spatially dependent advective velocity.
\begin{equation}
\frac{\partial f}{\partial s} = -v(s)\frac{\partial f}{\partial s}
\end{equation}
When the advective velocity $v(s)$ is positive the distribution function will move toward the left boundary, and the following scheme is used (n denotes the cell of interest, see figure \ref{boundarydiffs} for schematic),

\begin{eqnarray}
\mbox{Boundary cell} &-& \mbox{Second order scheme} - \frac{\partial f}{\partial s} = \frac{-\frac{3}{2}f_{n} + 2f_{n+1} -\frac{1}{2}f_{n+2}}{ds}\nonumber\\
\mbox{Adjacent cell} &-& \mbox{Third order} - \frac{\partial f}{\partial s} = \frac{-\frac{1}{3}f_{n-1} -\frac{1}{2}f_{n} + f_{n+1} - \frac{1}{6}f_{n+2}}{ds}\nonumber
\end{eqnarray}
with the right hand boundary using the centrally differenced schemes.

\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{Boundarydiff.eps}
\caption{Schematic showing the cells of intest in the differencing scheme when considering the boundary cell (Hollow circles) and the adjacent cell (Filled circles).  The figure shows the left hand boundary, for right hand simply reflect in the y direction}
\label{boundarydiffs}
\end{center}
\end{figure}

When the advective velocity is negative the opposite is true and the right hand boundary is differenced as follows,

\begin{eqnarray}
\mbox{Boundary cell} &-& \mbox{Second order scheme} - \frac{\partial f}{\partial s} = \frac{\frac{3}{2}f_{n} - 2f_{n-1} +\frac{1}{2}f_{n-2}}{ds}\nonumber\\
\mbox{Adjacent cell} &-& \mbox{Third order} - \frac{\partial f}{\partial s} = \frac{\frac{1}{3}f_{n+1} +\frac{1}{2}f_{n} - f_{n-1} + \frac{1}{6}f_{n-2}}{ds}\nonumber
\end{eqnarray}

\begin{figure}
\begin{center}
\includegraphics[width=0.95\textwidth]{fluxcomp.eps}
\caption{Figure showing the effect of the new open boundary conditions at the boundary. The new boundary has reduced cell to cell fluctuations giving a smoother solution.}
\label{fluxcomp}
\end{center}
\end{figure}

\noindent
Currently there is no dissipation (Diffusion/super-diffusion) within this scheme.

When using the second order scheme is used only special consideration is needed for the boundary cell.  Depending on the direction of the advective velocity a first order backwinded scheme is used at the boundary while the opposite uses the second order central differenced scheme with a ghost cell whose value is determined by the \textbf{parallel\_boundary\_conditions} parameter.

The effect of these new open boundary conditions is to remove the need for ghost cells in the differencing scheme while maintaining a high order scheme. It has reduced the amplitude of cell to cell fluctuations refelecting off the boundary (See figure \ref{fluxcomp}), which in turn, has reduced the need for adding artificial dissipation to the system to damp away these modes.  The dissipation, however, can not be wholly turned off in non-linear runs.  Note also that the dealiasing in nonlinear runs also has some dissipative effect in the perpendicular spatial directions.

\section{Parallel velocity grid - Trapping condition}%Added by Will 29/6/2008
\index{trapping condition}
By contorting the velocity grid to conform to the electron trapping condition we are able to remove terms that involve derivatives in the parallel velocity direction.  This greatly speeds up linear run computation times but more significantly removes the need for numerical diffusion in the parallel velocity direction and lastly, provides resolution in the required areas.

This option is turned on by setting \textbf{vp\_trap} $= 0$ in the \textbf{CONTROL} section of the input deck.  A further parameter is needed in the same section which is named \textbf{n\_trapped} which is the number of grid points within the trapped region of the velocity grid.

One point of note is in the current implementation of the trapping condition the number of cells in the s direction and the number of points in the parallel velocity direction add certain constrictions to the number of points that can be placed within the trapped region.

Firstly, for ease of differencing the bounce points of a specific $v_{\parallel 0}$ (The velocity on the low-field side) must occur exactly half way between two grid-cells, thus the number of grid cells in the s direction restrict both the number and initial values of $v_{\parallel}$.  The maximum number of bounce points available is given by $ns/2 - 1$.  The value of ntrapped must be less than or equal to this.  If less is chosen then the code will select which points to use determined by the spacing between points (The closer to the boundary, the closer the bounce points become), the algorithm will iterate over the points, sequentially removing points until ntrapped are left.  ntrapped must also be less than nvpar.  Outside the trapping region the points are uniformly placed, the spacing is determined by the number nvpar $- 2\times$n\_trapped.

\begin{figure}[here!]
\begin{center}
\includegraphics[width=0.7\textwidth]{traplowfield.eps}
\caption{Figure of parallel velocity grid points as given by the trapping condition, the line corresponds to the trapped/untrapped boundary $v_{||0} = \sqrt{2\mu(B_{max}-B_{min})}$.  In the above data, N\_vpar\_grid $= 16$ (only one half of the grid is shown as it is symmetric),n\_trapped $= 4$ and $\mu = 6$. It must be noted that the total number of points within the trapped region (full grid) is $2\times$n\_trapped.}
\label{lowfieldsidetrap}
\end{center}
\end{figure}

This algorithm gives the grid at the low field side (For an example see figure \ref{lowfieldsidetrap}), this is then propagated around the torus by the equation,

\begin{equation}
v_{\parallel N} = \sqrt{v_{\parallel 0 N}^{2} + 2\mu_{N}(B - B_{0})}
\end{equation}
\noindent
Where $B_{0}$ is the magnetic field at the low field side.  When a bounce point is reached, the velocity becomes zero, all points after this along s are essentially set to zero and are ignored in the differencing.  The grid is also symmetric around $v_{\parallel}=0$.

Rules that must be followed, otherwise the code will throw you out,
\begin{itemize}
\item The value of ntrapped must be less than N\_vpar\_grid/2
\item ntrapped must also be less than or equal to ns/2$-1$
\item For optimal convergence $2\times$n\_trapped must be roughly half of nvpar for more information on convergence properties see figure \ref{}.
\end{itemize}

%WAH: \begin{figure}
%WAH: \begin{center}
%WAH: \includegraphics[width=0.9\textwidth]{Killer3.eps}
%WAH: \caption{The absolute error in growth rates as a function of the mean parallel velocity element.}
%WAH: \label{converge}
%WAH: \end{center}
%WAH: \end{figure}

\printindex

\end{document}

### 2022-05-03 gkw/doc/etc/gs2gkw.tex
(1) Added gkw Clone gwk  #3

\documentclass[a4paper,10pt]{article}
\usepackage{fullpage}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}
\newcommand{\File}[2][]{\texttt{#2}} %output files
\newcommand{\src}[1]{\href{http://code.google.com./p/gkw/source/browse/trunk/src/#1}{#1}}
\newcommand{\doc}[1]{\href{http://code.google.com./p/gkw/source/browse/trunk/doc/#1}{#1}}
\newcommand{\scripts}[1]{\href{http://code.google.com./p/gkw/source/browse/trunk/scripts/#1}{#1}}
\newcommand{\matlab}[1]{\href{http://code.google.com./p/gkw/source/browse/trunk/matlab/#1}{#1}}
\newcommand{\name}[1]{\textsl{#1}} %find a way to make this link to correct doxygen search?

%opening
\title{Conversions between GKW and GS2 inputs}
\author{F.J.~Casson}

\begin{document}

\maketitle

\begin{abstract}

Converting between GKW and GS2 inputs should be straightforward, but it always seems to consume more time than it should.  Here we try to ease the process by documenting some knowledge that has been accumulated over experience of running benchmarks between the codes.  Checking it requires digging into the definitions and documentation of both codes.  This you should do if you want to check anything here, but you can use this note as a quick reference.  Please update this if you find anything wrong, or anything that is not covered here.  If you've not run GKW before, you should also look at the rest of the fairly comprensive GKW documentation.

\end{abstract}

\section{Radial Coordinate}

GKW always uses the same normalised radial coordinate $\psi$:
\begin{equation}
 \label{eq:psi-def}
 \epsilon = \psi = \frac{R_{\rm max}-R_{\rm min}}{2R_{\rm ref}},
\end{equation}
which is almost the LFS minor radius $r_{\rm minor}$ (except for peculiar geometries) normalised to some length scale.  The length scale can be anything, but will determine the definitions of gradients and collisionality.  It is traditionally thought of as a major radius.  When using Miller geometry it must be the major radius of the flux surface centre, thus
\begin{equation}
 R^{\rm GKW-Miller}_{\rm ref} = R^{\rm GS2}_{\rm maj} \cdot L_{\rm ref}^{\rm GS2}
\end{equation}
When using numerical equilbiria in GKW $R_{\rm ref} =$ \name{R0EXP} as written in the hamada.dat file output by CHEASE.  This can be set in the CHEASE inputs (a trick to overrid the EQDSK is to set \name{R0EXP} in CHEASE input negative), but is usually taken from the EQDSK file (5th entry).  It should be chosen to be near to the magnetic axis, but it appears that often it is related to the centre of the limiter geometry. It is always best to check the value used directly in the \name{hamada.dat} file output by CHEASE and read by GKW.  More information on running CHEASE for GKW is found in the document \name{chease/RunChease.tex}. The field $B_{\rm ref}$ in the Larmor radius and $\beta$ is always defined at $R_{\rm ref}$.

\section{Radial gradients}

The GKW lengthscale inputs are therefore $R_{\rm ref}/ L_n$ etc.  Depending on which radial coordinate you are using in GS2, all radial derivatives may need to be converted.  All the species gradient quantities in GKW are defined on the radial coordinate $\psi$ as follows:
\label{eq:gradients}
\begin{equation}
{\rm rln} = {R_{\rm ref} \over L_n} = - {1\over n_{R_0}} {\partial n_{R_0} \over \partial \psi} ,
\qquad 
{\rm rlt} = {R_{\rm ref} \over L_T} = - {1 \over T} {\partial T \over \partial \psi} , 
\qquad 
{\rm uprim} = - {R_{\rm ref} \over v_{\rm thref}} {\partial \omega_\phi \over \partial \psi}  
\end{equation}
In GKW, \name{uprim} for every species is always normalised by the same thermal velocity.
This is not the same as GS2 \name{uprime} which is normalised by the thermal velocity of each species.  In GKW all species should have the same value for \name{uprim} if they all rotate with the plasma rotation.  The standard ExB shearing in GKW is enabled through the rotation namelist with \name{shear\_profile='wavevector\_remap'} and the shearing rate can be coupled to the \name{uprim} value by using \name{toroidal\_shear='use\_uprim'}.  To use this ExB shearing you must have multiple $k_x$ and therefore \name{mode\_box=.true.}.

\section{Grid sizes}

In GKW, \name{n\_s\_grid} is the \textbf{total} number of points along the field line.  In GS2, \name{ntheta+1} is the number of points \textbf{per turn}.  The number of poloidal turns, \name{2*nperiod-1} has the same meaning in both codes.  Linear GKW is usually run with only one $k_y$ and one $k_x$ and multiple \name{nperiod}, with \name{mode\_box=.false.}.  (No twist and shift boundary condition needed). The velocity grids in the two codes are quite different and not directly comparable.  If resolving the trapped passing boundary is important (e.g. Collisionless TEM), in GKW you need \name{n\_vpar\_grid=64} or more, but this requirement diminishes for ITG or if collisions smooth the velocity space structures.

\section{$\beta$ and $\beta^\prime$}

$\beta_{\rm ref}$ is defined the same in both codes, assuming you choose the same species to have the reference values $n_{\rm ref}$ and $T_{\rm ref}$, and use the same location for $B_{\rm ref}$ in each code.   Using \name{alpha\_input} with \name{bishop = 5} in GS2 corresponds to the \name{gradp\_type=alpha\_mhd} option in GKW (NEEDS CHECKING for non circular surfaces). Using \name{beta\_prime\_input} with \name{bishop = 4} in GS2 corresponds to the \name{gradp\_type=betaprime} option in GKW (only the radial coordinate must be converted in the derivative) and is the safer option (betaprime was incorrect in some versions of GKW before r3012).

\section{Collisions}

The full collision operators are likely not completely identical but very close agreement can be achieved between the pitch angle scattering only versions.  In GKW it is usually easiest to set $n_{\rm ref}$, $T_{\rm ref}$ and $R_{\rm ref}$ and the collisionality for every species will be calculated internally.  However, since the Coulomb logarithm used for this is not a fixed value, if you want an exact benchmark with GS2 one should use the frequency override and input the collision frequency directly into GKW (which is the ion-ion collision frequency normalised by $v_{\rm thref} / R_{\rm ref}$).  The conversion from GS2 \name{newk} therefore contains a mass ratio as well as the length scale and temperature ratio factors.  Thus, assuming  $T_{\rm ref}=T_i$ in both GKW and GS2 and $L^{GS2}_{\rm ref} = R^{GKW}_{\rm ref}$, 
%Assumes
\begin{verbatim}
gkwin.COLLISIONS.coll_freq = gs2in.electrons.vnewk*
                             (gs2in.electrons.temp)^(1.5)*sqrt(gs2in.electrons.mass)
\end{verbatim}

\section{Miller Parameters}

The Miller parameters are not defined the same in GKW and GS2.  GKW follows the definitions of Candy PoP 2009 and is thus almost exactly the same as the GYRO definitions (exceptions are the normalisation of the radial coordinate and $s_\delta$).  A conversion between GS2 and GYRO Miller parameters is given in Bravenec et Al PoP 2011 (http://dx.doi.org/10.1063/1.3671907).  Both GKW and GS2 use $r_{\rm minor}$ as the radial coordinate for Miller, and when $L^{GS2}_{\rm ref} = R^{GKW}_{\rm ref}$ the conversion is:
\begin{verbatim}
 gkwin.GEOM.kappa = gs2mil.akappa;
 gkwin.GEOM.skappa = gs2mil.akappri/gkwin.GEOM.kappa*gkwin.GEOM.eps;
 gkwin.GEOM.delta = sin(gs2mil.tri);
 gkwin.GEOM.sdelta = gs2mil.tripri*cos(gs2mil.tri)*gkwin.GEOM.eps/sqrt(1-gkwin.GEOM.delta^2);
 gkwin.GEOM.square = 0.;
 gkwin.GEOM.ssquare = 0.;
 gkwin.GEOM.dRmil = gs2mil.shift; 
 gkwin.GEOM.Zmil = 0.;
 gkwin.GEOM.dZmil = 0.;
 gkwin.GEOM.geom_type='miller';
 if isfield(gs2mil,'beta_prime_input')
   gkwin.GEOM.gradp_type='beta_prime';
   gkwin.SPCGENERAL.betaprime_type='ref';
   gkwin.SPCGENERAL.betaprime_ref=gs2mil.beta_prime_input;
 end
\end{verbatim}

We note that in GKW the gradp\_type alpha and alpha\_mhd definitions do not agree when kappa/=1,
and it is not yet clear which one (if either) is the equivalent to the gs2 definition, since the benchmarks
scans of alpha were performed with kappa=1.  It is therefore safer to use the gradp\_type=betaprime option in
GKW.

\section{$k_\theta \rho_i$}

The projection of the poloidal wavevector onto the bi-normal is probably the biggest pain.  In any geometry other than $s-\alpha$, the $k_\theta$ wavevectors are not projected the same in GS2 and GKW.   The $\rho_{\rm ref}$ part is the same in both codes, (now that GS2 has lost the $\sqrt{T / m}$  velocity normalisation option, it agrees with GKW with $T_{\rm ref} = {1 \over 2} m_{\rm ref} v_{\rm thref}^2$), but the projection of $k_\zeta$ onto the $\theta$ coordinate is not (the reason why is open for discussion).   
To understand: The GKW definition of the wave vector in normalised form uses the expression 
\begin{equation}
(k_\theta^{\rm GKW} \rho_{\rm ref} )^2 = g^{\zeta \zeta} k_\zeta^2  
\end{equation}
evaluated at the low field side midplane ($s=0$) to determine $k_\zeta$ from the value of $k_\theta \rho_{\rm ref}$ given as an input (here
$k_\theta$ is not a normalised quantity).  In GKW, $\zeta$ is the bi-normal coordinate.   Recall also that $\rho_{\rm ref}$ is defined using $B_{\rm ref}$ at the major radius $R_{\rm ref}$.  GS2 uses 
\begin{equation}
k_\theta^{\rm GS2} \rho_{\rm ref} = 2 \pi \frac{\partial \Psi}{\partial \epsilon} k_\zeta = \frac{k_\zeta}{2{\cal E}^{\epsilon \zeta}}  
\end{equation}
in the projection of $k_y$, where $\Psi$ is the poloidal flux, and $2{\cal E}^{\epsilon \zeta}$ is a geometry quantity in GKW\\
(see http://w3.pppl.gov/$\sim$hammett/work/gs2/docs/geometry/g\_short.pdf and the GKW manual).\\
In $s-\alpha$ geometry with $\alpha = 0$, these are equivalent.

Both quantities \name{kthnorm}=$\sqrt{g^{\zeta \zeta}(LFS)}$ and \name{e\_eps\_zeta}=${\cal E}^{\epsilon \zeta}$ (which is a flux function, but is given for every s point) are output by GKW in the GKW geom.dat file and the $k_\theta$ input conversion can then be written
\begin{verbatim}
gs2in.aky = gkwin.MODE.kthrho * gkwgeom.e_eps_zeta *2 / gkwgeom.kthnorm
\end{verbatim}

\section{Some other differences}

The explicit solver makes GKW less efficient than GS2 for linear runs.  It also means that the timestep must be appropriate in order to get a numerically stable solution.  One advantage of this is that once the timestep is small enough to give a physical result, no convergence testing in timestep is required (the implicit solver of GS2, always gives a result, but one should check it's convergence in the timestep).  The GKW linear timestep estimator will try to choose a sensible value based on the Courant condition for the Vlasov equation, but it does not presently include the fields, so it is not foolproof for some high frequency waves (e.g long wavelength electrostatic - it is better to run with $\beta =$ 3e-4 for electrostatic runs), or fast growing modes (e.g. ETG).  If GKW gives unphysically large growth rates, the first thing one should try is to reduce the timestep (no more than a factor of ten at a time).  If your problem requires a very small timestep (less than 1e-6), something else is likely to be wrong; if a small timestep is really required it may be best to run with \name{normalized=.false}..

\end{document}

### 2022-05-03 gkw/doc/manual/Appendixgeometry.tex
(1) Added gkw Clone gwk  #3

\chapter{Geometry - generalised}

\section{Field aligned Hamada coordinates} 
\index{Hamada coordinates}
\label{sec:hamada}
In this Appendix the straight field line coordinates as well as the shifted metric are 
described in detail. 
The development of the material in this section has benefited from, and closely follows the 
publications of B.D. Scott \cite{SCO98,SCO01}, and the references cited therein. 
This material is in addition to Section \ref{geometry} of this document (which describes the specific geometry models implemented); here we add
a generalised formulation for deriving the geometry tensors, and describe the shifted metric procedure for the nonspectral version of the code.
At some point in the future the two sections should be consolidated.

We start with an arbitrarily shaped, but toroidally symmetric geometry, for which the magnetic 
field can be written in the form \index{Magnetic field representation}
\begin{equation}
{\bf B} = s_B F \nabla \varphi + s_j \nabla \varphi \times \nabla \Psi,
\end{equation}
where $F = RB_t$ with $B_t > 0$ being the toroidal magnetic field strength. The quantities 
$s_B$ and $s_j$ represent the sign of the magnetic field and plasma current (both positive 
when in the direction of $\nabla \varphi$. Finally, $\Psi$ is the poloidal flux, not to 
be confused with the radial coordinate ($\psi$). Of course, $\psi = \psi(\Psi)$, and the 
choice $\psi = \Psi$ will sometimes be made below.

An orthogonal coordinate system $(\psi,\theta,\varphi)$ is assumed, where $\psi$ is the 'radial'
coordinate (i.e. ${\bf B}\cdot \nabla \psi = 0$), $\theta$ is the poloidal angle (upward 
on the outboard midplane), and $\varphi$ is the toroidal angle (clockwise when viewed from 
above). 
Furthermore, it is assumed that the magnetic field and quantities like the major radius are 
known in this coordinate system. 
Below two coordinate transformations are introduced. 
The first will make the field lines straight in the new coordinates, and the second will 
align one of the coordinates with the magnetic field.

In the first step we transform the poloidal and toroidal angle 
\begin{equation} 
s = s(\psi,\theta), \qquad \gamma = \gamma(\psi,\theta,\varphi)
\end{equation} 
in such a way that the contra-variant components of the magnetic field are flux functions 
in the new coordinate system ($B^s = B^s(\psi)$, $B^\gamma = B^\gamma(\psi)$). These 
coordinates are known as Hamada coordinates \cite{HAM62}.
Note that the new coordinate $\gamma$ is still an ignorable coordinate, since any function that is independent of $\varphi$
will be independent of $\gamma$ ($f(\theta,\psi) \rightarrow f(s,\psi)$) after the coordinate 
transformation. The transformation of the poloidal angle is chosen such that $B^s$ becomes 
a flux function 
\begin{equation} 
B^s = {\bf B} \cdot \nabla s = {\bf B}\cdot \nabla \theta {\partial s \over \partial \theta} 
+ {\bf B} \cdot \nabla \psi {\partial s \over \partial \psi} = {\bf B} \cdot \nabla \theta 
{\partial s \over \partial \theta}
\end{equation} 
From which the relation between $s$ and $\theta$ can be derived 
\begin{equation}
{\partial s \over \partial \theta}={B^{s} \over {\bf B \cdot \nabla \theta}}\Leftrightarrow
\int_{0}^{\theta}{\partial s \over \partial \theta}d\theta =s=B^{s}\int_{0}^{\theta}{d\theta \over 
{\bf B}\cdot \nabla \theta}
\end{equation}
since $B^{s}$ does not depend on $\theta$ (flux function) and
\begin{equation}
\oint {\partial s \over \partial \theta}d\theta =1=B^{s}\oint {d\theta \over {\bf B}\cdot \nabla 
\theta}\Leftrightarrow B^{s}={1 \over \oint {d\theta \over {\bf B}\cdot \nabla \theta} }
\end{equation}
which gives the following equation for s:
\begin{equation} 
s(\theta,\psi) = {\int_0^\theta {{\rm d} \theta^\prime \over {\bf B} \cdot \nabla \theta^\prime} 
\biggl /  \oint {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime}} \qquad 
{\rm with} \qquad 
B^s =  {1
\biggl /  \oint {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime}}
\end{equation}  
where the normalizing constant has been chosen such that the domain [-1/2,1/2] in $s$ corresponds 
to one poloidal turn.

The new coordinate is closely related to the flux surface average. This average is defined 
as the volume average between two flux surfaces in the limiting case in which the distance between 
the flux functions goes to zero. One can derive the following relation  
\begin{equation} 
\{ g \} = \oint {{\rm d} l_p \over B_p } g \biggl / \oint {{\rm d} l_p \over B_p }
\end{equation}
where $l_p$ is the length in the poloidal direction (${\rm d}l_p = \vert {\bf e}_\theta \vert 
{\rm d} \theta$), $B_p$ is the poloidal component of the magnetic field strength ($B_p = 
{\bf B}\cdot \nabla \theta / \vert \nabla \theta \vert$), and the integral is to be taken at 
constant $\psi$. Inserting the expressions for the 
poloidal length and the poloidal magnetic field strength and using $\vert {\bf e}_\theta\vert 
\vert \nabla \theta \vert = 1$ one obtains 
\begin{equation} 
\{ g \} = \oint {\rm d}s g 
\end{equation}

For the transformation of the toroidal angle we choose 
\begin{equation} 
\gamma = {\varphi \over 2 \pi} + g(\theta,\psi) 
\end{equation}
The $2\pi$ is added here such the domain [0,1] in $\gamma$ will correspond to one toroidal 
turn around the flux surface. The contra-variant component of the magnetic field is 
\begin{equation} 
B^\gamma = s_B {F \over 2\pi} {1\over R^2} + {\bf B} \cdot \nabla \theta {\partial g \over \partial 
\theta} 
\end{equation} 
where $F = R B_t = F(\psi)$ is a flux function. (Here $B_t$ is the toroidal magnetic field 
strength $B_t = {\bf B} \cdot \nabla \varphi / \vert \nabla \varphi \vert$ where $\vert \nabla \varphi 
\vert = 1/R$ with $R$ being the local major radius)
From the expression above one can derive 
\begin{equation} 
g(\theta,\psi) = \int_0^\theta {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime } 
\biggl [ B^\gamma - s_B{F \over 2\pi} {1 \over R^2} \biggr ] 
\end{equation} 
Demanding that $g$ is periodic
\begin{equation}
\oint {{\rm d}\theta^\prime \over {\bf B}\cdot 
\nabla \theta^\prime } 
\biggl [ B^\gamma - s_B{F \over 2\pi} {1 \over R^2} \biggr ] =0
\end{equation}
yields 
\begin{equation} 
B^\gamma = s_B {F\over 2 \pi }  \left \{ {1\over R^2} \right \} 
\end{equation} 
The coordinate $\gamma$ can then be expressed as 
\begin{equation} 
\gamma = {\varphi \over 2\pi} + s_B {F \over 2\pi} \int_0^\theta {{\rm d} \theta^\prime \over {\bf B}
\cdot \nabla \theta^\prime} \biggl [ \left \{ {1\over R^2} \right \} - {1\over R^2} 
\biggr ] 
\end{equation} 
The coordinates $(\psi,s,\gamma)$ are the Hamada coordinates. The expression above should 
allow these coordinates to be calculated for an arbitrary toroidally symmetric geometry.

\index{Field aligned coordinates}
We now go through the second coordinate transformation where one of the coordinates is 
aligned with the magnetic field, i.e. we demand 
\begin{equation} 
{\bf B} \cdot \nabla = B^s {\partial \over \partial s}
\end{equation}
Note that this does not mean that $\nabla s$ is in the direction of the magnetic field. 
In general it is not. The transformation is 
\begin{equation} 
\zeta = \zeta(s,\gamma,\psi) 
\end{equation}
In the new coordinates 
\begin{equation}
{\bf B}\cdot \nabla =B^{\psi}{\partial \over \partial \psi}+B^{s}{\partial \over \partial s}+
B^{\zeta}{\partial \over \partial \zeta}
\end{equation}
where $B^\psi$ is zero. The transformation to field aligned coordinates is the transformation 
for which $B^\zeta = 0$. The latter condition can be formulated as 
\begin{equation} 
B^\gamma {\partial \zeta \over \partial \gamma} + B^s {\partial \zeta \over \partial s} = 0
\end{equation} 
which is satisfied for the simple linear transformation
\begin{equation} 
\zeta =  q s - \gamma \qquad {\rm where} \qquad q = {B^\gamma \over B^s} 
\end{equation}
From the field line equation it follows that 
\begin{equation} 
{{\rm d} \gamma \over {\rm d} s} = {B^\gamma \over B^s} \qquad \rightarrow \qquad 
\oint {{\rm d} \gamma \over {\rm d} s} {\rm d} s = \oint {B^\gamma \over B^s} {\rm d} s = q 
\end{equation} 
i.e. $q = q(\psi)$ is indeed the safety factor.

Note that the coordinate transformation above flips the sign of the toroidal angle. 
The right handed coordinate system can therefore be defined as $(\psi,\zeta,s)$. 
The Jacobian of the new coordinate system can be expressed in terms of the original 
Jacobian through 
\begin{equation} 
(\nabla \psi \times \nabla \zeta) \cdot \nabla s = {1 \over J_{\psi  \zeta s} } = 
{1\over 2 \pi}{\partial s \over \partial \theta} {1 \over J_{\psi \theta \varphi}} 
\end{equation} 
From which it follows 
\begin{equation} 
J_{\psi  \zeta s} = 2\pi {\bf B} \cdot \nabla \theta \oint {{\rm d} \theta^\prime \over 
{{\bf B} \cdot \nabla \theta^\prime}} J_{\psi \theta \varphi}  = 
2 \pi {{\bf B} \cdot \nabla \theta \over B^s } J_{\psi \theta \varphi}
\end{equation} 
We note here that the coordinates $s$ and $\zeta$ are dimensionless, but that the 
dimension of $\psi$ is not yet defined. We will always use a normalized $\psi$ that 
is dimensionless.

\section{Caculating the Geometry tensors from the metric tensor and derivatives of coordinates
and magnetic field}

The various tensors that are used for the implementation of the geometry are not independent. 
${\cal E}$, ${\cal D}$, ${\cal F}$, ${\cal G}$, ${\cal H}$, ${\cal I}$, ${\cal J}$, ${\cal K}$ 
can be expressed in the metric tensor and derivatives of the 
coordinates as well as the magnetic field. We will assume here that the metric tensor $g^{\alpha 
\beta}$, the magnetic field strength ($B$) and its derivatives towards the coordinates (${\partial 
B / \partial s}, \partial B / \partial \psi$), the major radius $R$ and its derivatives 
($\partial R / \partial s, \partial R / \partial \psi$), the z-coordinate $Z$ and its 
derivatives ($\partial Z / \partial s, \partial Z / \partial \psi$) and finally the 
derivative of the poloidal flux towards the radial coordinate $\partial \Psi / \partial \psi$ 
are given. Of course, even these quatities are not independent, since the derivatives of $R,Z$ towards 
the coordinates can be used to calculate the metric tensor. 
The tensors can then be calculated as discussed below

Since we can write any vector ${\bf A}$ as 
\begin{equation} 
{\bf A} = \frac{1}{2}\epsilon_{ijk} A^i J_{x^i x^j x^k} \nabla x^j \times \nabla x^k, 
\end{equation}
one finds for the magnetic field, using $\psi = \Psi$, 
\begin{equation} 
\mathbf{B} = B^s J_{\Psi \zeta s} \nabla \Psi \times \nabla \zeta.
\end{equation}
Using 
\begin{equation}
B^s = \mathbf{B}\cdot\nabla s = s_b F \nabla \varphi \cdot \nabla s + s_j \nabla \varphi \times \nabla \Psi \cdot \nabla s
\end{equation}
and remembering that by definition of $\zeta$
\begin{equation}
\nabla \varphi \times \nabla \Psi \cdot \nabla s = -2\pi \nabla \zeta \times \nabla \Psi \cdot \nabla s,
\end{equation}
one gets
\begin{equation} 
 B^s = 2\pi s_j \frac{1}{J_{\Psi \zeta s}} 
\end{equation}
% YC 05.03.2012: I prefer the few lines above, but feel free to revert to the lines below if you find it clearer
%\begin{equation} 
%\mathbf{B} = B^s J_{\Psi \zeta s} \nabla \Psi \times \nabla \zeta = 2 \pi ({\bf B} \cdot \nabla \theta) J_{\Psi \theta \varphi} 
%\end{equation}
%Since $\nabla \Psi$, $\nabla \theta$ and $\nabla \varphi$ are orthogonal 
%\begin{equation} 
%J_{\Psi \theta \varphi} = { 1\over \vert \nabla \Psi \vert \vert \nabla \theta \vert \vert \nabla \varphi \vert} 
%\end{equation}
%And 
%\begin{equation} 
%{\bf B} = 2 \pi {R {\bf B} \cdot \nabla \theta \over \vert \nabla \Psi \nabla \vert \vert \nabla \theta \vert } 
%\nabla \Psi \times \nabla \zeta 
%\end{equation} 
%Finally using  
%\begin{equation} 
%\nabla \Psi = R B_p  \qquad {\bf B} \cdot \nabla \theta = s_j B_p \vert \nabla \theta \vert 
%\end{equation}
%
Finally, one obtains \index{Magnetic field representation} 
\begin{equation} 
{\bf B} = 2 \pi s_j \nabla \Psi \times \nabla \zeta 
\label{clebsch}
\end{equation}

The Clebsch representation of the magnetic field can directly be used to calculate the tensor ${\cal E}$ 
\begin{equation} 
{\cal E}^{\alpha \beta} = {1 \over 2 B^2 }{\bf B} \cdot ( \nabla x^\alpha \times \nabla x^\beta )  =
{\pi s_j \over B^2} {\partial \Psi \over \partial \psi} [ g^{\psi \alpha} g^{\zeta \beta} - g^{\zeta \alpha} g^{\psi \beta} ] 
\end{equation} 
With the ${\cal E}$ tensor and the derivatives of the magnetic field strength given, the ${\cal D}$ vector 
directly follows from 
\begin{equation} 
{\cal D}^\alpha = - 2 {\cal E}^{\alpha \beta} {1 \over B} {\partial B \over \partial x_\beta} 
\end{equation}
The scalars that have been given earlier in this document are 
\begin{equation} 
{\cal F} = {B^s \over B} \qquad {\cal G} = {{\cal F} \over B} {\partial B \over \partial s} 
\end{equation}
The angular rotation vector points in the $\nabla Z$ direction, and 
\begin{equation} 
{\bf \Omega}_\perp = \Omega {\bf b} \times (\nabla z \times {\bf b} ) 
\end{equation} 
This allows the ${\cal H}$ vector to be written in the form 
\begin{equation} 
{\cal H^\alpha} = - \frac{s_B}{B_N} \biggl [ {\partial Z \over \partial x_\beta } g^{\alpha \beta} - 
\biggl ( {B^s \over B} \biggr )^2 {\partial Z \over \partial s} \delta_{\alpha s} \biggr ] 
\end{equation}
The ${\cal I}$ tensor is straightforwardly 
\begin{equation} 
{\cal I}^\alpha = {1 \over 2 B} (\nabla x_\alpha \times \nabla R^2 ) \cdot {\bf b} = 2 R {\cal E}^{\alpha \beta} 
{\partial R \over \partial x_\beta} 
\end{equation}
The ${\cal J}$ and ${\cal K}$ tensor can be directly calculated from the known quantities.

\section{Periodicity}

\index{Periodicity on the torus}
The torus is periodic in both the toroidal as well as the poloidal angle. In the new coordinates
this means that for any function $f(\psi,\zeta,s)$ 
\begin{equation} 
f(\psi,\zeta+1,s) = f(\psi,\zeta,s)
\label{toroidalperiodic}
\end{equation}
\begin{equation}
f(\psi,\zeta+q,s+1) = f(\psi,\zeta,s) 
\label{poloidalperiodic}
\end{equation}
In a flux tube one, furthermore, demands that the solution is periodic in the radial direction
\begin{equation} 
f(\psi + L_\psi,\zeta,s) = f(\psi,\zeta,s) 
\label{radialperiodic}
\end{equation}

None of the equilibrium quantities is a function of $\zeta$. For these quantities Eq.~(\ref{toroidalperiodic})
is trivially satisfied, and the Eq. (\ref{poloidalperiodic}) reduces to 
\begin{equation} 
B(\psi,s+1) = B(\psi,s)
\end{equation}
where we have use the background magnetic field strength as example. The perturbed quantities (distribution
function as well as fields) for the $\zeta-$ direction are always represented by Fourier modes 
\begin{equation} 
f(\psi,\zeta,s) = \sum\nolimits_{k_\zeta} \hat f (\psi,k_\zeta,s) \exp [ {\rm i} k_\zeta \zeta / \rho_* ] 
\end{equation} 
The periodicity constraint of Eq. (\ref{poloidalperiodic}) then reduces to 
\begin{equation} 
\hat f(\psi, k_\zeta,s+1) = \hat f (\psi, k_\zeta, s) \exp [ - {\rm i} k_\zeta q / \rho_* ] 
\end{equation}

The flux tube treats the limit of $\rho_* \rightarrow 0$ to the lowest relevant order only. Even though the 
phase factor can in principle take any value, we can assume that a very small change in the safety factor can always be 
choosen such that 
\begin{equation} 
\exp [ - {\rm i} k_\zeta q / \rho_* ]  = 1 
\end{equation}
without changing the results. This choice can, however, only be made for one location since the safety factor is a 
function of the radial coordinate $\psi$. We choose the phase factor above to be zero for the centre of the box (presented
by $\psi = 0$ to obtain a periodicity constraint 
\begin{equation} 
\hat f(\psi, k_\zeta,s+1) = \hat f (\psi, k_\zeta, s) \exp \left [ - {\rm i} k_\zeta {1 \over  \rho_*}{{\rm d} q \over {\rm d} \psi }\psi \right ] 
\end{equation}

\section{Shifted metric}

In general the metric tensor has no zero elements. Especially the magnetic shear leads to a finite $g^{\zeta \psi}$ element that is 
a function of the coordinate along the field line. This means that an originally rectangular grid cell, deforms when moving along the 
magnetic field. This can potentially lead to problems \cite{SCO01}. A remedy, that does not cost anything computationally when using 
Fourier modes in $\zeta$ is the shifted metric procedure. In this procedure we change the coordinate $\zeta$ at every location in 
$s$. 
\begin{equation} 
\zeta_k = q s - \gamma - \alpha_k(\psi)
\end{equation}
It is important that $\alpha$ is not a function of $\zeta$. With this transformation one has a different set of coordinates at 
every grid point in $s$, with the different grid points denoted by the index k. 
The metric tensor in the new coordinates can be expressed as 
\begin{equation} 
g_k^{\zeta \zeta} = g^{\zeta \zeta} - 2 \alpha_k^\prime g^{\zeta \psi} + (\alpha_k^\prime)^2 g^{\psi \psi}
\end{equation}
\begin{equation} 
g_k^{\zeta \psi} = g^{\zeta \psi} - \alpha_k^\prime g^{\psi \psi}
\end{equation}
\begin{equation}
g_k^{\zeta s} = g^{\zeta s} - \alpha_k^\prime g^{\psi s} 
\end{equation}
where $\alpha_k^\prime = {\partial \alpha_k / \partial \psi}$. With this transformation one can 
set the $g^{\zeta \psi}$ element at the grid point $s = s_k$ locally to zero 
\begin{equation} 
\alpha_k^\prime = {g^{\zeta \psi}(s_k) \over g^{\psi \psi}(s_k)}
\end{equation}
In the flux tube the metric elements are evaluated at one surface only, and are not a function of 
the radial coordinate. The equation above can then be trivially integrated to obtain 
\begin{equation} 
\alpha_k = {g^{\zeta \psi}(s_k) \over g^{\psi \psi}(s_k)} \psi 
\end{equation}
In the new coordinates 
\begin{equation}  
g_k^{\zeta \zeta} = g^{\zeta \zeta} -  (g^{\zeta \psi})^2 / g^{\psi \psi}
\end{equation}
\begin{equation} 
g_k^{\zeta \psi} = 0
\end{equation}
\begin{equation} 
g_k^{\zeta s} = g^{\zeta s} - g^{\zeta \psi} g^{\psi s} / g^{\psi \psi} 
\end{equation}

The perturbed solution in the new coordinates has the form 
\begin{equation} 
\hat f_k (\psi, k_\zeta, s) \exp [ {\rm i} k_\zeta \zeta_k / \rho_* ] 
\end{equation}
We must demand that the solution is single valued, and therefore 
\begin{equation} 
\hat f_k (\psi, k_\zeta, s) \exp [ {\rm i} k_\zeta \zeta_k / \rho_* ] =
\hat f_{p} (\psi, k_\zeta, s) \exp [ {\rm i} k_\zeta \zeta_{p} / \rho_* ]
\end{equation}
Using 
\begin{equation} 
\zeta_{p} = \zeta_k + \alpha_k - \alpha_{p} 
\end{equation} 
We obtain 
\begin{equation} 
\hat f_k (\psi, k_\zeta, s)  =
\hat f_{p} (\psi, k_\zeta, s) \exp [ {\rm i} k_\zeta (\alpha_k - \alpha_{p}) / \rho_* ]
\end{equation}
The perturbed distribution function as well as the fields are calculated at the discrete points 
$s_k$ as $f_k$. When taking a parallel derivative one must use the transformation given above 
to transform $\hat f_p$ to $\hat f_k$.

Finally we must consider the possible radial periodicity of the domain. In the new coordinates 
Eq.~(\ref{radialperiodic}) becomes 
\begin{equation}  
f_k(\psi + L_\psi,\zeta_k + \alpha_k^\prime \psi + \alpha_k^\prime L_\psi,s) = 
f_k(\psi, \zeta_k + \alpha_k^\prime \psi, s) 
\end{equation}
Or 
\begin{equation} 
\hat f_k(\psi + L_\psi,k_\zeta,s) = \hat f_k (\psi, k_\zeta,s) \exp \left [ -
{\rm i} k_\zeta \alpha_k^\prime L_x \right ] 
\end{equation}

\section{Summary of the sign conventions in GKW} 
\label{signsappendix}
In GKW, the cylindrical coordinate system $(R,Z,\varphi)$ is right handed, which means that $\varphi$ is increasing clockwise when the torus is viewed from above.\\
The toroidal rotation is defined positive for a plasma flowing in the direction of ${\bf B}$.\\
The mode frequency is defined positive for a perturbation evolving in the direction opposite to $\nabla \zeta$. This corresponds to the ion $\nabla B$ drift direction if $s_{\rm j}=1$ and to the
electron $\nabla B$ drift direction if $s_{\rm j}=-1$.  See also Fig.~\ref{directions}\\
Coordinate system:
\begin{itemize}
 \item $\psi=\varepsilon=(R_{\rm max}-R_{\rm min})/(2R_{\rm ref})$ is always increasing from the plasma center to the plasma edge.
 \item $s$ is always increasing upwards from the low field side midplane. It is zero at the height of the magnetic axis, on the low field side midplane.
 \item $\zeta$ is always increasing in the direction opposite to $\varphi$ (i.e. anticlockwise when viewed from above) at constant $\psi$ and $s$. The direction of $\nabla \zeta$ in the poloidal plane is
given by $\textrm{sign}(\nabla s \cdot \nabla \zeta)=\textrm{sign}(\nabla 
\zeta \cdot \nabla \theta)=s_{\rm B}s_{\rm j}$
\end{itemize}

\begin{align}
\textrm{sign}({\bf B} \cdot \nabla \varphi)&\equiv& s_B \\
\textrm{sign}({\bf j} \cdot \nabla \varphi)&\equiv&s_j \\
\textrm{sign}({\bf B} \cdot \nabla \theta)&=& s_j \\
\textrm{sign}({\bf B} \cdot \nabla s)&=& s_j \\
\textrm{sign}(\nabla \varphi \cdot \nabla \zeta)&=& -1 \\
\textrm{sign}(\nabla s \cdot \nabla \theta)&=& 1\\
\textrm{sign}(\nabla \theta \cdot \nabla \zeta)&=& s_B s_j \\
\textrm{sign}(B_\theta \nabla \theta \cdot \nabla \zeta)&=& s_B \\
\textrm{sign}(\nabla s \cdot \nabla \zeta)&=& s_B s_j \\
{\bf B} \cdot \nabla \zeta&=& 0 \\
{\bf \Omega} &=& -s_B \Omega \nabla z \\
u &=& R \Omega 
\end{align}

\section{Flux surface average \label{sec.fsa}}
\subsection{General}
The general definition of the flux surface average is
\begin{equation}
 <A> = \frac{\int A(\mathbf{x}) \delta(r-r_0) d^3{x}}{\int\delta(r-r_0) d^3{x}}
\end{equation}
where $A$ is the quantity being averaged, $r_0$ is the flux surface label of flux surface on which the average is being 
performed, $r$ is the value of the flux surface label at position $\mathbf{x}$, $\delta$ is the Dirac function and the integral is being performed over the whole plasma (or entire world, it does not matter) volume.

\subsection{GKW coordinates}
In GKW coordinates $(r,\zeta,s)$, the elementary plasma volume is 
\begin{equation}
 d^3 x = dr d\zeta ds \mathcal{J}
\end{equation}
with $\mathcal{J}^{-1} = \nabla r \cdot \nabla \zeta \times \nabla s$. Actually the expression above is true for any coordinates, but
one of the specificity of GKW coordinate system is that $\mathcal{J}$ is a flux surface label:
\begin{equation}
 \mathcal{J}= \frac{\partial\Psi}{\partial r} \frac{s_j}{B^s}
\end{equation}
and $B^s$ is a flux surface label by construction. This implies that the flux surface average can be written as
\begin{equation}
 <A> = \frac{\oint_{r=r_0} A(r,\zeta,s) d\zeta ds}{\oint_{r=r_0} d\zeta ds}
\end{equation}
If $A$ is independent of the toroidal angle, one arrives at
\begin{equation}
 <A> = \oint A(r_0,s) ds
\end{equation}

\subsection{Alternative expression}
Noting that the surface element $\textrm{d} S$ on a flux surface is related to the volume element 
$\textrm{d}^3x$ by
\begin{equation}
 \textrm{d}^3x = \textrm{d} S \frac{\textrm{d}r}{|\nabla r|}
\end{equation}
with $r$ an arbitrary flux surface label, an alternative expression for the flux surface average is obtained:
\begin{equation} \label{eq:FS3}
 <A> = \frac{1}{V'} \oint A \frac{\textrm{d}S}{|\nabla r|}
\end{equation}
with 
\begin{equation}
 V'= \frac{\partial V}{\partial r} = \oint \mathcal{J} \textrm{d}\zeta \textrm{d}s
\end{equation}
and $V$ being the volume enclosed by the flux surface labelled by $r$.

\section{Use of fluxes in transport equations \label{sec.transport}}
In the following the density conservation equation is taken as an example, but the same can be applied to heat, momentum or any
conserved quantity.
The local density conservation equation is 
\begin{equation}
 \frac{\partial n}{\partial t} + \nabla \cdot \mathbf{\Gamma} = S_n
\end{equation}
with $n$ the density in $\textrm{m}^{-3}$, $\mathbf{\Gamma}$ the particle flux in $\textrm{m}^{-2}\textrm{s}^{-1}$ and $S_n$ the particle source density
in $\textrm{m}^{-3}\textrm{s}^{-1}$

Integrating this equation over the volume enclosed by a flux surface labelled by $r_0$ leads to:
\begin{equation}
 \frac{\partial }{\partial t} \int n \textrm{d}V + \oint \mathbf{\Gamma}\cdot\frac{\nabla r}{|\nabla r|} \textrm{d}S = 
\int S_n \textrm{d} V
\end{equation}
where the flux surface shape and position has been assumed to be constant in time. 
Using Eq.~(\ref{eq:FS3}), the continuity equation can then be written as
\begin{equation}
 \frac{\partial }{\partial t} \int n \,\textrm{d}V + V'\left<\mathbf{\Gamma\cdot\nabla r}\right> = 
\int S_n \textrm{d}V
\end{equation}
In steady-state we have:
\begin{equation}
 <\Gamma^r>=<\mathbf{\Gamma\cdot\nabla r}> = \frac{1}{V'}\int S_n \textrm{d}V = \frac{1}{V'}S^\textrm{int}_n
\end{equation}
This means that $\Gamma^r$ which is the quantity we calculate in GKW needs to be compared to the particle source inside the flux 
surface $S^\textrm{int}_n$ divided by $V'$. This is the obvious choice for comparisons with the experiments or transport codes.

Concerning the flux decomposition, the best choice is a direct decomposition of $<\Gamma^r>$:
\begin{equation}
 <\Gamma^r> = - D \frac{\partial n}{\partial r} + V n
\end{equation}
with possibly a normalising length if $r$ is a dimensionless flux label (e.g. GKW $\psi$).
%But it is also possible to rather decompose $\mathbf{\Gamma}$ as above and then a factor $<|\nabla r|^2>$ 
%arises in the definition of $D$ and $V$. 
For a benchmark or a comparison with experiment one needs to make sure that everybody uses the same definition.

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/Memo-LFS-FSA-density.tex
(1) Added gkw Clone gwk  #3

\chapter{Transformation of transport coefficients in presence of poloidal asymmetries}
% The ABC memo

%\documentclass[12pt]{article}
%\include{amsmath}
%\textwidth=167mm
%\textheight=245mm
%\oddsidemargin=-3mm
%\topmargin=-18mm
%\newcommand{\Gfs}{\langle \Gamma \rangle}
\newcommand{\Gfs}{\Gamma}
\newcommand{\exparg}{\exp\left\{-\frac{Ze\Phi(r,\theta)}{T(r)} +\frac{m \Omega^2(r)}{2T(r)} \left(R(r,\theta)^2 - R_0(r)^2\right)\right\}}
\newcommand{\nparder}{\frac{\partial n(r,\theta)} {\partial r}}
\newcommand{\Pparder}{\frac{\partial \Phi(r,\theta)} {\partial r}}
\newcommand{\noder}{\frac{d n_0(r)} {d r}}
\newcommand{\nder}{\frac{d n(r)} {d r}}
\newcommand{\Tlnder}{\frac{1}{T(r)} \frac{d T(r)} {d r}}
\newcommand{\Rsqdiff}{\left(R(r,\theta)^2 - R_0(r)^2\right)}
\newcommand{\dnoovdnoder}{n_0(r) \frac{dr}{d n_0(r)} }
\newcommand{\Ecf} E
\newcommand{\expargs}{\exp\left\{-{\Ecf}(r,\theta)\right\}}
\newcommand{\calGr}{{\cal G}_r}
%\begin{document}

\begin{center} 
{\bf Memo on the transformation of the transport
coefficients\\ 
from LFS density to flux-surface averaged density}\\[0.3cm]
C. Angioni, E. Belli and F.J. Casson\\[0.2cm]
Please report comments and corrections to C. Angioni
\end{center}

Note: The notation used in this Appendix does not attempt to be consistent with the rest of the GKW manual.

In the presence of a poloidally asymmetric density $n(r,\theta)$
of a particle species like W, 
the flux surface averaged particle flux $\Gfs$
can be expressed as a function of the flux surface averaged density
$n(r) = \langle n(r,\theta) \rangle$
or as a function of densities evaluated at specific locations
or poloidal angles. A natural choice, for instance,
is $n_0(r) = n(r,\theta = 0)$, evaluated at the LFS $\theta = 0$ location,
which is adopted by codes like GKW and NEO.

In the present memo we provide the equations to transform 
the corresponding transport coefficients from one description to the other.
This is practical for many applications, in particular when the output
of codes which can include 2D poloidal asymmetries has to be
used inside usual 1D transport codes, which consider only 1D
(or in any case flux surface averaged) quantities.

The flux surface averaged particle flux is expressed in the two forms
\[ \Gfs = - D \frac {d n}{d r} + V n, \]
and
\[ \Gfs = -D_0 \frac {d n_0}{d r} + V_0 n_0. \]
and we state that these two forms are equivalent.
In the present memo, we shall made this equivalence explicit.

We note that the first expression for $\Gfs$
involves the flux surface averaged density 
$n(r) = \langle n(r,\theta) \rangle$,
and the corresponding transport coefficients $D$ and $V$.
In contrast, the second expression involves the LFS density $n_0 = n(r,\theta = 0)$
and the corresponding transport coefficients $D_0$ and $V_0$.

We remind that in general
\[ n(r,\theta) = n_0(r) \, \exparg \]
where $\Phi$ is the background electrostatic potential,
defined in such a way that $\Phi(r,\theta = 0) = 0$,
and $R_0(r)$ is the major radius of the location where $n_0$ is
evaluated (usually LFS $\theta = 0$).

We introduce also the auxiliary (species dependent) normalized energy
\[
\Ecf(r,\theta) = \frac{Ze\Phi(r,\theta)}{T(r)} - \frac{m \Omega^2(r)}{2T(r)} \left(R(r,\theta)^2 - R_0(r)^2\right)
\]
For clarity, with circular concentric flux surfaces,
$R(r,\theta) = R_{\rm geo}+r \cos \theta$ and $R_0(r) = R_{\rm geo}+r$.

In general flux tube geometry, $\theta$ is a generalized poloidal angle,
which describes the distance along the field line.
Considering the right handed system of coordinates $(r,\theta, \phi)$,
the flux surface average of the quantity $n(r,\theta)$ is given by
\[
\langle f \rangle = \frac{1}{V^\prime} \oint d\theta d\phi \; \sqrt{g} \; n(r, \theta),
\]
where $g = (\nabla r \times \nabla \theta \cdot \nabla \phi)^{-2}$ is
the determinant of the metric tensor, and ${V^\prime} = d V/ dr$, where
$V(r)$ is the plasma volume up to the flux surface $r$.

Then, the following relationship holds,
\[
\frac {d \langle n(r, \theta) \rangle}{d r} = 
\left \langle{ \frac {\partial n(r, \theta)}{ \partial r} } \right \rangle + n_0(r) \, \calGr,
\]
where we have defined 
\[
\calGr =  \left [ - \frac{1}{V^\prime} \frac{d^2 V}{d r ^2} \langle{n(r, \theta)}\rangle 
+ \frac{1}{V^\prime} \oint { d\theta d\phi \;  n(r, \theta) \, \frac{\partial{\sqrt{g}}}{\partial r}}
\right ] \, n_0(r)^{-1}
\]

that is
\[
\calGr =  - \frac{1}{V^\prime} \frac{d^2 V}{d r ^2} \, \left \langle \expargs \right \rangle 
+ \left \langle \expargs \frac{\partial \log(\sqrt{g})}{\partial r} \right \rangle
\]
Partial derivatives versus the minor radius $r$ are intended to 
be performed at constant $\theta$.
We note that $\calGr = 0$ with a Hamada coordinate system 
(as used in GKW).

In order to derive the relationship between the two pairs
$(D,V)$ and $(D_0,V_0)$, we proceed by expressing the
flux surface averaged density $n(r)$ and its radial
derivative $d n(r)/d(r)$ in terms of $n_0(r)$ and 
$d n_0(r)/d(r)$.
The relationship between $n(r)$ and $n_0(r)$ is straightforward
\[ n(r) = n_0(r) \, \left \langle \exparg \right \rangle \]
In order to express $d n(r)/d(r)$ in terms of $n_0(r)$ and 
$d n_0(r)/d(r)$, we proceed with the computation of 
$ \partial n(r,\theta) / \partial r$.
\begin{eqnarray}
\nparder &=& \exparg \; \left\{ \noder \; + {n_0(r)} \, \cdot \right. \nonumber \\
& \cdot & \left[-\frac{Ze}{T(r)} \Pparder + \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r} \Rsqdiff + \frac{m\Omega^2}{2T(r)}\frac{\partial}{\partial r} \Rsqdiff \right]
\nonumber \\
& + & \left. n_0(r) \Tlnder \left[\frac{Ze}{T(r)} \Phi(r,\theta) - \frac{m\Omega^2}{2T(r)} \Rsqdiff \right] \right\}
\label{eq.nderiv}
\end{eqnarray}

We take now the flux surface average, and recalling
that
\[
\frac{d n(r)}{d r} = \left \langle { \frac{ \partial n(r,\theta)}{\partial r}} \right \rangle + n_0(r)\,\calGr,
\]
we find, 
\begin{eqnarray*}
\nder &=& \left \langle \expargs \right \rangle \noder \,+\\
& + & n_0(r) \left\{ - \left \langle \expargs \, \frac{Ze}{T(r)} \Pparder \right \rangle \right. \\
& + & \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r}\\
& + & \left \langle \expargs \, \frac{\partial}{\partial r} \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)}\\
& + & \left \langle \expargs \, \frac{Ze}{T(r)} \Phi(r,\theta) \right \rangle
\Tlnder \\
& - & \left. \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)} \Tlnder
\right\}
+ n_0(r)\,\calGr.
\end{eqnarray*}

At this point, we replace the expressions of $d n(r) /d r$ and $n(r)$ in the expression for $\Gfs$ and
we obtain
\begin{eqnarray*}
\Gfs &=& -D \left \langle \expargs \right \rangle \noder  + \\
& & - D\, n_0(r)\,\left\{ - \left \langle \expargs \, \frac{Ze}{T(r)} \Pparder \right \rangle \right. \\
& & + \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r}\\
& & + \left \langle \expargs \, \frac{\partial}{\partial r} \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)}\\
& &
+ \left \langle \expargs \, \frac{Ze}{T(r)} \Phi(r,\theta) \right \rangle
\Tlnder \\
& & \left. - \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)} \Tlnder
+\calGr \right\}\\
& & + V \,n_0(r) \, \left \langle \expargs \right \rangle
\end{eqnarray*}

Finally, we compare with equation 
\[ \Gfs = -D_0 \frac {d n_0}{d r} + V_0 n_0, \]
and we identify the following relationships:

\[ D_0 = D \left \langle \expargs \right \rangle, \]
and
\begin{eqnarray*}
V_0 &=& - D \left\{ - \left \langle \expargs \, \frac{Ze}{T(r)} \Pparder \right \rangle \right. \\
& + &  \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r}\\
& + &  \left \langle \expargs \, \frac{\partial}{\partial r} \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)}\\
& + & \left \langle \expargs \, \frac{Ze}{T(r)} \Phi(r,\theta) \right \rangle
\Tlnder \\
& - & \left. \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)} \Tlnder
\,+\, \calGr
\right\}\\
& + &  V \left \langle \expargs \right \rangle.
\end{eqnarray*}

The final relationships follow directly
\[ D = D_0 \left \langle \expargs \right \rangle^{-1} = D_0 \frac{n_0}{n}, \]

\begin{eqnarray*}
V &=& V_0 \left \langle \expargs \right \rangle^{-1} +
D_0 \left\{ - \left \langle \expargs \, \frac{Ze}{T(r)} \Pparder \right \rangle \right. \\
& + &  \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r}\\
& + & \left \langle \expargs \, \frac{\partial}{\partial r} \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)}\\
& + & \left \langle \expargs \, \frac{Ze}{T(r)} \Phi(r,\theta) \right \rangle
\Tlnder \\
& - & \left. \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)} \Tlnder 
\,+\, \calGr \right\} \, \left \langle \expargs \right \rangle^{-2}
\end{eqnarray*}

We define the following conversion factor for the convection,
\begin{eqnarray*}
{\cal V} &=& a \left\{ - \left \langle \expargs \, \frac{Ze}{T(r)} \Pparder \right \rangle \right. \\
& + &  \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r}\\
& + & \left \langle \expargs \, \frac{\partial}{\partial r} \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)}\\
& + & \left \langle \expargs \, \frac{Ze}{T(r)} \Phi(r,\theta) \right \rangle
\Tlnder \\
& - & \left. \left \langle \expargs \, \Rsqdiff \right \rangle \frac{m\Omega^2}{2T(r)} \Tlnder 
\,+\, \calGr
\right\} \, \left \langle \expargs \right \rangle^{-1},
\end{eqnarray*}

by which the transformation from the pair $(D_0, V_0)$ into the pair $(D, V)$
simply reads
\begin{eqnarray} 
D &=& \left \langle \expargs \right \rangle^{-1} D_0 = \frac{n_0}{n} D_0, \\
a V &=&  \left \langle \expargs \right \rangle^{-1} \left( a V_0 \,+\,
D_0 \cal{V} \right).
\label{eq.DVtransform}
\end{eqnarray}

Here we have introduced an arbitrary normalizing length $a$
(usually the geometrical major radius $R_{geo}$ in GKW and 
the minor radius $a$ in NEO).

The quantity $\cal{V}$ can be directly computed inside codes
which treat poloidal aysmmetries (like GKW and NEO).
It is suggested that the quantity $\cal{V}$ is computed for
each species independently.

Thereby, for each kinetic species, codes like GKW and NEO 
can provide directly in output the two flux surface averaged 
quantities $n_0 / n$ (already present in output actually)
and the convection conversion factor $\cal{V}$.
In this way the equation to be applied in 1D 
transport codes to pass from $(D_0, V_0)$ to $(D, V)$ is
very simple.

Optionally, codes like GKW and NEO can also provide in output 
the entire set of flux surface averaged quantities which are required to
perform the transformation from the pair $(D_0, V_0)$ into the pair $(D, V)$,
\[ e_0 = \left \langle \expargs \right \rangle ,\]
\[ e_1 = \left \langle \expargs \, \frac{Ze}{T(r)} \Phi(r,\theta) \right \rangle, \]
\[ e_2 = a \left \langle \expargs \, \frac{Ze}{T(r)} \Pparder \right \rangle, \]
\[ e_3 = a^{-2} \left \langle \expargs \, \Rsqdiff \right \rangle, \]
\[ e_4 = a^{-1} \left \langle \expargs \, \frac{\partial}{\partial r} \Rsqdiff \right \rangle, \]
\[ e_5 = a \, \calGr = a \left [
- \frac{1}{V^\prime} \frac{d^2 V}{d r ^2} \, e_0
\,+\, \left \langle \expargs \frac{\partial \log(\sqrt{g})}{\partial r} \right \rangle
\right] 
\]
where we included normalizations to an arbitrary length $a$, which can be chosen
in order to be consistent with the other normalizations used in the codes
(that is, e.g., major radius in GKW, and minor radius in NEO). We remind
that $e_5 = 0$ with Hamada coordinates, as used in GKW.
Then, the convection conversion factor reads
\[
{\cal V} = \frac{1}{e_0} \left\{ - e_2 
+ e_3 a^3 \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r}
+ e_4 a^2 \frac{m\Omega^2}{2T(r)} 
+ e_1 a \Tlnder 
- e_3 a^3 \frac{m\Omega^2}{2T(r)} \Tlnder + e_5
\right\}.
\]
and the transformation equations read
\[ D =  \frac{D_0}{e_0} = \frac{n_0}{n} D_0,\]
\[ a V =  {\left( a V_0 \,+\,D_0 \cal{V} \right)}\,{e_0}^{-1}.\]

We notice that the convection conversion factor $\cal{V}$ has
been defined in such a way that
\[ \frac{a V}{D} = \frac{a V_0}{D_0}+{\cal V}, \]
which can also be regarded as a direct consequence of the following relationship
\[ \frac{a}{n} \frac{d n}{d r} = \frac{a}{n_0} \frac{d n_0}{d r} + {\cal V}. \]

\subsection{Anisotropic minorities (Reinke model) \label{sec.memo-icrh}}

In the Reinke model (see Sec. \ref{sec.icrh}), the poloidal density variation for the anisotropic minority is
\begin{equation}
 n(r,\theta) = n_0 \left( \frac{B}{B_0} \right)^{-\eta} \exp(-{\cal E}_{\Omega}/T_\parallel) = n_0 \exp(-{\cal E}_{\Omega,\eta}/T_\parallel)
\end{equation}
The radial deriviative then has additional terms 
\begin{equation}
\frac{\partial n(r,\theta)}{\partial r} = n_0
%\left(\frac{B_0}{B}\right)^\eta e_\eta 
\left[ -\ln(B/B_0)\frac{\partial \eta}{\partial r} - \frac{\eta}{B}\frac{\partial B}{\partial r} + \frac{\eta}{B_0}\frac{\partial B_0}{\partial r} \right]
\exp(-{\cal E}_{\Omega,\eta}/T_\parallel) + \cdots
\end{equation}
where $\cdots$ represents all the terms on the RHS of equation \ref{eq.nderiv} with the generalised normalised energy $E = {\cal E}_{\Omega,\eta}/T_\parallel $
in place of the former $E = {\cal E}_{\Omega}/T $ (as defined in Eq. \ref{eq.poloidal-energy}).  The coefficients $e_0,e_1,e_2,e_3,e_4,e_5$ are also generalised 
using this generalised energy in their definition.
Then, the transformation identities above all still hold if $\cal{V}$ is extended to
\begin{equation}
{\cal V} = \frac{1}{e_0} \left\{ - e_2 
+ e_3 a^3 \frac{m\Omega}{T(r)} \frac{d \Omega(r)}{d r}
+ e_4 a^2 \frac{m\Omega^2}{2T(r)} 
+ e_1 a \Tlnder 
- e_3 a^3 \frac{m\Omega^2}{2T(r)} \Tlnder + e_5 + e_\eta
\right\}.
\label{eq.vcalex}
\end{equation}
where the additional coefficient $e_\eta$ is defined as
\[
e_\eta = a \left \langle \exp\left\{-{\cal E}_{\Omega,\eta}(r,\theta)/T_\parallel\right\} \left[ -\ln(B/B_0)\frac{\partial \eta}{\partial r} - \frac{\eta}{B}\frac{\partial B}{\partial r}
+ \frac{\eta}{B_0}\frac{\partial B_0}{\partial r} \right] \right \rangle.
\].

\subsection{Anisotropic minorities (Bilato model)}
%spellcheck me
In the Bilato model (see Sec. \ref{sec.icrh2}), the poloidal density variation for the anisotropic minority is
\begin{equation}
 n_m = n_0 \frac{T_{\perp}(\theta)}{T_{\perp 0}} \exp(-{\cal E}_{\Omega}/T_\parallel) = n_0 \exp(-{\cal E}_{\Omega,\eta}/T_\parallel)
\end{equation}
where
\begin{equation}
 \frac{T_{\perp}(\theta)}{T_{\perp 0}} = 
\left [ \frac{T_{\perp 0}}{T_{\parallel}} + \left(1 - \frac{T_{\perp 0}}{T_{\parallel}} \right )\frac{B_{0}}{B} \right ]^{-1}.
\end{equation}
The radial deriviative then has additional terms 
\begin{equation}
\frac{\partial n(r,\theta)}{\partial r} = -n_0
%\left(\frac{B_0}{B}\right)^\eta e_\eta 
 \left[ \frac{\partial (T_{\perp 0}/T_\parallel)}{\partial r} \left(1- \frac{B_{0}}{B} \right )
+ \left( -\frac{B_{0}}{B^2}\frac{\partial B}{\partial r} 
+ \frac{1}{B}\frac{\partial B_{0}}{\partial r} \right) 
\left(1- \frac{T_{\perp 0}}{T_\parallel}\right)
 \right] \frac{T_\perp(\theta)}{T_{\perp 0}} 
\exp(-{\cal E}_{\Omega,\eta}/T_\parallel) + \cdots
\end{equation}
where $\cdots$ represents all the terms on the RHS of equation \ref{eq.nderiv} with the generalised normalised energy $E = {\cal E}_{\Omega,\eta}/T_\parallel $
in place of the former $E = {\cal E}_{\Omega}/T $ (as defined in Eq. \ref{eq.poloidal-energy2}).  The coefficients $e_0,e_1,e_2,e_3,e_4,e_5$ are also generalised 
using this generalised energy in their definition.
Then $\cal{V}$ is extended as in Eq. \ref{eq.vcalex} with the additional coefficient $e_\eta$ is defined as
\[
e_\eta = -a \left \langle \exp\left\{-{\cal E}_{\Omega,\eta}(r,\theta)/T_\parallel\right\} 
 \left[ \frac{\partial (T_{\perp 0}/T_\parallel)}{\partial r} \left(1- \frac{B_{0}}{B} \right )
+ \left( -\frac{B_{0}}{B^2}\frac{\partial B}{\partial r} 
+ \frac{1}{B}\frac{\partial B_{0}}{\partial r} \right) 
\left(1- \frac{T_{\perp 0}}{T_\parallel}\right)
 \right] \frac{T_\perp(\theta)}{T_{\perp 0}} 
 \right \rangle.
\]

\pagebreak

%\end{document}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/appendix-vel-nonlin.tex
(1) Added gkw Clone gwk  #3

\chapter{Velocity Nonlinearity and Conservation Properties}
\section{The velocity nonlinearity} 
\label{sec:velocity-nonlinearity} 
\index{Velocity nonlinearity}

There is a nonlinearity called the velocity nonlinearity, which is often neglected in the $\delta f$ formalism since it is one order smaller 
in the normalized Larmor radius $\rho_*$ compared with the leading order terms. The
expression for the velocity nonlinearity can be obtained by considering the evolution equation for 
of the parallel velocity (Here without rotation and only electro-static) 
\begin{equation}
mv_\parallel \d{v_\parallel}{t} = - \d{\bf X}{t}\cdot \left [
Ze \nabla \langle \phi \rangle + \mu \nabla B \right ] 
\end{equation}
In the $\delta f$ formalism the terms that are proportional to both the field $\phi$ as well as the perturbed 
distribution $f$ are neglected. 
Note that these terms do include the term due to ${\bf v}_E \cdot \mu \nabla B$, which cancels 
${\bf v}_{\nabla B} \cdot Z e \nabla \langle \phi \rangle$. The remaining terms are 
\begin{equation}
 \d{v_\parallel}{t} = - \frac{Ze }{ m v_\parallel} [ v_\parallel {\bf b} + {\bf v}_{D-} ] \cdot 
\nabla \langle \phi \rangle  
\end{equation}
where ${\bf v}_{D-}$ is the drift velocity minus the grad-B drift.

The parallel velocity nonlinearity therefore adds an additional term to the gyro-kinetic equation in the 
form 
\begin{equation}
\pd{f}{ t} \overset{+}{=} \frac{Ze }{ m v_\parallel} [ v_\parallel {\bf b} + {\bf v}_{D-} ] \cdot 
\nabla \langle \phi \rangle  \pd{ f }{v_\parallel} 
\end{equation}
Normalizing this contribution gives for the parallel motion and the drift, respectively 
\begin{equation}
\pd{f_N}{t_N} \overset{+}{=} \frac{Z \rho_*}{2 \sqrt{T_{GN} m_N}}
 {\cal F} \pd{ \langle \phi_N \rangle }{s} 
\pd{ f_N }{v_{\parallel N}} 
\end{equation}
\begin{equation}
\pd{ f_N}{t_N} \overset{+}{=}  
\frac{\rho_*^2 }{ 2} \left [ v_{\parallel N} {\cal D}^\alpha + v_{\parallel N} \beta^\prime {\cal E}^{1 \alpha} + 
2 H^\alpha \sqrt{\frac{m_N }{ T_{GN}}} \Omega_N \right ] 
\pd{ \langle \phi_N \rangle}{ x^\alpha} \pd{ f_N }{v_{\parallel N}} 
\label{eq:parallel-velocity-nonlinearity-drift}
\end{equation}
(here the Coriolis drift is kept which is somewhat inconsistent with the fact that it has been neglected in the 
equation above, but it is work in progress). 
The drift terms involve derivatives in the perpendicular plane that scale as $1/\rho_*$. Therefore, both 
terms are of the order $\rho_*$ and are usually neglected.

As discussed in section \ref{energy-conservation-local-limit}, in the case of the local limit the velocity nonlinearity 
must be implemented as 
\begin{equation}
\pd{ f}{ t} \overset{+}{=} \pd{}{ v_\parallel} \left [ \frac{Ze }{ m v_\parallel} 
[ v_\parallel {\bf b} + {\bf v}_{D-} ] \cdot \nabla \langle \phi \rangle  f \right ]  
\end{equation}
in order to assure phase space convervation. In the latter case an additional term appears in Eq.~(\ref{eq:parallel-velocity-nonlinearity-drift})
which must be implemented as 
\begin{align}
\pd{ f_N}{ t_N} &\overset{+}{=}
\frac{\rho_*^2 }{ 2} \left [ v_{\parallel N} {\cal D}^\alpha + v_{\parallel N} \beta^\prime {\cal E}^{1 \alpha} + 
2 H^\alpha \sqrt{\frac{m_N}{T_{GN}}} \Omega_N \right ] 
\pd{ \langle \phi_N \rangle}{ x^\alpha} \pd{ f_N }{v_{\parallel N}} \cr 
\noalign{\vskip 0.2 truecm} \\
& + \frac{\rho_*^2 }{ 2} \left [ {\cal D}^\alpha + \beta^\prime {\cal E}^{1 \alpha} \right ] \pd{ \langle \phi \rangle }{x^\alpha} f 
\end{align}

\section{Basic equation}

{\bf Warning: not all is correct in this section}

The equations in GKW are related to, but do not directly follow from a Lagrangian description. 
Further approximations to the Lagrange equations of motion are made in order to obtain a system that 
is easier to solve numerically. The approximations all involve neglecting higher order $\rho_*$ terms,
and since the Lagrangian from which the equations are derived itself is accurate up to terms linear in $\rho_*$ 
the neglection of higher order terms is consistent with accuracy of the Lagrange description. 
Care has to be taken not to lose the properties that come with the Lagrange description, like conservation 
of energy and momentum. Below it will be shown that these properties still hold for the model equations 
used by GKW.

Starting point of the derivation is the system Lagrangian \index{Lagrangian} 
\begin{equation}
L = \sum_{sp} \int {\rm d}^3{\bf X} {\rm d}^3 {\bf v} \, f L_p + \sum_{sp} \int {\rm d}^3 {\bf X} 
{\rm d}^3 {\bf v} \, \frac{Z_{sp}^2 e^2}{2 T_{sp} } [ \phi^2 - \langle \phi \rangle^2 ] F_M + \sigma_a \int {\rm d}^3 {\bf X} \,  L_{\rm adiabatic} 
+ \int {\rm d}^3 {\bf X} \, \frac{1 }{ 2 \mu_0} \vert \nabla A_\parallel \vert^2 
  \label{eq:lagrangian}
\end{equation}
with the particle-Lagrange density 
\begin{equation}
L_p = \Gamma_a \dot z^a -H  = \left ( Z e {\bf A} + m {\bf u}_0 +  m v_\parallel {\bf b} + Z e \langle A_\parallel \rangle  {\bf b} \right ) \cdot \d{\bf X}{t} 
- \left ( \frac{1 }{ 2} m v_\parallel^2 + \mu B + Z e \langle \phi \rangle + \frac{1 }{ 2} m u_0^2 \right ) ,
\label{eq:plagrangian}
\end{equation}
where $L_p$ is written without the species index. The quantity $\Gamma$ refers to the discussion in section~\ref{sec:theoryLagrangian}, eq.~\eqref{eq:gyrocenterlagrangian}.
The expressions above describe a gyrocenter lagrangian in the sense that the rapid gyro-motion has been removed from the Lagrangian. The indices in the Einstein 
summation convention are $a = 1,2,3,4$ with $a = 4$ refering to the parallel velocity (Note that $\gamma_4$ is zero though). 
The brackets $\langle \rangle$ correspond to the gyro-average operator, and the sum with index sp is over all 
kinetic species. 
The vector potential ${\bf A}$ describes the (constant) background magnetic field, while the perturbed magnetic field is 
described by $A_\parallel$, the perturbed parallel component of the vector potential.

In the case of adiabatic electrons, the electron contribution is not included in the species-sum $\sum_{sp}$ but is treated by $\sigma_a = 1$ in a special term in eq. \eqref{eq:lagrangian}.
(If electrons are meant to be kinetic, $\sigma_a = 0$ takes that term out.)
The Lagrangian density associated with adiabatic electrons in that term is 
\begin{equation}
L_{\rm adiabatic} = - \frac{e^2 n_e }{ 2 T_e} [ \phi - \{ \phi \} ]^2 ,  
\end{equation}
where $\{ \}$ indicates the flux surface average.

\nomenclature[1L]{$L$}{The Lagrangian}
\nomenclature[1Lp]{$L_p$}{The particle Lagrangian}
\nomenclature[1Z]{$Z$}{The charge number}
\nomenclature[1X]{${\bf X}$}{The gyro-centre spatial coordinate}
\nomenclature[1v]{${\bf v}$}{The velocity} 
\nomenclature[1f]{$f$}{The perturbed distribution} 
\nomenclature[1e]{$e$}{The unit charge}
\nomenclature[1T]{$T$}{The temperature} 
\nomenclature[2f]{$\phi$}{The perturbed potential} 
\nomenclature[3b]{$\langle \rangle $}{The gyro-average operation} 
\nomenclature[3c]{$\langle \rangle^\dagger$}{The conjugate transpose of the gyro-average operator} 
\nomenclature[1fm]{$F_M$}{The Maxwell distribution} 
\nomenclature[1H]{$H$}{The Hamiltonian}

In the following, We occasionally split the Hamiltonian $H$ from \eqref{eq:plagrangian} into a field part $H_1 = Z e \langle \phi \rangle$ and the rest $H_0$ (which is time independent),
\begin{align}
  H &= \underbrace{\frac{1 }{ 2} m v_\parallel^2 + \mu B +  \frac{1 }{ 2} m u_0^2}_{H_0} +\underbrace{Z e \langle \phi \rangle}_{H_1} = H_0(\mathbf{X}, v_\parallel, \mu) + H_1(\mathbf{X}, t, \mu) 
\label{eq:Hsplit}
\end{align}

Variation of $L$, which is already integrated over phase space, towards $\phi$
\begin{align}
  \frac{\delta L}{\delta \phi} &\overset{!}{=} 0  \\
   &= - \sum_{sp}\int\ud^3v 
\frac{\delta}{\delta\phi}
\left(\int\ud^3X f_{s,tot}
    \left[\underbrace{Ze\ga{\phi}}_{H_1} + \frac{Z_s^2e^2}{2T_s}\left(\ga{\phi}^2 - \phi^2\right)\right]
 \right)
 - \sigma_a\frac{\delta}{\delta\phi} \int \ud\mathbf{X} \frac{e^2 n_e }{ 2 T_e} [ \phi - \{ \phi \} ]^2
\label{eq:Lfuncderivphi}
\end{align}
yields the Poisson equation \index{Poisson equation}.
\ifmoredetails
To see how, consider the derivative of the first term in \eqref{eq:Lfuncderivphi}, which becomes
\begin{align}
  &\d{}{\epsilon}\left[\int\ud^3X^\prime f_{tot}(\mathbf{X}^\prime) Ze\ga{\phi(\mathbf{X}^\prime) + \epsilon\delta(\mathbf{X}^\prime - \mathbf{X})}\right]_{\epsilon=0}  \nonumber\\
{}={}&\d{}{\epsilon}\left[\int\ud^3X^\prime \iga{ f_{tot}}(\mathbf{X}^\prime,\mu) Ze(\phi(\mathbf{X}^\prime) + \epsilon\delta(\mathbf{X}^\prime - \mathbf{X}))\right]_{\epsilon=0}  \nonumber\\
{}={}& \int\ud^3X^\prime \iga{f_{tot}}(\mathbf{X}^\prime,\mu) Ze \delta(\mathbf{X}^\prime - \mathbf{X}) \nonumber\\
{}={}& Ze \iga{ f_{tot}}(\mathbf{X},\mu) 
\label{eq:Lfuncderivphi1stterm}
\end{align}
Here, the gyroaverage operator was written as $\ga{}$ and must be distinguished from its adjoint $\iga{}$.
The second term of \eqref{eq:Lfuncderivphi} becomes
  \begin{align}
   &\d{}{\epsilon}\Big[\int\ud^3X^\prime F_M\ga{\phi + \epsilon\delta(\mathbf{X}^\prime - \mathbf{X})}^2 &&- (\phi + \epsilon\delta(\mathbf{X}' - \mathbf{X}))^2\Big]_{\epsilon=0} \nonumber\\
   {}={}& \int\ud^3X^\prime F_M 2\ga{\phi + \epsilon\delta(\mathbf{X}^\prime - \mathbf{X})}\ga{\delta(\mathbf{X}^\prime - \mathbf{X})}\Bigg|_{\epsilon=0} &&+ 2F_M\phi(\mathbf{X})\nonumber\\
   {}={}& \int\ud^3X^\prime F_M 2\ga{\phi}\ga{\delta(\mathbf{X}^\prime - \mathbf{X})}&&+ 2F_M\phi(\mathbf{X})\nonumber\\
   {}={}& \int\ud^3X^\prime \iga{F_M 2\ga{\phi}}\delta(\mathbf{X}^\prime - \mathbf{X})&&+ 2F_M\phi(\mathbf{X})\nonumber\\
   {}={}&  F_M 2\iga{\ga{\phi}}(\mathbf{X},\mu)+ 2F_M\phi(\mathbf{X}) \label{eq:Lfuncderivphi2ndterm}
   \end{align}
and the third one works analogously to the simpler part in the right column of \eqref{eq:Lfuncderivphi2ndterm}.
Derived this way, the Poission equation is then
\fi
\begin{equation}
\sum_{sp} Z_{sp} e \int {\rm d}^3 {\bf v} \, \langle f \rangle^\dagger -\sigma_a \frac{n_e e^2 }{ T_e} ( \phi - \{ \phi \} ) 
= \sum_{sp} \frac{Z_{sp}^2 e^2}{T_{sp}} \int {\rm d}^3 {\bf v} \, [ \phi - \langle \langle \phi \rangle \rangle^\dagger ] F_{Msp} 
\label{eq:poissonequation}
\end{equation}
Furthermore, variation towards $A_\parallel$ yields Ampere's law. Again, we work each term separately:
\begin{align}
 \frac{\delta L}{\delta A_\parallel} &\overset{!}{=} 0  \\  
&= \frac{\delta}{\delta A_\parallel}
 \sum_{sp} \int {\rm d}^3{\bf X} {\rm d}^3 {\bf v} \, f Ze\ga{A_\parallel}\mathbf{b}
+ \frac{\delta}{\delta A_\parallel}\int {\rm d}^3 {\bf X} \, \frac{1 }{ 2 \mu_0} \vert \nabla A_\parallel \vert^2 
\end{align}
The first term is evaluated just like \eqref{eq:Lfuncderivphi1stterm}.
\begin{align}
   \frac{\delta}{\delta A_\parallel}
 \sum_{sp} \int {\rm d}^3{\bf X} {\rm d}^3 {\bf v} \, f Ze\ga{A_\parallel}\mathbf{b}\cdot\d{\mathbf{X}}{t}
&= 
\sum_{sp}\int\ud^3\mathbf{v}
\d{}{\epsilon}\left[\int\ud^3X^\prime \f_{tot}(\mathbf{X}^\prime) Ze\ga{A_\parallel(\mathbf{X}^\prime) + \epsilon\delta(\mathbf{X}^\prime - \mathbf{X})}\mathbf{b}\cdot\d{\mathbf{X}}{t}\right]_{\epsilon=0}  \nonumber\\
&= 
\sum_{sp}\int\ud^3\mathbf{v}
Ze\iga{f_{tot}\mathbf{b}\cdot\d{\mathbf{X}}{t}}(\mathbf{X},\mu)\\
&= 
\sum_{sp}\int\ud^3\mathbf{v}
Ze\iga{f_{tot}}(\mathbf{X},\mu)\mathbf{b}\cdot\d{\mathbf{X}}{t}
\end{align}
The second term is already more interesting due to the gradient.
\begin{align}
  \frac{\delta}{\delta A_\parallel}\int {\rm d}^3 {\bf X} \, \frac{1 }{ 2 \mu_0} \vert \nabla A_\parallel \vert^2 
&=\frac{1}{2\mu_0}\d{}{\epsilon}\left[\int\ud^3X^\prime |\nabla \left(A_\parallel(\mathbf{X}^\prime)+ \epsilon\delta(\mathbf{X}^\prime - \mathbf{X})\right))|^2\right]_{\epsilon=0}  \nonumber\\
&=\frac{1}{2\mu_0}\d{}{\epsilon}\left[\int\ud^3X^\prime (\nabla \left(A_\parallel(\mathbf{X}^\prime)+ \epsilon\delta(\mathbf{X}^\prime - \mathbf{X})\right))^2\right]_{\epsilon=0}  \nonumber\\
&=\frac{1}{2\mu_0}2\left[\int\ud^3X^\prime (\nabla\left( A_\parallel(\mathbf{X}^\prime)+ \epsilon\delta(\mathbf{X}^\prime - \mathbf{X})\right))\nabla\delta(\mathbf{X}^\prime - \mathbf{X})\right]_{\epsilon=0}  \nonumber\\
&=\frac{1}{\mu_0}\int\ud^3X^\prime (\nabla A_\parallel(\mathbf{X}^\prime))\nabla\delta(\mathbf{X}^\prime - \mathbf{X})  \nonumber\\
&= \underbrace{\frac{1}{\mu_0}\int_\partial\ud^2X'(\nabla A_\parallel(\mathbf{X}^\prime))\delta(\mathbf{X}^\prime - \mathbf{X})}_0
- \frac{1}{\mu_0}\int\ud^3X^\prime (\nabla^2 A_\parallel(\mathbf{X}^\prime))\delta(\mathbf{X}^\prime - \mathbf{X})  \nonumber\\
&= -\frac{1}{\mu_0}\nabla^2A_\parallel
\end{align}
%FIXME SRG how does the boundary term vanish? For a periodic box, ok, but generally?
That is, this variation yields Ampere's law.
\begin{equation}
- \frac{1}{\mu_0}\nabla^2 A_\parallel = \sum_{sp} Z_{sp} e \int {\rm d}^3 {\bf v} \, \d{\bf X}{t} \cdot {\bf b} \langle f \rangle^\dagger  
\label{eq:ampereslaw}
\end{equation}

GKW uses the Poisson equation and Ampere's law above, i.e. they follow directly from the Lagrange description. 
Approximations, however, are made in the equations of motion of the particles. The Lagrange description yields the 
equations of motion through the variation of the action 
\begin{equation}
\delta S = \delta \int_{t_0}^{t_1} L_p {\rm d} t  
= \int_{t_0}^{t_1} \pd{ \gamma_a}{ z^b} \delta z_b \dot z^a + \gamma_a \delta \dot z^a - 
\pd{ H}{ z^a} \delta z^a {\rm d} t . 
\label{eq:eqmotionderivation}
\end{equation}
Integrating the term with $\delta \dot z^a$ once partial towards the time yields
\ifmoredetails
\begin{align}
\delta S = \int_{t_0}^{t_1}  \gamma_a \delta \dot z^a  {\rm d} t 
&=  \gamma_a \delta z^a\Big|_{t_0}^{t_1} - \int_{t_0}^{t_1} \dot\gamma_a \delta z^a  {\rm d} t \\
&=  - \int_{t_0}^{t_1} \Big(\underbrace{\pd{\gamma_a}{z^b}}_{\gamma_{a,b}}\pd{z^b}{t} - \pd{\gamma_a}{t}\Big)\delta z^a  {\rm d} t
\end{align}
and rename $a\leftrightarrow b$ in the first term of \eqref{eq:eqmotionderivation}, to get
\fi
  the equations of motion in the form 
\begin{align} 
[ \gamma_{b,a} - \gamma_{a,b} ] \d{z^b}{ t} &= \pd{ H}{ z^a} + \pd{ \gamma_a}{ t} \label{Lagrangemotion} 
\end{align}
The equation above can be split into components $(1,2,3)$ and the fourth component as follows 
\begin{align}
\Gamma = \left ( Z e {\bf A} + m {\bf u}_0 +  m v_\parallel {\bf b} + Z e \langle A_\parallel \rangle  {\bf b} \right )  
\end{align}
\begin{align}
(\nabla \times {\bf A}^*) \times \d{\bf X}{t} + \pd{{\bf A}^*}{v_\parallel} \d{v_\parallel}{t} 
&= - \nabla H - \pd{{\bf A}^*}{t} \label{eqmotion} &&\text{eq. of motion for $v_\parallel$}\\
\pd{{\bf A}^*}{v_\parallel} \cdot \d{\bf X}{t} &= \pd{H}{v_\parallel} \label{eqmotion2} &&\text{eq. of motion for position space coordinates }
\end{align}
where 
\begin{align}
Z e {\bf A}^* &= Z e {\bf A} + {\bf u}_0 + m v_\parallel {\bf b} + A_\parallel {\bf b} 
\label{eq:Astar}
\\
{\bf B}^* = \nabla \times {\bf A}^*
\end{align}
% FIXME so $Ze\mathbf{A}^* \equiv \mathbf{\Gamma}$ from \eqref{eq:plagrangian}?
% FIXME is there a gyroavg. around the $A_\parallel$ or not? in \eqref{eq:plagrangian}?

The equations \eqref{eqmotion} and \eqref{eqmotion2} above determine both the evolution of both the position as well as the parallel velocity of a gyro-centre, in given fields. 
A set of reduced equations can be derived by taking the inner product of the first equation \eqref{eqmotion} with $\,{{\rm d} {\bf X} / {\rm d} t}$ to 
obtain 
\begin{align}
0 + \left (\d{\bf X}{t} \cdot  \pd{{\bf A}^*}{v_\parallel} \right ) \d{v_\parallel}{t} 
&= -  \d{\bf X}{t} \cdot \nabla H -  \d{\bf X}{t} \cdot \pd{{\bf A}^*}{t} 
\label{motionred1}
\\
 \pd{{\bf A}^*}{v_\parallel} \cdot \d{\bf X}{t} 
&= \pd{H}{v_\parallel} 
\label{motionred2} 
\end{align}
and by inserting \eqref{motionred2} into \eqref{motionred1} 
\begin{align}
\pd{H}{v_\parallel} \d{v_\parallel}{t} 
+  \d{\bf X}{t} \cdot \nabla H 
&= 
 -  \d{\bf X}{t} \cdot \pd{{\bf A}^*}{t} 
\label{motionred3}
\end{align}

The essential observation here is that these equations will guarantee
energy conservation independent of the choice of
${{\rm d} {\bf X} / {\rm d} t}$. To proof this we will use that
\begin{align}
%\d{J_v(t,\mathbf{X}, v_\parallel, \mu)}{t} &= 0 \qquad |\cdot \frac{1 }{ J_v} 
%\label{eq:phasespacecons}
%\\
\frac{1 }{ J_v} \pd{J_v}{ t} + \frac{1 }{ J_v} (\nabla \cdot  J_v) \d{\bf X}{t} 
 + \frac{1 }{ J_v}\pd{J_v}{ v_\parallel}  \d{v_\parallel}{ t} 
 + \frac{1 }{ J_v}\pd{J_v}{ \mu}  \underbrace{\d{\mu}{t}}_0
&= 0,
\label{eq:phasespacecons2}
\end{align}
where $J_v$ is the Jacobian of the velocity space integration. 
%FIXME I do not really understand how the second equation is constructed. Is it really from the first line?|
This condition \eqref{eq:phasespacecons2} expresses conservation of the gyrocenter phase space volume of an element in the flow.
While this condition is perhaps not under all circumstances necessary to obtain
exact energy conservation, it is in our derivation an additional constraint 
that must be satisfied for the gyrocentre velocity.  
This observation allows one to make approximations to ${{\rm d}{\bf X}/ {\rm d}t}$, removing higher order $\rho_*$ corrections, without 
breaking energy conservation.

Adding the equations \eqref{motionred3} and \eqref{eq:phasespacecons2} above we have 
\begin{align}
\pd{ H}{ v_\parallel} \d{v_\parallel}{t}
+  \d{\bf X}{ t} \cdot \nabla H 
&+&H
%\left
(
\frac{1 }{ J_v} \pd{J_v}{ t} 
+ \frac{1 }{ J_v} \nabla \cdot  [ J_v \d{\bf X}{ t}  ] 
+ \frac{1 }{ J_v}\d{}{ v_\parallel}  [ J_v \d{v_\parallel}{ t}  ] 
%\right
)
&=
- \d{\bf X}{t} \cdot \pd{ {\bf A}^*}{ t} 
\end{align}
\begin{align}
\pd{H}{v_\parallel} \d{v_\parallel}{t} 
+ H \frac{1 }{ J_v} \underbrace{\pd{J_v}{ t}}_0
+ H \frac{1 }{ J_v}\pd{}{ v_\parallel} \left [ J_v \d{v_\parallel}{ t} \right ]
&+& 
  \d{\bf X}{t} \cdot \nabla H 
+ H \frac{1 }{ J_v} \nabla \cdot \left [ J_v \d{\bf X}{t} \right ]
&= 
-  \d{\bf X}{t} \cdot \pd{{\bf A}^*}{t} 
\\
\frac{1 }{ J_v} \pd{}{ v_\parallel} \left [J_v H \d{v_\parallel}{t} \right ] 
&+&
\frac{1 }{ J_v} \nabla \cdot \left [ J_v H \d{\bf X}{t} \right ] 
&= - \d{\bf X}{t} \cdot \pd{{\bf A}^*}{t} 
\end{align}

Then we split the Hamiltonian \eqref{eq:Hsplit} and use that the field part $H_1 = Z e \langle \phi \rangle$ is time dependent, whereas the rest $H_0$ is not.
\begin{align}
 \d{H}{t} &= \d{H_0}{t} + \d{H_1}{t} \\
%\pd{H}{v_\parallel}\d{v_\parallel}{t} + \nabla H \d{X}{t} + \cancel{\pd{H}{t}} &= \pd{H_0}{v_\parallel}\d{v_\parallel}{t} + \nabla H_0 \d{X}{t} + \nabla H_1 \d{X}{t} + \cancel{\pd{H_1}{t}}
\pd{H}{v_\parallel}\d{v_\parallel}{t} + \nabla H \d{X}{t} + \pd{H}{t} &= \pd{H_0}{v_\parallel}\d{v_\parallel}{t} + \nabla H_0 \d{X}{t} + \nabla H_1 \d{X}{t} + \pd{H_1}{t}
\pd{H}{v_\parallel}\d{v_\parallel}{t} + \nabla H \d{X}{t}&= \pd{H_0}{v_\parallel}\d{v_\parallel}{t} + \nabla H_0 \d{X}{t} + \nabla H_1 \d{X}{t} 
\end{align}
to obtain 
\begin{align} 
\d{ H_0 }{t} = 
\frac{1 }{ J_v} \pd{}{ v_\parallel} \left [J_v H_0 \d{v_\parallel}{t} \right ] + 
\frac{1 }{ J_v} \nabla \cdot \left [ J_v H_0 \d{\bf X}{t} \right ] 
&= - \d{\bf X}{t} \cdot \left [  \nabla H_1 + Z e \pd{{\bf A}^*}{t} \right ]  
\end{align}
Using furthermore that $\pd{f}{t} = 0$ we arrive at an energy balance equation. 
\begin{equation} 
\d{ }{ t} \left [ H_0 f \right ] 
= -  \d{\bf X}{t} \cdot \left [  \nabla H_1 + Z e \pd{{\bf A}^*}{t} \right ] f 
\label{eq:energyconsParticles}
\end{equation} 
The left hand side of this equation represents the change of kinetic energy.
The right hand side represents the inner product of
current and electric field. It is the transfer term between kinetic
and field energy.

We now would like to investigate the transferm term in more detail by deriving the corresponding energy balance from the field equations.
We consider first the Poisson equation, afterwards the Ampere equation.
Multiplying the time derivative of the Poisson equation \eqref{eq:Lfuncderivphi} with $\phi$ we obtain

\begin{align}
0 &= - \phi \pd{}{t}\sum_{sp}\int\ud^3v 
\frac{\delta}{\delta\phi}
\left(\int\ud^3X f_{s,tot}
    \left[\underbrace{Ze\ga{\phi}}_{H_1} + \frac{Z_s^2e^2}{2T_s}\left(\ga{\phi}^2 - \phi^2\right)\right]
 \right)
 - \sigma_a\phi \pd{}{t}\frac{\delta}{\delta\phi} \int \ud\mathbf{X} L_a
\end{align}
The functional derivative of the first term can be executed, and the rest is put to the right hand side.
We may also substract the neutrality equation for the background $F_{sp,M}$ and continue with $f_{sp}$ in place of $f_{sp,tot} = F_{sp,M} + f_{sp}$.
\begin{align}
\sum_{sp} Z_{sp} e \phi \pd{}{t} \int {\rm d}^3 {\bf v} \, \langle f \rangle^\dagger 
&=
 % \sum_{sp} \frac{Z_{sp}^2 e^2}{T_{sp}}\phi\pd{}{t} \int {\rm d}^3 {\bf v} \, [ \phi - \langle \langle \phi \rangle \rangle^\dagger ] F_{Msp} 
% +
\phi \pd{}{t}\frac{\delta}{\delta\phi}\sum_{sp}\int\ud^3v \int\ud^3X f_{s,tot}\frac{Z_s^2e^2}{2T_s}\left(\ga{\phi}^2 - \phi^2\right)
+
\sigma_a\phi \pd{}{t}\frac{\delta}{\delta\phi} \int \ud\mathbf{X} L_a
\end{align}
If we consider only quantities which are first order in $\rho_*$ then $f_{s,tot}$ is only $F_{s,M}$ and the fluctuation part of the distribution is neglegted for this term.
\begin{align}
 \sum_{sp}\int {\rm d}^3 {\bf v} \, Ze\phi \pd{\iga{f}}{t} 
&=  \phi \pd{}{ t} \left [ \frac{\delta H_{2E}}{\delta \phi} \right ] 
\end{align} 
where 
\begin{equation} 
H_{2E} = \sum_{sp}\int {\rm d}^3 {\bf v} \int\ud\mathbf{X}\, \frac{Z_{sp}^2 e^2}{2 T_{sp} } [ \phi^2 - \langle \phi \rangle^2 ] F_M + \sigma_a \int\ud\mathbf{X}L_{adiabatic} 
\end{equation} 
By putting both sides under a position space integral $\int\ud^3\mathbf{X}$ again, we can roll the gyroaverage over to $\phi$ and identify $H_1$.
\begin{align}
 \sum_{sp} \int {\rm d}^3 {\bf X} \int {\rm d}^3 {\bf v} \, H_1 \pd{ f}{ t} 
&=\int{\rm d}^3 {\bf X} \phi \pd{}{ t} \left [ \frac{\delta H_{2E}}{\delta \phi} \right ] 
\end{align}
And using ${\rm d} f / {\rm d} t = 0$ the first term in the equation can be rewritten in the form 
\begin{equation} 
- \sum_{sp}\int {\rm d}^3 {\bf X} \int {\rm d}^3 {\bf v} \, \d{\bf X}{t} \cdot \nabla H_1\, f  
= \int{\rm d}^3 {\bf X} \phi \pd{}{ t} \left [ \frac{\delta H_{2E}}{\delta \phi} \right ] 
\label{eq:energyconsPoisson}
\end{equation} 
%FIXME SRG I dont see how getting from $\pd{}{t}$ to $\d{}{t}$ is supposed to work.|
On the left hand side of this equation we can already identify a part of the energy transfer term.
Lets also evaluate the right hand side and construct time derivatives of quadratic quantities.
\begin{align}
- \sum_{sp}\int {\rm d}^3 {\bf X} \int {\rm d}^3 {\bf v} \, \d{\bf X}{t} \cdot \nabla H_1\, f  
= &
\int\ud\mathbf{X}
\phi \pd{}{ t}
\Big(
\sum_{sp}
 \int {\rm d}^3 {\bf v} \, \frac{Z_{sp}^2 e^2}{T_{sp}} [ \phi - \iga{\ga{\phi}} ] F_{M,sp}+{} \nonumber\\
&{}+ \sigma_a \frac{n_e e^2 }{ T_e} (\phi - \{ \phi \} ) 
\Big)\\
=& 
\int\ud\mathbf{X}
\pd{}{ t}
\Big(
\sum_{sp}
 \int {\rm d}^3 {\bf v} \, \frac{Z_{sp}^2 e^2}{T_{sp}} \frac{1}{2}[ \phi^2 - \ga{\phi}^2 ] F_{M,sp}
\Big)
+{} \nonumber\\
&{}+ \sigma_a \int\ud\mathbf{X}\frac{n_e e^2 }{ T_e} (\frac{1}{2}\pd{\phi^2}{t} - \phi\pd{}{t}\{ \phi \} ) 
\end{align} 
In order to arrive at the form of \eqref{eq:energyconservation}, we
rewrite the term with the flux surface averaged potential $\fsa{\phi}$
as follows. We take advantage of the fact that taking a flux surface avg. is executed by
integrating over $s$ like $\fsa{.} = \int . \ud s$, and taking further averages has no effect $\fsa{\fsa{\phi}}= \fsa{\phi}$.
\begin{align}
\int\ud s\frac{1}{2}\pd{}{t}(\phi - \fsa{\phi})^2  
&=\fsa{\frac{1}{2}2(\phi-\fsa{\phi})\pd{}{t}(\phi-\fsa{\phi})}
\\
&=\fsa{\phi\pd{}{t}\phi - \phi\pd{}{t}\fsa{\phi}-\fsa{\phi}\pd{}{t}\phi + \fsa{\phi}\pd{}{t}\fsa{\phi}
}
\\
&=\fsa{\phi\pd{}{t}\phi - \phi\pd{}{t}\fsa{\phi}}
-\fsa{\phi}\pd{}{t}\fsa{\phi} + \fsa{\fsa{\phi}\pd{}{t}\fsa{\phi}}
\\
&=\int \pd{}{t}\frac{1}{2}\phi^2-\phi\pd{}{t}\fsa{\phi} \ud s
\end{align}

We proceed analogously for Ampere's law \eqref{eq:ampereslaw} by multiplying it with $ A_\parallel\pd{}{t}$ and integrating over the whole of position space. 
\begin{align}
 \int\ud\mathbf{X}A_\parallel\pd{}{t}
 \sum_{sp} Z_{sp} e \int {\rm d}^3 {\bf v} \, \d{\bf X}{t} \cdot {\bf b} \langle f \rangle^\dagger  
&=
 -\int\ud\mathbf{X}A_\parallel\pd{}{t}\frac{1}{\mu_0} \nabla^2 A_\parallel
\\
\sum_{sp} Z_{sp} e\int\ud\mathbf{X}\ud^3\mathbf{v}\ga{A_\parallel}\pd{}{t}
   \d{\bf X}{t} \cdot {\bf b}  f 
&=
 -\frac{1}{\mu_0}\int_\partial\ud^2\mathbf{X} A_\parallel\nabla A_\parallel + \frac{1}{\mu_0} \int\ud\mathbf{X}(\nabla A_\parallel)\pd{}{t}\nabla A_\parallel
\end{align}
% FIXME SRG I dont see how to find $\mathbf{A}^*$ here.
% FIXME SRG There must be a gyroavg around some $A$ in the energy balance, too, somewhere.
We obtain 
\begin{equation} 
\sum_{sp} Z_{sp} e \int {\rm d}^3 {\bf X} {\rm d}^3 {\bf v} \,\d{\bf X}{t} \cdot \pd{{\bf A}^*}{t}f 
=
 \pd{}{t}\frac{1 }{ 2 \mu_0} \int {\rm d}^3 {\bf X} \, \vert \nabla A_\parallel \vert^2   
\label{eq:energyconsAmpere}
\end{equation}
Consequently the energy conservation theorem results from adding
\eqref{eq:energyconsParticles}, \eqref{eq:energyconsPoisson} and \eqref{eq:energyconsAmpere}, which yields
\index{Noether's theorem, Energy conservation}  
\begin{align}
  \d{}{t}\int\ud^3\mathbf{X}\ud^3\mathbf{v}H_0f 
&= - \int\ud^3\mathbf{X}\ud^3\mathbf{v} \d{\bf X}{t} \cdot \left [  \nabla H_1 + Z e \pd{{\bf A}^*}{t } \right ] f 
\\
&= \int{\rm d}^3 {\bf X} \phi \pd{}{ t} \left [ \frac{\delta H_{2E}}{\delta \phi} \right ] +  \frac{1 }{ 2 \mu_0} \int {\rm d}^3 {\bf X} \, \vert A_\parallel \vert^2   
\\
\pd{E}{t} &= 0\\
E &=  \sum_{sp} \int {\rm d}^3 {\bf X} {\rm d}^3 {\bf v} \, \left ( \frac{1 }{ 2} m v_\parallel^2 + \mu B  + \frac{1 }{ 2} m u_0^2 \right ) f + \nonumber{}\\
&+
\sum_{sp} \int {\rm d}^3 {\bf X} {\rm d}^3 {\bf v} \, \frac{Z_{sp}^2 e^2}{2 T_{sp} } [ \phi^2 - \langle \phi \rangle^2 ] 
F_{Msp} + \sigma_a \int {\rm d}^3 {\bf X} \, \frac{n_e e^2 }{ 2 T_e} (\phi - \{ \phi \} )^2 + \frac{1 }{ 2 \mu_0} \int {\rm d}^3 {\bf X} \, \vert \nabla A_\parallel \vert^2 
\label{eq:energyconservation}
\end{align}
where the first part is the kinetic energy while the second part is the field energy. The Noether theorem on the original Maxwellian 
gives exactly the same energy theorem. Furthermore, note that no assumption has been made on the form of ${\rm d} {\bf X} / {\rm d} t$ other 
than that the equations and ?? must be satisfied.

Taking the cross product of eq.~(\ref{eqmotion}) with the unit vector along the magnetic field $\mathbf{b}$ one obtains the velocity of the gyro-centre 
\begin{align} 
\mathbf{b}\times\left(\mathbf{B}^* \times \d{\bf X}{t} + \pd{{\bf A}^*}{v_\parallel} \d{v_\parallel}{t} \right)
&= - \mathbf{b}\times\left(\nabla H - \pd{{\bf A}^*}{t}\right)
\\
\mathbf{B}^*(\mathbf{b}\cdot\d{\bf X}{t}) - \d{\bf X}{t}\underbrace{(\mathbf{b}\cdot\mathbf{B}^*)}_{B^*_\parallel}
+ \mathbf{b}\times\pd{{\bf A}^*}{v_\parallel} \d{v_\parallel}{t} 
&= - \mathbf{b}\times\left(\nabla H - \pd{{\bf A}^*}{t}\right)
\\
\d{ {\bf X} }{ t}
&= 
\frac{\mathbf{B}^*}{B^*_\parallel}(\mathbf{b}\cdot\d{ {\bf X} }{ t})
 + \frac{\mathbf{b}}{B^*_\parallel}\times\nabla H 
+ \frac{\mathbf{b}}{B^*_\parallel}\times\underbrace{\pd{\mathbf{A}^*}{v_\parallel}}_{m\mathbf{b}} \d{v_\parallel}{t} 
- \frac{\mathbf{b}}{B^*_\parallel}\times\underbrace{\pd{\mathbf{A}^*}{t}}_{\pd{A_\parallel}{t}\mathbf{b}}
\\
\d{\bf X}{t} &= \frac{{\bf B}^*}{B^*_{\parallel}} v_\parallel + \frac{{\bf b}}{B^*_\parallel} \times \nabla H 
\label{eq:gyrocentrevelocity}
\end{align}
where
\begin{equation} 
B_\parallel^* = {\bf b} \cdot {\bf B}^* = {\bf b} \cdot \nabla \times {\bf A}^* 
\label{eq:Bparallelstar}
\end{equation}
and $J_v = B_\parallel^*$. 
It was also used that the parts of $\mathbf{A}^*$ which are dependent of $v_\parallel$ and $t$ are parallel to $\mathbf{b}$, cf \eqref{eq:Astar}.

When used in the form of \eqref{eq:Bparallelstar}, $B_\parallel^*$ is a function of the field $\langle A_\parallel \rangle$, and therefore of time. 
Since ${\rm d} {\bf X} / {\rm d} t$ depends nonlinearly on $B_\parallel^*(t)$ this complicates the numerical solution. At each 
time point the gyro-centre velocity would have to be constructed explicitly and the various terms in the equation can not 
be precalculated and stored in matrix format.

The approximation made in GKW is the approximation of $B_\parallel^*$ by $B$. Working out ${\bf B}^*$ 
with the Hamiltonian from \eqref{eq:Hsplit}
we then obtain 
\begin{align} 
\d{{\bf X}}{t}  &\approx 
\frac{{\bf B}^* }{ B} v_\parallel 
+ \frac{{\bf b}}{ B} \times \nabla H\\
% \left( \frac{1}{2} m v_\parallel^2 + \mu B +  \frac{1 }{ 2} m u_0^2 +Z e \langle \phi \rangle\right) 
&=
\frac{1}{B} v_\parallel \nabla\times\left({\bf A} + \frac{1}{Ze}{\bf u}_0 + \frac{m v_\parallel}{Ze} {\bf b} + \frac{1}{Ze}A_\parallel {\bf b} \right)
+ \frac{{\bf b}}{ B} \times \nabla H
\\
&=v_\parallel {\bf b} + \frac{m v_\parallel^2}{ZeB} \nabla \times {\bf b} + \frac{m v_\parallel}{ZeB} {\bf \Omega}
 +  \frac{v_\parallel}{B} \nabla \times [ \langle A_\parallel \rangle {\bf b} ] 
+ \frac{1 }{ B} {\bf b} \times \nabla H 
\label{eq:gyrocentrevelocity2}
\end{align}
%FIXME where $\mathbf{\Omega} = \frac{1}{m}\nabla\times\mathbf{u}_0$?
With the approximation $J_v = B^*_\parallel \approx B$ it can be shown that eq.~\eqref{eq:phasespacecons2} is satisfied.
%FIXME SRG What is $J_v$?

A few remarks should be made here. First, the approximation
\begin{equation} 
\nabla \times \ga{ A_\parallel} {\bf b}
= 
\nabla \ga{A_\parallel}\times\mathbf{b} + \ga{A_\parallel}\nabla\times\mathbf{b}
 \approx - {\bf b} \times \nabla \langle A_\parallel \rangle 
\label{eq:approxtofindExBdrift2}
\end{equation}
is used in \eqref{eq:approxtofindExBdrift} to split the gyrocentre velocity into grad-B and ExB drifts.
This approximation is problematic for exact energy conservation since in this case eq.~\eqref{eq:phasespacecons2} is not satisfied. 
%FIXME how does this enter \eqref{eq:phasespacecons2}?|

\subsection{Local limit}

\label{energy-conservation-local-limit}
A brief look ahead for the local limit approximation may be in order. The local limit corresponds to a further approximation also in the gyro-centre velocity and must be 
separately discussed. 
Essentially, the local limit is the limit where the scale length of perpendicular dynamics, which is of the order of the Larmor radius $\rho_\tref$, is much much smaller than the equilibrium scale length $R_\tref$.
\begin{align}
  \rho_\tref \ll R_\tref
\end{align}
that means
\begin{align}
  \rho_* = \frac{\rho_\tref}{R_\tref} \ll 1
\end{align}
This implies further assumptions:
\begin{itemize}
\item all background and geometry quantities are assumed to be a function of the parallel coordinate ($s$) only.
\item periodic poundary conditions
\end{itemize}

The gyro-centre velocity \eqref{eq:gyrocentrevelocity2} does not depend on the coordinates perpendicular to the field and, therefore the divergence of this velocity is different from the expression derived above. 
% FIXME ?? Sure it does, there is $\nabla\ga{\phi}$, and this has dependencies on perp coords in leading order?!
% FIXME Which divergence expression above? There is none!?
The gyrocentre velocity \eqref{eq:gyrocentrevelocity2} can be
understood to consist of parallel motion and several drift motions.
\begin{align}
\d{\mathbf{X}}{t}
&=
v_\parallel {\bf b} + \frac{m v_\parallel^2 }{ ZeB} \nabla \times {\bf b} + \frac{m v_\parallel }{ ZeB} {\bf \Omega}
+ \frac{v_\parallel }{ B} \nabla \times [ \langle A_\parallel \rangle {\bf b} ] 
+ \frac{1 }{ B} {\bf b} \times \nabla \left ( \frac{1 }{ 2} m v_\parallel^2 + \mu B  + \frac{1 }{ 2} m u_0^2 
+ Z e \ga{\phi}
\right ) ,
\\
&= v_\parallel\mathbf{b}  
+ {\bf v}_D 
+  \underbrace{\frac{v_\parallel }{ B} \nabla \times [ \langle A_\parallel \rangle {\bf b} ]  + \frac{1 }{ B} {\bf b} \times \nabla Z e \langle \phi \rangle}_{\mathbf{v}_\chi}
\end{align}
where $\mathbf{v}_D$ is a perpendicular drift velocity due to the background magnetic fields and reference frame rotation. A drift of the form of an ExB drift can be identified if the additional approximation \eqref{eq:approxtofindExBdrift} is made:
\begin{equation} 
{\bf v}_\chi = \frac{{\bf b} \times \nabla \chi}{B} \qquad 
\textrm{with} \qquad \chi = \langle \phi \rangle - v_\parallel \langle A_\parallel \rangle ,
\end{equation}
As mentioned above at \eqref{eq:approxtofindExBdrift2}, this approximation makes the system violate phase space conservation, though.

Nevertheless, in the local limit, i.e. the limit of small $\rho_*$, phase space conservation is valid again:
In this limit (i.e. to lowest order in $\rho_*$) the drift velocity ${\bf v}_D$ and ExB velocity ${\bf v}_E$ do not have a component in the direction of $\mathbf{e}^s\equiv\nabla s$, i.e. ${\bf v}_D\cdot \nabla s = v_D^s = {\bf v}_E \cdot \nabla s =v_E^s= 0$. 
This is also strictly imposed in the code when the finite $\rho_*$ parallel derivatives are turned off.

With the velocity space jacobian
\begin{align}
  J_v \propto B
\end{align}
%FIXME I do not find an expression for the velspace jacobian anywhere.
we consider
\begin{align}
B\d{\mathbf{X}}{t}\cdot\nabla f = B ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E ) \cdot \nabla f 
&=
\nabla \cdot [ B ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E ) f ] - f\nabla \cdot [ B ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E )] 
\\
&=
 ... - f\nabla B(s)\cdot ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E ) - f B(s)\nabla\cdot ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E )
% \\
% &=
%  ... - f(\mathbf{e}^s\pd{}{s} + \mathbf{e}^\psi\pd{}{\psi} + \mathbf{e}^\zeta\pd{}{\zeta} ) \cdot [ B(s) ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E )]
\\
&=
 ... - f(\nabla s\pd{B(s)}{s}) \cdot ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E )
\\
&\qquad\qquad
 - fB(s)\frac{1}{J}\left(
\pd{J(v_\parallel {\bf b} + {\bf v}_D)^s}{s} + \pd{J({\bf v}_E)^\psi}{\psi} + \pd{J({\bf v}_E)^\zeta}{\zeta} 
\right) 
\\
&=
 ... - f\pd{B(s)}{s} v_\parallel \underbrace{\mathbf{e}^s\cdot\frac{\mathbf{B^s\mathbf{e}_s}}{B}}_{0}
\\
&\qquad\qquad
 - fB(s)\frac{1}{J}\left(
\pd{J(v_\parallel {\bf b} + {\bf v}_D + \mathbf{v}_E)^s}{s} + \pd{J({\bf v}_E)^\psi}{\psi} + \pd{J({\bf v}_E)^\zeta}{\zeta} 
\right) 
\\
&=
 - fB(s)\frac{1}{J}\left(
\pd{Jv_\parallel \frac{\mathbf{B^s\mathbf{e_s}}}{B}\cdot\mathbf{e}^s}{s} + \pd{J({\bf v}_E)^\psi}{\psi} + \pd{J({\bf v}_E)^\zeta}{\zeta} 
\right) 
\\
&=
 - fB(s)\frac{1}{J}\left(
\pd{J({\bf v}_E)^\psi}{\psi} + \pd{J({\bf v}_E)^\zeta}{\zeta} 
\right) 
\\
&= ? FIXME
\\
&= \nabla \cdot [ B ( v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E ) f ]
\end{align}
i.e. the rewrite above shows that the whole gyro-centre velocity satisfies 
\begin{equation}
%\frac{1 }{ B}
 \nabla \cdot \left [ B \d{\bf X}{t} \right ] \propto \nabla \cdot \left [ J_v \d{\bf X}{t} \right ] = 0
\end{equation}
The phase space conservation \eqref{eq:phasespacecons2} is then reduced to
\begin{align}
  \frac{1 }{ J_v} \pd{J_v}{ t} 
+ \frac{1 }{ J_v} (\nabla \cdot  J_v) \d{\bf X}{t} 
 + \frac{1 }{ J_v}\pd{J_v}{ v_\parallel}  \d{v_\parallel}{ t} 
 &= 0,
\end{align}
Then 
\begin{equation} 
\pd{}{ v_\parallel } \left [ \d{v_\parallel}{t} \right ] \quad \rightarrow \quad 
\pd{}{ v_\parallel } \left [ - \frac{1 }{ m} {\bf b} \cdot \mu \nabla B \right ] = 0 
\end{equation}
must be zero in order for phase space conservation to apply. 
% FIXME why? how does this enter the phasespacecons?
% FIXME free floating term here? What is this supposed to tell?
% FIXME Rico writes one has to implement 
% \begin{align}
%   \pd{}{v_\parallel}\left[\d{v_\parallel}{t}f\right]
% \end{align}
% rather than
% \begin{align}
%   \d{v_\parallel}{t}\pd{f}{v_\parallel}
% \end{align}
% but this is not the same as the term left of the arrow.
% FIXME nobody says why, I suppose this is to implement a conservative
% method - as far as I understand the matter (and the book of Toro on
% Riemand Solvers for CFD), however, (1) a conservative method only
% helps to conserve the quantitity the equation conserves ($f$ here) and
% (2) if the solution is sufficiently smooth (i.e. no shock
% discontinuities) it does not matter.

The terms on the right (the trapping term) represent the only term of ${\rm d} v_\parallel / {\rm d} t$ that is kept for the perturbed distribution $f$. 
Indeed for this term the parallel velocity derivative is zero and, consequently the phase space volume is conserved in the evolution of $f$.

The local limit is obtained through an expansion in $\rho_*$, and neglection of all terms which are of order $\rho_*^2$ or higher. In doing so, the velocity nonlinearity is neglected.
%FIXME What is the precise expression for the velocity nonlinearity?
A more correct approach 
%(FIXME cite Scott) 
would be to neglect contribution already at the level of the Lagrangian and make no approximation thereafter. This would ensure energetic consistency between particle and field dynamics.
Following such a calculation, it turns out that neglection of certain nonlinear terms (the ``velocity nonlinearity'') is not allowed, i.e. they must be kept in the model to achieve energy conservation although they are formally $\orderof (\rho_*^2)$ terms in the particle dynamics equations.

In fact, in the local limit with periodic boundary conditions,
conservation of the energy \eqref{eq:energyconservation} is
problematic since the total field energy is always positive, whereas
the total kinetic energy
\begin{equation} 
\int {\rm d}^3 {\bf X} \int {\rm d}^3 {\bf v} \, \left [ \frac{1 }{ 2} mv_\parallel^2 + \mu B \right ] f 
\end{equation} 
integrates to zero for all Fourier modes except the mode that has a zero wave vector in both perpendicular directions (below this mode is refered to as the $(0,0)$ mode). 
The $(0,0)$ mode is often not kept in the evolution equation (particularly in linear
 simulations), and it is then not possible to conserve the energy since the positive 
field energy can not be balanced with a 
negative total kinetic energy.

Thus, it is necessary to keep both the $(0,0)$ mode
and the velocity nonlinearity in the evolution equations, in order to be able to achieve energy conservation (there are further obstacles though).
It is beneficial to implement  the velocity nonlinearity in the form of \eqref{eq:vel-nonlin-implementation} rather than \eqref{eq:vel-nonlin-not-implementation}.
\begin{align} 
\pd{f}{t} &\overset{+}{=} -\pd{}{v_\parallel} \left [ \d{v_\parallel}{t} f \right ] 
\label{eq:vel-nonlin-implementation}
\\
&= -\d{}{t}(\pd{v_\parallel}{v_\parallel}) f - \d{v_\parallel}{t} \pd{f}{v_\parallel}\nonumber\\
&= - \d{v_\parallel}{t} \pd{f}{v_\parallel}
\label{eq:vel-nonlin-not-implementation}
\end{align}
The form \eqref{eq:vel-nonlin-implementation} ensures phase space conservation in the local limit where the divergence of the gyro-centre velocity is zero. 
In the equation above the change in the parallel velocity is considered only through the electro-magnetic field. 
It can easily be shown that when integrated over the whole computational domain the term only contributes to the $(0,0)$ mode. 
The $(0,0)$ mode has no spatial derivatives perpendicular to the field, and except through the velocity nonlinearity it does not couple to any other mode. 
The $(0,0)$ mode therefore only acts as a reservoir for the kinetic energy essuring that the balance equation for the energy holds.

%FIXME this sounds as if it was a success. According to Rico it is not - but Rico does not show data either!

\subsection{Toroidal momentum conservation}

The equations of motion (Eq.~(\ref{Lagrangemotion}) can be written in the form \eqref{eq:Lagrangemotion-other-form}.
\begin{align} 
[ \gamma_{b,a} - \gamma_{a,b} ] \d{z^b}{ t} &= \pd{ H}{ z^a} + \pd{ \gamma_a}{ t}
\nonumber\\
\gamma_{b,a}\d{z^b}{ t} - \gamma_{a,b} \d{z^b}{ t} &= \pd{ H}{ z^a} + \pd{ \gamma_a}{ t}
\nonumber\\
 - \pd{ \gamma_a}{ t} - \gamma_{a,b} \d{z^b}{ t} &= \pd{ H}{ z^a} - \gamma_{b,a}\d{z^b}{ t}
\nonumber\\
\pd{ \gamma_a}{ t} + \gamma_{a,b} \d{z^b}{ t} &= -\pd{ H}{ z^a} + \gamma_{b,a}\d{z^b}{ t}
\nonumber\\
\d{\gamma_a}{t} &= - \pd{H}{z^a} - \gamma_{b,a} \d{z^b}{t} 
\label{eq:Lagrangemotion-other-form}
\end{align} 
%FIXME I get a different sign here, + instead of the - that was here before.
To have a symmetry in coordinate $z^a$ means that Hamiltonian $H$ and $\gamma_b$ are independent of $z^a$.
\begin{equation} 
\pd{H}{z^a} = 0 \qquad 
\gamma_{b,a} = 0 
\end{equation} 
and from \eqref{eq:Lagrangemotion-other-form} follows then
\begin{equation} 
\d{ \gamma_a }{ t} = 0 
\end{equation}
This represents a conservation equation, in the sense that $\gamma_a$ is constant along along the trajectory of a particle.

The case of interest here is the toroidal symmetry of the tokamak (and the consequent 
conservation of toroidal angular momentum). To discuss this, note that both $\langle \phi \rangle$ and 
$\langle A_\parallel \rangle$ are a function of the toroidal angle $\varphi$, but background quantities such as $\mathbf{A}$, $\mathbf{u_0}$, $\mathbf{b}$, and $H_0$ (cf. \eqref{eq:Hsplit}) are not.
Choosing the toroidal angle $\varphi$ as coordinate one then
obtains from \eqref{eq:Lagrangemotion-other-form}
\begin{align} 
\d{\gamma_\varphi}{t} &= - \pd{H_1}{\varphi} + Ze \pd{\langle A_\parallel \rangle}{\varphi } \mathbf{b}\cdot \d{\bf X}{t} 
\nonumber\\
&= - \pd{H_1}{\varphi} + Ze \pd{\langle A_\parallel \rangle}{\varphi } v_\parallel \\
\d{\gamma_\varphi}{t} &= - \pd{H_1}{\varphi} - Ze \pd{\langle A_\parallel \rangle}{\varphi } v_\parallel \d{\bf X}{t} 
\end{align}
%FIXME Mistake in second term here, right?|
Multiplying this equation with $f$ and integrating over the whole phase space (using ${{\rm d} f / {\rm d} t} = 0$ we obtain 
\begin{align} 
\int {\rm d}^3 {\bf X} {\rm d}^3\d{\gamma_\varphi}{t}f &= - \int {\rm d}^3 {\bf X} {\rm d}^3\mathbf{v}\left [
\pd{H_1}{\varphi} - Ze \pd{\langle A_\parallel \rangle}{\varphi } v_\parallel
 \right]f
\\
 \d{}{ t} \int {\rm d}^3 {\bf X} {\rm d}^3 {\bf v} \, \gamma_\varphi f&= - \int {\rm d}^3 {\bf X} {\rm d}^3\mathbf{v}\left [
\pd{H_1}{\varphi} - Ze \pd{\langle A_\parallel \rangle}{\varphi } v_\parallel
 \right]f
\\
\pd{}{ t} \int {\rm d}^3 {\bf X} {\rm d}^3 {\bf v} \, \gamma_\varphi f &= 
- \int {\rm d}^3 {\bf X} {\rm d}^3 {\bf v} \,\left [ \pd{H_1}{\varphi} + Z e \pd{\langle A_\parallel\rangle}{ \varphi} v_\parallel \right ] f 
\end{align}
The right hand side of this equation can be shown to vanish
%FIXME text missing?|

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/benchmarks.tex
(1) Added gkw Clone gwk  #3

\chapter{Benchmarks}

% Benchmarks to Add (even if just the reference):

% Edge shortfall case: AUG
% Edge shortfall case: DIII-D
% Casson MAST cases
% Dickinson MAST MTM case (EPS 2013)
% Henderson MAST case (EPS 2013)
% Guttenfelder / Hornsby MTM case
% Guttenfelder / Buchholz NSTX case
% Cottier QualizKiz cases
% Self-Benchmarks for the flux tube nonspectral version:
%   Linear, RH, nonlinear
% NEMORB global benchmarks

In this section we describe a set of benchmarks. 
The benchmark cases have been chosen to maximise the effect of each of the different implemented 
terms as much as possible.
Together they give evidence for the correct implementation of the entire model. 
Linear benchmarks have been made against the GS2 code which is well established. 
Indeed the model equations are equivalent, with the exception that GKW includes the effect of a 
rigid body toroidal rotation, while GS2 does not. The implementation is to some extent similar;
both 
use a spectral approach for the plane perpendicular to the magnetic field, and finite differences for 
the other directions. 
The main difference lies in the velocity space discretisation where GKW uses $(v_\parallel, \mu)$
coordinates, whereas GS2 uses pitch angle and energy.  The GS2 velocity grid allows an efficient
implicit time integrator, while the default grid of GKW is suitable for an explicit time integrator
which facilitates parallelisation.  Also the finite difference scheme used
for GKW in the benchmarks is fourth order while GS2 has a second order scheme. 
In this section all quantities have been converted to use gyro-Bohm units.

\section{Linear Benchmarks\label{linearbenchmarks}}

The standard benchmark for linear problems is the growth rate of the Ion Temperature Gradient 
mode (ITG) as a function of the poloidal wave vector $k_\theta \rho_s$ for the so-called 
Cyclone base case \cite{DIM00} ($q = 1.4$, $\hat s = 0.78$, $\epsilon = 0.19$, $R / L_T = 6.9$, 
$R/L_n = 2.2$, $T_e / T_i = 1$, electro-static (zero $A_\parallel$), and adiabatic electrons). 
The growth rate for this case is shown in the right panel of Fig.~\ref{fig:cyclone-linear} as a function 
of $k_\theta\rho_s$ for various values of the normalised ion temperature gradient ($R/L_T = 6.9$,
8.28 10.35, 12.44, and 15.18) calculated with both GKW and GS2. 
As can be seen from the figure, good agreement is obtained. 
The left panel, furthermore, shows the comparison of the potential as a function of the coordinate 
along the field line ($s$) for both codes.

\begin{figure}[htb] 
\begin{center}
\includegraphics[width=14truecm]{cyclone-base-linear3.eps}
\caption{Left: mode structure of the Cyclone base case. The full line is the 
real part and the dotted line is the imaginary part of the potential calculated with GS2. 
The corresponding results of GKW are given by the diamonds. Right: growth rate as a function 
of $k_\theta \rho_s$ for various values of $R/L_{Ti}$ (numbers given in the figure). The full 
lines give the results of GS2 while the diamonds are the corresponding results of GKW.} 
\label{fig:cyclone-linear}
\end{center}
\end{figure}

For nonlinear runs the proper response of the zonal flows is of utmost importance. 
The standard benchmark \cite{CAN03,JEN05,JOL07} that addresses the physics of the zonal flow / geo-acoustic 
mode is the Rosenbluth-Hinton test which was described analytically in Ref. \cite{HINROS}. 
In this benchmark the initial condition is an ion density 
perturbation with a finite (small) radial wave vector, and no dependence on either $s$ or $\zeta$.  
The adiabatic electron response is used, keeping the correction due to the flux surface average 
of the potential. The density perturbation generates a potential perturbation and excites the so
called geo-acoustic mode. This mode is damped and a small residual poloidal flow remains. Fig. 
\ref{zonalfig} gives the potential perturbation as a function of time. The relevant parameters used for this 
benchmark are $q = 1.3$, $\epsilon = 0.05$, $k_\psi \rho_s = 0.02$, electrostatic, collisionless. 
Rather large grid sizes ($N_s = 128$ and $N_{v_\parallel} = 128$) are used to avoid the recurrence 
problem \cite{CAN03}. The residual potential is given by the equation 
\begin{equation} 
{ \phi (t= \infty) \over \phi (t= 0) } = {1 \over 1 + q^2 \Theta / \epsilon^2} \qquad 
{\rm with} \qquad \Theta = 1.6 \epsilon^{3/2} + 0.5 \epsilon^2 + 0.36 \epsilon^{5/2} 
\end{equation}
The original derivation of Ref. \cite{HINROS} is accurate to lowest order in $\epsilon$ only, i.e. 
it retained only the first term $(1.6 \epsilon^{3/2})$ in $\Theta$. 
The higher order terms have been calculated in Ref.~\cite{XIA06}. 
For the parameters of the simulation shown in the left panel the residue is $\phi(t = \infty) / \phi(t= 0) 
= 0.0713$ smaller than the result of Ref. \cite{HINROS} (0.0764) but in very good agreement with the 
results of Ref. \cite{XIA06} (0.0710). The right panel shows the residue as a function of $\epsilon$. 
Good agreement with the analytic theory is obtained provided the finite $\epsilon$ effects are kept.  
\begin{figure}[htb] 
\begin{center}
\includegraphics[width=14truecm]{RHbench.eps}
\caption{Zonal flow benchmark. Left: The potential perturbation (normalised to the potential at 
$t = 0$) as a function of normalised time, for $q = 1.3$, $\epsilon = 0.05$. The horizontal line gives the 
residual (0.0711) as predicted in Ref.~\cite{XIA06}. For this value of $\epsilon$ a numerical residual 
of 0.0713 is obtained, i.e. in agreement with the analytic result to within 0.3 \%
Right: Residual $\phi (t = \infty) / \phi(0) $ as a function of $\epsilon$ for 
$q = 1.3$. The circles are the result of GKW, the dotted line is the Rosenbluth-Hinton result \cite{HINROS} and 
the full line gives the analytic result of Xiao-Catto \cite{XIA06}. \label{zonalfig}} 
\label{overall1}
\end{center}
\end{figure}

The Cyclone base case with adiabatic electrons does not check the correct implementation of the 
kinetic electron response, and in general is not very sensitive to the effects associated with trapped 
particles. 
The physics effects associated with kinetic electrons can be well benchmarked by considering a Trapped Electron 
Mode. A benchmark with GS2 is shown on the left of Fig.~\ref{overall2}. The parameters are 
taken from a discharge in the ASDEX Upgrade tokamak \cite{PEE05}: $\hat s = 1.07$, 
$q = 1.57$, $\epsilon = 0.177$, $T_e / T_i = 3$, $R/L_{Ti} = 0$, electro-static, collisionless, 
and with an ion to electron mass ratio of a Deuterium plasma. 
The values of the density gradient as well as the temperature gradient are scanned. 
The agreement for the growth rates is good.

\begin{figure}[htb] 
\begin{center}
\includegraphics[width=16truecm]{overallfig-bench.eps}
\caption{Left: Benchmark of a trapped electron mode case. Shown is the normalised growth 
rate $R \gamma / c_s$ as a function of $R/L_{Te}$ for various values of $R/L_n$ indicated in 
the figure. Blue lines are the results of GS2 while red diamonds give the results of GKW. 
Middle: A benchmark of the growth rates of the ITG / Kinetic ballooning mode versus the plasma beta. 
Blue lines and circles 
are results from GS2 while red diamonds give the results of GKW. Right: The fluxes, normalised
to the ion heat flux as a function of beta. Lines give the results of GS2, while the diamonds give
the results of GKW.} 
\label{overall2}
\end{center}
\end{figure}

Finite beta effects are tested for an ITG case increasing the plasma beta. The middle panel of 
Fig. \ref{overall2} shows the normalised growth rate as a function of the electron beta $\beta = n_e T_e 
/ (B^2 / 2 \mu_0)$ for the Waltz standard case $R/L_{Ti} = R/ L_{Te} = 9$, $R/ L_n = 3$, $q = 2$
$\hat s = 1$, $\epsilon = 0.166$ $T_e/T_i = 1$, and with a mass ratio of a Deuterium plasma. The ITG is 
stabilised by the finite beta effects and, at sufficient high values of beta, a kinetic ballooning 
mode is destabilised. The agreement for the growth rate between GKW and GS2 is again good. 
The electro-magnetic case also provides for a good benchmark of the calculation of the fluxes, since 
all fluxes (also the ones due to the magnetic flutter) are nonzero in this case. The right panel of 
Fig.~\ref{overall2} shows the comparison of the fluxes calculated by GS2 and GKW. Since for 
a linear problem the amplitude of the mode and, therefore, the magnitude of the fluxes is arbitrary, 
all fluxes have been normalised to the ion heat flux. This allows for a straightforward comparison 
between the codes. All calculated fluxes are again in good agreement.

Effects of magnetic field compression are also tested with the linear Waltz standard test case $R/L_{Ti} = R/ L_{Te} = 9$, $R/ L_n = 3$, $q = 2$, $\hat s = 1$, $\epsilon = 0.166$, $T_e/T_i = 1$, the mode with the fastest electrostatic growth rate has been used: $k_{\perp} \rho_s = 0.3$. Fig. \ref{bpar-bm} shows that the ITG mode is hardly affected by the perturbation $B_{1 \parallel}$ but the kinetic ballooning mode is further destabilised with increasing $\beta$. Both with the growth rates and the parallel structure a good agreement with GS2 has been obtained.

\begin{figure}[htb] 
\begin{center}
\includegraphics[scale=0.25]{bpar-growthrates-waltz-linear.eps}
\includegraphics[scale=0.25]{bpar-parallelstructure-waltz-linear.eps}
\caption{Waltz linear test case. Left: Benchmark of growth rates of ITG and kinetic ballooning modes with magnetic field compression on and off. Right: Parallel structure of $B_{1 \parallel}$ in ITG and KBM regimes.} 
\label{bpar-bm}
\end{center}
\end{figure}

A collisionality benchmark of the pitch angle scattering against GS2 for a TEM is presented in Fig.~\ref{colls-tem}.  
To isolate the effect of e-e collisions and e-i collisions, the collisions input
variable $Z_{\rm eff}$
(a multiplier for the e-i collision rate) was varied.  Scattering of ions (off either species) made no difference for this benchmark 
(turning them on / off did not change the result of either code). This result was obtained with the Arakawa scheme and zero dissipation in $v_\parallel$. At low collisionalities, GKW convergence for a TEM requires many $v_\parallel$ points (the 64 used for this benchmark are not quite sufficient).

\begin{figure}[htb] 
\begin{center}
\includegraphics[scale=0.8]{bench-ee-sa-all.eps}
\caption{Collisions benchmark (pitch angle scattering) for a TEM with varying
$Z_{\rm eff}$} 
\label{colls-tem}
\end{center}
\end{figure}

The pitch angle collisions have also been benchmarked against GS2 in Ref.~\cite{PEE09scmode}, with good agreement 
obtained for the particle flux as a function of collisionality. 
The treatment of arbitrary toroidal geometry has been benchmarked against GS2 by performing an elongation scan. The parameters considered are $R/L_{Ti} = R/ L_{Te} = 8.9$, $R/ L_n = 2.85$, $q = 1.42$ $\hat s = 1.25$, $\epsilon = 0.182$, $T_e/T_i =1$ for deuterium ions and adiabatic electrons. 
The elongation of the last closed flux surface is varied from $\kappa=1.2$ to $\kappa=1.6$ and the corresponding MHD equilibrium is calculated using the CHEASE code \cite{LUT96}.
This equilibrium is then used for the calculation of the metric tensors both in GKW and GS2. 
The results for the linear growth rate as a function of $k_\theta\rho_s$ are shown in Fig.~\ref{geom-bench} showing a good agreement between the two codes.  The wavevector
normalisation (projection) in the two codes in general differs (GKW uses
$g^{\zeta\zeta}(LFS)$ while GS2 uses ${\cal E}^{\psi\zeta}$ \cite{dorland-note} which are only equivalent in `$s-\alpha$' geometry) and must
be corrected for (in this case the GKW $k_\theta$ inputs were rescaled). 
Note that the results are significantly different from the ones obtained with the simplified `$s-\alpha$' equilibrium shown by 
the dashed line (obtained with GKW).
\begin{figure}[htb] 
\begin{center}
\includegraphics[width=6truecm]{geom-bench.eps}
\caption{Benchmark of the geometry treatment. The normalised growth 
rate $R \gamma / c_s$ is shown as a function of GS2 $k_\theta\rho_s$ for various values of the last closed flux surface elongation $\kappa$ indicated in 
the figure. Blue lines are the results of GS2 while red diamonds give the results of GKW. The results
obtained for the simplified `$s-\alpha$' equilibrium
with GKW are indicated by the red dashed line.} 
\label{geom-bench}
\end{center}
\end{figure}

The effects of impurities have been benchmarked against GS2 using again the Waltz standard test case with
`$s-\alpha$' geometry $R/L_{Ti} = R/ L_{Te} = 9$, $R/ L_n = 3$, $q = 2$, $\hat s = 1$, $\epsilon = 0.166$, $T_e/T_i = 1$. Collisions were not included in this test. The left panel of Fig. \ref{impurities} shows the ITG-TEM part of the linear electro-static growth rate spectra for 4 different lithium impurity concentration. The increasing impurity density was balanced by lowering the amount of deuterium in order the preserve quasi-neutrality. This means that the increasing $n_{Li}$ leads to larger values of $Z_{\mathrm{eff}}$ which is accountable for the stabilisation of the TEM modes.

On the middle panel of Fig. \ref{impurities} the effect of varying the type of the impurity species can be seen. The impurity density is kept at a constant 1\% in this test. Fully ionized lithium, carbon and neon and partially ionized (charge number Z=10 for both cases) iron and tungsten ions have been used. Quasi-neutrality was maintained by adjusting the deuterium density. Here, as well, the increasing $Z_{eff}$ has a stabilizing effect on the TEM modes.

And finally, the right panel of Fig. \ref{impurities} shows the effect of the two electro-magnetic perturbations, the magnetic flutter $A_{\parallel}$ and the magnetic compression $B_{\parallel}$, obtained with a lithium impurity density of 10\%. The growth rates are significantly increased compared to the electro-static cases. Apart from the longest wavelength modes a good agreement is found between the two codes.

\begin{figure}[htb]
\begin{center}
\includegraphics[scale=0.21]{waltz-gr-es-sa-nocoll.eps}
\includegraphics[scale=0.21]{waltz-gr-es-sa-Zimp-nocoll.eps}
\includegraphics[scale=0.21]{waltz-gr-em-sa-nocoll.eps}
\caption{Benchmark of impurity treatment, growth rates of the Waltz standard test case with `$s-\alpha$' equilibrium without collisions. Line: GS2, diamonds: GKW. Left panel: effect of varying the Li impurity concentration, electro-static. Middle panel: effect of changing the impurity species, electro-static. Right panel: effect of the electro-magnetic perturations. Green line: fully electro-magnetic, both $A_{\parallel}$ (magnetic flutter) and $B_{\parallel}$ (magnetic compression) are retained. Blue line: only $A_{\parallel}$ is included.}
\label{impurities}
\end{center}
\end{figure}

The Miller geometry has been benchmarked against GS2, also for the adiabatic Waltz Standard case, with $\epsilon$ changed to 0.266.  As for the general geometry benchmark, the
different wavevector normalisations must be corrected (in this case the GS2 $k_\theta$ inputs were rescaled), after which very good agreement is found (Fig. \ref{miller}.).  The squareness and elevation have not been
benchmarked since GS2 does not have these parameters.  Some of the Miller parameters are defined differently in the two codes and require a simple conversion.

\begin{figure}[htb]
\begin{center}
\includegraphics[scale=0.65]{geom-miller.eps}
\caption{Benchmark of the Miller geometry for the growth rates of the adiabatic
Waltz standard case with $\epsilon=0.266$ and GKW $k_\theta \rho_i=0.43$. Line: GS2, Circles: GKW.  The Miller
parameters shown on the horizontal axes refer to the GKW definitions.}
\label{miller}
\end{center}
\end{figure}

\section{Nonlinear benchmarks}

\begin{figure}[thb] 
\begin{center}
\includegraphics[width=12 truecm,clip]{heatflux.eps}
\caption{GKW results of the electron heat flux expressed as a heat conduction coefficient (normalised
to $\rho_e^2  c_e / L_T$ with $c_e = \sqrt{T_e/m_e}$) as a function of normalised time ($t_{NN} = tc_e/L_T$)
for the Nevins ETG benchmark}
\label{cyclone-nl} 
\end{center}
\end{figure}

For the nonlinear solution there are two well established benchmark cases for which several codes have 
been compared: the cyclone base case, and the Nevins ETG benchmark. 
GKW has been benchmarked against the European gyrokinetic codes for the Cyclone case 
in Ref.~\cite{Euro-bench}.

Here we  describe the alternative ETG benchmark of Nevins \cite{NEV08}. The 
parameters for this case are the same as those of the Cyclone base case ($R/L_T = 6.9$, 
$R/L_n = 2.2$, $q = 1.4$, $T_e / T_i = 1$, $\epsilon = 0.18$) with only the magnetic shear 
changed to $\hat s = 0.1$. For this case the electron dynamics is simulated while the 
ions are assumed adiabatic. The latter response does not include the flux surface average 
of the electro-static potential, i.e. the response is proportional to $n_i \phi / T_i$.  
To obtain the same results as published in Ref.~\cite{NEV08} we have used (and must use)
the same number of bi-normal modes $N_{\rm mod} = 8$, and mode spacing $(k_\perp \rho_s)_{\rm max} 
= 0.69$. The radial grid spacing is $\Delta (k_r \rho_s) =0.0619$ with $N_x = 41$ (both 
positive and negative wave vectors counted).  Finally, $N_\mu = 8$, $N_{v_\parallel} = 
16$, and $N_s = 16$.

\begin{figure}[thb] 
\begin{center}
\includegraphics[width=12truecm]{shear-scan3.eps}
\caption{GKW results of the electron heat conduction coefficient for different values
of the magnetic shear}
\label{cyclone-nl2} 
\end{center}
\end{figure}

Figure~\ref{cyclone-nl} shows the normalised electron heat flux as a function of the normalised
time. Averaging over the time interval ($1000 < t c_e / L_T < 19500$), 
GKW predicts a heat diffusion coefficient $\chi_e = 3.08 \pm 0.19$ in good agreement with the 
value $\chi_e = 2.95\pm 0.15$ reported in Ref. \cite{NEV08}. The ETG case is 
sensitive to the magnetic shear. For values $\hat s > 0.3$ no convergence is found in the simulations 
reported in Ref. \cite{NEV08}. GKW reproduces this result as shown in Fig. \ref{cyclone-nl2}.

\begin{figure}[htb] 
\begin{center}
\subfigure[Benchmark with GYRO]{\includegraphics[height=145pt]{exb-benchmark}}
\subfigure[Benchmark with GS2]{\includegraphics[height=150pt]{GKW-GS2-exb}}
\caption{Left: GKW benchmark of background ExB shearing with GYRO for Waltz standard case parameters. GKW results (diamonds) for
 adiabatic electrons, kinetic electrons, and adiabatic with $\hat s = -0.5$, are compared to the equivalent GYRO results (circles) from Tables II, III, IV and V  of Ref. \cite{KIN05}. The dashed lines include coupled
parallel velocity shear for purely toroidal rotation with $u^\prime=12\gamma_E$.  Right: GKW benchmark of background ExB shearing with GS2 results reported in Ref. \cite{ROA09} for Cyclone case parameters.  Both codes were run including coupled parallel velocity shear for purely toroidal rotation.  $\gamma_{\rm max}$ is the maximum linear growth rate at $\gamma_E=0$.}
\label{exb-gyro} 
\end{center}
\end{figure}

The background ExB shear stabilisation 
has been successfully benchmarked against GYRO results as reported in Ref. \cite{CAS09}.  This result for the Waltz standard case 
is included again here for convenience in Fig. \ref{exb-gyro}.  GKW has also been checked against 
the GS2 results reported in Ref. \cite{ROA09} for the Cyclone case, also shown in Fig. \ref{exb-gyro}.  Both of these comparisions were against existing results, i.e. no further attempt to examine the convergence of the GYRO and GS2 results has been made.  Detailed convergence testing for the GKW results is described in Refs. \cite{Casson-Thesis} and \cite{exb-dissip}.  This study revealed that it is preferable to use some radial hyperdissipation and minimal parallel dissipation at high shear rates (see also Sec. \ref{sec.dissip}.) (not the case for these figures) which can improve the agreement at high shear rates to better than that shown in Fig. \ref{exb-gyro}.

Towards the edge, nonlinear and electromagnetic physics can be significantly more complex than
the ITG dominated Cyclone and GA-STD cases,
and benchmarking in this area is an active area of research.  
For L-mode ASDEX-Upgrade plasmas, GKW has been sucessfully benchmarked with GENE 
for four nonlinear electromagnetic cases using full geometry, and full collisions.  The results 
are reported in in Refs. \cite{FAB13} and \cite{TOL13}.  
Benchmarking for the the DIII-D `shortfall' case is in progress and 
is reported in ITPA presentations of C. Holland and Y. Camenen.

\subsection{Global Benchmarks}

Can be found in the Ringberg 2012 presentation of A.G Peeters on the GKW webpages under Talks..

\vskip 0.2 truecm

\noindent{\bf Acknowledgements} The authors want to thank W. Dorland and M. Kotschenreuther 
for making the GS2 code available, as well as A. Marinoni for his contribution to the geometry benchmark.
One of the authors (AGP) would also like to thank Bruce D. Scott for 
his tutorials and insights into the subject which he has been able to follow over many years.

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/bibliography.tex
(1) Added gkw Clone gwk  #3

\begin{thebibliography}{10}
%Appear in preface, in this order, copy needs to match.
\bibitem{CPC-paper} A.G. Peeters, Y. Camenen, F.J. Casson, W.A. Hornsby, A.P. Snodin, D. Strintzi, G. Szepesi,
Computer Physics Communications, {\bf 180}, 2650, (2009). doi: 10.1016/j.cpc.2009.07.001.
\bibitem{PEE04} A.G. Peeters, D. Strintzi, Phys. Plasmas {\bf 11} (2004) 3748 
\bibitem{PEE07} A.G. Peeters, C. Angioni, D. Strintzi, Phys. Rev. Lett. {\bf 98} (2007) 265003

% Particle code references 
\bibitem{PAR93} S.E. Parker, W.W. Lee, R.A. Sandtoro, Phys. Rev. Lett. {\bf 71} (1993) 2042 
\bibitem{LIN98} Z. Lin, T.S. Hahm, W.W. Lee, et al., Science {\bf 281} (1998) 1835
\bibitem{IDO03} Y. Idomura, S. Tokuda, Y. Kishimoto, Nucl. Fusion {\bf 43} (2003) 234  
\bibitem{HEI04} J.A. Heikkinen, J.S. Janhunen, T.P. Kiviniemi, et al., Contr. Plas. Phys. {\bf 44} (2004) 13
\bibitem{BOT07} A. Bottino, A.G. Peeters, R. Hatzky, et al., Phys. Plasmas {\bf 14} (2007) 010701 
\bibitem{JOL07} S. Jolliet, A. Bottino, P. Angelino, et al., Comp. Phys. Comm. {\bf 177} (2007) 409
\bibitem{GRA06} V. Grandgirard, M. Brunetti, P. Bertrand, et al., J. Comp. Physics {\bf 217} (2006) 395

% Vlasov references 
\bibitem{KOT95} M. Kotschenreuther, G. Rewoldt, W.M. Tang., Comp. Phys. Comm. {\bf 88} (1995) 128
\bibitem{DOR00} W. Dorland, F. Jenko, M. Kotschenreuther, et al., Phys. Rev. Lett. {\bf 85} (2000) 5579 
\bibitem{JEN00} F. Jenko, Comp. Phys. Comm. {\bf 125} (2000) 196  
\bibitem{DAN05} T. Dannert, F. Jenko, Phys Plasmas {\bf 12} (2005) 072309 
\bibitem{CAN03} J. Candy, R.E. Waltz, J. Comp. Phys. {\bf 186} (2003) 545 
\bibitem{SCO06} B.D. Scott {\sl delta-FEFI}

% Papers on toroidal momentum transport. 
\bibitem{PEE05} A.G. Peeters, C. Angioni, Phys. Plasmas {\bf 12} (2005) 072515   
\bibitem{PEE06} A.G. Peeters, C. Angioni, A. Bottino, et al. Plasma Phys. Contr. Fusion {\bf 48} (2006) B413 
\bibitem{PEE09com} A.G. Peeters, C. Angioni, D. Strintzi, Phys. Plasmas {\bf 16} (2009) 034703
\bibitem{PEE09} A.G. Peeters, D. Strintzi, C. Angioni, et al., Phys. Plasmas {\bf 16} (2009) 042310 (Note also the Erratum to this paper) 
\bibitem{CAM09} Y. Camenen, A.G. Peeters, C. Angioni, Phys. Rev. Lett. {\bf 102} (2009) 125001  
\bibitem{PEE09scmode} A.G. Peeters, C.Angioni, Y. Camenen, et al., Phys. Plasmas. {\bf 16} (2009) 062311
\bibitem{CAM09pop} Y. Camenen, A.G. Peeters, C. Angioni, et al., Phys. Plasmas {\bf 16} (2009) 062501  
\bibitem{CAS09} F.J. Casson, A.G. Peeters, Y. Camenen, et al., Phys. Plasmas {\bf 16} (2009) 092303
\bibitem{CAM09rot} Y. Camenen, A.G. Peeters, C. Angioni, et al., Phys. Plasmas {\bf 16} (2009) 012503
\bibitem{CAS10} F.J. Casson, A.G. Peeters, C. Angioni, et al., Phys. Plasmas {\bf 17} (2010) 102305 
\bibitem{HOR10} W.A. Hornsby, A.G. Peeters, A.P. Snodin, et. al., Phys. Plasmas {\bf 17} (2010) 092301 
\bibitem{HOR10epl} W.A. Hornsby, A.G. Peeters, E. Poli, et al., Euro. Phys. Let. {\bf 91} (2010) 45001

% References on gyrokinetic theory 
\bibitem{LIT83} R.G. Littlejohn, J. Plasma Phys. {\bf 29} (1983) 111
\bibitem{HAH88} T.S. Hahm, Phys. Fluids {\bf 31} (1988) 043207 
\bibitem{BRI88} A. Brizard, Phys. Plasmas {\bf 41} (1988) 541 
\bibitem{SUG00} H. Sugama, Phys. Plasmas {\bf 7} (2000) 466
\bibitem{BRI08} A. Brizard, Rev. Modern Phys.{\bf 79} (2007) 421 
\bibitem{BEER} M. Beer, PhD thesis, Princeton University 1995

% Coriolis drift 
\bibitem{BRI95} A.~J. Brizard, Phys. Plasmas {\bf 2} (1995) 459

% Geometry references 
\bibitem{HAM62} S. Hamada, Kakuyugo Kenkyu {\bf 1} (1958) 542
\bibitem{MIL98} R.~L. Miller, M.~S. Chu, J.~M. Greene, Y.~R. Lin-Liu, R.~E. Waltz, Phys. Plasmas {\bf 5} (1998) 973
\bibitem{CAN09} J.Candy, Plasma Phys. Control. Fusion {\bf 51} (2009) 105009
\bibitem{THYA06} A.Thyagaraja, K.G.McClements, Phys. Plasmas 13, 062502 (2006)
\bibitem{SCO98} B.~D. Scott, Phys Plasmas {\bf 5} (1998) 2334 
\bibitem{SCO01} B.~D. Scott, Phys Plasmas {\bf 8} (2001) 447 
\bibitem{LUT96} H. L\"utjens, A. Bondeson and O. Sauter, Comput. Phys. Commun.{\bf 97} (1996) 219 
\bibitem{CON} J.~W. Connor,  R .~J. Hastie and J.~B. Taylor, Phys. Rev. Let., {\bf 40} (1978) 396 
\bibitem{Lapillonne}  X. Lapillonne, S. Brunner, T. Dannert, S. Jolliet, A. Marinoni, L. Villard, T. Goerler, F. Jenko, and F. Merz,  Phys. Plasmas, {\bf 16}, 032308, (2009). 
\bibitem{dorland-note} W. Dorland, M. Kotschenreuther, \textit{Notes on Local
Equilibrium Implementation}\\
available from \href{http://w3.pppl.gov/~hammett/work/gs2/docs/geometry/g_short.pdf}{http://w3.pppl.gov/~hammett/work/gs2/docs/geometry/g_short.pdf}

% Collision operator  
\bibitem{KAR86} C.~F.~F. Karney, Comp. Phys. Reports {\bf 4} (1986) 183

% Convergence study
\bibitem{VET10} C. Veth, C. Angioni, 2011, \textit{Report on grid convergence with GKW}\\
available from \href{http://www.gkw.org.uk/tikiwiki/Manual}{http://www.gkw.org.uk/tikiwiki/Manual}

% Nonlinear benchmarks 
\bibitem{DIM00} A.M. Dimits, Phys. Plasmas {\bf 7} (2000) 969  
\bibitem{Euro-bench}
G.~L. Falchetto, B.~D Scott, P. Angelino, A. Bottino, T. Dannert, V. Grandgirard,
  S. Janhunen, F. Jenko, S. Jolliet, A. Kendl, B.~F. McMillan, V. Naulin, A.~H. Nielsen,
  M. Ottaviani, A.~G. Peeters, M.~J. Pueschel, D. Reiser, T.~T. Ribeiro, and
  M. Romanelli, Plasma Physics and Controlled Fusion {\bf 50} (2008) 124015 
\bibitem{ROA09} C.~M. Roach, et. Al. Plasma Phys. Control. Fusion {\bf 51}, 124020, (2009).
\bibitem{Casson-Thesis} F.~J. Casson, PhD Thesis, University of Warwick, 2011. \href{http://wrap.warwick.ac.uk/36765/}{http://wrap.warwick.ac.uk/36765/}
\bibitem{exb-dissip} F.~J. Casson, 2011, \textit{GKW benchmark of background ExB shear with GYRO}\\
 available from \href{http://www.gkw.org.uk/tikiwiki/Manual}{http://www.gkw.org.uk/tikiwiki/Manual}
\bibitem{KIN05} J. E. Kinsey, R. E. Waltz, and J. Candy, Phys. Plasmas {\bf 12}, 062302 (2005).
\bibitem{FAB13} Fable, E., C. Angioni, F. J. Casson, D. Told, et al., Plasma Physics and Controlled Fusion {\bf 55}, 124028 (2013)
\bibitem{TOL13} D. Told, F. Jenko, T. Goerler, F. J. Casson, E. Fable, and ASDEX Upgrade Team., Phys. Plasmas {\bf 20}, 122312 (2013)

% FFTW3
\bibitem{FFT05} M. Frigo, S.~G. Johnson, Proc. IEEE {\bf 93} (2005) 216

% slpec manual
\bibitem{SLE12} C. Campos, J.~E. Roman, E. Romero and A. Tomas, SLEPc Users Manual {\bf DSIC-II/24/02 - Revision 3.3},
  D. Sistemes Inform\`atics i Computaci\'o, Universitat Polit\`ecnica de Val\`encia, 2012.
  \href{http://www.grycap.upv.es/slepc/documentation/manual.htm}{http://www.grycap.upv.es/slepc/documentation/manual.htm}

% E x B shear method
\bibitem{Hahm95}
T.~S. Hahm and K.~H. Burrell.
\newblock {\em Phys. Plasmas}, {\bf 2} (1995) 1648

\bibitem{Burrell97}
K.~H. Burrell.
\newblock Phys. Plasmas, {\bf 4} (1997) 1499

\bibitem{Waltz94}
R.~E. Waltz, G.~D. Kerbel, and J.~Milovich, Phys. Plasmas {\bf 1} (1994) 2229

\bibitem{Miller94}
R.~L. Miller and R.~E. Waltz, Phys. Plasmas {\bf 1} (1994) 2835

\bibitem{Baron82}
F.~Baron. {PhD} in physics, (Univ. Pierre et Marie Curie, Paris, France, 1982)

\bibitem{Schumann85}
U.~{Schumann}, {Algorithms for direct numerical simulation of shear-periodic
  turbulence}, in {\em Lecture Notes in
  Physics {\bf 218}} (Berlin Springer Verlag 1985) pages 492

\bibitem{Gerz89}
T.~Gerz, U.~Schumann, and S.~E. Elghobashi, Journal of Fluid Mechanics Digital Archive {\bf 200} 
(1989) 563

\bibitem{Rogallo81}
R.~S. {Rogallo}, {\em Numerical experiments in homogeneous turbulence}.
\newblock {\em NASA STI/Recon Technical Report N} (1981) 81:31508

\bibitem{Zang88}
T.~A. {Zang}, Applied Numerical Mathematics {\bf 7} (1991) 27

\bibitem{Pumir96} Alain Pumir, Phys. Fluids, {\bf 8} (1996) 3112

\bibitem{HammettAPS06}
G.~W. {Hammett}, W.~{Dorland}, N.~F. {Loureiro}, and T.~{Tatsuno}.
\newblock {Implementation of Large Scale E x B Shear Flow in the GS2
  Gyrokinetic Turbulence Code}.
\newblock {\em APS Meeting Abstracts} (2006) pages 1136

\bibitem{JEN05} F. Jenko, T. Dannert, C. Angioni, Plasma Phys. Contr. Fusion {\bf 47} (2005) B195
\bibitem{LEE01} W.W. Lee, J.L.V. Lewandowski, T.S. Hahm, and Z. Lin, Phys. Plasmas {\bf 8}, 4435 (2001)

\bibitem{MER10}  F. Merz and F. Jenko, Nucl. Fusion {\bf 50}, 054005 (2010)

\bibitem{HINROS} F.~L. Hinton, M.~N. Rosenbluth, Phys. Rev. Lett. {\bf 80} (1998) 724

\bibitem{XIA06} Y. Xiao, P.~J. Catto, Phys. Plasmas {\bf 13} (2006) 102311

\bibitem{PEE05b} A.~G. Peeters, C. Angioni, M. Apostoliceanu et al., Phys. Plasma {\bf  12} (2005) 022505

\bibitem{NEV08} W. Nevins, et al., Phys. Plasmas {\bf 13} (2006) 122306

\bibitem{A66} A. Arakawa, J. Comp. Phys. {\bf 1} (1966) 119 
%(reprinted J. Comp. Phys. (1997) 135:103-14).

% Citations in the section on the entropy diagnostic
\bibitem{DUN07} J.Dunkel, P.Talkner, B. H\"anggi, New Journal of Physics {\bf 9} (2007) 114

% ICRH minority poloidal asymmetry
\bibitem{Reinke12} M. L. Reinke et al, Plasma Phys. Contr. Fusion {\bf 54} (2012) 045004
\bibitem{Bilato14} R. Bilato, O. Maj, C. Angioni, Nucl. Fusion (letter), {\bf 56} (2014)

% The rho* contributions to the parallel derivatives 
\bibitem{SUN13} T. Sung, R. Buchholz, F.J. Casson, E. Fable, S.R. Grosshauser, W.A. Hornsby, P. Migliano and A.G. Peeters, 
Phys. Plasmas {\bf 20}, 042506 (2013)

%Neoclassical corrections
\bibitem{NEO1} E A Belli and J Candy {\it Plasma Phys. Control. Fusion} {\bf 51} 075018  (2009)

\bibitem{NEO2} E A Belli and J Candy {\it Plasma Phys. Control. Fusion} {\bf 54} 015015 (2012)

\bibitem{Cyclone} C.M. Greenfield {\it et al} 1997 {\it Nuclear Fusion} {\bf 37} 1215

\bibitem{BarnesParra13} M. Barnes {\it et al} {\it Phys. Rev. Lett.} {\bf 111} 055005 (2013)

\end{thebibliography}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/buildandrun.tex
(1) Added gkw Clone gwk  #3

\chapter{Obtaining, building and running the code} \label{usage}

\section{Getting the code}

To obtain the latest version of the code (which should approximately
correspond to the latest version of this documentation), you should use the
git revision control system (\href{http://git-scm.com/.}
{http://git-scm.com/.}). The git program is installed on many
systems as the \name{git} command. Assuming you are using \name{git}, you can
check out the code with the command
%git clone http://gkw@bitbucket.org/gkw/gkw.git gkw-read-only
\begin{verbatim}
  git clone git@bitbucket.org:gkw/gkw.git gkw
\end{verbatim}
To push changes, you will need a bitbucket account to be given write permissions.

% Not yet sure how this will be relvant to bitbucket:
%[Some advice on which tagged versions to use, or how to download without git.]%

% This clone does not allow changes to be written back to the repository.
% Instructions for developers with commit permissions 
% can be found at \href{http://code.google.com/p/gkw/source/checkout}{
% http://code.google.com/p/gkw/source/checkout}.

% Alternatively, a reasonably stable version of the code can be obtained as a
% tarball from the downloads section
% (\href{http://code.google.com/p/gkw/downloads/list}
% {http://code.google.com/p/gkw/downloads/list}).
% At present, the version of the code found there is based on the CPC release
% (associated with the CPC paper \cite{CPC-paper}). The tarball is not often
% updated and will not have the latest code features.

\section{Building the source code}

% If using a version obtained in a tarball, it is suggested to look at the
% \File{README} file found within for specific compiling instructions. The
% guidance of the following sections pertains primarily to the latest version of
% the code obtained via Subversion.

There are two methods provided to build the code, involving various
makefiles. The use of the `simple' makefile, \File{src/makefile}, is intended
only as a fallback if the standard makefiles, requiring \name{GNU make},
cannot be used.  The standard GNU makefiles
allow one to create build settings on a per user/machine/compiler basis,
without having to make any modifications to the original makefiles\footnote{
The standard makefiles also result in executables which can
provide information about themselves and how they were produced. For example,
the executable will be able to report its revision number in various ways,
which can be quite useful to know.}.
Both methods are outlined below.

With the latest version of the source code, it will often be possible to
build the code by doing very little other than running \name{make} in
the top-level directory containing the \File{GNUmakefile} file.
Although this may appear successful, one should be aware that the resulting
executable may not contain all the desired functionality.

\subsection{Prerequisites and suggestions}

A Fortran 95 compiler is essential for building the code. Most Fortran
compilers are able to pre-process the source files containing C
pre-processor directives\footnote{In the unlikely event that this is not the case,
and an appropriate pre-processor is not available, some editing of the
source files may be necessary before compilation.}.

A minimum of a Unix-like system, together with the POSIX shell and standard
\name{make} utility is suggested.\footnote{Without these basics, the various source files will have to be
compiled in the correct order (or concatenated into a single file in the the
correct compilation order beforehand).}
%We are typically developing the code under
%GNU/Linux, which meets this minimum requirement -- as do many other operating
%systems.

For building the latest version of the code from the git repository,
\href{http://www.gnu.org./software/make/}{\name{GNU make}} is recommended.
This should be the default
\name{make} program installed on GNU operating systems, such as GNU/Linux.
On some systems, the command may be \name{gmake}, in order to distinguish it
from the system default \name{make}.

To enable parallelisation, you will require an MPI implementation and/or a
Fortran compiler which supports OpenMP directives.  To allow nonlinear runs, 
the FFTW3 library is required (more details in \Sec{thecode}).

A C compiler is {\it not} required to compile the code, but will be required
to compile the optional libraries includes with GKW.

More information on linking optional libraries is given in 
section \ref{sec:additional-libraries}.

\subsection{Using the GNU makefile}

Instructions within the top-level makefile, \File{GNUmakefile}, should be the
first point of reference for this method. Ultimately, you should just need to
run \name{make} (GNU make) in this directory and you will end up with an
executable in the \File{run/} directory. However, when building for the first
time on a given machine, it is suggested to create at least one new makefile
which contains settings specific to the local machine and user (and perhaps
compiler if there are multiple possibilities). Whenever the code is then built
again on this machine, those specific makefiles will be used
automatically when running make.

All host-specific makefiles are located in the \File{config/} directory and
subdirectories. The variables contained in \File{global_defaults.mk} will
always be used, unless host-specific variables override them. You can use
\File{template.mk}, which documents almost all the variables one may wish
to change. If you are creating a new file for machine \name{hostname}, it
should usually be saved to \File{defaults.mk} within the \File{hostname/}
subdirectory of \File{config/}. You will likely want to set at least the
Fortran compiler, external libraries, and the required precision. One may
wish to look at existing files for examples; template files for Intel, GNU
and IBM compilers are also provided in \File{config/templates}.

The code object files are built in different subdirectories of \File{/obj}, 
depending on hostname, compiler and precision, among other things. The
resulting executable name will represent which git
version of the code was used, which compiler was used, and if the
executable is single (SP) or double precision (DP)\footnote{Beyond these 
distinctions, it is possible to clobber an
existing executable with one of the same name that does something different.
%(this shortcoming may change in the future).
}. 
If you run \texttt{make clean}, ALL object files will removed.
\texttt{make cleanloc} will remove only object files for the compiler and
 precision subfolder\footnote{Note that if you pass variables to
the make program by specifying them after the command (rather than
changing them in the appropriate file), you should be aware that existing
objects might be re-used to complete the build, such that the desired
properties of the program are only partly present. For example, say you
compile the code without OpenMP, then edit some files, then type
\texttt{make SMP=OPENMP}, then it might be the case that only some of the
code has OpenMP enabled in it. To avoid such problems, use the configuration
files, or type \texttt{make clean}, before
 (re)building to make sure you get what you expect.}. 
If your make program supports it, there should be no problem in
building the code in parallel (GNU make does this via the `-j' flag).

If everything completes correctly, the compiled executable will appear in the 
\File{/run} directory.

%\pagebreak

\subsection{Using the `simple' makefile}

If using the GNU makefiles is problematic, one may wish to build the code
using the simple makefile, \File{src/makefile}. You will need to run
\name{make} from within the \File{src/} directory, but there is a good chance
you will have to edit the makefile beforehand -- instructions for doing this
can be found in the file itself. Any files created during compilation will be
in this directory, together with the source files, apart from the resulting
executable, which by default, will be found in the top-level
directory.

\subsection{Some important makefiles}

There are two files common to both methods which are found in the \File{src/}
directory. The \File{objfiles.mk} file contains the list of object files to be
created before linking to produce the executable. It should only be necessary
to modify this file when adding new source files to the code.
The file \File{deps.mk}
lists the dependencies for the objects found in \File{objfiles.mk}, so that
the objects/modules may be built in the correct order. This should be updated
automatically with the standard makefiles, but may need to be modified manually
when using the `simple' makefile. It {\it may} be possible to update it via

\begin{verbatim}
  ./scripts/mkdeps > deps.mk
\end{verbatim}

from within the source directory.

The file \File{src/gkw.mk} is the main GNU makefile which should not
need to be modified in general, unless new variables are being introduced
to the code. Otherwise, all settings should be made via the files in
\File{config/}.

\subsection{Potential compilation issues}

Most compilers tested to date have no problems compiling the code.
Occasionally, some non-standard features or bugs get into the trunk version, creating
problems. However, these tend to get removed rapidly after they are discovered.
The Intel and GNU compilers are in most frequent use and are least likely to find unexpected new
issues when compiling and running.   On Hector, the PGI compiler has historically caused less problems than the Cray compiler.

Some compilers, such as those found on Fujitsu machines, may not like very
large source files; if such a compiler is used, it may be necessary to edit
the problematic source file and remove code which is not needed for the
particular run. %It is hoped that certain large source code files will be
%significantly broken up in the near future, thus preventing this issue.

Some compilers (e.g. Intel) do not by default include the present working directory
in the include path.  If \code{make} gives an error that \File{gkw_info.h} cannot be found,
when it already exists in the \File{/obj} directory, you should first try adding \code{I./} to 
\code{FFLAGS_INC} to your \File{config} file.

% ----------------------------------------------------------------------------
% ----------------------------------------------------------------------------

\subsection{\label{sec:additional-libraries}Additional libraries}

Here we list some third party libraries that can be used to extend the
functionality of the code.

\subsubsection{MPI-2}

An MPI-2 implementation is required for any distributed memory
parallel run.\footnote{The code might compile and run (with
  limited MPI-I/O and nonspectral functionality) with MPI-1 but this
  has not been attempted for a long time, and it is likely that some
  MPI-2 function calls would need to be commented to make this
  possible.}  MPI is enabled through preprocessing by setting
\code{MPI=mpi2} in the \File{config} file.  For most MPI
implementations, the include files and linking are automatically
handled by a compiler wrapper script such as \code{mpif90} or
\code{mpiifort}.

% ----------------------------------------------------------------------------

\subsubsection{FFTW3}

For nonlinear runs a Fourier transform library is
needed. At the moment only the FFTW3 library is supported.
If this library is installed it can be linked against by setting
\code{FFTLIB = FFT_FFTW3} in the corresponding \File{config} file of gkw.
Depending on the system you may also need to set \code{FFLAGS_INC} and \code{LDFLAGS}
to link to \code{-lfftw3 (-lfftw3f)} for double (single) precision.

% ----------------------------------------------------------------------------

\subsubsection{AMD and UMFPACK}
The two libraries \File{amd} and \File{umfpack} are needed for the (experimental) 
implicit time integration scheme and the nonspectral and global cases.
They are provided with the GKW trunk and must be compiled before GKW.

At present these libraries can only be compiled in double precision. 
If GKW is compiled in single precision, then type conversions are made 
wherever these libraries are used in the code.  
To compile and link these libraries to GKW one must add to the GKW config file
\begin{verbatim}
IMPLICIT = umfpack      # Compile GKW with umfpack (64 bit) or umfpack32 (32 bit)
\end{verbatim}
In most cases this should be sufficient; the libraries should then be built and linked automatically.   
One can also explicitly require compilation and linking of umfpack by setting
\begin{verbatim}
UMFPACK = compile       # Compile the UMFpack provided with GKW
LD_UMFPACK = -lumfpack  # link GKW with UMFpack
\end{verbatim}

% ----------------------------------------------------------------------------

\subsubsection{\label{subsubsec:slepc}SLEPc and PETSc}

For the eigenvalue solver the two libraries {\sc slep}c and {\sc pets}c are required
(where the former depends on the latter).
Note that these two libraries have to be compiled for using complex numbers
to work. This can be achieved with the option
\code{--with-scalar-type=complex} for the configure script of {\sc pets}c. {\sc slep}c
inherits these settings automatically from {\sc pets}c (if configured afterwards).
To link these libraries, one must add the paths for the libraries to the \code{LDFLAGS} together
with some subset of these (-lslepc -lpetsc -llapack -lblas -lX11 -lmkl) and
you must set \code{SLEPC = HAVE_SLEPC} in your \File{config} file for GKW.  
If the headers are not installed in a standard location for Fortran
you may also have to add the library include paths to \code{FFLAGS_INC}.  
It might also be necessary to add the path to petscconf.h to \code{FFLAGS_INC} 
(probably under an \File{include} subdirectory for each architecture).

% ----------------------------------------------------------------------------

\subsubsection{HDF5}\label{sec:hdf5}

The HDF5 library is needed to write output in the HDF5 data format.
Note that at the moment GKW expects the HDF5 library to be compiled in
its \emph{serial} version.\\
To link this library, one must add the paths ot the \code{LDFLAGS} and
\code{FFLAGS_INC} variables. It is a good idea to make use of the
\code{h5fc} wrapper which is provided by the HDF5 library to make
compilation easier (e.g. see the helios config).  Furthermore,
set \code{IO_LIB = HAVE_HDF5} in your \File{config} file for GKW, this
will then enable GKW's HDF5 support.

Note that the default output format of GKW depends on the availability
of HDF5. If GKW was compiled without HDF5 it then \code{io_format='mixed'} is the default, but if
compiled with HDF5 the default becomes \code{io_format = 'hdf5+ascii'},
see also section \ref{sec.diagnos-data-format}.

Note also that some datasets, in particular higher-dimensional (3D+)
diagnostic data (using MPI-IO) is not output in HDF5 format at the
moment, because parallel HDF5 is not yet supported by GKW.  If such
data is requested, it will be output in fortran binary output instead of HDF5.

% ----------------------------------------------------------------------------
% ----------------------------------------------------------------------------

\section{Running GKW}

Before you do any run it is important to realize that it is not all that hard to produce nonsense. 
The code is not under all conditions stable, and it is not always immediately obvious if a 
result is wrong. It is therefore important to diagnose your runs to make sure they are O.K.. 
It is your responsibility as a user to produce physically meaningful results.

Here it is assumed you have the necessary MPI environment set up and
running on the machine (consult your MPI library documentation) and that any run time
dependencies can be satisfied.

In order to run GKW, the code requires a single input file \File{input.dat} to be present
in the run directory. Optionally, a file containing equilibrium geometry information from
the CHEASE code can be provided as input.
To get started, there are a few example input files found in the \doc{input} directory,
together with corresponding \File{README} files. The file \doc{input.dat.sample} lists every input switch
together with a description of its purpose. \doc{input.dat.minimum} is a bare minimum input file containing
the essential inputs required for the code to run.

To run the code in parallel, you will probably need to use the program provided by the
MPI implementation; this is usually called \name{mpirun} or \name{mpiexec}.

GKW will try to select an appropriate parallelisation given the gridsize input but this can be overridden 
(by setting \name{N_procs_[sp|mu|vpar|s]} in the gridsize namelist).  Nevertheless it is
sensible to have in mind a parallelisation setup when you set the gridsize, as the user controls the number of 
processors the code will be run on and should choose a number that fits with the
gridsize.  For the largest parallel runs, the user is recommended to manually choose the parallel setup.

The rest of this section describes a basic input file.
Many of the input switches are optional and default values will be used if they are not provided.
The input variables are provided in Fortran namelists, for example the namelist `control' contains
the main switches for the code. The namelist examples presented here provide you with a basic set 
needed to run the code.  The namelists are sensitive to bad formatting, in which case the code should 
inform you which namelist has the problem.

GKW tries to catch as many invalid input combinations as possible. Most of the time if you input a set of 
incompatible switches the code will tell you and abort if necessary. For more minor errors the code will generate 
a warning that it is ignoring an input parameter and continue.  The actual input parameters the code runs with are 
written to \File{input.out}.  By renaming \File{input.out} to \File{input.dat} the code should re-run without the warnings
(if the compiler correctly formats namelist outputs).

\subsection{Modes}

The spacing and location of the wavevectors is determined by the `mode' namelist.
For \name{mode_box=.false.} there is only one radial wavevector considered, but the field line can be extended to 
take more than one poloidal turn by setting \name{nperiod}. The total number of full poloidal turns is then 2 \name{nperiod} - 1. 
The bi-normal modes are determined by the input \name{kthrho} which specifies the value(s) of
$k_\theta\rho_{\rm ref}=\sqrt{g^{\zeta\zeta}(s=0)}k_\zeta$, see Eq.~(\ref{kperp}). For {mode_box=.false.} the bi-normal modes are given as a list i.e.
 \name{kthrho=0.1,0.2,0.3} (the variable \name{krhomax} is ignored).
The radial mode is determined by the input \name{krrho} which specifies the value of $k_r\rho_\textrm{ref}=\sqrt{g^{\psi\psi}(s=0)}k_\psi$.
Alternatively, a poloidal shift of the balooning angle can be specified using the input \name{CHIN}. A finite value of $k_\psi$ will then be
calculated to minimise $k_\perp\rho_{\rm ref}$ at the poloidal angle $\theta = \name{CHIN}$. The value of $k_\psi$ is given by
$$k_\psi = - k_\zeta g^{\psi\zeta}/g^{\psi\psi}$$
where the metric elements are evaluated at the poloidal location specified by CHIN (poloidal angle, from the magnetic axis, positive counterclockwise, in radian and
zero at the low field side midplane) and $k_\zeta=k_\theta\rho_{\rm ref} / \sqrt{g^{\zeta\zeta}(s=0)}$, as usual.

For \name{mode_box=.true.}, the field line must have one poloidal turn, i.e. one must set \name{nperiod=1}. In this case 
the radial modes are coupled at the end of the field line according to the ballooning shift connected with the
magnetic shear.  
The bi-normal modes will be equally spaced from $k_\theta\rho_{\rm ref} =0$ to \name{krhomax} (the \name{kthrho} input is ignored).  
The integer $i_k$ in Eq.~(\ref{ikxspace}) is input into the code as \name{ikxspace} and determines the balance between radial 
resolution and radial box size.  \name{ikspace} should be adjusted to give an approximately square box (the code will output LX and LY).
\begin{table}[h!]
\begin{center}
\begin{tabular}{c|c|c|c|c}
 NS & ikxspace & NX & NPERIOD & mode_box \\
\hline
 80      &    N/A   &  1  &    3    & false    \\
 16      &     1    &  5  &    1    & true     \\ 
\hline
 $n_s.(2n_p-1)$ &    N/A   &  1  &    $n_p$    & false    \\
 $n_s$    &     1    & $2n_p-1$ &  1   & true   \\

\end{tabular}
\caption{Equivalent formulations example with \name{mode_box} true and false}
\end{center}
\end{table}\\

\subsection{Dissipation \label{sec.dissip}}

The parameter \name{disp_vp} sets the dissipation on the trapping term. 
It acts a bit like a collisionality but the normalization is different: The dissipation decreases at fixed \name{disp_vp}
when \name{N_vpar_grid} is increased, so that convergence should always be reached.  For all collisionless problems
that depend critically on the trapping (TEM modes, particle fluxes) one should not set the dissipation in velocity space too high.
For these cases a lower \name{disp_vp} (say 0.01) will allow a lower resolution convergence of the particle fluxes.  
When collisions are included, \name{disp_vp} can and should be small (0.2).

The parameter \name{disp_par} sets the dissipation on the streaming term, and 0.2 is a reasonable value for most linear runs.
For nonlinear runs, values in the range 0.5 - 2.0 are typically required for numerical stability, 
although stable results can be achieved with lower dissipation by using the Arakawa scheme.
If parallel dynamics must be finely resolved, higher \name{N_s_grid} bay also allow lower \name{disp_par}.
The default form of the parallel dissipation uses \name{idisp = 2}, which does not vary over the velocity grid
and is usually more stable for electromagnetic runs .  However, 
in some physical cases with a sharp parallel structure (usually at low $k_y$),
excessive dissipation appears to drive a linear numerical instability;
in these cases the stability may be improved by using  \name{idisp = 1}. (See \issue{201}).

The circular geometry tends to be more stable than the $s-\alpha$ geometry if numerical problems arise in the zonal mode \cite{CAS10}.
For runs with strong $E \times B$ shear, \name{disp_par} should be minimised, and some radial dissipation (\name{disp_x}=1.0) should 
be used (see Ref. \cite{Casson-Thesis} or \cite{exb-dissip}).  \name{disp_x > 0.0} is also required for the finite differences of the
nonspectral version.  \name{disp_y > 0.0} can be useful in some nonlinear runs to cut out high k physical instabilities (e.g ETG modes) and 
supress high-k spectral pileup.

\subsection{Gridsizes}

A detailed GKW linear convergence study including many figures is available in Ref. \cite{VET10}, of which the numbers below are a summary.
The gridsizes required for a converged result will vary for each problem, but here we give some guidance assuming \name{disp_vp}=0.01 and circular geometry.  For adiabatic runs, $[N_s,N_\mu,N_{v_\parallel}]=[16,8,16]$ is often sufficient (Here, $N_s$ is points \textbf{per poloidal turn}, but \name{N_s_grid} in the input file is the total number.). 
Collisionless ITG modes with kinetic electrons will give growth rate and frequency converged to within 5\% for $[N_s,N_\mu,N_{v_\parallel}]=[16,8,32]$, 
but if the particle fluxes are to be examined, gridsizes of $[N_s,N_\mu,N_{v_\parallel}]=[40+,8,48]$ 
may be required (this is because the trapped passing boundary must be accurately resolved).

In general, trapped electron modes require more resolution to get accurate growth rate and frequencies, 
but once these are resolved, the particle fluxes should be also.  A TEM mode is converged to with 5\%
for $[N_s,N_\mu,N_{v_\parallel}]=[30,8,64]$  Including collisions can reduce the values of $N_{v_\parallel}$ required 
substantially ($N_{v_\parallel} = 16-32$), with the lower figures for greater collisionality.

For a linear run, \name{Nperiod} will depend on the type of mode you are looking at.  For some ITG cases with high magnetic shear, 
3 would be sufficient, but for some TEM cases or cases with lower magnetic shear, as many as 10 may be required.  Remember to adjust 
\name{N_s_grid} accordingly.

For a nonlinear run, \name{Nmod} should be at least 16.  In the past a
value of \name{nmod}=21 (or 43, 86,...) was used for particularly
efficient FFT computations, but this is no longer significant.  For
kinetic runs \name{NX}=167 is a sensible size.  For adiabatic runs
\name{NX}=83 may be sufficient.  For cases with very low magnetic
shear or high E x B shear, \name{NX}=339 is sometimes used for
convergence tests.

For $N_s$, more points are required if using complex general geometry.  
The boundaries of the velocity space do not usually need to be increased from the defaults.

\subsection{Species}
There is no default species data, hence the user must provide some. The species input must satisfy quasi-neutrality. This requires the user to set ratios of densities and charge to satisfy
\begin{equation}
 \sum_s Z_s n_s =0, \qquad \sum_s Z_s n_s {1 \over L_{n_s}} = 0
\end{equation}
Multiple ion species can be input but only one negatively charged species is allowed.  The number of species read in is controlled by the number of species parameter in the `gridsize' namelist and
hence it is possible that later species data you specify will not be read in if this is set incorrectly.  The \name{number_of_species} input variable does not include adiabatic electrons (the
species data for the negative charge species will still be read).

Note that for a trace species, $n_s=0$ is acceptable. The density only appears in the field equations and for scattering species in collisions.  A trace particle does not contribute to the fields and does not act as a scattering species, so setting $n_s=0$ will give the desired result.

\subsection{Stability}
The timestep required for numerical stability can vary greatly depending on input parameters, and for kinetic electrons will need to be much smaller.
For a linear run, there is a time step estimate made at the start of the run and the code will post a warning if
the chosen timestep is larger than this. For a non linear run, the minimum time step is estimated based on the
non linear terms, then the time step is adapted as necessary. A minimum time step size can be set in the code (\name{dt_min}) so
that the run stops when the time step becomes too small. The parallel structure of the solution is a good
indicator of a the stability of a solution.

For linear runs a stable solution will converge to a constant growth rate. A tolerance for the growth rate can be
specified in the code input (parameter \name{gamatol}), so that the run ends when the standard deviation of the growth rate computed over the last six big time steps is less than \name{gamatol}. When the tolerance on the growth rate is used to end the run, it is advised to keep the big time steps long enough to have a meaningful standard deviation. Typically, if the big time step length is such that $\name{DTIM}\times\name{NAVERAGE}>0.2$, setting $\name{gamatol}=10^{-6}$ will usually ensure convergence.

GKW outputs the global growth rate
in \File{time.dat} so a growth rate spectrum can be generated from a batch of linear runs with 
\name{NMOD=1} and \name{mode_box=.false.}.  Typical grid sizes can be seen in the example input files.

\subsection{\texorpdfstring{$\beta$ and $\beta'$}{beta and beta prime}}
\label{Sec:betabetapr}
The plasma $\beta$ enters the field equations (parallel and perpendicular components of the Ampere's law, Eq.~\ref{eq:AmpPar} and
\ref{eq:AmpPerp}) through the parameter $\beta_{ref}$ defined in Eq.~\ref{eq:betaref} and recalled here:
\begin{equation}
 \beta_{ref} =  \frac{n\ind{ref}T\ind{ref}}{B^2\ind{ref}/2\mu_0}.
\end{equation}
The parameter $\beta_{ref}$ is linked to the total plasma $\beta$ at the reference magnetic field as follows:
\begin{align}
\label{eq:betadef}
 \beta_0 &= \sum\ind{species} 2\mu_0\frac{n_sT_s}{B^2\ind{ref}} = \beta_{ref} \sum\ind{species} \frac{n_s}{n\ind{ref}}\frac{T_s}{T\ind{ref}} = \beta_{ref}
\sum\ind{species} n_{N_s} T_{N_s}\\
& = \beta_{ref} \sum\ind{species} \beta_{N,s}\qquad\text{with}\quad \beta_{N,s} := n_{N_s} T_{N_s}
\end{align}
The user can specify the value of $\beta_{ref}$ used by GKW within the namelist \name{SPCGENERAL}. Two options are implemented and controled by
the switch \name{BETA_TYPE}:
\begin{itemize}
 \item \name{BETA_TYPE=`ref'}\\
  The value of $\beta_{ref}$ is given as an input in \name{BETA_REF}.
 \item \name{BETA_TYPE=`eq'}.\\
 The total plasma $\beta$ at the reference magnetic field is read from the MHD equilibrium file and $\beta_{ref}$ is calculated using Eq.~\ref{eq:betadef}
and the species parameters.  The value given as an input in \name{BETA_REF} is not used and replaced by the calculated value. Note that this option is
only available if \name{GEOM_TYPE='chease'} in namelist \name{GEOM}. If it is not the case a warning is displayed and $\beta_{ref}=0$ is used instead. 
\end{itemize}

In addition, the radial derivative of the plasma $\beta$ modifies the curvature drift, as seen in Eq.~\ref{eq:drifts}. Here again, the value at the
reference magnetic field is needed (the dependence on the magnetic field is introduced directly in \File{linear_terms.F90}): 
\begin{equation}
\label{eq:betapr0def}
 \beta'_0 = \frac{2\mu_0}{B\ind{ref}^2}\sum\ind{species}\frac{\partial p_s}{\partial \psi} = - \beta_{ref} \sum\ind{species}  n_{N_s} T_{N_s}
[R/L_{n_s} + R/L_{T_s}]
\end{equation}
Three ways of inputing $\beta'_0$ are implemented:
\begin{itemize}
 \item \name{BETAPRIME_TYPE=`ref'}\\
The value of $\beta'_0$ is given as an input in \name{BETAPRIME_REF}.
 \item \name{BETAPRIME_TYPE=`sp'}\\
The value of $\beta'_0$ is made consistent with the species parameters and the value of $\beta_{ref}$ defined above
using Eq.~\ref{eq:betapr0def}. The value given as an input in \name{BETAPRIME_REF} is not used and replaced by the calculated value
 \item \name{BETAPRIME_TYPE=`eq'}. \\
The value of $\beta'_0$ is read from the MHD equilibrium file. The value given as an input in \name{BETAPRIME_REF}
is not used and replaced by the calculated value.
\end{itemize}

Note that when the species parameters are used, ALL species are considered including the adiabatic one.

Reminder: if \name{GEOM_TYPE='chease'}, the values of $R\ind{ref}$ and $B\ind{ref}$ used in GKW are the values \name{R0EXP}
and \name{B0EXP} given as an input to CHEASE. It is advised that \name{R0EXP} and \name{B0EXP} are chosen to be the radius of the magnetic axis and
the value of the magnetic field at this position respectively for better convergence in CHEASE.

\subsection{Generating a linear growth rate spectrum}
GKW currently outputs the global growth rate (and frequency) in \File{time.dat} so the easiest way to generate a growth rate spectrum is do a batch of linear runs with \name{NMOD=1}, \name{mode_box=.false.}, and \name{NX=1}, and scan over \name{kthrho} and take the growth rate from each. 
% If \name{mode_box=.true.} and you have mode than one radial mode, 
% scan over \name{krhomax} to set the toroidal wavevectors (to keep the kx spacing the same you also have to vary ikxspace with krhomax); 
The matlab scripts \matlab{create_gkwscan.m} and \matlab{growth_rate_spec.m} help with creating the input and plotting the output; 
script \scripts{gkwnlin} runs the batch jobs.

Alternatively, a growth rate (frequency) spectrum can be generated by running with \name{mode_box=.true.} and setting
\name{lgrowth_rates=.true.} (\name{lfrequencies=.true.})in the \name{DIAGNOSTIC} namelist.  For each toroidal mode,
the growth rate and of the mode with $k_\psi=0$ will be output.  To get this spectrum to exactly match a spectrum
generated by the first method at higher $k_\zeta$ will require large \name{NX} (and low \name{ikxspace}) to fully resolve 
the mode along the field line.  The excercise of matching the spectra generated by these two methods
will require the user to understand the mode coupling over the parallel boundary condition
and will indicate the \name{nx} resolution required for \name{mode_box=true} and nonlinear runs.

\pagebreak

\subsection{Restarts \label{sec.restart} and checkpoints}

On a successful exit, GKW writes a binary file \File{FDS} which
contains the distribution function needed for a restart.  Along with the
input file, this is the only file required to restart the code.  
For seamless continuation, some other quantities such as file counters 
and normalized time are also written to the ASCII file \File{FDS.dat} which
contains Fortran namelist which the code reads.   The binary restart file 
is written using MPI-IO and is independent of processor layout.  This means that
runs can be restarted with different parallelizations from the original run.

GKW will only attempt to restart from the above files if \name{read_file=T} in the \name{CONTROL}
namelist.  If it does not find any files, it will initialize a new run.  Under the default behaviour,
when this subsequent run completes, a restart files \File{FDS} and \File{FDS.dat} will be overwritten.
This behavior can be altered by setting the integer \name{irun} in the \name{CONTROL} namelist which
allows incremental numbering of restart files.  If \name{irun=1},  
\File{FDS} is read if present and \File{FD1} is written.   If \name{irun=2},  
\File{FD1} is read and \File{FD2} is written, and so on.

For systems on which MPI-IO is not available or problematic\footnote{
For large problems with more than about 64 processes,
there can be problems/errors when reading/writing the restart files in
parallel, depending the filesystem and the MPI implementation.
This has been observed on some Cray XT4 machines, where in
some cases it was necessary to modify environment variables to
successfully write restart files. There is also a workaround in the
code which must be set at compile time via adjustments to two variables in
\File{switches.F90}: Rather than writing to file with all processes
simultaneously, this allows writes with subsets of processes
at a time, where the file is closed then re-opened between subsets.
Increasing the number of subsets reduces overheads and can allow files to
be written successfully. This might also improve the I/O performance on
machines, where such operations already work. 
However, if the number of subsets is too large, the I/O
performance will suffer. The variable \name{min_fdisi_io_subsets} selects
the minimum number of process subsets to use. When problems are encountered, 
it is suggested to first try setting this value to the value of \name{n_procs_sp}, 
then to \name{n_procs_sp*n_procs_s} if problems persist. Using larger values will
likely give very poor performance, but may be necessary. The other variable,
\name{min_procs_io_split}, sets the minimum number of processes for which
the read/write split should be used.},
GKW can also be made to write
a legacy restart file format, in which each processor writes its own file 
\File{FDSXXXXXX} numbered by processor (option \name{restart_file_version=1} in \name{CONTROL}).  
In this case, restarts may not alter the parallelization.

Intermediate checkpoints can also be written by setting the integer \name{n_dump_ts=50} in the \name{CONTROL} namelist,
to write, for instance a checkpoint every 50 large timesteps.  This is strongly advised for any expensive nonlinear simulation.  
Two restart files \File{DM1/2} are alternately overwritten by a newer one each time.  These files have exactly the 
same format as the final restart file \File{FDS} and can be used for a restart by simple renaming.
One should note that if restarting an interrupted run from a checkpoint,  some the time trace files may have duplicated lines, 
which can be easily verified by checking or plotting column 1 of \File{time.dat}.

A clean exit can be requested at anytime while the code is running by creating a
file named \name{gkw.stop} in the run directory. An additional checkpoint 
file can be created by instead creating a file named \name{gkw.dump}, after which
the code will continue to run.

Some batch systems will automatically resubmit a job which is terminated by a hardware problem; 
to allow GKW to automatically restart from the \File{DM1/2} files (and take precedence over FDS files) if present, 
set \name{auto_restart=T} in the \name{CONTROL} namelist.   GKW could also be made to interface with batch 
systems which are provide accounting flexibility for codes able to halt on request,
by means of the \name{gkw.stop} file creation mechanism.

\pagebreak

\subsection{Tips and tricks}

%Should this be a wiki page?

This section aims to document other useful things to know when running the
code. Please add here anything else useful to the user that we have yet to
document properly. Received wisdom, etc.

\begin{itemize}

\item The linear timestep estimator is based on the Courant condition for the
       Vlasov equation, and the frequency of shear-Alfv\`{e}n wave \cite{LEE01}, so it is not
       foolproof for some high frequency waves (e.g. ETG). If GKW gives unphysicaly large growth
       rates, the first thing one should try is to reduce the timestep (no more
       than a factor of two at a time). If your problem requires a very small
       timestep (less than 1e-6), something else is likely to be wrong, but if
       a very small timestep is really required it may be best to run with
       \name{normalized=.false}. Equally, for most efficient resource usage,
       one should try to find the upper limit of the stable timestep.

\item For nonlinear kinetic runs, adding a small electromagnetic component e.g.
       $\beta=0.03\%$ can stabilise some long wavelength instabilities
       (electrostatic shear-Alfv\`{e}n waves), allowing a much larger stable
       timestep, as suggested in \cite{MER10}. 
       %Same trick is mentioned in Jenko 2010 paper

\item For kinetic electron runs (\name{adiabatic\_electrons=.false.}) the default 
      initialization option \name{finit='cosine2'} may lead to 
      initial bursts. A different initialization of the distribution function, 
      e.~g., \name{finit='cosine5'}, may then be more appropriate.

\item The nonlinear timestep estimator rarely manages to prevent blowup in the
       event of numerical instability, but can be useful for strongly nonlinear
       physics.   For very large numbers of processors it may cause an overhead
       which reduces the scaling efficiency.  It can therefore be disabled
       by setting. \name{nl_dtim_est=F} in the \name{CONTROL} namelist, which
       will cause the code will abort rather than continue if the timestep
       estimate is violated.  The code could then be restarted from the previous
       dumpfile with the nonlinear timestep estimator enabled, or with a smaller
       timestep.

\item The nonlinear timestep estimator is often too conservative.  
      If your problem is limited by the nonlinear timestep, 
      it may be possible to save CPU resources by increasing it (no more than $90\%$) 
      by setting e.g. \name{fac_nl_est = 1.6} in \name{CONTROL}.

\item Don't be tempted to use more cores than you need.  Pushing the parallelisation to its limit 
       (particularly less than 4 points per processor in parallel velocity) will result in inefficient scaling.

\item Be aware of the effects of the boundary condition at the end of the field line. If not running a box of modes, you should
       check that \name{nperiod} is large enough so that the results are insensitive to the free boundary. If running with
       \name{mode_box}, care should be taken with the radial mode resolution so that most toroidal modes have several periods and
       do not feel the effect of the boundary. This is particularly important when kinetic electrons are used.

\end{itemize}

\subsection{Hybrid scheme \label{sec:hybrid}}

Hybrid parallelism combining distributed memory MPI parallelism with shared memory OpenMP parallelism
has been implemented to allow the code to scale further, and more efficiently. 
Many modern high performance machines now have an architecture with a number of shared memory CPU cores on each node (2-24), 
and a large number of these nodes connected by a fast network interconnect.  
The communication between nodes is much slower than between cores, and a well optimised and well
utilised MPI implementation will take advantage of this by putting the most intensive communication on the local node.
  In cases where this is not done, or cases where a subset of 
the problem grid does not fit nicely on a single node, the efficiency of the inter-node communication may be increased by reducing 
the number of MPI processes involved in the intra-node communication.  For example, on a machine with 8 cores per node (e.g. HPC-FF), 
1024  processors can be utilised with the combinations 
(number of MPI processes, number of OpenMP threads per node) = (1024,1), (512,2), (256,4), or (128,8), 
where the lower number of MPI processes in the latter case may result in a more efficient reduction sum 
(though in theory the amount of inter-node data communicated by the ideal MPI implementation would remain the same).

In GKW, hybrid parallelism (for the spectral scheme only) is achieved by subdividing the most CPU intensive loops amongst OpenMP threads.  
The slowest of these is the loop in nonlinear terms, where for each local grid point, 5-7 FFTS in 2D must be performed. 
At the maximum limit on the number of MPI processes, 
there will always be at least 2 (4 with collisions) points in this loop (it contains \name{NMU*NSP*NS} local grid points) 
which can be divided between threads.  
(The code could be rewritten here to increase the number of threads that can be used in this section, 
if the hybrid scheme is demonstrated to bring a clear benefit at these numbers of threads).  
The second largest loop which must be parallelised is the sparse matrix-vector multiply involved in the linear terms. 
Due to the sparse matrix storage format, this loop cannot be vectorised, but cache efficiency can be maximised 
by an appropriate sorting of the matrix.  After this sort, the matrix elements are subdivided between threads at carefully chosen boundaries,  
to ensure there is no data race in the update of the right hand side.  The same technique is used for the fields calculation. 
All possible smaller loops ($<$5\% runtime) are also parallelised with OpenMP, 
but here there are diminishing returns here due to the overhead involved in thread branching.  
This inherently serial element of the code will set the limit of the number of threads that can efficiently be used (likely four at present).

The hybrid scheme with OpenMP parallelism can allow GKW to scale better in certain situations for large runs.  
Tests on HPC-FF and Babel Bluegene machines (combined with some reasoning), suggest that the cases in which it is most likely to be useful 
are those in which the MPI communications are the bottleneck.  When this occurs depends on the problem and the hardware.  
These situations are likely to include one or more of the following:
\begin{itemize}
 \item Problems with large numbers of modes
 \item Problems with collisions (4 threads has been demonstrated to bring a benefit here)
 \item Runs on very large numbers of processsors, near the limit of GKW MPI capability
 \item Hardware with slow communication (HPC-FF or BlueGene as opposed to Hector)
\end{itemize}

Preliminary tests with the hybrid scheme on up to 1536 processors (on HPC-FF: IBM-Intel-Xeon-Infiniband) demonstrated up to a 5\% 
performance increase using two OpenMP threads per node.  More extensive scaling tests on Bluegene/P BABEL have
demonstrated a scaling efficiency increase of up to 20\% with 4 threads when running a collisional case on 8192 cores and above 
(for more details see Sec. \ref{sec:bluegene}).
%ALSO, RECREATE HECTOR 2a or new HECTOR2b or HECTOR 3 scaling test to see in open MP can bring a benefit there.
%Want to do a large collisional scaling test without the diagnostics on.
%Want to know how much is due to momentum conservation (this was not even integrated into hybrid shcheme at time of test).

You are advised to do some brief checks for your problem and hardware to determine
if the hybrid scheme is an advantage.  Best results will probably be obtained using the hybrid scheme 
with no more than four OpenMP threads per MPI process.
 For acceptable results, the number of threads must be a factor of the local processor gridsize 
 \name{NMU*NSP*NS} (which is also the maximum number of threads), as a loop of this size is parallelised in the nonlinear terms.

To use the hybrid scheme, you need a compiler that supports OpenMP.  You will need to compile with
SMP=OPENMP preprocessor flag, openmp compiler flag, and link with the correct libraries (by using the same compiler flag). 
In the standard makefile build process, this is usually achived by setting 
\begin{verbatim}
SMP=OPENMP
FFLAGS_OMP=-openmp (for intel, or -fopenmp for gnu)
\end{verbatim}
Using the compile optimisations specific to the processor is also recommended.

\subsection{A brief guide to moving to global simulations}
\label{sec:brief-guide-moving}

\emph{This guide needs improvement, please add to it.}

This section assumes you can already cope with running non_spectral.
Running global simulations is somewhat more complicated:

The input switches for the global version are not yet very well
documented, and the input checks and error messages may not yet be
very comprehensive or informative (there is a job there for anyone 
that wants to be helpful).

It is probably easier to start from an existing global test case, 
but to move from local to global you would need to add at least 
the following switches.

\paragraph{CONTROL}
\subparagraph{\name{spectral_radius = .false.}}
Since the radial direction is no longer 
homogeneous you can not use a spectral 
method 
\subparagraph{\name{flux_tube = .false.}}
This is the switch that tells the code 
it is in fact a global run 
\subparagraph{\name{disp_x = 0.05}}
You need this (or more). Since there 
are radial derivatives in the equation 
one needs to put some dissipation.

\paragraph{GRIDSIZE}
\subparagraph{\name{vpmax = 4.0}} The velocity grid does not vary with radius, 
and so it must be able to resolve both the low 
as well as the high temperatures in the radial 
domain. This means that one often has to increase 
both the maximum value of the grid (vpmax) as 
well as the velocity space resolution (n_vpar_grid)
The value chosen depends on how much the temperature 
varies over the radial domain.
\subparagraph{\name{n_vpar_grid = 64}} You must often choose this larger than a fluxtube 
run (see comment vpmax)
\subparagraph{\name{mumax = 8.0}} You must often make the perpendicular velocity 
grid larger (see comment vpmax) 
\subparagraph{\name{n_mu_grid = 16}} also the resolution in the perpendicular velocity
must be larger usually (see comment vpmax)
\subparagraph{\name{psil = 0.02}} inner boundary eps=r/Rref
\subparagraph{\name{psih = 0.3}} outer boundary eps=r/Rref
\subparagraph{\name{nx = 128}} Better to be even in nonspectral. Also this value 
must be chosen high enough such that you resolve 
at least 2 times the ion Larmor radius (1 is better
and kinetic electron runs might require more 
resolution (this is true for nonspectral as well) 
\subparagraph{\name{n_procs_x = 4}} Now you can parallelize here too

\paragraph{MODE}
\subparagraph{\name{n_spacing = 2}} For a run with one mode this is the toroidal mode 
number (kthrho / krhomax are not used). For a 
nonlinear run with many modes this is the mode 
spacing, i.e. n_spacing = 2 with nmod = 4 means 
that the toroidal modes 0, 2, 4, and 6 are kept 
in the simulation (this is the half-torus)

\paragraph{SPCGENERAL}
\subparagraph{\name{rhostar = 1e-3}} rho_ref/R_ref

\paragraph{SPECIES}
\subparagraph{\name{dens_prof_type = 'exp_tanh'}}
\subparagraph{\name{temp_prof_type = 'exp_tanh'}}
\subparagraph{\name{temp_prof_coef = 0.0,0.0,0.0,0.0,0.0}}
\subparagraph{\name{dens_prof_coef = 0.0,0.0,0.0,0.0,0.0}}
These define the radial profile of temperature/density, and through the
derviatives also the profiles of the density gradient lengths. For possible
profile types and the meaning of the coefficients for those, see first part of
section \ref{secprofilefunctions}.

\paragraph{GEOM}
\subparagraph{\name{prof_type = 'parabolic'}}
\subparagraph{\name{qprof_coef = 0.0,0.0,0.0,0.0,0.0}}
These define the radial profile of the safety factor, and through the
derivative also the shear profile. For possible profile types and the meaning
of the coefficients for those, see second part of section \ref{secprofilefunctions}.

The chosen profiles are output in prof_back.dat 
(and there is a matlab script to read them).

If you are going to attempt a nonlinear simulation, then you will
probably also want the krook operator to prevent profile relaxation.
See section \ref{seckrookoperator}, for more details.

\paragraph{KROOK}
\subparagraph{\name{krook_option=1}}, global krook preserving momentum
\subparagraph{\name{gammak=0.05}}, global krook damping rate (should be << linear gr)
\subparagraph{\name{nlkrook=.true.}}, global krook
for boundary krook cells, see input.dat.sample

You should also take a look at the \name{GYROAVERAGE} namelist, which
is fairly well documented in \doc{input.dat.sample}.

\pagebreak

\section{Example GKW input file}
\vskip 0.2 truecm
Further example input files are provided with the code, including a comprehensive input file 
\doc{input.dat.sample}, a minimal input \doc{input.dat.minimum}, 
and input files for some common benchmark cases in \doc{input/}.
\begin{scriptsize}
\begin{verbatim}
 &CONTROL                   !Namelist read in module CONTROL
  non_linear = .false.      !Include the non linear terms, computationally expensive fft
  zonal_adiabatic = .false.,!If zonal flows correction included for adiabatic electrons
  order_of_the_scheme = 'fourth_order'  !or 'second_order' in space
  dtim   = 0.02          !Time step size (normalised time)
  ntime  = 25,           !Number of iterations of NAVERAGE.
  naverage = 100,        !Number of timesteps between re-normalisation, data output
  nlapar = .false.,      !true = keep parallel electro-magnetic potential.(electrostatic=false)
  nlbpar = .false.,      !true = keep compressional magnetic field (electrostatic=false) 
  collisions = .false.   !Turn on the collision operator
  disp_par = 1.0         !Dissipation coefficient for parallel derivatives.
  disp_vp  = 0.0         !Dissipation coefficient for parallel velocity space
  max_seconds = 1000.    !Internal program kill time.
  gamatol     = 0.       !tolerance in the linear growth rate for which the code should stop
  io_format   = 'ascii'  !chose output format to be ascii
 /
 &GRIDSIZE        !Namelist read in module GRID
  nx = 83,        !Number of radial wave vectors - needs to be an odd number for fft
  n_s_grid = 50,  !Number of grid points along the field line 
  nperiod = 1,    !Integer - the field line makes 2*nperiod - 1 poloidal turns. 
  n_mu_grid   = 8,        !Number of magnetic moment grid points
  n_vpar_grid = 32,       !Number of grid points for parallel velocity 
  nmod = 21,              !Number of bi-normal modes (k_zeta)
  number_of_species = 1   !Number of kinetic species. 
 /
 &MODE                !Namelist read in module MODE
  kthrho  = 0.5,1.0,  !'Poloidal' wavevector(s) if mode_box=.false.
  mode_box = .true.,  !Determines if there is a 2D grid of modes. If true use nperiod = 1
                      !Also determines if radial modes are coupled at end of field line.
  krhomax = 1.0,      !For mode_box, the maximum 'poloidal' wavevector (k_perp rho_i) used
                      !k_perp evaluated on the low field side of the outboard midplane
                      !rho_i evaluated on the flux surface at the major radius of magnetic axis
  ikxspace = 7,       !Spacing between radial modes (resolution against box size)
 /
 &SPCGENERAL                   !Namelist read in module COMPONENTS
  beta_ref = 0.000,            !Plasma beta (not used if both nlapar, nlbpar are .false.)
  adiabatic_electrons = .true. !Kinetic electrons require a smaller timestep
 /
 &SPECIES       !Namelist read in module COMPONENTS
  mass  = 1.0,  !Species mass in terms of reference mass
  z     = 1.0,  !Species charge. If negative, assumed to be the electrons.
  temp  = 1.,   !Temperature of species scaled by reference temperature 
  dens  = 1.,   !Density of species scaled by reference density 
  rlt   = 8,    !Temperature gradient R/LT
  rln   = 3.5,  !Density gradient R/Ln
  uprim = 0.0,  !Gradient of the toroidal rotation Rref^2 grad Omega / v_thref
 / 
 &SPECIES 
  mass  = 2.72e-4, !Electrons
  z     = -1.0,    !Only one negatively charged species is allowed.
  temp  = 1.0, 
  dens  = 1.0,
  rlt   = 7.9,
  rln   = 3.5, 
  uprim = 0.0,
 /
 &GEOM          !Namelist read in module GEOM, s-alpha is default geometry
  shat = 1.0,   !Magnetic shear
  q    = 2.0,   !Safety factor
  eps  = 0.1    !Radial Coordinate
 /
 &ROTATION                   !Namelist read in module ROTATION
  vcor = 0.0                 !Rotation of the plasma vcor = Rref Omega / vthref = \Omega_N
  shear_rate = 0.0           !Normalised shearing rate for the E x B perpendicular shear 
  shear_profile  = 'none'    !To include a shear flow, use `wavevector_remap'
  cf_trap = .true.           !Include the centrifugal trapping effects           
  cf_drift = .true.          !Include the centrifugal drift
 /
 &COLLISIONS    !Namelist read in module COLLISONOP if collisions = .true.
  rref = 1  !reference major radius used in collision operator 
  tref = 1  !reference temperature in units of kev used for the collision operator
  nref = 1  !reference density in units 10^19 m^-3 used for the collision operator
  mom_conservation = .false.     !Use the correction to conserve momentum
  mass_conserve = .true.,        !Forces zero flux boundary on velocity space so no out flow of mass   
  freq_override=.false.          !Manually specify ii collision frequency to override ref values 
  coll_freq=6.515E-5             !Ion-ion collision frequency used if freq_override=.true.
 /
\end{verbatim} 
\end{scriptsize}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/collisions.tex
(1) Added gkw Clone gwk  #3

\chapter{Collisions}

The previous sections have neglected the collision operator. 
In this section we describe the collisions as implemented in GKW.  The present implementation neglects the 
finite Larmor radius effects. 
The collisions enter the evolution equation as an additional term 
\begin{equation} 
{\partial \hat{f_a} \over \partial t} = C(\hat{f_a}) .
\end{equation} 
Here we have added the index $a$ to denote the species. 
Useful expressions for the collision operator ($C$) have been published in the literature, usually in the 
($v,\theta$) coordinate system, where $v$ is the velocity and $\theta$ is the pitch angle.
Since the perturbed distribution function is small $f \approx \rho_* F_M$, GKW uses a linearised 
collision operator with the Maxwellian as background. 
This operator has the following form \cite{KAR86} 
\begin{equation} 
C(\hat f_a) = \sum_b {1\over v^2} {\partial \over \partial v} \biggl [ v^2 \biggl ( 
D_{vv}^{a/b} {\partial \hat{f_a} \over \partial v} - F_v^{a/b}\hat{f_a} \biggr ) \biggr ] + {1 \over v 
\sin \theta } {\partial \over \partial \theta} \biggl [ \sin \theta  D_{\theta \theta}^{a/b} 
{1\over v}{\partial \hat{f_a} \over \partial \theta} \biggr ] .
\end{equation} 
Here the sum is over all species $b$. 
$D_{\theta \theta}$ represents the pitch angle scattering, while $D_{vv}$ is the energy scattering
and $F_v$ is the slowing down force. For convenience we split this collision 
operator in three parts 
\begin{equation} 
C(\hat{f}) = C_\theta(\hat{f}) + C_{vv}(\hat{f}) + C_v(\hat{f}),
\end{equation}
where the first part is due to $D_{\theta \theta}$, the second due to $D_{vv}$ and the third 
due to $F_v$. The normalisation of the collision operator follows from the normalisation of the 
gyrokinetic equation as set out in section \ref{sec:normalisation}, giving 
$C_N(\hat{f}_a) = {R_{\rm ref} v_{tha}^3 \over n_a \rho_* v_{\rm thref}} C(\hat{f}_a)$ with the diffusion 
and friction force coefficients made dimensionless accordingly.

The coefficients can be obtained from the literature \cite{KAR86} (and the references cited therein) and written in normalised form, suppressing the subscript N for the normalised quantities:	
\begin{equation} 
D_{\theta \theta}^{a} = \sum_b {\Gamma^{a/b} \over 4 v_{a}} \biggl [ \biggl ( 2 - {1\over v_{b}^2} \biggr ) 
{\rm erf}(v_{b}) + {1 \over v_{b}} {\rm erf}^\prime (v_{b}) \biggr ] ,
\end{equation}
\begin{equation} 
D_{vv}^{a} = \sum_b { \Gamma^{a/b}  \over 2 v_{a}} \biggl [ {1 \over v_{b}^2} {\rm erf}(v_{b}) - 
{1\over v_{b}}{\rm erf}^\prime (v_{b}) \biggr ] ,
\end{equation} 
\begin{equation} 
F_{v}^{a} = - \sum_b {\Gamma^{a/b} \over v_{a}^2} {m_{Ra} \over m_{Rb}} [ {\rm erf}(v_{b}) - v_{b} {\rm erf}^\prime (v_{b})  ]  .
\end{equation} 
where ${\rm erf}$ is the standard definition of the error function.  The normalised velocity in these equations is defined as 
\begin{equation}
v_{\gamma} =  {v_{Ra} \over v_{R\gamma}}[ v_{\parallel }^2 + 2 \mu B ]^{1/2} ,
\end{equation}
with $\gamma = a$ or $\gamma = b$.  The error function is  
\begin{equation} 
{\rm erf}(u) = {2 \over \sqrt{\pi}} \int_0^u \exp [ -x^2] {\rm d} x 
\end{equation} 
\begin{equation} 
{\rm erf}^\prime(u) = {2 \over \sqrt{\pi}} \exp [ -u^2] 
\end{equation}
Finally the constant $\Gamma^{a/b}$ is given by 
\begin{align} 
\Gamma^{a/b} = {R_{\rm ref} n_b Z_a^2 Z_b^2 e^4 \ln \Lambda^{a/b} \over 4 \pi \epsilon_0^2 m_a^2 v_{tha}^4} = 
6.5141\cdot 10^{-5} {R_{\rm ref} n_{\rm ref}^{19} \over (T_{\rm ref}^k)^2}  {n_{Rb} Z_a^2 Z_b^2 \ln \Lambda^{a/b} 
\over T_{Ra}^2} ,
\label{eq:refquantities-collisions-constraint}
\end{align}
where $R_{\rm ref}$ must be given in meters, $n_{\rm ref}^{19}$ is the reference density in units of $10^{19}$
$m^{-3}$, and $T_{\rm ref}^k$ is the reference temperature in units keV. The Coulomb logarithm is calculated as
\begin{eqnarray}
\ln \Lambda^{e/e} &=& 14.9 - \frac{1}{2}\ln[0.1n_{\rm R}^e n_{\rm ref}^{19}] + \ln[T_{\rm R}^e T_{\rm ref}^k]\\
\ln \Lambda^{e/i} &=& 17.2 - \frac{1}{2}\ln[0.1Z^2n_{\rm R}^e n_{\rm ref}^{19}] + 1.5\ln[T_{\rm R}^e T_{\rm ref}^k] \qquad \textrm{if} \,\, T_e < 10Z^2\,\rm{eV}\\
\ln \Lambda^{e/i} &=& 14.8 - \frac{1}{2}\ln[0.1n_{\rm R}^e n_{\rm ref}^{19}] + \ln[T_{\rm R}^e T_{\rm ref}^k] \qquad \textrm{if} \,\, T_e \geq 10Z^2\,\rm{eV}\\
\ln \Lambda^{i/e} &=& \ln \Lambda^{e/i} \\
\ln \Lambda^{i_1/i_2} &=& 17.3 -\ln[\frac{Z_1Z_2(m_1+m_2)}{m_1T_{\rm R2}T_{\rm ref}^k+m_2T_{\rm R1}T_{\rm ref}^k}] - \frac{1}{2}\ln[0.1\frac{n_{\rm ref}^{19}}{T_{\rm ref}^k}] -\frac{1}{2}\ln[\frac{n_{\rm R1}Z_1^2}{T_{\rm R1}}+\frac{n_{\rm R2}Z_2^2}{T_{\rm R2}}]
\end{eqnarray}
where the superscripts $e$ and $i$ stand for electron and ion, respectively.

Note that the reference density and temperature are defined independently for the collisions and do not need to be equal to the reference density and temperature used in Sec.~\ref{sec:normalisation} for the species parameters. However, the electron density entering the calculation of the Coulomb logarithm is taken to be $n_{\rm R}^e n_{\rm ref}^{19}$ and therefore depends on the normalised electron density given as in an input in the species namelist. The same comments pertains for the temperatures entering the calculation of the Coulomb logarithm.

Using the relation between ($v,\theta$) and the normalised coordinates ($v_{\parallel},\mu$) used in GKW, and performing 
the coordinate transform, one arrives at the explicit expressions given below. The pitch angle scattering term is 
\begin{align}
C_{\theta}  &\rightarrow & 
{2 v_R} {\partial \over \partial v_{\parallel}} \biggl [ {\mu D_{\theta \theta} \over 
v_{\parallel}^2 + 2 \mu B} \biggl ( B {\partial \hat{f}\over \partial v_{\parallel}} - v_{\parallel} 
{\partial \hat{f}\over \partial \mu} \biggr ) \biggr ] \cr 
\noalign{\vskip 0.2 truecm} 
&+& {2 v_R \over  B} {\partial \over \partial \mu} 
\biggl [ {v_{\parallel} \mu D_{\theta \theta} \over v_{\parallel}^2 + 2 \mu B} \biggl ( - B 
{\partial \hat{f} \over \partial v_{\parallel}} + v_{\parallel} {\partial \hat{f}\over \partial \mu} \biggr ) \biggr ] .
\end{align}
The energy scattering yields 
\begin{align}
C_{vv} & \rightarrow & 
v_R {\partial \over \partial v_{\parallel} } 
\biggl [  {v_{\parallel} D_{vv} \over v_{\parallel}^2 + 2 \mu B} \biggl ( v_{\parallel} {\partial 
\hat{f} \over \partial v_{\parallel}} + 2 \mu {\partial \hat{f} \over \partial \mu} \biggr ) \biggr ] \cr 
\noalign{\vskip 0.2 truecm}
&+& v_R {\partial \over \partial \mu} \biggl [ {2 \mu D_{vv} \over v_{\parallel }^2 + 
2 \mu B} \biggl ( v_{\parallel} {\partial \hat{f} \over \partial v_{\parallel}} + 2 \mu {\partial \hat{f} \over 
\partial \mu}\biggr ) \biggr ] .
\end{align}
The slowing down can be written as 
\begin{align}
C_{v}  &\rightarrow & -
v_R {\partial \over \partial v_{\parallel} } \biggl [  {v_{\parallel} F_v \over [ v_{\parallel}^2 
+ 2 \mu B ]^{1/2}} \hat{f}  \biggr ] \cr
\noalign{\vskip 0.2 truecm}
& & - v_R {\partial \over \partial \mu} \biggl [ {2 \mu F_v \over [v_{\parallel }^2 + 
2 \mu B]^{1/2}} \hat{f} \biggr ] . 
\end{align}

The linearised form of the Fokker-Planck collision operator above conserves particle number but does not conserve like particle momentum or 
energy. 
The conservation of parallel momentum and energy can be reintroduced using the following simplified model.  
We are interested in momentum conservation for like particle collisions. 
The change in parallel momentum due to the collision operator above is 
\begin{align}
&& 2 \pi B \int {\rm d}v_\parallel \int {\rm d} \mu v_{\parallel} C(\hat{f}) 
\end{align}
This loss of momentum can be corrected for by adding an ad hoc correction term 
\begin{equation} 
{\partial \hat{f} \over \partial t} = C_{mom} v_{\parallel} F_{M},
\end{equation}
but the energy term isn't simply
\begin{equation} 
{\partial \hat{f} \over \partial t} = C_{ene} (v^{2}_{\parallel} + 2\mu B) F_{M}
\end{equation}
as this would have the effect of adding particles, conservation of which is accounted for by
the differencing and boundary conditions.  Therefore we use
\begin{equation} 
{\partial \hat{f} \over \partial t} = C_{ene}\left( (v^{2}_{\parallel} + 2\mu B) - A\right)F_{M}
\end{equation}
where $A$ is calculated by taking the zeroth moment of the equation
\begin{equation}
\frac{\partial \hat{f}}{\partial t} =  C(\hat{f}) + C_{ene}\left( (v^{2}_{\parallel} + 2\mu B) - A\right)F_{M}
\end{equation}
and setting the change in density to zero.  Since particle conservation is determined by the differencing and boundary conditions, then the coefficient $A$ turns out to be simply:
\begin{equation}
A = \frac{\int {\rm d}v_\parallel \int {\rm d} \mu v^2 F_{M} }{\int {\rm d}v_\parallel \int {\rm d} \mu F_{M}}
\end{equation}
which analytically can be shown to be $\frac{3T}{2}$, but in GKW is calculated numerically to ensure highest precision conservation.

The constant $C_{mom}$ then follows from working out the momentum change due to this term. 
\begin{equation} 
2 \pi B C_{mom} \int {\rm d} v_{\parallel} {\rm d} \mu v_{\parallel}^2 F_{M} +  2 \pi B \int {\rm d}v_\parallel \int {\rm d} \mu v_{\parallel} C(\hat{f}) = 0.
\end{equation} 
while the constant $C_{ene}$ is determined by taking the energy moment $\frac{mv^{2}}{2}$, 
\begin{equation} 
2 \pi B C_{ene} \int {\rm d} v_{\parallel}{\rm d} \mu v^{2}(v^{2}-A) F_{M} +  2 \pi B \int {\rm d}v_\parallel \int {\rm d} \mu (v_{\parallel}^2 + 2\mu B) C(\hat{f}) = 0.
\end{equation}
here we have suppressed the $m/2$ factors as they cancel.

The present form of the collision operator drives the distribution 
towards a Maxwellian with the specified temperature. In the local problem one specifies the temperature as an input 
parameter and no evolution in the temperature is retained. A collision operator that maintains this temperature can therefore 
be considered acceptable.  The steady state distribution function for various velocity space resolutions is plotted in Fig.\ref{maxwellcoll}.

To switch on the conservation terms, which can be switched on or off independently, the parameters
 \name{mom_conservation} $=$  \texttt{.true.} and \name{ene_conservation} $=$  \texttt{.true.}.  However it must be noted that for energy conservation to machine accuracy, the parameter \name{mass_conserve} $=$  \texttt{.true.} must also be set, to enforce conservation at the velocity boundaries (although this may not be the most physical boundary condition).

In GKW the $\mu$-direction grid is non-uniform which represents a problem when differencing the collision terms, however it is
seen that while non-uniform in $\mu$ it is uniform in the $v_{\perp}$ coordinate ($mu = v_{\perp}^2/2B$) and as such the collision operator 
is solved using this as the coordinate as this ensures machine accurate flux conservation and momentum conservation.  To get the collision operator
in this form the following simple transformation must be performed:
\begin{eqnarray}
2{\mu}B &=& v_{\perp}^{2}\\
2\mu\partial/\partial\mu &=& \partial/\partial{v_{\perp}}.
\end{eqnarray}

\section{Collisions tutorial}

In the previous section we derived the form of the three terms of the Fokker-Plank operator.  
Here we will show, using data produced by GKW and our implementation of the previous to show what effect the operator has on the velocity space structure.

When running GKW, collisions are included if \name{collisions} $=$  \texttt{.true.} in the CONTROL namelist

\subsection{Pitch-angle scattering}

This effect is switched on by setting \name{pitch_angle} $=$ \texttt{.true.} in the collisions input deck.  Pitch-angle scattering conserves momentum and therefore will have the effect of isotropising the distribution function along arcs of constant velocity $v = \sqrt{v_{\parallel}^{2} + 2\mu B}$.

If we initialise the code with narrow, shifted gaussian in velocity space, then as time progresses then pitch-angle scattering will have the effect of producing a ring in velocity space (As seen in figure \ref{pitch-angle}).

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{pitchangle.eps}
\caption{Snapshot of velocity space as three different time intervals showing the effect of pitch angle scattering on the velocity space.  Coordinates are $v_{\parallel}$ horizontally and $v_{\perp}$ vertically. All data shown in this chapter was run with 100 parallel velocity points and 50 $\mu$ points non-equally spaced.\label{pitch-angle}}
\end{center}
\end{figure}

\subsection{Energy scattering}

This effect is switched on by setting \name{en_scatter} $=$ \name{.true.} in the collisions input deck.  The energy scattering term spreads the distribution function in energy.

\subsection{Collisional friction}

This effect is switched on by setting \name{friction_coll} $=$ \name{.true.} in the collisions input deck.  The collisional friction term advects the distribution function towards the origin in velocity space.

When you combine all three effects we get an evolution of the distribution function towards a Maxwellian centred on zero and with a width of a single normalised thermal velocity.

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{fulloper.eps}
\caption{Snapshot of velocity space as four different time intervals showing the effect of all the terms in the collision operator on the velocity space.  The final state is a Maxwellian centred on zero.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.65\textwidth]{../benchmarks/collisions/CollisionsMaxwellian.eps}
\caption{Figure showing the logarithm of the distribution function as a function of $v^2$ for varying velocity space resolutions.  A reference 
Maxwellian is also plotted for comparison.}
\label{maxwellcoll}
\end{center}
\end{figure}

\section{Velocity space boundary conditions}

There are currently two options for the boundary conditions in velocity space.  The input deck parameter \name{mass_conserve} changes between open boundary conditions (when set to \texttt{.false.}, default) and zero flux boundary conditions (when set to \texttt{.true.}).  The former is more physical, but the latter is useful for testing of conservation properties to machine precision. The difference in mass conservation between the two is very small.
\begin{figure}
\begin{center}
\includegraphics[width=0.45\textwidth]{../benchmarks/collisions/MassConservation.eps}
\includegraphics[width=0.45\textwidth]{../benchmarks/collisions/MomentumCons.eps}
\caption{(left) Figure shows the electron particle mass for when (blue) flux conserving boundaries and (red) open boundaries are used.  When the former is chosen the particle number is conserved to machine precision.  
Note the y-axis scale.  (Right) Shows the parallel momentum for runs with the same initial conditions for (blue) fully conservative (green) self collisions conservative
and (red) no conservation at all.  Again, the momentum is conserved to machine precision.  }
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{../benchmarks/collisions/EnergyConvergence.eps}
\caption{Figure shows the energy in the electron distribution for three different initial conditions (green and blue had the same initial width but with a different parallel velocity sign, red was gaussian centered different initial parallel velocity).  
All three converge on the same value. (Inlay is the parallel momentum)}
\end{center}
\end{figure}

\section{User defined collision-frequencies}

The collision frequency can be defined in two ways.  Firstly by specifying \name{nref} which is the particle density in units of $10^{19}$ particles per $m^{3}$, \name{tref}, the temperature in keV and \name{rref} which is the toroidal major radius.  These parameters are what the corresponding parameters are normalised to in the code and are never usually specified.  The normalised collision frequency $\Gamma^{a/b}=\nu_{ab} R_{\rm ref}/v_{\rm tha}$ is given by,

\begin{equation} 
\Gamma^{a/b}_N = {R_{\rm ref} n_b Z_a^2 Z_b^2 e^4 \ln \Lambda^{a/b} \over 4 \pi \epsilon_0^2 m_a^2 v_{tha}^4} = 
6.5141\cdot 10^{-5} {R_{\rm ref} n_{\rm ref}^{19} \over (T_{\rm ref}^k)^2}  {n_{Rb} Z_a^2 Z_b^2 Z^{a/b}_{\rm eff}\ln \Lambda^{a/b} 
\over T_{Ra}^2}
\end{equation}
where $Z^{a/b}_{\rm eff} = 1.0 $ unless $b = i$ (main ion species) as determined by the species density and $a = e$ as determined by the charge on the species.
The value of $Z^{e/i}_{\rm eff}$ is adjusted by the parameter \name{zeff} in the collisions namelist which is 1.0 by default. 
Note that if impurity species are included kinetically, then the scaling factor will be automatically modified to be 
\begin{equation}
 Z^{e/i}_{\rm eff} = 1 + (\name{zeff} - Z_\textrm{eff}^\textrm{sp})\frac{n_e}{n_iZ_i^2}
\end{equation}
with $Z_\textrm{eff}^\textrm{sp}=\frac{1}{n_e}\sum_s n_sZ_s^2$ and the sum on 's' performed on all ion species

However if \name{freq_override} $=$ \texttt{.true.} then,

\begin{equation}
\Gamma^{a/b}_N =  Z_a^2 Z_b^2 \nu_{\rm ref} \frac{n_{Rb}}{T_{Ra}^2} \frac{\ln \Lambda^{a/b}}{\ln \Lambda^{i/i}}
\end{equation}

Where $\nu_{\rm ref}$ is the user defined frequency defined as \name{coll_freq} in the input deck.  This is assumed to be the singly charged ion - ion (proton or deuteron) collision frequency.  To get agreement with the first method this can be set to $6.5141\cdot 10^{-5}R_{\rm ref}n_{\rm ref}^{19}\ln \Lambda^{i/i}/(T_{\rm ref}^k)^2$.  This value is then scaled to all the other species respectively.  For this to work the first species in the input deck should be singly charged ions.

\section{Direct benchmarks}

Here outlined is a direct analytical benchmark of the collision operator. The frictional slowing down of a particle can be calculated 
analytically, and accuratly approximated in the code by using a very narrow, high energy gaussian in velocity space.  The velocity
of this pulse can then be tracked and compared with the analytic expression.

The slowing down time of a particle can be written as:
\begin{equation}
\tau^{s} = - \frac{U}{\frac{\partial U}{\partial t}}
\end{equation}
where $\tau^{s} = 1/\nu_{S}$, which is defined as (we consider only ion-ion collisions):

\begin{equation}
\nu^{a/b}_{S} = - \left[ (1 + \frac{m_{a}}{m_{b}})\psi(x)\right]\nu_{0}^{a/b}
\end{equation}
where $\nu_{0}^{a/b}=\Gamma^{a/b}_N = {R_{\rm ref} n_b Z_a^2 Z_b^2 e^4 \ln \Lambda^{a/b} \over 4 \pi \epsilon_0^2 m_a^2 v_{a}^3} $,
where $\psi(x)$ is the Maxwell integral and can be shown to be $\psi(x) = erf(x) - x \frac{\partial erf(x)}{\partial x}$.  
and x is defined as $v^{a}/v_{th}^{b}$.  This ordinary differential equation can then be solved using standard methods.  The comparison between
GKW and this result is shown in Fig.~\ref{slowdown}.
\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.60\textwidth]{../benchmarks/collisions/SlowingDownComparison2.eps}
\caption{Measurement of the velocity of a highly localised pulse in velocity space as a function of time.  The result from GKW is shown
utilising two methods, one is the first moment of the distribution function, while the other is simply a trace of the spike.  The blue line is 
the solution of the ODE described above.}
\label{slowdown}
\end{center}
\end{figure} 
Agreement between the two curves is very good, implying that the friction and energy scatter terms are operating correctly
in the current implementation.  A benchmark of the pitch angle scattering against another code is presented in Sec. \ref{linearbenchmarks}.

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/diagnostics.tex
(1) Added gkw Clone gwk  #3

\chapter{Diagnostics}

In GKW, each diagnostic constitutes a separate module \File{diagnos_foobar}. The distinction
between modules is mainly made by the physical quantities they compute
- and usually not their dimensionality.

A template with some typical code snippets and comments on writing new
diagnostics can be found in \File{diagnos_template.F90}.

\section{Mode growth rates and frequencies (module \texttt{diagnos\_growth\_freq})}
For a linear run, the amplitude is defined as:
\begin{equation}
\mathcal{A}(k_\psi,k_\zeta,t) = \sqrt{\frac{\sum_{k_\psi,k_\zeta} \int |\hat \phi((k_\psi,k_\zeta,t)|^2 + |\hat A_\parallel((k_\psi,k_\zeta,t)|^2 + |\hat B_\parallel((k_\psi,k_\zeta,t)|^2 \, {\rm d}s}{\int \textrm{ds}} }
\end{equation}
For \name{mode_box=F}, there is only one $k_\psi$ and one $k_\zeta$ included in the simulation and the $s$ integral spans several poloidal turns. The linear mode amplitude is computed in the \texttt{normalise} module at each time step and used to compute the mode growth rate:

\begin{equation}
\gamma(t)= \frac{1}{\Delta t}\ln \left[ \frac{\mathcal{A}(t)}{\mathcal{A}(t-\Delta t)} \right] 
\end{equation}

The corresponding total mode frequency is calculated as
%\begin{equation}
%\omega(t)=\frac{1}{\Delta t}\left(\frac{\sum_s |\hat \phi(t)| \mathrm{Arg}({\hat \phi}(t))}{\sum_s|\hat \phi(t)|} - \frac{\sum_s |\hat \phi(t-\Delta t)|\mathrm{Arg}({\hat
%\phi}(t-\Delta t))}{\sum_s |\hat \phi(t-\Delta t)|}\right).
%\end{equation}
\begin{equation} 
\omega(t)=\frac{1}{\Delta t}\left[\mathrm{Arg}\left(\sum_{k_\psi,k_\zeta} \int \hat \phi(t) + \hat A_\parallel(t) + \hat B_\parallel(t) {\rm d}s  \right)-
\mathrm{Arg}\left(\sum_{k_\psi,k_\zeta} \int \hat \phi(t-\Delta t) + \hat A_\parallel(t-\Delta t) + \hat B_\parallel(t-\Delta t) {\rm d}s\right)\right] .\
\end{equation}
These values are output in columns 2 and 3 of \File{time.dat} respectively.
Positive values of $\omega(t)$ indicate propagation in the direction of $B \times \nabla B$ (opposite to $\nabla \zeta$), which for $s_{\rm j}=1$ and $R/L_n>0$ is in the ion diamagnetic drift direction.
In a run with \name{mode_box=F} (including nonlinear runs), per mode growth rates and frequencies are also produced in a similar way in the files \File{growth.dat} and \File{frequencies.dat}.  For these
diagnostics, however, only the $k_\psi$ modes connected to the $k_\psi = 0$ by the condition Eq. \eqref{eq:mb-connect} mode are included in the sum, which makes these mode
properties directly comparable to a run with \name{mode_box=F} and $\theta_0$=\name{CHIN=0}.

%Note that in an electromagnetic run, each sum in the growth rate and mode frequency expressions above becomes a
%double sum, the additional sum being over the two fields $\hat \phi$ and $\hat A_{\parallel}$.

In the case of runs with \name{mode_box=T}, the values above (always written to \File{time.dat}) correspond to a global growth rate and frequency
summed over all modes, which usually corresponds closely to the fastest growing mode.  In addition, the growth rate and
frequencies for each individual mode connected to $k_\psi = 0$ are also calculated and written to the files \File{growth.dat} and
\File{frequencies.dat}.

For a normalised linear run, the solution is normalised by $\mathcal{A}(k_\psi,k_\zeta,t)$.

\section{Fluxes (module \texttt{diagnos\_fluxes})}
\subsection{Co-moving frame}
The fluxes are calculated as guiding centre fluxes (flux surface averaged) and are given by the equations 
\begin{equation} 
\{\vec{\Gamma}_i \cdot \nabla \psi\} = \Gamma_i^\psi  = \biggl \{ \int {\rm d}^3 {\bf v} \tilde {\bf v}_E\cdot \nabla \psi  \alpha_i f \biggr \} + \biggl \{ \int {\rm d}^3 {\bf v} \tilde {\bf
v}_{\delta B}\cdot \nabla \psi  \alpha_i f \biggr \} 
+ \biggl \{ \int {\rm d}^3 {\bf v} \tilde {\bf v}_{\nabla B_{1 \parallel}} \cdot \nabla \psi  \alpha_i f \biggr \}
\end{equation}
for the electro-static and electro-magnetic (due to magnetic flutter and compression) fluxes, respectively.  
In the above equation all quantities are dimensional and unnormalised 
(with the exception of the dimensionless coordinate $\psi=\epsilon$), 
and $i = (1,2,3)$ with $\alpha_i = (1,m_sv^2/2,\frac{s\ind{B}RB_t}{B}m_sv_\parallel)$, i.e.\ $i = 1$ is the normalised particle flux, $i = 2$ is the heat flux, $i = 3$ is the toroidal angular momentum
flux (with the toroidal flow counted positive in the toroidal direction, clockwise from above).  The toroidal angular momentum is the quantity that is conserved in tokamaks,  and toroidal angular momentum flux is what enters the 1D transport equation (note that the contributions of the Maxwell and Reynold stresses are missing here). 
This flux is positive for outward flux of plasma rotating in the toroidal direction (clockwise from above), 
which differs from the parallel direction when $s_B = -1 $.

% add the expansion of the terms as discussed on r3223
% factor 2 missing the fluxes below ?
Both the potential and the distribution function are represented as a sum over 
Fourier modes.  Since the flux contains an average over space it will be zero unless 
the wave vectors of the potential and distribution function add up to be zero. In the 
representation of modes used in this manuscript this means that a particular wave 
vector of the representation of $f$ must be combined with the complex conjugate of 
the same wave vector in the representation of $\phi$ or $A_{\parallel}$ (sec. \ref{sec.parseval-correction}), i.e.\  
\begin{equation} 
\langle \hat v_E f \rangle  = 2 \sum_m {\rm Re}  [ \langle \hat {\bf v}_{Em}^\dagger 
\cdot \nabla \psi  \hat f_m \rangle ],
\end{equation} 
where 
$\sum_m = \sum_{k_{\zeta,{\rm min}}}^{k_{\zeta,{\rm max}}} \sum_{-k_{\psi,{\rm max}}}^{k_{\psi,{\rm max}}}$ is the sum over the modes as kept in the code
(noting that zero-mode does not contribute to the fluxes), 
$\hat v_{Em}$ and $\hat f_m$ are the Fourier amplitudes of the $E \times B$ velocity and distribution function of the mode $m$ respectively, 
and the dagger denotes the complex conjugate.
Substituting the various definitions one can derive the expression for the normalised fluxes (electro-static and electro-magnetic) 
% Factors of two finally added below.  See issue 237
% should correspond to what is implemented the code.
\begin{align}
\label{eq.fluxes-def}
{\cal I}_i &=& \sum_m \biggl \{ 2 {\cal E}^{\psi \zeta} k_\zeta \int 2 \pi B ~{\rm d} 
\mu {\rm d} v_\parallel \,  \hat{\alpha}_i {\rm Im} [ \langle \hat \phi_m \rangle^\dagger \hat f_m ] \biggr \} = \sum_m \biggl \{ 
\bar{\cal I}_i(k_\psi,k_\zeta,s) \biggr \},\\
{\cal J}_i &=& -2 \sum_m \biggl \{ 2 {\cal E}^{\psi \zeta} k_\zeta \int 2 \pi B ~{\rm d} 
\mu {\rm d} v_\parallel \,  \hat{\alpha}_i v_R v_\parallel {\rm Im} [ \langle \hat A_{\parallel m} \rangle^\dagger \hat f_m ] \biggr \} = \sum_m
\biggl \{ \bar{\cal J}_i(k_\psi,k_\zeta,s) \biggr \},\\
{\cal K}_i &=& 2 \sum_m \biggl \{ 2 {\cal E}^{\psi \zeta} k_\zeta \int 2 \pi B ~{\rm d} 
\mu {\rm d} v_\parallel \,  \hat{\alpha}_i \frac{\mu T_R}{Z_{sp}} {\rm Im} [ \langle \hat B_{\parallel m} \rangle^\dagger \hat f_m ] \biggr \} =
\sum_m \biggl \{ \bar{\cal K}_i(k_\psi,k_\zeta,s) \biggr \}
\end{align}
where here all quantities are normalised and $\sum_m \big \{\big \}=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} {\rm d}s$ is a sum over all modes 
and a flux surface average over $s$, and where $\hat{\alpha}_i = (1, v^2,\frac{s\ind{B}RB_t}{B}v_\parallel)$.  
(In the code, the factor $2\pi$ from the velocity space Jacobian is included in the array \name{intmu}=$2\pi \Delta \mu$).

The fluxes in physical units can be obtained from 
\begin{equation} 
R_{\rm ref}\Gamma^\psi_s  = n_{R_0,s} \rho_*^2 v_{\rm thref} ({\cal I}_1+{\cal J}_1+{\cal K}_1),
\end{equation} 
\begin{equation} 
R_{\rm ref}Q_s^\psi = n_{R_0,s} T_s \rho_*^2 v_{\rm thref} ({\cal I}_2+{\cal J}_2+{\cal K}_2) ,
\end{equation} 
%parallel velocity flux: Only valid in s-alpha geometry
%\begin{equation} 
%R_{\rm ref} \Gamma^\psi_{\phi s} = m_s n_{R_0,s} v_{ths} \rho_*^2 v_{\rm thref} ({\cal I}_3+{\cal J}_3+{\cal K}_3) .
%\end{equation}
\begin{equation} 
R_{\rm ref}\Pi^\psi_{\phi s} = m_s n_{R_0,s} v_{ths} R_{\rm ref} \rho_*^2 v_{\rm thref} ({\cal I}_3+{\cal J}_3+{\cal K}_3),
\end{equation}
for particle, heat, and angular momentum fluxes respectively ($\Gamma^\psi_s,Q_s^\psi,\Pi^\psi_{\phi s}$ are the contravariant components of the flux and therefore have units of particle, heat, angular momentum flux per unit length respectively).

Since for a normalised linear run, the solution is normalised by $\mathcal{A}(k_\psi,k_\zeta,t)$, the outputs in \File{fluxes.dat} correspond to
\begin{equation}
 ( {\cal I}_i, {\cal J}_i, {\cal K}_i ) / \mathcal{A}^2
\end{equation}
and can therefore be used directly in quasi-linear models.  When comparing with nonlinear or unnormalised runs (at least in the linear phase), 
these quasi-linear normalised values in \File{fluxes.dat} should be directly comparable to the values in the fluxes spectra divided by 
the sum of the per mode values in the files \File{kyspec} and \File{kyspec_em}.

Note:  In some GKW publications the flux of parallel momentum $\Gamma_\parallel$ has been used.  This is meaningful only for the $s-\alpha$ geometry in which the finite $\epsilon$ terms are neglected, in which case $\Pi = R_{\rm ref} \Gamma_\parallel $.

Note that when calculating momentum diffusivity,
the definition of $u^\prime$ relative to $v_{\rm thref}$ means that the factor $v_R=v_{thi}/v_{\rm thref}$ must be inserted if $T_{\rm ref} \ne T_i$. The Prandtl number (for example) is therefore calculated as
\begin{equation}
Pr = {{\cal I}_3 \over {\cal I}_2} {R/L_T \over {u^\prime}} v_R.
\end{equation}

The suprema of the fluxes (for which we use the notation ${\hat {\cal I}_i}, {\hat {\cal J}_i}, {\hat {\cal K}_i}$), 
are defined the same way as the fluxes in equation \ref{eq.fluxes-def} etc, 
but replacing $\hat f_m \rightarrow -i |\hat f_m|$ and $\hat \phi \rightarrow |\hat \phi|$.  The fluxes suprema represent 
the largest possible fluxes for the fluctuation amplitudes, which would occur if the fluctuations were out of phase by $\pi/2$.
The ratio between the fluxes and the flux suprema
\begin{equation}
\frac{{\cal I}_i}{\hat {\cal I}_i} = \sin \{ \hat \theta \}
\end{equation}
then contains the cross phase information, where $\{ \hat \theta \}$ is the average phase angle between the fluctuating quantites in the flux (note, 
that from this ratio one cannot distiguish phase angles outside the principal value).

\subsection{Laboratory frame}
The conversion from the co-moving frame (superscript ''co'') to the Laboratory frame (superscript ''L'') is obtained from the following transformations:
\begin{align}
 \mathbf{v}^\textrm{co} &=& \mathbf{v}^\textrm{L} -\mathbf{u_0}\\
 f^\textrm{co}(\mathbf{v}) &=& f^\textrm{L}(\mathbf{v}+\mathbf{u_0})
\end{align}
which in particular implies $\mathbf{v}_\chi^\textrm{co}\cdot \nabla \psi =\mathbf{v}_\chi^\textrm{L}\cdot \nabla \psi$. Applying these transformations to the (unnormalised) fluxes yields:
\begin{align}
\{\vec{\Gamma}_1 \cdot \nabla \psi\}^\textrm{L} &=& \{\vec{\Gamma}_1 \cdot \nabla \psi\}^\textrm{co} \\
\{\vec{\Gamma}_2 \cdot \nabla \psi\}^\textrm{L} &=& \{\vec{\Gamma}_2 \cdot \nabla \psi\}^\textrm{co} + \biggl \{ \int {\rm d}^3 {\bf v} \tilde {\bf v}_\chi\cdot \nabla \psi \, m_sv_\parallel\Omega\frac{RB_t}{B} f_s \biggr \} + \biggl \{ \int {\rm d}^3 {\bf v} \tilde {\bf v}_\chi\cdot \nabla \psi \, \frac{1}{2}m_s(R\Omega)^2 f_s \biggr \} \\
\{\vec{\Gamma}_3 \cdot \nabla \psi\}^\textrm{L} &=& \{\vec{\Gamma}_3 \cdot \nabla \psi\}^\textrm{co} + \biggl \{ \int {\rm d}^3 {\bf v} \tilde {\bf v}_\chi\cdot \nabla \psi \, s_bm_s\Omega\left(\frac{RB_t}{B}\right)^2 f_s \biggr \}
\end{align}
and the corresponding normalised components are modified as follows:
\begin{equation}
(\mathcal{I},\mathcal{J},\mathcal{K})_i^\textrm{L} = (\mathcal{I},\mathcal{J},\mathcal{K})_i^\textrm{co} + (d\mathcal{I},d\mathcal{J},d\mathcal{K})_i
\end{equation}
with
\begin{align}
(d\mathcal{I},d\mathcal{J},d\mathcal{K})_1 &=& 0 \\
(d\mathcal{I},d\mathcal{J},d\mathcal{K})_2 &=&  2s_bu\frac{v_\textrm{thref}}{v_\textrm{ths}}(\mathcal{I},\mathcal{J},\mathcal{K})_3^\textrm{co} + \frac{1}{2}m u^2\frac{T_\textrm{ref}}{T_s}\int_{s} {\rm d}s R^2 \frac{\partial }{\partial s}(\mathcal{I},\mathcal{J},\mathcal{K})_1^\textrm{co}\\
(d\mathcal{I},d\mathcal{J},d\mathcal{K})_3 &=& s_bu\frac{v_\textrm{thref}}{v_\textrm{ths}} \int_{s} {\rm d}s \left(R\frac{B_t}{B}\right)^2 \frac{\partial }{\partial s}(\mathcal{I},\mathcal{J},\mathcal{K})_1^\textrm{co} 
\end{align}

In the experiments, the momentum flux is usually computed in the Laboratory frame (i.e. it includes the momentum carried by the particle flux) and should therefore be compared to $\{\vec{\Gamma}_2 \cdot \nabla \psi\}^\textrm{L}$. The experimental heat flux, however, usually does not include the convective contributions and should therefore be compared to  $\{\vec{\Gamma}_3 \cdot \nabla \psi\}^\textrm{co}$ (i.e. experimentally one is more interested in the thermal energy flux than the total kinetic energy flux since the primary interest is on the achievable temperature).
The total kinetic energy flux $\{\vec{\Gamma}_3 \cdot \nabla \psi\}^\textrm{L}$ is of potential interest for comparisons with codes giving the Laboratory frame fluxes in output.

\section{Fluxes in more detail (module \texttt{diagnos\_fluxes\_vspace})}

There is the possibilibty to output the unintegrated fluxes by setting
the switch \name{lfluxes_detail}, with contributions of all field
fluctuations which are considered by the run.

\begin{align}
\label{eq.fluxes-detail}
\bar{\cal I}_i(k_\psi,k_\zeta,s,\mu,v_\parallel) &=& 2 {\cal E}^{\psi \beta} k_{\beta m }  2 \pi B ~{\rm d} 
\mu {\rm d} v_\parallel \,  \hat{\alpha}_i {\rm Im} [ \langle \hat \phi_m \rangle^\dagger \hat f_m ] \ud^3X\\
\bar{\cal J}_i(k_\psi,k_\zeta,s,\mu,v_\parallel) &=& -2 \cdot 2 {\cal E}^{\psi \beta} k_{\beta m } 2 \pi B ~{\rm d} 
\mu {\rm d} v_\parallel \,  \hat{\alpha}_i v_R v_\parallel {\rm Im} [ \langle \hat A_{\parallel m} \rangle^\dagger \hat f_m ] \ud^3X \\
\bar{\cal K}_i(k_\psi,k_\zeta,s,\mu,v_\parallel) &=& 2 \cdot 2 {\cal E}^{\psi \beta} k_{\beta m } 2 \pi B ~{\rm d} 
\mu {\rm d} v_\parallel \,  \hat{\alpha}_i \frac{\mu T_R}{Z_{sp}} {\rm Im} [ \langle \hat B_{\parallel m} \rangle^\dagger \hat f_m ] \ud^3X
\end{align}

\section{Mode structure (module \texttt{diagnos\_mode\_struct}) \label{sec.parallel}}
The file \File{parallel.dat} contains some diagnostics connected with the parallel mode structure. Its output is controlled with the switch \name{lparallel_output}. 
For every mode then species the following quantities are (sequentially) written to file, in this column order:
\begin{small}
\begin{verbatim}
1     sgr         the length along the field line 
2-3   phi         the perturbed potential 
4-5   apar        the parallel component of the vector potential 
6-7   dens        the perturbed density 
8-9   epar        the perturbed parallel energy 
10-11 eperp       the perturbed perpendicular energy 
12-13 wflow       the perturbed parallel flow velocity
14-15 bpar        the perturbed parallel (compressional) magnetic field
\end{verbatim}
\end{small}
Note only \name{sgr} is real. The other quantities are complex and fill 2 columns in the output file. For a normalised run,
all quantities are normalised and rotated in the complex plane so that the maximum of the potential for each mode is $1+0i$.  
Therefore the output quantity is ${\hat l}_E(s)$ where
\begin{equation}
{\hat l}(s,t)=\exp(\gamma t + i \omega t){\hat l}_E(s).
\end{equation}
The file will have $N_{\rm MOD} N_x N_s N_{\rm sp}$ rows (with data ordered by $\zeta$ mode, then $\psi$ mode, then species). 
For a run with multiple modes, extracting data of interest can be a little involved, matlab script \matlab{paralleldat.m}, helps with this.\\

\section{Mode structure (modules \texttt{diagnos\_fields} and \texttt{diagnos\_moments}) \label{sec.kykxs}}
For spectral runs, the parallel structure of the fields and of various moments of the distribution function can be given in output. These diagnostics are controlled by the switches \name{kykxs_phi}, \name{kykxs_apar}, \name{kykxs_bpar}, \name{kykxs_moments}, \name{kyks_j0_moments} and \name{kyks_j1_moments}. The output occurs at every large time step if the meta-switch \name{xy_estep} is true and otherwise at the last time step only. 
The fields ($\phi$, $A_\parallel$ and $B_\parallel$) and the moments are given as a function of the $k_\zeta$, $k_\psi$, $s$ and $t$. 
One binary file is created per quantity and time step, for the real and imaginary parts. The file name convention for the fields is \File{'prefix'_kykxs'file_number'_'suffix'} with the prefix being one of \File{Phi}, \File{Apa} or \File{Bpa} and the suffix one of \File{real} or \File{imag}. The file number is incremented at each large time step and its value stored in \File{file_count.dat} for keeping track of the correspondance with the simulation time base. The file name convention for the moments is 
\File{'prefix'_kykxs'species_number'_'file_number'_'suffix'} where the species number has been added since the moments are species dependent. The various moments of the gyrocenters distribution function are defined as follows:
\begin{equation}
\mathcal{\hat{M}}_s(k_\zeta,k_\psi,s,\alpha_s) = \int \alpha_s(v_\parallel,v_\perp)\hat{f}_s\,\mathbf{\textrm{d}v}
\end{equation}
with $\mathbf{\textrm{d}v}=2\pi B_N\textrm{d}\mu_N\textrm{d}v_{\parallel N}$ and $\alpha_s(v_\parallel,v_\perp)$ a generic prefactor whose values are listed in the table below. 
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
File prefix & $\alpha_s$ & Namelist switch \\
\hline
\File{dens} & $1$ &  \\
\File{vpar} & $v_\parallel$ &  \\
\File{Tpar} & $v_\parallel^2$ &  \\
\File{Tperp} & $\frac{1}{2}v_\perp^2$ & \name{kykxs_moments} \\
\File{Qpar} & $v_\parallel v^2$ &  \\
\File{M12}  & $v_\parallel v_\perp^2$ & \\
\File{M24}  & $v_\perp^2 v^2$ &  \\
\hline
\File{dens_ga} & $J_0$ &  \\
\File{vpar_ga} & $v_\parallel J_0$ &  \\
\File{Tpar_ga} & $v_\parallel^2J_0$ &  \\
\File{Tperp_ga} & $\frac{1}{2}v_\perp^2J_0$ & \name{kykxs_j0_moments} \\
\File{Qpar_ga} & $v_\parallel v^2 J_0$ &  \\
\File{M12_ga}  & $v_\parallel v_\perp^2 J_0$ & \\
\File{M24_ga}  & $v_\perp^2 v^2J_0$ &  \\
\hline
\end{tabular}
\end{center}
Note that the prefixes \File{Tpar} and \File{Tperp} are misleading since the computed quantities are not the perturbed temperatures, but related to the perturbed parallel and perpendicular pressure. 
More precisely, the data in the file \File{Tpar} is $\frac{\hat{p}_\parallel}{2T_sn_{s}}$ and that in the file \File{Tperp} is $\frac{\hat{p}_\perp}{2T_sn_{s}}$ (here $\hat{p}_\parallel$ and $\hat{p}_\perp$ are unnormalised) 
The total perturbed pressure obtained for $\alpha_s=\frac{1}{3}m_sv^2$ is given by $\hat{p}=\frac{1}{3}(\hat{p}_\parallel+2\hat{p}_\perp)$.

\section{Hermitian symmetry of binormal spectra}
%\section{Conjugate represnetation of binormal spectra}
%\section{A note about integrals over binormal spectra}
\label{sec.parseval-correction}

Many diagnostics calculate quantities which are quadratic in the
perturbations and output them either as $k_\zeta$ spectra or
integrated over $k_\zeta$.  According to what is often called
Parseval's Theorem one can integrate the Fourier-space field.
\begin{align}
  \int \ud\zeta |f(\zeta)|^2 &= \int \ud k_\zeta |\hat f(k_\zeta)|^2 \nonumber\\
  \int \ud\zeta f(\zeta)\phi(\zeta) &= \int \ud k_\zeta \hat f^*(k_\zeta) \hat\phi(k_\zeta)
\end{align}
The Fourier representation of a real-valued field possesses Hermitian symmetry, as
\begin{align}
  \hat f(k_\zeta) = \frac{1}{2\pi} \int\ud\zeta f(\zeta) e^{i k_\zeta \zeta}
\end{align}
one finds
\begin{align}
  \hat f(-k_\zeta) = \frac{1}{2\pi} \int\ud\zeta f(\zeta) e^{i (-k_\zeta) \zeta} = \hat f^*(k_\zeta) \qquad\text{and also}\quad \hat f(-k_\zeta, -k_\psi) = \hat f^*(k_\zeta, k_\psi)
\end{align}
This symmetry allows the FFT libraries and GKW to save unnecessary
memory for $k_\zeta$-space quantities. For a position space grid
having \name{n\_y\_grid} elements, the Fourier space representation
has in principle also \name{n\_y\_grid} coefficients, but due to the
symmetry, only
$\name{floor}(\frac{\name{n\_y\_grid}}{2}) + 1 = \name{nmod}$ of them
are stored in memory (Note that in contrast to this, radially spectral
quantities have \name{n_x_grid} elements).  When it comes to
integration, the sum $\int\ud k_\zeta$ is to be taken over all
\name{n_y_grid} coefficients, however.
\begin{align}
  \int \ud k_\zeta \hat f^*(k_\zeta) \hat\phi(k_\zeta) &\hat{=} 
\hat f^*(0) \hat\phi(0) + 
\sum_{-k_{\zeta,max}}^{-\Delta k_\zeta}\hat f^*(k_\zeta) \hat\phi(k_\zeta) + 
\sum_{\Delta k_\zeta}^{k_{\zeta,max}}\hat f^*(k_\zeta) \hat\phi(k_\zeta)  \nonumber\\
&\hat{=} 
\hat f^*(0) \hat\phi(0) + 
\sum_{k_{\zeta,max}}^{\Delta k_\zeta}\hat f^*(-k_\zeta) \hat\phi(-k_\zeta) + 
\sum_{\Delta k_\zeta}^{k_{\zeta,max}}\hat f^*(k_\zeta) \hat\phi(k_\zeta)  \nonumber\\
&\hat{=} 
\hat f^*(0) \hat\phi(0) + 
\sum_{k_{\zeta,max}}^{\Delta k_\zeta}\left(\hat f^*(k_\zeta) \hat\phi(k_\zeta)\right)^* + 
\sum_{\Delta k_\zeta}^{k_{\zeta,max}}\hat f^*(k_\zeta) \hat\phi(k_\zeta)  \nonumber\\
&\hat{=} 
\hat f^*(0) \hat\phi(0) + 
\sum_{k_{\zeta,max}}^{\Delta k_\zeta} 2 \Re\left\{\hat f^*(k_\zeta) \hat\phi(k_\zeta)\right\} 
\end{align}
Hence an integral over a modulus square has to be computed as
\begin{align}
  \int \ud k_\zeta |f(k_\zeta)|^2 &\hat{=} 
\hat f^*(0) \hat f(0) + 
\sum_{k_{\zeta,max}}^{\Delta k_\zeta} 2 |\hat f(k_\zeta)|^2
\end{align}
To simplify the code and to make the factor 2 explicit,
the array \name{parseval_correction} was introduced. Here is an
example of how it is used:
\begin{verbatim}
  real :: integral = 0.0
  integer :: ix = 1, i = 2, j = 3, k = 4, is = 1
  ! integrate |f|**2 over zeta, with all other coordinates fixed to some value
  do imod = 1, nmod
    integral = integral + parseval_correction(imod) * 
                 abs(get_f_from_g(imod,ix,i,j,k,is,fdisi))**2
  end do
\end{verbatim}

Several diagnostics output binormal spectra. Note that some of these
contain the factor 2 (e.g. the entropy \name{entr} from
\name{diagnos_energetics}), while others do not (e.g. the intensity
spectrum \name{kyspec} from \name{diagnos_fields}).  While for some
considerations the factor 2 in front of the non-zonal-mode
contributions may not matter much for the qualitative picture, it is
usually important to consider it if the quantity has to fulfill
certain consistency conditions.

\section{Tips for unit conversions}

Although GKW uses the thermal velocity ($ \sqrt{2 T /m}$) for normalisation, it is common 
in the literature to use the gyro-Bohm (GB) units of sound speed $c_s = \sqrt{T_e / m_i}$ and the
corresponding Larmor radius $\rho_s = c_s / \omega_{ci}$, where $\omega_{ci} = e B_{\rm ref} / m_i$ is the ion
cyclotron frequency, and the index i (e) corresponds to the ions (electrons).  
The user should note that the definition of $B$ used to define $c_s$ and $\rho_s$ may vary between codes and publications.

In the usual case where
$m_i=m_{\rm ref}$, the sound speed is $c_s=v_{\rm
thref}\sqrt{T_{Re}/2}$, the Larmor radius is $\rho_s=\rho_{\rm ref}\sqrt{T_{Re}/2}$, and the GKW
wavevectors have $k_\perp \rho_{\rm ref} = k_\perp \rho_s\sqrt{T_{Re}/2}$.
\footnote{Care should also be taken when comparing the GKW input wavenumber ``$k_\theta$''
with other codes. It is defined by Eq. \ref{kperp}, and the relationship
to the toroidal mode number is described by Eq. \ref{modenumber}.}

The time normalisation converts to GB units as
\begin{equation}
t_N = {v_{\rm thref} \over R_{\rm ref}} t = \sqrt{2 \over T_{Re}}{c_s \over R_{\rm ref}}
 = {\sqrt{2 \over T_{Re}} a \over R_{\rm ref}} \left({c_s \over a}\right) t ,
\end{equation}
and the heat fluxes convert to a gyro-Bohm diffusivity as 
\begin{equation}
\chi^N={{\cal I}_2 \over R_{\rm ref}/L_T \{g^{\psi \psi N}\}}={R_{\rm ref} \over \rho_{\rm ref}^2 v_{\rm thref}} \chi 
= \left({T_{Re} \over 2}\right)^{3/2}{R_{\rm ref} \over a}\left({a \over \rho_s^2 c_s}\right) \chi .
\end{equation}
%where $\chi_{GB} = {\rho_s^2 c_s/ a}$
Note that $N$ in the two previous expressions is used to denote a normalised quantity, as used in most of this
manual, and the equivalent quantity without the $N$ is a dimensional one.  The dimensional diffusivity in $m^2 /s$ 
is given by \begin{equation}
\chi=\frac{{\cal I}_2  \rho_*^2 v_{\rm thref} R_{\rm ref}}{\{g^{\psi \psi N}\} R_{\rm ref}/L_T}
\end{equation}
and a dimensional power flux in $J/s$ is obtained (see \ref{sec.transport}) by
\begin{equation}
Q = \{\vec{\Gamma}_2 \cdot \nabla \psi\} V^\prime
\end{equation}
where $V$ is the dimensional volume of the plasma with the local flux surface (see \ref{sec.fsa}) and
\begin{equation}
V^\prime = {\partial V \over \partial \epsilon} = R_{\rm ref}^3 J^N_{\psi \zeta s}.
\end{equation}

In the flux tube, one should realize that the \textit{local} radial coordinate $x_r$ appearing as the scale for the XY diagnostics (in file \File{xphi}) differs by a factor of $\rho_*$ from the \textit{global} radial coordinate $\psi=\epsilon$.  Thus $\Delta x_r = \Delta \psi / \rho_* = \Delta r / \rho_{\rm ref}$.  This is required when combining background and perturbed quantities in the flux tube to create a profile as in Sec. \ref{profiles}.

\section{Output grids (module diagnos_grid)}

The $\zeta$ grid output is re-scaled to the input scale of $k_\theta
\rho_{\rm ref}$, so the two perpendicular coordinate scales are equal
(Eq.~\ref{kperp}).

\begin{figure}[hp]

 \begin{center}
 \includegraphics[width=0.3\textwidth]{directions.eps}
 \caption{\label{directions}Coordinate directions on the outboard
   plane ($s=0$) for $s_{\rm B}=1$, as output by the potential slices
   diagnostic \File{phi\#} with the coordinate labels for each point
   as output by \File{xphi} and \File{yphi} (the yphi coordinate label
   inverts the plot). For $s_{\rm B}=1$, the poloidal field is
   `upwards' ${\rm sign}(B_p \nabla \theta \cdot \nabla \zeta)= s_{\rm
     B}$. If in doubt, see Sec.~\ref{signs}.  Note that the `centre'
   of the flux tube is in the bottom left corner when output from the
   FFTs.}
 \end{center}
 \end{figure}
%This means as the matlab script mkmovie plots it.

\section{Summary: diagnostic outputs and their data format}
\label{sec.diagnos-data-format}

In GKW, most diagnostics are calculated inside the code at runtime (as
opposed to in post-processing).  The advantage of this approach is
that one does not need to store the full 5 dimensional distribution
functions at every timestep (which occupy a lot of disk space and are
difficult to transfer).  The disadvantage is that if a time dependent
diagnostic is not enabled or implemented before the run, that
information cannot later be retrieved without re-running the entire
simulation (final state diagnostics can always be retrieved by
restarting the code if the restart file is available).  One should
therefore think carefully about which diagnostics are required before
launching any expensive simulation.  One could argue that this
approach is suited to a world in which computation is cheap, but
storage and transfer are expensive.

The default set of time dependent diagnostics contains scalar time
traces (fluxes, growth rates) and 1D spectra, but no 2D or 3D
quantities.

GKW does contain a wrapper which allows to write output data
in various formats, according to the setting of \name{io_format} in
the input file.
\begin{description}
\item[\name{io_format='ascii'}] (also called 'formatted' output) All
  data is written to a collection of text files. Those can be read and
  analysed easily, and they are very portable. The ASCII format
  becomes impractical for higher dimensional and large amounts of data.

  For ascii and binary formats, diagnostic meta-data is output by the
  code to a simple text file \File{gkwdata.meta}.

\item[\name{io_format='binary'}] (also called 'raw' output) All data
  is written to binary files. Binary data consumes typically by a
  factor of about 6-10 less disk space than ascii data. Binary or hdf5
  (see below) format is therefore recommended if large amounts of data
  are produced.
\item[\name{io_format='hdf5'}] (also called 'structured' output), if
  \name{io_format='hdf5'} is set. This structured binary format is
  available if the code was compiled with the HDF5 library. If the
  HDF5 library was compiled with szip encoding support, then the GKW
  output data can be compressed (for now, there is only a hardcoded
  switch, see the comments).
  \\
%Issue 69    
  The HDF5 format allows to output diagnostic meta-data by associating
  attributes to datasets. These can be read by analysis
  scripts.  More and more metadata will be added in the near future, for
  discussions see \issue{69}. Note that
  \begin{itemize}
  \item If one tries to access HDF5 files during run, they may turn
    out to be invalid (not always though).
  \item Aborted runs may leave a corrupted HDF5 file. Usually then
    there are just no parameters in the file, sometimes datasets are
    missing and in rarer cases even the file as a whole is invalid.
  \item To produce the 3d fields and moments data and the 4d mode
    structure data, a MPI gather operation is performed and the output
    is written serially by just one process. For large runs with many
    processes this is likely not very performant. Instead, it is
    recommended to use MPI-IO binary output format then.
  \item At the time moment GKW supports only serial
    HDF5.
  \end{itemize}
\end{description}

In order to take advantage of the benefits of certain formats, GKW
knows also about mixed output:
\begin{description}
\item[\name{io_format='mixed'}] Most of the data (mainly scalar time
  traces and 1D spectra, \File{parallel.dat} final state
  eigenfunctions) is output to text files, whereas higher dimensional
  data is output to binary files.
\item[\name{io_format='hdf5+ascii'}] All data is output as with the
  'hdf5' setting. In addition, lowerdimensional data is output also to
  text files.
\end{description}

As long as there is not enough descriptive metadata attached to the
GKW diagnostic output data, we here summarize most of the GKW
diagnostic outputs in tables \ref{tab:diag-part1} and
\ref{tab:diag-part2} for 1D time traces, final state outputs, and 2D
time trace outputs.  A full listing of available diagnostics and
switches is documented in \doc{input.dat.sample}.  In the case of less
standard diagnostics it may be necessary to examine the source and
comments in the respective diagnostic module \src{diagnos_foo.F90} to
determine the exact order and meaning of the output quantities.

Generally, 2D diagnostics can be output by GKW at every timestep, or
at the end (``R or E'' in tables \ref{tab:diag-part1} and
\ref{tab:diag-part2}, switched by \name{xy_estep}). \\
When successive binary records for each timestep are stacked into one
file, this happens with (32 bit / 4 byte, usually)
integers before and after each record for the size of the binary
record (which will have Endianness of the hardware it was written on).
The data records can be
read into matlab with the script \matlab{read_file2d.m}.\\

Some slices are output at the point nearest the LFS, for
these a different location can be set with switch
\name{xy_slice_ipar}.

\paragraph*{Reorganisation of diagnostics and the \name{io_legacy} switch:}
Note that there currently efforts to clean up GKW's diagnostics. This
includes renaming of some files, split up of datasets and change of
the order of the data, where necessary. The goal is to output data in
a more natural way which is easy to understand and to analyse.

In order to provide users for some time the possibility to keep using
their old analysis tools without larger changes, the switch
\name{io_legacy} was introduced.

\begin{landscape}
\begin{table}[hp!]
\begin{footnotesize}
\vspace{-2cm}
\begin{adjustwidth}{-1cm}{-2cm}
\centering 
\begin{tabular}{l|l|l|l|l|l|l}
 Filename / dataset name & Description & Quantity (norm.=normalised) & Size 
& when
& \parbox[t]{1cm}{Grid\\Files}
& \parbox[t]{1.5cm}{Namelist\\Switch}\\
\hline
%\File{foo} & [description] & [expression] & [rows,cols,..]  &B/R/E/O & [gridfiles] & [switch]\\
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_growth_freq:}} \\
\File{dominant_growth_rate} & Dominant linear mode growth rate & $\gamma_{N}^{\rm max}$ & $N_t$  & R & & \\
\File{dominant_real_freq} & Dominant linear mode frequency & $\omega_N$ belonging to $\gamma_{N}^{\rm max}$& $N_t$  & R & & \\
\File{growth_rates.dat} & Growth rates of all linear modes & $\gamma_N$ &   & R & & \\
\File{frequencies.dat} & Frequencies of all linear modes & $\omega_N$ &   & R & & \\
\File{amplitudes.dat} & Amplitudes of all linear modes &  &  & R & & \\

\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_grid:}} \\
\File{time.dat} & timestamp (large time steps) & $t_N$ & $N_t$  & R & & \\
\File{sgrid} & s grid (parallel coordinate) & $s$ & $N_s$  & B & & \\
\File{krho} & Bi-normal wavevector grid & $(q / 2 \pi \psi) k_\zeta \rho_{\rm ref} = k_\perp(k_\psi=0,s=0) \rho_{\rm ref}$ & $N_x, N_{\rm mod}$& B & & \\
\File{kxrh} & Radial wavevector grid & $k_\psi \rho_{\rm ref}$ & $N_x, N_{\rm mod}$& B & & \\
\File{yphi} & Real space bi-normal grid & $ 2 \pi \psi (\zeta-\zeta_{\rm min})/q\rho_{\rm ref} = k_\theta \rho_{\rm ref}$ & $M_x,
M_{\rm mod}$ & B & & \\
\File{xphi} & Real space radial grid & $(\psi-\psi_{\rm min})/\rho_* = (r-r_{\rm min})/\rho_{\rm ref} $ & $M_x, M_{\rm mod}$ & B & & \\
\File{kx_connect} & Parallel boundary connections & Labelled in file & N/A& B & & \\
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_mode_struct:}} \\
\File{phi} & es. potential & See Sec. \ref{sec.parallel}. & $N_{\rm MOD},  N_s, N_x$ & E and O & & \\
\File{Apar} &  & See Sec. \ref{sec.parallel}. & $N_{\rm MOD},  N_s, N_x$ & E and O & & \\
\File{Bpar} &  & See Sec. \ref{sec.parallel}. & $N_{\rm MOD},  N_s, N_x$ & E and O & & \\
\File{dens} & density fluctuations & See Sec. \ref{sec.parallel}. & $N_{\rm sp}, N_{\rm MOD},  N_s, N_x$ & E and O & & \\
\File{Tpar} & parall. temperature fluct. & See Sec. \ref{sec.parallel}. & $N_{\rm sp}, N_{\rm MOD},  N_s, N_x$ & E and O & & \\
\File{Tperp} & perp. temp. fluctuations & See Sec. \ref{sec.parallel}. & $N_{\rm sp}, N_{\rm MOD},  N_s, N_x$ & E and O & & \\
\File{vpar} & parallel flow fluctuations & See Sec. \ref{sec.parallel}. & $N_{\rm sp}, N_{\rm MOD},  N_s, N_x$ & E and O & & \\

\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_fields:}} \\
\File{kyspec} & Bi-normal mode potential spectrum &  $\sum_{k_\psi} \int_{s} |\hat \phi(k_\psi,k_\zeta,s)|^2 \, {\rm d}s $    &  $N_t,N_{\rm mod}$& R & & \\
\File{kyspec_em} & (Equivalent for $A_\parallel$ field) & $\sum_{k_\psi} \int_{s} |\hat A_\parallel(k_\psi,k_\zeta,s)|^2 \, {\rm d}s $ & $N_t,N_{\rm mod}$ & R\\
\File{kxspec(_em)} & Radial mode potential spectrum &  $\sum_{k_\zeta} \int_{s} |\hat \phi(k_\psi,k_\zeta,s)|^2 {\rm d}s $  & $N_t,N_x$  & R & & \\
\File{kyvort} & Bi-normal vorticity spectra &  $\sum_{k_\psi,sp} \int Z_{sp}|\hat f_{sp}(k_\psi,k_\zeta,s)|^2 \, {\rm d}^3 v \, {\rm d}s $  & $N_t,N_{\rm mod}$ & R & & \\

\File{kxvort} & Radial vorticity spectra &  $\sum_{k_\zeta,sp} \int Z_{sp} |\hat f_{sp}(k_\psi,k_\zeta,s)|^2 \, {\rm d}^3 v \, {\rm d}s $  & $N_t,N_x$ & R & & \\
\hline
\File{spc\#}  & ES potential spectrum (LFS) & $|\hat \phi(k_\psi,k_\zeta,s\approx0)|$ \ (norm.)& &R or E & \File{kxrh, krho} & \name{xy_phi} \\

\File{phi\#}  & ES potential slice (LFS) &  $\phi(\psi,\zeta,s\approx0)$ \ (norm.)&  &R or E & \File{xphi, yphi} & \name{xy_phi} \\

\File{apc\#}  & EM potential spectrum (LFS) &   $|\hat A_\parallel (k_\psi,k_\zeta,s\approx0)|$ \ (norm.)& &R or E & \File{kxrh, krho} & \name{xy_apar} \\

\File{apa\#}  & EM potential slice (LFS) &  $A_\parallel(\psi,\zeta,s\approx0)$\ (norm.) & &R or E & \File{xphi, yphi} & \name{xy_apar} \\

\File{bpc\#}  & EM field spectrum (LFS) &  $|\hat B_\parallel(k_\psi,k_\zeta,s\approx0)|$ \ (norm.)& &R or E & \File{kxrh, krho} & \name{xy_bpar} \\

\File{bpa\#}  & EM field slice (LFS) &  $B_\parallel(\psi,\zeta,s\approx0)$ \ (norm.)& &R or E & \File{xphi, yphi} & \name{xy_bpar} \\

\File{vok\#}  & Vorticity spectrum (FSA) &  $\sum_{s} \int |\hat f_{s}(k_\psi,k_\zeta)| d^3 v ds$ \ (norm.)& &R or E & \File{kxrh, krho} & \name{xy_vort}\\

\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_moments:}} \\
\File{ene[sp\#]_\#}  & Temp. pert. moment (LFS) &  $\sum_{sp} \int |\hat f_{s} v^2| d^3 v $ \ (norm.)& &R or E & \File{xphi, yphi} & \name{xy_temp} \\

\File{den[sp\#]_\#}  & Density pert. moment (LFS) &  $\int |\hat f_{s}| d^3 v $ \ (norm.)& &R or E & \File{xphi, yphi} & \name{xy_dens} \\

\File{pac[sp\#]_\#}  & Current pert. moment (LFS) &  $\int |\hat f_{s} | v_\parallel d^3 v $ \ (norm.)& &R or E & \File{xphi, yphi} & \name{xy_current} \\

\File{pac[sp\#]_\#}  & Current$^2$ pert. moment (LFS) &  $\int |\hat f_{s} | v_\parallel^2 d^3 v $ \ (norm.) & &R or E & \File{xphi, yphi} & \name{xy_current2} \\

\File{{den/ene}_spectra} & \parbox[t]{3.5cm}{Bi-normal spectral perturbation moments for each species} & $\sum_{k_\psi}\int_s \int \hat \alpha_{1,2} |\hat f_{sp}| {\rm d}^3v \, {\rm d}s $ & $N_{\rm mod},N_{sp}$ & & & \\

\end{tabular}
\end{adjustwidth}
\end{footnotesize}
\caption{
  Diagnostics output by GKW (Part 1), with \name{io_legacy=false}. The
  abbreviations denote output at the beginning (B), repeatedly (R), at the end
  (E) or other (O). FSA denotes Flux Surface Averaged quantities. LFS denotes quantities evaluated at the low field side, i.e. in the center of the $s$-grid.
}
\label{tab:diag-part1}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}[hp!]
\begin{footnotesize}
\vspace{-2cm}
\begin{adjustwidth}{-1cm}{-1cm}
\centering 
\begin{tabular}{l|l|l|l|l|l|l}
 Filename / dataset name & Description & Quantity (norm.=normalised) & Size 
& \parbox[t]{2cm}{beginning (B),\\ repeatedly (R),\\ at the end (E)}
& \parbox[t]{1cm}{Grid\\Files}
& \parbox[t]{1.5cm}{Namelist\\Switch}\\
\hline
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_moments:}} \\
\File{dens_kyzero_xs[sp\#]_\#}    & Density pert. moment ($k_\zeta = 0$) &  $\int |\hat f_{sp}| d^3 v $ \ (norm.)            & &R  & \File{xphi, sgrid} & \name{xs_kyzero_dens} \\
\File{current_kyzero_xs[sp\#]_\#} & Current pert. moment ($k_\zeta = 0$) &  $\int |\hat f_{sp}| v_\parallel d^3 v $ \ (norm.)& &R  & \File{xphi, sgrid} & \name{xs_kyzero_current} \\
\File{ene_kyzero_xs[sp\#]_\#}     & Energy pert. moment ($k_\zeta = 0$)  &  $\int |\hat f_{sp}| v^2 d^3 v $ \ (norm.)& &R  & \File{xphi, sgrid} & \name{xs_kyzero_ene} \\
\File{ene_par_kyzero_xs[sp\#]_\#}     & Parallel energy pert. moment ($k_\zeta = 0$)  &  $\int |\hat f_{sp}| v_\parallel^2 d^3 v $ \ (norm.)& &R  & \File{xphi, sgrid} & \name{xs_kyzero_ene_par} \\
\File{ene_perp_kyzero_xs[sp\#]_\#}     & Perpendicular energy pert. moment ($k_\zeta = 0$)  &  $\int |\hat f_{sp}| v_\perp^2 d^3 v $ \ (norm.)& &R  & \File{xphi, sgrid} & \name{xs_kyzero_ene_perp} \\
\File{phi_ga_fm_kyzero_xs[sp\#]_\#}     & Gyro-averaged (using $F_M$) el.-stat. potential  ($k_\zeta = 0$)  &  $\int |\hat F_{M,sp}| \langle \hat \phi \rangle d^3 v $ \ (norm.)& &R  & \File{xphi, sgrid} & \name{xs_kyzero_phi_ga_fm} \\

\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_fluxes:}} \\
%\File{fluxes.dat} & Total $v_E$ fluxes $(i=1,2,3)$ by species  & ${\cal I}_i=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal I}_i \, {\rm d}s$  & $3N_{sp}$ & R & & \\
\File{p/e/vflux_es} & Total $v_E$ particle/heat/momentum flux by species  & ${\cal I}_i=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal I}_i \, {\rm d}s$  & $N_t,N_{sp}$ & R & \File{time} & \\
\File{p/e/vflux_es_lab} & Corresponding Lab frame fluxes & ${\cal I}_i+d{\cal I}_i$  & $N_t,N_{sp}$ & R & \File{time} & \\
%\File{eflux_es} & Total $v_E$ heat flux by species  & ${\cal I}_2=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal I}_2 \, {\rm d}s$  & $N_t,N_{sp}$ & R & \File{time} & \\
%\File{vflux_es} & Total $v_E$ momentum flux by species  & ${\cal I}_3=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal I}_3 \, {\rm d}s$  & $N_t,N_{sp}$ & R & \File{time} & \\
\File{p/e/vflux_apar} & Total $A_\parallel$ particle/heat/momentum flux by species  & ${\cal J}_i=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal J}_i \, {\rm d}s$  & $N_t,N_{sp}$ & R &  \File{time} & \\
\File{p/e/vflux_apar_lab} &  Corresponding Lab frame fluxes & ${\cal J}_i+d{\cal J}_i$  & $N_t,N_{sp}$ & R &  \File{time} & \\
%\File{eflux_apar} & Total $A_\parallel$ heat flux by species  & ${\cal J}_2=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal J}_2 \, {\rm d}s$  & $N_t,N_{sp}$ & R &  \File{time} & \\
%\File{vflux_apar} & Total $A_\parallel$ momentum flux by species  & ${\cal J}_3=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal J}_3 \, {\rm d}s$  & $N_t,N_{sp}$ & R &  \File{time} & \\
\File{p/e/vflux_bpar} & Total $\nabla B_\parallel$ particle/heat/momentum flux by species  & ${\cal K}_i=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal K}_i \, {\rm d}s$  & $N_t,N_{sp}$ & R &  \File{time} & \\
\File{p/e/vflux_bpar_lab} & Corresponding Lab frame fluxes  & ${\cal K}_i + d{\cal K}_i$  & $N_t,N_{sp}$ & R &  \File{time} & \\
%\File{eflux_bpar} & Total $\nabla B_\parallel$ heat flux by species  & ${\cal K}_2=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal K}_2 \, {\rm d}s$  & $N_t,N_{sp}$ & R &  \File{time} & \\
%\File{vflux_bpar} & Total $\nabla B_\parallel$ momentum flux by species  & ${\cal K}_3=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal K}_3 \, {\rm d}s$  & $N_t,N_{sp}$ & R &  \File{time} & \\

\File{{p/e/v}flux_spectra} & Bi-normal spectral $v_E$ flux for each species & $\sum_{k_\psi} \int_{s} \bar {\cal I}_i(k_\psi,k_\zeta) \, {\rm d}s$ & $N_t,N_{\rm mod},N_{sp}$ & R & & \\

\File{{p/e/v}flux_xspec}  & Radial spectral $v_E$ fluxes for each species & $\sum_{k_\zeta} \int_{s} \bar {\cal I}_i(k_\psi,k_\zeta) \, {\rm d}s$  & $N_t,N_{\rm mod},N_{sp}$   & R & & \\

\File{{p/e}flux_(em_)sup} & \parbox[t]{3.5cm}{Bi-normal spectral flux suprema for each species} & $\hat {\cal I}_i(k_\zeta), \hat {\cal J}_i(k_\zeta) $ & $N_t,N_{\rm mod},N_{sp}$ & R & & \\

\hline
\File{PFlesr[sp\#]_\#}  & ES radial particle flux (LFS) & ${\cal I}_1(\psi,\zeta,s\approx0)$\ (norm.)& & R or E & \File{xphi, yphi} & \name{xy_fluxes_P} \\

\File{EFlesr[sp\#]_\#}  & ES radial energy flux (LFS) & ${\cal I}_2(\psi,\zeta,s\approx0)$ \ (norm.)& & R or E & \File{xphi, yphi} & \name{xy_fluxes} \\

\File{VFlesr[sp\#]_\#}  & ES radial momentum flux (LFS) & ${\cal I}_3(\psi,\zeta,s\approx0)$ \ (norm.)& & R or E & \File{xphi, yphi} & \name{xy_fluxes_V} \\

\File{[P/E/V]Fl\underline{em}r[sp\#]_\#}  & EM flutter fluxes as above (LFS) &  $J_i(\psi,\zeta,s\approx0)$ \ (norm.) & & R or E & \File{xphi, yphi} & \name{+ xy_fluxes_em} \\

\File{[P/E/V]Fl\underline{bp}r[sp\#]_\#}  & EM comp. fluxes as above (LFS) &  $K_i(\psi,\zeta,s\approx0)$ \ (norm.)& & R or E & \File{xphi, yphi} & \name{+ xy_fluxes_bpar} \\

\File{[P/E/V]F\underline{S}[es/em/bp]r[sp\#]_\#}  & All fluxes as abv., but (FSA) &  $\int I_i/J_i/K_i(\psi,\zeta) ds$ \ (norm.)& & R or E & \File{xphi, yphi} & \name{+ xy_fluxes_fsa} \\

\File{[P/E/V]F[l/S][es/em/bp]\underline{p}[sp\#]_\#}  & Binormal fluxes as abv. (LFS/FSA)&  N/A & & R or E & \File{xphi, yphi} & \name{+ xy_fluxes_bi} \\

\File{[P/E/V]Fl[es/em/bp]\underline{k}[sp\#]_\#}  & Radial fluxes spectra (FSA) &  $\int I_i/J_i/K_i(k_\psi,k_\zeta) ds$ \ (norm.)& & R or E & \File{kxrh, krho} &  \name{+ xy_fluxes_k} \\

\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_fluxes_vspace:}} \\
\File{fluxes_det.dat} & \parbox[t]{4cm}{5D fluxes for each species, \\including $\ud^3X$ and $\ud^3v$}& $\Big(\bar{\cal I}_i + \bar{\cal J}_i +  \bar{\cal K}_i\Big)(k_\psi,k_\zeta,s,\mu,v_\parallel)$  & \parbox[t]{2cm}{binary, $N_{v_\parallel} N_\mu N_s\\ N_\psi N_{mod} N_{sp} i,$\\$i=1,2,3$}& E & & \name{lfluxes_detail} \\
\hline

\File{[p/e/v]fluxes_vspace[sp\#]_\#}  & Fluxes in velocity space (FSA) &  $\int I_i/J_i/K_i(v_\parallel,\mu) ds$ \ (norm.)& &R or E & \File{distr[1/2]} & \name{lfluxes_vspace} \\

%\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_rad:}} \\
%\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_velspace:}} \\
%\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_energetics:}} \\
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_f:}} \\
\File{distr*.dat}   & 2D velocity space output of 1 mode. & 1,2 grids, 3 Re$({\hat g}(s=0))$, 4 Im$({\hat g}(s=0))$ & $N_\mu$, $N_{v_\parallel}$ & E & & \\
%\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_stresses:}} \\
%\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_zfshear:}} \\
%\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_eng:}} \\
%\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_nonlin_transfer:}} \\

\hline\multicolumn{7}{ |c| }{\textbf{Output by other parts of the code:}} \\
\File{parfun.dat} & Perpendicular wavevector & $k_\perp\rho_{\rm ref}$ & $2,N_s N_{\rm mod} N_x$& B & & \\
\File{par.dat} & Curvature and Coriolis functions & $k_\perp\rho_{\rm ref}$ , ${\cal D}^\psi k_\psi + {\cal
D}^\zeta k_\zeta$, ${\cal H}^\psi k_\psi + {\cal H}^\zeta k_\zeta$, & $3,N_s N_{\rm mod} N_x$& B & & \\
\File{geom.dat} & Geometry tensors & Labelled in file & N/A& B & & \\
\File{FDS}        & Restart distribution function  & binary (see Sec. \ref{sec.restart})& E & & \\
\File{FDS.dat}    & Additional restart information & Ascii namelists (see Sec. \ref{sec.restart}) & E & & \\
\File{Coll_params.dat} & Collision frequencies & Labelled in file & N/A & B & & \\
\File{cfdens.dat} & Centrifugal poloidal density variations & $s$, $R/L_{\rm n, sp}^E(s)$, $exp(-\cfenn)_{\rm sp}(s), \theta$ &  $2 N_{\rm sp} + 2$ & B & & \\
\end{tabular}
\end{adjustwidth}
\end{footnotesize}
\caption{
  Diagnostics output by GKW (Part 2) , with \name{io_legacy=false}.
}
\label{tab:diag-part2}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}[hp!]
\begin{footnotesize}
\vspace{-2cm}
\begin{adjustwidth}{-1cm}{-2cm}
\centering 
\begin{tabular}{l|l|l|l|l|l|l}
 Filename & Description & Quantity (norm.=normalised) & Cols,(Rows) 
& \parbox[t]{2cm}{beginning (B),\\ repeatedly (R),\\ at the end (E)}
& \parbox[t]{1cm}{Grid\\Files}
& \parbox[t]{1.5cm}{Namelist\\Switch}\\
\hline
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_growth_freq:}} \\
\File{time.dat} & Dominant linear mode growth rate & $t_N, \gamma_{N}^{\rm max}, \omega_N$ & 3  & R & & \\
\File{growth.dat} & Growth rates of all linear modes & $\gamma_N$ &   & R & & \\
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_mode_struct:}} \\
\File{parallel.dat} & Parallel mode structure & See Sec. \ref{sec.parallel}. & $N_{\rm MOD} N_x N_s N_{\rm sp}$ & E & & \\
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_moments:}} \\
\File{{den/ene}_spectra} & \parbox[t]{3.5cm}{Bi-normal spectral perturbation  moments for each species} & $\sum_{k_\psi}\int_s \int \hat \alpha_{1,2} |\hat f_{sp}| {\rm d}^3v \, {\rm d}s $ & $N_{\rm mod}N_{sp}$ & & & \\
\hline\multicolumn{7}{ |c| }{\textbf{Module diagnos_fluxes:}} \\
\File{fluxes.dat} & Total $v_E$ fluxes $(i=1,2,3)$ by species  & ${\cal I}_i=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal I}_i \, {\rm d}s$  & $3N_{sp}$ & R & & \\
\File{fluxes_lab.dat} & Total Lab frame $v_E$ fluxes $(i=1,2,3)$ by species  & ${\cal I}_i + d{\cal I}_i$  & $3N_{sp}$ & R & & \\
\File{fluxes_em.dat} & Total $\delta B$ fluxes $(i=1,2,3)$ by species  & ${\cal J}_i=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal J}_i \, {\rm d}s$  & $3N_{sp}$ & R & & \\
\File{fluxes_em_lab.dat} & Total Lab frame $\delta B$ fluxes $(i=1,2,3)$ by species  & ${\cal J}_i+d{\cal J}_i$  & $3N_{sp}$ & R & & \\
\File{fluxes_bpar.dat} & Total $\nabla B_{1 \parallel}$ fluxes $(i=1,2,3)$ by species  & ${\cal K}_i=\sum_{k_\zeta} \sum_{k_\psi} \int_{s} \bar{\cal K}_i \, {\rm d}s$  & $3N_{sp}$ & R & & \\
\File{fluxes_bpar_lab.dat} & Total Lab frame $\nabla B_{1 \parallel}$ fluxes $(i=1,2,3)$ by species  & ${\cal K}_i+d{\cal K}_i$  & $3N_{sp}$ & R & & \\
\end{tabular}
\end{adjustwidth}
\end{footnotesize}
\caption{
  Legacy diagnostics output by GKW, with \name{io_legacy=true}. This table shows
  output datasets which differ from the ones described in
  tables \ref{tab:diag-part1} and \ref{tab:diag-part2} which
  are associated with \name{io_legacy=false}.
}
\label{tab:diag-legacy}
\end{table}

\end{landscape}

\newpage

\section{Island torque and stabilization by the parallel current}

A diagnostic especially useful for tearing mode dynamics is the one giving the torque and the stabilization
 by the parallel current, namely $T_\varphi$ and $\Delta_{\rm pol}$, respectively. These quantities are defined as follows, under
the assumption of a non rotating island
\begin{equation}
T_\varphi \propto \int_{-L_x/2}^{L_x/2}\, dx\, \int_{-\pi}^{\pi}\, d\zeta_0\,\,J_\parallel\sin\left(\kappa_{\zeta,\rm isl}\zeta_0\right)
\label{torque}
\end{equation}
\begin{equation}
\Delta_{\rm pol} \propto \int_{-L_x/2}^{L_x/2}\, dx\, \int_{-\pi}^{\pi}\, d\zeta_0\,\,J_\parallel\cos\left(\kappa_{\zeta,\rm isl}\zeta_0\right)
\label{deltaprime}
\end{equation}
where $L_x$ is the width of the simulation box in the radial direction, the variable $\zeta_0$ is the
helical co-ordinate in the absence of rotation and $J_\parallel$ is the gyro-centre parallel
current, given by
\begin{equation}
J_\parallel = \sum_{s}eZ_s\int_{\mathbb{R}^3} \, d^3\mathbf{v}\, v_\parallel \delta f_s
\end{equation}
with $\delta f_s$ the perturbed distribution function of the gyro-centres of species $s$.
Note that multiplying equation \ref{torque} by $-i$
and adding the resulting expression to equation \ref{deltaprime} yields
\begin{equation}
\Delta_{\rm pol} - i T_\varphi \propto \sum_{s}eZ_s\int_{-L_x/2}^{L_x/2}\, dx\, \int_{-\pi}^{\pi}\, d\zeta_0\, \int_{\mathbb{R}^3} \, d^3\mathbf{v} \, v_\parallel \delta f_s\, \exp\left(-i\kappa_{\zeta,\rm isl}\zeta_0\right)
\end{equation}
where we can distinguish the Fourier mode $\hat{\delta f_s}_{isl}\equiv\hat{\delta
f_s}\left(\kappa_\zeta=\kappa_{\zeta,\rm isl}\right)\propto\int_{-\pi}^{\pi}\, d\zeta_0\, \delta f_s \, \exp\left(-i\kappa_{\zeta,\rm isl}\zeta_0\right)$, with $\kappa_{\zeta,\rm isl}$ being the mode
 of the island in the binormal direction. Therefore, we can write
\begin{equation}
\Delta_{\rm pol} - i T_\varphi \propto \sum_{s}eZ_s\int_{-L_x/2}^{L_x/2}\, dx\, \int_{\mathbb{R}^3} \, d^3\mathbf{v} \, v_\parallel \hat{\delta f_s}_{\rm isl}
\end{equation}
A rotating island is usually imposed or obtained self-consistently in GKW. This means that the island is not
 always in the middle of the simulation box in the binormal direction. However, the above expressions are
 obtained assuming that the island is centered and therefore need to be corrected by multiplying
 by the phase $\exp\left(i\int_0^t \,dt' \omega_{\rm isl}\right)$, where $\omega_{\rm isl}$ is the rotation of the island, which results
 in the definition of the co-ordinate $\zeta = \zeta_0 - \int_0^t\, dt'\omega_{\rm isl}$. Note that in the absence
 of rotation, the amplitude of the vector potential $A_\parallel$ is real, due to the fact that a pure
 $\cos\left(\kappa_\zeta \zeta\right)$ is represented in GKW by a real amplitude of $A_\parallel$. Therefore,
 any departure from a real amplitude in $A_\parallel$ implies a rotation $\omega_{\rm isl}$, which can be calculated as
\begin{equation}
\int_0^t \, dt' \omega_{\rm isl} = \arctan\left(\frac{\Im\left(A_\parallel\right)}{\Re\left(A_\parallel\right)}\right)
\end{equation}
leading to the expression
\begin{equation}
\mathcal{T}\equiv\Delta_{\rm pol} - i T_\varphi \propto \sum_{s}eZ_s\int_{-L_x/2}^{L_x/2}\, dx\, \int_{\mathbb{R}^3} \, d^3\mathbf{v} \, v_\parallel \hat{\delta f_s}_{\rm isl}\exp\left(i\int_0^t \, dt' \omega_{\rm isl}\right)
\end{equation}

Finally, the torque and instability parameter are calculated as
\begin{equation}
T_\varphi = -\Im\left(\mathcal{T}\right)
\end{equation}
\begin{equation}
\Delta_{\rm pol} = \Re\left(\mathcal{T}\right)
\end{equation}

These two quantities are output in the files \File{torque.dat} and \File{deltaprime.dat}, respectively. The first column corresponds to
 the values for the ions and the second column to the values for the electrons.

\pagebreak

\section{Entropy (module \texttt{diagnos\_energetics} and \texttt{diagnos\_eng})}

% To cite ?
% Candy PoP 13, 032310 (2006):  Does this not relate directly to what we have ?
% Krommes 94

\subsection{Definition of the Total Entropy of the Simulated Box}\label{sec:entr-simul-box}
Let $p(\mathbf{X},v_\parallel, \mu) \in [0,1]$ be a probability density function.

One can define the functional
\begin{equation}
  \label{eq:bgs-entropy}
  S = -\int\!\ud^3x\ud^3v p\ln(p\cdot c_d),\qquad c_d = \mathrm{const.}
\end{equation}
and show that this expression has properties of the thermodynamical entropy (see e.g.\ the book of Jelitto 1989). It is therefore called ``statistical entropy'' or Shannon-Jaynes-entropy or Pauli-entropy or Boltzmann-Shannon-Gibbs (BGS) entropy. 
 A constant $c_d$ is formally needed in \ref{eq:bgs-entropy} to make the logarithm dimensionless \cite{DUN07} but it can be set equal to 1. Note that the integral in \ref{eq:bgs-entropy} is over the whole system, i.e.\ derived integrals below, on quantities of the simulated model, are taken over the whole simulation box.

A ``relative entropy'' functional can be defined in the form
\begin{equation}
  \label{eq:relative-entropy}
  S = -\int\!\ud^3x\ud^3v\ p\ln \frac{p}{q}
\end{equation}
This functional characterizes the distribution $p$ with respect to a certain reference probability density function $q \in [0,1]$ .

Maximization of the relative entropy $S$ under the constraint
\begin{align}\label{eq:maximizationSconstraint}
  1 = \int\ud\mathbf{x}\ud\mathbf{v}\,p(\mathbf{x},\mathbf{v})
\end{align}
leads to the condition
\begin{align}
  0 = 1 + ln\frac{p}{q} + \alpha
\end{align}
and thus
\begin{align}\label{eq:maximizationSp}
  p = q\cdot\exp[-(1+\alpha)]\
\end{align}
the distribution which corresponds to the extremum of the functional depends on the reference. With the assumption that the reference PDF $q$ fulfills
\ref{eq:maximizationSconstraint}, one applies this constraint \ref{eq:maximizationSconstraint} to \ref{eq:maximizationSp} and finds then $\alpha = -1$ and so
\begin{align}
  p = q ,
\end{align}
that means the entropy has its extremum for $p=q$.
Variation of the relative entropy under other constraints, using a reference which fulfills those, leads to the condition $p = q$ by the same argument.

% Is it not the neoclassical canoncial Maxwellian that is really the maximum entropy distribution ?
For a PDF describing an ensemble of classical particles, the distribution which maximizes the entropy under two contraints (normalisation of the distribution and given kinetic energy average) is the Maxwellian. 
For our purposes, we put therefore $f_{tot}$ in place of $p$ and the reference
distribution is chosen to be the Maxwell distribution in velocity space and to
describe particle trapping (and possibly centrifugal trapping) in position space.
This Maxwellian distribution function is denoted with $F_M$ and is given in equation \eqref{eq:maxwell}.
\begin{equation}
  \label{eq:relative-entropy-with-f}
  S = -\int\!\ud^3X\ud^3v f_{tot}\ln \frac{f_{tot}}{F_M}
\end{equation}
A state of maximum entropy then corresponds to $f_{tot} = F_M$. In the $\delta f$ model, where the distribution is split into $f_{tot} = F_M + f$, this means that the distribution perturbation vanishes in the thermodynamic equilibrium: $f = 0$.

Furthermore, it can be shown \cite{DUN07} that in contrast to the BGS entropy \ref{eq:bgs-entropy} given first, maximizing the relative entropy \ref{eq:relative-entropy} leads to the same distribution function, independently of the coordinate system. This is an important property if we want to investigate systems relaxing towards equilibrium in the field aligned coordinate system in which the gyrokinetic model is formulated. The invariance under coordinate transformations of the maximum-entropy principle associated with the relative entropy notion \ref{eq:relative-entropy} guarantees that the distribution $f_{tot} = F_M$ really corresponds to the maximum value of the entropy, and not another distribution, e.g.\ by multiplication with a factor.

Reduction of the perturbation $|f|$ means growth of the entropy $S$, i.e.\ the system moves towards thermodynamic equilibrium. On the contrary, processes that grow fluctuations $|f|$ effectively diminish the entropy of the system and keep it away from equilibrium, which is by definition the macrostate of maximum entropy.

We set $f_{tot} = F_M + f$ and expand around $f=0$: 
\begin{align}
  (F_M + f)\ln\frac{F_M + f}{F_M} &= (F_M + f)\left[0 + \frac{1}{F_M + f}|_{f=0}f^1 + \frac{1}{2}\frac{-1}{(F_M + f)^2}|_{f=0}f^2 + \ldots\right] \nonumber\\
  &= 0 + f + \frac{f^2}{F_M} - \frac{1}{2}\frac{f^2}{F_M} + \ldots \nonumber\\
  &= 0 + f + \frac{1}{2}\frac{f^2}{F_M} + \ldots
                     % \ln\frac{F_M + f}{F_M} &=  0 + \frac{f}{F_M} - \frac{1}{2}\frac{f^2}{F_M^2}\ldots + \ldots
\end{align}
Then \ref{eq:relative-entropy-with-f} becomes
\begin{align}
  \label{eq:def-entropy}
  S \approx -\int\!\ud^3X\ud^3v (f + \frac{1}{2}\frac{f^2}{F_M}) = -\frac{1}{2}\int\!\ud^3X\ud^3v\frac{f^2}{F_M}
\end{align}
where the integral of the fluctuation $f$ vanishes. 
$S$ in \ref{eq:def-entropy} is a small-amplitude approximation to the plasma entropy.  % \cite{krommes94}
 Again, one can see that reduction of $|f|$ means increase of the entropy $S$.

In its normalized form  \ref{eq:def-entropy} is 
\begin{equation}
  \label{eq:entropy}
  S \approx -\frac{1}{2}\int\!\ud^3X\ud^3v\frac{f^2}{F_M} = \rho_*^2 \frac{v_{thref}}{R_{ref}}n_{R_0}\underbrace{\left(-\frac{1}{2}\int\!\ud^3X\ud^3v_N\frac{f_N^2}{F_{MN}}\right)}_{S_N}
\end{equation}
In the normalized balance equation \ref{eq:entropy-balance} the reference quantities are dropped, but the smallness parameter $\rho_*$ is kept for illustration.

The quantity $H = -S$ is a measure of the intensity of the fluctuations in the distribution.% \cite{candywaltz06}
\begin{equation}
  \label{eq:intensity}
  H \approx \frac{1}{2}\int\!\ud^3X\ud^3v\frac{f^2}{F_M} = \rho_*^2 \frac{v_{thref}}{R_{ref}}n_{R_0}\underbrace{\left(\frac{1}{2}\int\!\ud^3X\ud^3v_N\frac{f_N^2}{F_{MN}}\right)}_{H_N}
\end{equation}
%Being a positive functional of the distribution, the intensity $H$ is an energy-like quantity \cite{krommes94}.
%FIXME This may be interesting when looking at the growth of unstable fluctuations? \cite{krommes94}

To emphasise, the fluctuation intensity $H$ is found to be \emph{increasing} in the simulations of a system driven out of equilibrium by applied thermodynamic forces (temperature and density gradient). Accordingly the entropy $S=-H$ is found to decrease.

\subsection{Entropy Balance\label{sec:entr-balance}}

\subsubsection*{General procedure} 
One starts from the normalized gyrokinetic equation given in section \ref{equations}, multiplies every term with $\frac{f}{F_M}$ and integrates over the box volume and velocity space.

In this section, all quantities are normalized, but we omit the index $N$ here. Moreover, only the electrostatic case is considered, so that $g = f$ for the purposes of this section.

\begin{align}\label{eq:entropy-balance}  
\rho_*^2\int\!\ud^3X\ud^3v\frac{f}{F_M}\Big[\pd{g}{t} +\mathrm{I} + \mathrm{II} + \mathrm{III} + \mathrm{IV} + \mathrm{V} + \mathrm{VI} + \mathrm{VII} + \mathrm{VIII}\Big] = 0 
\end{align}

\subsubsection*{Regard the terms separately}
One can recover $\pd{\ }{t} H$ in the very first integral.
\begin{align}
  \label{eq:dSdt}
  \rho_*^2\int\!\ud^3X\ud^3v\frac{f}{F_M}\pd{g}{t} &= \rho_*^2\int\!\ud^3X\ud^3v\frac{1}{F_M}\frac{1}{2}\pd{f^2}{t} \nonumber\\
&= \pd{H}{t} = - \pd{S}{t}
\end{align}

The linear $\mathbf{E}\times\mathbf{B}$-drift term V yields
\begin{align} 
  \rho_*^2\int\!\ud^3X\ud^3v\frac{f}{F_M}\ \mathrm{V} & = \rho_*^3\int\!\ud^3X \Bigg[-\frac{1}{L_n}\Gamma_N^\psi - Q_N^\psi\frac{1}{L_T}\nonumber\\
  &\quad +\frac{3}{2}\frac{1}{L_T}\Gamma^\psi - \frac{\mathcal{E}_R}{T_R}\frac{1}{L_T}\Gamma^\psi +\nonumber\\
  &\quad - \frac{2}{T_R}\Pi_\varphi^\psi u^\prime
  - 2\frac{m_R}{T_R}\Omega_N^2\mathcal{L}\,\Gamma^\psi \Bigg]
\end{align}
At the moment these six terms are output separately by the code, but
very likely they will be combined in the near future to fewer, more
symmetric quantities.

The Landau-Damping terms VI and VII give non-vanishing contributions,
too. In combination with the Poisson equation, they can be
rewritten as
\begin{align}
  \rho_*^2\int\!\ud^3X\ud^3v\frac{f}{F_M}\Big[\mathrm{VII} + \mathrm{VIII}\Big] &= \int\ud^3X \pd{}{t} \sum_s\int\ud^3v \frac{Z_s^2}{2T_s^2}F_M\left(\phi^2(\mathbf{x}) - \Ga{\phi}^2 \right) \\
 \text{and $W_s$ is defined such that} &= \pd{}{t}\sum_sW_s \\
 W_{N,s} = {}& \int\ud^3X_N \int\ud^3v_N \frac{Z_s^2}{2T_{R,s}^2}F_{MN}\left(\phi_N^2 -  \Ga{\phi_N}^2\right)   \label{eq:W}
\end{align}
If only ion dynamics is simulated and electrons are treated
adiabatically, the index $s$ above only runs over ion species and the
term $W_{N,e}$ for the electrons is different.
\begin{align}
\pd{}{t}W_{N,e,adia}
{}={}& \pd{}{t}\int\ud^3X_N \phi_N \int\ud^3v_N\frac{Z_e}{T_{R,e}}\iGa{f_{eN}} \nonumber\\
{}={}& \pd{}{t}\int\ud^3X_N \phi_N \frac{Z_e}{T_{R,e}}n_R\frac{Z_e}{T_{R,e}}\left(\phi_N - \fsa{\phi_N} \right) \nonumber\\
{}={}& \int\ud^3X_N  n_R\frac{Z_e^2}{2T_{R,e}^2}\left(
 \pd{\phi_N^2}{t} - 2\phi_N\pd{}{t}\fsa{\phi_N}
 \right)
\end{align}
And with $\fsa{.}=\fsa{.}(\psi,\zeta)$ and $T_{R,e}^2$ being constant
with respect to the parallel direction, one can write
\begin{align}
{}={}& \int\ud\psi\ud\zeta J_X   n_R\frac{Z_e^2}{2T_{R,e}^2}\left(
\pd{\fsa{\phi_N^2}}{t} -
 2\fsa{\phi_N}\pd{}{t}\fsa{\phi_N}
 \right)\\
{}={}& \int\ud\psi\ud\zeta J_X   n_R\frac{Z_e^2}{2T_{R,e}^2}\left(
 \pd{\fsa{\phi_N^2}}{t} -
\pd{\fsa{\phi_N}^2}{t}
  \right) \\
{}={}& \pd{}{t}\int\ud\psi\ud\zeta J_X  n_R\frac{Z_e^2}{2T_{R,e}^2}\left(
  \fsa{\phi_N^2} -
 \fsa{\phi_N}^2 
\right) 
\ .
\label{eq:W-corr-adia}
\end{align}

% A rotating plasma knows about a further correction for the adiabatic
% electron contribution, also due to a modification of the Poisson
% equation.
% \begin{align}
%  \pd{}{t} W_{N,e,adia,rot}{}={}& \pd{}{t} \int\ud\psi\ud\zeta J_X   n_R\frac{Z_e}{T_{R,e}^2}\left(
% % Z_e\frac{1}{2}\fsa{\phi_N}^2
% %  \right.\nonumber\\&\qquad\qquad \left.
% % -Z_e\frac{1}{2}\fsa{\phi_N^2}
% + \fsa{\phi_N\mathcal{E}_{R,e}}
% \right)
% \end{align}

\subsection{Resulting entropy balance equation\label{sec:result-entropy-balance}}

The terms of the entropy balance equation which were very briefly presented
in section \ref{sec:entr-balance}, and the numerical diffusion terms,
summed over all species make up the following balance equation.
\begin{align}
  0 &= \sum_s\pd{}{t}(H_s + \overbrace{W_s}^{\text{\makebox[0pt]{from VII and VIII}}}) + \sum_s\Big(\overbrace{\text{contrib. from turb. fluxes}}^{\text{\makebox[0pt]{from linear ExB drift V;}}} + \overbrace{\text{\ \ contrib. from neoclass. fluxes}}^{\text{\makebox[0pt]{from VI, but this is ignored and not implemented}}}  + \nonumber \\
  &\quad{}-\mathcal{D}_{f\parallel} - \mathcal{D}_{v_\parallel} - \mathcal{D}_{\perp} \Big) - \sum_s\int\!\ud^3X\ud^3v_N\frac{f_N}{F_{MN}}\mathcal{C}_N
\end{align}
To be more explicit, in terms of entropy this equation reads
\begin{align}  \label{eq:result-entropy-balance}
 \sum_s\pd{}{t}(S_s - W_s) =&\sum_s\Big(\int\!\ud^3X \Bigg[-\frac{1}{L_n}\Gamma_N^\psi - Q_N^\psi\frac{1}{L_T}
  +\frac{3}{2}\frac{1}{L_T}\Gamma^\psi - \frac{\mathcal{E}_R}{T_R}\frac{1}{L_T}\Gamma^\psi 
  - \frac{2}{T_R}\Pi_\varphi^\psi u^\prime
  - 2\frac{m_R}{T_R}\Omega_N^2\mathcal{L}\,\Gamma^\psi \Bigg]\nonumber\\
 &\quad\quad-\mathcal{D}_{f\parallel} -\mathcal{D}_{v_\parallel} -\mathcal{D}_{\perp}\Big)
- \sum_s\int\!\ud^3X\ud^3v_N\frac{f_N}{F_{MN}}\mathcal{C}_N
\end{align}
and in terms of free energy
\begin{align}
  & \pd{}{t}\sum_s(-T_{R,s}S_{N,s} + E_{\chi,N,s}) = {}\nonumber\\
&\quad\sum_sT_{R,s}\Big(\int\!\ud^3X \Bigg[\frac{1}{L_n}\Gamma_N^\psi + Q_N^\psi\frac{1}{L_T}
  -\frac{3}{2}\frac{1}{L_T}\Gamma^\psi + \frac{\mathcal{E}_R}{T_R}\frac{1}{L_T}\Gamma^\psi
  + \frac{2}{T_R}\Pi_\varphi^\psi u^\prime
  + \frac{m_R}{T_R}\Omega_N^2\mathcal{L}\,\Gamma^\psi \Bigg]\nonumber\\
 &\quad\quad+\mathcal{D}_{f\parallel} +\mathcal{D}_{v_\parallel} +\mathcal{D}_{\perp}\Big)
 %+ \sum_sT_{R,s}\int\!\ud^3X_N\ud^3v_N\frac{h_N}{F_{MN}}\mathcal{C}_N
+ \sum_s\mathcal{D}_{C,s}
\end{align}
where $E_{\chi,S}=W_sT_{R,s}$.

In a system driven out of equilibrium the fluctuations grow, that is,
the entropy gets smaller (i.e.\ more negative). The time derivative on
the left side is therefore negative. This corresponds to the fluxes
and the gradient lengths defined here being positive and the numerical
dissipation terms $\mathcal{D}_i$ being negative.

In a statistically steady state, the left side of this equation
vanishes. The entropy sinks from the fluxes are then balanced by the
numerical dissipation terms.

\subsection{Output quantities of the diagnostic}\label{sec:entr-output}

If \texttt{lcalc_energetics} in the diagnostics namelist is switched
on, the code will produce a number of files related to the entropy
balance equation:

The quantities in the entropy balance equation \ref{eq:result-entropy-balance}
 correspond directly to the output of the \texttt{energetics()} diagnostic.

\begin{align}
   \mathtt{dt\_entr} + \mathtt{dt\_entr\_field} =& -\Big(\mathtt{entr\_src01} + \mathtt{entr\_src02} + \mathtt{entr\_src03} \nonumber\\
& \quad + \mathtt{entr\_src04} + \mathtt{entr\_src05} + \mathtt{entr\_src06}
\nonumber\\
 &\quad\quad + \mathtt{entr\_num\_dis} + \mathtt{entr\_num\_vp} + \mathtt{entr\_num\_perp} \nonumber\\
& \quad\quad + \mathtt{entr\_coll}\Big)
\end{align}
To clarify the signs, note the definition
\begin{align}
  \pd{}{t}\sum_s\left(-W_{N,s}\right) = \mathtt{dt\_entr\_field} \ .
\end{align}

In order to check the balance one can sum both sides and see how well
they eliminate each other.

\begin{align}
&   \mathtt{dt\_entr} + \mathtt{dt\_entr\_field} +\Big(\mathtt{entr\_src01} + \mathtt{entr\_src02} + \mathtt{entr\_src03} \nonumber\\
& \quad + \mathtt{entr\_src04} + \mathtt{entr\_src05} + \mathtt{entr\_src06}
\nonumber\\
 &
%\underbrace{
\quad\quad + \mathtt{entr\_num\_dis} + \mathtt{entr\_num\_vp} + \mathtt{entr\_num\_perp} + \mathtt{entr\_coll}\Big)
%}_{\mathtt{total}} 
\overset{?!}{=} 0
\end{align}

\pagebreak

\section{Zonal potential evolution (module \texttt{diagnos\_zonal\_evo})}
\label{zonal-potential-evolution}

The zonal potential is the flux-surface averaged potential and is defined by
\begin{equation}
\{\phi \} = \sum_{k_\psi} \underbrace{ \int \mathrm{d}s~\hat \phi (k_\psi, k_\zeta=0,s) }_{\{\hat \phi \}(k_\psi)} \exp(\mathrm{i} k_\psi \psi / \rho_\ast),
\end{equation}
where the spectral representation (see Sec.~\ref{sec:spectral}) is applied and $\{\hat \phi \}(k_\psi)$ defines the complex Fourier coefficient of the zonal potential with radial wave vector $k_\psi$. 
An evolution equation for the zonal potential can be derived from the Poisson equation (\ref{Poisson}).
The time derivative of the inverted and flux-surface averaged $k_\zeta = 0$ part of this equation reads
\begin{equation}
\label{eq:dt-inverted-poisson}
\frac{\partial \{ \hat \phi \}(k_\psi)}{\partial t} = -\int \mathrm{d} s ~ \frac{1}{\mathcal{P}}  \sum_{sp} Z_{sp} n_{R,sp} 2 \pi B  \int \mathrm{d} v_{\parallel} \int \mathrm{d} \mu ~ \biggl[J_0(k_\perp\rho_{sp}) \frac{\partial \hat g_{sp}}{\partial t} \biggr]_{\mathbf{k}=(k_\psi,k_\zeta=0)}
\end{equation}
where
\begin{equation}
\mathcal{P} = \sum_{sp} \frac{n_{R,sp} Z_{sp}^2}{T_{R,sp}} [ \Gamma_0(b_{sp}) -1].
\end{equation}
Here, the notation $[ \hat G ]_{\mathbf{k}=(k_\psi,k_\zeta=0)}$ means that the argument $\hat G$ has to be evaluated at $k_\zeta = 0$. \\
The module \texttt{diagnos\_zonal\_evo} calculates various contributions to the time evolution of the zonal potential.
For this purpose, the time derivative of the perturbed modified distribution function $\partial_t \hat g_{sp}$, entering the right hand side of Eq.~(\ref{eq:dt-inverted-poisson}), is substituted by the individual terms (I-XI) on the right hand side of the gyrokinetic equation (\ref{eqs:complete-set}).
This procedure yields several linear and nonlinear contributions to the evolution of the zonal potential.
The corresponding diagnostic output is discussed in Sec.~\ref{zevo-output-quantities}.

\subsection{Vanishing terms}
This part will be more specific about individual linear terms entering through the time derivative of the modified distribution.
Closer inspection shows that several terms vanish.\\
%
First, the velocity space integral of the trapping term (term IV) is carried out:
\begin{align}
& \int \mathrm{d} \mu \int \mathrm{d} v_\parallel ~ J_0[k_\perp \rho(\mu) ] v_{R} \left(\mu B {\cal G} + {1 \over 2} \pd{\cfenn}{s} {\cal F} \right) \pd{\hat f}{v_\parallel} \\
= & \int \mathrm{d} \mu  ~ J_0[k_\perp \rho(\mu) ]  v_{R} \left(\mu B {\cal G} + {1 \over 2} \pd{\cfenn}{s} {\cal F} \right) \int \mathrm{d} v_\parallel \pd{\hat f}{v_\parallel} \\
= & \int \mathrm{d} \mu ~  J_0[k_\perp \rho(\mu) ]  v_{R} \left(\mu B {\cal G} + {1 \over 2} \pd{\cfenn}{s} {\cal F} \right) \biggl[ \hat f\biggr]_{-\infty}^{+\infty} \\
= & ~0.
\end{align}
It has been used that $\partial_{v_\parallel} \hat f$ is the only quantity with $v_\parallel$-dependence, allowing for a separatation of the velocity space integration. 
Since the perturbed distribution has to vanish for $v_\parallel \rightarrow \pm \infty$ the integration over parallel velocity space yields zero. \\
%
Second, the inner product of the ExB-drift and the parallel motion along perturbed field lines, i.~e., $\mathbf{v}_\chi$, with the gradient of the maxwellian (term V) is concentrated on (neglecting the Sung terms that are higher order in $\rho_\ast$):
\begin{align}
& \int \mathrm{d} \mu \int \mathrm{d} v_\parallel ~J_0(k_\perp \rho) \mathrm{i}k_\alpha \hat \chi|_{k_\zeta = 0} {\cal E}^{\alpha \psi} \biggl [ {1 \over L_n} + {m_R \Omega ^2 \over T_R}{\cal L} + E_T {1 \over L_T} + {2 v_{\parallel} \over  v_{R}} {R B_t \over B} u^\prime \biggr ] F_{M} = \\
= & \int \mathrm{d} \mu \int \mathrm{d} v_\parallel ~ J_0(k_\perp \rho) \mathrm{i}k_\psi \hat \chi|_{k_\zeta = 0} {\cal E}^{\psi \psi} \biggl [ {1 \over L_n} + {m_R \Omega ^2 \over T_R}{\cal L} + E_T {1 \over L_T} + {2 v_{\parallel} \over  v_{R}} {R B_t \over B} u^\prime \biggr ] F_{M} = \\
= & ~ 0.
\end{align}
It has been used that the $k_\zeta = 0$ part of the modified potential is considered, i.~e., $\hat \chi|_{k_\zeta = 0} = \hat \chi (k_\psi, k_\zeta=0, s)$, such that the multiplication with $k_\zeta$ vanishes.
Furthermore, ${\cal E}^{\psi \psi} = 0$. \\
In a well resolved numerical simulation the above described terms should vanish.
The remaining terms are non-zero, in general, and may contribute to the evolution of the zonal potential.

\subsection{Constructing $k_\psi$-$k_\zeta'$-spectra from Hermitian symmetry}
\label{zevo-kxky-spectra}
The nonlinear terms of the gyrokinetic equation allow for the evolution of the zonal potential through the nonlinear interaction of perturbations with finite $k_\zeta'$.
Here, $k_\zeta'$ refers to the binormal wave vector of modes that interact through the nonlinearity.
It has to be distinguished from the wave vector $k_\zeta$ used in Eq.~(\ref{eq:dt-inverted-poisson}) that refers to the perturbation that is generated by the nonlinear interaction, i.~e., here the $k_\zeta = 0$ mode.
The Hermitian symmetry of the spectrum allows to construct $k_\psi$-$k_\zeta'$-spectra of the contribution of the nonlinear terms to the zonal potential evolution. \\
The nonlinear term entering Eq.~(\ref{eq:dt-inverted-poisson}) can be formally written in position space by
    \begin{equation}
    \mathcal{N}(\psi) = \frac{1}{L_\zeta} \int_0^{L_\zeta} \mathrm{d} \zeta A(\psi, \zeta) B(\psi,\zeta),
    \label{eq:formal-nonlinearity}
    \end{equation}
where $L_\zeta$ is the box size in the $\zeta$-direction.
For simplicity the $s$-dependence and integration are neglected.
Since the domain is periodic in $\zeta$, only the constant part of the integrand contributes.
This is equivalent to the condition $k_\zeta = 0$ on the right hand side of Eq.~(\ref{eq:dt-inverted-poisson}). \\
The perturbed fields entering the nonlinearity are real quantities and, therefore, their Fourier respresentation satisfies Hermitian symmetry (see also Sec.~\ref{sec.parseval-correction})
    \begin{equation}
    G(\psi, \zeta) = \hat G (\psi,k_\zeta'=0) + \sum_{k_\zeta' > 0} \biggl[ \hat G (\psi,k_\zeta') \exp(\mathrm{i} k_\zeta' \zeta/\rho_\ast) +  \hat G^\ast (\psi,k_\zeta') \exp(-\mathrm{i} k_\zeta' \zeta/\rho_\ast) \biggr],
    \label{eq:hermitian-spectrum}
    \end{equation}
where $G \in [A, B]$ and the superscript "$\ast$" denotes the complex conjugate.
Since the part that is constant in $\zeta$ contributes to the integral in Eq.~(\ref{eq:formal-nonlinearity}) only, one finds
    \begin{align}
    \mathcal{N}(\psi)  = & \frac{1}{L_\zeta} \int_0^{L_\zeta} \mathrm{d} \zeta \sum_{k_\zeta'>0} \sum_{k_\zeta''>0} \biggl[ \hat A(\psi,k_\zeta') \hat B^\ast(\psi,k_\zeta'') \exp[\mathrm{i} (k_\zeta' - k_\zeta'')\zeta'/\rho_\ast] \nonumber \\
                         & + \hat A^\ast(\psi,k_\zeta') \hat B(\psi,k_\zeta'') \exp[-\mathrm{i} (k_\zeta' - k_\zeta'')\zeta'/\rho_\ast] \biggr] \label{eq:zevo-hermitian-1}\\
            = & \sum_{k_\zeta'>0} \biggl[ \hat A(\psi,k_\zeta') \hat B^\ast(\psi,k_\zeta') + \hat A^\ast(\psi,k_\zeta') \hat B(\psi,k_\zeta') \biggr] \label{eq:zevo-hermitian-2}\\
            = & 2 \sum_{k_\zeta'>0} \mathrm{Re}[\hat A(\psi,k_\zeta') \hat B^\ast(\psi,k_\zeta')].\label{eq:zevo-hermitian-3}
    \end{align}
Terms proportional to $\exp[\mathrm{i}(k_\zeta'+k_\zeta'') \zeta'/\rho_\ast]$ in expression (\ref{eq:zevo-hermitian-1}) vanish due to periodicity in $\zeta$ and due to $k_\zeta',~k_\zeta'' > 0$.
In expression (\ref{eq:zevo-hermitian-2}) it has been used that $\int_0^{L_\zeta} \mathrm{d}\zeta~\exp[\mathrm{i}(k_\zeta'-k_\zeta'') \zeta'/\rho_\ast] = \delta_{k_\zeta',k_\zeta''}$, where the latter is the Kronecker delta.
The factor $2$ in expression (\ref{eq:zevo-hermitian-3}) is what is called \texttt{parseval\_correction} in the code. \\
In case of the nonlinearity of the gyrokinetic equation
    \begin{equation}
    A = \rho_\ast^2 \mathcal{E}^{\psi \zeta} \pd{\chi}{\zeta} \hspace{20pt} B = \pd{g}{\psi}
    \end{equation}
or 
    \begin{equation}
    A = -\rho_\ast^2 \mathcal{E}^{\psi \zeta}\pd{\chi}{\psi} \hspace{20pt} B = \pd{g}{\zeta},
    \end{equation}
where $\chi$ is defined in Eq.~(\ref{eq:chi-normalized}).
Finally, the time evolution of the $\mathbf{k}=(k_\psi,k_\zeta=0)$ mode of the distribution function due to the nonlinearity then reads
    \begin{equation}
    \label{eq:zevo-hermitian}
    \Biggl[ \pd{\hat g}{t}\Big|_\mathrm{NL}\Biggr]_{\mathbf{k}=(k_\psi,k_\zeta=0)} (k_\psi) = \sum_{k_\zeta'} \underbrace{\mathcal{T}_\psi \biggl\{\rho_\ast 2 \mathcal{E}^{\psi \zeta} \mathrm{Re} \biggl[ \mathrm{i} k_\zeta' \hat \chi(\psi,k_\zeta') \biggl( \pd{\hat g (\psi,k_\zeta')}{\psi} \biggr)^\ast - \pd{\hat \chi(\psi,k_\zeta')}{\psi} \biggl( \mathrm{i} k_\zeta' \hat g(\psi,k_\zeta') \biggr)^\ast \biggr]\biggr\}}_{[\partial_t \hat g |_\mathrm{NL}]_{\mathbf{k}=(k_\psi,k_\zeta=0)}  (k_\psi,k_\zeta')}.
    \end{equation}
Here, $\mathcal{T}_\psi$ denotes a Fourier transform in $\psi$. 
The factor $\rho_\ast$ in Eq.~(\ref{eq:zevo-hermitian}) vanishes after the Fourier transform, since $\partial_\psi \hat{g} \rightharpoondown (\mathrm{i} k_\psi /\rho_\ast) \hat{g}$ and $\partial_\psi \hat{\chi} \rightharpoondown (\mathrm{i} k_\psi /\rho_\ast) \hat{\chi}$ . 
Note that Eq.~(\ref{eq:zevo-hermitian}) also defines the contribution from interacting modes with wave vector $k_\zeta'$, i.~e., $[\partial_t \hat g |_\mathrm{NL}]_{\mathbf{k}=(k_\psi,k_\zeta=0)}  (k_\psi,k_\zeta')$.
The dependency on $k_\psi$ (not $k_\psi'$) is explicitly given within parentheses and refers to the radial wave vector of the mode that is generated nonlinearly. \\
Following the procedure described in Sec.~\ref{zonal-potential-evolution}, Eq.~(\ref{eq:zevo-hermitian}) is substituted for $\partial_t \hat g_{sp}$ on the right hand side of Eq.~(\ref{eq:dt-inverted-poisson}).
%Note that the condition $k_\zeta = 0$ in Eq.~(\ref{eq:dt-inverted-poisson}) translates to the requirement that only the constant part of the integrant in Eq.~(\ref{eq:formal-nonlinearity}) contributes.
%It is, therefore, implicitly satisfied by the expression given in Eq.~(\ref{eq:zevo-hermitian}).
It is worth to stress again that $k_\zeta'$ in Eq.~(\ref{eq:zevo-hermitian}) specifies the wave vector of the modes that interact nonlinearly, while $k_\zeta$ correponds to the mode that is generated by this nonlinear interaction process, i.~e., here a mode with $k_\zeta = 0$.

\subsection{Output quantities}
\label{zevo-output-quantities}
Per default (\texttt{lcalc\_zonal\_evo}) the diagnostic outputs $k_\psi$-spectra of the zonal potential $\{\hat \phi \}$, the time derivative of the zonal potential $\partial_t \{\hat \phi \}$, the contribution of the sum of all linear terms of the gyrokinetic equation (including numerical dissipation and collisions, if active) and the individual nonlinear contributions (if active) to the zonal potential evolution (\ref{eq:dt-inverted-poisson}).
See Tab.~\ref{tab:diagnos-zonal-evo-output} for a list of the output files.
The output file \File{zevo03\_phi\_kx\_[real,imag]} is associated to the ExB-nonlinearity [$\phi$-term of $\chi$ in Eq.~(\ref{eq:chi-normalized})], the output file \File{zevo03\_apar\_kx\_[real,imag]} to the magnetic flutter nonlinearity [$A_\parallel$-term of $\chi$ in Eq.~(\ref{eq:chi-normalized})] and the output file \File{zevo03\_bpar\_kx\_[real,imag]} to the magnetic field compression nonlinearity [$B_\parallel$-term of $\chi$ in Eq.~(\ref{eq:chi-normalized})].
A numerical simulation should always satisfy
\begin{align}
\File{dt\_zphi\_kx\_[real,imag]} & = \File{zevo\_lin\_kx\_[real,imag]} + \File{zevo03\_phi\_kx\_[real,imag]} \nonumber \\
                                 & + \File{zevo03\_apar\_kx\_[real,imag]} + \File{zevo03\_bpar\_kx\_[real,imag]},
\end{align}
provided the time step is small enough for an accurate numerical estimation of the time derivative of $\{\hat \phi \}$ and provided numerical instabilities are absent. \\
A detailed output of $k_\psi$-spectra of the individual linear contributions to the zonal potential evolution is provided by the namelist switch \texttt{zevo\_detail}.
The nomenclature of the output files (see Tab.~\ref{tab:diagnos-zonal-evo-output}) follows the numbering of the individual terms of the complete set of equations described in Sec.~\ref{equations}.  
The sum of the individual linear contributions yields the combined linear output
\begin{align}
\File{zevo\_lin\_kx\_[real,imag]} & = \File{zevo01\_kx\_[real,imag]} + \File{zevo02\_kx\_[real,imag]} \nonumber \\
                                  & + \File{zevo04\_kx\_[real,imag]} + \File{zevo05\_kx\_[real,imag]} \nonumber \\
                                  & + \File{zevo07\_kx\_[real,imag]} + \File{zevo08\_kx\_[real,imag]} \nonumber \\
                                  & + \File{zevo10\_kx\_[real,imag]} + \File{zevo11\_kx\_[real,imag]} \nonumber \\
                                  & + \File{zevo\_dpar\_kx\_[real,imag]} + \File{zevo\_dvp\_kx\_[real,imag]} \nonumber \\
                                  & + \File{zevo\_dperp\_kx\_[real,imag]} + \File{zevo\_coll\_kx\_[real,imag]}.
\end{align}
\textit{WARNING: The calculation of the individual linear contributions is memory consuming, since it requires the allocation of multiple linear terms matrices! Depending on the problem size, the number of processors may have to be increased.} \\
The namelist switch \texttt{zevo\_xy\_spec} allows for the output of $k_\psi$-$k_\zeta'$-spectra of the nonlinear contributions to the zonal potential evolution (see Tab.~\ref{tab:diagnos-zonal-evo-output} for the output file naming).
This feature exploits the Hermitian symmetry of the binormal spectrum (see also Sec.~\ref{zevo-kxky-spectra}).
The corresponding output relates to the default output through
\begin{align}
\File{zevo03\_phi\_kx\_[real,imag]} & = \sum_{k_\zeta'} \File{zevo03\_phi\_kxky\_[real,imag]} \\
\File{zevo03\_apar\_kx\_[real,imag]} & = \sum_{k_\zeta'} \File{zevo03\_apar\_kxky\_[real,imag]} \\
\File{zevo03\_bpar\_kx\_[real,imag]} & = \sum_{k_\zeta'} \File{zevo03\_bpar\_kxky\_[real,imag]}.
\end{align}
In the above equations the expression $\sum_{k_\zeta'}$ denotes a sum over the $k_\zeta'$ dimension of the respective output file.
\textit{WARNING: The calculation of the $k_\psi$-$k_\zeta'$-spectral output is computationally expensive. 
If not mandatory, run performance might be improved by deactivation of} \texttt{zevo_xy_spec}.

\begin{landscape}
\begin{table}
\begin{center}
\begin{tabular}{c|c|c|c|c}
\textbf{Namelist switch}    & \textbf{File name}                     & \textbf{Additional requirements}   & \textbf{Size} & \textbf{Grid Files}\\
\hline \hline
 \texttt{lcalc\_zonal\_evo} & \File{zphi\_kx\_[real,imag]}           &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh} \\
                            & \File{dt\_zphi\_kx\_[real,imag]}       &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo\_lin\_kx\_[real,imag]}      &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo03\_phi\_kx\_[real,imag]}    & \texttt{nonlinear}+\texttt{nlphi}  & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo03\_apar\_kx\_[real,imag]}   & \texttt{nonlinear}+\texttt{nlapar} & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo03\_bpar\_kx\_[real,imag]}   & \texttt{nonlinear}+\texttt{nlbpar} & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
 \hline
 \texttt{lcalc\_zonal\_evo} & \File{zevo01\_kx\_[real,imag]}         &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
 +\texttt{zevo\_detail}     & \File{zevo02\_kx\_[real,imag]}         &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo04\_kx\_[real,imag]}         &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo05\_kx\_[real,imag]}         &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo07\_kx\_[real,imag]}         &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo08\_kx\_[real,imag]}         &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo10\_kx\_[real,imag]}         & \texttt{nlbpar}                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo11\_kx\_[real,imag]}         & \texttt{nlbpar}                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo\_dpar\_kx\_[real,imag]}     &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo\_dvp\_kx\_[real,imag]}      &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo\_dperp\_kx\_[real,imag]}    &                                    & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\
                            & \File{zevo\_coll\_kx\_[real,imag]}     & \texttt{collisions}                & $N_\mathrm{t},~N_x$ & \File{time}, \File{kxrh}\\ 
 \hline
 \texttt{lcalc\_zonal\_evo} & \File{zevo03\_phi\_kxky\_[real,imag]}  & \texttt{nonlinear}+\texttt{nlphi}  & $N_\mathrm{t},~N_x,~N_\mathrm{mod}$ & \File{time}, \File{kxrh}, \File{krho} \\
 +\texttt{zevo\_xy\_spec}   & \File{zevo03\_apar\_kxky\_[real,imag]} & \texttt{nonlinear}+\texttt{nlapar} & $N_\mathrm{t},~N_x,~N_\mathrm{mod}$ & \File{time}, \File{kxrh}, \File{krho}\\
                            & \File{zevo03\_bpar\_kxky\_[real,imag]} & \texttt{nonlinear}+\texttt{nlbpar} & $N_\mathrm{t},~N_x,~N_\mathrm{mod}$ & \File{time}, \File{kxrh}, \File{krho}\\
\end{tabular}
\caption{This table lists the output of the module \name{diagnos\_zonal\_evo}.}
\label{tab:diagnos-zonal-evo-output}
\end{center}
\end{table}
\end{landscape}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/eiv.tex
(1) Added gkw Clone gwk  #3

\chapter{Eigenvalue solver}
\label{ch:eiv}

Here we give some information on the usage of the eigenvalue solver, implemented
in GKW, as well as some technical details. 
Further information can be found in the presentation of R. Buchholz on the GKW webpages under Talks.

As the system size of the linear gyrokinetic equation is too big for a direct numerical solution, projection
methods are used.
For a brief introduction to projection methods, see the {\sc slepc} manual \cite{SLE12} and references therein.

\section{Usage}
\label{sec:eivusage}
First, GKW must be compiled with {\sc slepc/petsc}, see \ref{subsubsec:slepc}
for details.

The easiest way to create an input file for the eigenvalue solver is the following:
First, set up an input file which uses explicit time
integration.
As second step then replace \name{METHOD = 'EXP'} with \name{METHOD='EIV'}
and  set \name{METH=1}. It is recommended to set a low value \name{NAVERAGE} like \name{NAVERAGE=1}.
Setting \name{METH=2} or \name{NAVERAGE$\gg$1} works too, but the run is then usually slower.

The third and final step is to set the parameters for the eigenvalue solver
itself. This is done in the namelist \name{eiv_integration},
an example of which is given below.
One has to select the portion of the spectrum to be sought with the parameter \name{which_eigenvalues}. Available options are listed in the sample input file.
In addition, you can choose between two methods for extracting eigenvalues,
\name{type_extraction = 'harmonic'}, and \name{type_extraction = 'ritz'}.
Note that with \name{type_extraction = 'harmonic'} the target values \name{growthrate} and \name{freq}) are always used.

Unfortunately it is difficult to recommend optimal settings in advance,
or even settings which will guarantee to find physically interesting eigenmodes of the system.

One rule seems to be valid in general:
If you use harmonic extraction and search for the mode with largest
eigenvalue, you should keep the target frequency \name{freq=0.0}.  
If you don't find an eigenvalue, increase the target \name{growthrate}, 
while if you just find stable high frequency eigenmodes, decrease it.

To share some experiences, fig. \ref{fig:eivharmonicgrowth}--\ref{fig:eivtargetsa} 
show scans for two sets of parameters.
\begin{figure}[htp!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{HarmonicGrowth}
    \caption{\label{fig:eivharmonicgrowth} Scan over growth rate with
    frequency set to zero. Dashed lines depict negative values of
    \name{growthrate}. Normalization is done with the maximum number of
    iterations for the first/second eigenvalue/total number of iterations,
    respectively.}
  \end{center}
\end{figure}
\begin{figure}[htp!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{HarmonicFreq}
    \caption{\label{fig:eivharmonicfreq} Scan over frequency with growth rate
      set to zero. Dashed lines depict negative values of \name{freq}. The
      point at $10^{-7}$ actually had zero frequency. Normalization is as for
      fig. \ref{fig:eivharmonicgrowth}. The dashed-dotted line depicts the
      minimum value for the total iterations of fig. \ref{fig:eivharmonicgrowth}
      (only counting points where both unstable eigenmodes are found).}
  \end{center}
\end{figure}
\begin{figure}[htp!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{HarmonicNcv}
    \caption{\label{fig:eivharmonicncv} Scan over size of subspace  (\name{nr\_column\_vec}), for one
    point of each of the other two scans. First case: \name{growthrate = 0.0},
    freq = 0.45, searching for 5 eigenpairs (default value: 20). Second case: growthrate = 2.50E-3,
    freq = 0.0, searching for 2 eigenpairs (default value: 17).}
  \end{center}
\end{figure}

The input parameters for fig. \ref{fig:eivtargets1}--\ref{fig:eivtargetsa} are
based on the \File{STD_linear_ITG} and \File{simple_TEM} input files found in
\File{doc/input}, with $R/L_{T_i} = 6$, which is near the ITG-TEM transition.
Two unstable eigenmodes are therefore expected.
Scans were made over the target \name{growthrate} and \name{freq} for the harmonic extraction.
The fixed parameters for the eigenvalue solver were
\begin{verbatim}
max_iterations = 30000
tolerance = 1.0e-6
type_solver = 'krylovschur'
type_extraction = 'harmonic'
number_eigenvalues = 3 ! just in case there are more unstable ones than expected
nr_column_vec      = 20
mat_vec_routine    = 1
which_eigenvalues =  1 ! largest magnitude
\end{verbatim}

\begin{figure}[htp!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{Targets}
    \caption{\label{fig:eivtargets1} A scan over target values near the ITG,
      searching for eigenvalues with largest magnitude.
      Plus signs refer to actual found eigenvalues, crosses to 'conjugated'
      eigenvalues, stars to the projections on the growth rate axis. Color
      of the dots refers to the number of found eigenvalues. Please note, that
      a value of 1(or more) requires that the fastest growing eigenmode was found and
      a value of 2(or more) requires that both fastest growing eigenmodes where found.
      Harmonic extraction was used.      
}
  \end{center}
\end{figure}
\begin{figure}[htp!]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{Targets2}
    \caption{\label{fig:eivtargets2} A scan analogous to \ref{fig:eivtargets1}, 
      but target values are near the TEM eigenvalue.
      For the meaning of the symbols please refer to the caption of fig.
      \ref{fig:eivtargets1}.}
  \end{center}
\end{figure}
\begin{figure}
  \begin{center}[htp!]
    \includegraphics[width=0.8\textwidth]{TargetsA}
    \caption{\label{fig:eivtargetsa} A scan over the target values searching for eigenvalue with largest magnitude, with harmonic extraction.
      Scan along \name{freq = 0}, some of the other two scans can also be
      seen.
      For the meaning of the symbols please refer to the caption of fig.
      \ref{fig:eivtargets1}.}
  \end{center}
\end{figure}

If you just want to find the fastest growing
eigenmode, there is not much variation in the speed a solution is found.
 In the examples here, 900-1000 iterations were
needed (if a mode was found at all). This is different though for the violet points,
 which needed 6000-8000 iterations.
For the second eigenvalue the situation is
quite samilar. It is typically found 100-150 iterations after the first one in
the case of fig. \ref{fig:eivtargets1}, whereas for fig. \ref{fig:eivtargets2}
it is found even quicker. Finding a third (stable) eigenvalue may then take several thousand further iterations.  
Wasting computing time unnecessarily can be avoided by setting a suitable value for max_iterations,
setting \name{max_seconds} or by observing the log and placing a \name{gkw.stop} if desired.\\

Robustness and speed of the solver are very dependent on the initial condition.
We found \name{finit='cosine4'} to be a good choice, but when
doing a parameter scan, it can also be useful to restart with the output of the previous
run.  The batch launcher \scripts{gkwnlin} provides the feature (\name{-restart_chain}) for doing 
this.

The diagnostic output is only written when the solver finishes,
i.e. as soon as the requested number of eigenvalues is found, the
maximum number of iterations is reached, 
or the solver stops due to \name{gkw.stop} or \name{max_seconds}. 
No diagnostic output is produced if the code is killed preliminarly by the queuing system.

The diagnostic output differs from the output obtained with \name{method = 'EXP'}.
The main difference is that eigenmode index takes the place of the time in time-dependent datasets.\\
 For example, the first line in \File{fluxes.dat} will contain
the fluxes computed for the first eigenmode, the second line those for
the second eigenmode, and so on.

You will also notice that instead of a single \File{parallel.dat} or restart file, one for each found eigenmode is produced.

Furthermore, there is an diagnostic output \File{eigenvalues} that contains the growth
rate and frequency determined by {\sc slepc} (first column acts as an index).

If eigenvalues are found, you should check that these belong to a
reasonable, physical mode. If
the eigenvalues computed by the \name{diagnos_growth_freq} diagnostics and the one
determined by slepc (found in \File{eigenvalues.dat}) differ this is a sign
that the result may be unphysical. 
Another possibility to check this is to plot the mode structure.
Note that some of the eigenvalues found could be physical while others
are not. In particular for stable eigenmodes, different parallel decompositions 
 may result in different eigenmodes being found,

%-----------------------------------------------------------------------------

\section{Technical details}
\label{sec:eivtechdetails}
The eigenvalue solver in {\sc gkw} is mainly a interface for the library
 {\sc slepc} (which relies on {\sc petsc}).

Most parameters of the \name{eiv_integration} namelist are directly passed down to
 {\sc slepc}.

Note that no matrix is given to {\sc slepc}, as the so called matrix-free approach
is used. This means instead of a matrix we define the action of the matrix
on a vector via a function, provided to {\sc slepc}.
%This approach is used as the actual matrix is not available.
Depending on the settings in the input file, either \code{exp_integration} 
(\name{mat_vec_routine = 1}) is used
to determine the result of the matrix vector multiplication or just the subroutine
\code{calculate_rhs} (\name{mat_vec_routine = 2}).

%-----------------------------------------------------------------------------

\subsection{The actual eigenvalue problem}
\label{subsec:eivproblem}
In general an eigenvalue problem is formulated as
\begin{equation}
  A x = \lambda x, \label{eq:eivgen}
\end{equation}
where $A$ is the operator of the problem and $\lambda$ is the (complex)
eigenvalue.
The \code{exp_integration()} routine computes
\begin{equation}
  f_{t+1} = f_{t} + dt A f_{t} + dt B \phi
\end{equation}
\begin{equation}
  0 = C f_{t} + D \phi
\end{equation}
where $A$, $B$, $C$ and $D$ are matrices and the vectors $f$ and $\phi$
represent all the fields that may be present in the simulation.
Note that GKW works with a vector \code{fdisi} = ($f, \phi$).

In principle it is possible to solve the field equation for $\phi$ and insert
it into the evolution equation to get
\begin{equation}
  f_{t+1} = f{t} + dt A f_{t} - dt B D^{-1} C f_{t}.
\end{equation}
The disadvantage of this approach is, that the inversion of the matrix $D$
and the matrix matrix multiplication most probably would destroy the sparsity
structure of the complete operator. 
This would make the determination of the eigenvalues much more expensive.
For this reason the actual matrix of this problem is not explicitly formulated.

%-----------------------------------------------------------------------------

\subsection{Transformation between eigenvalue and growth rate/frequency}
\label{subsec:eivtrafoeivgf}
The transformation between the complex eigenvalue computed by SLEPC and the growth
rate and frequency in terms of GKW can be derived as follows.
In comparison to the eigenvalue problem formulated in \eqref{eq:eivgen}, 
the eigenvalue problem in GKW is
\begin{equation}
  l f =  (1 + dt N)^n f
\end{equation}
where $N$ stands for the effective matrix of evolution and field equation and $n$ is \code{naverage}.
Identifying $M$ and $(1 + dt N)^n$ results in
\begin{equation}
  \lambda = (1 + dt N) = l
\end{equation}
\begin{equation}
  \ln \lambda = n \ln l
\end{equation}

\begin{equation}
  \frac{\ln \lambda}{n dt} = \frac{\ln l}{n dt}
\end{equation}
The right hand side of these equation is the same relation as between the
change in the norm and the growth rate in the code. We take this as reason to
propose this as general transformation between the eigenvalue found by slepc and
the growth rate and frequency as commonly used in GKW.

\newcommand{\pder}[2]{\frac{\partial#1}{\partial#2}}
The transformation for the tolerance $a$ is
\begin{align}
	a_{gkw} & = & \sqrt{\left( \pder{\lambda_{gkw}}{\lambda_{slepc}} |\lambda_{slepc}|a_{slepc}\right)^2} \\
	& = & \sqrt{\left( \pder{(\log(\lambda_{slepc})/(n_{average} \Delta t))}{\lambda_{slepc}} |\lambda_{slepc}|a_{slepc}\right)^2} \\
	& = & \sqrt{\left( \frac{1}{n_{average} \Delta t} a_{slepc}\right)^2} \\
	& = & \frac{a_{slepc}}{n_{average} \Delta t}
\end{align}

%-----------------------------------------------------------------------------

\section{\label{sec:eivfaq}FAQ}

\paragraph{What is a good starting point for the parameters?}

For the Ritz extraction
to find the two most unstable modes you could try
\begin{verbatim}
&CONTROL
 fac_dtim_est=0.5, naverage= 1, method= 'EIV', meth= 1,
 read_file = .true., irun = 2   ! to restart from previous dominant mode
 ! using gkwnlin -restart_chain
 max_seconds = 10000 ! adjust the runtime
 /
&EIV_INTEGRATION
 max_iterations = 90000
 tolerance=1.0e-4
 number_eigenvalues=2
 type_extraction = 'ritz'
 which_eigenvalues=3 ! search for most unstable modes
 /
\end{verbatim}
It seems that the timestep must not be larger than that required for explicit time integration
but that setting a very small timestep makes the convergence take longer.
Setting the tolerance too small can also cause convergence to take longer, and can even prevent finding
the correct modes.

For the harmonic extraction, you can try the following.
\begin{verbatim}
 &CONTROL
 fac_dtim_est=0.5,
 naverage= 1,
 method= 'EIV',
 meth= 1,
 max_seconds = 10000 ! adjust the runtime
 /
 &EIV_INTEGRATION
 max_iterations = 90000
 tolerance = 1.0e-4
 type_extraction = 'harmonic'
 number_eigenvalues = 2
 nr_column_vec      = 20
 mat_vec_routine    = 1
 which_eigenvalues  = 8 ! look for eigenvalues near the target growthrate
 growthrate = 0.5000 
 freq = 0.0000
 /
 &SPCGENERAL
 finit="cosine4"
 /
 \end{verbatim}
If you increase \code{number_eigenvalues} you may also have to increase
\code{nr_column_vec}.

\paragraph{Does \code{nr_column_vec = 0} (letting 'PETSC_DECIDE') work?}

From the scan done over the size of the subspace (see fig. \ref{fig:eivharmonicncv}
we would expect, that it should work well for at least \code{number_eigenvalues = 1-2}.
For bigger values it might be not optimal.\\
Experience shows that even for \code{number_eigenvalues = 15} eigenvalues 
were found.

\paragraph{I try to compile in single precision/default real but
eiv_integration.F90 fails./Testcases fail after I fix the errors.}
You probably have to configure {\sc pets}c/{\sc slep}c
 \code{--with-precision=single} (as of 3.6.2). It has not been checked if the
code then actually works, though.
Just fixing the bugs that prevent compiling is not enough as there will still
be a type mismatch between the vectors used by {\sc slep}c and the arrays
used by gkw.

\paragraph{How Do I find the number of iterations at which an eigenmode was found?}
You can find this information in the screen output as soon as
converged modes are found. You may stop the run at any time
by creating a \name{gkw.stop} file.

% \paragraph{Is there a way to output eigenmodes as they are found, instead
% of waiting until the end or the maximum number of iterations ?}

% \paragraph{Doesn't the optimal nr_colum_vec increase with the problem size ?}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/global.tex
(1) Added gkw Clone gwk  #3

\chapter{The non-spectral and global version of the code}

This chapter discusses the non-spectral and global version of the code.  (Work in 
progress)

Non-spectral \index{Non-spectral} here refers to the treatment of the radial direction through the use of finite 
difference methods, and is switched on through the switch 
\begin{equation}
\hbox{\tt spectral_radius = .false.}
\end{equation}
in the control namelist. The non-spectral representation can be used both in the local limit
(which we will refer to as flux-tube) as well as in the global case. 
The term \emph{global} \index{Global simulations} is used as synonym for allowing for
profile effects, i.e. allowing plasma and geometry parameters to 
be a function of the radius. A global run is selected by setting 
\begin{equation}
\hbox{\tt flux_tube = .false.}
\end{equation}
in the \name{CONTROL} namelist. Naturally, global simulations can only be performed with a non-spectral 
representation. Global simulations can still be $\delta f$ simulations, i.e. the perturbed distribution 
function can be assumed small compared to the background. In this case energy is not conserved since
the parallel velocity nonlinearity (cf. section \ref{sec:velocity-nonlinearity})is neglected. Energy conserving simulations will be referred to as 
full-f simulations below\index{full-f simulations}, and can be run setting 
\begin{equation}
\hbox{\tt lpar_vel_nl = .true.} 
\end{equation}
in the \name{CONTROL} namelist.

Two points should be noted: First, although you can run with the parallel velocity nonlinearity 
in a flux-tube run, there is no point since the expansion around a local flux surface means that energy 
cannot be conserved (It can easily be shown that the integrated kinetic energy in the perturbed 
distribution is always zero in the local limit, while the integrated energy in the field must be postive. 
The two can then not balance each other). 
Therefore, global full-f, global $\delta f$, and local $\delta f$ make sense, but local 
full-f does not, unless one wants to study some distinct property of the velocity nonlinearity.

Second, care must also be taken using global $\delta f$. The turbulence will lead to a 
rapid profile evolution that is represented by the perturbed distribution. The condition 
$\delta f = {\cal O}[ \rho_* F_M ]$ is then easily violated. 
One way to keep $\delta f$ small is through the Krook operator (see the section below), which damps the 
perturbed distribution on a specified timescale and acts through this damping as a source / sink of energy. 
But even with the use of the Krook operator the ordering is not necessarily satisfied.

\section{Complete Set of Equations}
\label{sec:global-compl-set-equat}

With the new normalization as described in \ref{sec:normalisation-global-specialities} the equations of motion are
\begin{equation}
\label{eqs:global-complete-set}
\pd{g}{t} = {\rm I} + {\rm II} + {\rm III} + {\rm IV} + {\rm V} + {\rm VI}+ {\rm VII}+ {\rm VIII},
\end{equation}
with 
\pagebreak
%\abovedisplayskip
%\belowdisplayskip
%\begin{small}
\begin{align}
\label{eq:gI}{\rm I} &= - v_\parallel {\bf b} \cdot \nabla f \rightarrow - w v_{\parallel} {\cal F} \pd{f}{s}, \\
\noalign{\vskip 0.4 truecm} 
%
\label{eq:gII}{\rm II} &= - {\bf v}_D \cdot \nabla f \rightarrow \cr 
\noalign{\vskip 0.4 truecm} 
& \hspace{0.5cm} -\frac{\rho_*}{Z} \left[ T_G E_D {\cal D}^\alpha + T_G v_{\parallel}^2 \beta^\prime  {\cal E}^{\psi \alpha} + 2 m w v_{\parallel}\Omega{\cal
H}^\alpha + m \Omega^2 I^\alpha + Z{\cal E}^{\beta \alpha} \pd{\Phi}{x_\beta}\right] 
\pd{ f}{x_\alpha}  \\
\noalign{\vskip 0.4 truecm} 
%
\label{eq:gnl-term}
{\rm III} &= - {\bf v}_\chi \cdot \nabla g \rightharpoonup - \rho_*^2 \pd{\chi}{x_\beta} 
{\cal E}^{\beta \alpha} \pd{g}{x_\alpha} \\
\noalign{\vskip 0.4 truecm} 
%
\label{eq:gIV}{\rm IV} &= + \frac{\bf b}{ m }\cdot(\mu \nabla B + \nabla \cfen) \pd{ f }{v_\parallel} \rightarrow
w \left(\mu B {\cal G} + \pd{T}{2 T_G} {\cal F} \pd{ {\cal E}_\Omega }{s} \right) 
\pd{  f }{v_{\parallel}},\\
\noalign{\vskip 0.4 truecm} 
%
\label{eq:gV}{\rm V} &= - {\bf v}_\chi \cdot \nabla F_M \rightarrow  
\rho_* \pd{ \chi }{x_\alpha}  
 {\cal E}^{\alpha \psi} \biggl [ \frac{1}{L_n} + E_T \frac{1 }{ L_T} + \biggl ( \frac{2 m w v_{\parallel}}{T} 
 \frac{R B_t }{ B} + \frac{2 m \Omega }{ T} {\cal J} \biggr ) u^\prime + \frac{m \Omega^2 }{ T}{\cal L} \biggr ] F_{M} \\
\noalign{\vskip 0.4 truecm} 
%
\label{eq:gVI}{\rm VI} &= - {\bf v}_D \cdot \nabla F_M \rightarrow  \frac{1 }{ Z} \left[ T_G E_D {\cal D}^\psi + 2 m w v_{\parallel}
\Omega{\cal H}^\psi + m \Omega^2 I^\psi + Z{\cal E}^{s \psi} \pd{ \Phi }{s}\right] \\
\noalign{\vskip 0.4 truecm}
& \hspace{2.3cm} \times \biggl [ \frac{1 }{ L_n} + E_T \frac{1 }{ L_T} + \biggl ( \frac{2 m w v_{\parallel}}{ T} 
 \frac{R B_t }{ B} + \frac{2 m \Omega }{ T} {\cal J} \biggr ) u^\prime + \frac{m \Omega^2 }{ T}{\cal L} \biggr ] F_{M} ,\\ 
\noalign{\vskip 0.4 truecm} 
%
\label{eq:gVII}{\rm VII} &= - \frac{Z e }{ T} v_\parallel {\bf b} \cdot \nabla \langle \phi \rangle F_M \rightarrow - 
\frac{Z }{ T} w v_{\parallel} {\cal F} \pd{ {\langle \phi \rangle} }{s} F_{M} ,\\
\noalign{\vskip 0.4 truecm} 
%
\label{eq:gVIII}{\rm VIII} &= - \frac{Z e }{ T}{\bf v}_D \cdot \nabla \langle \phi \rangle F_M  \rightarrow \cr 
\noalign{\vskip 0.4 truecm} 
& \hspace{0.2cm} - \frac{\rho_* }{ T} \left[ T_G E_D {\cal D}^\alpha + T_G \beta^\prime v_{\parallel}^2 {\cal E}^{\psi \alpha} + 
{2 m w}v_{\parallel}\Omega{\cal H}^\alpha + {m\Omega^2 }{\cal I}^\alpha 
+ {Z} {\cal E}^{\beta \alpha} \pd{ \Phi }{x_\beta} \right] 
\pd{ \langle \phi \rangle }{x_\alpha} \, F_{M} \cr 
\noalign{\vskip 0.4 truecm} 
%
\end{align}
Here, all quantities are normalized, i.e. $w$ in these equations 
refers to $w_N$, etc.
Moreover we use the generalised potential 
\begin{equation} 
\chi = {\langle \phi \rangle} - 2 w v_{\parallel} {\langle A_{\parallel} \rangle}  %+ \frac{2 T_G }{ Z} \mu B_{\parallel} 
\end{equation} 
and 
\begin{equation} 
g = f + \frac{2 Z }{ T} w v_{\parallel} {\langle A_{\parallel} \rangle} F_{M}, 
\end{equation}
\begin{equation}
E_D = v_{\parallel}^2 + \mu B , \qquad 
E_T = \frac{T }{ T_G} \left [ v_{\parallel}^2 + 2 \mu B  + {\cal E}_\Omega \right ]  - \frac{3}{ 2} .
\end{equation}
\begin{equation} 
\frac{1 }{ L_N} \equiv - \frac{1 }{ n}\pd{ n }{\psi} \qquad 
\frac{1 }{ L_T} \equiv - \frac{1 }{ T}\pd{ T }{\psi} \qquad 
u^\prime      \equiv - \pd{ \omega_\phi }{\psi} 
\end{equation}
The equations above apply to each of the species individually.

In the equations the spatial derivatives of the perturbed distribution function appear, and the Einstein summation convention has been used. 
To shorten the notation, the sum over $\alpha$ also includes $\alpha = s$.
The directions $s$ and $\psi$ are treated in position space, i.e. finite differences discretise  $\pd{}{x^\alpha}$.
\begin{align}
  \pd{}{\psi} &= \frac{1}{R_\tref} \pd{}{\psi_N} \\
\pd{}{s} &\sim \orderof(1)\text{$s$ is naturally normalised}
\label{eq:global-s-normalisation}
\end{align}
Note that the ordering nevertheless assumes that flucuations vary on the lengthscale $\rho_\tref$, not $R_\tref$.
\begin{align}
\pd{\hat f_N}{\psi}
&= \frac{1}{\rho_\tref} \pd{\hat f_N}{\psi_N} 
= \frac{R_\tref}{\rho_\tref} \frac{1}{R_\tref}\pd{\hat f_N}{\psi_N} \\
&= \frac{1}{\rho_*}\left(\pd{\hat f_N}{\psi}\right)_N
\label{eq:global-dpsi-normalisation}
\end{align}
For the $\zeta$ coordinate principally we have
\begin{align}
  \pd{}{\zeta} \sim \orderof(1)\text{$\zeta$ is naturally normalised}
\end{align}
but actually $\zeta$-derivatives only of fluctuating quantities appear
(not of the background - this is tokamak-only)
where the ordering assumes
\begin{align}
  \pd{\hat f_N}{\zeta} = \frac{1}{\rho_*}\left(\pd{\hat f_N}{\zeta}\right)_N
\end{align}
In the code, the $\zeta$ grid is treated in spectral representation, and we write
\begin{align} 
\pd{\hat f_N}{\zeta} &\rightarrow {\rm i} k_\zeta \hat f_N = {\rm i} \frac{1}{\rho_*}k_{\zeta N} \hat f_N
\label{eq:global-kzeta-normalisation}
\\
\left(\pd{\hat f_N}{\zeta}\right)_N &\rightarrow {\rm i} k_{\zeta N} \hat f_N
\end{align}

In contrast to the perpendicular ones \eqref{eq:global-dpsi-normalisation} and \eqref{eq:global-kzeta-normalisation}, the terms containing
$s$-derivatives \eqref{eq:global-s-normalisation} are $\rho_*$ smaller than the derivatives
perpendicular to the magnetic field.
For this reason they are usually neglected on the level of the evolution equations (although this may break physical system properties, see section \ref{sec:velocity-nonlinearity}).
These ``higher order $\rho_*$ terms''  have been studied in connection with momentum transport in Ref.~\cite{SUN13}, and must be explicitly switched on through the input file, to involve them.

\section{Missing implementation}
\label{sec:global-missing}

Not all physics which is available for flux-tube systems in GKW has already been implemented for the global case or makes sense.

\subsection*{Plasma Rotation}
\label{sec:global-plasma-rotation}

The frame rotation $\Omega$ is in both flux-tube and global case not a function of the radius. 
This is more than just a choice. 
The beauty of a corotating frame in the flux-tube case lies in the fact that rotational 
phenomena can be more clearly adressed when the background rotation is zero with respect to the frame.

Now, if in the global case the frame rotation is a function of the radius, one could have the idea to choose a frame which also rotates differentially.
The issue with such a frame is, that in general two stationary points in the co-moving frame 
shear apart in the laboratory frame. 
In short, the metric tensor would be a function of time, which is highly undesirable.

By choosing a frame which rotates with at single frequency $\Omega$, the frame is not a comoving frame everywhere.
Such a frame is equivalent to the laboratory frame and will not simplify the examination of the system.

The co-moving frame description of the rotation is, therefore, useful for the flux tube case, but less so 
for the global case. 
The use of $\Omega$ in the global case is allowed, but the implementation is not entirely consistent (the 
${\cal E}_\Omega$ quantity, for instance, is not a function of the radius). 
The user is therefore strongly discouraged from using the plasma rotation $\Omega$ in the global case. 
Momentum fluxes due to the $u^\prime$ however can be calculated. 
In the future a rotation profile will be implemented through a shift of the perturbed distribution, rather than 
a coordinate transformation.

\section{Krook operator}
\label{seckrookoperator}
\index{Krook operator} 
In a global run the profiles relax. Without any sources or sinks this leads 
to a decay of the turbulence since it is no longer driven. One of the methods
to obtain a 'stationary' turbulence run is to enable the Krook operator via the 
settings in the \texttt{KROOK} namelist. The implementation for \texttt{krook_option=1} 
picks an operator which conserves density and parallel velocity, but not 
energy. \\
The Krook operator $\mathcal{C}_K$ takes the shape of a source or damping term in the gyrokinetic equation.
\begin{align}
  \pd{ f (\mathbf{X}, v_\parallel,\mu)}{ t} + ... &= \mathcal{C}_K
\end{align}
% Below, the change of a quantity due to the Krook operator will be denoted by $|_K$, like this
% \begin{align}
%   \pd{ f (\mathbf{X}, v_\parallel,\mu)}{ t}\biggr|_K &= \mathcal{C}_K
% \end{align}

In the following, the operator implemented for \texttt{krook_option=1} is described. 
First the density moment is calculated and stored in the array \texttt{sss} 
\begin{equation}
\mathtt{sss}  = \tilde n = \int f \, {\rm d}^3 {\bf v}
\end{equation}
FIXME this was not the density moment...?|
This density moment is used in the calculation of the damping term 
\begin{equation}
%\pd{ f (v_\parallel,\mu)}{ t} \biggr \vert_{K}  
\mathcal{C}_K
= - \gamma_K \biggl [   
\frac{1 }{ 2} [ f(v_\parallel,\mu) + f(-v_\parallel,\mu)] -  \tilde n \bar F_M(v_\parallel,\mu) \biggr ] 
\label{eq:krook-option1}
\end{equation}
The Maxwellian $\bar F_M$ here is not precisely identical to $F_M$ from \eqref{eq:maxwell}, but defined such that it satisfies 
\begin{equation}
\int \bar F_M \, {\rm d}^3 {\bf v} = 1 \qquad \int \frac{1}{ 2} m v^2 \bar F_M \, {\rm d}^3 {\bf v} = \frac{3 }{ 2} T 
\end{equation}
where T is the local background Temperature. 
Integration of \eqref{eq:krook-option1} over velocity space shows that this operator does not constitute a source or sink of particles.
\begin{align}
\pd{}{ t} \biggr \vert_{K} \int f \, {\rm d}^3 {\bf v} =
 \int \mathcal{C}_K \, {\rm d}^3 {\bf v}
 &=   - \gamma_K\int \, {\rm d}^3 {\bf v}  \biggl [   
\frac{1 }{ 2}[f(v_\parallel,\mu) + f(-v_\parallel,\mu)] -  \tilde n \bar F_M(v_\parallel,\mu) \biggr ] 
\nonumber\\
&=   - \gamma_K\biggl [   
\frac{1 }{ 2}\int \, {\rm d}^3 {\bf v}   f(v_\parallel,\mu) 
+\frac{1 }{ 2} \int \, {\rm d}^3 {\bf v}  f(-v_\parallel,\mu)] 
-  \int \, {\rm d}^3 {\bf v}  \tilde n \bar F_M(v_\parallel,\mu) \biggr ] 
\nonumber\\
&=   - \gamma_K [ \tilde n  - \tilde n] 
\nonumber\\
&= 0
\end{align}
FIXME for me this is not 0, so I suppose $\tilde n$ is not supposed to be a density moment?|g
That is, under the action of the term \eqref{eq:krook-option1} density is locally conserved.

The parallel velocity moment is
\begin{equation}
%\pd{}{ t} \biggr \vert_{K} \int v_\parallel f \, {\rm d}^3 {\bf v} = 0 
\int v_\parallel \mathcal{C}_K \, {\rm d}^3 {\bf v} = 0 
\end{equation}
i.e. also the parallel momentum is locally conserved. Of course, the energy is not 
\begin{equation}
%\pd{}{ t} \biggr \vert_{K} \int \frac{1 }{ 2 } m v^2 f {\rm d}^3 {\bf v}  = 
\int \frac{1 }{ 2 } m v^2 \mathcal{C}_K {\rm d}^3 {\bf v}  = 
- \gamma_K \int \left [ \frac{1 }{ 2} m v^2 - \frac{3 }{ 2} T_0 \right ] f \, {\rm d}^3 {\bf v} 
\end{equation}
Since density is conserved the perturbed distribution can not be uniformly (in velocity space) damped. 
The expression above makes clear that the distribution function is reduced for energies more than the 
thermal energy of the bulk, whereas it is growing for smaller energies.

\texttt{krook_option=3} is similar to the above apart from the lack of velocity space symmetrisation, i.e
\begin{equation}
\pd{ f (v_\parallel,\mu)}{ t} \biggr \vert_{K}  = - \gamma_K \biggl [   
\frac{1 }{ 2} [ f(v_\parallel,mu)  -  \tilde n \bar F_M(v_\parallel,\mu) \biggr ]. 
\end{equation}
This version does not conserve the parallel momentum and damps away any parallel flows.

\texttt{krook_option=4} is a hybrid of these two.  The momentum conserving operator is applied to all non-zero toroidal modes, while the parallel flow damping operator is applied to the $n=0$ mode (the damping rate can be modified by a prefactor gamkpre in the input file) so the perturbed equilibrium toroidal flow is damped away without modifying heavily the turbulence.

\section{Profile functions in GKW}
\label{secprofilefunctions}
\index{Profile functions} 
In the global version one can choose various analytic profiles through the 
switches '\name{dens_prof_type}' and '\name{temp_prof_type}' in the 
\name{SPECIES} namelist. Allowed are at 
present the option '\name{const}', '\name{cosh2}', '\name{tanh}', 
'\name{exp_tanh}', '\name{orb}', '\name{orb3}', '\name{exp_poly3}', '\name{exp_poly6}' as well as '\name{file}'. 
Specifying the latter will make GKW read the profile data from the file \name{input.prof}, whereas all the others denote an analytic expression for the respective profile.

The parameters needed for the analytic profile expression are given 
in the arrays 'dens_prof_coef' and 'temp_prof_coef'. At present a maximum 
of five parameters is available.

Below the quantity $G$ stands for either density or temperature and $G'$ is its normalised gradient.
\begin{equation}
G' = \frac{1 }{ G} \d{ G }{ \psi} 
\end{equation}

Besides the profiles of density and temperature one can also specify the 
$q$ profile. To select a specific profile one sets the \name{prof_type} 
option in the \name{GEOM} namelist. The magnetic shear is calculated 
consistently with the profile specified. Valid options are at present: 
'\name{parabolic}', '\name{parabolic2}', '\name{orb}', '\name{wesson}', 
'\name{rexp}', '\name{mishchenko}'. The parameters needed for the analytic $q$ 
profile are given to the code through \name{qprof_coef} array in the namelist \name{GEOM}.

The figures \ref{comparisonProfile} and \ref{comparisonDprofile} show some of
the temperature/density profiles and the corresponding gradient length, respectively.
All the profiles use the same parameter set $(1.0, 5.0, 0.15, 0.1, 0.1$ and
thus might not be physical.\\
Examples for the safety factor and the corresponding shear are shown in fig.~\ref{comparisonQprofile}
and \ref{comparisonSprofile}, respectively. All the profiles use the same
parameter set $(0.9, 5, 3)$ and thus might not be physical.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{comparisonProfile}
    \caption{\label{comparisonProfile}Comparsion of some possible profiles for density/temperature. All use the same set of parameters (see text) and thus might be nonphysical.}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{comparisonDprofile}
    \caption{\label{comparisonDprofile}Comparsion the gradients length of some possible profiles for density/temperature. All use the same set of parameters (see text) and thus might be nonphysical.}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{comparisonQprofile}
    \caption{\label{comparisonQprofile}Comparison of some possible safety factor profiles. All use the same set of parameters (see text) and thus might be nonphysical.}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{comparisonSprofile}
    \caption{\label{comparisonSprofile}Comparison of the shear for some possible safety factor profiles. All use the same set of parameters (see text) and thus might be nonphysical.}
  \end{center}
\end{figure}

\subsection{Density/Temperature profile option: cosh2}
Allows to have a plateau like gradient profile, i.e. almost constant in the
middle and going to zero at the border.

\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
w      = {\rm dens\_prof\_coef}(4)   &\hbox{the normalized width of the plateau $\Delta x / R_0$}\cr
\Delta = {\rm dens\_prof\_coef}(5)   &\hbox{Width that determines the rise of the $R/L_T$ profile}
\end{eqnarray*}
\begin{equation}
X = {\rm min}({\rm max}(X_0-w/2,\psi),X_0+w/2.)  
\end{equation}
\begin{equation}
G = G_0 \exp\biggl [ -\frac{R }{ L_G} (X-X_0) + \frac{R}{ L_G} \Delta \tanh \biggl ( \frac{X - X_0 + w / 2 }{ 
\Delta} \biggr ) + \frac{R}{ L_G} \Delta \tanh \biggl ( \frac{X - X_0 - w / 2 }{ 
\Delta} \biggr ) \biggr ] 
\end{equation}
\begin{equation}
G^\prime = \frac{R}{ L_G} \biggl ( 1 - \frac{1 }{ \cosh^2 ((X - X_0 + w / 2)/\Delta)} - 
 \frac{1 }{ \cosh^2 ((X - X_0 - w / 2)/\Delta)}\biggr ) 
\end{equation}
\begin{equation}
G_{\rm norm} = G_0   
\end{equation}

\subsection{Density/Temperature profile option: 'const'}

\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature }
\end{eqnarray*} 
\begin{equation}
G = G_0 
\end{equation}
\begin{equation}
G^\prime = 0
\end{equation}
\begin{equation}
G_{\rm norm} = G_0 
\end{equation}

\subsection{Density/Temperature profile option: 'exp_tanh'}

\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
w      = {\rm dens\_prof\_coef}(4)   &\hbox{the normalized width of the gradient profile $\Delta x / R_0$}\cr
\end{eqnarray*} 
\begin{equation}
G = G_0 \exp \biggl ( - \frac{R}{ L_G} w \tanh \biggl ( \frac{\psi-X_0 }{ w} \biggr ) \biggr ) 
\end{equation}
\begin{equation}
G^\prime = \frac{R }{ L_G} \frac{1 }{ \cosh^2 ((\psi-X_0)/w)}
\end{equation}
\begin{equation}
G_{norm} = G_0
\end{equation}

\subsection{Density/Temperature profile option: '1_m_exp_tanh'}
\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
w      = {\rm dens\_prof\_coef}(4)   &\hbox{the normalized width of the gradient profile $\Delta x / R_0$}\cr
\end{eqnarray*}
\begin{equation}
G = 1 - G_0 \exp \biggl ( - \frac{R}{L_G} w \tanh \biggl ( \frac{\psi-X_0}{w} \biggr ) \biggr )
\end{equation}
\begin{equation}
G^\prime = \frac{R}{L_G} \frac{1}{\cosh^2 \frac{\psi-X_0}{w}}
\end{equation}
\begin{equation}
G_{norm} = 1 - G_0
\end{equation}

\subsection{Density/Temperature profile option: exp_poly3}

Results in a quadratic profile of the gradient length. Limiting cases of linear 
(for a = 0) and constant (for a=0 and b=0) are included. Some care has to be taken 
to get physical profiles. To get a gradient length that is positive over the whole range, 
the inequality $b^2 < 4a$ has to be fullfilled.
\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
a      = {\rm dens\_prof\_coef}(4)   &\hbox{constant for the quadratic term}\cr
b      = {\rm dens\_prof\_coef}(5)   &\hbox{constant for the linear term}
\end{eqnarray*}
\begin{equation}
G = G_0  \exp \left[ - \frac{R }{ L_G} \left(  \frac{a}{3} (\psi - X_0)^3
+ \frac{b}{2} (\psi - X_0)^2  + (\psi - X_0) \right) \right]
\end{equation}
\begin{equation}
G^\prime = \frac{R }{ L_G} \left[ a (\psi - X_0)^2 + b (\psi - X_0) + 1 \right]
\end{equation}
\begin{equation}
G_{\rm norm} = G_0 
\end{equation}

\subsection{Density/Temperature profile option: exp_poly6}

Results in a fifth and third order term as well as a constant for the
gradient length.
\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
a      = {\rm dens\_prof\_coef}(4)   &\hbox{constant for the fifth order term}\cr
b      = {\rm dens\_prof\_coef}(5)   &\hbox{constant for the third order term}
\end{eqnarray*}
\begin{equation}
G = G_0 \exp \left[ - \frac{R }{ L_G} \left(  \frac{a}{6} (\psi - X_0)^6
                                          + \frac{b}{4} (\psi - X_0)^4
                                          +             (\psi - X_0) \right) \right]
\end{equation}
\begin{equation}
G^\prime = \frac{R }{ L_G} \left[  a (\psi - X_0)^5
                           +     b (\psi - X_0)^3
                           + 1 \right]
\end{equation}
\begin{equation}
G_{\rm norm} = G_0
\end{equation}

\subsection{Density/Temperature profile option : orb}

This produces profiles equivalent to one of the options used in ORB5 (NEMORB). 
The calculation is a little more involved because these orb profiles are 
defined as function of a different radial coordinate.

\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
w      = {\rm dens\_prof\_coef}(4)   &\hbox{Width of the profile}\cr
X_e    = {\rm dens\_prof\_coef}(5)   &\hbox{minor radius of the plasma edge $X_e = a /R_0$}
\end{eqnarray*}
The values of $\bar {q}$ as used in ORB at the edge and in the centre 
\begin{equation} q_a = q(a) \sqrt{1 - X_e^2} \end{equation}
\begin{equation} q_0    = q(0) \end{equation}
Rescale the gradient length to have exactly the input value at the $s_0$ 
coordinate used in ORB
\begin{equation}
\frac{R}{ L_{G*}} = \frac{R/L_G }{ 1 - 1 / \cosh^2(X_0/w)} 
\end{equation}
Calculate the $s$ coordinate used in ORB 
\begin{equation} 
s = \sqrt{\left ( \frac{\log(1 + (q_a - q_0)\psi^2/(q_0 X_e^2)) }{ 
\log(1 + (q_a - q_0)/ q_0 )} \right ) }
\end{equation}
\begin{equation}
D(X) = \exp \left ( \frac{R }{ L_{G*}} \frac{s^2 }{ \cosh(x_0/w)^2} - 2\frac{R }{ L_{G*}} 
s w \tanh \left ( \frac{s - X_0 }{ w} \right ) \right ) 
\end{equation}
\begin{equation}
G(\psi) = G_0 D(\psi) / D(X_0) 
\end{equation}
\begin{equation}
G^\prime = \frac{R }{ L_{G*}} \left ( \frac{1 }{ \cosh^2((s-X_0)/w)} - 
\frac{1 }{ \cosh^2(x_0/w) } \right ) 
\end{equation}
\begin{equation}
G_{\rm norm} = G_0 
\end{equation}

\subsection{Density/Temperature profile option: 'orb3'}

One of the profiles used in ORB (NEMORB) 
\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
w      = {\rm dens\_prof\_coef}(4)   &\hbox{Width of the profile}\cr
\Delta = {\rm dens\_prof\_coef}(5)   &\hbox{width that determines the rise of the profile}
\end{eqnarray*}
%\begin{equation}
%X = {\rm min}({\rm max}(\psi,X_0-w/2),X_0+w/2)
%\end{equation}
\begin{equation}
G = G_0 \exp \left ( - \frac{1}{2}\frac{R }{ L_G} \Delta \log \left ( 
\frac{\cosh((X - X_0 + w )/ \Delta) }{ \cosh((X-X_0-w)/\Delta) } \right ) \right)
\end{equation}
\begin{equation}
G^\prime = \frac{1}{2} \frac{R}{ L_G} \left ( \tanh \left ( \frac{X-X_0+w}{ \Delta} \right) 
- \tanh \left ( \frac{X - X_0 - w }{ \Delta} \right ) \right) 
\end{equation}
\begin{equation}
G_{\rm norm} = G_0 
\end{equation}

\subsection{Density/Temperature profile option: 'tanh'}
Profile is Taylor approximation of orb3, the gradient is chosen to be the same?\\
The tanh profile 
\begin{eqnarray*}
G_0    = {\rm dens\_prof\_coef}(1)   &\hbox{the density / temperature at $x = X_0$} \cr
R/L_G  = {\rm dens\_prof\_coef}(2)   &\hbox{gradient length R/L_G at $x = X_0$}\cr
X_0    = {\rm dens\_prof\_coef}(3)   &\hbox{Radial location of the maximum gradient}\cr
w      = {\rm dens\_prof\_coef}(4)   &\hbox{Width of the profile}\cr
\Delta = {\rm dens\_prof\_coef}(5)   &\hbox{width that determines the rise of the profile}
\end{eqnarray*}
\begin{equation}
G = G_0 \left[ 1.- \frac{\Delta}{2}\frac{R }{ L_G} \left ( \log \left (\cosh \left (\frac{\psi-X_0+w/2}{ 
\Delta} \right ) \right) - \log \left ( \cosh \left (\frac{\psi-X_0-w/2}{ \Delta} \right ) \right ) \right ) 
\right ] 
\end{equation}
\begin{equation}
G^\prime = \frac{1}{2} \frac{R }{ L_G} \left ( \tanh \left ( \frac{\psi-X_0+w/2)}{ \Delta} \right ) -  
\tanh \left ( \frac{\psi-X_0-w/2}{ \Delta} \right )  \right )
\end{equation}
The norm is the value of the profile at $X_0$, not the $G_0$ value (which should be equal to $G_0$)
\begin{equation}
G_{\rm norm} = G(X_0) 
\end{equation}

\subsection{q-profile option: parabolic}

Parabolic q-profile 
\begin{eqnarray*}
q_0    = {\rm qprof\_coef}(1)   &\hbox{$q$ value at the axis} \cr
a      = {\rm qprof\_coef}(2)   &\hbox{coefficient for the quadratic term}
\end{eqnarray*}
\begin{equation}
q = q_0 + a \psi^2 
\end{equation}
with $\psi = r /R_0$ in circular geometry. 
\begin{equation}
\hat s = \frac{ 2 a \psi^2 }{ q_0 + a \psi^2} 
\end{equation}

\subsection{q-profile option: parabolic2}

Second degree polynomial 
\begin{eqnarray*}
q_0    = {\rm qprof\_coef}(1)   &\hbox{$q$ value at the axis} \cr
a      = {\rm qprof\_coef}(2)   &\hbox{coefficient for the linear term}\cr
b      = {\rm qprof\_coef}(3)   &\hbox{coefficient for the quadratic term}
\end{eqnarray*}

\begin{equation}
q = q_0 + a \psi + b \psi^2 
\end{equation}
\begin{equation}
\hat s = \frac{a \psi + 2 b \psi^2}{q}
\end{equation}

\subsection{q-profile option: orb}

For circular geometry one of the options in ORB is 
\begin{eqnarray*}
q_0    = {\rm qprof\_coef}(1)   &\hbox{$q$ value at the axis} \cr
a      = {\rm qprof\_coef}(2)   &\hbox{coefficient for the quadratic term}
\end{eqnarray*}
\begin{equation}
q = \frac{q_0 + a \psi^2}{\sqrt{1 - \psi^2}} 
\end{equation}
\begin{equation}
\hat s = \frac{2 a \psi^2 }{ q_0 + a \psi^2} - \frac{\psi^2 }{ 1 - \psi^2} 
\end{equation}

\subsection{q-profile option: wesson}

The q-profile used in Wesson. For this profile $qprof_coef(1) \psi^2 < 1$ has to hold
over the entire profile, otherwise the term in the logarithm gets negative. 
If this happens, the simulation will abort.
\begin{eqnarray*}
D_w    = {\rm qprof\_coef}(1)   &\hbox{The aspect ratio squared}  (R/a)^{2} \cr
\nu    = {\rm qprof\_coef}(2)   &\hbox{Exponent of the q-profile.}\cr
q_0    = {\rm qprof\_coef}(3)   &\hbox{q value on the axis}\cr
\end{eqnarray*}
\begin{equation}
q = \frac{q_0 \psi^2 D_w }{ 1 - \exp((\nu+1)\log(1 - D_w \psi^2)) } 
\end{equation}
\begin{equation}
\hat s = 2 \left (1 - \frac{q(\nu + 1)}{q_0}\exp\left ( \nu \log(1-D_w \psi^2) \right) \right)
\end{equation}

\subsection{q-profile option: rexp}

Exponential q-profile, with a cut off $q_c$. Values of the exponent lower than 
$q_c$ are set to $q_c$ with the shear then set to zero. 
\begin{eqnarray*}
a      = {\rm qprof\_coef}(1)   &\hbox{1/length for exponential increase } \cr
q_v    = {\rm qprof\_coef}(2)   &\hbox{Multiplication factor}\cr
q_c    = {\rm qprof\_coef}(3)   &\hbox{Cut off value}
\end{eqnarray*}
\begin{equation}
q = {\rm max}(q_v \psi \exp(a \psi),q_c) 
\end{equation}
\begin{equation}
\hat s = 1 + \psi a \quad \hbox{for $q>q_c$} 
\end{equation}

\subsection{q-profile option: mishchenko}

This has for example been used in Mishchenko and Zocco, Phys. Plasmas 19. 
\begin{eqnarray*}
q_0      = {\rm qprof\_coef}(1)   &\hbox{The q-value on the axis} \cr
a        = {\rm qprof\_coef}(2)   &\hbox{Norm factor for $\psi$}\cr
\nu      = {\rm qprof\_coef}(3)   &\hbox{Exponent}
\end{eqnarray*}
\begin{equation}
q = q_0 + (1 - q_0) (\psi/ a)^\nu
\end{equation}
\begin{equation}
\hat s = \frac{\nu (1 - q_0) (\psi / a)^\nu }{ q} 
\end{equation}

\subsection{q-profile option: file}

By specifying \name{prof_type='file'}, the $q$-profile is read from the file '\name{input.prof}'.
The block of geometry data in the file '\name{input.prof}' must be preceeded by a line starting with the text \verb|#Geometry|.
If Miller geometry is chosen, the data is expected to be given in 13 columns and \name{n_x_grid} rows.
The quantities in each row are \name{xgr},\name{qx},\name{shatx},\name{kappax},\name{deltax},\name{squarex},\name{ skappax},\name{sdeltax},\name{ssquarex},\name{Zmilx},\name{dRmilx},\name{dZmilx},\name{gradpx}, in this order.
\begin{verbatim}
#Geometry
  0.01022949   1.26822967   0.01021214   1.29296310   0.00248834  -0.00528476   0.00009362   0.00247875  -0.00003630   0.03824134  -0.00964105  -0.00147389   0.00000000
  0.01068848   1.26882114   0.01104428   1.29296896   0.00260034  -0.00528594   0.00010279   0.00259195  -0.00004001   0.03824021  -0.01007150  -0.00154127   0.00000000
  0.01114746   1.26943302   0.01189906   1.29297521   0.00271252  -0.00528711   0.00011236   0.00270567  -0.00004393   0.03823903  -0.01050195  -0.00160874   0.00000000
  0.01160645   1.27006485   0.01277532   1.29298176   0.00282480  -0.00528831   0.00012233   0.00281975  -0.00004801   0.03823780  -0.01093237  -0.00167609   0.00000000
...
\end{verbatim}
Otherwise, the data is expected to be given in 3 columns and \name{n_x_grid} rows.
The quantities in each row are \name{xgr},\name{qx},\name{shatx}, in this order.
\begin{verbatim}
#Geometry
  0.01022949   1.26822967   0.01021214
  0.01068848   1.26882114   0.01104428
  0.01114746   1.26943302   0.01189906
  0.01160645   1.27006485   0.01277532
...
\end{verbatim}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/implementation.tex
(1) Added gkw Clone gwk  #3

\chapter{Numerical implementation}

In this section we outline the details of the numerical solution of the equations. GKW uses a combination of finite difference and spectral methods.
The turbulence in the plane perpendicular to the magnetic field is homogeneous and the
solution in the $\zeta, \psi$ plane is represented by Fourier modes, as mentioned
previously. All other directions are treated using finite difference techniques.
In what follows $N_{\rm mod}$ and $N_x$ are the number of bi-normal ($\zeta$ direction) and radial ($\psi$ direction) modes
respectively, $N_{sp}$ is the number of kinetic species, $N_s$ the number of grid points along the field line. $N_{v\parallel}$ is the number of grid points 
in the parallel velocity direction and $N_\mu$ the number of magnetic moment grid points.

\section{Derivatives along the magnetic field line and in the parallel velocity
direction}
\label{sec:finite-differences}
Terms I, IV and VII on the right hand side of \Eq{eqs:complete-set}, involving a derivative along the magnetic field line or in the parallel velocity direction are considered
here. These terms correspond to an advection with a spatially dependent advective velocity
and can be written in the form $-v(s)\frac{\partial g}{\partial s}$. The second and fourth-order centred-differences used for these
terms can then be written
% \label{general_advection}
%\end{equation}
\begin{equation} 
v {\partial g \over \partial s} \rightarrow v_i {- g_{i-1} + g_{i+1} \over 2 \Delta s} 
- D \vert v_i \vert {g_{i-1} - 2 g_i + g_{i+1} \over 2 \Delta s},
\label{eq:s-2nd-order}
\end{equation} 
\begin{align}
v {\partial g \over \partial s}  \rightarrow  v_i {g_{i-2} - 8 g_{i-1} + 8 g_{i+1} -g_{i+2} \over 
12 \Delta s}
 %\cr \noalign{\vskip 0.2 truecm} & &
 - D \vert v_i \vert {- g_{i-2} + 4 g_{i-1} - 6 g_i + 4 g_{i+1}-g_{i+2} \over 12 \Delta s},  
\label{eq:s-4th-order}
\end{align}
where the terms involving a dissipation coefficient $D$ correspond to (hyper)diffusive upwind dissipation. This dissipation is
not included in the case of term VII (containing derivatives of $\phi$) for numerical stability reasons.

\ifmoredetails Merely to keep the code simple, the finite difference
expression for the 4th derivative in \eq{s-4th-order} has only second
order accuracy, as a higher order would involve more than just 5
gridpoints.

The fact that in Fourier representation the $\partial_s^4$ becomes
$ik_s^4$ illustrates that the dissipation term affects in particular
the small scales.\\
Note that only a small dissipative contribution is desired, the
smaller the higher the grid resolution is. Hence it makes sense to
prepend the dissipation term with a factor $\Delta s^n$, where $n$ is
an arbitrary power. This should make an appropriate choice of the
coefficient $D$ (parameters \name{disp\_par} and \name{disp\_vp}) more
obvious.
\begin{align}
 - D \Delta s^n\vert v_i \vert {- g_{i-2} + 4 g_{i-1} - 6 g_i + 4 g_{i+1}-g_{i+2} \over 12 \Delta s^4}\nonumber
\end{align}
One may consider a grid-scale oscillation, i.e. $g_{i} = 1$ for even
indices $i$ and $-1$ for odd $i$. This is just the kind of oscillation
that the artifical dissipation term is supposed to suppress.  In this
example, the finite difference in the dissipation term yields
$1/\Delta s^4$ while that of the advection yields $1/\Delta s$.  If
one prepends the dissipation with $\Delta s^3$ then for coefficients
$D \sim 1$ the dissipation term is comparable to the advection term
and will indeed damp the grid-scale oscillation effectively. This is
why in \eq{s-4th-order} only $\Delta s$ appears, not $\Delta s^4$.
% For a large scale oscillation may be illustrated by
% $g_i = -2, -1, 0, 1, 2$ on five neighbouring points. Then the
% dissipation term yields $2 - 4 - 0 + 4 - 2=0$ while the advection
% gives $1/\Delta s$.
\fi

Rather than apply this central differencing scheme separately to terms I and IV, the code user can choose an
alternative scheme, which combines the derivative along the field line
with the trapping term. Defining $H(s,v_{\parallel}) = \frac{1}{2} v_{\parallel}^2 + \mu B+\frac{1}{2}{\cal E}_R$, and noting that
$B {\cal G} = {\cal F} {{\partial B}\over{\partial s}}$ in IV, we can combine terms
I and IV to give
\begin{equation}
\label{eq:arakawa}
v_R {\cal F} \bigg( \mu \frac{\partial B}{\partial s}\frac{\partial \hat{f}}{\partial v_{\parallel}}  -
v_{\parallel}\frac{\partial \hat{f}}{\partial s} \bigg) =
v_R {\cal F} ~\{H,\hat{f}\},
\end{equation}
where $\{H,\hat{f}\}$ is a Poisson bracket (note that we have neglected the redundant
dependence of $\mu$ in the definition of $H$ for simplicity). Following the second and
fourth-order schemes of Arakawa \cite{A66}, we allow the Poisson bracket to be
differenced directly. The terms containing $D$ in
\Eqs{eq:s-2nd-order}{eq:s-4th-order} are also added to allow some dissipation. The
advantage of this scheme over the separate differencing scheme is that usually zero (or relatively small) dissipation is
needed to obtain a stable solution. Presently this is implemented only as a fourth-order scheme.

\section{Boundary conditions along a field line}
We here discuss the implementation of the boundary conditions in the parallel direction $s$, that are presented in Sec.~\ref{sec:spectral}.
After one poloidal turn the Fourier mode connects to a mode with a different radial wave vector.
The grid in the $s$ direction is therefore constructed such that at the end of the field line it connects smoothly to 
the start of the field line (i.e. the grid spacing remains constant). Of course, the radial wave vector to which 
the mode has to connect might not be represented on the grid. At the end (or beginning) of the field line 
a boundary condition is then employed. GKW allows for a choice in this boundary condition, setting the 
perturbed distribution either to zero (NO LONGER AVAILABLE) or to have a zero derivative (default, \name{parallel_boundary_conditions='open'}) in the direction upwind with respect
to the convection.\footnote{The option \name{parallel_boundary_conditions='Dirichlet'} may be different to both of these and is experimental - not documented further here.} 
The down wind direction is always treated with a zero derivative boundary condition, allowing the distribution to 
'flow off the grid', without reflections. 
Close to the end point boundaries a staggered change in differencing order accuracy is used to remove 
ghost cells and to allow the flow of the distribution function out of the domain.  
Reflection or the allowed build up at the upper velocity boundaries is seen as unphysical and therefore undesirable.  
Because of the use of up-winding the boundary condition is dependent on the sign of the advective velocity.

When the advective velocity $v(s)$ is positive the distribution function will move toward the left boundary, and the 
following scheme is used
\begin{eqnarray}
\mbox{Boundary cell}: & & \mbox{Second order scheme} - \frac{\partial g}{\partial s} \rightarrow \frac{-\frac{3}{2}g_{n} + 2g_{n+1} -\frac{1}{2}g_{n+2}}{\Delta s},\nonumber\\
\mbox{Adjacent cell}: & & \mbox{Third order} - \frac{\partial g}{\partial s} \rightarrow \frac{-\frac{1}{3}g_{n-1} -\frac{1}{2}g_{n} + g_{n+1} - \frac{1}{6}g_{n+2}}{\Delta s}. \nonumber
\end{eqnarray}
with the right hand boundary using the centrally differenced schemes.  
When the advective velocity is negative the opposite is true and the right hand boundary is differenced in an identical way.  
To obtain the scheme for the opposing boundary simply reverse the signs of the equation above.
When using the second order scheme, special consideration is needed only for the boundary cell.  
Depending on the direction of the advective velocity, a first order back-winded scheme is used at the boundary. 
The effect of these open boundary conditions is to remove the need for ghost cells in the differencing scheme while maintaining 
a high order scheme. 
Compared to other alternatives we have found that the chosen scheme removes rapid fluctuations near the end points, and 
therefore allows for a minimum dissipation. In the Poisson bracket formulation \eq{arakawa}, the above convention is integrated into the scheme at the boundaries to have similar properties. Less (or no) dissipation (Diffusion/super-diffusion) is needed for this scheme, but the dissaption is implemented as separate routine using the normal differencing above, allowing identical dissipation to be applied to both schemes.

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{Boundarydiff.eps}
\caption{Schematic showing the cells of intest in the differencing scheme when considering the boundary cell (Hollow circles) and the adjacent cell (Filled circles).  The figure shows the left hand boundary, for right hand simply reflect in the y direction}
\label{boundarydiffs}
\end{center}
\end{figure}

% Does new refer to zero-deriv and old to zero ?
\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{fluxcomp.eps}
\caption{Figure showing the effect of the new open boundary conditions at the boundary. The new boundary has reduced cell to cell fluctuations giving a smoother solution.}
\label{fluxcomp}
\end{center}
\end{figure}

\noindent

The effect of the open boundary conditions is to remove the need for ghost cells in the differencing scheme while maintaining a high order scheme. It has reduced the amplitude of cell to cell fluctuations refelecting off the boundary (See figure \ref{fluxcomp}), which in turn, has reduced the need for adding artificial dissipation to the system to damp away these modes.  The dissipation, however, can not be wholly turned off in non-linear runs.  Note also that the dealiasing in nonlinear runs also has some dissipative effect in the perpendicular spatial directions.

\section{Parallel velocity grid - Trapping condition} %Added by Will 29/6/2008
(STILL EXPERIMENTAL AND UNDER DEVELOPMENT)
By contorting the velocity grid to conform to the electron trapping condition we are able to remove terms that involve derivatives in the parallel velocity direction.  This greatly speeds up linear run computation times but more significantly removes the need for numerical diffusion in the parallel velocity direction and lastly, provides resolution in the required areas.

This option is turned on by setting \name{vp\_trap} $= 1$ in the \name{CONTROL} namelist of the input deck.  A further parameter is needed in the \name{GRIDSIZE} namelist which is named \name{n\_trapped} which is the number of grid points within the trapped region of the velocity grid.

In the current implementation of the trapping condition, the number of cells in the s direction and the number of points in the parallel velocity direction add certain restrictions to the number of points that can be placed within the trapped region.

Firstly, for ease of differencing the bounce points of a specific $v_{\parallel 0}$ (The velocity on the low-field side) must occur exactly half way between two grid-cells, thus the number of grid cells in the s direction restrict both the number and initial values of $v_{\parallel}$.  The maximum number of bounce points available is given by \name{n\_s\_grid}$/2 - 1$.  The value of \name{n\_trapped} must be less than or equal to this.  If less is chosen then the code will select which points to use determined by the spacing between points (The closer to the boundary, the closer the bounce points become), the algorithm will iterate over the points, sequentially removing points until \name{n\_trapped} are left.  \name{n\_trapped} must also be less than \name{n\_vpar\_grid}.  Outside the trapping region the points are uniformly placed, the spacing is determined by the number \name{n\_vpar\_grid} $- 2\times$ \name{n\_trapped}.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.4\textwidth]{traplowfield.eps}
\caption{Figure of parallel velocity grid points as given by the trapping condition, the line corresponds to the trapped/untrapped boundary $v_{||0} = \sqrt{2\mu(B_{max}-B_{min})+\cfen^{H}-\cfen^{L}}$.  In the above data, N_vpar_grid $= 16$ (only one half of the grid is shown as it is symmetric),n_trapped $= 4$ and $\mu = 6$. It must be noted that the total number of points within the trapped region (full grid) is $2\times$n_trapped.}
\label{lowfieldsidetrap}
\end{center}
\end{figure}

This algorithm gives the grid at the low field side (For an example see figure \ref{lowfieldsidetrap}), this is then propagated around the torus by the equation
\begin{equation}
v_{\parallel N} = \sqrt{v_{\parallel 0 N}^{2} + 2\mu_{N}(B - B_{0}) + \cfen- \cfen^{\rm L}}
\end{equation}
\noindent
where $B_{0}$ is the magnetic field at the low field side, and $\cfen^L$ is the centrifugal energy at the low field side.  When a bounce point is reached, the velocity becomes zero, and all velocity grid points after this along $s$ are set to zero and are ignored in the differencing.  The grid is also symmetric around $v_{\parallel}=0$.

Rules that must be followed, otherwise the code will throw you out,
\begin{itemize}
\item The value of \name{n\_trapped} must be less than \name{n\_vpar\_grid}/2
\item \name{n\_trapped} must also be less than or equal to \name{n\_s\_grid}/2$-1$
\item For optimal convergence $2\times$\name{n\_trapped} must be roughly half of \name{n\_vpar\_grid}. (This may no longer be true: See \issue{8} for latest updates on Gaussian integration (and section \ref{sec:gaussian-integration})
\end{itemize}

\section{Gaussian Integration}
\label{sec:gaussian-integration}

This is a short note on how the Gaussian integration is implemented in GKW.

We wish to calculate the integral over the normalized parallel velocity $v_\parallel$ 
\begin{equation}
I = \int\nolimits_{v_{tp}}^{v_{pm}} g(v_\parallel) {\rm d} v_\parallel 
\end{equation} 
The integral is over (half of the) passing domain with the lower limit the trapped-passing 
boundary ($v_{tp}$) which is a function of $\mu$. We use Gaussian integration 
in combination with ${\rm vp\_trap = 1}$, in which case the parallel velocity grid must 
satisfy 
\begin{equation} 
v_{\parallel k}^2 = v_{\parallel 0}^2 + 2 \mu (B_0 - B_k) 
\end{equation} 
where $v_{\parallel k}$ is the parallel velocity at grid point k along the magnetic field, 
with $B_k$ being the local magnetic field, and the index 0 indicates the low field 
side position. We can therefore define only one set of parallel velocity gird points 
with the grid at different locations along the field line satisfying the equation 
above. The integral is therefore written in low field side coordinates 
\begin{equation} 
I = \int\nolimits_{v_{pt0}}^{v_{pm0}} g(v_\parallel ( v_{\parallel 0}) ) {{\rm d} v_\parallel \over 
{\rm d} v_{\parallel 0}} \biggr \vert_\mu \, {\rm d} v_{\parallel 0} 
\end{equation} 
From the relation between $v_\parallel$ and $v_{\parallel 0}$ it follows 
\begin{equation} 
{{\rm d} v_\parallel \over {\rm d} v_{\parallel 0}} \biggr \vert_\mu  = {v_{\parallel 0} \over 
v_\parallel } 
\end{equation}
We furthermore choose to write $g$ as 
\begin{equation}
{v_{\parallel 0} \over v_\parallel} g = \exp [-v_{\parallel 0}^2 ] z(v_{\parallel 0} )
\label{zdef} 
\end{equation} 
This choice is of course motivated by the expected dependence of $g$ on $v_\parallel$. The 
Maxwellian background will lead to the strong exponential decay. It is this exponential 
dependence that we would like to 'transform away' in order to obtain an efficient integration
algorithm.

With the choices discussed above the integral is 
\begin{equation}
I = \int\nolimits_{v_{tp0}}^{v_{pm0}} \exp [ - v_{\parallel 0}^2 ] z(v_{\parallel 0}) {\rm d} v_{\parallel 0} 
\end{equation} 
Using the transformation 
\begin{equation} 
y = {\rm erf} (v_{\parallel 0}) \qquad {\rm with}\qquad {\rm erf}(v_{\parallel 0} ) = {2 \over \sqrt{\pi}} 
\int_0^{v_{\parallel 0}} \exp [-x^2] {\rm d} x
\end{equation} 
one obtains 
\begin{equation} 
I = {\sqrt\pi \over 2 } \int\nolimits_{ya}^{yb} z(y(v_{\parallel 0})) {\rm d} y 
\end{equation} 
with 
\begin{equation} 
ya = {\rm erf}(v_{tp0}) \qquad yb = {\rm erf}(v_{pm0} ) 
\end{equation} 
Finally, we transform the integral to the interval [-1,1], using the transformation 
\begin{equation} 
\alpha = {2 y \over yb - ya} - {ya + yb \over yb - ya} 
\end{equation} 
to obtain 
\begin{equation} 
I = {1\over 4} \sqrt\pi (yb - ya) \int_{-1}^1 z(\alpha(y)) {\rm d} \alpha 
\end{equation} 
The latter integral can then be solved by a standard Gauss Legendre integration using the points 
$\alpha_i$ and weights $w_i$ (which can be obtained from publicly available software) 
\begin{equation} 
\int_{-1}^1 z(\alpha(y)) {\rm d} \alpha = \sum\nolimits_{i = 1}^m  w_i z(\alpha_i) 
\end{equation}

For the implementation we transform back to $v_{\parallel}$ 
\begin{equation} 
I = {1 \over 4} \sqrt\pi (yb-ya) \sum\nolimits_{i = 1}^m w_i \exp[ + v_{\parallel 0i}^2] {v_{\parallel 0 i} \over v_\parallel (v_{\parallel 0i} ) } 
g(v_\parallel(v_{\parallel 0i})) 
\end{equation} 
with the points of the grid $v_{\parallel 0 i}$ determined by the relation 
\begin{equation} 
{\rm erf}(v_{\parallel 0 i} ) = ya + (yb - ya)(\alpha_i + 1 )/2 
\end{equation}

\section{The collision operator}
\label{numerics:collisions}
The collision operator can be thought of as a convection/diffusion equation in velocity space and thus is differenced accordingly.
The velocity space boundary conditions are taken to be zero flux, which occurs naturally at the $\mu =0$ boundary and is artificially
imposed on the other three boundaries.  These boundary conditions guarantee the conservation of mass.  
Considering that the diffusion coefficient is velocity space dependent, and the grid can be non-uniform in the $\mu$-direction, the differencing scheme of the equations is chosen as follows for the parallel velocity direction
\begin{equation}
\frac{\partial}{\partial v_\parallel}\left(D\frac{\partial \hat{f}}{\partial v_{\parallel}}\right) \rightarrow
\frac{1}{v_{i+1/2\parallel}^{j}-v_{i-1/2\parallel}^{j}}\Big( D_{i+1/2}^{j}\left(\frac{f_{i+1}^{j}-f_{i}^{j}}{v_{i+1\parallel}^{j}-v_{i\parallel}^{j}}\right) - 
D_{i-1/2}^{j}\left(\frac{f_{i}^{j}-f_{i-1}^{j}}{v_{i\parallel}^{j}-v_{i-1\parallel}^{j}}\right)\Big),
\end{equation}
\noindent
where i denotes the point in the parallel velocity direction and j the $\mu$ direction.  When considering the cross-terms a four point 
interpolation method is used to calculate the half point values.  Four points are needed on a two-dimensional grid.
The friction term is differenced in the same way, but with a two point interpolation.

To bypass the complications caused by non-uniform distribution of $\mu$ points (which in uniform in the $v_\perp$ coordinate), the code chooses the suitable version of the
collision operator that is uniform, and as such, can conserve mass (and parallel momentum) to machine accuracy.

A further switch \textbf{selfcollcon} in the collisions input deck changes the momentum conservation from just the self-collisions for each species (selfcollcon = .true.) to absolute conservation (selfcollcon=.false, default=.true.). Note that the absolute conservation option is unphysical and only implemented for tests and benchmarks purposes.

\section{The Fourier representation}

In a nonlinear run a rectangular grid of wave vectors ($k_\zeta,k_\psi$) is used. 
The nonlinearity is treated using fast Fourier transforms (FFT).
Different choices of FFT libraries are possible, but at present we use mostly the FFTW library \cite{FFT05}. 
For the nonlinear term a de-aliasing method is applied, i.e. the transformation to real space is constructed on a grid finer 
(number of grid points at least 3/2 larger in real space) than the Fourier modes represent. 
The de-aliasing method is non-conservative and has a dissipative effect on the finer grid scales, with the consequence that a 
numerical dissipation term in the spectral directions is generally not required.

Since the distribution function is a real quantity the number of
complex Fourier modes can be reduced by a factor two and a half plane
of wavevectors with $k_\zeta \ge 0$ is used.  Grid sizes in real space
are chosen so that they have only small prime factors. Such sizes lead
to particularly efficient computations with the standard FFT
libraries. There are no significantly more efficient grid sizes any
more, as it used to be in the past.

\section{Time integration}

GKW has several options for the explicit time integration: modified midpoint, fourth order Runga Kutta (recommended), 
and a second order scheme that is stable for hyperbolic equations. An implicit scheme is also implemented, but is still under development
and not yet recommended for general use (For latest updates see \issue{10}).

\section{Poisson equation \label{Poisson-detail}}

The Poisson equation \ref{Poisson-adiabatic} is split into three linear terms which are stored in 
the main solution matrix,  but evaluated separately from the other linear terms. 
The calculation of the fields takes place in \File{fields.F90} subroutine \name{calculate_fields}.

The Poisson equation given in \ref{Poisson-adiabatic} can be rewritten.
\ifmoredetails
\begin{align}
  &\sum_{sp , ions }  Z_{sp} n_{R_0,sp} \biggl [ 
  2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp} + {Z_{sp} \over T_{Rsp}} [ \Gamma(b_{sp}) -1]\exp(-\cfennsp{sp}) \hat \phi 
  \biggr ] =  { n_{R_0,e}\exp(-\cfennsp{e})
    \over T_{Re}} ( \hat \phi - {\fsa{\hat \phi }}) \nonumber\\
  &
  \sum_{sp , ions }  Z_{sp} n_{R_0,sp} \biggl [ 
  2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp}
  \biggr ]
  + { n_{R_0,e}\exp(-\cfennsp{e}) \over T_{Re}} {\fsa{\hat \phi }} 
  = {} \nonumber\\
  &\qquad {} = 
  -\sum_{sp , ions }  Z_{sp} n_{R_0,sp} \biggl [ 
  {Z_{sp} \over T_{Rsp}} [ \Gamma(b_{sp}) -1]\exp(-\cfennsp{sp}) \hat \phi 
  \biggr ]
  + { n_{R_0,e}\exp(-\cfennsp{e}) \over T_{Re}} \hat \phi
 \nonumber\\
\end{align}
Solving for $\hat\phi$
\begin{align}
  &
  \sum_{sp , ions }  Z_{sp} n_{R_0,sp} \biggl [ 
  2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp}
  \biggr ]
  + { n_{R_0,e}\exp(-\cfennsp{e}) \over T_{Re}} {\fsa{ \hat\phi}} 
  = {} \nonumber\\
  &\qquad {} = 
  \left[-\sum_{sp , ions }  Z_{sp} n_{R_0,sp} \biggl [ 
  {Z_{sp} \over T_{Rsp}} [ \Gamma(b_{sp}) -1]\exp(-\cfennsp{sp})
  \biggr ]
  + { n_{R_0,e}\exp(-\cfennsp{e}) \over T_{Re}} \right] \hat \phi
 \nonumber\\
\end{align}
yields
\fi
\begin{align}
\hat{\phi }= -{1 \over A} \left[ \sum_{sp , ions }  \underbrace{Z_{sp} n_{R_0,sp} 2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp})}_{\name{poisson_int()}\rightarrow M^3} \hat g_{sp} + \underbrace{{n_{Re}\exp(-\cfennsp{e}) \over T_{Re}}
 \fsa{\hat\phi}}_{\name{poisson_zf()}\rightarrow \text{eq. \ref{eq:zonalmatrixform}}}
\right]
\label{eq:poisson2}
\end{align}
where the factor
\begin{equation}
 A =  \sum_{\rm sp, ions} n_{sp} Z_{sp} \left[ {Z_{sp} \over T_{Rsp}} ( \Gamma(b_{sp}) -1)\exp(-\cfennsp{sp}) - {\exp(-\cfennsp{e}) \over T_{Re} } \right] = M^4
\end{equation}
is evaluated in subroutine \name{poisson_dia}. The symbols $M^3$ and
$M^4$ number the matrices presented in this paragraph (and are not
``powers of $M$'' or so) and refer to sections of \name{mat}. In this
sense also $M^1$ and $M^2$ which contain the linear terms of the GKE
and are mentioned elsewhere denote sections 1 and 2 of the matrix
\name{mat}. The quantity $\fsa{\hat\phi}$ can be eliminated from
\ref{eq:poisson2} by taking the flux surface average of equation
\ref{eq:poisson2}.  
\ifmoredetails
\begin{align}
\fsa{\hat{\phi}} &= \fsa{-{1 \over A} \left[ 
    \sum_{sp , ions }  Z_{sp} n_{R_0,sp} 2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp} 
    + {n_{Re}\exp(-\cfennsp{e}) \over T_{e}} \fsa{\hat \phi}
  \right]} \nonumber\\
 &= \fsa{-{1 \over A} 
    \sum_{sp , ions }  Z_{sp} n_{R_0,sp} 2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp} 
    }
    + \fsa{-{1 \over A}{n_{Re}\exp(-\cfennsp{e}) \over T_{Re}} \fsa{\hat \phi}  
}
\end{align}
The expressions containing $\fsa{\hat\phi}$ are drawn together:
\begin{align}
\fsa{\hat\phi} - \frac{n_{Re}}{T_{Re}}\fsa{\frac{-\exp(-\cfennsp{e})}{A}} \fsa{\hat \phi}
  &= \fsa{-{1 \over A} 
    \sum_{sp , ions }  Z_{sp} n_{R_0,sp} 2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp} 
}\nonumber
\end{align}
\begin{align}
\fsa{\hat\phi}\left(
  1
  -
  \frac{n_{Re}}{T_{Re}}\fsa{\frac{-\exp(-\cfennsp{e})}{A}}\right)
  &=\nonumber\\
\frac{n_{Re}}{T_{Re}}\exp(-\cfennsp{e})\fsa{\hat\phi}\left(
  \frac{T_{Re}}{n_{Re}}\frac{1}{\exp(-\cfennsp{e})}
  -
  \fsa{\frac{-1}{A}}\right)
  &=
\end{align}
\fi
The zonal correction term can then be written as a function of the
distribution function, parameters and background quantities.
\begin{align}
{n_{Re}\exp(-\cfennsp{e}) \over T_{Re} } \fsa{\hat\phi}
&= { \fsa{
-{1 \over A} \sum_{sp, ions}  Z_{sp} n_{R_0,sp} 2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp})\hat g_{sp}
}
 \over
 {1 \over \exp(-\cfennsp{e})} \left({T_Re \over n_Re} + \fsa{{\exp(-\cfennsp{e}) \over A}}\right)} 
\label{eq:zonal}
\end{align}
The subroutine \name{poisson_zf} prepares matrices according to the definitions
\begin{align}
  Z &= -\frac{\ud s}{A} \\
  Y &= \frac{1}{\exp(-\cfennsp{e})} \left({T_Re \over n_Re} + \fsa{{\exp(-\cfennsp{e}) \over A}}\right)
\end{align}
which are used to express the right hand side of \ref{eq:zonal} in the form
\begin{align}
{n_{Re}\exp(-\cfennsp{e}) \over T_{Re} } \fsa{\hat\phi}
  &= {\sum_s Z M^3 {\hat g} \over Y}\ .
\label{eq:zonalmatrixform}
\end{align}
\ifmoredetails
Thanks to \ref{eq:zonal}, \ref{eq:poisson2} allows the subroutine
\name{calculate_fields} to calculate $\hat\phi$ explicitely.
\begin{align}
\hat{\phi } & = \sum_s Z M^3 {\hat g} + {\sum_s Z M^3 {\hat g} \over Y} \\
            & = {(Y + 1)\sum_s Z M^3 {\hat g} \over Y}\ .
\end{align}
\fi

To calculate $\hat\phi$ from the distribution function, first the integral
part \name{poisson_int} is evaluated using section $M^3$ of the
matrix.
\begin{equation}
a_i(k_y,k_x,s) = M^{3}_{ij} {\hat g}_j.
\end{equation}
Next, the zonal flow correction term, equation \ref{eq:zonal}, is evaluated
separately, using the special ``matrices'' $Z$ and $Y$.  First, the flux surface
average in the numerator is evaluated.
\begin{equation}
b_i(k_x) = \sum_s Z_{ij} a_j 
\end{equation}
Then each element is divided by the denominator from equation \ref{eq:zonal}.
\begin{equation}
c_i(0,k_x,s)) = b_i / Y_i(s) 
\end{equation}
The zonal correction term is then subtracted from the minuend in
equation \ref{eq:poisson2}.
\begin{equation}
d_i(k_y,,k_x,s)=a_i-c_i 
\end{equation}
Finally, all elements are divided by the factor $A$ ($A$ being identical
to $M^4$, the section 4 of ``the matrix of the linear terms'').
\begin{equation}
{\hat \phi}_i(k_y,k_x,s) = d_i / M^4_i
\end{equation}
The other linear terms are stored in sections 1 and 2 of the matrix and evaluated as 
$\hat{f_i}^\prime = \hat{f_i} + \delta t \cdot \hat{f_j} M^{1,2}_{ij}$ in \name{calculate_rhs}
after the evaluation of the fields.

\section{Notes on individual routines}

This section discusses the individual routines and gives some details on the physics 
processes implemented. Here, not all the input and output or all the variables are 
discussed. It merely gives some details that should allow the reader to understand
the code better.

\subsection{\src{linear_terms.f90}} %Added by Francis Jan 2008, not recently updated.
NOT INCLUDED here: centrifugal effects\\

In the treatment so far, the terms as implemented in the code were described in \ref{eqs:complete-set} as terms I-VIII. Here we link these terms to the routines in the code, as well as to \ref{Vlasov}, the Vlasov equation for the modified perturbation distribution function $g$.
\begin{equation} 
\label{1}
{\partial g \over \partial t} + (\underbrace{v_\parallel {\bf b}}_{I} + \underbrace{{\bf v}_D}_{II}) \cdot \nabla f + \underbrace{{\bf v}_\chi}_{III} \cdot \nabla g  
-\underbrace{{\mu B \over m}{{\bf B}\cdot \nabla B \over B^2}{\partial f \over \partial v_\parallel}}_{IV} = S. 
\end{equation}

Where the source from the Maxwellian background is given by eqn \ref{source}
\begin{equation} 
\label{2}
S =  - \underbrace{({\bf v}_\chi}_{V} + \underbrace{{\bf v}_D)}_{VI} \cdot \nabla_p F_M
-  {Ze \over T} \underbrace{[ v_\parallel {\bf b}}_{VII} + \underbrace{{\bf v}_D ]}_{VIII} \cdot \nabla \chi F_M  
\end{equation}

We also need the expansion of the drift velocity \ref{eq:drifts}
\begin{align}
\label{3}
{\bf v}_D &=& \underbrace{{1\over Ze} \biggl [ \underbrace{{m v_\parallel^2\over B}}_{1} + \underbrace{\mu}_{2} \biggr ] {{\bf B} \times \nabla B
\over B^2}}_{A} + \underbrace{{m v_\parallel^2 \over 2 Z e B} \beta^\prime {\bf b} \times \nabla \psi}_{3} 
 + \underbrace{{ 2 m v_\parallel \over Z e B } {\bf \Omega}_\perp}_{4} \cr 
\noalign{\vskip 0.2 truecm} 
&-& \underbrace{{m \Omega^2 R \over Z e B }  {\bf b} \times {\nabla R}}_{5}, 
\end{align}

In \src{linear_terms.f90} the various terms from these equations are input into the matrix in individual routines which will be described below.  Most of the computation in the code is done in the wave vector domain. All derivatives in the perpendicular direction are in Fourier space; but derivatives along the field line are done in real space (terms I and VII)

\par
Some routines in \src{linear_terms.f90} have two versions (identified by a postfix to the routine name), one for each of the second order and fourth order methods.  We concentrate here on the fourth order method since the other is out of order. The routines do the same thing with regards to which terms are implemented; the implementation is different for each method.  The linear terms all operate on the the distribution $g$ (often called $f$ in the code comments!), because the linear terms matrix is applied to \textbf{the array called \name{fdisi} which always contains $g$}.

\par
\name{calc_linear_terms} is the master routine that calls the subroutines which put the terms I-XI (excluding term III) into the matrix.  Broadly, the first section puts the terms on the LHS of \ref{1}.  The second section calculates the source terms due to the maxwell background.\ref{2}.  The third section of \name{calc_linear_terms} calls the routines that input the field equations into the matrix. These are the equations that force the solution the be self consistent with the potentials calculated from the distribution function.  After each section is completed there is a call to \name {finish_matrix_section} in module \name {matdat} which stores the k index labels of the matrix for each section..

\par
Hence \name{calc_linear_terms} covers all the terms I-VII in \ref{eqs:complete-set} ( also labelled as the underbraced numbers in \ref{1}, \ref{2}, and \ref{3}), with the exception of term III, which is implemented in \src{non_linear_terms.f90}.

\subsubsection{\name{vpar_grad_df}, term I}
\name{vpar_grad_df} calculates the convection parallel to the field.  This routine also adds parallel velocity dissipation, which can be required for a numerically stable solution in nonlinear runs.
The dissipation is controlled by the input parameter \name{disp_par}.  This term is the free streaming motion along the field line, with derivative calculated in real space.

\subsubsection{\name{vdgradf}, term II}
\name{vdgradf} calculates the drift in the gradient of the eikonal term.

\subsubsection{\name{trapdf}, terms IV}
\name{trapdf} calcuates the mirror trapping terms dependent on the switch variable \name{trapping}.   This routine also adds perpendicular velocity dissipation, which can be required for a numerically stable solution in nonlinear runs. 
The dissipation is controlled by the input parameter \name{disp_vp}.

\subsubsection{\name{ve_grad_fm}, term V}
\name{ve_grad_fm} calculates the ${\bf E} \times {\bf B}$ drift due to the background distribution, as well as (optionally according to logical \name{nlapar}) the electromagnetic correction ${\bf v}_{\delta B_\perp}.$

\subsubsection{\name{neoclassical}, term VI}
\name{neoclassical} optionally includes the neoclassical terms according to the switch \name{neoclassics}.  Note many gyrokinetics codes disregard this term.

\subsubsection{\name{vpgrphi}, term VII}
\name{vpgrphi} calculates the landau damping term, dependent on the switch variable \name{landau}.  This is an acceleration along the field line with parallel derivative in real space.

\subsubsection{\name{vd_grad_phi_fm}, term VIII}
\name{vd_grad_phi_fm} calculates the  drift in the gradient of phi times the velocity derivative of the Maxwell background.  Note that Equation terms  \ref{3}.1 and \ref{3}.2 are combined as \ref{3}.A in the $E_D$.multiplier as defined in \ref{eqs:complete-set}.

\subsection{\src{non_linear_terms.F90}} %Added by Francis Feb 2008, needs checking again.

CHECK CONSISTENCY OF THIS SECTION

The Fourier amplitudes of the potential (and perturbed distribution function) are defined as 
\begin{equation} 
\phi_N(x) = \sum_k \phi_{Nk} \exp [ {\rm i} k x ] + c.c.  
\end{equation} 
Note that because the inverse discrete Fourier transform ($H_l$) is defined such that 
the values in real space ($h_n$) are given by  
\begin{equation} 
h_n = {1\over N} \sum_{l=0}^{N-1} H_l \exp [ {\rm i} 2\pi l n / N ] 
\end{equation} 
where $N$ is the total number of grid points of the discrete transform. The relation between the 
amplitudes and the discrete Fourier transform, therefore, is 
\begin{equation} 
H_l = N\phi_{Nk}
\end{equation} 
The wave vector $k$ is given by the standard relation  
\begin{equation} 
k = {2 \pi l \over N \Delta} 
\end{equation} 
where $\Delta$ is the grid distance. This relation is used to determine the size of the box 
in real space. It follows that the lowest non-zero wave vector $k_1$ determines $\Delta$ and
the size of the box $L = N\Delta$ 
\begin{equation} 
L = {2\pi \over k_1} 
\end{equation}

\subsubsection{\name{add_non_linear_terms}, term III} 
\index{Non linear terms}
This routine implements the full term III with electromagnetic corrections.  Hence the routine finds the ${\bf E} \times {\bf B}$ drift of the perturbed distribution function, and adds the term:
\begin{equation} 
\label{Edrift}
{\bf v}_\chi \cdot \nabla g = {{\bf b} \times \nabla \hat{\langle \chi \rangle} \over B} \cdot \nabla g
\end{equation}
Without the elctromagnetic corrections, Term III of \ref{eqs:complete-set} becomes
\begin{equation}
{\rm III} = - {\bf v}_\phi \cdot \nabla g  \mathcal{T}\Big({\cal E}^{\psi \zeta}\left[\mathcal{T}^{-1}(ik_{\zeta} \hat{ \phi }) \mathcal{T}^{-1}(ik_{\psi} \hat g) -
\mathcal{T}^{-1}(ik_{\zeta} \hat g) \mathcal{T}^{-1}(ik_{\psi} \hat{ \phi })\right]\Big),
\end{equation}

where the expansion of the tensor sum uses the anti-symmetry of ${\cal E}^{\beta \alpha}$.  Note also that the diagonal elements of ${\cal E}^{\beta \alpha}$ are zero, and derivatives along $s$ are neglected.

The gyroaveraged potential $\langle \phi_N \rangle$ is first calculated in the wavevector domain by multiplying the potential by the Bessel function $J_0$ (\ref {Bessel}). In k space the derivative of the potential is obtained by multiplying by $ik_\alpha$.

Hence the implementation in the code is
\begin{equation}
a = i J_{0}(k_{\perp}\rho) k_{N,\zeta} \phi_{N,k} = ik_{\zeta}\rho_{\rm ref}{\langle \phi_{N,k} \rangle} 
\end{equation}
\begin{equation}
b = i J_{0}(k_{\perp}\rho) k_{N,\psi} \phi_{N,k} = ik_{\psi}\rho_{\rm ref}{\langle \phi_{N,k} \rangle} 
\end{equation}
Note the $\rho$ in the Bessel function is species dependent.  Also the normalization of the k vector $k = {k_N / \rho_{\rm ref}}$ introduces a multiplier of $\rho_{\rm ref}$
Hence the inverse Fourier transformation gives
\begin{equation}
ar = \mathcal{F}^{-1}(a) =  \rho_{\rm ref} {\partial {\langle \phi_N \rangle} \over \partial \zeta_{N}} = {\rho_{\rm ref} \over R_{\rm ref}} {\partial {\langle \phi_N \rangle} \over \partial \zeta} =  \rho_* {\partial {\langle \phi_N \rangle} \over \partial \zeta}
\end{equation}
\begin{equation}
br = \mathcal{F}^{-1}(b) =  \rho_* {\partial {\langle \phi_N \rangle} \over \partial \psi}
\end{equation}
%Incorrect, not all corrdinates are normalised the same.
Since the coordinates are normalised by $x_{\alpha,N} = {x_{\alpha} \slash R_{\rm ref}}$.  Here we have explicitly linked the normalizations of the k vector and coordinates to show that under the Fourier transform; $\rho_* {\partial \over \partial x_\alpha} \rightarrow {\rm i} k_\alpha$, as was intended in its definition. Put simply, a factor of $\rho_{*}$ appears after the inverse Fourier transform.

Similarly, the gradient of the distribution function in k space is found $a = i k_\psi g_{N,k}$, $b = i k_\zeta g_{N,k}$
and applying the inverse Fourier transform gives
\begin{equation}
cr = \mathcal{F}^{-1}(a) = \rho_* {\partial g_N \over \partial \psi}
\end{equation}
\begin{equation}
dr = \mathcal{F}^{-1}(b) = \rho_* {\partial g_N \over \partial \zeta}
\end{equation}

The tensor ${\cal E}^{\alpha \beta}$ (see \ref{efun}) is defined in \src{geom.f90} with name  {\name efun}. Note that although the tensor has two indices, each component varies along the field line due to the gradient in $B$. The array therfore has 3 indices with $efun{(i,2,1)}$ being the ${\cal E}(s)^{\zeta \psi}$ component. Calling this element of the array we now have all the components needed to calculate \ref{eq:nl-term}.  The factor of $\rho_*^2$ comes from the 2 inverse Fourier transforms.  Term III then appears in the code as
\begin{equation}
efun(i,2,1)*({ar*cr-br*dr})
\end{equation}

Finally this is Fourier transformed back to the wavevector domain and $\mathcal{F}({\bf v}_E \cdot \nabla g)$ is added to the RHS of the equations -  which were passed to the subroutine as an argument along with the distribution function when it was called from \src{exp_integration.F90}.

\par
The routine also calculates maximum velocities to be used in the timestep estimator.  The first time this routine is called it does some initialisations which are not repeated in later calls:  The sizes of the boxes in real space are calculated, the max k-vectors are calculated and the location of the $k_x = 0$ mode is determined.  Index arrays are also set up to relate the storage regime on the fast Fourier transform (fft) routine to the GKW storage regime. The remainder of the routine as described above is executed every time.

\section{The code}\label{thecode}

The package GKW consists of a README file (where a description of the full contents can be found),
Fortran 95 source files, various makefiles and related scripts, some documentation, some sample code input files,
some scripts associated with running and testing the code and tools to pre/post process data and perform visualisation.
This section describes what the code can do in terms of parallelisation and how it is structured
with respect to the solution of the equations presented in the previous sections. In order to do
this, we follow the various tasks the code performs as it runs and discuss the relevant modules to
that task. Instructions for building, installing and running the code can be found in section \ref{usage}.

Each source code file contains a module, except the main program file \src{gkw.f90}.
Each source file has either the extension \File{.f90} or \File{.F90}, the latter denoting the need to be pre-processed.
For the modules, each module name corresponds directly to the file name (which is always lower case), minus the suffix.
Therefore, in the following subsections, when referring to the module \name{GRID}, the corresponding source code can be found
in the file \src{grid.F90}.

The code does not come with any routines to perform FFTs, which are required for the computation of the non-linear term in Eq.~\ref{eq:nl-term}, and to transform some of the diagnostics into real space. No FFT's are required for a linear run and the code can be compiled without FFT functionality. The FFT's are presently performed via the external FFTW3 library \cite{FFT05}, for which the
\name{FFT} module provides an interface. Although more interfaces for various other portable Fortran FFT libraries could be
provided, the performance of the FFTW3 library has been found to exceed that of other alternatives significantly. In addition,
the FFTW3 library is fairly portable, so to date there has been no good reason to consider alternatives.

GKW is parallelised primarily using the Message Passing Interface (MPI)
\footnote{Shared memory parallelisation, implemented via OpenMP directives, is
now possible with the code and is documented in a later subsection}. Each
parallel process is responsible for solving the equations over a sub domain of
the space and a subset of the species. A process usually only has knowledge of
the part of the solution it is responsible for, unless explicitly communicated
between processes using the MPI library. The parallelisation is discussed in
more detail in a later section.

\subsection{Initialisation}

At the start of a run, the code will first call various basic MPI routines.
These give each process an individual rank (\name{processor_number}) and inform
each process of the total number of processors available for the computation
(\name{number_of_processors}). Checks are then performed to see if various
output or restart files are present.

After this, the main input file is read. A single processor
is responsible for doing this. The input file contains several Fortran namelists, which
are described later. The first namelist is called \name{control} and is read in the module
\name{CONTROL}, and is responsible for controlling the basic switches for the physics, numerics and
some of the code output. Each module requiring input is responsible for reading its input
from this file. After each namelist is read, the processor that read the file broadcasts
the information to all the other processors via MPI. All processors can then individually
check that the input is allowed or satisfactory, and if necessary, abort the run.

\subsection{Parallelisation: local grid}

The sizes of the local and global grids and associated quantities are dealt with in
the module \name{GRID}. The namelist \name{gridsize} contains the input
variables for the grid sizes. The code can decompose the calculation over the
$s$, $v_\parallel$ and $\mu$ coordinates and the species. Given the number of
processors as input, the routines of \name{GRID} decide how the global solution
will be subdivided between the processors. Alternatively, a user can specify
explicitly the number of processors over which to subdivide the problem in each
divisible direction. 
%There are some restrictions for the case of collisions, and the Arakawa style differencing scheme. 
If a user explicitly sets 
an invalid combination, the code will abort at this point or earlier.
For the decomposition over the species or the $\mu$ grid, 1 point per processor is the minimum.
The minimum number of grid points per processor in the $s$ and $v_\parallel$ directions is limited to either 1 (when using the
second-order finite differences) or 2 (when using the fourth-order scheme). 
These minimum grid sizes dictate the maximum number of processors that can be used
in a given direction; the maximum is either the number of grid points in that direction (for the second-order scheme)
or half that value (for the fourth-order scheme). 
The parallelisation is discussed in more detail in Sect.~\ref{parallelisation}.

A processor is provided with various logical variables denoting if it is at a particular
boundary in the global domain together with the ranks of the processors that it
will need to communicate with (which are usually the ones responsible for adjacent
parts of the computational domain).

Once the local grid sizes have been confirmed, some MPI communicators are set up
to facilitate communication between processes responsible various subsets of
the whole domain. These are found in the \name{MPICOMMS} module.
They are particularly useful when performing a sum over a subset of the coordinates
when all the relevant points are not present on the local processor.

\subsection{Memory layout and allocation}
\label{memorylayout}
In GKW, the solution pertubation distribution
$g(k_\zeta,k_\psi,s,\mu,v_\parallel)$ for each species, together with the
various fields, are stored in the one-dimensional array \name{fdisi},
found in the module \name{DIST}. These quantities are defined at discrete grid
points, so can most conveniently be thought of as multidimensional arrays of
dimension 6 for the distribution function (three dimensions of space, two of
velocity, and one for the species) or 3 for the fields (three dimensions of
space). In the code we want to address them as multidimension arrays, via the
various grid indices, in order to perform initialisation and output
calculations. However, these arrays are embedded into the array \name{fdisi} in
order to facilitate the implementation of different types of numerical schemes
and optimisations.

The modules \name{DIST} and \name{INDEX_FUNCTION} are responsible for the tasks
of mapping the multidimensional arrays into \name{fdisi} and providing a means
to reference them. The \name{DIST} module is primarily responsible for deciding
where within \name{fdisi} those arrays should be stored (as a contiguous
block), and also making sure that each array is in a distinct part of
\name{fdisi}. The \name{INDEX_FUNCTION} module determines how the indices of the
multidimensional array are mapped to a single unique index for the one
dimensional array. This mapping can be
adjusted in order to improve parallel efficiency and (possibly) improve cache
efficiency. The default ordering is at present quite optimal, but manual
adjustments within the \name{INDEX_FUNCTION} module may yield further
improvement in some cases. These two modules are quite complicated and one
should look into those modules in the code.
% or at Sect.~\ref{appendix:fdisi} for a detailed explanation.
%FIXME Describe them somewhere?

The function which provides the array location within \name{fdisi} is
\name{indx()}.  This is used in many modules throughout the code, but
specifically {\it not} the time integration modules, which are designed not
to require it. As an example of how it is used, the value \name{phi} of the
potential at a radial grid point \name{ix}, a binormal grid point \name{imod}
and an $s$ grid point \name{i} is obtained in the code as
\begin{Verbatim}
  phi = fdisi(indx(iphi,imod,ix,i))
\end{Verbatim}
where \name{iphi} is an identifier for the potential.

In addition to the local solution $g$ and the various fields, the \name{fdisi_tmp}
array contains parts of the solution from other processors which have been
communicated via MPI datatypes. In the case of parallelising over the $s$-grid, the
fields are also decomposed, so parts of the potential from other processors
must also be present in order to calculate the derivative of
$\langle \phi \rangle$. The \name{indx()} function deals with referencing the
right part of \name{fdisi} so that other parts of the code do not need to make
special arrangements for processor boundary points.

The distribution function $f$ only ever exists temporarily in the array
\name{fdis_tmp}, and in diagnostics and elsewhere \textbf{a conversion from
$g$ to $f$ must be made (usually using the routine \name{get_f_from_g}, except
in optimised routines) when it is necessary to work with $f$}.

In general, each module is responsible for allocating any arrays they require.
The code attempts to do as much memory allocation as necessary early on, so
that if it will exceed the memory limits of a given machine, not much time will
have been wasted before the code aborts. When diagnostics are performed at the
end of a run, memory allocated for time integration is freed.

%Something about cache efficiency in hybrid scheme?

\subsection{Linear terms}

The linear terms are calculated in the \name{LINEAR_TERMS} and \name{COLLISIONOP} module and put into a matrix which
is in \name{MATDAT}. The \name{LINEAR_TERMS} module is designed so that more new terms can be added by only
adding one new routine and calling it from within that module.

The linear terms all operate on the the distribution $g$ (often called $f$ in the code comments!), 
because the linear terms matrix is applied to \textbf{the array called \name{fdisi} which always contains $g$}.

\section{Code parallelisation}
\label{parallelisation}

At present, the parallelisation of the code is done primarily via decomposition
of the distribution function grid and the various species in a given run, using
MPI. There are now shared memory OpenMP directives implemented into the code
which enable more efficient parallelisation in some cases with a large number
of parallel processes. This is because performing parallel reduction operations
between a large number of processes becomes expensive; at some point, when
doubling the number of available processor cores, work sharing the local data
between 2 cores can become more efficient than doubling the number of MPI
processes.
The implementation and testing of the OpenMP parallelisation can be
found in Sec. \ref{sec:hybrid}. Here we describes the various
MPI parallelisation options and their potential consequences for code
performance. In the next section, a scaling test is shown for a Cray XT4.

\subsubsection{Simple parallelisation via the fields}

The equation for the distribution function
(without collisions) contains no derivatives in the magnetic moment $\mu$, which means that the distribution function for any
point in the $\mu$ grid can be updated independently of the other points in the $\mu$ grid. Similarly, the distribution
function for each species can be updated independently. Therefore, one can run up to
{\it number of grid points in mu} $\times$ {\it number of species} parallel processes (typically 32-64 processes) to update
the distribution function if parallelising over these two `directions'.
However, these points are coupled via the fields, which must be updated before the distribution
function can be calculated in an explicit time step. The field equations 
involve a summation over the whole of the velocity space and the species, for which the code requires
an inter-process reduction sum. In parallelising over these `trivial' directions, a process
can first perform a partial sum over local species and $\mu$-grid points, before participating in a sum
involving all the parallel processes. For each field, this sum must be performed at every point in the
3-dimensional space, and the result must be then known to every process. As the number of utilised processor cores increases, the
amount of data per core that is involved in the sum remains constant. How much the parallel efficiency of the code decreases
as the number of utilised processors is increased will depend on the algorithm used by MPI_ALLREDUCE and underlying system.
In this simple parallelisation case (and potentially in other cases) this ALLREDUCE operation is the bottleneck in parallel performance.

\subsubsection{Parallelisation over the $v_\parallel$ grid}

The equations have derivatives in parallel velocity, which in their numerical implementation involve points of the
distribution function along lines in the parallel velocity grid. The derivative at the local point requires at most
the value of the distribution function at 1 or 2 adjacent points (second or fourth-order scheme) in each direction along the parallel
velocity grid (see Sect~\ref{sec:finite-differences}). When decomposing the $v_\parallel$ grid, the calculation of a
$v_\parallel$ derivative at the end of a local grid will require 1 or 2 more points managed by another process.
These points need to be communicated from the
adjacent processor before the derivatives are performed. Typically, the 2 points at either end of the local parallel velocity
grid must be communicated to the corresponding adjacent process, and 2 points must be received from it.
Therefore, if we parallelise over the $v_\parallel$ grid we have this communication to perform in addition to that required
to calculate the fields as discussed in the previous subsection. On systems which support performing computation at the same
time as communication (such as a Cray XT4), this communication can be overlapped with the relatively
expensive nonlinear terms computation. However, in such a case we would still be left with the same field calculation
bottleneck as the number of processors responsible for the parallel velocity grid is increased.

\subsubsection{Parallelisation over the $s$ grid}

In terms of the communication required for the derivatives along the $s$ direction, parallelisation over the $s$ grid is much
the same as for the $v_\parallel$ grid. However, splitting up the $s$ direction decomposes the space over which the fields
are calculated. This means that the ALLREDUCE operation for a given processor only needs to be performed between processors
responsible for the same part of the $s$ grid. If the number of processors over which the $s$ grid is decomposed is doubled,
then the data involved in the reduction sum is halved, and the number of processors that partake in the sum remains the same.
A consequence of this is that when parallelising only in the $s$ direction, no communication is needed to calculate
the fields. This is typically very favourable for parallel performance; one potential bottleneck would be effectively
removed (and the other can usually be overlapped with computation).

\subsubsection{Parallelisation over the $\mu$ (and $v_\parallel$) grid with collisions}

If the collision operator is used, the parallelisation over the $\mu$ grid is not simply via the fields as derivatives are
present. The second-order local derivative for the collision operator uses all the adjacent points in the velocity space
(see Sect. \ref{numerics:collisions}). Although only 1 point must be exchanged between adjacent processors, any 1 processor
may need to perform this exchange between 8 processors. If parallelising over the $v_\parallel$ and $\mu$ grids at the same time, this would mean
communication between 6 additional processors (when the grid is decomposed over more than 2 processors in each direction) when compared to a run without
collisions. Ideally we would like to avoid parallelising over both the $v_\parallel$ and $\mu$ grids till other options have been exhausted.
Parallelising over the $v_\parallel$ grid, but not the $\mu$ grid should be equally expensive with or without collisions, since there is no difference
in the grid points communicated. Parallelising instead over just the $\mu$ grid, but not the $v_\parallel$ grid would be more expensive with collisions than without, although typically cheaper (when using the default
fourth-order finite-differences) than the previous case using only $v_\parallel$.

\section{Parallel performance (pure MPI), Cray XT4 \label{scaling-orig}}

\begin{figure}[h] 
\begin{center}
\includegraphics[width=0.5\textwidth,clip]{performance}
\caption{
  Parallel performance for a typical sized problem on HECToR (Cray XT4), both without and 
  with the collision operator turned on. Due to machine memory constraints, this problem can run with
  a minimum of 64 cores without collisions and 128 cores with. The performance/speedup is based on the
  time taken to perform the main time integration loop of the code and is scaled such that it is equal
  to the number of cores when the smallest possible number of cores is used. The parallel efficiency,
  based on $100$\% for the smallest number of cores, is noted at several points on the plot.
}
\label{scaling}
\end{center}
\end{figure}
Here we make a simple assessment of the parallel performance of the code on a Cray XT4 (HECTOR). We define the performance
as the amount of main time loop computation done by the code per unit time. In the ideal case, a doubling in the
number of processor cores used will double the amount of computation done per unit time (or halve the time required
to do a fixed number of time steps).
We consider a non linear problem of fairly typical size that we believe to be reasonably well converged with respect to grid
resolution. As mentioned in Sect.~\ref{memorylayout}, we can change the data memory layout to potentially
tweak the parallelisation and cache efficiency of the code. We have experienced some variation in performance due to these
adjustments. However, we do not yet have enough experience of how to optimally prescribe the layout depending on grid size and parallelisation options,
so for the purpose of this test, we keep the layout fixed to the default. We perform a test with a resolution of 83 radial wave vectors, 21 bi-normal modes, 4 species, 16 $s$ grid points, 16 $\mu$ grid points and 64 $v_\parallel$ grid
points, using fourth-order finite-differences in double precision. Two cases are considered, one using the collision operator and one
without. Due to the memory available per processor core, a minimum of 64 processors are required for the former case and 128 for the latter. 
Using a second-order finite-difference scheme and/or a single precision calculation can reduce this minimum. From the parallel algorithm
perspective, the upper limit on the number of parallel processes used for this case is 16384. Using second-order finite differences would increase this maximum to 65536 cores.
However, only up to 8192 cores were available on the hardware at the time of testing.

Before deciding how to producing a scaling plot for this testcase (such as in Fig.~\ref{scaling}), some preliminary investigations on
the effects of parallelising in each direction were made. Using the `trivial' directions alone, the test case parallelises over the minimum
64 processors for the default case. The difference in performance
between this parallelisation and parallelising primarily over the parallel velocity grid is fairly small, the `trivial' directions giving
less than $10$\% better performance than with the $v_\parallel$ direction. Parallelising primarily over the $s$ grid at
64 cores (using the maximum number of cores in the $s$ direction before using the simple direction) yields a performance which is almost identical
to that using only the `trivial' directions. It seems that for 64 processors, the communication needed for the derivatives has little (if any) effect
on the parallel performance. To get an idea of the relative expense of parallelising over $s$ and $v_\parallel$, two tests were made by starting with 64
cores (using only the trivial parallelisations), then varying either the number of processors in the $v_\parallel$ direction
or the $s$ direction until a total of 512 cores were used. The reduction in efficiency with increasing processors was greater with $v_\parallel$, with the $s$
parallelisation performing around $6\%$ better at 512 cores (at $>95\%$ of the efficiency at 64 processors).

A scaling test was performed for the collisionless case up to 8192 cores. Fig.~\ref{scaling} shows how the performance increases with the number of cores.
For the first point at 64 cores, only the `trivial' directions are used for the parallelisation, since that combination provides the best performance 
(but only by around $1\%$). Since the $s$ direction is the preferred parallelisation direction (least expensive in terms of performance), the number of 
processors in $s$ doubled between each point on the graph up to 512 cores, where the maximum of 8 processors in $s$ is used. Beyond that point, the more expensive $v_\parallel$ grid is decomposed until
8192 cores are used. At 8192 cores, the code runs at $63\%$ of the efficiency at 64 cores.

Following the discussion in the previous subsections, we should expect that with collisions, parallelisation over the $\mu$ direction
is more expensive than over the $s$ direction. Therefore, the scaling test for collisions (as shown in Fig.~\ref{scaling}) starts at 128 processors fully parallelised over the $s$ direction
and over the species, using only 4 processors in $\mu$. Then, the number of processors in $\mu$ is doubled, until 512 cores are used. At 1024 cores, it is cheaper to use 32
processors in $v_\parallel$ and only $1$ in $\mu$, since this avoids using parallelisation in both those directions together. For the final 2 points on the graph, the 
number of processors in the $v_\parallel$ and $\mu$ are either identical, or there are twice as many in $v_\parallel$. Since these
parallelisation combinations give the best performance (considerably better than using the same ones as without collisions), careful consideration should be given to
the number of processors used in each direction when using collisions. At 4096 cores the efficiency is $63\%$ of that at 64 cores.

In both cases considered we see decreasing performance gains when we begin to parallelise in
less favourable directions. Running the tests in single precision would mean less data communicated, so may improve
scaling (but at the same time would likely reduce the time between communication, or the time available for overlapping communication with computation). Running with the second
order finite difference scheme would allow double the number of cores in the $s$ direction and may improve the efficiency
at a large number of cores. If different test cases are considered, it may well be that more resolution in a particular direction
is needed, which would then lend itself to further scaling.

% Something about cartesian topology ordering
For hardware in which not all processor interconnects are equal in speed, the ordering of the MPI communicators within the cartesian topology 
can have a significant effect on parallel performance.  Current high performance machines appear with multiple shared memory cores on interconnected nodes are one example.  Given the above discussion of the bottleneck in global reduction operations, it is preferable that the $s$ direction decomposition should be over the slowest communication direction.  The GKW default setup assumes that MPI rank ordering is done with adjacent physical processors adjacent in the rank ordering.  Testing of a collisionless case on a Cray XT6 (HECTOR phase-2b with 24 cores per node) with 1344 cores indicated that the default dimension ordering (Species, $\mu$, $v_\parallel$, $s$) was optimal.  The configuration ($v_\parallel$, Species, $\mu$, $s$), was 30\% slower, even with exact multiples of the $v_\parallel$ grid (presumably) on a single node.  This is counter-intuitive, since one would expect putting the $v_\parallel$ direction first to be optimal. Configurations with $s$ first were up to 80\% slower. The ordering can be simply reconfigured near the top of \File{grid.F90}.  The optimal configuration will depend on hardware and problem size, and conducting tests prior to expensive runs is recommended.

\section{Parallel performance (hybrid, MPI+OpenMP), BlueGene/P \label{sec:bluegene}}
The performance of pure MPI parallelisation has also been compared to hybrid parallelisation MPI+OpenMP (see Sec.~\ref{sec:hybrid} for details on the hybrid scheme) for a typical non-linear electromagnetic case with kinetic electrons. The problem size is almost the same as in the previous section: 83 radial wave vectors, 20 bi-normal modes, 4 species, 32 $s$ grid points, 8 $\mu$ grid points and 64 $v_\parallel$ grid. A case without collisions and a case with collisions ($R_\textrm{ref}=3$, $T_\textrm{2}$ and $n_\textrm{ref}=6$, was investigated.  In contrast to the previous section, the full collision operator \textit{with} mass and momentum conservation was used\footnote{How much the scaling is affected by the communication required for the momentum conservation has not yet been fully investigated by comparisons on the same machine. Limited tests on BABEL indicated that momentum conservation caused a consistent 4\% slowdown up to 4096 cores, but it could become more (or less) important at higher number of cores (note that it is overlapped with derivatives communications when the MPI implementation supports this).  The version of GKW tested on BABEL did not yet have OpenMP implemented in the calculation of the momentum conserving field. Comparison of timings for the regular fields indicate that implementing this would give a 4\% (2\%) speedup at 4 (2) threads without MPI, but with MPI it will likely make no difference if the scaling bottleneck is in the overlapped derivatives communication.}.

The tests were conducted on the BlueGene/P BABEL (4 cores at 850MHz and 2GB of memory per node \url{http://www.idris.fr/eng/User_Support/bg_support.html}) for which the number of OpenMP threads per node can be 1, 2 or 4. The executable was compiled from revision 2076 of GKW source with O4 optimisation level (typically $10\%$ faster that O2/O3 optimisation for a compilation without OpenMP and $5\%$ faster for a compilation with OpenMP) and double precision. Due to the small amount of memory per node available on BABEL at least 512 cores were required to run the problem under consideration (to be accurate, the run fits on 256 cores but the execution is then very slow). For all the runs in this section, the number of MPI processes in the species and $s$ direction was \name{n_procs_sp}=4 and \name{n_procs_s}=16, respectively. The number of processes in the $\mu$ direction was \name{n_procs_mu}=8 for the collisionless case and \name{n_procs_mu}=4 for the case with collisions (except when 4 threads were used at 512 cores where \name{n_procs_mu} was reduced to 4 and 2, respectively). This means that the MPI parallelisation over the three most favorable directions was maximised to start with and the scaling of the least favorable direction, $v_\parallel$, was then compared to the OpenMP scaling by increasing either \name{n_procs_vpar} or the number of threads per node. For instance, a collisonless run on 2048 cores was performed on 1 thread with \name{n_procs_vpar}=4 or on 2 threads with \name{n_procs_vpar}=2.  The tests were conducted with a typical set of diagnostics with NAVERAGE=100, and with the nonlinear timestep estimator on, both were found to have little effect on the scaling performance (for the single threaded case at least).

The results of the tests are summarised in Table~\ref{table:nocol} (no collisions) and Table~\ref{table:col} (with collisions). The corresponding speed-up and total main loop time is displayed in Fig.~\ref{scalingBABEL} and \ref{scalingBABEL:total}, respectively. Hybrid parallelism with 4 threads is especially useful with collisions and allows to recover the scaling obtained without collisions near the MPI limits. 
\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
cores & 512 & 1024 & 2048 & 4096 & 8192 & 16384\\
\hline
1 thread - no OMP& 203.5h & 212.8h & 227.9h& 255.3h & 304.3h & 408.4h \\
(time, efficiency) & 100\% & 95.6\%  & 89.3\%  & 79.7\%&66.9\% & 49.8\%\\
\hline
1 thread & 214.4h & 222.1h & 237.0h& 263.6h & 316.6h & 551.0h \\
(time, efficiency) & 100\% & 96.5\%  & 90.5\%  & 81.3\%& 67.7\% & 38.9\%\\
\hline
2 threads &221.1h & 225.5h & 233.3h& 255.1h & 313.2h & 493.1h \\
(time, efficiency) & 100\% & 98.0\%  & 94.8\%  & 86.7\% & 70.6\% & 44.8\%\\
\hline
\end{tabular}
\caption{Case without collisions. Total main loop time (in hours) for 600 time steps and strong scaling efficiency based on the case with 512 cores.}
\label{table:nocol}
\end{center}
\end{table}
Compilation without OpenMP is slightly faster (due to different cache efficiencies in the matrix ordering). This was also observed on HECTOR. (This has since been changed, both now use the same sort ordering so there is no longer any cache penalty for openmp)
\begin{small}
\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
cores & 512 & 1024 & 2048 & 4096 & 8192 & 16384\\
\hline
1 thread& - & 290.0h & 317.9h & 373.1h & 519.2h & - \\
(time, efficiency) & (100\%) & (92.6\%)  & (84.4\%)  & (71.9\%)& (51.7\%) & - \\
\hline
2 threads & 268.4h & 283.0h & 300.4h& 323.4h & 480.6h & 713.6h \\
(time, efficiency) & 100\% & 94.8\%  & 89.4\%  & 83.0\%& 55.9\% & 37.6\%\\
\hline
4 threads & 281.7h & 283.5h & 294.9h & 314.6h & 380.5h & 506.1h \\
(time, efficiency) & 100\% & 99.4\%  & 95.5\%  & 89.5\% & 74.0\% & 55.7\%\\
\hline
\end{tabular}
\caption{Case with collisions (including momentum conservation). Total main loop time (in hours) for 600 time steps and strong scaling efficiency based on the case with 512 cores. On 1 thread, the run at 512 cores was out of memory and the strong scaling efficiency was calculated using the results on 2 threads. }
\label{table:col}
\end{center}
\end{table}
\end{small}

\begin{figure}
\begin{center}
\includegraphics[width=0.4\textwidth]{speedupnocollisions.eps}%
\includegraphics[width=0.4\textwidth]{speedupcollisions.eps}
\caption{ Parallel performance for a typical sized problem on BABEL (BlueGene/P), both without (left plot) and 
  with (right plot) the collision operator (with momentum conservation) turned on. The performance/speedup is based on the
  time taken to perform the main time integration loop of the code and is scaled such that it is equal
  to the number of cores when the smallest possible number of cores is used.}
\label{scalingBABEL}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{totaltime.eps}
\caption{ Total main loop time in hours for the cases studied in this section.}
\label{scalingBABEL:total}
\end{center}
\end{figure}

A test has also been performed with the number of modes increased to NX=183 and NMOD=32 with the results summarised in Table~\ref{table:bigcol}.
\begin{small}
\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
cores & 2048 & 4096 & 8192 & 16384\\
\hline
1 thread& 1166.4h & 1369.1h & 1731.7 h & - \\
(time, efficiency) & 100\% & 85.2\%  & 67.4\%  & -\\
\hline
2 threads & 1094.9h& 1164.4h & 1412.5h & 2317.3h \\
(time, efficiency) & 100\% & 94.0\%& 77.5\% & 47.3\%\\
\hline
4 threads &1081.7h & 1136.5h & 1305.4h & 1692.5h \\
(time, efficiency) & 100\% & 95.2\% & 82.9\% & 63.9\%\\
\hline
\end{tabular}
\caption{Case with collisions (with momentum conservation), NX=167 and NMOD=32. Total main loop time (in hours) for 600 time steps and strong scaling efficiency based on the case with 2048 cores. }
\label{table:bigcol}
\end{center}
\end{table}
\end{small}

\pagebreak

\section{Parallel nonspectral performance (pure MPI), Helios}

The nonspectral version of the code additionally parallelizes over the radial (x) direction (with the exception an implicit solve for the polarization term).  The nonspectral 
version of the code spends proportionally more time in the fields calculations than the spectral version, since the multi-point averages use for the gyroaverages are much slower 
than the Bessel functions of the spectral version. (The nonlinear terms, by contrast, are much faster since they only require 1D FFTs.)

Since the ``diagonal part'' of the fields solve is only 3 dimensional, it cannot parallelise over the velocity space or species.   This is true in both the spectral and nonspectral
versions, but in the spectral version, the diagonal part of the fields solve takes a neglible of CPU time.  To allow efficient scaling for the nonspectral version, the fields solve 
work is additionally parallelised over the toroidal modes using the same communicators that are elsewhere used to distribute work over the velocity space and species. 
This feature also reduces the fraction of time spent in the implicit polarization solve, which is radially global. Two allgather communications are therefore required in the fields
solve: The first in x before the polarization solve, and the second in nmod after the polarization solve so that every processor obtains the result for every toroidal mode. 
Since these are both gather operations of 3d quantities, they are relatively fast compared to the reductions and communications of ghost cells.

Parallelization over the x direction brings the same benefits for reducing the reduction operations in the field solve as the parallelism over the s direction. The additional
penalties of extra ghost cell and allgather communications in the fields solve are outweighed by the benefit that this brings.  The result is that parallelism over x is always more
efficient than parallelism over parallel velocity (which in the nonspectral case performs much worse than in the spectral case due to the larger fraction of time in the field
solve).  For more details see \raw{doc/GKW_MPI_nonspectral.pdf}.

Scaling tests of the radially nonspectral version of the code (Fig. \ref{scalingHelios}) were performed on the Bull IFERC Helios machine (Intel Sandy Bridge with fat tree topology Infiniband
communication network).  As with the scaling test described in Sec. \ref{scaling-orig}, the `trivial' parallel directions were used first, followed by s, then x (although not to maximum), then finally vpar.
Two cases were chosen to represent a medium and a large physics case simulation, with gridsizes
\begin{verbatim}
NX = 256, N_s_grid = 32,  N_mu_grid = 16,  N_vpar_grid = 32,  NMOD = 16, number_of_species = 2
\end{verbatim}
for the medium case, and 
\begin{verbatim}
NX = 512, N_s_grid = 64,  N_mu_grid = 16,  N_vpar_grid = 64,  NMOD = 43, number_of_species = 2
\end{verbatim}
for the large case.

The radial box size was \texttt{lx = 128} or \texttt{lx = 256} respectively, which resulted in 6 x ghost points required for the gyroaverages (only two are communicated for the derivatives).  Since the number of ghost points must be less than the number of local points, this sets the maximum limit of \texttt{n_procs_x = 32} or \texttt{64} repsectively.

\begin{figure}[htb!]
\begin{center}
\includegraphics[width=0.48\textwidth]{scaling-helios-med3}
\includegraphics[width=0.49\textwidth]{scaling-helios-large3}
\caption{ Strong scaling performance for the medium and large sized nonspectral problems on Helios. 
          The speedup is based on the time taken to perform the main time integration loop 
          of the code and is scaled such that it is equal to the number of cores when the smallest possible number of cores is used.
          BullX MPI seems to perform better at ghost cell communications of derived datatypes while Intel MPI seems to perform better at the large
collective operations.  Which is better therefore depends on which MPI communication is dominant for a given problem.}  
\label{scalingHelios}
\end{center}
\end{figure}

For the medium problem size, the maximum possible \texttt{n_procs_x=32, n_procs_vpar=16} would allow a theoretical maximum of 131,072 cores (65,536 with collisions).  For the large problem size, the maximum possible \texttt{n_procs_x=64, n_procs_vpar=32} allows a theortical maximum of 1,048,576 cores (524,288 with collisions).  These results compare favorably with the spectral scaling tests: 
The parallelism over x removes any scaling performance penalty associated with the use of the collision operator
(which adds 20\% to all runtimes).

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/neoclassics.tex
(1) Added gkw Clone gwk  #3

%\fancyfoot[L]{$ $Date$ $}
%\fancyfoot[R]{$ $Revision$ $}

\chapter{Neoclassical background}

By default the background distribution function in considered to be a Maxwellian.  However due to background flows the temperature and density 
gradients, coupled with the geometry and collisions, the distribution function can deviate from the Maxwellian by a small amount.   This section describes the implementation of the interface between NEO and GKW so that small deviations from background Maxwellian can be included.

\section{Form of the equations for arbitrary background}

The corrections to the background Maxwellian are calculated, at the moment, by the Eulerian
Neoclassical transport code {\small NEO} \cite{NEO1,NEO2}.

We start with the gyro-kinetic Vlasov equation in (${\bf X}, v_\parallel, 
\mu$) coordinates,
\begin{equation}
{\partial f_{\rm tot} \over \partial t} + {{\rm d} {\bf X} \over {\rm d} t} \cdot \nabla_\mu 
f_{\rm tot} + {{\rm d} v_\parallel \over {\rm d} t} {\partial f_{\rm tot} \over \partial v_\parallel} = 0 .
\label{eq:neoclassics-vlasov}
\end{equation}

Finally
\begin{eqnarray}
\frac{\partial f}{\partial t} + \underbrace{v_{||}{\bf b}\cdot\nabla f}_{I} + \underbrace{{\bf v}_D\cdot\nabla f}_{II} + \underbrace{{\bf v}_\chi \cdot\nabla f}_{III} - \underbrace{\frac{{\bf b}}{m}\cdot (\mu\nabla B + \nabla\epsilon_\Omega)\frac{\partial f}{\partial v_{||}}}_{IV} = \\
-v_{||}{\bf b}\cdot\nabla F_{0} -\underbrace{({\bf v}_D + {\bf v}_\chi)\cdot\nabla F_{0}}_{VI + V} + \underbrace{\frac{Ze}{mv_{||}}(v_{||}{\bf b} + {\bf v}_D)\cdot\nabla\phi \frac{\partial F_{0}}{\partial v_{||}}}_{VII + VIII} + \frac{({\bf v}_D + {\bf v}_\chi)}{mv_{||}}\cdot\mu\nabla B\frac{\partial F_{0}}{\partial v_{||}} + C(f)
\end{eqnarray}
Where the first term on the right hand side and Term VI are the terms in the drift-kinetic equation that is solved by NEO:
\begin{equation}
{\bf v_{D}}\cdot\nabla F_{0}+ v_{||}{\bf b}\cdot\nabla F_{0} -\frac{\mu}{m}{\bf b}\cdot\nabla B\frac{\partial F_{0}}{\partial v_{||}}= C(F_{0}).
\end{equation}
This equation determines the form of the perturbation to the background distribution function.

The background distribution function is split into two stationary components, $F_0 = F_{M} + F_{\mbox{neo}}$, the first term is a lowest order Maxwellian and the second is its neoclassical correction.  The Maxwellian distribution function and its derivatives have the form:

\begin{eqnarray}
F_{M} = \frac{n_{0}}{\pi^{3/2}v_{th i}^{3}}\exp{\left(-\frac{(v_{||}-RB_{t}\omega_{\phi}/B)^{2} + 2\mu B/m}{v_{th i}^{2}}\right)}\\
\nabla F_{M} = \nabla P F_{M} - \frac{\mu\nabla B}{T}F_{M} \\
\partial F_{M}/\partial v_{||} = -\frac{mv_{||}}{T}F_{M} \\
\nabla P = \frac{\nabla n}{n} + \left(\frac{v_{||}^{2}}{v_{th}^{2}} + \frac{\mu B}{T} - \frac{3}{2}\right)\frac{\nabla T}{T} - \frac{mv_{||}}{T}\frac{RB_{t}}{B}\nabla\omega_{\phi} \nonumber\\
\end{eqnarray}
where Eq.17 is the Maxwellian derivative considering only the thermodynamic quantities, the gradient in the magnetic field we leave separate.  Derivatives of the perturbed background are taken numerically using fourth order central differencing.  There is a subtlety in the equations where, when considering a Maxwellian background the gradient in the magnetic field terms cancel 
exactly with the trapping force term acting on the Maxwellian. 
The background Maxwellian has only a radial derivative and $\omega_{\phi}$ is the angular toroidal rotation frequency.

The source term for the Maxwellian component of the background for each species, $s$, ($S(F_{0}) = S(F_{M,s})+S(F_{\mbox{neo},s})$) is given by:

\begin{eqnarray}
S(F_{M,s}) = -\frac{1}{Z_{s}eB}\left(\frac{mv_{||}^{2}}{B}+\mu\right)\left(\frac{{\bf b}\times\nabla B}{B}\right)\cdot F_{M,s}\nabla P -\left(\frac{{\bf b}\times\nabla \phi}{B}\right)\cdot F_{M,s}\nabla P \nonumber\\
-\frac{v_{||}Ze}{T_{s}}{\bf b}\cdot\nabla\phi F_{M,s} -\frac{1}{T}\left(\frac{mv_{||}^{2}}{B}+\mu\right){\bf b}\cdot\frac{\nabla B\times\nabla\phi}{B}F_{M,s}.
\end{eqnarray}

Terms involving gradients of the magnetic field, $\mu\nabla B$  cancel out, leaving only terms involving $\nabla P$.  In contrast, for the neoclassical correction this cancellation does not occur, the source term has the form,

\begin{eqnarray}
S(F_{\mbox{neo}},s) = -\frac{1}{Z_{s}eB}\left(\frac{mv_{||}^{2}}{B}+\mu\right)\left(\frac{{\bf b}\times\nabla B}{B}\right)\cdot \nabla F_{\mbox{neo},s} -\left(\frac{{\bf b}\times\nabla \phi}{B}\right)\cdot\nabla F_{\mbox{neo},s}+ \nonumber\\ \frac{Z_{s}e}{m}{\bf b}\cdot\nabla\phi \frac{\partial F_{\mbox{neo},s}}{\partial v_{||}} + \frac{1}{mv_{||}}\left(\frac{mv_{||}^{2}}{B}\right){\bf b}\cdot\frac{\nabla B\times\nabla\phi}{B} \frac{\partial F_{\mbox{neo},s}}{\partial v_{||}}.
\end{eqnarray}

\section{Equation modifications}

For the sake of completeness and to document exactly
which equations are being solved, the modifications to the equations in the electrostatic limit are written here.
The equations, as in the previous sections, can be written in the form 
\begin{equation}
\label{eqs:neoclassics-complete-set}
{\partial f \over \partial t} = {\rm III} + {\rm IV} + {\rm V} + {\rm VII}+ {\rm VIII}.
\end{equation}
Only terms that are modified by the
neoclassical background are described here, then each term is, in its normalised form:

\begin{eqnarray}
{\rm III} &\rightharpoonup& - {\bf v}_\chi \cdot \nabla f =  \frac{{\bf b}\times\nabla\phi}{B}\cdot\nabla{f} \rightarrow 
\imath k_{\zeta}\varepsilon^{\psi\zeta} \frac{\partial\phi_{\mbox{neo}}}{\partial\psi} f + \imath k_{\zeta}\varepsilon^{s\zeta} \frac{\partial\phi_{\mbox{neo}}}{\partial s} f + \imath k_{\psi}\varepsilon^{s\psi} \frac{\partial\phi_{\mbox{neo}}}{\partial s} f \nonumber\\
\noalign{\vskip 0.6 truecm}
{\rm IV} &\rightharpoonup& -\frac{1}{mv_{||}}\left(v_{||}{\bf b}\cdot\left(\mu\nabla B + Ze\nabla\phi\right)\right)\frac{\partial f}{\partial v_{||}} 
= -\frac{1}{m}\left({\bf b}\cdot\left(\mu\nabla B + Ze\nabla\phi_{\mbox{neo}}\right)\right)\frac{\partial f}{\partial v_{||}} \nonumber\\ &\rightarrow& -\left(\mu_{N}v_{s}\mathcal{F} \frac{\partial B_{N}}{\partial s} + \frac{v_{s}Z}{2T_{s}}\mathcal{F}\frac{\partial\phi_{\mbox{neo},N}}{\partial s}\right)\frac{\partial f}{\partial v_{|| N}}\nonumber\\
\noalign{\vskip 0.6 truecm}
{\rm V} &\rightharpoonup& - {\bf v}_\chi \cdot \nabla F_0 =\frac{{\bf b}\times\nabla x^{\alpha}}{B}\imath k_{\alpha}\phi\cdot\nabla F_0 \rightarrow
\varepsilon^{\alpha\psi}\imath \rho_{i}k_{\alpha}\cdot\nabla\psi_{N} \frac{\partial F_{M}}{\partial\psi}\phi_{N} + \imath\varepsilon^{\alpha\beta} \rho_{i}k_{\alpha}\phi_{N}\frac{\partial F_{\mbox{neo}}}{\partial x^{\beta}}\nonumber\\
\noalign{\vskip 0.6 truecm}
{\rm VII} &\rightharpoonup& \frac{Ze}{m_{s}} {\bf b}\cdot\nabla\phi \frac{\partial F}{\partial v_{||}}  \rightarrow -\frac{Zv_{s}}{T_{s}} v_{||N} \mathcal{F}\frac{\partial\phi_{N}}{\partial s} F_{M} + \frac{Zv_{s}}{2T_{s}} \mathcal{F}\frac{\partial\phi_{N}}{\partial s} \frac{\partial F_{\mbox{neo}}}{\partial v_{N||}} \nonumber\\
\noalign{\vskip 0.6 truecm}
{\rm VIII} &\rightharpoonup& -\frac{Ze}{v_{||}m_{s}} \vec{v_{D}}\cdot\nabla\phi \frac{\partial F}{\partial v_{||}} \rightarrow-i\frac{k_{\alpha}Z}{T_{s}}\vec{v_{D}}\cdot\nabla x^{\alpha}\phi_{N}F_{M}+\frac{\imath k_{\alpha}Z}{2T_{s}v_{||N}} \vec{v_{D}}\cdot\nabla x_{a}\phi_{N} \frac{\partial F_{\mbox{neo}}}{\partial v_{||N}}\nonumber\\
\end{eqnarray}

\section{The NEO code}

{\small NEO} outputs the neoclassical distribution function, $F_{NEO}$, and the neoclassical electrostatic 
potential, $\phi_{NEO}$.   To run {\small NEO} in a way that can be read by {\small GKW} the following input files are needed for  {\small NEO}:
\begin{itemize}
\item {\bf input.neo} - Input file
\item {\bf input.profiles} - The radial profiles of experimental data.  Needed for the normalisation of the velocities.
\end{itemize}
When run the following files must be in the folder that {\small GKW} runs in,
\begin{itemize}
\item {\bf out.neo.equil} - Equilibrium quantities per species
\item {\bf out.neo.f} - The distribution functions in spectral form
\item {\bf out.neo.grid} - The poloidal and radial grid sizes and points
\item {\bf out.neo.expnorm} - The normalising quantities
\end{itemize}
The latter file is only produced when {\small NEO} is run in profile mode which requires a relevant {\bf input.profiles} file.  Files for the Cyclone base case benchmark outlined in the next section can be found below and in the repository.

The neoclassical component of the background distribution function, $F_{\mbox{neo}}$ is calculated and output by {\small NEO}, which is then read into GKW and transformed from a harmonic expansion in velocity space into real velocity space as utilised by GKW.  In {\small NEO} the non-adiabatic part of the distribution function is represented by.
\begin{eqnarray}
G_{\mbox{neo},s} = F_{M,s}\sum_{l=0}^{N_\xi}\sum_{m=0}^{N_{E}}\hat{g}_{l,m}P_{l}(\xi)L^{k(l)+1/2}_{m}(v_{s}^{2}/v_{th}^{2})(v_{s}/v_{th})^{k(l)},\nonumber\\
\end{eqnarray}
where $P_{l}(\xi)$ are Legendre polynomials and $L_{m}^{\alpha}(v)$ are Laguerre polynomals and $\hat{g}_{l,m}$ are the amplitudes and is related to the distribution function, $F_{\mbox{neo},s}$ by, $F_{\mbox{neo},s}=G_{\mbox{neo,s}}-F_{Ms}\phi_{\mbox{neo}}e/T_{s}$.  As a benchmark of this process, Figure \ref{neo:flows} shows the flux averaged parallel flow, $\langle u_{||}B\rangle = \int ds B\int B dv_{||}d\mu v_{||}f$, and critically, its radial gradient as calculated by NEO and by GKW after the transformation is performed for the parameters as described in the next section.  It can be seen that agreement is excellent, with the values of the flow matching to within 2\% error. 
\begin{figure}
\begin{centering}
\includegraphics[width=8cm,clip]{GKWvNEOflow2.eps}
\caption{The first order parallel flow(left) , $\langle u_{||}B\rangle$ and its radial gradient (right) as calculated by NEO (green) and then from the neoclassical distribution function once it has been read and transformed into GKW coordinates (blue) as a function of collisionality.  Agreement in flow magnitude is to within 2\%. }
\label{neo:flows}
\end{centering}
\end{figure}

\section{Benchmark}

In this section a benchmark of the independent implementations of the neoclassical background in GKW and GS2 \cite{DOR00} is outlined.  Then benchmark of the interface is based on the standard benchmark for linear problems.  The parameters used are based on those described by the Cyclone Base Case \cite{Cyclone}: A circular flux surface equilibrium \cite{Lapillonne}, the safety factor, $q=1.4$, magnetic shear, $\hat{s}=0.8$, inverse aspect ratio of the $r/a=0.5$ flux surface, $\epsilon = 0.18$, $R/a = 2.78$.  The logarithmic density and temperature gradients for both the bulk ions and kinetic electrons are $R/L_{Ti} = R/L_{Te} = 6.9$, where $R/L_{Ti} = R\partial \ln{T_{i}}/\partial r$ and $R/L_{ni} = R/L_{ne} = 2.2$. The ratio of the temperatures is, $T_{e}/T_{i}=1.0$.  The value of the normalised gyro-radius (in the case of GKW normalised to the major radius, R) is $\rho_{*}=\rho_{i}/R = 0.01$.

In {\small GKW} the calculation of the radial derivatives requires five equally spaced flux surface calculations for fourth order centrally differenced radial derivatives to be performed.  Here, these are chosen to be, $r/a = [0.49\ 0.495\ 0.5\ 0.505\ 0.51]$.

The grid resolutions for simulations performed with NEO were, $N_{E}=10$, $N_{\xi}=19$, $N_{\theta}=41$ for the energy, angular polynomials and in the poloidal directions respectively. The Full Fokker-Plank collision operator was used.  In GKW, $N_{v_{||}}=64$, $N_{\mu}=16$, $N_{s}= 30$ in the parallel velocity, magnetic moment and parallel coordinate directions respectively,  $N_{x}=21$ radial modes were used.

The diamagnetic component of the toroidal rotation frequency is given by the expression \cite{BarnesParra13},
\begin{eqnarray}
\frac{\omega_{\zeta,d}v_{th i}}{R_{0}} =
\sum_{s} \left\{ m_{s}R\left(v_{||}\frac{\vec{B}\cdot\nabla\zeta}{B}\right)F_{\mbox{neo}}\right\}/\sum_{s}m_{s}n_{s}\{ R^{2}\}.
\end{eqnarray}
\begin{figure}
\begin{centering}
\includegraphics[width=9cm,clip]{Flowscombined.eps}
\includegraphics[width=9cm,clip]{GKWGS2QLbenchmark.eps}
\caption{(top) The ratio of the quasi-linear radial flux of toroidal momentum ($\Pi_{i}$) with the radial heat flux ($Q_{i}$) as a function of collisionality for both GKW-NEO (red) and GS2-NEO (blue) at $\rho_{*}=0.089$.  Also plotted (black dashed line) is $a(1-k_{a})$ where a is fitted to the value from GKW-NEO at the zero collisionality limit.  Black crosses represent data from non-linear simulations, the error bars represent the standard deviation of the fluctuations over the statistical steady state part of the simulation.  (bottom) The toroidal diamagnetic frequency, $\omega_{\zeta,d}$ (left) and its radial gradient, $\partial\omega_{\zeta,d}/\partial r$ (right) for runs with GKW-NEO (blue) and GS2-NEO (red) as a function of the normalised collisionality, GS2-NEO data taken from \cite{BarnesParra13} for $\rho_{*}=0.01$. }
\label{neo:gkwgs2comp}
\end{centering}
\end{figure}

Figure~\ref{neo:gkwgs2comp} (left) shows the diamagnetic flow and its gradients as a function of the normalised collision frequency for both GS2-NEO and GKW-NEO for the above parameters, showing good agreement between the two codes.   Figure \ref{neo:gkwgs2comp} shows the turbulent parallel momentum flux ($\Pi_{i}$) normalised to the radial heat flux ($Q_{i}$) for a scan in the collisionality as calculated by GKW-NEO (Red) and GS2-NEO (Blue) for a series of linear simulations at the single toroidal wave number of $k_{\zeta}\rho_{i} =0.4242$.

\section{Sample input}

A sample NEO input file for the Cyclone base case run, most of the parameters here are ignored as they are read from the input.profile file found in the benchmarks folder in the GKW repository

This uses the code in profile mode and the full Fokker-Plank collision operator.  Most of the parameters are overridden by those in the input profile file.

\begin{verbatim}
PROFILE_MODEL=2
PROFILE_EQUILIBRIUM_MODEL=1
EQUILIBRIUM_MODEL=1
COLLISION_MODEL=4
ROTATION_MODEL=1
N_ENERGY=10
N_XI=19
N_THETA=41
N_RADIAL=5
N_SPECIES=2
Z_1=1
Z_2=-1
MASS_2=  0.00027245
DENS_2=1
RMIN_OVER_A=0.49000000
RMIN_OVER_A_2=0.51000000
RMAJ_OVER_A=3.40730000
KAPPA=  1.28760000
S_KAPPA=  0.01117700
DELTA=  0.07159200
S_DELTA=  0.02895700
SHIFT= -0.10774000
Q=  1.24770000
SHEAR=  0.34412000
OMEGA_ROT=  0.05153600
OMEGA_ROT_DERIV=  0.03058600
RHO_STAR=  0.00776240
DLNNDR_1=  0.0
DLNTDR_1=  0.0
DENS_1=  0.82927000
NU_1=  0.00118530
TEMP_2=  1.17130000
DLNNDR_2=  0.0
DLNTDR_2=  0.0
TEMP_1=1
\end{verbatim}

In the GKW input file the following amendments need to be made to read in the NEO files correctly.

\begin{verbatim}
 &LINEAR_TERM_SWITCHES
 lneo_equil = .true.
 /
 \end{verbatim}

 It is possible to apply the neoclassical background to all or a simply a subset of the species in GKW.  Each species in GKW can be pointed to a specific species in NEO.  For example when

 \begin{verbatim}
&LINEAR_TERM_SWITCHES
 lneo_equil = .true.
 neo_equil_parse_sp_seq = 1,2,3,3
 /
 \end{verbatim}

 is used, for a four species run in GKW for a three species NEO run, then GKW species 3 and 4 use the flow from species 3 in NEO.  This allows the effects of flows on trace species with different parameters to be studied.

### 2022-05-03 gkw/doc/manual/practise.tex
(1) Added gkw Clone gwk  #3

\chapter{Gyrokinetic practice}

\section{Normalisations} 
\label{sec:normalisation}

Of course, all quantities in the code are normalised to typical reference quantities. 
This section will first describe the normalisation for the local (fluxtube) case. Afterwards differences for the global case are pointed out.

The reference quantities chosen for GKW are\\
\begin{tabular}{rl}
  \label{eq:refquantities}
  $\rho_\tref $ : & reference larmorradius \\
  $m_\tref $ : & reference masse \\
  $v_{thref}$ : & reference therm. velocity \\
  $n_\tref $ : & reference number density \\
  $T_\tref $ : & reference temperature \\
  $B_\tref $ : & reference magnetic field \\
  $R_\tref $ : & reference major radius \\
\end{tabular}

These are related with each other by 
\begin{align}
T_{\rm ref} &= \frac{1}{2} m_{\rm ref} v_{\rm thref}^2  
\label{eq:normalise-Tref}
\\
\rho_{\rm ref} &= {m_{\rm ref} v_{\rm thref} \over e B_{\rm ref}} = \frac{2T_\tref}{eB_\tref v_{th\tref}}.
\end{align}
The unit of temperature is energy. Then the symbol $k$ is used for
wavenumbers.  Naturally, the code uses only normalized quantities, and
physical units can only be obtained through the reference quatities.
These quantities are never explicitly specified. The numerical
solution is valid for any choice of these quantities. Note though that
the collision frequency leads to an additional constraint 
\eqref{eq:refquantities-collisions-constraint} between the
quantities mentioned above.
A further constraint exists with finite $\beta = \beta_\tref \beta_N$, where 
\begin{equation} 
\label{eq:betaref}
\beta_{ref} = {n_{\rm ref} T_{\rm ref} \over B_{\rm ref}^2 / 2 \mu_0} .
\end{equation}

There are a couple of normalised dimensionless quantities that are independent of the species. These are the
normalised Larmor radius
\begin{equation} 
\rho_* = \rho_{\rm ref} / R_{\rm ref} \qquad \leadsto  \rho_* = \frac{\rho_\tref }{R_\tref } = \frac{m_\tref v_{thref}}{eB_\tref R_\tref } .
\label{eq:normalise-rho}
\end{equation}
and the fields 
\begin{align} 
\phi = \rho_* {T_{\rm ref} \over e} \phi_N ,
\qquad 
A_\parallel = B_{\rm ref} R_{\rm ref} \rho_*^2 A_{\parallel N} ,
\qquad 
B_{1\parallel} = B_{\rm ref} \rho_* B_{1\parallel N}
\qquad
\chi = \rho_* {T_{\rm ref} \over e} \chi_N ,
\qquad
\Phi = {T_{\rm ref} \over e} \Phi_N ,
\label{eq:phi-norm}
\end{align} 
Here the index $N$ refers to a normalised quantity. 
Note that factors of $\rho_*$ have been added in the definitions of the normalised perturbed fields. 
These factors are chosen such that the normalised quantities are of the order 1.

Also the parameters time $t$, magnetic field $B$, reference frame angular rotation frequency $\Omega$, and the major radius $R$, are made dimensionless using the reference values 
\begin{align}
&t = { R_{\rm ref} } t_N / v_{\rm thref} , \qquad &B = B_{\rm ref} B_N , \cr 
\noalign{\vskip 0.2 truecm} 
&\Omega = {v_{\rm thref} } \Omega_N / R_{\rm ref},  \qquad &R = R_{\rm ref} R_N .
\label{eq:normalise-parameters}
\end{align}

On the other hand, there are quantities which describe species.
Here, the reference values are used to define, for each species, a dimensionless mass $m_R$, a dimensionless thermal 
velocity $v_R$, a dimensionless density $n_R$, a dimensionless temperature $T_R$, background flow velocity $R\omega$ and a dimensionless centrifugal energy $\cfenn$.
The index $R$ refers to a normalised quantity with species dependence. 
\begin{align} 
m &= m_R m_{\rm ref}
\\
v_{\rm th} &=  v_R v_{\rm thref}
\\
n_{R_0} &= n_R n_{\rm ref} 
\\
T &= T_R T_{\rm ref} \qquad\leadsto T_R = m_Rv_R^2
\label{eq:normalise-T}
\\
\omega &= \frac{v_{\rm thref} }{R_\tref} \omega_N
\end{align}
Since it is not as obvious as the above, we elaborate on the centrifugal potential $\cfen$ with more detail:
\begin{align}
\\
 \cfen &= Ze \Phi - \frac{1}{2}m\Omega^2(R^2 - R_0^2)
\nonumber\\ 
& = Ze\frac{T_\tref }{e}\Phi_N - \frac{1}{2}m_\tref m_R\frac{v_{thref}^2}{R_\tref ^2}\Omega_N^2R_\tref ^2(R_N^2-R_{0N}^2) \nonumber\\
 & =  \left(Z\Phi_N - m_R\Omega_N^2(R_N^2 - R_{0N}^2)\right) T_\tref\nonumber\\
 \cfen &=\cfenn T_{\rm ref}
\label{eq:normalise-cfen}
\end{align} 
The GKW code defines a ``centrifugal energy'' array \texttt{cfen} which contains a temperature factor
\begin{align}
  \mathtt{cfen} = \frac{\mathcal{E}_R}{T_R}
\label{eq:cfen-vs-ER}
\end{align}
such that
\begin{align}
  \mathcal{E}_\Omega = \mathtt{cfen}\cdot T_RT_\tref
\end{align}

The velocity space coordinates are made dimensionless with the help of 
the thermal velocity $v_{th}$ of the respective species 
\begin{align}
  \label{eq:normalisation-velocityspace}
  v_\parallel &= v_{th}v_{\parallel N} = v_{thref} v_R v_{\parallel N} \\
\mu &= \frac{1}{2}\frac{mv_\perp^2}{B}  \nonumber\\
&= \frac{1}{2}\frac{m_\tref m_R(v_{thref}v_R)^2v_{\perp N}^2}{B_\tref B_N} \nonumber\\
&= \frac{m_\tref m_Rv_{thref}^2v_R^2}{B_\tref } \frac{1}{2}\frac{v_{\perp N}^2}{B_N}
= \frac{mv_{th}^2}{B_\tref } \mu_N 
= \frac{2T_\tref T_R}{B_\tref } \mu_N 
\label{eq:normalisation-velocityspace-mu}
\\
J_v &= \frac{2\pi B}{m} = \frac{B_\tref}{m} 2\pi B_N \quad\leadsto  J_{vN} = 2\pi B_N\\
\ud^3 v &= J_v \ud v_{\parallel} \ud\mu = v_{th}^3 \ J_{v,ref}J_{vN}\ \ud v_{\parallel N}\ \frac{m}{B_\tref }\ud\mu_N \quad\text{and with } J_{v,ref}\frac{m}{B_\tref } = 1 \nonumber\\
&= v_{th}^3 J_{vN}\ \ud v_{\parallel N}\ \ud\mu_N = v_{th}^3\ud^3v_N
\end{align}
It is worth to stress that the velocity is normalised by the thermal velocity of each of the species. 
This has the advantage that the grid in velocity space can always be defined relative to the thermal 
velocity. 
Quantities like the potential in \eqref{eq:phi-norm}, however, must be normalised by a reference temperature since it is 
species independent. 
A similar argument applies to the vector potential and the time.

Note that the magnetic moment $\mu$ is normalised such that the
particle mass $m_R$ (which differs greatly between electrons and ions)
is not part of $\mu_N$. This makes the grid used in the code be
actually a $mu_N/m$ grid, which has the same values for all
species. When a quantity is integrated over velocity space, this does
not alter the result.
FIXME The code not even uses \texttt{bn} for $B_N$, but puts a 1?|

It is then convenient to normalise the distribution functions according to 
\begin{equation} 
f = \rho_* {n_{R_0} \over v_{\rm th}^3} f_N ,
\qquad 
F_M = {n_{R_0} \over v_{\rm th}^3} F_{MN} .
\end{equation} 
The gradient of density and temperature are made dimensionless using the density and 
temperature of the species under consideration, but the plasma rotation 
is largely a bulk motion of all the species together and its gradient is normalised using  
the reference thermal velocity $v_{\rm thref}$
\begin{equation} 
\label{eq:gradients}
{1 \over L_{n,N}} = {R_{\rm ref} \over L_n} = - {1\over n_{R_0}} \pd{ n_{R_0} }{\psi} ,
\qquad 
{1 \over L_{T,N}} = {R_{\rm ref} \over L_T} = - {1 \over T} \pd{ T }{\psi} , 
\qquad 
u^\prime_N = - {R_{\rm ref} \over v_{\rm thref}} \pd{ \omega_\phi}{\psi} .
\end{equation}
The gradient length scales are normalised with $R_{\rm ref}$, but note that the same quantities are often written 
as $R/L_n$ and $R/L_T$ elsewhere in the literature.  Note also that the radial coordinate $\psi$ is normalised:
\begin{equation}
 \label{eq:psi-def}
 \psi = \frac{R_{\rm max}-R_{\rm min}}{2R_{\rm ref}},
\end{equation}
where $R_{\rm max}$ ($R_{\rm min}$) is the maximum (minimum) major radius of the flux 
surface. 
For circular surfaces, for instance, $\psi = r / R_{\rm ref} = \epsilon$, 
where $r$ is the minor radius.  For the $s-\alpha$ geometry, you can take $R_{\rm ref}$ to have any 
value you like, but it is simplest to always use the magnetic axis. For the Circ, Miller and Chease $R_{\rm ref}$ 
has a specific meaning defined by the geometry.
All gradients are normalised using the major radius $R_{\rm ref}$,
i.e.
\begin{equation} 
\nabla = {1\over R_{\rm ref}} \nabla_{N} .\qquad 
\end{equation} 
The poloidal flux $\Psi$ used in the derivation of the metric tensors is normalised according to
\begin{equation} 
 \Psi = R_{\rm ref}^2 B_{\rm ref}\Psi_N .
\end{equation}

The strength of the electromagnetic effects is determined by the plasma beta $\beta$.  
Although $\beta$ is dimensionless we define a different dimensionless beta $\beta_{ref}$ for the 
use in the code. This $\beta_{ref}$ is directly related to the reference values via \eqref{eq:betaref}.

Finally, the wave vectors introduced in the spectral representation arise from the perpendicular
gradient of a fluctuating quantity and will therefore be normalised to $\rho_{\rm ref}$:
\begin{align} 
k_\psi = \frac{k_{\psi N}}{\rho_{\rm ref}} \qquad 
k_\zeta = \frac{k_{\zeta N}}{\rho_*} \qquad 
\label{eq:kN}
\end{align}
In the later sections we will, of course, use only the normalised quantities. 
The index $N$ is then dropped for convenience.

\subsection{Normalisation for the Global Case}
\label{sec:normalisation-global-specialities}

\subsubsection*{Grid Temperature and Grid Density}
\label{sec:grid-temp-grid-dens}

In the local version of GKW the velocity coordinates are normalized to the thermal velocity 
of the considered species background. \index{Normalization} 
This has the advantage that the Maxwellian can be represented with the same accuracy for 
each of the species even if the temperatures of the species are very different. 
When the background temperature is a function of the radial coordinate, however, the thermal velocity is no 
longer a constant. 
Using the local thermal velocity for normalization would have the advantage that the Maxwell 
is represented with the same relative resolution at each radial point, but would make the 
normalization constant a function of radius, and would lead to cross terms in all radial 
derivatives. A choice has therefore been made to use a fixed temperature, often the temperature at 
one particular radial location, for the normalization. This has the advantage that it does not change the structure 
of the equations, but also has disadvantages. In the global case the grid must be, both set up 
such that it is big enough to represent the maximum temperature ($T_{\rm max}$) in the domain, as well as 
have sufficient resolution to resolve the distribution function of the smallest temprature
($T_{\rm min}$). Because the velocity scales as the square root of the temperature this increases 
the computational cost by a factor $T_{\rm max} / T_{\rm min}$ over a local simulation.

The temperature used for the normalization of the velocity coordinates
is the 'grid temperature' $T_G$, instead of $T_\tref$ directly. 
$T_G$ is not dimensionless but has physical units, just like $T_\tref$. $T_G$ is
not a function of the radius but, unlike $T_{\rm ref}$, is different
for each species.  The corresponding thermal velocity used for normalization is then $w$.
\begin{equation}
T_G = \frac{1}{2} m w^2 \qquad w = \sqrt{\frac{2 T_G}{m} }
\end{equation}
Where $m$ is the particle mass of the respective species. 
That is, while in the local case the velocity is normalised to the same $v_{th\tref}$ (related to a $T_\tref$)
for all species, for the global case $v_{th\tref}$ is replaced by a reference velocity $w$ (related to 
a new temperature, the species grid temperature $T_G$).

Species background temperature $T$ and species grid temperature $T_G$ are normalized with $T_{\rm ref}$
\begin{equation}
T_G = T_{GR} T_\tref \qquad T = T_R T_\tref
\end{equation}
and since $T_{\rm ref}$ is connected to the reference thermal velocity $v_{th\tref}$ via \eqref{eq:normalise-Tref} we have 
\begin{align}
w &= w_R v_{thref} \\
w_R &= \frac{w}{v_{thref}} 
=\frac{1}{v_{thref}} \sqrt{\frac{2T_G}{m}}
\nonumber\\
&=\frac{1}{v_{thref}} \sqrt{\frac{2T_{GR}T_\tref}{m}}
\\
&=\sqrt{\frac{ m_\tref}{2T_\tref} } \sqrt{\frac{2T_{GR}T_\tref}{ m_Rm_\tref} } 
= \sqrt{\frac{T_{GR}}{ m_R} } \qquad \leadsto T_{GR} = m_Rw_R^2
\end{align}
By normalising the species thermal velocity to $w$ we have
\begin{align}
  v_{th} &= v_R w\\
&= v_Rw_Rv_{th\tref} 
= \sqrt{\frac{T_R}{m_R}}\sqrt{\frac{T_{GR}}{m_R}}v_{th\tref}
= \frac{\sqrt{T_RT_{GR}}}{m_R}v_{th\tref}
\end{align}
The velocity space coordinates for the respective species are
\begin{align}
v_\parallel = v_{\parallel N} v_R w  \qquad \mu = \frac{ mw^2}{B_{\rm ref}} \mu_N 
\end{align}

To see some relations this implies, consider the normalisation of the
species temperature $T$:
\begin{align}
  T &= \frac{1}{2}m v_{th}^2 \\
&= \frac{1}{2}m_Rm_\tref v_R^2w_R^2v_{th\tref}^2  
 = v_R^2 m_Rw_R^2 \frac{1}{2}m_\tref v_{th\tref}^2  
\\
 &= \frac{T_R}{m_R} T_{GR} T_\tref
\end{align}
Another relationship is
\begin{align}
  \frac{T_R}{T_{GR}} &= \frac{T_RT_\tref}{T_{GR}T_\tref} = \frac{T}{T_G}
\\
&= \frac{\frac{1}{2}mv_{th}^2}{\frac{1}{2}mw_{th}^2}
= \frac{v_R^2w_R^2v_{th\tref}^2}{w_R^2v_{th\tref}^2}
= v_R^2
\label{eq:normalise-vR-Tratio}
\end{align}

Note that only the velocity space coordinates are normalized with the grid thermal velocity $w$. 
Species independent quantities and parameters, like time, rotation velocity and normalized Larmor radius 
continue to be normalized using 
$v_{\rm thref}$ , in the global case, according to \eqref{eq:normalise-rho} and \eqref{eq:normalise-parameters}.
Accordingly, the fields are also normalized with $T_{\rm ref}$ like in \eqref{eq:phi-norm}.

Besides the temperature, a similar normalization is used for the density. Each of the species 
is normalized to the 'grid density' $n_G$ which is not a function of the radius but, unlike $n_\tref$ is different for each species. 
\begin{equation}
n_G = n_{\rm ref}n_{GR}  \qquad n_{R_0} = n_{R}n_{\rm ref}
\end{equation}

The the potential $\cfen$ contains the frame rotation frequency $\Omega$ which we would like to keep a konstant by all means (cf. section \ref{sec:global-plasma-rotation}). Likewise, $\omega$ continues to be normalised like in \eqref{eq:normalise-parameters}.
\begin{align}
\omega &= \frac{v_{th\tref}}{R_\tref} \omega_N 
= \frac{w}{w} \frac{v_{th\tref}}{R_\tref} \omega_N 
= \frac{w}{w_Rv_{th\tref}} \frac{v_{th\tref}}{R_\tref} \omega_N 
= w \frac{1}{w_R} \frac{1}{R_\tref} \omega_N 
= \frac{w}{R_\tref}  \sqrt\frac{m_R}{T_{GR}} \omega_N 
\end{align}

\emph{Please note, the implementation of plasma rotation $\Omega$ in
  the global case is not entirely consistent and using the plasma
  rotation $\Omega$ in the global case is strongly discouraged. Please
  see also the comments on this in
  section~\ref{sec:global-normalisation}.}

% \begin{equation}
%  {\cal E}_\Omega = {\cal E}_{\Omega N} T_{R} 
% \label{eq:normalise-cfen-global}
% \end{equation}
% FIXME This text used to list $\Omega = v_{\rm thref} \Omega_N / R_\tref$ and ${\cal E}_\Omega = {\cal E}_{\Omega N} T_{G}$,i.e. the frame rotation is species 
% independent, just like for fluxtube. If we then normalise $\cfen$ with $T_G$, this introduces a 
% species-dependence in the normalisation of $\Omega$ inside $\cfen$ while there is none for $\Omega$ 
% elsewhere, which looks inconsistent to SRG. However, this seems to be used in the Maxwellian below.|

% FIXME $\beta_\tref$ depends on $T_\tref n_\tref$. Is this replaced by a kind of 'grid beta' or not, and does one need an argument for it?|

It is to be noted that the 'old' flux-tube normalization is included in the global normalization above if one sets $T_G \equiv T_\tref$ and $n_G \equiv n_\tref$.

\paragraph{Example: Normalisation of the Maxwellian Background Distribution} 
\label{sec:global-normalisation}
In the global case, there are a couple of difficulties concerning the normalisation, so that the normalisation scheme is not totally identical to that of the local (fluxtube) case.
This is described in detail in section~\ref{sec:normalisation-global-specialities}.

For the global case, with its adapted normalization, the Maxwellian from \eqref{eq:maxwell}

\begin{align} 
F_{M} &= {n_{R_0} \over \pi^{3/2} v_{\rm th}^3 } \exp 
\biggl [ - {(v_\parallel-(R B_t/B)\omega_\phi )^2 + 2 \mu B / m \over v_{\rm th}^2} - \cfen/T \biggr ]
\end{align}
becomes \index{Maxwell distribution} 
\begin{align}
  F_{M,\mathrm{global}} 
 &=\frac{n_Rn_\tref }{ \pi^{3/2} (v_Rw)^{3} } \exp \left [ - \frac{ 
\left(v_{\parallel N}v_Rw - R_\tref R_N B_{tN} \frac{w}{R_\tref}\sqrt\frac{m_R}{T_{GR}}\omega_{\phi N} / B_N\right)^2 + 2 \frac{mw^2}{B_\tref}\mu_N B_\tref B_N\frac{1}{m} 
}{ (v_R w)^2}
- 
\frac{T_\tref\mathcal{E}_R}{T_\tref T_R}  \right ] 
\nonumber\\
 &=\frac{n_\tref}{w^3} \frac{n_R }{ \pi^{3/2} v_R^3 } \exp \left [ - \frac{ 
\left(v_{\parallel N}v_R - \sqrt\frac{m_R}{T_{GR}}R_N B_{tN} \omega_{\phi N} / B_N\right)^2 + 2 \mu_N B_N
}{ v_R ^2}
- 
\frac{\mathcal{E}_R}{T_R}  \right ] 
\\
F_{M,\mathrm{global}}&=\frac{n_\tref}{w^3} \frac{n_R }{ \pi^{3/2} v_R^3 } \exp \left [ - \frac{ 
\left(v_{\parallel N}v_R - \sqrt\frac{m_R}{T_{GR}}R_N B_{tN} \omega_{\phi N} / B_N\right)^2 + 2 \mu_N B_N
}{ v_R ^2}
- 
\frac{\mathcal{E}_R}{T_R}  \right ] 
\end{align}
We replace $v_R$ with \eqref{eq:normalise-vR-Tratio}.
\begin{align}
  F_{M,\mathrm{global}}&=\frac{n_G}{w^3}\frac{n_R }{n_{GR} \pi^{3/2} (T_R/T_{GR})^{3/2} } \exp \left [ - \frac{ 
\left(v_{\parallel N}v_R - \sqrt\frac{m_R}{T_{GR}}R_N B_{tN} \omega_{\phi N} / B_N\right)^2 + 2 \mu_N B_N
}{ T_R/T_{GR}}
- 
\frac{\mathcal{E}_R}{T_R}  \right ] 
\label{eq:maxwellian-global}
\end{align}
In order to draw a grid density $n_G$ out of the distribution, we have expanded the prefactor with $n_{GR}$ here.
In the Maxwellian \eqref{eq:maxwellian-global} above, the quantities
$n_R$, $T_R$, $R_N$, $\omega_{\phi N}$, $\mathcal{E}_R$ and
$B_{tN},B_N$ are functions of radius.

In this sense, the distribution functions are normalized as follows 
\begin{align} 
F_M &= 
\frac{n_G}{ w^3} F_{MN} \\
f &= \rho_* \frac{n_G}{w^3} f_N
\end{align}

Due to the implementation of \texttt{cfen} from \eqref{eq:cfen-vs-ER}, the code has
\begin{align}
  F_{M,\mathrm{global}}&=\frac{n_G}{w^3}\frac{n_R }{n_{GR} \pi^{3/2} (T_R/T_{GR})^{3/2} } \exp \left [ - \frac{ 
\left(v_{\parallel N}v_R - \sqrt\frac{m_R}{T_{GR}}R_N B_{tN} \omega_{\phi N} / B_N\right)^2 + 2 \mu_N B_N
}{ T_R/T_{GR}}
- 
\mathtt{cfen} \right ] 
\end{align}

(This is the previous version of the global normalisation of the
Maxwellian given here, but SRG thinks there is something wrong with it:
\begin{align}
F_{MN,orig}&= \frac{n_{GN}}{1} \frac{n_N }{ \pi^{3/2} (T_N / T_{GN})^{3/2} } \exp \left [ - \frac{ 
(v_{\parallel N} - \sqrt{m_N/T_{GN}} R_N B_{tN} \omega_{\phi N} / B_N)^2 + 2 \mu_N B_N + 
{\cal E}_{\Omega N} }{ T_N / T_{GN}} \right ] 
\end{align}
)

Just like \eqref{eq:maxwell}, the Maxwellian here in
\eqref{eq:maxwellian-global} integrates to the background density and
background temperature. The exp. factor modifies the momenta because
the distribution has a dependence on the poloidal angle $\theta$ if the frame rotates.
\begin{align}
\int F_M \, {\rm d}^3 {\bf v} &= n_{R_0} \exp\left(-\frac{\mathcal{E}_\Omega (\theta)}{T}\right) & \int F_{MN} \, {\rm d}^3 {\bf v} &= \frac{n_{R}}{n_{GR}} \exp\left(-\frac{\mathcal{E}_\Omega}{T}\right)
\\
\int \frac{1}{ 2} m v^2 F_M \, {\rm d}^3 {\bf v} &= \frac{3 }{ 2} T \exp\left(-\frac{\mathcal{E}_R(\theta)}{T_R}\right)
&
\int \frac{1}{ 2} m v^2 F_{MN} \, {\rm d}^3 {\bf v} &= \frac{3 }{ 2} T_R \exp\left(-\frac{\mathcal{E}_R}{T_R}\right)
\end{align}

\section{Geometry}
\label{geometry}
GKW is formulated in field aligned Hamada coordinates \cite{HAM62}, i.e. the contravariant
components of both the poloidal $B^s$ and toroidal magnetic field $B^\gamma$
are flux functions. The safety factor is then determined by the ratio $q = B^\gamma / B^s$.  
Starting from an orthogonal coordinate system $(\psi,\theta,\varphi)$, where $\psi$ is the radial
coordinate (i.e. ${\bf B}\cdot \nabla \psi = 0$), $\theta$ is the poloidal angle (upward 
on the outboard midplane), and $\varphi$ is the toroidal angle (clockwise when viewed from 
above), and using the transformations 
\begin{equation} 
s = s(\psi,\theta), \qquad \gamma = \gamma(\psi,\theta,\varphi),
\label{transformations}
\end{equation} 
one can derive \cite{SCO98,SCO01} (see Appendix~\ref{sec:hamada} for more details):
\begin{equation} 
s(\theta,\psi) = {\int_0^\theta {{\rm d} \theta^\prime \over {\bf B} \cdot \nabla \theta^\prime} 
\biggl /  \oint {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime}} ,
\end{equation}
\begin{equation} 
\gamma = {\varphi \over 2\pi} + s_{\rm B}{R B_t \over 2\pi} \int_0^\theta {{\rm d} \theta^\prime \over {\bf B}
\cdot \nabla \theta^\prime} \biggl [ \biggl \{ {1\over R^2} \biggr \} - {1\over R^2} 
\biggr ] ,
\end{equation} 
with 
\begin{equation} 
B^s =  {1 \biggl /  \oint {{\rm d}\theta^\prime \over {\bf B}\cdot \nabla \theta^\prime}} \qquad 
B^\gamma = s_{\rm B}{R B_t\over 2 \pi } \biggl \{ {1\over R^2} \biggr \}. 
\end{equation}  
In the equations above, the magnetic field is decomposed as
\begin{equation}
{\bf B} = s_{\rm B}RB_t\nabla \varphi + s_{\rm j}\nabla\varphi\times\nabla\Psi ,
\end{equation}
where $B_t>0$ is the toroidal component of the magnetic field, $s_{\rm B}=\pm 1$ and $s_{\rm j}=\pm 1$ the sign of 
the magnetic field and plasma current (positive in the direction of $\nabla \varphi$) and $\Psi$ the normalised poloidal 
flux ($\nabla \Psi$ points from the magnetic axis to the plasma edge).
The brackets $\{ \}$ denote the flux surface average, which in the transformed coordinates 
can be written (more details in \ref{sec.fsa}) as 
\begin{equation} 
\{ g \} = \oint g \,{\rm d}s. 
\end{equation} 
The normalising constants have been chosen such that the domain [-1/2,1/2] in $s$ corresponds 
to one poloidal turn and the domain [-1/2,1/2] in $\gamma$ corresponds to one toroidal turn.
Note that the transformations of Eq.~(\ref{transformations}) leave the angle $\gamma$ to be 
an ignorable coordinate, since any quantity that is independent of $\varphi$ will also be 
independent of $\gamma$ after the transformation.

The coordinates given above are transformed using a simple linear transformation 
\begin{equation} 
\zeta =  q s - \gamma \qquad {\rm where} \qquad q = {B^\gamma \over B^s} 
\end{equation}
to make them field aligned, i.e. ${\bf B} \cdot \nabla = B^s \pd{}{s}$ as $B^\zeta = 0$ and $B^\psi = 0$. 
Note that the coordinate transformation above flips the sign of the toroidal angle. 
The right handed coordinate system can therefore be defined as $(\psi,\zeta,s)$. 
The Jacobian of the new coordinate system can be expressed in terms of the original 
Jacobian through 
\begin{equation} 
J_{\psi  \zeta s} = 2\pi {\bf B} \cdot \nabla \theta \oint {{\rm d} \theta^\prime \over 
{{\bf B} \cdot \nabla \theta^\prime}} J_{\psi \theta \varphi} .
\end{equation} 
We note here that the coordinates $\psi$, $s$ and $\zeta$ are chosen dimensionless with $\psi$ being a normalised 'minor radius' of the flux surface, see Eq.~(\ref{eq:psi-def}).
We shall refer to $\zeta$ as the `bi-normal' coordinate since it is neither strictly toroidal or poloidal.

Close inspection of the gyrokinetic equation~(\ref{Vlasov}) shows that it contains parallel derivatives 
\begin{equation} 
{\bf b} \cdot \nabla  = {{\cal F}} \pd{}{s} \quad \rightarrow \quad 
{\cal F} = {B^s \over B} ,
\end{equation}
and perpendicular drifts, with the latter all involving a cross product with the magnetic field. 
It is then convenient to define the tensor 
\begin{equation} 
\label{efun}
{\cal E}^{\alpha \beta} = {1 \over 2 B} (\nabla x_\alpha \times \nabla x_\beta) \cdot {\bf b}.
\end{equation}

The convection connected with the velocity ${\bf v}_\chi$ can be directly expressed in 
this tensor times the derivatives of the potential towards $x_\alpha$ and the distribution towards
$x_\beta$.
\begin{align}
  v_\chi\cdot \nabla f &= \frac{(-\nabla\phi)\times \mathbf{b}}{\mathbf{B}} \cdot \nabla f\\
&= \frac{\mathbf{b}\times \frac{1}{\rho_*R_{ref}} 
\nabla_N\rho_*\frac{T_{ref}}{e}\phi_N}{B_{ref} B_N} 
\cdot \frac{1}{\rho_*R_{ref}}\nabla_N \rho_* \frac{n_{R0}}{v_{th}^3}f_N\\
&= \frac{\frac{1}{2}m_{ref}v_{th,ref}^2}{eR_{ref}B_{ref}}
\frac{\mathbf{b}\times\nabla_N\phi_N}{B_N}
\cdot \frac{1}{R_{ref}}\nabla_N \frac{n_{R0}}{v_{th}^3}f_N\\
&= \frac{\rho_{ref}}{R_{ref}}v_{th,ref}
\frac{1}{2}\frac{\mathbf{b}\times\nabla_N\phi_N}{B_N}
\cdot \frac{1}{R_{ref}}\nabla_N \frac{n_{R0}}{v_{th}^3}f_N
\qquad
\text{use } \nabla_N = \mathbf{e}^\alpha \partial_\alpha = \nabla_N x^\alpha\pd{}{ x^\alpha}
\\
&= \rho_*\frac{v_{th,ref}}{R_{ref}}\frac{n_{R0}}{v_{th}^3}
\frac{1}{2}\frac{\mathbf{b}\times\nabla_N x^\alpha \pd{}{ x^\alpha} \phi_N}{B_N}
\cdot \nabla_N x^\beta\pd{}{ x^\beta} f_N\\
&= \rho_*\frac{1}{t_{ref}}\frac{n_{R0}}{v_{th}^3}
\frac{1}{2}\frac{\nabla_N x^\alpha\times\nabla_N x^\beta}{B_N}
\cdot \mathbf{b}  \pd{\phi_N}{x^\alpha} \pd{f_N}{x^\beta} \\
&= \rho_*\frac{1}{t_{ref}}\frac{n_{R0}}{v_{th}^3}
\mathcal{E}^{\alpha\beta}
\pd{\phi_N}{x^\alpha} \pd{f_N}{x^\beta}
\end{align}

It is worth noting that the magnetic field can be expressed as 
\begin{equation}
\label{eq:efun-cst}
 \mathbf{B} = 2\pi s_{\rm j} \nabla \Psi \times \nabla \zeta,
\end{equation}
which immediately leads to 
\begin{equation} 
{\cal E}^{\psi \zeta} = \frac{s_{\rm j}}{4\pi}\pd{\psi}{\Psi},
\label{eq:efun-flux}
\end{equation} 
showing that ${\cal E}^{\psi \zeta}$ is constant on a flux surface. For convenience we define two more quantities 
\begin{equation} 
{\cal D}^\alpha = - 2 {\cal E}^{\alpha \beta} {1 \over B} \pd{ B}{x_\beta}, \qquad 
{\cal G} = {\cal F} \pd{ \ln B}{s} ,
\end{equation}
with ${\cal D}$ related to the grad B drift, and ${\cal G}$ to the trapping. Finally, the Coriolis drift and centrifugal drift will enter the 
equations through the one forms ${\cal H}$ and ${\cal I}$ respectively
\begin{equation} 
{\cal H}^\alpha = -\frac{s_{\rm B}}{B\Omega}{\bf \Omega}_\perp \cdot \nabla x_\alpha, 
\qquad 
{\cal I}^\alpha =  {s_{\rm B} \over 2 B} (\nabla x_\alpha \times \nabla R^2) \cdot {\bf b}
\end{equation}
and the centrifugal potential is calculated (see Sec. \ref{cfphi}) using the quantities
\begin{equation} 
{\cal J} = R^2 - R_0^2, 
\qquad 
{\cal K} = \pd{ {\cal J}}{\psi}\biggr|_s,
%=R\pd{ R}{\psi} + R_0\pd{ R_0}{\psi}
\qquad
{\cal L} = \pd{ (R_0^2)}{\psi}\biggr|_s.
\end{equation}
For transformations of the shearing rate we also define 
\begin{equation}
{\cal M} = \pdd{\Psi}{\psi} = {s_{\rm j} \over 4 \pi}\pd{}{\psi}\left({1 \over {\cal E}^{{\psi \zeta}}}\right).
\end{equation}
The tensor elements defined above all have similar magnitude. It must be realised though that these elements
are multiplied with the derivatives towards the respective coordinate. Since the derivatives along the field 
are much smaller than the derivatives in the perpendicular plane, the tensor elements containing the coordinate
$s$ are usually neglected (i.e. ${\cal D}^s = {\cal E}^{s \psi} = {\cal E}^{s \zeta} = {\cal H}^s = 0$), 
except in the case of those multiplied with the background potential $\Phi$, or when higher $\rho^*$ effects are included.

GKW uses the local limit. Central to this approximation is the small scales of the turbulent fluctuations. 
All background plasma parameters are assumed homogeneous across the simulation domain perpendicular to 
the field, i.e. there is no dependence of the parameters on the coordinates $\zeta$ and $\psi$.  The turbulence is homogeneous 
in the plane perpendicular to the magnetic field and periodic boundary conditions apply.  
Note that the dependence of equilibrium quantities on the 'parallel' coordinate $s$ are kept. 
All the tensors ${\cal D, E, F, G, H, I, J, K, M}$ as well as the metric tensor, are therefore, functions of the
parallel coordinate $s$ only.

\subsection{\texorpdfstring{s-$\alpha$}{s-alpha}}
The simplified `$s-\alpha$' equilibrium (with $\alpha=0$) is directly implemented in the code through pre-defined tensor elements. 
The latter model equilibrium is the simplest choice of the geometry of a tokamak with the flux surfaces being 
circular and having a small inverse aspect ratio $\epsilon = r / R \ll 1$, with $r$ being the 
minor radius of the surface and $R=R_{\rm ref}$.  
Only the lowest order in an $\epsilon$ expansion is kept in the definition of all the tensors. 
The coordinates can then be approximated as 
\begin{equation} 
\psi = \epsilon = {r \over R}, \qquad 
\zeta =  {s_{\rm B} s_{\rm j} \over 2 \pi} [|q| \theta - \varphi ], \qquad 
s = {\theta \over 2 \pi },  \qquad
\end{equation}
with the (normalised) gradients  
\begin{equation} 
\nabla \psi = {\bf e}_r, \qquad 
\nabla \zeta = {s_{\rm B} s_{\rm j} |q| \over 2 \pi \epsilon} [ {\bf e}_\theta + \hat s \theta {\bf e}_r ],  \qquad 
\nabla s = {1 \over 2 \pi \epsilon} {\bf e}_\theta , 
\end{equation} 
where $\hat s = ({r / q}) {\rm d} q / {\rm d} r$ is the magnetic shear. 
Note that the gradient of the toroidal angle is ordered small compared with the gradient of the poloidal
angle (times q) and is neglected in the gradient of $\zeta$. The contravariant components of the 
magnetic field are 
%FC: write ordering as equation: $\nabla \varphi \ll q \nabla \theta$?
\begin{equation} 
B^s = s_{\rm j}{B_p \over 2 \pi r} \qquad 
B^\zeta = -s_{\rm B}{B_t \over 2 \pi R} , 
\end{equation} 
where $B_p$ is the poloidal magnetic field. 
The gradient of the magnetic field strength is assumed to be in the direction of $\nabla R$ and the 
drift term is approximated as 
\begin{equation} 
{\bf B} \times \nabla B = -s_{\rm B}B^2 [ \cos \theta {\bf e}_\theta + \sin \theta {\bf e}_r ] .
\end{equation} 
Using these expressions one can evaluate the tensors to be 
\begin{equation} 
{\cal F} = {s_{\rm j}\over 2 \pi |q|},  \qquad  
{\cal D^\psi} = - s_{\rm B} \sin (2 \pi s), \qquad
%\end{equation} 
%\begin{equation} 
{\cal D}^\zeta = - {s_{\rm j}|q| \over 2 \pi \epsilon } [ \cos (2 \pi s) + 2 \pi \hat s s \sin (2 \pi s) ] ,
\end{equation}
\begin{equation}
{\cal I}^\psi = {\cal H}^\psi = {\cal D^\psi},
\qquad
{\cal I}^\zeta = {\cal H}^\zeta = {\cal D^\zeta},
\qquad
{\cal G} =  {s_{\rm j}\epsilon \over |q|} \sin ( 2 \pi s), 
\end{equation}
\begin{equation}
{\cal E}^{\psi \zeta} = s_{\rm j}{|q| \over 4 \pi \epsilon  } = - {\cal E}^{\zeta \psi},
\qquad
{\cal E}^{\psi s} =  {s_{\rm B} \over 4 \pi \epsilon},
\qquad
{\cal E}^{\zeta s} =  { s_{\rm j}|q| \over 4 \pi \epsilon^2} \hat s s,
\qquad
{\cal M} = {1 \over q}-{{\hat s} \over q}
\end{equation}
\begin{equation}
{\cal J} = 2 \epsilon \cos( 2 \pi s), 
\qquad
{\cal K} = 2 \cos(2 \pi s)- {\cal L}.
\end{equation}
where the two choices for $R_0$ (mag. axis $R$ and $R(s=0)$, see \ref{cfphi}) give
\begin{equation}
R_0=1 \Rightarrow {\cal L}=0, \qquad  
R_0=1+\epsilon \Rightarrow {\cal L}=2.
\end{equation}
respectively.

Finally the metric elements are
\begin{equation} 
g^{\zeta \zeta} = \biggl ( {q \over 2 \pi \epsilon} \biggr )^2 + \biggl ( {q \over\epsilon} \hat{s} s \biggr )^2 ,  \qquad 
g^{\zeta \psi} = {s_{\rm B}s_{\rm j}|q| \over \epsilon} \hat s s, \qquad 
g^{\psi \psi} = 1 .
\end{equation} 
All numerical results presented in this manual have been obtained with the geometry coefficients given above, except the benchmark
of the full geometry and Miller parametrisation treatment.

\subsection{Circular geometry}
The $s-\alpha$ model has been shown to be inconsistent in the $\epsilon$ ordering of the different terms involved \cite{Lapillonne} and for some nonlinear runs can also be numerically unstable for the zonal mode.  
In GKW the improved ad-hoc circular equilibrium model of Ref. \cite{Lapillonne} is also implemented. 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{circFS.eps}
\caption{Circular flux surface centred on $R=R\ind{ref}$, $Z=0$ and $(r,\theta,\varphi)$ coordinates.\label{circFS}}
\end{center}
\end{figure}

In this section, unless explicitly denoted by a $N$ subscript, the various quantities are not normalised. The starting assumption of the model is that the flux surfaces are circular and concentric
with the poloidal flux being a function of the radial coordinate only $\Psi=\Psi(r)$, see
figure~\ref{circFS}. This is a valid assumption in the case of small $\epsilon$ and small $\beta$. It simply means that the terms of order $\epsilon^2$ are neglected and that the Shafranov shift is
considered to be of order $\epsilon^2$.
For simplicity, the various terms are not expanded in $\epsilon$ in the following calculation, but one has to remember that the results are only valid up to order $\epsilon$.
The second assumption of the model is that the radial derivative of the poloidal flux is given by:
\begin{equation}
 \pd{\Psi}{r}=\frac{r B\ind{ref}}{\bar{q}}
\end{equation}
where $B\ind{ref}$ is the value of the magnetic field at $R\ind{ref}$ taken to be the major radius of the magnetic axis (i.e. the centre of the flux surfaces) and $\bar{q}>0$ is a parameter directly
related to the safety factor $q$. The flux surfaces are parametrised by $R=R\ind{ref}+r \cos \theta$ and $Z=r \sin\theta$ in the $(r,\theta,\varphi)$ coordinates system. The inverse aspect ratio is
defined as $\epsilon=r/R\ind{ref}$ and it might be worth reminding that:
\begin{equation}
 \nabla \epsilon = \frac{1}{R\ind{ref}} \mathbf{e_r}, \qquad \nabla \theta = \frac{1}{\epsilon R\ind{ref}}\mathbf{e_\theta}, \qquad \nabla \varphi = \frac{1}{R} \mathbf{e_\varphi}
\end{equation}
Another useful quantity is the Jacobian of the $(\Psi,\theta,\varphi)$ coordinates system which can be expressed as
\begin{equation}
 {J_{\Psi\theta\varphi}}^{-1}=\nabla\Psi\times\nabla\theta\cdot\nabla\varphi=\pd{\Psi}{r}{J_{r\theta\varphi}}^{-1}=\pd{\Psi}{r}\frac{1}{rR}=\frac{B\ind{ref}}{\bar{q}R}
\end{equation}

As usual, the magnetic field is decomposed into its toroidal and poloidal components:
\begin{equation}
 \mathbf{B} = s\ind{B}F(\Psi)\nabla \varphi + s\ind{j} \nabla \varphi \times \nabla \Psi
\end{equation}
One furthermore assumes that $F(\Psi)=R\ind{ref}B\ind{ref}$ and the magnetic field can then be written as
\begin{equation}
 \mathbf{B} = R\ind{ref}B\ind{ref}\left[ s\ind{B} \nabla \varphi + s\ind{j}\frac{1}{\bar{q}}\frac{\epsilon^2}{1+\epsilon\cos\theta}\nabla \theta\right]
\end{equation}
The definition of the safety factor is
\begin{equation}
 q=\frac{1}{2\pi}\int_0^{2\pi}\frac{\mathbf{B}\cdot\nabla\varphi}{\mathbf{B}\cdot\nabla\theta}\,\textrm{d}\theta
\end{equation}
which in the model considered here leads to
\begin{equation}
 q=\frac{1}{2\pi}\int_0^{2\pi}s\ind{B}s\ind{j}J_{\Psi\theta\varphi}\frac{R\ind{ref}B\ind{ref}}{R^2}\,\textrm{d}\theta=s\ind{B}s\ind{j}\frac{\bar{q}}{2\pi}\int_0^{2\pi}
\frac{1}{1+\epsilon\cos\theta}\,\textrm{d}\theta=s\ind{B}s\ind{j}\frac{\bar{q}}{\sqrt{1-\epsilon^2}}
\end{equation}
which makes explicit the relationship between $q$ and $\bar{q}$.

We now have all the ingredients to derive the field aligned coordinates $(\psi,\zeta,s)$. In GKW, $\psi$ is defined to be $\epsilon$ and we will use this notation in the following to avoid possible
confusion with the poloidal flux $\Psi$. The definition of the parallel coordinate $s$ is
\begin{equation}
 s = \int_0^\theta \frac{\textrm{d}\theta}{\mathbf{B}\cdot\nabla\theta} \,\,\Big{/} \int_0^{2\pi} \frac{\textrm{d}\theta}{\mathbf{B}\cdot\nabla\theta}
\end{equation}
So, we first calculate
\begin{equation}
 \int_0^\theta \frac{\textrm{d}\theta}{\mathbf{B}\cdot\nabla\theta} = \int_0^\theta s\ind{j}J_{\Psi\theta\varphi}\,\textrm{d}\theta = s\ind{j}\bar{q}\frac{R\ind{ref}}{B\ind{ref}} \int_0^\theta
[1+\epsilon\cos\theta]\,\textrm{d}\theta = s\ind{j}\bar{q}\frac{R\ind{ref}}{B\ind{ref}} [\theta + \epsilon \sin\theta],
\end{equation}
to get
\begin{equation}
 s=\frac{1}{2\pi}[\theta + \epsilon\sin\theta]
\label{eq:s:def}
\end{equation}
and
\begin{eqnarray}
 \pd{s}{\epsilon} &=& \frac{1}{2\pi}\sin\theta, \\
 \pd{s}{\theta} &=& \frac{1}{2\pi}[1+\epsilon\cos\theta],\\
 \pd{s}{\varphi} &=& 0.
\end{eqnarray}
The definition of the $\zeta$ coordinate is:
\begin{equation}
 \zeta = - \frac{\varphi}{2\pi} + s\ind{B}s\ind{j}\frac{F}{2\pi}\int_0^\theta \frac{J_{\Psi\theta\varphi}}{R^2}\,\textrm{d}\theta 
\end{equation}
which in the ad-hoc circular model gives
\begin{equation}
 \zeta = - \frac{\varphi}{2\pi} + s\ind{B}s\ind{j}\frac{\bar{q}}{2\pi}\int_0^\theta \frac{\textrm{d}\theta}{1+\epsilon\cos\theta}
\end{equation}
Noticing that
\begin{equation}
 \int_0^\theta \frac{\textrm{d}\theta}{1+\epsilon\cos\theta} = \frac{2}{\sqrt{1-\epsilon^2}}\arctan\left[ \sqrt{\frac{1-\epsilon}{1+\epsilon}}\tan \frac{\theta}{2}\right],
\end{equation}
one gets
\begin{equation}
 \zeta = - \frac{\varphi}{2\pi} + s\ind{B}s\ind{j}\frac{|q|}{\pi}\arctan\left[ \sqrt{\frac{1-\epsilon}{1+\epsilon}}\tan \frac{\theta}{2}\right]
 \label{eq:def:zeta}
\end{equation}
and
\begin{eqnarray}
 \pd{\zeta}{\epsilon} &=& s\ind{B}s\ind{j}\frac{|q|}{\epsilon\pi}\left[ \hat{s}\arctan\left[ \sqrt{\frac{1-\epsilon}{1+\epsilon}}\tan \frac{\theta}{2}\right] -
\frac{\epsilon\tan\frac{\theta}{2}}{\sqrt{1-\epsilon^2}\left(1+\epsilon+(1-\epsilon)\tan^2\frac{\theta}{2}\right)} \right], \\
 \pd{\zeta}{\theta} &=& s\ind{B}s\ind{j}\frac{|q|}{2\pi}\frac{\sqrt{1-\epsilon^2}}{1+\epsilon\cos\theta},\\
 \pd{\zeta}{\varphi} &=& -\frac{1}{2\pi}.
\end{eqnarray}
From the decomposition of $\nabla s$ and $\nabla \zeta$ as a function of $\nabla \epsilon$, $\nabla \theta$ and $\nabla\varphi$
\begin{eqnarray}
 \nabla s &=& \pd{s}{\epsilon} \nabla\epsilon +  \pd{s}{\theta} \nabla\theta +  \pd{s}{\varphi} \nabla\varphi \\
 \nabla \zeta &=& \pd{\zeta}{\epsilon} \nabla\epsilon +  \pd{\zeta}{\theta} \nabla\theta +  \pd{\zeta}{\varphi} \nabla\varphi ,
\end{eqnarray}
the calculation of the metric tensor is straightforward:
\begin{eqnarray}
 g^{\epsilon\epsilon}_N &= R\ind{ref}^2 g^{\epsilon\epsilon} &= 1,\\
 g^{\epsilon\zeta}_N &= R\ind{ref}^2 g^{\epsilon\zeta} &= \pd{\zeta}{\epsilon},\\
 g^{\epsilon s}_N &=R\ind{ref}^2 g^{\epsilon s} &= \frac{1}{2\pi}\sin\theta,\\
 g^{\zeta\zeta}_N &= R\ind{ref}^2 g^{\zeta\zeta} &= \left[\pd{\zeta}{\epsilon}\right]^2 + \left[\frac{1}{2\pi}\frac{1}{1+\epsilon\cos\theta} \right]^2
\left[1+\frac{q^2}{\epsilon^2}(1-\epsilon^2)\right]\\
 g^{\zeta s}_N &= R\ind{ref}^2 g^{\zeta s} &= \frac{\sin\theta}{2\pi}\pd{\zeta}{\epsilon} + s\ind{B}s\ind{j}\frac{|q|}{(2\pi\epsilon)^2}\sqrt{1-\epsilon^2}\\
 g^{s s}_N &= R\ind{ref}^2 g^{s s} &= \frac{1}{4\pi^2}\left[ 1 + \frac{2}{\epsilon}\cos\theta + \frac{1}{\epsilon^2}\right].
\end{eqnarray}
To calculate the other metric elements one needs the norm of the magnetic field
\begin{equation}
 B=\sqrt{\mathbf{B}\cdot\mathbf{B}}=\frac{B\ind{ref}}{1+\epsilon\cos\theta}\left[ 1+ \frac{1}{q^2}\frac{\epsilon^2}{1-\epsilon^2} \right]^{1/2}
\end{equation}
and its gradient
\begin{equation}
 \nabla B = \pd{B}{\epsilon} \nabla \epsilon + \pd{B}{\theta} \nabla \theta
\end{equation}
which involves
\begin{eqnarray}
 \pd{B}{\epsilon} &=&  \left[ -\frac{\cos\theta}{1+\epsilon\cos\theta} + \frac{\epsilon}{\epsilon^2+q^2(1-\epsilon^2)}(1-\hat{s}+\frac{\epsilon^2}{1-\epsilon^2})\right]B\\
 \pd{B}{\theta} &=& \frac{\epsilon\sin\theta}{1+\epsilon\cos\theta}B
\end{eqnarray}
From the definition 
\begin{equation}
 \mathcal{D}^\alpha_N = \frac{R\ind{ref}^2}{B^2B_N}\mathbf{B}\times\nabla B \cdot \nabla x^\alpha
\end{equation}
and some straightforward algebra (which has to be understood as ``I am by far too lazy to write down all the steps in \LaTeX'') we get the components of the tensor related to the curvature and
$\nabla B$ drift:
\begin{eqnarray}
 \mathcal{D}^{\epsilon}_N &=& -s\ind{B} \sin\theta \frac{1}{1+\frac{1}{q^2}\frac{\epsilon^2}{1-\epsilon^2}} \\
 \mathcal{D}^{\zeta}_N &=& \left[\frac{s\ind{j}}{2\pi} \frac{1}{B}\pd{B}{
\epsilon}\left(\frac{|q|\sqrt{1-\epsilon^2}}{\epsilon}+\frac{\epsilon}{|q|\sqrt{1-\epsilon^2}}\right)-s\ind{B}\pd{\zeta}{\epsilon}\sin\theta \right]
\frac{1}{1+\frac{1}{q^2}\frac{\epsilon^2}{1-\epsilon^2}}\\
 \mathcal{D}^{s}_N &=& \frac{s\ind{B}}{2\pi}\left[\frac{1}{\epsilon B}\pd{B}{\epsilon}(1+\epsilon\cos\theta)^2 -\sin^2\theta\right]
\frac{1}{1+\frac{1}{q^2}\frac{\epsilon^2}{1-\epsilon^2}}\\
\end{eqnarray}
The tensor related to the $\mathbf{E}\times\mathbf{B}$ drift can be obtained from the metric tensor elements:
\begin{equation}
 \mathcal{E}^{\alpha\beta}_N = s\ind{j}\frac{\pi R\ind{ref}^2}{BB_N}\pd{\Psi}{\epsilon}(g^{\alpha\epsilon} g^{\beta\zeta} - g^{\alpha\zeta} g^{\beta\epsilon})
\end{equation}
which for the ad-hoc model gives:
\begin{equation}
 \mathcal{E}^{\alpha\beta}_N = s\ind{j}\frac{\pi }{B_N^2}\frac{\epsilon}{|q|\sqrt{1-\epsilon^2}}(g^{\alpha\epsilon}_N g^{\beta\zeta}_N - g^{\alpha\zeta}_N g^{\beta\epsilon}_N)
\end{equation}
We the calculate the tensor related to the parallel derivative 
\begin{equation}
 \mathcal{F}_N = R\ind{ref} \frac{\mathbf{B}\cdot\nabla s}{B} = \frac{s\ind{j}}{2\pi |q| B_N}\frac{1}{\sqrt{1-\epsilon^2}}
\end{equation}
and the tensor related to the trapping
\begin{equation}
 \mathcal{G}_N = \frac{R\ind{ref}\mathbf{B}\cdot\nabla B}{B^2} = s\ind{j}\frac{\epsilon B_N \sin \theta}{|q|\sqrt{1-\epsilon^2}}\frac{1}{1+\frac{1}{q^2}\frac{\epsilon^2}{1-\epsilon^2}}.
\end{equation}
For the tensor related to the Coriolis drift
\begin{equation}
 \mathcal{H}^\alpha_N=\frac{R\ind{ref}}{B_N\Omega}\mathbf{\Omega}_\perp\cdot\nabla x^\alpha,
\end{equation}
we first decompose $\mathbf{\Omega}$ as
\begin{equation}
 \mathbf{\Omega} = -s\ind{B}\Omega\nabla Z = -s\ind{B}R\ind{ref}\Omega\left[ \sin\theta \nabla \epsilon + \epsilon \cos \theta \nabla \theta\right]
\end{equation}
to get
\begin{equation}
 \mathbf{\Omega}\cdot\mathbf{b} = -s\ind{B}s\ind{j}\Omega\frac{\epsilon\cos\theta}{|q|\sqrt{1-\epsilon^2}}\frac{R\ind{ref}B\ind{ref}}{RB} 
\end{equation}
and finally obtain
\begin{eqnarray}
 \mathcal{H}^\epsilon_N &=& -\frac{s\ind{B}}{B_N}\sin\theta \\
 \mathcal{H}^\zeta_N &=& -\frac{s\ind{B}}{B_N}\sin\theta\pd{\zeta}{\epsilon} -
                         \frac{s\ind{j}|q|\cos\theta}{2\pi\epsilon B_N}\frac{\sqrt{1-\epsilon^2}}{1+\epsilon\cos\theta}\\
 \mathcal{H}^s_N &=& -\frac{s\ind{B}}{2\pi B_N}\sin^2\theta + \frac{s\ind{B}}{2\pi B_N}(1+\epsilon\cos\theta)\left[ -\frac{\cos\theta}{\epsilon} + \frac{\epsilon\cos\theta}{q^2(1-\epsilon^2)}
\frac{1}{1+\frac{1}{q^2}\frac{\epsilon^2}{1-\epsilon^2}}\right].
\end{eqnarray}
Using
\begin{equation}
 \nabla R = R\ind{ref} \left[\cos\theta \nabla \epsilon -\epsilon\sin\theta\nabla\theta \right] 
\end{equation}
the tensor related to the centrifugal drift, can be expressed as
\begin{equation}
 \mathcal{I}^\alpha_N = s\ind{j}\frac{2\pi}{B_N^2}\frac{\epsilon(1+\epsilon\cos\theta)}{|q|\sqrt{1-\epsilon^2}}\left[g^{\epsilon\alpha}_N\left[\cos\theta\pd{\zeta}{\epsilon} -
s\ind{B}s\ind{j}\frac{|q|}{2\pi\epsilon}\frac{\sin\theta\sqrt{1-\epsilon^2}}{1+\epsilon\cos\theta}\right] -g^{\zeta\alpha}_N\cos\theta\right]
\end{equation}
We also have
\begin{equation}
 {\cal M} = {1 \over q \sqrt{1-\epsilon^2}}\left(1-{\epsilon \over 1-\epsilon^2}-{\hat s \over q}\right).
\end{equation}
Finally, the elements used to calculate the centrifugal potential are 
\begin{equation}
 \mathcal{J}_N= (1+\epsilon\cos\theta)^2-\frac{R_0^2}{R\ind{ref}^2} \qquad \mathcal{K}_N=\pd{\mathcal{J}_N}{
\epsilon} \biggr |_s =2(\cos\theta+\epsilon)-{\cal L}_N
\end{equation}
where the transformation between partial deriviatives at constant $\theta$ and $s$ gives
\begin{equation}
\frac{1}{R\ind{ref}} \pd{R}{\epsilon} \biggr |_s= \cos(\theta) + \frac{\epsilon \sin^2(\theta)}{1+\epsilon \cos(\theta)}
\end{equation}
which has been used to calculate ${\cal K}$.
The two choices for $R_0$ (magnetic axis value or $R(s=0)$) give
\begin{equation}
R_0=R\ind{ref} \Rightarrow {\cal L}_N=0, \quad  
R_0=(1+\epsilon)R\ind{ref} \Rightarrow {\cal L}_N =2(1+\epsilon)
\end{equation}
respectively.

% \subsection{Shifted circular surface geometry} 
% (NOT IMPLEMENTED, NOT CONSISTENT WITH OTHER SECTIONS - not normalised, definitions of Rref and r0 differ)\\
% 
% The shifted circular surface geometry is defined in the ($R,Z$) plane (using cylindrical coordinates
% $(R,\phi,z)$. 
% \begin{equation} 
% R = r \cos \theta + \Delta(r) + R_0 \qquad {\rm with} \qquad \Delta(0) = 0 
% \end{equation}
% \begin{equation} 
% Z = r \sin \theta
% \end{equation}
% Surfaces of constant $r$ are assumed to be flux surfaces. One can confusingly define two major 
% radii. One is the radius of the magnetic axis $R_0$, and one is the major radius of the centre 
% of the surface $R_c$ 
% \begin{equation} 
% R_c = (R(\theta =0) + R(\theta = \pi) )/2 = R_0 + \Delta(r) 
% \end{equation}
% Since the first one is a constant while $R_c$ depends on $r$, we will choose to normalise all 
% the radii with $R_0$, i.e. 
% \begin{equation} 
% R_{\rm ref} = R_0
% \end{equation}
% The normalised radial coordinate then is 
% \begin{equation} 
% \psi = r / R_0 = \epsilon
% \end{equation}
% and we will define a normalised major radius that gives the centre of the surface as 
% \begin{equation} 
% R_{cN} = 1 + {\Delta (r) \over R_0} 
% \end{equation}
% 
% From the equations above one can work out
% \begin{equation} 
%   \bbmatrix
%     \nabla r      \\
%     \nabla \theta
%   \ebmatrix
%   = {1\over r (1 + \Delta^\prime \cos \theta)} 
%   \bbmatrix
%     r \cos \theta & r \sin \theta \\
%     - \sin \theta & \Delta^\prime + \cos \theta
%    \ebmatrix
%    \bbmatrix
%      \nabla R \\
% 	 \nabla Z
%    \ebmatrix 
% \end{equation}
% where 
% \begin{equation}
%   \Delta^\prime = {{\rm d} \Delta \over {\rm d} r} 
% \end{equation}
% The poloidal magnetic field can be written as 
% \begin{equation} 
% {\bf B}_p = B_p {{\bf e}_\theta \over \vert {\bf e}_\theta \vert} \qquad 
% {\bf e}_\theta = \pd{{\bf x}}{\theta} =
% \bbmatrix
%   -r \sin \theta \\
%    r \cos \theta
% \ebmatrix  
% \end{equation} 
% From which it folows that 
% \begin{equation} 
% {\bf e}_\theta \cdot \nabla \theta = 1 \qquad {\bf B} \cdot \nabla \theta = B_p / r 
% \end{equation} 
% From flux conservation 
% \begin{equation} 
% R B_p {\rm d}l_\perp = {R B_p \over \vert \nabla r \vert } {\rm d} r = {\rm constant} 
% \end{equation}
% It follows that  
% \begin{equation} 
% {B_p \over B_p(\pi/2)} = {R_{cN}(\psi) \over (1 + \Delta^\prime(\psi) \cos \theta) (R_{cN}(\psi) + \psi \cos \theta)}
% \end{equation}
% The integral to obtain $s$ and $B^s$ can be easily performed 
% \begin{equation} 
% B^s = {B_p(\pi/2)\over 2 \pi ( 1 + \psi \Delta^\prime / 2 R_{Nc} )}
% \end{equation}
% \begin{equation} 
% s = {\theta \over 2 \pi} + {2 \psi \sin \theta + 2 \Delta^\prime R_{cn} \sin \theta + \Delta^\prime \psi 
% \cos \theta \sin \theta \over 4 \pi R_{cn} ( 1 + \Delta^\prime \psi / 2 R_{cn})}
% \end{equation}
% After some manipulations also the flux surface average of $1/R^2$ can be expressed as 
% \begin{equation} 
% \biggl \langle {1 \over R^2} \biggr \rangle = {1\over R_0^2} {1 \over R_{cN} + \psi \Delta^\prime / 2} 
% \biggl [ {1 - \Delta^\prime \over \sqrt{R_{cN}^2 - \psi^2} } + {\Delta^\prime \over \psi} \biggr ] 
% \end{equation}

\subsection{Miller parametrisation}
Miller parametrisation of the flux surface shape is a good compromise between the flexibility of the circ geometry and the precision of an equilibrium obtained from the solutin of the Grad-Shafranov equation.
In this section the Miller parametrisation first done in \cite{MIL98} is described. Then the derivation of the metric tensor in GKW coordinate system ($\psi,\zeta,s$) and the derivatives of the magnetic field are shown using the implementation method of \cite{CAN09}. \\
To avoid any confusion between the poloidal magnetic flux and the radial coordinate the latter will be called $r$ through the entire section.  For the miller geometry, $R_{\rm ref} = (R_{\rm max} + R_{\rm min})/2 $ which is the geometric axis of the flux surface in question.

\subsubsection{Parametrisation (Input List)}
The following parameters are used to describe a magnetic flux surface in (R,Z) plane : $\kappa$ (elongation), $\delta$ (triangularity), $\zeta$ (squareness), $R_{\rm mil}$, $Z_{\rm mil}$, and their radial derivatives $s_\kappa, s_\delta ($definition of$ \cite{MIL98}), s_\zeta, {\mathrm{d} R_{
\rm mil} \over \mathrm{d}r}$, ${\mathrm{d}Z_{\rm mil} \over \mathrm{d}r}$ defined such that:  
\begin{equation}
s_\kappa  =  {r \over \kappa}{\mathrm{d}\kappa \over \mathrm{d}r}, \qquad
s_\delta  =  r{{\mathrm{d}\delta \over \mathrm{d}r} \over \sqrt{1-\delta^2}}, \qquad
s_\zeta  = r{\mathrm{d}\zeta \over \mathrm{d}r}
\end{equation}
Parameters such as ${\mathrm{d} p \over \mathrm{d} \Psi}$ (pressure gradient), q (safety factor), $\hat s$ (magnetic shear) and $\epsilon$ (aspect ratio $= {r \over R_{
\rm mil}}$) are also used.
Flux surfaces geometry are then given by :
\begin{eqnarray}
R &=& R_{\rm mil} + r \cos{(\theta + \arcsin \delta \sin \theta)} \\
Z &=& Z_{\rm mil} + r \kappa \sin{(\theta + \zeta \sin{2\theta})}
\end{eqnarray}
and the radial derivatives of R and Z by :
\begin{eqnarray}
\pd{ R}{r} &=& {\mathrm{d}R_{\rm mil} \over \mathrm{d}r} + \cos{(\theta + \arcsin \delta \sin \theta)} - s_\delta \sin{\theta} \sin{(\theta + \arcsin \delta \sin \theta)} \\
\pd{ Z}{r} &=& {\mathrm{d}Z_{\rm mil} \over \mathrm{d}r} + \kappa \sin{(\theta + \zeta \sin{2\theta})} + \kappa s_\kappa \sin{(\theta + \zeta \sin{2\theta})} + \kappa s_\zeta \sin{(2\theta)} \cos{(\theta + \zeta \sin{2\theta})}
\end{eqnarray}
Examples how the parameters affect the cross section of a flux surface can be
seen in Fig.~\ref{fig:examplemillergeom}.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.49\textwidth]{millergeomkappa}
    \includegraphics[width=0.49\textwidth]{millergeomdelta}
    \includegraphics[width=0.49\textwidth]{millergeomzeta}
    \includegraphics[width=0.49\textwidth]{millergeomall}
    \caption[Examples Miller geometry]{\label{fig:examplemillergeom}
    Examples for the effects of $\kappa$, $\delta$ and $\zeta$ in Miller
    geometry. In all plots shown in this drawing, the dashed line is a circular flux surface.
    Top left: $\kappa=2.0$, $\delta=\zeta=0.0$. Top right: The
    full/dashed line has $\delta=\pm 0.5$, $\kappa=1.0$ and $\zeta=0.0$. Bottom
    left: The full/dashed line has $\zeta=\pm 0.5$, $\kappa=1$ $\delta=0.0$.
    Bottom right: The full line has $\kappa=2$, $\delta=0.5$ and $\zeta=0.5$.
    The dashed line is an example taken from an actual experiment
    with $\kappa=2.25$, $\delta=0.25$, $\zeta=-0.016$.}
  \end{center}
\end{figure}
The pressure gradient can be computed from different parameters through the input \textit{gradp\_type}.
The input list is also defined in the file \doc{input.dat.sample}.

The normalisation is made such that:
\begin{eqnarray}
R_N &=& 1 + \epsilon \cos (\theta + \arcsin \delta \sin \theta) \\
F_N &=& 1
\end{eqnarray}
We have then $R_{\rm ref} = R_{\rm mil}$ and $B_{\rm ref} = B_t(R_{\rm mil})$. \\
The normalisation for the plasma pressure gradient is given here using :
\begin{equation}
\left({\mathrm{d} p \over \mathrm{d} \Psi}\right)_N = {2 \mu_0 R_{\rm ref}^2\over B_{\rm ref}} {\mathrm{d} p \over \mathrm{d} \Psi}
\end{equation}
The normalised pressure gradient used in the code can be given by :
\begin{eqnarray}
\left({\mathrm{d} p \over \mathrm{d} \Psi}\right)_N &=& - \alpha {\left(2\pi^2\right) \over \left(\partial_{\Psi}V\right)_N } \left({V_N \over 2\pi ^2}\right)^{-1/2}\\
\left({\mathrm{d} p \over \mathrm{d} \Psi}\right)_N &=& {\alpha_{MHD}\over \epsilon ^2}\left({\mathrm{d} \Psi \over \mathrm{d} \epsilon}\right)_N \\
\left({\mathrm{d} p \over \mathrm{d} \Psi}\right)_N &=& {\beta ' \over 2 {\mathrm{d} \Psi_N \over \mathrm{d} \epsilon}}
\end{eqnarray}
with $\alpha$ taken from equation 42 in \cite{MIL98}, $\alpha_{MHD}$ from equation 141 \cite{CAN09} and $\beta '$ is defined in section \ref{Sec:betabetapr}. \\
For centrifugal effects on the pressure gradient go to equation \ref{eq:rota}. \\
The equations used below are not normalised however they are in the code. \\
In the coordinate system $(r,\theta,\varphi)$ where the magnetic flux surfaces are described, the elements of the metric tensor are :
\begin{eqnarray}
g_{rr} &=& \left(\pd{ R}{r}\right)^2 + \left(\pd{ Z}{r}\right)^2, \\
g_{r \theta} &=& \pd{ R}{r}\pd{ R}{\theta} + \pd{ Z}{r}\pd{ Z}{\theta} \\
g_{\theta \theta} &=& \left(\pd{ R}{\theta}\right)^2 + \left(\pd{ Z}{\theta}\right)^2, \\
g_{\varphi \varphi} &=& R^2
\end{eqnarray}
The contravariant metric tensor, used to calculate the contravariant metric tensor of the GKW basis, is deduced from the relation $g_{ij} \cdot g^{ij} = I$ with I the identity matrix.
\subsubsection{Mercier-Luc coordinate system}
An expansion of the poloidal flux $\Psi$ is needed. A coordinate system (introduced by Mercier-Luc) is defined such that the expansion is easy to use (orthogonal system ($\rho,l,\varphi$)).
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{mercier-luc.eps}
\caption{Mercier-Luc coordinate system.\label{merc-luc}...from \cite{CAN09}}
\end{center}
\end{figure}
In addition to the plasma shape, 
the gradient of the plasma pressure must also be specified, which in GKW can be done independently for the geometry, or by coupling to $\beta^\prime$.

\begin{eqnarray}
\Psi &=& \Psi_s + \rho \Psi_1 + \rho^2 \Psi_2 \\
R &=& R_s + \rho \cos{u} \\
Z &=& Z_s + \rho \sin{u}
\end{eqnarray}
with 
\begin{eqnarray}
\cos u &=& \pd{ Z_s}{l} \\
\sin u &=& - \pd{ R_s}{l}
\end{eqnarray}
where the subscript s means the value on the surface of consideration. \\
The metric elements are: 
\begin{eqnarray}
g_{\rho \rho} &=& 1 \\
g_{ll} &=& \left(1 + {\rho \over r_c} \right) \\
g_{\varphi \varphi} &=& R^2
\end{eqnarray}
The Jacobian is then:
\begin{equation}
J_\rho = R \left(1 + {\rho \over r_c}\right) 
\end{equation}
with $r_c$ being the curvature radius :
\begin{equation}
r_c = g_{\theta \theta}^{3/2} \left(\pd{ R}{\theta}\pdd{Z}{ \theta}-\pd{ Z}{\theta}\pdd{R}{\theta}\right)^{-1} 
\end{equation}
To switch from $(\rho,l,\varphi)$ to $(r,\theta,\varphi)$ the following relations are important (all quantities are taken on the flux surface)  :
\begin{eqnarray}
\pd{ \rho}{r} &=& \cos u \pd{ R}{r} + \sin u \pd{ Z}{r} \\
\pd{ l}{r} &=& \cos u \pd{ Z}{r} - \sin u \pd{ R}{r} \\
\pd{ \rho}{\theta} &=& 0 \\
\pd{ l}{\theta} &=& \sqrt{g_{\theta \theta}}
\end{eqnarray}
From the definition of the poloidal magnetic field:
\begin{equation}
B_p^2 = |\nabla \varphi|^2|\nabla \Psi|^2 = {\Psi_1 ^2 \over R_s ^2 }
\end{equation}
one finds
\begin{equation}
\Psi_1 = R_s B_{ps}
\end{equation}
$\Psi_2$ is obtained from the Grad-Shafranov equation for the expansion of $\Psi$:
\begin{equation}
R \nabla \cdot{\nabla \Psi \over R^2} = -\mu_0 R^2 p'(\Psi) - FF'(\Psi)
\end{equation}
with ' corresponding to the derivative toward the poloidal magnetic flux.
One can easily obtain: 
\begin{equation}
\Psi_2 = {1 \over 2}\left(B_p \cos u - {B_p R \over r_c} - \mu_0 R^2 p' - FF'\right)
\end{equation}
From this point we see that we have other parameters : $F$ (see the normalisation section), $F'$, $p'$, ${d \Psi \over dr}$ that we need to link to our first set of parameters.
The relation for $F'$ will be expressed later (it is not trivial).
\subsubsection{Relations for $p'$ and ${d\Psi \over dr}$}
${d\Psi \over dr}$ is obtained from the definition of $q$:
\begin{equation}
q = {1 \over 2 \pi} \int_0^{2\pi} {{{\bf B} \cdot \nabla \varphi \over {\bf B} \cdot \nabla \theta}\mathrm{d}\theta} 
\end{equation}
with 
\begin{equation}
{\bf B} = s_B F \nabla \varphi + s_J \nabla \varphi \times \nabla \Psi
\end{equation}
It follows : 
\begin{equation}
q = s_B s_J{F \over 2 \pi {\mathrm{d}\Psi \over \mathrm{d}r}} \int_0^{2\pi}{{J_r \over R^2}\mathrm{d}\theta}
\end{equation}
yields 
\begin{equation}
{\mathrm{d}\Psi \over \mathrm{d}r} = s_B s_J{F \over 2 \pi q} \int_0^{2\pi}{{J_r \over R^2}\mathrm{d}\theta}
\end{equation}
Numerical integrals are implemented using Simpson's method. \\
$p'$ is given from the relation in \cite{MIL98}:
\begin{equation}
\alpha = -{2\pd{ V}{\Psi} \over (2 \pi)^2}\left({V \over 2 \pi^2 R_{
\rm mil}}\right)^{1/2} \mu_0 p^\prime
\end{equation}
V is the volume defined by the magnetic flux surface :
\begin{equation}
V = 2 \pi S R_G
\end{equation}
with
\begin{eqnarray}
R_G &=& {\int_0^{2\pi} {R \sqrt{(\pd{ R}{\theta})^2 + (\pd{ Z}{\theta})^2} \mathrm{d}\theta} \over \int_0^{2\pi} {\sqrt{(\pd{ R}{\theta})^2 + (\pd{ Z}{\theta})^2} \mathrm{d}\theta}} \\
S &=& \int_0^{2\pi} {Z\pd{ R}{\theta}}
\end{eqnarray}
$\pd{ V}{\Psi}$ can easily be defined analytically by differentiating towards $r$ and using ${\mathrm{d}\Psi \over \mathrm{d}r}$ \\
\\

\subsubsection{Effect of toroidal rotation on the magnetic equilibrium}
Force balance equation for species a:

\begin{align}
m_an_a\left(\pd{}{t}+V_a\cdot \nabla \right)V_a = -\nabla p_a - \nabla \cdot \pi_a + e_a n_a\left(E+V_a\times B\right)+R_a
\end{align}

Viscosity $\nabla \cdot \pi_a$ and friction $R_a$ are being neglected. \\
A two species model with singly (for now) charged ions is considered (trace species can be added but won't affect the magnetic equilibrium: $m_tn_t \ll m_in_i$, i being the main ion species). Using $m_e \ll m_i$ and taking the dot product of force balance equation with $\mathbf b$, one obtains:

\begin{align}
n_e &= n_{R_0,e}\exp{\left({e \Phi \over T_e}\right)} \\
n_i &= n_{R_0,i}\exp{\left({-e \Phi \over T_i }+ {m_i\Omega^2(R^2 - R_0^2)\over 2T_i}\right)}
\end{align}

$T_i$ is supposed to be a flux function (might not be always the case \cite{THYA06}) \\
Using Quasineutrality and $T_e \sim T_i$:
\begin{align}
e\phi = {T_e \over T_e + T_i}{m_i \over T_i}{1 \over 2}\Omega^2\left(R^2-R_0^2\right)
\end{align}

The total pressure is then:
\begin{align}
p = Tn\exp{\left({m_i\Omega^2 \over 2 T}(R^2-R_0^2)\right)}
\end{align}

with $T = T_e + T_i$.\\
Furthermore the Grad-Shafranov is modified ($\Psi$ and $R^2$ are regarded as independant variables \cite{THYA06}):
\begin{equation}
R^2 \nabla \cdot \left({\nabla \Psi \over R^2}\right) = -\mu_0 R^2 \pd{ p \left(\Psi,R^2\right)}{\Psi} - FF'
\end{equation}

In normalised units one gets 
\begin{align} \label{eq:rota}
\left(\pd{ p}{\Psi}\right)_N &= {1 \over 2}\left[\beta' + \beta_{ref} n_Rm_{R,i}
\left(\left(R_N^2-R_0^2\right)\left(-2u_N' \Omega + {T_{R,e}{R\over L_{T_e}} + T_{R,i}{R \over L_{T_i}}\over \left(T_{R,e}+T_{R,i}\right)}\right)\right.\right. \nonumber \\
&\left.\left.\left. -R_0\pd{ R_0}{\epsilon}\Omega^2\right)\right)\right]\left({d \psi \over d \Psi}\right)_N\exp{\left({m_{R,i}\Omega_N^2 \over \left(T_{R,e}+T_{R,i}\right)}\left(R_N^2-R_0^2\right)\right)}
\end{align}

From this definition, it enters normalised Grad-Shafranov equation through:
\begin{equation}
\left(R^2 \nabla \cdot \left({\nabla \Psi \over R^2}\right)\right)_N = - R_N^2 \left(\pd{ p \left(\Psi,R^2\right)}{\Psi}\right)_N - F_NF_N'
\end{equation}

$\beta'$ is given in SPCGENERAL. $\beta_{ref}$ can be taken from SPCGENERAL but can also be given using the GEOM parameter \textit{beta\_rota\_miller\_type = 'geom'} and specifying the value in \textit{beta\_rota\_miller}. The other parameters are taken from SPECIES and ROTATION namelists.

\textbf{Modification of the curvature drift}

The magnetic curvature can be recast as:
\begin{eqnarray}
\kappa &=& - \mathbf b \times \left(\nabla \times {\mathbf B \over B} \right) \\
\kappa &=& - {1 \over B} \mathbf{b}\times \left(\nabla \times \mathbf{B}\right) - \mathbf{b} \times \left[\nabla\left({1 \over B}\right) \times \mathbf{B}\right] \\
\kappa &=& {\mu_0 J\times B \over B^2}+{\nabla_\bot \mathbf B \over B} \\
\end{eqnarray}

We will now only consider:
\begin{align}
\kappa_1 = \mu_0 {J \times B \over B^2}
\end{align}
since the term $\nabla_\bot B$ is not modified by rotation.\\

Without rotation 
\begin{align}
J\times B = \nabla p
\end{align}
which leads to the $\beta'$ correction.

For strong toroidal rotation one obtains for a two fluid model with singly charged ions:

\begin{align}
J\times B = -m_in_iR\Omega^2\nabla R + \pd{ p}{\Psi} {d \Psi \over d \psi}\nabla \psi + \pd{ p}{R^2} \nabla R^2
\end{align}

The curvature enters Vlasov equation through
\begin{align}
\left(\kappa_1 \times \nabla f\right)\cdot \mathbf{b} = {\mu_0 \over B^2}\left[\left(m_in_i\Omega ^2-2\pd{ p}{R^2}\right) \left(Bs_B\pd{ f}{x_\alpha}\mathcal{I}^\alpha\right)+2\pd{ p}{\Psi}{d \Psi \over d \psi} \pd{ f}{x_\beta} \mathcal{E}^{\psi \beta}\right]
\end{align}

\subsubsection{Determination of $\nabla s$}
We start from the expression of the magnetic field to express the contravariant component $B^s$ in the coordinate system $(\Psi,l,\varphi)$.
\begin{eqnarray}
B^s(\Psi) &=& B \cdot \nabla s \\
&=& B \cdot \nabla l \pd{ s}{l} \\
\pd{ s}{l} &=& B^s {1 \over B \cdot \nabla l}
\end{eqnarray}
Furthermore : 
\begin{equation}
\int_0^L {\pd{ s}{l'}\mathrm{d}l'} = 1 = B^s \int_0^L {{1 \over B \cdot \nabla l'}\mathrm{d}l'}
\end{equation}
with L being the arclength corresponding to $\theta = 2 \pi$.
Hence
\begin{equation}
B^s = {1 \over \int_0^L {{1 \over B \cdot \nabla l'}\mathrm{d}l'}}
\end{equation}
One gets the expression of $s$ in this basis:
\begin{equation}
s = B^s \int_0^l {{1 \over B \cdot \nabla l'}\mathrm{d}l'}
\end{equation}
Using the expansion of the magnetic flux:
\begin{eqnarray}
B \cdot \nabla l &=& s_B F\nabla \varphi \cdot \nabla l + s_j \nabla \varphi \times \left[(\Psi_1 + 2 \rho \Psi_2)\nabla \rho + \rho \pd{ \Psi_1}{l} \nabla l\right] \cdot \nabla l \\
B \cdot \nabla l &=& s_j {\Psi_1 + 2 \rho \Psi_2 \over J_\rho} \\
B \cdot \nabla l &=& s_j {\Psi_1 + 2 \rho \Psi_2 \over (R_s + \rho \cos u)(1 + {\rho \over r_c})}
\end{eqnarray}
Then 
\begin{equation}
{1 \over B \cdot \nabla l} = {s_j \over \Psi_1} \left[R_s + \rho \left(\cos u + {R_s \over r_c} - 2 R_s {\Psi2 \over \Psi1}\right)\right]
\end{equation}
One can write $\nabla s$ as
\begin{equation}
\nabla s = \pd{ s}{\Psi} \nabla \Psi + \pd{ s}{l} \nabla l
\end{equation}
We need then to determine $\pd{ s}{\Psi}$. Using the Mercier-Luc expansion of $\Psi$ one gets the second order equation in $\rho$:
\begin{equation}
\Psi_2 \rho^2 + \Psi_1 \rho + \Psi_s - \Psi = 0
\end{equation}
At $\Psi = \Psi_s$, $\rho = 0$. We have then $\rho$ as a function of $\Psi$
\begin{equation}
\rho = {-\Psi_1 + \sqrt{\Psi_1^2 - 4\Psi_2 (\Psi_s - \Psi)} \over 2 \Psi_2}
\end{equation}
For $\Psi = \Psi_s$, $\pd{ \rho}{\Psi} = {1 \over \Psi_1}$.
We define the two following integrals:
\begin{eqnarray}
A_1 (l) &=& \int_0^l {s_j{R_s \over \Psi_1}\mathrm{d}l'} \\
A_2 (l) &=& \int_0^l {{s_j \over \Psi_1}\pd{ \rho}{\Psi}\left(\cos u + {R_s \over r_c} - 2{R_s \Psi_2 \over \Psi_1}\right)\mathrm{d}l'}
\end{eqnarray}
At $\Psi = \Psi_s$ we then have:
\begin{equation}
\pd{ s}{\Psi} = {A_2 (l) \over A_1 (L)} - {A_2 (L) A_1 (l) \over A_1 (L)^2}
\end{equation}
$\nabla s$ is then given on the flux surface by :
\begin{equation}
\nabla s = \left(\pd{ s}{\Psi} {\mathrm{d}\Psi}{\mathrm{d}r} + \pd{ s}{l}\pd{ l}{r} \right) \nabla r + \pd{ s}{l}\pd{ l}{\theta} \nabla \theta
\end{equation}
\subsubsection{Determination of $\nabla \zeta$}
The magnetic field can be written as:
\begin{equation}
{\bf B} = s_J 2 \pi \nabla \Psi \times \nabla \zeta
\end{equation}
Using the following expansion for the zeta coordinate \cite{CAN09}:
\begin{equation}
\zeta = \zeta_s (l) + \rho \zeta_1 (l) - {\varphi \over 2 \pi} + \mathcal{O}(\rho^2)
\end{equation}
one gets
\begin{eqnarray}
F &=& F_s + \rho \Psi_1 F_s' + \mathcal{O}(\rho^2) \\
\nabla \Psi &=& \left(\Psi_1 + 2 \rho \Psi_2 \right) \nabla \rho + \rho \pd{ \Psi_1}{l} \nabla l + \mathcal{O}(\rho^2) \\
\nabla \zeta &=& \zeta_1 \nabla \rho + \left(\pd{ \zeta_s}{l} + \rho \pd{ \zeta_1}{l} \right) \nabla l + {\nabla \varphi \over 2 \pi} + \mathcal{O}(\rho^2)
\end{eqnarray}
From the two expressions of ${\bf B}$:
\begin{equation}
2 s_J \pi \nabla \Psi \times \nabla \zeta = s_B F \nabla \varphi + s_J \nabla \Psi \times \nabla \varphi
\end{equation}
dotting with $\nabla \varphi$ yields
\begin{equation}
{2 \pi s_J \over J_\rho} \left[(\Psi_1 + 2 \rho \Psi_2)\left(\pd{ \zeta_s}{l} + \rho \pd{ \zeta_1}{l} \right) - \rho \zeta_1 \pd{ \Psi_1}{l} \right] = s_B {F_s + \rho \Psi_1 F_s' \over \left(R_s + \rho \cos u \right)^2}
\end{equation}
At zeroth order in $\rho$:
\begin{equation}
\pd{ \zeta_s}{l} = s_J s_B {F_s \over 2 R_s \Psi_1 \pi}
\end{equation}
at first order :
\begin{equation}
2 s_J s_B \pi \left(\Psi_1 \pd{ \zeta_1}{l} - \zeta_1 \pd{ \Psi_1}{l} + 2 \Psi_2 \pd{ \zeta_s}{l}\right) = {F_s \over R_s} \left({1 \over r_c} - {\cos u \over R_s} \right) + {F_s' \Psi_1 \over R_s}
\end{equation}
hence
\begin{equation}
\Psi_1^2 s_J s_B \pd{}{l} \left({\zeta_1 \over \Psi_1} \right) = {1 \over 2 \pi} \left[{F_s \over R_s} \left({1 \over r_c} - {\cos u \over R_s} \right) + {F_s' \Psi_1 \over R_s} \right] - 2 \Psi_2 \pd{ \zeta_s}{l}
\end{equation}
$\zeta_s$ and $\zeta_1$ are then defined by:
\begin{eqnarray}
\zeta_s (l) &=& s_J s_B \int_0^l {{F_s \over 2 \pi R_s \Psi_1}\mathrm{d}l'} \\
\zeta_1 (l) &=& s_J s_B \Psi_1 \int_0^l {\left({1 \over 2 \pi} \left[{F_s \over R_s}\left({1 \over r_c} - {\cos u \over R_s} \right) + {F_s' \Psi_1 \over R_s} \right] - {\Psi_2 F_s \over \pi R_s \Psi_1} \right) {1 \over \Psi_1^2} \mathrm{d}l'}
\end{eqnarray}
replacing $\Psi_2$ by its value one can define the following integrals:
\begin{eqnarray}
D_0 (l) &=& \int_0^l {{F_s \over \pi \Psi_1^2 R_s} \left({1 \over r_c} - {\cos u \over R_s} \right) \mathrm{d}l'} \\
D_1 (l) &=& \int_0^l {\left({F_s^2 \over 2 \pi R_s \Psi_1^3} + {1 \over 2 \pi R_s \Psi_1}\right)\mathrm{d}l'} \\
D_2 (l) &=& \int_0^l {{\mu_0 R_s F_s \over 2 \pi \Psi_1^3} \mathrm{d}l'} 
\end{eqnarray}
$\zeta_1 (l)$ can be then written in the form:
\begin{equation}
\zeta_1 (l) = s_J s_B \Psi_1 \left(D_0(l) + D_1(l)F' + D_2(l)p' \right)
\end{equation}
Warning: If the plasma rotation is taken into account through $p'$, it will depend on $R$ and therefore on $l$. $p'$ is then in $D_2 (l)$ \\
$F'$ has not yet been linked to the input parameters. Once we have $F'$ we see that $\pd{ \zeta}{l} = \zeta_s$ and $\pd{ \zeta}{\rho} = \zeta_1$. \\
$\nabla \zeta$ is then given on the flux surface by:
\begin{equation}
\nabla \zeta = \left(\zeta_s \pd{ l}{r} + \zeta_1 \pd{ \rho}{r} \right) \nabla r + \zeta_s \pd{ l}{\theta} \nabla \theta
\end{equation}
\subsubsection{Relation for F'}
Using the expression of q and the specific boundary condition \cite{CAN09}: 
\begin{equation}
\zeta(\Psi,0,\varphi) = FIX ME
\end{equation}
one gets
\begin{equation}
\zeta(\Psi,2 \pi,\varphi) = -2\pi q(\Psi)
\end{equation}
using the expansion of $\Psi$ we have the following equation:
\begin{equation}
-2 \pi (q_s + q_s' \rho \Psi_1) + \mathcal{O}(\rho^2) = \zeta_s (L) + \rho \zeta_1 (L) + \mathcal{O}(\rho^2)  
\end{equation}
We then have:
\begin{equation}
F_s' = {2 \pi \hat{s}{q \over r} - D_0 (L) - D_2 (L) p' \over D_1(L)}
\end{equation}
\subsubsection{Contravariant metric tensor of $(r,\zeta,s)$}
We can then deduce the contravariant metric tensor in the GKW basis.
\begin{eqnarray} 
g^{r \zeta} &=& g^{\zeta r} = \pd{ \zeta}{r}g^{rr} + \pd{ \zeta}{\theta}g^{r \theta} \\
g^{r s} &=& g^{s r} = \pd{ s}{r}g^{rr} + \pd{ s}{\theta}g^{r \theta} \\
g^{\zeta \zeta} &=& \left(\pd{ \zeta}{r} \right)^2 g^{rr} + \left(\pd{ \zeta}{\theta} \right)^2 g^{\theta \theta} + 2 \pd{ \zeta}{r} \pd{ \zeta}{\theta} g^{r \theta} + {1 \over 4 \pi^2} g^{\varphi \varphi} \\
g^{s s} &=& \left(\pd{ s}{r} \right)^2 g^{rr} + \left(\pd{ s}{\theta} \right)^2 g^{\theta \theta} + 2 \pd{ s}{r} \pd{ s}{\theta} g^{r \theta} \\
g^{\zeta s} &=& g^{s \zeta} = \left(\pd{ \zeta}{r}\pd{ s}{r} \right)g^{rr} + \left(\pd{ \zeta}{\theta}\pd{ s}{\theta} \right) g^{\theta \theta} + \left(\pd{ \zeta}{r}\pd{ s}{\theta} + \pd{ \zeta}{\theta}\pd{ s}{r} \right) g^{r \theta} 
\end{eqnarray}

\subsubsection{Derivatives of the magnetic field strength}
To obtain the derivatives of the magnetic field strength B we start from \cite{CAN09}:
\begin{eqnarray}
B_p^2 &=& |\nabla \varphi|^2 |\nabla \Psi|^2 \\
B_t^2 &=& \left|{F(\Psi) \over R}\right|
\end{eqnarray}
Using the expansion of the magnetic flux one obtains: 
\begin{eqnarray}
R^2 B_p^2 &=& (\Psi_1 + 2 \rho \Psi_2)^2 g^{\rho \rho} + \left(\rho \pd{ \Psi_1}{l}^2 g^{ll}\right) \\
&=& \Psi_1^2 + 4 \rho \Psi_1 \Psi_2 + \mathcal{O}(\rho^2) \\
R^2 B_t^2 &=& F_s^2 + 2 \rho \Psi_1 F_s F_s' + \mathcal{O}(\rho^2)
\end{eqnarray}
Using now the expansion on $R$ and differentiating by $\rho$ and $l$ 
\begin{eqnarray}
\pd{ B_p^2}{\rho} &=& {1 \over R_s^2}\left(4 \Psi_1 \Psi_2 - 2 {\Psi_1^2 \over R_s} \cos u\right) \\
\pd{ B_t^2}{\rho} &=& {1 \over R_s^2}\left(2 F_s F_s' \Psi_1 - 2 {F_s^2 \over R_s} \cos u\right) \\
\pd{ B_p^2}{l} &=& {1 \over R_s^2}\left(2 \Psi_1 \pd{ \Psi_1}{l} - 2 \pd{ R_s}{l} {\Psi_1^2 \over R_s}\right) \\
\pd{ B_t^2}{l} &=& - {2 \over R_s^2} \pd{ R_s}{l} {F^2 \over R_s}
\end{eqnarray}
We can then deduce the derivatives of B by summing the previous equations
\begin{eqnarray}
\pd{ (B^2)}{\rho} &=& {1 \over R_s^2}\left(4 \Psi_1 \Psi_2 - 2 \cos u {\Psi_1 \over R_s^2} + 2 F_s F_s' \Psi_1 - 2 {F_s^2 \over R_s \cos u}\right) \\
\pd{ (B^2)}{l} &=& {1 \over R_s^2}\left(2 \Psi_1 \pd{ \Psi_1}{l} - 2 \pd{ R_s}{l}\left({\Psi_1^2 \over R_s}-{F^2 \over R_s}\right)\right)
\end{eqnarray}

A successful benchmark of the Miller geometry against GS2 is presented in Sec. \ref{linearbenchmarks}.

\subsection{Fourier parametrisation}
An alternative to the Miller parametrisation, particularly convenient for flux-tube gyrokinetic codes was introduced by J. Candy in \cite{CAN09}. It has the same advantages as the Miller parametrisation in terms of flexibility and accuracy but has the additional advantage to handle arbitrary plasma cross-sections. It relies on the Fourier transform of the flux surface minor radius as a function of the poloidal angle.\\
Starting from a reference point $(R_0,Z_0)$, the poloidal angle $\theta$ is defined as being zero at the low field side miplane and increasing anti-clockwise (same convention as Miller parametrisation). The flux surfaces are then described by:
\begin{eqnarray}
 R_N(\psi,\theta) &=& \frac{R_0}{R_\textrm{ref}} + a_N(\psi,\theta)\cos{\theta}\\
 Z_N(\psi,\theta) &=& \frac{Z_0}{R_\textrm{ref}} + a_N(\psi,\theta)\sin{\theta}
\end{eqnarray} 
where $\psi=r/R_\textrm{ref}$ as usual and 
\begin{equation}
a_N(\psi,\theta)=\sum_{n=1}^{N_\textrm{shape}} c_{nN}(\psi)\cos((n-1)\theta) + s_{nN}(\psi)\sin((n-1)\theta) 
\end{equation}
By convention, when using the Fourier parametrisation in GKW, $R_0=R_\textrm{ref}=(R_{\rm max} + R_{\rm min})/2$ and $B_\textrm{ref}=\frac{RB_t}{R_0}$  (same conventions as the Miller parametrisation implementation). The choice and value of $Z_0$ does strictly matter for GKW: it will impact the parametrisation but ultimately lead to the same metric tensors, the only difference will be the position of $s=0$ which will be at $Z=Z_0$. It is recommended, however, to use $Z_0=(Z_{\rm max} + Z_{\rm min})/2$ (IMAS definition) for consistency in the location of $s=0$. Note that the reference point $(R_0,Z_0)$ is independent on $\psi$. In practice, it means that when computing the $c_N$, $s_N$, $\partial c_N/\partial \psi$ and $\partial s_N/\partial \psi$ values from contours of the poloidal flux, a single reference point $(R_0,Z_0)$ must be used, i.e. one MUST NOT redefine the reference point $(R_0,Z_0)$ for each flux surface. \\
The usage of the Fourier parametrisation is switched on by setting \name{GEOM_TYPE='fourier'} in the \name{GEOM} namelist. One then needs to specify the number of Fourier modes in \name{N_shape}, the values of the $c_N$ and $s_N$ coefficients, as well as their radial derivatives $\partial c_N/\partial \psi$ and $\partial s_N/\partial \psi$. These inputs together with the specification of the pressure gradient in \name{gradp} and \name{gradp_type}, \name{eps}, \name{q}, \name{shat}, \name{signb} and \name{signj} provide a full description of the local equilibrium.\\
To have a self-consistent treatment of the pressure gradient in the curvature drift and in the other metric tensors, it is recommended to use \name{gradp_type=beta_prime} and specify the pressure gradient in the \name{betaprime_ref} parameter of the \name{SPCGENERAL} namelist, with \name{betaprime_type='ref'}.\\
In practice, the Fourier parametrisation is implemented in GKW by using the inputs $c_N$, $s_N$, $\partial c_N/\partial \psi$ and $\partial s_N/\partial \psi$ to compute $R_N$, $Z_N$ and their radial derivatives. 
From then on, the computation of the metric tensors is done similarly to that of the Miller parametrisation and shares the same routine.\\
Note that to use the Fourier parametrisation for global runs, one would need to add $R_0(\psi)$ and $Z_0(\psi)$ as new inputs and relax the $R_0=R_\textrm{ref}$ assumption used implicitly in the calculation of the metric coefficients.

\subsection{Sheared slab geometry}

For consistency with the way the geometry and boundary conditions are treated in the code
here we derive Hamada field aligned slab coordinates.

We require that $\vec{B} \cdot \nabla s = B^{s}$ is a flux function.

The normalised coordinate along the field line is defined similary to the toroidal case, where:
\begin{equation}
s = \frac{z}{l_s}
\end{equation}
Where s varied between -1/2 and 1/2.  The components of the magnetic field are written as:
\begin{eqnarray}
\vec{B} \cdot \nabla s &=& \vec{B} \cdot \nabla x\pd{s}{x} + \vec{B} \cdot \nabla y\pd{s}{y} + \vec{B} \cdot \nabla z\pd{s}{z}  \\
\vec{B} \cdot \nabla z &=& B^z \pd{s}{z} = B^z /l_{z} = B^s \\
B^s &=& B^z / l_z
\end{eqnarray}
Where $l_x$,$l_y$ and $l_z$ are the computational box sizes in each of the three directions.

The perpendicular component has the form:
\begin{equation}
\vec{B} \cdot \nabla\zeta =  \vec{B} \cdot \nabla y \pd{\zeta}{y} + \vec{B} \cdot \nabla z \pd{\zeta}{z} = 0
\end{equation}
Which gives:
\begin{equation}
B^y \pd{\zeta}{y} + B^z \pd{\zeta}{z} = 0
\end{equation}

If we say that the binormal coordinate is a linear function of the x and z directions
\begin{equation}
\zeta = \alpha y + \beta z
\end{equation}
we choose $\alpha=-1$, which gives:
\begin{equation}
\beta = \frac{B^y}{B^z}
\end{equation}

The field aligned, sheared slab coordinate transform from cartesian coordinates {x,y,z} can be 
written as:

\begin{eqnarray}
\psi &=& x \nonumber\\
\zeta &=& -y + \frac{B^y}{B^z} z \nonumber\\
s  &=& z
\end{eqnarray}
The negative sign in the definition of $\zeta$ is used to obtain
the same right handed set $(\psi,\zeta,s)$ as in the toroidal geometry. CAUTION: This means that the set $(x,y,z)$ is left handed.
Here $y$ corresponds to $\gamma$.

We expand the y component ($B^{y}_{0} = 0$) giving:
\begin{equation}
B^{y} = (B_{y}^{'} )\psi
\end{equation} 
where the prime denotes the radial derivative, and as such we define
\begin{equation}
\hat{s} = \frac{1}{\vec{B} \cdot \nabla z}\pd{}{ x} (\vec{B} \cdot \nabla y)
\end{equation}
When we normalise the y direction, we get a perpendicular component of the form:
\begin{equation}
\zeta_{N} = -\frac{y}{l_y} - \hat{s}_N\frac{l_{x}l_{z}}{l_y}\psi s
\end{equation}
\noindent
(signs ???) where
\begin{equation}
\hat{s}_{N} = \frac{l_{x}l_{z}}{l_y}\hat{s}
\end{equation}
This is the same as having the magnetic field of the form:
\begin{equation}
\vec{B} = B_{0} ( \hat{z} + \hat{s}x\hat{y})
\end{equation}
\noindent
With this form it can be shown that:
\begin{equation}
\vec{b} \cdot \nabla = \pd{}{s}
\end{equation}
The differentials have the form:
\begin{eqnarray}
\nabla\psi &=& \pd{x'}{x} \hat{x} = \hat{x} \\
\nabla\zeta &=& -\hat{y} + \hat{s}z \hat{x} + \hat{s}x\hat{z} \\
\nabla s  &=& \hat{z}
\end{eqnarray}

In the slab model, the input parameters $\epsilon$ and $q$ do not affect the geometry
but must both be set to 1 for correct $k_x$ mode spacing with \name{mode\_box=T}.
It can be seen that to obtain correspondance with the toroidal geometries, 
we must set $q=\hat{s} \psi$.

\subsubsection{Parallel Boundary Conditions}

The slab is always defined to be periodic in the 'radial' (x) and 'poloidal' (y) directions.
We may also wish to maintain the parallel (z) periodicity and shear boundary conditions as in the toroidal
case (Sec. \ref{sec:spectral}), i.e.
\begin{equation}
f(\psi,\zeta,s+1/2) = f(\psi,\zeta,s-1/2)
\end{equation}
The mode amplitudes
\begin{equation}
A\exp{\imath k_\zeta \zeta} = A_{-1/2} \exp{\imath k_{\zeta}\zeta_{0} - \imath k_\zeta 1/2\hat{s}x} = A_{1/2} \exp{\imath k_{\zeta}\zeta_{0} + \imath k_\zeta 1/2\hat{s}x}
\end{equation}
and as such
\begin{equation}
\Delta k_x = k_{\zeta} {\hat{s} \over i_k}
\end{equation}
where $i_k$ is an integer.  To allow the slab periodicity condition to be equivalent with the condition in toroidal geometry , we must set $q=1$ and $\epsilon=1$ in the slab case (c.f  eq. \ref{ikxspace}).   This makes the toroidal $\hat{s}$ equivalent to the one defined above,
allowing the same numerical implementation for all geometries.  The periodic slab is selected in GKW as the geometry `slab_periodic', for use with mode box only.  See also Hammett et Al 1993.

Alternatively the slab model can be chosen to be not periodic but infinite in the parallel direction. In GKW 
the `slab' model has no periodicity (the boundary condition is flappy, the same as at the ``end'' of the toroidal field line).   
For `slab' only,  nperiod $> 1$ can therefore be run with mode box.  Without mode box, `slab' and `slab_periodic' are identical.

\subsubsection{Geometry tensors}

WARNING:  The signs in this section are incorrect and need to be fixed.
The $S_B$ and $S_j$ were added by anaolgy with the toroidal case and are not kept in the derivation above.

We now calculate the geometry tensors, first the diagonal terms are as follows,

\begin{eqnarray}
g^{\psi\psi} &=& \nabla\psi \cdot \nabla\psi = 1 \\
g^{\zeta\zeta} &=& \nabla\zeta \cdot \nabla\zeta = 1 + (\hat{s}s)^{2} + (sx)^2 = 1 + (\hat{s}s)^{2} \\
g^{ss}   &=& \nabla s \cdot \nabla s = 1 \\
\end{eqnarray}
then the off diagonal:
\begin{eqnarray}
g^{\psi\zeta} &=& \nabla\psi . \nabla\zeta = -s_B s_j \hat{s} s = g^{\zeta\psi} \\
g^{\psi s} &=& g^{s \psi} = 0 \\
g^{\zeta s} &=& g^{s\zeta} = -\hat{s}x = 0
\end{eqnarray}
finally the derived tensors
\begin{eqnarray}
{\cal D}={\cal G}={\cal H}={\cal I} = 0 \\
{\cal F}= s_j, \qquad {\cal E}^{\psi \zeta} = -{\cal E}^{\zeta \psi} = -{s_j \over 2}\\
\end{eqnarray}
with all other components of ${\cal E}=0$.

We maintain the local approximation and the above tensors are evaluated at the centre of the
computational domain, which is set to $x=0$.

\subsection{General geometry}
The metric elements for general toroidal MHD equilibria can be obtained from the CHEASE code \cite{LUT96}. The sign of the magnetic field and plasma current are not taken into account into CHEASE
which always consider that ${\bf B}$ and ${\bf j}$ are in the direction opposite to $\nabla \varphi$ (here
$\nabla \varphi = \nabla \varphi_{\rm GKW}$, as everywhere in this document). The sign of ${\bf B}$ and ${\bf j}$ are therefore specified in GKW, in the \name{GEOM} namelist
using the variables \name{SIGNB} and \name{SIGNJ} defined as in the previous section by $s_{\rm B}=\textrm{sign}(\bf{B}\cdot\nabla\varphi)$ and $s_{\rm j}=\textrm{sign}(\bf{j}\cdot\nabla\varphi)$. The flux
surface of interest is also specified in the \name{GEOM} namelist using the \name{EPS} parameter. Depending on the value (1 or 2) of the \name{EPS_TYPE} parameter, \name{EPS} is either the local
inverse aspect ratio (i.e. the $\psi$ coordinate of GKW) or $\rho_\Psi$ the square root of the normalised poloidal flux ($\rho_\Psi=0$ on the axis and $\rho_\Psi=1$ on the last closed flux surface).
When CHEASE is used, the absolute value of the safety factor and magnetic shear (used for the periodicity condition and to set the radial wave vector grid) is taken from the equilibrium
reconstruction and the values specified in the namelist \name{GEOM} are not read.

When wished, the values of $\beta$ and $\beta'$ consistent with the MHD equilibrium can be used as inputs in GKW as detailed in
Sec.~\ref{Sec:betabetapr}.

For the CHEASE geometry, $R_{\rm ref} = R0EXP $ which is specified by CHEASE. It may be near the magnetic axis, but
you should always check the value used in the \name{hamada.dat} output of CHEASE.  The separate document
\doc{chease/RunChease.tex} contains more information on running CHEASE for GKW.

The $R_0=R_{\rm axis}$ option for the centrifugal terms is not quite exact for the general geometry, 
since an $s$ point on the axis may not exist.
The nearest $s$ point to $\theta=\pi/2$ is taken (the TOP intersection of the flux surface with the
magnetic axis).  For heavy impurities, this approximation could be innaccurate.  
The $R_0=R_{\rm LFS}$ option is exact for $R_0=R(s=0)$, which may not be at $\theta=0$
The value of the poloidal angle used is output to screen, and the actual $R_0$ used is written to \File{geom.dat}.

\subsection{Summary of the sign conventions in GKW} 
\label{signs}
In GKW, the cylindrical coordinate system $(R,Z,\varphi)$ is right handed, which means that $\varphi$ is increasing clockwise when the torus is viewed from above.\\
The toroidal rotation is defined positive for a plasma flowing in the direction of ${\bf B}$.\\
The mode frequency is defined positive for a perturbation evolving in the direction opposite to $\nabla \zeta$. This corresponds to the ion $\nabla B$ drift direction if $s_{\rm j}=1$ and to the
electron $\nabla B$ drift direction if $s_{\rm j}=-1$.  See also Fig.~\ref{directions}\\
Coordinate system:
\begin{itemize}
 \item $\psi=\varepsilon=(R_{\rm max}-R_{\rm min})/(2R_{\rm ref})$ is always increasing from the plasma center to the plasma edge.
 \item $s$ is always increasing upwards from the low field side midplane. It is zero at the height of the magnetic axis, on the low field side midplane.
 \item $\zeta$ is always increasing in the direction opposite to $\varphi$ (i.e. anticlockwise when viewed from above) at constant $\psi$ and $s$. The direction of $\nabla \zeta$ in the poloidal plane is
given by $\textrm{sign}(\nabla s \cdot \nabla \zeta)=\textrm{sign}(\nabla 
\zeta \cdot \nabla \theta)=s_{\rm B}s_{\rm j}$
\end{itemize}

\begin{align}
\textrm{sign}({\bf B} \cdot \nabla \varphi)&\equiv& s_{\rm B} \\
\textrm{sign}({\bf j} \cdot \nabla \varphi)&\equiv&s_{\rm j} \\
\textrm{sign}({\bf B} \cdot \nabla \theta)&=& s_{\rm j} \\
\textrm{sign}({\bf B} \cdot \nabla s)&=& s_{\rm j} \\
\textrm{sign}(\nabla \varphi \cdot \nabla \zeta)&=& -1 \\
\textrm{sign}(\nabla s \cdot \nabla \theta)&=& 1\\
\textrm{sign}(\nabla \theta \cdot \nabla \zeta)&=& s_{\rm B} s_{\rm j} \\
\textrm{sign}(B_\theta \nabla \theta \cdot \nabla \zeta)&=& s_{\rm B} \\
\textrm{sign}(\nabla s \cdot \nabla \zeta)&=& s_{\rm B} s_{\rm j} \\
{\bf B} \cdot \nabla \zeta&=& 0 \\
{\bf \Omega} &=& -s_{\rm B} \Omega \nabla z \\
u &=& R \Omega 
\end{align}
%The direction of zeta is flipped by both sign_j reversal and sign_B reversal.

\section{Spectral representation}
\label{sec:spectral}
GKW uses a combination of finite difference techniques and
pseudo-spectral methods (similar to GS2 \cite{KOT95,DOR00}).  In the
local limit (also called fluxtube approximation), the turbulence is of
homogeneous nature which motivates periodic boundary conditions. This
allows to represent all perturbed quantities in the plane
perpendicular to the magnetic field by a sum over
Fourier modes 
\begin{equation}
\label{eq:local-limit}
f(\psi,\zeta,s) = \sum_{k_\zeta k_\psi} \hat f (k_\psi, k_\zeta,s) \exp [ {\rm i} k_\zeta  \zeta /\rho_* + {\rm i} k_\psi  \psi / \rho_* ] = {\mathcal T}^{-1}(\hat f),
\end{equation}
where the normalised Larmor radius ($\rho_*$) has been added for convenient normalisation so that
\begin{equation}
\rho_* \pd{}{x_\alpha} \rightarrow {\rm i} k_\alpha .
\end{equation}
Because the distribution function is real, the Fourier amplitude of
the mode with wave vector ($k_\zeta, k_\psi$) is equal to the complex
conjugate of the mode with the wave vector ($-k_\zeta, -k_\psi$) (see
also section \ref{sec.parseval-correction}). This symmetry makes some
fourier coefficients redundant and therefore the code calculates the
evolution of the modes internally with $k_\zeta \geq 0$ only, whereas
both positive as well as negative radial wave vectors ($k_\psi$) are
used (this is in accordance to the interface of the FFT library
routines). For $k_\zeta > 0$ both signs of the wave vector are,
therefore, not represented internally, whereas for $k_\zeta = 0$ they
are. Consequently the real valued distribution function, when
expressed in the modes evaluated in the code is
\begin{eqnarray} 
f(\psi,\zeta,s) &=& \sum_{k_\zeta>0, k_\psi}  \biggl [ \hat f (k_\psi, k_\zeta,s) \exp [ {\rm i} k_\zeta  \zeta /\rho_* + {\rm i} k_\psi  \psi / \rho_* ]  + 
\hat f^\dagger (k_\psi, k_\zeta,s) \exp [ -{\rm i} k_\zeta  \zeta /\rho_* + {\rm i} k_\psi  \psi / \rho_* ] \biggr ] \cr 
\noalign{\vskip 0.2 truecm}
& & + \sum_{k_\zeta = 0, k_\psi}  \hat f (k_\psi, k_\zeta = 0,s) \exp [  {\rm i} k_\psi  \psi / \rho_* ], 
\end{eqnarray}
where the dagger denotes the complex conjugate. 
For the correct calculation of quadratic quantities like the fluxes, for instance, this representation must be considered and can lead to additional 
factors of two, when compared with the straightforward Fourier representation (see also section \ref{sec.parseval-correction}).

According to the switch \name{mode\_box} GKW is able to either setup a 2d grid of $k_\zeta,k_\psi$ given \name{krhomax} and \name{ikxspace} and or just use a single vector $(k_\zeta,k_\psi)$.
In what follows a ~$\widehat\cdot$~ indicates the Fourier representation of a quantity and ${\mathcal T}$ and ${\mathcal T}^{-1}$ 
represent the forward and inverse Fourier transform  operations, respectively.

The use of Fourier harmonics means that the boundary conditions in the perpendicular plane are periodic.  
The periodicity of the torus is, however, not automatically satisfied but leads to the shear-periodic boundary condition \eqref{eq:mb-connect} below.\\
The condition of toroidal periodicity for can be formulated for $f$ as follows, which is fulfilled by the Fourier harmonics on this $k_\zeta$ grid
\begin{equation} 
f(\psi,\zeta+ 1,s) = f(\psi,\zeta,s)  \quad \rightarrow \quad
{k_\zeta \over 2 \pi \rho_*} = N , 
\end{equation}
where $N$ is an integer. 
Since $\rho_*$ is small, this condition can in general be satisfied with very small changes to $k_\zeta$ or $\rho_*$. 
In the local limit employed here, the final equations are independent of $\rho_*$ and it 
is assumed that the relation above is satisfied. 
The poloidal periodicity translates to
\begin{equation} 
f(\psi,\zeta + q/2, 1/2) = f(\psi,\zeta- q/2,-1/2) .
\end{equation}
This condition translates to 
\begin{align}
\sum_{\bf k} \hat f( k_\psi, k_\zeta, {1 \over 2} ) \exp \biggl [ {{\rm i} k_\zeta \over \rho_*} + {{\rm i} k_\psi 
\psi \over \rho_*} + {{\rm i} q k_\zeta \over 2 \rho_*} \biggr ] \cr 
\noalign{\vskip 0.2 truecm} 
= \sum_{\bf k} \hat f ( k_\psi, k_\zeta, - {1 \over 2} )
 \exp \biggl [ {{\rm i} k_\zeta \over \rho_*} + {{\rm i} k_\psi 
\psi \over \rho_*} - {{\rm i} q k_\zeta \over 2 \rho_*} \biggr ]
\end{align}
Expanding the safety factor around a reference value $q_R$ (value at the centre of the radial domain) 
\begin{equation} 
{q k_\zeta \over \rho_*} = {q_R k_\zeta \over \rho_*} + k_\zeta \pd{ q}{\psi}{\psi \over \rho_*} 
+ {1 \over 2} k_\zeta \rho_* \pdd{q}{\psi} \biggl ( {\psi \over \rho_*} \biggr )^2 ,
\end{equation}
and neglecting the second derivative correction (which gives only a $\rho_*$ variation over the size of the box) one 
can see that this condition can be satisfied if also $q_R k_\zeta / 2 \pi \rho_*$ is assumed to be an integer and if 
\begin{equation} 
\hat f(k_\psi, k_\zeta, {1\over 2} ) = \hat f( k_\psi + k_\zeta \pd{ q}{\psi} , k_\zeta, -{1 \over 2} ) ,
\label{eq:mb-connect}
\end{equation}
i.e. on the boundary of the $s$ domain one simply connects the mode with the appropriate higher $k_\psi$ mode. 
In the code, this boundary condition is implemented in the routine \name{connect_parallel}.
This formulation is close to the `ballooning transform' \cite{CON}, and has previously been employed in GS2 \cite{KOT95,DOR00}. 
The above boundary condition for the Fourier amplitudes implies that a convenient choice for the spacing of 
the $k_\psi$ modes in any discrete Fourier representation is 
\begin{equation} 
\label{ikxspace}
\Delta k_\psi = k_{\zeta,{\rm min}} \pd{ q}{\psi} {1 \over i_k} ,
\end{equation}
where $i_k$ is some integer (\name{IKXSPACE} in the code). With this spacing, the wavenumbers necessary for the parallel boundary condition \eqref{eq:mb-connect} exist on the grid. The integer $i_k$ (parameter \name{ikxspace}) allows for a control over the spacing between the radial modes which otherwise 
would be dictated by the magnetic shear.

In order to specify the wave vector in normalised form, the expression 
\begin{equation} 
(k_\theta \rho_{\rm ref} )^2 = g^{\zeta \zeta} k_\zeta^2 
\label{kperp}
%This is referred to in the diagnostics section
\end{equation}
is evaluated at the low field side midplane ($s=0$) to determine $k_\zeta$ from the value of $k_\theta \rho_{\rm ref}$ given as an input. Note that $\rho_{\rm ref}$ is defined on the flux surface at the major radius of the magnetic axis.  Here $k_\theta$ is an unnormalised dimensional quantity, but $k_\zeta$ and $g^{\zeta \zeta}$ are normalised.

The $k_\theta$ notation used is historical but sometimes misleading; strictly speaking it would be more accurate to describe $k_\theta$ as
``$k_\perp$ at the low field side for a radial mode''.  It is related to the toroidal mode number $n$ by
\begin{equation}
n = \frac{k_\theta \rho_{\rm ref}}{2 \pi \rho_* \sqrt{g^{\zeta \zeta}|_{s=0}}} = \frac{k_\zeta}{2 \pi \rho_*} 
\label{modenumber}
\end{equation}
where $\sqrt{g^{\zeta \zeta}|_{s=0}}$ is output in \File{geom.dat} as \name{kthnorm}.  
Note that only in the $s-\alpha$ geometry does $k_\theta \rho_* = n q / r$, in other cases particular 
care should be taken  when comparing with other codes, with the toroidal mode number the least ambiguous quantity for comparison.

The full perpendicular wave vector is given by 
\begin{equation} 
k_\perp^2 = g^{\alpha \beta} k_\alpha k_\beta .
\end{equation}

One of the main advantages of the Fourier representation is that the gyro-average becomes
an algebraic operation
\begin{equation} 
\widehat{\langle \phi \rangle} = J_0 ( k_\perp \rho) \widehat \phi ,
\label{Bessel}
\end{equation}
where $J_0$ is the Bessel function and 
\begin{equation} 
(k_\perp \rho)^2 = 2\mu B {m_R T_R \over Z^2 B^2} g^{\alpha \beta} k_\alpha k_\beta .
\end{equation}

%FJC -> GSZ Describe gyroaverage with J 1 of B here

In the equation above $Z$ is the charge number of the species considered. GKW can run with 
an arbitrary number of species, which have arbitrary mass and charge.

\newpage

\section{The complete set of equations}
\label{equations}
In this section we give the complete set of equations for a collisionless rotating plasma based on the material introduced 
in the previous sections. The gyrokinetic equation can be written in the form 
\begin{equation}
\label{eqs:complete-set}
\pd{g}{t} = {\rm I} + {\rm II} + {\rm III} + {\rm IV} + {\rm V} + {\rm VI}+ {\rm VII}+ {\rm VIII},
\end{equation}
with 
\begin{align}
\label{eq:I}{\rm I} &= - v_\parallel {\bf b} \cdot \nabla f \rightarrow - v_R v_{\parallel} {\cal F} \pd{\hat f}{s}
%\cr \noalign{\vskip 0.4 truecm} 
%\end{equation}
%\begin{equation} 
\\
\label{eq:II}{\rm II} &= - {\bf v}_D \cdot \nabla f \rightarrow \cr 
\nonumber\\
& \hspace{0.5cm} -\frac{{\rm i}}{Z} \left[ T_R E_D {\cal D}^\alpha + T_R v_{\parallel}^2 \beta^\prime  {\cal E}^{\psi \alpha} + 2 m_R v_R v_{\parallel}\Omega{\cal
H}^\alpha + m_R \Omega^2 I^\alpha + Z{\cal E}^{\beta \alpha} \pd{\Phi}{x_\beta}\right] 
k_\alpha \hat f+ \cr 
\nonumber\\
& \hspace{0.5cm} -\frac{\rho_*}{Z} \left[ T_R E_D {\cal D}^s + T_R v_{\parallel}^2 \beta^\prime  {\cal E}^{\psi s} + 2 m_R v_R v_{\parallel}\Omega{\cal
H}^s + m_R \Omega^2 I^s + Z{\cal E}^{\beta s} \pd{\Phi}{x_\beta}\right] \pd{\hat f}{s}\\
\label{eq:nl-term}
{\rm III} &= - {\bf v}_\chi \cdot \nabla g \rightharpoonup - \rho_*^2 \pd{\chi}{x_\beta} {\cal E}^{\beta \alpha} \pd{g}{x_\alpha}
= {\rho_*^2 {\cal E}^{\psi \zeta}}\left({\pd{\chi}{\zeta} \pd{g}{\psi} - \pd{g}{\zeta}\pd{{ \chi }}{\psi}}\right) \cr 
\nonumber\\
& \hspace{1.5cm} \rightharpoondown \mathcal{T}\Big({\cal E}^{\psi \zeta}\left[\mathcal{T}^{-1}(ik_{\zeta} \hat{ \chi }) \mathcal{T}^{-1}(ik_{\psi} \hat g) -
\mathcal{T}^{-1}(ik_{\zeta} \hat g) \mathcal{T}^{-1}(ik_{\psi} \hat{ \chi })\right]\Big),
\\
\label{eq:IV}{\rm IV} &= + {{\bf b} \over m }\cdot(\mu \nabla B + \nabla \cfen) \pd{f}{v_\parallel} \rightarrow
v_R \left(\mu B {\cal G} + {1 \over 2} \pd{{\cal E}_R}{s} {\cal F} \right) \pd{\hat f}{v_{\parallel}}
\\
\label{eq:V}{\rm V} &= - {\bf v}_\chi \cdot \nabla F_M \rightarrow  
{\rm i}k_\alpha \hat \chi 
 {\cal E}^{\alpha \psi} \biggl [ {1 \over L_n} + {m_R \Omega ^2 \over T_R}{\cal L} + E_T {1 \over L_T} + {2 v_{\parallel} \over v_R} {R B_t \over B} u^\prime \biggr ] F_{M} \cr 
\nonumber\\
& \hspace{2.3cm} +\rho_* \pd{\hat\chi}{s}{\cal E}^{s\psi} \biggl [ {1 \over L_n} + {m_R \Omega ^2 \over T_R}{\cal L} + E_T {1 \over L_T} + {2 v_{\parallel} \over v_R}{ R B_t \over B} u^\prime \biggr ] F_{M}
\\
\label{eq:VI}{\rm VI} &= - {\bf v}_D \cdot \nabla F_M \rightarrow  {1 \over Z} \left[ T_R E_D {\cal D}^\psi + 2 m_R v_R v_{\parallel}
\Omega{\cal H}^\psi + m_R \Omega^2 I^\psi + Z{\cal E}^{s \psi} \pd{\Phi}{s}\right] \cr 
\nonumber\\
& \hspace{2.3cm} \times \biggl [ {1 \over L_n} + {m_R \Omega ^2 \over T_R}{\cal L} + E_T {1 \over L_T} + {2 v_{\parallel} \over v_R} {R B_t \over B} u^\prime \biggr ] F_{M}
\\
\label{eq:VII}{\rm VII} &= - {Z e \over T} v_\parallel {\bf b} \cdot \nabla \langle \phi \rangle F_M \rightarrow - {Z \over T_R} v_R v_{\parallel} 
{\cal F} \pd{\widehat{\langle \phi \rangle}}{s} F_{M}
\\
\label{eq:VIII}{\rm VIII} &= - {Z e \over T}{\bf v}_D \cdot \nabla \langle \phi \rangle F_M  \rightarrow \cr 
\nonumber\\
& \hspace{0.5cm} - {\rm i} \left[ E_D {\cal D}^\alpha + \beta^\prime v_{\parallel}^2 {\cal E}^{\psi \alpha} + 
{2 m_R v_R \over T_R}v_{\parallel}\Omega{\cal H}^\alpha + {m_R\Omega^2 \over T_R}{\cal I}^\alpha 
+ {Z \over T_R} {\cal E}^{\beta \alpha} \pd{\Phi}{x_\beta} \right] 
k_\alpha \, \widehat{\langle \phi \rangle} \, F_{M}+ \cr 
\nonumber\\
& \hspace{0.5cm} - \rho_*\left[ E_D {\cal D}^s + \beta^\prime v_{\parallel}^2 {\cal E}^{\psi s} + 
{2 m_R v_R \over T_R}v_{\parallel}\Omega{\cal H}^s + {m_R\Omega^2 \over T_R}{\cal I}^s 
+ {Z \over T_R} {\cal E}^{\beta s} \pd{\Phi}{x_\beta} \right] 
\, \pd{\widehat{\langle \phi \rangle}}{s} \, F_{M} \cr 
 & &\qquad 
%\cr \noalign{\vskip 0.4 truecm} 
%\end{equation}
%\begin{equation}
%{\rm IX} &= - {Z e \over T} {\mu B \over m} \biggl ( {{\bf B} \cdot \nabla {\bf B} \over B^2}  \biggr ) \langle A_\parallel \rangle F_M  \rightarrow 
%- {2 Z \over m_R } \mu B \langle A_{\parallel} \rangle {\cal G} F_{M}
\end{align}
where the arrows represent the transformations to normalised ($\rightharpoonup$) and Fourier ($\rightharpoondown$)  quantities and where 
\begin{equation} 
\label{eq:chi-normalized}
\hat \chi = \widehat {\langle \phi \rangle} - 2 v_{R} v_{\parallel} \widehat{\langle A_{\parallel} \rangle},
\end{equation} 
and 
\begin{equation} 
\hat g = \hat f + {2 Z \over T_{\rm R}} v_{\rm R} v_{\parallel} \widehat{\langle A_{\parallel} \rangle} F_{M}, 
\end{equation}
\begin{equation}
E_D = v_{\parallel}^2 + \mu B , \qquad 
E_T = v_{\parallel}^2 + 2 \mu B + \cfenn - {3\over 2} .
\end{equation}
The equations above apply to each of the species individually.

In presence of magnetic field compression $\hat \chi$ becomes
\begin{equation}
	 \hat \chi = \widehat {\langle \phi \rangle}  + \frac{2 \mu T_R}{Z} \widehat{\langle {B}_{1\parallel} \rangle} - 2 v_{R} v_{\parallel} \widehat{\langle A_{\parallel} \rangle}
\end{equation}
and there are two additional linear terms which take the form
\begin{alignat}{3}
\label{eq:X} {\rm X} &\ =\ -\frac{F_M}{T} v_{\parallel} \mathbf{b} \cdot \left( \mu \nabla \langle {B}_{1\parallel} \rangle \right) & &\ \to\ & &-2 v_R v_{\parallel} \mu F_{M} \mathcal{F} \pd{\widehat{\langle{B}_{1\parallel}\rangle}}{s} \\
\noalign{\vskip 0.1 truecm}
\label{eq:XI} {\rm XI} &\ =\ -\frac{F_M}{T} \mathbf{v}_D \cdot \left( \mu \nabla \langle{B}_{1\parallel} \rangle \right) & &\ \to\ & &-\frac{i}{Z} F_{M} 2 T_R  \mu \left( E_D \mathcal{D}^{\beta} + v_{\parallel}^2 \beta^\prime \mathcal{E}^{\psi \beta} \right) k_{\beta} \widehat{{\langle{B}}_{1\parallel}\rangle} \\
\noalign{\vskip 0.1 truecm}
&&&&& -\rho_*\frac{1}{Z} F_{M} 2 T_R \mu \left( E_D \mathcal{D}^{s} + v_{\parallel}^2 \beta^\prime \mathcal{E}^{\psi s} \right) k_{\beta} \widehat{{\langle{B}}_{1\parallel}\rangle} \nonumber
% Please do not change the numbering here to IX, there is another term later in the manual
% which has this number.  Also the numbers here correspond to numbers used in the code. But there is an exception
% The terms \propto rho_* in II is calculated in I and any term containing rho_* and (d phi/ds) are calculated in VII
\end{alignat}

The only nonlinearity in the equations is the ${\bf v}_\chi \cdot \nabla g$ term (term III, Eq.~(\ref{eq:nl-term})). Linear stability
can be investigated by removing this term. 
In the linear case the bi-normal modes are decoupled, but different radial modes can still be coupled over the parallel boundary conditions. 
An unstable mode will grow to infinite amplitude, and GKW therefore applies a scaling of the 
equations at every time-step to obtain a stationary solution. 
The nonlinear term (III) is implemented using FFTs. 
The velocity ${\bf v}_\chi$ and $\nabla g$ are calculated in Fourier space, then transformed to real space and multiplied. 
The product is transformed back into Fourier space.

The fields $\phi$ and $A_\parallel$ have to be obtained through a solution of the Poisson equation and Ampere's law. 
These can be found in the literature \cite{BRI08}. Using the normalisations of GKW the Poisson equation can be written in the form 
\begin{align}
\label{Poisson}
 \sum_{sp}  Z_{sp} n_{R_0,sp} \biggl [ 2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp} + {Z_{sp} \over T_{Rsp}} [ \Gamma(b_{sp}) -1]\exp(-\cfennsp{sp}) \hat \phi \biggr ] = 0 , 
\end{align}
where 
\begin{equation} 
\label{bsp}
b_{sp} = {1 \over 2} {m_R T_R \over Z^2 B^2} k_{\perp}^2 \qquad\leadsto  \overbrace{ {1 \over 2} {m^2 v_{\rm th}^2 \over Z^2 e^2 B^2} k_\perp^2}^{\textrm{\scriptsize in original units}} . %\textrm{   (in original units)} \bigg\} 
\end{equation}
Note that the quantities succeeding the arrow in Eq.~(\ref{bsp}), i.~e., which are denoted by 'in original units', are not normalized, while the quantities preceding the arrow are dimensionless.
Eq.~(\ref{Poisson}) differs from the literature \cite{BRI08} only in the polarisation term which includes the rotational density variation in the Maxwellian \cite{Casson-Thesis}.  An often employed approximation for the electron dynamics is the so called adiabatic response. 
In the adiabatic electron limit only the ion dynamics is simulated and the Poisson equation above is evaluated as 
\begin{align}
\label{Poisson-adiabatic}
 \sum_{sp , ions }  Z_{sp} n_{R_0,sp} \biggl [ 2 \pi B \int {\rm d} v_{\parallel} {\rm d} \mu J_0(k_\perp\rho_{sp}) \hat g_{sp} + {Z_{sp} \over T_{Rsp}} [ \Gamma(b_{sp}) -1]\exp(-\cfennsp{sp}) \hat \phi \biggr ] =  { n_{R_0,e}\exp(-\cfennsp{e})
\over T_{Re}} ( \hat \phi - \widehat{\{ \phi \}}) ,
\end{align}
where the index $e$ refers to the electrons, and the sum on the left hand side of the equation is over all ion species.   Further details on the evaluation of this equation are given in Section \ref{Poisson-detail}.

The vector potential must be calculated from Ampere's law 
\begin{align}
\label{Ampere}
\biggl ( k_{\perp}^2 + \beta \sum_{sp} {Z_{sp}^2 n_{R_0,sp} \over m_{Rsp}}\exp(\cfennsp{sp})  \Gamma(b_{sp}) \biggr ) \hat A_{\parallel}
= \beta \sum_{sp} Z_{sp} v_{Rsp} n_{R_0,sp} \times 2 \pi B \int {\rm d} v_{\parallel} \int {\rm d} \mu v_{\parallel} J_0 \hat g_{sp}.
\end{align}
Note that both field equations are formulated for $\hat g$ rather than $f$. The reason is that it is $g$ which is integrated 
forward in time. Only after the fields are calculated for the new time point, can $f$ be reconstructed. 
The Maxwell correction in $g$ leads to the second term in the brackets on the left hand side of the equation above. 
This term can lead to a cancellation problem, and the integral over the Maxwellian is therefore performed numerically 
in a similar manner as the term on the right hand side \cite{CAN03} rather than the analytic form which is given above.

The field equations in the fully electromagnetic case can be found in section \ref{field-equations}.

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/preface.tex
(1) Added gkw Clone gwk  #3

\chapter*{Preface}

This documentation comes with the GKW code. It is largely based on our 2009 paper in Computer Physics Communications Ref. \cite{CPC-paper},
but contains additional material and descriptions of additional features that the code has aquired since that paper.
In this manual GKW is presented 
with the aim of documenting the exact equations solved, presenting the essential benchmarks which 
document the proper numerical solution, and explaining the code structure and installation.  
This document is evolving along with the code, and 
currency and completeness presently take precedence over integration and polish of all sections.  
GKW developers are strongly urged to update this document (\doc{tex/doc.tex}) to reflect changes made in the code repository.

\section*{What is GKW?}

Gyrokinetic Workshop (GKW) is a code for the simulation of microinstabilities and turbulence in a magnetically confined plasma.
%The code incorporates all physics effects that can be expected from a state of the art gyrokinetic simulation code in the 
%local limit: kinetic electrons, electromagnetic effects, collisions, full general geometry with a coupling to a MHD equilibrium code, 
%and ExB shearing.
%In addition the physics of plasma rotation has been implemented through a formulation of the gyrokinetic equation in the co-moving %system. 
The gyrokinetic model is is five dimensional partial differential equation and its solution requires a massively parallel approach. 
GKW is written in Fortran 95 and has been parallelised using MPI, which enables it to scale well up to 32768+ cores.
The code was initially developed at the University of Warwick in 2007 through the extension of the linear code
LINART \cite{PEE04}, and can be used both in the linear as well as in the nonlinear regime. In 2016 it has been renamed from 'Gyrokinetics at Warwick' to 'Gyrokinetic Workshop', to reflect the nonlocality of its developers and users.
The code has been freely available online since 2010,
and has been hosted at \href{https://bitbucket.org/gkw/gkw}{https://bitbucket.org/gkw/gkw},
since 2015, where it is actively maintained.

\section*{Can I use GKW?}

You can obtain and use GKW for any purpose with minimal restrictions.
Essentially you can make any changes to the source, and are not obliged to inform us or communicate 
these changes back to us (although we would welcome it if you did).  
Furthermore, you are free to redistribute the code to others under the terms of the
license (see the LICENSE file), which is the GNU General Public License version 3.

We politely request that the effort made by us is recognised. 
In any publication that uses the results of GKW, either directly or indirectly,
you should cite the CPC paper \cite{CPC-paper}. (If you want to be kind to us you could additionally
cite the original LINART paper Ref. \cite{PEE04} and/or the first paper in which GKW has been
used Ref. \cite{PEE07}).

%N.B. This section only works because of the present document style and the ordering of the references above.
\subsection*{References}

\begin{enumerate}
\item A.G. Peeters, Y. Camenen, F.J. Casson, W.A. Hornsby, A.P. Snodin, D. Strintzi, G. Szepesi,\\
Computer Physics Communications, {\bf 180}, \href{http://dx.doi.org/10.1016/j.cpc.2009.07.001}{2650}, (2009)
\item A.G. Peeters, D. Strintzi, Phys. Plasmas {\bf 11},  \href{http://dx.doi.org/10.1063/1.1762876}{3748}, (2004)
\item A.G. Peeters, C. Angioni, D. Strintzi, Phys. Rev. Lett. {\bf 98},  \href{http://dx.doi.org/10.1103/PhysRevLett.98.265003}{265003}, (2007)
\end{enumerate}

\pagebreak

\section*{What does GKW do?}

GKW aims at solving the turbulent transport problem arising in tokamak plasmas. It solves
the gyrokinetic (the kinetic Vlasov equation averaged over the fast gyro-motion) equation on a fixed
grid in the 5-dimensional space using a combination of finite difference and pseudo spectral
methods. The solution of the kinetic equations governing this problem is an active area of 
research, with many groups worldwide contributing. The two approaches used are Particle in Cell 
(PIC) codes \cite{PAR93,LIN98,IDO03,HEI04,BOT07,JOL07,GRA06}, and Vlasov codes 
\cite{KOT95,DOR00,JEN00,DAN05,CAN03}. GKW obviously falls in the latter category, and 
reaches a standard that can be expected from a modern Vlasov code, i.e. it includes kinetic 
electrons, a general description of the geometry with a coupling to a MHD equilibrium solver, 
electro-magnetic effects, collisions, and the effect of ExB shearing. In addition the physics of 
toroidal plasma rotation is included through a formulation of the equations in the co-moving system.   
The physics of rotation is currently an active area of research and many of the applications of 
GKW have concentrated on the transport of toroidal momentum \cite{PEE07,PEE05,PEE06,
PEE09com,PEE09,CAM09,PEE09scmode,CAM09pop,CAS09}, the influence of rotation on 
other transport channels \cite{CAM09rot,CAS10}, and the interaction between turbulence and islands \cite{HOR10,HOR10epl}.
% Update the references above

\section*{Code summary}

{\bf Licensing provisions:} GNU GPLv3

{\bf Programming language:} Fortran 95

{\bf RAM:} $\sim$~128MB-1GB for a linear run; 25GB for typical nonlinear kinetic run (30 million grid points).

{\bf Number of processors used:} 1, although the program can efficiently utilise 4096+ processors (spectral), depending on problem and available computer. 128 processors is reasonable for a typical nonlinear kinetic run on the latest x86-64 machines.  The nonspectral scheme can efficiently scale to 32768+ processors.

{\bf External routines/libraries:}  None required, although the functionality of the program is somewhat
limited without a MPI implementation (preferably MPI-2) and the FFTW3 library.  In addition, the nonspectral and
implicit schemes require the UMFPACK and AMD libraries, and the eigenvalue solver requires the SLEPc and PETSc libraries. Furthermore, GKW can be compiled to support the HDF5 output format.

{\bf Standard solution method:} Pseudo-spectral and finite difference with explicit time integration.

{\bf Running time:} (On recent x86-64 hardware) $\sim$5 minutes for a short linear problem; 4000
hours for typical nonlinear kinetic run.

{\bf Geometry Interface:}  The MHD equilibrium code CHEASE \cite{LUT96} is used for the general geometry calculations. This code has been developed in CRPP Lausanne and
is not distributed together with GKW, but can be downloaded separately. The geometry module of GKW is based on the version 7.3 of CHEASE, which includes the output for Hamada coordinates.\\

\setlength{\parskip}{0pt}
\setlength{\parindent}{0pt}

\tableofcontents

\setlength{\parskip}{8pt}
\setlength{\parindent}{0pt}

\newpage

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/rotation.tex
(1) Added gkw Clone gwk  #3

\chapter{Rotation and poloidal asymmetries}

The co-moving system employed in GKW corresponds to a frame that rotates as a rigid body with constant frequency 
$\Omega$. The Coriolis drift, centrifugal drift, centrifugal trapping and centrifugal potential 
due to the rotation of the plasma are all incorporated in the complete set of equations described above. 
The possible radial gradient in the toroidal rotation is  treated through a radial gradient in the averaged parallel velocity $u^\prime$ 
of the background.  The radial gradient of the perpendicular velocity ($E \times B$ shear flow) has so far been ignored, 
but is known to play an important role in the saturation of the turbulence.

In this chapter we describe how the centrifugal potential is calculated from the species inputs, 
and how the  the $E \times B$ shearing is treated in GKW.  Much of this Chapter appears in greater detail in Ref. \cite{Casson-Thesis}.

%Describe further:
% Origin of cf potential
% Why some terms vd.grad remain and others not from ordering.
% Maxwellian
% Make note in arawaka scheme about s periodicity.
% Modification to collisions
% Choice of n_{R_0 location}
% Feild equations do not need modifying
% deriv wrt zeta zero due to axisymmetry

%Do
%Check Arakawa scheme is working correctly with mode_box

%Discuss
% No gyroaverage of phi

\section{Centrifugal potential\label{cfphi}}
We recall that the rotation of the plasma leads to an equilibrium density variation within a flux surface
\begin{equation}
  n(s)=n_{R_0,s} \exp\left({-Z_s \langle\Phi\rangle\over T_{R,s}}+{m_s\Omega^2(R^2-R_0^2)\over T_{R,s}}\right) 
\end{equation}
(normalised units). The background equilibrium potential $\Phi$ in the rotating frame is found by applying the quasi-neutrality condition over all the species
\begin{equation}
  Q(\Phi)=\sum_s Z_s n_{R_0,s} \exp\left({-Z_s \langle\Phi\rangle\over T_{R,s}}\right)\exp\left({m_s\Omega^2(R^2-R_0^2)\over T_{R,s}}\right)=0
  \label{eq:cf-quasineutral}
\end{equation}
Since there is only one negative species and the exponential function is monotonically increasing, the equation $Q(\Phi)=0$ will always have exactly one root.  
In GKW, $\Phi$ is calculated numerically for arbitrary species combinations from the quasi-neutrality condition by an iterative root finding bisection algorithm.  The calculation is done once for each point in $s$ during the initialisation phase of the code. 
The terms II, VI, and VIII (\ref{eq:II}, \ref{eq:VI}, \ref{eq:VIII}) also require the $\psi$ and $s$ derivatives of $\Phi$, which are calculated using a fourth-order finite difference.

%Describe bisection method
The bisection algorithm begins with an upper limit estimate $\Phi_a$ for $\Phi$.  The initial value used in GKW is (somewhat arbitrarily) taken as
\begin{equation}
  \biggr[\max\{\log(1+|Z_s n_s|)\}+\max\{m_s\log(1+n_s)\}\biggr]\Omega^2 (R^2-R_0^2) + 0.1
\end{equation}
which should always contain the root, but be small enough to prevent exponentiation overflows.  It is possible that extreme
species input data could break the solver, and the upper limit estimate might need to be adjusted for those cases.
% The earlier condition below caused an overflow for tungsten impurities:
% \begin{equation}
% \Phi_a = 5\max(T_s)\left[\log\left(\left|{\max(Z_s)\max(m_s) \over \min(T_s)}\right|\right)+\log\left(|\Omega^2(R^2-R_0^2)|\right)\right]+0.1
% \end{equation}

The starting lower limit is estimated as $\Phi_b=-\Phi_a$. The value is chosen to ensure that the root lies in the range $(\Phi_b,\Phi_a)$. In each step the mid value $\Phi_{\rm est}=(\Phi_a+\Phi_b)/2$ is taken and the value of $Q(\Phi_{\rm est})$ is calculated.  If $Q(\Phi_{\rm est})>0$, then the upper estimate is updated ($\Phi_a=\Phi_{\rm est}$), and if $Q(\Phi_{\rm est})<0$, then the lower estimate is updated ($\Phi_b=\Phi_{\rm est}$).  The process repeats until the solution has converged to within machine accuracy.  The bisection algorithm, while not the fastest possible, is guaranteed to converge as long as the bounds are appropriate and the function $Q$ has only one root in the initial range.  In practice, the convergence occurs in about 55 iterations using the initial estimates specified above.  The function $Q(\Phi,s,\psi)$ is evaluated in module \name{components} as \name{cf_quasineutral} called from the bisection algorithm \name{centrifugal_energy} in module \name{rotation}.

In the local flux tube model, the gyrokinetic equation is solved only on one flux surface (i.e. at one position in $\psi$), keeping radial derivatives of equilibrium quantities.  To consistently calculate the radial derivative of $\Phi$, equation \ref{eq:cf-quasineutral} is also solved numerically on adjacent flux surfaces using $Q(\Phi,s,\psi+\Delta \psi)$, using the radial derivatives of the species quantities to calculate the temperature and densities on adjacent flux surfaces.

The choice of $\Delta \psi$ used in the code is motivated by a compromise between machine precision when dividing by small numbers, and the accuracy of the finite difference as $\Delta \psi \rightarrow 0$.  The discrepancy between the root finding algorithm and the analytic solution (Eq. \ref{eq:deriv-scions}) is plotted in figure \ref{fig:cfphi-acc}. The value of $\Delta \psi = 1e-4$ is implemented as a permanent choice, which for the case investigated minimises the error to less than $10^{-11}$ in double precision.

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{cfphi-accuracy.eps}
\caption{Numerical experiment to determine optimum $\Delta \psi$ for calculating the radial derivative of the centrifugal potential $\Phi$ for the singly charged ions Waltz standard case. The experiment was conducted in double precision and the fractional error is calculated against the analytic result Eq. \ref{eq:deriv-scions}.}

\label{fig:cfphi-acc}
\end{center}
\end{figure}

In the case of a plasma of only singly charged ions, Eq. \ref{eq:cf-quasineutral} solves exactly to give
\begin{equation}
 \Phi = \underbrace{{T_e T_i \over T_e + T_i} \left({m_i \over T_i} - {m_e \over T_e}\right)}_{\name{cf_mass_weight}} \Omega^2 (R^2-R_0^2).
\label{eq:scions}
\end{equation}
with radial derivative 
\begin{equation}
 {\partial \Phi \over \partial \psi} = \underbrace{\left[{m_i {\partial T_e \over \partial \psi}- m_e {\partial T_i \over \partial \psi}\over T_e + T_i} + 
{m_e T_i - m_i T_e \over (T_e + T_i)^2}\left({\partial T_e \over \partial \psi}+{\partial T_i \over \partial \psi}\right)\right]}_{\name{cf_dtf1}}\Omega^2(R^2-R_0^2) 
+2 \Omega^2 \underbrace{\left[{m_i T_e - m_e T_i \over T_e + T_i}\right]}_{\name{cf_mass_weight}} 
\left(R {\partial R \over \partial \psi}-R_0 {\partial R_0 \over \partial \psi}\right)
\label{eq:deriv-scions}
\end{equation}
%ADD UPRIME PIECE HERE, and in COMPLETE SET OF EQUATIONS, and in solver above
If the bulk ion species is singly charged, these quantities are evaluated in the code in \name{rotation} with the labelled function provided by \name{components} to provide a check on the solution of the root finding algorithm.

The density $n_R$ and density gradient $1/L_n=(1 / L_n)_{R_0}$ (Eq. \ref{eq:gradients}) are defined for each species at the point on the flux surface at which $R=R_0$.  Two choices for the definition of $R_0$ are available in the code, which are selected by the input variable \name{R0\_loc} in the \name{geom} namelist.  The option \texttt{R0\_loc='axis'} sets $R_0$ as the major radius of the magnetic axis (but the reference point is not on the axis itself), whilst the default \texttt{R0_loc='LFS'} sets $R_0$ to be the value of $R$ at $s=0$, in the plane of the magnetic axis at the low field side (for general geometry this may \textit{not} be equivalent to the maximum $R$).  This choice alters only the definition of the geometry quantities $\mathcal{J}$ and $\mathcal{K}$ .  In the rotating case, the effective density gradient varies over the flux surface and there is no longer one obvious definition of the density gradient as a dimensionless quantity.  For instance
\begin{equation}
{1 \over L_n^{E} } = -{1 \over n} {\partial n \over \partial \psi},\qquad
{1 \over L_n^{e1} } = -{1 \over \{ n \} } {\partial n \over \partial \psi},\qquad
{1 \over L_n^{e2} } = -{1 \over n_{R_0}} {\partial n \over \partial \psi}, \qquad
\label{eq.rlndiags}
\end{equation}
can each be argued to be a relevant quantity. The most appropriate may be determined by the nature of the particular instability being examined. These definitions can be evaluated to be

\begin{equation}
{1 \over L_n^{e} } = {\partial \cfenn \over \partial \psi} + 
\cfenn {1 \over L_T}+  {1 \over L_n} \biggr|_{R_0}, \qquad
{1 \over L_n^{e1} } = {1 \over L_n^{e} } { \exp(-\cfenn) \over
\big\{ \exp(-\cfenn) \big\} }, \qquad
{1 \over L_n^{e2} } = {1 \over L_n^{e} }  \exp(-\cfenn)
\label{eq.rlndiags2}
\end{equation}

Since the density gradient is an important parameter for determining the stability of many modes, GKW calculates ${1 / L^E_N }$ and ${1 / L^{e1}_n }$ and $n/n_{R_0}$ for each species as a function along the field, which is written to the file \File{cfdens.dat}, and as
a flux surface average and at the outboard midplane, which is written to screen during the initialisation phase of the code. ${1 / L^{e2}_n }$ can be trivially calculated from the other two.  It should be noted that whilst the input ${1 / L_n} |_{R_0}$ must always be the same for every species in order to satisfy quasi-neutrality, the effective  ${1 / L_n^E}$ can differ for each species.  For typical parameters of a deuterium plasma, the variation in effective density and gradient can be $30\%$ at $\Omega=1$, increasing rapidly for $\Omega>1$.

\section{Background \texorpdfstring{$E \times B$}{E x B} shear flow}

The sheared $E\times B$ velocity leads to a stabilisation of turbulence through the breaking up of eddy structures. 
This $E\times B$ shearing is always present for a purely toroidally rotating plasma but can, of course, also be related 
to a sheared poloidal rotation.

%FJC: CHECK AGAIN FACTORS OF RHO* What happens if normalise as background potential?

The equilibrium $E \times B$ rotation 
\begin{equation}
{\bf v}_s(\psi)= {{\bf b} \times \nabla \bar{\Phi} \over B} 
\end{equation}
lies inside the flux surface.  The shearing rate in the normalised units is defined to be 
\begin{equation}
\gamma_{E}^N = {1 \over 2}\rho_*^2 {\partial^2 \bar{\Phi}_N \over \partial \psi^2} ,
\label{gamma-E}
\end{equation}
where the normalisation $\bar{\Phi}= \rho_* {T_{\rm ref} \over e}\bar{\Phi}_N$ is as for the
perturbed potential Eq.~(\ref{eq:phi-norm}).  The factor $1/2$ is present due to the definition of the
reference thermal velocity.  Since GKW treats the $E \times B$ shear independently from the centrifugal effects,
we use $\bar{\Phi}$ to distinguish the background potential of the shearing from the background potential of the rotating frame.
In physical units the shearing rate is
\begin{equation}
\gamma_E = {v_{\rm thref} \over R_{\rm ref}} \gamma_E^N = 
{1 \over B_{\rm ref}}{\partial^2 \bar{\Phi} \over \partial r^2} ,
%Contingent on \phi=r/R_{ref}
%True if r = R_{max}-R_{min}
\end{equation}
where $r=(R_{\rm max}-R_{\rm min})/2$.  The GKW shearing rate is assumed
radially constant and is defined as a flux function.  At the radial centre of the flux tube
$\bf{v_s}$ is chosen to be zero, with the result that there is no net flow over the domain.  In the
local limit with the approximations of the `$s-\alpha$' model geometry
the shearing rate is then equivalent to the familiar definition \cite{Hahm95,Burrell97}:
\begin{equation}
\gamma_E \approx {(RB_p)^2 \over B_t}{\partial^2 \bar{\Phi} \over \partial \Psi^2}
\approx {d v_s \over d r} .
\end{equation}
The flow velocity is normalised as $v_s^N=v_s / v_{\rm thref}$, and as before the index $N$ is dropped in what follows.
The sign convention here is opposite to Eq.~(\ref{eq:gradients}), so $\gamma_E<0$ corresponds to 
$\nabla E_0$ radially outwards.

Only the Doppler shift of the background $E \times B$ rotation is kept in the description, i.e. the
background $E\times B$ rotation is 
added as an additional convective term for the perturbed distribution (${\bf v}_s \cdot \nabla g$) in the gyrokinetic equation~(\ref{eqs:complete-set}), 
and we neglect the acceleration due to the background potential (${\bf v}_D \cdot \nabla \bar{\phi} F_M$). 
By omitting the latter term, the flow is implemented in conservative form and provides no drive to the turbulence, so that the 
stabilising effect may be studied in isolation.  If the acceleration term were kept, it would provide a Kelvin-Helmholtz drive to
the turbulence.  This drive is likely to be small compared to the parallel shear drive (which is kept in term VIII), but could be the subject of future work.

With zero net flow across the flux tube one obtains
\begin{equation}
{\rm III.B} = -{\bf v}_s \cdot \nabla g \rightharpoonup - \rho_*^2 {\partial \bar{\Phi} \over \partial
\psi}{\mathcal E}^{\psi \zeta}{\partial g \over \partial \zeta} =
 - 2 {\gamma_E \psi}{\mathcal E}^{\psi \zeta}  {\partial g \over \partial \zeta} .
\end{equation}
Defining $\bar \gamma_E= 2 \gamma_E {\mathcal E}^{\psi \zeta} $ and taking the Fourier transform, the term can be written as a derivative in
Fourier space
\begin{equation} 
{\mathcal T}(-{\bar \gamma_E \psi}{\partial g \over \partial \zeta}) = \bar \gamma_E
k_\zeta {\partial \hat g \over \partial k_\psi} .
\end{equation}
The derivative represents a continuous advection (and shearing) in Fourier space in $k_\psi$.

There are a number of subtleties associated with numerical evaluation of this derivative. A finite difference approximation cannot resolve the fine 
scale structure in Fourier space.  The `harmonic derivative' \cite{Waltz94,Miller94} is a full convolution over all modes and is equivalent to a
pseudo-spectral implementation using FFTs (as was used for the nonlinear term in Eq.~(\ref{eq:nl-term})).  
Although it correctly implements the $E \times B$ shearing term inside the computational domain, it does not represent a homogeneous shear 
flow because the flow profile does not satisfy the periodicity of the discrete Fourier transform; any turbulent structure passing through the radial
boundary will experience a discontinuity in the flow profile (which in the periodic domain has a sawtooth form).  
Since the local limit requires that the turbulence has no preference for the boundary, periodicity at the boundary must be preserved. 
The `shear-periodic' boundary condition \cite{Baron82,Schumann85,Gerz89} that moves with the mean flow accomplishes this for a finite 
difference representation in the radial direction, but does not naturally translate to the spectral representation.

In coordinates that move with the flow  \cite{Rogallo81,Zang88,Pumir96}, the `radial' wavenumbers become time dependent:
\begin{align}
\psi^\prime = \psi , &\qquad& \qquad \zeta^\prime=\zeta-\psi \bar \gamma_E t ,\\
k_\zeta^\prime = k_\zeta , &\qquad& \qquad  k_\psi^\prime = k_\psi - k_\zeta \bar \gamma_E t .
\end{align}
In these coordinates, the derivative in Fourier space becomes part of the time derivative.

For a gyrokinetic code, a time dependent wave vector requires the re-evaluation of the linear terms and Bessel functions at every time-step
and would be computationally prohibitive. 
By discretising the time dependence of the wavenumbers as a remapping of the solution vector between the original fixed wavevectors, 
this expensive re-evaluation can be avoided \cite{HammettAPS06}. 
Using only the fixed wavevector grid, the advection in Fourier space occurs in jumps only when the boundary periodicity is satisfied for a 
given mode. 
Explicitly, the remapping is implemented by tracking the number of times each wavevector has been remapped $i_r(k_\zeta)$, 
and (for $\bar{\gamma}_E > 0$) remapping the solution vector
\begin{equation}
\hat g (k_\psi, k_\zeta,s) \rightarrow \hat g (k_\psi-\Delta k_\psi, k_\zeta,s),
\end{equation}
 when the inequality
\begin{equation}
 {\rm Int}( k_\zeta \bar{\gamma}_E t / \delta k_\psi ) > i_r (k_\zeta),
\end{equation} 
is satisfied.  Here, ${\rm Int}$ is the function which returns the \textit{nearest} integer.
At the low $k_\psi$ limit the solution vector is discarded.  
This means that the method is non-conservative, but since the turbulence is characterised by a peaked spectrum 
(Fig.~\ref{fig:cyclone-linear}), the losses are negligible if the range of radial wavenumbers is suitably wide.  
For numerical stability, the remapping must occur at the same time for all points along the flux tube. 
As the metric tensor ${\mathcal E}^{\psi \zeta}$ is a flux function, see Eq.~(\ref{eq:efun-cst}), this condition is always satisfied for a 
constant shear rate along the field line.

The method in GKW differs from the standard spectral methods to model homogeneous shear flows in fluid turbulence of
Refs.~\cite{Rogallo81,Zang88,Pumir96}, where the wavenumbers must be recalculated and re-meshing occurs for all wavevectors 
at the same time. 
In GKW the method used is the one proposed by Hammett et al. in Ref.~\cite{HammettAPS06}.  
Here the `re-meshing' occurs continually (and at different times for each $k_\zeta$), and the wavenumbers stay on their original fixed grid.  
The accuracy of the method has been argued to be second order on average \cite{HammettAPS06} and
is able to capture the physics effects of a 
background shear flow whilst allowing the desirable features of the flux tube model to be retained.

Convergence of the remap method in $L_x / L_y$ should be checked (particularly by for the modes with lowest $k_\zeta$) by decreasing 
$\Delta k_\psi / k_{\zeta {\rm min}}$ (increasing $N_x$ and $p$ (IKXSPACE) whilst holding $N_{\rm mod}$ constant (defined later)).

\section{Purely toroidal sheared rotation in general geometry}

For this section only, we adopt the superscript $^L$ to represent quantities in the laboratory frame.  
All quantities without a superscript should be interpreted as being in the rotating frame (as in all other sections). 
The same coupling condition is derived by two routes to make clear the relationship between the frames.

The rotating frame is constructed \cite{PEE09} such that rigid body rotation $\Omega$ of the frame 
matches the plasma rotation $\omega_\phi^L$ on the local flux surface.  The angular rotation then transforms
as 
\begin{equation}
\omega_{\phi}(\psi)=\omega_{\phi}^L(\psi)-\Omega 
\label{eq:transform}
\end{equation}
such that $\omega_{\phi}(\psi)=0$ (and hence ${\bf
v_s}=0$) at the centre of the radial domain.

\subsection{In comoving frame}

For a purely toroidal rotation, decomposing the toroidal flow into its parallel and perpendicular components gives
\begin{equation}
u_\parallel \mathbf{b} +  \mathbf{v_s} = s_{\rm B} R \omega_\phi(\psi) R \nabla \varphi,
\label{eq:toroidal}
\end{equation}
where $\varphi$ is the toroidal angle. Since \eq{toroidal} is written in the comoving frame,
 it reads $0+0=0$ in the centre of the flux tube.
From the normalisations above, one can show that in the normalised units
\begin{equation}
 \mathbf{v_s} = {1 \over 2} \rho_*^2 {\mathbf{b} \times \nabla \psi \over
B} {\partial \bar{\Phi} \over \partial \psi}.
\end{equation}
Taking the bi-normal component of \eq{toroidal} one finds
\begin{equation}
 u_{\parallel} \underbrace{\mathbf{b} \cdot \nabla \zeta}_{=0} +  {1 \over 2}
\rho_*^2 \underbrace{{\mathbf{b} \times \nabla \psi \over B}\cdot \nabla
	\zeta}_{=2\mathcal{E}^{\psi \zeta}}{\partial \bar{\Phi} \over \partial \psi} = s_{\rm B} R^2 \omega_{\phi}(\psi)
\underbrace{\nabla \phi \cdot \nabla \zeta}_{=-1/2\pi R^2},
\end{equation}
hence
\begin{equation}
 {1 \over 2} \rho_*^2 {\partial \bar{\Phi} \over \partial \psi} = - {s_{\rm B} \over 4 \pi}{1 \over
\mathcal{E}^{\psi \zeta}}\omega_{\phi}(\psi),
\label{eq:comoving}
\end{equation}
which upon differentiation gives the coupling relation
\begin{equation}
 \underbrace{{1 \over 2} \rho_*^2 {\partial^2 \bar{\Phi} \over \partial \psi^2}}_{\gamma_E} = - {s_{\rm B} \over 4 \pi}\left[{1 \over
\mathcal{E}^{\psi \zeta}}\underbrace{\partial \omega_{\phi} \over \partial \psi}_{\rm u^\prime} +
\underbrace{\omega_\phi}_{=0}{\partial \over \partial \psi} \left({1 \over \mathcal{E}^{\psi \zeta}}\right)\right].
\label{eq:coupling}
\end{equation}
Note that the second term on the right is zero when evaluated at the centre of the flux tube.
All quantities in the above equation are flux functions, hence the shear rate will always be a
flux function, as required for the discrete remapping method.

\subsection{Relation to lab frame}

In the laboratory frame \eq{toroidal} becomes 
\begin{equation}
u_\parallel^L \vec{b} + {\bf v}_s^L  = s_{\rm B} R \omega_\phi^L(\psi) R \nabla \varphi.
\label{eq:toroidal2}
\end{equation}
Taking the binormal component leads to
\begin{equation}
 {1 \over 2} \rho_*^2 {\partial \bar{\Phi}^L \over \partial \psi} = - {s_{\rm B} \over 4 \pi}{1 \over
\mathcal{E}^{\psi \zeta}}\omega^L_{\phi}(\psi),
\end{equation}
and taking the radial derivative gives
\begin{equation}
 {1 \over 2} \rho_*^2 {\partial^2 \bar{\Phi}^L \over \partial \psi^2} = - {s_{\rm B} \over 4 \pi}\left[{1 \over
\mathcal{E}^{\psi \zeta}}{\partial \omega^L_{\phi} \over \partial \psi} + \omega^L_{\phi}{\partial \over \partial
\psi} \left({1 \over \mathcal{E}^{\psi \zeta}}\right)\right],
\end{equation}
which has the same form as \eq{coupling} but different values.

To make explicit the relationships between the quantities, the previous equation 
is rewritten (using \eq{transform}) in terms of the rotating frame quantities.
Since the frame rotates rigidly, ${\partial \Omega \over
\partial \psi}=0$ and ${\partial \omega / \partial \psi} = {\partial \omega^L / \partial \psi} = u^\prime$. 
The tensor component $\mathcal{E}^{\psi \zeta}$ is invariant under the transformation.
The electric field transforms \cite{PEE09} as
\begin{equation}
\bar{\Phi}=\bar{\Phi}^L + s_{\rm B} s_{\rm j}{2 \over \rho_*^2}\Psi \Omega ,
\end{equation}
where $\Psi$ is the (frame independent) poloidal flux with $\nabla \Psi$ radially outward, 
and the factor of $2 / \rho_*^2$ arises from the normalisation of $\bar{\Phi}$ (see Sec. \ref{signs} for clarification of signs).  
The coupling condition can then be written as 
\begin{equation}
 \underbrace{{1 \over 2} \rho_*^2 {\partial^2 \bar{\Phi} \over \partial \psi^2}
-s_{\rm B} s_{\rm j}  {\Omega}{\partial^2 \Psi \over
\partial \psi^2}}_{\gamma_E^L} = - {s_{\rm B} \over 4 \pi}\left[{1 \over
\mathcal{E}^{\psi \zeta}}{\partial \omega_{\phi} \over \partial \psi} +
\Omega{\partial \over \partial \psi} \left({1 \over \mathcal{E}^{\psi \zeta}}\right)\right],
\end{equation}
when evaluated in the centre of the flux tube where $\omega_\psi = 0$.
Since $\Psi=f(\psi)$ only, from the relation \ref{eq:efun-flux}
% \begin{equation} 
% {\cal E}^{\psi \zeta} = \frac{s_{\rm j}}{4\pi}\frac{\partial \psi}{\partial \Psi}
% \label{eq:efun-cst}
% \end{equation}
it follows that
\begin{equation}
 {\partial \over \partial \psi} \left({1 \over \mathcal{E}^{\psi \zeta}}\right) = s_{\rm j} 4 \pi 
{\partial^2 \Psi \over \partial \psi^2},
\end{equation}
hence cancellation gives 
\begin{equation}
 \underbrace{{1 \over 2} \rho_*^2 {\partial^2 \bar{\Phi} \over \partial \psi^2}}_{=\gamma_E}
 = - {s_{\rm B} \over 4 \pi}{1 \over
\mathcal{E}^{\psi \zeta}}\underbrace{\partial \omega_{\phi} \over \partial \psi}_{=u^\prime},
\end{equation}
which is the same as \eq{coupling} when evaluated at the centre of the flux tube.  The shearing rates in the two frames are therefore related by
\begin{equation}
 \gamma_E^L = \gamma_E - s_{\rm B} s_{\rm j}{\Omega}\underbrace{{\partial^2 \bar{\Psi} \over \partial \psi^2}}_{\cal{M}},
\end{equation}
but presently only $\gamma_E$ is implemented in the code.

\section{Poloidal asymmetries induced by anisotropic minorities}

In this section, all units are dimensional, except those indicated with a superscript $N$, which denotes 
a dimensionless quantity using the GKW normalisations defined in Section \ref{sec:normalisation}.

Recent experiments have observed poloidal asymmetries in heavy impurities in the presence of ion cyclotron
resonance heating (ICRH) of a light minority ion \cite{Reinke12}.  This phenomena is believed to be due to 
the anisotropy of the heated species, which sets up an equilibrium potential.

\subsection{Reinke model \label{sec.icrh}}

 In the rotating frame of reference, the parallel
force balance for an anisotropic species $m$ (generalisation of Eq 3 of Ref. \cite{Reinke12} with a sign correction for the second term)
is 
\begin{equation}
%\vec{b} \cdot (\nabla \cdot \vec{P} n_m e Z_m \nabla_\parallel \Phi) = 
\nabla_\parallel p_\parallel - \frac{p_\parallel - p_\perp}{B}\nabla_\parallel B + n_m Z_m e \nabla_\parallel \Phi - n_m m_m \Omega^2 R \nabla R = 0
\end{equation}  
Assuming $\nabla_\parallel T_\parallel = 0$, one finds on integration
\begin{equation}
n_m = A B^{-\eta} \exp\left(-\frac{e Z_m \Phi}{T_\parallel} + \frac{m_m \Omega^2 R^2}{2 T_\parallel}\right)
\end{equation}
where $\eta = T_\perp/T_\parallel - 1$, and $A$ is a constant of integration (with dimensions of $[n B^{\eta}]$) to be determined. 
In Ref. \cite{Reinke12}, $A = \{n\}/\{B^{-\eta}\}$, but this choice is actually 
inconsistent with the definition of the flux surface average $\{ \}$; the correct definition (when $\Omega = 0$) would actually require
\begin{equation}
A_{\rm FSA}  = \frac{\{n_m\}}{\{B^{-\eta}\exp(-{e Z_m \Phi / T_\parallel})\}}. 
\label{eq.correctnorm}
\end{equation}
Although the difference between this and the normalisation of Eq. 3 of Ref. \cite{Reinke12} is small for typical parameters, the deviation
is enough to break quasi-neutrality and so the approximate normalisation of Ref. \cite{Reinke12} is not possible
to use in a code which contains a field solver.  Furthermore, defining the constant $A$ correctly as in Eq. \ref{eq.correctnorm} 
would require repeated iterations of the root finding quasi-neutrality solver condition over the whole flux surface, 
which although possible is probably unnecessary for this problem.
Instead, to be consistent with the previous definition of $n_{R0}$ as the density at $R=R_{0}$, we must have $A = n_{R0} B_{R0}^\eta$  
and thus we have
\begin{equation}
n_m = n_{R0,m} \left(\frac{B}{B_{R0}}\right)^{-\eta} \exp\left(-\frac{e Z_m \Phi}{T_\parallel} + \frac{m_m \Omega^2 (R^2-R_0^2)}{2 T_\parallel}\right)
\end{equation}

For the minority species it is also convenient to define a generalised poloidal asymmetry energy
\begin{equation}
  {\cal E}_{\Omega,\eta} = {\cal E}_{\Omega} + T_\parallel \eta \ln({B/B_{R_0}})
  \label{eq.poloidal-energy}
\end{equation}
which in the GKW normalisations is 
\begin{equation}
  {\cal E}^N_{R,\eta} = {\cal E}^N_R + T^N_\parallel \eta \ln({B/B_{R_0}})
\end{equation}
where $T^N_\parallel = T_\parallel / T^N_R T_{\rm ref}$  (set $ = 1 $ in the approximations below).  This 
fits into the previous formalism such that 
\begin{equation}
  n_m = n_{R0} \exp(-{\cal E}_{\Omega,\eta}/T_\parallel)
\label{eq.cfen-gen}
\end{equation}
The definition of $\cfen$ is unchanged for all other species (although the value of $\Phi$ inside $\cfen$ will be).

\subsection{Bilato model \label{sec.icrh2}}
The Reinke model makes another approximation, by taking $p_\perp$ as a flux function.  The self-consistent solution worked out by Bilato \cite{Bilato14} gives a reduction in the asymmetry compared to the Reinke model roughly equivalent to transforming $\eta \rightarrow \eta^{3/4}$.  This model is also implemented in GKW, and is selected by using an input value of $\frac{T_{\perp R0}}{T_{\parallel}} < 0$, where the absolute value is used by the code.

The exact solution is 
\begin{equation}
 n_m = n_{R0,m} \frac{T_{\perp}(\theta)}{T_{\perp R0}}  \exp\left(-\frac{e Z_m \Phi}{T_\parallel} + \frac{m_m \Omega^2 R^2}{2 T_\parallel}\right)
\end{equation}
where
\begin{equation}
 \frac{T_{\perp}(\theta)}{T_{\perp R0}} = \left [ \frac{T_{\perp
 R0}}{T_{\parallel}} + \left(1 - \frac{T_{\perp R0}}{T_{\parallel}} \right )\frac{B_{R0}}{B} \right ]^{-1}.
\end{equation}
The generalised asymmetry energy for the minority is then defined such that Eq. \ref{eq.cfen-gen} still holds, with
\begin{equation}
    {\cal E}_{\Omega,\eta} = {\cal E}_{\Omega} + T_\parallel \ln({T_{\perp R0}}/{T_{\perp}(\theta)})
\label{eq.poloidal-energy2}
\end{equation}
which in the GKW normalisations is 
\begin{equation}
   {\cal E}^N_{R,\eta} = {\cal E}^N_R + T^N_\parallel \ln({T_{\perp R0}}/{T_{\perp}(\theta)})
\end{equation}
where $T^N_\parallel = T_\parallel / T^N_R T_{\rm ref}$  (set $ = 1 $ in the approximations below).

\subsection{Transformations of input parameters}

If, as is usual in GKW, we take $R_0 = R_{\rm LFS}$, then the consequence is that the minority
fraction $f_{\rm min} = \{n_m\}/\{n_e\}$ differs strongly from the value of $n_{R0,m} / n_{R0,e}$ in the presence of anisotropy.

However, in some cases, (particularly those of a strongly anistropic minority species)
it is desirable that the code density inputs can be specified in the form of flux surface averages,
because this is the canoncial conserved quantity in the absence of radial transport.  Such an input specification 
also removes the need for post-processing of the output quantities as dsecribed in \ref{sec.memo-icrh}.

For the reasons given in the preceding section, defining $n_{\rm input} = \{n\}$ and $rln_{\rm input} = R_{\rm ref}/L_{\{n\}} = 1/\{n\} \partial \{n\} / \partial \psi$
using the flux surface averaged densities requires an extra layer of iterative numerical solutions.

ITERATIVE transformation scheme under development, to be documented shortly....

\subsection{Implementation}

Using the generalised quasi-neutrality root finding solver for $Q(\Phi) = 0$ described in Sec. \ref{cfphi} above, it is then straightforward to 
insert the additional $B^{-\eta}$ dependence for a minority species.  Since GKW also requires and computes radial derivatives of $\Phi$, 
a maximum of 7 parameters are potentially required: 
\begin{equation}
( n_{\rm input,m}/n_{\rm ref},\qquad {T_\perp R0}/T_\parallel,\qquad {\partial (T_{\perp R0}/T_\parallel) / \partial \psi},\qquad
 T_\parallel,\qquad  \partial T_\parallel / \partial \psi,\qquad Z_m,\qquad m_m ).
\end{equation}
We remind the reader that $\psi = r/R_{\rm ref}$ is the dimensionless radial coordinate used in GKW.  Note that the minority mass is not required as an input in the case when $\Omega = 0$.

In addition, it can be useful to include the asymmetric potential generated by the minority species without simulating it as a separate gyrokinetic species (combination method: NOT PRESENTLY WORKING CORRECTLY).
For this purpose, although treated separately in the background quasi-neutrality solve, the minority species is after-wards combined with the bulk ion species ($i$).  
Since $Z e \Phi / T_\parallel \ll 1$ and $f_{\rm min} \ll 1$ only ever appear in combination if the exponential is expanded as series, 
the values of $T_\parallel$ and $\partial T_\parallel / \partial \psi$ make a negligible impact on the results 
(tested for $f_{\rm min} = 0.1$ and $T_\parallel/T_{\perp R0} = 10.0 $). 
For simplicity, these parameters are therefore neglected in the combination method, and where required, the code makes the assumption that $T_i = T_\parallel$.  (This assumption is required to have a value in the code, but in reality it does not affect the results). Only one mass is used for the combined species.  Since the mass does not appear in the anisotropy effect, the average mass of the minority plus bulk should be given
to model the centrifugal and kinetic effects correctly, but using the assumption $m_m = m_i$ will likely give almost identical results 
for most parameters.  Outside the quasi-neutrality solve, the two species are combined, with $n_{R0,t} = n_{R0,i} + n_{R0,m} Z_m / Z_i$ such that $n_t = n_i + n_m Z_m / Z_i $ everywhere, which gives a resulting asymmetry energy of
\begin{equation}
% Line below only when the species have the same Z
%  {\cal E}_{\Omega,\eta,t} =  {\cal E}_{\Omega,\eta,m} - \ln\left(n_{RO,i} (B/B_{\rm ref})^\eta + n_{RO,m}\right)
   {\cal E}_{\Omega,\eta,t} =  -\ln\left(n_{R_0,i}\exp(-{\cal E}_{\Omega,i}) + \frac{n_{R_0,m} Z_m}{Z_i}\exp(-{\cal E}_{\Omega,\eta,m})\right) + \ln(n_{R_0,t})
\end{equation}
such that quasi-neutrality is still satisfied everywhere for the reduced set of species.

Because the combination option means that the asymmetry does not require an additional gyrokinetic species, the 
input parameters $n_{\rm input,m}/n_{\rm ref}, T_{\perp R0}/T_\parallel, \partial (T_{\perp R0}/T_\parallel) / \partial \psi^N, Z_{\rm min}$ are 
separated from the individual species inputs and appear in the \name{SPCGENERAL} namelist as the array \name{icrh_params(1,2,3,4)}.
respectively.

If instead the minority is treated as a full gyrokinetic species (independent method), the species is identified with
\name{background='ICRH'} in the species namelist. This allows the fully general case without the (mild) 
assumptions on the parameters $m_m, T_\parallel$ and $\partial T_\parallel / \partial \psi^N$ required for 
the combination option.  In this case, $n_{\rm input,m}/n_{\rm ref}$ and $Z_{\rm min}$ are taken from the species inputs \name{dens} and \name{Z} respectively
(the first and last inputs \name{icrh_params(1,4)} are not used), and the species inputs \name{temp} and 
\name{rlt} describe $T_\parallel^N$. In the gyrokinetic equation, this species is then treated as a Maxwellian with temperature $T=T_\parallel$
(i.e., there is no bi-Maxwellian option implemented for $F_0$). Strictly, a more accurate approximation would be a Maxwellian with $T=(T_\parallel T_\perp^2)^{1/3}$, which might be more accurate for collisions or neoclassics.

After the solver completes, the code runs self-consistency checks on the quasi-neutrality of the gradients of the flux surface averages, which is a strong test on the consistency of the solver, using both the diagnostics in Eq. \ref{eq.rlndiags}, and the transformations described in Sec. \ref{sec.memo-icrh}.  The gradient of the flux surface average
\begin{equation}
-\frac{1}{\{n\}}\frac{\partial \{n\}}{\partial \psi} =  \left \{1/L_n^{e1}\right\}
\end{equation}
for the minority is calcuated using 
\begin{equation}
 {1 \over L_n^{e,N} } = {\partial {\cal E}_{R,\eta} \over \partial \psi} + 
{\cal E}_{R} {1 \over L^N_T}+  {1 \over L^N_n} \biggr|_{R_0}
\label{eq.rlndiag3}
\end{equation}
(note, the second term does NOT use the generalised energy $ {\cal E}_{R,\eta}$) 
with
\begin{equation}
 {1 \over L_n^{e1,N} } = {1 \over L_n^{e,N} } { \exp(-\cfenn) \over
\big\{ \exp(-\cfenn) \big\} }, \qquad
\end{equation}
which is the generalisation of Eq. \ref{eq.rlndiags2}).  The first term in Eq. \ref{eq.rlndiag3} is evaluated with the derivatives appearing in Sec. \ref{sec.memo-icrh}, 
%or Sec. \ref{sec.memo-icrh2}, 
and the results of the checks are output to screen.

\pagebreak

\subsubsection{Examples of anisotropic minority input setup}

Since the quasi-neutrality solver is only activated for centrifugal effects, the switch \name{cf_trap=T} must also be set
even in the absence of rotation.  Here we present two example input setups for a 
$5\%$ helium-4 minority with $T_\perp/T_\parallel = 4$ and $\partial (T_\perp/T_\parallel) / \partial \psi = -8.5$.
First treating the minority species as a gyrokinetic species, (independent method) we have:
\begin{footnotesize}
\begin{verbatim}
&GRIDSIZE
number_of_species = 4
/
&SPCGENERAL
icrh_params = 999, 4.0, -8.5, 999  ! First and last not used
/
&SPECIES ! D ions
mass = 1.0000, dens = 0.90, z = 1.00, rlt = 5.5, rln = 2.5, temp = 1.0, uprim = 0.0
/
&SPECIES ! electrons
mass = 2.7e-4, dens = 1.00, z = -1.0, rlt = 4.5, rln = 2.5, temp = 1.2, uprim = 0.0
/
&SPECIES ! Molby trace impurity
mass = 48.000, dens = 0.00, z = 32.0, rlt = 6.5, rln = 1.5, temp = 1.0, uprim = 0.0
/
&SPECIES ! He ICRH minority
mass = 2.0000, dens = 0.05, z = 2.00, rlt = 9.9, rln = 2.5, temp = 2.0, uprim = 0.0
background = 'ICRH'
/
&ROTATION
cf_trap = .true.
/
\end{verbatim}
\end{footnotesize}
or, alternatively, treating the minority in combination with the bulk ions (combination method),
and giving practically identical results for $\Phi$ \texttt{and its deriviatives}: 
\begin{footnotesize}
\begin{verbatim}
&GRIDSIZE
number_of_species = 3
/
&SPCGENERAL
icrh_params = 0.05, 4.0, -8.5, 2.0  ! minority fraction, Tp/T||, d(Tperp/T||)/dpsi, Z_m
/
&SPECIES ! D ions + He minority (treated with Z = 1 in Possion and GK equations)
mass = 1.0527, dens = 1.00, z = 1.00, rlt = 5.5, rln = 2.5, temp = 1.0, uprim = 0.0
/
&SPECIES ! electrons
mass = 2.7e-4, dens = 1.00, z = -1.0, rlt = 4.5, rln = 2.5, temp = 1.2, uprim = 0.0
/
&SPECIES ! Molby trace impurity
mass = 48.000, dens = 0.00, z = 32.0, rlt = 6.5, rln = 1.5, temp = 1.0, uprim = 0.0
/
&ROTATION
cf_trap = .true.
/
\end{verbatim}
\end{footnotesize}

\subsubsection{Tests}

\begin{itemize}

\item The combination method and the independent method give identical asymmetries.
\item Quasi-neutrality is always satisfied everywhere, with all $\eta$, and all $\Omega$, all geometries, for both $R_0$ choices, 
and arbitrary non-trace species combinations.
\item Quasi-neutrality in the gradients is satisfied at all locations, and in the flux surface average density, checked using the formulae of Sec. \ref{sec.memo-icrh}.
\item The $\Phi$ asymmetry scales linearly with $f_m$, $Z_m$ and $\eta$, and is practically independent of $T_\parallel$.
\item High-Z impurity asymmetries are of the same magnitude as in Ref. \cite{Reinke12} for the given parameters.
\
%STILL TO DO:

%\item Investigate sensitivities of the radial derivatives
%\item Fix radial derivatives for combination method

\end{itemize}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/tearing.tex
(1) Added gkw Clone gwk  #3

% In here should add the integral of Mattias hand written notes and corrections
% that David has found to signs and amplitudes

% FIX NOTATION OF \psi -> \Psi for poloidal flux.  \psi is the radial coord = r/R only.
% REMOVE CGS units

\chapter{Tearing modes}

Here are some brief notes that describe how an imposed magnetic island is included in GKW 
(self-consistent tearing modes can occur naturally given correct inputs and thus do not 
require a description of their implementation here). 
The previous version (Refs~\cite{HOR10,HOR10epl}) assumed that the island is always initialised
in the largest poloidal mode.  To have the island in a larger wavevector there are some subtleties to the initialisation.   
For a recent description of the implementation and benchmarks in slab geometry, see 
Zarzoso et al,. PoP, 2015.

\section{Spectral implementation of magnetic islands in toroidal geometry}
If we initially assume a constant poloidal flux perturbation with a given helicity 
\begin{equation} 
A_\parallel = A_{\parallel0} \exp [ 2 \pi {\rm i} (m s - n \gamma ) ] 
\end{equation} 
where $s$ and $\gamma$ are the poloidal and toroidal angles, respectively.  Introducing the helical coordinate $\zeta = qs - \gamma$ and expanding the safety factor up to first order
\begin{equation} 
q = m / n + \Delta\psi\partial q /\partial\psi
\end{equation} 
in the centre of the box (i.e. the mode is resonant at this position). We need to transform this
representation to the ($\psi, \zeta, s)$ coordinate system which gives,
\begin{equation} 
A_\parallel = A_{\parallel0} \exp [ 2 \pi {\rm i} ( n \zeta - {\partial q \over \partial \psi} n s \Delta\psi ) ] 
\end{equation} 
From which it directly follows that 
\begin{equation} 
k_\zeta = 2 \pi n \rho_* 
\end{equation}
is the toroidal mode of the island.  However for generality we assume that this may not be the largest wavelength within the domain, this is $k_\zeta^{min} = 2\pi\rho_*$.

In the helical field line co-ordinate system the x-points only appear in the $\zeta$ direction.  In the toroidal and poloidal angles, x-points appear in both.  
From the periodicity constraint connected with the parallel boundary conditions it follows that the radial wavevectors are given by
\begin{equation} 
 k_\psi = (p/i_k) k_\zeta {\partial q \over \partial \psi} = (p/i_k) 2 \pi \rho_* {\partial q \over \partial \psi} 
\end{equation} 
where $p$ is a set of integers in the range $[-N_x/2, N_x/2]$ and $i_k$ is the parameter that determines the radial mode spacing.  
From $p = 1$ one directly obtains the radial size of the box 
\begin{equation} 
\psi = \biggl [ - { i_k \over 2 \partial q / \partial \psi} , {i_k \over 2 \partial q / \partial \psi} \biggr ] \cdot {1 \over k_\zeta^{min}} 
\end{equation}

The Fourier amplitudes of the different radial modes then can be calculated by integrating over 
the radial domain 
\begin{align}
\hat{A}(k_\psi, k_\zeta, s) &=& A_{\parallel0} \exp{(i\ n k_\zeta^{min}\zeta)}\int {\rm d} \Delta\psi \exp{\biggl [i\Delta\psi( k_\psi + n s k_\zeta^{min} dq/d\psi)   \biggr ]} \nonumber\\
&=& {A_{\parallel0} i_k \over \pi n \partial q / \partial \psi} {\sin [ \pi (n i_k s + p) ] \over n i_k s + p}  \nonumber\\
&=& C{i_k \over n} {\sin [ \pi s^\circ ] \over s^\circ} 
\end{align}
where we have used the ballooning angle $s^\circ = n i_k s + p$ in the last step and $n$ is the poloidal wave mode (determined by the input parameter \name{isl_mode}).

One point to note at this point is that this perturbation is not periodic in the radial domain of GKW for all values of $s$.  This manifests itself as a discontinuity in the parallel vector potential at the radial edge of the computational domain.  A solution to this problem is to relax the constant flux approximation and therefore write the perturbation as,
\begin{equation} 
A_\parallel = C{i_k \over n} \exp{i k_{\zeta}\zeta}\sum_{p=N_x/2}^{N_x/2} A^{p}(s)\exp{i pk_{\psi}\Delta\psi}
\end{equation}  
Where $A^{p}(s)$ includes a damping function.  We have chosen a Gaussian profile to reduce any ringing.  This gives
\begin{equation}
A^{p}(s) = \exp{\left((-(n i_k s + p)^{2}/L_s^{2})\right)}\frac{\sin{\pi(n i_k s + p)}}{(n i_k s + p)}
\end{equation}
Where L is half width of the gaussian envelope.  This approximation leads to a more satisfactory shape for the vector potential, where the value of L can be tuned to that the damping of the small k modes is negligible and the of high k modes sufficiently strongly to prevent the occurance of a jump at the boundary.

Thus the full implemented vector potential has the form,
\begin{equation}
A_{||} = C {i_k \over n} \exp{(i k_{\zeta}\zeta - i\omega t)}\sum_{p=N_x/2}^{N_x/2}\exp{\left(-\frac{(n i_k s + p)^2}{L_s^2}\right)}\frac{\sin{\pi((n i_k s + p))}}{(n i_k s + p)}\exp{(i p k_{\psi}\Delta\psi)}
\end{equation}
where $\omega$ is the poloidal rotation frequency of the island.  Note that only a single binormal mode is used.  The island half-width is defined in dimensional units (Wesson) as 
\begin{equation}
W = 2\sqrt{\frac{rq \tilde B_{r}}{m B_p (dq/dr)}} = 2\sqrt{\frac{r \tilde A_\parallel}{B_p {\hat s}}} = 2\sqrt{\frac{q R \tilde A_\parallel}{B_t {\hat s}}}
\end{equation}
where $B_p$ is the poloidal component of the background magnetic field\footnote{The correspondance with the literature on the sheared slab by Smolyakov and others is as follows: They have $W^2 = 4 L_s A_{\parallel} / B_0$, but $\psi$ is the vector potential in their notation.  The toroidal result used in GKW can be obtained by inserting $1/L_s = (1/q)(dq/dr)$, and $B_0 = B_p$.}.  The island width determines the amplitude of the island perturbation:  The factor $dq/dr = q {\hat s} / r$ is obtained from the magnetic shear and $m$ is the poloidal mode number.  Using the fact that the peturbed radial magnetic field is prescribed as $\tilde{B}_{r}= m\tilde{\Psi} /rR$, and that for the $s-\alpha$ model equilibrium the perturbed magnetic flux is given by $\tilde{\Psi} = R\tilde{A}_{\parallel 0}$, and $q=r B_t / R B_p$, the amplitude of the island vector potential (now in GKW normalised units) is given by
\begin{equation}
\tilde{A}_{\parallel0N} = {W_N^2 B_{tN} {\hat s} \over 4 q R_N}, \qquad  C = {W_N^2 B_{tN} \epsilon \over 4 \pi q^2 R_N}.
\label{aparamp}
\end{equation} 
where $B_t^N$ is taken at the low field side in order to define the island width at the same location, and $W = W_N \rho_{\rm ref}$. Note that the amplitude of C given in equation \ref{aparamp} is divided by 2, due to the form of the Fourier representation described in Sec. \ref{sec:spectral}.  % HOW IS C OBTAINED - THE BOX SIZE with dq/dpsi = shat q /eps - eqn 5.6 enters   ?
%the fact that
%the island is actually a perturbation in real space (not in Fourier space), which must be of the form
%$\cos\left[2\pi\left(ms-n\gamma\right)\right]$.

\begin{figure}
\begin{center}
\includegraphics[width=0.3\textwidth]{MagIslGeom.eps}
\caption{Magnetic Island Geometry for an island with a half width of 30 ion gyro-radii}
\label{islandgeom}
\end{center}
\end{figure}

To run GKW with a magnetic island present the following options must be used:  Firstly, the code must be set to run electro-magnetically by setting \name{nlapar=.true.}.  In the \name{SPCGENERAL} namelist, \name{tearingmode} must be set to {\it .true.}, which adds the magnetic island as a perturbation of the equilibrium field.  The half width of the island is determined by the input parameter \name{wstar} which is the island width normalised to $\rho_{ref}$ (usually the ion gyro-radius).  A point of note is that this island width can not exceed (and should be smaller than) the radial extent of the computational box due to the periodicity constraints of a pseudo-spectral code. The rotation frequency of the island ($\omega$) is set by the parameter \name{isl\_rot\_freq}, normalised as other frequencies in the code BUT with opposite sign (positive is in the $\nabla \zeta$ direction, which is the electron grad-B drift direction when $s_j = 1$ and rln > 0.  By default the island is placed in the smallest non-zero poloidal wavevector (\name{isl\_mode=2}), however this can be changed by setting \name{isl\_mode} to another integer (e.g. \name{isl\_mode=3} will give 2 islands within the domain as the zero mode is always the first mode).

We recommend initialising the distribution function with \name{finit='zero'} as the island itself will seed the turbulence, and this gives faster convergence compared to non-zero distribution function initialisations.  When plotting XY profile outputs, setting {\it lisl_follow} in the \name{DIAGNOSTIC} namelist will follow the magnetic island so that the O-point is always in the centre of the box.  We do not recommend using the options to initialise the island after the turbulence, as growing the island too quickly drives a unphysical and large zonal flow, which initally supresses the turbulence and then takes a long time to decay.

\begin{small}
\begin{verbatim}
 &SPCGENERAL
 beta = 0.0001,  ! not used
 tearingmode=.true.,  wstar = 5.0,  finit = 'zero',  adiabatic_electrons = .false.
 Ls = 2.0,  isl_mode = 2, isl_rot_freq = 0.1   
 /
\end{verbatim}
\end{small}

\section{Spectral implementation of islands in 2D sheared slab geometry}

An alternative way to deal with 2D islands is also implemented, in which
parallel dynamics are projected into the $y$ direction, 
as is usually done in slab analytics.  This allows GKW islands to be compared
directly with analytical results in the slab.   Whilst this setup is intended for use with the slab geometry, 
it might also be applicable for a LFS toroidal case.
The island amplitude is decoupled from geometric magnetic shear, and 
a new parameter, \name{isl\_shear} is used in determing the island width.
To use this option the geometric shear \name{shat} is set to 0 (which
means reducing the coordinate system to purely Cartesian) and the code is
run with a single s point and without parallel dynamics 
(a limit in which the geometrical magnetic shear plays no role).
This also implies that the minimum $k_x$ and $k_y$ now can be set completely
independently, but they are chosen to be the same (to give a square box).

The perturbed $A_\parallel$ in this case then also contains the part of the island that
normally comes from the background magnetic field (in the 2D case the background B
plays no role in the geometry and theoretically only enters via the Larmor
radius). The slab geometry is defined in this case through a Cartesian set of
coordinates, $x$,$y$ and $z$, where $z$, the direction of the guide magnetic
field, is meant to be a symmetry coordinate for the whole system,
i.e. $\partial/\partial z = 0$, accordingly to the existing dedicated literature. 
The total magnetic field in this case reads
\begin{equation}
{\bf B} = B_z \nabla z -\nabla \psi \times \nabla z,
\end{equation}
with 
\begin{equation}
\psi = -\frac{i_s x^2}{2}B_z+\tilde{\psi}\cos( k_y y),
\label{psiisl}
\end{equation}
where $B_z$ is a constant and where  $i_s$ corresponds to \name{isl\_shear},
the (inverse of the) scale of variation of $B_y$, which plays a role analogous
to the magnetic shear in toroidal geometry. The $y$-dependent term represents
the island.  The amplitude $\tilde{\psi}$, assumed to be constant as a
consequence of the well-known ``constant-$\psi$ approximation'', is  linked to the island width $w$ through the relation
\begin{equation}
w^2=\frac{4 \tilde{\psi}}{i_s B_z}.
\end{equation}

In the 3D case, only the latter ``island''-part of $\psi$ is imposed as a
perturbation, entering the gyrokinetic equation through $v_\chi$ in the nonlinear term. 
In this 2D case, in constrast, we consider a shearless slab, with equilibrium 
\begin{equation}
{\bf B}= B_z \nabla z
\end{equation}
(constant) and we include the whole $\psi$ of Eq. \ref{psiisl}
as a perturbation. In Fourier space this extra island component
(additional to the usual part in the island $k_y$ mode)
is applied in the $k_y=0$ mode, and has the form 
\begin{equation}
A_\parallel^+(p) =  {4 (-1)^p i_s \over {\bar k}_x(p)^2 } \exp(-(p/Ls)^2)
\end{equation} 
where $p$ is again the radial mode number, and an exponential damping has again 
be added to the exact Fourier transformed quantity. The overbar on ${\bar k}_x$
represents the modes in the code, to distinguish them from the island wavenumbers $k_x$ and $k_y$
used above.  While it is apparent that $\psi$ is periodic on $y$, the
quadratic dependence on $x$ cannot be exactly reproduced in Fourier space, as
its derivative is not the same on the two boundaries. Thus, analogously to
the toroidal case, we again employ the exponential damping for the high $k_x$ modes in
order to remove boundary discontinuities.  This approximation has the effect of creating a second
magnetic island at the boundary which in full 3D geometry is obscured by parallel coupling, ballooning
and curvature. The effect of this boundary island can be minimised by 
choosing an appropriate combination of the parameters box size, shear, island width, damping length
(which differ in slab and torodial geometry) but it can never be completely eliminated.

In the 2D case, there is however an additional island option, to add \textit{two}
identical islands radially with perfect periodicity.  This option is activated
when \name{shat} = 0 and \name{isl\_shear} $> 0$ by selecting a
negative \name{wstar}. By choosing this option,  $\psi$ is
approximated by 
\begin{equation}
\psi = {\hat C}\cos{k_x x}+\tilde{\psi}\cos{k_y y},
\end{equation}
where $\hat{C}= \pi B_z i_s/ k_x^2$ is a constant chosen in order to have the
same amplitude as the other case for $B_y= \partial \psi/\partial$ at the edge
of the box. In this case, the additional nonzero component only appears in the $k_y=0$, $p= -1,1$ modes
with value $A_\parallel^+ = \pi i_s / k_x^2$.  The sinusoidal dependence can of course be implemented without any
approximation, or artificial damping, leading therefore to no parasitic boundary island
formation. However, the presence of both a maximum and a minimum in $x$
corresponds to the presence of two identical islands with $\pi/2$-shift in the
$y$-direction. By carefully choosing the box width, one is able to reduce the
mutual interaction of the two islands.

Summarizing, for slab geometry GKW  has three options for islands:
\begin{itemize}
\item To run with shat $>$ 0 in 3D, which is expensive (and where
discontinuities at the boundary create numerical problems) and which does not
reproduce the boundary condition $\partial / \partial z =0$ used in many analytic models.
\item To run with shat = 0 and wstar $>$ 0. These are 2D runs, much faster, but which still
employ the artificial damping on high- ${\bar k}_x$ modes, and therefore require
specific attention to the discontinuity at the boundary, as in the full 3D cases.
\item To run with shat = 0 and wstar $<$  0. These are 2D runs, with no
artificial damping, but which contain two identical out of phase magnetic
islands at two radial locations.
\end{itemize}

Example 2D setup:
\begin{small}
\begin{verbatim}
 &GRIDSIZE
 NX = 167,  N_mu_grid = 8,  n_vpar_grid = 12, number_of_species = 2,
 N_s_grid = 1,  NMOD = 20,  nperiod = 1,
 /
 &MODE
 mode_box = .true., krhomax = 0.45, ikxspace = 1,
 /
 &GEOM
 GEOM_TYPE = 'slab_periodic', SHAT = 0.0,  Q = 1.0, EPS = 1.0,
 /
 &SPCGENERAL
 finit = 'zero',  tearingmode=.true.,
 wstar = 10.0, isl_rot_freq = -0.02,
 adiabatic_electrons = .false.
 ls=8.0, isl_shear = 0.1
\end{verbatim}
\end{small}

\section{Nonspectral implementation of a magnetic island}

If the radial direction is not treated in Fourier space, the implementation is much more direct, since we need only to introduce at $t=0$ the term
\begin{equation}
A_\parallel = A_{\parallel 0}\cos\left(2 \pi {\rm i} (m s - n \gamma )\right)
\end{equation}
which can be expressed by
\begin{equation}
A_\parallel = {1\over 2}A_{\parallel 0}\left[e^{i2\pi n\zeta}\left(\cos{\partial q \over \partial \psi} n s \Delta\psi - i\sin{\partial q \over \partial \psi} n s \Delta\psi\right)
+ e^{-i2\pi n\zeta}\left(\cos{\partial q \over \partial \psi} n s \Delta\psi + i\sin{\partial q \over \partial \psi} n s \Delta\psi\right)\right]
\end{equation}

In the binormal direction we initialise only one mode, namely $e^{i2\pi n\zeta}$ and assume the existence of the other mode in other to get a real quantity once the inverse Fourier transform is performed. Therefore,
 we need to implement only the potential
 \begin{equation}
A_{\parallel,+} = {1\over 2}A_{\parallel 0}\left(\cos{\partial q \over \partial \psi} n s \Delta\psi - i\sin{\partial q \over \partial \psi} n s \Delta\psi\right)
\end{equation}

In order to decouple the island from the boundary of the simulation box, we multiply the previous potential by a radial envelope having the explicit expression
\begin{equation}
A_r\left(\Delta\psi\right) = {1\over 4}\left(1+\tanh\left({\Delta\psi+\Delta\psi_0\over \delta\psi_0}\right)\right)\left(1-\tanh\left({\Delta\psi-\Delta\psi_0\over \delta\psi_0}\right)\right)
\end{equation}
where $\Delta\psi_0$ and $\delta\psi_0$ are defined in the tearingmode namelist of the input file of GKW by the variables \File{psi_0} and \File{delta_psi_0}, respectively.

In a 2D sheared slab geometry, the previous implementation reduces to
\begin{equation}
A_\parallel = {1\over 2}A_{\parallel 0}A_r\left(\Delta\psi\right)
\end{equation}
and the background magnetic shear is introduced together with this perturbation by means of the mode $\kappa_\zeta=0$. Therefore, the total parallel vector potential reads in 2D sheared slab geometry
\begin{equation}
A_\parallel = {B_zi_s\over 2}\left[\Delta\psi^2-\left(L_x/2\right)^2\right]+{1\over 2}A_{\parallel 0}A_r\left(\Delta\psi\right)e^{i2\pi n\zeta}
\end{equation}
where the additional term ${B_zi_s\over 2}\left(L_x/2\right)^2$ is introduced to impose a vanishing potential at the boundary in the nonspectral version. We therefore need to introduce the term ${B_zi_s\over 2}\left[\Delta\psi^2-\left(L_x/2\right)^2\right]$ in the $\kappa_\zeta=0$ component and the term ${1\over 2}A_{\parallel 0}A_r\left(\Delta\psi\right)$ in the $\kappa_\zeta = \kappa_{\rm isl}$ component.

\section{Reconstruction of flux tube radial profiles \label{profiles} and profile flattening in GKW}

The benchmark of the magnetic islands implementation includes the flattening of the density and pressure profiles. The equilbrium profiles are linear with respect to the radial coordinate and are present in the source terms.  Since we are solving the gyrokinetic equation for the perturbed part of the distribution function, the solution should respond to the source terms due to the background. In particular, the magnetic island should give a perturbation that cancels the background out inside the separatrix. Note that runs must be non-linear for the correct level of flattening to be obtained. To reconstruct the radial profiles the XY diagnostics must enabled in the diagnostics namelist.

% \begin{verbatim}
% &DIAGNOSTIC
%  lisl_follow = .false.,
%  lphi_diagnostics = .true.,
%  xy_phi = .true.,
%  xy_apar = .true.,
%  xy_fluxes = .true.,
%  xy_fluxes_em = .true.,
%  xy_dens = .true.
%  xy_current = .true.,
%  xy_temp = .true.,
%  / 
% \end{verbatim}

\subsection{Density profiles}

The files named \File{den0\#_000*_*} hold the normalised XY perturbed density data ($\tilde{n}_{N,s}$) for each large time step. The background density profile can be reconstructed by using the local radial cordinate $x_r = \psi / \rho_* = r / \rho_{\rm ref}$ which is written in the file named \name{xphi} (with the same dimensions as the XY slices) and the input parameter \name{RLN_s} = $-(1/n_{R_0,s}) {\partial n_{R_0,s} \over \partial \psi}$. The total density of species $s$ is $n_s = n_{s,\rm eq} + \tilde{n}_s$, where $n_{s,\rm eq} = n_{R_0,s}\left(1-{r-r_0\over L_{n,s}}\right)$ is the global density of the Maxwellian equilibrium written in the local limit, i.e. $n_{R_0,s}$ at the resonant position $r_0$ plus the linear variation due to the gradient at that position. Note that in the expression of $n_{s,\rm eq}$, the quantity $n_{R_0,s}$ does not exhibit any radial dependence since it is considered in the local limit. The perturbed density of species $s$ is

\begin{equation}
\tilde{n}_s = \int_{\mathbb R^3} d^3\mathbf{v}\delta f_s = \rho_*n_{R_0,s}\int_{\mathbb R^3} d^3\mathbf{v}_{N,s}\delta f_{N,s} = \rho_*n_{R_0,s}\tilde{n}_{N,s}
\end{equation}

Therefore, the total density for species $s$ reads

\begin{equation}
n_s = n_{R_0,s}\left(1-{r-r_0\over L_{n,s}}\right) + \rho_*n_{R_0,s}\tilde{n}_{N,s} = n_{R_0,s}\left(1 - {\left(x_r-x_{r0}\right)\rho_{\rm ref}\over {L_{n,s}}}\right) + \rho_*n_{R_0,s}\tilde{n}_{N,s}
\end{equation}

We can now normalise the density for species $s$ as $n_{N,s} = n_s / n_{R_0,s}$

\begin{equation}
n_{N,s} = \left(1 - {\left(x_r-x_{r0}\right)\rho_{\rm ref}\over {L_{n,s}}}\right) + \rho_*\tilde{n}_{N,s}
\end{equation}
and define the change in the density for species $s$ as $\Delta n_{N,s} = n_{N,s} - 1$

\begin{equation}
 \Delta n_{N,s} = - \left(x_r-x_{r0}\right){\rho_{\rm ref}R_{\rm ref}\over {L_{n,s}R_{\rm ref}}} + \rho_*\tilde{n}_{N,s}  =  - \rho_* \left(x_r-x_{r0}\right) {R_{\rm ref}\over L_{n,s}} + \rho_*\tilde{n}_{N,s}
\end{equation}
or alternatively
\begin{equation}
{\Delta n_{N,s} \over \rho_*} = \tilde{n}_{N,s} - {R_{\rm ref} \over L_{n,s}} \left(x_r-x_{r0}\right)
\end{equation}
and the total density gradient (background plus perturbed density) can be computed as
\begin{equation}
{R_{\rm ref} \over L_{n_{\rm eq}+\tilde{n}}}  =  -{1 \over n_{R_0,s}} {\partial (n_{s,\rm eq}+\tilde{n}_s) \over \partial \psi} = 
 -{\partial \tilde{n}_{N,s} \over \partial x_r} + {R_{\rm ref} \over L_{n,s}}.
\end{equation}

%PREVIOUS VERSION needs a factor rho_* to get to a normalised total density (i.e. it has the normalisation of delta f, not F_M)
%\begin{equation}
%n({\Delta\psi}) = \tilde{n} - RLn*xphi
%\end{equation}

\subsection{Pressure profiles}
To obtain the pressure profiles we need to use the files \File{ene0\#_000*_*}, which hold the normalised second moments of the distribution function, namely
\begin{equation}
\tilde E_{N,s} = \int_{\mathbb{R}^3}d^3\mathbf{v}_{N,s}\delta f_{N,s}v_{N,s}^2
\end{equation}

It can be shown that the pressure perturbation is written in terms of this quantity as follows
\begin{equation}
\tilde{p}_s = {1\over 3}m_{\rm ref}v_{th,s}^2\rho_* n_{R_0,s}m_{R,s}\tilde E_{N,s}
\end{equation}

Taking into account the definition of the reference temperature $T_{\rm ref}$, one can write

\begin{equation}
\tilde{p}_s = {2\over 3}\rho_* n_{R_0,s}T_{\rm ref}m_{R,s}v_{R,s}^2\tilde E_{N,s}
\end{equation}

In addition, $m_{R,s}v_{R,s}^2=T_{R,s}$ and $T_s = T_{\rm ref}T_{R,s}$, which leads to
\begin{equation}
\tilde{p}_s = {2\over 3}\rho_* n_{R_0,s}T_s\tilde E_{N,s}
\end{equation}

In the same way as has been done with the density, the total pressure is defined as $p_s = p_{s,\rm eq} + \tilde{p}_s$, where $p_{s,\rm eq} = n_{R_0,s}T_s\left(1-{r-r_0\over L_{n,s}}\right)\left(1-{r-r_0\over L_{T,s}}\right)$. In the local limit we can neglect terms of second order in $\rho_*$ and therefore, the equilibrium pressure profile reads
\begin{equation}
p_{s,\rm eq} \approx n_{R_0,s}T_s\left(1-\rho_*\left(x_r-x_{r0}\right){R_{\rm ref}\over L_{n,s}}-\rho_*\left(x_r-x_{r0}\right){R_{\rm ref}\over L_{T,s}}\right)
\end{equation}

We can now divide by $n_{R_0,s}T_s$ and define the change in the normalized pressure as

\begin{equation}
\Delta p_{N,s} = {2\over 3}\rho_*\tilde E_{N,s}-\rho_*\left(x_r-x_{r0}\right){R_{\rm ref}\over L_{n,s}}-\rho_*\left(x_r-x_{r0}\right){R_{\rm ref}\over L_{T,s}}
\end{equation}
or equivalently
\begin{equation}
{\Delta p_{N,s} \over \rho_*} = {2\over 3}\tilde E_{N,s}-\left(x_r-x_{r0}\right){R_{\rm ref}\over L_{n,s}}-\left(x_r-x_{r0}\right){R_{\rm ref}\over L_{T,s}}
\end{equation}

This quantity is plotted in figure \ref{fig:pressureflat}. Note that the flattening is observed in the moments of the distribution function, namely the density and the pressure. One can define a temperature for the unperturbed profiles, since these unperturbed quantities are obtained from a Maxwellian equilibrium. However, one cannot say that the total pressure is $p=nT$, where $n=n_{\rm eq} + \tilde{n}$ and $T_{\rm eq} + \tilde{T}$. For this reason, the flattening of the profiles can only be analysed for the density and the pressure. An example of unexpected results that one can obtain when defining a perturbed temperature is evidenced when introducing a magnetic island in a plasma with flat temperature profile but finite density gradient. One expects that the temperature is not modified, since it is already flat. However, by defining the temperature as $T=p/n$ one obtains a modification which actually corresponds to the modifications of both density and pressure.

\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{pressflat.eps}
\caption{Left: 2D colormap of the pressure profiles for ions and electrons in the presence of a magnetic island. The isocontours of the vector potential are shown only with the electron colormap. Right: radial profiles of the pressure through the O (black line) and X (blue line) points.}
\label{fig:pressureflat}
\end{center}
\end{figure}

The total pressure gradient (background plus perturbed pressure) reads
\begin{equation}
{R_{\rm ref} \over L_{p_{\rm eq}+\tilde{p}}}  =  -{1 \over n_{R_0,s}T_s} {\partial (p_{s,\rm eq}+\tilde{p}_s) \over \partial \psi} = -{\partial\over\partial x_r}{\Delta p_{N,s} \over \rho_*} = 
 -{2\over 3}{\partial\tilde E_{N,s}\over\partial x_r} + {R_{\rm ref} \over L_{n,s}} + {R_{\rm ref} \over L_{T,s}}
\end{equation}

\section{Frozen electrons}

Due to the small inertia of electrons, they are considered as frozen particles to the magnetic field lines. This means in particular that the rotation frequency of electrons must equal the rotation
frequency of the island, namely
\begin{equation}
\omega_{\rm isl}\equiv\omega_{\rm TOT,e} = \omega_{*,e} + \omega_{E\times B}
\end{equation}
which constitutes the second benchmark after implementation of a magnetic island.

Therefore, the electric field must always compensate for the difference between the diamagnetic drift produced by any pressure gradient inside the island and the rotation of the island itself. In particular,
if the island does not rotate, for a sufficiently large island resulting in a complete flattening, the electrostatic potential must vanish inside the island. To verify the previous identity, one must
 conveniently quantify from the GKW outputs the diamagnetic frequency resulting from the total pressure profile and the $E\times B$ frequency.
These two quantities are derived hereafter assuming a slab geometry and considering that the binormal (in the $y$ direction) structure is determined by the island. The diamagnetic frequency in the $y$ direction reads
\begin{equation}
\omega_{*,s} = k_y{{\mathbf{b}\times\nabla p_s}\over{e_sn_sB}}\cdot\mathbf{e}_y\approx k_y{1\over{e_sn_{s,\rm eq}B}}{\partial p_s\over{\partial r}}
\end{equation}

Using the normalizations of GKW, we can write
\begin{equation}
\omega_{*,s} \approx k_y{1\over{e_{N,s}en_{R_0,s}B_NB_{\rm ref}}}{{n_{R_0,s}T_s\rho_*}\over{\rho_{\rm ref}}}{\partial\over{\partial x_r}}{\Delta p_{N,s} \over \rho_*}
\end{equation}

The temperature $T_s$ can now be expressed in terms of $T_{\rm ref}$ and $T_{R,s}$ and the reference Larmor radius used to write
\begin{equation}
\omega_{*,s} \approx k_y{1\over 2}v_{\rm th,ref}{T_{R,s}\over e_{N,s}B_N}\rho_*\left({2\over 3}{\partial\tilde E_{N,s}\over\partial x_r} - {R_{\rm ref} \over L_{n,s}} - {R_{\rm ref} \over L_{T,s}}\right)
\end{equation}

Under the assumption that the $y$ structure is determined by the island, one can write the wave number is that direction as $k_y=\pi/y_{\rm max}$, where $y_{\rm max}$ is the maximum value of the $y$ coordinate. Expressing $\rho_*$ as $\rho_*=\rho_{\rm ref}/R_{\rm ref}$ and normalizing the diamagnetic frequency to the transit frequency $\omega_t = v_{\rm th,ref}/R_{\rm ref}$ one gets the normalized diamagnetic frequency in slab geometry
\begin{equation}
\omega_{*,N,s} \approx {\pi\over y_{\rm max,N}}{1\over 2}{T_{R,s}\over e_{N,s}B_N}\left({2\over 3}{\partial\tilde E_{N,s}\over\partial x_r} - {R_{\rm ref} \over L_{n,s}} - {R_{\rm ref} \over L_{T,s}}\right)
\end{equation}

In a similar way, one can show that the $E\times B$ frequency in GKW units reads
\begin{equation}
\omega_{E,N,s} \approx {\pi\over y_{\rm max,N}}{1\over 2}{1\over B_N}{\partial\phi_{N}\over\partial x_r}
\end{equation}

Figure \ref{fig:freqsmatching} represents the profiles of the frequency for each species in the presence of an island rotating in the ion diamagnetic direction with a frequency $\omega_{\rm isl} = -0.006$. This frequency is represented by a solid black line, which must overlap the solid green line, corresponding to the total frequency of electrons. We observe this overlap everywhere inside the separatrix, but not around the X-point.

\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{freqsmatching.eps}
\caption{Left: 2D colormap of the frequency profiles for ions (up) and electrons (down) in the presence of a magnetic island with $R/L_n=0.7$ and $R/L_T=0$. Right: radial profiles of the frequency through the O-point (up) and binormal profiles of the frequency through the O and X points (down). Electrons are represented by green curves and ions by blue curves. The dotted lines represent the diamagnetic frequency and the solid lines the total frequency. For the radial profiles, the black dashed line represents the $E\times B$ frequency and the solid black line the rotation frequency of the island. For the binormal profiles, the dashed red line indicates the position of the X-point and the solid black line the rotation frequency of the island.}
\label{fig:freqsmatching}
\end{center}
\end{figure}

\section{Dependence of the electrostatic potential on the island rotation frequency}

The third and last benchmark needed after the implementation of a magnetic island constitutes the dependence of the electrostatic potential on the island rotation frequency. This dependence is calculated assuming that
 electrons inside the island short out any variation of the parallel electric field, i.e. we need to verify that $\nabla_\parallel\phi=-c^{-1}\partial_t \psi$ inside the island.
  For this we need to write an explicit expression for the parallel gradient, which is $\nabla_\parallel = \mathbf{b}\cdot\nabla$, where $\mathbf{b}=B^{-1}\mathbf{B}$ is the unit vector in the direction of the magnetic field.
  Let us perform the calculation, for the sake of simplicity, in sheared 2D slab geometry, where the magnetic field reads
  \begin{equation}
  \mathbf{B} = B_z\nabla z - \nabla\psi\times\nabla z = B_z\nabla z + {B_z x \over L_s}\nabla y - \tilde\psi k_y\sin\left(k_yy-\omega_{\rm isl}t\right)\nabla x
  \end{equation}

Under the assumption that the amplitude of the magnetic field is not significantly modified by the presence of the magnetic island and considering no derivatives in the $z$-direction (slab model), we can write
\begin{equation}
\nabla_\parallel = - {\tilde\psi k_y\over B_z}\sin\left(k_yy-\omega_{\rm isl}t\right)\partial_x + {x \over L_s}\partial_y
\end{equation}

The equation for the electrostatic potential is written as
\begin{equation}
{\tilde\psi k_y\over B_z}\sin\left(k_yy-\omega_{\rm isl}t\right){\partial\phi\over\partial x} - {x \over L_s}{\partial\phi\over\partial y} = {1\over c}{\partial \psi\over\partial t}
\end{equation}

Noticing that $x$ and $y$ coordinates must be related to each other on a flux surface for a given time $t$ by the flux label
\begin{equation}
\Omega = {x^2B_z \over 2\tilde\psi L_s} - \cos\left(k_yy-\omega_{\rm isl}t\right)
\end{equation}
we can rewrite the equation satisfied by the electrostatic potential as follows
\begin{equation}
- {x \over L_s}{\partial\phi\over\partial y} = {1\over c}{\partial \psi\over\partial t},\,\,\,\,\,{\rm for}\,\,d\Omega=0
\end{equation}
or equivalently
\begin{equation}
- {x \over L_s}\left.{\partial\phi\over\partial y}\right\vert_{\Omega} = {1\over c}{\partial \psi\over\partial t}
\end{equation}
The derivative of $x$ with respect to $y$, for $x\ne 0$, reads
\begin{equation}
{dx\over dy} = -{\tilde\psi L_s\over B_z x}\sin\left(k_yy-\omega_{\rm isl}t\right)
\end{equation}
and therefore, the equation satisfied by the electrostatic potential becomes, for $k_yy\not\equiv_\pi\omega_{\rm isl}t$
\begin{equation}
{k_y\over B_z}\left.{\partial\phi\over\partial x}\right\vert_{\Omega} = {\omega_{\rm isl}\over c}
\end{equation}
which gives, after integration
\begin{equation}
\phi = {\omega_{\rm isl}B_z\over k_y c}\left(x-h\left(\Omega\right)\right)
\end{equation}
where $h\left(\Omega\right)$ is a constant of integration for each flux surface. Note that the expression we have obtained for the electrostatic potential is only valid for $x\ne 0$ and $k_yy\not\equiv_\pi\omega_{\rm isl}t$. The function $h\left(\Omega\right)$ usually vanishes inside the island, corresponding to a complete flattening of radial profiles. Therefore, inside the separatrix and for $x\ne 0$ we can write, in GKW units
\begin{equation}
\phi_N = 2{\omega_{\rm isl,N}y_{\rm max,N}\over \pi}\Delta x_r
\end{equation}
where $\Delta x_r$ is the distance to the resonant surface. The extension to $x=0$ and $k_yy\equiv_\pi\omega_{\rm isl}t$ can be done by assuming that the electrostatic potential is continuous. Therefore, $\phi(x=0)=\lim_{x\rightarrow 0^\pm}\phi = 0$, $\phi(x=w,k_yy\equiv_\pi\omega_{\rm isl}t)=\lim_{x\rightarrow w^+}\phi = 2{\omega_{\rm isl,N}y_{\rm max,N}\over \pi}w$ and $\phi(x=-w,k_yy\equiv_\pi\omega_{\rm isl}t)=\lim_{x\rightarrow -w^-}\phi = -2{\omega_{\rm isl,N}y_{\rm max,N}\over \pi}w$. The left panel of figure \ref{fig:phiseparatrix} shows a remarkable agreement between the expected values for the electrostatic potential at the separatrix ($\Delta x_r = w$)
 from the previous expression (solid black line) and the values obtained from five simulations with GKW (red asterisks). These simulations have been performed with $w=10$, $y_{\rm max,N}=121.565$ and flat profiles.
\begin{figure}
\begin{center}
\includegraphics[width=0.4\textwidth]{phiseparatrix.eps}
\includegraphics[width=0.4\textwidth]{phiseparatrixw2.eps}
\label{fig:phiseparatrix}
\caption{Expected values of the electrostatic potential at the separatrix (solid black line) and values obtained from simulations with GKW (red asterisks). Left: $w=10$. Right: $w=2$}
\end{center}
\end{figure}

When the island width is comparable to or smaller than the ion Larmor radius, the ion pressure profile is little modified, whereas the electron pressure profile inside the island can still be modified.
 Of course, in the absence of background density and temperature profiles, this means that the ion pressure remains almost flat, whereas the electron pressure is only modified around the separatrix. However, this
  modification is of the order of the ion Larmor radius, meaning that for an island width of the order of the ion Larmor radius the electron pressure profile will be modified inside the island.
   In this case, the equation satisfied by the electrostatic potential should include the term related to the parallel gradient of the electrons pressure inside the island, namely
\begin{equation}
\nabla_\parallel\phi=-c^{-1}\partial_t \psi + {1\over en_e}\nabla_\parallel p_e
\end{equation}

Integrating this equation leads straightforwardly to the expression
\begin{equation}
\phi = {\omega_{\rm isl}B_z\over k_y c}\left(x-h\left(\Omega\right)\right) + {1\over en_e}p_e
\end{equation}
or equivalently, in GKW units
\begin{equation}
\phi_{N,\rm sep} = 2{\omega_{\rm isl,N}y_{\rm max,N}\over \pi}w + \left.{\Delta p_{N,s} \over \rho_*}\right\vert_{\rm sep}
\end{equation}

The right panel of figure \ref{fig:phiseparatrix} shows quite a good agreement, but some differences are observed. More analysis is needed for a complete understanding in the case of small islands.

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/manual/theory.tex
(1) Added gkw Clone gwk  #3

\chapter{Gyrokinetic theory}

\section{The gyrokinetic framework}
\label{sec:framework}
The starting point of the equation for the evolution of the distribution function is the 
gyrokinetic theory formulated in Refs \cite{LIT83,HAH88,BRI88,SUG00}. In GKW these equations 
are solved up to first order in the Larmor radius over the major radius $\rho_* \ll 1 $ 
(with the exception of the polarisation which enters in the field equations). 
In order to determine which of the terms in the original formulation of the theory (which is
accurate up to second order in $\rho_*$) must be kept, we will here briefly discuss the ordering. 
More details can be found in Ref.~\cite{PEE09}.

A thermal velocity ($v_{\rm th}$) and a major radius ($R$) are used to normalise the length and 
time scales. For the gradients we assume 
\begin{equation} 
R \nabla_\parallel  \approx 1 \qquad 
R \nabla_\perp \approx 1/ \rho_*,
\end{equation} 
i.e. the solution is assumed to have a length scale of the order of the system size along the 
field, but can have a length scale of the order of the Larmor radius perpendicular to the 
magnetic field. 
This difference in length scales means that the convecting velocities must be evaluated to a 
different order parallel and perpendicular to the field. 
The velocity along the field is to be evaluated in lowest order only (i.e. of the order of $v_{\rm th}$), but
the perpendicular velocity must be evaluated up to first order (order $\rho_* v_{\rm th}$). 
In GKW the perturbations in the electro-static potential $\phi$ as well as the vector potential ${\bf A}$ are 
retained.
For the latter it is, however, assumed that 
\begin{equation} 
{\bf A} = A_\parallel {\bf b} ,
\end{equation}
where ${\bf b}$ is the unit vector in the direction of the unperturbed magnetic field. 
The assumption above means that no compressional effects of the magnetic field are retained in the 
description.
Compressional effects are currently implemented in the code but are not yet described here.  
Note that the vector potential above is the perturbed vector potential not to be confused with the one 
connected with the equilibrium magnetic field.
The equations will be formulated in the co-moving system of a toroidally rotating plasma. 
The $E \times B$ drift velocity connected with this toroidal rotation is of the order of the thermal 
velocity in the laboratory frame, but the transformation to the co-moving frame removes this largest order $E \times B$ velocity from the 
equations \cite{PEE09,Casson-Thesis}. The transformation to the co-moving frame also introduces a background centrifugal potential $\Phi$.
The perturbed potential $\phi$, perturbed vector potential $A_\parallel$, and background potential $\Phi$
in the co-moving frame can then be taken to be of the order 
\begin{equation} 
\phi \approx {T \over e} \rho_* \qquad 
A_\parallel \approx {T \over e v_{\rm th}} \rho_*
\qquad
\Phi \approx {T \over e} \qquad 
\label{fieldordering}
\end{equation}
where $T$ is the temperature, and $e$ is the elementary charge. 
Since $\Phi$ is an equilibrium quantity, the length scale of any variation is assumed large compared
to the Larmor radius, whilst the perturbed quantities may have perpendicular gradient length scales on the order of the Larmor radius.  
The field gradients can then be shown to be of the same order, with $\nabla_\perp \phi \approx \nabla_\perp \Phi$.  
From these assumptions it can be derived that the $E \times B$ velocity is of the order $\rho_* v_{\rm th}$.

Gyro-centre coordinates (${\bf X},\mu,v_\parallel$) are used, with ${\bf X}$ being the gyro-centre position, $v_\parallel$ the parallel (to the magnetic field) velocity, and $\mu$ the magnetic moment $\mu = {m v_\perp^2 / 2 B} $.  
Here $v_\perp$ is the velocity component perpendicular to the equilibrium magnetic field ${\bf B}$, $m$
is the particle mass and $B$ is the magnetic field strength. 
The magnetic moment is an invariant of motion, but the parallel velocity changes due to the electric field 
acceleration and mirror force. 
It turns out that for consistency one needs to keep both the lowest as well as the first order (in $\rho_*$) 
contribution to the acceleration.

%FJC -> GSZ Add Bpar ordering here?

Finally, the approximation known as the $\delta f$ approximation is employed. 
This approximation assumes that the perturbed distribution $f$ is much smaller than the background distribution 
$F$, but has a perpendicular length scale of the order of the Larmor radius whereas the background varies 
on the length scale of the system size:
\begin{equation} 
f \approx \rho_* F \qquad 
{\partial f \over \partial v_\parallel} \approx \rho_* {\partial F \over \partial v_\parallel}  \qquad 
\nabla f \approx \nabla F .
\end{equation}
The ordering assumption above removes the `velocity non-linearity' (combinations of the 
electro-magnetic fields and the velocity space derivative of the perturbed distribution) from
the evolution equation of the perturbed distribution. 
The disadvantage is  that strict energy conservation no longer applies.

With the help of these approximations the gyrokinetic equation of Refs. \cite{LIT83,BRI88,SUG00}
can be rewritten.

\section{Lagrangian}
\label{sec:theoryLagrangian}
(This section is not in the CPC paper which instead refers to Ref. \cite{PEE09} which contains the full derivation with centrifugal effects.)
(CHECK CONSISTENCY)

The starting point is the Lagrangian \index{Lagrangian}
\begin{equation}
  \gamma=\left({e\over c}{\bf A}+{e\over c}A_{\parallel}{\bf b}+m{\bf v} \right)\cdot d{\bf x}-\left({m\over 2}v^{2}+e\phi \right)dt
\end{equation}
The coordinates are transformed into guiding center coordinates. The electrostatic guiding center
Lagrangian is known, and can be found in Hahm 1988 \cite{HAH88}. For the magnetic perturbations we take that
\begin{equation}
  A_{\parallel}({\bf X}+{\bf \rho}){\bf b} \cdot d({\bf X}+\rho)
\end{equation}
where $\rho =\rho({\bf X},{\mu},{\theta})$. All the derivatives of $\rho$ will give vectors
perpendicular to the magnetic field. Thus if there is not a perpendicular magnetic potential
perturbation, the only term that survives is
$$A_{\parallel}{\bf b}\cdot d{\bf X}$$
Then the guiding center Lagrangian is
\begin{equation}
\gamma =\left({e\over c}{\bf A}+{e\over c}A_{\parallel}{\bf b}+mv_{\parallel}{\bf b} \right)\cdot 
d{\bf X}+\mu d\theta -\left({m\over 2}v_{\parallel}^{2}+\mu B+e\phi \right)dt
\end{equation}
This Lagrangian is transformed in the gyrocenter coordinates with the Lie transforms.
To the first order, if there are no perpendicular perturbations in the magnetic potential,
we can just substitute the perturbations of the fields with the respective gyroaveraged quantities. 
This can also be derived rigorously with the Lie transforms. The new Lagrangian will be
\begin{equation}
\Gamma =\left({e\over c}{\bf A}+{e\over c}\ga{A_{\parallel}}{\bf b}+mv_{\parallel}{\bf b} \right)\cdot 
d{\bf X}+\mu d\theta -\left({m\over 2}v_{\parallel}^{2}+\mu B+e\ga{\phi} \right)dt
\label{eq:gyrocenterlagrangian}
\end{equation}
where now ${\bf X},\mu,v_{\parallel},\theta$ are the gyrocenter coordinates. From this one can derive 
the equations of motion using the Euler Lagrange equations
\begin{equation}
\omega_{ij}{dz^{j}\over dt}={\partial H \over \partial z^{i}}-{\partial \gamma_{i}\over \partial t}
\end{equation}
FIXME what is $\mathrm{d}\gamma$ or $\partial\gamma$ supposed to mean if $\gamma$ is already a differential (cf. \eqref{eq:gyrocenterlagrangian})?
where $H$ is the Hamiltonian and 
\begin{equation}
\omega_{ij}={\partial \gamma_{j}\over \partial z^{i}}-{\partial \gamma_{i}\over \partial z^{j}}
\end{equation}
are the Lagrange brackets.

The Lagrange brackets that are of interest are
\begin{equation}
\omega_{X_{i}X_{j}}={e\over c}\epsilon_{ijk}{\hat B}_{k}
\end{equation}
\begin{equation}
\omega_{{\bf X}U}=-m{\bf b}
\end{equation}
where 
\begin{equation}
{\hat {\bf B}}=\nabla \times {\bf A}+\nabla \times A_{\parallel}{\bf b}+{mc \over e}v_{\parallel}\nabla \times {\bf b}
\end{equation}

Choosing $i=U$ we get that
\begin{equation}
m{\bf b}\cdot {\dot {\bf X}}=mv_{\parallel}
\end{equation}
Choosing $i={\bf X}$ we get that
\begin{equation}
m{\dot v}_{\parallel}{\bf b}-{e\over c}{\dot {\bf X}}\times {\hat{\bf B}}=-e\nabla \ga{\phi}-\mu \nabla B-{e\over c}
{\partial \ga{A_{\parallel}}\over \partial t}{\bf b}
\end{equation}
and taking that
\begin{equation}
{\bf E}=-\nabla \phi -{1\over c}{\partial \ga{A_{\parallel}}\over \partial t}{\bf b}
\end{equation}
we get that
\begin{equation}
m{\dot v}_{\parallel}{\bf b}-{e\over c}{\dot {\bf X}}\times {\hat{\bf B}}=e{\bf E}-\mu \nabla B
\end{equation}

We can cross this equation with ${\bf b}$ to derive the equation for $\dot X$. To derive the equation of GKW we have
to approximate
\begin{equation}
\nabla \times \ga{ A_\parallel} {\bf b}
= 
\nabla \ga{A_\parallel}\times\mathbf{b} + \ga{A_\parallel}\nabla\times\mathbf{b}
 \approx - {\bf b} \times \nabla \langle A_\parallel \rangle 
\label{eq:approxtofindExBdrift}
\end{equation}
Then the equation of $\dot X$ is.
\begin{equation} 
{{\rm d} {\bf X} \over {\rm d} t} = v_\parallel {\bf b} + {\bf v}_D + {\bf v}_E + {\bf v}_{\delta B_\perp}, 
\end{equation}
To derive the equation for $v_{\parallel}$ we dot with $\dot X$. This gives
\begin{equation}
mv_{\parallel}{\dot v}_{\parallel}={\dot {\bf X}}\cdot \left(e{\bf E}-\mu \nabla B\right)
\end{equation}

\section{The gyrokinetic equation}

In a rigidly rotating frame, 
the evolution of equation for the gyro-centre position ${\bf X}$, and the parallel velocity are \cite{PEE09}
\begin{equation} 
{{\rm d} {\bf X} \over {\rm d} t} = v_\parallel {\bf b} + {\bf v}_D + {\bf v}_\chi ,
\qquad
\label{eq:drifts}
%\end{equation}
%\begin{equation} 
m v_\parallel {{\rm d} v_\parallel \over {\rm d} t} =   
{{\rm d} {\bf X} \over {\rm d} t} \cdot \left[ Z {e }{\bf E}  - \mu 
 \nabla B + m\Omega^2 R \nabla R \right],
%\end{equation}
%where $\mu$ is a constant of motion. 
%\begin{equation} 
\qquad 
{{\rm d} \mu \over {\rm d} t} = 0 .
\end{equation}
Here $Z$ is the particle charge, $R$ the local major radius, and $\Omega$ the frame rotation frequency.  
$\bf E$ is the gyro-averaged perturbed electric field plus the inertial electric field
\begin{equation} 
{\bf E} = -\nabla \langle \phi \rangle  - {\partial \langle A_\parallel \rangle \over \partial t} {\bf b}  -\nabla \langle \Phi \rangle ,
\end{equation}
where the angle brackets denote the gyro-average. 
Since $\Phi$ is an equilibrium quantity, the scale lengths of any variation are assumed large compared
to the Larmor radius and we neglect the gyroaverage, taking $\Phi \approx \langle\Phi\rangle$.  (The 
full details of the calculation of $\Phi$ are given in Sec. \ref{cfphi}).

The velocities in \eq{drifts} are from left to right: the parallel motion along the 
unperturbed field ($v_\parallel {\bf b}$), the drift motion due to the inhomogeneous field 
(${\bf v}_D$), and 
\begin{equation} 
\label{eq:v-chi}
{\bf v}_\chi = {{\bf b} \times \nabla \chi \over B} \qquad 
\textrm{with} \qquad \chi = \langle \phi \rangle - v_\parallel \langle A_\parallel \rangle ,
\end{equation}
which is the combination of the $E \times B$ velocity 
(${\bf v}_E = {\bf b} \times \nabla \langle \phi \rangle / B$) and the parallel motion along 
the perturbed field line (${\bf v}_{\delta B} = - {{\bf b} \times \nabla v_\parallel 
\langle A_\parallel \rangle /B}$).
The drifts due to the inhomogeneous magnetic field and inertial terms can be written in the form
\begin{equation}
\label{eq:driftsinhomogeneousinterial}
{\bf v}_D = 
{1\over Ze} \biggl [ {m v_\parallel^2\over B} + \mu \biggr ] {{\bf B} \times \nabla B \over B^2}  
+ {m v_\parallel^2 \over 2 Z e B} \beta^\prime {\bf b} \times \nabla \psi
+ { 2 m v_\parallel \over Z e B } {\bf \Omega}_\perp + {1 \over ZeB} {\bf b} \times \nabla \cfen
\end{equation}
In the equation above $\psi$ is a radial coordinate (flux label) and ${\bf \Omega}_\perp$ is the angular
(toroidal) rotation vector perpendicular to the field, i.e. ${\bf \Omega}_\perp = {\bf \Omega} 
- ({\bf \Omega } \cdot {\bf b}) {\bf b}$.
The first term on the right is the combination of the grad-B drift and curvature drift 
in the low beta approximation, whereas the second term, involving
\begin{equation}
\label{eq:defbetapr}
\beta' = \frac{2\mu_0}{B^2}\frac{\partial p}{\partial \psi},
\end{equation}
is the correction to the 
curvature drift due to the modification of the equilibrium associated with the pressure
gradient. The penultimate term is the Coriolis drift derived in Ref.~\cite{PEE07} 
using the formulation of the gyrokinetic equations in the co-moving frame \cite{BRI95}.
The formulation of Ref.~\cite{PEE07} was extended in Refs.~\cite{PEE09,Casson-Thesis} to include 
the centrifugal force, and the first centrifugal results were published in  Refs.~\cite{CAS10,Casson-Thesis}.  
The last term combines the centrifugal drift and background potential 
in the (species dependent) centrifugal energy
\begin{equation}
\cfen=Ze\Phi - {1 \over 2} m \Omega^2 (R^2-R_0^2)
\end{equation}
where $R_0$ is the major radius of the point on the flux surface at which the densities are defined.

From the equations above the gyrokinetic equation in (${\bf X}, v_\parallel, 
\mu$) coordinates can be derived: \cite{LIT83,PEE09}
\begin{equation} 
{\partial f_{\rm tot} \over \partial t} + {{\rm d} {\bf X} \over {\rm d} t} \cdot \nabla_\mu 
f_{\rm tot} + {{\rm d} v_\parallel \over {\rm d} t} {\partial f_{\rm tot} \over \partial v_\parallel} = 0 .
\end{equation} 
The distribution function $f_{\rm tot}$ in this equation is split in a background $F$ and a perturbed 
distribution $f$. As discussed in Section \ref{sec:framework} the $\delta f$ approximation 
will be employed. 
Consequently, the equation for $f$ can be written in the form \cite{PEE09}
\begin{equation} 
{\partial f \over \partial t} + (v_\parallel {\bf b} + {\bf v}_D + {\bf v}_\chi) \cdot \nabla f  
-{{\bf b} \over m}\cdot(\mu \nabla B + \nabla \cfen){\partial f \over \partial v_\parallel} = S. 
\end{equation}
where $S$ is determined by the background distribution function.
The latter is assumed to be a Maxwellian ($F_M$), with
temperature ($T$) and mean parallel velocity ($u_\parallel$) being functions of the radial coordinate only.
In the case of a rotating plasma the density in the flux surface varies as $n(\theta)=n_{R_0} \exp(-\cfen(\theta)/T)$
and is kept in the solution of the equilibrium equation \cite{PEE09}.
\begin{equation} 
\label{eq:maxwell}
F_M = {n_{R_0} \over \pi^{3/2} v_{\rm th}^3 } \exp 
\biggl [ - {(v_\parallel-(R B_t/B)\omega_\phi )^2 + 2 \mu B / m \over v_{\rm th}^2} - \cfen/T \biggr ] . 
\end{equation} 
Here $v_{\rm th}\equiv \sqrt{ 2 T / m}$ is the thermal velocity (note that in this area of 
research the thermal velocity is usually defined without the $\sqrt{2}$ factor).
The appearance of an explicit rotation velocity $u_\parallel$ in the Maxwellian may be confusing since the equations are formulated
in the co-moving system in which the rotation vanishes. 
However, under experimental conditions the plasma rotation $\omega(\phi)$ has a radial gradient while the frame is chosen
to rotate as a rigid body with a constant angular frequency $\Omega$. 
Indeed a differential rotation of the frame would be impractical since the distance between two
fixed points in the co-moving 
system would increase in time, i.e. the metric would be time dependent.

The local model constructed here chooses the rotation of the frame to be equal to the plasma rotation at one particular radius. 
Therefore $\omega_\phi$ in the above Maxwellian is zero at the radial location considered, but has a finite radial gradient. This 
gradient will enter the equations as $\nabla \omega_\phi$. (for more details see Ref.~\cite{PEE09}).

The term containing the derivative of the vector potential towards time presents a numerical difficulty. 
In the $\delta f$ formalism it can, however, easily be combined with the time derivative of $f$. If one defines a `new' distribution $g$ 
\begin{equation} 
g = f + {Z e \over T} v_\parallel \langle A_\parallel \rangle F_M ,
\end{equation}
and replaces the time derivative of $f$ in the gyrokinetic equation with the time derivative of $g$, the cumbersome 
time derivative of $\langle A_\parallel \rangle $ is removed. 
Using the equations above one arrives at the following equation for the distribution function $g$
\begin{equation} 
{\partial g \over \partial t} + {\bf v}_\chi \cdot \nabla g  + (v_\parallel {\bf b} + {\bf v}_D) \cdot \nabla f  
-{{\bf b} \over m}\cdot(\mu \nabla B + \nabla \cfen){\partial f \over \partial v_\parallel} = S. 
\label{Vlasov}
\end{equation}
where $S$ is given by
\begin{equation}
S =  - ({\bf v}_\chi + {\bf v}_D) \cdot \biggl [ {\nabla n_{R_0} \over n_{R_0}} + \biggl ( {v_\parallel^2 \over v_{\rm th}^2 } 
+ {(\mu B + \cfen) \over T} - {3 \over 2} \biggr ) {\nabla T \over T} + {m v_\parallel \over T} {R B_t \over B }\nabla \omega_\phi \biggr ] F_M -  
{Ze \over T} [ v_\parallel {\bf b} + {\bf v}_D ] \cdot \nabla \langle \phi \rangle  F_M .
\label{source} 
\end{equation}
Note that both $g$ and $f$ appear in this equation. Since any form of dissipation on the field variables 
leads to a numerical instability \cite{CAN03}, it is preferable to use $f$ in any term that requires 
dissipation for stable numerical implementation.  
It should also be noted here that the Maxwellian is not an exact solution of the gyrokinetic equation in the limit of a zero 
perturbed electro-magnetic field. 
From the equation above it follows that $S$ is nonzero due to the drifts connected with the magnetic field inhomogeneity 
in the gradients of density temperature and velocity. 
The solution of the evolution equation will determine the deviation away from the Maxwellian as a nonzero perturbed
distribution $f$. 
This perturbation is of the order $\rho_* F_M$, and neglecting it as part of the background distribution is consistent with 
the ordering adopted. 
The drift term is ${\bf v}_D$ in the first term of the equation above is nevertheless implemented in GKW since the perturbation 
it generates is responsible for the neo-classical transport. All physics and diagnostic of the neo-classical fluxes have been 
implemented in the code, but the implementation has not been benchmarked and should therefore not be used.

To close the set of equations, one needs to calculate the potential and vector potential through the (low 
frequency ordering) Maxwell equations. These have been formulated in the literature, and 
will be given in Section \ref{eqs:complete-set} when the full set of equations in normalised form is discussed.

\section{Fully electromagnetic gyrokinetic equation}
In high beta tokamak plasmas the amplitude of the magnetic perturbations can become significant compared to the fluctuations of the electro-static potential. In such cases electro-magnetic effects should be taken into account in the Vlasov equation for an accurate gyrokinetic simulation. In this section the gyrokinetic Lagrangian, equations of motion, the Vlasov and Maxwell equations are summarized in the fully electro-magnetic, collisionless case in a rotating frame of reference. The detailed derivation of the equations can be found in \doc{tex/derivation.tex}.

The gyrokinetic Lagrangian is
\begin{eqnarray}
	\bar{\Gamma} &=& \left( m v_{\parallel} \mathbf{b}(\mathbf{X}) + Z e \mathbf{A}_0(\mathbf{X}) + m \mathbf{u}_0 + Z e \langle \mathbf{A}_1 \rangle(\mathbf{X}) \right) \cdot \mathrm{d} \mathbf{X} + \frac{2 \mu m}{Z e} \mathrm{d} \theta - \nonumber\\
	&& \left( \frac{1}{2} m \left(v_{\parallel}^2 - u_0^2 \right) + Z e \left(\Phi_0(\mathbf{X}) + \langle \phi \rangle(\mathbf{X}) \right) + \mu \left(B_0(\mathbf{X}) + \langle B_{1 \parallel} \rangle(\mathbf{X})\right) \right) \mathrm{d}t
\end{eqnarray} 
where $\langle \phi \rangle = J_0(\lambda) \phi$, $\langle \mathbf{A}_1 \rangle = J_0(\lambda) \mathbf{A}_1$ and $\langle B_{1 \parallel} \rangle = \hat{J}_1(\lambda) B_{1 \parallel}$ are the gyroaveraged field perturbations. $\mathbf{b}$ is the unit vector along the equilibrium magnetic field, $\mathbf{u}_0$ is the velocity of the rotating frame of reference, and $\Phi_0$ is the equilibrium electro-static potential due to the plasma rotation. The bar over $\bar{\Gamma}$ distinguishes the gyro-centre Lagrangian from its guiding-centre version.

The equations of motion can be derived from the Euler--Lagrange equations and are written as
\begin{eqnarray}
	\dot{\mathbf{X}} &=& \mathbf{b} v_{\parallel} + \frac{m v_{\parallel}^2}{Z e B_0} \left( \nabla \times \mathbf{b} \right)_{\perp} + \frac{\langle \mathbf{B}_{1 \perp} \rangle} {B_0} v_{\parallel} - \frac{1}{B_0} \langle \mathbf{E} \rangle \times \mathbf{b} + \nonumber \\
	&& \frac{\mu}{Z e B_0} \nabla \left( B_0 + \langle B_{1 \parallel} \rangle \right) \times \mathbf{b} + \frac{2 m v_{\parallel}}{Z e B_0} \mathbf{\Omega}_{\perp} - \frac{m R \Omega^2}{Z e B_0} \nabla R \times \mathbf{b} \nonumber \\
	&=& \mathbf{b} v_{\parallel} + \underbrace{\mathbf{v}_{\langle \mathbf{B}_{1 \perp} \rangle} + \mathbf{v}_{\langle \mathbf{E} \rangle \times \mathbf{B}_0} + \mathbf{v}_{\nabla \langle B_{1\parallel} \rangle}}_{\mathbf{v}_{\chi}} + \underbrace{\mathbf{v}_{\st{C}} + \mathbf{v}_{\nabla \mathbf{B}_0} + \mathbf{v}_{\st{co}} + \mathbf{v}_{\st{cf}}}_{\mathbf{v}_{\st{D}}}
\end{eqnarray}
where $\mathbf{\Omega}$ is the angular momentum  vector associated with the rotation of the reference frame, and
\begin{equation}
	\dot{v}_{\parallel} = \frac{\dot{\mathbf{X}}}{m v_{\parallel}} \cdot \left( Z e \langle \mathbf{E} \rangle - \mu \nabla (B_0 + \langle B_{1 \parallel} \rangle) + \frac{1}{2}m \nabla u_0^2 \right).
\end{equation}
The quantity $\chi$, related to the drifts due to perturbations of the fields, can be now written as
\begin{equation}
\chi = \underbrace{\Phi_0 + \langle \phi \rangle}_{\langle \Phi \rangle} - v_{\parallel} \langle {A}_{1 \parallel} \rangle + \frac{\mu}{Z e} \langle B_{1 \parallel} \rangle 
\end{equation}
and
\begin{equation}
\mathbf{v}_{\chi} = \frac{\mathbf{b} \times \nabla \chi}{B_0} = \mathbf{v}_{\langle \mathbf{B}_{1 \perp} \rangle} + \mathbf{v}_{\langle \mathbf{E} \rangle \times \mathbf{B}_0} + \mathbf{v}_{\nabla \langle B_{1 \parallel} \rangle}. 
\end{equation}

The Vlasov equation in presence of magnetic field compression takes the form
\begin{equation}
	\frac{\partial g}{\partial t} + \mathbf{v}_{\chi} \cdot \nabla g + \left( v_{\parallel} \mathbf{b} + \mathbf{v}_{\st{D}} \right) \cdot \nabla \delta f - \frac{\mu}{m} \mathbf{b} \cdot \left(\nabla \Phi_0 - m R \Omega^2 \nabla R + \nabla B_0 \right) \frac{\partial \delta f}{\partial v_{\parallel}} = S 
	\label{eq:vlasov}
\end{equation}
with the source term being
\begin{equation}
	S = - \left( \mathbf{v}_{\chi} + \mathbf{v}_{\st{D}} \right) \cdot \nabla_p F_M + \frac{F_M}{T} \left( v_{\parallel} \mathbf{b} + \mathbf{v}_{D} \right) \cdot \left( - Z e \nabla \langle \phi \rangle - \mu \nabla \langle B_{1 \parallel} \rangle \right)
	\label{eq:vlasov-source}
\end{equation}
where $g$ is the modified distribution function
\begin{equation}
	g = \delta f + \frac{Z e v_{\parallel}}{T} \langle A_{1 \parallel} \rangle F_{\st{M}}.
	\label{eq:g}
\end{equation}

Equations \ref{eq:vlasov} and \ref{eq:vlasov-source} show that there
are four new terms appearing due to the inclusion of
$B_{1 \parallel}$. These are related to the convection of the
perturbed distribution function in phase space due to the grad-B drift
in the fluctuating magnetic field (non-linear), the same convection of
the equilibrium distribution function (linear), and a couple of
additional mirror force terms in the direction of the parallel and
drift velocities (linear). The normalised form of these terms can be
found in section \ref{equations}. The former two is included in the
modified definition of $\chi$ appearing in terms III and V (equations
\ref{eq:nl-term} and \ref{eq:V}), the latter two are explicitly written in
terms X and XI (equations \ref{eq:X} and \ref{eq:XI}).

\subsection{Gyrokinetic field equations} 
\label{field-equations}
The normalised gyrokinetic field equations in presence of $B_{1 \parallel}$ are listed here. The normalizing assumptions can be found in section \ref{sec:normalisation}. The equations are written in Fourier-space, the Fourier components of the fields and the distribution function are denoted by $\hat{(.)}$. $\Gamma_0$ and $\Gamma_1$ are modified Bessel-functions, $b_{\st{sp}}$ denotes their species dependent attributes in the Fourier-space.

The quasi-neutrality equation is
\begin{eqnarray}
	\hspace{-5mm}
	&& \sum_{\st{sp}} Z_{\st{sp}} n_{\st{R,sp}} \left[ 2 \pi B_{\st{N}} \int J_0(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d} v_{\parallel N} \mathrm{d} \mu_{\st{N}} + \right. \nonumber \\ 
	\hspace{-5mm}
	&& \left. \frac{Z_{\st{sp}}}{T_{\st{R,sp}}} \hat{\phi}_{\st{N}} \left( \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \Gamma_0(b_{\st{sp}}) - 1 \right) + \frac{\hat{B}_{1 \parallel \st{N}}}{B_{\st{N}}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} (\Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}})) \right] = 0
	\label{eq:poisson}
\end{eqnarray}
where $\mathcal{E}$ is a combined energy term including the kinetic energy of the toroidal rotation of the plasma and the energy stored in the equilibrium electric field:
\begin{equation}
 \mathcal{E} = Z e \langle \Phi_0 \rangle - \frac{1}{2} m \omega_{\varphi}^2 (R^2-R_0^2).
\end{equation}
The parallel component of Amp\`ere's law gives
\begin{eqnarray}
	&& \left[ k_{\perp \st{N}}^2 + \beta_{\st{ref}} \sum_{\st{sp}} \frac{Z_{\st{sp}}^2 n_{\st{R,sp}}}{m_{\st{R,sp}}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \Gamma_0(b_{\st{sp}}) \right] \hat{A}_{1 \parallel \st{N}} = \nonumber \\
	&& 2 \pi B_{\st{N}} \beta_{\st{ref}} \sum_{\st{sp}} Z_{\st{sp}} n_{\st{R,sp}} v_{\st{R,sp}} \int v_{\parallel N} J_0(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d} v_{\parallel N} \mathrm{d} \mu_{N}
	\label{eq:apar} \label{eq:AmpPar}
\end{eqnarray}
where $\beta_{\st{ref}}$ is the reference plasma beta:
\begin{equation}
	\beta_{\st{ref}} = \frac{2 \mu_0 n_{\st{ref}} T_{\st{ref}}}{B_{\st{ref}}^2}.
	\label{eq:beta}
\end{equation}
The perpendicular component of Amp\`ere's law gives
\begin{eqnarray}
	&& \hspace{-1cm} \left[ 1 + \sum_{\st{sp}} \frac{T_{\st{R,sp}} n_{\st{R,sp}}}{B_{\st{N}}^2} \beta_{\st{ref}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \left( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) \right) \right] \hat{B}_{1 \parallel \st{N}} = \nonumber \\
	&& \hspace{1cm} - \sum_{\st{sp}} \beta_{\st{ref}} \left[ 2 \pi B_{\st{N}} T_{\st{R,sp}} n_{\st{R,sp}} \int \mu_{N} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d}v_{\parallel N} \mathrm{d} \mu_{\st{N}} \right. \nonumber\\
	&& \hspace{1cm} + \left. \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \left( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) \right) \frac{Z_{\st{sp}} n_{\st{R,sp}}}{2 B_{\st{N}}} \hat{\phi}_{\st{N}} \right].
	\label{eq:bpar} \label{eq:AmpPerp}
\end{eqnarray}
where $\hat{J}_1(z) = \frac{2}{z} J_1(z)$ is a modified first order Bessel function of the first kind.

\subsection{Field equations in GKW} \label{sec:gkw}
The gyrokinetic Poisson equation \ref{eq:poisson} and perpendicular Amp\`ere's law \ref{eq:bpar} are coupled through the fluctuating electro-static potential and magnetic compression appearing in both of them. For simpler numerical treatment the two equations have to be decoupled. By introducing the notations
\begin{eqnarray*}
	I_{1 \st{sp}} &=& Z_{\st{sp}} n_{\st{R,sp}} 2 \pi B_{\st{N}} \int J_0(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d} v_{\parallel N} \mathrm{d} \mu_{\st{N}} \\
	F_{1 \st{sp}} &=& \frac{Z^2_{\st{sp}} n_{\st{R,sp}}}{T_{\st{R,sp}}} \left( \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} \Gamma_0(b_{\st{sp}}) - 1 \right) \\
	B_{1 \st{sp}} &=& \frac{Z_{\st{sp}} n_{\st{R,sp}}}{B_{\st{N}}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} (\Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}})) \\
	I_{2 \st{sp}} &=& \beta_{\st{ref}} 2 \pi B_{\st{N}} T_{\st{R,sp}} n_{\st{R,sp}} \int \mu_{N} \hat{J}_1(k_{\perp} \rho_{\st{sp}}) \hat{g}_{\st{N,sp}} \mathrm{d}v_{\parallel N} \mathrm{d} \mu_{\st{N}} \\
	F_{2 \st{sp}} &=& \frac{\beta_{\st{ref}} Z_{\st{sp}} n_{\st{R,sp}}}{2 B_{\st{N}}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} ( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) )  \\
	B_{2 \st{sp}} &=& \frac{T_{\st{R,sp}} n_{\st{R,sp}}}{B_{\st{N}}^2} \beta_{\st{ref}} \st{e}^{\frac{-\mathcal{E}_{\st{N,sp}}}{T_{\st{R,sp}}}} ( \Gamma_0(b_{\st{sp}}) - \Gamma_1(b_{\st{sp}}) )
\end{eqnarray*}
equations \ref{eq:poisson} and \ref{eq:bpar} take the form
\begin{eqnarray*}
	\sum_{\st{sp}} I_{1 \st{sp}} + \hat{\phi}_{\st{N}} \sum_{\st{sp}} F_{1 \st{sp}} + \hat{B}_{1 \parallel \st{N}} \sum_{\st{sp}} B_{1 \st{sp}} &=& 0 \\
	\left( 1 + \sum_{\st{sp}} B_{2 \st{sp}} \right) \hat{B}_{1 \parallel \st{N}} + \sum_{\st{sp}} I_{2 \st{sp}} + \hat{\phi}_{\st{N}} \sum_{\st{sp}} F_{2 \st{sp}} &=& 0.
\end{eqnarray*}
Accordingly, the decoupled field equations for $\hat{\Phi}_1$ and $\hat{B}_{1 \parallel \st{N}}$ can be written as
\begin{eqnarray}
	&& \left[ \sum_{\st{sp}} F_{1 \st{sp}} \left(1 + \sum_{\st{sp}} B_{2 \st{sp}} \right) - \sum_{\st{sp}} F_{2 \st{sp}} \sum_{\st{sp}} B_{1 \st{sp}} \right] \hat{\phi}_{\st{N}} + \nonumber \\
	&& \left[ \sum_{\st{sp}} I_{1 \st{sp}} \left(1 + \sum_{\st{sp}} B_{2 \st{sp}} \right) - \sum_{\st{sp}} I_{2 \st{sp}} \sum_{\st{sp}} B_{1 \st{sp}} \right] = 0
	\label{eq:poisson-dec}
\end{eqnarray}
and
\begin{eqnarray}
	&& \left[ \sum_{\st{sp}} F_{1 \st{sp}} \left(1 + \sum_{\st{sp}} B_{2 \st{sp}} \right) - \sum_{\st{sp}} F_{2 \st{sp}} \sum_{\st{sp}} B_{1 \st{sp}} \right] \hat{B}_{1 \parallel \st{N}} + \nonumber \\
	&& \left[ \sum_{\st{sp}} I_{2 \st{sp}} \sum_{\st{sp}} F_{1 \st{sp}} - \sum_{\st{sp}} I_{1 \st{sp}} \sum_{\st{sp}} F_{2 \st{sp}} \right] = 0.
	\label{eq:bpar-dec}
\end{eqnarray}
The implementation of the field equations in GKW follows the above notation.

%%Please see Appendix A for full derivation of the feild equations

% (Optional Section - NOT IN CPC PAPER, NEED TO CHECK CONSISTENCY)
% (YC: why not rather use here the derivation done by Gabor, which is consistent and done within the same formalism as the GK equations?)
% The quasi-neutrality equation (also referred to as Poisson equation) and the Amp\`{e}re Law are used to calculate the perturbed electrostatic potential $\phi$ and vector potential $A_\parallel$ from
% the perturbed distribution function. These equations apply in real space while the perturbed distribution function $f$ is calculated in the gyrocenter phase space. We therefore need to express the
% distribution function in the real phase space $f^{*}_{tot}(\mathbf{x};\mu,v_\parallel,\gamma)$ as a function of the distribution in the gyrocenter phase space $f_{tot}(\mathbf{X};\mu,v_\parallel)$:
% \index{field equations}
% \begin{eqnarray}
% \label{eq:f-real}
%  f^{*}_{tot}(\mathbf{x};\mu,v_\parallel,\gamma) &=& f_{b}(\mathbf{X};\mu,v_\parallel) + f(\mathbf{X};\mu,v_\parallel) \\
%  &+& \frac{Ze}{B}\frac{\partial f_{b}^{*}}{\partial \mu}(\mathbf{x}) \left[ \phi(\mathbf{x}) - \left< \phi\right>(\mathbf{X}) - \frac{v_\parallel}{c}\left(A_\parallel(\mathbf{x}) - \left<
% A_\parallel\right>(\mathbf{X})\right) \right] \nonumber
% \end{eqnarray}
% In this expression, $<.>$ denotes the gyroaverage and $\mathbf{X}$ is the gyrocenter of the particle in $\mathbf{x}$ with velocity coordinates $(\mu,v_\parallel,\gamma)$, where $\gamma$ is the
% gyroangle. The total distribution functions (subscript \textit{tot}) have been split as usual in a background part (subscript \textit{b}) and a perturbed part (no subscript) with the assumption that
% the background distribution is approximately constant over a gyro-orbit. The three contributions to the real space distribution function in Eq.~(\ref{eq:f-real}) are therefore, in that order, the
% background, the gyroaveraged perturbed part and the gyroangle dependent perturbed part (which arises from the variation of the potential and vector potential over a gyro-orbit). In the following,
% the background distribution is taken to be a Maxwellian: 
% $$f_{b}^{*}(\mathbf{x};\mu,v_\parallel,\gamma) = f_{b}(\mathbf{X};\mu,v_\parallel) = F_M(\mathbf{X};\mu,v_\parallel)$$
% Using for the potential the local limit approximation already used in Eq.~(\ref{eq:local-limit}) for the distribution function, one can write:
% \begin{equation}
%  \phi(\mathbf{x}) = \sum_\mathbf{k_\perp} \phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}} = \sum_\mathbf{k_\perp}
% \phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}}\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{\rho}}
% \end{equation}
% where $\mathbf{x} = \mathbf{X} + \mathbf{\rho}$  has been used, $\mathbf{\rho}$ being the gyroradius vector. Note that $\phi_{k_\perp}(\mathbf{X})$ is a slowly varying amplitude that can be
% apporximated by $\phi_{k_\perp}(\mathbf{x})$. The gyro-average leads to 
% \begin{equation}
% \label{eq:phi-gyro}
%  \left<\phi\right>(\mathbf{X}) = \sum_\mathbf{k_\perp}
% \phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}}\left<\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{\rho}}\right> = \sum_\mathbf{k_\perp}
% \phi_{k_\perp}(\mathbf{X})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}}J_0(k_\perp\rho)
% \end{equation}
% where the Bessel function $J_0$ arises from: 
% \begin{equation}
%  \left<\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{\rho}}\right> = \frac{1}{2\pi}\oint \textrm{e}^{ik_\perp\rho\cos(\gamma-\gamma_{\rm{\rm ref}})} \,\textrm{d}\gamma= J_0(k_\perp\rho)
% \end{equation}
% From Eq.~(\ref{eq:f-real}) and Eq.~(\ref{eq:phi-gyro}) one can then calculate the density in real space:
% \begin{eqnarray*}
% n(\mathbf{x}) &=& \int f^{*}_{tot}(\mathbf{x};\mu,v_\parallel,\gamma) B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma \\
%  &=& \int f_{b}(\mathbf{X};\mu,v_\parallel) B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma + \int \sum_\mathbf{k_\perp} 
% f_{k_\perp}(\mathbf{x};\mu,v_\parallel)\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}}  \textrm{e}^{-i\mathbf{k_\perp}\cdot\mathbf{\rho}} B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma \\
%  &-& \frac{Ze}{T} \int F_M(\mathbf{X};\mu,v_\parallel) \sum_\mathbf{k_\perp} \left(\phi_{k_\perp}(\mathbf{X}) - \frac{v_\parallel}{c}A_{k_\perp}(\mathbf{X})\right)
% \textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}} \left(1-\textrm{e}^{-i\mathbf{k_\perp}\cdot\mathbf{\rho}}J_0(k_\perp\rho)\right)     B\,\mathrm{d}\mu\,\mathrm{d}v_\parallel\,\mathrm{d}\gamma
% \end{eqnarray*}
% Note that the integral over $\gamma$ is NOT a gyroaverage but the sum of the contribution to the local density coming from all the
% gyrocenters $\mathbf{X}$ around $\mathbf{x}$. The reference point is $\mathbf{x}$ and the perturbed distribution function has been expressed as
% $$f(\mathbf{X}) = \sum_\mathbf{k_\perp} f_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{X}} = \sum_\mathbf{k_\perp}
% f_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}} \textrm{e}^{-i\mathbf{k_\perp}\cdot\mathbf{\rho}}$$
% Performing the velocity space integrals, one obtains:
% \begin{eqnarray*}
% \label{eq:density}
% n(\mathbf{x}) &=& \overbrace{n_0 + 2\pi B \int \sum_{k_\perp} f_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}}J_0(k_\perp\rho)
% \,\mathrm{d}\mu\,\mathrm{d}v_\parallel}^\textrm{\scriptsize guiding center density (background+perturbed)}\\* 
% & & - \underbrace{n_0\frac{Ze}{T}\sum_{k_\perp} \phi_{k_\perp}(\mathbf{x})\textrm{e}^{i\mathbf{k_\perp}\cdot\mathbf{x}}\left(1-\Gamma(b)\right)}_\textrm{\scriptsize
% polarisation density}
% \end{eqnarray*}
% where $n_0$ and $T$ are the density and temperature of the Maxwellian and where the following relationship has been used:
% $$\frac{1}{v_{\rm th}^2}\int_0^\infty \mathrm{e}^{-v_\perp^2/v_{\rm th}^2}J_0^2(k_\perp v_\perp \frac{ZeB}{m}) v_\perp \,\mathrm{d}v_\perp = \Gamma(b)=\textrm{e}^{-b}I_0(b) \qquad \mathrm{with} \qquad b={1
% \over 2} {k_\perp^2 m^2 v_{\rm th}^2 \over Z^2e^2 B^2}$$
% with $I_0$ the modified Bessel function. The gyro-kinetic quasi-neutrality equation is then obtained by summing over the different species and assuming quasi-neutrality for the background (see also
% Ref. \cite{BEER} Eq. (2.55)). Writing the result for a single wave vector to keep the expressions shorter, one obtains:
% \begin{equation} 
% \sum_s Z_s n_s(\mathbf{x}) = 0 = \sum_s  Z_s \left[2\pi B \int  f_s(\mathbf{x})J_0(k_\perp\rho_s)\,\mathrm{d}\mu\,\mathrm{d}v_\parallel + {Z_s e n_s \over T_s} ( \Gamma(b_s) - 1)\phi \right] 
% \end{equation} 
% The quasi-neutrality equation in normalised form is 
% \begin{equation} 
% \sum_s  Z_s n_{Rs} \biggl [ 2 \pi B_N \int {\rm d} v_{\parallel N} {\rm d} \mu_N J_0(k_\perp\rho_s) g_{Ns} 
% + {Z_s \over T_{Rs}} [ \Gamma(b_s) -1]\phi_N \biggr ] = 0 
% \end{equation}
% \blue From here on still needs to be checked \black 
% with the argument of the Bessel function being 
% \begin{equation} 
% k_\perp \rho_s = { k_\perp \rho_* R_{\rm ref} \over Z} \sqrt{ 2 m_R T_R \mu_N \over B_N}
% \end{equation}
% 
% The vector potential must be calculated from Ampere's law 
% \begin{equation} 
% \nabla^2 A_\parallel = \mu_0 {\bf J} = \mu_0 \sum_s Z_s e \int {\rm d}^3 {\bf v} \, 
% v_\parallel J_0 f_s 
% \end{equation} 
% Since we the distribution $g$ forward in time, we have to express this equation 
% in terms of $g$ 
% \begin{equation} 
% \nabla^2 A_\parallel + \mu_0 \sum_s {Z_s^2 e^2 \over T} \int {\rm d}^3 {\bf v} \, 
% v_\parallel^2 J_0^2 A_\parallel F_M 
%  =  \mu_0 \sum_s Z_s e \int {\rm d}^3 {\bf v} \, v_\parallel J_0 g_s  
% \end{equation} 
% The integral over the Maxwell distribution can be performed analytically with the 
% help of the integrals 
% \begin{equation} 
% {1 \over \sqrt{\pi}} \int_{-\infty}^\infty {\rm d} v_\parallel  v_\parallel^2 
% \exp \biggl [ - {v_\parallel^2 \over v_{\rm th}^2} \biggl ] = {1\over 2} v_{\rm th}^3 
% \end{equation} 
% \begin{equation} 
% \int_0^\infty {\rm d} x J_0(\sqrt{2 b x}) \exp [ -x ] = \Gamma(b) = \exp [ -b ] I_0(b) 
% \end{equation} 
% where $I$ is the modified Bessel function. 
% Using these expressions one obtains 
% \begin{equation}
% \int {\rm d}^3 {\bf v}\, v_\parallel^2 J_0^2 F_M = {n v_{\rm th}^2 \over 2} \Gamma(b)
% \end{equation} 
% with 
% \begin{equation} 
% b = {1 \over 2} {k_\perp^2 m^2 v_{\rm th}^2 \over e^2 B^2} = {1 \over 2} m_R T_R (k_\perp 
% \rho_* R_{\rm ref} / Z B_N^2)^2
% \end{equation}
% Replacing furthermore $\nabla_\perp^2$ with $-k_\perp^2$ Ampere's law for $g_s$ can be 
% written as 
% \begin{equation} 
% -k_\perp^2 A_\parallel + \mu_0 \sum_s {Z_s^2 e^2 n \over m} \Gamma(b) A_\parallel =
% \mu_0 \sum_s Z_s e \int {\rm d}^3 {\bf v} v_\parallel J_0^2 g_s
% \end{equation}
% The normalization can then be applied to yield 
% \begin{align}
% \biggl [ -k_{\perp N}^2 + \beta \sum_s {Z_s^2 n_{rat,s} \over m_{rat,s}} \Gamma(b_s) \biggr ] A_{\parallel N}
% = \beta \sum_s Z_s v_{thrat,s} n_{rat,s} \times \cr 
% \noalign{\vskip 0.2 truecm} 
% 2 \pi B_N \int {\rm d} v_{\parallel N} \int {\rm d} \mu_N
% v_{\parallel N} J_0 g_{N,s}
% \end{align}
% where $g_N$ is given by 
% \begin{equation} 
% g_N = f_N + 2 Z {v_{\rm thrat} \over T_{\rm rat}} v_{\parallel N} J_0 A_{\parallel N} F_{MN}
% \end{equation}

% related to auctex mode and latex-preview-mode in Emacs:
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "doc"
%%% End:

### 2022-05-03 gkw/doc/benchmarks/cyclone/dimits_shift/mkcyc.m

@@ -1,0 +1,33 @@
+load cycgene ;
+load cycgkw_highres;
+
+gkw = cycgkw_highres;
+gene = cycgene; 
+
+figure(99);
+hold on; 
+
+plot(gkw(:,1),gkw(:,2),'rs')
+plot(gkw(:,1),gene(:,1),'bo')
+
+ns = size(gkw,1);
+for i = 1: ns 
+  x = [gkw(i,1) gkw(i,1)];
+  y = [gkw(i,2)-gkw(i,3) gkw(i,2)+gkw(i,3)];
+  plot(x,y,'r-')
+end;
+
+x = linspace(6,20,100);
+y = 15.4*(1 - 6./x);
+axis([0 20 0 10]);
+plot(x,y,'k');
+
+figure(98) 
+hold on; 
+y = y / 2. / sqrt(2.)
+plot(x,y)
+plot(gkw(:,1),gkw(:,2)/2/sqrt(2.),'ro');
+axis([0 16 0 4])
+xlabel('R/L_T')
+ylabel('\chi_i (\rho_i^2 v_{th} / L_N)')
+

### 2022-05-03 gkw/doc/scaling/nonspectral2/mkfigs.m

@@ -1,0 +1,10 @@
+read_perf('nonspec_scaling2b','pa_*','pa_64_mu16sp2s2',1)
+read_perf('nonspec_scaling2b','ca_*','ca_64_mu8sp2s4',1)
+
+%legend('Med ES, No Coll','linear','Med ES Coll','linear')
+
+read_perf('nonspec_scaling3','s*','s1024_mu16sp2s32',1)
+read_perf('nonspec_scaling3_clean_coll','ac*','ac_2048_mu8sp2s32x4',1)
+read_perf('nonspec_scaling3_clean','a*','a2_1024_mu16sp2s32',1)
+
+%legend('Large ES, No Coll','linear','Large EM, Coll','linear','Large EM, No Coll','linear')

### 2022-05-03 gkw/matlab/average_error.m

@@ -1,0 +1,42 @@
+function [av aver flmst] = fl_cal(in,nstart,nend) 
+% function [av aver flmst] = fl_cal(in,nstart,nend) 
+% 
+% For a nonlinear run caculates the average of a signal 
+% and determines the error on the average by deviding 
+% the interval in several sub intervals over which the 
+% average is performed. 
+% 
+% Input in      : the array of signal 
+%       nstart  : starting point of the average 
+%       nend    : end point of the average 
+% Output av     : The averaged 
+%        aver   : Error on the average
+%        flstd  : Standard dev on the average
+
+
+% Set the number of intervals 
+ninter = 4; 
+
+if (nend-nstart < ninter); 
+    % The average 
+    av = mean(in(nstart:nend));
+    aver=NaN;
+    flmst=NaN;    
+    return; 
+end
+
+% The average 
+av = mean(in(nstart:nend));
+
+% The standard deviation 
+flmst = std(in(nstart:nend));
+
+% The real estimate of the error 
+nstep = round((nend-nstart)/ninter);
+nstart = round(nstart);
+
+for i = 1:ninter-1 
+  val(i) = mean(in(nstart+(i-1)*nstep:nstart+i*nstep)); 
+end; 
+val(ninter) = mean(in(nstart+(ninter-1)*nstep:nend)) ;
+aver = min(std(val),flmst); 

### 2022-05-03 gkw/matlab/chease/eqdsk_miller.m

@@ -1,0 +1,278 @@
+function [gkwmil eql yMill]= eqdsk_miller(eqdsk_file,eqdsk_path,epsilon,optplot,eqm);
+%
+% function [gkwinput gs2input gs2_mill_all_surfs] = (eqdsk_file,eqdsk_path,epsilon,optplot,eqm);
+%
+% builds the Miller equilibrium from eqdsk file (or hamada, in principle)
+%
+% Requires interpos, available from https://crppwww.epfl.ch/~sauter/interpos/x
+%
+% WARNING: this produces quite different results to hamada_miller
+% due to either better smoothing / interpolation or averaging over top and bottom halves
+%
+% Derives GS2 miller parameters then converts to GKW at the end
+% in output, GKW miller parameters
+%
+% optplot = 0 : no plots
+% optplot = 1 : plot eqdsk surfaces
+% optplot = 2 : plot eqdsk + contourc
+% optplot = 3 : plot miller surfaces + eqdsk + contourc (default)
+%
+% eqm = 'eqdsk' read eqdsk file (default)
+% eqm = 'hamada' read gkw-chease hamada file (not yet working)
+%
+% TO DO: also calculate q and shat from eqdsk
+%
+% cla, 11.01.2007
+% fjc, 31.07.2012
+%
+
+if ~exist('optplot'); optplot = 3; end
+if ~exist('eqm'); eqm='eqdsk'; end
+
+%number or surfaces to remove (adjust if some give problems)
+bnd=3;
+
+%% Get the flux surfaces into same format
+%% and calculate the q profile
+% yout = RZsurfs(eqdg, shot, t1,t2,1, expeq, editeq); close(gcf);
+%Zmag = yout.Zin(1);
+
+if (isequal(eqm,'hamada'))
+    error('hamada option still has bugs')
+    ham = read_hamada(eqdsk_file,eqdsk_path);
+    Zmag=mean(ham.z(:,1));
[…]

### 2022-05-03 gkw/matlab/chease/hamada_miller.m

@@ -1,0 +1,278 @@
+% Calculate the miller's parameters from an hamada file and write them in a GKW input file
+%	function [H,M] = hamada_miller(ham_file,ham_path,gkw_file,proj,rw,square_sw,zmil_sw,epsilon)
+% Inputs :
+%	ham_file : name of the hamada file
+%	ham_path : path of the hamada file
+%	gkw_file : name of the GKW input file
+%	proj : name of the project where the GKW input file is
+%	rw : 'r' if you only want to calculate the Miller's parameters, 'rw' if you also want to write them in a GKW file 
+%   square_sw: (optional) integer switch, use 0 to set squareness and deriv to zero
+%   zmil_sw: (optional) integer switch, use 0 to set Zmil and deriv to zero
+%   epsilon (optional) select eps surface (chease value normalised by R_axis) to select
+%
+% Outputs :
+%	H : contains the informations of hamada file
+%	M : contains all the Miller's parameters
+%
+% WARNING : this version works fine with zeta ~ 0 (for numerical surfaces). High zeta produces surfaces that cannot be resolved using zmax etc... could be done with an fminsearch but not sure if it is worth.
+% There is no known problems with surfaces perfectly described using miller's parameters.
+% For up-down asymmetric equilibiria it is more common to calculate upper and lower values of
+% each parameter, and then take the mean.
+%
+%
+% YC 22.11.2012: 
[…]

### 2022-05-03 gkw/matlab/chease/read_eqdsk.m

@@ -1,0 +1,110 @@
+% reads the content of an EQDSK file (all in MKSA units)
+%	function [out]=read_eqdsk(flnm,flpth)
+% Inputs:
+%		flnm	file name
+%		flpth file path, if no path is entered, the default path is used
+% This routine is not written at all in a general way, therefore it only works for a specific output file structure
+
+function [out]=read_eqdsk(flnm,flpth)
+
+ 
+% default path
+if ~exist('flpth')||isempty(flpth)
+%	flpth='/home/space/phsgbe/runs/chease/output/';
+	flpth='./';
+end
+
+% reads file
+if unix(['test -e ' flpth flnm])==0
+	fid = fopen([flpth flnm], 'r');
+else	
+	error(['The file ' flpth flnm ' does not exist' ])
+end
+frewind(fid);
+
+% comments and date
+%out.comments=fscanf(fid,'%c',48);
+%out.comments=fscanf(fid,'%c',43);
+
+out.comments = fgetl(fid);
+stst = fseek(fid, -8,0); 
+
+% % switch
+% out.switch=fscanf(fid,'%i5');
+% if out.switch~=3,
+% 	disp(['Warning, i3=' num2str(out.string) ', unknown file structure...'])
+% 	return
+% end
+
+% box size
+out.nrbox=fscanf(fid,'%5i',1);  % number of R points
+out.nzbox=fscanf(fid,'%5i',1);	% number of Z points
+
+% first line
+out.rboxlength=fscanf(fid,'%e',1);	% bow width 
+out.zboxlength=fscanf(fid,'%e',1);	% box height
+out.R0EXP=fscanf(fid,'%e',1);	% Normalizing radius in CHEASE
+out.rboxleft=fscanf(fid,'%e',1);	% left position of the R grid
+fscanf(fid,'%e',1); % nothing
+
+% second line
+out.Raxis=fscanf(fid,'%e',1);	% Raxis
+out.Zaxis=fscanf(fid,'%e',1);	% Zaxis
+out.psiaxis=fscanf(fid,'%e',1);	% psi_axis-psi_edge
+out.psiedge=fscanf(fid,'%e',1);	% psi_edge-psi_edge (=0)
+out.B0EXP=fscanf(fid,'%e',1);	% Normalizing magnetic field in CHEASE
[…]

### 2022-05-03 gkw/matlab/chease/read_eqdsk_cla2.m

@@ -1,0 +1,226 @@
+function yyy = read_eqdsk_cla2(filename, pth, optplot, numcont);
+%
+% function yyy = read_eqdsk_cla2(filename, pth, optplot, numcont);
+%
+% by cla; fjc: added flux surface computation by contourc
+%
+% optplot = 0 : no plots
+% optplot = 1 : plot eqdsk surfaces (default)
+% optplot = 2 : plot eqdsk + contourc
+%
+
+if ~exist('optplot'); optplot = 1; end
+
+if ~exist('numcont')
+numcont = 200;
+end
+if isempty(numcont);
+numcont = 200;
+end
+
+
+pthflnm = [pth '/' filename]
+%
+fptr = fopen(pthflnm);
+
+frewind(fptr);
+aaa = fgetl(fptr);
+stst = fseek(fptr, -8,0);
+
+%aaa = fscanf(fptr, '%f',1);
+nw = fscanf(fptr, '%f',1)
+nh = fscanf(fptr, '%f',1)
+%
+rwid = fscanf(fptr, '%f',1);
+zhei = fscanf(fptr, '%f',1);
+rcentr = fscanf(fptr, '%f',1);
+rleft  = fscanf(fptr, '%f',1);
+zmid = fscanf(fptr, '%f',1);
+%
+rmaxis = fscanf(fptr, '%f',1);
+zmaxis = fscanf(fptr, '%f',1);
+simag  = fscanf(fptr, '%f',1);
+sibry  = fscanf(fptr, '%f',1);
+bcentr = fscanf(fptr, '%f',1);
+%
+current = fscanf(fptr, '%f',1);
+simag  = fscanf(fptr, '%f',1);
+xdum   = fscanf(fptr, '%f',1);
+rmaxis = fscanf(fptr, '%f',1);
+xdum   = fscanf(fptr, '%f',1);
+%
+zmaxis = fscanf(fptr, '%f',1);
+xdum   = fscanf(fptr, '%f',1);
+sibry  = fscanf(fptr, '%f',1);
+xdum   = fscanf(fptr, '%f',1);
+xdum   = fscanf(fptr, '%f',1);
+%
+fpol   = fscanf(fptr, '%f',nw);
+pres   = fscanf(fptr, '%f',nw);
+ffprim = fscanf(fptr, '%f',nw);
[…]

### 2022-05-03 gkw/matlab/chease/read_expeq.m

@@ -1,0 +1,53 @@
+% reads the content of an EXPEQ file (CHEASE units)
+%	function [out]=read_expeq(flnm,flpth)
+% Inputs:
+%		flnm	file name for the EXPEQ file
+%		flpth file path, if no path is entered, the default path is used
+
+
+
+function [out]=read_expeq(flnm,flpth)
+
+% default path
+if ~exist('flpth')||isempty(flpth)
+	%flpth='/home/space/phsgbe/runs/chease/input/';
+        flpth='./'
+end
+
+% reads file
+if unix(['test -e ' flpth flnm])==0
+	fid = fopen([flpth flnm], 'r');
+else	
+	error(['The file ' flpth flnm ' does not exist' ])
+end
+frewind(fid);
+
+
+
+% aspect ratio (Rmax-Rmin)/Rref/2
+out.aspct=fscanf(fid,'%e',1);
+% Zref/Rref
+out.ZRrat=fscanf(fid,'%e',1);
+% edge pressure
+out.pedge=fscanf(fid,'%e',1);
+
+% nb of points for LCFS
+out.N_LCFS=fscanf(fid,'%5i',1);
+% LCFS description
+RZ=fscanf(fid,'%e',[2,out.N_LCFS])';
+out.R=RZ(:,1)';
+out.Z=RZ(:,2)';
+
+% nb of points for profiles
+out.N_rho=fscanf(fid,'%5i',1);
+% NSTTP
+out.NSTTP=fscanf(fid,'%5i',1);
+% rho (rho_psi)
+out.rho=fscanf(fid,'%e',out.N_rho);
+% P'
+out.pprime=fscanf(fid,'%e',out.N_rho);
+% current 
+out.j=fscanf(fid,'%e',out.N_rho);
+
+
+fclose(fid);

### 2022-05-03 gkw/matlab/chease/rz2miller.m

@@ -1,0 +1,242 @@
+% Computes the Miller parameters from  R,Z flux surfaces description
+% 
+%     [r,Rmil,Zmil,k,d,z,dRmildr,dZmildr,sk,sd,sz,Rout,Zout,chi2]=rz2miller(R,Z,code,doplots)
+%
+% Inputs:
+%  R,Z:   flux surfaces description in [m]
+%         Size of the array assumed to be (nrho,npol), take care of it: no check performed!
+%         The array is also assumed to have no double points
+%  code:  Switch between different code conventions.
+%         Available are: 'gkw', 'gyro', 'tglf' 
+%  doplots: if 1 (default) perform plots to check the quality of the parametrisation
+%
+% Outputs:
+%  r,Rmil,Zmil,k,d,z,dRmildr,dZmildr,sk,sd,sz: Miller parameters, as described in the GKW manual
+%  Rout, Zout:  parametrised flux surfaces in [m]
+%  chi2: parametrisation quality (chi2>0.05 starts to be quite bad, switch to full equilibrium advised!)
+%
+% Triangularity and squareness obtained by minimising the distance between the input flux surfaces and the 
+% parametrised solutions.
+%
+% Requires interpos, available from https://crppwww.epfl.ch/~sauter/interpos/
+%
+%
+% YC 22.11.2012
+
+function [r,Rmil,Zmil,k,d,z,dRmildr,dZmildr,sk,sd,sz,Rout,Zout,chi2]=rz2miller(R,Z,code,doplots)
+
+if (~exist('doplots','var')||isempty(doplots))
+ doplots=1;
+end 
+
+if (~exist('code','var')||isempty(code))
+ code='gkw';
+end 
+
+if ~(strcmp(code,'gkw')|strcmp(code,'gyro')|strcmp(code,'tglf'))
+ disp(['Miller parametrisation not available for code=' code])
[…]

### 2022-05-03 gkw/matlab/fluxes_ratio.m

@@ -1,0 +1,162 @@
+%This script makes a matlab function that extracts the RATIO of AVERAGE fluxes information 
+%from the fluxes folder for multiple runs in a project, averages the last data points, and plots against a scanned variable
+%
+%Usage: out=fluxes_ratio(proj,flux1,flux2,variable,filename,n_av,etype,etype2,t_av)
+%
+%proj is a string with your project name
+%Scanned variable in the form of the structure created by read_gkwinput
+%Filename (optional) to select which runs of project to plot (accepts
+%wildcards).  Default = All.
+%n_av is the (optional) number of final data points in time.dat to average (default 100 if not provide)
+%t_av is optional time to start averaging from (overrides n_av)
+%You are advised to check convergence of your runs first using the time_traces function
+%
+%Flux chooses the column number to look at
+%Flux2 chooses the column number to normalize with
+%
+%etype and etype2 chooses from electro-static ('es'), electromagnetic ('em')
+%or compressional ('bpar') or neoclassical (neoc, not output yet) fluxes.
+%The default is electro-static.
+%
+%Example: fluxes_ratio('latest_project',1,2,'ROTATION.shear_rate','files*',100,'em','es')
+%
+%This script requires gkw_path to be set up correctly for your projects directory.
+
+function [flux_rat]=fluxes(proj,flux,flux2,variable,filename,n_av,etype,etype2,t_av)
+
+if ~exist('flux')
+	flux=2;
+    disp('Energy Flux for species 1')
+end
+
+if ~exist('flux2')
+	flux=5;
+    disp('Over Energy flux for species 2')
+end
+
+if ~exist('proj')
+	proj='default';
+    disp('You must provide the project name')
+    %No default value
+end
+if ~exist('n_av')
+    n_av=100;
+    disp('Average over last 100 data points')
+end
+if ~exist('variable')
+    variable='ROTATION.shear_rate';
+end
+if ~exist('filename')
+    filename='';
+end
+if ~exist('etype')
+    etype='es';
+end
+if ~exist('etype2')
+    etype2='es';
+end
+
+count=0;            %Counter for output files
+no_files=0;         %Counter for input files
+
+files=dir([gkwpath('input',proj) filename]);
+%files=dir(gkwpath('input',proj));
+total = size(files);
+
+switch etype
+    case 'es'
+        fluxes_path='fluxes'
+    case 'em'
[…]

### 2022-05-03 gkw/matlab/gkw_gui.m

@@ -1,0 +1,843 @@
+function varargout = gkw_gui(varargin)
+% GKW_GUI M-file for gkw_gui.fig
+%      GKW_GUI, by itself, creates a new GKW_GUI or raises the existing
+%      singleton*.
+%
+%      H = GKW_GUI returns the handle to a new GKW_GUI or the handle to
+%      the existing singleton*.
+%
+%      GKW_GUI('CALLBACK',hObject,eventData,handles,...) calls the local
+%      function named CALLBACK in GKW_GUI.M with the given input arguments.
+%
+%      GKW_GUI('Property','Value',...) creates a new GKW_GUI or raises the
+%      existing singleton*.  Starting from the left, property value pairs are
+%      applied to the GUI before gkw_gui_OpeningFcn gets called.  An
+%      unrecognized property name or invalid value makes property application
+%      stop.  All inputs are passed to gkw_gui_OpeningFcn via varargin.
+%
+%      *See GUI Options on GUIDE's Tools menu.  Choose "GUI allows only one
+%      instance to run (singleton)".
+%
+% See also: GUIDE, GUIDATA, GUIHANDLES
+
+% Edit the above text to modify the response to help gkw_gui
+
+% Last Modified by GUIDE v2.5 13-Feb-2010 22:26:22
+
+% Begin initialization code - DO NOT EDIT
+gui_Singleton = 1;
+gui_State = struct('gui_Name',       mfilename, ...
+                   'gui_Singleton',  gui_Singleton, ...
+                   'gui_OpeningFcn', @gkw_gui_OpeningFcn, ...
+                   'gui_OutputFcn',  @gkw_gui_OutputFcn, ...
+                   'gui_LayoutFcn',  [] , ...
+                   'gui_Callback',   []);
+if nargin && ischar(varargin{1})
+    gui_State.gui_Callback = str2func(varargin{1});
+end
+
+if nargout
+    [varargout{1:nargout}] = gui_mainfcn(gui_State, varargin{:});
+else
+    gui_mainfcn(gui_State, varargin{:});
+end
+% End initialization code - DO NOT EDIT
+
+
+
+% --- Executes just before gkw_gui is made visible.
+function gkw_gui_OpeningFcn(hObject, eventdata, handles, varargin)
+% This function has no output args, see OutputFcn.
+% hObject    handle to figure
+% eventdata  reserved - to be defined in a future version of MATLAB
+% handles    structure with handles and user data (see GUIDATA)
[…]

### 2022-05-03 gkw/matlab/growth_rate_spec_ntor.m

@@ -1,0 +1,153 @@
+% This file defines a matlab function which outputs the linear spectra
+% against toroidal mode number.  Uses read_gkw_all to get rhostar from inputs
+% Can read the data from either
+% [a batch of linear runs with nmod=1] or [a single linear run with nmod > 1].
+% This script assumes the naming  conventions of the script gkwnlin,
+% and requires gkwpath.m to be setup.
+%
+% out = growth_rate_spec_ntor(proj,runname,n_av,optgr)
+%
+% WHERE:
+%
+% proj and runname are strings for gkwpath.m; use wildcard in runname
+% n_av is optional number of final points to average over (default 1)
+% optgr = 0 (optional, default) takes set of single growth rates from time.dat
+% optgr = 1 takes set of single growth rates from growth.dat
+% optgr = 2 takes spectrum of growth rates from single growth.dat
+%
+% OUTPUT:
+%
+% Column 1 is toroidal mode number
+% Column 2 is growth rate in c_s/a
+% Column 3 is mode frequency in c_s/a (negative is electron direction)
+% Column 4 is the parity parameter (absolute value > 0.5 for microtearing)
+% Column 5 is the QL ion flux, normalised to (|phi|^2 + |Apar|^2)
+% Column 6 is the QL electron flux, normalised to (|phi|^2  + |Apar|^2)
+%
+%
+% EXAMPLE: grdat = growth_rate_spec_ntor('proj','runname*',5,0)
+
+function [growth_rate]=growth_rate_spec_ntor(proj,filename,n_av,optgr)
+
+if ~exist('proj')
+    proj='default';
+    disp('You must provide the project name')
+    return;
+    %No default value
+end
+if ~exist('filename')
+    disp('You must provide the run name')
+    filename='default';
+    return;
+end
+if ~exist('n_av')
+    n_av=1;
+end
+if ~exist('optgr')
+    optgr=0;
+end
+
+count=0;            %Counter for output files
+no_files=0;         %Counter for input files
+
+files=dir([gkwpath('time',proj) filename])
+total = size(files);
+
+disp(files(1).name)
+dat=read_gkw_all(files(1).name,proj,1);
+
+grfac=1/(dat.rova*sqrt(dat.SPECIES(2).temp/2));
+
+% ntor 
+kyfac=1/(2*pi*dat.rhostar*dat.geom.kthnorm);   
+
[…]

### 2022-05-03 gkw/matlab/growth_rates.m

@@ -1,0 +1,69 @@
+% matlab function that extracts the growth rate information from the time folder
+% for multiple (linear) runs in a project and plots against a scanned variable
+%
+% Usage: growth_rates(proj,,file,variable,n_av)
+%
+% proj is a string with your project name
+% Scanned variable in the form of the structure created by read_gkwinput
+% n_av is the (optional) number of final data points in time.dat to average
+% Default is 1.  For most linear runs time averaging should not be necssary. You are adivsed to check convergence of your runs first!
+%
+% Example: growth_rates('latest_project','runs*','ROTATION.shear_rate',100)
+%
+% This script interfaces with Yann's scripts 
+% It requires gkw_path to be set up correctly for your projects directory.
+
+function [growth_rat]=growth_rates(proj,filename,variable,n_av)
+
+
+if ~exist('proj')
+	proj='default';
+    disp('You must provide the project name')
+    %No default value
+end
+if ~exist('variable')
+    variable='ROTATION.shear_rate';
+end
+if ~exist('n_av')
+    n_av=1;
+end
+    
+count=0;            %Counter for output files
+no_files=0;         %Counter for input files
+
+%Currently you have to give a prefix for the filenames. 
+%files = dir('input/*');
+files=dir([gkwpath('input',proj) filename]);
+%Could be made neater to only list all files (not directories)
+total = size(files);
+
+for i = 1:total(1)
+    
+   if(files(i).isdir==0)
+        no_files=no_files+1;
+        input=read_gkwinput(files(i).name,proj);
+       
+        if (exist([gkwpath('time',proj) files(i).name],'file')==2)
+            count = count+1;
+            str=sprintf('growth_rat(count,1)=input.%s;',variable);
+            eval(str);
+            time=load([gkwpath('time',proj) files(i).name]);
+            disp(['Loaded ' gkwpath('time',proj) files(i).name])
+            dim=size(time);
+            growth_rat(count,2)=mean(time(end-n_av+1:end,2));
+         else
+            disp(['Skipped missing', gkwpath('time',proj) files(i).name])
+        end
+        %cd ..
+   end
+end
+
+disp(sprintf('Loaded %i output files for %i input files', count, no_files))
+growth_rat=sortrows(growth_rat,1);
+
+plot(growth_rat(:,1),growth_rat(:,2),'+-','DisplayName',filename);
+xlabel(variable);
+ylabel('\gamma (R/v_{th})');
+%legend(sprintf('%s, n av = %i',proj,n_av));
+
+end

### 2022-05-03 gkw/matlab/gs2tools/gs2gkw_input.m

@@ -1,0 +1,78 @@
+function [gkwin gs2in] = gs2gkw_input(gs2_file,gkw_template,gkw_proj);
+%
+% function [gkwin gs2in] = gs2gkw_input(gs2_file,gkw_template,gkw_proj);
+%
+% Convert a GS2 input file to a GKW input file using a template for GKW input
+%
+% e.g. give in input gs2_file = 'xAUG25832_6.775_k_3_CxUB_tr_0_up1';
+%
+% Only for simple geometries - other not included (they are in gkw2gs2_input)
+% Exactly two species only (more are allowed in gkw2gs2_input)
+%  Species 1 is ions (assumed deuterium)
+%  Species 2 is electrons
+%
+% Use with care - check the output manually !
+%
+% GS2 file directory is './' (hard coded for now)
+%
+% FJC 01.10.11
+
+gs2_pthin = './';
+[gs2in.eql gs2in.spc gs2in.prkn gs2in.aky] = read_cgs2input(gs2_file, gs2_pthin);
+
+gkwin = read_gkwinput(gkw_template,gkw_proj);
+
+if gs2in.prkn.fapar == 1; gkwin.CONTROL.nlapar = '.true.'; end;
+
+gkwin.MODE.kthrho = gs2in.aky.aky_min.*sqrt(2);
+
+gkwin.GEOM.SHAT = gs2in.eql.s_hat_input;
+gkwin.GEOM.Q = gs2in.eql.qinp;
+gkwin.GEOM.EPS = gs2in.eql.rhoc;
+gkwin.GEOM.GEOM_TYPE= 'circ';
+
+gkwin.SPCGENERAL.beta = gs2in.prkn.beta;
+
+gkwin.SPECIES(1).MASS= gs2in.spc(1).mass;
+gkwin.SPECIES(1).Z= gs2in.spc(1).z;
+gkwin.SPECIES(1).TEMP= gs2in.spc(1).temp;
+gkwin.SPECIES(1).dens= 1.0;
+gkwin.SPECIES(1).rlt= gs2in.spc(1).tprim;
+gkwin.SPECIES(1).rln= gs2in.spc(2).fprim;
+%This assumes Ti=Tref in GS2
+gkwin.SPECIES(1).uprim= gs2in.spc(1).uprim;
+
[…]

### 2022-05-03 gkw/matlab/gs2tools/read_gs2output.m

@@ -1,0 +1,179 @@
+function [aky, omavg, qheat, pflux, vflux, qmheat, pmflux] = read_gs2output(flnm, pth, opt)
+%
+% function [aky, omavg, qheat, pflux, vflux, qmheat, pmflux] = read_gs2output(flnm, [pth], opt)
+%
+% Input.
+%       flnm :  filename
+%       pth  :  path (optional: by default looks in user's directory ~/gs2run/output)
+%       if pth == 1 looks in user's directory ~/gs2run/gs2archive/output)
+% Output.
+%       aky   : vector of k_theta values
+%       omavg : matrix with complex frequency for each k_theta
+%               omavg(:,1) : real frequency
+%               omavg(:,2) : growth rate
+%               same normalisations as in gs2 output !
+%       qheat and pflux : matrix with heat and particle fluxes
+%                         (same normalisations as in gs2)
+%
+% Calls read_gs2outtime to check time evolution if not converged
+%
+% CLA 10.08.03
+%
+
+usrnm = find_usrnm;
+%
+if ~exist('pth')
+    pth = ['/afs/ipp/home/' usrnm(1) '/' usrnm '/gs2run/output/'];
+    %pth = ['/scratch/' usrnm '/GS2scratch/output/'];
+end
+
+if ~exist('opt'); opt = 1; end
+
+totflnm = [pth flnm];
+fptr = fopen(totflnm, 'r');
+frewind(fptr);
+%
+
+iconv = 1;
+yk = 0;
+while yk == 0
+    sss = fscanf(fptr,'%s', 1);
+    if isempty(sss); break; end;
+    if strcmp(sss, 'converged'), yk = 1; end;
+end
+
+if yk == 0
+    disp([flnm ' not converged']);
+    iconv = 0;
+    if opt == 1
+        ync = input('Do you want to check the output ? (0/1)  ');
+        if ~ync
+            aky = NaN;
+            omavg = [NaN NaN];
+            qheat = [NaN NaN];
+            pflux = [NaN NaN];
+            vflux = [NaN NaN];
+            qmheat = [NaN NaN];
+            pmflux = [NaN NaN];
+            vmflux = [NaN NaN];
+        else
+            [aky, omavg, qheat, pflux, vflux, qmheat, pmflux, vmflux, ky] = read_gs2outtime(flnm, pth);
+        end
+    else
+        aky = NaN;
+        omavg = [NaN NaN];
+        qheat = [NaN NaN];
+        pflux = [NaN NaN];
+        vflux = [NaN NaN];
[…]

### 2022-05-03 gkw/matlab/list_gkwscan.m

@@ -1,0 +1,52 @@
+% list the GKW scan file available for a given project and print their description
+%	function []=list_gkwscan(proj)
+% Inputs:
+%	proj:	project name (optional)
+%		path for the input files is obtained from the gkwpath function with "proj" in argument
+%		If you give proj='ALL', all the scans of all the projects are listed
+
+function []=list_gkwscan(proj)
+
+% default 
+if ~exist('proj')
+	proj='ALL';
+end
+
+if strcmp(proj,'ALL')
+
+	D=dir(gkwpath('top'));
+	I=find([D.isdir]);
+	for ii= 1:length(I)
+		if ~(strcmp(D(I(ii)).name,'.')|strcmp(D(I(ii)).name,'..'))
+		  fprintf('******************** Project %s*******************\n',D(I(ii)).name)
+		  list_gkwscan(D(I(ii)).name)
+		  fprintf('******************************************************\n\n',D(I(ii)).name)
+		end
+	end
+
+else
+
+	flpth=gkwpath('scan',proj);
+	fl=dir(flpth);
+	for ii=1:length(fl),
+		if ~isempty(strfind(fl(ii).name,'.mat')),
+			load([flpth fl(ii).name],'sinfo');
+			sss=[fl(ii).name '\t' sinfo.date];
+			for jj=1:sinfo.nbvar
+				sss=[sss '\t' sinfo.var.name{jj}];	
+				if isfield(sinfo.var,'coupled')
+				  for kk=1:sinfo.var.coupled(jj)
+					sss=[sss '+' sinfo.var.cpl(jj).name{kk}];
+				  end
+				end
+			end
+			sss=[sss '\n' sinfo.comments '\n\n'];
+            
+            
+			fprintf(sss)
+		end
+    end
+    
+    dir(gkwpath('time',proj))
+
+end

### 2022-05-03 gkw/matlab/mkmovie_f.m

@@ -1,0 +1,94 @@
+load xphi;
+load yphi;
+xmin = min(min(xphi));
+xmax = max(max(xphi));
+ymin = min(min(yphi));
+ymax = max(max(yphi));
+clear N;
+close all; 
+maxv = 0; 
+minv = 0;
+figure(1);
+%colormap('bone')
+figure(2);
+hold on ;
+for j= 1: nframes 
+  i = j-1;
+  s = int2str(i);
+  if (i<10) 
+    name = strcat('fdi00000',s)
+  else 
+    if (i<100) 
+      name = strcat('fdi0000',s)
+    else
+      if (i<1000) 
+        name = strcat('fdi000',s)
+      else
+        if (i<10000) 
+          name = strcat('fdi00',s)
+        else
+          if (i<100000) 
+            name = strcat('fdi0',s)
+          else
+            name = strcat('fdi',s)
+          end
+        end
+      end
+    end
+  end   
+  ll = ['load ' name ];
+  eval(ll);
+  dd = ['maxv=max(maxv,max(max(' name ')))'];
+  eval(dd);
+  dd = ['minv=min(minv,min(min(' name ')))'];
+  eval(dd);
+  dd = ['norm = sum(sum(' name '))'];
+  eval(dd);
+  %figure(2);
+  %plot([i],[norm],'bo');
+end;
+maxv = max(maxv,abs(minv));
+vv = linspace(-1.1*maxv, 1.1*maxv, 20); 
+figure(1); 
+load fdi000000
+contourf(xphi,yphi,fdi000000,vv)
+axis equal; 
+shading flat; 
+for j= 1: nframes
+  i = j-1
+  s = int2str(i);
+  if (i<10) 
+    name = strcat('fdi00000',s)
+  else 
+    if (i<100) 
+      name = strcat('fdi0000',s)
+    else
+      if (i<1000) 
+        name = strcat('fdi000',s)
+      else
+        if (i<10000) 
+          name = strcat('fdi00',s)
+        else
+          if (i<100000) 
+            name = strcat('fdi0',s)
+          else
+            name = strcat('fdi',s)
+          end
+        end
+      end
+    end
+  end   
+  ll = ['load ' name ];
+  eval(ll);
+  figure(1);
+  axis equal; 
+  axis([xmin,xmax,ymin,ymax]); 
+  pp = strcat('contourf(xphi,yphi,',name,',20)'); 
[…]

### 2022-05-03 gkw/matlab/mode_box_calculator.m

@@ -1,0 +1,52 @@
+% Function to calculate derived paramaters for GKW mode_box inputs
+% Returns structure with derived mode quantities
+%
+% Usage:
+%   out=mode_box_calculator(nmod,nx,krhomax,ikxspace,shat,eps,q)
+%
+% Alternatively takes a gkw input file structure as a single input
+%
+
+function [out in]=mode_box_calculator(varargin)
+
+  if (nargin==1)
+    strin=cell2mat(varargin(1));
+    
+    in.nmod     = strin.GRIDSIZE.nmod;
+    in.nx       = strin.GRIDSIZE.nx;
+    in.krhomax  =  strin.MODE.krhomax;
+    in.ikxspace =  strin.MODE.ikxspace;
+    in.shat     =  strin.GEOM.shat;
+    in.eps      =  strin.GEOM.eps;
+    in.q        =  strin.GEOM.q;
+    
+  elseif (nargin==7)
+    in.nmod     = cell2mat(varargin(1));
+    in.nx       = cell2mat(varargin(2));
+    in.krhomax  = cell2mat(varargin(3));
+    in.ikxspace = cell2mat(varargin(4));
+    in.shat     = cell2mat(varargin(5));
+    in.eps      = cell2mat(varargin(6));
+    in.q        = cell2mat(varargin(7));
+  else
+    in=NaN;
+    out=NaN;
+    disp('Incorrect number of input arguments')
+    return  
+  end
+
+  disp('WARNING: kthnorm and derived quantities only exact for s-alpha')
+
+  out.krho_min = in.krhomax/max(in.nmod-1,1);
+  out.kthnorm  = in.q/(2*pi*in.eps);
+  out.ky_max   = in.krhomax/out.kthnorm;
+  out.ky_min   = out.krho_min/out.kthnorm;
+  out.kx_min   = 2*pi*in.shat*out.ky_min/in.ikxspace;
+  out.kx_max   = out.kx_min*in.nx/2;
+  out.lyn      = 2*pi/out.ky_min/out.kthnorm;
+  out.lx       = 2*pi/out.kx_min;
+  % Obsolete and deprecated
+  % out.fft_y    = 2^floor(log2(1.5*(2*in.nmod-2))+1);
+  % out.fft_x    = 2^floor(log2(1.5*(in.nx+1))+1);
+
+end

### 2022-05-03 gkw/matlab/old/fluxes_spec.m

@@ -1,0 +1,128 @@
+% This file defines a matlab function which plots the time averaged fluxes
+% spectrum, useful for nonlinear runs.  This script assumes the naming
+% conventions of the script gkwnlin, requires gkw_path.m to be setup.
+% 
+% fluxes_spec(proj,runname,flux,species,n_av,normalise)
+%
+% WHERE:
+%
+% proj and runname are strings for gkwpath.m
+%
+% species (integer) corresponds to the species no. for which you want the flux,
+% in the order of the input file
+%
+% n_av (integer) is the number of points to average over working backwards from
+% the last point in the file. It is useful to check the saturation by eye
+% in fluxes first.
+%
+% flux (integer) selects the flux you want. The script assumes there are exactly 3 fluxes for each species:
+%   1) particle flux
+%   2) heat flux
+%   3) parallel momenutum flux
+%
+% normalize (integer) = 1 will normalize the output to total flux.
+%
+% EXAMPLE: fluxes_spec('proj','runname',2,2,300,1)
+%
+% The spectrum in file is plotted against  runname.krho values in /spectrum/kyspec
+
+
+function []=fluxes_spec(proj,filename,flux,species,n_av,normalize)
+%function []=fluxes(proj,flux,variable,n_av)
+
+
+if ~exist('proj')
+	proj='default';
+    disp('You must provide the project name')
+    return;
+    %No default value
+end
+if ~exist('filename')
+    disp('You must provide the run name')
+    filename='default';
+    return;
+end
+if ~exist('flux')
+	flux='eflux_spec';
+    disp('Plotting energy flux spectra')
+end
+if ~exist('species')
+	species=1;
+    disp('Plotting first species')
+end
+if ~exist('n_av')
+    n_av=100;
+    disp('Average over last 100 data points')
+end
+if ~exist('normalize')
+    normalize=1;
+    disp('Will normalize data to total flux')
+end
+
+%Related column of fluxes.dat containing totals
+%This assumes three fluxes for each species
+flux_no=flux+3*(species-1);
+n_av=n_av-1;
+
+switch(flux)
+    case(2)
+    flux='eflux_spec' 
+    case(3)
+    flux='vflux_spec'
+    case(1)
+    flux='pflux_spec'
+    otherwise
[…]

### 2022-05-03 gkw/matlab/old/kyspec_label.m

@@ -1,0 +1,42 @@
+%This script defines a function which plots kyspec with a legend 
+%The legend is taken from the krho file
+%
+%Run from a single run output folder
+%Or if you are using a project kyspec folder first copy a krho file to it then run from there.
+%
+%Useage kyspec_label(krhomax,kyspec_file)
+%
+%Where kyspec_file is the kyspec filename (optional)
+%
+%You should provide krhomax since the values in file krho 
+%have been normalised by kthnorm -a function of q and eps
+%
+%
+
+function []=kyspec_label(krhomax,ky_file)
+
+%Default values
+if ~exist('krhomax')
+	krhomax=1;
+    disp('Used krhomax = 1')
+end
+if ~exist('ky_file')
+    ky_file='kyspec';
+end
+
+set(0,'DefaultAxesLineStyleOrder','-|--|:')    
+kyspec=load(ky_file);
+plot(kyspec)
+str=[ky_file '.krho'];
+
+%load krho;
+krho=load(str);
+krho=krho*krhomax/krho(size(krho, 1));
+
+for i=1:size(krho,1)
+    labels(i)={num2str(krho(i,1))};
+end
+
+legend(labels)
+
+end

### 2022-05-03 gkw/matlab/old/sum_kxspec.m

@@ -1,0 +1,40 @@
+%This routine will extract and plot the kxspec information from the spc output files
+%As with mkmovie you must first define nframes.
+%Run from a run folder with spcXXXXX files
+%
+
+for j= 1: nframes 
+  i = j-1;
+
+s = int2str(i);
+  if (i<10) 
+    name = strcat('spc00000',s);
+  else 
+    if (i<100) 
+      name = strcat('spc0000',s);
+    else
+      if (i<1000) 
+        name = strcat('spc000',s);
+      else
+        if (i<10000) 
+          name = strcat('spc00',s);
+        else
+          if (i<100000) 
+            name = strcat('spc0',s);
+          else
+            name = strcat('spc',s);
+          end
+        end
+      end
+    end
+  end
+  
+  spc=load(name);
+
+  kxspec(j,:) = sum(spc,1);
+    
+  plot(kxspec)
+  
+end
+  
+  

### 2022-05-03 gkw/matlab/parallel_phi_movie.m

@@ -1,0 +1,61 @@
+% This file defines a matlab function which plots the parallel potential as
+% a movie against time.  This script assumes the naming conventions of the script gkwnlin.
+% 
+% parallel_phi(proj,runname)
+%
+% WHERE:
+%
+% proj and runname are strings for gkwpath.m
+%
+% n_av (integer) is the number of points to average over working backwards from
+% the last point in the file. It is useful to check the saturation by eye
+% in fluxes first.
+
+function []=parallel_phi_movie(proj,filename)
+%function [M]=parallel_phi_movie(proj,filename)
+
+if ~exist('proj')
+	proj='default';
+    disp('You must provide the project name')
+    return;
+    %No default value
+end
+if ~exist('filename')
+    disp('You must provide the run name')
+    filename='default';
+    return;
+end
+
+input=read_gkwinput(filename,proj);
+ns=input.GRIDSIZE.n_s_grid
+scale=[1:1:ns];
+scale=-0.5+(scale-0.5)/ns;
+
+data_file=load([gkwpath('parallel_phi',proj) filename]);
+%data_file=load('parallel_phi.dat')
+disp(['Loaded ' gkwpath('parallel_phi',proj) filename])
+
+%figure(99)
+max1=max(data_file);
+max2=max(max1);
+min1=min(data_file);
+min2=min(min1);
+
+axis manual
+
+for j=1:size(data_file,1)
+
+start=1;
+finish=ns;    
+    
+label=[proj,' ',filename,' Norm:\phi'];
+plot(scale(start:finish),data_file(j,start:finish),'+-','DisplayName',label)
+axis([-0.5,0.5,min2,max2])
+
+%M(j) = getframe;
+ 
+pause(0.05)
+
+end
+
+end

### 2022-05-03 gkw/matlab/paralleldat.m

@@ -1,0 +1,152 @@
+% This file defines a matlab function which can extract data for a single mode from the mess
+% that is parallel.dat with multiple modes and plots.
+% This script assumes the naming conventions of the script gkwnlin and
+% requires gkwpath.m to be configured.
+% 
+% out=paralleldat(proj,runname,imod,ix,is,col,optplot,opteiv)
+%
+% col contents
+% 1: sgr         the length along the field line 
+% 2-3: phi       the potential: (re and im)
+% 4-5: apar      the parallel component of the vector potential (re and im)
+% 6-7: dens      the perturbed density (re and im)
+% 8-9: tpar      the perturbed parallel temperature (re and im)
+% 10-11: tperp   the perturbed perpendicular temperature (re and im)
+% 12-13: wflow   the perturbed parallel flow velocity (re and im)
+% 14-15: bpar    the perturbed parallel (compressional) magnetic feild (re and im)
+%
+% negative 2,4,6,8,10,12,14 will give abs(complex)
+%
+% WHERE:
+%
+% proj and runname are strings for gkwpath.m
+% optplot 1 or zero turns the plot on or off
+%
+% YC: you may want to consider the modifications commented below. 
+% The idea is to reshape data_file once for all to avoid loading multiple 
+% instance of parallel.dat when you want to look at different modes
+%
+% FC: alternatively, use read_gkwparallel for reading in and reshaping the file once.
+
+
+function [out]=paralleldat(proj,filename,imod,ix,is,col,optplot,opteiv)
+
+%YC [kxrh,krho,out]=paralleldat(proj,filename)
+
+if ~exist('proj')
+	proj='default';
+    disp('You must provide the project name')
+    return;
+    %No default value
+end
+if ~exist('filename')
+    disp('You must provide the run name')
+    filename='default';
+    return;
+end
+
+if ~exist('optplot')
+   optplot=1
+end
+
+% if ~exist(imod)
+%     disp('You must select both modes')
+%     return;
+% end
+% if ~exist(ix)
[…]

### 2022-05-03 gkw/matlab/plot_back_prof.m

@@ -1,0 +1,55 @@
+
+profiles = read_prof; 
+
+ns = size(profiles.de,1); 
+
+figure; 
+for i = 1: ns 
+
+  subplot(ns,5,5*(i-1)+1); 
+  plot(profiles.xgr,profiles.de(i,:),'bx-'); 
+  xlabel('\psi')
+  ylabel(['density species ' int2str(i)]); 
+  title('density'); 
+
+end; 
+
+for i = 1: ns 
+
+  subplot(ns,5,5*(i-1)+2); 
+  plot(profiles.xgr,profiles.tmp(i,:),'rx-'); 
+  xlabel('\psi')
+  ylabel(['temperature species ' int2str(i)]); 
+  title('Temperature')
+
+end; 
+
+for i = 1: ns 
+
+  subplot(ns,5,5*(i-1)+3); 
+  plot(profiles.xgr,profiles.rln(i,:),'bo-'); 
+  xlabel('\psi')
+  ylabel(['density gradient species ' int2str(i)]); 
+  title('R/L_N')
+
+end; 
+
+for i = 1: ns 
+
+  subplot(ns,5,5*(i-1)+4); 
+  plot(profiles.xgr,profiles.rlt(i,:),'rx-'); 
+  xlabel('\psi')
+  ylabel(['temperature gradient species ' int2str(i)]); 
+  title('R/L_T');
+
+end; 
+
+for i = 1: ns 
+
+  subplot(ns,5,5*(i-1)+5); 
+  plot(profiles.xgr,profiles.uprim(i,:),'mx-'); 
+  xlabel('\psi')
+  ylabel(['velocity gradient species ' int2str(i)]); 
+  title('u^\prime'); 
+
+end; 

### 2022-05-03 gkw/matlab/plot_flux.m

@@ -1,0 +1,119 @@
+% Plots the fluxes of a certain structure 
+%
+% Usage : plot_flux(in) 
+% where 'in' is the structure that contains the fluxes 
+
+function plot_flux(in) 
+
+  % Get the screensize and generate the figure 
+  scrsz = get(0,'ScreenSize');
+  figure('Position',[1 3*scrsz(4)/4 3*scrsz(3)/4 scrsz(4)/2])
+
+  % The symbols of the lines use (append if necessary)
+  symbol = ['r-' 'b-' 'g-' 'm-' 'k-' 'r--' 'b--' 'g--' 'm--' 'k--' 'r.-' 'b.-' 'g.-' 'm.-' 'k.-' ]; 
+  
+  % Plot the heat fluxes (Electro-static is always assumed to exist)
+  subplot(2,2,1); 
+  nspec = size(in.qfluxes,2); 
+  for i = 1: nspec 
+    plot(in.time(:,1),in.qfluxes(:,i),symbol(2*i-1:2*i)); 
+    npnt = size(in.qfluxes,1); 
+    label = ['ES' int2str(i)]; 
+    text(in.time(npnt,1),in.qfluxes(npnt,i),label); 
+    hold on;
+  end 
+  % Would be easier in future just to check for existance of file
+  if isfield(in.CONTROL,'nlapar'); 
+    if (in.CONTROL.nlapar(1)==1) 
+      for i = 1: nspec 
+        plot(in.time(:,1),in.qfluxem(:,i),symbol(2*i-1+2*nspec:2*i+2*nspec)); 
+        npnt = size(in.qfluxem,1); 
+        label = ['EM' int2str(i)]; 
+        text(in.time(npnt,1),in.qfluxem(npnt,i),label); 
+      end 
+    end 
+  end 
+  if isfield(in.CONTROL,'nlbpar'); 
+    if (in.CONTROL.nlbpar(1)==1) 
+      for i = 1: nspec 
+        plot(in.time(:,1),in.qfluxbpar(:,i),symbol(2*i-1+4*nspec:2*i+4*nspec)); 
+        npnt = size(in.qfluxbpar,1); 
+        label = ['BP' int2str(i)]; 
+        text(in.time(npnt,1),in.qfluxbpar(npnt,i),label); 
+      end 
+    end    
+  end 
+  title('Heat fluxes'); 
[…]

### 2022-05-03 gkw/matlab/read_gkw_all.m

@@ -1,0 +1,100 @@
+function out=read_gkw_all(file,proj,optnstart,optnend,optamin,optBref)
+% Attempts to read everything possible from a project folder
+% into a the ULTIMATE single structure for a GKW NL run
+%
+% Usage: out=read_gkw_all(file,proj,optnstart,optnend,optamin,optBref)
+%
+% optnstart = timestep to begin averaging (default: mid point)
+% optnend   = timestep to end averaging (default: end of run)
+%
+% Calculates the transport coefficients in GKW units, GB units, and SI units
+% Uses values of Tref, nref and Rref from COLLISIONS for conversions
+% Calculation of SI units requires Bref (from hamada file, or optBref)
+% Calculation of GB units requires rminor a (from hamada file, or optamin)
+%
+% WARNING: The [q/v]fluxes in the structure are not identical to those 
+% in the gkw fluxes file, here they are corrected by T_R/v_R factors
+%
+% WARNING: The dimensional values should be used with caution, 
+% and the user should understand how they are calculated: 
+% It is advised to perform independent checks on the numbers returned.
+
+%Read input.out file for all namelists
+out=read_gkwinput(file,proj,1);
+out.file=file;
+out.proj=proj;
+
+% Alternate way to provide the dVdr information
+% Read input file for comments in GEOM_PLUS namelist
+% in=read_gkwinput(file,proj,0);
+% if isfield(in,'GEOM_PLUS')
+%   out.GEOM_PLUS=in.GEOM_PLUS;
+% else
+%   out.GEOM_PLUS.dvdr=NaN; 
+% end
+
+if exist('opta','var'); out.amin=optamin; end
+if exist('optBref','var'); out.Bref=optBref; end  
+if exist('optnstart','var');  out.nstart=optnstart; end
+if exist('optnend','var');  out.nend=optnend; end
+
+out=read_gkwflux(file,proj,out);
+
+% Use radial fluxes subset instead of averaged flux
+%out.qfluxes=out.qfluxes_rav;
+
+try
+  %out.hamada=read_hamada(out.GEOM.eqfile(14:end),gkwpath('hamada',proj));
[…]

### 2022-05-03 gkw/matlab/read_gkw_version.m

@@ -1,0 +1,10 @@
+function [integer string] = read_gkw_version(file,proj)
+% [integer string] =read_gkw_version(file,proj)
+% reads the GKW version number used for a run from input.out
+% uses the gkwnlin folder structure
+    
+    [err str]=system(['grep -i "GKW version" ' gkwpath('input_out',proj) file]);
+    string = str(40:end);
+    integer = str2num(str(40:44));
+
+end

### 2022-05-03 gkw/matlab/read_gkwcfphi.m

@@ -1,0 +1,49 @@
+function outdens = read_gkwcfphi(sim_name, pth);
+%
+% function outdens = read_gkwcfdens(sim_name, pth);
+%
+
+if ~exist('pth');
+[a pth] = unix('echo $GKWMAT_DIR');
+end
+if isempty(pth);
+clear pth
+[a pth] = unix('echo $GKWMAT_DIR');
+end
+pthsim = [pth(1:end-1) '/cfphi/'];
+pthsimgeo = [pth(1:end-1) '/geom/'];
+
+clear flnm
+flnm = [pthsim sim_name];
+fptr = fopen(flnm,'r');
+if fptr == -1
+disp(['No such simulation in ' pth]);
+disp(['These are the simulations available']);
+[a avsims] = unix(['ls ' pthsim]);
+disp(avsims)
+ystr = -1;
+error(sim_name)
+%return;
+end;
+fclose(fptr);
+
+eval(['load ' flnm ';']);
+eval(['fff = ' sim_name ';']);
+%
+size(fff);
+%
+%outdens.s = fff(:,1);
+%
+[a b] = size(fff);
+%nb_of_species = (b-2)./2;
+outdens.phi = fff(:,1);
+outdens.dphi_dpsi= fff(:,2);
+%
+outdens.sim_name = sim_name;
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+
+
+
+

### 2022-05-03 gkw/matlab/read_gkwfluxes.m

@@ -1,0 +1,62 @@
+% read the detailed fluxes output for a GKW run
+% function [grids,pflux,eflux,vflux]=read_gkwfluxes(flnm,proj)
+% Inputs:
+%	flnm:	file name
+%	proj:	project name (optional)
+%		path for the input files is obtained from the gkwpath function with "proj" in argument
+%	full:	full=1 -> load time arrays, full=0 -> load only the last time step
+%
+% Warning: in this version, all quantities are normalised as in GKW !!!!!!!
+
+function [grids,pflux,eflux,vflux]=read_gkwfluxes(flnm,proj)
+
+if ~exist('proj') 
+	proj=[];
+end
+if exist('full')~=1 
+	full=0;
+end
+
+flpth=gkwpath('fluxes_det',proj);
+
+% read grids
+if unix(['test -e ' flpth flnm '.g'])==0
+	fid = fopen([flpth flnm '.g'], 'r');
+else	
+	error(['The file ' flpth flnm '.g does not exist' ])
+end
+frewind(fid);
+
+nmod=fread(fid,1,'long');
+nx=fread(fid,1,'long');
+nsp=fread(fid,1,'long');
+nmu=fread(fid,1,'long');
+nvpar=fread(fid,1,'long');
+ns=fread(fid,1,'long');
+
+grids.kthrho = fread(fid,nmod,'double');
+grids.kxrho = fread(fid,nx,'double');
+grids.mu = fread(fid,nmu,'double');
+grids.vpar = reshape(fread(fid,ns*nmu*nvpar,'double'),ns,nmu,nvpar);
+grids.s = reshape(fread(fid,nx*ns,'double'),nx,ns);
+
+grids.intmu = fread(fid,nmu,'double');
+grids.intvp = reshape(fread(fid,ns*nmu*nvpar,'double'),ns,nmu,nvpar);
+grids.ints = fread(fid,ns,'double');
+
+fclose(fid);
+
+
+% read file
+if unix(['test -e ' flpth flnm])==0
+	fid = fopen([flpth flnm], 'r');
+else	
+	error(['The file ' flpth flnm ' does not exist' ])
+end
+frewind(fid);
+
+pflux = reshape(fread(fid,nmod*nx*nsp*ns*nmu*nvpar,'double'),nmod,nx,nsp,ns,nmu,nvpar);
[…]

### 2022-05-03 gkw/matlab/read_gkwoutput.m

@@ -1,0 +1,99 @@
+% read the outputs for a linear gkw run
+%	function [time, anrate, freq, s, phi, apar, n, Tpar, Tper, wflow, thetasq_av, kperp_av,  GKWin, pflux, eflux, vflux]=read_gkwoutput(flnm,proj,full)
+% Inputs:
+%	flnm:	file name
+%	proj:	project name (optional)
+%		path for the input files is obtained from the gkwpath function with "proj" in argument
+%	full:	full=1 -> load time arrays, full=0 -> load only the last time step
+%Note1: the fluxes are still semi-normalised, i.e. the species dependent normalisation (factors n_s, n_s*T_s and m_s*n_s*vth_s for pflux, eflux, vflux, respectively) has been removed but the vthref*rhostar^2 factor is still there. 
+%Note2: the electrostatic potential is rotated in the complex plane such as to have real(phi(s=0))=max(real(phi)) and imag(phi(s=0))=0. Same representation than GS2.
+% The non rotated potential (i.e. the one given as a GKW output) is given by (phi(:,1) - i*phi(:,2))./phi_norm
+
+function [time, anrate, freq, s, phi, apar, n, Tpar, Tper, wflow, thetasq_av, kperp_av,  GKWin, pflux, eflux, vflux,phi_norm]=read_gkwoutput(flnm,proj,full)
+
+if ~exist('proj') 
+	proj=[];
+end
+if exist('full')~=1 
+	full=0;
+end
+if full==1,
+	range='1:end';
+else
+	range='end';
+end
+
+
+tmp=load([gkwpath('time',proj)  flnm]);
+eval(['time=tmp(' range ',1);']);
+%Nonlinear runs don't have a growth rate
+if size(tmp,2)>1
+eval(['anrate=tmp(' range ',2);']);
+else
+  anrate = time;
+end
+
+if size(tmp,2)>2 
+  eval(['freq=tmp(' range ',3);'])
+else
+  freq = NaN.*time;
+end
+
[…]

### 2022-05-03 gkw/matlab/rh_plus.m

@@ -1,0 +1,12 @@
+%  rh_plus(eps,q)
+%
+%  Returns Xiao-Catto result for Rosenbluth Hinton GAM residual given q and eps
+%  use with (e.g.) fplot(@(x) rh_plus(x,1.5),[0 1])  
+%
+function [residual] = rh_plus(eps,q)
+ 
+  Odum= 1.6*eps^(3/2)+0.5*eps*eps+0.36*eps^(5/2);
+  
+  residual= 1/(1+q^2*Odum/(eps*eps));
+
+end

### 2022-05-03 gkw/matlab/spectra.m

@@ -1,0 +1,120 @@
+% This file defines a matlab function which plots the averaged spectrum
+% against time on a log plot, useful for nonlinear runs.
+% 
+% usage: spectra('proj,'file','spectrum',sp,tstart,tend,optline)
+%
+% Where spectrum is one of
+%
+% kyspec, eflux, pflux, vflux, kyspec_em, eflux_em, pflux_em, vflux_em
+% kxspec, eflux_kx, pflux_kx, vflux_kx, kxspec_em, eflux_em_kx, pflux_em_kx, vflux_em_kx
+%
+% sp is the species number
+%
+% tstart and tend are optional data points to average from (default last 2)
+%
+% The spectrum in file is plotted against  file.krho values
+% Hence this script assumes the naming conventions of the script gkwnlin and gkwpath.m
+
+
+function []=spectra(proj,file,spec,sp,from,to,line)
+
+if ~exist('spec','var')
+   spectrum='kyspec'
+else
+   spectrum=spec
+end
+
+if ~exist('line','var')
+   line='-';
+end
+
+input=read_gkwinput(file,proj,0);
+nx=input.GRIDSIZE.nx;
+nmod=input.GRIDSIZE.nmod;
+
+switch spectrum
+    case {'kxspec', 'eflux_kx', 'pflux_kx', 'vflux_kx','kxspec_em','eflux_em_kx','pflux_em_kx','vflux_em_kx','kxvort'}
+    scale=['spectrum/kxspec/' file '.kxrh'];
+    legend='k_\psi \rho_i';
+    if exist([gkwpath('root',proj) scale],'file')
+     sc=importdata([gkwpath('root',proj) scale])/sqrt(2);
+    elseif exist([gkwpath('kxrh',proj) file],'file')
+     sc=importdata([gkwpath('kxrh',proj) file])/sqrt(2);
+    else
+     lx=input.GRIDSIZE.lx;
+     kxmin=2*pi/lx
+     kxmax=kxmin*nx/2
+     sc(1:nx/2)=linspace(0,kxmax,nx/2);
+     sc(nx/2+1:nx)=linspace(-kxmax,0,nx/2);
+    end
+        
+    sc=sc';
+    semilog=1;
+    sign=1
+    %sign=-1
+    
+    otherwise
+    scale=['spectrum/kyspec/' file '.krho'];
[…]

### 2022-05-03 gkw/matlab/spectra_ratio.m

@@ -1,0 +1,82 @@
+% This file defines a matlab function which plots the RATIO of time averaged
+% spectra, useful for nonlinear runs.  This script assumes the naming
+% conventions of the script gkwnlin.
+% 
+% spectra_ratio(proj,runname,flux,flux2,species,species2,first,last)
+%
+% WHERE:
+%
+% proj and runname are strings for gkwpath.m
+%
+% species (integer) corresponds to the species no. for which you want the flux,
+% in the order of the input file
+%
+% first and last select the timesstep range for averaging
+%
+% flux selects the flux you want (same names as spectra diagnostic). 
+% Species selects the species you want
+%
+% normalize (integer) = 1 will normalize the output to total flux.
+%
+% EXAMPLE: spectra_ratio('proj','runname','pflux','eflux',2,2,100,200)
+%
+% The spectrum in file is plotted against  runname.krho values in /spectrum/kyspec
+
+
+function spectra_ratio(proj,filename,flux,flux2,species,species2,first,last)
+
+if ~exist('proj')
+	proj='default';
+    disp('You must provide the project name')
+    return;
+    %No default value
+end
+if ~exist('filename')
+    disp('You must provide the run name')
+    filename='default';
+    return;
+end
+if ~exist('flux')
+    disp('You must choose flux')
+    return;
+end
+if ~exist('flux2')
+	disp('You must choose flux2')
+    return;
+end
+
+input=read_gkwinput(filename,proj);
+nmod=input.GRIDSIZE.nmod
+
+%Select columns for chosen species
+start=1+(nmod*(species-1));
+finish=nmod*species-1;
+
+start2=1+(nmod*(species2-1));
+finish2=nmod*species2-1;
+
+scale=load([gkwpath('kyspec',proj) filename '.krho']);
+%scale=2*scale;
+disp(['Loaded ' gkwpath('kyspec',proj) filename '.krho'])
+flux_file=importdata([gkwpath('spectrum',proj) flux '/' filename]);
+flux_file2=importdata([gkwpath('spectrum',proj) flux2 '/' filename]);
+
+a=mean(flux_file(first:last,start:finish));
+err_a=std(flux_file(first:last,start:finish))/sqrt(last-first);
+
+b=mean(flux_file2(first:last,start2:finish2));
[…]

### 2022-05-03 gkw/matlab/time_traces.m

@@ -1,0 +1,119 @@
+%This script makes a matlab function that plots a set of time traces from
+%GKW output onto one plot.  Requires gkwpath to be configured.
+%
+%Usage: time_traces('proj','filename','data',column,optplot,optline)
+%
+%proj is a string with your project name
+%Filename to select which runs of project to plot (accepts wildcards).
+%Data file to plot (default is fluxes) - as goes into gkwpath input
+%Column to plot (default is 2)
+%optplot selects the time vs timestep plot  (optional)
+%optline is the plot line type e.g. '-')   (optional)
+%
+%Example: time_traces('my_proj','runs*','fluxes',2)
+
+function[]=time_traces(proj,filename,data,col,optplot,line)
+ 
+if ~exist('proj','var')
+	proj='default';
+    disp('You must provide the project name')
+    %No default value
+end
+if ~exist('filename','var')
+    %list all
+    filename='';
+end
+if ~exist('data','var')
+    data='fluxes'
+end
+if ~exist('optplot','var')
+    optplot=0;
+end
+if ~exist('line','var')
+    line='-';
+end
+
+
+count=0;            %Counter for output files
+no_files=0;         %Counter for input files
+
+% if ~exist('names')
+%  %files=dir([gkwpath('root',proj) 'good/' filename]);
+  files=dir([gkwpath('time',proj) filename]);
+  total = size(files)
+% else
+%  files=names 
+%  [files(1:size(files')).isdir]=deal(0);
+%  total = size(files')
+% end
+
+if optplot ==0; h=figure; end
+hold all
+
+for i = 1:total(1)
+   
+   if(files(i).isdir==0)
+        no_files=no_files+1;
+        
+        %input=read_gkwinput(files(i).name,proj);
+        %gkwpath(data,proj)
+        disp([gkwpath(data,proj) files(i).name]);
+        if (exist([gkwpath(data,proj) files(i).name],'file')==2)
+            files(i).name;
+            count = count+1;
+            time=load([gkwpath('time',proj) files(i).name]);
+            datafile=load([gkwpath(data,proj) files(i).name]); 
+                        
+            len=min(length(time),length(datafile));
+            
+            %These lines renormalise to gyrobohm if desired
+            %time(:,1)=time(:,1)/(sqrt(2)*3);
+            %data(:,col)=data(:,col)/9.545;
+            
+            hold all;
+            
+            %Add the plot
+            if (length(time) > 1)
+                %if (~isnan(datafile(end,col)) && datafile(end,col) < 1e3) 
[…]

### 2022-05-03 gkw/octave/FL_Diff.m

@@ -1,0 +1,75 @@
+%*********************************************************************************************************
+%This program calculates the diffusion coefficient for a single fieldline and averages over all fieldlines
+%The output is a plot of the average against the number of turns.
+%It was written by Benjamin Schobert.
+%*********************************************************************************************************
+
+%Inputs:                Poincare1.mat (as created by FL_Tracing.m)
+
+%Outputs:               Diffusion1.mat (contains all diffusion coefficients and the average)
+%                       figure(1) (plot of average against number of turns)
+
+
+clear;
+clf(figure(2));
+
+
+%Set to 1 for saving all plots as eps files with latex code
+save_plot = 0;
+
+
+poincare = load('Poincare1.mat');
+ly = poincare.ly;
+diff_psi_n = poincare.diff_psi_n;               %This array contains all differences for psi for each turn
+diff_zeta_n = poincare.diff_zeta_n.*ly;
+size = size(diff_psi_n);
+num_fl = size(1);                               %number of field lines
+nturns = size(2);                               %number of toriodal turns
+
+psi_dico_n(num_fl,nturns) = 0;                  %stores all the diffusion coefficients
+zeta_dico_n(num_fl,nturns) = 0;
+psi_dico_av(nturns) = 0;                        %stores the average of the diffusion coefficients
+zeta_dico_av(nturns) = 0;
+
+psi_dico_n = (diff_psi_n .* diff_psi_n);
+zeta_dico_n = (diff_zeta_n .* diff_zeta_n);
+
+for k=1:nturns
+  psi_dico_n(:,k) = psi_dico_n(:,k)/(k);
+  zeta_dico_n(:,k) = zeta_dico_n(:,k)/(k);
+end
+
+psi_dico_av = mean(psi_dico_n);                 %average for each turn
+zeta_dico_av = mean(zeta_dico_n);
+
+figure(2);                                      %plot the average
+hold on;
+plot(psi_dico_av,'r');
+plot(zeta_dico_av,'k');
+%plot(zeta_dico_n([55,72],:)');
+%plot(zeta_dico_n');
+%plot(psi_dico_n','k');
+legend ({'$\psi$', '$\zeta$'});
+xlabel('number of turns')
+ylabel('normalized diffusion coefficient');
+
+save('Diffusion1.mat','psi_dico_n','psi_dico_av','zeta_dico_n','zeta_dico_av');
+if(save_plot)
[…]

### 2022-05-03 gkw/octave/FL_Tracing2.m

@@ -1,0 +1,583 @@
+%*******************************************************************************************************************
+%This programm integrates and thereby traces fieldlines in a tokamak reactor.
+%It creates a Poincare plot in the outboard midplane and stores the 3d cartesian coordinates of a specific fieldline.
+%It was written by Benjamin Schobert
+%*******************************************************************************************************************
+
+%Inputs:        xphi
+%               geom.dat
+%               Apara00000***
+%               kzeta                   (only for local runs)
+%               lxn                     (only for local runs)
+
+%Outputs:       Poincare1.mat           contains:
+%               psi_n                   all psi coordinates for each turn
+%               zeta_n                  all zeta coordinates for each turn
+%               coord_store_cart        the cartesian coordinates for a specific field plot
+%               figure(1)               the Poincare plot in the outboard midplane
+
+clear;                                  %clear memory
+clf(figure(1));                         %clear figure(1) Poincare plot
+clf(figure(3));                         %clear psimaxdisp plot
+
+rhostar = 5e-3;                         %normalised gyro radius (only for global runs)
+factor_s=1;                             %refinement factor in s direction
+factor_r=2;                             %refinement factor in psi and zeta direction
+nmod = 21;                              %number of modes
+mn=1;                                   %only for linear runs
+nturns = 100;                           %number of turns
+fl_3d = 5;                              %this is the number of the field line, whose 3d coordinates will be saved
+dotsize = 2;                            %this is the dotsize in the Poincare plot. I recommend 2 for octave plots
+                                        %and 0.5 for png oder pdf or eps
+nfields = 1;                            %number of different A_para fields that are used (only for time dependency)
+napara = 200;                           %number of the Apara file (Apara00000***)
+                                        
+%Flags
[…]

### 2022-05-03 gkw/octave/create_plot_comparison_profiles.m

@@ -1,0 +1,84 @@
+graphics_toolkit('gnuplot')
+
+# radial grid
+psi=0.03:0.001:0.3;
+
+# parameters for the profiles
+p=[1.0 5.0 0.15 0.1 0.1];
+
+x=min(max(psi, p(3)-p(4)/2), p(3)+p(4)/2);
+
+# defining the profiles (temp./density) and the derivatives
+cos2 = p(1)*exp(-p(2)*(x-p(3)) + p(2)*p(5)*tanh((x-p(3)+p(4)/2)/p(5)) + p(2)*p(5)*tanh((x-p(3)-p(4)/2)/p(5)));
+dcos2 = p(2)*(1 - 1./(cosh((x-p(3)+p(4)/2)/p(5))).^2 - 1./(cosh((x-p(3)-p(4)/2)/p(5))).^2);
+
+const = p(1)*ones(size(psi));
+dconst = zeros(size(psi));
+
+exptanh = p(1)*exp(-p(2)*p(4)*tanh((psi-p(3))/p(4)));
+dexptanh= p(2)*1./(cosh((psi-p(3))/p(4))).^2;
+
+mexptanh = 1 - p(1)*exp(-p(2)*p(4)*tanh((psi-p(3))/p(4)));
+dmexptanh= -p(2)*1./(cosh((psi-p(3))/p(4))).^2;
+
+exppoly3=p(1)*exp(p(2)*(p(4)/3*(psi-p(3)).^3 + p(5)/3*(psi-p(3)).^2 + (psi-p(3))));
+dexppoly3= p(2)*(p(4)*(psi-p(3)).^2 + p(5)*(psi-p(3)) + 1);
+
+exppoly6=p(1)*exp(p(2)*(p(4)/6*(psi-p(3)).^6 + p(5)/4*(psi-p(3)).^4 + (psi-p(3))));
+dexppoly6=p(2)*(p(4)*(psi-p(3)).^5 + p(5)*(psi-p(3)).^3 + 1);
+
+%orb=;
+%dorb=;
+
+orb3= p(1)*exp(-0.5*p(2)*p(5)*log(cosh((psi-p(3)+p(4))/p(5))./cosh((psi-p(3)-p(4))/p(5))));
+dorb3=0.5*p(2)*(tanh((psi-p(3)+p(4)/2)/p(5)) - tanh((psi-p(3)-p(4)/2)/p(5)));
+
[…]

### 2022-05-03 gkw/octave/plot_matrix.m

@@ -1,0 +1,90 @@
+#!/usr/bin/env octave
+
+% This is a standalone GNU Octave script.
+
+
+% You should add
+%   addpath("~/gkw/octave");
+% to your ~/.octaverc
+% or whatever your GKW_HOME folder is, in order to use the
+% files where often used functions are defined.
+libgkwoct;
+
+init()
+
+set(0,'defaultfigurepaperposition',[0 0 660 450]*0.5)
+
+%%%%%%%% LOAD DATA
+pkg load hdf5oct
+
+for dsetname = {'mat','mat_maxwll_background', 'mat_poisson','mat_field_diag'}
+    rank = 0
+    while(true)
+      try
+        dname=[dsetname{1},'_',num2str(rank,'%03d')]
+        ## irs = reshape(irs,[],1);
+        mat = complex(h5read("gkwdata.h5",["/diagnostic/diagnos_matrix/",dname,"_values_real"]),
+	              h5read("gkwdata.h5",["/diagnostic/diagnos_matrix/",dname,"_values_imag"]));
+        nmat=length(mat)
+        jj = h5read("gkwdata.h5",["/diagnostic/diagnos_matrix/",dname,"_jj"]);
+        length(jj)
+        catch
+          break
+      end
+
+        try
+          irs = h5read("gkwdata.h5",["/diagnostic/diagnos_matrix/",dname,"_irs"]);
+
+          nrows = length(irs)-1
+          ## irs(end+1) = 0;
+          [w,iw] = max(irs)
+
+          row = [];
+          col = [];
+          val = [];
+
+          for i = 1:nrows
+            for j = irs(i):(irs(i+1)-1)
+              %disp(['element',num2str(i),num2str(jj(j)),'is', num2str(mat(j))])
+              try
+                row(end+1) = i;
+                col(end+1) = jj(j);
+                val(end+1) = mat(j);
+              catch
+                lasterror
+                exit
+              end
+            end
+          end
+
+        catch
+          ii = h5read("gkwdata.h5",["/diagnostic/diagnos_matrix/",dname,"_ii"]);
+          row = ii;
+          col = jj;
+          val = mat;
+          lasterror
+        end
+        sparse_mat = sparse(row,col,val);
+
+        %% display some information
+        %spstats(sparse_mat)
+        disp('Matrix type:')
+        matrix_type(sparse_mat)
+
+
+        ## spy(real(sparse_mat))
[…]

### 2022-05-03 gkw/python/convert_clip_global_rad_domain.py

@@ -1,0 +1,190 @@
+#!/usr/bin/env python3
+import numpy as np
+import sys
+import h5py
+import argparse
+import shutil
+import os
+import scipy.interpolate
+
+try:
+    import namelist_python
+except:
+    # if the GKW_HOME/python folder is not contained in the PYTHONPATH
+    import os
+    sys.path.append(os.path.join(os.environ['GKW_HOME'],'python'))
+    import namelist_python
+
+import gkwlib
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser(
+        description="Clip the radial domain of a GKW global input file and all input profiles in the associated input.prof file.")
+    parser.add_argument('--input-file',
+                        help="GKW local input file (input file)", 
+                        type=str,
+                        default='input.dat'
+    )
+    parser.add_argument('--inputprof-file',
+                        help="GKW input.prof file (input file)", 
+                        type=str,
+                        default='input.prof'
+    )
+    parser.add_argument('--psil',
+                        help="lower boundary of the radial grid", 
+                        type=float,
+                        default=None
+    )
+    parser.add_argument('--psih',
+                        help="upper boundary of the radial grid", 
+                        type=float,
+                        default=None
+    )
+    parser.add_argument('--keep-n-gridpoints',
+                        help="Set True, to keep the original number of radial gridpoints. Set False, to keep the original resolution, i.e. decrease the number of gridpoints proportionally. ", 
+                        type=bool,
+                        default=False
+    )
+
+    parser.add_argument('--avoid-primes',
+                        help="Set True, to adjust psil or psih automatically by a small amount, to avoid a prime number of gridpoints. Note that the given (not the adjusted!) values will be used to name the output folder", 
+                        type=bool,
+                        default=True
+    )
+
+    parser.add_argument('--output-folder',
+                        help="output folder (will be created if not existent). This is where the new input.dat and input.prof will be put. You may use {n_x_grid} and {factor} to format n_x_grid or the factor into the folder name.", 
+                        type=str,
+                        default=None
+    )
+    
+    args = parser.parse_args()
+
+    if(args.output_folder is None):
+        args.output_folder = ""
+        if(args.psil is not None):
[…]

### 2022-05-03 gkw/python/convert_global_rad_resolution.py

@@ -1,0 +1,121 @@
+#!/usr/bin/env python3
+import numpy as np
+import sys
+import h5py
+import argparse
+import shutil
+import os
+import scipy.interpolate
+
+try:
+    import namelist_python
+except:
+    # if the GKW_HOME/python folder is not contained in the PYTHONPATH
+    import os
+    sys.path.append(os.path.join(os.environ['GKW_HOME'],'python'))
+    import namelist_python
+
+import gkwlib
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser(
+        description="Change the radial resolution of a GKW global input file, and its profiles in the associated input.prof file.")
+    parser.add_argument('--input-file',
+                        help="GKW local input file (input file)", 
+                        type=str,
+                        default='input.dat'
+    )
+    parser.add_argument('--inputprof-file',
+                        help="GKW input.prof file (input file)", 
+                        type=str,
+                        default='input.prof'
+    )
+    group = parser.add_mutually_exclusive_group(required=True)
+    group.add_argument('--n_x_grid',
+                        help="new number of radial grid points", 
+                        type=int,
+                        default=None
+    )
+    group.add_argument('--factor',
+                        help="factor, to augment number of radial grid points", 
+                        type=float,
+                        default=None
+    )
+
+    parser.add_argument('--output-folder',
+                        help="output folder (will be created if not existent). This is where the new input.dat and input.prof will be put. You may use {n_x_grid} and {factor} to format n_x_grid or the factor into the folder name.", 
+                        type=str,
+                        default=None
+    )
+    
+    args = parser.parse_args()
+
+    if(args.n_x_grid is not None):
+        if(args.output_folder is None):
+            args.output_folder = 'n_x_grid_{n_x_grid}'
+        foldername = args.output_folder.format(n_x_grid=args.n_x_grid, factor=args.factor)
+    else:
+        if(args.output_folder is None):
+            args.output_folder = 'n_x_grid_factor_{factor}'
+        foldername = args.output_folder.format(n_x_grid=args.n_x_grid, factor=args.factor)
+
+    # parse the file into a Namelist object, and take the groups member of that.
+    nlist = namelist_python.read_namelist_file(args.input_file)
[…]

### 2022-05-03 gkw/python/convert_inputfile_localglobal.py

@@ -1,0 +1,224 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import sys
+import h5py
+import argparse
+import shutil
+import os
+
+try:
+    import namelist_python
+except:
+    # if the GKW_HOME/python folder is not contained in the PYTHONPATH
+    import os
+    sys.path.append(os.path.join(os.environ['GKW_HOME'],'python'))
+    import namelist_python
+
+import gkwlib
+
+TINY_NUMBER = 1.0e-10
+    
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser(
+        description="Convert a GKW global input file to a local input file.")
+    parser.add_argument('--local-input-file',
+                        help="GKW local input file (output file)", 
+                        type=str,
+                        default='-'
+    )
+    parser.add_argument('--global-input-file',
+                        help="GKW local input file (input file)", 
+                        type=str,
+                        default='input.dat'
+    )
+    parser.add_argument('--gkwdatah5-file',
+                        help="a GKW HDF5 data file (input file)", 
+                        type=str,
+                        default='gkwdata.h5'
+    )
+    parser.add_argument('--geomdat-file',
+                        help="GKW geom.dat file (input file)", 
+                        type=str,
+                        default='geom.dat'
+    )
+    parser.add_argument('--inputprof-file',
+                        help="GKW input.prof file (input file)", 
+                        type=str,
+                        default='input.prof'
+    )
+    group = parser.add_mutually_exclusive_group(required=True)
+    group.add_argument('--radial-ix',
+                        help="radial index (0-based)", 
+                        type=int,
+                        default=None
+    )
+    group.add_argument('--radial-x',
+                        help="radial x position", 
+                        type=float,
+                        default=None
+    )
+    args = parser.parse_args()
+
+    if(args.local_input_file == '-'):
+        output = sys.stdout
+    else:
+        output = open(args.local_input_file, 'w')
+
+    # parse the file into a Namelist object, and take the groups member of that.
+    nlist = namelist_python.read_namelist_file(args.global_input_file)
+    n = nlist.groups
+    # that object is an ordered case-insensitive dictionary
+
+    # sanity check
[…]

### 2022-05-03 gkw/python/convert_mass_ratio.py

@@ -1,0 +1,239 @@
+#!/usr/bin/env python3
+import numpy as np
+import sys
+import h5py
+import argparse
+import shutil
+import os
+
+try:
+    import namelist_python
+except:
+    # if the GKW_HOME/python folder is not contained in the PYTHONPATH
+    import os
+    sys.path.append(os.path.join(os.environ['GKW_HOME'],'python'))
+    import namelist_python
+
+import gkwlib
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser(
+        description="Change the electron/ion mass ratio of a GKW global input file, and the associated input.prof file.")
+    parser.add_argument('--input-file',
+                        help="GKW local input file (default: input.dat)", 
+                        type=str,
+                        default='input.dat'
+    )
+    parser.add_argument('--inputprof-file',
+                        help="GKW input.prof file (default: input.prof)", 
+                        type=str,
+                        default='input.prof'
+    )
+    ELECTRON_SPECIES = "electron"
+    ION_SPECIES = "ion"
+    c = [ELECTRON_SPECIES,ION_SPECIES]
+    parser.add_argument('--species',
+                       help="which species to change. default 'electron' changes electron mass, and leaves ion mass as it is", 
+                       type=str,
+                       choices=c,
+                       default=ELECTRON_SPECIES
+    )
+    
+    group = parser.add_mutually_exclusive_group(required=True)
+    group.add_argument('--mass',
+                        help="new mass", 
+                        type=float,
+                        default=None
+    )
+    group.add_argument('--mass-factor',
+                        help="factor, to change mass", 
+                        type=float,
+                        default=None
+    )
+    group.add_argument('--mass-ratio',
+                        help="change mass, to meet given mass ratio of first species and electron species", 
+                        type=float,
+                        default=None
+    )
+    group.add_argument('--larmor',
+                        help="change mass, to meet given larmor radius", 
+                        type=float,
+                        default=None
+    )
+    group.add_argument('--larmor-factor',
+                        help="change mass, to meet larmor radius times this factor.", 
+                        type=float,
[…]

### 2022-05-03 gkw/python/gkw/__init__.py

@@ -1,0 +1,2 @@
+# vtk is needed for vtktorus, so don't import by default
+#from gkw.vtk.vtktorus import *

### 2022-05-03 gkw/python/intensity.py

@@ -1,0 +1,248 @@
+#!/usr/bin/env python3
+
+import matplotlib.pyplot as plt
+import numpy as np
+import h5py
+
+try:
+    # derive the plot object from the class which is part of the
+    # gkwgui package, if it is available.
+    import gkwgui.plottable
+    Plottable = gkwgui.plottable.Plottable
+except ImportError:
+    Plottable = object
+
+def find_convergence_time(h5filename, data):
+    f = h5py.File(h5filename, "r")
+    non_linear = f['/'].attrs['control.non_linear'][0] == b'T'
+    time = f['/grid/time'].value.flatten()
+    import scipy.interpolate
+    # obtain a smoothed cubic spline representation of the curve
+    smoothing_cond = 1.0e0
+    tck = scipy.interpolate.splrep(time,data, k=3, s=smoothing_cond)
+    # evaluate the profile
+    smoothed_data = scipy.interpolate.splev(time, tck, der=0)
+    smoothed_data_1stderiv = scipy.interpolate.splev(time, tck, der=1)
+    # approximately determine the time where the step occurs. this is where the first derivative its large bump
+    sm_dat_1stderiv_center_of_mass = time.dot(smoothed_data_1stderiv)/smoothed_data_1stderiv.sum()
+    data = smoothed_data
+
+    # now fit a Gaussian bell curve to the derivative of the smoothed
+    # timetrace, in order to determine its width, and to avoid the
+    # linear timeintervall properly.
+    
+    def gauss_bell(x, mu, sigma, norm): return norm/(sigma*np.sqrt(2*np.pi)) * np.exp(-0.5*((x-mu)/sigma)**2)
+    
+    import scipy.optimize
+
+    # now fit a gauss bell
+    time_crit_0 = sm_dat_1stderiv_center_of_mass
+    if(time_crit_0 >= time[-1]):
+        time_crit_0 = time[0]
+    amp_0 = data[-1]
+    sigma_0 = 10.0
+    
+    popt, pcov = scipy.optimize.curve_fit(gauss_bell, time, smoothed_data_1stderiv,
+                                          p0=(time_crit_0, sigma_0, amp_0),
+                                          # sigma=data*0.0,
+                                          # absolute_sigma=True,
+                                          #bounds=([0.0,0], [time[-1],np.max(data)]),
+    )
+    time_crit_opt, sigma_opt, amp_opt = popt
+
+    time_crit_opt += sigma_opt*3.0
[…]

### 2022-05-03 gkw/python/namelist_python/__init__.py

@@ -1,0 +1,3 @@
+
+__all__ = ['namelist']
+from .namelist import *

### 2022-05-03 gkw/python/namelist_python/namelist.py

@@ -1,0 +1,640 @@
+#!/usr/bin/python3
+
+import unittest
+try:
+    from collections import OrderedDict
+except ImportError:
+    from .utils import OrderedDict
+
+import re
+import sys
+
+def read_namelist_file(filename):
+    return Namelist(open(filename, 'r').read())
+
+# trick for py2/3 compatibility
+if 'basestring' not in globals():
+   basestring = str
+
+class AttributeMapper():
+    """
+    Simple mapper to access dictionary items as attributes.
+    """
+
+    def __init__(self, obj):
+        self.__dict__['data'] = obj
+
+    def __getattr__(self, attr):
+        if attr in self.data:
+            found_attr = self.data[attr]
+            if isinstance(found_attr, dict):
+                return AttributeMapper(found_attr)
+            else:
+                return found_attr
+        else:
+            raise AttributeError
+
+    def __setattr__(self, attr, value):
+        if attr in self.data:
+            self.data[attr] = value
+        else:
+            raise NotImplementedError
+
+    def __dir__(self):
+        return self.data.keys()
+
+class CaseInsensitiveDict(OrderedDict):
+    """
+    This is an ordered dictionary which ignores the letter case of the
+    keys given (if the respective key is a string).
+
+    Many thanks to user m000 who answered
+    https://stackoverflow.com/questions/2082152/case-insensitive-dictionary
+    so helpfully.
+    """
+    
+    @classmethod
+    def _k(cls, key):
+        return key.lower() if isinstance(key, basestring) else key
+
+    def __init__(self, *args, **kwargs):
+        super(CaseInsensitiveDict, self).__init__(*args, **kwargs)
+        self._convert_keys()
+    def __getitem__(self, key):
+        return super(CaseInsensitiveDict, self).__getitem__(self.__class__._k(key))
+    def __setitem__(self, key, value):
+        super(CaseInsensitiveDict, self).__setitem__(self.__class__._k(key), value)
+    def __delitem__(self, key):
+        return super(CaseInsensitiveDict, self).__delitem__(self.__class__._k(key))
+    def __contains__(self, key):
+        return super(CaseInsensitiveDict, self).__contains__(self.__class__._k(key))
+    def has_key(self, key):
+        return super(CaseInsensitiveDict, self).has_key(self.__class__._k(key))
+    def pop(self, key, *args, **kwargs):
[…]

### 2022-05-03 gkw/python/profiles.py

@@ -1,0 +1,266 @@
+#!/usr/bin/env python3
+
+from mpl_toolkits.mplot3d import Axes3D
+from matplotlib import cm
+from matplotlib.ticker import LinearLocator, FormatStrFormatter
+import matplotlib.pyplot as plt
+import numpy as np
+from numpy import pi
+import h5py
+
+try:
+    # derive the plot object from the class which is part of the
+    # gkwgui package, if it is available.
+    import gkwgui.plottable
+    Plottable = gkwgui.plottable.Plottable
+except ImportError:
+    Plottable = object
+
+import gkwlib
+
+class FlutubeInBackgroundMarkerPlot(Plottable):
+    quantities = ['dens',
+              'temp',
+              'rln',
+              'rlt',
+          ]
+    
+    @classmethod
+    def isPlottable(cls,h5filename):
+        return True
+
+    def __init__(self, h5filename):
+        super().__init__(h5filename)
+
+    def plot(self,ax):
+        import os
+        try:
+            import namelist_python
+        except:
+            # if the GKW_HOME/python folder is not contained in the PYTHONPATH
+            sys.path.append(os.path.join(os.environ['GKW_HOME'],'python'))
+            import namelist_python
+
+        input_dat_filename = os.path.join(os.path.dirname(self.h5filename), 'input.dat')
+        nlist = namelist_python.read_namelist_file(input_dat_filename)
+        n = nlist.groups
+
+        import re
+        with open(input_dat_filename, 'r') as f:
+            m = re.search(r'! *x = ([0-9.+-e]+)', f.read());
+            if(m is None):
+                x = float(self.arguments["x"].get())
+            else:
+                x = float(m.group(1))
+                self.arguments["x"].set("%f (parsed from input.dat)" % x)
+                
+        for isp in range(n['gridsize']['number_of_species']):
+                ax.plot(x, n['species'][isp][self.arguments["dsetname"].get()],
+                        marker='.',
+                        color='black')
+
+        
+    def specifyArguments(self, parser):
+        self.parser.add_argument("--dsetname", help="dataset name",
+                                 choices=self.quantities,
+                                 type=str)
+        self.parser.add_argument("--x", help="flux surface label",
+                                 default=0.5,
+                                 type=float)
+        
+
+class BackgroundInputProfPlot(Plottable):
+    BACKGROUND_CHARGE_DENS = 'background charge density'
+    BACKGROUND_CHARGE_DENS_GRADIENT = 'background charge density gradient'
[…]

### 2022-05-03 gkw/python/reduce_gradient_profile.py

@@ -1,0 +1,144 @@
+#!/usr/bin/env python3
+
+import sys
+try:
+    import gkwlib
+except:
+    # if the GKW_HOME/python folder is not contained in the PYTHONPATH
+    import os
+    sys.path.append(os.path.join(os.environ['GKW_HOME'],'python'))
+    import gkwlib
+
+import os
+import shutil
+import numpy as np
+import scipy.interpolate
+import scipy.integrate
+import copy
+
+import matplotlib.pyplot as plt
+
+def clip_electron_background(complete_input_prof_data, background_dset, max_value):
+    reduction_ratio_profile = None
+    reduction_ratio_grad_profile = None
+    for iblock in range(len(complete_input_prof_data)):
+        if(complete_input_prof_data[iblock][0].startswith("Species")):
+            # check if this is indeed the input.prof entry for this species:
+            mass = complete_input_prof_data[iblock][1][0,0]
+            Z = complete_input_prof_data[iblock][1][0,1]
+            if(Z < 0):
+                # due to the not-so-great file format of
+                # input.prof there is no header line specified
+                # to address the block containing the
+                # background profiles. Instead, it is the next block!
+                xgrid = complete_input_prof_data[iblock+1][1][:,gkwlib.input_prof_background_prof_columns.index("xgr")]
+                icol = gkwlib.input_prof_background_prof_columns.index(background_dset)
+                profile = complete_input_prof_data[iblock+1][1][:,icol]
+                profile = np.clip(profile, -max_value, +max_value)
+                # obtain a smoothed cubic spline representation of the curve
+                smoothing_cond = 1.0e0
+                tck = scipy.interpolate.splrep(xgrid,profile, k=3, s=smoothing_cond)
+                # evaluate the profile
+                new_profile = scipy.interpolate.splev(xgrid, tck, der=0)
+        
+                if(background_dset.startswith("rl")):
+                    # having modified the gradient, now reintegrate the profile
+                    
+                    scalar_profile = complete_input_prof_data[iblock+1][1][:,icol-2]
+                    leftmost_value = scalar_profile[0]
+                    print("Re-integrate the profile starting from leftmost value %f..." % leftmost_value)
+                    
+                    
+                    # Right now, there seems to be a bug in cumtrapz
[…]

### 2022-05-03 gkw/python/rm-data-from-previous-run.py

@@ -1,0 +1,55 @@
+#!/usr/bin/env python
+
+# This little script is useful if one has accidentally appended data
+# to gkwdata.h5 by continuing a run.
+
+# This script first looks for a jump in the time grid, and
+# then it removes the data before that jump from the timetraces,
+# making use of another python script.
+
+
+import h5py
+import bisect
+import sys
+import subprocess
+import os
+
+h5filename = "gkwdata.h5"
+
+
+if __name__ == "__main__":
+    f = h5py.File("gkwdata.h5", "r+")
+
+    try:
+        time_dset = "/grid/time"
+        time = f[time_dset][0,:]
+    except:
+        # it is a (1,n) dimensional array
+        time_dset = "/diagnostic/diagnos_growth_freq/time"
+        time = f[time_dset][0,:]
+
+    try:
+        gkw_home = os.environ['GKW_HOME']
+    except:
+        print("GKW_HOME environment variable is not set.")
+        exit(1)
+
+    # find index where the time jumps to a lower value
+    for i in range(1,len(time)):
+        if(time[i]< time[i-1]):
+            # ok, nailed it.
+            i-=1
+            print("timegrid:" + str(time[i-5:i+5]))
+            print("position:" + str(i))
+            # now copy the data to a backup file
+            print(subprocess.call(["cp", "-v","gkwdata.h5","backup_gkwdata.h5"]))
+            print("Data backup written to : backup_gkwdata.h5")
+            # and remove the chunk of data in front of the time grid jump
+            print("call:    " + " ".join([gkw_home+"/python/clip_gkwdatah5.py", "--remove-before","-n",str(i)]))
+            print(subprocess.call([gkw_home+"/python/clip_gkwdatah5.py", "--remove-before","-n",str(i)]))
+            exit(0)
+
+    print("No jump in " + time_dset + " was found.")
+    
+    
+    

### 2022-05-03 gkw/python/yc/example.py

@@ -1,0 +1,81 @@
+cd /home/yann/codes/gkw/python/yc
+import main 
+import importlib #to reload modules by doing importlib.reload(main)
+
+# create an instance of the run class
+tt=main.gkwrun()
+
+# specify the type of file storage and get the corresponding paths
+gkw_home='/home/yann/runs1/gkw/'
+proj='PFS_Samuele'
+flnm='Waltz_disp_1_1_1'  # linear run
+
+gkw_home='/home/yann/runs1/gkw/'
+proj='TCVREV_3'
+flnm='p2LINhigh_2_2'    # linear run
+
+gkw_home='/home/yann/runs1/gkw/'
+proj='JET_KBMbis'
+flnm='j15_NLfiW_rlte_exb_1'     # non-linear run
+
+
+gkw_home='/home/yann/runs1/gkw/'
+proj='GKDB_TEST'
+flnm='ref_fourier'     # linear run, fourier
+
+tt.set_storage(gkw_home+proj,'multiple_folders',flnm)
+
+# load input files (input.dat and input_out.dat)
+tt.get_inputs()
+
+# load geom
+tt.get_geom()
+
+# load grids 
+tt.get_grids()
+
+# load output file
+tt.get_info()
+
+# load eigenvalues
+tt.get_eigenvalues()
+
+# load collisions
+tt.get_collisions()
+
+# load background density
+tt.get_cfdens()
+
+# load fluxes
+tt.get_fluxes()
+
+# load fields and moments from parallel.dat
+tt.get_parallel()
+
+# load kykxs fields and moments 
+tt.get_kykxs('phi',[103.5,104.2])
+tt.get_kykxs({'apar','Tperp','Tperp_J0'},[103.5,104.2])
+
+# make a GK IDS from GKW inputs and outputs
+tt.gkw2ids([])
+
+
+
+# to do: put all IDS related routines in a separate class
+
+
+# temporary, look how a JSON file is loaded in python
+import json
+with open('/home/yann/projects/gkdb/GENE_GKW/GKW-case-1-1.json') as f:
+ ids=json.load(f)
+
+
+# structure represented as a combination of dicts and lists of dicts
+type(ids['wavevector']) # list
+type(ids['wavevector'][0]) # list
+
+
+    # ids properties (to fill later)
+    self.ids['ids_properties']['comment']=""
+    self.ids['ids_properties']['provider']=""
+    self.ids['ids_properties']['creation_date']=""

### 2022-05-03 gkw/scripts/gkw_daily_tests.sh

@@ -1,0 +1,771 @@
+#!/bin/bash
+
+##############################################################################
+##############################################################################
+#
+##############################################################################
+# Summary:
+# This is a script for enabling periodic tests (e.g. daily) of gkw, when run
+# in a cronjob. See below for an example how to create those.
+#
+##############################################################################
+# Description
+#
+# This script is intended for automatic testing of gkw.
+# For this purpose it is capable of automaticly cloning the git repository
+# (read only)/pulling it. The develop branch is checked out. Then either the
+# default config file is used or looped over all of them for the machine.
+# The standard tests are run. At the end it is tried to make the
+# documentation.
+# The results of all the makes and tests are logged. The results of the tests
+# are also mailed.
+# Command line arguments can alter the behaviour in some respects. See the
+# 'Options' section below.
+#
+##############################################################################
+# Basic usage:
+# Create a cronjob at the machine where the tests should run, for an example
+# how to do this see below.
+# A .bashrc in the home directory is expected to set the needed environment
+# variables. Besides of default ones like PATH,... also the gkw specific
+# GKW_HOME, has to be set.
+# The latter is used to determine, where to put the folder for the tests
+# ($GKW_HOME/../gkw-daily-tests
+# Additionaly also EMAIL_ADDRESS should be set, to an adress, where to send
+# the mails with the result of this script.
+#
+# Comands/programs that must be available, besides of those required by gkw:
+# grep
+# mail
+# rm
+# sed
+#
+##############################################################################
+# Options
+#
+# \attention Long option names not supported yet.
+#
+# -a --all-branches   Check all the branches.
+# -b --branch-blacklist file  Check every branch, except those listed in the
+#                     file given as argument.
+# -c --all-config     Check with every config file in the used config-path.
+#                     Suppose the makefile will use as default
+#                     config/name/default.mk, then every config file in
+#                     config/name/ will be checked for compiling and testing.
+# -h --help           Print help for usage and quit.
+# -q --quiet          Quiet mode, no general logfile.
+# -u --just-update    Just update the branches instead of forcing a clone.
+#    --no-clone
+# -w --branch-whitelist file  Check every branch listed in the file given as
+#                     argument. The filename
+#
[…]

### 2022-05-03 gkw/scripts/gkw_source_in_single_file.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+one_file='gkw_in_one_file.F90'
+rm -f $one_file
+
+# concatenate lines ending with backslash | grep certain line | remove everything up to the equal sign
+for objfile in $(perl -p -e 's/\\\n//' $GKW_HOME/src/objfiles.mk  | grep 'F90OBJ =' | sed 's/^.*=//')
+do
+    cat ${objfile%o}[fF]90 >> $one_file
+done
+
+echo "All GKW source files were concatenated to $one_file"

### 2022-05-03 gkw/scripts/gkw_upgrade_inputfile.pl

@@ -1,0 +1,171 @@
+#!/bin/sh
+#  -*-Perl-*-  (for Emacs)    vim:set filetype=perl:  (for vim)
+#======================================================================#
+# Run the right perl version:
+if [ -x /usr/local/bin/perl ]; then
+  perl=/usr/local/bin/perl
+elif [ -x /usr/bin/perl ]; then
+  perl=/usr/bin/perl
+else
+  perl=`which perl| sed 's/.*aliased to *//'`
+fi  
+#   
+exec $perl -x -S $0 "$@";     # -x: start from the following line
+#======================================================================#
+#! /Good_Path/perl -w 
+# line 17
+#
+#  Update the GKW input files to work with a the code when some switches have
+#  been moved from the CONTROL namelist to DIAGNOSTIC. Takes the local file
+#  `input.dat' and outputs the update as `input.dat.new'.
+#  
+
+use strict;
+
+my $control_nml  = '^\s*\&control\s*';
+my $diagnostic_nml  = '^\s*\&diagnostic\s*';
+my $nml_end = '^\s*\/\s*';
+my @lines_to_move;
+my @moved_vars = qw(
+       output3d
+       lwrite_output1
+       lphi_diagnostics
+       screen_output
+       lfinal_output
+       lcalc_fluxes
+       lcalc_freq
+   );
+my @deleted_vars = qw(
+       ldiagnostic_namelist
+   );
+my $input_file = 'input.dat';
+my $new_file_suffix = '.new';
+
+
+# first find the lines in control to move
+find_lines_to_move($input_file);
+
+# create a new file with the removed/moved lines
+create_new_file($input_file);
+
+
+
+#----------------------------------------------------------------------------
+# find the lines of &CONTROL to move to &DIAGNOSTIC
+# 
+sub find_lines_to_move{
+    my $file = shift;
+    my $check_line = 0;
+	
+    open(INPUT,"<$file") || die "can not open file $file\n";
+	
+	# read through the whole file a line at a time
+    while( my $line = <INPUT> ) {
+		
+        chop($line);
+		# check if we are at the control nml
+        $check_line = 1 if ($line =~ m/$control_nml/i );
+	    $check_line = 0 if ($line =~ m/$nml_end/i);
+		
+		# make a list of the items to move from control
+	    if ($check_line) {
+            foreach my $item (@moved_vars) {
+	            if ($line =~ m/^\s*$item/i ) {
+		            push @lines_to_move,$line;
[…]

### 2022-05-03 gkw/scripts/gkwlib.sh

@@ -1,0 +1,35 @@
+#!/bin/bash
+
+# This file collects useful functions which are used by several of the
+# other scripts shipped with GKW.
+
+# Use it by "sourcing" it:
+# Either write
+#   . $GKW_HOME/scripts/gkwlib.sh
+# or
+#   source $GKW_HOME/scripts/gkwlib.sh
+# at the beginning of your script.
+
+find_latest_executable(){
+    #after calling this function, the variable $latest_executable is set.
+
+    # need gkw home
+    if [ -z "${GKW_HOME}" ]; then
+	echo "You need to set GKW_HOME!"
+	exit 1
+    fi
+
+    
+    # get the network node hostname and strip everything behind the first dot
+    HN=`uname -n | sed -e 's/\..*$//'`
+    # get the first three letters
+    HN3=$(echo ${HN} | awk '{ string=substr($1, 1, 3); print string; }')
+    
+    latest_executable=$(ls -t ${GKW_HOME}/run/gkw*${HN3}*.x | head -1)
+
+    if [ -z "${latest_executable}" ]; then
+	echo "No suitable executables found for ${HN3}"
+	exit 1
+    fi
+}
+

### 2022-05-03 gkw/src/diagnos_eng.f90

@@ -1,0 +1,255 @@
+!------------------------------------------------------------------------------
+!> Energy transfer diagnostic. Perhaps temporary 
+!------------------------------------------------------------------------------
+module diagnos_eng
+
+  implicit none 
+
+  public :: set_default_nml_values
+  public :: init, bcast, check, finalize, allocate_mem
+  public :: initial_output, final_output
+  public :: output
+
+
+  private 
+
+  !> switch for a simple diagnostic on the mode energy (in the potential only)
+  logical, save, public :: lmode_energy 
+
+
+  !> Temporary storage for the distribution function and its time derivative 
+  complex, save, allocatable :: fstore(:), dfdt(:) 
+
+  !> zfs contains the time derivative of abs(phi)^2
+  real,    save, allocatable :: zfs(:), duma(:) 
+
+  ! the file identifier 
+  integer, save :: lun1, lun2  
+
+contains
+
+  !--------------------------------------------------------------------
+  !> Set reasonable default values for the namelist items this
+  !> diagnostic provides. 
+  !--------------------------------------------------------------------
+  subroutine set_default_nml_values()
+    lmode_energy = .false.
+
+    
+  end subroutine set_default_nml_values
+
+
+!------------------------------------------------------------------------------
+!> Change in mode energy 
+!------------------------------------------------------------------------------
+subroutine mode_energy
+
+  use dist,             only : fdisi, nsolc, iphi, ifdis
+  use grid,             only : nmod, nx, ns, parallel_s, parallel_x, nmu 
+  use grid,             only : nvpar, nsp 
+  use index_function,   only : indx 
+  use io,               only : xy_fmt, ascii_fmt, append_chunk
+  use non_linear_terms, only : add_non_linear_terms
+  use mpicomms,         only : COMM_S_NE_X_NE
+  use mpiinterface,     only : mpiallreduce_sum
+  use velocitygrid,     only : intvp, intmu 
+  use geom,             only : bn 
+
+  integer imm, imn, i, imod, ix, j, k, is 
+  complex :: dndt 
+
+  ! outer loop over the toroidal mode number. This mode number is what is 
+  ! kept in the ExB velocity and f 
+  do imm = 1, nmod; do imn = 1, nmod 
+
[…]

### 2022-05-03 gkw/src/diagnos_fluxes_vspace.F90

@@ -1,0 +1,700 @@
+!------------------------------------------------------------------------------
+!> Computes and outputs fluxes in velocity space or full 5d.  
+!> Should be consolidated with module fluxes, calculation presently duplicated.
+!------------------------------------------------------------------------------
+module diagnos_fluxes_vspace
+
+  implicit none
+
+  private
+
+#define OLD_FLUX_OUTPUT
+
+  public :: set_default_nml_values
+  public :: init, bcast, check, finalize, allocate_mem
+  public :: initial_output, final_output
+  public :: output
+
+  public :: calc_fluxes_full_detail
+  public :: pflux_det, eflux_det, vflux_det
+  public :: allocate_flux_det
+
+  !> switch for output of the particle, momentum and heat fluxes on
+  !> the 6D grid
+  logical, save, public :: lfluxes_detail
+
+  !> switches for output of the particle, momentum and heat fluxes on
+  !> velspace, due to any field
+  logical, save, public :: lfluxes_vspace
+  !> switches for output of the particle, momentum and heat fluxes on
+  !> velspace, due to any field
+  logical, save, public :: lfluxes_vspace_phi
+  logical, save, public :: lfluxes_vspace_em
+  logical, save, public :: lfluxes_vspace_bpar
+
+  logical, parameter :: lfluxes_detail_estep = .false.
+
+  !> the particle flux  pflux(nmod,nx,number_of_species,ns,nmu,nvpar)
+  !> due to the ExB motion
+  real, target, save, allocatable, dimension(:,:,:,:,:,:) :: pflux_det
+
+  !> the energy flux  eflux(nmod,nx,number_of_species,ns,nmu,nvpar)
+  !> due to the ExB motion
+  real, target, save, allocatable, dimension(:,:,:,:,:,:) :: eflux_det
+
+  !> the toroidal angular momentum flux
+  !> vflux(nmod,nx,number_of_species,ns,nmu,nvpar)
+  !> due to the ExB motion
+  real, target, save, allocatable, dimension(:,:,:,:,:,:) :: vflux_det
+
+  !> the fluxes  [p/e/v]flux(number_of_species,nmu,nvpar) due
+  !> to the ExB motion 
+  real, save, allocatable, dimension(:,:,:) :: pflux_vspace, eflux_vspace
+  real, save, allocatable, dimension(:,:,:) :: vflux_vspace
+
[…]

### 2022-05-03 gkw/src/diagnos_growth_freq.f90

@@ -1,0 +1,1110 @@
+!------------------------------------------------------------------------------
+!> Calculates growth rates, amplitudes, and frequencies for both linear
+!> and nonlinear runs, using the sum of the perturbed fields.  
+!> 
+!> Uses mode amplitudes calculated by the normalise module.
+!> Calculations are both per-mode and global.  In the per-mode case, 
+!> connected kx modes are identified using the mode_label array.
+!> The 1D outputs (as a function of krho) give only the kx = 0 mode.
+!>
+!> Performs growth rate convergence checks to trigger stop in linear runs.
+!------------------------------------------------------------------------------
+module diagnos_growth_freq
+  implicit none
+
+  private
+
+  public :: set_default_nml_values
+  public :: init, bcast, check, finalize, allocate_mem
+  public :: initial_output, final_output
+  public :: output, screen_output, read_last_data
+  public :: calc_largestep, calc_smallstep
+
+  public :: growth_rates, real_frequency
+
+  !> Switch, to calculate the mode frequencies.
+  !> output the frequencies of individual toroidal modes
+  logical, save, public :: lfrequencies
+  
+  !> Switch to calculate and output the growth rate of the dominant
+  !> mode (also called "global growth rate"), as well as the growth
+  !> rates of individual modes.
+  logical, save, public :: lgrowth_rates
+  
+  !> calculate and output the amplitudes of individual modes
+  logical, save, public :: lamplitudes
+
+  !> mode amplitudes
+  real, save, allocatable, dimension(:,:) :: amplitudes
+  !> mode amplitudes at previous time; used for growth rates
+  real, save, allocatable, dimension(:,:) :: last_amplitudes
+  logical, save, allocatable, dimension(:,:) :: amplitude_is_tiny
+  !> mode amplitudes temporary array
+  real, save, allocatable, dimension(:,:) :: amplitudes_G
+  !> growth rates, per toroidal mode
+  real, save, allocatable, dimension(:) :: growth_rates
+
+  integer, parameter :: history_size = 6
+  real, allocatable, save :: growth_rate_history(:,:)
+
+  real, allocatable, save :: std_gr(:), mean_gr(:)
+
+  !> the growth rate (or maximum growth rate) as reported in time.dat
+  real, save :: growth_rate
+
+  !> The real frequency of the mode.
+  real, allocatable, save :: real_frequency(:)
[…]

### 2022-05-03 gkw/src/diagnos_jdote.f90

@@ -1,0 +1,943 @@
+!------------------------------------------------------------------------------
+!>
+!> 
+!>
+!------------------------------------------------------------------------------
+module diagnos_jdote
+
+  implicit none
+
+  private
+  
+  public :: set_default_nml_values, init, bcast, check, allocate_mem
+  public :: calc_smallstep
+  public :: finalize
+  public :: output
+  
+  !> True if the flux surface averaged J.E diagnostic is switched on
+  logical, save, public :: lcalc_jdote_fs
+
+  !> True if the J.E diagnostic is switched on
+  logical, save, public :: lcalc_jdote
+
+  !> time between call of calc_smallstep() and output()
+  real, save :: delta_time_jdote = 1.
+  
+  !> last apar field, for real calculation of the
+  !> time derivative of apar.
+  complex, save, allocatable, dimension(:,:,:,:,:) :: last_aparga
+
+  !> the range of tags to be used by this diagnostic
+  integer, save :: tag_range_start, tag_range_end_inkl
+  integer, save :: tag_range_start2, tag_range_end_inkl2
+  integer, parameter :: n_velocity_slices = 6
+contains
+
+  !--------------------------------------------------------------------
+  !> Set reasonable default values for the namelist items this
+  !> diagnostic provides. 
+  !--------------------------------------------------------------------
+  subroutine set_default_nml_values()
+
+    lcalc_jdote_fs = .false.
+    lcalc_jdote = .false.
+
+  end subroutine set_default_nml_values
+
+  !--------------------------------------------------------------------
+  !> Broadcast all namelist items of this diagnostic to all processes.
+  !--------------------------------------------------------------------
+  subroutine bcast()
+    use mpiinterface, only : mpibcast
+
+    call mpibcast(lcalc_jdote,1)
+    call mpibcast(lcalc_jdote_fs,1)
+  end subroutine bcast
+
+  !--------------------------------------------------------------------
+  !> Check the diagnostic parameters and if the setup is compatible
+  !> with this diagnostic.
+  !--------------------------------------------------------------------
+  subroutine check()
+    use general, only : gkw_abort, gkw_warn
+    use mode, only : mode_box, lxn, lyn
+    use diagnos_velspace, only : npointsvel, psi_velspace, zeta_velspace
+    integer :: ipoint
[…]

### 2022-05-03 gkw/src/diagnos_kinenergy_trappas.f90

@@ -1,0 +1,1278 @@
+!------------------------------------------------------------------------------
+!> This diagnostic computes the various contributions to the time
+!> derivative of the kinetic energy (i.e the work of the particles in
+!> the electric field), similar to ORB5.
+!>
+!> It outputs the potential and magnetic energy, and the time variation of 
+!> kinetic energy for all the velocity contributions: gradB drifts,
+!> parallel velocity...
+!> The growth rate can be computed from these outputs. Thus,
+!> this diagnostic can reveal details about the nature of an
+!> instability.
+!>
+!> For a linear and electrostatic instability the fluctuations grow
+!> exponentially.
+!>   phi(k,t) = phi_0(k) exp[- i omega t + gamma t]
+!>   f(k,t) = f_0(k) exp[- i omega t + gamma t]
+!> where phi_0 and f_0 are complex amplitudes, so that phi and f may have a
+!> phase difference.
+!> In what follows, an asterisk means complex conjugation.
+!>
+!> The potential energy, being quadratic in fluctuations,
+!> growths accordingly:
+!>   Epot = \sum_k \int dv Z e f_0 exp[- i omega t + gamma t] phi_0* exp[+ i omega t + gamma t]
+!>        = \sum_k \int dv Z e f_0 phi_0* exp[2 gamma t]
+!> A typical linear run with nmod=1 does not have a zeromode, thus
+!>        = 2 real (\int dv Z e f_0 phi_0* exp[2 gamma t] )
+!>
+!> Note that Epot is real (see also the section 'Hermitian symmetry of binormal
+!> spectra' in the GKW manual).
+!> Its time derivative is then
+!> 
+!>   dEpot/dt = 2 gamma Epot
+!>
+!> We assume then that total energy is conserved (again, Ekin is a real valued
+!> quantity)
+!>   dEpot/dt = -dEkin/dt
+!>
+!> Total energy conservation yields that the growth rate is given by:
+!>   gamma = (dEpot/dt) / (2 Epot)  = (-dEkin/dt) / (2 Epot)
+!>
+!> The change in kinetic energy can be computed: it is the dotproduct
+!> of current and E-field.
+!>
+!> dEkin/dt = \sum_{species} \sum_k \int dv
+!>   Z_s e f (v_||\mathbf{b} + \mathbf{v}_D ) \cdot \mathbf{E}
+!>
[…]

### 2022-05-03 gkw/src/diagnos_matrix.f90

@@ -1,0 +1,247 @@
+!------------------------------------------------------------------------------
+!>
+!> DESCRIPTION
+!> 
+!> This diagnostic serves to output the matrix of the linear terms, as
+!> computed in GKW.
+!>
+!------------------------------------------------------------------------------
+module diagnos_matrix
+
+  implicit none
+
+  private
+  
+  public :: set_default_nml_values, bcast, check
+  public :: initial_output
+
+  !> The general on/off switch for this diagnostic
+  logical, save, public :: loutput_matrix
+
+contains
+
+  !--------------------------------------------------------------------
+  !> Set reasonable default values for the namelist items this
+  !> diagnostic provides. 
+  !--------------------------------------------------------------------
+  subroutine set_default_nml_values()
+    ! Usually the diagnostic should be switched
+    ! off by default.
+
+    loutput_matrix = .false.
+
+  end subroutine set_default_nml_values
+
+  !--------------------------------------------------------------------
+  !> Broadcast all namelist items of this diagnostic to all processes.
+  !--------------------------------------------------------------------
+  subroutine bcast()
+    use mpiinterface, only : mpibcast
+
+    call mpibcast(loutput_matrix,1)
+  end subroutine bcast
+
+  !--------------------------------------------------------------------
+  !> Check the diagnostic parameters and if the setup is compatible
+  !> with this diagnostic.
+  !--------------------------------------------------------------------
+  subroutine check()
+
+    if(.not. loutput_matrix) return
+
+  end subroutine check
+
+  !--------------------------------------------------------------------
+  !> This routine is called at the beginning of each run (after
+  !> restarts, too).
+  !--------------------------------------------------------------------
+  subroutine initial_output()
+    use matdat, only : mat, mat_maxwll_background, mat_poisson, mat_field_diag
+    use io, only : output_array, xy_fmt, binary_fmt, ascii_fmt, attach_metadata
+    use io, only : description_key, comments_key, phys_unit_key, not_avail
[…]

### 2022-05-03 gkw/src/diagnos_moments.f90

@@ -1,0 +1,1651 @@
+!------------------------------------------------------------------------------
+!> Output moments (density, flow, temperature perturbations) in 3D and in
+!> 2D perpendicular slices (spectral and real).
+!------------------------------------------------------------------------------
+module diagnos_moments
+  use diagnos_generic, only : DENSITY_MOMENT, TEMPERATURE_MOMENT
+  use diagnos_generic, only : PAR_TEMPERATURE_MOMENT, PERP_TEMPERATURE_MOMENT
+  use diagnos_generic, only : PAR_TEMPERATURE_GA_MOMENT, PERP_TEMPERATURE_GA_MOMENT
+  use diagnos_generic, only : PERP_TEMPERATURE_J1_MOMENT
+  use diagnos_generic, only : CURRENT_MOMENT, CURRENT_SQ_MOMENT, VPERP_SQ_MOMENT
+  use diagnos_generic, only : CURRENT_GA_MOMENT
+  use diagnos_generic, only : PASSING_DENSITY_MOMENT, TRAPPED_DENSITY_MOMENT
+  use diagnos_generic, only : DENSITY_GA_MOMENT, DENSITY_POLAR_MOMENT
+  use diagnos_generic, only : PHI_GA_FM_MOMENT, PHI_GA_DELTAF_MOMENT
+  use diagnos_generic, only : QPAR_MOMENT, M12_MOMENT, M24_MOMENT
+  use diagnos_generic, only : QPAR_GA_MOMENT, M12_GA_MOMENT, M24_GA_MOMENT
+
+  implicit none
+
+  private
+
+  public :: set_default_nml_values
+  public :: init, bcast, check, finalize, allocate_mem
+  public :: initial_output, final_output
+  public :: output
+
+  public :: gather_moment_4d_serial
+  public :: get_4d_moment
+
+  !> the range of tags to be used by this diagnostic
+  integer, save :: tag_range_start, tag_range_end_inkl
+
+  !> For output of moments spectra
+  real, save, allocatable, dimension(:,:) :: moment_spec
+
+  !> logical unit numbers
+  integer, save :: i_denspec = -1, i_enespec = -1
+
+  !> logical unit numbers for 2d output of various moments
+  !> Note: please remove the second dimension if io_legacy is ever removed
+  integer, save, allocatable, dimension(:,:) :: lun_moment_spec
+  integer, save, allocatable, dimension(:,:) :: lun_moment_xy
[…]

### 2022-05-03 gkw/src/diagnos_rad.f90

@@ -1,0 +1,1573 @@
+!------------------------------------------------------------------------------
+!> Outputs radial profiles for non-spectral runs.
+!> A mixture of fields, moments, energetics, and background profiles.
+!> Could be logically re-organised into other relevant modules.
+!------------------------------------------------------------------------------
+module diagnos_rad
+
+  implicit none
+
+  public :: set_default_nml_values
+  public :: init, bcast, check, finalize, allocate_mem
+  public :: initial_output, output
+
+  private
+
+  !> the luns for the radial profile files
+  integer, save, allocatable :: i_delta_n(:), i_delta_v(:), i_delta_t(:), i_delta_e(:)
+  integer, save :: i_fields
+  !> luns for turbulence and zonal mode intensity
+  integer, save :: i_f_int, i_zf_int
+  !> luns for turbulence and zonal mode entropy
+  integer, save :: i_f_entropy, i_zf_entropy
+  !> luns for  turbulence and zonal mode entropy fields
+  integer, save :: i_f_entropy_fields, i_zf_entropy_fields
+  !> luns for kparallel diagnostic and kparallel^2 and theta_0
+  integer, save :: i_kpar, i_kpar2, i_theta0, i_kr, i_kr2
+
+  !> zonal mode entropy flux ExB particles and fields real part
+  integer, save :: i_zf_entropy_flux_re, i_zf_entropy_flux_fields_re
+  !> zonal mode entropy flux ExB particles and fields imag part
+  integer, save :: i_zf_entropy_flux_im, i_zf_entropy_flux_fields_im
+  !> perturbations entropy flux ExB 
+  integer, save :: i_f_entropy_flux
+  !> zonal mode entropy flux drift particles and fields 
+  integer, save :: i_zf_entropy_flux_drift, i_zf_entropy_flux_fields_drift
+  !> perturbations entropy flux drift particles and fields 
+  integer, save :: i_f_entropy_flux_drift, i_f_entropy_flux_fields_drift
+
+  !> Help arrays
+  real, save, allocatable  :: dum_G(:)            !< array global in x
+  real, save, allocatable  :: dum(:),dum2(:),dum_buf(:)  !< array local in x
+
+  complex, save, allocatable :: cdum_G(:)         !< array global in x
+  complex, save, allocatable :: cdum(:)           !< array local in x
+  complex, save, allocatable :: cdum2(:)           !< array local in x
+  complex, save, allocatable :: cdum3(:)           !< array local in x
[…]

### 2022-05-03 gkw/src/diagnos_template.f90

@@ -1,0 +1,382 @@
+!------------------------------------------------------------------------------
+!>  This is a template for a GKW diagnostic module.
+!>
+!> Before writing a new diagnostic or module, you should in
+!> general try to find the most similar existing diagnostic in terms
+!> of quantity calculated.  If at all possible,
+!> you should generalize an existing diagnostic for your needs,
+!> rather than copy/pasting and adapting it.
+!>
+!> To write a new diagnostic, first read some of the other
+!> diagnostics, to learn how things are done.
+!> Then fill the subroutines of this template with life and hook them
+!> into the appropriate subroutines of diagnostic.F90 . You may delete
+!> unneeded routines of this skeleton and not hook them into the
+!> diagnostic.F90 . All routines are 'collective' in the MPI sense.
+!> If a block has to be executed by only a single processor, you have
+!> to write something like
+!>    if(root_processor) then ... end if
+!> or
+!>    if(proc_subset(0,1,1,1,gsp(is))) then ... end if
+!> Be aware of the parallelization in particular when you want to write
+!> output data.
+!>
+!> Please try to encapsulate new diagnostics as much as possible, so
+!> that they can be easily understood by others.
+!> This means that you should not use public variables/routines of the
+!> other diagnos_* modules, except from diagnos_generic.
+!> Routines and switches which are used by more than one diagnos_* module
+!> (e.g. generally useful output routines) should be placed
+!> into diagnos_generic.
+!>
+!> It is really important put comments, *immediately*, at the time of
+!> writing the code, to save other readers (and yourself) precious time
+!> when they need to understand your code. Others should not have to waste
+!> time thinking about ambiguities or non-trivial algorithms.
+!>
+!> The lifecycle of a module diagnos_foo during a run of the explicit
+!> solver is usually
+!>   1.  The namelist items provided by diagnos_foo are initialised
+!>       with default values.
+!>   2.  The diagnostic namelist is parsed and the default values of namelist
+!>       items are overwritten.
+!>   3.  bcast() is called.
+!>   4.  check() is called.
+!>   5.  allocate_mem() is called.
+!>   6.  read_last_data() is called.
+!>   7.  init() is called.
+!>   8.  initial_output() is called.
+!>   9.  output() is called repeatedly.
[…]

### 2022-05-03 gkw/src/diagnos_timetrace.F90

@@ -1,0 +1,290 @@
+!------------------------------------------------------------------------------
+!> 
+!------------------------------------------------------------------------------
+module diagnos_timetrace
+
+  implicit none
+
+  private
+
+  !>==== DESCRIPTION ==================================================
+  !>
+  !> This diagnostic outputs a time trace, which can be subject to a
+  !> signal analysis technique at postprocessing such as
+  !> time-frequency analysis, higher-order spectral analysis,
+  !> wavelet analysis or whatever makes sense.
+  !>
+  !>---- FURTHER NOTES ------------------------------------------------
+  !>
+  !> see also figure 5a in
+  !> http://iopscience.iop.org/0029-5515/50/5/054005/
+  !> or
+  !> D. Told et al., PoP 20, 122312 (2013)
+  !>   http://dx.doi.org/10.1063/1.4858899
+  !>
+  !>---- LIMITATIONS --------------------------------------------------
+  !>
+  !> This diagnostic does not care about at-runtime refinement of the
+  !> timestep by any of the timestep estimators, and the analysis
+  !> scripts probably have a hard time handling such cases (see
+  !> comments there).
+  !>
+  !>===================================================================
+
+  public :: set_default_nml_values, init, bcast, check, allocate_mem
+  public :: finalize
+  public :: initial_output, read_last_data
+  public :: output
+
+  public :: calc_smallstep
+
+  !> The general on/off switch for this diagnostic
+  logical, save, public :: lcomplex_timetrace
+
+  !> array for quick look up of the fields
+  integer, save, dimension(:), allocatable :: lindx
+
+  !> buffers
+  complex, save, dimension(:,:), allocatable :: cbuf
+
+  integer :: lun_csignal
+
+contains
+
+  !--------------------------------------------------------------------
+  !> Set reasonable default values for the namelist items this
+  !> diagnostic provides. 
+  !--------------------------------------------------------------------
+  subroutine set_default_nml_values()
+
+    lcomplex_timetrace = .false.
+
+  end subroutine set_default_nml_values
+
+  !--------------------------------------------------------------------
+  !> Broadcast all namelist items of this diagnostic to all processes.
+  !--------------------------------------------------------------------
+  subroutine bcast()
+    use mpiinterface, only : mpibcast
+
+    call mpibcast(lcomplex_timetrace,1)
+
[…]

### 2022-05-03 gkw/src/diagnos_zfshear.f90

@@ -1,0 +1,609 @@
+!------------------------------------------------------------------------------
+!> This diagnostic calculates the shear rate of the zonal flows.
+!!    \f$ \omega = \frac{\partial}{\partial \psi} v_{E,\zeta}
+!!          = \frac{\partial^2}{\partial \psi^2} \ga{\phi} \f$
+!!
+!! It is used for example in
+!!    * Dannert and Jenko, PoP 12, 072309 (2005)
+!!    * Merz and Jenko, Nucl. Fusion 50, 054005 (2010)
+!!
+!! Other interesting works in this context include
+!!    * Hahm et al., PoP 6, No. 3, 922 (1999)
+!!
+!<------------------------------------------------------------------------------
+module diagnos_zfshear
+
+  implicit none
+
+  private
+
+  public :: set_default_nml_values, init, bcast, check, allocate_mem
+  public :: finalize
+  public :: initial_output, final_output
+  public :: output
+
+  !> The general on/off switch for this diagnostic
+  logical, save, public :: lcalc_zfshear
+
+  !> The logical unit numbers, used to output data
+  integer, save, allocatable :: i_zfshear_ga(:), i_zfflow_ga(:), i_zfphi_ga(:)
+  integer :: i_zfshear, i_zfflow, i_zfphi
+
+  !> Help array for first and second radial derivative of the electrostatic
+  !> potential; the corresponding functions (dfielddx, d2fielddx2) only
+  !> accept complex arrays, although the potential is a real quantity.
+  !> Therefore, in the calculation of the individual quantities below, dum
+  !> is casted to a real value always.
+  complex, allocatable, dimension(:) :: dum
+  
+  !> Help arrays for the fluxsurface average as well as velocity space integral
+  real, allocatable, dimension(:) :: local_x_line_ga, local_x_line
+  
+  !> Global arrays in radial direction containing the gyro-averaged quantities
+  real, allocatable, dimension(:) :: zfshear_ga_G, zfflow_ga_G, zfphi_ga_G
+  
+  !> Global arrays in radial direction containing the non-gyro-averaged quantities
+  real, allocatable, dimension(:) :: zfshear_G, zfflow_G, zfphi_G
+  
+  logical, parameter :: test_x_derivatives = .false.
+  
+  !> mpi datatypes 
+  integer, save :: mpi_dtype_radial_line
+
+contains
+
[…]

### 2022-05-03 gkw/src/diagnos_zonal_evo.f90

@@ -1,0 +1,1972 @@
+!>------------------------------------------------------------------------------
+!> ZONAL EVO MODULE DESCRIPTION
+!> 
+!>----DESCRIPTION---------------------------------------------------------------
+!> 
+!>   This diagnostic computes the time derivative of the zonal (flux surface
+!>   averaged) electrostatic potential and the individual terms of the zonal
+!>   evolution equation. 
+!>
+!>----OUTPUT--------------------------------------------------------------------
+!>
+!>   This diagnostic outputs several quantities connected to the evolution of
+!>   the zonal (flux surface averaged) electrostatic potential.
+!>
+!>   The following output files have dimensions (time, kx) and
+!>   contain the complex Fourier amplitudes of the respective quantity.
+!> 
+!>   If lcalc_zonal_evo = .true.
+!>   - the zonal potential: zphi_kx
+!>   - the time derivative of the zonal potential: dt_zphi_kx
+!>   - all linear terms added: zevo_lin_kx
+!>   - the nonlinear terms: zevo03_[phi,apar,bpar]_kx
+!>
+!>   If zevo_detail = .true.
+!>   - the numerical damping terms: zevo_disp_[par,vp,perp]_kx
+!>   - the contributions from linear terms: 
+!>     zevo[01,02,04,05,07,08,10,11]_kx
+!>   - the contribution from collisions: zevo_coll_kx
+!>
+!>   The following output files have dimensions (time, kx, ky) and
+!>   contain the complex Fourier amplitudes of the respective quantity.
+!> 
+!>   If zevo_xy_spec = .true.
+!>   - the nonlinear terms: zevo03_[phi,apar,bpar]_kxky
+!>    
+!>----FURTHER NOTES------------------------------------------------------------- 
+!>
+!>   Per default only the sum of all linear terms is calculated.
+!>   More details about individual linear terms is obtaind by setting
+!>   
+!>       zevo_detail = .true.
+!>
+!>   kxky-spectra of nonlinear terms are output for
+!> 
+!>       zevo_xy_spec = .true.
+!>
+!-------------------------------------------------------------------------------
+module diagnos_zonal_evo
+  use control, only : lcalc_zonal_evo, zevo_detail
+  
+  implicit none
+
[…]

### 2022-05-03 gkw/src/fft.F90

@@ -1,0 +1,2397 @@
+!----------------------------------------------------------------------
+!> Contains generalised FFT interfaces for fftw3 and fftw3
+!> Also contains a lot of stuff that is not currently used by GKW
+!----------------------------------------------------------------------
+module fft
+#if defined(FFT_MKL)
+  ! UPPERCASE USE to deliberately exclude from mkdeps script
+  USE mkl_dfti
+#endif
+  
+  use global, only : i8
+  implicit none
+
+  private
+
+  public :: fourcol, four1D_real, four2D_real
+
+  integer, parameter, public :: FFT_FORWARD = -1
+  integer, parameter, public :: FFT_INVERSE = 1
+
+  ! define the maximum number of plans/work arrays/descriptor handles.
+  integer, parameter :: MXPLAN = 64
+  integer, parameter :: N_FFT_ROUTINES = 9
+
+  interface fourcol
+    module procedure fourcol_ca
+    module procedure fourcol_caa
+    module procedure fourcol_ra_ca
+    module procedure fourcol_raa_caa
+  end interface fourcol
+
+#if defined(FFT_FFTW3)
+!=============================================================================
+! FFTW v3
+!=============================================================================
+
+  !> if this interface compiles, FFTW should work
+  logical, parameter, public :: WORKING_FFT_LIBRARY=.true.
+  character (len=5), parameter, public :: fftlib = 'FFTW3'
+
+  ! UPPERCASE INCLUDE to deliberately exclude from mkdeps script
+  INCLUDE 'fftw3.f'
+
+  !> plans for FFT transforms
+  integer (kind=i8), dimension(MXPLAN,N_FFT_ROUTINES), save :: plans
+  integer, dimension(MXPLAN,N_FFT_ROUTINES,4), save :: plan_properties = 1
+  !> number of plans saved
+  integer, dimension(N_FFT_ROUTINES), save :: nplans=0
+
+contains
+
+  !--------------------------------------------------------------------
+  !>
+  !--------------------------------------------------------------------
+  pure function get_plan_id(arr_shape, isign, num) result(id)
[…]

### 2022-05-03 gkw/src/fields.F90

@@ -1,0 +1,879 @@
+!------------------------------------------------------------------------------
+!> Module containing routines for returning gyroaveraged fields
+!> Will contain more soon, if not could be merged with gyro_average
+!------------------------------------------------------------------------------
+module fields 
+  
+  use global, only : rumf 
+  
+  implicit none 
+
+  public :: get_averaged_phi, get_averaged_apar, get_averaged_bpar 
+  public :: fields_allocate, fields_deallocate, calculate_fields
+  public :: field_solve_nonspec_wrap, calculate_cons, gyro_average_fields
+
+  private 
+
+  !> help array for the mpi
+  complex, allocatable, save :: bufphi(:)
+
+  !> help array for the nonspectral polarisation of the potential 
+  real(rumf), allocatable, save :: z_buf(:),  z(:)
+  complex(rumf), allocatable :: phi_dum(:)
+
+  !>  help array to store the additional potential which is of the order rhostar
+  real(rumf), allocatable :: phi1(:)
+
+contains 
+
+!------------------------------------------------------------------------------
+!> This function gives the value of the gyro-averaged potential for a given 
+!> grid point. If nlphi = .false. a zero is returned 
+!------------------------------------------------------------------------------
+function get_averaged_phi(imod,ix,i,j,is,fdis)
+  use control,        only : nlphi, spectral_radius 
+  use index_function, only : indx 
+  use functions,      only : besselj0_gkw
+  use dist,           only : iphi, iphi_ga 
+
+  integer, intent(in)  :: imod, ix, i, j, is
+  complex, intent(in)  :: fdis(:)
+  complex :: get_averaged_phi 
+
+  if (nlphi) then 
+    if (spectral_radius) then 
+      get_averaged_phi = besselj0_gkw(imod,ix,i,j,is) * &
+                       & fdis(indx(iphi,imod,ix,i)) 
+    else 
+      get_averaged_phi = fdis(indx(iphi_ga,imod,ix,i,j,is)) 
+    endif 
+  else 
+    get_averaged_phi = (0.E0,0.E0) 
+  endif 
+ 
+end function get_averaged_phi 
+
+!------------------------------------------------------------------------------
[…]

### 2022-05-03 gkw/src/functions.f90

@@ -1,0 +1,248 @@
+!-----------------------------------------------------------------------------
+!> Contains mathematical functions required by GKW.
+!> Bessel and Gamma functions
+!-----------------------------------------------------------------------------
+module functions
+
+  private
+
+  public :: gamma_gkw, gamma1_gkw, besselj0_gkw, mod_besselj1_gkw, mod_sinh_gkw
+  public :: legendre, leguerre, cheby
+
+contains
+
+!++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
+!++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
+!> This function calculates the Gamma function needed in Poissons and
+!> Ampere's equation
+!----------------------------------------------------------------------------
+
+function gamma_gkw(imod,ix,i,is)
+
+  use specfun,    only : expbessi0
+  use components, only : mas, vthrat, signz  
+  use geom,       only : bn
+  use mode,       only : krloc
+
+  integer, intent(in) :: imod, ix, i, is
+
+  real :: gamma_gkw, dum1
+
+  dum1 = 0.5*(mas(is)*vthrat(is)*krloc(imod,ix,i)/(signz(is)*bn(ix,i)))**2 
+  gamma_gkw = expbessi0(dum1)
+
+end function gamma_gkw
+
+!++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
+!++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
+!> This function calculates the first order Gamma function needed in Poissons
+!> and Ampere's equation
+!----------------------------------------------------------------------------
+
+function gamma1_gkw(imod,ix,i,is)
+
+  use specfun,    only : expbessi1
+  use components, only : mas, vthrat, signz  
+  use geom,       only : bn
+  use mode,       only : krloc
+
+  integer, intent(in) :: imod, ix, i, is
+
+  real :: gamma1_gkw, dum1
+
+  dum1 = 0.5*(mas(is)*vthrat(is)*krloc(imod,ix,i)/(signz(is)*bn(ix,i)))**2
+  gamma1_gkw = expbessi1(dum1)
+
+end function gamma1_gkw
+
[…]

### 2022-05-03 gkw/src/general.f90

@@ -1,0 +1,239 @@
+!*****************************************************************************
+!> General purpose routines, having no dependencies other than things
+!> in mpiinterface, should be in here.
+!*****************************************************************************
+module general
+
+  use global, only : int2char
+
+  implicit none
+
+  private
+
+  public :: general_init
+  public :: gkw_abort, gkw_warn, gkw_exit
+  public :: matout
+  public :: gkw_is_nan
+  public :: max_factor
+
+  contains
+
+!-----------------------------------------------------------------------------
+!> initialize anything for general
+!-----------------------------------------------------------------------------
+subroutine general_init()
+
+end subroutine general_init
+
+!-----------------------------------------------------------------------------
+!> Abort the code safely (mpiabort) with an optional error message.
+!> This routine can be used to abort from a module higher in the hierarchy
+!> than IO.
+!-----------------------------------------------------------------------------
+subroutine gkw_abort(abort_message)
+  use mpiinterface, only : processor_number
+  use io, only : finalize_and_abort
+  
+  character (len=*),intent(in), optional :: abort_message
+
+  if (present(abort_message)) then
+    write (*,'(A,A,A,I4)') ' GKW_ABORT -- ', abort_message,                  &
+                         & '  ', processor_number
+  else
+    write (*,'(A,A,A,I4)') ' GKW_ABORT -- ', '(no message given!)',          &
+                         & '  ', processor_number
+  end if
+
+  call finalize_and_abort
+
+end subroutine gkw_abort
+
+!-----------------------------------------------------------------------------
+!> Clean abort that must only be called from ALL procs. 
+!> This is only to be used when we are sure that all processors
+!> will call the routine (check params routines) 
+!> Otherwise, gkw_abort should be used as a safe option to avoid MPI deadlock
+!-----------------------------------------------------------------------------
+subroutine gkw_exit(abort_message)
+  use mpiinterface, only : root_processor
+  use io, only : finalize_and_exit
+  
+  character(len=*), intent(in), optional :: abort_message
+
+  if (root_processor) then
+    if (present(abort_message)) then
[…]

### 2022-05-03 gkw/src/global.F90

@@ -1,0 +1,576 @@
+!+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
+!> Various unchanging parameters needed in most parts of the code.
+!> Furthermore some very small functions and subroutines which do not depend
+!> on any other module and are also needed in many parts of the code.
+!+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
+
+module global
+
+  implicit none
+
+  private
+
+  !
+  ! Parameters for capabilities.
+  !
+#ifdef HAVE_SLEPC
+  logical, parameter, public :: compiled_with_slepc = .true.
+#else
+  logical, parameter, public :: compiled_with_slepc = .false.
+#endif
+
+#ifdef HAVE_HDF5
+  logical, parameter, public :: compiled_with_hdf5 = .true.
+#else
+  logical, parameter, public :: compiled_with_hdf5 = .false.
+#endif
+
+#ifdef HAVE_LIBRSB
+  logical, parameter, public :: compiled_with_librsb = .true.
+#else
+  logical, parameter, public :: compiled_with_librsb = .false.
+#endif
+
+#ifdef HAVE_MKL
+  logical, parameter, public :: compiled_with_mkl = .true.
+#else
+  logical, parameter, public :: compiled_with_mkl = .false.
+#endif
+
+#if !defined(mpi2)
+  !> parameter for setting some dummy integers when the code runs without MPI
+  integer, parameter, public :: I_RUN_WITHOUT_MPI = -64523
+  logical, parameter, public :: compiled_with_mpi = .false.
+#else
+  logical, parameter, public :: compiled_with_mpi = .true.
+#endif
+
+#ifdef _OPENMP
+  !> reporting label
+  logical, parameter, public :: compiled_with_openmp = .true.
+  integer, parameter, public :: openmp_date = _OPENMP
+#else
+  !> reporting label
+  logical, parameter, public :: compiled_with_openmp = .false.
+  integer, parameter, public :: openmp_date = 0
+#endif
+
+ !> reporting label
+#ifdef umfpack
+#ifdef kernel32bit
+  integer, parameter, public :: compiled_with_umfpack = 32
+#else
+  integer, parameter, public :: compiled_with_umfpack = 64
+#endif
+#else
+   integer, parameter, public :: compiled_with_umfpack = 0
+#endif
+
+
+  !
+  ! Precision
+  !
[…]

### 2022-05-03 gkw/src/krook.f90

@@ -1,0 +1,505 @@
+!-----------------------------------------------------------------------------
+!> Module containing the source connected with a global run 
+!-----------------------------------------------------------------------------
+module krook 
+
+  implicit none 
+
+  private 
+
+  public :: krook_read_nml, krook_bcast_nml 
+  public :: krook_write_nml, add_krook, krook_init 
+  public :: nlkrook, nlbound, bwidth, gammab, krook_option, fdis_nvpar 
+
+  !> Logical that determines if the Krook operator is used 
+  logical, save :: nlkrook 
+
+  !> Damping rate used in the Krook operator 
+  real, save :: gammak 
+  
+  !> For krook option 4, prefactor of n=0 mode damping rate.  I.e
+  !> can selectively chose to damp the zonal mode stronger than the 
+  !> turbulence
+  real, save :: gamkpre
+
+  !> Logical that determines in damping is applied in the boundary layers 
+  logical, save :: nlbound 
+
+  !> width of the boundary layers (in grid points) 
+  real, save :: bwidth 
+
+  !> damping rate of the boundary layers 
+  real, save :: gammab 
+
+  !> option for the various forms of the krook operator 
+  integer, save :: krook_option 
+
+  interface krook_write_nml
+    module procedure krook_read_nml
+  end interface
+
+  ! arrays for the krook operator 
+  integer, save              :: nmats 
+  integer, save, allocatable :: iss(:), jss(:), id1(:), id2(:)
+  real,    save, allocatable :: mats1(:), mats2(:), mats3(:), mats4(:)
+  real,    save, allocatable :: pref(:), prefgam(:)
+  complex, save, allocatable :: sss(:), sst(:), fdis_nvpar(:) !, sss_tmp(:)
+
+contains 
+
+!------------------------------------------------------------------------------
+!> Subroutine that reads the namelist 
+!------------------------------------------------------------------------------
+subroutine krook_read_nml(lun,io_stat,lwrite)
+  use io, only : write_run_parameter
+  use mpiinterface, only : root_processor
+
+  integer, intent(in)  :: lun
+  integer, intent(out) :: io_stat
+
+  logical, optional, intent(in) :: lwrite
+
[…]

### 2022-05-03 gkw/src/linear_terms.f90

@@ -1,0 +1,4752 @@
+!-----------------------------------------------------------------------------
+!> This module calculates all the linear terms and puts them in the matrix
+!> All terms I-XI except III and IX.
+!> only used during the initialisation of the code.
+!> The matrix is stored in matdat.  The matrix vector multiply is
+!> performed during the time integration. The vector is stored in dist.
+!-----------------------------------------------------------------------------
+module linear_terms
+
+  implicit none
+
+  private 
+
+  public :: init_linear_terms, calc_linear_terms, linear_terms_read_nml 
+  public :: linear_terms_bcast_nml, linear_terms_check_params 
+  public :: linear_terms_write_nml, differential_scheme, drift
+  
+  public :: lpoisson, lvpgrphi, lampere, lneorotsource
+  public :: lvdgradf, lvpar_grad_df, lneo_equil_switch_default
+
+  public :: dmaxwel
+  
+  !> switches for hard disabling of calls
+  logical, parameter :: linear_term_switch_default = .true.
+  logical, parameter :: lneo_equil_switch_default = .false.
+  integer, dimension(128), save :: apply_on_imod = 0
+  logical, save :: lvpar_grad_df          = linear_term_switch_default
+  logical, save :: lvdgradf               = linear_term_switch_default
+  logical, save :: ltrapdf                = linear_term_switch_default
+  logical, save :: lve_grad_fm            = linear_term_switch_default
+  logical, save :: lvd_grad_phi_fm        = linear_term_switch_default
+  logical, save :: lvpgrphi               = linear_term_switch_default
+  logical, save :: lpoisson               = linear_term_switch_default
+  logical, save :: lg2f_correction        = linear_term_switch_default
+  logical, save :: lpoisson_zf            = linear_term_switch_default
+  logical, save :: lampere                = linear_term_switch_default
+  logical, save :: lneoclassical          = linear_term_switch_default
+  logical, save :: lbpar                  = linear_term_switch_default
+  logical, save :: lneorotsource          = lneo_equil_switch_default
+  logical, save, public :: lneo_equil     = lneo_equil_switch_default
+  logical, save :: lneo_trap              = linear_term_switch_default
[…]

### 2022-05-03 gkw/src/matconv.F90

@@ -1,0 +1,192 @@
+!-----------------------------------------------------------------------
+!> module that contains routines for converting matrix formats
+!-----------------------------------------------------------------------
+module matconv
+
+  use global, only : iumf, rumf
+  use general, only : gkw_abort
+
+implicit none
+
+#if defined(umfpack)
+
+public :: crstoreal, crstoreal_umf
+
+private
+
+contains
+!-----------------------------------------------------------------
+!>     this routine converts the complex row storage format to
+!>     a real row storage format
+!>
+!>     np     number of columns ("rows") in the complex matrix
+!>     nm     maximum dimension of mat and jam
+!>     iam    integer array that points to the beginning of 
+!>            the next row in the mat array
+!>     jam    index array that points to the column entry
+!>     mat    the matrix elements
+!>     nmr    maximum dimension of a and ja
+!>  
+!>     output 
+!>     ia     same as iam but for the real matrix (twice the size)
+!>     ja     same as jam but for the real matrix (4 times in size)
+!>     a      same as mat but for the real matrix (4 times in size)
+!-------------------------------------------------------------------
+!> This routine could be further optimised to deal with real 
+!> entries in the matrix (which give zeros on some of the 
+!> diagonals
+subroutine crstoreal(np,nm,iam,jam,mat,nmr,ia,ja,a)
+  use global, only : gkw_a_equal_b_accuracy
+integer          :: np, nm, iam(np+1), jam(nm) 
+integer(kind=iumf)        :: nmr, itel 
+integer(kind=iumf)        :: ia(2*np+1), ja(nmr)
+complex          :: mat(nm) 
+real(kind=rumf)  :: a(nmr)
+
+integer(kind=iumf) i, j, istart
+
+! start the conversion 
+do i = 1, np 
+  ia(2*i) = 2*(iam(i+1)-iam(i))
+  ia(2*i+1) = 2*(iam(i+1)-iam(i))
+end do 
+ia(1) = 1
+do i = 2, 2*np+1
+  ia(i) = ia(i) + ia(i-1)
+end do 
+itel = 1
+do i = 1, np
+  do j = iam(i), iam(i+1)-1
+    ja(itel) = 2*jam(j)-1
[…]

### 2022-05-03 gkw/src/matdat.F90

@@ -1,0 +1,1926 @@
+!-------------------------------------------------------------------------------
+!> The module matdat contains all the quantities associated with the matrix
+!! that contains the linear terms, and sorting and compression routines.
+!! 
+!! The solution vector fdisi in GKW contains both the distribution function g
+!! as well as all the fields. A single matrix is built for the implicit scheme, 
+!! although only the distribution function is integrated forward in time. 
+!! One can think of the matrix consisting of 4 sections
+!! [ g^n+1 ]  = [ M  N ] [ g^n ]
+!! [ 0     ]    [ P  Q ] [ phi,apar,bpar,mom ]  
+!!
+!! M contains the parallel derivatives, the drifts and the trapping terms 
+!! N contains the source due to the Maxwell background 
+!! (which is always proportional to one of the fields)
+!! P is the integral part of the field equations, i.e. the integral over g 
+!! Q  is the polarization (diagonal field) terms. 
+!! 
+!! The whole matrix [M,N,P,Q] is stored in mat, with the indices n1,n2,n3,n4 
+!! delineating the different areas: M(1,n1), N(n1+1,n2), P(n2+1,n3), Q(n3+1,n4). 
+!!
+!! The explicit scheme does not allow mixing of any of the areas. 
+!! In the non-parallel version of the code the fields are always at the end 
+!! of the solution vector, although in the parallel version of the code
+!! all the ghost cell buffers appear after the end of the main solution vector.
+!<-------------------------------------------------------------------------------
+module matdat
+  use matrix_format, only : sparse_matrix
+  use structures, only : matrix_element
+
+  implicit none
+  
+  private
+
+  public :: finish_matrix_section, abort_if_bad_element
+  public :: put_source, add_source
+  public :: get_f_from_g, connect_parallel 
+  public :: pos_par_grid, connect_rad, compress_piece, get_estimated_timestep
+
+  public :: iac, matdat_allocate, maty
+  public :: mat, mat_maxwll_background, mat_poisson, mat_field_diag
+  public :: matz, matg2f, source
+  public :: add_element, set_indx
[…]

### 2022-05-03 gkw/src/matrix_format.F90

@@ -1,0 +1,1514 @@
+!===========================================================================
+!> This module provides the abstracted interface for the matrix format.
+!> It is an interface wrapper around swappable modules matrix_format_* which
+!> perform the actual sparse matrix computations for different formats.
+!>
+!> The recognized output options (selected by input variable io_format) are
+!> gkw-crs     : self implemented compressed row storage (called CRS,
+!>               CSR or Yale format)
+!> librsb      : the publicly available library written by M.Martone
+!>
+!===========================================================================
+module matrix_format
+
+#define FIRST_TOUCH
+
+#define CRS_FASTDIAGONAL_FORMAT
+
+#ifdef HAVE_MKL
+! UPPERCASE USE to deliberately exclude from mkdeps script
+  ! USE f95_precision, only: wp => dp
+
+  !lowercase 'use', to include mkl_sblas module as an additional
+  ! dependency.  SRG: it seems to me that installing the MKL does not
+  ! compile a .o and .mod files for sparse BLAS. One has to compile
+  ! them on it's own.
+  USE mkl_spblas
+
+  ! USE ISO_C_BINDING
+  ! INCLUDE 'mkl_spblas.fi'
+  ! INCLUDE 'mkl_sparse_handle.fi'
+#endif
+
+  implicit none
+
+  public
+
+  !> coordinate format. This is the simplest and most flexible format.
+  integer, parameter :: FORMAT_GKWCOO = 0
+
+  character(len=10) :: matrix_format_gkwcrs = 'gkw-crs'
+  integer, parameter :: FORMAT_GKWCRS = 1
+  character(len=10) :: matrix_format_librsb = 'librsb'
+  integer, parameter :: FORMAT_LIBRSB = 2
+  character(len=10) :: matrix_format_mklcrs = 'mkl-crs'
+  integer, parameter :: FORMAT_MKLCRS = 3
+  
+  type :: sparse_matrix
+    ! Initially, every sparse matrix is constructed using the arrays below
+    ! and a simple coordinate format. As soon as the matrix is filled,
+    ! it can be transformed to a (potentially) more efficient sparse
+    ! matrix format.
+
+    ! FIXME find out if it does have any significance to put the
+    ! elements of this struct in a certain order (e.g. scalars first)
+    ! (maybe to help for good alignment or so).
+
+    !> the total number of nonzero elements
+    integer :: nmat = 0
+    !> counter for the number of compressions
+    integer :: compressions = 0
[…]

### 2022-05-03 gkw/src/mode.f90

@@ -1,0 +1,1252 @@
+!------------------------------------------------------------------------------
+!> Module contains most of the information on the modes kept in the
+!> simulation. Sets up 2D mode grid and connections.
+!------------------------------------------------------------------------------
+module mode
+
+  use global, only : lenswitch
+  
+  implicit none
+
+  private
+
+  public :: mode_read_nml, mode_write_nml, mode_bcast_nml, mode_init
+  public :: mode_check_params, krbal, kgrid, mode_box_recon, mode_allocate
+  public :: kgrid_nonspec_rad, parallel_phase_shift 
+
+  !> The array of 'toroidal' (zeta) wave vectors, krho(nmod).
+  !> These wavevectors are perpendicular to the field line in the plane of
+  !> the flux surface. Confusion can arise because they are called both
+  !> 'toroidal' and 'poloidal' at various points! This direction is also often
+  !> referred to as the y direction.
+  real, allocatable, public, save :: krho(:)
+
+  !> The perpendicular wave vector (as a function along the field line)
+  real, allocatable, public, save :: krloc(:,:,:)
+
+  !> The array of radial (psi) wave vectors, kxrh(nx).
+  real, allocatable, public, save :: kxrh(:)
+
+  !> Integers that determine to which kx mode the mode is connected to
+  !> through the parallel boundary conditions. ixplus(nmod,nx) is for the
+  !> boundary connection in the positive s-direction?
+  integer, allocatable, public, save :: ixplus(:,:)
+  integer, allocatable, public, save :: ixminus(:,:) !< Opposite boundary to ixplus.
+
+  !> Logical that determines if there is a 2D grid of ky,kx. Used in
+  !> conjunction with nperiod = 1 (necessary for non linear runs).
+  logical, public, save :: mode_box
+
+  !> For mode_box, the integer that determines the spacing between the
+  !> different kx modes.
+  integer, public, save :: ikxspace
+  
+  !> LX/LY for zero shear case, spectral only
+  real, save :: rkxspace
+
+  !> The location of the kx=0 mode.
+  integer, public, save :: ixzero
+
+  !> The location of the ky=0 mode.  Needed by linear terms.
+  integer, public, save :: iyzero
+
+  !> The spacing of the toroidal mode numbers. Only used in global simulations 
[…]

### 2022-05-03 gkw/src/mpidatatypes.F90

@@ -1,0 +1,257 @@
+!-----------------------------------------------------------------------------
+!> provide derived datatypes for communication of processor boundaries and
+!> parallel io etc.
+!> 
+!> Derived datatypes for ghost cell communciation are initialised in dist.
+!> The communications using these datatypes are managed by module mpighosts.
+!-----------------------------------------------------------------------------
+
+module mpidatatypes
+  
+  implicit none
+
+  private
+
+  ! --- Public Routines ---
+  public :: create_subarray_datatype,free_datatype
+  
+  ! --- Public --- !
+
+  ! In the comments below, "fields" refers only to NON gyro-averaged fields.
+  ! All ghost cells have two points, used for finite difference derivatives.
+
+  ! Non-spectral scheme only:
+  ! "ga-fields" refers to gyro-averaged fields, which exist only for the non-spectral scheme.
+  ! The x ghost cells may have > 2 points, used only for taking gyro-averages.
+
+  integer, public, save :: TYPE_NEXT_MU     !< dist. only. Communicated in fdis_tmp using gc1
+  integer, public, save :: TYPE_PREV_MU     !< dist. only. Communicated in fdis_tmp using gc1
+  integer, public, save :: TYPE_NEXT_VPAR   !< dist. only. Communicated in fdis_tmp using gc1
+  integer, public, save :: TYPE_PREV_VPAR   !< dist. only. Communicated in fdis_tmp using gc1
+  integer, public, save :: TYPE_NEXT_S      !< dist., fields, and ga-fields.  Communicated in fdis_tmp using gc1
+  integer, public, save :: TYPE_PREV_S      !< dist., fields, and ga-fields.  Communicated in fdis_tmp using gc1
+  integer, public, save :: TYPE_NEXT_X_PHI  !< fields only (all gp). Communicated in fdis_tmp2 using gc2x_phi
+  integer, public, save :: TYPE_PREV_X_PHI  !< fields only (all gp). Communicated in fdis_tmp2 using gc2x_phi
+  integer, public, save :: TYPE_NEXT_X_PGA  !< ga-fields only (all gp). Communicated in fdis_tmp2 using gc2x_pga
+  integer, public, save :: TYPE_PREV_X_PGA  !< ga-fields only (all gp). Communicated in fdis_tmp2 using gc2x_pga
[…]

### 2022-05-03 gkw/src/non_linear_terms.F90

@@ -1,0 +1,2800 @@
+!-----------------------------------------------------------------------------
+!> Calculates the nonlinear term III with FFT pseudospectral method.
+!> For nonlinear runs, the code spends 40% (or much more) of time here.
+!> This module is optimised, contains OpenMP, and (optional) perflib timings
+!-----------------------------------------------------------------------------
+module non_linear_terms
+
+  implicit none
+
+  private
+
+  public :: add_non_linear_terms, jind, jinv, mphi, mphiw3, mrad
+  public :: nonlinear_allocate, nonlinear_init, nl_initialised
+  public :: entropy_radial
+  public :: get_extended_firstdim_fft_size
+  public :: mbd, ibd
+  
+  !> Index array for the storage of the kx modes in the arrays for the FFT
+  !> jind(nx)
+  integer, save, allocatable :: jind(:)
+
+  !> Inverse index array for the storage of the kx modes in the arrays for the
+  !> FFT. jinv(mrad)
+  integer, save, allocatable :: jinv(:)
+
+  !>Nested loop indexing array
+  integer, save, allocatable :: i3loop(:,:)
+  ! arrays for the FFT
+
+  ! Throughout the code:
+  ! * x is used to refer to the radial (psi) direction
+  ! * y is used to refer to the perpendicular (zeta) direction within the flux
+  !     surface, This direction can be called 'polodial' but is also referred
+  !     to as 'toroidal'
+  ! * Gradient d/dx in k space is an ik_x multiplier
+
+  !> a(mphiw3,mrad) In the spectral method 
+  !> 1st usage: a = grad_y_k <phi_k> = zeta gradient of the gyroaverage of
+  !>                potential in k space.
+  !> 2nd usage: a = grad_y_k <A||_k>, electromagnetic terms
+  !> 3rd usage: a = grad_y_k <B||_k>, compressional terms
+  !> 4th usage: a = grad_x f_k = radial gradient of the distribution function
+  !>                in k space.
+  !> a(mphiw3,-1:nx+2) In the non_spectral method 
+  !> 1st usage  a = phi  
+  complex, save, allocatable :: a(:,:,:)
+  complex, save, allocatable :: aa(:,:,:,:)
+
+  !> b(mphiw3,mrad) In the spectral method 
+  !> 1st usage: b = grad_x_k <phi_k> = radial gradient of the gyroaverage of
[…]

### 2022-05-03 gkw/src/umfpack_wrapper.c

@@ -1,0 +1,991 @@
+/* Minimal UMFPACK wrapper for GKW using non-standard calls to C from Fortran */
+/* $Id:$                                                                      */
+/* This is a slightly modified and simplified version of the umf4_f77wrapper
+   and umf4_f77zwrapper provided with UMFPACK. It should be used when the
+   Fortran 2003 standard ISO_C_BINDING is not available. Note that some of the
+   original comments below may not apply modern Fortran.                      */
+  
+/* ========================================================================== */
+/* === umf4_f77wrapper ====================================================== */
+/* ========================================================================== */
+
+/* -------------------------------------------------------------------------- */
+/* UMFPACK Copyright (c) Timothy A. Davis, CISE,                              */
+/* Univ. of Florida.  All Rights Reserved.  See ../Doc/License for License.   */
+/* web: http://www.cise.ufl.edu/research/sparse/umfpack                       */
+/* -------------------------------------------------------------------------- */
+
+/* FORTRAN interface for the C-callable UMFPACK library (double / int version
+ * only and double / UF_long versions only).  This is HIGHLY non-portable.  You
+ * will need to modify this depending on how your FORTRAN and C compilers
+ * behave.  This has been tested in Linux, Sun Solaris, SGI IRIX, and IBM AIX,
+ * with various compilers.  It has not been exhaustively tested on all possible
+ * combinations of C and FORTRAN compilers.  The UF_long version works on
+ * Solaris, SGI IRIX, and IBM AIX when the UMFPACK library is compiled in
+ * 64-bit mode.
+ *
+ * Only a subset of UMFPACK's capabilities are provided.  Refer to the UMFPACK
+ * User Guide for details.
+ *
+ * For some C and FORTRAN compilers, the FORTRAN compiler appends a single
+ * underscore ("_") after each routine name.  C doesn't do this, so the
+ * translation is made here.  Other FORTRAN compilers treat underscores
+ * differently.  For example, a FORTRAN call to a_b gets translated to a call
+ * to a_b__ by g77, and to a_b_ by most other FORTRAN compilers.  Thus, the
[…]

### 2022-05-03 gkw/src/version.F90

@@ -1,0 +1,188 @@
+!-----------------------------------------------------------------------
+!> This module provides gkw version information from gkw_info.h
+!> (created from compile-time git describe) and reports information on 
+!> executable capabilities, as selected by the preprocessing options.
+!> 
+!> This module depends on the gkw_info.h header, which changes whenever
+!> the git version hash changes, hence this module is near the top of 
+!> the dependency tree to enable faster compilations.
+!-----------------------------------------------------------------------
+
+module version
+
+  implicit none
+  
+  private
+  
+  ! To limit depdencies, should not be used other than in init  
+  public :: gkw_info, output_header
+  
+  ! provide "character (len=8) :: GKW_REV = <git version>",
+  ! generated by git describe, called via the makefile
+#ifdef NOVERSION
+  character (len=32),  parameter, public :: GKW_REV = 'NOVERSION'
+  character (len=32),  parameter, public :: GKW_EXE = 'gkw.x'
+  character (len=128), parameter, public :: GKW_FC =  'fortran_compiler'
+  character (len=128), parameter, public :: MPIRUNCMD =  'none'
+#else
+  include 'gkw_info.h'
+#endif
+
+contains
+
+  !-----------------------------------------------------------------------------
+  !> Report the basic code capabilities then abort. This routine is called when
+  !> no input file is present.
+  !-----------------------------------------------------------------------------
+  subroutine gkw_info()
+    use global,       only : compiled_with_slepc, compiled_with_hdf5
+    use global,       only : compiled_with_openmp, openmp_date
+    use global,       only : int2char, compiled_with_umfpack
+    use global,       only : have_double_precision, rp
+    use fft,          only : fftlib, working_fft_library
+    use mpiinterface, only : mpifinalize, mpistandard, root_processor
+#ifdef HAVE_HDF5
+    use io_hdf5, only : get_libversion
+#endif
+    character (len=6) :: real_precision
+    character (len=12) :: have_openmp
+    character (len=1) :: have_slepc
+    character (len=1) :: usable_eiv, usable_nonspec, usable_imp, usable_nonlin
+    character (len=15) :: have_hdf5
+
+    if (have_double_precision) then
[…]

### 2022-09-01 publication/publication.tex
#37: Latex Template

% ****** Start of file aipsamp.tex ******
%
%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
%   Version 4.1 of REVTeX, October 2009
%
%   Copyright (c) 2009 American Institute of Physics.
%
%   See the AIP README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
% 
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex  aipsamp
%  2)  bibtex aipsamp
%  3)  latex  aipsamp
%  4)  latex  aipsamp
%
% Use this file as a source of example code for your aip document.
% Use the file aiptemplate.tex as a template for your document.
\documentclass[%
 aip,
% jmp,
% bmf,
% sd,
% rsi,
 amsmath,amssymb,
%preprint,%
 reprint,%
%author-year,%
%author-numerical,%
% Conference Proceedings
]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{etoolbox}

%% Apr 2021: AIP requests that the corresponding 
%% email to be moved after the affiliations
\makeatletter
\def\@email#1#2{%
 \endgroup
 \patchcmd{\titleblock@produce}
  {\frontmatter@RRAPformat}
  {\frontmatter@RRAPformat{\produce@RRAP{*#1\href{mailto:#2}{#2}}}\frontmatter@RRAPformat}
  {}{}
}%
\makeatother
\begin{document}

\preprint{AIP/123-QED}

\title[Sample title]{Sample Title:\\with Forced Linebreak}
% Force line breaks with \\
\author{A. Author}
 \altaffiliation[Also at ]{Physics Department, XYZ University.}%Lines break automatically or can be forced with \\
\author{B. Author}%
 \email{Second.Author@institution.edu.}
\affiliation{ 
Authors' institution and/or address%\\This line break forced with \textbackslash\textbackslash
}%

\author{C. Author}
 \homepage{http://www.Second.institution.edu/~Charlie.Author.}
\affiliation{%
Second institution and/or address%\\This line break forced% with \\
}%

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified

\begin{abstract}
An article usually includes an abstract, a concise summary of the work
covered at length in the main body of the article. It is used for
secondary publications and for information retrieval purposes. 
\end{abstract}

\maketitle

\begin{quotation}
The ``lead paragraph'' is encapsulated with the \LaTeX\ 
\verb+quotation+ environment and is formatted as a single paragraph before the first section heading. 
(The \verb+quotation+ environment reverts to its usual meaning after the first sectioning command.) 
Note that numbered references are allowed in the lead paragraph.
%
The lead paragraph will only be found in an article being prepared for the journal \textit{Chaos}.
\end{quotation}

\section{\label{sec:level1}First-level heading:\protect\\ The line
break was forced \lowercase{via} \textbackslash\textbackslash}

This sample document demonstrates proper use of REV\TeX~4.1 (and
\LaTeXe) in manuscripts prepared for submission to AIP
journals. Further information can be found in the documentation included in the distribution or available at
\url{http://authors.aip.org} and in the documentation for 
REV\TeX~4.1 itself.

When commands are referred to in this example file, they are always
shown with their required arguments, using normal \TeX{} format. In
this format, \verb+#1+, \verb+#2+, etc. stand for required
author-supplied arguments to commands. For example, in
\verb+\section{#1}+ the \verb+#1+ stands for the title text of the
author's section heading, and in \verb+\title{#1}+ the \verb+#1+
stands for the title text of the paper.

Line breaks in section headings at all levels can be introduced using
\textbackslash\textbackslash. A blank input line tells \TeX\ that the
paragraph has ended.

\subsection{\label{sec:level2}Second-level heading: Formatting}

This file may be formatted in both the \texttt{preprint} (the default) and
\texttt{reprint} styles; the latter format may be used to 
mimic final journal output. Either format may be used for submission
purposes; however, for peer review and production, AIP will format the
article using the \texttt{preprint} class option. Hence, it is
essential that authors check that their manuscripts format acceptably
under \texttt{preprint}. Manuscripts submitted to AIP that do not
format correctly under the \texttt{preprint} option may be delayed in
both the editorial and production processes.

The \texttt{widetext} environment will make the text the width of the
full page, as on page~\pageref{eq:wideeq}. (Note the use the
\verb+\pageref{#1}+ to get the page number right automatically.) The
width-changing commands only take effect in \texttt{twocolumn}
formatting. It has no effect if \texttt{preprint} formatting is chosen
instead.

\subsubsection{\label{sec:level3}Third-level heading: Citations and Footnotes}

Citations in text refer to entries in the Bibliography;
they use the commands \verb+\cite{#1}+ or \verb+\onlinecite{#1}+. 
Because REV\TeX\ uses the \verb+natbib+ package of Patrick Daly, 
its entire repertoire of commands are available in your document;
see the \verb+natbib+ documentation for further details.
The argument of \verb+\cite+ is a comma-separated list of \emph{keys};
a key may consist of letters and numerals.

By default, citations are numerical; \cite{feyn54} author-year citations are an option. 
To give a textual citation, use \verb+\onlinecite{#1}+: (Refs.~\onlinecite{witten2001,epr,Bire82}). 
REV\TeX\ ``collapses'' lists of consecutive numerical citations when appropriate. 
REV\TeX\ provides the ability to properly punctuate textual citations in author-year style;
this facility works correctly with numerical citations only with \texttt{natbib}'s compress option turned off. 
To illustrate, we cite several together \cite{feyn54,witten2001,epr,Berman1983}, 
and once again (Refs.~\onlinecite{epr,feyn54,Bire82,Berman1983}). 
Note that, when numerical citations are used, the references were sorted into the same order they appear in the bibliography.

A reference within the bibliography is specified with a \verb+\bibitem{#1}+ command,
where the argument is the citation key mentioned above. 
\verb+\bibitem{#1}+ commands may be crafted by hand or, preferably,
generated by using Bib\TeX. 
The AIP styles for REV\TeX~4 include Bib\TeX\ style files
\verb+aipnum.bst+ and \verb+aipauth.bst+, appropriate for
numbered and author-year bibliographies,
respectively. 
REV\TeX~4 will automatically choose the style appropriate for 
the document's selected class options: the default is numerical, and
you obtain the author-year style by specifying a class option of \verb+author-year+.

This sample file demonstrates a simple use of Bib\TeX\ 
via a \verb+\bibliography+ command referencing the \verb+aipsamp.bib+ file.
Running Bib\TeX\ (in this case \texttt{bibtex
aipsamp}) after the first pass of \LaTeX\ produces the file
\verb+aipsamp.bbl+ which contains the automatically formatted
\verb+\bibitem+ commands (including extra markup information via
\verb+\bibinfo+ commands). If not using Bib\TeX, the
\verb+thebibiliography+ environment should be used instead.

\paragraph{Fourth-level heading is run in.}%
Footnotes are produced using the \verb+\footnote{#1}+ command. 
Numerical style citations put footnotes into the 
bibliography\footnote{Automatically placing footnotes into the bibliography requires using BibTeX to compile the bibliography.}.
Author-year and numerical author-year citation styles (each for its own reason) cannot use this method. 
Note: due to the method used to place footnotes in the bibliography, \emph{you
must re-run BibTeX every time you change any of your document's
footnotes}.

\section{Math and Equations}
Inline math may be typeset using the \verb+$+ delimiters. Bold math
symbols may be achieved using the \verb+bm+ package and the
\verb+\bm{#1}+ command it supplies. For instance, a bold $\alpha$ can
be typeset as \verb+$\bm{\alpha}$+ giving $\bm{\alpha}$. Fraktur and
Blackboard (or open face or double struck) characters should be
typeset using the \verb+\mathfrak{#1}+ and \verb+\mathbb{#1}+ commands
respectively. Both are supplied by the \texttt{amssymb} package. For
example, \verb+$\mathbb{R}$+ gives $\mathbb{R}$ and
\verb+$\mathfrak{G}$+ gives $\mathfrak{G}$

In \LaTeX\ there are many different ways to display equations, and a
few preferred ways are noted below. Displayed math will center by
default. Use the class option \verb+fleqn+ to flush equations left.

Below we have numbered single-line equations, the most common kind: 
\begin{eqnarray}
\chi_+(p)\alt{\bf [}2|{\bf p}|(|{\bf p}|+p_z){\bf ]}^{-1/2}
\left(
\begin{array}{c}
|{\bf p}|+p_z\\
px+ip_y
\end{array}\right)\;,
\\
\left\{%
 \openone234567890abc123\alpha\beta\gamma\delta1234556\alpha\beta
 \frac{1\sum^{a}_{b}}{A^2}%
\right\}%
\label{eq:one}.
\end{eqnarray}
Note the open one in Eq.~(\ref{eq:one}).

Not all numbered equations will fit within a narrow column this
way. The equation number will move down automatically if it cannot fit
on the same line with a one-line equation:
\begin{equation}
\left\{
 ab12345678abc123456abcdef\alpha\beta\gamma\delta1234556\alpha\beta
 \frac{1\sum^{a}_{b}}{A^2}%
\right\}.
\end{equation}

When the \verb+\label{#1}+ command is used [cf. input for
Eq.~(\ref{eq:one})], the equation can be referred to in text without
knowing the equation number that \TeX\ will assign to it. Just
use \verb+\ref{#1}+, where \verb+#1+ is the same name that used in
the \verb+\label{#1}+ command.

Unnumbered single-line equations can be typeset
using the \verb+\[+, \verb+\]+ format:
\[g^+g^+ \rightarrow g^+g^+g^+g^+ \dots ~,~~q^+q^+\rightarrow
q^+g^+g^+ \dots ~. \]

\subsection{Multiline equations}

Multiline equations are obtained by using the \verb+eqnarray+
environment.  Use the \verb+\nonumber+ command at the end of each line
to avoid assigning a number:
\begin{eqnarray}
{\cal M}=&&ig_Z^2(4E_1E_2)^{1/2}(l_i^2)^{-1}
\delta_{\sigma_1,-\sigma_2}
(g_{\sigma_2}^e)^2\chi_{-\sigma_2}(p_2)\nonumber\\
&&\times
[\epsilon_jl_i\epsilon_i]_{\sigma_1}\chi_{\sigma_1}(p_1),
\end{eqnarray}
\begin{eqnarray}
\sum \vert M^{\text{viol}}_g \vert ^2&=&g^{2n-4}_S(Q^2)~N^{n-2}
        (N^2-1)\nonumber \\
 & &\times \left( \sum_{i<j}\right)
  \sum_{\text{perm}}
 \frac{1}{S_{12}}
 \frac{1}{S_{12}}
 \sum_\tau c^f_\tau~.
\end{eqnarray}
\textbf{Note:} Do not use \verb+\label{#1}+ on a line of a multiline
equation if \verb+\nonumber+ is also used on that line. Incorrect
cross-referencing will result. Notice the use \verb+\text{#1}+ for
using a Roman font within a math environment.

To set a multiline equation without \emph{any} equation
numbers, use the \verb+\begin{eqnarray*}+,
\verb+\end{eqnarray*}+ format:
\begin{eqnarray*}
\sum \vert M^{\text{viol}}_g \vert ^2&=&g^{2n-4}_S(Q^2)~N^{n-2}
        (N^2-1)\\
 & &\times \left( \sum_{i<j}\right)
 \left(
  \sum_{\text{perm}}\frac{1}{S_{12}S_{23}S_{n1}}
 \right)
 \frac{1}{S_{12}}~.
\end{eqnarray*}
To obtain numbers not normally produced by the automatic numbering,
use the \verb+\tag{#1}+ command, where \verb+#1+ is the desired
equation number. For example, to get an equation number of
(\ref{eq:mynum}),
\begin{equation}
g^+g^+ \rightarrow g^+g^+g^+g^+ \dots ~,~~q^+q^+\rightarrow
q^+g^+g^+ \dots ~. \tag{2.6$'$}\label{eq:mynum}
\end{equation}

A few notes on \verb=\tag{#1}=. \verb+\tag{#1}+ requires
\texttt{amsmath}. The \verb+\tag{#1}+ must come before the
\verb+\label{#1}+, if any. The numbering set with \verb+\tag{#1}+ is
\textit{transparent} to the automatic numbering in REV\TeX{};
therefore, the number must be known ahead of time, and it must be
manually adjusted if other equations are added. \verb+\tag{#1}+ works
with both single-line and multiline equations. \verb+\tag{#1}+ should
only be used in exceptional case - do not use it to number all
equations in a paper.

Enclosing single-line and multiline equations in
\verb+\begin{subequations}+ and \verb+\end{subequations}+ will produce
a set of equations that are ``numbered'' with letters, as shown in
Eqs.~(\ref{subeq:1}) and (\ref{subeq:2}) below:
\begin{subequations}
\label{eq:whole}
\begin{equation}
\left\{
 abc123456abcdef\alpha\beta\gamma\delta1234556\alpha\beta
 \frac{1\sum^{a}_{b}}{A^2}
\right\},\label{subeq:1}
\end{equation}
\begin{eqnarray}
{\cal M}=&&ig_Z^2(4E_1E_2)^{1/2}(l_i^2)^{-1}
(g_{\sigma_2}^e)^2\chi_{-\sigma_2}(p_2)\nonumber\\
&&\times
[\epsilon_i]_{\sigma_1}\chi_{\sigma_1}(p_1).\label{subeq:2}
\end{eqnarray}
\end{subequations}
Putting a \verb+\label{#1}+ command right after the
\verb+\begin{subequations}+, allows one to
reference all the equations in a subequations environment. For
example, the equations in the preceding subequations environment were
Eqs.~(\ref{eq:whole}).

\subsubsection{Wide equations}
The equation that follows is set in a wide format, i.e., it spans
across the full page. The wide format is reserved for long equations
that cannot be easily broken into four lines or less:
\begin{widetext}
\begin{equation}
{\cal R}^{(\text{d})}=
 g_{\sigma_2}^e
 \left(
   \frac{[\Gamma^Z(3,21)]_{\sigma_1}}{Q_{12}^2-M_W^2}
  +\frac{[\Gamma^Z(13,2)]_{\sigma_1}}{Q_{13}^2-M_W^2}
 \right)
 + x_WQ_e
 \left(
   \frac{[\Gamma^\gamma(3,21)]_{\sigma_1}}{Q_{12}^2-M_W^2}
  +\frac{[\Gamma^\gamma(13,2)]_{\sigma_1}}{Q_{13}^2-M_W^2}
 \right)\;. \label{eq:wideeq}
\end{equation}
\end{widetext}
This is typed to show the output is in wide format.
(Since there is no input line between \verb+\equation+ and
this paragraph, there is no paragraph indent for this paragraph.)
\section{Cross-referencing}
REV\TeX{} will automatically number sections, equations, figure
captions, and tables. In order to reference them in text, use the
\verb+\label{#1}+ and \verb+\ref{#1}+ commands. To reference a
particular page, use the \verb+\pageref{#1}+ command.

The \verb+\label{#1}+ should appear in a section heading, within an
equation, or in a table or figure caption. The \verb+\ref{#1}+ command
is used in the text where the citation is to be displayed.  Some
examples: Section~\ref{sec:level1} on page~\pageref{sec:level1},
Table~\ref{tab:table1},%
\begin{table}
\caption{\label{tab:table1}This is a narrow table which fits into a
text column when using \texttt{twocolumn} formatting. Note that
REV\TeX~4 adjusts the intercolumn spacing so that the table fills the
entire width of the column. Table captions are numbered
automatically. This table illustrates left-aligned, centered, and
right-aligned columns.  }
\begin{ruledtabular}
\begin{tabular}{lcr}
Left\footnote{Note a.}&Centered\footnote{Note b.}&Right\\
\hline
1 & 2 & 3\\
10 & 20 & 30\\
100 & 200 & 300\\
\end{tabular}
\end{ruledtabular}
\end{table}
and Fig.~\ref{fig:epsart}.

\section{Figures and Tables}
Figures and tables are typically ``floats''; \LaTeX\ determines their
final position via placement rules. 
\LaTeX\ isn't always successful in automatically placing floats where you wish them.

Figures are marked up with the \texttt{figure} environment, the content of which
imports the image (\verb+\includegraphics+) followed by the figure caption (\verb+\caption+).
The argument of the latter command should itself contain a \verb+\label+ command if you
wish to refer to your figure with \verb+\ref+.

Import your image using either the \texttt{graphics} or
\texttt{graphix} packages. These packages both define the
\verb+\includegraphics{#1}+ command, but they differ in the optional
arguments for specifying the orientation, scaling, and translation of the figure.
Fig.~\ref{fig:epsart}%
\begin{figure}
%\includegraphics{fig_1}% Here is how to import EPS art
\caption{\label{fig:epsart} A figure caption. The figure captions are
automatically numbered.}
\end{figure}
is small enough to fit in a single column, while
Fig.~\ref{fig:wide}%
\begin{figure*}
%\includegraphics{fig_2}% Here is how to import EPS art
\caption{\label{fig:wide}Use the \texttt{figure*} environment to get a wide
figure, spanning the page in \texttt{twocolumn} formatting.}
\end{figure*}
is too wide for a single column,
so instead the \texttt{figure*} environment has been used.

The analog of the \texttt{figure} environment is \texttt{table}, which uses
the same \verb+\caption+ command.
However, you should type your caption command first within the \texttt{table}, 
instead of last as you did for \texttt{figure}.

The heart of any table is the \texttt{tabular} environment,
which represents the table content as a (vertical) sequence of table rows,
each containing a (horizontal) sequence of table cells. 
Cells are separated by the \verb+&+ character;
the row terminates with \verb+\\+. 
The required argument for the \texttt{tabular} environment
specifies how data are displayed in each of the columns. 
For instance, a column
may be centered (\verb+c+), left-justified (\verb+l+), right-justified (\verb+r+),
or aligned on a decimal point (\verb+d+). 
(Table~\ref{tab:table4}%
\begin{table}
\caption{\label{tab:table4}Numbers in columns Three--Five have been
aligned by using the ``d'' column specifier (requires the
\texttt{dcolumn} package). 
Non-numeric entries (those entries without
a ``.'') in a ``d'' column are aligned on the decimal point. 
Use the
``D'' specifier for more complex layouts. }
\begin{ruledtabular}
\begin{tabular}{ccddd}
One&Two&\mbox{Three}&\mbox{Four}&\mbox{Five}\\
\hline
one&two&\mbox{three}&\mbox{four}&\mbox{five}\\
He&2& 2.77234 & 45672. & 0.69 \\
C\footnote{Some tables require footnotes.}
  &C\footnote{Some tables need more than one footnote.}
  & 12537.64 & 37.66345 & 86.37 \\
\end{tabular}
\end{ruledtabular}
\end{table}
illustrates the use of decimal column alignment.)

Extra column-spacing may be be specified as well, although
REV\TeX~4 sets this spacing so that the columns fill the width of the
table.
Horizontal rules are typeset using the \verb+\hline+
command.
The doubled (or Scotch) rules that appear at the top and
bottom of a table can be achieved by enclosing the \texttt{tabular}
environment within a \texttt{ruledtabular} environment.
Rows whose columns span multiple columns can be typeset using \LaTeX's
\verb+\multicolumn{#1}{#2}{#3}+ command
(for example, see the first row of Table~\ref{tab:table3}).%
\begin{table*}
\caption{\label{tab:table3}This is a wide table that spans the page
width in \texttt{twocolumn} mode. It is formatted using the
\texttt{table*} environment. It also demonstrates the use of
\textbackslash\texttt{multicolumn} in rows with entries that span
more than one column.}
\begin{ruledtabular}
\begin{tabular}{ccccc}
 &\multicolumn{2}{c}{$D_{4h}^1$}&\multicolumn{2}{c}{$D_{4h}^5$}\\
 Ion&1st alternative&2nd alternative&lst alternative
&2nd alternative\\ \hline
 K&$(2e)+(2f)$&$(4i)$ &$(2c)+(2d)$&$(4f)$ \\
 Mn&$(2g)$\footnote{The $z$ parameter of these positions is $z\sim\frac{1}{4}$.}
 &$(a)+(b)+(c)+(d)$&$(4e)$&$(2a)+(2b)$\\
 Cl&$(a)+(b)+(c)+(d)$&$(2g)$\footnote{This is a footnote in a table that spans the full page
width in \texttt{twocolumn} mode. It is supposed to set on the full width of the page, just as the caption does. }
 &$(4e)^{\text{a}}$\\
 He&$(8r)^{\text{a}}$&$(4j)^{\text{a}}$&$(4g)^{\text{a}}$\\
 Ag& &$(4k)^{\text{a}}$& &$(4h)^{\text{a}}$\\
\end{tabular}
\end{ruledtabular}
\end{table*}

The tables in this document illustrate various effects.
Tables that fit in a narrow column are contained in a \texttt{table}
environment.
Table~\ref{tab:table3} is a wide table, therefore set with the
\texttt{table*} environment.
Lengthy tables may need to break across pages.
A simple way to allow this is to specify
the \verb+[H]+ float placement on the \texttt{table} or
\texttt{table*} environment.
Alternatively, using the standard \LaTeXe\ package \texttt{longtable} 
gives more control over how tables break and allows headers and footers 
to be specified for each page of the table.
An example of the use of \texttt{longtable} can be found
in the file \texttt{summary.tex} that is included with the REV\TeX~4
distribution.

There are two methods for setting footnotes within a table (these
footnotes will be displayed directly below the table rather than at
the bottom of the page or in the bibliography).
The easiest
and preferred method is just to use the \verb+\footnote{#1}+
command. This will automatically enumerate the footnotes with
lowercase roman letters.
However, it is sometimes necessary to have
multiple entries in the table share the same footnote.
In this case,
create the footnotes using
\verb+\footnotemark[#1]+ and \verb+\footnotetext[#1]{#2}+.
\texttt{\#1} is a numeric value.
Each time the same value for \texttt{\#1} is used, 
the same mark is produced in the table. 
The \verb+\footnotetext[#1]{#2}+ commands are placed after the \texttt{tabular}
environment. 
Examine the \LaTeX\ source and output for Tables~\ref{tab:table1} and 
\ref{tab:table2}%
\begin{table}
\caption{\label{tab:table2}A table with more columns still fits
properly in a column. Note that several entries share the same
footnote. Inspect the \LaTeX\ input for this table to see
exactly how it is done.}
\begin{ruledtabular}
\begin{tabular}{cccccccc}
 &$r_c$ (\AA)&$r_0$ (\AA)&$\kappa r_0$&
 &$r_c$ (\AA) &$r_0$ (\AA)&$\kappa r_0$\\
\hline
Cu& 0.800 & 14.10 & 2.550 &Sn\footnotemark[1]
& 0.680 & 1.870 & 3.700 \\
Ag& 0.990 & 15.90 & 2.710 &Pb\footnotemark[2]
& 0.450 & 1.930 & 3.760 \\
Au& 1.150 & 15.90 & 2.710 &Ca\footnotemark[3]
& 0.750 & 2.170 & 3.560 \\
Mg& 0.490 & 17.60 & 3.200 &Sr\footnotemark[4]
& 0.900 & 2.370 & 3.720 \\
Zn& 0.300 & 15.20 & 2.970 &Li\footnotemark[2]
& 0.380 & 1.730 & 2.830 \\
Cd& 0.530 & 17.10 & 3.160 &Na\footnotemark[5]
& 0.760 & 2.110 & 3.120 \\
Hg& 0.550 & 17.80 & 3.220 &K\footnotemark[5]
&  1.120 & 2.620 & 3.480 \\
Al& 0.230 & 15.80 & 3.240 &Rb\footnotemark[3]
& 1.330 & 2.800 & 3.590 \\
Ga& 0.310 & 16.70 & 3.330 &Cs\footnotemark[4]
& 1.420 & 3.030 & 3.740 \\
In& 0.460 & 18.40 & 3.500 &Ba\footnotemark[5]
& 0.960 & 2.460 & 3.780 \\
Tl& 0.480 & 18.90 & 3.550 & & & & \\
\end{tabular}
\end{ruledtabular}
\footnotetext[1]{Here's the first, from Ref.~\onlinecite{feyn54}.}
\footnotetext[2]{Here's the second.}
\footnotetext[3]{Here's the third.}
\footnotetext[4]{Here's the fourth.}
\footnotetext[5]{And etc.}
\end{table}
for an illustration.

All AIP journals require that the initial citation of
figures or tables be in numerical order.
\LaTeX's automatic numbering of floats is your friend here:
just put each \texttt{figure} environment immediately following 
its first reference (\verb+\ref+), as we have done in this example file.

\begin{acknowledgments}
We wish to acknowledge the support of the author community in using
REV\TeX{}, offering suggestions and encouragement, testing new versions,
\dots.
\end{acknowledgments}

\section*{Data Availability Statement}

AIP Publishing believes that all datasets underlying the conclusions of the paper should be available to readers. Authors are encouraged to deposit their datasets in publicly available repositories or present them in the main manuscript. All research articles must include a data availability statement stating where the data can be found. In this section, authors should add the respective statement from the chart below based on the availability of data in their paper.

\begin{center}
\renewcommand\arraystretch{1.2}
\begin{tabular}{| >{\raggedright\arraybackslash}p{0.3\linewidth} | >{\raggedright\arraybackslash}p{0.65\linewidth} |}
\hline
\textbf{AVAILABILITY OF DATA} & \textbf{STATEMENT OF DATA AVAILABILITY}\\  
\hline
Data available on request from the authors
&
The data that support the findings of this study are available from the corresponding author upon reasonable request.
\\\hline
Data available in article or supplementary material
&
The data that support the findings of this study are available within the article [and its supplementary material].
\\\hline
Data openly available in a public repository that issues datasets with DOIs
&
The data that support the findings of this study are openly available in [repository name] at http://doi.org/[doi], reference number [reference number].
\\\hline
Data openly available in a public repository that does not issue DOIs
&
The data that support the findings of this study are openly available in [repository name], reference number [reference number].
\\\hline
Data sharing not applicable – no new data generated
&
Data sharing is not applicable to this article as no new data were created or analyzed in this study.
\\\hline
Data generated at a central, large scale facility
&
Raw data were generated at the [facility name] large scale facility. Derived data supporting the findings of this study are available from the corresponding author upon reasonable request.
\\\hline
Embargo on data due to commercial restrictions
&
The data that support the findings will be available in [repository name] at [DOI link] following an embargo from the date of publication to allow for commercialization of research findings.
\\\hline
Data available on request due to privacy/ethical restrictions
&
The data that support the findings of this study are available on request from the corresponding author. The data are not publicly available due [state restrictions such as privacy or ethical restrictions].
\\\hline
Data subject to third party restrictions
&
The data that support the findings of this study are available from [third party]. Restrictions apply to the availability of these data, which were used under license for this study. Data are available from the authors upon reasonable request and with the permission of [third party].
\\\hline
\end{tabular}
\end{center}

\appendix

\section{Appendixes}

To start the appendixes, use the \verb+\appendix+ command.
This signals that all following section commands refer to appendixes
instead of regular sections. Therefore, the \verb+\appendix+ command
should be used only once---to set up the section commands to act as
appendixes. Thereafter normal section commands are used. The heading
for a section can be left empty. For example,
\begin{verbatim}
\appendix
\section{}
\end{verbatim}
will produce an appendix heading that says ``APPENDIX A'' and
\begin{verbatim}
\appendix
\section{Background}
\end{verbatim}
will produce an appendix heading that says ``APPENDIX A: BACKGROUND''
(note that the colon is set automatically).

If there is only one appendix, then the letter ``A'' should not
appear. This is suppressed by using the star version of the appendix
command (\verb+\appendix*+ in the place of \verb+\appendix+).

\section{A little more on appendixes}

Observe that this appendix was started by using
\begin{verbatim}
\section{A little more on appendixes}
\end{verbatim}

Note the equation number in an appendix:
\begin{equation}
E=mc^2.
\end{equation}

\subsection{\label{app:subsec}A subsection in an appendix}

You can use a subsection or subsubsection in an appendix. Note the
numbering: we are now in Appendix~\ref{app:subsec}.

\subsubsection{\label{app:subsubsec}A subsubsection in an appendix}
Note the equation numbers in this appendix, produced with the
subequations environment:
\begin{subequations}
\begin{eqnarray}
E&=&mc, \label{appa}
\\
E&=&mc^2, \label{appb}
\\
E&\agt& mc^3. \label{appc}
\end{eqnarray}
\end{subequations}
They turn out to be Eqs.~(\ref{appa}), (\ref{appb}), and (\ref{appc}).

\nocite{*}
\bibliography{aipsamp}% Produces the bibliography via BibTeX.

\end{document}
%
% ****** End of file aipsamp.tex ******

### 2022-10-14 publication/publication.tex
#37: Word Counter function
#37: graphics function, cleaned document

@@
+% AUTHOR ===================================================================================================================
+% Manuel Lippert (GitHub: ManeLippert (https://github.com/ManeLippert))
+% ==========================================================================================================================
+
+% PREAMBLE =================================================================================================================
+
 % ****** Start of file aipsamp.tex ******
@@
 % Use the file aiptemplate.tex as a template for your document.
-\documentclass[%
- aip,
-% jmp,
-% bmf,
-% sd,
-% rsi,
- amsmath,amssymb,
-%preprint,%
- reprint,%
-%author-year,%
-%author-numerical,%
-% Conference Proceedings
-]{revtex4-1}
 
+% DOCUMENT =================================================================================================================
+
+\documentclass[aip, amsmath, amssymb, reprint, twocolumn]{revtex4-1}
+
+% documentclass revtex4-1 options:
+% - aip,
+% - jmp,
+% - bmf,
+% - sd,
+% - rsi,
+% - amsmath,amssymb,
+% - preprint,%
+% - reprint,%
+% - author-year,%
+% - author-numerical,%
+% - Conference Proceedings
+
+\preprint{AIP/123-QED}
+
+% PACKAGES =================================================================================================================
+
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{mathptmx}
+\usepackage{etoolbox}
 \usepackage{graphicx}% Include figure files
@@
 
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{mathptmx}
-\usepackage{etoolbox}
+%% Additional
+\usepackage{caption}
+%\usepackage{hyperref}
+
+% FUNCTIONS, INPUT =========================================================================================================
+
+\input{wordcount.tex}% Word & page count for document
+
+\graphicspath{{../pictures/}}% Path for pictures
 
@@
 \makeatletter
-\def\@email#1#2{%
+\def\@email#1#2{
 	\endgroup
@@
 		{}{}
-}%
+}
 \makeatother
+
+%% Include graphic for one column with specific place 
+\newcommand{\includegraphicsOneCol}[3]{
+	\begin{figure}[h]
+		\includegraphics[width=\linewidth]{#1}
+		\caption{#2}
+  	\end{figure}
+	\label{#3}
+  	\increasecounter{fig}{1}
+}
+
+%% Include graphic for two column with specific place, figrue* places graphic anywhere...
+\newcommand{\includegraphicsTwoCol}[3]{
+	\onecolumngrid
+	\begin{center}
+		\captionsetup{type=figure}
+    	\includegraphics[width=\textwidth]{#1}
+		\captionof{figure}{#2}
+	\end{center}
+	\twocolumngrid
+	\label{#3}
+	\increasecounter{fig}{2}
+}
+
+% MAIN =====================================================================================================================
+
 \begin{document}
 
-\preprint{AIP/123-QED}
+%% TITLE, INFO =============================================================================================================
 
-\title[Sample title]{Sample Title:\\with Forced Linebreak}
-% Force line breaks with \\
-\author{A. Author}
- \altaffiliation[Also at ]{Physics Department, XYZ University.}%Lines break automatically or can be forced with \\
-\author{B. Author}%
- \email{Second.Author@institution.edu.}
-\affiliation{ 
-Authors' institution and/or address%\\This line break forced with \textbackslash\textbackslash
-}%
+%\title[Convergence of shearingrate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize]{
+%	Convergence of shearingrate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize
+%}
 
-\author{C. Author}
- \homepage{http://www.Second.institution.edu/~Charlie.Author.}
-\affiliation{%
-Second institution and/or address%\\This line break forced% with \\
-}%
+%\author{M. Lippert}
+%	\altaffiliation{GitHub Repository for files reagarding this paper: \\ 
+%					https://github.com/ManeLippert/Bachelorthesis-ZonalFlows}
+%\author{F. Rath}
+%	\altaffiliation{Author to whom correspondence should be addressed: \\ 
+%					Florian.Rath1@uni-bayreuth.de}
+%	\email{Florian.Rath1@uni-bayreuth.de}
+%\author{A. G. Peeters}
+%\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
 
-\date{\today}% It is always \today, today,
-             %  but any date may be explicitly specified
+%\date{\today}
 
-\begin{abstract}
-An article usually includes an abstract, a concise summary of the work
-covered at length in the main body of the article. It is used for
-secondary publications and for information retrieval purposes. 
-\end{abstract}
+%% ABSTRACT ================================================================================================================
 
-\maketitle
+%\begin{abstract}
+%	An article usually includes an abstract, a concise summary of the work
+%	covered at length in the main body of the article. It is used for
+%	secondary publications and for information retrieval purposes. 
+%\end{abstract}
 
-\begin{quotation}
-The ``lead paragraph'' is encapsulated with the \LaTeX\ 
-\verb+quotation+ environment and is formatted as a single paragraph before the first section heading. 
-(The \verb+quotation+ environment reverts to its usual meaning after the first sectioning command.) 
-Note that numbered references are allowed in the lead paragraph.
-%
-The lead paragraph will only be found in an article being prepared for the journal \textit{Chaos}.
-\end{quotation}
+%\maketitle
 
-\section{\label{sec:level1}First-level heading:\protect\\ The line
-break was forced \lowercase{via} \textbackslash\textbackslash}
+%% TEXT ====================================================================================================================
 
-This sample document demonstrates proper use of REV\TeX~4.1 (and
-\LaTeXe) in manuscripts prepared for submission to AIP
-journals. Further information can be found in the documentation included in the distribution or available at
-\url{http://authors.aip.org} and in the documentation for 
-REV\TeX~4.1 itself.
+\section{\label{sec:level1}Introduction}
 
-When commands are referred to in this example file, they are always
-shown with their required arguments, using normal \TeX{} format. In
-this format, \verb+#1+, \verb+#2+, etc. stand for required
-author-supplied arguments to commands. For example, in
-\verb+\section{#1}+ the \verb+#1+ stands for the title text of the
-author's section heading, and in \verb+\title{#1}+ the \verb+#1+
-stands for the title text of the paper.
+\includegraphicsOneCol{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
 
-Line breaks in section headings at all levels can be introduced using
-\textbackslash\textbackslash. A blank input line tells \TeX\ that the
-paragraph has ended.
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize2x1-2-3x1-3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{Test}{fig:test}
 
-\subsection{\label{sec:level2}Second-level heading: Formatting}
+\ref{fig:test}
 
-This file may be formatted in both the \texttt{preprint} (the default) and
-\texttt{reprint} styles; the latter format may be used to 
-mimic final journal output. Either format may be used for submission
-purposes; however, for peer review and production, AIP will format the
-article using the \texttt{preprint} class option. Hence, it is
-essential that authors check that their manuscripts format acceptably
-under \texttt{preprint}. Manuscripts submitted to AIP that do not
-format correctly under the \texttt{preprint} option may be delayed in
-both the editorial and production processes.
+\section*{Data Availability}
+The data that support the findings of this study are available from the corresponding author upon reasonable request.
 
-The \texttt{widetext} environment will make the text the width of the
-full page, as on page~\pageref{eq:wideeq}. (Note the use the
-\verb+\pageref{#1}+ to get the page number right automatically.) The
-width-changing commands only take effect in \texttt{twocolumn}
-formatting. It has no effect if \texttt{preprint} formatting is chosen
-instead.
+% WORDCOUNT ================================================================================================================
 
-\subsubsection{\label{sec:level3}Third-level heading: Citations and Footnotes}
+\wordcountmessage
 
-Citations in text refer to entries in the Bibliography;
-they use the commands \verb+\cite{#1}+ or \verb+\onlinecite{#1}+. 
-Because REV\TeX\ uses the \verb+natbib+ package of Patrick Daly, 
-its entire repertoire of commands are available in your document;
-see the \verb+natbib+ documentation for further details.
-The argument of \verb+\cite+ is a comma-separated list of \emph{keys};
-a key may consist of letters and numerals.
-
-By default, citations are numerical; \cite{feyn54} author-year citations are an option. 
-To give a textual citation, use \verb+\onlinecite{#1}+: (Refs.~\onlinecite{witten2001,epr,Bire82}). 
-REV\TeX\ ``collapses'' lists of consecutive numerical citations when appropriate. 
-REV\TeX\ provides the ability to properly punctuate textual citations in author-year style;
-this facility works correctly with numerical citations only with \texttt{natbib}'s compress option turned off. 
-To illustrate, we cite several together \cite{feyn54,witten2001,epr,Berman1983}, 
-and once again (Refs.~\onlinecite{epr,feyn54,Bire82,Berman1983}). 
-Note that, when numerical citations are used, the references were sorted into the same order they appear in the bibliography.
-
-A reference within the bibliography is specified with a \verb+\bibitem{#1}+ command,
-where the argument is the citation key mentioned above. 
-\verb+\bibitem{#1}+ commands may be crafted by hand or, preferably,
-generated by using Bib\TeX. 
-The AIP styles for REV\TeX~4 include Bib\TeX\ style files
-\verb+aipnum.bst+ and \verb+aipauth.bst+, appropriate for
-numbered and author-year bibliographies,
-respectively. 
-REV\TeX~4 will automatically choose the style appropriate for 
-the document's selected class options: the default is numerical, and
-you obtain the author-year style by specifying a class option of \verb+author-year+.
-
-This sample file demonstrates a simple use of Bib\TeX\ 
-via a \verb+\bibliography+ command referencing the \verb+aipsamp.bib+ file.
-Running Bib\TeX\ (in this case \texttt{bibtex
-aipsamp}) after the first pass of \LaTeX\ produces the file
-\verb+aipsamp.bbl+ which contains the automatically formatted
-\verb+\bibitem+ commands (including extra markup information via
-\verb+\bibinfo+ commands). If not using Bib\TeX, the
-\verb+thebibiliography+ environment should be used instead.
-
-\paragraph{Fourth-level heading is run in.}%
-Footnotes are produced using the \verb+\footnote{#1}+ command. 
-Numerical style citations put footnotes into the 
-bibliography\footnote{Automatically placing footnotes into the bibliography requires using BibTeX to compile the bibliography.}.
-Author-year and numerical author-year citation styles (each for its own reason) cannot use this method. 
-Note: due to the method used to place footnotes in the bibliography, \emph{you
-must re-run BibTeX every time you change any of your document's
-footnotes}.
-
-\section{Math and Equations}
-Inline math may be typeset using the \verb+$+ delimiters. Bold math
-symbols may be achieved using the \verb+bm+ package and the
-\verb+\bm{#1}+ command it supplies. For instance, a bold $\alpha$ can
-be typeset as \verb+$\bm{\alpha}$+ giving $\bm{\alpha}$. Fraktur and
-Blackboard (or open face or double struck) characters should be
-typeset using the \verb+\mathfrak{#1}+ and \verb+\mathbb{#1}+ commands
-respectively. Both are supplied by the \texttt{amssymb} package. For
-example, \verb+$\mathbb{R}$+ gives $\mathbb{R}$ and
-\verb+$\mathfrak{G}$+ gives $\mathfrak{G}$
-
-In \LaTeX\ there are many different ways to display equations, and a
-few preferred ways are noted below. Displayed math will center by
-default. Use the class option \verb+fleqn+ to flush equations left.
-
-Below we have numbered single-line equations, the most common kind: 
-\begin{eqnarray}
-\chi_+(p)\alt{\bf [}2|{\bf p}|(|{\bf p}|+p_z){\bf ]}^{-1/2}
-\left(
-\begin{array}{c}
-|{\bf p}|+p_z\\
-px+ip_y
-\end{array}\right)\;,
-\\
-\left\{%
- \openone234567890abc123\alpha\beta\gamma\delta1234556\alpha\beta
- \frac{1\sum^{a}_{b}}{A^2}%
-\right\}%
-\label{eq:one}.
-\end{eqnarray}
-Note the open one in Eq.~(\ref{eq:one}).
-
-Not all numbered equations will fit within a narrow column this
-way. The equation number will move down automatically if it cannot fit
-on the same line with a one-line equation:
-\begin{equation}
-\left\{
- ab12345678abc123456abcdef\alpha\beta\gamma\delta1234556\alpha\beta
- \frac{1\sum^{a}_{b}}{A^2}%
-\right\}.
-\end{equation}
-
-When the \verb+\label{#1}+ command is used [cf. input for
-Eq.~(\ref{eq:one})], the equation can be referred to in text without
-knowing the equation number that \TeX\ will assign to it. Just
-use \verb+\ref{#1}+, where \verb+#1+ is the same name that used in
-the \verb+\label{#1}+ command.
-
-Unnumbered single-line equations can be typeset
-using the \verb+\[+, \verb+\]+ format:
-\[g^+g^+ \rightarrow g^+g^+g^+g^+ \dots ~,~~q^+q^+\rightarrow
-q^+g^+g^+ \dots ~. \]
-
-\subsection{Multiline equations}
-
-Multiline equations are obtained by using the \verb+eqnarray+
-environment.  Use the \verb+\nonumber+ command at the end of each line
-to avoid assigning a number:
-\begin{eqnarray}
-{\cal M}=&&ig_Z^2(4E_1E_2)^{1/2}(l_i^2)^{-1}
-\delta_{\sigma_1,-\sigma_2}
-(g_{\sigma_2}^e)^2\chi_{-\sigma_2}(p_2)\nonumber\\
-&&\times
-[\epsilon_jl_i\epsilon_i]_{\sigma_1}\chi_{\sigma_1}(p_1),
-\end{eqnarray}
-\begin{eqnarray}
-\sum \vert M^{\text{viol}}_g \vert ^2&=&g^{2n-4}_S(Q^2)~N^{n-2}
-        (N^2-1)\nonumber \\
- & &\times \left( \sum_{i<j}\right)
-  \sum_{\text{perm}}
- \frac{1}{S_{12}}
- \frac{1}{S_{12}}
- \sum_\tau c^f_\tau~.
-\end{eqnarray}
-\textbf{Note:} Do not use \verb+\label{#1}+ on a line of a multiline
-equation if \verb+\nonumber+ is also used on that line. Incorrect
-cross-referencing will result. Notice the use \verb+\text{#1}+ for
-using a Roman font within a math environment.
-
-To set a multiline equation without \emph{any} equation
-numbers, use the \verb+\begin{eqnarray*}+,
-\verb+\end{eqnarray*}+ format:
-\begin{eqnarray*}
-\sum \vert M^{\text{viol}}_g \vert ^2&=&g^{2n-4}_S(Q^2)~N^{n-2}
-        (N^2-1)\\
- & &\times \left( \sum_{i<j}\right)
- \left(
-  \sum_{\text{perm}}\frac{1}{S_{12}S_{23}S_{n1}}
- \right)
- \frac{1}{S_{12}}~.
-\end{eqnarray*}
-To obtain numbers not normally produced by the automatic numbering,
-use the \verb+\tag{#1}+ command, where \verb+#1+ is the desired
-equation number. For example, to get an equation number of
-(\ref{eq:mynum}),
-\begin{equation}
-g^+g^+ \rightarrow g^+g^+g^+g^+ \dots ~,~~q^+q^+\rightarrow
-q^+g^+g^+ \dots ~. \tag{2.6$'$}\label{eq:mynum}
-\end{equation}
-
-A few notes on \verb=\tag{#1}=. \verb+\tag{#1}+ requires
-\texttt{amsmath}. The \verb+\tag{#1}+ must come before the
-\verb+\label{#1}+, if any. The numbering set with \verb+\tag{#1}+ is
-\textit{transparent} to the automatic numbering in REV\TeX{};
-therefore, the number must be known ahead of time, and it must be
-manually adjusted if other equations are added. \verb+\tag{#1}+ works
-with both single-line and multiline equations. \verb+\tag{#1}+ should
-only be used in exceptional case - do not use it to number all
-equations in a paper.
-
-Enclosing single-line and multiline equations in
-\verb+\begin{subequations}+ and \verb+\end{subequations}+ will produce
-a set of equations that are ``numbered'' with letters, as shown in
-Eqs.~(\ref{subeq:1}) and (\ref{subeq:2}) below:
-\begin{subequations}
-\label{eq:whole}
-\begin{equation}
-\left\{
- abc123456abcdef\alpha\beta\gamma\delta1234556\alpha\beta
- \frac{1\sum^{a}_{b}}{A^2}
-\right\},\label{subeq:1}
-\end{equation}
-\begin{eqnarray}
-{\cal M}=&&ig_Z^2(4E_1E_2)^{1/2}(l_i^2)^{-1}
-(g_{\sigma_2}^e)^2\chi_{-\sigma_2}(p_2)\nonumber\\
-&&\times
-[\epsilon_i]_{\sigma_1}\chi_{\sigma_1}(p_1).\label{subeq:2}
-\end{eqnarray}
-\end{subequations}
-Putting a \verb+\label{#1}+ command right after the
-\verb+\begin{subequations}+, allows one to
-reference all the equations in a subequations environment. For
-example, the equations in the preceding subequations environment were
-Eqs.~(\ref{eq:whole}).
-
-\subsubsection{Wide equations}
-The equation that follows is set in a wide format, i.e., it spans
-across the full page. The wide format is reserved for long equations
-that cannot be easily broken into four lines or less:
-\begin{widetext}
-\begin{equation}
-{\cal R}^{(\text{d})}=
- g_{\sigma_2}^e
- \left(
-   \frac{[\Gamma^Z(3,21)]_{\sigma_1}}{Q_{12}^2-M_W^2}
-  +\frac{[\Gamma^Z(13,2)]_{\sigma_1}}{Q_{13}^2-M_W^2}
- \right)
- + x_WQ_e
- \left(
-   \frac{[\Gamma^\gamma(3,21)]_{\sigma_1}}{Q_{12}^2-M_W^2}
-  +\frac{[\Gamma^\gamma(13,2)]_{\sigma_1}}{Q_{13}^2-M_W^2}
- \right)\;. \label{eq:wideeq}
-\end{equation}
-\end{widetext}
-This is typed to show the output is in wide format.
-(Since there is no input line between \verb+\equation+ and
-this paragraph, there is no paragraph indent for this paragraph.)
-\section{Cross-referencing}
-REV\TeX{} will automatically number sections, equations, figure
-captions, and tables. In order to reference them in text, use the
-\verb+\label{#1}+ and \verb+\ref{#1}+ commands. To reference a
-particular page, use the \verb+\pageref{#1}+ command.
-
-The \verb+\label{#1}+ should appear in a section heading, within an
-equation, or in a table or figure caption. The \verb+\ref{#1}+ command
-is used in the text where the citation is to be displayed.  Some
-examples: Section~\ref{sec:level1} on page~\pageref{sec:level1},
-Table~\ref{tab:table1},%
-\begin{table}
-\caption{\label{tab:table1}This is a narrow table which fits into a
-text column when using \texttt{twocolumn} formatting. Note that
-REV\TeX~4 adjusts the intercolumn spacing so that the table fills the
-entire width of the column. Table captions are numbered
-automatically. This table illustrates left-aligned, centered, and
-right-aligned columns.  }
-\begin{ruledtabular}
-\begin{tabular}{lcr}
-Left\footnote{Note a.}&Centered\footnote{Note b.}&Right\\
-\hline
-1 & 2 & 3\\
-10 & 20 & 30\\
-100 & 200 & 300\\
-\end{tabular}
-\end{ruledtabular}
-\end{table}
-and Fig.~\ref{fig:epsart}.
-
-\section{Figures and Tables}
-Figures and tables are typically ``floats''; \LaTeX\ determines their
-final position via placement rules. 
-\LaTeX\ isn't always successful in automatically placing floats where you wish them.
-
-Figures are marked up with the \texttt{figure} environment, the content of which
-imports the image (\verb+\includegraphics+) followed by the figure caption (\verb+\caption+).
-The argument of the latter command should itself contain a \verb+\label+ command if you
-wish to refer to your figure with \verb+\ref+.
-
-Import your image using either the \texttt{graphics} or
-\texttt{graphix} packages. These packages both define the
-\verb+\includegraphics{#1}+ command, but they differ in the optional
-arguments for specifying the orientation, scaling, and translation of the figure.
-Fig.~\ref{fig:epsart}%
-\begin{figure}
-%\includegraphics{fig_1}% Here is how to import EPS art
-\caption{\label{fig:epsart} A figure caption. The figure captions are
-automatically numbered.}
-\end{figure}
-is small enough to fit in a single column, while
-Fig.~\ref{fig:wide}%
-\begin{figure*}
-%\includegraphics{fig_2}% Here is how to import EPS art
-\caption{\label{fig:wide}Use the \texttt{figure*} environment to get a wide
-figure, spanning the page in \texttt{twocolumn} formatting.}
-\end{figure*}
-is too wide for a single column,
-so instead the \texttt{figure*} environment has been used.
-
-The analog of the \texttt{figure} environment is \texttt{table}, which uses
-the same \verb+\caption+ command.
-However, you should type your caption command first within the \texttt{table}, 
-instead of last as you did for \texttt{figure}.
-
-The heart of any table is the \texttt{tabular} environment,
-which represents the table content as a (vertical) sequence of table rows,
-each containing a (horizontal) sequence of table cells. 
-Cells are separated by the \verb+&+ character;
-the row terminates with \verb+\\+. 
-The required argument for the \texttt{tabular} environment
-specifies how data are displayed in each of the columns. 
-For instance, a column
-may be centered (\verb+c+), left-justified (\verb+l+), right-justified (\verb+r+),
-or aligned on a decimal point (\verb+d+). 
-(Table~\ref{tab:table4}%
-\begin{table}
-\caption{\label{tab:table4}Numbers in columns Three--Five have been
-aligned by using the ``d'' column specifier (requires the
-\texttt{dcolumn} package). 
-Non-numeric entries (those entries without
-a ``.'') in a ``d'' column are aligned on the decimal point. 
-Use the
-``D'' specifier for more complex layouts. }
-\begin{ruledtabular}
-\begin{tabular}{ccddd}
-One&Two&\mbox{Three}&\mbox{Four}&\mbox{Five}\\
-\hline
-one&two&\mbox{three}&\mbox{four}&\mbox{five}\\
-He&2& 2.77234 & 45672. & 0.69 \\
-C\footnote{Some tables require footnotes.}
-  &C\footnote{Some tables need more than one footnote.}
-  & 12537.64 & 37.66345 & 86.37 \\
-\end{tabular}
-\end{ruledtabular}
-\end{table}
-illustrates the use of decimal column alignment.)
-
-Extra column-spacing may be be specified as well, although
-REV\TeX~4 sets this spacing so that the columns fill the width of the
-table.
-Horizontal rules are typeset using the \verb+\hline+
-command.
-The doubled (or Scotch) rules that appear at the top and
-bottom of a table can be achieved by enclosing the \texttt{tabular}
-environment within a \texttt{ruledtabular} environment.
-Rows whose columns span multiple columns can be typeset using \LaTeX's
-\verb+\multicolumn{#1}{#2}{#3}+ command
-(for example, see the first row of Table~\ref{tab:table3}).%
-\begin{table*}
-\caption{\label{tab:table3}This is a wide table that spans the page
-width in \texttt{twocolumn} mode. It is formatted using the
-\texttt{table*} environment. It also demonstrates the use of
-\textbackslash\texttt{multicolumn} in rows with entries that span
-more than one column.}
-\begin{ruledtabular}
-\begin{tabular}{ccccc}
- &\multicolumn{2}{c}{$D_{4h}^1$}&\multicolumn{2}{c}{$D_{4h}^5$}\\
- Ion&1st alternative&2nd alternative&lst alternative
-&2nd alternative\\ \hline
- K&$(2e)+(2f)$&$(4i)$ &$(2c)+(2d)$&$(4f)$ \\
- Mn&$(2g)$\footnote{The $z$ parameter of these positions is $z\sim\frac{1}{4}$.}
- &$(a)+(b)+(c)+(d)$&$(4e)$&$(2a)+(2b)$\\
- Cl&$(a)+(b)+(c)+(d)$&$(2g)$\footnote{This is a footnote in a table that spans the full page
-width in \texttt{twocolumn} mode. It is supposed to set on the full width of the page, just as the caption does. }
- &$(4e)^{\text{a}}$\\
- He&$(8r)^{\text{a}}$&$(4j)^{\text{a}}$&$(4g)^{\text{a}}$\\
- Ag& &$(4k)^{\text{a}}$& &$(4h)^{\text{a}}$\\
-\end{tabular}
-\end{ruledtabular}
-\end{table*}
-
-The tables in this document illustrate various effects.
-Tables that fit in a narrow column are contained in a \texttt{table}
-environment.
-Table~\ref{tab:table3} is a wide table, therefore set with the
-\texttt{table*} environment.
-Lengthy tables may need to break across pages.
-A simple way to allow this is to specify
-the \verb+[H]+ float placement on the \texttt{table} or
-\texttt{table*} environment.
-Alternatively, using the standard \LaTeXe\ package \texttt{longtable} 
-gives more control over how tables break and allows headers and footers 
-to be specified for each page of the table.
-An example of the use of \texttt{longtable} can be found
-in the file \texttt{summary.tex} that is included with the REV\TeX~4
-distribution.
-
-There are two methods for setting footnotes within a table (these
-footnotes will be displayed directly below the table rather than at
-the bottom of the page or in the bibliography).
-The easiest
-and preferred method is just to use the \verb+\footnote{#1}+
-command. This will automatically enumerate the footnotes with
-lowercase roman letters.
-However, it is sometimes necessary to have
-multiple entries in the table share the same footnote.
-In this case,
-create the footnotes using
-\verb+\footnotemark[#1]+ and \verb+\footnotetext[#1]{#2}+.
-\texttt{\#1} is a numeric value.
-Each time the same value for \texttt{\#1} is used, 
-the same mark is produced in the table. 
-The \verb+\footnotetext[#1]{#2}+ commands are placed after the \texttt{tabular}
-environment. 
-Examine the \LaTeX\ source and output for Tables~\ref{tab:table1} and 
-\ref{tab:table2}%
-\begin{table}
-\caption{\label{tab:table2}A table with more columns still fits
-properly in a column. Note that several entries share the same
-footnote. Inspect the \LaTeX\ input for this table to see
-exactly how it is done.}
-\begin{ruledtabular}
-\begin{tabular}{cccccccc}
- &$r_c$ (\AA)&$r_0$ (\AA)&$\kappa r_0$&
- &$r_c$ (\AA) &$r_0$ (\AA)&$\kappa r_0$\\
-\hline
-Cu& 0.800 & 14.10 & 2.550 &Sn\footnotemark[1]
-& 0.680 & 1.870 & 3.700 \\
-Ag& 0.990 & 15.90 & 2.710 &Pb\footnotemark[2]
-& 0.450 & 1.930 & 3.760 \\
-Au& 1.150 & 15.90 & 2.710 &Ca\footnotemark[3]
-& 0.750 & 2.170 & 3.560 \\
-Mg& 0.490 & 17.60 & 3.200 &Sr\footnotemark[4]
-& 0.900 & 2.370 & 3.720 \\
-Zn& 0.300 & 15.20 & 2.970 &Li\footnotemark[2]
-& 0.380 & 1.730 & 2.830 \\
-Cd& 0.530 & 17.10 & 3.160 &Na\footnotemark[5]
-& 0.760 & 2.110 & 3.120 \\
-Hg& 0.550 & 17.80 & 3.220 &K\footnotemark[5]
-&  1.120 & 2.620 & 3.480 \\
-Al& 0.230 & 15.80 & 3.240 &Rb\footnotemark[3]
-& 1.330 & 2.800 & 3.590 \\
-Ga& 0.310 & 16.70 & 3.330 &Cs\footnotemark[4]
-& 1.420 & 3.030 & 3.740 \\
-In& 0.460 & 18.40 & 3.500 &Ba\footnotemark[5]
-& 0.960 & 2.460 & 3.780 \\
-Tl& 0.480 & 18.90 & 3.550 & & & & \\
-\end{tabular}
-\end{ruledtabular}
-\footnotetext[1]{Here's the first, from Ref.~\onlinecite{feyn54}.}
-\footnotetext[2]{Here's the second.}
-\footnotetext[3]{Here's the third.}
-\footnotetext[4]{Here's the fourth.}
-\footnotetext[5]{And etc.}
-\end{table}
-for an illustration.
-
-All AIP journals require that the initial citation of
-figures or tables be in numerical order.
-\LaTeX's automatic numbering of floats is your friend here:
-just put each \texttt{figure} environment immediately following 
-its first reference (\verb+\ref+), as we have done in this example file.
-
-\begin{acknowledgments}
-We wish to acknowledge the support of the author community in using
-REV\TeX{}, offering suggestions and encouragement, testing new versions,
-\dots.
-\end{acknowledgments}
-
-\section*{Data Availability Statement}
-
-AIP Publishing believes that all datasets underlying the conclusions of the paper should be available to readers.
-Authors are encouraged to deposit their datasets in publicly available repositories or present them in the main manuscript.
-All research articles must include a data availability statement stating where the data can be found.
-In this section, authors should add the respective statement from the chart below based on the availability of data in their paper.
-
-\begin{center}
-\renewcommand\arraystretch{1.2}
-\begin{tabular}{| >{\raggedright\arraybackslash}p{0.3\linewidth} | >{\raggedright\arraybackslash}p{0.65\linewidth} |}
-\hline
-\textbf{AVAILABILITY OF DATA} & \textbf{STATEMENT OF DATA AVAILABILITY}\\  
-\hline
-Data available on request from the authors
-&
-The data that support the findings of this study are available from the corresponding author upon reasonable request.
-\\\hline
-Data available in article or supplementary material
-&
-The data that support the findings of this study are available within the article [and its supplementary material].
-\\\hline
-Data openly available in a public repository that issues datasets with DOIs
-&
-The data that support the findings of this study are openly available in [repository name] at http://doi.org/[doi], reference number [reference number].
-\\\hline
-Data openly available in a public repository that does not issue DOIs
-&
-The data that support the findings of this study are openly available in [repository name], reference number [reference number].
-\\\hline
-Data sharing not applicable – no new data generated
-&
-Data sharing is not applicable to this article as no new data were created or analyzed in this study.
-\\\hline
-Data generated at a central, large scale facility
-&
-Raw data were generated at the [facility name] large scale facility. Derived data supporting the findings of this study are available from the corresponding author upon reasonable request.
-\\\hline
-Embargo on data due to commercial restrictions
-&
-The data that support the findings will be available in [repository name] at [DOI link] following an embargo from the date of publication to allow for commercialization of research findings.
-\\\hline
-Data available on request due to privacy/ethical restrictions
-&
-The data that support the findings of this study are available on request from the corresponding author.
-The data are not publicly available due [state restrictions such as privacy or ethical restrictions].
-\\\hline
-Data subject to third party restrictions
-&
-The data that support the findings of this study are available from [third party].
-Restrictions apply to the availability of these data, which were used under license for this study.
-Data are available from the authors upon reasonable request and with the permission of [third party].
-\\\hline
-\end{tabular}
-\end{center}
-
-\appendix
-
-\section{Appendixes}
-
-To start the appendixes, use the \verb+\appendix+ command.
-This signals that all following section commands refer to appendixes
-instead of regular sections. Therefore, the \verb+\appendix+ command
-should be used only once---to set up the section commands to act as
-appendixes. Thereafter normal section commands are used. The heading
-for a section can be left empty. For example,
-\begin{verbatim}
-\appendix
-\section{}
-\end{verbatim}
-will produce an appendix heading that says ``APPENDIX A'' and
-\begin{verbatim}
-\appendix
-\section{Background}
-\end{verbatim}
-will produce an appendix heading that says ``APPENDIX A: BACKGROUND''
-(note that the colon is set automatically).
-
-If there is only one appendix, then the letter ``A'' should not
-appear. This is suppressed by using the star version of the appendix
-command (\verb+\appendix*+ in the place of \verb+\appendix+).
-
-\section{A little more on appendixes}
-
-Observe that this appendix was started by using
-\begin{verbatim}
-\section{A little more on appendixes}
-\end{verbatim}
-
-Note the equation number in an appendix:
-\begin{equation}
-E=mc^2.
-\end{equation}
-
-\subsection{\label{app:subsec}A subsection in an appendix}
-
-You can use a subsection or subsubsection in an appendix. Note the
-numbering: we are now in Appendix~\ref{app:subsec}.
-
-\subsubsection{\label{app:subsubsec}A subsubsection in an appendix}
-Note the equation numbers in this appendix, produced with the
-subequations environment:
-\begin{subequations}
-\begin{eqnarray}
-E&=&mc, \label{appa}
-\\
-E&=&mc^2, \label{appb}
-\\
-E&\agt& mc^3. \label{appc}
-\end{eqnarray}
-\end{subequations}
-They turn out to be Eqs.~(\ref{appa}), (\ref{appb}), and (\ref{appc}).
+% BIBLOGRAPHY ==============================================================================================================
 
@@
 \end{document}
-%
+% END MAIN =================================================================================================================
 % ****** End of file aipsamp.tex ******

### 2022-10-20 publication/publication.tex
#37: Picture insert und word count

@@
 	\begin{figure}[h]
-		\includegraphics[width=\linewidth]{#1}
+		\includegraphics[width=0.9\linewidth]{#1}
 		\caption{#2}
@@
 
-%\title[Convergence of shearingrate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize]{
-%	Convergence of shearingrate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize
+%\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize]{
+%	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize
 %}
@@
 
+\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
+
+\includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{Test}{fig:eflux}
+
+\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.2_boxsize2x2_Ns16_Nvpar48-64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
+
 \includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize2x1-2-3x1-3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{Test}{fig:test}
+
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{Test}{fig:test}
+
+%\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{Test}{fig:test}
+
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison.pdf}{Test}{fig:test}
+
+\increasecounter{fig}{2}
 
@@
 
-% BIBLOGRAPHY ==============================================================================================================
+% BIBLIOGRAPHY ==============================================================================================================
 

### 2022-11-05 bachelorthesis/01-introduction.tex
#39:  New structure of files and references

% 1. Introduction

\chapter{Introduction}
\label{chap:intro}

### 2022-11-05 bachelorthesis/02-theory.tex
#39:  New structure of files and references

% 4. Theory

\chapter{Theory}
\label{chap:theory}

% Text

% etc.

### 2022-11-05 bachelorthesis/03-methods.tex
#39:  New structure of files and references

% 3. Methods and MAterial

\chapter{Methods and Material}
\label{chap:methods}

### 2022-11-05 bachelorthesis/04-results.tex
#39:  New structure of files and references

% 4. Results

\chapter{Results}
\label{chap:results}

% Text

% etc.

### 2022-11-10 briefcommunicaton/briefcommunication.tex
#37: Renamed folder and main file
#37: wordcount with remaining word counter

--- 2022-10-20 publication/publication.tex
@@
 \usepackage{caption}
-%\usepackage{hyperref}
+\usepackage{hyperref}
 
@@
 
-\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
+%\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
 
@@
 
-\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.2_boxsize2x2_Ns16_Nvpar48-64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
+\includegraphicsOneCol{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{Test}{fig:eflux}
+
+%\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.2_boxsize2x2_Ns16_Nvpar48-64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
 
@@
 \nocite{*}
-\bibliography{aipsamp}% Produces the bibliography via BibTeX.
+\bibliography{references.bib}% Produces the bibliography via BibTeX.
 

### 2022-11-14 briefcommunicaton/briefcommunication.tex
#37: Introduction (1)

@@
 
-\section{\label{sec:level1}Introduction}
+%%% INTRODUCTION ===========================================================================================================
+
+It is known that radially sheared zonal flows plays a significant role in nonlinear stabilization in tokamak plasmas. \cite{WACooper1988,doi:10.1063/1.859529,doi:10.1063/1.873896}. 
+Through advection on the sheared zonal flows the turbulent structure in plasma gets deformed and tilted that causes an $E \times B$ nonlinearty.
+\cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}
+Zonal flows mediate spectral energy transfer to larger radial wave vectors. \cite{doi:10.1063/1.3033206, doi:10.1063/1.3675855, PhysRevLett.120.175002}
+The strength of the shearing process is the $E \times B$ shearing rate $\omega_{\mathrm{E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
+\cite{doi:10.1063/1.871313, doi:10.1063/1.872847}
+The shearing rate $\omega_{\mathrm{E \times B}}$ is defined as 
+\begin{equation}
+	\omega_{\mathrm{E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
+	\label{eq:shearingrate}
+\end{equation}
+where $\langle \Phi \rangle$ is the zonal electrostatic potential and $\psi$ the radial coordinate that labels the flux surfaces.
+\cite{doi:10.1063/1.4952621, doi:10.1063/1.5035184, doi:10.1063/1.3005380}
+It was shown that the nonlinear threshold for turbulence is directly related to shear stabilization. \cite{doi:10.1063/1.873896} 
+Often the shear stabilization is expressed in the empirical Waltz rule $\omega_{\mathrm{E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
+In the discovered zonal flows, also known as $E \times B$ staircase \cite{PhysRevE.82.025401}, exhibit amplitudes in terms of the $E \times B$ shearing rate satisfying the stabilization criteria.
+\cite{doi:10.1063/1.4952621, doi:10.1063/1.4961231}
 
@@
 
-\ref{fig:test}
-
 \section*{Data Availability}
@@
 
-% BIBLIOGRAPHY ==============================================================================================================
+% BIBLIOGRAPHY =============================================================================================================
 
-\nocite{*}
+%\nocite{*}
 \bibliography{references.bib}% Produces the bibliography via BibTeX.

### 2022-11-22 briefcommunicaton/briefcommunication.tex
#37: Title, Github Repository etc.

@@
 
-%\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize]{
-%	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize
-%}
+\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation]{
+	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation
+}
 
-%\author{M. Lippert}
-%	\altaffiliation{GitHub Repository for files reagarding this paper: \\ 
-%					https://github.com/ManeLippert/Bachelorthesis-ZonalFlows}
-%\author{F. Rath}
-%	\altaffiliation{Author to whom correspondence should be addressed: \\ 
-%					Florian.Rath1@uni-bayreuth.de}
-%	\email{Florian.Rath1@uni-bayreuth.de}
-%\author{A. G. Peeters}
-%\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
+\author{M. Lippert}
+	\altaffiliation{GitHub Repository of this work: \\ 
+					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
+\author{F. Rath}
+	\altaffiliation{Author to whom correspondence should be addressed: \\ 
+					Florian.Rath1@uni-bayreuth.de}
+	\email{Florian.Rath1@uni-bayreuth.de}
+\author{A. G. Peeters}
+\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
 
-%\date{\today}
+\date{\today}
 
@@
 
-%\begin{abstract}
-%	An article usually includes an abstract, a concise summary of the work
-%	covered at length in the main body of the article. It is used for
-%	secondary publications and for information retrieval purposes. 
-%\end{abstract}
+\begin{abstract}
+	An article usually includes an abstract, a concise summary of the work
+	covered at length in the main body of the article. It is used for
+	secondary publications and for information retrieval purposes. 
+\end{abstract}
 
-%\maketitle
+\maketitle
 

### 2022-11-23 briefcommunicaton/briefcommunication.tex
#37: Fixed Plots, Started writing

--- 2022-11-14 briefcommunicaton/briefcommunication.tex
@@
 %% Include graphic for two column with specific place, figrue* places graphic anywhere...
+%\newcommand{\includegraphicsTwoCol}[3]{
+%	\onecolumngrid
+%	\begin{center}
+%		\captionsetup{type=figure}
+%    	\includegraphics[width=\textwidth]{#1}
+%		\captionof{figure}{#2}
+%	\end{center}
+%	\twocolumngrid
+%	\label{#3}
+%	\increasecounter{fig}{2}
+%}
+
 \newcommand{\includegraphicsTwoCol}[3]{
-	\onecolumngrid
-	\begin{center}
-		\captionsetup{type=figure}
+	\begin{figure*}
     	\includegraphics[width=\textwidth]{#1}
-		\captionof{figure}{#2}
-	\end{center}
-	\twocolumngrid
+		\caption{#2}
+	\end{figure*}
 	\label{#3}
@@
 
-%\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize]{
-%	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize
+%\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation]{
+%	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation
 %}
@@
 %\author{M. Lippert}
-%	\altaffiliation{GitHub Repository for files reagarding this paper: \\ 
-%					https://github.com/ManeLippert/Bachelorthesis-ZonalFlows}
+%	\altaffiliation{Repository of this work: \\ 
+%					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
 %\author{F. Rath}
@@
 
+\section{Introduction}
+\label{sec:intro}
+
+This brief communication focused on continuing the work Rath et. al \cite{doi:10.1063/1.4961231}. At the end Section IV it was 
+discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper. To gain more
+knowledge if the staircase can develop with the chosen box size from Ref 1.
+This paper will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.
+
+%%% THEORY =================================================================================================================
+
+\section{Theory}
+\label{sec:theory}
+
 It is known that radially sheared zonal flows plays a significant role in nonlinear stabilization in tokamak plasmas. \cite{WACooper1988,doi:10.1063/1.859529,doi:10.1063/1.873896}. 
-Through advection on the sheared zonal flows the turbulent structure in plasma gets deformed and tilted that causes an $E \times B$ nonlinearty.
+Through advection on the sheared zonal flows the turbulent structure in plasma gets deformed and tilted that causes an E$\times$B nonlinearty.
 \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}
 Zonal flows mediate spectral energy transfer to larger radial wave vectors. \cite{doi:10.1063/1.3033206, doi:10.1063/1.3675855, PhysRevLett.120.175002}
-The strength of the shearing process is the $E \times B$ shearing rate $\omega_{\mathrm{E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
+The strength of the shearing process is the E$\times$B shearing rate $\omega_{\mathrm{E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
 \cite{doi:10.1063/1.871313, doi:10.1063/1.872847}
@@
 \begin{equation}
-	\omega_{\mathrm{E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
+	\omega_{\mathrm{E \times B}} = \frac{1}{2B} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
 	\label{eq:shearingrate}
 \end{equation}
-where $\langle \Phi \rangle$ is the zonal electrostatic potential and $\psi$ the radial coordinate that labels the flux surfaces.
-\cite{doi:10.1063/1.4952621, doi:10.1063/1.5035184, doi:10.1063/1.3005380}
+where $\langle \Phi \rangle$ is the zonal electrostatic potential and $\psi$ the radial coordinate that labels the flux surfaces. \cite{doi:10.1063/1.4952621, doi:10.1063/1.3005380}
 It was shown that the nonlinear threshold for turbulence is directly related to shear stabilization. \cite{doi:10.1063/1.873896} 
 Often the shear stabilization is expressed in the empirical Waltz rule $\omega_{\mathrm{E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
-In the discovered zonal flows, also known as $E \times B$ staircase \cite{PhysRevE.82.025401}, exhibit amplitudes in terms of the $E \times B$ shearing rate satisfying the stabilization criteria.
+In the discovered zonal flows, also known as E$\times$B staircase \cite{PhysRevE.82.025401}, exhibit amplitudes in terms of the E$\times$B shearing rate satisfying the stabilization criteria.
 \cite{doi:10.1063/1.4952621, doi:10.1063/1.4961231}
 
-\includegraphicsOneCol{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
+\includegraphicsOneCol{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial and binormal increased boxsizes
+}{fig:eflux-1x1-2x2-3x3-comparison}
 
-%\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
+\includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
+	Shearing rate $\omega_{\mathrm{ExB}}$ for different time intervals in which heat conduction is almost zero but staircase has not fully developed for boxsize 3$\times$1
+}{fig:wexb-3x1-selection}
 
-\includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{Test}{fig:eflux}
+\includegraphicsOneCol{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
+	Stablized shearing rate $\omega_{\mathrm{ExB}}$ for boxsize 3$\times$3
+}{fig:wexb-3x3-stable}
 
-\includegraphicsOneCol{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{Test}{fig:eflux}
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize2x1-2-3x1-3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
+	Comparison of time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for boxsize 2$\times$1 compared to 2$\times$2 and 3$\times$1 compared to 3$\times$3
+}{fig:eflux-2x1-2-3x1-3-comparison}
 
-%\includegraphicsOneCol{Comparison/Gradient-Length/S6_rlt6.0-6.2_boxsize2x2_Ns16_Nvpar48-64_Nmu9_eflux_comparison.pdf}{Test}{fig:eflux}
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{
+	Comparison of shearing rate $\omega_{\mathrm{ExB}}$ for radial increased boxsizes. The staircase structure got 
+	shifted for better visibility.
+}{fig:wexb-1-2-3-4x1-stable-comparison}
 
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize2x1-2-3x1-3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{Test}{fig:test}
-
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{Test}{fig:test}
-
-%\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{Test}{fig:test}
-
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison.pdf}{Test}{fig:test}
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased boxsizes \linebreak
+	\textbf{(b)} Time traces of $\omega_{\mathrm{ExB}}^{\mathrm{max}}$ for radial increased boxsizes
+}{fig:wexb-eflux-1-2-3-4x1-comparison}
 

### 2022-11-24 briefcommunicaton/briefcommunication.tex
#37: Fixed plots spacing and wexb_max y label
#37: Introduction, Theory, Setup

@@
 \newcommand{\includegraphicsOneCol}[3]{
-	\begin{figure}[h]
+	\begin{figure}[ht]
 		\includegraphics[width=0.9\linewidth]{#1}
@@
 
-This brief communication focused on continuing the work Rath et. al \cite{doi:10.1063/1.4961231}. At the end Section IV it was 
-discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper. To gain more
-knowledge if the staircase can develop with the chosen box size from Ref 1.
-This paper will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.
+This brief communication focused on continuing the work of Rath et. al \cite{doi:10.1063/1.4961231}. 
+The paper does mainly concentrate on gradient driven flux-tub simulations close to the non-linear threshold and the occurrence of the E$\times$B staircase structure and its formation over the simulation.
+At the end Section IV it was discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper.
+To gain more knowledge if the staircase structure can fully develop the following paper will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.
 
@@
 
-\section{Theory}
-\label{sec:theory}
+%\section{Theory}
+%\label{sec:theory}
 
@@
 \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}
-Zonal flows mediate spectral energy transfer to larger radial wave vectors. \cite{doi:10.1063/1.3033206, doi:10.1063/1.3675855, PhysRevLett.120.175002}
 The strength of the shearing process is the E$\times$B shearing rate $\omega_{\mathrm{E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
@@
 Often the shear stabilization is expressed in the empirical Waltz rule $\omega_{\mathrm{E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
-In the discovered zonal flows, also known as E$\times$B staircase \cite{PhysRevE.82.025401}, exhibit amplitudes in terms of the E$\times$B shearing rate satisfying the stabilization criteria.
-\cite{doi:10.1063/1.4952621, doi:10.1063/1.4961231}
+In the discovered zonal flows, also known as E$\times$B staircase \cite{PhysRevE.82.025401}, exhibit amplitudes in terms of the E$\times$B shearing rate satisfying the stabilization criteria and for a fully developed staircase structure the E $\times$ B shearing rate $\omega_{\mathrm{E \times B}} = \gamma$.
+\cite{doi:10.1063/1.4961231, doi:10.1063/1.4952621}
+
+\section{Simulation Setup}
+\label{sec:setup}
+
+The plasma parameters are nearly the same as in Ref \onlinecite{doi:10.1063/1.4961231} with the cyclone base case: safety factor $q = 1.4$, magnetic shear $\hat{s} = 0.78$, inverse aspect ratio $\epsilon = 0.19$, density gradient $R/L_n = 2.2$ and electron to ion temperature ratio $T_e/T_i = 1$.
+The maximum of the velocity grid is three times the thermal velocity for both the parallel and the perpendicular velocities and adiabatic electrons with neglected collisions were investigated.
+
+For this paper the resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table \ref{tab:resolution} with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduces the runtime of simulations.
+Simulations showed that this correction in grid points does not affect the simulation itself.
+
+The simulations are performed with the flux tube version of the non-linear gyro-kinetic code GKW.\cite{Peeters20092650}
+Futher informations regarding the simulation can be found in Ref \onlinecite{doi:10.1063/1.4961231}.
+
+\begin{center}
+	\captionsetup{type=table}
+	\begin{ruledtabular}
+		\begin{tabular}{l | ccccc | ccccc | c | cc}
+			& $N_m$ & $N_x$ & $N_s$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_d$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
+			\hline
+			S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
+		\end{tabular}
+	\end{ruledtabular}
+	\captionof{table}{
+		Resolution used in this paper (for more informations about the parameter read Ref \onlinecite{doi:10.1063/1.4961231}) %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient oftoroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+	}
+	\label{tab:resolution}
+\end{center}
+
+\section{Results}
+\label{sec:results}
 
@@
 \includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Shearing rate $\omega_{\mathrm{ExB}}$ for different time intervals in which heat conduction is almost zero but staircase has not fully developed for boxsize 3$\times$1
+	Shearing rate $\omega_{\mathrm{E \times B}}$ for different time intervals in which heat conduction is almost zero but staircase has not fully developed for boxsize 3$\times$1
 }{fig:wexb-3x1-selection}
@@
 \includegraphicsOneCol{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Stablized shearing rate $\omega_{\mathrm{ExB}}$ for boxsize 3$\times$3
+	Stablized shearing rate $\omega_{\mathrm{E \times B}}$ for boxsize 3$\times$3
 }{fig:wexb-3x3-stable}
@@
 \includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{
-	Comparison of shearing rate $\omega_{\mathrm{ExB}}$ for radial increased boxsizes. The staircase structure got 
+	Comparison of shearing rate $\omega_{\mathrm{E \times B}}$ for radial increased boxsizes. The staircase structure got 
 	shifted for better visibility.
@@
 	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased boxsizes \linebreak
-	\textbf{(b)} Time traces of $\omega_{\mathrm{ExB}}^{\mathrm{max}}$ for radial increased boxsizes
+	\textbf{(b)} Time traces of $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}}$ for radial increased boxsizes
 }{fig:wexb-eflux-1-2-3-4x1-comparison}
@@
 \increasecounter{fig}{2}
+
+\section{Conclusion}
+\label{sec:conclusion}
 

### 2022-11-25 briefcommunicaton/briefcommunication.tex
#24: Some problems with pending status

@@
 The maximum of the velocity grid is three times the thermal velocity for both the parallel and the perpendicular velocities and adiabatic electrons with neglected collisions were investigated.
-
 For this paper the resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table \ref{tab:resolution} with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduces the runtime of simulations.
@@
 	\captionof{table}{
-		Resolution used in this paper (for more informations about the parameter read Ref \onlinecite{doi:10.1063/1.4961231}) %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient oftoroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+		Resolution used in this paper (for more informations about the parameter read Ref \onlinecite{doi:10.1063/1.4961231}) %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
 	}

### 2022-12-22 briefcommunicaton/briefcommunication.tex
#37: First Correction

--- 2022-11-23 briefcommunicaton/briefcommunication.tex
@@
 %% Additional
-\usepackage{caption}
+%\usepackage{caption}
 \usepackage{hyperref}
@@
 
-\section{Introduction}
-\label{sec:intro}
+%\section{Introduction}
+%\label{sec:intro}
 
-This brief communication focused on continuing the work Rath et. al \cite{doi:10.1063/1.4961231}. At the end Section IV it was 
-discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper. To gain more
-knowledge if the staircase can develop with the chosen box size from Ref 1.
-This paper will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.
+This brief communication focuses on continuing the work of Rath et. al \cite{doi:10.1063/1.4961231}. 
+Their paper elaborates on gradient driven flux-tub simulations close to the non-linear threshold and the occurrence of the E$\times$B staircase structure and its formation over the simulation.
+At the end of Section IV it was discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper.
+To gain further insights on whether the staircase structure can fully develop the following brief communication will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.\bigskip
 
@@
 
-\section{Theory}
-\label{sec:theory}
+%\section{Theory}
+%\label{sec:theory}
 
-It is known that radially sheared zonal flows plays a significant role in nonlinear stabilization in tokamak plasmas. \cite{WACooper1988,doi:10.1063/1.859529,doi:10.1063/1.873896}. 
-Through advection on the sheared zonal flows the turbulent structure in plasma gets deformed and tilted that causes an E$\times$B nonlinearty.
+It is known that radially sheared zonal flows play a significant role in nonlinear stabilization in tokamak plasmas. \cite{WACooper1988,doi:10.1063/1.859529,doi:10.1063/1.873896}. 
+Through advection on the sheared zonal flows the turbulent structure in plasma gets deformed and tilted, which causes an E$\times$B nonlinearity.
 \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}
-Zonal flows mediate spectral energy transfer to larger radial wave vectors. \cite{doi:10.1063/1.3033206, doi:10.1063/1.3675855, PhysRevLett.120.175002}
 The strength of the shearing process is the E$\times$B shearing rate $\omega_{\mathrm{E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
@@
 \begin{equation}
-	\omega_{\mathrm{E \times B}} = \frac{1}{2B} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
+	\omega_{\mathrm{E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
 	\label{eq:shearingrate}
@@
 It was shown that the nonlinear threshold for turbulence is directly related to shear stabilization. \cite{doi:10.1063/1.873896} 
-Often the shear stabilization is expressed in the empirical Waltz rule $\omega_{\mathrm{E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
-In the discovered zonal flows, also known as E$\times$B staircase \cite{PhysRevE.82.025401}, exhibit amplitudes in terms of the E$\times$B shearing rate satisfying the stabilization criteria.
-\cite{doi:10.1063/1.4952621, doi:10.1063/1.4961231}
+The shear stabilization is often expressed in the empirical Waltz rule $\omega_{\mathrm{E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
+The discovered zonal flows, also known as E$\times$B staircase \cite{PhysRevE.82.025401}, exhibit amplitudes, which satisfy the stabilization criteria in terms of the E$\times$B shearing rate.
+For a fully developed staircase structure the E $\times$ B shearing rate $\omega_{\mathrm{E \times B}} = \gamma$.
+\cite{doi:10.1063/1.4961231, doi:10.1063/1.4952621}\bigskip
+
+%%% METHODS ==================================================================================================================
+
+%\section{Simulation Setup}
+%\label{sec:setup}
+
+The plasma parameters are closely modelled after those in Rath et al.
+\cite{doi:10.1063/1.4961231} with the cyclone base case: safety factor $q = 1.4$, magnetic shear $\hat{s} = 0.78$, inverse aspect ratio $\epsilon = 0.19$, density gradient $R/L_n = 2.2$ and electron to ion temperature ratio $T_e/T_i = 1$.
+The maximum of the velocity grid is three times the thermal velocity for both the parallel and the perpendicular velocities.
+Adiabatic electrons were investigated, while neglecting collisions.
+
+For this paper the resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduces the runtime of simulations.
+Simulations showed that this correction in grid points does not affect the results itself.
+
+The simulations are performed with the flux tube version of the non-linear gyro-kinetic code GKW \cite{Peeters20092650} with periodic boundary conditions.
+Further information regarding the simulation can be found in Ref \onlinecite{doi:10.1063/1.4961231}.
+
+\begin{table}
+	\begin{ruledtabular}
+		\begin{tabular}{l | ccccc | ccccc | c | cc}
+			& $N_m$ & $N_x$ & $N_s$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_d$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
+			\hline
+			S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
+		\end{tabular}
+	\end{ruledtabular}
+	\caption{
+		Resolution used in this paper for further information read Rath et al.
+\cite{doi:10.1063/1.4961231} %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+	}
+	\label{tab:resolution}
+\end{table} \bigskip
+
+%%% RESULTS ================================================================================================================
+
+%\section{Results}
+%\label{sec:results}
+
+The data is normalized with the heat conduction coefficient $\chi$ in gyro-Bohm units ($\rho^2\nu_{\mathrm{th}}/R)$, where $\rho = m_\mathrm{i}\nu_{\mathrm{th}}/eB$ is the ion Larmor radius, $\nu_{\mathrm{th}} = \sqrt{2T/m_\mathrm{i}}$ is the thermal velocity, $T$ the background temperature, e is the unit charge and $R$ is the major radius.
 
@@
 \includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Shearing rate $\omega_{\mathrm{ExB}}$ for different time intervals in which heat conduction is almost zero but staircase has not fully developed for boxsize 3$\times$1
+	Shearing rate $\omega_{\mathrm{E \times B}}$ for different time intervals in which heat conduction is almost zero but the E$\times$B staircase has not fully developed for boxsize 3$\times$1
 }{fig:wexb-3x1-selection}
@@
 \increasecounter{fig}{2}
+
+%%% CONCLUSION =============================================================================================================
+
+%\section{Conclusion}
+%\label{sec:conclusion}
 
@@
 
-\wordcountmessage
+%\wordcountmessage
 
@@
 
-%\nocite{*}
+\nocite{}
 \bibliography{references.bib}% Produces the bibliography via BibTeX.

### 2022-12-29 briefcommunicaton/briefcommunication.tex
#37: First complete draft

@@
 %% Additional
-%\usepackage{caption}
+\usepackage{caption}
 \usepackage{hyperref}
@@
 	\begin{figure}[ht]
-		\includegraphics[width=0.9\linewidth]{#1}
+		\includegraphics[width=0.90\linewidth]{#1}
 		\caption{#2}
+		\label{#3}
   	\end{figure}
-	\label{#3}
   	\increasecounter{fig}{1}
@@
 %% Include graphic for two column with specific place, figrue* places graphic anywhere...
+\newcommand{\includegraphicsTwoCol}[4]{
+	\onecolumngrid
+	\begin{center}
+		\captionsetup{type=figure}
+    	\includegraphics[width=#4\textwidth]{#1}
+		\captionof{figure}{#2}
+    	\label{#3}
+	\end{center}
+	\twocolumngrid
+	\increasecounter{fig}{2}
+}
+
 %\newcommand{\includegraphicsTwoCol}[3]{
-%	\onecolumngrid
-%	\begin{center}
-%		\captionsetup{type=figure}
-%    	\includegraphics[width=\textwidth]{#1}
-%		\captionof{figure}{#2}
-%	\end{center}
-%	\twocolumngrid
+%	\begin{figure*}
+%    	\includegraphics[width=0.9\textwidth]{#1}
+%		\caption{#2}
 %		\label{#3}
+%	\end{figure*}
 %	\increasecounter{fig}{2}
@@
 
-\newcommand{\includegraphicsTwoCol}[3]{
-	\begin{figure*}
-    	\includegraphics[width=\textwidth]{#1}
-		\caption{#2}
-	\end{figure*}
-	\label{#3}
-	\increasecounter{fig}{2}
+\newcommand{\words}{
+	\immediate\write18{texcount -sum -1 \jobname.tex > wordcount.txt}
 }
@@
 
-%\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation]{
-%	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation
-%}
+\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with box size in gradient driven simulation]{
+	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with box size in gradient driven simulation
+}
 
-%\author{M. Lippert}
-%	\altaffiliation{Repository of this work: \\ 
-%					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
-%\author{F. Rath}
-%	\altaffiliation{Author to whom correspondence should be addressed: \\ 
-%					Florian.Rath1@uni-bayreuth.de}
-%	\email{Florian.Rath1@uni-bayreuth.de}
-%\author{A. G. Peeters}
-%\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
+\author{M. Lippert}
+	\altaffiliation{Repository of this work: \\ 
+					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
+\author{F. Rath}
+	\altaffiliation{Author to whom correspondence should be addressed: \\
+					Florian.Rath1@uni-bayreuth.de}
+	\email{Florian.Rath1@uni-bayreuth.de}
+\author{A. G. Peeters}
+\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
 
-%\date{\today}
+\date{\today}
 
@@
 
-%\begin{abstract}
-%	An article usually includes an abstract, a concise summary of the work
-%	covered at length in the main body of the article. It is used for
-%	secondary publications and for information retrieval purposes. 
-%\end{abstract}
+\begin{abstract}
+	The gradient driven simulation of Rath et al.\cite{doi:10.1063/1.4961231} gets revisited.
+Its gets confirmed that the wavelength of the longest wavelength zonal mode converges with increased box sizes in radial and binormal direction.
+Specially the E$\times$B staircase structure repeats itself multiple times in the given box size.
+\end{abstract}
 
-%\maketitle
+\maketitle
 
@@
 Their paper elaborates on gradient driven flux-tub simulations close to the non-linear threshold and the occurrence of the E$\times$B staircase structure and its formation over the simulation.
-At the end of Section IV it was discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper.
+At the end of Section IV it was discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper and if the longest wavelength changes with increasing the box size.
 To gain further insights on whether the staircase structure can fully develop the following brief communication will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.\bigskip
@@
 \end{equation}
+\increasecounter{eq}{1}{1}
 where $\langle \Phi \rangle$ is the zonal electrostatic potential and $\psi$ the radial coordinate that labels the flux surfaces. \cite{doi:10.1063/1.4952621, doi:10.1063/1.3005380}
@@
 
+This paper covers the increase of the box size in radial direction and in binormal direction with periodic boundary conditions for the box itself.
+Which box size was used can be seen in each graphic in form of $N_\mathrm{R}\times N_\mathrm{B}$, where $N_\mathrm{R}$ stands for the times the radial box size was increased and $N_\mathrm{B}$ for the binormal box size.
+To validate a stabilized turbulence and if the E$\times$B Staircase structure has fully developed a closer look at the long wavelength zonal flow mode $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}}$ is necessary.
+The long wavelength zonal flow mode is the absolute of the fourier transform of the shearing rate $\omega_{\mathrm{E \times B}}$ for each time step, where $k_\mathrm{i}$ is the wave vector connected to the long wavelength zonal flow mode with index i and is defined as following
+\begin{equation}
+	k_\mathrm{i} = \frac{2\pi \mathrm{i}}{l_\psi}
+\end{equation}
+\increasecounter{eq}{1}{1}
+with $l_\psi$ as the length of the box size.
+In addition, the maximum amplitude of the long wavelength zonal flow mode  $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}} = \gamma $.
+\cite{doi:10.1063/1.4961231} It is important that only for one wave vector $k_i$ the zonal mode reaches its maximum amplitude any other mode has to be close by zero.
+The mode with the maximum amplitude is called the longest wavelength zonal mode.
+\bigskip
+
 %%% METHODS ==================================================================================================================
@@
 
-The plasma parameters are closely modelled after those in Rath et al.
-\cite{doi:10.1063/1.4961231} with the cyclone base case: safety factor $q = 1.4$, magnetic shear $\hat{s} = 0.78$, inverse aspect ratio $\epsilon = 0.19$, density gradient $R/L_n = 2.2$ and electron to ion temperature ratio $T_e/T_i = 1$.
+The plasma parameters are closely modelled after those in Rath et al. \cite{doi:10.1063/1.4961231} with the cyclone base case.
+%safety factor $q = 1.4$, magnetic shear $\hat{s} = 0.78$, inverse aspect ratio $\epsilon = 0.19$, density gradient $R/L_n = 2.2$ and electron to ion temperature ratio $T_e/T_i = 1$.
 The maximum of the velocity grid is three times the thermal velocity for both the parallel and the perpendicular velocities.
 Adiabatic electrons were investigated, while neglecting collisions.
-
-For this paper the resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduces the runtime of simulations.
+The resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table \ref{tab:resolution} with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduces the runtime of simulations.
 Simulations showed that this correction in grid points does not affect the results itself.
-
-The simulations are performed with the flux tube version of the non-linear gyro-kinetic code GKW \cite{Peeters20092650} with periodic boundary conditions.
+Furthermore, the gradient length $R/L_\mathrm{T}$ was set to the value 6.0 nearby the finite heat flux threshold shown in Fig.
+4 in Rath et al \cite{doi:10.1063/1.4961231}.
+The simulations are performed with the flux tube version of the non-linear gyro-kinetic code GKW \cite{Peeters20092650} with periodic boundary conditions and circular geometry.
 Further information regarding the simulation can be found in Ref \onlinecite{doi:10.1063/1.4961231}.
 
-\begin{table}
+\begin{table}[ht]
 	\begin{ruledtabular}
@@
 	\label{tab:resolution}
-\end{table} \bigskip
+\end{table}
+
+The data is normalized with the heat conduction coefficient $\chi$ in gyro-Bohm units ($\rho^2\nu_{\mathrm{th}}/R)$, where $\rho = m_\mathrm{i}\nu_{\mathrm{th}}/eB$ is the ion Larmor radius, $\nu_{\mathrm{th}} = \sqrt{2T/m_\mathrm{i}}$ is the thermal velocity, $T$ the background temperature, e is the unit charge and $R$ is the major radius.
+\bigskip
 
@@
 
-The data is normalized with the heat conduction coefficient $\chi$ in gyro-Bohm units ($\rho^2\nu_{\mathrm{th}}/R)$, where $\rho = m_\mathrm{i}\nu_{\mathrm{th}}/eB$ is the ion Larmor radius, $\nu_{\mathrm{th}} = \sqrt{2T/m_\mathrm{i}}$ is the thermal velocity, $T$ the background temperature, e is the unit charge and $R$ is the major radius.
+At the beginning the box size was only increased in radial direction with values for $N_\mathrm{R} = 1, 2, 3, 4$.
+The simulation are showing that the longest wavelength zonal mode is connected with $N_\mathrm{R}$ in terms of index $\mathrm{i} = N_\mathrm{R}$.
+This behavior can be seen in Fig.
+\ref{fig:wexb-eflux-1-2-3-4x1-comparison} where the longest wavelength zonal mode for the box size 1$\times$1 has the wave vector $k_1$, for 2$\times$1 the wave vector $k_2$ and so on.
+Also, it is worth mentioning that after a long run of the simulation the turbulence can still stabilize, which was not certain because of the numerical dissipation of the simulation itself.
+\cite{doi:10.1063/1.4961231}
+
+Furthermore, it is clearly visible that the longest wavelength zonal mode reaches its maximum amplitude (Fig.
+\ref{fig:wexb-eflux-1-2-3-4x1-comparison} (b)) when the turbulence is stabilized  (Fig.
+\ref{fig:wexb-eflux-1-2-3-4x1-comparison} (a)) but further investigation yields the result that the simulations has not stabilized yet.
+For example for the box size 3$\times$1 is not stable although between [28000, 31000] normalized time units the turbulence subdued.
+This can be observed in Fig.
+\ref{fig:wexb-3x1-selection} where for [28000, 31000] normalized time units the staircase structure has a jump at ~40 normalized radial coordinates.
+In comparison, an unstable staircase structure has the form of a Sawtooth wave also visible in Fig.
+\ref{fig:wexb-3x1-selection} for [10000, 12000] normalized time units.
+This occurrence is due to the previous mentioned condition that only one zonal mode has a maximal amplitude and any other mode is closely by zero.
+So as additional verification process one could visualize the shearing rate $\omega_{\mathrm{E \times B}}$ to determine if the staircase structure has fully developed.
+
+When the turbulence is stabilized and the longest wavelength zonal mode has developed the wavelength of the shearing rate $\omega_{\mathrm{E \times B}}$ does converge with the box size.
+The E$\times$B staircase structure does repeat itself exactly $N_\mathrm{R}$ times.
+In Fig.
+\ref{fig:wexb-1-2-3-4x1-stable-comparison} is shown how the wavelength converges for different box sizes for different time intervals and the remarkable result of the repetition of the staircase structure.
+This can also be linked to the wave vector of the longest wavelength zonal mode for example for the box size 2$\times$1 the wave vector would be $k_2$ (Fig.
+\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b)) and the staircase structure repeats itself 2 times.
+It is also visibility that the Waltz rule does even hold for radial increased box sizes.
+Note that for clearer visibility the staircase structure for box size 1$\times$1 and for 4$\times$1 got shifted to the left so that the staircase structures do overlapped.
+This does not affect the physical result because of the applied periodic boundary conditions of the box.
+
+\includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
+	Shearing rate $\omega_{\mathrm{E \times B}}$ for different time intervals in which heat conduction is almost zero but the E$\times$B staircase has not fully developed for box size 3$\times$1
+}{fig:wexb-3x1-selection}
+
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased box sizes
+	\textbf{(b)} Time traces of $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}}$ for radial increased box sizes
+}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
+
+\increasecounter{fig}{2}
+
+\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{
+	Comparison of shearing rate $\omega_{\mathrm{E \times B}}$ for radial increased box sizes. The staircase structure got 
+	shifted for better visibility.
+}{fig:wexb-1-2-3-4x1-stable-comparison}{}
+
+As next step the box size got increased in radial and binormal direction the same time.
+For this simulation is it noticeable that the turbulence got stabilized and fully developed after 1500 normalized time unit seen in Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}.
+This leads to a much faster computation of zonal flow structures if the box size has to be increased.
+In comparison, the box size 3$\times$1 get stabilized and fully developed after 42000 normalized time units.
 
 \includegraphicsOneCol{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial and binormal increased boxsizes
+	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial and binormal increased box sizes
 }{fig:eflux-1x1-2x2-3x3-comparison}
 
-\includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Shearing rate $\omega_{\mathrm{E \times B}}$ for different time intervals in which heat conduction is almost zero but the E$\times$B staircase has not fully developed for boxsize 3$\times$1
-}{fig:wexb-3x1-selection}
+Even for this type of box sizes the wavelength converges with the box size and the Waltz rule is still satisfied.
+But it is to mentioned that for the 3$\times$3 box size the E$\times$B staircase structure repeats itself four times which can be seen in Fig.
+\ref{fig:wexb-3x3-stable} which stand in contradiction to the repetition of the radial increased box size.
+One could argue that the repetition of the staircase structure of a radial and binormal increased box size is greater equal $N_\mathrm{R}$.
+This could lead to the question if E$\times$B staircase structure has a different repetition for different box sizes, although this question is beyond of the scope of this paper.
 
@@
 
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize2x1-2-3x1-3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
-	Comparison of time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for boxsize 2$\times$1 compared to 2$\times$2 and 3$\times$1 compared to 3$\times$3
-}{fig:eflux-2x1-2-3x1-3-comparison}
-
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{
-	Comparison of shearing rate $\omega_{\mathrm{E \times B}}$ for radial increased boxsizes. The staircase structure got 
-	shifted for better visibility.
-}{fig:wexb-1-2-3-4x1-stable-comparison}
-
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased boxsizes \linebreak
-	\textbf{(b)} Time traces of $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}}$ for radial increased boxsizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}
-
-\increasecounter{fig}{2}
+%\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize2x1-2-3x1-3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
+%	Comparison of time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for box size 2$\times$1 compared to 2$\times$2 and 3$\times$1 compared to 3$\times$3
+%}{fig:eflux-2x1-2-3x1-3-comparison}
 
@@
 
+To conclude the wavelength does converge with the chosen box size and the Waltz rule holds in every simulation.
+Long runs can still result in stabilized turbulence.
+Further investigation of the long wavelength zonal mode has to be made to determine if the turbulence has been stabilized.
+For that it is important that only one long wavelength zonal mode has a maximal amplitude for given wave vector $k_i$ which is also the longest wavelength zonal mode.
+The E$\times$B staircase structure repeats itself multiple times in increased box sizes where for radial increased box size the repetition is linked to the times the box size was increased radial, although this does not hold for radial and binormal increased box sizes.
+\bigskip
+
+%% APPENDIX ================================================================================================================
+
 \section*{Data Availability}
 The data that support the findings of this study are available from the corresponding author upon reasonable request. 
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
 

### 2023-01-01 briefcommunicaton/briefcommunication.tex
#24: Corrected Graphic
#37: Correction CL, Dominik

@@
 
-\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with box size in gradient driven simulation]{
-	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with box size in gradient driven simulation
+\title[Convergence of shearing rate $\omega_{\mathrm{\:E}\times\mathrm{B}}$ with box size in gradient driven simulation]{
+	Convergence of shearing rate $\omega_{\mathrm{\:E}\times\mathrm{B}}$ with box size in gradient driven simulation
 }
@@
 	The gradient driven simulation of Rath et al.\cite{doi:10.1063/1.4961231} gets revisited.
-Its gets confirmed that the wavelength of the longest wavelength zonal mode converges with increased box sizes in radial and binormal direction.
-Specially the E$\times$B staircase structure repeats itself multiple times in the given box size.
+Its is confirmed that the wavelength of the longest wavelength zonal flow mode converges with increased box sizes in radial and binormal direction.
+Especially the E$\times$B staircase structure repeats itself multiple times in the given box size.
+%why is this intresting?
 \end{abstract}
@@
 
-This brief communication focuses on continuing the work of Rath et. al \cite{doi:10.1063/1.4961231}. 
+This brief communication focuses on continuing the work of Rath et al. \cite{doi:10.1063/1.4961231}. 
 Their paper elaborates on gradient driven flux-tub simulations close to the non-linear threshold and the occurrence of the E$\times$B staircase structure and its formation over the simulation.
-At the end of Section IV it was discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper and if the longest wavelength changes with increasing the box size.
-To gain further insights on whether the staircase structure can fully develop the following brief communication will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.\bigskip
+At the end section IV of the Rath et al.
+paper it was discussed that the circumstances for which the staircase can fully develop, like the change of the longest wavelength with increasing box size, were beyond the scope of the original paper.
+To gain further insights on whether the staircase structure can fully develop the following brief communication will focus on the effects of the box size on the E$\times$B staircase structure and whether the wavelength converges with the box size.\bigskip
 
@@
 \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}
-The strength of the shearing process is the E$\times$B shearing rate $\omega_{\mathrm{E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
+The strength of the shearing process is the E$\times$B shearing rate $\omega_{\mathrm{\:E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
 \cite{doi:10.1063/1.871313, doi:10.1063/1.872847}
-The shearing rate $\omega_{\mathrm{E \times B}}$ is defined as 
+The shearing rate $\omega_{\mathrm{\:E \times B}}$ is defined as 
 \begin{equation}
-	\omega_{\mathrm{E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
+	\omega_{\mathrm{\:E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
 	\label{eq:shearingrate}
@@
 where $\langle \Phi \rangle$ is the zonal electrostatic potential and $\psi$ the radial coordinate that labels the flux surfaces. \cite{doi:10.1063/1.4952621, doi:10.1063/1.3005380}
-It was shown that the nonlinear threshold for turbulence is directly related to shear stabilization. \cite{doi:10.1063/1.873896} 
-The shear stabilization is often expressed in the empirical Waltz rule $\omega_{\mathrm{E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
+It was shown that the nonlinear threshold for turbulence is directly related to shear stabilization. \cite{doi:10.1063/1.873896} % Dimits et al. showed
+The shear stabilization is often expressed in the empirical Waltz rule $\omega_{\mathrm{\:E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
 The discovered zonal flows, also known as E$\times$B staircase \cite{PhysRevE.82.025401}, exhibit amplitudes, which satisfy the stabilization criteria in terms of the E$\times$B shearing rate.
-For a fully developed staircase structure the E $\times$ B shearing rate $\omega_{\mathrm{E \times B}} = \gamma$.
+For a fully developed staircase structure the E $\times$ B shearing rate $\omega_{\mathrm{\:E \times B}} = \gamma$.
 \cite{doi:10.1063/1.4961231, doi:10.1063/1.4952621}\bigskip
@@
 Which box size was used can be seen in each graphic in form of $N_\mathrm{R}\times N_\mathrm{B}$, where $N_\mathrm{R}$ stands for the times the radial box size was increased and $N_\mathrm{B}$ for the binormal box size.
-To validate a stabilized turbulence and if the E$\times$B Staircase structure has fully developed a closer look at the long wavelength zonal flow mode $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}}$ is necessary.
-The long wavelength zonal flow mode is the absolute of the fourier transform of the shearing rate $\omega_{\mathrm{E \times B}}$ for each time step, where $k_\mathrm{i}$ is the wave vector connected to the long wavelength zonal flow mode with index i and is defined as following
+To validate a stabilized turbulence and if the E$\times$B Staircase structure has fully developed a closer look at the long wavelength zonal flow mode $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{i}}$ is necessary.
+The long wavelength zonal flow mode is the absolute of the Fourier transform of the shearing rate $\omega_{\mathrm{\:E \times B}}$ for each time step, where $k_\mathrm{i}$ is the wave vector connected to the long wavelength zonal flow mode with index i and is defined as following
 \begin{equation}
@@
 with $l_\psi$ as the length of the box size.
-In addition, the maximum amplitude of the long wavelength zonal flow mode  $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}} = \gamma $.
+In addition, the maximum amplitude of the long wavelength zonal flow mode  $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{i}} = \gamma $.
 \cite{doi:10.1063/1.4961231} It is important that only for one wave vector $k_i$ the zonal mode reaches its maximum amplitude any other mode has to be close by zero.
@@
 The plasma parameters are closely modelled after those in Rath et al. \cite{doi:10.1063/1.4961231} with the cyclone base case.
-%safety factor $q = 1.4$, magnetic shear $\hat{s} = 0.78$, inverse aspect ratio $\epsilon = 0.19$, density gradient $R/L_n = 2.2$ and electron to ion temperature ratio $T_e/T_i = 1$.
+%safety factor $q = 1.4$, magnetic shear $\widehat{s} = 0.78$, inverse aspect ratio $\epsilon = 0.19$, density gradient $R/L_n = 2.2$ and electron to ion temperature ratio $T_e/T_i = 1$.
 The maximum of the velocity grid is three times the thermal velocity for both the parallel and the perpendicular velocities.
 Adiabatic electrons were investigated, while neglecting collisions.
-The resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table \ref{tab:resolution} with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduces the runtime of simulations.
-Simulations showed that this correction in grid points does not affect the results itself.
-Furthermore, the gradient length $R/L_\mathrm{T}$ was set to the value 6.0 nearby the finite heat flux threshold shown in Fig.
-4 in Rath et al \cite{doi:10.1063/1.4961231}.
+The resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table \ref{tab:resolution} with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduce the runtime of simulations.
+Simulations showed no effect of the correction in grid points on the produced results.
+Furthermore, the gradient length $R/L_\mathrm{T}$ was set to the value 6.0 close to the finite heat flux threshold shown in Fig.
+4 in Rath et al.
+\cite{doi:10.1063/1.4961231}.
 The simulations are performed with the flux tube version of the non-linear gyro-kinetic code GKW \cite{Peeters20092650} with periodic boundary conditions and circular geometry.
@@
 
-The data is normalized with the heat conduction coefficient $\chi$ in gyro-Bohm units ($\rho^2\nu_{\mathrm{th}}/R)$, where $\rho = m_\mathrm{i}\nu_{\mathrm{th}}/eB$ is the ion Larmor radius, $\nu_{\mathrm{th}} = \sqrt{2T/m_\mathrm{i}}$ is the thermal velocity, $T$ the background temperature, e is the unit charge and $R$ is the major radius.
-\bigskip
+The data is normalized with the heat conduction coefficient $\chi$ in gyro-Bohm units ($\rho^2\nu_{\mathrm{th}}/R)$, where $\rho = m_\mathrm{i}\nu_{\mathrm{th}}/eB$ is the ion Larmor radius, $\nu_{\mathrm{th}} = \sqrt{2T/m_\mathrm{i}}$ is the thermal velocity, $T$ the background temperature, $e$ is the unit charge and $R$ is the major radius.
 
@@
 At the beginning the box size was only increased in radial direction with values for $N_\mathrm{R} = 1, 2, 3, 4$.
-The simulation are showing that the longest wavelength zonal mode is connected with $N_\mathrm{R}$ in terms of index $\mathrm{i} = N_\mathrm{R}$.
+The simulations show that the longest wavelength zonal mode is connected with $N_\mathrm{R}$ in terms of index $\mathrm{i} = N_\mathrm{R}$.
 This behavior can be seen in Fig.
 \ref{fig:wexb-eflux-1-2-3-4x1-comparison} where the longest wavelength zonal mode for the box size 1$\times$1 has the wave vector $k_1$, for 2$\times$1 the wave vector $k_2$ and so on.
-Also, it is worth mentioning that after a long run of the simulation the turbulence can still stabilize, which was not certain because of the numerical dissipation of the simulation itself.
+Despite numerical dissipation of the simulation producing an unlikeliness of turbulence stabilization, it may still stabilize after the simulation is run for a longer amount of time.
 \cite{doi:10.1063/1.4961231}
 
-Furthermore, it is clearly visible that the longest wavelength zonal mode reaches its maximum amplitude (Fig.
+Furthermore, it is visible that the longest wavelength zonal mode reaches its maximum amplitude (Fig.
 \ref{fig:wexb-eflux-1-2-3-4x1-comparison} (b)) when the turbulence is stabilized  (Fig.
 \ref{fig:wexb-eflux-1-2-3-4x1-comparison} (a)) but further investigation yields the result that the simulations has not stabilized yet.
-For example for the box size 3$\times$1 is not stable although between [28000, 31000] normalized time units the turbulence subdued.
+For example for the box size 3$\times$1 the turbulence has not been stabilized between [28000, 31000] normalized time units although the heat conduction coefficient $\chi$ subdued.
 This can be observed in Fig.
@@
 \ref{fig:wexb-3x1-selection} for [10000, 12000] normalized time units.
-This occurrence is due to the previous mentioned condition that only one zonal mode has a maximal amplitude and any other mode is closely by zero.
-So as additional verification process one could visualize the shearing rate $\omega_{\mathrm{E \times B}}$ to determine if the staircase structure has fully developed.
+This occurrence is due to the previously mentioned condition that only one zonal mode has a maximal amplitude while any other mode has an amplitude close to zero.
+So as additional verification process one could visualize the shearing rate $\omega_{\mathrm{\:E \times B}}$ to determine whether the staircase structure has fully developed.
 
-When the turbulence is stabilized and the longest wavelength zonal mode has developed the wavelength of the shearing rate $\omega_{\mathrm{E \times B}}$ does converge with the box size.
+After the turbulence is stabilized and the longest wavelength zonal mode has developed the wavelength of the shearing rate $\omega_{\mathrm{\:E \times B}}$ converges with the box size.
 The E$\times$B staircase structure does repeat itself exactly $N_\mathrm{R}$ times.
 In Fig.
-\ref{fig:wexb-1-2-3-4x1-stable-comparison} is shown how the wavelength converges for different box sizes for different time intervals and the remarkable result of the repetition of the staircase structure.
-This can also be linked to the wave vector of the longest wavelength zonal mode for example for the box size 2$\times$1 the wave vector would be $k_2$ (Fig.
-\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b)) and the staircase structure repeats itself 2 times.
-It is also visibility that the Waltz rule does even hold for radial increased box sizes.
-Note that for clearer visibility the staircase structure for box size 1$\times$1 and for 4$\times$1 got shifted to the left so that the staircase structures do overlapped.
-This does not affect the physical result because of the applied periodic boundary conditions of the box.
+\ref{fig:wexb-1-2-3-4x1-stable-comparison} shows how the wavelength converges for different box sizes for different time intervals and the for this paper significant staircase structure repetition pattern.
+This can be linked to the wave vector of the longest wavelength zonal mode, for example for the box size 2$\times$1 the wave vector would be $k_2$ (Fig.
+\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b)) and the staircase structure repeats itself 2 times which shows a convergence of the structure to the box size.
+It is also visible that the Waltz rule applies for radially increased box sizes as well.
+Note that for clearer visibility the staircase structure for box size 1$\times$1 and for 4$\times$1 got shifted to the left in Fig.
+\ref{fig:wexb-1-2-3-4x1-stable-comparison} so that the staircase structures do overlapped.
+This shift does not affect the physical result because of the applied periodic boundary conditions of the box.
 
 \includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Shearing rate $\omega_{\mathrm{E \times B}}$ for different time intervals in which heat conduction is almost zero but the E$\times$B staircase has not fully developed for box size 3$\times$1
+	Shearing rate $\omega_{\mathrm{\:E \times B}}$ for different time intervals in which heat conduction is almost zero but the E$\times$B staircase has not fully developed for box size 3$\times$1
 }{fig:wexb-3x1-selection}
@@
 	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased box sizes
-	\textbf{(b)} Time traces of $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}}$ for radial increased box sizes
+	\textbf{(b)} Time traces of $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{i}}$ for radial increased box sizes
 }{fig:wexb-eflux-1-2-3-4x1-comparison}{}
@@
 \includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{
-	Comparison of shearing rate $\omega_{\mathrm{E \times B}}$ for radial increased box sizes. The staircase structure got 
+	Comparison of shearing rate $\omega_{\mathrm{\:E \times B}}$ for radial increased box sizes. The staircase structure got 
 	shifted for better visibility.
@@
 
-As next step the box size got increased in radial and binormal direction the same time.
-For this simulation is it noticeable that the turbulence got stabilized and fully developed after 1500 normalized time unit seen in Fig.
+In the next step the box size was increased in radial and binormal direction at the same time.
+For this simulation the turbulence got stabilized and fully developed after 1500 normalized time unit seen in Fig.
 \ref{fig:eflux-1x1-2x2-3x3-comparison}.
-This leads to a much faster computation of zonal flow structures if the box size has to be increased.
-In comparison, the box size 3$\times$1 get stabilized and fully developed after 42000 normalized time units.
+This leads to a much faster computation of zonal flow structures if the box size had to be increased.
+In comparison, the box size 3$\times$1 gets stabilized and fully developed after 42000 normalized time units.
 
@@
 Even for this type of box sizes the wavelength converges with the box size and the Waltz rule is still satisfied.
-But it is to mentioned that for the 3$\times$3 box size the E$\times$B staircase structure repeats itself four times which can be seen in Fig.
-\ref{fig:wexb-3x3-stable} which stand in contradiction to the repetition of the radial increased box size.
+For the 3$\times$3 box size the E$\times$B staircase structure repeats itself four times which can be seen in Fig.
+\ref{fig:wexb-3x3-stable}.
+This stands in contradiction to the repetition of the radial increased box size.
 One could argue that the repetition of the staircase structure of a radial and binormal increased box size is greater equal $N_\mathrm{R}$.
-This could lead to the question if E$\times$B staircase structure has a different repetition for different box sizes, although this question is beyond of the scope of this paper.
+This poses the question whether E$\times$B staircase structure has a different repetition for different box sizes, for different box sizes, but this will need further research to be conducted.
 
 \includegraphicsOneCol{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Stablized shearing rate $\omega_{\mathrm{E \times B}}$ for boxsize 3$\times$3
+	Stablized shearing rate $\omega_{\mathrm{\:E \times B}}$ for boxsize 3$\times$3
 }{fig:wexb-3x3-stable}
@@
 
-To conclude the wavelength does converge with the chosen box size and the Waltz rule holds in every simulation.
-Long runs can still result in stabilized turbulence.
-Further investigation of the long wavelength zonal mode has to be made to determine if the turbulence has been stabilized.
-For that it is important that only one long wavelength zonal mode has a maximal amplitude for given wave vector $k_i$ which is also the longest wavelength zonal mode.
-The E$\times$B staircase structure repeats itself multiple times in increased box sizes where for radial increased box size the repetition is linked to the times the box size was increased radial, although this does not hold for radial and binormal increased box sizes.
 \bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+To conclude, the wavelength does converge with the chosen box size and the Waltz rule holds in every simulation.
+Long runs can produce stabilized turbulence.
+Further investigation of the long wavelength zonal mode has to be made to determine if and under what conditions the turbulence has been stabilized.
+Therefore, it is important that only one long wavelength zonal mode has a maximal amplitude for given wave vector $k_i$ which is also the longest wavelength zonal mode.
+The E$\times$B staircase structure repeats itself multiple times in increased box sizes whereas radial increased box size shows a link between repetition and the times of radial box size increase, although this does not hold for radial and binormal increased box sizes.
 

### 2023-01-30 presentation/presentation.tex
Presentation + Theme

% Author         : Pedro Ferreira (up201806093@up.pt)
% Version        : 1.0
% Created on     : 12.08.2022
% Last Edited on : 14.08.2022

% Adapted from HSRM theme by Benjamin Weiss
% Copyright      : Copyright (c) 2013-2014 by Benjamin Weiss. All rights reserved.
% License        : This file may be distributed and/or modified under the
%                  GNU Public License.
% Description    : HSRM beamer theme demonstration. Also includes a short 
%                  Tutorial regarding the beamer class.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                    !!! Compile with LuaLaTex !!!
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[compress,aspectratio=169]{beamer}

\usepackage{sleektheme/beamerthemesleek}
\usetheme{sleek}
\usepackage{sleektheme/dtklogos} % must be loaded after theme

%--------------------------------------------------------------------------
% Common packages
%--------------------------------------------------------------------------
\usepackage[ngerman]{babel}

\usepackage{graphicx}
\usepackage{multicol}
\usepackage{multimedia}
\usepackage{media9}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{tikz}
\usetikzlibrary{mindmap,backgrounds}
\usepackage{booktabs}
\newcommand{\tabitem}{~~\llap{\textbullet}~~}

% Advanced table functions
\usepackage{tabularx,ragged2e}
\usepackage{booktabs}
% Listings extension
\usepackage{listings}
\lstset{ %
language=[LaTeX]TeX,
basicstyle=\normalsize\ttfamily,
keywordstyle=,
numbers=left,
numberstyle=\tiny\ttfamily,
stepnumber=1,
showspaces=false,
showstringspaces=false,
showtabs=false,
breaklines=true,
frame=tb,
framerule=0.5pt,
tabsize=4,
framexleftmargin=0.5em,
framexrightmargin=0.5em,
xleftmargin=0.5em,
xrightmargin=0.5em
}

%--------------------------------------------------------------------------
% Functions
%--------------------------------------------------------------------------

\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}

%--------------------------------------------------------------------------
% Document
%--------------------------------------------------------------------------

\title{Turbulenzen in Plasma}
\subtitle{Beispiel anhand von Plasma im Tokamak Reaktor}
\author{Manuel Lippert}
\institute{Physik (Master of Science)}

\begin{document}

	\maketitle

	\begin{frame}{Gliederung}
		\tableofcontents[hideallsubsections]
	\end{frame}

	\centering

	\section{Magnetohydrodynamische (MHD) Gleichungen}
	%\subsection{Magnetohydrodynamische Gleichungen}
	%\begin{frame}
	%	\frametitle{Magnetohydrodynamische Gleichungen}
	%	\begin{gather*}
	%		\begin{aligned}
	%			\partial_t \rho + \nabla \cdot (\rho \vect{v}) &= 0 \\
	%			\rho \partial_t \vect{v} + \rho (\vect{v} \cdot \nabla) \vect{v} + \nabla p &= \frac{\vect{J} \times \vect{B}}{c} + \rho \vect{g} + \vect{F} \\
	%			\nabla \times \vect{B} &= \frac{4 \pi}{c} \vect{J} \\
	%			\nabla \times \vect{E} &= -\frac{1}{c}\partial_t\vect{B} \\
	%			\nabla \cdot \vect{B} &= 0 \\
	%			\vect{J} &= \sigma \left(\vect{E} + \frac{\vect{v} \times \vect{B}}{c}\right)
	%		\end{aligned}
	%	\end{gather*}
	%\end{frame}

	\begin{frame}
		\frametitle{Magnetohydrodynamische (MHD) Gleichungen}
		\begingroup
			\setbeamercolor{block title}{bg=Sec3Dark}
			\setbeamercolor{block body}{bg=Sec3}
			\begin{block}{Kontinuitätsgleichung}
				\centering
				$\partial_t \rho + \nabla \cdot (\rho \vect{v}) = 0$
			\end{block}
		\endgroup
		\begingroup
			\setbeamercolor{block title}{bg=Sec2CompDark}
			\setbeamercolor{block body}{bg=Sec2Comp}
			\begin{block}{Navier-Stokes Gleichung}
				\centering
				$\rho \partial_t \vect{v} + \rho (\vect{v} \cdot \nabla) \vect{v} = - \nabla p + \frac{\vect{J} \times \vect{B}}{c} + \rho \vect{g} + \vect{F}$
			\end{block}
		\endgroup
		\begingroup
			\setbeamercolor{block title}{bg=Sec1Dark}
			\setbeamercolor{block body}{bg=Sec1}
			\begin{block}{Maxwell Gleichungen}
				\centering
				\begin{tabular}{l l}
					$\nabla \cdot \vect{B} = 0$ & $\nabla \times \vect{B} = \frac{4 \pi}{c} \vect{J}$ \\[0.3cm]
					$\vect{J} = \sigma \left(\vect{E} + \frac{\vect{v} \times \vect{B}}{c}\right)$ & $\nabla \times \vect{E} = -\frac{1}{c}\partial_t\vect{B}$ \\
				\end{tabular}
			\end{block}
		\endgroup
	\end{frame}
	\begin{frame}
		\frametitle{Beschreibung von Plasma im Tokamak Reaktor}
		\includegraphics[scale = 1]{pictures/Zonal_Flow.PNG}
		\begin{alertblock}{Achtung Instabilität am Plasmarand!}
			\centering
			Ionen Temperaturgradient Instabilität $\approx$ Rayleigh-Taylor Instabilität
		\end{alertblock}
	\end{frame}

	\section{Ionen Temperaturgradient (ITG) Instabilität}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_00.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_01.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_02.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_03.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_04.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_05.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_06.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_07.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_08.pdf}

	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_09.pdf}
	\end{frame}
	\begin{frame}
		\frametitle{ITG Instabilität am Plasmarand}
		\includegraphics{pictures/ITG_Instability_10.pdf}
	\end{frame}

	\begin{frame}
		\frametitle{Zonal Flow Bildung}
		\begin{minipage}{0.6\textwidth}\raggedleft
			\centering
			\begin{itemize}
				\item ITG Instabilität erzeugt turbulente Wirbelströmung
				\item Wirbelströmung breitet sich über die Breite des Plasmatorus aus
				\item Entstehung von \alert{Zonal Flows}
				\item Namensgebung aus der Atmosphärenphysik (Bsp. Wolkenbänder von Jupiter)
			\end{itemize}
		\end{minipage}
		\begin{minipage}{0.3\textwidth}
			\begin{figure}%[4]
				\href{https://www.youtube.com/watch?v=O6tUHzfj-zM}{\includegraphics[width=\textwidth]{pictures/turbulence.png}}
			\end{figure}
			{\tiny Simulation eines Plasmaschlauch von Jeff Candy und Ron Waltz mit Code GYRO}
		\end{minipage}
	\end{frame}

	\section{Turbulenz im Plasma stabilisieren}
	\begin{frame}
		\frametitle{Anpassung der Magnetfeldlinienrichtung}
		Verdrillen der magnetischen Flussdichte $\vect{B}$ überführt Plasma in stabilen Zustand \\[0.2cm]
		\includegraphics[scale=0.25]{pictures/Stabilzation_Mechanism_black.png}
	\end{frame}

	\begin{frame}
		\frametitle{Turbulenzen verringern}
		Turbulenzen können mit negativer magnetischen Scherung verringert werden.\\[0.2cm]		
		\begin{minipage}{0.6\textwidth}\raggedleft
			\begin{itemize}
				\item Wirbelströmung verursachende Partikel folgen meistens Feldlinien
				\item Umkehrung der magnetischen Scherung verdreht Wirbelströmung in kleinen Abständen in Richtung der "Guten Krümmung"
				\item Umgekehrte magnetische Scherung wird erzeugt durch Quetschung des Magnetfelds bei hohen Plasmadruck
				\item Durch die Verformung des Plasmas kann sich Scherung lokal verändern
			\end{itemize}
		\end{minipage}
		\begin{minipage}{0.3\textwidth}
			\includegraphics[scale = 0.2]{pictures/Magnetic_Shear_black.png}
		\end{minipage}
	\end{frame}

	\begin{frame}
		Wirbelströmung, welche größere Längenskalen in \doublequoted{Schlechte Krümmung} transportieren, werden mit Scherung reduziert bis sie sich komplett unterdrücken.
		\frametitle{Regelung oder Unterdrückung von Turbulenzen}
		\includegraphics[scale = 1.3]{pictures/Shearing.pdf}
	\end{frame}

	\begin{frame}
		\frametitle{Regelung oder Unterdrückung von Turbulenzen}
		{\tiny Waltz, Kerbel, Phys. Plasmas 1994 w/ Hammett, Beer, Dorland, Waltz Gyrofluid Eqs., Numerical Tokamak Project, DoE Computational Grand Challenge}
		\begin{tabular}{c c}
			\includegraphics[scale = 0.2]{pictures/Stablized_Plasma_Interaction.png} & \includegraphics[scale = 0.2]{pictures/Stablized_Plasma_LongScale.png} \\
			Dominante nichtlineare Interaction  & Zusätzlicher verscherter Zonal Flow \\
			zwischen turbulenter Wirbelströmung & unterdrückt Turbulenz komplett \\
			und kontrollierten Zonal Flow & \\
		\end{tabular}
	\end{frame}

	%\section{Parallelen zur Atmosphären-Physik}
	%\begin{frame}
	%	\frametitle{Zonal Flow auf Jupiter vs. Tokamak Reaktor}	
	%\end{frame}

	\begin{frame}
		\frametitle{Danke für die Aufmerksamkeit!}
		\includegraphics[scale = 0.35]{pictures/Fusion_Meme.jpeg} \\
	\end{frame}

\end{document}

### 2023-01-30 presentation/presentation.tex
Correction of the presentation

@@
 		\frametitle{Turbulenzen verringern}
-		Turbulenzen können mit negativer magnetischen Scherung verringert werden.\\[0.2cm]		
 		\begin{minipage}{0.6\textwidth}\raggedleft
@@
 				\item Wirbelströmung verursachende Partikel folgen meistens Feldlinien
-				\item Umkehrung der magnetischen Scherung verdreht Wirbelströmung in kleinen Abständen in Richtung der "Guten Krümmung"
-				\item Umgekehrte magnetische Scherung wird erzeugt durch Quetschung des Magnetfelds bei hohen Plasmadruck
+				\item Turbulenzen können mit negativer magnetischen Scherung verringert werden
+				\item Verdrehung in Richtung der "Guten Krümmung"
+				%\item Umkehrung der magnetischen Scherung verdreht Wirbelströmung in kleinen Abständen in Richtung der "Guten Krümmung"
+				%\item Umgekehrte magnetische Scherung wird erzeugt durch Quetschung des Magnetfelds bei hohen Plasmadruck
 				\item Durch die Verformung des Plasmas kann sich Scherung lokal verändern
@@
 
-	\begin{frame}
-		Wirbelströmung, welche größere Längenskalen in \doublequoted{Schlechte Krümmung} transportieren, werden mit Scherung reduziert bis sie sich komplett unterdrücken.
-		\frametitle{Regelung oder Unterdrückung von Turbulenzen}
-		\includegraphics[scale = 1.3]{pictures/Shearing.pdf}
-	\end{frame}
+	%\begin{frame}
+	%	Wirbelströmung, welche größere Längenskalen in \doublequoted{Schlechte Krümmung} transportieren, werden mit Scherung reduziert bis sie sich komplett unterdrücken.
+	%	\frametitle{Regelung oder Unterdrückung von Turbulenzen}
+	%	\includegraphics[scale = 1.3]{pictures/Shearing.pdf}
+	%\end{frame}
 

### 2023-02-13 briefcommunicaton/briefcommunication.tex
#37: 1st Correction Florian Rath

--- 2022-12-22 briefcommunicaton/briefcommunication.tex
@@
 
-\newcommand{\includegraphicsTwoCol}[3]{
-	\begin{figure*}
-    	\includegraphics[width=\textwidth]{#1}
-		\caption{#2}
-	\end{figure*}
-	\label{#3}
-	\increasecounter{fig}{2}
+\newcommand{\words}{
+	\immediate\write18{texcount -sum -1 \jobname.tex > wordcount.txt}
 }
+
+\newcommand{\wexb}{\omega_{\mathrm{\:E \times B}}}
+\newcommand{\exb}{\mathrm{\:E \times B}}
+\newcommand{\hatwexbkvec}{\hat{\omega}_{\mathrm{\:E \times B},\mathbf{k}}}
+\newcommand{\NR}{N_\mathrm{R}}
+\newcommand{\NB}{N_\mathrm{B}}
+\newcommand{\vthref}{v_\mathrm{th,ref}}
+\newcommand{\Rref}{R_\mathrm{ref}}
+\newcommand{\rlt}{R/L_T}
+\newcommand{\rhothref}{\rho_\mathrm{th,ref}}
 
@@
 
-%\title[Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation]{
-%	Convergence of shearing rate $\omega_{\mathrm{E}\times\mathrm{B}}$ with boxsize in gradient driven simulation
-%}
+\title[Wavelength convergence of the ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
+{Wavelength convergence of the ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
 
-%\author{M. Lippert}
-%	\altaffiliation{Repository of this work: \\ 
-%					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
-%\author{F. Rath}
-%	\altaffiliation{Author to whom correspondence should be addressed: \\ 
-%					Florian.Rath1@uni-bayreuth.de}
-%	\email{Florian.Rath1@uni-bayreuth.de}
-%\author{A. G. Peeters}
-%\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
+\author{M. Lippert}
+	\altaffiliation{Repository of this work: \\ 
+					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
+\author{F. Rath}
+	\altaffiliation{Author to whom correspondence should be addressed: \\
+					Florian.Rath1@uni-bayreuth.de}
+	\email{Florian.Rath1@uni-bayreuth.de}
+\author{A. G. Peeters}
+\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
 
-%\date{\today}
+\date{\today}
 
@@
 
-%\begin{abstract}
-%	An article usually includes an abstract, a concise summary of the work
-%	covered at length in the main body of the article. It is used for
-%	secondary publications and for information retrieval purposes. 
-%\end{abstract}
+\begin{abstract}
+    The wavelength convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
+    Its is shown that a mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rhothref$ is inherent to ITG driven turbulence with Cyclone Base Case paramters in the local limit.
 
-%\maketitle
+\end{abstract}
+
+\maketitle
 
@@
 
-This brief communication focuses on continuing the work of Rath et. al \cite{doi:10.1063/1.4961231}. 
-Their paper elaborates on gradient driven flux-tub simulations close to the non-linear threshold and the occurrence of the E$\times$B staircase structure and its formation over the simulation.
-At the end of Section IV it was discussed that the circumstances for which the staircase can fully develop are beyond the scope of the paper.
-To gain further insights on whether the staircase structure can fully develop the following brief communication will focus on the effects of the box size on the E$\times$B staircase structure and if the wavelength converges with the box size.\bigskip
+Ion temperature gradient driven turbulence close to marginal stability often exhibits zonal flow pattern formation on mesoscales \textemdash so-called ${\mathrm{\:E}\times\mathrm{B}}$ staircase structures \cite{pradalier2010}.
+Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{peeters2016, weikl2017, rath2021} as well as global gradient-driven \cite{mcmillan2009, villard2013, seo2022} and global flux-driven \cite{pradalier2010, pradalier2015, wang2020, kim2022, kishimoto2023} studies.
+In global studies, usually spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern wavelength of several ten Larmor radii.
+% By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
+By contrast, in the aformentioned local studies the wavelength of ${\mathrm{\:E}\times\mathrm{B}}$ staircase structures is always found to converge to the radial box size of the flux tube domain.
+The above observations lead to the question: 
+Does the basic pattern wavelength always converges to the box size, or is there a typical mesoscale wavelength inherent to staircase structures also in a local flux-tube description?
+The latter case would imply that it is not necessarily global physics, i.e., profile effects, that sets (i) the wavelength of the ExB staircase pattern and (ii) the scale of avalanche-like transport events, usually restricted to ${\mathrm{\:E}\times\mathrm{B}}$ staircase structures and considered as a nonlocal transport mechnaism\cite{pradalier2010}.
+In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\cite{peeters2016}.\bigskip
 
-%%% THEORY =================================================================================================================
+%%% SIMULATION SET-UP ======================================================================================================
 
-%\section{Theory}
-%\label{sec:theory}
-
-It is known that radially sheared zonal flows play a significant role in nonlinear stabilization in tokamak plasmas. \cite{WACooper1988,doi:10.1063/1.859529,doi:10.1063/1.873896}. 
-Through advection on the sheared zonal flows the turbulent structure in plasma gets deformed and tilted, which causes an E$\times$B nonlinearity.
-\cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}
-The strength of the shearing process is the E$\times$B shearing rate $\omega_{\mathrm{E \times B}}$ which is the radial derivative of the advecting zonal flow velocity.
-\cite{doi:10.1063/1.871313, doi:10.1063/1.872847}
-The shearing rate $\omega_{\mathrm{E \times B}}$ is defined as 
-\begin{equation}
-	\omega_{\mathrm{E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2}
-	\label{eq:shearingrate}
-\end{equation}
-where $\langle \Phi \rangle$ is the zonal electrostatic potential and $\psi$ the radial coordinate that labels the flux surfaces. \cite{doi:10.1063/1.4952621, doi:10.1063/1.3005380}
-It was shown that the nonlinear threshold for turbulence is directly related to shear stabilization. \cite{doi:10.1063/1.873896} 
-The shear stabilization is often expressed in the empirical Waltz rule $\omega_{\mathrm{E \times B}} \sim \gamma$, \cite{doi:10.1063/1.870934, doi:10.1063/1.872847} where $\gamma$ is defined as the maximum linear growth rate in the unstable mode.
-The discovered zonal flows, also known as E$\times$B staircase \cite{PhysRevE.82.025401}, exhibit amplitudes, which satisfy the stabilization criteria in terms of the E$\times$B shearing rate.
-For a fully developed staircase structure the E $\times$ B shearing rate $\omega_{\mathrm{E \times B}} = \gamma$.
-\cite{doi:10.1063/1.4961231, doi:10.1063/1.4952621}\bigskip
-
-%%% METHODS ==================================================================================================================
-
-%\section{Simulation Setup}
-%\label{sec:setup}
-
-The plasma parameters are closely modelled after those in Rath et al.
-\cite{doi:10.1063/1.4961231} with the cyclone base case: safety factor $q = 1.4$, magnetic shear $\hat{s} = 0.78$, inverse aspect ratio $\epsilon = 0.19$, density gradient $R/L_n = 2.2$ and electron to ion temperature ratio $T_e/T_i = 1$.
-The maximum of the velocity grid is three times the thermal velocity for both the parallel and the perpendicular velocities.
-Adiabatic electrons were investigated, while neglecting collisions.
-
-For this paper the resolution "Standard resolution with 6th order (S6)" from Ref \onlinecite{doi:10.1063/1.4961231} was used and is given in Table with changes in $N_{\nu_\parallel}$ from 64 grid points to 48 to reduces the runtime of simulations.
-Simulations showed that this correction in grid points does not affect the results itself.
-
-The simulations are performed with the flux tube version of the non-linear gyro-kinetic code GKW \cite{Peeters20092650} with periodic boundary conditions.
-Further information regarding the simulation can be found in Ref \onlinecite{doi:10.1063/1.4961231}.
-
-\begin{table}
+\begin{table}[ht]
 	\begin{ruledtabular}
@@
 	\caption{
-		Resolution used in this paper for further information read Rath et al.
-\cite{doi:10.1063/1.4961231} %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+		Resolution used in this paper for further information read Ref.~\cite{peeters2016}.
+%: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
 	}
 	\label{tab:resolution}
-\end{table} \bigskip
+\end{table}
+
+Here, gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} in the adiabatic electron approximation.
+In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $R/L_T = 6.0$ (unless stated otherwise) and circular concentric flux surfaces.
+The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
+It has been carefully verified that this modification preserves the same physical outcome as the orginial study.
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.
+\\
+The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)~\rho_\mathrm{th,ref}$, in the radial and binormal direction is always indicated by the real parametes $N_\mathrm{R}$ and $N_\mathrm{B}$, respectively.
+Here, $\rho_\mathrm{th,ref}$ is the reference thermal Larmor radius and the nomenclature $N_\mathrm{R}\times N_\mathrm{B}$ is applied throughout this work to indicate the increase in box size.  
+Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution (Tab.~\ref{tab:resolution}) and standard box size.
+\bigskip
+
+%%% DIAGNOSTICS ================================================================================================================
+The ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern is manifest as radial structure formation in the ${\mathrm{\:E}\times\mathrm{B}}$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
+\begin{equation}
+	\omega_{\mathrm{\:E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2},
+	\label{eq:shearingrate}
+\end{equation}
+\increasecounter{eq}{1}{1}
+where $\langle \Phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T_\mathrm{ref}/e$ ($\rho_\ast = \rho_\mathrm{th,ref}/R_\mathrm{ref}$ is the reference thermal Larmor radius normalized with the reference major radius $R_\mathrm{ref}$, $T_\mathrm{ref}$ is a reference temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rho_\mathrm{th,ref}$.
+The E$\times$B shearing rate $\omega_{\mathrm{\:E \times B}}$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
+Here, the zonal potential is estimated from the electrostatic potential $\phi$ on the two dimensional $\psi$-$\zeta$-plane at the low field side according to
+\begin{equation}
+\langle \Phi \rangle = \frac{1}{l_\zeta} \int_0^{l_\zeta} \mathrm{d}\zeta ~ \phi(\psi,\zeta,s=0).
+\end{equation}
+In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the ${\mathrm{\:E}\times\mathrm{B}}$ shearing rate is considered.
+It is defined by
+\begin{equation}
+	\omega_{\mathrm{\:E \times B}} = \sum_{k_j} \hat \omega_{\mathrm{\:E \times B}}(k_j,t) \times \exp(\mathrm{i} k_j \psi),
+	\label{eq:shearingrate_fourier}
+\end{equation}
+\increasecounter{eq}{1}{1}
+were $\hat \omega_{\mathrm{\:E \times B}}$ is the complex Fourier coefficient and 
+% use j as zonal flow mode number to distinguish it from the imaginary unit i
+\begin{equation}
+	k_\mathrm{j} = \frac{2\pi \mathrm{j}}{l_\psi}
+\end{equation}
+\increasecounter{eq}{1}{1}
+% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
+defines the zonal flow wave vector with the zonal flow mode number $j$ ranging in $-(N_\psi -1)/2 \leq j \leq (N_\psi -1)/2 $.
+% define the zonal flow modes' amplitude in terms of the shearing rate
+Based on the defintions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{j}} = 2 |\hat \omega_{\mathrm{\:E \times B}}|$.
+% introduces the term basic mode of the pattern for later usage
+In general, the zonal flow mode that dominates the ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern, \textemdash also referred to as the \textit{basic mode } of the pattern in this work \textemdash exhibits the maximum amplitude in the spectrum $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{j}}$.
+Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rho_\mathrm{th,ref}^2 v_{\mathrm{th,ref}}/R_\mathrm{ref}$ ($v_{\mathrm{th,ref}} = \sqrt{2 T_\mathrm{ref}/m_\mathrm{ref}}$ is the reference thermal velocity and $m_\mathrm{ref}$ is the reference mass) throughout this work.
+\bigskip
 
@@
 
-The data is normalized with the heat conduction coefficient $\chi$ in gyro-Bohm units ($\rho^2\nu_{\mathrm{th}}/R)$, where $\rho = m_\mathrm{i}\nu_{\mathrm{th}}/eB$ is the ion Larmor radius, $\nu_{\mathrm{th}} = \sqrt{2T/m_\mathrm{i}}$ is the thermal velocity, $T$ the background temperature, e is the unit charge and $R$ is the major radius.
+\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased box sizes
+	\textbf{(b)} Time traces of $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{i}}$ for radial increased box sizes
+}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
 
-\includegraphicsOneCol{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial and binormal increased boxsizes
+% \increasecounter{fig}{1}
+
+\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
+	Comparison of shearing rate $\omega_{\mathrm{\:E \times B}}$ for radial increased box sizes. The staircase structure got 
+	shifted for better visibility.
+}{fig:wexb-1-2-3-4x1-stable-comparison}{}
+
+% what kind of analysis
+In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size.
+The scan covers the realizations $\NR\times\NB \in [ 1\times1,~2\times1,~3\times1,~4\times1]$.
+% short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost supressed turbulence (see Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}).
+% introduce the fully developed staircase structure
+The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-1-2-3-4x1-stable-comparison}. 
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable liner ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+% insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
+Fig.~\ref{fig:wexb-1-2-3-4x1-stable-comparison} shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
+% yes, the wavelength converges; but even stronger result: the standard box size already allows for the correct staircase wavelength
+Hence, the basic wavelength of the pattern not only converges with increasing radial box size, the converged wavelength turns out to agree with the standard radial box size of Ref.~\cite{peeters2016} \\
+% cautios words
+Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-1-2-3-4x1-stable-comparison} represent the real converged pattern in a statistical sense.
+% very long initial quasi-stationary states
+Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4~\Rref/\vthref$.
+% the j=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
+During this period the zonal flow mode with $j = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3~\Rref/\vthref$.
+Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000~\Rref/\vthref$ and $t \approx 18000~\Rref/\vthref$.
+% first important point: competition between the j = 3 and j = 4 mode -> the j = 4 mode is of shorter length scale compared to the j = 3 mode, hence, no convergence to the box scale
+It is the $j = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that domiantes the shear spectrum $|\hatwexbkvec|$ in the latter two phases (not shown), demonstrating a competition between the $j = 3$ and $j = 4$ modes.
+% second important point: no secular growth of the box scale zonal mode
+Most importantly, no secular growth of the $j = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [dotted line in panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}]].
+% summary
+The above discussion indicates that although the $j = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27~\rhothref$ (i.~e., $j = 4,~3$ in the $3\times1$ case).
+\\
+
+% what kind of analysis
+Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simulataneously.
+This scan covers the realizations $\NR\times\NB \in [1\times1,~2\times2,~3\times3]$.
+% mention shorter times scales for stabilization
+Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~\Rref/\vthref$ always (see Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}), i.~e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
+% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
+%
+% the main result of this test: confirmation of wave length convergence
+As shown in Fig.~\ref{fig:wexb-3x3-stable} also this test confirms the convergence of the staircase pattern wavelength to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+% discuss sumewhat smaller wavelength
+By contrast to the radial box size scan, however, here the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.~e., a somewhat smaller pattern wavelength.
+Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]~\rhothref$ as observed in the first test is adressed in the next paragraph.
+\bigskip
+
+% what kind of analysis
+In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
+This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
+% -> outcome?
+\bigskip
+
+% what kind of test
+In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
+Since suppression of turbulence usually occurs at later times when approaching the finite heat flux treshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
+This scan covers realizations with $\rlt \in [6.0,~6.2,~6.4]$.
+% what is the finite heat flux threshold
+In the case of $\rlt = ???$ turbulence suppression is observed for $t > ???$, while stationary turbulence during the entire simulation time trace of $???~\Rref/\vthref$ is found for $\rlt = ???$.
+The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = ??? \pm ???$.
+% Which stationay wavelength occurs????
+Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.0$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $j = ???$ establishes.
+% Does it confirm our previous tests
+Again, the $j = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
+Also this test confirmes the statistical soundness of the converged pattern wavelength of $\sim ???$.
+
+\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial and binormal increased box sizes
 }{fig:eflux-1x1-2x2-3x3-comparison}
 
-\includegraphicsOneCol{S6_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x1_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Shearing rate $\omega_{\mathrm{E \times B}}$ for different time intervals in which heat conduction is almost zero but the E$\times$B staircase has not fully developed for boxsize 3$\times$1
-}{fig:wexb-3x1-selection}
-
-\includegraphicsOneCol{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection.pdf}{
-	Stablized shearing rate $\omega_{\mathrm{E \times B}}$ for boxsize 3$\times$3
+\includegraphicsOneCol{{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection}.pdf}{
+	Stablized shearing rate $\omega_{\mathrm{\:E \times B}}$ for boxsize 3$\times$3. \textcolor{red}{Add $1\times1$, $2\times2$!}
 }{fig:wexb-3x3-stable}
-
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize2x1-2-3x1-3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
-	Comparison of time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for boxsize 2$\times$1 compared to 2$\times$2 and 3$\times$1 compared to 3$\times$3
-}{fig:eflux-2x1-2-3x1-3-comparison}
-
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}{
-	Comparison of shearing rate $\omega_{\mathrm{E \times B}}$ for radial increased boxsizes. The staircase structure got 
-	shifted for better visibility.
-}{fig:wexb-1-2-3-4x1-stable-comparison}
-
-\includegraphicsTwoCol{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased boxsizes \linebreak
-	\textbf{(b)} Time traces of $|\hat{\omega}_{\mathrm{E\times B}}|_{k_\mathrm{i}}$ for radial increased boxsizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}
-
-\increasecounter{fig}{2}
 
@@
 
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+
+% what has been done
+Through careful tests this briefcommunication confirms the wavelength convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
+% mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
+A mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rhothref$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
+% compare to other studies
+This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10^1~\rhothref$ \cite{pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
+% relevance of the outcome
+The occurrence of this mesoscale implies that non-locality is inherent to ITG driven turbulence, since avalanches are usually spatially organized by the $\exb$ staircase pattern \cite{mcmillan2009, pradalier2010, rath2016, peeters2016}.
+
+%% APPENDIX ================================================================================================================
+
 \section*{Data Availability}
 The data that support the findings of this study are available from the corresponding author upon reasonable request. 
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
+\bigskip
 
@@
 
-%\wordcountmessage
+\wordcountmessage
 

### 2023-02-21 briefcommunicaton/briefcommunication.tex
#37: widehat for FT{wexb}

@@
 \newcommand{\exb}{\mathrm{\:E \times B}}
-\newcommand{\hatwexbkvec}{\hat{\omega}_{\mathrm{\:E \times B},\mathbf{k}}}
+\newcommand{\hatwexbkvec}{\widehat{\omega}_{\mathrm{\:E \times B},\mathbf{k}}}
 \newcommand{\NR}{N_\mathrm{R}}
@@
 \begin{equation}
-	\omega_{\mathrm{\:E \times B}} = \sum_{k_j} \hat \omega_{\mathrm{\:E \times B}}(k_j,t) \times \exp(\mathrm{i} k_j \psi),
+	\omega_{\mathrm{\:E \times B}} = \sum_{k_j} \widehat \omega_{\mathrm{\:E \times B}}(k_j,t) \times \exp(\mathrm{i} k_j \psi),
 	\label{eq:shearingrate_fourier}
@@
 \increasecounter{eq}{1}{1}
-were $\hat \omega_{\mathrm{\:E \times B}}$ is the complex Fourier coefficient and 
+were $\widehat \omega_{\mathrm{\:E \times B}}$ is the complex Fourier coefficient and 
 % use j as zonal flow mode number to distinguish it from the imaginary unit i
@@
 % define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the defintions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{j}} = 2 |\hat \omega_{\mathrm{\:E \times B}}|$.
+Based on the defintions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{j}} = 2 |\widehat \omega_{\mathrm{\:E \times B}}|$.
 % introduces the term basic mode of the pattern for later usage

### 2023-02-26 briefcommunicaton/briefcommunication.tex
#37: Changed binormal heatflux plot
#37: New plots in paper

@@
 
-% \increasecounter{fig}{1}
-
-\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
+\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
 	Comparison of shearing rate $\omega_{\mathrm{\:E \times B}}$ for radial increased box sizes. The staircase structure got 
@@
 
-\includegraphicsOneCol{{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection}.pdf}{
-	Stablized shearing rate $\omega_{\mathrm{\:E \times B}}$ for boxsize 3$\times$3. \textcolor{red}{Add $1\times1$, $2\times2$!}
-}{fig:wexb-3x3-stable}
+\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for binormal increased box sizes
+}{fig:eflux-1x1-2x2-3x3-comparison}
+
+%\includegraphicsOneCol{{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection}.pdf}{
+%	Stablized shearing rate $\omega_{\mathrm{\:E \times B}}$ for boxsize 3$\times$3. \textcolor{red}{Add $1\times1$, $2\times2$!}
+%}{fig:wexb-3x3-stable}
 

### 2023-03-19 briefcommunicaton/briefcommunication.tex
#37: removed _ref, renamed correction

@@
 %% Include graphic for two column with specific place, figrue* places graphic anywhere...
-\newcommand{\includegraphicsTwoCol}[4]{
-	\onecolumngrid
-	\begin{center}
-		\captionsetup{type=figure}
-    	\includegraphics[width=#4\textwidth]{#1}
-		\captionof{figure}{#2}
+%\newcommand{\includegraphicsTwoCol}[4]{
+%	\onecolumngrid
+%	\begin{center}
+%		\captionsetup{type=figure}
+%    	\includegraphics[width=#4\textwidth]{#1}
+%		\captionof{figure}{#2}
+%    	\label{#3}
+%	\end{center}
+%	\twocolumngrid
+%	\increasecounter{fig}{2}
+%}
+
+\newcommand{\includegraphicsTwoCol}[3]{
+	\begin{figure*}
+    	\includegraphics[width=0.9\textwidth]{#1}
+		\caption{#2}
 		\label{#3}
-	\end{center}
-	\twocolumngrid
+	\end{figure*}
 	\increasecounter{fig}{2}
 }
-
-%\newcommand{\includegraphicsTwoCol}[3]{
-%	\begin{figure*}
-%    	\includegraphics[width=0.9\textwidth]{#1}
-%		\caption{#2}
-%		\label{#3}
-%	\end{figure*}
-%	\increasecounter{fig}{2}
-%}
 
@@
 \newcommand{\NB}{N_\mathrm{B}}
-\newcommand{\vthref}{v_\mathrm{th,ref}}
-\newcommand{\Rref}{R_\mathrm{ref}}
 \newcommand{\rlt}{R/L_T}
-\newcommand{\rhothref}{\rho_\mathrm{th,ref}}
 
@@
     The wavelength convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-    Its is shown that a mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rhothref$ is inherent to ITG driven turbulence with Cyclone Base Case paramters in the local limit.
+    Its is shown that a mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rho_\mathrm{th}$ is inherent to ITG driven turbulence with Cyclone Base Case paramters in the local limit.
 
@@
 It has been carefully verified that this modification preserves the same physical outcome as the orginial study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.
-\\
-The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)~\rho_\mathrm{th,ref}$, in the radial and binormal direction is always indicated by the real parametes $N_\mathrm{R}$ and $N_\mathrm{B}$, respectively.
-Here, $\rho_\mathrm{th,ref}$ is the reference thermal Larmor radius and the nomenclature $N_\mathrm{R}\times N_\mathrm{B}$ is applied throughout this work to indicate the increase in box size.  
-Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution (Tab.~\ref{tab:resolution}) and standard box size.
-\bigskip
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.\\
 
 %%% DIAGNOSTICS ================================================================================================================
+
 The ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern is manifest as radial structure formation in the ${\mathrm{\:E}\times\mathrm{B}}$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
@@
 \increasecounter{eq}{1}{1}
-where $\langle \Phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T_\mathrm{ref}/e$ ($\rho_\ast = \rho_\mathrm{th,ref}/R_\mathrm{ref}$ is the reference thermal Larmor radius normalized with the reference major radius $R_\mathrm{ref}$, $T_\mathrm{ref}$ is a reference temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rho_\mathrm{th,ref}$.
+where $\langle \Phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rho/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is a temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rho$.
+Furthermore quantities $\rho$, $R$, $T$ are referenced from Ref (\textbf{?}).
+
 The E$\times$B shearing rate $\omega_{\mathrm{\:E \times B}}$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
@@
 In general, the zonal flow mode that dominates the ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern, \textemdash also referred to as the \textit{basic mode } of the pattern in this work \textemdash exhibits the maximum amplitude in the spectrum $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{j}}$.
-Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rho_\mathrm{th,ref}^2 v_{\mathrm{th,ref}}/R_\mathrm{ref}$ ($v_{\mathrm{th,ref}} = \sqrt{2 T_\mathrm{ref}/m_\mathrm{ref}}$ is the reference thermal velocity and $m_\mathrm{ref}$ is the reference mass) throughout this work.
+
+The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)\,\rho$, in the radial and binormal direction is always indicated by the real parametes $N_\mathrm{R}$ and $N_\mathrm{B}$, respectively.
+Here, the nomenclature $N_\mathrm{R}\times N_\mathrm{B}$ is applied throughout this work to indicate the increase in box size.  
+Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution (Tab.~\ref{tab:resolution}) and standard box size.
+
+Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rho^2 v_{\mathrm{th}}/R$ ($v_{\mathrm{th}} = \sqrt{2 T/m}$ is the reference thermal velocity and $m$ is the reference mass) throughout this work.
 \bigskip
@@
 % very long initial quasi-stationary states
-Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4~\Rref/\vthref$.
+Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4~R/v_\mathrm{th}$.
 % the j=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
-During this period the zonal flow mode with $j = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3~\Rref/\vthref$.
-Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000~\Rref/\vthref$ and $t \approx 18000~\Rref/\vthref$.
+During this period the zonal flow mode with $j = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3~R/v_\mathrm{th}$.
+Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000~R/v_\mathrm{th}$ and $t \approx 18000~R/v_\mathrm{th}$.
 % first important point: competition between the j = 3 and j = 4 mode -> the j = 4 mode is of shorter length scale compared to the j = 3 mode, hence, no convergence to the box scale
-It is the $j = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that domiantes the shear spectrum $|\hatwexbkvec|$ in the latter two phases (not shown), demonstrating a competition between the $j = 3$ and $j = 4$ modes.
+It is the $j = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that domiantes the shear spectrum $|\widehat{\omega}_{\mathrm{\:E \times B}}|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $j = 3$ and $j = 4$ modes.
 % second important point: no secular growth of the box scale zonal mode
@@
 % summary
-The above discussion indicates that although the $j = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27~\rhothref$ (i.~e., $j = 4,~3$ in the $3\times1$ case).
+The above discussion indicates that although the $j = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27~\rho_\mathrm{th}$ (i.~e., $j = 4,~3$ in the $3\times1$ case).
 \\
@@
 % mention shorter times scales for stabilization
-Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~\Rref/\vthref$ always (see Fig.
+Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~R/v_\mathrm{th}$ always (see Fig.
 \ref{fig:eflux-1x1-2x2-3x3-comparison}), i.~e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
@@
 By contrast to the radial box size scan, however, here the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.~e., a somewhat smaller pattern wavelength.
-Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]~\rhothref$ as observed in the first test is adressed in the next paragraph.
+Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]~\rho_\mathrm{th}$ as observed in the first test is adressed in the next paragraph.
 \bigskip
@@
 % what is the finite heat flux threshold
-In the case of $\rlt = ???$ turbulence suppression is observed for $t > ???$, while stationary turbulence during the entire simulation time trace of $???~\Rref/\vthref$ is found for $\rlt = ???$.
+In the case of $\rlt = ???$ turbulence suppression is observed for $t > ???$, while stationary turbulence during the entire simulation time trace of $???~R/v_\mathrm{th}$ is found for $\rlt = ???$.
 The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = ??? \pm ???$.
@@
 % mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
-A mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rhothref$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
+A mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rho_\mathrm{th}$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
 % compare to other studies
-This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10^1~\rhothref$ \cite{pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
+This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10^1~\rho_\mathrm{th}$ \cite{pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
 % relevance of the outcome

### 2023-03-19 briefcommunicaton/briefcommunication.tex
#37: fixed ref, formula commands and half space between number and formula

@@
 \newcommand{\wexb}{\omega_{\mathrm{\:E \times B}}}
-\newcommand{\exb}{\mathrm{\:E \times B}}
+\newcommand{\hatwexb}{\widehat{\omega}_{\mathrm{\:E \times B}}}
+\newcommand{\exb}{\mathrm{\:E}\times\mathrm{B}}
 \newcommand{\hatwexbkvec}{\widehat{\omega}_{\mathrm{\:E \times B},\mathbf{k}}}
@@
 \newcommand{\rlt}{R/L_T}
+\newcommand{\rhoth}{\rho_\mathrm{th}}
+\newcommand{\vth}{v_{\mathrm{th}}}
 
@@
 
-\title[Wavelength convergence of the ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
-{Wavelength convergence of the ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
+\title[Wavelength convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
+{Wavelength convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
 
@@
     The wavelength convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-    Its is shown that a mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rho_\mathrm{th}$ is inherent to ITG driven turbulence with Cyclone Base Case paramters in the local limit.
+    Its is shown that a mesoscale pattern wavelength of \linebreak $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
 
@@
 
-Ion temperature gradient driven turbulence close to marginal stability often exhibits zonal flow pattern formation on mesoscales \textemdash so-called ${\mathrm{\:E}\times\mathrm{B}}$ staircase structures \cite{pradalier2010}.
+Ion temperature gradient driven turbulence close to marginal stability often exhibits zonal flow pattern formation on mesoscales \textemdash so-called $\exb$ staircase structures \cite{pradalier2010}.
 Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{peeters2016, weikl2017, rath2021} as well as global gradient-driven \cite{mcmillan2009, villard2013, seo2022} and global flux-driven \cite{pradalier2010, pradalier2015, wang2020, kim2022, kishimoto2023} studies.
@@
 % By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
-By contrast, in the aformentioned local studies the wavelength of ${\mathrm{\:E}\times\mathrm{B}}$ staircase structures is always found to converge to the radial box size of the flux tube domain.
+By contrast, in the aforementioned local studies the wavelength of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
 The above observations lead to the question: 
 Does the basic pattern wavelength always converges to the box size, or is there a typical mesoscale wavelength inherent to staircase structures also in a local flux-tube description?
-The latter case would imply that it is not necessarily global physics, i.e., profile effects, that sets (i) the wavelength of the ExB staircase pattern and (ii) the scale of avalanche-like transport events, usually restricted to ${\mathrm{\:E}\times\mathrm{B}}$ staircase structures and considered as a nonlocal transport mechnaism\cite{pradalier2010}.
+The latter case would imply that it is not necessarily global physics, i.e., profile effects, that sets (i) the wavelength of the ExB staircase pattern and (ii) the scale of avalanche-like transport events, usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism\cite{pradalier2010}.
 In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\cite{peeters2016}.\bigskip
@@
 
+Here, gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} in the adiabatic electron approximation.
+In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $R/L_T = 6.0$ (unless stated otherwise) and circular concentric flux surfaces.
+The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
+It has been carefully verified that this modification preserves the same physical outcome as the orginial study.
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.
 \begin{table}[ht]
@@
 	\caption{
-		Resolution used in this paper for further information read Ref.~\cite{peeters2016}.
+		Resolution used in this paper for further information the author links to Ref.~\cite{peeters2016}.
 %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
@@
 	\label{tab:resolution}
-\end{table}
-
-Here, gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} in the adiabatic electron approximation.
-In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $R/L_T = 6.0$ (unless stated otherwise) and circular concentric flux surfaces.
-The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
-It has been carefully verified that this modification preserves the same physical outcome as the orginial study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.\\
+\end{table}\bigskip
 
@@
 
-The ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern is manifest as radial structure formation in the ${\mathrm{\:E}\times\mathrm{B}}$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
+The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
 \begin{equation}
-	\omega_{\mathrm{\:E \times B}} = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2},
+	\wexb = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2},
 	\label{eq:shearingrate}
@@
 \increasecounter{eq}{1}{1}
-where $\langle \Phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rho/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is a temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rho$.
-Furthermore quantities $\rho$, $R$, $T$ are referenced from Ref (\textbf{?}).
+where $\langle \Phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is a temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rhoth$.
 
-The E$\times$B shearing rate $\omega_{\mathrm{\:E \times B}}$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
+The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
 Here, the zonal potential is estimated from the electrostatic potential $\phi$ on the two dimensional $\psi$-$\zeta$-plane at the low field side according to
@@
 \end{equation}
-In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the ${\mathrm{\:E}\times\mathrm{B}}$ shearing rate is considered.
+In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
 It is defined by
 \begin{equation}
-	\omega_{\mathrm{\:E \times B}} = \sum_{k_j} \widehat \omega_{\mathrm{\:E \times B}}(k_j,t) \times \exp(\mathrm{i} k_j \psi),
+	\wexb = \sum_{k_j} \hatwexb(k_j,t) \times \exp(\mathrm{i} k_j \psi),
 	\label{eq:shearingrate_fourier}
@@
 \increasecounter{eq}{1}{1}
-were $\widehat \omega_{\mathrm{\:E \times B}}$ is the complex Fourier coefficient and 
+were $\hatwexb$ is the complex Fourier coefficient and 
 % use j as zonal flow mode number to distinguish it from the imaginary unit i
@@
 % define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the defintions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{j}} = 2 |\widehat \omega_{\mathrm{\:E \times B}}|$.
+Based on the defintions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\hatwexb|_{k_\mathrm{j}} = 2 |\hatwexb|$. 
 % introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the ${\mathrm{\:E}\times\mathrm{B}}$ staircase pattern, \textemdash also referred to as the \textit{basic mode } of the pattern in this work \textemdash exhibits the maximum amplitude in the spectrum $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{j}}$.
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, \textemdash also referred to as the \textit{basic mode } of the pattern in this work \textemdash exhibits the maximum amplitude in the spectrum $|\hatwexb|_{k_\mathrm{j}}$.
 
-The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)\,\rho$, in the radial and binormal direction is always indicated by the real parametes $N_\mathrm{R}$ and $N_\mathrm{B}$, respectively.
-Here, the nomenclature $N_\mathrm{R}\times N_\mathrm{B}$ is applied throughout this work to indicate the increase in box size.  
+The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parametes $\NR$ and $\NB$, respectively.
+Here, the nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
 Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution (Tab.~\ref{tab:resolution}) and standard box size.
 
-Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rho^2 v_{\mathrm{th}}/R$ ($v_{\mathrm{th}} = \sqrt{2 T/m}$ is the reference thermal velocity and $m$ is the reference mass) throughout this work.
-\bigskip
+Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass) throughout this work.
+Furthermore quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref (\textbf{?}).\bigskip
 
@@
 	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased box sizes
-	\textbf{(b)} Time traces of $|\widehat{\omega}_{\mathrm{\:E\times B}}|_{k_\mathrm{i}}$ for radial increased box sizes
+	\textbf{(b)} Time traces of $|\hatwexb|_{k_\mathrm{i}}$ for radial increased box sizes
 }{fig:wexb-eflux-1-2-3-4x1-comparison}{}
@@
 \includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
-	Comparison of shearing rate $\omega_{\mathrm{\:E \times B}}$ for radial increased box sizes. The staircase structure got 
+	Comparison of shearing rate $\wexb$ for radial increased box sizes. The staircase structure got 
 	shifted for better visibility.
@@
 % very long initial quasi-stationary states
-Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4~R/v_\mathrm{th}$.
+Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
 % the j=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
-During this period the zonal flow mode with $j = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3~R/v_\mathrm{th}$.
-Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000~R/v_\mathrm{th}$ and $t \approx 18000~R/v_\mathrm{th}$.
+During this period the zonal flow mode with $j = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
+Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
 % first important point: competition between the j = 3 and j = 4 mode -> the j = 4 mode is of shorter length scale compared to the j = 3 mode, hence, no convergence to the box scale
@@
 % summary
-The above discussion indicates that although the $j = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27~\rho_\mathrm{th}$ (i.~e., $j = 4,~3$ in the $3\times1$ case).
+The above discussion indicates that although the $j = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.~e., $j = 4,~3$ in the $3\times1$ case).
 \\
@@
 % mention shorter times scales for stabilization
-Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~R/v_\mathrm{th}$ always (see Fig.
+Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~R/\vth$ always (see Fig.
 \ref{fig:eflux-1x1-2x2-3x3-comparison}), i.~e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
@@
 By contrast to the radial box size scan, however, here the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.~e., a somewhat smaller pattern wavelength.
-Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]~\rho_\mathrm{th}$ as observed in the first test is adressed in the next paragraph.
+Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is adressed in the next paragraph.
 \bigskip
@@
 % what is the finite heat flux threshold
-In the case of $\rlt = ???$ turbulence suppression is observed for $t > ???$, while stationary turbulence during the entire simulation time trace of $???~R/v_\mathrm{th}$ is found for $\rlt = ???$.
+In the case of $\rlt = ???$ turbulence suppression is observed for $t > ???$, while stationary turbulence during the entire simulation time trace of $???~R/\vth$ is found for $\rlt = ???$.
 The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = ??? \pm ???$.
@@
 
-\bigskip
-\bigskip
-\bigskip
-\bigskip
-\bigskip
-\bigskip
-
 % what has been done
@@
 % mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
-A mesoscale pattern wavelength of $\sim 57.20 - 76.27~\rho_\mathrm{th}$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
+A mesoscale pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
 % compare to other studies
-This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10^1~\rho_\mathrm{th}$ \cite{pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
+This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
 % relevance of the outcome
@@
 The data that support the findings of this study are available from the corresponding author upon reasonable request.
-\bigskip
-\bigskip
-\bigskip
-\bigskip
-\bigskip
-\bigskip
 
@@
 
-\nocite{}
+%\nocite{}
 \bibliography{references.bib}% Produces the bibliography via BibTeX.

### 2023-03-19 briefcommunicaton/briefcommunication.tex
#37: correction grammar, spelling

@@
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
-It has been carefully verified that this modification preserves the same physical outcome as the orginial study.
+It has been carefully verified that this modification preserves the same physical outcome as the original study.
 A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.
@@
 The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
-Here, the zonal potential is estimated from the electrostatic potential $\phi$ on the two dimensional $\psi$-$\zeta$-plane at the low field side according to
+Here, the zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\psi$-$\zeta$-plane at the low field side according to
 \begin{equation}
@@
 % define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the defintions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\hatwexb|_{k_\mathrm{j}} = 2 |\hatwexb|$. 
+Based on the definitions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\hatwexb|_{k_\mathrm{j}} = 2 |\hatwexb|$. 
 % introduces the term basic mode of the pattern for later usage
@@
 
-The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parametes $\NR$ and $\NB$, respectively.
+The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parameter $\NR$ and $\NB$, respectively.
 Here, the nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
@@
 Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass) throughout this work.
-Furthermore quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref (\textbf{?}).\bigskip
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref (\textbf{?}).\bigskip
 
@@
 % short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost supressed turbulence (see Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}).
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence (see Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}).
 % introduce the fully developed staircase structure
@@
 % first important point: competition between the j = 3 and j = 4 mode -> the j = 4 mode is of shorter length scale compared to the j = 3 mode, hence, no convergence to the box scale
-It is the $j = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that domiantes the shear spectrum $|\widehat{\omega}_{\mathrm{\:E \times B}}|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $j = 3$ and $j = 4$ modes.
+It is the $j = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that dominates the shear spectrum $|\widehat{\omega}_{\mathrm{\:E \times B}}|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $j = 3$ and $j = 4$ modes.
 % second important point: no secular growth of the box scale zonal mode
@@
 % what kind of analysis
-Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simulataneously.
+Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
 This scan covers the realizations $\NR\times\NB \in [1\times1,~2\times2,~3\times3]$.
@@
 % discuss sumewhat smaller wavelength
-By contrast to the radial box size scan, however, here the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.~e., a somewhat smaller pattern wavelength.
-Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is adressed in the next paragraph.
+By contrast to the radial box size scan, however, here the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern wavelength.
+Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
 \bigskip
@@
 In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
-Since suppression of turbulence usually occurs at later times when approaching the finite heat flux treshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
+Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
 This scan covers realizations with $\rlt \in [6.0,~6.2,~6.4]$.
@@
 Again, the $j = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also this test confirmes the statistical soundness of the converged pattern wavelength of $\sim ???$.
+Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim ???$.
 
@@
 % what has been done
-Through careful tests this briefcommunication confirms the wavelength convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
+Through careful tests this brief communication confirms the wavelength convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
 % mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)

### 2023-03-19 briefcommunicaton/briefcommunication.tex
#37: Readjust some parts of introduction

@@
 
-Ion temperature gradient driven turbulence close to marginal stability often exhibits zonal flow pattern formation on mesoscales \textemdash so-called $\exb$ staircase structures \cite{pradalier2010}.
+Ion temperature gradient driven turbulence close to marginal stability often exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{pradalier2010}.
 Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{peeters2016, weikl2017, rath2021} as well as global gradient-driven \cite{mcmillan2009, villard2013, seo2022} and global flux-driven \cite{pradalier2010, pradalier2015, wang2020, kim2022, kishimoto2023} studies.
@@
 The above observations lead to the question: 
-Does the basic pattern wavelength always converges to the box size, or is there a typical mesoscale wavelength inherent to staircase structures also in a local flux-tube description?
-The latter case would imply that it is not necessarily global physics, i.e., profile effects, that sets (i) the wavelength of the ExB staircase pattern and (ii) the scale of avalanche-like transport events, usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism\cite{pradalier2010}.
+\textit{Does the basic pattern wavelength always converges to the box size, or is there a typical mesoscale wavelength inherent to staircase structures also in a local flux-tube description?}
+The latter case would imply that it is not necessarily global physics which means profile effects are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism.
+This profile effects set (i) the wavelength of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events \cite{pradalier2010}.
 In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\cite{peeters2016}.\bigskip

### 2023-03-19 briefcommunicaton/briefcommunication.tex
#37: Correction simulation set up

@@
 Here, gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} in the adiabatic electron approximation.
-In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $R/L_T = 6.0$ (unless stated otherwise) and circular concentric flux surfaces.
+In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $R/L_T = 6.0$ and circular concentric flux surfaces.
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.

### 2023-03-19 briefcommunicaton/briefcommunication.tex
#37: Hamada coord with explanation, Readjust diagnostics, N_\psi and N_\zeta, \rho_th to \rho

@@
 \newcommand{\hatwexbkvec}{\widehat{\omega}_{\mathrm{\:E \times B},\mathbf{k}}}
-\newcommand{\NR}{N_\mathrm{R}}
-\newcommand{\NB}{N_\mathrm{B}}
+%\newcommand{\NR}{N_\mathrm{R}}
+%\newcommand{\NB}{N_\mathrm{B}}
+\newcommand{\NR}{N_\psi}
+\newcommand{\NB}{N_\zeta}
 \newcommand{\rlt}{R/L_T}
-\newcommand{\rhoth}{\rho_\mathrm{th}}
+%\newcommand{\rhoth}{\rho_\mathrm{th}}
+\newcommand{\rhoth}{\rho}
 \newcommand{\vth}{v_{\mathrm{th}}}
@@
 \begin{equation}
-	\wexb = \frac{1}{2} \frac{\partial^2 \langle \Phi \rangle}{\partial \psi^2},
+	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \psi^2},
 	\label{eq:shearingrate}
@@
 \increasecounter{eq}{1}{1}
-where $\langle \Phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is a temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rhoth$.
+where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rhoth$.
+The zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\psi$-$\zeta$-plane at the low field side according to
+\begin{equation}
+\langle \phi \rangle = \frac{1}{l_\zeta} \int_0^{l_\zeta} \mathrm{d}\zeta ~ \phi(\psi,\zeta,s=0),
+\end{equation}
+where $\zeta$ labels the field lines and is an approximate binormal or toroidal coordinate, and $s$ parameterizes the length along the field lines and represents a poloidal coordinate which represents together with the radial coordinate $\psi$ the Hamada coordinates $(\psi,\zeta,s)$.
+\cite{rath2021} (\textbf{REFERENCE Hamada})
+The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
 
-The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
-Here, the zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\psi$-$\zeta$-plane at the low field side according to
-\begin{equation}
-\langle \Phi \rangle = \frac{1}{l_\zeta} \int_0^{l_\zeta} \mathrm{d}\zeta ~ \phi(\psi,\zeta,s=0).
-\end{equation}
 In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
@@
 \begin{equation}
-	\wexb = \sum_{k_j} \hatwexb(k_j,t) \times \exp(\mathrm{i} k_j \psi),
+	\wexb = \sum_{k_j} \hatwexb(k_j,t) \, \exp(\mathrm{i} k_j \psi),
 	\label{eq:shearingrate_fourier}
@@
 \increasecounter{eq}{1}{1}
-were $\hatwexb$ is the complex Fourier coefficient and 
-% use j as zonal flow mode number to distinguish it from the imaginary unit i
-\begin{equation}
-	k_\mathrm{j} = \frac{2\pi \mathrm{j}}{l_\psi}
-\end{equation}
-\increasecounter{eq}{1}{1}
+where $\hatwexb$ is the complex Fourier coefficient and \linebreak $k_\mathrm{j} = 2\pi \mathrm{j}/l_\psi$
 % explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
@@
 Based on the definitions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\hatwexb|_{k_\mathrm{j}} = 2 |\hatwexb|$.
+
 % introduces the term basic mode of the pattern for later usage
@@
 Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass) throughout this work.
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref (\textbf{?}).\bigskip
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref (\textbf{REFERENCE Quantities}).\bigskip
 

### 2023-03-19 briefcommunicaton/briefcommunication.tex
#37: removed \textemdash

@@
 % introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, \textemdash also referred to as the \textit{basic mode } of the pattern in this work \textemdash exhibits the maximum amplitude in the spectrum $|\hatwexb|_{k_\mathrm{j}}$.
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work,exhibits the maximum amplitude in the spectrum $|\hatwexb|_{k_\mathrm{j}}$.
 

### 2023-03-20 briefcommunicaton/briefcommunication.tex
#37: Placement graphics, Rearrangements of Diagnostics, j to n_ZF and k_j to k_ZF

@@
 \usepackage{hyperref}
+\usepackage{float}
 
@@
 \newcommand{\includegraphicsOneCol}[3]{
-	\begin{figure}[ht]
-		\includegraphics[width=0.90\linewidth]{#1}
+	\begin{figure}[hbt!]
+		\centering
+		\includegraphics[width=0.9\linewidth]{#1}
 		\caption{#2}
@@
 %% Include graphic for two column with specific place, figrue* places graphic anywhere...
-%\newcommand{\includegraphicsTwoCol}[4]{
-%	\onecolumngrid
-%	\begin{center}
-%		\captionsetup{type=figure}
-%    	\includegraphics[width=#4\textwidth]{#1}
-%		\captionof{figure}{#2}
+\newcommand{\includegraphicsTwoCol}[4]{
+	\onecolumngrid
+	\begin{center}
+		\centering
+		\captionsetup{type=figure}
+    	\includegraphics[width=#4\textwidth]{#1}
+		\captionof{figure}{#2}
+    	\label{#3}
+	\end{center}
+	\twocolumngrid
+	\increasecounter{fig}{2}
+}
+
+%\newcommand{\includegraphicsTwoCol}[3]{
+%	\begin{figure*}
+%    	\includegraphics[width=0.9\textwidth]{#1}
+%		\caption{#2}
 %		\label{#3}
-%	\end{center}
-%	\twocolumngrid
+%	\end{figure*}
 %	\increasecounter{fig}{2}
 %}
-
-\newcommand{\includegraphicsTwoCol}[3]{
-	\begin{figure*}
-    	\includegraphics[width=0.9\textwidth]{#1}
-		\caption{#2}
-		\label{#3}
-	\end{figure*}
-	\increasecounter{fig}{2}
-}
 
@@
 \newcommand{\vth}{v_{\mathrm{th}}}
+\newcommand{\nzf}{n_\mathrm{ZF}}
+\newcommand{\kzf}{k_\mathrm{ZF}}
 
@@
 Here, gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} in the adiabatic electron approximation.
-In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $R/L_T = 6.0$ and circular concentric flux surfaces.
+In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
@@
 	\label{tab:resolution}
-\end{table}\bigskip
+\end{table}
 
 %%% DIAGNOSTICS ================================================================================================================
+
+The increase of the box size, relative to the standard box size $(L_\psi,~L_\zeta) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parameter $\NR$ and $\NB$, respectively.
+Here, the nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
+Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution (Tab.~\ref{tab:resolution}) and standard box size.
+\bigskip
 
@@
 where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rhoth$.
+Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref(\textbf{?}).
 The zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\psi$-$\zeta$-plane at the low field side according to
 \begin{equation}
-\langle \phi \rangle = \frac{1}{l_\zeta} \int_0^{l_\zeta} \mathrm{d}\zeta ~ \phi(\psi,\zeta,s=0),
+\langle \phi \rangle = \frac{1}{L_\zeta} \int_0^{L_\zeta} \mathrm{d}\zeta ~ \phi(\psi,\zeta,s=0),
 \end{equation}
+\increasecounter{eq}{1}{1}
 where $\zeta$ labels the field lines and is an approximate binormal or toroidal coordinate, and $s$ parameterizes the length along the field lines and represents a poloidal coordinate which represents together with the radial coordinate $\psi$ the Hamada coordinates $(\psi,\zeta,s)$.
-\cite{rath2021} (\textbf{REFERENCE Hamada})
+\cite{rath2021}(\textbf{?})
 The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
@@
 \begin{equation}
-	\wexb = \sum_{k_j} \hatwexb(k_j,t) \, \exp(\mathrm{i} k_j \psi),
+	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \psi),
 	\label{eq:shearingrate_fourier}
@@
 \increasecounter{eq}{1}{1}
-where $\hatwexb$ is the complex Fourier coefficient and \linebreak $k_\mathrm{j} = 2\pi \mathrm{j}/l_\psi$
+where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\psi$
 % explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
-defines the zonal flow wave vector with the zonal flow mode number $j$ ranging in $-(N_\psi -1)/2 \leq j \leq (N_\psi -1)/2 $.
+defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\psi -1)/2 \leq \nzf \leq (N_\psi -1)/2 $.
 % define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the definitions above, the shear carried by the zonal flow mode with wave vector $k_j$ is defined by $|\hatwexb|_{k_\mathrm{j}} = 2 |\hatwexb|$.
-
+Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $|\hatwexb|_{\kzf} = 2 |\hatwexb|$. 
 % introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work,exhibits the maximum amplitude in the spectrum $|\hatwexb|_{k_\mathrm{j}}$.
-
-The increase of the box size, relative to the standard box size $(l_\psi,~l_\zeta) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parameter $\NR$ and $\NB$, respectively.
-Here, the nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
-Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution (Tab.~\ref{tab:resolution}) and standard box size.
-
-Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass) throughout this work.
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref (\textbf{REFERENCE Quantities}).\bigskip
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work,exhibits the maximum amplitude in the spectrum $|\hatwexb|_{\kzf}$.
+\bigskip
 
@@
 %\label{sec:results}
-
-\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial increased box sizes
-	\textbf{(b)} Time traces of $|\hatwexb|_{k_\mathrm{i}}$ for radial increased box sizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
-
-\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
-	Comparison of shearing rate $\wexb$ for radial increased box sizes. The staircase structure got 
-	shifted for better visibility.
-}{fig:wexb-1-2-3-4x1-stable-comparison}{}
 
@@
 % introduce the fully developed staircase structure
-The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-1-2-3-4x1-stable-comparison}. 
+The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
 This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable liner ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
 % insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
-Fig.~\ref{fig:wexb-1-2-3-4x1-stable-comparison} shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
+Fig.~\ref{fig:wexb-stable-comparison} shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
 % yes, the wavelength converges; but even stronger result: the standard box size already allows for the correct staircase wavelength
@@
 % cautios words
-Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-1-2-3-4x1-stable-comparison} represent the real converged pattern in a statistical sense.
+Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
 % very long initial quasi-stationary states
 Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
-% the j=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
-During this period the zonal flow mode with $j = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
+% the \nzf=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
+During this period the zonal flow mode with $\nzf = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
 Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
-% first important point: competition between the j = 3 and j = 4 mode -> the j = 4 mode is of shorter length scale compared to the j = 3 mode, hence, no convergence to the box scale
-It is the $j = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that dominates the shear spectrum $|\widehat{\omega}_{\mathrm{\:E \times B}}|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $j = 3$ and $j = 4$ modes.
+% first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
+It is the $\nzf = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that dominates the shear spectrum $|\widehat{\omega}_{\mathrm{\:E \times B}}|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
 % second important point: no secular growth of the box scale zonal mode
-Most importantly, no secular growth of the $j = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [dotted line in panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}]].
+Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [dotted line in panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}]].
 % summary
-The above discussion indicates that although the $j = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.~e., $j = 4,~3$ in the $3\times1$ case).
+The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.~e., $\nzf = 4,~3$ in the $3\times1$ case).
 \\
@@
 % the main result of this test: confirmation of wave length convergence
-As shown in Fig.~\ref{fig:wexb-3x3-stable} also this test confirms the convergence of the staircase pattern wavelength to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+As shown in Fig.~\ref{fig:wexb-stable-comparison} also this test confirms the convergence of the staircase pattern wavelength to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+
+\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes
+	\textbf{(b)} Time traces of $|\hatwexb|_{\kzf}$ for radial increased box sizes
+}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
+
+\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
+	Comparison of shearing rate $\wexb$ for radial increased box sizes. The staircase structure got 
+	shifted for better visibility.
+}{fig:wexb-stable-comparison}{}
+
 % discuss sumewhat smaller wavelength
@@
 In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
-This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
+This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$. \bigskip
 % -> outcome?
-\bigskip
+
+\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial and binormal increased box sizes
+}{fig:eflux-1x1-2x2-3x3-comparison}
+
+\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
+}{fig:eflux-3x1.5-2.5-3-5-comparison}
 
@@
 % Which stationay wavelength occurs????
-Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.0$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $j = ???$ establishes.
+Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.0$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = ???$ establishes.
 % Does it confirm our previous tests
-Again, the $j = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim ???$.
-
-\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for radial and binormal increased box sizes
-}{fig:eflux-1x1-2x2-3x3-comparison}
-
-\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $R/L_T = 6.0$ for binormal increased box sizes
-}{fig:eflux-1x1-2x2-3x3-comparison}
-
-%\includegraphicsOneCol{{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_wexb_selection}.pdf}{
-%	Stablized shearing rate $\omega_{\mathrm{\:E \times B}}$ for boxsize 3$\times$3. \textcolor{red}{Add $1\times1$, $2\times2$!}
-%}{fig:wexb-3x3-stable}
+Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
+Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim ???$.\bigskip
 

### 2023-03-20 briefcommunicaton/briefcommunication.tex
#37: Binormal box scan, Fig captions, Ref in text,

@@
 
-\documentclass[aip, amsmath, amssymb, reprint, twocolumn]{revtex4-1}
+\documentclass[aip, amsmath, amssymb, reprint, twocolumn, floatfix]{revtex4-1}
 
@@
 %% Additional
-\usepackage{caption}
+\usepackage[format=hang]{caption}
 \usepackage{hyperref}
-\usepackage{float}
 
@@
 	\altaffiliation{Author to whom correspondence should be addressed: \\
-					Florian.Rath1@uni-bayreuth.de}
-	\email{Florian.Rath1@uni-bayreuth.de}
+					Florian.Rath@uni-bayreuth.de}
+	\email{Florian.Rath@uni-bayreuth.de}
 \author{A. G. Peeters}
@@
 Here, the nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
-Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution (Tab.~\ref{tab:resolution}) and standard box size.
+Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
 \bigskip
@@
 % short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence (see Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}).
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
 % introduce the fully developed staircase structure
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable liner ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
 % insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
-Fig.~\ref{fig:wexb-stable-comparison} shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
+Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
 % yes, the wavelength converges; but even stronger result: the standard box size already allows for the correct staircase wavelength
@@
 % cautios words
-Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
+Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
 % very long initial quasi-stationary states
@@
 % the \nzf=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
-During this period the zonal flow mode with $\nzf = 3$, i.~e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
+During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
 Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
 % first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
-It is the $\nzf = 4$ zonal flow mode, i.~e., the next shorter radial scale mode, that dominates the shear spectrum $|\widehat{\omega}_{\mathrm{\:E \times B}}|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
+It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $|\hatwexb|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
 % second important point: no secular growth of the box scale zonal mode
-Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [dotted line in panel (b) of Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}]].
+Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
 % summary
-The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.~e., $\nzf = 4,~3$ in the $3\times1$ case).
+The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
 \\
@@
 % mention shorter times scales for stabilization
-Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~R/\vth$ always (see Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}), i.~e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
+Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~R/\vth$ always [Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
 % Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
@@
 % the main result of this test: confirmation of wave length convergence
-As shown in Fig.~\ref{fig:wexb-stable-comparison} also this test confirms the convergence of the staircase pattern wavelength to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern wavelength to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 
 \includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
 	\textbf{(b)} Time traces of $|\hatwexb|_{\kzf}$ for radial increased box sizes
@@
 \includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
-	Comparison of shearing rate $\wexb$ for radial increased box sizes. The staircase structure got 
-	shifted for better visibility.
+	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
+The staircase structures got shifted till alignment for better visibility.\\
+	\begin{tabular}{l l l l l l l l l}
+		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$ & $\in [26000, 28000]$ \\
+		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
+		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [11000, 12000]$, & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$ & $\in [2000, 4000]$    &                 &                      \\
+	\end{tabular}
 }{fig:wexb-stable-comparison}{}
@@
 In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
-This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$. \bigskip
-% -> outcome?
+This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats always after $\sim 1000\,R/\vth$ although the turbulence of realization $3\times 1.5$ does suppress after $\sim 11000\,R/\vth$ [Fig.
+\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
+The convergence of staircase pattern can be seen in Fig \ref{fig:wexb-stable-comparison}\,(c) and the test confirms again the typical mesoscale distinct from $\NR > 1$.
+Fig.
+\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+The zonal flow mode number varies between $\nzf = [3,4]$ which can be seen in Fig.
+\ref{fig:wexb-stable-comparison}\,(c) for the realization $3\times 2.5$ where the staircase structure has an (PATTERN).
+The Fourier analysis yields in this case no definitely basic mode rather than two dominate modes with $\nzf = 3, 4$ which split the maximum amplitude $|\hatwexb|_{\kzf}$ between them.
 
@@
 
-%\nocite{}
+\nocite{}
 \bibliography{references.bib}% Produces the bibliography via BibTeX.

### 2023-03-21 briefcommunicaton/briefcommunication.tex
#37: Binormal scan paragraph, graphic placement, threshold analysis

@@
 \newcommand{\includegraphicsOneCol}[3]{
-	\begin{figure}[hbt!]
+	\begin{center}
 		\centering
+		\captionsetup{type=figure}
 		\includegraphics[width=0.9\linewidth]{#1}
-		\caption{#2}
+		\captionof{figure}{#2}
 		\label{#3}
-  	\end{figure}
+	\end{center}
   	\increasecounter{fig}{1}
 }
+
+%\newcommand{\includegraphicsOneCol}[3]{
+%	\begin{figure}[hbt!]
+%		\centering
+%		\includegraphics[width=0.9\linewidth]{#1}
+%		\caption{#2}
+%		\label{#3}
+%  	\end{figure}
+%  	\increasecounter{fig}{1}
+%}
 
@@
 %\newcommand{\NB}{N_\mathrm{B}}
-\newcommand{\NR}{N_\psi}
-\newcommand{\NB}{N_\zeta}
+\newcommand{\NR}{N_\mathrm{R}}
+\newcommand{\NB}{N_\mathrm{B}}
 \newcommand{\rlt}{R/L_T}
@@
 \newcommand{\kzf}{k_\mathrm{ZF}}
+\newcommand{\xcoord}{x}
+\newcommand{\ycoord}{y}
 
@@
 
-The increase of the box size, relative to the standard box size $(L_\psi,~L_\zeta) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parameter $\NR$ and $\NB$, respectively.
+The increase of the box size, relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parameter $\NR$ and $\NB$, respectively.
 Here, the nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
@@
 \begin{equation}
-	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \psi^2},
+	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
 	\label{eq:shearingrate}
@@
 \increasecounter{eq}{1}{1}
-where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge) and $\psi$ is the radial coordinate that labels the flux surfaces normalized by $\rhoth$.
+where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge) and $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by $\rhoth$.
 Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
 Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref(\textbf{?}).
-The zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\psi$-$\zeta$-plane at the low field side according to
+The zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to
 \begin{equation}
-\langle \phi \rangle = \frac{1}{L_\zeta} \int_0^{L_\zeta} \mathrm{d}\zeta ~ \phi(\psi,\zeta,s=0),
+\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0),
 \end{equation}
 \increasecounter{eq}{1}{1}
-where $\zeta$ labels the field lines and is an approximate binormal or toroidal coordinate, and $s$ parameterizes the length along the field lines and represents a poloidal coordinate which represents together with the radial coordinate $\psi$ the Hamada coordinates $(\psi,\zeta,s)$.
+where $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate, and $s$ parameterizes the length along the field lines and represents a poloidal coordinate which represents together with the radial coordinate $\xcoord$ the Hamada coordinates $(\xcoord,\ycoord,s)$.
 \cite{rath2021}(\textbf{?})
@@
 \begin{equation}
-	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \psi),
+	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord),
 	\label{eq:shearingrate_fourier}
@@
 \increasecounter{eq}{1}{1}
-where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\psi$
+where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
 % explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
-defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\psi -1)/2 \leq \nzf \leq (N_\psi -1)/2 $.
+defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
 % define the zonal flow modes' amplitude in terms of the shearing rate
@@
 		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
-		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [11000, 12000]$, & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$ & $\in [2000, 4000]$    &                 &                      \\
+		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$, & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$ & $\in [1000, 3000]$    &                 &                      \\
 	\end{tabular}
 }{fig:wexb-stable-comparison}{}
+
+\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
+}{fig:eflux-1x1-2x2-3x3-comparison}
 
@@
 This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats always after $\sim 1000\,R/\vth$ although the turbulence of realization $3\times 1.5$ does suppress after $\sim 11000\,R/\vth$ [Fig.
-\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
-The convergence of staircase pattern can be seen in Fig \ref{fig:wexb-stable-comparison}\,(c) and the test confirms again the typical mesoscale distinct from $\NR > 1$.
-Fig.
-\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
-The zonal flow mode number varies between $\nzf = [3,4]$ which can be seen in Fig.
-\ref{fig:wexb-stable-comparison}\,(c) for the realization $3\times 2.5$ where the staircase structure has an (PATTERN).
-The Fourier analysis yields in this case no definitely basic mode rather than two dominate modes with $\nzf = 3, 4$ which split the maximum amplitude $|\hatwexb|_{\kzf}$ between them.
-
-\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial and binormal increased box sizes
-}{fig:eflux-1x1-2x2-3x3-comparison}
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats always after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
+The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and the test confirms again the typical mesoscale distinct from $\NR > 1$.
+Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+The zonal flow mode number varies between $\nzf = [3,4]$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c).
+For the realization $3\times 2.5$ the staircase structure has an pattern between $3$ and $4$ repetitions which get represented in the second repetition with no $+ \gamma$ interval.
+Instead the pattern returns immediately after reaching $+ \gamma$ to the $- \gamma$ interval of the third repetition.
+The Fourier analysis yields in this case no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $|\hatwexb|_{\kzf}$ each.
 
@@
 % what is the finite heat flux threshold
-In the case of $\rlt = ???$ turbulence suppression is observed for $t > ???$, while stationary turbulence during the entire simulation time trace of $???~R/\vth$ is found for $\rlt = ???$.
-The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = ??? \pm ???$.
+In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 12000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
+The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.4 \pm 0.1$.
 % Which stationay wavelength occurs????
-Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.0$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = ???$ establishes.
+Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.0$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
 % Does it confirm our previous tests
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim ???$.\bigskip
+Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$.\bigskip
 

### 2023-03-21 briefcommunicaton/briefcommunication.tex
#37: Corrections, colors in plots, rearangements, references

@@
 \newcommand{\exb}{\mathrm{\:E}\times\mathrm{B}}
-\newcommand{\hatwexbkvec}{\widehat{\omega}_{\mathrm{\:E \times B},\mathbf{k}}}
+\newcommand{\hatwexbampvec}{|\hatwexb|_\mathbf{n}}
+\newcommand{\hatwexbamp}{|\hatwexb|_{\nzf}}
 %\newcommand{\NR}{N_\mathrm{R}}
@@
 \textit{Does the basic pattern wavelength always converges to the box size, or is there a typical mesoscale wavelength inherent to staircase structures also in a local flux-tube description?}
-The latter case would imply that it is not necessarily global physics which means profile effects are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism.
-This profile effects set (i) the wavelength of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events \cite{pradalier2010}.
+The latter case would imply that it is not necessarily global physics, i.e., profile effects that are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism.
+These profile effects set (i) the wavelength of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events \cite{pradalier2010}.
 In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\cite{peeters2016}.\bigskip
@@
 
-Here, gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} in the adiabatic electron approximation.
+The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} with adiabatic electron approximation.
 In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
@@
 The increase of the box size, relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parameter $\NR$ and $\NB$, respectively.
-Here, the nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
+Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate.
+Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates.(\textbf{?})
+The nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
 Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
@@
 \increasecounter{eq}{1}{1}
-where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge) and $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by $\rhoth$.
-Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref(\textbf{?}).
-The zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to
+where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
+The zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
 \begin{equation}
-\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0),
+\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
 \end{equation}
 \increasecounter{eq}{1}{1}
-where $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate, and $s$ parameterizes the length along the field lines and represents a poloidal coordinate which represents together with the radial coordinate $\xcoord$ the Hamada coordinates $(\xcoord,\ycoord,s)$.
-\cite{rath2021}(\textbf{?})
 The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
+\\
+Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\cite{peeters2016,peeters2009}.
 
@@
 % define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $|\hatwexb|_{\kzf} = 2 |\hatwexb|$. 
+Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb|$. 
 % introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work,exhibits the maximum amplitude in the spectrum $|\hatwexb|_{\kzf}$.
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.
 \bigskip
@@
 % first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
-It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $|\hatwexb|_{\mathbf{k}}   $ in the latter two phases (not shown), demonstrating a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
+It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbampvec$ in the latter two phases (not shown).
+This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
 % second important point: no secular growth of the box scale zonal mode
@@
 % mention shorter times scales for stabilization
-Interestingly, here, suppression of the turbulence by the emergence of a fully developed staircase pattern occurs after $\sim 1000~R/\vth$ always [Fig.
+Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern always occurs after $\sim 1000~R/\vth$ [Fig.
 \ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
@@
 		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
-		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$, & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$ & $\in [1000, 3000]$    &                 &                      \\
+		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    & $t_{3\times 5}$ & $\in [1000, 3000]$   \\
 	\end{tabular}
@@
 % discuss sumewhat smaller wavelength
-By contrast to the radial box size scan, however, here the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern wavelength.
+By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern wavelength.
 Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
@@
 This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats always after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
-The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and the test confirms again the typical mesoscale distinct from $\NR > 1$.
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
+The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a wavelength of a typical mesoscale.
 Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
-The zonal flow mode number varies between $\nzf = [3,4]$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c).
-For the realization $3\times 2.5$ the staircase structure has an pattern between $3$ and $4$ repetitions which get represented in the second repetition with no $+ \gamma$ interval.
-Instead the pattern returns immediately after reaching $+ \gamma$ to the $- \gamma$ interval of the third repetition.
-The Fourier analysis yields in this case no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $|\hatwexb|_{\kzf}$ each.
+The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
+The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no positive shear.
+Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
+The Fourier analysis of this case yields no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
 
@@
 % Which stationay wavelength occurs????
-Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.0$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
+Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
 % Does it confirm our previous tests
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$.\bigskip
+Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$.\\
 

### 2023-03-21 python/color.py

@@ -1,15 +1,27 @@
 import seaborn as sns
 
-color = sns.color_palette("hls", 9).as_hex()
+#color = sns.color_palette("hls", 9).as_hex()
 
-print(color)
+color = sns.color_palette("viridis", 4).as_hex()
 
-pal_1 = ['#db5f57', '#dbb757', '#a7db57', '#57db5f', '#57dbb7', '#57a7db', '#5f57db', '#b757db', '#db57a7']
+#print(color)
 
-pal_2 = ['#0173b2', '#de8f05', '#029e73', '#d55e00', '#cc78bc', '#ca9161', '#fbafe4', '#ece133', '#56b4e9']
+hls           = ['#db5f57', '#dbb757', '#a7db57', '#57db5f', '#57dbb7', '#57a7db', '#5f57db', '#b757db', '#db57a7']
+colorblind    = ['#0173b2', '#de8f05', '#029e73', '#d55e00', '#cc78bc', '#ca9161', '#fbafe4', '#ece133', '#56b4e9']
+spectral      = ['#d43d4f', '#f46d43', '#fdad60', '#fee08b', '#ffffbe', '#e6f598', '#aadca4', '#66c2a5', '#3387bc']
+seagreen_dark = ['#232724', '#24332a', '#254031', '#274c37', '#28593e', '#2a6644', '#2b724a', '#2d7f51', '#2e8b57']
+rocket_r      = ['#4c1d4b', '#a11a5b', '#e83f3f', '#f69c73']
+viridis       = ['#414487', '#2a788e', '#22a884', '#7ad151']
 
-pal_3 = ['#d43d4f', '#f46d43', '#fdad60', '#fee08b', '#ffffbe', '#e6f598', '#aadca4', '#66c2a5', '#3387bc']
+colors = [['#a11a5b', '#029e73', '#de8f05', '#0173b2'],
+ 		             ['#66c2a5', '#dbb757', '#0173b2'],
+		  ['#87429b', '#66c2a5', '#d55e00', '#56b4e9']]
 
+color_rad = colors[0][::-1]
+
+print(color_rad)
+
+['#0173b2', '#de8f05', '#029e73', '#a11a5b']
 
 '''
 1 1x1

### 2023-03-21 presentation/presentation.tex
Presentation pictures in picture folder

@@
 }
+
+\graphicspath{{../pictures/}}
 
@@
 		\frametitle{Beschreibung von Plasma im Tokamak Reaktor}
-		\includegraphics[scale = 1]{pictures/Zonal_Flow.PNG}
+		\includegraphics[scale = 1]{Presentation/Zonal_Flow.PNG}
 		\begin{alertblock}{Achtung Instabilität am Plasmarand!}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_00.pdf}
+		\includegraphics{Presentation/ITG_Instability_00.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_01.pdf}
+		\includegraphics{Presentation/ITG_Instability_01.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_02.pdf}
+		\includegraphics{Presentation/ITG_Instability_02.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_03.pdf}
+		\includegraphics{Presentation/ITG_Instability_03.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_04.pdf}
+		\includegraphics{Presentation/ITG_Instability_04.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_05.pdf}
+		\includegraphics{Presentation/ITG_Instability_05.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_06.pdf}
+		\includegraphics{Presentation/ITG_Instability_06.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_07.pdf}
+		\includegraphics{Presentation/ITG_Instability_07.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_08.pdf}
+		\includegraphics{Presentation/ITG_Instability_08.pdf}
 
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_09.pdf}
+		\includegraphics{Presentation/ITG_Instability_09.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{pictures/ITG_Instability_10.pdf}
+		\includegraphics{Presentation/ITG_Instability_10.pdf}
 	\end{frame}
@@
 			\begin{figure}%[4]
-				\href{https://www.youtube.com/watch?v=O6tUHzfj-zM}{\includegraphics[width=\textwidth]{pictures/turbulence.png}}
+				\href{https://www.youtube.com/watch?v=O6tUHzfj-zM}{\includegraphics[width=\textwidth]{Presentation/turbulence.png}}
 			\end{figure}
@@
 		Verdrillen der magnetischen Flussdichte $\vect{B}$ überführt Plasma in stabilen Zustand \\[0.2cm]
-		\includegraphics[scale=0.25]{pictures/Stabilzation_Mechanism_black.png}
+		\includegraphics[scale=0.25]{Presentation/Stabilzation_Mechanism_black.png}
 	\end{frame}
@@
 		\begin{minipage}{0.3\textwidth}
-			\includegraphics[scale = 0.2]{pictures/Magnetic_Shear_black.png}
+			\includegraphics[scale = 0.2]{Presentation/Magnetic_Shear_black.png}
 		\end{minipage}
@@
 	%	\frametitle{Regelung oder Unterdrückung von Turbulenzen}
-	%	\includegraphics[scale = 1.3]{pictures/Shearing.pdf}
+	%	\includegraphics[scale = 1.3]{Presentation/Shearing.pdf}
 	%\end{frame}
@@
 		\begin{tabular}{c c}
-			\includegraphics[scale = 0.2]{pictures/Stablized_Plasma_Interaction.png} & \includegraphics[scale = 0.2]{pictures/Stablized_Plasma_LongScale.png} \\
+			\includegraphics[scale = 0.2]{Presentation/Stablized_Plasma_Interaction.png} & \includegraphics[scale = 0.2]{Presentation/Stablized_Plasma_LongScale.png} \\
 			Dominante nichtlineare Interaction  & Zusätzlicher verscherter Zonal Flow \\
@@
 		\frametitle{Danke für die Aufmerksamkeit!}
-		\includegraphics[scale = 0.35]{pictures/Fusion_Meme.jpeg} \\
+		\includegraphics[scale = 0.35]{Presentation/Fusion_Meme.jpeg} \\
 	\end{frame}

### 2023-03-22 briefcommunicaton/briefcommunication.tex
#37: x_max=3000 of eflux plots, removed titles from reference, Corrections Florian Rath, Gradient Length 6.3+/-0.1, Mean sheraing rate in plot, \hatwexbamp in caption.

@@
 		\captionsetup{type=figure}
-		\includegraphics[width=0.9\linewidth]{#1}
+		\includegraphics[width=0.95\linewidth]{#1}
 		\captionof{figure}{#2}
@@
 \newcommand{\exb}{\mathrm{\:E}\times\mathrm{B}}
-\newcommand{\hatwexbampvec}{|\hatwexb|_\mathbf{n}}
+%\newcommand{\hatwexbampvec}{|\hatwexb|_\mathbf{n}}
 \newcommand{\hatwexbamp}{|\hatwexb|_{\nzf}}
@@
 Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate.
-Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates.(\textbf{?})
+Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
 The nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
-Note that the number of modes in the respective direction is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
-\bigskip
-
+Note that the number of modes in the respective direction, i.~e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
+\\
 The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
@@
 \end{equation}
-\increasecounter{eq}{1}{1}
 where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
@@
 \end{equation}
-\increasecounter{eq}{1}{1}
 The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
@@
 Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\cite{peeters2016,peeters2009}.
-
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\cite{peeters2016,peeters2009}.\\
 In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
@@
 \end{equation}
-\increasecounter{eq}{1}{1}
 where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
@@
 % introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.
-\bigskip
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\
 
@@
 % first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
-It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbampvec$ in the latter two phases (not shown).
+It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbamp$ in the latter two phases (not shown).
 This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
@@
 	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
-	\textbf{(b)} Time traces of $|\hatwexb|_{\kzf}$ for radial increased box sizes
+	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
 }{fig:wexb-eflux-1-2-3-4x1-comparison}{}
@@
 \includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
-	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
+	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
 The staircase structures got shifted till alignment for better visibility.\\
@@
 By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern wavelength.
-Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
-\bigskip
+Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.\bigskip
 
@@
 In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 12000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
-The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.4 \pm 0.1$.
+The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$.
 % Which stationay wavelength occurs????
@@
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$.\\
+Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$.\bigskip
 

### 2023-04-04 briefcommunicaton/briefcommunication.tex
#37: Correction Prof Arthur Peeters

@@
 		\captionsetup{type=figure}
-		\includegraphics[width=0.95\linewidth]{#1}
+		\includegraphics[width=0.87\linewidth]{#1}
 		\captionof{figure}{#2}
@@
 
-\title[Wavelength convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
-{Wavelength convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
+\title[Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
+{Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
 
@@
 					Florian.Rath@uni-bayreuth.de}
-	\email{Florian.Rath@uni-bayreuth.de}
 \author{A. G. Peeters}
@@
 \begin{abstract}
-    The wavelength convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-    Its is shown that a mesoscale pattern wavelength of \linebreak $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
+    The radial size convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
+    Its is shown that a mesoscale pattern size of \linebreak $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
 
@@
 
-Ion temperature gradient driven turbulence close to marginal stability often exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{pradalier2010}.
+Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{pradalier2010}.
 Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{peeters2016, weikl2017, rath2021} as well as global gradient-driven \cite{mcmillan2009, villard2013, seo2022} and global flux-driven \cite{pradalier2010, pradalier2015, wang2020, kim2022, kishimoto2023} studies.
-In global studies, usually spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern wavelength of several ten Larmor radii.
+In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
 % By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
-By contrast, in the aforementioned local studies the wavelength of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
+By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
 The above observations lead to the question: 
-\textit{Does the basic pattern wavelength always converges to the box size, or is there a typical mesoscale wavelength inherent to staircase structures also in a local flux-tube description?}
-The latter case would imply that it is not necessarily global physics, i.e., profile effects that are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism.
-These profile effects set (i) the wavelength of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events \cite{pradalier2010}.
-In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\cite{peeters2016}.\bigskip
+\textit{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}
+The latter case would imply that it is not necessarily global physics, i.e., profile effects, that set (i) the radial size of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events.
+These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}.
+In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\onlinecite{peeters2016}.\bigskip
 
@@
 The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} with adiabatic electron approximation.
-In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
+In agreement with Ref.~\onlinecite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
 It has been carefully verified that this modification preserves the same physical outcome as the original study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\onlinecite{peeters2009, peeters2016}.
 \begin{table}[ht]
@@
 	\caption{
-		Resolution used in this paper for further information the author links to Ref.~\cite{peeters2016}.
+		Resolution used in this paper for further information the author links to Ref.~\onlinecite{peeters2016}.
 %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
@@
 
-The increase of the box size, relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$, in the radial and binormal direction is always indicated by the real parameter $\NR$ and $\NB$, respectively.
+In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
 Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate.
 Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
-The nomenclature $\NR\times \NB$ is applied throughout this work to indicate the increase in box size.  
-Note that the number of modes in the respective direction, i.~e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
+The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
+Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
 \\
@@
 where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
-The zonal potential is estimated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
+The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
 \begin{equation}
@@
 \\
-Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\cite{peeters2016,peeters2009}.\\
+Consistent with Ref.~\onlinecite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\onlinecite{peeters2016,peeters2009}.
+\newpage
 In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
@@
 Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
-% yes, the wavelength converges; but even stronger result: the standard box size already allows for the correct staircase wavelength
-Hence, the basic wavelength of the pattern not only converges with increasing radial box size, the converged wavelength turns out to agree with the standard radial box size of Ref.~\cite{peeters2016} \\
+% yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
+Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\cite{peeters2016} \\
 % cautios words
@@
 % the main result of this test: confirmation of wave length convergence
-As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern wavelength to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 
@@
 	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
-The staircase structures got shifted till alignment for better visibility.\\
+The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
 	\begin{tabular}{l l l l l l l l l}
@@
 
-% discuss sumewhat smaller wavelength
-By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern wavelength.
-Whether this is related to a possible pattern wavelength dependence on the binormal box size or to the competition between patterns with the two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.\bigskip
-
+% discuss sumewhat smaller radial size
+By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
+Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
+\newpage
 % what kind of analysis
@@
 As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
-The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a wavelength of a typical mesoscale.
-Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two wavelengths $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
+Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
 The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
@@
 The Fourier analysis of this case yields no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
-
 \includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
 	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
-}{fig:eflux-3x1.5-2.5-3-5-comparison}
+}{fig:eflux-3x1.5-2.5-3-5-comparison}\bigskip
 
@@
 The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$.
-% Which stationay wavelength occurs????
+% Which stationay radial size occurs????
 Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
@@
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$.\bigskip
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.\\
+
+\bigskip
+\bigskip
+\bigskip
 
@@
 % what has been done
-Through careful tests this brief communication confirms the wavelength convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
+Through careful tests this brief communication confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
 % mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
-A mesoscale pattern wavelength of $\sim 57.20 - 76.27\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
+A mesoscale pattern size of $\sim 57.20 - 76.27\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
 % compare to other studies
@@
 % relevance of the outcome
-The occurrence of this mesoscale implies that non-locality is inherent to ITG driven turbulence, since avalanches are usually spatially organized by the $\exb$ staircase pattern \cite{mcmillan2009, pradalier2010, rath2016, peeters2016}.
+The occurrence of this mesoscale implies that non-locality, in terms of Ref.
+\onlinecite{pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{mcmillan2009, pradalier2010, rath2016, peeters2016}.
 

### 2023-04-04 bachelorthesis/01-introduction.tex
#39: Include brief communication in thesis

@@
 \label{chap:intro}
+
+Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{pradalier2010}.
+Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{peeters2016, weikl2017, rath2021} as well as global gradient-driven \cite{mcmillan2009, villard2013, seo2022} and global flux-driven \cite{pradalier2010, pradalier2015, wang2020, kim2022, kishimoto2023} studies.
+In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
+% By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
+By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
+The above observations lead to the question: 
+\textit{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}
+The latter case would imply that it is not necessarily global physics, i.e., profile effects, that set (i) the radial size of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events.
+These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}.
+In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\cite{peeters2016}.

### 2023-04-04 bachelorthesis/02-theory.tex
#39: Include brief communication in thesis

@@
-% 4. Theory
+% 2. Theory
 
@@
 
-% Text
-
-% etc.
+In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
+Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate.
+Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
+The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
+Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
+\\
+The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
+\begin{equation}
+	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
+	\label{eq:shearingrate}
+\end{equation}
+where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
+The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
+\begin{equation}
+\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
+\end{equation}
+The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
+\\
+Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\cite{peeters2016,peeters2009}.
+\newpage
+In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
+It is defined by
+\begin{equation}
+	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord),
+	\label{eq:shearingrate_fourier}
+\end{equation}
+where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
+% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
+defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
+% define the zonal flow modes' amplitude in terms of the shearing rate
+Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb|$. 
+% introduces the term basic mode of the pattern for later usage
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\

### 2023-04-04 bachelorthesis/03-methods.tex
#39: Include brief communication in thesis

@@
 \label{chap:methods}
+
+\section{Simulation Setup}
+\label{sec:simulation}
+
+The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} with adiabatic electron approximation.
+In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
+The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
+It has been carefully verified that this modification preserves the same physical outcome as the original study.
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.
+\begin{center}
+	\captionsetup{type=table}
+	\begin{tabular}{l | ccccc | ccccc | c | cc}
+		& $N_m$ & $N_x$ & $N_s$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_d$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
+		\hline
+		S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
+	\end{tabular}
+	\captionof{table}{
+		Resolution used in this paper: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+	}
+	\label{tab:resolution}
+\end{center}
+
+\section{btrzx1 Cluster}
+\label{sec:cluster}
+
+\section{Restart Script for Simulation btrzx1 Cluster}
+\label{sec:restartscript}

### 2023-04-04 bachelorthesis/04-results.tex
#39: Include brief communication in thesis

@@
 
-\chapter{Results}
+\chapter{Results and Discussion}
 \label{chap:results}
 
-% Text
+\section{Variation of Computational Resolution}
+\label{sec:variationsofresolution}
 
-% etc.
+At the beginning of this work the goal is to change the used resolutuin to a minimum. The goal behind this research is to reduce Computational time and costs for the computation of the simulation.
+
+\section{Convergenve of Wavelength}
+\label{sec:convergence}
+
+%%% RESULTS ================================================================================================================
+
+\subsection{Radial increased Box Size}
+\label{sub:radial}
+
+% what kind of analysis
+In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size.
+The scan covers the realizations $\NR\times\NB \in [ 1\times1,~2\times1,~3\times1,~4\times1]$.
+% short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
+% introduce the fully developed staircase structure
+The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+% insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
+Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
+% yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
+Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\cite{peeters2016} \\
+% cautios words
+Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
+% very long initial quasi-stationary states
+Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
+% the \nzf=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
+During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
+Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
+% first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
+It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbamp$ in the latter two phases (not shown).
+This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
+% second important point: no secular growth of the box scale zonal mode
+Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
+% summary
+The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
+\\
+
+\subsection{Isotropic increased Box Size}
+\label{sub:isotropic}
+
+% what kind of analysis
+Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
+This scan covers the realizations $\NR\times\NB \in [1\times1,~2\times2,~3\times3]$.
+% mention shorter times scales for stabilization
+Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern always occurs after $\sim 1000~R/\vth$ [Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
+% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
+%
+% the main result of this test: confirmation of wave length convergence
+As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
+	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
+}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
+
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
+	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
+	\begin{tabular}{l l l l l l l l l}
+		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$ & $\in [26000, 28000]$ \\
+		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
+		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    & $t_{3\times 5}$ & $\in [1000, 3000]$   \\
+	\end{tabular}
+}{fig:wexb-stable-comparison}{}
+
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
+}{fig:eflux-1x1-2x2-3x3-comparison}{}
+
+% discuss sumewhat smaller radial size
+By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
+Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
+
+\subsection{Binormal increased Box Size}
+\label{sub:binormal}
+
+% what kind of analysis
+In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
+This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
+The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
+Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
+The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no positive shear.
+Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
+The Fourier analysis of this case yields no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
+}{fig:eflux-3x1.5-2.5-3-5-comparison}{}\bigskip
+
+\section{The finite heat flux threshold}
+\label{sub:threshold}
+
+% what kind of test
+In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
+Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
+This scan covers realizations with $\rlt \in [6.0,~6.2,~6.4]$.
+% what is the finite heat flux threshold
+In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 12000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
+The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$.
+% Which stationay radial size occurs????
+Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
+% Does it confirm our previous tests
+Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.\\

### 2023-04-04 bachelorthesis/05-closure.tex
#39: Include brief communication in thesis

--- 2023-04-04 briefcommunicaton/briefcommunication.tex
@@
-% AUTHOR ===================================================================================================================
-% Manuel Lippert (GitHub: ManeLippert (https://github.com/ManeLippert))
-% ==========================================================================================================================
+% 6. Closure
 
-% PREAMBLE =================================================================================================================
-
-% ****** Start of file aipsamp.tex ******
-%
-%   This file is part of the AIP files in the AIP distribution for REVTeX 4.
-%   Version 4.1 of REVTeX, October 2009
-%
-%   Copyright (c) 2009 American Institute of Physics.
-%
-%   See the AIP README file for restrictions and more information.
-%
-% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
-% as well as the rest of the prerequisites for REVTeX 4.1
-% 
-% It also requires running BibTeX. The commands are as follows:
-%
-%  1)  latex  aipsamp
-%  2)  bibtex aipsamp
-%  3)  latex  aipsamp
-%  4)  latex  aipsamp
-%
-% Use this file as a source of example code for your aip document.
-% Use the file aiptemplate.tex as a template for your document.
-
-% DOCUMENT =================================================================================================================
-
-\documentclass[aip, amsmath, amssymb, reprint, twocolumn, floatfix]{revtex4-1}
-
-% documentclass revtex4-1 options:
-% - aip,
-% - jmp,
-% - bmf,
-% - sd,
-% - rsi,
-% - amsmath,amssymb,
-% - preprint,%
-% - reprint,%
-% - author-year,%
-% - author-numerical,%
-% - Conference Proceedings
-
-\preprint{AIP/123-QED}
-
-% PACKAGES =================================================================================================================
-
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{mathptmx}
-\usepackage{etoolbox}
-\usepackage{graphicx}% Include figure files
-\usepackage{dcolumn}% Align table columns on decimal point
-\usepackage{bm}% bold math
-%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
-%\linenumbers\relax % Commence numbering lines
-
-%% Additional
-\usepackage[format=hang]{caption}
-\usepackage{hyperref}
-
-% FUNCTIONS, INPUT =========================================================================================================
-
-\input{wordcount.tex}% Word & page count for document
-
-\graphicspath{{../pictures/}}% Path for pictures
-
-%% Apr 2021: AIP requests that the corresponding 
-%% email to be moved after the affiliations
-\makeatletter
-\def\@email#1#2{
-	\endgroup
-	\patchcmd{\titleblock@produce}
-		{\frontmatter@RRAPformat}
-		{\frontmatter@RRAPformat{\produce@RRAP{*#1\href{mailto:#2}{#2}}}\frontmatter@RRAPformat}
-		{}{}
-}
-\makeatother
-
-%% Include graphic for one column with specific place 
-\newcommand{\includegraphicsOneCol}[3]{
-	\begin{center}
-		\centering
-		\captionsetup{type=figure}
-		\includegraphics[width=0.87\linewidth]{#1}
-		\captionof{figure}{#2}
-		\label{#3}
-	\end{center}
-  	\increasecounter{fig}{1}
-}
-
-%\newcommand{\includegraphicsOneCol}[3]{
-%	\begin{figure}[hbt!]
-%		\centering
-%		\includegraphics[width=0.9\linewidth]{#1}
-%		\caption{#2}
-%		\label{#3}
-%  	\end{figure}
-%  	\increasecounter{fig}{1}
-%}
-
-%% Include graphic for two column with specific place, figrue* places graphic anywhere...
-\newcommand{\includegraphicsTwoCol}[4]{
-	\onecolumngrid
-	\begin{center}
-		\centering
-		\captionsetup{type=figure}
-    	\includegraphics[width=#4\textwidth]{#1}
-		\captionof{figure}{#2}
-    	\label{#3}
-	\end{center}
-	\twocolumngrid
-	\increasecounter{fig}{2}
-}
-
-%\newcommand{\includegraphicsTwoCol}[3]{
-%	\begin{figure*}
-%    	\includegraphics[width=0.9\textwidth]{#1}
-%		\caption{#2}
-%		\label{#3}
-%	\end{figure*}
-%	\increasecounter{fig}{2}
-%}
-
-\newcommand{\words}{
-	\immediate\write18{texcount -sum -1 \jobname.tex > wordcount.txt}
-}
-
-\newcommand{\wexb}{\omega_{\mathrm{\:E \times B}}}
-\newcommand{\hatwexb}{\widehat{\omega}_{\mathrm{\:E \times B}}}
-\newcommand{\exb}{\mathrm{\:E}\times\mathrm{B}}
-%\newcommand{\hatwexbampvec}{|\hatwexb|_\mathbf{n}}
-\newcommand{\hatwexbamp}{|\hatwexb|_{\nzf}}
-%\newcommand{\NR}{N_\mathrm{R}}
-%\newcommand{\NB}{N_\mathrm{B}}
-\newcommand{\NR}{N_\mathrm{R}}
-\newcommand{\NB}{N_\mathrm{B}}
-\newcommand{\rlt}{R/L_T}
-%\newcommand{\rhoth}{\rho_\mathrm{th}}
-\newcommand{\rhoth}{\rho}
-\newcommand{\vth}{v_{\mathrm{th}}}
-\newcommand{\nzf}{n_\mathrm{ZF}}
-\newcommand{\kzf}{k_\mathrm{ZF}}
-\newcommand{\xcoord}{x}
-\newcommand{\ycoord}{y}
-
-% MAIN =====================================================================================================================
-
-\begin{document}
-
-%% TITLE, INFO =============================================================================================================
-
-\title[Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
-{Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
-
-\author{M. Lippert}
-	\altaffiliation{Repository of this work: \\ 
-					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
-\author{F. Rath}
-	\altaffiliation{Author to whom correspondence should be addressed: \\
-					Florian.Rath@uni-bayreuth.de}
-\author{A. G. Peeters}
-\affiliation{Physics Department, University of Bayreuth, 95440 Bayreuth, Germany}
-
-\date{\today}
-
-%% ABSTRACT ================================================================================================================
-
-\begin{abstract}
-    The radial size convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-    Its is shown that a mesoscale pattern size of \linebreak $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
-
-\end{abstract}
-
-\maketitle
-
-%% TEXT ====================================================================================================================
-
-%%% INTRODUCTION ===========================================================================================================
-
-%\section{Introduction}
-%\label{sec:intro}
-
-Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{pradalier2010}.
-Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{peeters2016, weikl2017, rath2021} as well as global gradient-driven \cite{mcmillan2009, villard2013, seo2022} and global flux-driven \cite{pradalier2010, pradalier2015, wang2020, kim2022, kishimoto2023} studies.
-In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
-% By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
-By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
-The above observations lead to the question: 
-\textit{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}
-The latter case would imply that it is not necessarily global physics, i.e., profile effects, that set (i) the radial size of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events.
-These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}.
-In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\onlinecite{peeters2016}.\bigskip
-
-%%% SIMULATION SET-UP ======================================================================================================
-
-The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} with adiabatic electron approximation.
-In agreement with Ref.~\onlinecite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
-The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
-It has been carefully verified that this modification preserves the same physical outcome as the original study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\onlinecite{peeters2009, peeters2016}.
-\begin{table}[ht]
-	\begin{ruledtabular}
-		\begin{tabular}{l | ccccc | ccccc | c | cc}
-			& $N_m$ & $N_x$ & $N_s$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_d$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
-			\hline
-			S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
-		\end{tabular}
-	\end{ruledtabular}
-	\caption{
-		Resolution used in this paper for further information the author links to Ref.~\onlinecite{peeters2016}.
-%: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
-	}
-	\label{tab:resolution}
-\end{table}
-
-%%% DIAGNOSTICS ================================================================================================================
-
-In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
-Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate.
-Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
-The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
-Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
-\\
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
-\begin{equation}
-	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
-	\label{eq:shearingrate}
-\end{equation}
-where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
-The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
-\begin{equation}
-\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
-\end{equation}
-The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
-\\
-Consistent with Ref.~\onlinecite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\onlinecite{peeters2016,peeters2009}.
-\newpage
-In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
-It is defined by
-\begin{equation}
-	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord),
-	\label{eq:shearingrate_fourier}
-\end{equation}
-where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
-% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
-defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
-% define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb|$. 
-% introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\
-
-%%% RESULTS ================================================================================================================
-
-%\section{Results}
-%\label{sec:results}
-
-% what kind of analysis
-In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size.
-The scan covers the realizations $\NR\times\NB \in [ 1\times1,~2\times1,~3\times1,~4\times1]$.
-% short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
-% introduce the fully developed staircase structure
-The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
-% insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
-Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
-% yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\cite{peeters2016} \\
-% cautios words
-Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
-% very long initial quasi-stationary states
-Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
-% the \nzf=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
-During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
-Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
-% first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
-It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbamp$ in the latter two phases (not shown).
-This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
-% second important point: no secular growth of the box scale zonal mode
-Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
-% summary
-The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
-\\
-
-% what kind of analysis
-Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
-This scan covers the realizations $\NR\times\NB \in [1\times1,~2\times2,~3\times3]$.
-% mention shorter times scales for stabilization
-Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern always occurs after $\sim 1000~R/\vth$ [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
-% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
-%
-% the main result of this test: confirmation of wave length convergence
-As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
-
-\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
-	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
-
-\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
-	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
-The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
-	\begin{tabular}{l l l l l l l l l}
-		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$ & $\in [26000, 28000]$ \\
-		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
-		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    & $t_{3\times 5}$ & $\in [1000, 3000]$   \\
-	\end{tabular}
-}{fig:wexb-stable-comparison}{}
-
-\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
-}{fig:eflux-1x1-2x2-3x3-comparison}
-
-% discuss sumewhat smaller radial size
-By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
-Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
-\newpage
-% what kind of analysis
-In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
-This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
-The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
-Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
-The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
-The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no positive shear.
-Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
-The Fourier analysis of this case yields no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
-\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
-}{fig:eflux-3x1.5-2.5-3-5-comparison}\bigskip
-
-% what kind of test
-In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
-Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
-This scan covers realizations with $\rlt \in [6.0,~6.2,~6.4]$.
-% what is the finite heat flux threshold
-In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 12000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
-The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$.
-% Which stationay radial size occurs????
-Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
-% Does it confirm our previous tests
-Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.\\
-
-\bigskip
-\bigskip
-\bigskip
-
-%%% CONCLUSION =============================================================================================================
-
-%\section{Conclusion}
-%\label{sec:conclusion}
+\chapter{Closure}
+\label{chap:close}
 
@@
 The occurrence of this mesoscale implies that non-locality, in terms of Ref.
-\onlinecite{pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{mcmillan2009, pradalier2010, rath2016, peeters2016}.
-
-%% APPENDIX ================================================================================================================
-
-\section*{Data Availability}
-The data that support the findings of this study are available from the corresponding author upon reasonable request.
-
-% WORDCOUNT ================================================================================================================
-
-%\wordcountmessage
-
-% BIBLIOGRAPHY =============================================================================================================
-
-\nocite{}
-\bibliography{references.bib}% Produces the bibliography via BibTeX.
-
-\end{document}
-% END MAIN =================================================================================================================
-% ****** End of file aipsamp.tex ******
+\cite{pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{mcmillan2009, pradalier2010, rath2016, peeters2016}.

### 2023-04-04 bachelorthesis/04-results.tex
#39: Adjust chapter style, Seperate chapters in latex files, further Adjustment of brief commmunication

@@
 \label{chap:results}
+\thispagestyle{empty}
+\newpage
 
@@
 
-\section{Convergenve of Wavelength}
+\section{Size Convergence of $\exb$ staircase pattern}
 \label{sec:convergence}
 
-%%% RESULTS ================================================================================================================
+This chapter is an further iteration of the brief communication published in \enquote{Physics of Plasma}.
+It provides additional plots that was not necessary for the publication and some other formatting as well.
+The brief communication can be find in the appendix of this thesis or under \textbf{REF PAPER GITHUB}
+%TODO
 
@@
 
-% what kind of analysis
-In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size.
-The scan covers the realizations $\NR\times\NB \in [ 1\times1,~2\times1,~3\times1,~4\times1]$.
-% short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
+In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size. The scan covers the realizations: 
+\begin{gather*}
+	\boxed{\NR \times \NB \in [1\times1,~2\times1,~3\times1,~4\times1]}~.
+\end{gather*}
 Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
-% introduce the fully developed staircase structure
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
@@
 % yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\cite{peeters2016} \\
+Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\cite{peeters2016} \bigskip
+
 % cautios words
@@
 
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
+	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
+}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
+
+\newpage
 \subsection{Isotropic increased Box Size}
@@
 Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
-This scan covers the realizations $\NR\times\NB \in [1\times1,~2\times2,~3\times3]$.
+This scan covers the realizations:
+\begin{gather*}
+	\boxed{\NR\times\NB \in [1\times1,~2\times2,~3\times3]}~.
+\end{gather*}
 % mention shorter times scales for stabilization
@@
 
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
-	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
-
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
-	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
-The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
-	\begin{tabular}{l l l l l l l l l}
-		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$ & $\in [26000, 28000]$ \\
-		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
-		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    & $t_{3\times 5}$ & $\in [1000, 3000]$   \\
-	\end{tabular}
-}{fig:wexb-stable-comparison}{}
-
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
 	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
@@
 
+\newpage
 \subsection{Binormal increased Box Size}
@@
 In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
-This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
+This test covers the realizations:
+\begin{gather*}
+	\boxed{\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]}~.
+\end{gather*}
 As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
@@
 The Fourier analysis of this case yields no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
 	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
 }{fig:eflux-3x1.5-2.5-3-5-comparison}{}\bigskip
+
+\subsection{Staircase structures in Comparison}
+\label{sub:comparisonwexb}
+
+\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
+	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
+	\begin{tabular}{l l l l l}
+		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, \\
+		                        & $t_{3\times 1}$   & $\in [43000, 45000]$, & $t_{4\times 1}$   & $\in [26000, 28000]$  \\
+		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   \\
+		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    &                   &                       \\
+		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   \\
+		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    & $t_{3\times 5}$   & $\in [1000, 3000]$    \\
+	\end{tabular}
+}{fig:wexb-stable-comparison}{}
 
@@
 
-% what kind of test
 In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
 Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
-This scan covers realizations with $\rlt \in [6.0,~6.2,~6.4]$.
+This scan covers realizations with:
+\begin{gather*}
+	\boxed{\rlt \in [6.0,~6.2,~6.4]}~.
+\end{gather*}
 % what is the finite heat flux threshold
 In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 12000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
-The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$.
+The finite heat flux threshold, hence, is:
+\begin{gather*}
+	\boxed{\rlt|_\mathrm{finite} = 6.3 \pm 0.1}~.
+\end{gather*}
 % Which stationay radial size occurs????

### 2023-04-05 briefcommunicaton/briefcommunication.tex
#37: Update link to repository
#37: Correction 2 Florian Rath, Cover letter, reviewers choice

@@
 	\altaffiliation{Repository of this work: \\ 
-					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Wavelength}
+					https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Convergence}
 \author{F. Rath}
@@
 In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
-Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate.
-Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
+Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
+Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
 The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
@@
 % define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb|$. 
+Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$. 
 % introduces the term basic mode of the pattern for later usage
@@
 % yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\cite{peeters2016} \\
+Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\onlinecite{peeters2016}.
+\\
 % cautios words
@@
 This test covers the realizations $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$.
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
 The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
@@
 The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
-The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no positive shear.
+The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no signifciant plateau at positive shear.
 Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
-The Fourier analysis of this case yields no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
+The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
 \includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
@@
 % what is the finite heat flux threshold
-In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 12000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
-The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$.
+In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
+The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$ in accordance to Ref. \onlinecite{peeters2016}.
 % Which stationay radial size occurs????
-Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
+Although the initial quasi-stationary turbulence in the former case is significantly longer compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
 % Does it confirm our previous tests

### 2023-04-05 bachelorthesis/02-theory.tex
#39: Adjust chapter style, Seperate chapters in latex files, further Adjustment of brief commmunication
#39: Correction of brief communication

@@
 \label{chap:theory}
+\thispagestyle{empty}
+\newpage
 
 In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
-Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal or toroidal coordinate.
-Together with the coordinate $s$ which parameterizes the length along the field lines and represents the poloidal coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
+Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
+Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
 The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.

### 2023-04-05 bachelorthesis/04-results.tex
#39: Correction of brief communication

@@
 \end{gather*}
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern formats after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
 The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
@@
 The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
-The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no positive shear.
+The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no signifciant plateau at positive shear.
 Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
-The Fourier analysis of this case yields no definitely basic mode rather two dominate modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
+The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
 \includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
@@
 % what is the finite heat flux threshold
-In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 12000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
+In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
 The finite heat flux threshold, hence, is:
 \begin{gather*}
-	\boxed{\rlt|_\mathrm{finite} = 6.3 \pm 0.1}~.
+	\boxed{\rlt|_\mathrm{finite} = 6.3 \pm 0.1}
 \end{gather*}
+in accordance to Ref. \cite{peeters2016}.
 % Which stationay radial size occurs????
-Although the initial quasi-stationary turbulence in the former case is significantly larger compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
+Although the initial quasi-stationary turbulence in the former case is significantly longer compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
 % Does it confirm our previous tests

### 2023-04-05 bachelorthesis/01-introduction.tex
#39: Adjust chapter style, Seperate chapters in latex files, further Adjustment of brief commmunication
#39: Corrected inline citation

@@
 \label{chap:intro}
+\thispagestyle{empty}
+\newpage
 
@@
 These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}.
-In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\cite{peeters2016}.
+In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.
+\citenum{peeters2016}.

### 2023-04-05 bachelorthesis/02-theory.tex
#39: Corrected inline citation

@@
 \\
-Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\cite{peeters2016,peeters2009}.
+Consistent with Ref.
+\citenum{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.
+\citenum{peeters2016,peeters2009}.
 \newpage

### 2023-04-05 bachelorthesis/03-methods.tex
#39: Adjust chapter style, Seperate chapters in latex files, further Adjustment of brief commmunication
#39: Corrected inline citation

@@
 \label{chap:methods}
+\thispagestyle{empty}
+\newpage
 
@@
 The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} with adiabatic electron approximation.
-In agreement with Ref.~\cite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
+In agreement with Ref.
+\citenum{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
 It has been carefully verified that this modification preserves the same physical outcome as the original study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\cite{peeters2009, peeters2016}.
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.
+\citenum{peeters2009, peeters2016}.
 \begin{center}

### 2023-04-05 bachelorthesis/04-results.tex
#39: Corrected inline citation

@@
 % yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\cite{peeters2016} \bigskip
+Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.
+\citenum{peeters2016}.
+\bigskip
 
@@
 \end{gather*}
-in accordance to Ref. \cite{peeters2016}.
+in accordance to Ref. \citenum{peeters2016}.
 % Which stationay radial size occurs????

### 2023-04-06 briefcommunicaton/briefcommunication.tex
#37: Minor correction with onlinecites

@@
 It has been carefully verified that this modification preserves the same physical outcome as the original study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.~\onlinecite{peeters2009, peeters2016}.
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Ref.~\onlinecite{peeters2009, peeters2016}.
 \begin{table}[ht]
@@
 Consistent with Ref.~\onlinecite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\onlinecite{peeters2016,peeters2009}.
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.
+\onlinecite{peeters2016,peeters2009}.
 \newpage

### 2023-04-08 bachelorthesis/10-introduction.tex
#39: Split results part in separate, Plots for thesis, result part finished

--- 2023-04-05 bachelorthesis/01-introduction.tex
@@
 By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
-The above observations lead to the question: 
-\textit{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}
-The latter case would imply that it is not necessarily global physics, i.e., profile effects, that set (i) the radial size of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events.
-These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}.
-In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.
+The above observations lead to the question: \bigskip
+
+\textit{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}\bigskip
+
+The latter case would imply that it is not necessarily global physics, i.e., profile effects, that set 
+\begin{itemize}
+	\item[(i)] the radial size of the $\exb$ staircase pattern
+	\item[(ii)] the scale of avalanche-like transport events.
+\end{itemize}
+These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}. \bigskip
+
+In this bachlor thesis the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.
 \citenum{peeters2016}.
+%TODO More Text

### 2023-04-08 bachelorthesis/20-theory.tex
#39: Split results part in separate, Plots for thesis, result part finished

--- 2023-04-05 bachelorthesis/02-theory.tex
@@
 \\
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
+The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, peeters2016}
 \begin{equation}

### 2023-04-08 bachelorthesis/40-resolution.tex
#39: Split results part in separate, Plots for thesis, result part finished

% 4. Results

\chapter{Results and Discussion}
\label{chap:results}
\thispagestyle{empty}
\newpage

\section{Variation of Computational Resolution}
\label{sec:variationsofresolution}

At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without fearing numerical dissipation. Numerical dissipatation can therefore result to no formation of zonal flow stuctures, which cause an permanent turbulent state of the simulation. The goal behind this testing is to reduce \textbf{calculation time} and \textbf{costs} of the simulation. \\
Because of that, the criteria for the best resolution should be:
\begin{enumerate}
	\item[\textbf{(1)}] Subdued turbulence after \textbf{short} time periods
	\item[\textbf{(2)}] Stability for \textbf{long} time periods 
\end{enumerate}
and the following procedure will be applied for verification:
\begin{enumerate}
	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
	\item Reduce to knowm minimum number of grid points to verify result in general.
\end{enumerate}

\subsection{Benchmark}
\label{sub:benchmark}

Starting from the \enquote{Standard resolution with 6th order (S6)} [Tab. \ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref. \citenum{peeters2016} in Section IV. Note that, because of selected circular geometry the used gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$. In Fig. \ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is simular to the results from Ref. \citenum{peeters2016} with subdued turbulence after $\sim 3000 R/\vth$.

\includegraphicsHere{S6_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for benchmark
}{fig:eflux-ns16-nvpar64-nmu9}{}

As next step an approach to the finite heat flux threshold were made to verify the selection of the gradient length $\rlt$. As in Ref. \citenum{peeters2016} in Section V conclude is the finite heat flux threshold approximately located at a grendent length of $\rlt = 6.3$ for circular geometry [FIG. 4 of Ref. \citenum{peeters2016}]. Therefore following parameters were used:
\begin{gather*}
	\rlt \in [6.0, 6.3]~.
\end{gather*}
As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occur in the whole time domain, which is in agreement with Ref. \citenum{peeters2016}.

\includegraphicsHere{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ and $\rlt = 6.3$ for benchmark
}{fig:eflux-gradient-length}{}

\subsection{Reduction of parallel velocity grid points $\Nvpar$}
\label{sub:reduceNvpar}

In the following the number of grid points for the parallel velocity $\Nvpar$ is reduced to:
\begin{gather*}
	\Nvpar \in [16, 32, 48, 64]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visable that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period. But resolution $\Nvpar = 32$ is as well not acceptable according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000 R/\vth$.\\
So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution will be:
\begin{gather*}
	\boxed{\Nvpar = 48}
\end{gather*}
With this number of grid points the time till turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500 R/\vth$.

\includegraphicsHere{Comparison/Resolution/Nvpar/S6_rlt6.0_boxsize1x1_Ns16_Nvpar16-32-48-64_Nmu9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced parallel velocity grid points $\Nvpar$
}{fig:eflux-ns16-nvpar16-32-48-64-nmu9}{}

\subsection{Reduction of magnetic moment grid points $\Nmu$}
\label{sub:reduceNmu}

As next step the number of grid points for the magnetic moment $\Nmu$ were reduced with:
\begin{gather*}
	\Nmu \in [6, 9] ~.
\end{gather*}
As in Fig. \ref{fig:eflux-ns16-nvpar64-nmu6-9} shown, the reduction of grid points for the magnetic moment does not significantly impact the suppression of turbulence. The turbulence enters the stationary state in both cases after $\sim 3000 R/\vth$.

\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu6-9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced magnetic moment grid points $\Nmu$
}{fig:eflux-ns16-nvpar64-nmu6-9}{}

To conclude a curial result the number of grids point for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}. In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with the both criteria in mind, to the following resolution:
\begin{gather*}
	\boxed{\Nmu = 9}
\end{gather*}

\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar48_Nmu6-9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ and $\Nvpar = 48$ for reduced magnetic moment grid points $\Nmu$
}{fig:eflux-ns16-nvpar48-nmu6-9}{}

\subsection{Reduction of magnetic field grid points $\Ns$}
\label{sub:reduceNs}

In the final step the number of grid points for the magnetic field $\Ns$ get reduced with the following parameters:
\begin{gather*}
	\Ns \in [12, 16]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns12-16-nvpar64-nmu9} is clearly visible that the reduction to $\Ns = 12$ does not satisfy the criteria \textbf{(2)} because the the stationary state of the turbulence gets instabil after $\sim 2500 R/\vth$. This concludes to the following resolution:
\begin{gather*}
	\boxed{\Ns = 16}
\end{gather*}

\includegraphicsHere{Comparison/Resolution/Ns/S6_rlt6.0_boxsize1x1_Ns12-16_Nvpar64_Nmu9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced magnetic field grid points $\Ns$
}{fig:eflux-ns12-16-nvpar64-nmu9}{}

\subsection{Final Resolution for Simulation}
\label{sub:resolution}

Together with the results of Chapter \ref{sub:reduceNvpar}, \ref{sub:reduceNmu} and \ref{sub:reduceNs} the final resolution used in the upcoming simulations are displayed in Tab. \ref{tab:resolution} with reduced number of grid points for the parallel velocity $\Nvpar$.
\begin{center}
	\captionsetup{type=table}
	\begin{tabular}{l | ccccc | ccccc | c | cc}
		& $N_\mathrm{m}$ & $N_x$ & $N_\mathrm{s}$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_\mathrm{d}$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
		\hline
		S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
	\end{tabular}
	\captionof{table}{
		Resolution used in this paper: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
	}
	\label{tab:resolution}
\end{center}

### 2023-04-08 bachelorthesis/41-boxsizescan.tex
#39: Split results part in separate, Plots for thesis, result part finished

--- 2023-04-05 bachelorthesis/04-results.tex
@@
 
-\chapter{Results and Discussion}
-\label{chap:results}
-\thispagestyle{empty}
 \newpage
-
-\section{Variation of Computational Resolution}
-\label{sec:variationsofresolution}
-
-At the beginning of this work the goal is to change the used resolutuin to a minimum. The goal behind this research is to reduce Computational time and costs for the computation of the simulation.
-
 \section{Size Convergence of $\exb$ staircase pattern}
@@
 This chapter is an further iteration of the brief communication published in \enquote{Physics of Plasma}.
-It provides additional plots that was not necessary for the publication and some other formatting as well.
-The brief communication can be find in the appendix of this thesis or under \textbf{REF PAPER GITHUB}
+It provides additional plots and paragraphs that were not necessary for the publication.
+the format was changed as well because the publication was written in REV\TeX~4.1 and this bachelor thesis in \texttt{scrreprt}.
+The brief communication can be found in the appendix of this thesis or under \textbf{REF PAPER GITHUB}.
 %TODO
@@
 \begin{gather*}
-	\boxed{\NR \times \NB \in [1\times1,~2\times1,~3\times1,~4\times1]}~.
+	\NR \times \NB \in [1\times1,~2\times1,~3\times1,~4\times1]~.
 \end{gather*}
@@
 During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
+\\
 Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
@@
 The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
-\\
-
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
+\includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}{
 	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
@@
 \begin{gather*}
-	\boxed{\NR\times\NB \in [1\times1,~2\times2,~3\times3]}~.
+	\NR\times\NB \in [1\times1,~2\times2,~3\times3]~.
 \end{gather*}
@@
 Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern always occurs after $\sim 1000~R/\vth$ [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}\,(a)], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
 % Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
@@
 
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
+\includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes\\
+	\textbf{(b)} Time traces of $\hatwexbamp$ for isotropic increased box sizes
 }{fig:eflux-1x1-2x2-3x3-comparison}{}
@@
 % discuss sumewhat smaller radial size
-By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
+By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size [Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}\,(b), Fig.
+\ref{fig:wexb-stable-comparison}\,(b)].
 Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
@@
 \begin{gather*}
-	\boxed{\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]}~.
+	\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]~.
 \end{gather*}
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}\,(a)].
 The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
@@
 Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
-The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
+The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each [Fig.
+\ref{fig:eflux-3x1.5-2.5-3-5-comparison}\,(b)].
+\includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}{
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes\\
+	\textbf{(b)} Time traces of $\hatwexbamp$ for binormal increased box sizes
 }{fig:eflux-3x1.5-2.5-3-5-comparison}{}\bigskip
@@
 
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
+\includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison_thesis.pdf}{
 	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
@@
 }{fig:wexb-stable-comparison}{}
-
-\section{The finite heat flux threshold}
-\label{sub:threshold}
-
-In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
-Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
-This scan covers realizations with:
-\begin{gather*}
-	\boxed{\rlt \in [6.0,~6.2,~6.4]}~.
-\end{gather*}
-% what is the finite heat flux threshold
-In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
-The finite heat flux threshold, hence, is:
-\begin{gather*}
-	\boxed{\rlt|_\mathrm{finite} = 6.3 \pm 0.1}
-\end{gather*}
-in accordance to Ref. \citenum{peeters2016}.
-% Which stationay radial size occurs????
-Although the initial quasi-stationary turbulence in the former case is significantly longer compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
-% Does it confirm our previous tests
-Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.\\

### 2023-04-08 bachelorthesis/42-heatfluxthreshold.tex
#39: Split results part in separate, Plots for thesis, result part finished

--- 2023-04-05 bachelorthesis/04-results.tex
@@
 
-\chapter{Results and Discussion}
-\label{chap:results}
-\thispagestyle{empty}
 \newpage
-
-\section{Variation of Computational Resolution}
-\label{sec:variationsofresolution}
-
-At the beginning of this work the goal is to change the used resolutuin to a minimum. The goal behind this research is to reduce Computational time and costs for the computation of the simulation.
-
-\section{Size Convergence of $\exb$ staircase pattern}
-\label{sec:convergence}
-
-This chapter is an further iteration of the brief communication published in \enquote{Physics of Plasma}.
-It provides additional plots that was not necessary for the publication and some other formatting as well.
-The brief communication can be find in the appendix of this thesis or under \textbf{REF PAPER GITHUB}
-%TODO
-
-\subsection{Radial increased Box Size}
-\label{sub:radial}
-
-In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size. The scan covers the realizations: 
-\begin{gather*}
-	\boxed{\NR \times \NB \in [1\times1,~2\times1,~3\times1,~4\times1]}~.
-\end{gather*}
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
-The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
-% insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
-Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
-% yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.
-\citenum{peeters2016}.
-\bigskip
-
-% cautios words
-Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
-% very long initial quasi-stationary states
-Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
-% the \nzf=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
-During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
-Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
-% first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
-It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbamp$ in the latter two phases (not shown).
-This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
-% second important point: no secular growth of the box scale zonal mode
-Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
-% summary
-The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
-\\
-
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
-	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
-
-\newpage
-\subsection{Isotropic increased Box Size}
-\label{sub:isotropic}
-
-% what kind of analysis
-Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
-This scan covers the realizations:
-\begin{gather*}
-	\boxed{\NR\times\NB \in [1\times1,~2\times2,~3\times3]}~.
-\end{gather*}
-% mention shorter times scales for stabilization
-Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern always occurs after $\sim 1000~R/\vth$ [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
-% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
-%
-% the main result of this test: confirmation of wave length convergence
-As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
-
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
-}{fig:eflux-1x1-2x2-3x3-comparison}{}
-
-% discuss sumewhat smaller radial size
-By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
-Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
-
-\newpage
-\subsection{Binormal increased Box Size}
-\label{sub:binormal}
-
-% what kind of analysis
-In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
-This test covers the realizations:
-\begin{gather*}
-	\boxed{\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]}~.
-\end{gather*}
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
-The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
-Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
-The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
-The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no signifciant plateau at positive shear.
-Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
-The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
-}{fig:eflux-3x1.5-2.5-3-5-comparison}{}\bigskip
-
-\subsection{Staircase structures in Comparison}
-\label{sub:comparisonwexb}
-
-\includegraphicsHere{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
-	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
-The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
-	\begin{tabular}{l l l l l}
-		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, \\
-		                        & $t_{3\times 1}$   & $\in [43000, 45000]$, & $t_{4\times 1}$   & $\in [26000, 28000]$  \\
-		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   \\
-		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    &                   &                       \\
-		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   \\
-		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    & $t_{3\times 5}$   & $\in [1000, 3000]$    \\
-	\end{tabular}
-}{fig:wexb-stable-comparison}{}
-
 \section{The finite heat flux threshold}
@@
 \begin{gather*}
-	\boxed{\rlt \in [6.0,~6.2,~6.4]}~.
+	\rlt \in [6.0,~6.2,~6.4]~.
 \end{gather*}
 % what is the finite heat flux threshold
-In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
+In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$ [Fig.
+\ref{fig:eflux-6.0-6.2-6.4-comparison}].
 The finite heat flux threshold, hence, is:
@@
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.\\
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.
+
+\includegraphicsHere{Comparison/Gradient-Length/S6_rlt6.0-6.2-6.4_boxsize3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
+	Time traces of the heat conduction coefficient $\chi$ for different gradient lengths $\rlt$
+}{fig:eflux-6.0-6.2-6.4-comparison}{}

### 2023-04-08 evaluation/compare_boxsize_eflux_wexb_max.py

@@ -25,7 +25,10 @@ colors = [['#a11a5b', '#029e73', '#de8f05', '#0173b2'],
 # RADIAL ============================================================================================================================================
 
 
-plot.parameters(True, 32, (24,8), 300)
+#plot.parameters(32, (24,8), 300)
+
+#Thesis
+plot.parameters(54, (30,14), 300, tickwidth = 3)
 
 # File import and Create picture folder
 data = 'S6_rlt6.0'
@@ -43,18 +46,20 @@ if not os.path.exists(picDir):
 
 
 # Compare eflux and amplitude in time domain
-fig, (ax_eflux, ax_wexb_max) = plt.subplots(2, 1, figsize = (24,8), sharex=True)
+fig, (ax_eflux, ax_wexb_max) = plt.subplots(2, 1, sharex=True)
 
 boxsize = [r'1$\times$1', r'2$\times$1', r'3$\times$1', r'4$\times$1']
 
 #ax_eflux.set_title(r'$N_s$ = 16,   $N_{\mathrm{vpar}}$ = 48,   $N_{\mathrm{\mu}}$ = 9', pad=20)
 #ax_eflux.set_xlabel(r'$t~[R/ \nu_{\mathrm{th}}]$')
 ax_eflux.set_ylabel(r'$\chi~[\rho^2 \nu_{\mathrm{th}} / R]$')
-ax_eflux.yaxis.set_label_coords(-0.045,0.5)
+#ax_eflux.yaxis.set_label_coords(-0.045,0.5)
+ax_eflux.yaxis.set_label_coords(-0.065,0.5)
 
 ax_wexb_max.set_xlabel(r'$t~[R/ \nu_{\mathrm{th}}]$')
 ax_wexb_max.set_ylabel(r'$|\widehat{\omega}_{\mathrm{E \times B}}|_{n_\mathrm{ZF}}~[\nu_{\mathrm{th}}/R]$')
-ax_wexb_max.yaxis.set_label_coords(-0.045,0.5)
+#ax_wexb_max.yaxis.set_label_coords(-0.045,0.5)
+ax_wexb_max.yaxis.set_label_coords(-0.065,0.5)
 
 
 fourier_index = 1
@@ -95,8 +100,9 @@ for i, n, k, c in zip(f, boxsize, max_index, color_rad):
     ax_wexb_max.set_xlim(xmin=0, xmax=x_max)
     ax_wexb_max.set_ylim(ymin=0, ymax=0.29)
 
[…]

### 2023-04-08 evaluation/compare_boxsize_wexb.py

@@ -28,8 +28,6 @@ lineargrowth = pd.read_csv(homepath + 'data/linear_growthrate_ITG.dat', index_co
 lineargrowth_rlt = lineargrowth['gamma_N'][6.0]
 lineargrowth_rlt_color = '#949494'
 
-plot.parameters(True, 40, (24,8), 300, linewidth=2)
-
 #colors = [['#db5f57', '#dbb757', '#a7db57', '#57db5f'],
 # 		             ['#57a7db', '#57dbb7', '#57db5f'],
 #		  ['#5f57db', '#57a7db', '#b757db', '#db57a7']]
@@ -135,7 +133,10 @@ def box_plot(fig, time_label, plot_label,
             
             if YLABEL:
                 ax.set_ylabel(r'$\langle\omega_{\mathrm{E \times B}}\rangle~[\nu_{\mathrm{th}}/R]$')
-                ax.yaxis.set_label_coords(-0.25/box_max, 0.5, transform=ax.transAxes)
+                #ax.yaxis.set_label_coords(-0.25/box_max, 0.5, transform=ax.transAxes)
+                
+                # Thesis
+                ax.yaxis.set_label_coords(-0.35/box_max, 0.5, transform=ax.transAxes)
 
 
         ax.set_xlim(xmin=0, xmax = rad_boxsize)
@@ -210,6 +211,8 @@ path_rad = ['boxsize4x1/Ns16/Nvpar48/Nmu9', 'boxsize3x1/Ns16/Nvpar48/Nmu9', 'box
 filename_rad = [homepath + 'data/'+data+'/'+i+'/data.h5' for i in path_rad]
 file_rad = [h5py.File(i,"r+") for i in filename_rad]
 
+plot.parameters(40, (24,8), 300, linewidth=2)
+
 boxes_rad   = [4, 3, 2, 1]
 shifts_rad  = [20, 2, -5, 60]
 boxsize_rad = [r'4 \times 1\;\:\:', r'3 \times 1\;\:\:', r'2 \times 1\;\:\:', r'1 \times 1\;\:\:']
@@ -243,7 +246,7 @@ path_iso = ['boxsize3x3/Ns16/Nvpar48/Nmu9', 'boxsize2x2/Ns16/Nvpar48/Nmu9', 'box
 filename_iso = [homepath + 'data/'+data+'/'+i+'/data.h5' for i in path_iso]
[…]

### 2023-04-08 evaluation/eval.py

@@ -21,11 +21,11 @@ print('\n<------------------------- Evaluation ------------------------->\n')
 
 # PARAMETER =========================================================================================================================================
 
-plot.parameters(True, 22, (24,8), 300)
+plot.parameters(40, (24,8), 300, linewidth = 4, tickwidth = 3)
 
-rlt, boxsize, Ns, Nvpar, Nmu = 6.2, '3x3', 16, 48, 9
+rlt, boxsize, Ns, Nvpar, Nmu = 6.0, '1x1', 16, 64, 9
 
-ALL, EVOLUTION, SELECTION = True, True, True
+ALL, EVOLUTION, SELECTION = False, False, False
 
 print('Input:', rlt, boxsize, Ns, Nvpar, Nmu, ALL, EVOLUTION, SELECTION, '\n')
 
@@ -84,7 +84,7 @@ lineargrowth_rlt_color = 'grey'
 
 eflux_data, time = zonalflow.get_eflux_time(f)
 
-plot.eflux_time(time, eflux_data, (24,8))
+plot.eflux_time(time, eflux_data, (24,8), xlim = (0, 6000), ylim = (0, 25))
 plt.savefig(picpath+data+'_'+resolution+'_eflux.pdf', bbox_inches='tight')
 
 # Omega_ExB =========================================================================================================================================

### 2023-04-08 python/plot.py

@@ -7,20 +7,31 @@ from cycler import cycler
 
 
 # Plot parameters
-def parameters(usetex, fontsize, figsize, dpi, 
-               linewidth = 1.5, ticklength = 10, ticksadditional = True, ticksdirc = 'in', tickspad = 10,
-               colorpalette = sns.color_palette("colorblind", as_cmap=True)):
+def parameters(fontsize, figsize, dpi, usetex = True,
+               linewidth = 3, colorpalette = sns.color_palette("colorblind", as_cmap=True),
+               tickwidth = 1.5, ticklength = 10, tickspad = 10, ticksdirc = 'in', ticksadditional = True,
+               legendpad = -2, legendontop = True):
+    
+    # TEXT ==========================================================
     
     plt.rcParams['text.usetex'] = usetex
     plt.rcParams['font.family'] = 'serif'
     plt.rcParams['font.size'] = fontsize
     
+    # FIGURE ========================================================
+    
     plt.rcParams['figure.figsize'] = figsize
     plt.rcParams['figure.dpi'] = dpi
     
+    # STYLE =========================================================
+    
+    plt.rcParams['axes.labelpad'] = 15
+    plt.rcParams['lines.linewidth'] = linewidth
     plt.rcParams['axes.prop_cycle'] = cycler('color', colorpalette)
     
-    plt.rcParams['axes.linewidth'] = linewidth
+    # TICKS AND FRAME ===============================================
+    
+    plt.rcParams['axes.linewidth'] = tickwidth
     
     plt.rcParams['xtick.top'] = ticksadditional
     plt.rcParams['xtick.bottom'] = ticksadditional
@@ -29,22 +40,30 @@ def parameters(usetex, fontsize, figsize, dpi,
     plt.rcParams['ytick.right'] = ticksadditional
     
     plt.rcParams['xtick.major.size'] = ticklength
-    plt.rcParams['xtick.major.width'] = linewidth
+    plt.rcParams['xtick.major.width'] = tickwidth
     
     plt.rcParams['ytick.major.size'] = ticklength
-    plt.rcParams['ytick.major.width'] = linewidth
+    plt.rcParams['ytick.major.width'] = tickwidth
     
[…]

### 2023-04-18 bachelorthesis/31-simulation.tex
#39: Seperate sections in tex files, New LaTeX Workshop commands
#39: Include paper, data, monitor code and status file in appendix, codestyle, chapter about methods, minor changes, flowdiagram style, new references

--- 2023-04-05 bachelorthesis/03-methods.tex
@@
-% 3. Methods and MAterial
-
-\chapter{Methods and Material}
-\label{chap:methods}
-\thispagestyle{empty}
-\newpage
-
 \section{Simulation Setup}
@@
 \citenum{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
-The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
-It has been carefully verified that this modification preserves the same physical outcome as the original study.
+The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference.
 A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.
@@
 	\begin{tabular}{l | ccccc | ccccc | c | cc}
-		& $N_m$ & $N_x$ & $N_s$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_d$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
+		& $N_\mathrm{m}$ & $N_x$ & $N_\mathrm{s}$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_\mathrm{d}$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
 		\hline
-		S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
+		S6   & 21    & 83    & 16    & 64                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
 	\end{tabular}
@@
 	}
-	\label{tab:resolution}
+	\label{tab:resolutionBenchmark}
 \end{center}
-
-\section{btrzx1 Cluster}
-\label{sec:cluster}
-
-\section{Restart Script for Simulation btrzx1 Cluster}
-\label{sec:restartscript}

### 2023-04-18 bachelorthesis/41-resolution.tex
#39: Seperate sections in tex files, New LaTeX Workshop commands
#39: Include paper, data, monitor code and status file in appendix, codestyle, chapter about methods, minor changes, flowdiagram style, new references

--- 2023-04-08 bachelorthesis/40-resolution.tex
@@
-% 4. Results
-
-\chapter{Results and Discussion}
-\label{chap:results}
-\thispagestyle{empty}
-\newpage
-
 \section{Variation of Computational Resolution}
@@
 \citenum{peeters2016} in Section IV.
-Note that, because of selected circular geometry the used gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$.
+Note that, because of selected circular cencentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref.
+\citenum{peeters2016} for $s$-$\alpha$ geometry.
 In Fig.

### 2023-04-18 bachelorthesis/62-data.tex
#39: Include paper, data, monitor code and status file in appendix, codestyle, chapter about methods, minor changes, flowdiagram style, new references

\section{Simulation parameter}
\label{append:data}

\begin{center}
    \captionsetup{type=table}
    \begin{tabular}{l | c | l | l l l l l l l}
            & $\rlt$ & box size  & $\Ns$ & $\Nvpar$ & $\Nmu$ & $d_\mathrm{tim}$ & $k\rho^\mathrm{max}$ & $N_\mathrm{mod}$ & $N_x$ \\ \hline 
        1   & 6.0    & 1x1       & 12    & 16       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
        2   & 6.0    & 1x1       & 12    & 32       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
        3   & 6.0    & 1x1       & 12    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        4   & 6.0    & 1x1       & 12    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        5   & 6.0    & 1x1       & 16    & 16       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        6   & 6.0    & 1x1       & 16    & 32       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        7   & 6.0    & 1x1       & 16    & 48       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
        8   & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        9   & 6.0    & 1x1       & 16    & 48       & 9      & 0.025            & 1.4                  & 21               & 83    \\ 
        10  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 0.7                  & 11               & 83    \\ 
        11  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 43    \\ 
        12  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 63    \\ 
        13  & 6.0    & 1x1       & 16    & 64       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
        14  & 6.0    & 1x1       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        15  & 6.0    & 2x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        16  & 6.0    & 2x2       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        17  & 6.0    & 3x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        18  & 6.0    & 3x1.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        19  & 6.0    & 3x1.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        20  & 6.0    & 3x2.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        21  & 6.0    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        22  & 6.0    & 3x5       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        23  & 6.0    & 4x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        24  & 6.2    & 2x2       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        25  & 6.2    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        26  & 6.3    & 1x1       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
        27  & 6.4    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
    \end{tabular}
    \captionof{table}{Parameters of simulations performed for this thesis}
    \label{tab:parametersSet}
\end{center}

\newpage

\begin{center}
    \captionsetup{type=table}
    \begin{tabular}{l | c | l | r r l | l l | l }
            & $\rlt$ & box size & time    & timestep    & error\_index & stable   & $\nzf$ & backup \\ \hline 
        1   & 6.0    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
        2   & 6.0    & 1x1      & 12000.0 & 20000       & ~            & False    & 0      & True   \\ 
        3   & 6.0    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
        4   & 6.0    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
        5   & 6.0    & 1x1      & 12000.0 & 20000       & ~            & False    & 0      & True   \\ 
        6   & 6.0    & 1x1      & 12000.0 & 20000       & ~            & False    & 0      & True   \\ 
        7   & 6.0    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
        8   & 6.0    & 1x1      &  6000.0 & 10000       & ~            & True     & 1      & True   \\ 
        9   & 6.0    & 1x1      &  7125.0 & 10000       & ~            & True     & 1      & True   \\ 
        10  & 6.0    & 1x1      &  6000.0 & 10000       & ~            & True     & 1      & True   \\ 
        11  & 6.0    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
        12  & 6.0    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
        13  & 6.0    & 1x1      & 12000.0 & 20000       & ~            & True     & 1      & True   \\ 
        14  & 6.0    & 1x1      & 12000.0 & 20000       & ~            & True     & 1      & True   \\ 
        15  & 6.0    & 2x1      & 18000.0 & 30000       & ~            & True     & 2      & True   \\ 
        16  & 6.0    & 2x2      & 18000.0 & 30000       & ~            & True     & 2      & True   \\ 
        17  & 6.0    & 3x1      & 48000.0 & 80000       & ~            & True     & 3      & True   \\ 
        18  & 6.0    & 3x1.5    & 16840.3 & 28068       & ~            & True     & 4      & True   \\ 
        19  & 6.0    & 3x1.5    & 14170.0 & 23618       & 23618        & True     & 3      & True   \\ 
        20  & 6.0    & 3x2.5    &  6000.0 & 10000       & ~            & True     & 3, 4   & True   \\ 
        21  & 6.0    & 3x3      &  7847.0 & 13083       & ~            & True     & 4      & True   \\ 
        22  & 6.0    & 3x5      &  4769.0 &  7958       & ~            & True     & 4      & True   \\ 
        23  & 6.0    & 4x1      & 30000.0 & 50000       & 47084        & True     & 4      & True   \\ 
        24  & 6.2    & 2x2      &  6000.0 & 10000       & ~            & True     & 2      & True   \\ 
        25  & 6.2    & 3x3      & 14682.7 & 24475       & 14473-16132  & True     & 3      & True   \\ 
        26  & 6.3    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
        27  & 6.4    & 3x3      & 12000.0 & 20000       & ~            & False    & 0      & True   \\ 
    \end{tabular}
    \captionof{table}{Time steps with index which should be exclude in data analysis. Additional the stable column which indicates if turbulence subdued in simulation with the corresponding zonal flow mode number $\nzf$ (0 stands for turbulence not stable) and if the data is saved on the NAS of TPV (backup column)}
    \label{tab:parametersTime}
\end{center}

\newpage

\begin{center}
    \captionsetup{type=table}
    \begin{tabular}{l | l }
            & path                                                              \\ \hline
        1   & ./data/S6\_rlt6.0/boxsize1x1/Ns12/Nvpar16/Nmu6                    \\ 
        2   & ./data/S6\_rlt6.0/boxsize1x1/Ns12/Nvpar32/Nmu6                    \\
        3   & ./data/S6\_rlt6.0/boxsize1x1/Ns12/Nvpar48/Nmu9                    \\ 
        4   & ./data/S6\_rlt6.0/boxsize1x1/Ns12/Nvpar64/Nmu9                    \\ 
        5   & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar16/Nmu9                    \\ 
        6   & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar32/Nmu9                    \\ 
        7   & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar48/Nmu6                    \\ 
        8   & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar48/Nmu9                    \\ 
        9   & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar48/Nmu9/dtim0.025          \\ 
        10  & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar48/Nmu9/krhomax0.70/Nmod11 \\ 
        11  & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar48/Nmu9/Nx43               \\ 
        12  & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar48/Nmu9/Nx63               \\ 
        13  & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu6                    \\ 
        14  & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9                    \\ 
        15  & ./data/S6\_rlt6.0/boxsize2x1/Ns16/Nvpar48/Nmu9                    \\ 
        16  & ./data/S6\_rlt6.0/boxsize2x2/Ns16/Nvpar48/Nmu9                    \\ 
        17  & ./data/S6\_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9                    \\ 
        18  & ./data/S6\_rlt6.0/boxsize3x1.5/Ns16/Nvpar48/Nmu9                  \\ 
        19  & ./data/S6\_rlt6.0/boxsize3x1.5/Ns16/Nvpar48/Nmu9/Broken           \\ 
        20  & ./data/S6\_rlt6.0/boxsize3x2.5/Ns16/Nvpar48/Nmu9                  \\ 
        21  & ./data/S6\_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
        22  & ./data/S6\_rlt6.0/boxsize3x5/Ns16/Nvpar48/Nmu9                    \\ 
        23  & ./data/S6\_rlt6.0/boxsize4x1/Ns16/Nvpar48/Nmu9                    \\ 
        24  & ./data/S6\_rlt6.2/boxsize2x2/Ns16/Nvpar64/Nmu9                    \\ 
        25  & ./data/S6\_rlt6.2/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
        26  & ./data/S6\_rlt6.3/boxsize1x1/Ns16/Nvpar64/Nmu9                    \\ 
        27  & ./data/S6\_rlt6.4/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
    \end{tabular}
    \captionof{table}{Data location for each simulation}
    \label{tab:parametersPath}
\end{center}

### 2023-04-29 bachelorthesis/01-dedication.tex
#39: Dedication, Acknowledgement, Abstract (ENG, DE), Eidstattliche Versicherung, Declaration

\thispagestyle{empty}

\begin{center}
    \vspace*{3cm}
    \begin{tikzpicture}
        \clip(0,0)circle[x radius=3cm, y radius=4cm];
        \node{\includegraphics[width=6cm]{blacky1_black_white.jpg}};
    \end{tikzpicture}
    \vspace{2cm}\\
    This thesis is dedicated to my cat \textbf{Blacky}\\ which is to this date 19 years old and followed me\\ since I was 5 years through my life.\\You were always a big part of my life and\\ I will miss you everyday when you are gone.\\
    \bigskip
    Manuel Lippert\\
    \today
\end{center}

### 2023-04-29 bachelorthesis/02-acknowledgement.tex
#39: Dedication, Acknowledgement, Abstract (ENG, DE), Eidstattliche Versicherung, Declaration

\chapter*{Acknowledgement}
\label{chap:acknowledgement}

First of all, I would like to express my greatest gratitude to my supervisor Arthur Peeters and Florian Rath. I would like to thank both of you for your guidance and support during the submission process of the briefcommunication and the helpful hands during the the progress of this thesis and the restart script. 
\bigskip
I would like to thank Bernhard Winkler and Markus Hilt for the great technical support regarding my questions to the \texttt{btrzx1}-cluster which helps me to develope the restart script as it is now. 
\bigskip
I would also like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend. In Addition to that I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time sharing the office together. 
\bigskip
I will also thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get better in spelling.
\bigskip
Outside academia, I would like to express my gratitude to my parents, my brother and his wife, my sister and Hanna Schütz for there encouragement, endurance and support during the time of this thesis.

### 2023-04-29 bachelorthesis/03-abstract.tex
#39: Dedication, Acknowledgement, Abstract (ENG, DE), Eidstattliche Versicherung, Declaration

\chapter*{Abstract}
\label{chap:abstractENG}

Ion temperature gradient driven turbulence (ITG) close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures. Such pattern formation has been observed in local gradient-driven flux-tube simulations as well as global gradient-driven and global flux-driven studies.

To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimal minimum resolution for the performed simulations.

For convenience a \texttt{python} script (\texttt{slurm\_monitor.py}) was written which monitors the simulation on the \texttt{btrzx1}-cluster and starts/restarts until the completion criteria is fulfilled.

Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit. This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG driven turbulence.

\chapter*{Zusammenfassung}
\label{chap:abstractDE}

Ionen-Temperaturgradienten getriebener Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal FLow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf. Solche Strukturbildungen wurden sowohl in lokal gradientengetriebene Flussschlauchsimulationen als auch in global gradientengetriebenen und global flussgetriebenen Untersuchungen entdeckt.

Um den Aufwand der Berechnungen der Simulationen zu reduzieren wurden mehrere kleinere Inputparameter für GKW (Gyro Kinetic Workshop) getestet um die optimal kleinste Auflösung zu finden ausgeführten Simulationen zu finden.

Zur Einfachheit wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist.

Weiterhin wurde doch mehrere radiale konvergierende Boxgrößen-Scans gezeigt, dass eine Mesoskalengröße von $\sim 57.20-76.27\,\rhoth$ inhärent zur ITG getriebenen Turbulenz mit Cyclone Base Parameter in lokalen Limit ist. Dieses Ergebnis impliziert auch, dass eine typische Skala für den lawinenenartig Transport inhärent für die ITG getriebene Turbulenz ist.

### 2023-04-29 bachelorthesis/04-declaration.tex
#39: Dedication, Acknowledgement, Abstract (ENG, DE), Eidstattliche Versicherung, Declaration

\chapter*{Declaration}
\label{chap:declaration}

The author declares that the material presented in this thesis is his own work, unless explicitly stated otherwise. Chapter \ref{sec:convergence} and \ref{sec:threshold} is based on
\bigskip
\begin{itemize}
    \item [] \bibentry{Lippert2023}
\end{itemize}
\bigskip
and is an further iteration of the brief communication published in the journal \enquote{Physics of Plasma}. It provides additional plots and paragraphs that were not necessary to be included in the final publication. The format was changed as well because the publication was written in REV\TeX~4.1 and this bachelor thesis in \texttt{scrreprt}. The brief communication can be found in Appendix \ref{append:paper}.

### 2023-04-29 bachelorthesis/42-boxsizescan.tex
#39: Include paper, data, monitor code and status file in appendix, codestyle, chapter about methods, minor changes, flowdiagram style, new references
#39: Dedication, Acknowledgement, Abstract (ENG, DE), Eidstattliche Versicherung, Declaration

--- 2023-04-08 bachelorthesis/41-boxsizescan.tex
@@
 
-This chapter is an further iteration of the brief communication published in \enquote{Physics of Plasma}.
-It provides additional plots and paragraphs that were not necessary for the publication.
-the format was changed as well because the publication was written in REV\TeX~4.1 and this bachelor thesis in \texttt{scrreprt}.
-The brief communication can be found in the appendix of this thesis or under \textbf{REF PAPER GITHUB}.
 %TODO

### 2023-04-29 bachelorthesis/70-eid.tex
#39: Dedication, Acknowledgement, Abstract (ENG, DE), Eidstattliche Versicherung, Declaration

\newpage
\chapter*{Eidesstattliche Erklärung}
\label{sec:eid}
\thispagestyle{empty}

\addcontentsline{toc}{chapter}{Eidesstattliche Erklärung}
\vspace*{0.1cm}
Hiermit erkläre ich, Manuel Lippert, dass ich die vorliegende Arbeit selbständig und ohne Benutzung anderer als der angegebenen Hilfsmittel angefertigt habe.
\bigskip
Alle Stellen, die wörtlich oder sinngemäß aus veröffentlichten oder nicht veröffentlichten Schriften entnommen wurden, sind als solche kenntlich gemacht.
\bigskip
Die Arbeit hat in gleicher oder ähnlicher Form noch keiner anderen Prüfungsbehörde vorgelegen.
\vspace{3cm}

\noindent Bayreuth, den \today
\begin{flushright}
$\overline{~~~~~~~~~~\mbox{Manuel Lippert}~~~~~~~~~~}$
\end{flushright}

### 2023-04-29 bachelorthesis/02-acknowledgement.tex
#39: Fixed Linebreak

@@
 \bigskip
+
 I would like to thank Bernhard Winkler and Markus Hilt for the great technical support regarding my questions to the \texttt{btrzx1}-cluster which helps me to develope the restart script as it is now.
 \bigskip
+
 I would also like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend.
@@
 \bigskip
-I will also thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get better in spelling.
-\bigskip
-Outside academia, I would like to express my gratitude to my parents, my brother and his wife, my sister and Hanna Schütz for there encouragement, endurance and support during the time of this thesis.
+
+I will also thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get better in spelling. \bigskip
+
+Outside academia, I would like to express my gratitude to my parents, my brother and his wife, my sister and my girlfriend for their encouragement, endurance and support during the time of this thesis.
+I know sometimes I am an workaholic and I need to be reminded outside my academic work.

### 2023-05-01 bachelorthesis/31-simulation.tex
#39: Reorganize Methods chapter

@@
 \end{center}
+
+Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
+Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
+
+Consistent with Ref.
+\citenum{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.
+\citenum{peeters2016,peeters2009}.

### 2023-05-01 bachelorthesis/32-diagnostics.tex
#39: Reorganize Methods chapter

--- 2023-04-05 bachelorthesis/02-theory.tex
@@
-% 2. Theory
+\newpage
+\section{Diagnostics}
+\label{sec:diagnostics}
 
-\chapter{Theory}
-\label{chap:theory}
-\thispagestyle{empty}
-\newpage
-
-In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
-Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
-Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
-The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
-Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
-\\
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
-\begin{equation}
-	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
-	\label{eq:shearingrate}
-\end{equation}
-where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
-The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
-\begin{equation}
-\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
-\end{equation}
-The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
-\\
-Consistent with Ref.~\cite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.\cite{peeters2016,peeters2009}.
-\newpage
 In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
@@
 % explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
-defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
+defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $ and the radial box size $L_\xcoord$.
 % define the zonal flow modes' amplitude in terms of the shearing rate

### 2023-05-01 bachelorthesis/33-cluster.tex
#39: Reorganize Methods chapter

\section{btrzx1 Cluster}
\label{sec:cluster}

The simulation itself will be performed on the \texttt{btrzx1}-cluster of the University Bayreuth. This cluster has a wall time from 24 hours, an total amount of 345 compute nodes with
two AMD Epyc Processors (2nd generation) with 16 cores each and 128 GB of main memory each node and as resource manager \texttt{Slurm}. \cite{btrzx1}

The simulations are performed on the \texttt{/scratch} directory because the \texttt{/home} folder has an disk space limit of \SI{80}{\giga\byte} which is not enough to perform long runs of GKW.

### 2023-05-01 bachelorthesis/34-restartscript.tex
#39: Reorganize Methods chapter

\section{Restart Script for Simulation}
\label{sec:restartscript}

As stated in Chapter \ref{sec:cluster} the \texttt{btrzx1}-cluster has a wall time of 24 hours but since simulations typical run longer than 24 hours one have to restart the simulation after the wall time gets exceeded. Some simulation need even longer so that the simulation itself needs multiple weeks to finish with a result which makes the restart of the simulation a daily task that could be easily forget. So to eliminate this minor inconveniences the following restart script were written.

As script language \texttt{python} was chosen over shell because of its easiness to write code and the variety of tools which can interact with the \texttt{bash}-shell. \texttt{bash} was also considered but after some issue occurred regarding the mechanism to check wether a file exist or not, \texttt{python} was selected.

The script was named \texttt{slurm\_monitor.py} to indicate that this script works with the the \texttt{Slurm} resource manager. The source code was pushed to the gkw repository on BitBucket \cite{slurmmonitor} or can be found in Appendix \ref{append:slurmcode}

\subsection{How to run Restart Script in Terminal}
\label{sub:codeRun}

To run the restart script \texttt{python3} has to be installed on the system and should be accessible by the command line. With the following command can be the monitoring started:
\begin{lstlisting}[language=Bash]
python3 -u slurm_monitor.py --job-name $JOBNAME
\end{lstlisting}
Additional parser options can be added to the command above which will be summarized in the upcoming Chapter \ref{sub:codeFeatures}. The script was build to run in the background of the terminal for which there are two approaches:
\begin{enumerate}
    \item \texttt{nohup}: Runs command in the background while terminal can still be used or closed. \cite{nohup} The command for this method would be: \\
\begin{lstlisting}[language=Bash]
nohup python3 -u slurm_monitor.py --job-name $JOBNAME &> /dev/null &
\end{lstlisting}
and to cancel the monitoring:
\begin{lstlisting}[language=Bash]
python3 -u slurm_monitor.py --job-name $JOBNAME --kill
\end{lstlisting}
    \item \texttt{screen}: Opens a virtual terminal session which is detached from the main terminal and can be entered and closed. \cite{screen} Here the command would be:
\begin{lstlisting}[language=Bash]
screen -S $SESSION #Create screen session
python3 -u slurm_monitor.py --job-name $JOBNAME --verbose
\end{lstlisting}
and to cancel the monitoring use the keyboard shortcut \keys{\ctrl + C} or kill screen itself with \keys{\ctrl + D}. To leave the screen use ( \keys{\ctrl + A} ) + \keys{D} and to enter the screen use:
\begin{lstlisting}[language=Bash]
screen -r $SESSION
\end{lstlisting}
\end{enumerate}

\newpage
\subsection{Features of Restart Script}
\label{sub:codeFeatures}

The restart script has several features which can be used via the command line interface as parser options. The following features with the corresponding parser option were implementing:
\begin{itemize}
    \item The script has no requirements because it runs on the standard \texttt{python3} library
    \item \textbf{\textcolor{red}{!Required!}} Set job name not longer than 8 characters to identify simulation\\
          \texttt{-j [JOBNAME], -\/-job-name [JOBNAME]}
    \item Creates jobscript file with defined content\\
          (look into variable \texttt{jobscriptContent} to add more option) \\
          \texttt{-\/-jobscript [JOBSCRIPTFILE]~~~~~~~(default=jobscript-create)}\\
          \texttt{-\/-nodes [NODES]~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~(default=3)}\\
          \texttt{-\/-ntask-per-node [TASKS]~~~~~~~~~~~~~~~~~~~~~~~~(default=32)}\\
          \texttt{-\/-walltime [WALLTIME]~~~~~~(d-hh:mm:ss)~(default=0-24:00:00)}
    \item Start/Restarts simulation until job criteria is suffused \\
          \texttt{-n [TIMESTEPS]~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\,\,(default=10000)}\\
          \texttt{-\/-restartfile [RESTARTFILE]~~~~~~~~~~~~~~~~(default=FDS.dat)}
    \item Makes backup after each run before Restart and Restore files after failed run\\
          (uses \texttt{rsync} command line utility)\\
          Quicklinks: local (simulation folder), home (home folder) \\
          \texttt{-b [LOCATION], -\/-backup [LOCATION]~~~~~~~~~~~\,(default=None)}
    \item Reset option after run fails and dump files were written\\
          (rely on \texttt{h5py}, \texttt{pandas} and \texttt{numpy} which get installed by the script itself)\\
          \texttt{-r, -\/-reset~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\,(default=False)}
    \item Creates status file with current status and progress bars and updates it dynamically\\
          \texttt{-\/-statusfile [STATUSFILE]~~~~~~~~~~~~~~~(default=status.txt)}
    \item Set form of the output table in status file\\
          Options: fancy (round box), universial (crossplattform), None (No frame)\\
          \texttt{-\/-format [FORMATTABLE]~~~~~~~~~~~~~~~~~~~~~~~~(default=None)}
    \item Sends mail at the beginning, end and restart with status file as attachment\\
          (\texttt{mailx} other equivalent has to be installed, look into \texttt{send\_mail} function for more info)\\
          \texttt{-\/-mail [MAILADDRESS]~~~~~~~~~(mail@server.de)~(default=None)}\\
          \texttt{-\/-restart-mail ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~(default=False)}
    \item Kills monitoring process by using the \texttt{PID} number\\
          \texttt{-\/-kill~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~(default=False)}
    \item Set sleep interval after which the status of the current simulation should be checked\\
          \texttt{-\/-refresh-rate [SLEEPTIME]~~~~~~~~~~~~~~~~~~~~~(default=300)}
    \item Print output of status file table to console\\
          \texttt{-v, -\/-verbose~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\,(default=False)}
\end{itemize}

%options:
%  -h, --help            show this help message and exit
%
%required arguments:
%  -j [JOBNAME], --job-name [JOBNAME]
%                        job name not longer than 8 character
%
%additional arguments:
%  -n [TIMESTEPS]        required timesteps                   (default=10000)
%  -r, --reset           Uses Dumpfiles to reset Simulation   (default=False)
%  -v, --verbose         activate output of script            (default=False)
%  -b [BACKUP], --backup [BACKUP]
%                        backup location for files            (default=None)
%                        - local (creates backup in simulation folder)
%                        - home  (creates backup in home folder)
%  --jobscript [JOBSCRIPTFILE]
%                        jobscript to run SLURM job           (default=jobscript-create)
%  --ntask-per-node [TASKS]
%                        MPI task per node                    (default=32)
%  --nodes [NODES]       number of nodes                      (default=3)
%  --walltime [WALLTIME]
%                        walltime of server (d-hh:mm:ss)      (default=0-24:00:00)
%  --mail [MAIL]         mail address (mail@server.de)        (default=None)
%  --restart-mail        mail after every restart             (default=False)
%  --statusfile [STATUSFILE]
%                        file with output from nohup command  (default=status.txt)
%  --restartfile [RESTARTFILE]
%                        restart file with data               (default=FDS.dat)
%  --format [FORMATTABLE]
%                        format of output table               (default=none)
%                        - fancy (round box)
%                        - universal (crossplattform)
%                        - none (No frame around box)
%  --refresh-rate [SLEEPTIME]
%                        time interval to check status in sec (default=300)
%  --kill                kills monitor process                (default=False)
%

\newpage

\subsection{General Structure}
\label{sub:codeStructure}

\newcommand{\distance}{2.5cm}

\begin{center}
    \begin{tikzpicture}[node distance=2cm]

        \node (start)        [startstop]                                            {Start};
        \node (init)         [init, below of = start]                               {Initialize};
        \node (startLoop)    [startstop, below of = init]                           {Loop};
        \node (checkStatus)  [process, below of=startLoop]                          {Check Status};
        \node (checkData)    [process, below of=checkStatus, yshift = -2cm]         {Check Data};
        \node (checkPending) [decision, below of=checkStatus, xshift=\distance]     {Pending};
        \node (checkRunning) [decision, below of=checkStatus, xshift=-\distance]    {Running};
        \node (error)        [decision, below of=checkData, xshift=-\distance]      {Errors};
        \node (checkPoint)   [decision, below of=checkData, xshift=\distance]       {Checkpoint};
        \node (backup)       [process,  below of=error, xshift=-\distance]          {Backup Data};
        \node (restore)      [process,  below of=checkPoint, xshift=\distance]      {Restore Data};
        \node (reset)        [process,  below of=checkData, yshift=-2cm]            {Reset Data};
        \node (checkStep)    [decision, below of = reset]                           {Enough Timesteps};
        \node (end)          [startstop, below of=checkStep, xshift=4cm]            {End};
        \node (restart)      [init, below of=checkStep, xshift=-4cm]                {Start/Restart};

        \draw [arrow] (start)          --                                           (init);
        \draw [arrow] (init)           --                                           (startLoop);
        \draw [arrow] (startLoop)      --                                           (checkStatus);
        \draw [arrow] (checkStatus)    -|                                           (checkRunning);
        \draw [arrow] (checkRunning)   -- node[anchor=south] {No}                   (checkPending);
        \draw [arrow] (checkPending)   -- node[anchor=north west] {Yes} + ( 3,0) |- (startLoop);
        \draw [arrow] (checkRunning)   -- node[anchor=north east] {Yes} + (-3,0) |- (startLoop);
        \draw [arrow] (checkPending)   |- node[anchor=north west] {No}              (checkData);
        \draw [arrow] (checkData)      -|                                           (error);
        \draw [arrow] (error)          -- node[anchor=south] {Yes}                  (checkPoint);
        \draw [arrow] (error)          -| node[anchor=south east] {No}              (backup);
        \draw [arrow] (checkPoint)     -| node[anchor=south west] {No}              (restore);
        \draw [arrow] (checkPoint)     |- node[anchor=north west] {Yes}             (reset);
        \draw [arrow] (backup)         --                                           (checkStep);
        \draw [arrow] (reset)          --                                           (checkStep);
        \draw [arrow] (restore)        --                                           (checkStep);
        \draw [arrow] (checkStep)      -| node[anchor=south east] {No}              (restart);
        \draw [arrow] (checkStep)      -| node[anchor=south west] {Yes}             (end);
        \draw [arrow] (restart)        -- + (-3,0) |-                               (startLoop);
        \draw [arrow] (end)            -- + ( 3,0) |-                               (start);

    \end{tikzpicture}
\end{center}

\newpage

\subsection{Status File created from Restart Script}
\label{sub:statusFile}

As stated in Chapter \ref{sub:codeFeatures} the restart script is writing a status file with the current status and progress bars of the simulation. The output is formatted like a table and can bei adjust in its appearance with the \texttt{-\/-format} parser option [Chapter \ref{sub:codeFeatures}]. The output types which are written in the status file are listed in Tab. \ref{tab:statusTypes}. To each output prints the restart script the date, time and the duration of the monitoring additionally into the table. 
\begin{center}
    \captionsetup{type=table}
    \begin{tabular}{l | l}
        Type              & Information                                                                                     \\\hline
        \texttt{STARTING} & Start of monitoring process or Start/Restart of simulation                                      \\
        \texttt{CONTINUE} & If the simulation has performed at least one run                                                \\\hline
        \texttt{CONTROL}  & Check of the current time steps of the simulation                                               \\
        \texttt{SUCCESS}  & Stop of monitoring because the time steps condition has been fulfilled                          \\\hline
        \texttt{IMPORT}   & Modules \texttt{h5py}, \texttt{pandas} and \texttt{numpy} will be loaded                        \\
        \texttt{INSTALL}  & Script installs the modules \texttt{h5py}, \texttt{pandas} and \texttt{numpy} on the user level \\
        \texttt{CHECK}    & Required modules to use the reset option are already installed                                  \\\hline
        \texttt{WAITING}  & Simulation is pending in the queue of Slurm                                                     \\
        \texttt{RUNNING}  & Simulation is running on the server                                                             \\\hline
        \texttt{BACKUP}   & The script performs the backup of the data to the predefined location                           \\
        \texttt{RESTORE}  & The script reloads the data from the backup folder if errors occur                              \\
        \texttt{RESET}    & The simulation gets reset with the use of dump files                                            \\\hline 
        \texttt{ERROR}    & During the monitoring does occur an error                                                       \\
                          & Types:                                                                                          \\
                          & No executable (\texttt{gkw.x}), walltime, timeout, no config, \texttt{hdf5}, string error       \\
    \end{tabular}
    \captionof{table}{Output types in status file}
    \label{tab:statusTypes}
\end{center}

To continue, the status file has different style options, e.g., \texttt{fancy}, \texttt{universal} and \texttt{None} which are displayed in Fig. \ref{fig:statusStyle}. 
\begin{center}
\captionsetup{type=figure}
\begin{verbatim}
  ╭───────────────╮    +---------------+    ------------------
  │     fancy     │    |   universal   |           None
  ├───────────────┤    +---------------+    ------------------
  │               │    |               |    
  ╰───────────────╯    +---------------+    ------------------
\end{verbatim}
\captionof{figure}{Styles of output table in status file}
\label{fig:statusStyle}
\end{center}
In addition to that, it writes down the progress of simulation to the end of the status file in form of a progress bar which gets updated dynamically. The progress bar contains the current time step of the simulation in contrast to the required time steps and the progress of the run itself in seconds in relation to the wall time. It also contains the current output from \texttt{squeue}, if the simulation is running or pending, else the important information of the simulation. An example progress bar ist shown in Fig. \ref{fig:statusBar}.
\begin{center}
\captionsetup{type=figure}
\begin{verbatim}
 +------------------------------------------------------------+
 |  JOBID  NAME     USER ST    TIME NODES  NODELIST(REASON)   |
 | 970455 3x1.5 bt712347 CG 3:16:01     3  r03n03,r05n[15-16] |
 +------------------------------------------------------------+
 | PROGRESS  [===================>.....]  ( 80%)  20000/25000 |
 | RUN 13    [===>.....................]  ( 13%)  11462/86400 |
 +------------------------------------------------------------+
\end{verbatim}
\captionof{figure}{Progress bar example}
\label{fig:statusBar}
\end{center}
The progress bar has the advantage that someone can easily access all important information at the glance. To boost productivity the following command could be quite handy:
\begin{lstlisting}
cd $DATA 
find . -name $STATUSFILENAME -exec tail -8 {} \;
\end{lstlisting}
which prints all progress bars of every simulation in the data folder. If one want to see the complete status file change \texttt{tail -8} to \texttt{cat}.
An example status file is displayed in the Appendix \ref{append:status}.

%\displayfonttable[hex-digits=head+foot, range-end=1FFFF,compare-with=New Computer Modern Math,compare-color=black,  compare-bgcolor=black!5, missing-glyph-color=black!50, color=black!75]{Latin Modern Math}

\subsection{Support for other Resource Managers}
\label{sub:generalSupport}

The general idea behind the restart script could be adapted for other resource managers. For example in the early stage of the development the script was changed to support the \texttt{PBS/Torque} resource manager which mainly runs on the \texttt{btrzx2}-cluster of the University of Bayreuth\cite{btrzx2} and supports the research of Dominik M{\"u}ller for his Bachelor Thesis\cite{Mueller2023} with the \texttt{RHMD}-code\cite{rmhd}. The script was named \texttt{torque\_monitor.py} and can be found in the GitHub Repository of this thesis\cite{torquemonitor} or in the Appendix \ref{append:torquecode}. But note that due to the early adaptation the code does lack many features of \texttt{slurm\_monitor.py} and was not maintained actively.

### 2023-05-01 bachelorthesis/04-declaration.tex
#39: Declaration corrected

@@
 
-The author declares that the material presented in this thesis is his own work, unless explicitly stated otherwise. Chapter \ref{sec:convergence} and \ref{sec:threshold} is based on
+The author declares that the material presented in this thesis is his own work, unless explicitly stated otherwise. This thesis is based on
 \bigskip
@@
 \bigskip
-and is an further iteration of the brief communication published in the journal \enquote{Physics of Plasma}.
+and is a further iteration of this publication (brief communication published in the journal \enquote{Physics of Plasma}).
 It provides additional plots and paragraphs that were not necessary to be included in the final publication.

### 2023-05-01 bachelorthesis/20-theory.tex
#39: Include paper, data, monitor code and status file in appendix, codestyle, chapter about methods, minor changes, flowdiagram style, new references
#39: New title style; corrected Chapter names

@@
 
-\chapter{Theory}
+\chapter{Plasma Physics, Zonal Flows and Gyrokinetic Theory}
 \label{chap:theory}
+
 \thispagestyle{empty}
 \newpage
-
-In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
-Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
-Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
-The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
-Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
-\\
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, peeters2016}
-\begin{equation}
-	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
-	\label{eq:shearingrate}
-\end{equation}
-where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
-The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
-\begin{equation}
-\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
-\end{equation}
-The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
-\\
-Consistent with Ref.
-\citenum{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
-Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.
-\citenum{peeters2016,peeters2009}.
-\newpage
-In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
-It is defined by
-\begin{equation}
-	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord),
-	\label{eq:shearingrate_fourier}
-\end{equation}
-where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
-% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
-defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
-% define the zonal flow modes' amplitude in terms of the shearing rate
-Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$. 
-% introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\

### 2023-05-01 bachelorthesis/41-resolution.tex
#39: New title style; corrected Chapter names

@@
 
-\subsection{Reduction of parallel velocity grid points $\Nvpar$}
+\subsection{Reduction of parallel Velocity Grid Points $\Nvpar$}
 \label{sub:reduceNvpar}
@@
 
-\subsection{Reduction of magnetic moment grid points $\Nmu$}
+\subsection{Reduction of Magnetic Moment Grid Points $\Nmu$}
 \label{sub:reduceNmu}
@@
 
-\subsection{Reduction of magnetic field grid points $\Ns$}
+\subsection{Reduction of Magnetic Field Grid Points $\Ns$}
 \label{sub:reduceNs}

### 2023-05-01 bachelorthesis/42-boxsizescan.tex
#39: New title style; corrected Chapter names

@@
 \newpage
-\section{Size Convergence of $\exb$ staircase pattern}
+\section{Size Convergence of $\exb$ Staircase Pattern}
 \label{sec:convergence}
 
-%TODO
+In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
+The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
+Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
+\\
 
@@
 
-\subsection{Staircase structures in Comparison}
+\subsection{Staircase Structures in Comparison}
 \label{sub:comparisonwexb}

### 2023-05-01 bachelorthesis/21-motion.tex
#39: Theory, Motion, ITG, Zonal Flow, Confinement partly

\section{Charged Particle Motion in Magnetic and Electric Field}
\label{sec:motion}

In magnetic confinement devices like the tokamak reactor, the charged particles experience magnetic and electric fields which results in distinct motion under tha associated force. Charged particles can be separated in species, e.g. electrons and ions, which will be later on indicated bei the index sp on the quantities.

\subsection{Particle Motion perpendicular to the magnetic field - The Gyro-Motion}
\label{sub:gyromotion}

Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp,\mathrm{sp}}$ undergo a circular motion in the plane perpendicular to the magnetic field. This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as:
\begin{gather}
    \omega_\mathrm{c,sp} = \frac{|q_\mathrm{sp}|B}{m_\mathrm{sp}}~,
\end{gather}
where $m_\mathrm{sp}$ and $q_\mathrm{sp}$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by:
\begin{gather}
    \rho_\mathrm{sp} = \frac{m_\mathrm{sp}v_{\perp,\mathrm{sp}}}{|q_\mathrm{sp}|B}
\end{gather}
which center is often as \textit{gyro-center}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.

Due to the Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical species thermal velocity is
\begin{gather}
    v_\mathrm{th,sp} = \sqrt{\frac{2T_\mathrm{sp}}{m_\mathrm{sp}}}~,
\end{gather}
where $T_\mathrm{sp}$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th,sp}$ the \textit{thermal Larmor radius} gets introduced as:
\begin{gather}
    \rho_\mathrm{th,sp} = \frac{m_\mathrm{sp}v_{\mathrm{th,sp}}}{|q_\mathrm{sp}|B}~.
\end{gather}

\newpage

\subsection{Particle Motion parallel to the magnetic field - The Gyro Motion}
\label{sub:parallelmotion}

In absence of forces in the parallel direction of the magnetic field the particles can stream freely parallel to the homogenous magnetic field. The velocity of this motion is connected to the thermal velocity $v_\mathrm{th,sp}$ and is dominated bei electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).

When an electric field with a component parallel to the magnetic field $E_\parallel$ acts on plasma the charged particles gets accelerated by the electric force
\begin{gather}
    F_{\parallel,E} = q_\mathrm{sp}E_\parallel~.
\end{gather}
The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.

Since magnetic fields are not always homogenous an inhomogeneous magnetic field with its gradient $\nabla B$ contains an component parallel to the magnetic field which is given by:
\begin{gather}
    \nabla_\parallel B = \frac{\vect{B}}{B} \cdot \nabla B~.
\end{gather}
This causes the force
\begin{gather}
    F_{\parallel,\nabla_\parallel B} = - \frac{m_\mathrm{sp}v^2_{\perp,\mathrm{sp}}}{2B} \nabla_\parallel B = \mu \nabla_\parallel B~;~~~~\mu = \frac{m_\mathrm{sp}v^2_{\perp,\mathrm{sp}}}{2B}
\end{gather}
with \textit{magnetic moment} $\mu$ which can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c,sp}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{sp}$. The caused force has it application in the mirror effect where a charged particle gets reflected due to this force.

\newpage

\subsection{Drifts in the Gyro-Center}
\label{sub:drift}

In the presence of a magnetic field (homogenous, inhomogeneous or perturbed) and electric fields the gyro-center undergoes slow (compared to the thermal velocity $\vth$) drift motions perpendicular to the magnetic field. There are serval examples for this drift motion. According to this thesis topic only the main three drift types will be covered in the following.

\begin{enumerate}
    \item \textbf{$\exb$ drift:}\\
    If an electric field $\vect{E}$ with an perpendicular component is present together with a the magnetic field $\vect{B}$ (both fields are homogenous) the acting Coulomb force and Lorentz force results to a drift of the gyro-center with
    \begin{gather}
        \vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}
        \label{eq:exb}
    \end{gather}
    which is called the $\exb$-drift. Since both acting forces direction depends on the species type the direction of the $\exb$-drift is for every species the same.\\
    \item \textbf{$\nabla B$ Drift:}\\
    Inhomogeneous magnetic field causes an gradient $\nabla B$ of the magnetic field. Because of that gradient the gyro-center undergoes an $\nabla B$-drift defined by
    \begin{gather}
        \vect{v}_{\nabla B} = \frac{m_\mathrm{sp} v^2_{\perp,\mathrm{sp}}}{2 q_\mathrm{sp}}\frac{\vect{B}\times \nabla B}{B^3}~.
        \label{eq:gradB}
    \end{gather}
    The $\nabla B$-drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$-drift depends on the species type. \\
    \item \textbf{Curvature drift:}\\
    Due to centrifugal force acting on the particle in a curved magnetic filed the gyro-center experience an curvature drift given by
    \begin{gather}
        \vect{v}_{\kappa} = \frac{m_\mathrm{sp} v^2_{\parallel,\mathrm{sp}}}{q_\mathrm{sp}}\frac{\vect{B}\times \kappa}{B^2} = \frac{m_\mathrm{sp} v^2_{\parallel,\mathrm{sp}}}{q_\mathrm{sp}} \frac{\vect{B}\times \nabla B}{B^3}~;~~~~\kappa = \frac{\nabla B}{B}~.
        \label{eq:curvature}
    \end{gather}
    To obtain the result for $\kappa$ in Eq. \ref{eq:curvature} the plasma pressure has to be small in comparison die the magnetic field strength $B$. In the form of Eq. \ref{eq:curvature} $\nabla B$- and curvature drift can be treated similarly.
\end{enumerate}

### 2023-05-01 bachelorthesis/22-confinement.tex
#39: Theory, Motion, ITG, Zonal Flow, Confinement partly

\newpage
\section{Magnetic Confinement in Tokamak}
\label{sec:confinement}

In a tokamak devices strong magnetic fields confine the hot plasma. As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion containing gyro motion and slow perpendicular gyro center drifts. Because of the much smaller size of the Larmor radius compared to the device size $R_0$ the particle and energy losses are caused by the gyro center drift. To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus. This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux sufaces} and with magnetic field lines which appearing on that surfaces.

### 2023-05-01 bachelorthesis/23-iontempgrad.tex
#39: Theory, Motion, ITG, Zonal Flow, Confinement partly

\newpage

\section{Ion Temperature Gradient (ITG) Driven Instability}
\label{sec:ITG}

The described electrostatic forces acting on the tokamak plasma causes a variety of collective phenomena. Microinstabilities is one of this phenomena which are instabilities on the spatial scale of the Larmor radius and frequencies much smaller than the cyclotron frequency. These instabilities are driven bei density and temperature gradients that are present in the tokamak devices. Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma Ref (10,11,12) which limits the confinement time of tokamak reactors. The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability Ref (13,14,15).

\includegraphicsHere{Theory/ITG-Turbulence.png}{
	Sketch of ion temperature gradient driven instability on the outboard low field sid of the tokamak
}{fig:ITG}{0.6}

The ITG driven is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard side of the tokamak. The mechanism will be briefly explained with the help of Fig. \ref{fig:ITG} Ref (18,17,16 and 4 in BA). An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature and $\nabla B$ dirft $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, grey cells). This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response. As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential causes an electric field which results in a $\exb$ drift $\vect{v}_E$. The $\exb$-drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.

The described mechanism causes perturbation on the outboard side of the tokamak this region has a \enquote{bad curvature} and instability happens there. On the inboard side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}. Furthermore such turbulent structures are called \textit{eddies} and the have a typical radial scale of serval Larmor radius. (Ref 19 BA). \bigskip

Through the explanation above is clear that the initial perturbation is growing when the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$. For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$. (Ref 3, 7 BA)

### 2023-05-01 bachelorthesis/24-zonalflows.tex
#39: Theory, Motion, ITG, Zonal Flow, Confinement partly

\section{Zonal Flows and Shearing Rate $\wexb$}
\label{sec:zonalflow}

Zonal flows are referred in the tokamak plasma to the $\exb$-flows [Eq. \ref{eq:exb}] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}. (Ref 28) Because of the variation only across flux-surfaces only the zonal flow is tangential to the flux-surfaces and do not contribute to the turbulent transport across flux-surfaces caused by the $\exb$-motion. 
Zonal flows are driven nonlinear through turbulent Reynolds stresses (Ref 32). For more about the mechanism behind zonal flow generation the reader is referred to Ref. (42,43,44,10). \bigskip

Furthermore zonal flows are connected with the radially alternating zonal potential what introduces an velocity shear,i.e., a radial variation of the zonal flow velocity. This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence (Ref 29, 67) especially the ion temperature gradient driven turbulence (Ref 22, 63, 64, 65, 66). The nonlinear generation mechanism causes an shift, also known as Dimits shift (Ref 67), of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to an later development of ITG driven turbulence. The shearing process tilted und stretches turbulent eddies, caused be the shear flow, until they break into smaller eddies with reduced radial correlation length which characterize the shearing as decorrelation process. Turbulence will then be reduced because of this process. (Ref 29,68,28)  \bigskip

When the zonal flows are strong enough to suppress turbulence completely due to the shearing process, pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. The $\exb$ staircase pattern has following properties:
\begin{enumerate}
	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10 \rhoth which is larger than the Larmor radius but smaller than the machine size $R_0$.
	\item[(2)] The structure are \textit{quasi stationary} in space and varies on time scales much longer than the typical turbulent time scales.
	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$-shearing rate of the order of $10^{-1} \vth/R_0$.
	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. (Ref 23)
\end{enumerate} \bigskip

The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, peeters2016}
\begin{equation}
	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
	\label{eq:shearingrate}
\end{equation}
where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
\begin{equation}
\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
\end{equation}
The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}. \\ \bigskip

As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation. Avalanches are initial when the$\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing (Ref 26,27). The fully developed avalanche resluts in almost completely quenched turbulence. Its development results in a discontinuous step int he dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold} (Ref 25,77)

### 2023-05-09 briefcommunicaton/briefcommunication.tex
#37: Update bibentrys

@@
 
-Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{pradalier2010}.
-Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{peeters2016, weikl2017, rath2021} as well as global gradient-driven \cite{mcmillan2009, villard2013, seo2022} and global flux-driven \cite{pradalier2010, pradalier2015, wang2020, kim2022, kishimoto2023} studies.
+Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
+Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Weikl2017, Rath2021} as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
 In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
@@
 The latter case would imply that it is not necessarily global physics, i.e., profile effects, that set (i) the radial size of the $\exb$ staircase pattern and (ii) the scale of avalanche-like transport events.
-These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}.
-In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\onlinecite{peeters2016}.\bigskip
+These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{Pradalier2010}.
+In this brief communication the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.~\onlinecite{Peeters2016}.\bigskip
 
@@
 
-The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} with adiabatic electron approximation.
-In agreement with Ref.~\onlinecite{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
+The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{Peeters2009} with adiabatic electron approximation.
+In agreement with Ref.~\onlinecite{Peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference, with a somewhat lowered number of parallel velocity grid points.
 It has been carefully verified that this modification preserves the same physical outcome as the original study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Ref.~\onlinecite{peeters2009, peeters2016}.
+A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Ref.~\onlinecite{Peeters2009, Peeters2016}.
 \begin{table}[ht]
@@
 	\caption{
-		Resolution used in this paper for further information the author links to Ref.~\onlinecite{peeters2016}.
+		Resolution used in this paper for further information the author links to Ref.~\onlinecite{Peeters2016}.
 %: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
@@
 Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
-Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{hamada1958}.
+Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}.
 The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
@@
 \\
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, rath2016, peeters2016}
+The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Rath2016, Peeters2016}
 \begin{equation}
@@
 where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
-The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{rath2021}
+The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
 \begin{equation}
@@
 \end{equation}
-The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
+The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}.
 \\
-Consistent with Ref.~\onlinecite{peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Consistent with Ref.~\onlinecite{Peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
 Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.
-\onlinecite{peeters2016,peeters2009}.
+\onlinecite{Peeters2016,Peeters2009}.
 \newpage
@@
 % short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
 % introduce the fully developed staircase structure
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
 % insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
@@
 % yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\onlinecite{peeters2016}.
+Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\onlinecite{Peeters2016}.
 \\
@@
 In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
-Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
+Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{Peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
 This scan covers realizations with $\rlt \in [6.0,~6.2,~6.4]$.
@@
 In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
-The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$ in accordance to Ref. \onlinecite{peeters2016}.
+The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$ in accordance to Ref. \onlinecite{Peeters2016}.
 % Which stationay radial size occurs????
@@
 % compare to other studies
-This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
+This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{Pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
 % relevance of the outcome
 The occurrence of this mesoscale implies that non-locality, in terms of Ref.
-\onlinecite{pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{mcmillan2009, pradalier2010, rath2016, peeters2016}.
+\onlinecite{Pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{McMillan2009, Pradalier2010, Rath2016, Peeters2016}.
 

### 2023-05-09 bachelorthesis/10-motivation.tex
#39: Introduction -> Motivation
#39: Theory finished and closure corrected, First Draft finished

--- 2023-04-08 bachelorthesis/10-introduction.tex
@@
 
-\chapter{Introduction}
-\label{chap:intro}
+\chapter{Motivation}
+\label{chap:motivation}
+
 \thispagestyle{empty}
@@
 \end{itemize}
-These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{pradalier2010}. \bigskip
+These transport events are usually restricted to $\exb$ staircase structures and considered as a nonlocal transport mechanism \cite{Pradalier2010}. \bigskip
 
 In this bachlor thesis the above question is addressed through a box size convergence scan of the same cases close to the nonlinear threshold for turbulence generation as studied in Ref.
-\citenum{peeters2016}.
+\citenum{Peeters2016}.
 %TODO More Text

### 2023-05-09 bachelorthesis/21-motion.tex
#39: Theory finished and closure corrected, First Draft finished

@@
 Charged particles can be separated in species, e.g.
-electrons and ions, which will be later on indicated bei the index sp on the quantities.
+electrons and ions, which will be later on not displayed in the governing equation.
+Throughout this thesis the charge $q$ or the mass $m$ indicate the quantities of a specific spiecies.
 
-\subsection{Particle Motion perpendicular to the magnetic field - The Gyro-Motion}
+\subsection{Particle Motion perpendicular to the magnetic field}
 \label{sub:gyromotion}
 
-Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp,\mathrm{sp}}$ undergo a circular motion in the plane perpendicular to the magnetic field.
+Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field.
 This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as:
 \begin{gather}
-    \omega_\mathrm{c,sp} = \frac{|q_\mathrm{sp}|B}{m_\mathrm{sp}}~,
+    \omega_\mathrm{c} = \frac{|q|B}{m}~,
 \end{gather}
-where $m_\mathrm{sp}$ and $q_\mathrm{sp}$ are the mass and the charge of the particle and $B$ the strength of the magnetic field.
-The radius, the so called \textit{Larmor radius}, of this motion is given by:
+where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by:
 \begin{gather}
-    \rho_\mathrm{sp} = \frac{m_\mathrm{sp}v_{\perp,\mathrm{sp}}}{|q_\mathrm{sp}|B}
+    \rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}
 \end{gather}
-which center is often as \textit{gyro-center}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.
+which center is often as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.
 
@@
 \begin{gather}
-    v_\mathrm{th,sp} = \sqrt{\frac{2T_\mathrm{sp}}{m_\mathrm{sp}}}~,
+    \vth = \sqrt{\frac{2T}{m}}~,
 \end{gather}
-where $T_\mathrm{sp}$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th,sp}$ the \textit{thermal Larmor radius} gets introduced as:
+where $T$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th}$ the \textit{thermal Larmor radius} gets introduced as \cite{Wesson2011}
 \begin{gather}
-    \rho_\mathrm{th,sp} = \frac{m_\mathrm{sp}v_{\mathrm{th,sp}}}{|q_\mathrm{sp}|B}~.
+    \rhoth = \frac{m\vth}{|q|B}~.
 \end{gather}
@@
 
-\subsection{Particle Motion parallel to the magnetic field - The Gyro Motion}
+\subsection{Particle Motion parallel to the magnetic field}
 \label{sub:parallelmotion}
@@
 In absence of forces in the parallel direction of the magnetic field the particles can stream freely parallel to the homogenous magnetic field.
-The velocity of this motion is connected to the thermal velocity $v_\mathrm{th,sp}$ and is dominated bei electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).
+The velocity of this motion is connected to the thermal velocity $\vth$ and is dominated bei electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).
 
@@
 \begin{gather}
-    F_{\parallel,E} = q_\mathrm{sp}E_\parallel~.
+    F_{\parallel,E} = qE_\parallel~.
 \end{gather}
@@
 
-Since magnetic fields are not always homogenous an inhomogeneous magnetic field with its gradient $\nabla B$ contains an component parallel to the magnetic field which is given by:
+Since magnetic fields are not always homogenous an inhomogeneous magnetic field with its gradient $\nabla B$ contains a component parallel to the magnetic field which is given by:
 \begin{gather}
@@
 \begin{gather}
-    F_{\parallel,\nabla_\parallel B} = - \frac{m_\mathrm{sp}v^2_{\perp,\mathrm{sp}}}{2B} \nabla_\parallel B = \mu \nabla_\parallel B~;~~~~\mu = \frac{m_\mathrm{sp}v^2_{\perp,\mathrm{sp}}}{2B}
+    F_{\parallel,\nabla_\parallel B} = - \frac{mv^2_{\perp}}{2B} \nabla_\parallel B = \mu \nabla_\parallel B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
 \end{gather}
-with \textit{magnetic moment} $\mu$ which can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c,sp}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{sp}$.
+with \textit{magnetic moment} $\mu$ which can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
 The caused force has it application in the mirror effect where a charged particle gets reflected due to this force.
+\cite{Wesson2011}
 
@@
 
-\subsection{Drifts in the Gyro-Center}
+\subsection{Drifts in the Gyrocenter}
 \label{sub:drift}
 
-In the presence of a magnetic field (homogenous, inhomogeneous or perturbed) and electric fields the gyro-center undergoes slow (compared to the thermal velocity $\vth$) drift motions perpendicular to the magnetic field.
+In the presence of a magnetic field (homogenous, inhomogeneous or perturbed) and electric fields the gyrocenter undergoes slow (compared to the thermal velocity $\vth$) drift motions perpendicular to the magnetic field.
 There are serval examples for this drift motion.
@@
 \begin{enumerate}
-    \item \textbf{$\exb$ drift:}\\
-    If an electric field $\vect{E}$ with an perpendicular component is present together with a the magnetic field $\vect{B}$ (both fields are homogenous) the acting Coulomb force and Lorentz force results to a drift of the gyro-center with
+    \item \textbf{$\exb$ Drift:}\\
+    If an electric field $\vect{E}$ with a perpendicular component is present together with the magnetic field $\vect{B}$ (both fields are homogenous) the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
     \begin{gather}
@@
     \end{gather}
-    which is called the $\exb$-drift. Since both acting forces direction depends on the species type the direction of the $\exb$-drift is for every species the same.\\
+    which is called the $\exb$-drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
     \item \textbf{$\nabla B$ Drift:}\\
-    Inhomogeneous magnetic field causes an gradient $\nabla B$ of the magnetic field. Because of that gradient the gyro-center undergoes an $\nabla B$-drift defined by
+    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$-drift defined by
     \begin{gather}
-        \vect{v}_{\nabla B} = \frac{m_\mathrm{sp} v^2_{\perp,\mathrm{sp}}}{2 q_\mathrm{sp}}\frac{\vect{B}\times \nabla B}{B^3}~.
+        \vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}~.
         \label{eq:gradB}
     \end{gather}
-    The $\nabla B$-drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$-drift depends on the species type. \\
-    \item \textbf{Curvature drift:}\\
-    Due to centrifugal force acting on the particle in a curved magnetic filed the gyro-center experience an curvature drift given by
+    The $\nabla B$-drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$-drift depends on the species type.
+    \item \textbf{Curvature Drift:}\\
+    Due to centrifugal force acting on the particle in a curved magnetic filed the gyrocenter experience a curvature drift given by
     \begin{gather}
-        \vect{v}_{\kappa} = \frac{m_\mathrm{sp} v^2_{\parallel,\mathrm{sp}}}{q_\mathrm{sp}}\frac{\vect{B}\times \kappa}{B^2} = \frac{m_\mathrm{sp} v^2_{\parallel,\mathrm{sp}}}{q_\mathrm{sp}} \frac{\vect{B}\times \nabla B}{B^3}~;~~~~\kappa = \frac{\nabla B}{B}~.
+        \vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{q}\frac{\vect{B}\times \vect{\kappa}}{B^2} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}~;\qquad\vect{\kappa} = -(\vect{b}\cdot \nabla)\vect{b} = \frac{\nabla B}{B}~,
         \label{eq:curvature}
     \end{gather}
-    To obtain the result for $\kappa$ in Eq.
-\ref{eq:curvature} the plasma pressure has to be small in comparison die the magnetic field strength $B$.
+    where $\vect{b}$ is the unit vector along the magnetic field.
+To obtain the result for the curvature $\vect{\kappa}$ in Eq.
+(\ref{eq:curvature}) the plasma pressure has to be small in comparison die the magnetic field strength $B$.
 In the form of Eq.
-\ref{eq:curvature} $\nabla B$- and curvature drift can be treated similarly.
+(\ref{eq:curvature}) $\nabla B$ and curvature drift can be treated similarly.
+\cite{Wesson2011}
 \end{enumerate}

### 2023-05-09 bachelorthesis/22-confinement.tex
#39: Theory finished and closure corrected, First Draft finished

@@
 This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux sufaces} and with magnetic field lines which appearing on that surfaces.
+For stability reasons the magnetic field has a toroidal and a poloidal component which also holds equivalence of the plasma pressure.
+\cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current.
+Together the components results in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}].
+To characterize the the quality of confinement the so-called \textit{plasma beta} is used and is given by
+\begin{gather}
+    \beta = \frac{nT}{\mu_0 B^2/2}~.
+\end{gather} 
+with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$.
+Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates when the plasma beta is smaller the confinement is better.
+In a tokamak reactor the plasma beta has a typical order of few percent.
+\cite{Wesson2011}
+\includegraphicsHere{Theory/Tokamak-Torus.pdf}{
+	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or zylinder coordinates ($Z$, $R$, $\phi$).
+\cite{Barton2015}
+}{fig:confinement}{0.7}

### 2023-05-09 bachelorthesis/23-iontempgrad.tex
#39: Theory finished and closure corrected, First Draft finished

@@
 \newpage
-
 \section{Ion Temperature Gradient (ITG) Driven Instability}
@@
 These instabilities are driven bei density and temperature gradients that are present in the tokamak devices.
-Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma Ref (10,11,12) which limits the confinement time of tokamak reactors.
-The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability Ref (13,14,15).
+Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors.
+The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability \cite{Coppi1967, Cowley1991, Rudakov1961}
 
-\includegraphicsHere{Theory/ITG-Turbulence.png}{
-	Sketch of ion temperature gradient driven instability on the outboard low field sid of the tokamak
-}{fig:ITG}{0.6}
+\includegraphicsHere{Theory/ITG-Turbulence.pdf}{
+	Sketch of ion temperature gradient driven instability on the outboard low field sid of the tokamak \cite{Casson_PHD}
+}{fig:ITG}{0.5}
 
@@
 The mechanism will be briefly explained with the help of Fig.
-\ref{fig:ITG} Ref (18,17,16 and 4 in BA).
-An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature and $\nabla B$ dirft $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, grey cells).
+\ref{fig:ITG} \cite{Beer_PHD, Casson_PHD, Dannert_PHD}.
+An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature and $\nabla B$ dirft $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells).
 This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response.
 As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential causes an electric field which results in a $\exb$ drift $\vect{v}_E$.
-The $\exb$-drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.
+The $\exb$ drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip
 
@@
 Furthermore such turbulent structures are called \textit{eddies} and the have a typical radial scale of serval Larmor radius.
-(Ref 19 BA).
-\bigskip
-
+\cite{Newins2006}.
 Through the explanation above is clear that the initial perturbation is growing when the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$.
 For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$.
-(Ref 3, 7 BA)
+\cite{Dimits2000, Isliker2010}

### 2023-05-09 bachelorthesis/24-zonalflows.tex
#39: Theory finished and closure corrected, First Draft finished

@@
 
-Zonal flows are referred in the tokamak plasma to the $\exb$-flows [Eq.
-\ref{eq:exb}] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}.
-(Ref 28) Because of the variation only across flux-surfaces only the zonal flow is tangential to the flux-surfaces and do not contribute to the turbulent transport across flux-surfaces caused by the $\exb$-motion.
-Zonal flows are driven nonlinear through turbulent Reynolds stresses (Ref 32). For more about the mechanism behind zonal flow generation the reader is referred to Ref. (42,43,44,10). \bigskip
+Zonal flows are referred in the tokamak plasma to the $\exb$ flows [Eq.( \ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}.
+\cite{Diamond2005} Because of the variation only across flux-surfaces only the zonal flow is tangential to the flux-surfaces and do not contribute to the turbulent transport across flux-surfaces caused by the $\exb$ motion.
+Zonal flows are driven nonlinear through turbulent Reynolds stresses \cite{Diamond1991}.
+For more about the mechanism behind zonal flow generation the reader is referred to Ref.
+\citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}.
+\bigskip
 
-Furthermore zonal flows are connected with the radially alternating zonal potential what introduces an velocity shear,i.e., a radial variation of the zonal flow velocity.
-This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence (Ref 29, 67) especially the ion temperature gradient driven turbulence (Ref 22, 63, 64, 65, 66).
-The nonlinear generation mechanism causes an shift, also known as Dimits shift (Ref 67), of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to an later development of ITG driven turbulence.
-The shearing process tilted und stretches turbulent eddies, caused be the shear flow, until they break into smaller eddies with reduced radial correlation length which characterize the shearing as decorrelation process.
+\includegraphicsHere{Theory/Zonal-Flow-Generation.pdf}{
+	Subdued turbulence because if stretched eddies which breaks in smaller structures caused by sheared $\exb$ drifts \cite{Hammett2009}
+}{fig:zonalflow}{0.9}
+
+Furthermore, zonal flows are connected with the radially alternating zonal potential what introduces a velocity shear, i.e., a radial variation of the zonal flow velocity.
+This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000} especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}.
+The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a later development of ITG driven turbulence.
+The shearing process tilted und stretches turbulent eddies, caused be the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig.
+\ref{fig:zonalflow}] which characterize the shearing as decorrelation process.
 Turbulence will then be reduced because of this process.
-(Ref 29,68,28)  \bigskip
+\cite{Biglari1990,Diamond2005,Burnell1997} \bigskip
+
+\newpage
 
 When the zonal flows are strong enough to suppress turbulence completely due to the shearing process, pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure.
-The $\exb$ staircase pattern has following properties:
+The $\exb$ staircase pattern has the following properties:
 \begin{enumerate}
@@
 	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$-shearing rate of the order of $10^{-1} \vth/R_0$.
-	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. (Ref 23)
+	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. \cite{McMillan2009}
 \end{enumerate} \bigskip
 
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{rath2016, doi:10.1063/1.3005380, peeters2016}
+The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Peeters2016}
 \begin{equation}
@@
 \end{equation}
-The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{doi:10.1063/1.871313, doi:10.1063/1.872847} and quantifies the zonal flow induced shearing of turbulent structures \cite{doi:10.1063/1.859529, doi:10.1063/1.871313, doi:10.1063/1.872367}.
+The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}.
 \\ \bigskip
@@
 As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation.
-Avalanches are initial when the$\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing (Ref 26,27).
+Avalanches are initial when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing \cite{Idomura2009,McMillan2009}.
 The fully developed avalanche resluts in almost completely quenched turbulence.
-Its development results in a discontinuous step int he dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold} (Ref 25,77)
+Its development results in a discontinuous step int he dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}.
+\cite{Peeters2016,Weikl2017}

### 2023-05-09 bachelorthesis/25-gyrokinetics.tex
#39: Theory finished and closure corrected, First Draft finished

\newpage
\section{Gyrokinetic Theory}
\label{sec:gyrokinetic}

\subsection{Vlasov Equation}
\label{sub:vlasov}

Because of the large number of particle in the fusion plasma a prediction only using Newton-Maxwell dynamics results in an impossible task for computers, but this problem can be solved with a statistical approach. For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles postion $\vect{x}$ and velocity $\vect{v}$ is needed. Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often considered \cite{Garbet2010} which gives thourgh evolution of the particle density distribution function the \textit{Vlasov equation}
\begin{gather}
	\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0~.
	\label{eq:vlasov}
\end{gather}
To obtain a closed system the Maxwell equations with the particle density $n$ and current density $j$ can be descripted the distribution function as follows:

\begin{gather}
	n = \int \mathrm{d}\vect{v}\,f(\vect{x}, \vect{v}, t) \qquad j = q \int \mathrm{d}\vect{v}\, v f(\vect{x}, \vect{v}, t)~,
	\label{eq:densitycurrent}
\end{gather}
which are substituted into the Maxwell equations
\begin{gather}
	\begin{aligned}
		\nabla \cdot \vect{B} &= 0 &\qquad \nabla \times \vect{B} &= \mu_0\left( \sum_\mathrm{species} j + \epsilon_0 \frac{\partial \vect{E}}{\partial t} \right) \\
		\nabla \cdot \vect{E} &= \frac{1}{\epsilon_0} \sum_\mathrm{species} qn &\qquad \nabla \times \vect{E} &= - \frac{\partial \vect{B}}{\partial t}~.
	\end{aligned}
	\label{eq:maxwell}
\end{gather}
The Vlasov equation (\ref{eq:vlasov}) together with the Maxwell equations (\ref{eq:densitycurrent}) and (\ref{eq:maxwell}) is the basis of the gyrokinetic model. \cite{Krommes2012}

\newpage

\subsection{Gyrokinetic Ordering}
\label{sub:gyroordering}

Dynamics in tokamak plasma is descripted by spatio-temporal scales which is also referred as \textit{gyrokinetic ordering}. This seperation is a comparison between the tuburbulence frequency $\omega$ and the cyclotron frequency $\omega_\mathrm{c}$ ($\omega/\omega_\mathrm{c} \sim 10^{-3})$ due to strong external magnetic fields which causes a fast gyro motion. The length scales of the turbulence are associated with the wavevector $\vect{k}$ which can be seperated into an perpendicular component $k_\perp = |\vect{k} \times \vect{\hat{b}}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{\hat{b}}|$ where $\vect{\hat{b}}$ is parallel to the poloidal component of the magnetic field. The perpendicular component $k_\perp$ scales with the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significant smaller as the equilibrium density $n_0$ what can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$. The gradient length compares with the machnine size which leads to the normalized Larmor radius $\rho_* = \rhoth/R_0 \sim 10^{-3} - 10^{-4}$ with $R_0$ as major radius of the tokamak. The parallel component $k_\parallel$ on the other hand scales with the machine size. Experiments also show that in the core plasma the fluctuation amplitude of the density perturbation $\delta n/n_0$ and magnetic field fluctuation $\delta B/B_0$, with $B_0$ as strength of the magnetic field in equilibrium, is of order $\sim 10^{-4}$. Together this seperations result in the following result
\begin{gather}
	\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}
	\label{eq:gyroordering}
\end{gather}
with $\epsilon_\mathrm{g} \ll 1$ which applies to the typical dynamics of a fusion core plasma. \cite{Brizard2007,Garbet2010} The gyrokinetic ordering allows the reduction of parameters in the Vlasov equation (\ref{eq:vlasov}).

\newpage

\subsection{Gyrokinetic Equation}
\label{sub:gyrokinetic}

For the description for charged particle in the tokamak device the \textit{gyrocenter coordinates} will be used [Fig. \ref{fig:gyrocenterCoords}]. In this set of coordinates the gyrocenter follows the magnetic field with the parallel velocity $v_{\parallel}$. The gyro motion will be descripted together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.

\includegraphicsHere{Theory/Gyro-Center-Coordinates.pdf}{
	Sketch of gyrocenter coordinates where the charged particle performs a circular motion around the gyrocenter. \cite{Krommes2000}
}{fig:gyrocenterCoords}{0.7}

In the next step the coordinates of the Vlasov equation (\ref{eq:vlasov}) will be transformed into the gyrocenter coordinates. For that the particle postion $\vect{x}$ will be described through the gyrocenter postion $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$ with the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$. The particle velocity $\vect{v}$ gets decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{\hat{b}}$ and 
a perpendicular component $v_\perp = |\vect{v} \times \vect{\hat{b}}|$ which relates to the magnetic moment $\mu$. Through the gyrokinetic ordering a phase space transformation can be performed which is based on the Lie perturbation theory, the Hamiltonian or Lagrangian formalism and an asymtotic expansion in the smallness parameter $\epsilon_\mathrm{g}$. In the lowest order is the magnetic moment an exact invariant which makes the gyro phase $\zeta$ an ignorable variable. \cite{Garbet2010,Cary1981,Cary1983} Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transferres the Vlasov equation into the \textit{gyrokinetic equation}
\begin{gather}
	\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0~.
	\label{eq:gyrokinetic}
\end{gather}
The time derivative of the magnetic moment $\mu$ is zero because the magnetic moment is an exact invariant.

\subsection{$\delta f$ Approximation and Local Limit}
\label{sub:application}

All simulation in this thesis are performed with GKW (Gyro Kinetic Workshop) uses in the application of the gyrokinetic equation the $\delta f$ approximation and local limit approximation. For the $\delta f$ approximation the particle density distribution function f gets seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}). This is given by
\begin{gather}
	f = f_0 + \delta f \qquad \delta f \sim \rho_* f_0~.
\end{gather}
The distribution function in the equilibrium $f_0$ is set in GKW to the Maxwell distribution function. The perturbed distribution function $\delta f$ is considered to vary perpendicular to the poloidal magnetic field in order of the Larmor radius while the perturbation changes with the system size which results in the ordering
\begin{gather}
	\nabla_\perp(\delta f) = \nabla_\perp (f_0)~.
\end{gather}
The $\delta f$ approximation applies for global and local description where the latter case is used in this thesis.
The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities can be taken constant over the radial extent of the simulation volume. Because of the tokamak symmetry all equilibrium constants are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general.

In addition to that in GKW the following coordinates get applied with the parameter $\vect{X} = \{\xcoord, \ycoord, s\}$. Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate. Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}. This results in the new coordinates $\{\xcoord, \ycoord, s, v_\parallel, \mu\}$. Combining the $\delta f$ approximation and the Hamada coordinates and the gyrokinetic equation can be formualted as follows
\begin{gather}
	\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S~,
\end{gather}
where $g$ is a function which contains the the perturbation $\delta f$ and the equilibrium $f_0$. Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{\chi}$ are drift velocities which are caused by the inhomogeneous magnetic field and inhomogeneous magnetic field (${v}_\mathrm{D}$) and through the $\exb$ drift. The source term $S$ combines from the distribution function in the equilibrium $f_0$ and a correction term for the collisions as wall as the energy injection term. Due to prediction boundary conditions in the radial direction the radially gradients cannot change in time which is also referred as \textit{gradient driven} approach. \cite{Peeters2009}

### 2023-05-09 bachelorthesis/31-simulation.tex
#39: Theory finished and closure corrected, First Draft finished

--- 2023-04-18 bachelorthesis/31-simulation.tex
@@
 
-The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{peeters2009} with adiabatic electron approximation.
+The gyrokinetic simulations are performed with the non-linear flux tube version of Gyrokinetic Workshop (GKW) \cite{Peeters2009} with adiabatic electron approximation. 
 In agreement with Ref.
-\citenum{peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
+\citenum{Peeters2016}, Cyclone Base Case (CBC) like parameters are chosen with an inverse background temperature gradient length $\rlt = 6.0$ and circular concentric flux surfaces.
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference.
 A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.
-\citenum{peeters2009, peeters2016}.
+\citenum{Peeters2009, Peeters2016}.
 \begin{center}
@@
 \end{center}
+
+Consistent with Ref.
+\citenum{Peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass).
+Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref.
+\citenum{Peeters2016,Peeters2009}.

### 2023-05-09 bachelorthesis/41-resolution.tex
#39: Theory finished and closure corrected, First Draft finished

@@
 \ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref.
-\citenum{peeters2016} in Section IV.
+\citenum{Peeters2016} in Section IV.
 Note that, because of selected circular cencentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref.
-\citenum{peeters2016} for $s$-$\alpha$ geometry.
+\citenum{Peeters2016} for $s$-$\alpha$ geometry.
 In Fig.
 \ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is simular to the results from Ref.
-\citenum{peeters2016} with subdued turbulence after $\sim 3000 R/\vth$.
+\citenum{Peeters2016} with subdued turbulence after $\sim 3000 R/\vth$.
 
@@
 As in Ref.
-\citenum{peeters2016} in Section V conclude is the finite heat flux threshold approximately located at a grendent length of $\rlt = 6.3$ for circular geometry [FIG.
+\citenum{Peeters2016} in Section V conclude is the finite heat flux threshold approximately located at a grendent length of $\rlt = 6.3$ for circular geometry [FIG.
 4 of Ref.
-\citenum{peeters2016}].
+\citenum{Peeters2016}].
 Therefore following parameters were used:
@@
 \end{gather*}
-As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occur in the whole time domain, which is in agreement with Ref. \citenum{peeters2016}.
+As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occur in the whole time domain, which is in agreement with Ref. \citenum{Peeters2016}.
 

### 2023-05-09 bachelorthesis/42-boxsizescan.tex
#39: Theory finished and closure corrected, First Draft finished

@@
 \end{gather*}
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{doi:10.1063/1.870934, doi:10.1063/1.872847} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
-% insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+% insofar the results of Peeters2016 are well reproduced; now turn to the description of the main result of this test
 Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
@@
 Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.
-\citenum{peeters2016}.
+\citenum{Peeters2016}.
 \bigskip

### 2023-05-09 bachelorthesis/43-heatfluxthreshold.tex
#39: Dedication, Acknowledgement, Abstract (ENG, DE), Eidstattliche Versicherung, Declaration
#39: New title style; corrected Chapter names
#39: Theory finished and closure corrected, First Draft finished

--- 2023-04-08 bachelorthesis/42-heatfluxthreshold.tex
@@
 \newpage
-\section{The finite heat flux threshold}
-\label{sub:threshold}
+\section{The finite Heat Flux Threshold}
+\label{sec:threshold}
 
 In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
-Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
+Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{Peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
 This scan covers realizations with:
@@
 \end{gather*}
-in accordance to Ref. \citenum{peeters2016}.
+in accordance to Ref. \citenum{Peeters2016}.
 % Which stationay radial size occurs????

### 2023-05-09 bachelorthesis/50-closure.tex
#39: Split results part in separate, Plots for thesis, result part finished
#39: New title style; corrected Chapter names
#39: Theory finished and closure corrected, First Draft finished

--- 2023-04-04 bachelorthesis/05-closure.tex
@@
 
+\thispagestyle{empty}
+\newpage
+
 % what has been done
-Through careful tests this brief communication confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
+Through careful tests this bachelor thesis confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
 % mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
@@
 % compare to other studies
-This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
+This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{Pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
 % relevance of the outcome
 The occurrence of this mesoscale implies that non-locality, in terms of Ref.
-\cite{pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{mcmillan2009, pradalier2010, rath2016, peeters2016}.
+\citenum{Pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{McMillan2009, Pradalier2010, Rath2016, Peeters2016}.

### 2023-05-09 bachelorthesis/50-closure.tex
#39: Closre with work of thesis

@@
-% 6. Closure
+% 5. Closure
 
@@
 \newpage
+
+In this thesis the minimal resolution for simulations with GKW in the Cyclone Base parameter were observed in which the number of grid points for the parallel velocity $\Nvpar$ could be reduced from 64 to 48.
+\bigskip
+
+Additional the active development of an restart script in \texttt{python} leads to further convenience during the performance of a simulation on the \texttt{btrzx1} cluster. \bigskip
 

### 2023-05-18 briefcommunicaton/responseletter.txt
#37: Update Graphic, Refs + Responseletter

Editorial Office Comments:

* Revised manuscript source files must be in Word or LaTeX/ReVTeX. Replace the current PDF file with an acceptable manuscript source file.

    > [ ] Remove graphics hack (?)

* Separate figure files (separate from text and tables) are required for production purposes. Submit a separate figure file for each figure cited in the manuscript, not just one file containing all figures. The allowed file formats for figure files are TIF, PS, EPS, JPEG, or high-quality PDF. Figures may remain embedded in the manuscript. Number your separate figure files as they appear in the manuscript. For example 1, 2, 3. 
NOTE: DOC or DOCX figure files are not permitted.

    > [ ] Rename figure files

Reviewer Comments:

Referee #1:

This manuscript presents a dedicated convergence study of zonal flow generation from ITG turbulence in the Cyclone Base Case. It is clearly written and well-organized. Although it doesn't contain any new physics by itself (ExB staircase in a local gyrokinetic simulation is nothing but a stylish expression of the zonal flow), it would be a valuable reference for a thorough understanding of the ExB staircase which contains rich local and non-local physics.

I recommend a minor revision by discussing the following issues before being accepted by Phys. Plasmas.

1. What would happen to the ExB staircase when the collisionality is included in the local simulation? It is well-known that residual zonal flow in a tokamak damps by collisional friction between trapped and passing ions [Hinton-Rosenbluth, PPCF '99], and it significantly affects the turbulence-zonal flow dynamics [Lin et al., PRL '99], [Diamond et al., PPCF '05] especially in the absence of nonlinear zonal flow damping [Kim-Diamond, PoP '02].

    > Not goal of ths paper to explain the generation of zonal flows
    > Brief communication has to be short so it was not the focus
    > Arne Weikl 2017 -> Heat flux threshold gets shifted under collision 
      [ ] Include statement of no collision in the "Setup" paragraph

2. What would happen to the ExB staircase when the mean (profile-driven) radial electric field is included in the local simulation? There have been analytic studies of the effect of the mean ExB flow on the zonal flows [Kim-Diamond, PRL '03], [Kagan-Catto, PoP '09], [Hsu-Diamond, PoP '15], but a discussion from the point of view of the ExB staircase would be helpful for readers, as the distinction between the mean and the zonal ExB flows becomes elusive in the ExB staircase problem.

    > From the paper it is not clear what will happen and we do also not know

3. As a related issue, I request a discussion on the zonal modes (structures, i.e., zonal density, etc) in the local simulation study of the ExB staircase, which could be interpreted as a quasi-periodic profile corrugation. The profile corrugation is thought to be an essential symptom of the presence of the ExB staircase [Hahm-Diamond, JKPS '18] with an argument based on the radial force balance, but a discussion based on a point of view of the zonal mode generation would be interesting.

    > No data to following issue was also not the goal of this paper

Referee #2:

This short paper addresses one of the fundamental questions in magnetized plasma turbulence and self-organization. The corrugations (so-called 'staircase' pattern) observed in several simulations pose a challenge of basic understanding. While the paradigm of turbulence suppression by ExB zonal flows has been established since a long time, the appearance of zonal structures, and in particular what determine their radial size, is still unclear.

The main conclusion of the paper is that the radial size of these structures (mesoscale) are, at least in part, due to 'local' physics - which of course does not mean that nonlocal effects are not important.

My main recommendation is that this paper should be accepted, but a few points should be addressed and a few more simulations are required to really make the case more convincing.

1. The zonal structures, once formed, completely quench turbulence in all cases shown here. The end state is thus a pure zonal flow, which persist indefinitely as they should in the collisionless limit. However, the question remains whether the end state depends on the initial condition. In this work, nothing is said about the latter. A proper statistical study would imply an ensemble of simulations with different initial conditions (meaning same equilibrium but different initial perturbations) to address this issue. The statistically meaningful prediction is that of the ensemble average of these runs. This referee strongly advises the authors to run such a set of ensemble runs for at least two most representative cases. This would really make the case much stronger.

    > [ ] Run simulations with different initial conditions (cosine5, noise, default = cosine2) (Florian Rath 2022 as inspiration)

2. The periodicity of the flux-tube implies that the structures tend to have a radial size equal to an integer divisor of the box size. The cases shown all take integer multiples of the initial radial box size. What if non-integer multiples of the radial box size (e.g. 1.5 x , 2.5 x, ...) are considered?

    > [ ] Run simulation with requested box sizes 1.5x1.5, 2.5x2.5
    > Selected box sizes are the mst convenient ones 
    > Not every box is allowed because of the border conditions

3. It would be helpful to plot the radial profiles of R/L_T. Are these profiles peaked at radial positions of zero ExB shearing rate? In the mind of the authors, is the corrugation of temperature gradient the result of what is essentially a turbulence-ExB zonal flows interaction, or is it an essential part of the establishment of the radial structures?

    > [ ] Additional diagnostics (xs_kyzero_dens, xs_kyzero_ene_perp, xs_kyzero_par) for simulation of point 1. of referee #2
    > [ ] Maybe include one sentence in paper
    > Turbulent transport defines R/L_T. When shearing rate is crossing zero a much smaller gradient is needed

4. It looks a bit strange to give a range of pattern size with 4 digits (abstract and throughout the paper).

    > [ ] Remove the last two digits from range

### 2023-05-18 briefcommunication/briefcommunication.tex
Rename folder typo

--- 2023-05-09 briefcommunicaton/briefcommunication.tex
@@
     The radial size convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-    Its is shown that a mesoscale pattern size of \linebreak $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
+    Its is shown that a mesoscale pattern size of \linebreak $\sim 57-76\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
 
@@
 	\end{tabular}
-}{fig:wexb-stable-comparison}{}
+}{fig:wexb-stable-comparison}{0.98}
 
@@
 % mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
-A mesoscale pattern size of $\sim 57.20 - 76.27\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
+A mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
 % compare to other studies

### 2023-05-18 briefcommunication/briefcommunication.tex
#37: Rename pictures and seperate in folder and remove last two digit of range

@@
 
-In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
+In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76,~89)\,\rhoth$ in the radial and binormal direction.
 Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
@@
 % summary
-The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
+The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57 - 76\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
 \\
@@
 
-\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison}.pdf}{
+\includegraphicsTwoCol{{Briefcommunication/1}.pdf}{
 	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
@@
 
-\includegraphicsTwoCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison}.pdf}{
+\includegraphicsTwoCol{{Briefcommunication/2}.pdf}{
 	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
@@
 
-\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+\includegraphicsOneCol{{Briefcommunication/3}.pdf}{
 	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
@@
 By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
-Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
+Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57,~ 76]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
 \newpage
@@
 The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
-Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57,~ 76]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
 The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
@@
 The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
-\includegraphicsOneCol{{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_eflux_comparison}.pdf}{
+
+\includegraphicsOneCol{{Briefcommunication/4}.pdf}{
 	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
@@
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.\\
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$.\\
 

### 2023-05-21 bachelorthesis/01-dedication.tex
#39: dedication with picture in black and white
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
         \clip(0,0)circle[x radius=3cm, y radius=4cm];
-        \node{\includegraphics[width=6cm]{blacky1_black_white.jpg}};
+        \node{\includegraphics[width=6cm]{Dedication/blacky1_black_white.jpg}};
     \end{tikzpicture}
     \vspace{2cm}\\
-    This thesis is dedicated to my cat \textbf{Blacky}\\ which is to this date 19 years old and followed me\\ since I was 5 years through my life.\\You were always a big part of my life and\\ I will miss you everyday when you are gone.\\
+    (\textborn) 29.05.2005 - (\textdied) 12.05.2023\\
+    \bigskip
+    This thesis is dedicated to my cat \textbf{Blacky}\\ which get nearly 19 years old and followed me\\ since I was 5 years through my life.\\You were always a big part of my life and\\ 
+    \bigskip
+    I will miss you.\\
     \bigskip

### 2023-05-21 bachelorthesis/02-acknowledgement.tex
#39: Removed name from dedication
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 
-First of all, I would like to express my greatest gratitude to my supervisor Arthur Peeters and Florian Rath.
-I would like to thank both of you for your guidance and support during the submission process of the briefcommunication and the helpful hands during the the progress of this thesis and the restart script.
+First, I would like to express my greatest gratitude to my supervisor Arthur Peeters and Florian Rath.
+I would like to thank both of you for your guidance and support during the submission process of the briefcommunication and the helping hands during the progress of this thesis and the restart script.
 \bigskip
@@
 
-I would also like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend.
-In Addition to that I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time sharing the office together.
+Also, I would like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend.
+In Addition to that I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time-sharing the office together.
 \bigskip
 
-I will also thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get better in spelling. \bigskip
+I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get good at english spelling. \bigskip
 
-Outside academia, I would like to express my gratitude to my parents, my brother and his wife, my sister and my girlfriend for their encouragement, endurance and support during the time of this thesis.
-I know sometimes I am an workaholic and I need to be reminded outside my academic work.
+Outside academia, I would like to express my gratitude to my parents, my brother, his wife and daughter, my sister and my girlfriend for their encouragement, endurance and support during the time of this thesis.
+I know sometimes I am a workaholic and I need to be reminded about a life outside my academic work.

### 2023-05-21 bachelorthesis/03-abstract.tex
#39: Fixed Linebreak
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 Such pattern formation has been observed in local gradient-driven flux-tube simulations as well as global gradient-driven and global flux-driven studies.
+\bigskip
 
-To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimal minimum resolution for the performed simulations.
+To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimal minimum resolution for the performed simulations. \bigskip
 
 For convenience a \texttt{python} script (\texttt{slurm\_monitor.py}) was written which monitors the simulation on the \texttt{btrzx1}-cluster and starts/restarts until the completion criteria is fulfilled.
+\bigskip
 
-Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57.20-76.27\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
+Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57-76\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
 This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG driven turbulence.
@@
 
-Ionen-Temperaturgradienten getriebener Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal FLow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf.
+Ionen-Temperaturgradienten getriebene Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal FLow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf.
 Solche Strukturbildungen wurden sowohl in lokal gradientengetriebene Flussschlauchsimulationen als auch in global gradientengetriebenen und global flussgetriebenen Untersuchungen entdeckt.
+\bigskip
 
-Um den Aufwand der Berechnungen der Simulationen zu reduzieren wurden mehrere kleinere Inputparameter für GKW (Gyro Kinetic Workshop) getestet um die optimal kleinste Auflösung zu finden ausgeführten Simulationen zu finden.
+Um den Aufwand der Berechnungen der Simulationen zu reduzieren wurden mehrere kleinere Inputparameter für GKW (Gyro Kinetic Workshop) getestet um die optimal kleinste Auflösung für die ausgeführten Simulationen zu finden.
+\bigskip
 
 Zur Einfachheit wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist.
+\bigskip
 
-Weiterhin wurde doch mehrere radiale konvergierende Boxgrößen-Scans gezeigt, dass eine Mesoskalengröße von $\sim 57.20-76.27\,\rhoth$ inhärent zur ITG getriebenen Turbulenz mit Cyclone Base Parameter in lokalen Limit ist.
+Weiterhin wurde doch mehrere radiale konvergierende Boxgrößen-Scans gezeigt, dass eine Mesoskalengröße von $\sim 57-76\,\rhoth$ inhärent zur ITG getriebenen Turbulenz mit Cyclone Base Parameter in lokalen Limit ist.
 Dieses Ergebnis impliziert auch, dass eine typische Skala für den lawinenenartig Transport inhärent für die ITG getriebene Turbulenz ist.

### 2023-05-21 bachelorthesis/04-declaration.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 \bigskip
-and is a further iteration of this publication (brief communication published in the journal \enquote{Physics of Plasma}).
+and is a further iteration of this publication (brief communication).
 It provides additional plots and paragraphs that were not necessary to be included in the final publication.
-The format was changed as well because the publication was written in REV\TeX~4.1 and this bachelor thesis in \texttt{scrreprt}.
 The brief communication can be found in Appendix \ref{append:paper}.

### 2023-05-21 bachelorthesis/21-motion.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 
-In magnetic confinement devices like the tokamak reactor, the charged particles experience magnetic and electric fields which results in distinct motion under tha associated force.
+In magnetic confinement devices like the tokamak reactor, the charged particles experience magnetic and electric fields which results in distinct motion under the associated force.
 Charged particles can be separated in species, e.g.
 electrons and ions, which will be later on not displayed in the governing equation.
-Throughout this thesis the charge $q$ or the mass $m$ indicate the quantities of a specific spiecies.
+Throughout this thesis the charge $q$ the mass $m$ or the temperature $T$ indicate the quantities of a specific species.
 
@@
 Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field.
-This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as:
+This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as
 \begin{gather}
@@
 \end{gather}
-where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by:
+where $m$ and $q$ are the mass and the charge of the particle and $B$ the strength of the magnetic field. The radius, the so called \textit{Larmor radius}, of this motion is given by
 \begin{gather}
@@
 \end{gather}
-which center is often as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.
+which center is often referred as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.\\\bigskip
 
-Due to the Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical species thermal velocity is
+Due to the Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical thermal velocity is
 \begin{gather}
@@
 \end{gather}
-where $T$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th}$ the \textit{thermal Larmor radius} gets introduced as \cite{Wesson2011}
+where $T$ represents the species temperature. Based on the thermal velocity $v_\mathrm{th}$ the \textit{thermal Larmor radius} gets introduced as
 \begin{gather}
-    \rhoth = \frac{m\vth}{|q|B}~.
+    \rhoth = \frac{m\vth}{|q|B}~. \cite{Wesson2011}
 \end{gather}
@@
 \end{gather}
-The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.
+The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.\\\bigskip
 
-Since magnetic fields are not always homogenous an inhomogeneous magnetic field with its gradient $\nabla B$ contains a component parallel to the magnetic field which is given by:
+Since magnetic fields are not always homogenous, an inhomogeneous magnetic field with its gradient $\nabla B$ containing a component parallel to the magnetic field which is given by
 \begin{gather}
-    \nabla_\parallel B = \frac{\vect{B}}{B} \cdot \nabla B~.
+    \nabla_\parallel B = \frac{\vect{B}}{B} \cdot \nabla B 
 \end{gather}
-This causes the force
+and causes the force
 \begin{gather}
@@
 \end{gather}
-with \textit{magnetic moment} $\mu$ which can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
+with \textit{magnetic moment} $\mu$.
+The magnetic moment $\mu$ can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
 The caused force has it application in the mirror effect where a charged particle gets reflected due to this force.
@@
     \item \textbf{$\exb$ Drift:}\\
-    If an electric field $\vect{E}$ with a perpendicular component is present together with the magnetic field $\vect{B}$ (both fields are homogenous) the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
+    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
     \begin{gather}
@@
     \end{gather}
-    which is called the $\exb$-drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
+    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
     \item \textbf{$\nabla B$ Drift:}\\
-    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$-drift defined by
+    Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
     \begin{gather}
@@
     \end{gather}
-    The $\nabla B$-drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$-drift depends on the species type.
+    The $\nabla B$ drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
     \item \textbf{Curvature Drift:}\\

### 2023-05-21 bachelorthesis/22-confinement.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 
-In a tokamak devices strong magnetic fields confine the hot plasma.
-As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion containing gyro motion and slow perpendicular gyro center drifts.
-Because of the much smaller size of the Larmor radius compared to the device size $R_0$ the particle and energy losses are caused by the gyro center drift.
+In tokamak devices strong magnetic fields confine the hot plasma.
+As mentioned in Chapter \ref{sec:motion} a magnetic field forces a perpendicular particle motion and a motion which contains the gyro motion and slow perpendicular gyro center drifts.
+Because of the much smaller size of the Larmor radius compared to the device size $R$ the particle and energy losses are caused by the gyro center drift.
 To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus.

### 2023-05-21 bachelorthesis/23-iontempgrad.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 The described electrostatic forces acting on the tokamak plasma causes a variety of collective phenomena.
-Microinstabilities is one of this phenomena which are instabilities on the spatial scale of the Larmor radius and frequencies much smaller than the cyclotron frequency.
-These instabilities are driven bei density and temperature gradients that are present in the tokamak devices.
+Microinstabilities is one of these phenomena which are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$.
+These instabilities are driven by density and temperature gradients that are present in the tokamak devices.
 Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors.
-The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability \cite{Coppi1967, Cowley1991, Rudakov1961}
+The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability.
+\cite{Coppi1967, Cowley1991, Rudakov1961}
 
 \includegraphicsHere{Theory/ITG-Turbulence.pdf}{
-	Sketch of ion temperature gradient driven instability on the outboard low field sid of the tokamak \cite{Casson_PHD}
+	Sketch of ion temperature gradient driven instability on the outboard low field side of the tokamak \cite{Casson_PHD}
 }{fig:ITG}{0.5}
 
-The ITG driven is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard side of the tokamak.
+The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak.
 The mechanism will be briefly explained with the help of Fig.
-\ref{fig:ITG} \cite{Beer_PHD, Casson_PHD, Dannert_PHD}.
-An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature and $\nabla B$ dirft $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells).
+\ref{fig:ITG} and Ref.
+\citenum{Beer_PHD, Casson_PHD, Dannert_PHD}.
+An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ dirft with velocity $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells).
 This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response.
-As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential causes an electric field which results in a $\exb$ drift $\vect{v}_E$.
+As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$.
 The $\exb$ drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip
 
-The described mechanism causes perturbation on the outboard side of the tokamak this region has a \enquote{bad curvature} and instability happens there.
-On the inboard side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}.
-Furthermore such turbulent structures are called \textit{eddies} and the have a typical radial scale of serval Larmor radius.
-\cite{Newins2006}.
+The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there.
+On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}.
+Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of serval Larmor radius.
+\cite{Newins2006}
 Through the explanation above is clear that the initial perturbation is growing when the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$.

### 2023-05-21 bachelorthesis/24-zonalflows.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 
-Zonal flows are referred in the tokamak plasma to the $\exb$ flows [Eq.( \ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}.
-\cite{Diamond2005} Because of the variation only across flux-surfaces only the zonal flow is tangential to the flux-surfaces and do not contribute to the turbulent transport across flux-surfaces caused by the $\exb$ motion.
-Zonal flows are driven nonlinear through turbulent Reynolds stresses \cite{Diamond1991}.
-For more about the mechanism behind zonal flow generation the reader is referred to Ref.
+Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.( \ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}.
+\cite{Diamond2005} Because of the variation only across flux-surfaces the zonal flow is tangential to those surfaces and do not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion.
+Zonal flows are driven nonlinear through turbulent Reynolds stresses.
+\cite{Diamond1991} For more details about the mechanism behind zonal flow generation the reader is referred to Ref.
 \citenum{Brizard2007, Dubin1983, Hasegawa1978, Krommes2000}.
@@
 \begin{enumerate}
-	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10 \rhoth which is larger than the Larmor radius but smaller than the machine size $R_0$.
+	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10\,$\rhoth$ which is larger than the Larmor radius but smaller than the machine size $R$.
 	\item[(2)] The structure are \textit{quasi stationary} in space and varies on time scales much longer than the typical turbulent time scales.
-	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$-shearing rate of the order of $10^{-1} \vth/R_0$.
+	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$ shearing rate of the order of $10^{-1} \vth/R$.
 	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. \cite{McMillan2009}
@@
 \begin{equation}
-	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
+	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}~,
 	\label{eq:shearingrate}
@@
 \begin{equation}
-\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
+\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)~.
 \end{equation}
@@
 As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation.
-Avalanches are initial when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing \cite{Idomura2009,McMillan2009}.
-The fully developed avalanche resluts in almost completely quenched turbulence.
-Its development results in a discontinuous step int he dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}.
+Avalanches are initial when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing.
+\cite{Idomura2009,McMillan2009} The fully developed avalanche results in almost completely quenched turbulence.
+Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}.
 \cite{Peeters2016,Weikl2017}

### 2023-05-21 bachelorthesis/25-gyrokinetics.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 Because of the large number of particle in the fusion plasma a prediction only using Newton-Maxwell dynamics results in an impossible task for computers, but this problem can be solved with a statistical approach.
-For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles postion $\vect{x}$ and velocity $\vect{v}$ is needed.
-Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often considered \cite{Garbet2010} which gives thourgh evolution of the particle density distribution function the \textit{Vlasov equation}
+For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles position $\vect{x}$ and velocity $\vect{v}$ is needed.
+Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often considered \cite{Garbet2010} which gives through evolution of the particle density distribution function the \textit{Vlasov equation}
 \begin{gather}
@@
 \end{gather}
-To obtain a closed system the Maxwell equations with the particle density $n$ and current density $j$ can be descripted the distribution function as follows:
+To obtain a closed system the Maxwell equations with the particle density $n$ and current density $j$ can be descripted the distribution function as follows
 
@@
 
-Dynamics in tokamak plasma is descripted by spatio-temporal scales which is also referred as \textit{gyrokinetic ordering}.
+Dynamics in tokamak plasma is described by spatio-temporal scales which is also referred as \textit{gyrokinetic ordering}.
 This seperation is a comparison between the tuburbulence frequency $\omega$ and the cyclotron frequency $\omega_\mathrm{c}$ ($\omega/\omega_\mathrm{c} \sim 10^{-3})$ due to strong external magnetic fields which causes a fast gyro motion.
-The length scales of the turbulence are associated with the wavevector $\vect{k}$ which can be seperated into an perpendicular component $k_\perp = |\vect{k} \times \vect{\hat{b}}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{\hat{b}}|$ where $\vect{\hat{b}}$ is parallel to the poloidal component of the magnetic field.
+The length scales of the turbulence are associated with the wavevector $\vect{k}$ which can be seperated into an perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field.
 The perpendicular component $k_\perp$ scales with the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significant smaller as the equilibrium density $n_0$ what can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
-The gradient length compares with the machnine size which leads to the normalized Larmor radius $\rho_* = \rhoth/R_0 \sim 10^{-3} - 10^{-4}$ with $R_0$ as major radius of the tokamak.
-The parallel component $k_\parallel$ on the other hand scales with the machine size.
+The gradient length compares with the machine size $R$ which leads to the normalized Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak.
+The parallel component $k_\parallel$ on the other hand scales with the machine size $R$.
 Experiments also show that in the core plasma the fluctuation amplitude of the density perturbation $\delta n/n_0$ and magnetic field fluctuation $\delta B/B_0$, with $B_0$ as strength of the magnetic field in equilibrium, is of order $\sim 10^{-4}$.
@@
 In this set of coordinates the gyrocenter follows the magnetic field with the parallel velocity $v_{\parallel}$.
-The gyro motion will be descripted together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.
+The gyro motion will be describd together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.
 
@@
 For that the particle postion $\vect{x}$ will be described through the gyrocenter postion $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$ with the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$.
-The particle velocity $\vect{v}$ gets decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{\hat{b}}$ and
-a perpendicular component $v_\perp = |\vect{v} \times \vect{\hat{b}}|$ which relates to the magnetic moment $\mu$.
+The particle velocity $\vect{v}$ gets decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{b}$ and
+a perpendicular component $v_\perp = |\vect{v} \times \vect{b}|$ which relates to the magnetic moment $\mu$.
 Through the gyrokinetic ordering a phase space transformation can be performed which is based on the Lie perturbation theory, the Hamiltonian or Lagrangian formalism and an asymtotic expansion in the smallness parameter $\epsilon_\mathrm{g}$.
 In the lowest order is the magnetic moment an exact invariant which makes the gyro phase $\zeta$ an ignorable variable.
-\cite{Garbet2010,Cary1981,Cary1983} Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transferres the Vlasov equation into the \textit{gyrokinetic equation}
+\cite{Garbet2010,Cary1981,Cary1983} Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transforms the Vlasov equation into the \textit{gyrokinetic equation}
 \begin{gather}
@@
 
-All simulation in this thesis are performed with GKW (Gyro Kinetic Workshop) uses in the application of the gyrokinetic equation the $\delta f$ approximation and local limit approximation.
+All simulation in this thesis are performed with GKW (Gyro Kinetic Workshop) which uses in application of the gyrokinetic equation the $\delta f$ approximation and local limit approximation.
 For the $\delta f$ approximation the particle density distribution function f gets seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}).
@@
 Because of the tokamak symmetry all equilibrium constants are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general.
+\\\bigskip
 
@@
 This results in the new coordinates $\{\xcoord, \ycoord, s, v_\parallel, \mu\}$.
-Combining the $\delta f$ approximation and the Hamada coordinates and the gyrokinetic equation can be formualted as follows
+Combining the $\delta f$ approximation and the Hamada coordinates the gyrokinetic equation can be formualted as follows
 \begin{gather}
@@
 where $g$ is a function which contains the the perturbation $\delta f$ and the equilibrium $f_0$.
-Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{\chi}$ are drift velocities which are caused by the inhomogeneous magnetic field and inhomogeneous magnetic field (${v}_\mathrm{D}$) and through the $\exb$ drift.
-The source term $S$ combines from the distribution function in the equilibrium $f_0$ and a correction term for the collisions as wall as the energy injection term.
-Due to prediction boundary conditions in the radial direction the radially gradients cannot change in time which is also referred as \textit{gradient driven} approach.
+Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{v}_\chi$ are drift velocities which are caused by the inhomogeneous magnetic field and the perturbed magnetic field ($\vect{v}_\mathrm{D}$) and through the $\exb$ drift ($\vect{v}_\chi$).
+The source term $S$ combines from the distribution function in the equilibrium $f_0$ and a correction term for the collisions as well as the energy injection term.
+Due to prediodic boundary conditions in the radial direction the radially gradients cannot change in time which is also referred as \textit{gradient driven} approach.
 \cite{Peeters2009}

### 2023-05-21 bachelorthesis/32-diagnostics.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 \begin{equation}
-	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord),
+	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord)~,
 	\label{eq:shearingrate_fourier}
 \end{equation}
-where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
+where $\hatwexb$ is the complex Fourier coefficient and $\kzf = 2\pi \nzf/L_\xcoord$
 % explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
@@
 % introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.

### 2023-05-21 bachelorthesis/34-restartscript.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 As stated in Chapter \ref{sec:cluster} the \texttt{btrzx1}-cluster has a wall time of 24 hours but since simulations typical run longer than 24 hours one have to restart the simulation after the wall time gets exceeded.
-Some simulation need even longer so that the simulation itself needs multiple weeks to finish with a result which makes the restart of the simulation a daily task that could be easily forget.
+Some simulation need even longer so that the simulation itself needs multiple weeks to finish which makes the restart of the simulation a daily task that could be easily forgotten.
 So to eliminate this minor inconveniences the following restart script were written.
 
-As script language \texttt{python} was chosen over shell because of its easiness to write code and the variety of tools which can interact with the \texttt{bash}-shell.
+As script language \texttt{python} was chosen because of its easiness to write code and the variety of tools which can interact with the \texttt{bash} shell.
 \texttt{bash} was also considered but after some issue occurred regarding the mechanism to check wether a file exist or not, \texttt{python} was selected.
+Furthermore, the script has no requirements because it runs on the standard \texttt{python3} library.
 
@@
 
-To run the restart script \texttt{python3} has to be installed on the system and should be accessible by the command line. With the following command can be the monitoring started:
+To run the restart script \texttt{python3} has to be installed on the system and should be accessible via the command line. The monitoring can be started with the following command:
 \begin{lstlisting}[language=Bash]
@@
 \begin{enumerate}
-    \item \texttt{nohup}: Runs command in the background while terminal can still be used or closed. \cite{nohup} The command for this method would be: \\
+    \item \texttt{nohup}:\\
+    Runs command in the background while terminal can still be used or closed. \cite{nohup} The command for this method would be:
 \begin{lstlisting}[language=Bash]
@@
 \end{lstlisting}
-    \item \texttt{screen}: Opens a virtual terminal session which is detached from the main terminal and can be entered and closed. \cite{screen} Here the command would be:
+    \item \texttt{screen}:\\
+    Opens a virtual terminal session which is detached from the main terminal (like a window) and can be entered and closed. \cite{screen} Here the command would be:
 \begin{lstlisting}[language=Bash]
@@
 and to cancel the monitoring use the keyboard shortcut \keys{\ctrl + C} or kill screen itself with \keys{\ctrl + D}.
-To leave the screen use ( \keys{\ctrl + A} ) + \keys{D} and to enter the screen use:
+To leave the screen use (~\keys{\ctrl + A}~) + \keys{D} and to enter the screen use:
 \begin{lstlisting}[language=Bash]
@@
 \begin{itemize}
-    \item The script has no requirements because it runs on the standard \texttt{python3} library
     \item \textbf{\textcolor{red}{!Required!}} Set job name not longer than 8 characters to identify simulation\\
@@
           \texttt{-\/-walltime [WALLTIME]~~~~~~(d-hh:mm:ss)~(default=0-24:00:00)}
-    \item Start/Restarts simulation until job criteria is suffused \\
+    \item Start/Restarts simulation until job criteria is suffused\\
           \texttt{-n [TIMESTEPS]~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\,\,(default=10000)}\\
@@
 \end{itemize}
+All boolean parameters are switches, i.e., the praser option only changes the value from \texttt{False} to \texttt{True}.
 

### 2023-05-21 bachelorthesis/41-resolution.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
-	\item Reduce to knowm minimum number of grid points to verify result in general.
+	\item Reduce to known minimum number of grid points to verify result in general.
 \end{enumerate}
@@
 \ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is simular to the results from Ref.
-\citenum{Peeters2016} with subdued turbulence after $\sim 3000 R/\vth$.
+\citenum{Peeters2016} with subdued turbulence after $\sim 3000\,R/\vth$.
 
 \includegraphicsHere{S6_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for benchmark
+	Time trace of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for benchmark
 }{fig:eflux-ns16-nvpar64-nmu9}{}
@@
 \citenum{Peeters2016}].
-Therefore following parameters were used:
+Therefore following parameters were used
 \begin{gather*}
@@
 \includegraphicsHere{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ and $\rlt = 6.3$ for benchmark
+	Time traces for $\rlt = 6.0$ and $\rlt = 6.3$ for benchmark
 }{fig:eflux-gradient-length}{}
@@
 
-In the following the number of grid points for the parallel velocity $\Nvpar$ is reduced to:
+In the following the number of grid points for the parallel velocity $\Nvpar$ is reduced to
 \begin{gather*}
@@
 In Fig.
-\ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visable that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period.
-But resolution $\Nvpar = 32$ is as well not acceptable according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000 R/\vth$.\\
-So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution will be:
+\ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visible that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period.
+But resolution $\Nvpar = 32$ is not acceptable as well according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000\,R/\vth$.\\
+So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution will be
 \begin{gather*}
-	\boxed{\Nvpar = 48}
+	\boxed{\Nvpar = 48}~.
 \end{gather*}
-With this number of grid points the time till turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500 R/\vth$.
+With this number of grid points the time till turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500\,R/\vth$.
 
 \includegraphicsHere{Comparison/Resolution/Nvpar/S6_rlt6.0_boxsize1x1_Ns16_Nvpar16-32-48-64_Nmu9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced parallel velocity grid points $\Nvpar$
+	Time traces for reduced parallel velocity grid points $\Nvpar$
 }{fig:eflux-ns16-nvpar16-32-48-64-nmu9}{}
@@
 
-As next step the number of grid points for the magnetic moment $\Nmu$ were reduced with:
+As next step the number of grid points for the magnetic moment $\Nmu$ were reduced with
 \begin{gather*}
@@
 \ref{fig:eflux-ns16-nvpar64-nmu6-9} shown, the reduction of grid points for the magnetic moment does not significantly impact the suppression of turbulence.
-The turbulence enters the stationary state in both cases after $\sim 3000 R/\vth$.
+The turbulence enters the stationary state in both cases after $\sim 3000\,R/\vth$.
 
 \includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu6-9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced magnetic moment grid points $\Nmu$
+	Time traces for reduced magnetic moment grid points $\Nmu$
 }{fig:eflux-ns16-nvpar64-nmu6-9}{}
@@
 To conclude a curial result the number of grids point for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}.
-In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with the both criteria in mind, to the following resolution:
+In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with the both criteria in mind, to the following resolution
 \begin{gather*}
-	\boxed{\Nmu = 9}
+	\boxed{\Nmu = 9}~.
 \end{gather*}
@@
 \includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar48_Nmu6-9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ and $\Nvpar = 48$ for reduced magnetic moment grid points $\Nmu$
+	Time traces for reduced magnetic moment grid points $\Nmu$
 }{fig:eflux-ns16-nvpar48-nmu6-9}{}
@@
 
-In the final step the number of grid points for the magnetic field $\Ns$ get reduced with the following parameters:
+In the final step the number of grid points for the magnetic field $\Ns$ get reduced with the following parameters
 \begin{gather*}
@@
 In Fig.
-\ref{fig:eflux-ns12-16-nvpar64-nmu9} is clearly visible that the reduction to $\Ns = 12$ does not satisfy the criteria \textbf{(2)} because the the stationary state of the turbulence gets instabil after $\sim 2500 R/\vth$.
-This concludes to the following resolution:
+\ref{fig:eflux-ns12-16-nvpar64-nmu9} can be seen that the reduction to $\Ns = 12$ does not satisfy the criteria \textbf{(2)} because the stationary state of the turbulence gets instabil after $\sim 2500\,R/\vth$.
+This concludes to the following resolution
 \begin{gather*}
-	\boxed{\Ns = 16}
+	\boxed{\Ns = 16}~.
 \end{gather*}
@@
 \includegraphicsHere{Comparison/Resolution/Ns/S6_rlt6.0_boxsize1x1_Ns12-16_Nvpar64_Nmu9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced magnetic field grid points $\Ns$
+	Time traces for reduced magnetic field grid points $\Ns$
 }{fig:eflux-ns12-16-nvpar64-nmu9}{}

### 2023-05-21 bachelorthesis/50-closure.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

--- 2023-05-09 bachelorthesis/50-closure.tex
@@
 \newpage
+
+In this thesis the minimal resolution for simulations with GKW in the Cyclone Base parameter were observed in which the number of grid points for the parallel velocity $\Nvpar$ could be reduced from 64 to 48 which halfs the time until suppression.
+\bigskip
+
+Additional the active development of a restart script in \texttt{python} leads to further convenience during the task of performing a simulation on the \texttt{btrzx1} cluster. \bigskip
 

### 2023-05-21 briefcommunication/briefcommunication.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

@@
 
-\title[Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
-{Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
+\title[Size convergence of the \boldmath{$\exb$} staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
+{Size convergence of the \boldmath{$\exb$} staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
 
@@
 \begin{equation}
-	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord),
+	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord)~,
 	\label{eq:shearingrate_fourier}

### 2023-05-31 briefcommunication/briefcommunication.tex
#42: Compare meso sclae for different initial conditions, #37: Include new box sizes in brief communictaion

--- 2023-05-18 briefcommunication/briefcommunication.tex
@@
 Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
-Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Weikl2017, Rath2021} as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
+Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Rath2021}, including collisions \cite{Weikl2017}, as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
 In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
@@
 Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
-This scan covers the realizations $\NR\times\NB \in [1\times1,~2\times2,~3\times3]$.
+This scan covers the realizations $\NR\times\NB \in [1\times1,~1.5\times1.5,~2\times2,~2.5\times2.5,~3\times3]$.
 % mention shorter times scales for stabilization
-Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern always occurs after $\sim 1000~R/\vth$ [Fig.
+Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern almost always occurs after \linebreak $\sim 1000~R/\vth$ [Fig.
 \ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
@@
 		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$     & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$     & $\in [26000, 28000]$ \\
-		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
-		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    & $t_{3\times 5}$ & $\in [1000, 3000]$   \\
+		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{1.5\times 1.5}$ & $\in [7000, 8000]$,   & $t_{2\times 2}$ & $\in [2000, 3000]$,   & $t_{2.5\times 2.5}$ & $\in [2000, 3000]$   \\
+		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    &                     &                       &                 &                       &                     &                      \\
+		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$     & $\in [1000, 3000]$   \\
 	\end{tabular}

### 2023-06-02 briefcommunication/briefcommunication.tex
#37: Old plots, Referees paragraphs, Rearangments

@@
 This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+The negative gradient of the perturbed perpendicular and parallel ion pressure (not shown) exhibit positive corrugations in regions with maximum $|\wexb|$ and negative corrugations at zero crossings of $\wexb$.
+A radial force balance analysis suggests that the structures in $\wexb$ as depicted in Fig.~\ref{fig:wexb-stable-comparison} are not a consequence of the pressure gradient corrugations as discussed elsewhere \cite{hahm2018}.
+Rather, the corrugations in the pressure gradient have to be interpreted as a consequence of the staircase structure in $\wexb$ due to the stabilizing zonal flow - turbulence interaction.
 % insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
@@
 Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\onlinecite{Peeters2016}.
-\\
 % cautios words
@@
 This scan covers the realizations $\NR\times\NB \in [1\times1,~1.5\times1.5,~2\times2,~2.5\times2.5,~3\times3]$.
-% mention shorter times scales for stabilization
-Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern almost always occurs after \linebreak $\sim 1000~R/\vth$ [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
-% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
-%
-% the main result of this test: confirmation of wave length convergence
-As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 
@@
 	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}{}
+}{fig:wexb-eflux-1-2-3-4x1-comparison}{0.96}
 
@@
 		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$, & $t_{2\times 1}$   & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$ & $\in [26000, 28000]$ \\
-		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{1.5\times 1.5}$ & $\in [7000, 8000]$,   & $t_{2\times 2}$ & $\in [2000, 3000]$,   & $t_{2.5\times 2.5}$ & $\in [2000, 3000]$   \\
-		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    &                     &                       &                 &                       &                     &                      \\
+		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$, & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
 		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$, & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$ & $\in [1000, 3000]$   \\
@@
 
+% mention shorter times scales for stabilization
+Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern almost always occurs after $\sim 1000~R/\vth$ [Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
+% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
+%
+% the main result of this test: confirmation of wave length convergence
+As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 % discuss sumewhat smaller radial size
@@
 Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57,~ 76]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
-\newpage
+
+Note that two additional realizations of the $3\times3$ case with different initial conditions and otherwise identical parameters confirm structure formation on scales within the range given above.\\
+
 % what kind of analysis
@@
 	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
-}{fig:eflux-3x1.5-2.5-3-5-comparison}\bigskip
+}{fig:eflux-3x1.5-2.5-3-5-comparison}
 
@@
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$.\\
-
-\bigskip
-\bigskip
-\bigskip
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$. \\
 

### 2023-06-06 briefcommunication/briefcommunication.tex
#37: Revised publication with cleaned up code and references

@@
 \usepackage[format=hang]{caption}
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
 
@@
 
-\input{wordcount.tex}% Word & page count for document
-
 \graphicspath{{../pictures/}}% Path for pictures
-
-%% Apr 2021: AIP requests that the corresponding 
-%% email to be moved after the affiliations
-\makeatletter
-\def\@email#1#2{
-	\endgroup
-	\patchcmd{\titleblock@produce}
-		{\frontmatter@RRAPformat}
-		{\frontmatter@RRAPformat{\produce@RRAP{*#1\href{mailto:#2}{#2}}}\frontmatter@RRAPformat}
-		{}{}
-}
-\makeatother
-
-%% Include graphic for one column with specific place 
-\newcommand{\includegraphicsOneCol}[3]{
-	\begin{center}
-		\centering
-		\captionsetup{type=figure}
-		\includegraphics[width=0.87\linewidth]{#1}
-		\captionof{figure}{#2}
-		\label{#3}
-	\end{center}
-  	\increasecounter{fig}{1}
-}
-
-%\newcommand{\includegraphicsOneCol}[3]{
-%	\begin{figure}[hbt!]
-%		\centering
-%		\includegraphics[width=0.9\linewidth]{#1}
-%		\caption{#2}
-%		\label{#3}
-%  	\end{figure}
-%  	\increasecounter{fig}{1}
-%}
-
-%% Include graphic for two column with specific place, figrue* places graphic anywhere...
-\newcommand{\includegraphicsTwoCol}[4]{
-	\onecolumngrid
-	\begin{center}
-		\centering
-		\captionsetup{type=figure}
-    	\includegraphics[width=#4\textwidth]{#1}
-		\captionof{figure}{#2}
-    	\label{#3}
-	\end{center}
-	\twocolumngrid
-	\increasecounter{fig}{2}
-}
-
-%\newcommand{\includegraphicsTwoCol}[3]{
-%	\begin{figure*}
-%    	\includegraphics[width=0.9\textwidth]{#1}
-%		\caption{#2}
-%		\label{#3}
-%	\end{figure*}
-%	\increasecounter{fig}{2}
-%}
-
-\newcommand{\words}{
-	\immediate\write18{texcount -sum -1 \jobname.tex > wordcount.txt}
-}
 
@@
 \newcommand{\exb}{\mathrm{\:E}\times\mathrm{B}}
-%\newcommand{\hatwexbampvec}{|\hatwexb|_\mathbf{n}}
 \newcommand{\hatwexbamp}{|\hatwexb|_{\nzf}}
-%\newcommand{\NR}{N_\mathrm{R}}
-%\newcommand{\NB}{N_\mathrm{B}}
 \newcommand{\NR}{N_\mathrm{R}}
@@
 \newcommand{\rlt}{R/L_T}
-%\newcommand{\rhoth}{\rho_\mathrm{th}}
 \newcommand{\rhoth}{\rho}
@@
 \end{abstract}
-
 \maketitle
-
-%% TEXT ====================================================================================================================
 
@@
 
-%\section{Introduction}
-%\label{sec:intro}
-
 Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
-Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Rath2021}, including collisions \cite{Weikl2017}, as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
+Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Rath2021}, including collisions \cite{Weikl2017} and background $\exb$ shear \cite{Rath2021}, local flux-driven realizations including mean electric field shear \cite{Seiferling2019}, as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
 In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
-% By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
 By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
@@
 	\caption{
-		Resolution used in this paper for further information the author links to Ref.~\onlinecite{Peeters2016}.
-%: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
-	}
+		Resolution used in this paper for further information the author links to Ref.~\onlinecite{Peeters2016}.}
 	\label{tab:resolution}
@@
 
-In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76,~89)\,\rhoth$ in the radial and binormal direction.
+In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.3,~89.8)\,\rhoth$ in the radial and binormal direction.
 Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
@@
 where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
-% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
 defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
-% define the zonal flow modes' amplitude in terms of the shearing rate
 Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$. 
-% introduces the term basic mode of the pattern for later usage
-In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\
+In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\bigskip
 
@@
 
-%\section{Results}
-%\label{sec:results}
-
-% what kind of analysis
 In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size.
 The scan covers the realizations $\NR\times\NB \in [ 1\times1,~2\times1,~3\times1,~4\times1]$.
-% short introduction of the turbulence - zonal flow dynamics; cite peeters2016 explicitly to avoid any discussions lateron
 Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
-% introduce the fully developed staircase structure
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
@@
 The negative gradient of the perturbed perpendicular and parallel ion pressure (not shown) exhibit positive corrugations in regions with maximum $|\wexb|$ and negative corrugations at zero crossings of $\wexb$.
-A radial force balance analysis suggests that the structures in $\wexb$ as depicted in Fig.~\ref{fig:wexb-stable-comparison} are not a consequence of the pressure gradient corrugations as discussed elsewhere \cite{hahm2018}.
+A radial force balance analysis suggests that the structures in $\wexb$ as depicted in Fig.~\ref{fig:wexb-stable-comparison} are not a consequence of the pressure gradient corrugations as discussed elsewhere \cite{Kosuga2013}.
 Rather, the corrugations in the pressure gradient have to be interpreted as a consequence of the staircase structure in $\wexb$ due to the stabilizing zonal flow - turbulence interaction.
-% insofar the results of peeters2016 are well reproduced; now turn to the description of the main result of this test
 Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
-% yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
 Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\onlinecite{Peeters2016}.
-% cautios words
 Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
-% very long initial quasi-stationary states
 Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
-% the \nzf=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
-During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
-Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
-% first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
+During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical timescale of several $\sim 10^3\,R/\vth$.
+Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,R/\vth$.
 It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbamp$ in the latter two phases (not shown).
 This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
-% second important point: no secular growth of the box scale zonal mode
 Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
-% summary
 The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57 - 76\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
-\\
+\bigskip
 
-% what kind of analysis
 Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
 This scan covers the realizations $\NR\times\NB \in [1\times1,~1.5\times1.5,~2\times2,~2.5\times2.5,~3\times3]$.
+Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern almost always occurs after $\sim 1000\,R/\vth$ [Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
+As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 
-\includegraphicsTwoCol{{Briefcommunication/1}.pdf}{
+%% Include graphic for two column with specific place
+\onecolumngrid
+\begin{center}
+	\centering
+	\captionsetup{type=figure}
+	\includegraphics[width=0.85\textwidth]{Briefcommunication/1.pdf}
+	\captionof{figure}{
 		\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
 		\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
-}{fig:wexb-eflux-1-2-3-4x1-comparison}{0.96}
+	}
+	\label{fig:wexb-eflux-1-2-3-4x1-comparison}
+\end{center}
+\twocolumngrid
 
-\includegraphicsTwoCol{{Briefcommunication/2}.pdf}{
+%% Include graphic for two column with specific place
+\onecolumngrid
+\begin{center}
+	\centering
+	\captionsetup{type=figure}
+	\includegraphics[width=0.98\textwidth]{Briefcommunication/2.pdf}
+	\captionof{figure}{
 		Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
@@
 		\end{tabular}
-}{fig:wexb-stable-comparison}{0.98}
+	}
+	\label{fig:wexb-stable-comparison}
+\end{center}
+\twocolumngrid
 
-\includegraphicsOneCol{{Briefcommunication/3}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes
-}{fig:eflux-1x1-2x2-3x3-comparison}
+%% Include graphic for one column with specific place 
+\begin{center}
+	\centering
+	\captionsetup{type=figure}
+	\includegraphics[width=0.87\linewidth]{Briefcommunication/3.pdf}
+	\captionof{figure}{Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes}
+	\label{fig:eflux-1x1-2x2-3x3-comparison}
+\end{center}
 
-% mention shorter times scales for stabilization
-Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern almost always occurs after $\sim 1000~R/\vth$ [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
-% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
-%
-% the main result of this test: confirmation of wave length convergence
-As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
-% discuss sumewhat smaller radial size
-By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
+By contrast to the radial box size scan, the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size. 
 Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57,~ 76]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
-
+The scale of structures developing in the $1.5\times1.5$ and $2.5\times2.5$ realizations (not included in Fig.~\ref{fig:wexb-stable-comparison} to preserve the clarity of this figure) also lie within the range given above.
 Note that two additional realizations of the $3\times3$ case with different initial conditions and otherwise identical parameters confirm structure formation on scales within the range given above.\\
 
-% what kind of analysis
 In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
@@
 The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no signifciant plateau at positive shear.
-Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
+Instead, the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
 The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each (not shown).
 
-\includegraphicsOneCol{{Briefcommunication/4}.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes
-}{fig:eflux-3x1.5-2.5-3-5-comparison}
+%% Include graphic for one column with specific place 
+\begin{center}
+	\centering
+	\captionsetup{type=figure}
+	\includegraphics[width=0.87\linewidth]{Briefcommunication/4.pdf}
+	\captionof{figure}{Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes}
+	\label{fig:eflux-3x1.5-2.5-3-5-comparison}
+\end{center}
 
-% what kind of test
 In the final test the inverse background temperature gradient length $\rlt$ is varied at fixed $3\times3$ box size.
@@
 This scan covers realizations with $\rlt \in [6.0,~6.2,~6.4]$.
-% what is the finite heat flux threshold
 In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
 The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$ in accordance to Ref. \onlinecite{Peeters2016}.
-% Which stationay radial size occurs????
 Although the initial quasi-stationary turbulence in the former case is significantly longer compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
-% Does it confirm our previous tests
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$. \\
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$. \bigskip
 
@@
 
-%\section{Conclusion}
-%\label{sec:conclusion}
-
-% what has been done
 Through careful tests this brief communication confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-% mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
 A mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
-% compare to other studies
 This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{Pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
-% relevance of the outcome
 The occurrence of this mesoscale implies that non-locality, in terms of Ref.
@@
 
-% WORDCOUNT ================================================================================================================
-
-%\wordcountmessage
-
 % BIBLIOGRAPHY =============================================================================================================
 
-\nocite{}
 \bibliography{references.bib}% Produces the bibliography via BibTeX.
@@
 \end{document}
+
 % END MAIN =================================================================================================================

### 2023-06-08 bachelorthesis/01-dedication.tex
#39: Correction Cornelia Lippert

@@
     \vspace{2cm}\\
-    (\textborn) 29.05.2005 - (\textdied) 12.05.2023\\
+    \gtrsymBorn~29.05.2005 - \gtrsymDied~12.05.2023\\
     \bigskip
-    This thesis is dedicated to my cat \textbf{Blacky}\\ which get nearly 19 years old and followed me\\ since I was 5 years through my life.\\You were always a big part of my life and\\ 
+    This thesis is dedicated to my cat \textbf{Blacky}\\ who almost reached the age of 18 years.\\ She followed me through life since I was 5 years old.\\You always were a big part of my life and\\ 
     \bigskip

### 2023-06-08 bachelorthesis/02-acknowledgement.tex
#39: Correction Cornelia Lippert

--- 2023-04-29 bachelorthesis/02-acknowledgement.tex
@@
 
-First of all, I would like to express my greatest gratitude to my supervisor Arthur Peeters and Florian Rath.
-I would like to thank both of you for your guidance and support during the submission process of the briefcommunication and the helpful hands during the the progress of this thesis and the restart script.
+First, I would like to extend my greatest gratitude to my supervisor Prof.
+Arthur Peeters and Dr.
+Florian Rath.
+I would like to thank both of you for your guidance and support during the submission process of the brief communication and the helping hands during the writing process of this thesis and development of the restart script.
 \bigskip
-I would like to thank Bernhard Winkler and Markus Hilt for the great technical support regarding my questions to the \texttt{btrzx1}-cluster which helps me to develope the restart script as it is now.
+
+I would like to thank Bernhard Winkler and Markus Hilt for the great technical support regarding my questions to the \texttt{btrzx1}-cluster which helped me develop the restart script as it is now.
 \bigskip
-I would also like to thank Paul Schwanitz for his helpfulness in many situations and the good discussions in and outside academia as my best friend.
-In Addition to that I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time sharing the office together.
+
+Also, I would like to thank my best friend, Paul Schwanitz, for his support in many situations and the good discussions in- and outside of academia.
+In addition I would like to thank Anna-Maria Pleyer, Dominik Müller and Stefan Barthelmann for the great time of sharing the office together.
 \bigskip
-I will also thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and briefcommunication which helps me to get better in spelling.
-\bigskip
-Outside academia, I would like to express my gratitude to my parents, my brother and his wife, my sister and Hanna Schütz for there encouragement, endurance and support during the time of this thesis.
+
+I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and brief communication, which helps me get better at English spelling. \bigskip
+
+Outside of academia, I would like to extend my gratitude to my parents, my brother, his wife and daughter, my sister and my girlfriend for their encouragement, endurance and support while writing this thesis.
+I am aware that sometimes I am a workaholic who needs to be reminded about the life outside of my academic work.

### 2023-06-08 bachelorthesis/03-abstract.tex
#39: Correction Cornelia Lippert

@@
 
-To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimal minimum resolution for the performed simulations. \bigskip
+To reduce the computational effort for the simulations lower input parameter of GKW (Gyro Kinetic Workshop) were tested to find the optimum of minimum resolution for the performed simulations.
+\bigskip
 
-For convenience a \texttt{python} script (\texttt{slurm\_monitor.py}) was written which monitors the simulation on the \texttt{btrzx1}-cluster and starts/restarts until the completion criteria is fulfilled.
+For convenience, a \texttt{python} script (\texttt{slurm\_monitor.py}) was written to monitor the simulation on the \texttt{btrzx1}-cluster and start/restart until the completion criterion is fulfilled.
 \bigskip
@@
 
-Ionen-Temperaturgradienten getriebene Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal FLow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf.
+Ionen-Temperaturgradienten getriebene Turbulenzen (ITG) weisen nahe niedriger Stabilität Zonal Flow Strukturbildungen, sogenannte $\exb$ Treppenstrukturen, auf Mesoskalen auf.
 Solche Strukturbildungen wurden sowohl in lokal gradientengetriebene Flussschlauchsimulationen als auch in global gradientengetriebenen und global flussgetriebenen Untersuchungen entdeckt.
@@
 
-Zur Einfachheit wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist.
+Der Einfachheit halber wurde ein \texttt{python}-Skript (\texttt{slurm\_monitor.py}) geschrieben, was Simulationen auf den \texttt{btrzx1}-Cluster überwacht und gegebenenfalls startet/neustartet bis das Kriterium zur Vollendung erfüllt ist.
 \bigskip

### 2023-06-08 bachelorthesis/04-declaration.tex
#39: Correction Cornelia Lippert

@@
 
-The author declares that the material presented in this thesis is his own work, unless explicitly stated otherwise. This thesis is based on
+The author declares that the materials presented in this thesis are his own work, unless explicitly stated otherwise. This thesis is based on
 \bigskip
@@
 and is a further iteration of this publication (brief communication).
-It provides additional plots and paragraphs that were not necessary to be included in the final publication.
+It provides additional plots and paragraphs that were included in the publication.
 The brief communication can be found in Appendix \ref{append:paper}.

### 2023-06-08 bachelorthesis/10-motivation.tex
#39: Correction Cornelia Lippert

@@
 Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
-Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Weikl2017, Rath2021} as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
+Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Rath2021}, including collisions \cite{Weikl2017} and background $\exb$ shear \cite{Rath2021}, local flux-driven realizations including mean electric field shear \cite{Seiferling2019}, as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
 In global studies, spanning a larger fraction of the minor radius, multiple radial repetitions of staircase structures are usually observed, with a typical pattern size of several ten Larmor radii.
-% By contrast, the radial domain size of local flux-tube simulations is often restricted to a similar mesoscale.
 By contrast, in the aforementioned local studies the radial size of $\exb$ staircase structures is always found to converge to the radial box size of the flux tube domain.
@@
 \citenum{Peeters2016}.
-%TODO More Text

### 2023-06-08 bachelorthesis/21-motion.tex
#39: Correction Cornelia Lippert

@@
 
-In magnetic confinement devices like the tokamak reactor, the charged particles experience magnetic and electric fields which results in distinct motion under the associated force.
+In magnetic confinement devices like the tokamak reactor, the charged particles experience forces caused by magnetic and electric fields which results in distinct motion under the associated force.
 Charged particles can be separated in species, e.g.
 electrons and ions, which will be later on not displayed in the governing equation.
-Throughout this thesis the charge $q$ the mass $m$ or the temperature $T$ indicate the quantities of a specific species.
+Throughout this thesis the charge $q$, the mass $m$ or the temperature $T$ indicate the quantities of a specific species, i.e., electrons or ions.
 
@@
 
-Due to the Lorentz force particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field.
-This type motion has circular frequency which is often referred as \textit{cyclotron frequency} and is defined as
+Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field.
+This type of motion has circular frequency, which is often referred to as \textit{cyclotron frequency} and is defined as
 \begin{gather}
@@
 \end{gather}
-which center is often referred as \textit{gyrocenter}. Note that since the Lorentz force depends on the species of the particle the circulation direction is opposed between electron in ions.\\\bigskip
+with the center often being referred to as \textit{gyrocenter}.
+Note that since the Lorentz force depends on the species of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip
 
-Due to the Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical thermal velocity is
+Due to Coulomb collisions the plasma gets thermalized. Together with the Maxwell-Boltzmann distribution the typical thermal velocity is
 \begin{gather}
@@
 
-In absence of forces in the parallel direction of the magnetic field the particles can stream freely parallel to the homogenous magnetic field.
-The velocity of this motion is connected to the thermal velocity $\vth$ and is dominated bei electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).
+In absence of forces in the parallel direction of the magnetic field the particles can move in parallel direction to the homogenous magnetic field.
+The velocity of this motion is connected to the thermal velocity $\vth$ and is dominated by electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).
 
-When an electric field with a component parallel to the magnetic field $E_\parallel$ acts on plasma the charged particles gets accelerated by the electric force
+When an electric field with a component parallel to the magnetic field $E_\parallel$ influences the plasma the charged particles are accelerated by the electric force
 \begin{gather}
@@
 \begin{gather}
-    \nabla_\parallel B = \frac{\vect{B}}{B} \cdot \nabla B 
+    \nabla_{\!\parallel} B = \frac{\vect{B}}{B} \cdot \nabla B 
 \end{gather}
-and causes the force
+causes the force
 \begin{gather}
-    F_{\parallel,\nabla_\parallel B} = - \frac{mv^2_{\perp}}{2B} \nabla_\parallel B = \mu \nabla_\parallel B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
+    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
 \end{gather}
@@
 The magnetic moment $\mu$ can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
-The caused force has it application in the mirror effect where a charged particle gets reflected due to this force.
+The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force.
 \cite{Wesson2011}
@@
 \begin{enumerate}
-    \item \textbf{$\exb$ Drift:}\\
+    \item {\boldmath $\exb$} \textbf{Drift:}\\
     If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
@@
     which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
-    \item \textbf{$\nabla B$ Drift:}\\
+    \item {\boldmath $\nabla B$} \textbf{Drift:}\\
     Inhomogeneous magnetic field causes a gradient $\nabla B$ of the magnetic field. Because of that gradient the gyrocenter undergoes a $\nabla B$ drift defined by
@@
     \item \textbf{Curvature Drift:}\\
-    Due to centrifugal force acting on the particle in a curved magnetic filed the gyrocenter experience a curvature drift given by
+    Due to centrifugal force acting on the particle in a curved magnetic field the gyrocenter experiences a curvature drift according to
     \begin{gather}
@@
 To obtain the result for the curvature $\vect{\kappa}$ in Eq.
-(\ref{eq:curvature}) the plasma pressure has to be small in comparison die the magnetic field strength $B$.
+(\ref{eq:curvature}) the plasma pressure has to be small compared to the magnetic field strength $B$.
 In the form of Eq.

### 2023-06-08 bachelorthesis/22-confinement.tex
#39: Correction Cornelia Lippert

@@
 To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus.
-This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux sufaces} and with magnetic field lines which appearing on that surfaces.
-For stability reasons the magnetic field has a toroidal and a poloidal component which also holds equivalence of the plasma pressure.
+This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux surfaces}, and magnetic field lines which appear on these surfaces.
+To maintain stability the magnetic field has a toroidal and a poloidal component which also is equivalant to the plasma pressure.
 \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current.
-Together the components results in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}].
-To characterize the the quality of confinement the so-called \textit{plasma beta} is used and is given by
+Together the components result in a magnetic field which follows helical trajectories [Fig \ref{fig:confinement}].
+To characterize the quality of confinement the so-called \textit{plasma beta} is used and is given as
 \begin{gather}
-    \beta = \frac{nT}{\mu_0 B^2/2}~.
+    \beta = \frac{nT}{\mu_0 B^2/2}~,
 \end{gather} 
 with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$.
-Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates when the plasma beta is smaller the confinement is better.
-In a tokamak reactor the plasma beta has a typical order of few percent.
+Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates that in cases when the plasma beta is smaller the confinement is better.
+For fusion devices the plasma beta has to be smaller than 1 ($\beta < 1$).
+In a tokamak reactor the plasma beta has a typical order of a few percent.
 \cite{Wesson2011}
 \includegraphicsHere{Theory/Tokamak-Torus.pdf}{
-	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or zylinder coordinates ($Z$, $R$, $\phi$).
+	Toroidal flux surfaces in tokamak plasma with helical magnetic field (\textcolor{Ubtgreen}{green} line) in torus coordinates ($\rho$ (radial), $\phi$ (toroidal), $\theta$ (poloidal)) or cylindrical coordinates ($Z$, $R$, $\phi$).
 \cite{Barton2015}

### 2023-06-08 bachelorthesis/23-iontempgrad.tex
#39: Correction Cornelia Lippert

@@
 
-The described electrostatic forces acting on the tokamak plasma causes a variety of collective phenomena.
-Microinstabilities is one of these phenomena which are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$.
-These instabilities are driven by density and temperature gradients that are present in the tokamak devices.
-Turbulence caused from microinstabilities are considered as the main reason for the loss of particles energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors.
-The most dominante instability in tokamak plasma is the ion temperature gradient (ITG) driven instability.
+The described electrostatic forces acting on the tokamak plasma cause a variety of collective phenomena.
+The occurence of Microinstabilities is one of these phenomena.
+They are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$, driven by density and temperature gradients present in the tokamak devices.
+Turbulence by microinstabilities are considered as the main reason for the loss of particle energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors.
+The most dominant instability in tokamak plasma is the ion temperature gradient (ITG) driven instability.
 \cite{Coppi1967, Cowley1991, Rudakov1961}
@@
 The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak.
-The mechanism will be briefly explained with the help of Fig.
+The mechanism is briefly explained with the help of Fig.
 \ref{fig:ITG} and Ref.
 \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}.
-An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ dirft with velocity $\vect{v}_\mathrm{d}$ which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells).
+An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$, which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells).
 This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response.
 As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$.
-The $\exb$ drifts direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip
+The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip
 
@@
 \cite{Newins2006}
-Through the explanation above is clear that the initial perturbation is growing when the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$.
+The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$.
 For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$.

### 2023-06-08 bachelorthesis/24-zonalflows.tex
#39: Correction Cornelia Lippert

@@
-\section{Zonal Flows and Shearing Rate $\wexb$}
+\section{Zonal Flows and Shearing Rate {\boldmath $\wexb$}}
 \label{sec:zonalflow}
 
-Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.( \ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}.
-\cite{Diamond2005} Because of the variation only across flux-surfaces the zonal flow is tangential to those surfaces and do not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion.
-Zonal flows are driven nonlinear through turbulent Reynolds stresses.
+Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.(\ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}.
+\cite{Diamond2005} Because of the variation exclusively across flux-surfaces, the zonal flow is tangential to those surfaces and does not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion.
+Zonal flows are driven nonlinearly by turbulent Reynolds stresses.
 \cite{Diamond1991} For more details about the mechanism behind zonal flow generation the reader is referred to Ref.
@@
 \includegraphicsHere{Theory/Zonal-Flow-Generation.pdf}{
-	Subdued turbulence because if stretched eddies which breaks in smaller structures caused by sheared $\exb$ drifts \cite{Hammett2009}
+	Subdued turbulence because of stretched eddies, which breaks in smaller structures caused by sheared $\exb$ drifts \cite{Hammett2009}
 }{fig:zonalflow}{0.9}
 
-Furthermore, zonal flows are connected with the radially alternating zonal potential what introduces a velocity shear, i.e., a radial variation of the zonal flow velocity.
-This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000} especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}.
+Furthermore, zonal flows are connected with the radially alternating zonal potential, which introduces a velocity shear, i.e., a radial variation of the zonal flow velocity.
+This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}.
 The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a later development of ITG driven turbulence.
-The shearing process tilted und stretches turbulent eddies, caused be the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig.
-\ref{fig:zonalflow}] which characterize the shearing as decorrelation process.
+The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig.
+\ref{fig:zonalflow}],  characterizing the shearing as a decorrelation process.
 Turbulence will then be reduced because of this process.
@@
 	\item[(1)] They occur on a \textit{radial mesoscale} of order of 10\,$\rhoth$ which is larger than the Larmor radius but smaller than the machine size $R$.
-	\item[(2)] The structure are \textit{quasi stationary} in space and varies on time scales much longer than the typical turbulent time scales.
-	\item[(3)] They have an \textit{typical amplitude} in terms of the $\exb$ shearing rate of the order of $10^{-1} \vth/R$.
+	\item[(2)] The structure are \textit{quasi stationary} in space and vary on time scales much longer than the typical turbulence time scales.
+	\item[(3)] They have a \textit{typical amplitude} in terms of the $\exb$ shearing rate of the order of $10^{-1} \vth/R$.
 	\item[(4)] Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate. \cite{McMillan2009}
@@
 As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation.
-Avalanches are initial when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards for positive shearing.
+Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing.
 \cite{Idomura2009,McMillan2009} The fully developed avalanche results in almost completely quenched turbulence.

### 2023-06-08 bachelorthesis/25-gyrokinetics.tex
#39: Correction Cornelia Lippert

@@
 
-Because of the large number of particle in the fusion plasma a prediction only using Newton-Maxwell dynamics results in an impossible task for computers, but this problem can be solved with a statistical approach.
+Because of the large number of particles in the fusion plasma a prediction only using Newton-Maxwell dynamics results in an impossible task for simulation based computations, but this problem can be solved with a statistical approach.
 For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles position $\vect{x}$ and velocity $\vect{v}$ is needed.
-Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often considered \cite{Garbet2010} which gives through evolution of the particle density distribution function the \textit{Vlasov equation}
+Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often preferred \cite{Garbet2010} which results through evolution of the particle density distribution function in the \textit{Vlasov equation}
 \begin{gather}
@@
 \end{gather}
-To obtain a closed system the Maxwell equations with the particle density $n$ and current density $j$ can be descripted the distribution function as follows
-
+To obtain a closed system the Maxwell equations with the particle density $n$ and current density $j$ can be described with the distribution function as follows
 \begin{gather}
@@
 \end{gather}
-which are substituted into the Maxwell equations
+which are then substituted into the Maxwell equations
 \begin{gather}
@@
 \end{gather}
-The Vlasov equation (\ref{eq:vlasov}) together with the Maxwell equations (\ref{eq:densitycurrent}) and (\ref{eq:maxwell}) is the basis of the gyrokinetic model. \cite{Krommes2012}
+The Vlasov equation (\ref{eq:vlasov}) in combination with the Maxwell equations (\ref{eq:densitycurrent}) and (\ref{eq:maxwell}) is the basis of the gyrokinetic model. \cite{Krommes2012}
 
@@
 
-Dynamics in tokamak plasma is described by spatio-temporal scales which is also referred as \textit{gyrokinetic ordering}.
-This seperation is a comparison between the tuburbulence frequency $\omega$ and the cyclotron frequency $\omega_\mathrm{c}$ ($\omega/\omega_\mathrm{c} \sim 10^{-3})$ due to strong external magnetic fields which causes a fast gyro motion.
-The length scales of the turbulence are associated with the wavevector $\vect{k}$ which can be seperated into an perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field.
-The perpendicular component $k_\perp$ scales with the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significant smaller as the equilibrium density $n_0$ what can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
+Dynamics in tokamak plasma is described by spatio-temporal scales which is also referred to as \textit{gyrokinetic ordering}.
+This seperation compares the tuburbulence frequency $\omega$ and the cyclotron frequency $\omega_\mathrm{c}$ ($\omega/\omega_\mathrm{c} \sim 10^{-3})$ due to strong external magnetic fields which causes a fast gyro motion.
+The length scales of the turbulence are associated with the wavevector $\vect{k}$ which can be seperated into a perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field.
+The perpendicular component $k_\perp$ scales with the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the equilibrium density $n_0$, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
 The gradient length compares with the machine size $R$ which leads to the normalized Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak.
 The parallel component $k_\parallel$ on the other hand scales with the machine size $R$.
-Experiments also show that in the core plasma the fluctuation amplitude of the density perturbation $\delta n/n_0$ and magnetic field fluctuation $\delta B/B_0$, with $B_0$ as strength of the magnetic field in equilibrium, is of order $\sim 10^{-4}$.
-Together this seperations result in the following result
+Experiments show that in the core plasma the fluctuation amplitude of the density perturbation $\delta n/n_0$ and magnetic field fluctuation $\delta B/B_0$, with $B_0$ as strength of the magnetic field in equilibrium, is of order $\sim 10^{-4}$.
+Together this seperation result in the following
 \begin{gather}
@@
 
-For the description for charged particle in the tokamak device the \textit{gyrocenter coordinates} will be used [Fig.
+For the description of charged particle behaviour in the tokamak device the \textit{gyrocenter coordinates} are used [Fig.
 \ref{fig:gyrocenterCoords}].
 In this set of coordinates the gyrocenter follows the magnetic field with the parallel velocity $v_{\parallel}$.
-The gyro motion will be describd together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.
+The gyro motion is described together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.
 
@@
 
-In the next step the coordinates of the Vlasov equation (\ref{eq:vlasov}) will be transformed into the gyrocenter coordinates.
-For that the particle postion $\vect{x}$ will be described through the gyrocenter postion $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$ with the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$.
-The particle velocity $\vect{v}$ gets decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{b}$ and
+In the next step the coordinates of the Vlasov equation (\ref{eq:vlasov}) are transformed into the gyrocenter coordinates.
+For that the particle postion $\vect{x}$ is described through the gyrocenter postion $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$ with the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$.
+The particle velocity $\vect{v}$ is decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{b}$ and
 a perpendicular component $v_\perp = |\vect{v} \times \vect{b}|$ which relates to the magnetic moment $\mu$.
 Through the gyrokinetic ordering a phase space transformation can be performed which is based on the Lie perturbation theory, the Hamiltonian or Lagrangian formalism and an asymtotic expansion in the smallness parameter $\epsilon_\mathrm{g}$.
-In the lowest order is the magnetic moment an exact invariant which makes the gyro phase $\zeta$ an ignorable variable.
+In the lowest order is the magnetic moment $\mu$ as an exact invariant which makes the gyro phase $\zeta$ an ignorable variable.
 \cite{Garbet2010,Cary1981,Cary1983} Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transforms the Vlasov equation into the \textit{gyrokinetic equation}
@@
 
-\subsection{$\delta f$ Approximation and Local Limit}
+\subsection{{\boldmath $\delta f$} Approximation and Local Limit}
 \label{sub:application}
 
-All simulation in this thesis are performed with GKW (Gyro Kinetic Workshop) which uses in application of the gyrokinetic equation the $\delta f$ approximation and local limit approximation.
-For the $\delta f$ approximation the particle density distribution function f gets seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}).
+All simulations in this thesis are performed with GKW (Gyro Kinetic Workshop) which applies the gyrokinetic equation the $\delta f$ approximation and local limit approximation.
+For the $\delta f$ approximation the particle density distribution function f is seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}).
 This is given by
@@
 \end{gather}
-The $\delta f$ approximation applies for global and local description where the latter case is used in this thesis.
-The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities can be taken constant over the radial extent of the simulation volume.
+The $\delta f$ approximation applies for global and local description where the latter is used in this thesis.
+The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities can be considered constant over the radial extent of the simulation volume.
 Because of the tokamak symmetry all equilibrium constants are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general.
@@
 
-In addition to that in GKW the following coordinates get applied with the parameter $\vect{X} = \{\xcoord, \ycoord, s\}$.
+In addition to this, in GKW the following coordinates get applied with the parameter $\vect{X} = \{\xcoord, \ycoord, s\}$.
 Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
 Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}.
-This results in the new coordinates $\{\xcoord, \ycoord, s, v_\parallel, \mu\}$.
+This results in the coordinates $\{\xcoord, \ycoord, s, v_\parallel, \mu\}$.
 Combining the $\delta f$ approximation and the Hamada coordinates the gyrokinetic equation can be formualted as follows
@@
 \end{gather}
-where $g$ is a function which contains the the perturbation $\delta f$ and the equilibrium $f_0$.
+where $g$ is a function which contains the perturbation $\delta f$ and the equilibrium $f_0$.
 Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{v}_\chi$ are drift velocities which are caused by the inhomogeneous magnetic field and the perturbed magnetic field ($\vect{v}_\mathrm{D}$) and through the $\exb$ drift ($\vect{v}_\chi$).
-The source term $S$ combines from the distribution function in the equilibrium $f_0$ and a correction term for the collisions as well as the energy injection term.
-Due to prediodic boundary conditions in the radial direction the radially gradients cannot change in time which is also referred as \textit{gradient driven} approach.
+The source term $S$ contains the distribution function in the equilibrium $f_0$ and a correction term for the collisions as well as the energy injection term.
+Due to prediodic boundary conditions in the radial direction the radial averaged gradients are fixed in time which is also referred to as \textit{gradient driven} approach.
 \cite{Peeters2009}

### 2023-06-08 bachelorthesis/32-diagnostics.tex
#39: Correction Cornelia Lippert

@@
 
-In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
+In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transformation of the $\exb$ shearing rate is considered. 
 It is defined by
@@
 where $\hatwexb$ is the complex Fourier coefficient and $\kzf = 2\pi \nzf/L_\xcoord$
-% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
 defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $ and the radial box size $L_\xcoord$.
-% define the zonal flow modes' amplitude in terms of the shearing rate
 Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$. 
-% introduces the term basic mode of the pattern for later usage
 In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.

### 2023-06-08 bachelorthesis/34-restartscript.tex
#39: Correction Cornelia Lippert

@@
 
-As stated in Chapter \ref{sec:cluster} the \texttt{btrzx1}-cluster has a wall time of 24 hours but since simulations typical run longer than 24 hours one have to restart the simulation after the wall time gets exceeded.
-Some simulation need even longer so that the simulation itself needs multiple weeks to finish which makes the restart of the simulation a daily task that could be easily forgotten.
-So to eliminate this minor inconveniences the following restart script were written.
+As stated in Chapter \ref{sec:cluster} the \texttt{btrzx1}-cluster has a wall time of 24 hours but since simulations typically run longer than 24 hours, the simulation has to be restarted after the wall time is exceeded.
+Some simulations need multiple weeks making a regular restart vital to obtaining complete sets of simulation data.
+The following restart script were written to outsource and simplify the process.
 
-As script language \texttt{python} was chosen because of its easiness to write code and the variety of tools which can interact with the \texttt{bash} shell.
-\texttt{bash} was also considered but after some issue occurred regarding the mechanism to check wether a file exist or not, \texttt{python} was selected.
-Furthermore, the script has no requirements because it runs on the standard \texttt{python3} library.
+As script language \texttt{python} was chosen because of its simple syntax and the variety of tools compatible with the \texttt{bash} shell.
+\texttt{bash} was also considered, but after some issues occurred regarding the mechanism to check wether a file exist or not, \texttt{python} was selected.
+Furthermore, the script uses the standard \texttt{python3} library and is therefore executabe with any \texttt{python3} enviroment.
 
-The script was named \texttt{slurm\_monitor.py} to indicate that this script works with the the \texttt{Slurm} resource manager.
+The script was named \texttt{slurm\_monitor.py} to indicate that this script works with the \texttt{Slurm} resource manager.
 The source code was pushed to the gkw repository on BitBucket \cite{slurmmonitor} or can be found in Appendix \ref{append:slurmcode}
@@
           \texttt{-j [JOBNAME], -\/-job-name [JOBNAME]}
-    \item Creates jobscript file with defined content\\
-          (look into variable \texttt{jobscriptContent} to add more option) \\
+    \item Can be started from everywhere with simulation folder path\\
+          \texttt{-d [DIRECTORY], -\/-dir [DIRECTORY]~~~~~~~~~~~~~(default=cwd)}
+    \item Creates jobscript file with defined content (variable \texttt{jobscriptContent})\\
           \texttt{-\/-jobscript [JOBSCRIPTFILE]~~~~~~~(default=jobscript-create)}\\
@@
     \item Makes backup after each run before Restart and Restore files after failed run\\
-          (uses \texttt{rsync} command line utility)\\
-          Quicklinks: local (simulation folder), home (home folder) \\
+          Quicklinks: \texttt{local} (simulation folder), \texttt{home} (home folder) \\
           \texttt{-b [LOCATION], -\/-backup [LOCATION]~~~~~~~~~~~\,(default=None)}
@@
 As stated in Chapter \ref{sub:codeFeatures} the restart script is writing a status file with the current status and progress bars of the simulation.
-The output is formatted like a table and can bei adjust in its appearance with the \texttt{-\/-format} parser option [Chapter \ref{sub:codeFeatures}].
-The output types which are written in the status file are listed in Tab.
+The output is formatted as a table and can be adjusted in its appearance with the \texttt{-\/-format} parser option [Chapter \ref{sub:codeFeatures}].
+The output types written in the status file are listed in Tab.
 \ref{tab:statusTypes}.
-To each output prints the restart script the date, time and the duration of the monitoring additionally into the table.
+To each output the restart script writeesthe date, time and the duration of the monitoring additionally into the table.
 \begin{center}
@@
         Type              & Information                                                                                     \\\hline
-        \texttt{STARTING} & Start of monitoring process or Start/Restart of simulation                                      \\
+        \texttt{STARTING} & Start monitoring process or Start/Restart of simulation                                         \\
         \texttt{CONTINUE} & If the simulation has performed at least one run                                                \\\hline
-        \texttt{CONTROL}  & Check of the current time steps of the simulation                                               \\
-        \texttt{SUCCESS}  & Stop of monitoring because the time steps condition has been fulfilled                          \\\hline
+        \texttt{CONTROL}  & Check the current time steps of the simulation                                                  \\
+        \texttt{SUCCESS}  & Stop monitoring because the time steps condition has been fulfilled                             \\\hline
         \texttt{IMPORT}   & Modules \texttt{h5py}, \texttt{pandas} and \texttt{numpy} will be loaded                        \\
@@
         \texttt{RUNNING}  & Simulation is running on the server                                                             \\\hline
-        \texttt{BACKUP}   & The script performs the backup of the data to the predefined location                           \\
+        \texttt{BACKUP}   & The script performs data backup to the predefined location                                      \\
         \texttt{RESTORE}  & The script reloads the data from the backup folder if errors occur                              \\
-        \texttt{RESET}    & The simulation gets reset with the use of dump files                                            \\\hline 
-        \texttt{ERROR}    & During the monitoring does occur an error                                                       \\
+        \texttt{RESET}    & The simulation restes with the use of dump files                                                \\\hline 
+        \texttt{ERROR}    & An error occur during monitoring                                                                \\
                           & Types:                                                                                          \\
@@
 \end{center}
-In addition to that, it writes down the progress of simulation to the end of the status file in form of a progress bar which gets updated dynamically.
+In addition to that, it dynamically updates the progress of simulation to the end of the status file in form of a progress bar.
 The progress bar contains the current time step of the simulation in contrast to the required time steps and the progress of the run itself in seconds in relation to the wall time.
@@
 \end{center}
-The progress bar has the advantage that someone can easily access all important information at the glance. To boost productivity the following command could be quite handy:
+The progress bar displays all important information at one glance. To boost productivity the following command is recommended:
 \begin{lstlisting}
@@
 \end{lstlisting}
-which prints all progress bars of every simulation in the data folder. If one want to see the complete status file change \texttt{tail -8} to \texttt{cat}.
+which prints all progress bars of every simulation to the data folder. If one want to see the complete status file change \texttt{tail -8} to \texttt{cat}.
 An example status file is displayed in the Appendix \ref{append:status}.

### 2023-06-08 bachelorthesis/41-resolution.tex
#39: Correction Cornelia Lippert

@@
 \citenum{Peeters2016} in Section IV.
-Note that, because of selected circular cencentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref.
+Note that, because of selected circular concentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref.
 \citenum{Peeters2016} for $s$-$\alpha$ geometry.
@@
 
-As next step an approach to the finite heat flux threshold were made to verify the selection of the gradient length $\rlt$.
-As in Ref.
-\citenum{Peeters2016} in Section V conclude is the finite heat flux threshold approximately located at a grendent length of $\rlt = 6.3$ for circular geometry [FIG.
+As next step to verify the selection of the gradient length $\rlt$ an approach to the finite heat flux threshold was made.
+As Ref.
+\citenum{Peeters2016} in Section V concludes the finite heat flux threshold is approximately located at a gradient length of $\rlt = 6.3$ for circular geometry [FIG.
 4 of Ref.
@@
 But resolution $\Nvpar = 32$ is not acceptable as well according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000\,R/\vth$.\\
-So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution will be
+So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution is
 \begin{gather*}
@@
 \end{gather*}
-With this number of grid points the time till turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500\,R/\vth$.
+With this number of grid points the time until turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500\,R/\vth$.
 
@@
 
-As next step the number of grid points for the magnetic moment $\Nmu$ were reduced with
+In the next step the number of grid points for the magnetic moment $\Nmu$ were reduced with
 \begin{gather*}
@@
 
-To conclude a curial result the number of grids point for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}.
-In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with the both criteria in mind, to the following resolution
+To conclude a curial result the number of grid points for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}.
+In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with both criteria in mind, to the following resolution
 \begin{gather*}
@@
 
-In the final step the number of grid points for the magnetic field $\Ns$ get reduced with the following parameters
+In the final step the number of grid points for the magnetic field $\Ns$ gets reduced with the following parameters
 \begin{gather*}
@@
 
-Together with the results of Chapter \ref{sub:reduceNvpar}, \ref{sub:reduceNmu} and \ref{sub:reduceNs} the final resolution used in the upcoming simulations are displayed in Tab.
-\ref{tab:resolution} with reduced number of grid points for the parallel velocity $\Nvpar$.
+Together with the results of Chapter \ref{sub:reduceNvpar}, \ref{sub:reduceNmu} and \ref{sub:reduceNs} the final resolution used in the upcoming simulations is displayed in Tab.
+\ref{tab:resolution}, with reduced number of grid points for the parallel velocity $\Nvpar$.
 \begin{center}

### 2023-06-08 bachelorthesis/42-boxsizescan.tex
#39: Correction by mayself, #37: Whitspace between formula and sign
#39: Correction Cornelia Lippert

@@
 
-In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.27,~89.76)\,\rhoth$ in the radial and binormal direction.
+In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.3,~89.8)\,\rhoth$ in the radial and binormal direction.
 The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
@@
 
-In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size. The scan covers the realizations: 
+In the first test the radial box size is increased while the binormal box size is kept fixed to the standard size. The scan covers the realizations 
 \begin{gather*}
@@
 This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
-% insofar the results of Peeters2016 are well reproduced; now turn to the description of the main result of this test
+The negative gradient of the perturbed perpendicular and parallel ion pressure (not shown) exhibit positive corrugations in regions with maximum $|\wexb|$ and negative corrugations at zero crossings of $\wexb$.
+A radial force balance analysis suggests that the structures in $\wexb$ as depicted in Fig.~\ref{fig:wexb-stable-comparison} are not a consequence of the pressure gradient corrugations as discussed elsewhere \cite{Kosuga2013}.
+Rather, the corrugations in the pressure gradient have to be interpreted as a consequence of the staircase structure in $\wexb$ due to the stabilizing zonal flow - turbulence interaction.
 Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
-% yes, the radial size converges; but even stronger result: the standard box size already allows for the correct staircase size
 Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.
@@
 
-% cautios words
 Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
-% very long initial quasi-stationary states
 Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
-% the \nzf=3 zonal flow mode does show a temporal evolution with a few cycles covered by the initial turbulent phase -> this allows for some degree of statistical argument
 During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
@@
 Hence, several of such cycles are covered by the initial turbulent phase, which is evident from the occurrence of phases with reduced amplitude around $t \approx 8000\,R/\vth$ and $t \approx 18000\,    R/\vth$.
-% first important point: competition between the \nzf = 3 and \nzf = 4 mode -> the \nzf = 4 mode is of shorter length scale compared to the \nzf = 3 mode, hence, no convergence to the box scale
 It is the $\nzf = 4$ zonal flow mode, i.e., the next shorter radial scale mode, that dominates the shear spectrum $\hatwexbamp$ in the latter two phases (not shown).
 This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
-% second important point: no secular growth of the box scale zonal mode
 Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
-% summary
-The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57.20 - 76.27\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
+The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57 - 76\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
 \includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}{
@@
 
-% what kind of analysis
 Since the radially elongated simulation domain might inhibit the development of isotropic turbulent structures, in the second test the radial and binormal box size is increased simultaneously.
-This scan covers the realizations:
+This scan covers the realizations
 \begin{gather*}
-	\NR\times\NB \in [1\times1,~2\times2,~3\times3]~.
+	\NR\times\NB \in [1\times1,~1.5\times1.5,~2\times2,~2.5\times2.5,~3\times3]~.
 \end{gather*}
-% mention shorter times scales for stabilization
-Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern always occurs after $\sim 1000~R/\vth$ [Fig.
+Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern almost always occurs after $\sim 1000~R/\vth$ [Fig.
 \ref{fig:eflux-1x1-2x2-3x3-comparison}\,(a)], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
-% Although the computational costs per time step is higher, the difference in the time scales renders this test computationally cheaper.
-%
-% the main result of this test: confirmation of wave length convergence
 As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
@@
 
-% discuss sumewhat smaller radial size
 By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size [Fig.
@@
 \ref{fig:wexb-stable-comparison}\,(b)].
-Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57.20,~ 76.27]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
+Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57,~ 76]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
+The scale of structures developing in the $1.5\times1.5$ and $2.5\times2.5$ realizations (not included in Fig.~\ref{fig:wexb-stable-comparison} and Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison} to preserve the clarity of this figure) also lie within the range given above.
+Note that two additional realizations of the $3\times3$ case with different initial conditions and otherwise identical parameters confirm structure formation on scales within the range given above.
 
@@
 
-% what kind of analysis
 In a third test the binormal box size is varied with the radial box size fixed to $\NR = 3$.
-This test covers the realizations:
+This test covers the realizations
 \begin{gather*}
@@
 \includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison_thesis.pdf}{
-	Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
+	Comparison of shearing rate $\wexb$ for each box size scan averaged over given a time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
 The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\

### 2023-06-08 bachelorthesis/43-heatfluxthreshold.tex
#39: Correction by mayself, #37: Whitspace between formula and sign
#39: Correction Cornelia Lippert

@@
 Since suppression of turbulence usually occurs at later times when approaching the finite heat flux threshold from below \cite{Peeters2016}, the analysis aims to lengthen the phase during which the zonal flow varies in time due to turbulent Reynolds stresses.
-This scan covers realizations with:
+This scan covers realizations with
 \begin{gather*}
@@
 \end{gather*}
-% what is the finite heat flux threshold
 In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$ [Fig.
 \ref{fig:eflux-6.0-6.2-6.4-comparison}].
-The finite heat flux threshold, hence, is:
+The finite heat flux threshold, hence, is
 \begin{gather*}
@@
 in accordance to Ref. \citenum{Peeters2016}.
-% Which stationay radial size occurs????
 Although the initial quasi-stationary turbulence in the former case is significantly longer compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
-% Does it confirm our previous tests
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.

### 2023-06-08 bachelorthesis/50-conclusion.tex
#39: Correction Cornelia Lippert

--- 2023-05-09 bachelorthesis/50-closure.tex
@@
-% 6. Closure
+% 5. Closure
 
-\chapter{Closure}
+\chapter{Conclusion}
 \label{chap:close}
@@
 \newpage
+
+In this thesis the minimal resolution for simulations with GKW in the Cyclone Base parameter were observed in which the number of grid points for the parallel velocity $\Nvpar$ could be reduced from 64 to 48, which halfed the time until suppression of turbulence.
+\bigskip
+
+Additionally, the active development of a restart script in \texttt{python} led to further convenience during the task of performing simulations on the \texttt{btrzx1} cluster. \bigskip
 

### 2023-06-11 bachelorthesis/62-data.tex
#39, #36: New date in thesis, csv and readme

@@
     \captionsetup{type=table}
-    \begin{tabular}{l | c | l | l l l l l l l}
-            & $\rlt$ & box size  & $\Ns$ & $\Nvpar$ & $\Nmu$ & $d_\mathrm{tim}$ & $k\rho^\mathrm{max}$ & $N_\mathrm{mod}$ & $N_x$ \\ \hline 
-        1   & 6.0    & 1x1       & 12    & 16       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
-        2   & 6.0    & 1x1       & 12    & 32       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
-        3   & 6.0    & 1x1       & 12    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        4   & 6.0    & 1x1       & 12    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        5   & 6.0    & 1x1       & 16    & 16       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        6   & 6.0    & 1x1       & 16    & 32       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        7   & 6.0    & 1x1       & 16    & 48       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
-        8   & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        9   & 6.0    & 1x1       & 16    & 48       & 9      & 0.025            & 1.4                  & 21               & 83    \\ 
-        10  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 0.7                  & 11               & 83    \\ 
-        11  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 43    \\ 
-        12  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 63    \\ 
-        13  & 6.0    & 1x1       & 16    & 64       & 6      & 0.02             & 1.4                  & 21               & 83    \\ 
-        14  & 6.0    & 1x1       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        15  & 6.0    & 2x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        16  & 6.0    & 2x2       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        17  & 6.0    & 3x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        18  & 6.0    & 3x1.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        19  & 6.0    & 3x1.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        20  & 6.0    & 3x2.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        21  & 6.0    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        22  & 6.0    & 3x5       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        23  & 6.0    & 4x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        24  & 6.2    & 2x2       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        25  & 6.2    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        26  & 6.3    & 1x1       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
-        27  & 6.4    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    \\ 
+    \begin{tabular}{l | c | l | l l l l l l l l | l}
+            & $\rlt$ & box size  & $\Ns$ & $\Nvpar$ & $\Nmu$ & $d_\mathrm{tim}$ & $k\rho^\mathrm{max}$ & $N_\mathrm{mod}$ & $N_x$ & ikx & finit   \\ \hline 
+        1   & 6.0    & 1x1       & 12    & 16       & 6      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        2   & 6.0    & 1x1       & 12    & 32       & 6      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        3   & 6.0    & 1x1       & 12    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        4   & 6.0    & 1x1       & 12    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        5   & 6.0    & 1x1       & 16    & 16       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        6   & 6.0    & 1x1       & 16    & 32       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        7   & 6.0    & 1x1       & 16    & 48       & 6      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        8   & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        9   & 6.0    & 1x1       & 16    & 48       & 9      & 0.025            & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        10  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 0.7                  & 11               & 83    & 5   & cosine2 \\ 
+        11  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 43    & 5   & cosine2 \\ 
+        12  & 6.0    & 1x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 63    & 5   & cosine2 \\ 
+        13  & 6.0    & 1x1       & 16    & 64       & 6      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        14  & 6.0    & 1x1       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\
+        15  & 6.0    & 1.5x1.5   & 16    & 48       & 9      & 0.02             & 1.4                  & 31               & 125   & 5   & cosine2 \\
+        16  & 6.0    & 2x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 10  & cosine2 \\ 
+        17  & 6.0    & 2x2       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        18  & 6.0    & 2.5x2.5   & 16    & 48       & 9      & 0.02             & 1.4                  & 51               & 207   & 5   & cosine2 \\ 
+        19  & 6.0    & 3x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 15  & cosine2 \\ 
+        20  & 6.0    & 3x1.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 10  & cosine2 \\ 
+        21  & 6.0    & 3x1.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 10  & cosine2 \\ 
+        22  & 6.0    & 3x2.5     & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 6   & cosine2 \\ 
+        23  & 6.0    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        24  & 6.0    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine5 \\
+        25  & 6.0    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & noise   \\
+        26  & 6.0    & 3x5       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 3   & cosine2 \\ 
+        27  & 6.0    & 4x1       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 20  & cosine2 \\ 
+        28  & 6.2    & 2x2       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        29  & 6.2    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        30  & 6.3    & 1x1       & 16    & 64       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
+        31  & 6.4    & 3x3       & 16    & 48       & 9      & 0.02             & 1.4                  & 21               & 83    & 5   & cosine2 \\ 
     \end{tabular}
@@
         14  & 6.0    & 1x1      & 12000.0 & 20000       & ~            & True     & 1      & True   \\
-        15  & 6.0    & 2x1      & 18000.0 & 30000       & ~            & True     & 2      & True   \\ 
-        16  & 6.0    & 2x2      & 18000.0 & 30000       & ~            & True     & 2      & True   \\ 
-        17  & 6.0    & 3x1      & 48000.0 & 80000       & ~            & True     & 3      & True   \\ 
-        18  & 6.0    & 3x1.5    & 16840.3 & 28068       & ~            & True     & 4      & True   \\ 
-        19  & 6.0    & 3x1.5    & 14170.0 & 23618       & 23618        & True     & 3      & True   \\ 
-        20  & 6.0    & 3x2.5    &  6000.0 & 10000       & ~            & True     & 3, 4   & True   \\ 
-        21  & 6.0    & 3x3      &  7847.0 & 13083       & ~            & True     & 4      & True   \\ 
-        22  & 6.0    & 3x5      &  4769.0 &  7958       & ~            & True     & 4      & True   \\ 
-        23  & 6.0    & 4x1      & 30000.0 & 50000       & 47084        & True     & 4      & True   \\ 
-        24  & 6.2    & 2x2      &  6000.0 & 10000       & ~            & True     & 2      & True   \\ 
-        25  & 6.2    & 3x3      & 14682.7 & 24475       & 14473-16132  & True     & 3      & True   \\ 
-        26  & 6.3    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
-        27  & 6.4    & 3x3      & 12000.0 & 20000       & ~            & False    & 0      & True   \\ 
+        15  & 6.0    & 1.5x1.5  &  6000.0 & 10000       & ~            & True     & 2      & True   \\
+        16  & 6.0    & 2x1      & 18000.0 & 30000       & ~            & True     & 2      & True   \\ 
+        17  & 6.0    & 2x2      & 18000.0 & 30000       & ~            & True     & 2      & True   \\ 
+        18  & 6.0    & 2.5x2.5  &  3402.1 &  5672       & ~            & True     & 3      & True   \\
+        19  & 6.0    & 3x1      & 48000.0 & 80000       & ~            & True     & 3      & True   \\ 
+        20  & 6.0    & 3x1.5    & 16840.3 & 28068       & ~            & True     & 4      & True   \\ 
+        21  & 6.0    & 3x1.5    & 14170.0 & 23618       & 23618        & True     & 3      & True   \\ 
+        22  & 6.0    & 3x2.5    &  6000.0 & 10000       & ~            & True     & 3, 4   & True   \\ 
+        23  & 6.0    & 3x3      &  7847.0 & 13083       & ~            & True     & 4      & True   \\ 
+        24  & 6.0    & 3x3      &  3454.8 &  5758       & 1840-1845    & True     & 4      & True   \\ 
+        25  & 6.0    & 3x3      &  5808.0 &  9680       & 1840-1842    & True     & 3      & True   \\ 
+        26  & 6.0    & 3x5      &  4769.0 &  7958       & ~            & True     & 4      & True   \\ 
+        27  & 6.0    & 4x1      & 30000.0 & 50000       & 47084        & True     & 4      & True   \\ 
+        28  & 6.2    & 2x2      &  6000.0 & 10000       & ~            & True     & 2      & True   \\ 
+        29  & 6.2    & 3x3      & 14682.7 & 24475       & 14473-16132  & True     & 3      & True   \\ 
+        30  & 6.3    & 1x1      &  6000.0 & 10000       & ~            & False    & 0      & True   \\ 
+        31  & 6.4    & 3x3      & 12000.0 & 20000       & ~            & False    & 0      & True   \\ 
     \end{tabular}
@@
         14  & ./data/S6\_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9                    \\
-        15  & ./data/S6\_rlt6.0/boxsize2x1/Ns16/Nvpar48/Nmu9                    \\ 
-        16  & ./data/S6\_rlt6.0/boxsize2x2/Ns16/Nvpar48/Nmu9                    \\ 
-        17  & ./data/S6\_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9                    \\ 
-        18  & ./data/S6\_rlt6.0/boxsize3x1.5/Ns16/Nvpar48/Nmu9                  \\ 
-        19  & ./data/S6\_rlt6.0/boxsize3x1.5/Ns16/Nvpar48/Nmu9/Broken           \\ 
-        20  & ./data/S6\_rlt6.0/boxsize3x2.5/Ns16/Nvpar48/Nmu9                  \\ 
-        21  & ./data/S6\_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
-        22  & ./data/S6\_rlt6.0/boxsize3x5/Ns16/Nvpar48/Nmu9                    \\ 
-        23  & ./data/S6\_rlt6.0/boxsize4x1/Ns16/Nvpar48/Nmu9                    \\ 
-        24  & ./data/S6\_rlt6.2/boxsize2x2/Ns16/Nvpar64/Nmu9                    \\ 
-        25  & ./data/S6\_rlt6.2/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
-        26  & ./data/S6\_rlt6.3/boxsize1x1/Ns16/Nvpar64/Nmu9                    \\ 
-        27  & ./data/S6\_rlt6.4/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
+        15  & ./data/S6\_rlt6.0/boxsize1.5x1.5/Ns16/Nvpar48/Nmu9                \\
+        16  & ./data/S6\_rlt6.0/boxsize2x1/Ns16/Nvpar48/Nmu9                    \\ 
+        17  & ./data/S6\_rlt6.0/boxsize2x2/Ns16/Nvpar48/Nmu9                    \\ 
+        18  & ./data/S6\_rlt6.0/boxsize2.5x2.5/Ns16/Nvpar48/Nmu9                \\
+        19  & ./data/S6\_rlt6.0/boxsize3x1/Ns16/Nvpar48/Nmu9                    \\ 
+        20  & ./data/S6\_rlt6.0/boxsize3x1.5/Ns16/Nvpar48/Nmu9                  \\ 
+        21  & ./data/S6\_rlt6.0/boxsize3x1.5/Ns16/Nvpar48/Nmu9/Broken           \\ 
+        22  & ./data/S6\_rlt6.0/boxsize3x2.5/Ns16/Nvpar48/Nmu9                  \\ 
+        23  & ./data/S6\_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
+        24  & ./data/S6\_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/cosine5            \\
+        25  & ./data/S6\_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/noise              \\
+        26  & ./data/S6\_rlt6.0/boxsize3x5/Ns16/Nvpar48/Nmu9                    \\ 
+        27  & ./data/S6\_rlt6.0/boxsize4x1/Ns16/Nvpar48/Nmu9                    \\ 
+        28  & ./data/S6\_rlt6.2/boxsize2x2/Ns16/Nvpar64/Nmu9                    \\ 
+        29  & ./data/S6\_rlt6.2/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
+        30  & ./data/S6\_rlt6.3/boxsize1x1/Ns16/Nvpar64/Nmu9                    \\ 
+        31  & ./data/S6\_rlt6.4/boxsize3x3/Ns16/Nvpar48/Nmu9                    \\ 
     \end{tabular}

### 2023-06-11 presentation/presentation.tex
#39: Developed own preseantation theme -> fancy

@@
-% Author         : Pedro Ferreira (up201806093@up.pt)
-% Version        : 1.0
-% Created on     : 12.08.2022
-% Last Edited on : 14.08.2022
-
-% Adapted from HSRM theme by Benjamin Weiss
-% Copyright      : Copyright (c) 2013-2014 by Benjamin Weiss. All rights reserved.
-% License        : This file may be distributed and/or modified under the
-%                  GNU Public License.
-% Description    : HSRM beamer theme demonstration. Also includes a short 
-%                  Tutorial regarding the beamer class.
+% Author         : Manuel Lippert
+% Version        : 1.1
+% Created on     : 11.06.2023
+% Last Edited on : 11.06.2023
 
@@
 
-\documentclass[compress,aspectratio=169]{beamer}
+\documentclass[compress,aspectratio=43]{beamer}
 
-\usepackage{sleektheme/beamerthemesleek}
-\usetheme{sleek}
-\usepackage{sleektheme/dtklogos} % must be loaded after theme
+%--------------------------------------------------------------------------
+% Theme
+%--------------------------------------------------------------------------
+
+\usepackage{theme/beamerthemefancy}
+\usetheme{fancy}
+
+\usepackage{theme/dtklogos} % must be loaded after theme
 
@@
 %--------------------------------------------------------------------------
+
 \usepackage[ngerman]{babel}
-
 \usepackage{graphicx}
@@
 \usepackage{tikz}
+\usepackage{booktabs}
+\usepackage{soul}
+
+%--------------------------------------------------------------------------
+% Functions, Graphics, Tikz, Underlines
+%--------------------------------------------------------------------------
+
 \usetikzlibrary{mindmap,backgrounds}
-\usepackage{booktabs}
-\newcommand{\tabitem}{~~\llap{\textbullet}~~}
-
-% Advanced table functions
-\usepackage{tabularx,ragged2e}
-\usepackage{booktabs}
-% Listings extension
-\usepackage{listings}
-\lstset{ %
-language=[LaTeX]TeX,
-basicstyle=\normalsize\ttfamily,
-keywordstyle=,
-numbers=left,
-numberstyle=\tiny\ttfamily,
-stepnumber=1,
-showspaces=false,
-showstringspaces=false,
-showtabs=false,
-breaklines=true,
-frame=tb,
-framerule=0.5pt,
-tabsize=4,
-framexleftmargin=0.5em,
-framexrightmargin=0.5em,
-xleftmargin=0.5em,
-xrightmargin=0.5em
-}
-
 \graphicspath{{../pictures/}}
 
-%--------------------------------------------------------------------------
-% Functions
-%--------------------------------------------------------------------------
+\newcommand{\tabitem}{~~\llap{\textbullet}~~}
+\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
 
-\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
+\setul{0.3ex}{0.2ex}
+
+\makeatletter 
+\let\UL\ul 
+\renewcommand\ul{
+	\let\set@color\beamerorig@set@color 
+	\let\reset@color\beamerorig@reset@color 
+	\UL} 
+\makeatother
 
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_00.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_00.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_01.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_01.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_02.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_02.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_03.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_03.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_04.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_04.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_05.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_05.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_06.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_06.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_07.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_07.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_08.pdf}
-
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_08.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_09.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_09.pdf}
 	\end{frame}
@@
 		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics{Presentation/ITG_Instability_10.pdf}
+		\includegraphics[scale=0.8]{Presentation/ITG_Instability_10.pdf}
 	\end{frame}
@@
 			\begin{itemize}
-				\item ITG Instabilität erzeugt turbulente Wirbelströmung
+				\item ITG \ul{Instabilität} erzeugt turbulente Wirbelströmung
 				\item Wirbelströmung breitet sich über die Breite des Plasmatorus aus
@@
 		\begin{minipage}{0.3\textwidth}
-			\includegraphics[scale = 0.2]{Presentation/Magnetic_Shear_black.png}
+			\includegraphics[scale = 0.15]{Presentation/Magnetic_Shear_black.png}
 		\end{minipage}
@@
 		\begin{tabular}{c c}
-			\includegraphics[scale = 0.2]{Presentation/Stablized_Plasma_Interaction.png} & \includegraphics[scale = 0.2]{Presentation/Stablized_Plasma_LongScale.png} \\
+			\includegraphics[scale = 0.15]{Presentation/Stablized_Plasma_Interaction.png} & \includegraphics[scale = 0.15]{Presentation/Stablized_Plasma_LongScale.png} \\
 			Dominante nichtlineare Interaction  & Zusätzlicher verscherter Zonal Flow \\
@@
 
+	\appendix
+	\section*{Appendix}
 	\begin{frame}
 		\frametitle{Danke für die Aufmerksamkeit!}
-		\includegraphics[scale = 0.35]{Presentation/Fusion_Meme.jpeg} \\
+		\includegraphics[scale = 0.3]{Presentation/Fusion_Meme.jpeg} \\
 	\end{frame}

### 2023-06-11 presentation/theme/blocks.tex
#39: Developed own preseantation theme -> fancy

%---------------------------------------------------------------------
% Blocks
%---------------------------------------------------------------------

\setbeamertemplate{block begin}
{
  \setbeamercolor{item}{parent=block body}
  \par\vskip\medskipamount%
  \begin{beamercolorbox}[sep=.5ex,dp=0.6ex,leftskip=0.5ex,rightskip=0.5ex]{block title}
    \usebeamerfont*{block title}\insertblocktitle%
  \end{beamercolorbox}%
  {\parskip0pt\par}%
  {\nointerlineskip\vskip-0.5pt}%
  \usebeamerfont{block body}%
  \begin{beamercolorbox}[sep=.5ex,dp=0.6ex,leftskip=0.5ex,rightskip=0.5ex,vmode]{block body}%
}
\setbeamertemplate{block end}  
{\end{beamercolorbox}\vskip\smallskipamount}

\setbeamertemplate{block alerted begin}
{
  \setbeamercolor{item}{parent=block body alerted}
  \par\vskip\medskipamount%
  \begin{beamercolorbox}[sep=.5ex,dp=0.6ex,leftskip=0.5ex,rightskip=0.5ex]{block title alerted}
    \usebeamerfont*{block title alerted}\insertblocktitle%
  \end{beamercolorbox}%
  {\parskip0pt\par}%
  {\nointerlineskip\vskip-0.5pt}%
  \usebeamerfont{block body alerted}%
  \begin{beamercolorbox}[sep=.5ex,dp=0.6ex,leftskip=0.5ex,rightskip=0.5ex,vmode]{block body alerted}%
}
\setbeamertemplate{block alerted end}
{\end{beamercolorbox}\vskip\smallskipamount}

\setbeamertemplate{block example begin}
{
  \par\vskip\medskipamount%
  \begin{beamercolorbox}[sep=.5ex,dp=0.6ex,leftskip=0.5ex,rightskip=0.5ex]{block title example}
    \usebeamerfont*{block title example}\insertblocktitle%
  \end{beamercolorbox}%
  {\parskip0pt\par}%
  {\nointerlineskip\vskip-0.5pt}%
  \usebeamerfont{block body example}%
  \begin{beamercolorbox}[sep=.5ex,dp=0.6ex,leftskip=0.5ex,rightskip=0.5ex,vmode]{block body example}%
}
\setbeamertemplate{block example end}
{\end{beamercolorbox}\vskip\smallskipamount}

### 2023-06-16 bachelorthesis/02-acknowledgement.tex
#39: Correction Florian Rath

@@
 
-I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and brief communication, which helps me get better at English spelling. \bigskip
+I thank Anna-Maria Pleyer and my sister Cornelia Lippert for proofreading this thesis and brief communication, which helps me getting better at English spelling. \bigskip
 
 Outside of academia, I would like to extend my gratitude to my parents, my brother, his wife and daughter, my sister and my girlfriend for their encouragement, endurance and support while writing this thesis.
-I am aware that sometimes I am a workaholic who needs to be reminded about the life outside of my academic work.
+% I am aware that sometimes I am a workaholic who needs to be reminded about the life outside of my academic work.

### 2023-06-16 bachelorthesis/21-motion.tex
#39: Correction Florian Rath

@@
 with the center often being referred to as \textit{gyrocenter}.
-Note that since the Lorentz force depends on the species of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip
+Note that since the Lorentz force depends on the species charge of the particle, the circulation direction is the opposite between electron in ions.\\\bigskip
 
@@
 
-In absence of forces in the parallel direction of the magnetic field the particles can move in parallel direction to the homogenous magnetic field.
-The velocity of this motion is connected to the thermal velocity $\vth$ and is dominated by electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).
+In absence of forces in the direction parallel to the magnetic field the particles can move freely in parallel direction to the homogenous magnetic field.
+The velocity of this motion is of order of the thermal velocity $\vth$ and is dominated by electrons due to their lighter mass compared to ions ($v_\mathrm{th,e}/v_\mathrm{th,i} = 60$).
 
@@
 with \textit{magnetic moment} $\mu$.
-The magnetic moment $\mu$ can be an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
+The magnetic moment $\mu$ is an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
 The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force.
@@
     \item {\boldmath $\exb$} \textbf{Drift:}\\
-    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results to a drift of the gyrocenter with
+    If an electric field $\vect{E}$ with a perpendicular component together with the magnetic field $\vect{B}$ (both fields are homogenous) is present the acting Coulomb force and Lorentz force results into a drift of the gyrocenter with
     \begin{gather}
@@
     \end{gather}
-    The $\nabla B$ drift varies thereby on scales which are larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
+    The gradient of the magnetic field $\nabla B$ varies thereby on scales larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
     \item \textbf{Curvature Drift:}\\

### 2023-06-16 bachelorthesis/22-confinement.tex
#39: Correction Florian Rath

@@
 Because of the much smaller size of the Larmor radius compared to the device size $R$ the particle and energy losses are caused by the gyro center drift.
-To avoid this at the end of the field lines the magnetic field in the tokamak devices is shaped like a torus.
-This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux surfaces}, and magnetic field lines which appear on these surfaces.
-To maintain stability the magnetic field has a toroidal and a poloidal component which also is equivalant to the plasma pressure.
+To avoid additional loss of particles because of the parallel motion the field lines of the magnetic field in the tokamak devices is shaped like a torus.
+This type of geometry has nested surfaces with constant magnetic flux, so-called \textit{flux surfaces}, and magnetic field lines which lie on these surfaces.
+To maintain stability the magnetic field has a toroidal and a poloidal component.
+According to the force balance the magnetic field is equivalent to the plasma pressure which means on flux-surfaces the plasma pressure is constant.
 \cite{Stroth2011, Wesson2011} The toroidal component is produced by external coils whereas the poloidal component is provided by the toroidal plasma current.
@@
 with $n$ the plasma density, $T$ as temperature, $\mu_0$ the permeability in vacuum and the magnetic field strength $B$.
-Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$ which indicates that in cases when the plasma beta is smaller the confinement is better.
-For fusion devices the plasma beta has to be smaller than 1 ($\beta < 1$).
+Respectively, the plasma beta compares the thermal plasma pressure $nT$ to the ambient magnetic field pressure $\mu_0 B^2/2$.
+For fusion devices the plasma beta has to be a bit smaller than 1 ($\beta < 1$) for optimal confinement.
 In a tokamak reactor the plasma beta has a typical order of a few percent.

### 2023-06-16 bachelorthesis/23-iontempgrad.tex
#39: Correction Florian Rath

@@
 The described electrostatic forces acting on the tokamak plasma cause a variety of collective phenomena.
-The occurence of Microinstabilities is one of these phenomena.
+The occurrence of microinstabilities is one of these phenomena.
 They are instabilities on the spatial scale of the Larmor radius $\rho_\mathrm{L}$ and frequencies much smaller than the cyclotron frequency $\omega_\mathrm{c}$, driven by density and temperature gradients present in the tokamak devices.
-Turbulence by microinstabilities are considered as the main reason for the loss of particle energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors.
+Turbulence by microinstabilities are considered as one of the main reason for the loss of particle energy and momentum from tokamak plasma \cite{Brizard2007, Garbet2010, Horton1999} which limits the confinement time of tokamak reactors.
 The most dominant instability in tokamak plasma is the ion temperature gradient (ITG) driven instability.
@@
 \citenum{Beer_PHD, Casson_PHD, Dannert_PHD}.
-An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the formation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$, which compresses the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells).
+An ion temperature perturbation ($T^\pm \gtrless 0$ \textcolor{blue}{blue} and \textcolor{red}{red} cells) on the outboard low field side of the tokamak causes the modulation of curvature drift and $\nabla B$ drift with velocity $\vect{v}_\mathrm{d}$.
+The divergence of the velocity $\vect{v}_\mathrm{d}$ is therefore not zero which results in compression of the plasma ($n_\pm \gtrless 0$, \textcolor{gray}{gray} cells).
 This relates to the perturbation of the potential ($\phi^\pm  \gtrless 0$) in the adiabatic electron response.
 As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$.
-The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($n^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($n^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip
+The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($T^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($T^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip
 
@@
 On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}.
-Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of serval Larmor radius.
+Furthermore, such turbulent structures are called \textit{eddies} which have a typical radial scale of several Larmor radii.
 \cite{Newins2006}
 The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$.
-For nonlinear turbulence the critical value is given by $R/L_\mathrm{T,c} \sim 6$.
+For nonlinear turbulence with Cyclone Base Case parameters and circular concentric flux-surfaces the critical value is given by $R/L_\mathrm{T,c} \sim 6$.
 \cite{Dimits2000, Isliker2010}

### 2023-06-16 bachelorthesis/24-zonalflows.tex
#39: Update headings with bold math
#39: Correction Florian Rath

@@
-\section{Zonal Flows and Shearing Rate {\boldmath $\wexb$}}
+\section{Zonal Flows and Shearing Rate $\wexb$}
 \label{sec:zonalflow}
 
-Zonal flows are linked in the tokamak plasma to the $\exb$ flows [Eq.(\ref{eq:exb})] that is connected to the on flux-surfaces constant electrostatic potential, also called \textit{zonal potential}.
+In the tokamak plasma zonal flows are linked to the $\exb$ flows [Eq.(\ref{eq:exb})] that are connected to the electrostatic potential that is constant on flux-surfaces, also called \textit{zonal potential}.
 \cite{Diamond2005} Because of the variation exclusively across flux-surfaces, the zonal flow is tangential to those surfaces and does not contribute to the turbulent transport across the flux-surfaces caused by the $\exb$ motion.
@@
 This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}.
-The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a later development of ITG driven turbulence.
+The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG driven turbulence at higher $\rlt$ compared to the linear threshold.
 The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig.
@@
 Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing.
-\cite{Idomura2009,McMillan2009} The fully developed avalanche results in almost completely quenched turbulence.
-Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient als known as \textit{finite heat flux threshold}.
+\cite{Idomura2009,McMillan2009} The fully developed staircase results in almost completely quenched turbulence.
+Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient also known as \textit{finite heat flux threshold}.
 \cite{Peeters2016,Weikl2017}

### 2023-06-16 bachelorthesis/25-gyrokinetics.tex
#39: Update headings with bold math
#39: Correction Florian Rath

@@
 
-Because of the large number of particles in the fusion plasma a prediction only using Newton-Maxwell dynamics results in an impossible task for simulation based computations, but this problem can be solved with a statistical approach.
+Because of the large number of particles in the fusion plasma a prediction on the basis of Newton-Maxwell dynamics results in an impossible task for simulation, but this problem can be solved with a statistical approach.
 For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles position $\vect{x}$ and velocity $\vect{v}$ is needed.
@@
 
-Dynamics in tokamak plasma is described by spatio-temporal scales which is also referred to as \textit{gyrokinetic ordering}.
-This seperation compares the tuburbulence frequency $\omega$ and the cyclotron frequency $\omega_\mathrm{c}$ ($\omega/\omega_\mathrm{c} \sim 10^{-3})$ due to strong external magnetic fields which causes a fast gyro motion.
-The length scales of the turbulence are associated with the wavevector $\vect{k}$ which can be seperated into a perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field.
-The perpendicular component $k_\perp$ scales with the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the equilibrium density $n_0$, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
+The typical spatio-temporal scales connected to the dynamics in a tokamak plasma allow for the so-called \textit{gyrokinetic ordering} as outlined below.
+The fast gyromotion is much smaller compared to typical  time scales connected to turbulence ($\omega/\omega_\mathrm{c} \sim 10^{-3})$.
+The length scales of the turbulence are associated with the wave vector $\vect{k}$ which can be seperated into a perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field.
+The perpendicular component $k_\perp$ is of the ordre of the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the scale on which the equilibrium density $n_0$ varies, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
 The gradient length compares with the machine size $R$ which leads to the normalized Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak.
@@
 with $\epsilon_\mathrm{g} \ll 1$ which applies to the typical dynamics of a fusion core plasma.
-\cite{Brizard2007,Garbet2010} The gyrokinetic ordering allows the reduction of parameters in the Vlasov equation (\ref{eq:vlasov}).
+\cite{Brizard2007,Garbet2010} The gyrokinetic ordering allows tallows to formulate reduced governing equations referred to as the gyrokinetic formalism.
 
@@
 
-For the description of charged particle behaviour in the tokamak device the \textit{gyrocenter coordinates} are used [Fig.
+For the description of charged particle behaviour in the tokamak device the \textit{guidingcenter coordinates} are used [Fig.
 \ref{fig:gyrocenterCoords}].
-In this set of coordinates the gyrocenter follows the magnetic field with the parallel velocity $v_{\parallel}$.
+In this set of coordinates the guidingcenter follows the magnetic field with the parallel velocity $v_{\parallel}$.
 The gyro motion is described together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.
+The gyrocenter position $\vect{X}$ is expressed with the particle position $\vect{x}$ and the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$ as $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$.
 
@@
 
-In the next step the coordinates of the Vlasov equation (\ref{eq:vlasov}) are transformed into the gyrocenter coordinates.
-For that the particle postion $\vect{x}$ is described through the gyrocenter postion $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$ with the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$.
-The particle velocity $\vect{v}$ is decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{b}$ and
-a perpendicular component $v_\perp = |\vect{v} \times \vect{b}|$ which relates to the magnetic moment $\mu$.
-Through the gyrokinetic ordering a phase space transformation can be performed which is based on the Lie perturbation theory, the Hamiltonian or Lagrangian formalism and an asymtotic expansion in the smallness parameter $\epsilon_\mathrm{g}$.
-In the lowest order is the magnetic moment $\mu$ as an exact invariant which makes the gyro phase $\zeta$ an ignorable variable.
-\cite{Garbet2010,Cary1981,Cary1983} Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transforms the Vlasov equation into the \textit{gyrokinetic equation}
+In the next step the unperturbed Lagrangian is expressed through the guidingcenter coordinates.
+The particle velocity $\vect{v}$ is decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{b}$ and a perpendicular component $v_\perp = |\vect{v} \times \vect{b}|$ which relates to the magnetic moment $\mu$.
+As final step the Lagrangian is transformed into gyrocenter phase space, based on the so-called Lie perturbation theory.
+This method allows to eliminate the gyrophase-dependent contributions, rendering the gyrophase $\zeta$ an ignorable coordinate and the magnetic moment $\mu$ an exact constant of motion.
+\cite{Garbet2010,Cary1981,Cary1983}
+Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transforms the Vlasov equation into the \textit{gyrokinetic equation}, describing the evolution of the gyrocenter distribution function.
 \begin{gather}
@@
 
-\subsection{{\boldmath $\delta f$} Approximation and Local Limit}
+\subsection{$\delta f$ Approximation and Local Limit}
 \label{sub:application}
 
-All simulations in this thesis are performed with GKW (Gyro Kinetic Workshop) which applies the gyrokinetic equation the $\delta f$ approximation and local limit approximation.
+All simulations in this thesis are performed with GKW (Gyro Kinetic Workshop) which solves the gyrokinetic equationin the $\delta f$ approximation and local limit.
 For the $\delta f$ approximation the particle density distribution function f is seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}).
@@
 \end{gather}
-The distribution function in the equilibrium $f_0$ is set in GKW to the Maxwell distribution function.
-The perturbed distribution function $\delta f$ is considered to vary perpendicular to the poloidal magnetic field in order of the Larmor radius while the perturbation changes with the system size which results in the ordering
+In GKW the equilibrium distribution function $f_0$ is set to the Maxwell distribution function.
+The perturbed distribution function $\delta f$ is considered to vary perpendicular to the poloidal magnetic field of order of the Larmor radius while the equilibrium changes with the system size which results in the ordering
 \begin{gather}
@@
 The $\delta f$ approximation applies for global and local description where the latter is used in this thesis.
-The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities can be considered constant over the radial extent of the simulation volume.
-Because of the tokamak symmetry all equilibrium constants are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general.
+The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities are considered constant over the radial extent of the simulation volume.
+Because of the tokamak symmetry all equilibrium quantities are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general.
 \\\bigskip
@@
 This results in the coordinates $\{\xcoord, \ycoord, s, v_\parallel, \mu\}$.
-Combining the $\delta f$ approximation and the Hamada coordinates the gyrokinetic equation can be formualted as follows
+Combining the $\delta f$ approximation and the Hamada coordinates the gyrokinetic equation can be formulated as follows
 \begin{gather}
@@
 where $g$ is a function which contains the perturbation $\delta f$ and the equilibrium $f_0$.
-Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{v}_\chi$ are drift velocities which are caused by the inhomogeneous magnetic field and the perturbed magnetic field ($\vect{v}_\mathrm{D}$) and through the $\exb$ drift ($\vect{v}_\chi$).
+Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{v}_\chi$ are drift velocities which are caused by the inhomogeneous magnetic field ($\vect{v}_\mathrm{D}$) and through the $\exb$ drift ($\vect{v}_\chi$).
 The source term $S$ contains the distribution function in the equilibrium $f_0$ and a correction term for the collisions as well as the energy injection term.

### 2023-06-16 bachelorthesis/33-cluster.tex
#39: Correction Florian Rath

@@
 
-The simulation itself will be performed on the \texttt{btrzx1}-cluster of the University Bayreuth. This cluster has a wall time from 24 hours, an total amount of 345 compute nodes with
+The simulation itself will be performed on the \texttt{btrzx1}-cluster of the University Bayreuth. This cluster has a wall time of 24 hours, an total amount of 345 compute nodes with
 two AMD Epyc Processors (2nd generation) with 16 cores each and 128 GB of main memory each node and as resource manager \texttt{Slurm}. \cite{btrzx1}

### 2023-06-16 bachelorthesis/34-restartscript.tex
#39: Correction Florian Rath

@@
 Some simulations need multiple weeks making a regular restart vital to obtaining complete sets of simulation data.
-The following restart script were written to outsource and simplify the process.
+The following restart script was written to outsource and simplify the process.
 
@@
 \end{lstlisting}
-which prints all progress bars of every simulation to the data folder. If one want to see the complete status file change \texttt{tail -8} to \texttt{cat}.
+which prints all progress bars of every simulation to the data folder. If one wants to see the complete status file change \texttt{tail -8} to \texttt{cat}.
 An example status file is displayed in the Appendix \ref{append:status}.

### 2023-06-16 bachelorthesis/41-resolution.tex
#39: Correction Florian Rath

@@
 
-At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without fearing numerical dissipation.
-Numerical dissipatation can therefore result to no formation of zonal flow stuctures, which cause an permanent turbulent state of the simulation.
+At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without numerical dissipation biasing the physical outcome.
+Numerical dissipation can therefore result to no formation of zonal flow structures, which cause an permanent turbulent state of the simulation.
 The goal behind this testing is to reduce \textbf{calculation time} and \textbf{costs} of the simulation.
@@
 In Fig.
-\ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is simular to the results from Ref.
+\ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is similar to the results from Ref.
 \citenum{Peeters2016} with subdued turbulence after $\sim 3000\,R/\vth$.
@@
 \citenum{Peeters2016}].
-Therefore following parameters were used
+Therefore, following parameters were used
 \begin{gather*}
@@
 \end{gather*}
-As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occur in the whole time domain, which is in agreement with Ref. \citenum{Peeters2016}.
+As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occurs in the whole time domain, which is in agreement with Ref. \citenum{Peeters2016}.
 
@@
 \ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visible that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period.
-But resolution $\Nvpar = 32$ is not acceptable as well according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000\,R/\vth$.\\
-So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution is
+But resolution $\Nvpar = 32$ is not acceptable as well according to criteria \textbf{(2)} since the suppressed turbulence state gain instability after $\sim 3000\,R/\vth$.\\
+So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set of criteria. Due to criteria \textbf{(1)} the selected resolution is
 \begin{gather*}
@@
 
-To conclude a curial result the number of grid points for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}.
+To conclude a final result the number of grid points for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}.
 In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with both criteria in mind, to the following resolution
@@
 
-\subsection{Reduction of Magnetic Field Grid Points $\Ns$}
+\subsection{Reduction Grid Points along of the Magnetic Field $\Ns$}
 \label{sub:reduceNs}
 
-In the final step the number of grid points for the magnetic field $\Ns$ gets reduced with the following parameters
+In the final step the number of grid points along the magnetic field $\Ns$ gets reduced with the following parameters
 \begin{gather*}
@@
 	\captionof{table}{
-		Resolution used in this paper: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+		Resolution used in this work: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
 	}

### 2023-06-20 briefcommunication/briefcommunication.tex
#37: Correction Journal

@@
 \begin{abstract}
-    The radial size convergence of the $\exb$ staircase pattern is adressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-    Its is shown that a mesoscale pattern size of \linebreak $\sim 57-76\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
+    The radial size convergence of the $\exb$ staircase pattern is addressed in local gradient-driven flux tube simulations of ion temperature gradient (ITG)-driven turbulence.
+    Its is shown that a mesoscale pattern size of \linebreak $\sim 57-76\,\rhoth$ is inherent to ITG-driven turbulence with Cyclone Base Case parameters in the local limit.
 
@@
 
-Ion temperature gradient driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
+Ion temperature gradient-driven turbulence close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures \cite{Pradalier2010}.
 Such pattern formation has been observed in local gradient-driven flux-tube simulations \cite{Peeters2016, Rath2021}, including collisions \cite{Weikl2017} and background $\exb$ shear \cite{Rath2021}, local flux-driven realizations including mean electric field shear \cite{Seiferling2019}, as well as global gradient-driven \cite{McMillan2009, Villard2013, Seo2022} and global flux-driven \cite{Pradalier2010, Pradalier2015, Wang2020, Kim2022, Kishimoto2023} studies.
@@
 It has been carefully verified that this modification preserves the same physical outcome as the original study.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Ref.~\onlinecite{Peeters2009, Peeters2016}.
+A summary of the numerical parameters is given in Table~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Ref.~\onlinecite{Peeters2009, Peeters2016}.
 \begin{table}[ht]
@@
 In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76.3,~89.8)\,\rhoth$ in the radial and binormal direction.
-Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
+Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$ and $\ycoord$ labels the field lines and is an approximate binormal coordinate.
 Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}.
 The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
-Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
+Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Table~\ref{tab:resolution}] and standard box size.
 \\
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Rath2016, Peeters2016}
+The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined as\cite{Rath2016, Pueschel2008, Rath2016, Peeters2016}
 \begin{equation}
@@
 \end{equation}
-where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
-The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
+where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature and $e$ is the elementary charge).
+The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to Ref.~\onlinecite{Rath2021}
 \begin{equation}
@@
 \end{equation}
-The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}.
+The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}.
 \\
@@
 The scan covers the realizations $\NR\times\NB \in [ 1\times1,~2\times1,~3\times1,~4\times1]$.
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(a)].
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} [$\gamma$ is the growth rate of the most unstable linear ion temperature gradient (ITG)-driven Eigenmode], connected by steep flanks where $\wexb$ crosses zero.
 The negative gradient of the perturbed perpendicular and parallel ion pressure (not shown) exhibit positive corrugations in regions with maximum $|\wexb|$ and negative corrugations at zero crossings of $\wexb$.
@@
 Rather, the corrugations in the pressure gradient have to be interpreted as a consequence of the staircase structure in $\wexb$ due to the stabilizing zonal flow - turbulence interaction.
-Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.~\onlinecite{Peeters2016}.
-Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
+Figure~\ref{fig:wexb-stable-comparison}(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
+Hence, the basic size of the pattern not only converges with increasing radial box size, but the converged radial size also turns out to at least roughly agree with the standard radial box size of Ref.~\onlinecite{Peeters2016}.
+Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
 Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
@@
 This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
-Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
+Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b) dotted line].
 The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57 - 76\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
@@
 \ref{fig:eflux-1x1-2x2-3x3-comparison}], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
-As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+As shown in Fig.~\ref{fig:wexb-stable-comparison}(b) this test also confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 
@@
 	\captionof{figure}{
-		\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
-		\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
+		\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes,\\
+		\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes.
 	}
@@
 	\captionof{figure}{
-		Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
-The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
+		Comparison of shearing rate $\wexb$ for each box sizes scan averaged over given time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG-driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility:\\
 		\begin{tabular}{l l l l l l l l l}
-			\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$, & $t_{2\times 1}$   & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$ & $\in [26000, 28000]$ \\
-			\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$, & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$    &                 &                      \\
-			\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$, & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$ & $\in [1000, 3000]$   \\
+			\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$, & $t_{2\times 1}$   & $\in [15000, 18000]$, & $t_{3\times 1}$ & $\in [43000, 45000]$, & $t_{4\times 1}$ & $\in [26000, 28000]$;   \\
+			\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$, & $t_{2\times 2}$   & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$;   &                 &                         \\
+			\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$, & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   & $t_{3\times 3}$ & $\in [2000, 3000]$,   & $t_{3\times 5}$ & $\in [1000, 3000]$.
+\\
 		\end{tabular}
@@
 	\includegraphics[width=0.87\linewidth]{Briefcommunication/3.pdf}
-	\captionof{figure}{Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes}
+	\captionof{figure}{Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes.}
 	\label{fig:eflux-1x1-2x2-3x3-comparison}
@@
 As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}].
-The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
-Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57,~ 76]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
-The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
-The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no signifciant plateau at positive shear.
+The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}(c) and confirms again a size of a typical mesoscale.
+Fig.~\ref{fig:wexb-stable-comparison}(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57,~ 76]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}(c) in the $3\times 2.5$ realization.
+The staircase structure has a pattern between three and four repetitions which get represented in the second repetition with no significant plateau at positive shear.
 Instead, the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
@@
 	\includegraphics[width=0.87\linewidth]{Briefcommunication/4.pdf}
-	\captionof{figure}{Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes}
+	\captionof{figure}{Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes.}
 	\label{fig:eflux-3x1.5-2.5-3-5-comparison}
@@
 In the case of $\rlt = 6.2$ turbulence suppression is observed for $t > 11000\,R/\vth$, while stationary turbulence during the entire simulation time trace of $12000\,R/\vth$ is found for $\rlt = 6.4$.
-The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$ in accordance to Ref. \onlinecite{Peeters2016}.
+The finite heat flux threshold, hence, is $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$ in accordance with Ref. \onlinecite{Peeters2016}.
 Although the initial quasi-stationary turbulence in the former case is significantly longer compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
@@
 
-Through careful tests this brief communication confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-A mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
+Through careful tests this brief communication confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG)-driven turbulence.
+A mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG-driven turbulence for Cyclone Base Case parameters.
 This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{Pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
 The occurrence of this mesoscale implies that non-locality, in terms of Ref.
-\onlinecite{Pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{McMillan2009, Pradalier2010, Rath2016, Peeters2016}.
+\onlinecite{Pradalier2010}, is inherent to ITG-driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{McMillan2009, Pradalier2010, Rath2016, Peeters2016}.
 

### 2023-06-20 pictures/Theory/Electric-Force.tex
#39:Additional Plots with tikz and cleand structure

% \usetikzlibrary{mindmap,backgrounds}
% \usetikzlibrary{decorations.pathmorphing}
% \usetikzlibrary{decorations.markings}
% \usetikzlibrary{arrows.meta,bending}

\begin{tikzpicture}
    % Charge
    \draw [fill=blue, blue] (0,1.5) circle (0.4) node[white, font=\boldmath]{$-$};
    \draw [fill=red, red] (0,-1.5) circle (0.4) node[white, font=\boldmath]{$+$};
    \draw [fill=blue, blue] (-0.5,0) circle (0.2) node[white, font=\boldmath]{$-$};
    \draw [fill=red, red] (0.5,0) circle (0.2) node[white, font=\boldmath]{$+$};
    % Electric field
    \draw[thick,Notablue, ->] (0,-1) -- (0,1) node[below left, Notablue] {$\vect{E}$};
    % Force
    \draw[thick, Notagrey, ->] (-0.5, -0.2) -- (-0.5, -1.2);
    \draw (-0.5, -1.2) node[left, Notagrey]{$\vect{F}^{-}_{\parallel,E}$};
    \draw[thick, Notagrey, ->] (0.5, 0.2) -- (0.5, 1.2);
    \draw (0.5, 1.2) node[right, Notagrey]{$\vect{F}^{+}_{\parallel,E}$};
\end{tikzpicture}

### 2023-06-20 pictures/Theory/ExB-Drift.tex
#39:Additional Plots with tikz and cleand structure

% \usetikzlibrary{mindmap,backgrounds}
% \usetikzlibrary{decorations.pathmorphing}
% \usetikzlibrary{decorations.markings}
% \usetikzlibrary{arrows.meta,bending}

\begin{tikzpicture}
    % Electron
    \draw[thick,decoration={segment length=2mm, amplitude=0.3cm, coil},decorate,<-,blue] (3,0) -- (0,0);
    \draw [fill=blue, blue] (0,0.2) circle (0.2) node[white, font=\boldmath]{$-$};
    % Ion
    \draw[thick,decoration={coil, aspect = -0.5, amplitude = -0.9cm, segment length=6mm},decorate,<-,red] (3,-2) -- (0,-2);
    \draw [fill=red, red] (0,-2.2) circle (0.2) node[white, font=\boldmath]{$+$};
    % Electric field
    \draw[thick,Notablue, ->] (4,-2) -- (4,0) node[above right, Notablue]{$\vect{E}$};
    % Drift velocity
    \draw[thick,Notaorange, ->] (0.5,-0.7) -- (2.5,-0.7) node[right, Notaorange]{$\vect{v}_{E}$};
    % Magnetic field
    \draw [thick, Notagreen] (4,-2.5) circle (0.2);
    \filldraw [Notagreen] (4,-2.5) circle (1.5pt);
    \draw (4.3,-2.2) node[Notagreen]{$\vect{B}$};
\end{tikzpicture}

### 2023-06-20 pictures/Theory/GradB-Drift.tex
#39:Additional Plots with tikz and cleand structure

--- 2023-06-20 pictures/Theory/ExB-Drift.tex
@@
     % Ion
-    \draw[thick,decoration={coil, aspect = -0.5, amplitude = -0.9cm, segment length=6mm},decorate,<-,red] (3,-2) -- (0,-2);
-    \draw [fill=red, red] (0,-2.2) circle (0.2) node[white, font=\boldmath]{$+$};
-    % Electric field
-    \draw[thick,Notablue, ->] (4,-2) -- (4,0) node[above right, Notablue]{$\vect{E}$};
+    \draw[thick,decoration={coil, aspect = -0.5, amplitude = -0.9cm, segment length=6mm},decorate,<-,red] (0,-2) -- (3,-2);
+    \draw [fill=red, red] (3,-1.8) circle (0.2) node[white, font=\boldmath]{$+$};
+    % Grad B field
+    \draw[thick,Notagreen, ->] (4,-2) -- (4,0) node[above right, Notagreen]{$\nabla B$};
     % Drift velocity
-    \draw[thick,Notaorange, ->] (0.5,-0.7) -- (2.5,-0.7) node[right, Notaorange]{$\vect{v}_{E}$};
+    \draw[thick,Notaorange, ->] (0.5,-0.6) -- (2.5,-0.6) node[right, Notaorange]{$\vect{v}^{-}_{\nabla B;\kappa}$};
+    \draw[thick,Notaorange, <-] (0.5,-0.8) node[left, Notaorange]{$\vect{v}^{+}_{\nabla B;\kappa}$} -- (2.5,-0.8);
     % Magnetic field

### 2023-06-20 pictures/Theory/ITG-Turbulence.tex
#39:Additional Plots with tikz and cleand structure

% \usetikzlibrary{mindmap,backgrounds}
% \usetikzlibrary{decorations.pathmorphing}
% \usetikzlibrary{decorations.markings}
% \usetikzlibrary{arrows.meta,bending}

\begin{tikzpicture}
    % Temperature background
    \shade [left color=red, right color= blue, rounded corners = 0.5cm, opacity = 0.7] (0,0) rectangle (12,6);
    % Magnetic field
    \draw [thick, white] (8,0.8) circle (0.2);
    \filldraw [white] (8,0.8) circle (1.5pt);
    \draw (8.3,1.1) node[white]{$\vect{B}$};
    % Density and Potencial cells
    \draw [thick, gray, opacity=1, fill = gray] (4.5,4.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
    \draw [thick, gray, opacity=1, fill = gray] (4.5,3.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
    \draw [thick, gray, opacity=1, fill = gray] (4.5,2.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
    \draw [thick, gray, opacity=1, fill = gray] (4.5,1.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
    % Temperature cells
    \draw [thick, red, opacity=1, fill = red]   (7.5,4) ellipse (2.5 and 0.4) node[white, opacity=1]{$T^{+}$};
    \draw [thick, blue, opacity=1, fill = blue] (7.5,3) ellipse (2.5 and 0.4) node[white, opacity=1]{$T^{-}$};
    \draw [thick, red, opacity=1, fill = red]   (7.5,2) ellipse (2.5 and 0.4) node[white, opacity=1]{$T^{+}$};
    % Temperature and density gradient
    \draw[thick, white, <-] (3,0.8) -- (6,0.8) node[midway, below, white]{$\nabla n~;\nabla T$};
    % Outline Density and Potencial cell
    \draw [thick, lightgray, dashed] (4.5,4.5) ellipse (2.5 and 0.4);
    \draw [thick, lightgray, dashed] (4.5,3.5) ellipse (2.5 and 0.4);
    \draw [thick, lightgray, dashed] (4.5,2.5) ellipse (2.5 and 0.4);
    \draw [thick, lightgray, dashed] (4.5,1.5) ellipse (2.5 and 0.4);
    % Grad B and curvature drift velocity
    \draw[thick, white, ->] (9.5,4.2) -- (9.5,3.5) node[midway, left, white]{$\vect{v}_\mathrm{d}$};
    \draw[thick, white, ->] (9.5,3) -- (9.5,2.6) node[midway, left, white]{$\vect{v}_\mathrm{d}$};
    \draw[thick, white, ->] (9.5,2.2) -- (9.5,1.5) node[midway, left, white]{$\vect{v}_\mathrm{d}$};
    % ExB drift velocity
    \draw[thick, white, <-] (10.5,5) -- (11.5,5) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (10.5,4) -- (11.5,4) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (10.5,3) -- (11.5,3) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (10.5,2) -- (11.5,2) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (10.5,1) -- (11.5,1) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (0.5,5) --   (1.5,5) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (0.5,4) --   (1.5,4) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (0.5,3) --   (1.5,3) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (0.5,2) --   (1.5,2) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (0.5,1) --   (1.5,1) node[midway, below, white]{$\vect{v}_{E}$};
\end{tikzpicture}

### 2023-06-20 pictures/Theory/Inhomogenous-Magnetic-Field.tex
#39:Additional Plots with tikz and cleand structure

% \usetikzlibrary{mindmap,backgrounds}
% \usetikzlibrary{decorations.pathmorphing}
% \usetikzlibrary{decorations.markings}
% \usetikzlibrary{arrows.meta,bending}

{
\def\ang{36}
\def\Rx{0.7}
\def\Ry{1.14}
\def\h{0.5}
\def\H{3}
\def\W{4}
\def\NB{2}

\begin{tikzpicture}

	\coordinate (O) at (0,0);
	\coordinate (N) at (0,0.24*\H);
	\coordinate (M) at (0,0.45*\H);
	\coordinate (B) at (\ang:\H);

	% Magnetic momentum (1/2)
	\draw [thick, Notared] (0,\Ry) arc (90:270:{\Rx} and {\Ry});
	% Magnetic field
	\foreach \i [evaluate={\y=(0.34*\H)*\i^2/\NB; \out=1*\i^2; \in=180+10*\i^2; \f=0.80-0.10*\i;}] in {1,...,\NB}{
	  \draw [thick, Notagreen, <-] (-0.95*\H, 0.7*\y/\H) to[out= \out,in= \in,looseness=0.8] (0.25*\W, \y);
	  \draw [thick, Notagreen, <-] (-0.95*\H,-0.7*\y/\H) to[out=-\out,in=-\in,looseness=0.8] (0.25*\W,-\y);
	}
	\node [Notagreen] at (0.26*\W,0.52*\H) {$\vect{B}$};
	% Force
	\draw [thick, Notagrey, ->] ( 95:{\Rx} and {\Ry}) --++ (-25:0.8*\Ry) node[right=0, Notagrey] {$\vect{F}_{\parallel,\nabla_{\!\parallel} B}$};
	\draw [thick, Notagrey, ->] (-95:{\Rx} and {\Ry}) --++ ( 25:0.8*\Ry) node[right=1, Notagrey] {$\vect{F}_{\parallel,\nabla_{\!\parallel} B}$};
	% Magnetic momentum (2/2)
	\draw [thick, Notared] (0,\Ry) arc (90:-90:{\Rx} and {\Ry});
	\draw [thick, Notared, ->] (0,0) --++ (-0.3*\W,0) node[left, Notared] {$\vect{\mu}$};
	% Particle
	\filldraw [Notablue] (-95:{\Rx} and {\Ry}) circle (2pt);
	\filldraw [Notablue] ( 95:{\Rx} and {\Ry}) circle (2pt);
	\draw (-3,1) node[above right, Notablue]{Particle};

\end{tikzpicture}}

### 2023-06-20 pictures/Theory/Lorentz-Force.tex
#39:Additional Plots with tikz and cleand structure

% \usetikzlibrary{mindmap,backgrounds}
% \usetikzlibrary{decorations.pathmorphing}
% \usetikzlibrary{decorations.markings}
% \usetikzlibrary{arrows.meta,bending}

\begin{tikzpicture}
    % Circle
    \draw (0,0) circle (1.5);
    \filldraw (0,0) circle (1.5pt);
    % Force
    \draw[thick, Notagrey, <-] (90:0.5) node[right, Notagrey]{$\vect{F}_{\perp, B}$} -- ++ (90:1.0) node[coordinate] (x) {};
    % Larmor radius
    \draw[thick, black, ->] (0,0) -- ++(0:1.5);
    \draw (0.6,-0.4) node {$\rho_\mathrm{L} = \frac{mv_{\perp}}{|q|B}$};
    % Velocity
    \draw[thick,Notaorange, ->] (x) -- ++(180:1cm) node[above, Notaorange]{$\vect{v}$};
    % Cyclontron freqency
    \draw[thick, ->, Notared] (200:1.1*1.5) arc(200:240:1.1*1.5);
    \draw (240:1.4*1.5) node[Notared] {$\omega_\mathrm{c} = \frac{|q|B}{m}$};
    % Particle
    \filldraw [Notablue] (x) circle (2pt);
    \draw (x) node[above right, Notablue]{Particle};
    % Magnetic field
    \draw [thick, Notagreen] (1.5,-1.5) circle (0.2);
    \filldraw [Notagreen] (1.5,-1.5) circle (1.5pt);
    \draw (1.8,-1.2) node[Notagreen]{$\vect{B}$};
\end{tikzpicture}

### 2023-06-20 bachelorthesis/03-abstract.tex
#39: Bachelor thesis finished

@@
 
-Ion temperature gradient driven turbulence (ITG) close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures.
+Ion temperature gradient-driven turbulence (ITG) close to marginal stability exhibits zonal flow pattern formation on mesoscales, so-called $\exb$ staircase structures.
 Such pattern formation has been observed in local gradient-driven flux-tube simulations as well as global gradient-driven and global flux-driven studies.
@@
 
-Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57-76\,\rhoth$ is inherent to ITG driven turbulence with Cyclone Base Case parameters in the local limit.
-This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG driven turbulence.
+Furthermore, it is shown by multiple box size convergence scans that a mesoscale pattern  size of $\sim 57-76\,\rhoth$ is inherent to ITG-driven turbulence with Cyclone Base Case parameters in the local limit.
+This outcome also implies that a typical scale for avalanche-like transport is inherent to ITG-driven turbulence.
 

### 2023-06-20 bachelorthesis/21-motion.tex
#39: Bachelor thesis finished

@@
 
-Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field.
+Due to the Lorentz force, particles with a velocity component perpendicular to the homogenous magnetic field $v_{\perp}$ undergo a circular motion in the plane perpendicular to the magnetic field [Fig.
+\ref{fig:perp-par-motion}(a)].
 This type of motion has circular frequency, which is often referred to as \textit{cyclotron frequency} and is defined as
@@
 \end{gather}
-The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type.\\\bigskip
+The parallel motion follows then from the equation of motion. Here the direction of the motion also depends on the species type [Fig. \ref{fig:perp-par-motion}(b)].\\\bigskip
 
@@
 \begin{gather}
-    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
+    F_{\parallel,\nabla_{\!\parallel} B} = - \frac{mv^2_{\perp}}{2B} \nabla_{\!\parallel} B = - \mu \nabla_{\!\parallel} B~;~~~~\mu = \frac{mv^2_{\perp}}{2B}
 \end{gather}
 with \textit{magnetic moment} $\mu$.
-The magnetic moment $\mu$ is an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron freqency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
-The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force.
+The magnetic moment $\mu$ is an adabatic invariant (constant of motion) if the variation of the magnetic field over time is smaller than the inverse of the cyclotron frequency $\omega^{-1}_\mathrm{c}$ and the spatial variation is larger the Larmor radius $\rho_\mathrm{L}$.
+The resulting force has its application in the mirror effect where a charged particle gets reflected due to this force [Fig.
+\ref{fig:perp-par-motion}(c)].
 \cite{Wesson2011}
+
+\newpage
+
+\begin{center}
+    \captionsetup{type=figure}
+    \begin{tabular}{c c c c c c}
+        \input{../pictures/Theory/Lorentz-Force_Thesis.tex} &
+        \input{../pictures/Theory/Electric-Force.tex} &
+        \input{../pictures/Theory/Inhomogenous-Magnetic-Field.tex}\\[0.3cm]
+        \textbf{(a)} & \textbf{(b)} & \textbf{(c)} \\
+    \end{tabular}
+    \captionof{figure}{Forces acting on a charged particle:\\
+        \begin{tabular}{l l}
+            \textbf{(a)} & Lorentz force $\vect{F}_{\perp, B}$ perpendicular to velocity $\vect{v}$ which causes,\\
+                         & circular motion with Lamor radius $\rho_\mathrm{L}$ and cyclotron frequency $\omega_\mathrm{c}$ \\
+            \textbf{(b)} & Electric force $\vect{F}_{\parallel,E}^\pm$,\\
+            \textbf{(c)} & Mirror effect with force $\vect{F}_{\parallel,\nabla_{\!\parallel} B}$ and magnetic moment $\mu$ \\
+                         & caused by an inhomogeneous magnetic field $\vect{B}$.  \\
+        \end{tabular}
+    }
+    \label{fig:perp-par-motion}
+\end{center}
 
@@
 In the presence of a magnetic field (homogenous, inhomogeneous or perturbed) and electric fields the gyrocenter undergoes slow (compared to the thermal velocity $\vth$) drift motions perpendicular to the magnetic field.
-There are serval examples for this drift motion.
+There are several examples for this drift motion.
 According to this thesis topic only the main three drift types will be covered in the following.
@@
     \end{gather}
-    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same.
+    which is called the $\exb$ drift. Since both acting forces direction depends on the species type the direction of the $\exb$ drift is for every species the same [Fig. \ref{fig:drift}(a)].
     \item {\boldmath $\nabla B$} \textbf{Drift:}\\
@@
     \end{gather}
-    The gradient of the magnetic field $\nabla B$ varies thereby on scales larger compared to the Larmor radius. The direction of the $\nabla B$ drift depends on the species type.
+    The gradient of the magnetic field $\nabla B$ varies thereby on scales larger compared to the Larmor radius.
+The direction of the $\nabla B$ drift depends on the species type [Fig.
+\ref{fig:drift}(b)].
     \item \textbf{Curvature Drift:}\\
@@
 \end{enumerate}
+
+\newpage
+
+\begin{center}
+    \captionsetup{type=figure}
+    \begin{tabular}{c c}
+        \input{../pictures/Theory/ExB-Drift.tex} &
+        \input{../pictures/Theory/GradB-Drift.tex} \\[0.3cm]
+        \textbf{(a)} & \textbf{(b)} \\
+    \end{tabular}
+    \captionof{figure}{Dirft motion in gyrocenter:\\
+        \begin{tabular}{l l}
+            \textbf{(a)} & $\exb$ Drift with drift velocity $\vect{v}_E$, electric field $\vect{E}$ and\\
+                         & magnetic field $\vect{B}$,\\
+            \textbf{(b)} & $\nabla B$ Drift/Curvature Drift with drift velocity $\vect{v}^{\pm}_{\nabla B;\kappa}$, magnetic\\
+                         & field $\vect{B}$ and gradient of the magnetic field $\nabla B$. \\
+        \end{tabular}
+    }
+    \label{fig:drift}
+\end{center}

### 2023-06-20 bachelorthesis/23-iontempgrad.tex
#39: Bachelor thesis finished

@@
 
-\includegraphicsHere{Theory/ITG-Turbulence.pdf}{
-	Sketch of ion temperature gradient driven instability on the outboard low field side of the tokamak \cite{Casson_PHD}
-}{fig:ITG}{0.5}
+\begin{center}
+	\captionsetup{type=figure}
+	\input{../pictures/Theory/ITG-Turbulence.tex}
+	\captionof{figure}{Sketch of ion temperature gradient driven instability on the outboard low field side of the tokamak. \cite{Casson_PHD}}
+	\label{fig:ITG}
+\end{center}
 
-The ITG driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak.
+The ITG-driven instability is present when a sufficiently large ion temperature gradient occurs and is driven by the so-called bad curvature of the outboard low field side of the tokamak.
 The mechanism is briefly explained with the help of Fig.
@@
 As stated in the equation for the electric field ($\vect{E} = -\nabla \phi$) the perturbation of the potential $\phi$ causes an electric field $\vect{E}$ which results in a $\exb$ drift with velocity $\vect{v}_E$.
-The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($T^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($T^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.\bigskip
-
+The $\exb$ drift direction transports hot plasma in the outer hot region of perturbation ($T^+ > 0$, \textcolor{red}{red} cell) and cooler plasma into inner cold region perturbation ($T^- < 0$, \textcolor{blue}{blue} cell) which reinforce the process and instability grows.
+\newpage
 The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there.

### 2023-06-20 bachelorthesis/24-zonalflows.tex
#39: Bachelor thesis finished

@@
+\newpage
 \section{Zonal Flows and Shearing Rate $\wexb$}
@@
 \includegraphicsHere{Theory/Zonal-Flow-Generation.pdf}{
-	Subdued turbulence because of stretched eddies, which breaks in smaller structures caused by sheared $\exb$ drifts \cite{Hammett2009}
+	Subdued turbulence because of stretched eddies, which breaks in smaller structures caused by sheared $\exb$ drifts. \cite{Hammett2009}
 }{fig:zonalflow}{0.9}
@@
 This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}.
-The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG driven turbulence at higher $\rlt$ compared to the linear threshold.
+The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG-driven turbulence at higher $\rlt$ compared to the linear threshold.
 The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig.
@@
 
-The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Peeters2016}
+The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined as\cite{Rath2016, Pueschel2008, Peeters2016}
 \begin{equation}
@@
 \end{equation}
-where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
-The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
+where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature and $e$ is the elementary charge).
+The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to Ref. \citenum{Rath2021}
 \begin{equation}
@@
 
-As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG driven turbulence in the adabatic electron approximation.
+As mentioned before the turbulent transport occurs through avalanches which also mediate the turbulence of the ITG-driven turbulence in the adabatic electron approximation.
 Avalanches are initiated when the $\exb$ sharing rate has a zero crossing with a steep flank and propagates outwards with negative shearing and inwards with positive shearing.

### 2023-06-20 bachelorthesis/25-gyrokinetics.tex
#39: Bachelor thesis finished

@@
 For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles position $\vect{x}$ and velocity $\vect{v}$ is needed.
-Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often preferred \cite{Garbet2010} which results through evolution of the particle density distribution function in the \textit{Vlasov equation}
+Because collisions are happening at much smaller frequencies than the characteristic frequencies connected to turbulence, the collisionless model is often preferred \cite{Garbet2010} which results through evolution of the particle density distribution function in the \textit{Vlasov equation}
 \begin{gather}
@@
 The length scales of the turbulence are associated with the wave vector $\vect{k}$ which can be seperated into a perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field.
-The perpendicular component $k_\perp$ is of the ordre of the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the scale on which the equilibrium density $n_0$ varies, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
+The perpendicular component $k_\perp$ is of the order of the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the scale on which the equilibrium density $n_0$ varies, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
 The gradient length compares with the machine size $R$ which leads to the normalized Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak.
@@
 Experiments show that in the core plasma the fluctuation amplitude of the density perturbation $\delta n/n_0$ and magnetic field fluctuation $\delta B/B_0$, with $B_0$ as strength of the magnetic field in equilibrium, is of order $\sim 10^{-4}$.
-Together this seperation result in the following
+Together this separation result in the following
 \begin{gather}
@@
 with $\epsilon_\mathrm{g} \ll 1$ which applies to the typical dynamics of a fusion core plasma.
-\cite{Brizard2007,Garbet2010} The gyrokinetic ordering allows tallows to formulate reduced governing equations referred to as the gyrokinetic formalism.
+\cite{Brizard2007,Garbet2010} The gyrokinetic ordering allows to formulate reduced governing equations referred to as the gyrokinetic formalism.
 
@@
 The gyro motion is described together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.
-The gyrocenter position $\vect{X}$ is expressed with the particle position $\vect{x}$ and the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$ as $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$.
+The gyrocenter position $\vect{X}$ is expressed with the particle position $\vect{x}$ and the position Larmor radius vector $\vect{\rho_\mathrm{L}}$ as $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$.
 
@@
 
-All simulations in this thesis are performed with GKW (Gyro Kinetic Workshop) which solves the gyrokinetic equationin the $\delta f$ approximation and local limit.
+All simulations in this thesis are performed with GKW (Gyro Kinetic Workshop) which solves the gyrokinetic equation in the $\delta f$ approximation and local limit.
 For the $\delta f$ approximation the particle density distribution function f is seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}).
@@
 The $\delta f$ approximation applies for global and local description where the latter is used in this thesis.
-The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities are considered constant over the radial extent of the simulation volume.
-Because of the tokamak symmetry all equilibrium quantities are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general.
+The \textit{local limit} on the other hand uses the spatial separation of equilibrium and perturbation and implies that all equilibrium and geometry quantities are considered constant over the radial extent of the simulation volume.
+Because of the tokamak symmetry all equilibrium quantities are also constant in the toroidal direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general.
 \\\bigskip
@@
 In addition to this, in GKW the following coordinates get applied with the parameter $\vect{X} = \{\xcoord, \ycoord, s\}$.
-Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate.
+Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$ and $\ycoord$ labels the field lines and is an approximate binormal coordinate.
 Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}.

### 2023-06-20 bachelorthesis/31-simulation.tex
#39: Bachelor thesis finished

@@
 The numerical resolution is compliant to the "Standard resolution with 6th order (S6)" set-up of the aforementioned reference.
-A summary of the numerical parameters is given in Tab.~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.
+A summary of the numerical parameters is given in Table~\ref{tab:resolution} and for more details about the definition of individual quantities the reader is referred to Refs.
 \citenum{Peeters2009, Peeters2016}.
@@
 	\captionof{table}{
-		Resolution used in this paper: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+		Resolution used in this work: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$.
 	}

### 2023-06-20 bachelorthesis/33-cluster.tex
#39: Bachelor thesis finished

@@
 
-The simulations are performed on the \texttt{/scratch} directory because the \texttt{/home} folder has an disk space limit of \SI{80}{\giga\byte} which is not enough to perform long runs of GKW.
+The simulations are performed on the \texttt{/scratch} directory because the \texttt{/home} folder has a disk space limit of \SI{80}{\giga\byte} which is not enough to perform long runs of GKW.

### 2023-06-20 bachelorthesis/34-restartscript.tex
#39: Bachelor thesis finished

@@
 As script language \texttt{python} was chosen because of its simple syntax and the variety of tools compatible with the \texttt{bash} shell.
-\texttt{bash} was also considered, but after some issues occurred regarding the mechanism to check wether a file exist or not, \texttt{python} was selected.
-Furthermore, the script uses the standard \texttt{python3} library and is therefore executabe with any \texttt{python3} enviroment.
+\texttt{bash} was also considered, but after some issues occurred regarding the mechanism to check whether a file exist or not, \texttt{python} was selected.
+Furthermore, the script uses the standard \texttt{python3} library and is therefore executable with any \texttt{python3} environment.
 
 The script was named \texttt{slurm\_monitor.py} to indicate that this script works with the \texttt{Slurm} resource manager.
-The source code was pushed to the gkw repository on BitBucket \cite{slurmmonitor} or can be found in Appendix \ref{append:slurmcode}
+The source code was pushed to the GKW repository on BitBucket \cite{slurmmonitor} or can be found in Appendix \ref{append:slurmcode}
 
@@
 \end{itemize}
-All boolean parameters are switches, i.e., the praser option only changes the value from \texttt{False} to \texttt{True}.
+All boolean parameters are switches, i.e., the parser option only changes the value from \texttt{False} to \texttt{True}.
 
@@
 The output is formatted as a table and can be adjusted in its appearance with the \texttt{-\/-format} parser option [Chapter \ref{sub:codeFeatures}].
-The output types written in the status file are listed in Tab.
-\ref{tab:statusTypes}.
-To each output the restart script writeesthe date, time and the duration of the monitoring additionally into the table.
+The output types written in the status file are listed in Table \ref{tab:statusTypes}.
+To each output the restart script writes the date, time and the duration of the monitoring additionally into the table.
 \begin{center}
@@
 \end{verbatim}
-\captionof{figure}{Styles of output table in status file}
+\captionof{figure}{Styles of output table in status file.}
 \label{fig:statusStyle}
@@
 It also contains the current output from \texttt{squeue}, if the simulation is running or pending, else the important information of the simulation.
-An example progress bar ist shown in Fig.
+An example progress bar is shown in Fig.
 \ref{fig:statusBar}.
@@
 \end{verbatim}
-\captionof{figure}{Progress bar example}
+\captionof{figure}{Progress bar example.}
 \label{fig:statusBar}

### 2023-06-20 bachelorthesis/41-resolution.tex
#39: Bachelor thesis finished

@@
 At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without numerical dissipation biasing the physical outcome.
-Numerical dissipation can therefore result to no formation of zonal flow structures, which cause an permanent turbulent state of the simulation.
+Numerical dissipation can therefore result to no formation of zonal flow structures, which cause a permanent turbulent state of the simulation.
 The goal behind this testing is to reduce \textbf{calculation time} and \textbf{costs} of the simulation.
@@
 	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
-	\item Reduce to known minimum number of grid points to verify result in general.
+	\item Reduce to known the minimum number of grid points to verify result in general.
 \end{enumerate}
@@
 
-Starting from the \enquote{Standard resolution with 6th order (S6)} [Tab.
-\ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref.
+Starting from the \enquote{Standard resolution with 6th order (S6)} [Table \ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref.
 \citenum{Peeters2016} in Section IV.
@@
 \includegraphicsHere{S6_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux.pdf}{
-	Time trace of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for benchmark
+	Time trace of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for benchmark.
 }{fig:eflux-ns16-nvpar64-nmu9}{}
@@
 \includegraphicsHere{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{
-	Time traces for $\rlt = 6.0$ and $\rlt = 6.3$ for benchmark
+	Time traces for $\rlt = 6.0$ and $\rlt = 6.3$ for benchmark.
 }{fig:eflux-gradient-length}{}
@@
 \includegraphicsHere{Comparison/Resolution/Nvpar/S6_rlt6.0_boxsize1x1_Ns16_Nvpar16-32-48-64_Nmu9_eflux_comparison.pdf}{
-	Time traces for reduced parallel velocity grid points $\Nvpar$
+	Time traces for reduced parallel velocity grid points $\Nvpar$.
 }{fig:eflux-ns16-nvpar16-32-48-64-nmu9}{}
@@
 \includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu6-9_eflux_comparison.pdf}{
-	Time traces for reduced magnetic moment grid points $\Nmu$
+	Time traces for reduced magnetic moment grid points $\Nmu$.
 }{fig:eflux-ns16-nvpar64-nmu6-9}{}
@@
 
-\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar48_Nmu6-9_eflux_comparison.pdf}{
-	Time traces for reduced magnetic moment grid points $\Nmu$
+\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu6-9_eflux_comparison.pdf}{
+	Time traces for reduced magnetic moment grid points $\Nmu$.
 }{fig:eflux-ns16-nvpar48-nmu6-9}{}
@@
 \includegraphicsHere{Comparison/Resolution/Ns/S6_rlt6.0_boxsize1x1_Ns12-16_Nvpar64_Nmu9_eflux_comparison.pdf}{
-	Time traces for reduced magnetic field grid points $\Ns$
+	Time traces for reduced magnetic field grid points $\Ns$.
 }{fig:eflux-ns12-16-nvpar64-nmu9}{}
@@
 
-Together with the results of Chapter \ref{sub:reduceNvpar}, \ref{sub:reduceNmu} and \ref{sub:reduceNs} the final resolution used in the upcoming simulations is displayed in Tab.
-\ref{tab:resolution}, with reduced number of grid points for the parallel velocity $\Nvpar$.
+Together with the results of Chapter \ref{sub:reduceNvpar}, \ref{sub:reduceNmu} and \ref{sub:reduceNs} the final resolution used in the upcoming simulations is displayed in Table \ref{tab:resolution}, with reduced number of grid points for the parallel velocity $\Nvpar$.
 \begin{center}
@@
 	\captionof{table}{
-		Resolution used in this work: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
+		Resolution used in this work: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$.
 	}

### 2023-06-20 bachelorthesis/42-boxsizescan.tex
#39: Bachelor thesis finished

@@
 The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
-Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size.
+Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Table~\ref{tab:resolution}] and standard box size.
 \\
@@
 \end{gather*}
-Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(a)].
+Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(a)].
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ITG driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
-The negative gradient of the perturbed perpendicular and parallel ion pressure (not shown) exhibit positive corrugations in regions with maximum $|\wexb|$ and negative corrugations at zero crossings of $\wexb$.
-A radial force balance analysis suggests that the structures in $\wexb$ as depicted in Fig.~\ref{fig:wexb-stable-comparison} are not a consequence of the pressure gradient corrugations as discussed elsewhere \cite{Kosuga2013}.
-Rather, the corrugations in the pressure gradient have to be interpreted as a consequence of the staircase structure in $\wexb$ due to the stabilizing zonal flow - turbulence interaction.
-Fig.~\ref{fig:wexb-stable-comparison}\,(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
-Hence, the basic size of the pattern not only converges with increasing radial box size, the converged radial size turns out to at least roughly agree with the standard radial box size of Ref.
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} [$\gamma$ is the growth rate of the most unstable linear ion temperature gradient (ITG)-driven Eigenmode], connected by steep flanks where $\wexb$ crosses zero.
+Figure~\ref{fig:wexb-1-2-3-4x1-compariso} or Figure~\ref{fig:wexb-stable-comparison}(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
+Hence, the basic size of the pattern not only converges with increasing radial box size, but the converged radial size also turns out to at least roughly agree with the standard radial box size of Ref.
 \citenum{Peeters2016}.
@@
 
-Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
+Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
 Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
-During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical time scale of several $\sim 10^3\,R/\vth$.
+During this period the zonal flow mode with $\nzf = 3$, i.e., the mode that dominates the staircase pattern in final suppressed phase, undergoes a long-term evolution with a typical timescale of several $\sim 10^3\,R/\vth$.
 \\
@@
 This demonstrates a competition between the $\nzf = 3$ and $\nzf = 4$ modes.
-Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}\,(b) dotted line].
+Most importantly, no secular growth of the $\nzf = 1$ (box scale) zonal flow mode is observed during the entire quasi-stationary turbulent phase [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b) dotted line].
 The above discussion indicates that although the $\nzf = 3,~4$ zonal modes compete, the pattern scale does not converge to the radial box scale but rather to a mesoscale of $\sim 57 - 76\,\rhoth$ (i.e., $\nzf = 4,~3$ in the $3\times1$ case).
 \includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes\\
-	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for radial increased box sizes,\\
+	\textbf{(b)} Time traces of $\hatwexbamp$ for radial increased box sizes.
 }{fig:wexb-eflux-1-2-3-4x1-comparison}{}
+
+\begin{center}
+	\captionsetup{type=figure}
+	\includegraphics[height = 0.18\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}
+	\captionof{figure}{
+		Comparison of shearing rate $\wexb$ for the radial box size scan averaged over given a time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG-driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility:\\
+		\begin{tabular}{l l l l}
+			$t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, \\
+			$t_{3\times 1}$   & $\in [43000, 45000]$, & $t_{4\times 1}$   & $\in [26000, 28000]$.  \\
+		\end{tabular}
+	}
+	\label{fig:wexb-1-2-3-4x1-compariso}
+\end{center}
 
@@
 Interestingly, suppression of the turbulence by the emergence of a fully developed staircase pattern almost always occurs after $\sim 1000~R/\vth$ [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}\,(a)], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
-As shown in Fig.~\ref{fig:wexb-stable-comparison}\,(b) also this test confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+\ref{fig:eflux-1x1-2x2-3x3-comparison}(a)], i.e., significantly faster compared to the $3\times1$ and $4\times1$ realizations.
+As shown in Fig.
+\ref{fig:wexb-1x1-2x2-3x3-comparison} or Fig.~\ref{fig:wexb-stable-comparison}(b) this test also confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 
 \includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes\\
-	\textbf{(b)} Time traces of $\hatwexbamp$ for isotropic increased box sizes
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for isotropic increased box sizes,\\
+	\textbf{(b)} Time traces of $\hatwexbamp$ for isotropic increased box sizes.
 }{fig:eflux-1x1-2x2-3x3-comparison}{}
@@
 By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}\,(b), Fig.
-\ref{fig:wexb-stable-comparison}\,(b)].
+\ref{fig:eflux-1x1-2x2-3x3-comparison}(b), Fig.
+\ref{fig:wexb-1x1-2x2-3x3-comparison} and Fig.
+\ref{fig:wexb-stable-comparison}(b)].
 Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57,~ 76]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
-The scale of structures developing in the $1.5\times1.5$ and $2.5\times2.5$ realizations (not included in Fig.~\ref{fig:wexb-stable-comparison} and Fig.
+The scale of structures developing in the $1.5\times1.5$ and $2.5\times2.5$ realizations (not included in Fig.~\ref{fig:wexb-stable-comparison}, Fig.~\ref{fig:wexb-1x1-2x2-3x3-comparison} and Fig.
 \ref{fig:eflux-1x1-2x2-3x3-comparison} to preserve the clarity of this figure) also lie within the range given above.
-Note that two additional realizations of the $3\times3$ case with different initial conditions and otherwise identical parameters confirm structure formation on scales within the range given above.
+
+\newpage
+\begin{center}
+	\captionsetup{type=figure}
+	\includegraphics[height = 0.18\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}
+	\captionof{figure}{
+		Comparison of shearing rate $\wexb$ for the isotropic box size scan ($1.5\times1.5$ not $2.5\times2.5$ included) averaged over given a time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG-driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility:\\
+		\begin{tabular}{l l l l l l}
+			$t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,  & $t_{3\times 3}$   & $\in [2000, 3000]$. \\
+		\end{tabular}
+	}
+	\label{fig:wexb-1x1-2x2-3x3-comparison}
+\end{center}
+
+Note that two additional realizations of the $3\times3$ case with different initial conditions and otherwise identical parameters confirm structure formation on scales within the range given above [Fig.~\ref{fig:ini-3x3-comparison}].
+
+\begin{center}
+	\captionsetup{type=figure}
+	\includegraphics[height = 0.18\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_cosine2-cosine5-noise_wexb_comparison.pdf}
+	\captionof{figure}{
+		Comparison of shearing rate $\wexb$ for the $3\times3$ box size with different initial conditions averaged over given a time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG-driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility:\\
+		\begin{tabular}{l l l l l l}
+			$t_\mathrm{cosine2}$   & $\in [2000, 3000]$,   & $t_\mathrm{cosine5}$   & $\in [2000, 3000]$,  & $t_\mathrm{noise}$     & $\in [4000, 5000]$. \\
+		\end{tabular}
+	}
+	\label{fig:ini-3x3-comparison}
+\end{center}
+
+\newpage
+The negative gradient of the perturbed perpendicular and parallel ion pressure [Fig.
+\ref{fig:ene-webx-3x3-comparison}] exhibit positive corrugations in regions with maximum $|\wexb|$ and negative corrugations at zero crossings of $\wexb$.
+A radial force balance analysis suggests that the structures in $\wexb$ as depicted in Fig.~\ref{fig:wexb-stable-comparison} are not a consequence of the pressure gradient corrugations as discussed elsewhere \cite{Kosuga2013}.
+Rather, the corrugations in the pressure gradient have to be interpreted as a consequence of the staircase structure in $\wexb$ due to the stabilizing zonal flow - turbulence interaction.
+
+\includegraphicsHere{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/noise/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_noise_wexb_dens_ene_selection.pdf}{
+	Time averaged radial proﬁles of (top) the $\exb$ shearing rate $\wexb$ and (bottom) the negative gradient of the perpendicular and parallel energy moment $-\langle \nabla_{\!\!x} E_{\perp, \mathrm{i}}\rangle$ and  $-\langle \nabla_{\!\!x} E_{\parallel, \mathrm{i}}\rangle$, respectively, as well as the electric ﬁeld term $e n_0 \langle \nabla_{\!\!x} \phi\rangle$ of the force balance (on the right axis).
+Shown is a $3\times3$ realization with $\rlt = 6.0$.
+Horizontal black and gray bars in the bottom panel indicate the modulation of the negative gradient of the perpendicular energy on the staircase scale.
+}{fig:ene-webx-3x3-comparison}{0.8}
 
@@
 \end{gather*}
-As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}\,(a)].
-The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) and confirms again a size of a typical mesoscale.
-Fig.~\ref{fig:wexb-stable-comparison}\,(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57,~ 76]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
-The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}\,(c) in the $3\times 2.5$ realization.
-The staircase structure has a pattern between $3$ and $4$ repetitions which get represented in the second repetition with no signifciant plateau at positive shear.
+As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}(a)].
+The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-3x1.5-2.5-3-5-comparison} or Fig.~\ref{fig:wexb-stable-comparison}(c) and confirms again a size of a typical mesoscale.
+Fig.~\ref{fig:wexb-stable-comparison}(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57,~ 76]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}(c) in the $3\times 2.5$ realization.
+The staircase structure has a pattern between three and four repetitions which get represented in the second repetition with no significant plateau at positive shear.
 Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
 The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each [Fig.
-\ref{fig:eflux-3x1.5-2.5-3-5-comparison}\,(b)].
+\ref{fig:eflux-3x1.5-2.5-3-5-comparison}(b)].
 \includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}{
-	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes\\
-	\textbf{(b)} Time traces of $\hatwexbamp$ for binormal increased box sizes
+	\textbf{(a)} Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for binormal increased box sizes,\\
+	\textbf{(b)} Time traces of $\hatwexbamp$ for binormal increased box sizes.
 }{fig:eflux-3x1.5-2.5-3-5-comparison}{}\bigskip
+
+\begin{center}
+	\captionsetup{type=figure}
+	\includegraphics[height = 0.18\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}
+	\captionof{figure}{
+		Comparison of shearing rate $\wexb$ for the binormal box size scan averaged over given a time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG-driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility:\\
+		\begin{tabular}{l l l l}
+			$t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   \\
+		    $t_{3\times 3}$   & $\in [2000, 3000]$,   & $t_{3\times 5}$   & $\in [1000, 3000]$.    \\
+		\end{tabular}
+	}
+	\label{fig:wexb-3x1.5-2.5-3-5-comparison}
+\end{center}
+
+\newpage
 
@@
 \includegraphicsHere{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison_thesis.pdf}{
-	Comparison of shearing rate $\wexb$ for each box size scan averaged over given a time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG driven Eigenmode.
-The staircase structures are radially shifted with respect to each over till alignment for better visibility.\\
+	Comparison of shearing rate $\wexb$ for each box size scan averaged over given a time interval and the growth rate $\pm \gamma$ of the most unstable linear ITG-driven Eigenmode.
+The staircase structures are radially shifted with respect to each over till alignment for better visibility:\\
 	\begin{tabular}{l l l l l}
 		\textbf{(a) radial:}    & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 1}$   & $\in [15000, 18000]$, \\
-		                        & $t_{3\times 1}$   & $\in [43000, 45000]$, & $t_{4\times 1}$   & $\in [26000, 28000]$  \\
+		                        & $t_{3\times 1}$   & $\in [43000, 45000]$, & $t_{4\times 1}$   & $\in [26000, 28000]$;  \\
 		\textbf{(b) isotropic:} & $t_{1\times 1}$   & $\in [2000, 5000]$,   & $t_{2\times 2}$   & $\in [2000, 3000]$,   \\
-		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    &                   &                       \\
+		                        & $t_{3\times 3}$   & $\in [2000, 3000]$;    &                   &                       \\
 		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   \\
-		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    & $t_{3\times 5}$   & $\in [1000, 3000]$    \\
+		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    & $t_{3\times 5}$   & $\in [1000, 3000]$.    \\
 	\end{tabular}

### 2023-06-20 bachelorthesis/50-conclusion.tex
#39: Bachelor thesis finished

@@
 
-Additionally, the active development of a restart script in \texttt{python} led to further convenience during the task of performing simulations on the \texttt{btrzx1} cluster. \bigskip
+Additionally, the active development of a restart script in \texttt{python3} led to further convenience during the task of performing simulations on the \texttt{btrzx1} cluster. \bigskip
 
-% what has been done
-Through careful tests this bachelor thesis confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG) driven turbulence.
-% mention the scale explicitly, and also that it is the scale compliant to CBC parameters (for other plasma parameters it might differ)
-A mesoscale pattern size of $\sim 57.20 - 76.27\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters.
-% compare to other studies
+Through careful tests this bachelor thesis confirms the radial size convergence of the $\exb$ staircase pattern in local gyrokinetic flux tube simulations of ion temperature gradient (ITG)-driven turbulence.
+A mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG-driven turbulence for Cyclone Base Case parameters.
 This length scale is somewhat larger compared to results from global studies with finite $\rho_\ast$, which report of a few $10\,\rhoth$ \cite{Pradalier2010}, and has to be considered the proper mesoscale in the local limit $\rho_\ast \rightarrow 0$.
-% relevance of the outcome
 The occurrence of this mesoscale implies that non-locality, in terms of Ref.
-\citenum{Pradalier2010}, is inherent to ITG driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{McMillan2009, Pradalier2010, Rath2016, Peeters2016}.
+\citenum{Pradalier2010}, is inherent to ITG-driven turbulence, since avalanches are spatially organized by the $\exb$ staircase pattern \cite{McMillan2009, Pradalier2010, Rath2016, Peeters2016}.

### 2023-06-20 presentation/presentation.tex
#39: Embedded videos
#39: Redefined footline and heading
#39: Presentation so far

@@
 
-\documentclass[compress,aspectratio=43]{beamer}
+\documentclass[compress,aspectratio=1610,noflama]{beamer}
 
@@
 
-\usepackage[ngerman]{babel}
+\usepackage[english]{babel}
 \usepackage{graphicx}
@@
 \usepackage{soul}
+%\usepackage[numbers, super]{natbib}
+\usepackage{tikz}
+\usepackage{multirow}
+
+%--------------------------------------------------------------------------
+% Variables
+%--------------------------------------------------------------------------
+
+\input{../bachelorthesis/colors.tex}
+\input{../bachelorthesis/flowdiagramm.tex}
+\input{../bachelorthesis/variables.tex}
+\input{../bachelorthesis/tikz.tex}
+
+%--------------------------------------------------------------------------
+% Variables
+%--------------------------------------------------------------------------
+
+\newcommand{\backgroundpicture}{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}
+\newcommand{\backgroundxshift}{0.27\paperwidth}
+\newcommand{\backgroundwidth}{0.7\paperwidth}
+
+\newcommand{\logopicture}{Uni_Logo_white_black}
 
@@
 
-\usetikzlibrary{mindmap,backgrounds}
 \graphicspath{{../pictures/}}
 
-\newcommand{\tabitem}{~~\llap{\textbullet}~~}
 \newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
@@
 
-\title{Turbulenzen in Plasma}
-\subtitle{Beispiel anhand von Plasma im Tokamak Reaktor}
+\title{Size convergence of the E$\times$B \\ Staircase pattern in flux tube \\ simulations of ion temperature \\ gradient driven turbulence}
+\subtitle{~}
 \author{Manuel Lippert}
-\institute{Physik (Master of Science)}
+\institute{Theoretical Physics V}
 
@@
 
-	\begin{frame}{Gliederung}
-		\tableofcontents[hideallsubsections]
+	%\begin{frame}{Gliederung}
+	%	\tableofcontents[hideallsubsections]
+	%\end{frame}
+
+	\section*{Motivation}
+	\begin{frame}
+		\frametitle{Motivation}
+
+		\begin{center}
+			\begin{itemize}
+				\item <2-> Ion temperature gradient driven turbulence close to marginal stability exhibits Zonal flow pattern formation on mesoscales 
+				\item <3-> Patterns are called $\exb$ staircase structures
+			\end{itemize}
+		\end{center}
+		\bigskip
+		\onslide<4-> Observed pattern formation in
+		\begin{center}
+			\begin{itemize}
+				\item <5-> local gradient-driven flux-tube simulations, including collisions and background $\exb$ shear,
+				\item <6-> local flux-driven realizations, including mean electric field shear,
+				\item <7-> global gradient-driven and global flux-driven studies.
+			\end{itemize}
+			\bigskip
+			\onslide<8-> \textbf{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}
+		\end{center}
 	\end{frame}
 
-	\centering
+	\section*{Theory}
 
-	\section{Magnetohydrodynamische (MHD) Gleichungen}
-	%\subsection{Magnetohydrodynamische Gleichungen}
+	\begin{frame}
+		\frametitle{Charged Particle Motion}
+
+		\begin{center}
+			\begin{tabular}{>{\onslide<2->}c<{\onslide}@{\hspace{0.5cm}} >{\onslide<3->}c<{\onslide}@{\hspace{0.5cm}} >{\onslide<4->}c<{\onslide}}
+				\textbf{Lorentz force} & 
+				\textbf{Electric force} & 
+				\textbf{Inhomogeneous magnetic field} \\
+				$F_{\perp, B} = |q| v_\perp B$ & 
+				$F_{\parallel,E} = qE_\parallel$ & 
+				$F_{\parallel,\nabla_{\!\parallel} B} = - \underbrace{\frac{mv^2_{\perp}}{2B}}_{\mu} \nabla_{\!\parallel} B$ \\
+				\input{../pictures/Theory/Lorentz-Force.tex} & 
+				\input{../pictures/Theory/Electric-Force.tex} & 
+				\input{../pictures/Theory/Inhomogenous-Magnetic-Field.tex} \\
+			\end{tabular}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Drift in the Gyrocenter}
+		\begin{center}
+			\begin{tabular}{>{\onslide<2->}c<{\onslide} >{\onslide<3->}c<{\onslide} >{\onslide<3->}c<{\onslide}}
+				{\boldmath $\exb$} \textbf{Drift} & 
+				{\boldmath $\nabla B$} \textbf{Drift} & 
+				\textbf{Curvature Drift} \\
+				$\vect{v}_{E} = \frac{\vect{E}\times\vect{B}}{B^2}$ & 
+				$\vect{v}_{\nabla B} = \frac{m v^2_{\perp}}{2 q}\frac{\vect{B}\times \nabla B}{B^3}$ & 
+				$\vect{v}_{\kappa} = \frac{m v^2_{\parallel}}{qB^2}\vect{B}\times \underbrace{\vect{\kappa}}_{\frac{\nabla B}{B}} = \frac{m v^2_{\parallel}}{q} \frac{\vect{B}\times \nabla B}{B^3}$\\[1cm]
+				\input{../pictures/Theory/ExB-Drift.tex} & 
+				\multicolumn{2}{>{\onslide<3->}c<{\onslide}}{\input{../pictures/Theory/GradB-Drift.tex}} \\
+			\end{tabular}
+		\end{center}
+
+	\end{frame}
+	\begin{frame}
+		\frametitle{Magnetic Confinement in Tokamak}
+		\begin{center}
+			\begin{tabular}{>{\onslide<2->}c<{\onslide} >{\onslide<2->}c<{\onslide}}
+				\includegraphics[width=0.6\paperwidth]{Theory/Tokamak-Torus.pdf} & $\beta = \frac{nT}{\mu_0 B^2/2}$
+			\end{tabular}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Ion Temperature Gradient (ITG) Driven Instability}
+		\begin{center}
+			\onslide<2->\input{../pictures/Theory/ITG-Turbulence.tex}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Zonal Flows}
+		\begin{center}
+			\onslide<2->\includegraphics[width = 0.8\paperwidth]{Theory/Zonal-Flow-Generation.pdf}
+			%\begin{itemize}
+			%	\item <3-> Zonal flows are linked to the $\exb$ flows
+			%	\item <4-> Zonal flow is tangential to the flux-surfaces and does not contribute to the turbulent transport across the flux-surfaces
+			%\end{itemize}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Shearing Rate}
+
+		\onslide<2-> Pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. \\\bigskip
+
+		Properties:
+		\begin{enumerate}
+			\item[(1)] <3-> They occur on a \textit{radial mesoscale} of order of $\rho_\mathrm{L} < 10\,\rhoth < R$
+			\item[(2)] <4-> The structure are \textit{quasi stationary} in space and vary on time scales much longer than the typical turbulence time scales.
+			\item[(3)] <5-> They have a \textit{typical amplitude} of the order of $10^{-1} \vth/R$.
+			\item[(4)] <6-> Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate.
+		\end{enumerate}
+		\onslide<7->{
+		\begin{center}
+			\begin{tabular}{c c}
+				$\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}$ & $\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)$ \\
+			\end{tabular}
+		\end{center}
+		}
+	\end{frame}
+
 	%\begin{frame}
-	%	\frametitle{Magnetohydrodynamische Gleichungen}
+	%	\frametitle{Vlasov Equation}
+	%	\begin{gather*}
+	%		\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0
+	%	\end{gather*}
+	%	\begin{gather*}
+	%		n = \int \mathrm{d}\vect{v}\,f(\vect{x}, \vect{v}, t) \qquad j = q \int \mathrm{d}\vect{v}\, v f(\vect{x}, \vect{v}, t)
+	%	\end{gather*}
 	%	\begin{gather*}
 	%		\begin{aligned}
-	%			\partial_t \rho + \nabla \cdot (\rho \vect{v}) &= 0 \\
-	%			\rho \partial_t \vect{v} + \rho (\vect{v} \cdot \nabla) \vect{v} + \nabla p &= \frac{\vect{J} \times \vect{B}}{c} + \rho \vect{g} + \vect{F} \\
-	%			\nabla \times \vect{B} &= \frac{4 \pi}{c} \vect{J} \\
-	%			\nabla \times \vect{E} &= -\frac{1}{c}\partial_t\vect{B} \\
-	%			\nabla \cdot \vect{B} &= 0 \\
-	%			\vect{J} &= \sigma \left(\vect{E} + \frac{\vect{v} \times \vect{B}}{c}\right)
+	%			\nabla \cdot \vect{B} &= 0 &\qquad \nabla \times \vect{B} &= \mu_0\left( \sum_\mathrm{species} j + \epsilon_0 \frac{\partial \vect{E}}{\partial t} \right) \\
+	%			\nabla \cdot \vect{E} &= \frac{1}{\epsilon_0} \sum_\mathrm{species} qn &\qquad \nabla \times \vect{E} &= - \frac{\partial \vect{B}}{\partial t}
 	%		\end{aligned}
@@
 	\begin{frame}
-		\frametitle{Magnetohydrodynamische (MHD) Gleichungen}
-		\begingroup
-			\setbeamercolor{block title}{bg=Sec3Dark}
-			\setbeamercolor{block body}{bg=Sec3}
-			\begin{block}{Kontinuitätsgleichung}
-				\centering
-				$\partial_t \rho + \nabla \cdot (\rho \vect{v}) = 0$
-			\end{block}
-		\endgroup
-		\begingroup
-			\setbeamercolor{block title}{bg=Sec2CompDark}
-			\setbeamercolor{block body}{bg=Sec2Comp}
-			\begin{block}{Navier-Stokes Gleichung}
-				\centering
-				$\rho \partial_t \vect{v} + \rho (\vect{v} \cdot \nabla) \vect{v} = - \nabla p + \frac{\vect{J} \times \vect{B}}{c} + \rho \vect{g} + \vect{F}$
-			\end{block}
-		\endgroup
-		\begingroup
-			\setbeamercolor{block title}{bg=Sec1Dark}
-			\setbeamercolor{block body}{bg=Sec1}
-			\begin{block}{Maxwell Gleichungen}
-				\centering
-				\begin{tabular}{l l}
-					$\nabla \cdot \vect{B} = 0$ & $\nabla \times \vect{B} = \frac{4 \pi}{c} \vect{J}$ \\[0.3cm]
-					$\vect{J} = \sigma \left(\vect{E} + \frac{\vect{v} \times \vect{B}}{c}\right)$ & $\nabla \times \vect{E} = -\frac{1}{c}\partial_t\vect{B}$ \\
+		\frametitle{Gyrokinetics}
+		\pause
+		\begin{tabular}{l l l}
+
+			$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0$ &
+			$f(\vect{x}, \vect{v}, t)$ & 
+			Vlasov Equation \pause \\[0.7cm]
+
+			\multicolumn{2}{l}{$\Downarrow$ \qquad $\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}$} &
+			Gyrokinetic Ordering \pause \\[0.7cm]
+
+			$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0$ &
+			$f(\vect{X}, v_{\parallel}, \mu)~;~\frac{\mathrm{d}\mu}{\mathrm{d}t} = 0$ & 
+			Gyrokinetic Formalism \pause \\[0.7cm]
+
+			\multicolumn{2}{l}{$\Downarrow$ \qquad $f = f_0 + \delta f$} & 
+			$\delta f$ Approx \& Local Limit \pause \\[0.7cm]
+
+			\multicolumn{2}{l}{$\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S$} &
+			Gyrokinetic Equation \\[0.7cm]
+			\multicolumn{3}{l}{$g \sim (f_0 + \delta f)\qquad S \sim (f + f_0)\qquad \vect{v}_\mathrm{D} \sim \nabla B \qquad \vect{v}_\chi \sim \exb \qquad f(\xcoord, \ycoord, s, v_\parallel, \mu)$} \\
 		\end{tabular}
-			\end{block}
-		\endgroup
-	\end{frame}
-	\begin{frame}
-		\frametitle{Beschreibung von Plasma im Tokamak Reaktor}
-		\includegraphics[scale = 1]{Presentation/Zonal_Flow.PNG}
-		\begin{alertblock}{Achtung Instabilität am Plasmarand!}
-			\centering
-			Ionen Temperaturgradient Instabilität $\approx$ Rayleigh-Taylor Instabilität
-		\end{alertblock}
 	\end{frame}
 
-	\section{Ionen Temperaturgradient (ITG) Instabilität}
+	\section*{Material and Methods}
+
 	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_00.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_01.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_02.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_03.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_04.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_05.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_06.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_07.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_08.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_09.pdf}
-	\end{frame}
-	\begin{frame}
-		\frametitle{ITG Instabilität am Plasmarand}
-		\includegraphics[scale=0.8]{Presentation/ITG_Instability_10.pdf}
+		\frametitle{Simulation Setup}
+
+		\begin{center}
+			\begin{itemize}
+				\item <2-> Non-linear flux tube version of Gyrokinetic Workshop (GKW) 
+				\item <3-> Circular concentric flux surfaces 
+				\item <4-> Standard resolution with 6th order (S6) 
+				\item <5-> Cyclone Base Case (CBC) parameters $\rlt = 6.0$ 
+				\item <6-> Turbulence level is quantified by the turbulent heat conduction coefficient $\chi$ (normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$)) 
+				\item <7-> Quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities
+				\item <8-> Standard box size $(L_\xcoord,~L_\ycoord) = (76.3,~89.8)\,\rhoth$
+			\end{itemize}
+			\onslide<9->{
+			\bigskip
+			\begin{tabular}{l | ccccc | ccccc | c | cc}
+				& $N_\mathrm{m}$ & $N_x$ & $N_\mathrm{s}$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_\mathrm{d}$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
+				\hline
+				S6   & 21    & 83    & 16    & 64                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
+			\end{tabular}}
+		\end{center}
 	\end{frame}
@@
 	\begin{frame}
-		\frametitle{Zonal Flow Bildung}
-		\begin{minipage}{0.6\textwidth}\raggedleft
-			\centering
+		\frametitle{Diagnostics}
+
+		\begin{center}
+			\onslide<2-> $\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord)$ \\[0.5cm]
+			\onslide<3-> $\kzf = 2\pi \nzf/L_\xcoord$ \\[0.5cm]
+			\onslide<4-> $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$ \\[0.5cm]
 			\begin{itemize}
-				\item ITG \ul{Instabilität} erzeugt turbulente Wirbelströmung
-				\item Wirbelströmung breitet sich über die Breite des Plasmatorus aus
-				\item Entstehung von \alert{Zonal Flows}
-				\item Namensgebung aus der Atmosphärenphysik (Bsp. Wolkenbänder von Jupiter)
+				\item <5-> Zonal flow mode that dominates the $\exb$ staircase pattern are called \textbf{basic mode} 
+				\item <6-> The basic mode exhibits the maximum amplitude in the spectrum $\hatwexbamp$
+			\end{itemize}
+		\end{center}
+	\end{frame}
+	\begin{frame}
+		\frametitle{Restart Script}
+		\pause
+
+		\newcommand{\distance}{2.5cm}
+
+		\begin{center}
+		    \begin{tikzpicture}[node distance=1.5cm]
+
+		        \node (init)         [init]                                                 {Start / Initialize};
+		        \node (startLoop)    [startstop, below of = init]                           {Loop};
+		        \node (checkStatus)  [process, below of=startLoop]                          {Check Status};
+		        \node (checkData)    [process, below of=checkStatus, yshift = -1.5cm]       {Check Data};
+		        \node (checkPending) [decision, below of=checkStatus, xshift=\distance]     {Pending};
+		        \node (checkRunning) [decision, below of=checkStatus, xshift=-\distance]    {Running};
+
+		        \draw [arrow] (init)           --                                           (startLoop);
+		        \draw [arrow] (startLoop)      --                                           (checkStatus);
+		        \draw [arrow] (checkStatus)    -|                                           (checkRunning);
+		        \draw [arrow] (checkRunning)   -- node[anchor=south] {No}                   (checkPending);
+		        \draw [arrow] (checkPending)   -- node[anchor=north west] {Yes} + ( 3,0) |- (startLoop);
+		        \draw [arrow] (checkRunning)   -- node[anchor=north east] {Yes} + (-3,0) |- (startLoop);
+		        \draw [arrow] (checkPending)   |- node[anchor=north west] {No}              (checkData);
+
+		    \end{tikzpicture}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Restart Script}
+
+		\newcommand{\distance}{2.5cm}
+
+		\begin{center}
+		    \begin{tikzpicture}[node distance=2cm]
+
+		        \node (checkData)    [process]       										{Check Data};
+		        \node (error)        [decision, below of=checkData, xshift=-\distance]      {Errors};
+		        \node (checkPoint)   [decision, below of=checkData, xshift=\distance]       {Checkpoint};
+		        \node (backup)       [process,  below of=error, xshift=-\distance]          {Backup Data};
+		        \node (restore)      [process,  below of=checkPoint, xshift=\distance]      {Restore Data};
+		        \node (reset)        [process,  below of=checkData, yshift=-2cm]            {Reset Data};
+		        \node (checkStep)    [decision, below of = reset]                           {Enough Timesteps};
+
+		        \draw [arrow] (checkData)      -|                                           (error);
+		        \draw [arrow] (error)          -- node[anchor=south] {Yes}                  (checkPoint);
+		        \draw [arrow] (error)          -| node[anchor=south east] {No}              (backup);
+		        \draw [arrow] (checkPoint)     -| node[anchor=south west] {No}              (restore);
+		        \draw [arrow] (checkPoint)     |- node[anchor=north west] {Yes}             (reset);
+		        \draw [arrow] (backup)         --                                           (checkStep);
+		        \draw [arrow] (reset)          --                                           (checkStep);
+		        \draw [arrow] (restore)        --                                           (checkStep);
+
+		    \end{tikzpicture}
+		\end{center}
+
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Restart Script}
+
+		\newcommand{\distance}{2.5cm}
+
+		\begin{center}
+		    \begin{tikzpicture}[node distance=1.5cm]
+
+				\node (init)         [init]                                                 {Start / Initialize};
+		        \node (startLoop)    [startstop, below of = init]                           {Loop};
+		        \node (checkStatus)  [process, below of=startLoop]                          {...};
+		        \node (checkStep)    [decision, below of = checkStatus]                     {Enough Timesteps};
+		        \node (end)          [startstop, below of=checkStep, xshift=4cm]            {End};
+		        \node (restart)      [init, below of=checkStep, xshift=-4cm]                {Start/Restart};
+
+		        \draw [arrow] (init)           --                                           (startLoop);
+		        \draw [arrow] (startLoop)      --                                           (checkStatus);
+		        \draw [arrow] (checkStep)      -| node[anchor=south east] {No}              (restart);
+		        \draw [arrow] (checkStep)      -| node[anchor=south west] {Yes}             (end);
+		        \draw [arrow] (restart)        -- + (-3,0) |-                               (startLoop);
+		        \draw [arrow] (end)            -- + ( 3,0) |-                               (init);
+
+		    \end{tikzpicture}
+		\end{center}
+
+	\end{frame}
+
+	\section*{Results}
+
+	\begin{frame}
+		\frametitle{Variation of Computational Resolution}
+		\pause
+		\textbf{Goals:}
+		\begin{itemize}
+			\item Estimate the minimal resolution without numerical dissipation
+			\item Reduce \textbf{calculation time} and \textbf{costs} of the simulation
+		\end{itemize}
+		\pause
+		\textbf{Criteria:}
+		\begin{enumerate}
+			\item[\textbf{(1)}] Subdued turbulence after \textbf{short} time periods
+			\item[\textbf{(2)}] Stability for \textbf{long} time periods 
+		\end{enumerate}
+		\pause
+		\textbf{Verification:}
+		\begin{enumerate}
+			\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
+			\item Reduce to known the minimum number of grid points to verify result in general.
+		\end{enumerate}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Benchmark}
+		\begin{center}
+			\only<2>{\includegraphics[width = 0.8\paperwidth]{S6_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux.pdf}}
+
+			\only<3>{\includegraphics[width = 0.8\paperwidth]{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Reduction of Grid Points}
+
+		\begin{center}
+			\only<2>{\includegraphics[width = 0.8\paperwidth]{Comparison/Resolution/Nvpar/S6_rlt6.0_boxsize1x1_Ns16_Nvpar16-32-48-64_Nmu9_eflux_comparison.pdf}}
+
+			\only<3>{\includegraphics[width = 0.8\paperwidth]{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu6-9_eflux_comparison.pdf}}
+
+			\only<4>{\includegraphics[width = 0.8\paperwidth]{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar48_Nmu6-9_eflux_comparison.pdf}}
+
+			\only<5>{\includegraphics[width = 0.8\paperwidth]{Comparison/Resolution/Ns/S6_rlt6.0_boxsize1x1_Ns12-16_Nvpar64_Nmu9_eflux_comparison.pdf}}
+
+			\only<6>{
+				\textbf{Final Resolution}\\
+				\bigskip
+				\begin{tabular}{l | ccccc | ccccc | c | cc}
+					& $N_\mathrm{m}$ & $N_x$ & $N_\mathrm{s}$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_\mathrm{d}$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
+					\hline
+					S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
+				\end{tabular}
+			}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Size Convergence of $\exb$ Staircase Pattern}
+		\begin{center}
+			\only<2>{\textbf{(1) Radial} \\[0.5cm] $\NR \times \NB \in [1\times1,~2\times1,~3\times1,~4\times1]$}
+
+			\only<3>{\includegraphics[width = 0.8\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}}
+
+			\only<4>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
+
+			\only<5>{\textbf{(2) Isotropic} \\[0.5cm] $\NR \times \NB \in [1\times1,~1.5\times1.5,~2\times2,~2.5\times2.5,~3\times3]$}
+
+			\only<6>{\includegraphics[width = 0.8\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}}
+
+			\only<7>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize1x1-2x2-3x3_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
+
+			\only<8>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_cosine2-cosine5-noise_wexb_comparison.pdf}}
+
+			\only<9>{\textbf{(3) Binormal} \\[0.5cm] $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$}
+
+			\only<10>{\includegraphics[width = 0.8\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}}
+
+			\only<11>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
+		\end{center}
+
+		\only<12>{
+			$\Rightarrow$ Mesoscale pattern size of $\sim 57 - 76\,\rhoth$ 
+			\pause
+			That means profile effects set 
+			\pause
+			\begin{itemize}
+				\item[(i)] the radial size of the $\exb$ staircase pattern
+				\pause
+				\item[(ii)] the scale of avalanche-like transport events.
+			\end{itemize}
+		}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{The finite Heat Flux Threshold}
+		\begin{center}
+			\onslide<2->\includegraphics[width = 0.8\paperwidth]{Comparison/Gradient-Length/S6_rlt6.0-6.2-6.4_boxsize3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}
+
+			\onslide<3->$\Rightarrow~\boxed{\rlt|_\mathrm{finite} = 6.3 \pm 0.1}$
+		\end{center}
+	\end{frame}
+
+	\section*{Conclusion}
+
+	\begin{frame}
+		\frametitle{Conclusion}
+
+		\begin{minipage}{0.4\paperwidth}\raggedleft
+			\begin{itemize}
+				\item <2-> Parallel velocity $\Nvpar$ could be reduced from 64 to 48, which halfed the time until suppression of turbulence
+				\item <3-> Restart Script with \texttt{python} led to further convenience during the task of performing simulations
+				\item <4-> Mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters
+				\item <5-> Finite heat flux threshold is located at $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$
 			\end{itemize}
 		\end{minipage}
-		\begin{minipage}{0.3\textwidth}
-			\begin{figure}%[4]
-				\href{https://www.youtube.com/watch?v=O6tUHzfj-zM}{\includegraphics[width=\textwidth]{Presentation/turbulence.png}}
-			\end{figure}
-			{\tiny Simulation eines Plasmaschlauch von Jeff Candy und Ron Waltz mit Code GYRO}
+		\begin{minipage}{0.4\paperwidth}
+			\onslide<4-> \includegraphics[width = 0.5\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison_thesis.pdf}
 		\end{minipage}
@@
 
-	\section{Turbulenz im Plasma stabilisieren}
-	\begin{frame}
-		\frametitle{Anpassung der Magnetfeldlinienrichtung}
-		Verdrillen der magnetischen Flussdichte $\vect{B}$ überführt Plasma in stabilen Zustand \\[0.2cm]
-		\includegraphics[scale=0.25]{Presentation/Stabilzation_Mechanism_black.png}
-	\end{frame}
-
-	\begin{frame}
-		\frametitle{Turbulenzen verringern}
-		\begin{minipage}{0.6\textwidth}\raggedleft
-			\begin{itemize}
-				\item Wirbelströmung verursachende Partikel folgen meistens Feldlinien
-				\item Turbulenzen können mit negativer magnetischen Scherung verringert werden
-				\item Verdrehung in Richtung der "Guten Krümmung"
-				%\item Umkehrung der magnetischen Scherung verdreht Wirbelströmung in kleinen Abständen in Richtung der "Guten Krümmung"
-				%\item Umgekehrte magnetische Scherung wird erzeugt durch Quetschung des Magnetfelds bei hohen Plasmadruck
-				\item Durch die Verformung des Plasmas kann sich Scherung lokal verändern
-			\end{itemize}
-		\end{minipage}
-		\begin{minipage}{0.3\textwidth}
-			\includegraphics[scale = 0.15]{Presentation/Magnetic_Shear_black.png}
-		\end{minipage}
-	\end{frame}
-
-	%\begin{frame}
-	%	Wirbelströmung, welche größere Längenskalen in \doublequoted{Schlechte Krümmung} transportieren, werden mit Scherung reduziert bis sie sich komplett unterdrücken.
-	%	\frametitle{Regelung oder Unterdrückung von Turbulenzen}
-	%	\includegraphics[scale = 1.3]{Presentation/Shearing.pdf}
-	%\end{frame}
-
-	\begin{frame}
-		\frametitle{Regelung oder Unterdrückung von Turbulenzen}
-		{\tiny Waltz, Kerbel, Phys. Plasmas 1994 w/ Hammett, Beer, Dorland, Waltz Gyrofluid Eqs., Numerical Tokamak Project, DoE Computational Grand Challenge}
-		\begin{tabular}{c c}
-			\includegraphics[scale = 0.15]{Presentation/Stablized_Plasma_Interaction.png} & \includegraphics[scale = 0.15]{Presentation/Stablized_Plasma_LongScale.png} \\
-			Dominante nichtlineare Interaction  & Zusätzlicher verscherter Zonal Flow \\
-			zwischen turbulenter Wirbelströmung & unterdrückt Turbulenz komplett \\
-			und kontrollierten Zonal Flow & \\
-		\end{tabular}
-	\end{frame}
-
-	%\section{Parallelen zur Atmosphären-Physik}
-	%\begin{frame}
-	%	\frametitle{Zonal Flow auf Jupiter vs. Tokamak Reaktor}	
-	%\end{frame}
-
-	\appendix
-	\section*{Appendix}
-	\begin{frame}
-		\frametitle{Danke für die Aufmerksamkeit!}
-		\includegraphics[scale = 0.3]{Presentation/Fusion_Meme.jpeg} \\
-	\end{frame}
-
 \end{document}

### 2023-06-21 bachelorthesis/23-iontempgrad.tex
#39: Correction Anna-Maria

@@
 \cite{Newins2006}
-The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_\mathrm{T,c}$.
-For nonlinear turbulence with Cyclone Base Case parameters and circular concentric flux-surfaces the critical value is given by $R/L_\mathrm{T,c} \sim 6$.
+The initial perturbation is growing if the normalized temperature gradient overcomes a critical value $R/L_{T,\mathrm{c}}$.
+For nonlinear turbulence with Cyclone Base Case parameters and circular concentric flux-surfaces the critical value is given by $R/L_{T,\mathrm{c}} \sim 6$.
 \cite{Dimits2000, Isliker2010}

### 2023-06-21 bachelorthesis/24-zonalflows.tex
#39: Correction Anna-Maria

@@
 This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}.
-The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG-driven turbulence at higher $\rlt$ compared to the linear threshold.
+The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_{T,\mathrm{c}}$ which leads to a development of ITG-driven turbulence at higher $\rlt$ compared to the linear threshold.
 The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig.

### 2023-06-21 bachelorthesis/43-heatfluxthreshold.tex
#39: Bachelor thesis finished
#39: Correction Anna-Maria

@@
 \end{gather*}
-in accordance to Ref. \citenum{Peeters2016}.
+in accordance with Ref. \citenum{Peeters2016}.
 Although the initial quasi-stationary turbulence in the former case is significantly longer compared to the $\rlt = 6.2$ realization discussed in the second test, a stationary pattern with basic zonal flow mode $\nzf = 3$ establishes.
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57.20 - 76.27\,\rhoth$.
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$.
 
 \includegraphicsHere{Comparison/Gradient-Length/S6_rlt6.0-6.2-6.4_boxsize3x3_Ns16_Nvpar48_Nmu9_eflux_comparison.pdf}{
-	Time traces of the heat conduction coefficient $\chi$ for different gradient lengths $\rlt$
+	Time traces of the heat conduction coefficient $\chi$ for different\\gradient lengths $\rlt$
 }{fig:eflux-6.0-6.2-6.4-comparison}{}

### 2023-06-21 evaluation/compare_eflux.py

@@ -51,7 +51,7 @@ def eflux_compare_plot(files, label, xlim = [0, None], ylim = [0, None], invert_
 	else:
 		ax.legend(handles, labels, ncol = len(files))
 
-'''
+#'''
 # R/L_T 6.0 & 6.3 ===================================================================================================================================
 
 # File import and Create picture folder
@@ -65,14 +65,14 @@ picDir = homepath + 'pictures/Comparison/Gradient-Length/'
 if not os.path.exists(picDir):
 	os.makedirs(picDir)
 
-resolution = [r'$R/L_{\mathrm{T}}$ = 6.0', r'$R/L_{\mathrm{T}}$ = 6.3']
+resolution = [r'$R/L_T$ = 6.0', r'$R/L_T$ = 6.3']
 
-eflux_compare_plot(files, resolution, xlim=(0, 6000), ylim=(0,25))
+eflux_compare_plot(files, resolution, xlim=(0, 6000), ylim=(0,25), invert_legend=False)
 
 plt.savefig(picDir + '/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf', bbox_inches='tight')
 #'''
 
-#'''
+'''
 # R/L_T 6.0 & 6.2 & 6.4  ============================================================================================================================
 
 # File import and Create picture folder
@@ -86,7 +86,7 @@ picDir = homepath + 'pictures/Comparison/Gradient-Length/'
 if not os.path.exists(picDir):
 	os.makedirs(picDir)
 
-resolution = [r'$R/L_{\mathrm{T}}$ = 6.0', r'$R/L_{\mathrm{T}}$ = 6.2',  r'$R/L_{\mathrm{T}}$ = 6.4']
+resolution = [r'$R/L_T$ = 6.0', r'$R/L_T$ = 6.2',  r'$R/L_T$ = 6.4']
 
 error_index = [[None, None],
                [14473,16133],

### 2023-06-21 evaluation/compare_gradient_length_eflux_wexb_max.py

@@ -19,7 +19,7 @@ import zonalflow, h5tools, derivative, plot
 
 plot.parameters(40, (24,16), 300)
 
-'''
+#'''
 # COMPARE: 6.0 & 6.3 ================================================================================================================================
 
 # File import and Create picture folder
@@ -38,7 +38,7 @@ if not os.path.exists(picDir):
 fig, (ax_eflux, ax_wexb_max) = plt.subplots(2, 1) #, sharex=True)
 #fig.suptitle('Comparsion of finite heat flux threshold of $R/L_{\mathrm{T}}$')
 
-resolution = [r'$R/L_{\mathrm{T}}$ = 6.0', r'$R/L_{\mathrm{T}}$ = 6.3']
+resolution = [r'$R/L_T$ = 6.0', r'$R/L_T$ = 6.3']
 
 #ax_eflux.set_title(r'$N_{\mathrm{vpar}}$ = 64; $N_{\mathrm{\mu}}$ = 9', pad=20)
 ax_eflux.set_xlabel(r'$t~[R/ \nu_{\mathrm{th}}]$')
@@ -116,7 +116,7 @@ if not os.path.exists(picDir):
 fig, (ax_eflux, ax_wexb_max) = plt.subplots(1, 2, figsize = (24,6)) #, sharex=True)
 #fig.suptitle('Comparsion of finite heat flux threshold of $R/L_{\mathrm{T}}$')
 
-resolution = [r'$R/L_{\mathrm{T}}$ = 6.0', r'$R/L_{\mathrm{T}}$ = 6.2']
+resolution = [r'$R/L_T$ = 6.0', r'$R/L_T$ = 6.2']
 
 #ax_eflux.set_title(r'$N_{\mathrm{vpar}}$ = 64; $N_{\mathrm{\mu}}$ = 9', pad=20)
 ax_eflux.set_xlabel(r'$t~[R/ \nu_{\mathrm{th}}]$')

### 2023-06-21 presentation/presentation.tex
#39: Presentation finished

@@
 		\frametitle{Gyrokinetics}
-		\pause
+
+		\only<4>{
+			\begin{center}
+				\includegraphics[width=0.85\paperwidth]{Theory/Gyro-Center-Coordinates.pdf}		
+			\end{center}
+		}
+
 		\begin{tabular}{l l l}
 
-			$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0$ &
-			$f(\vect{x}, \vect{v}, t)$ & 
-			Vlasov Equation \pause \\[0.7cm]
+			\onslide<2->{$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0$} &
+			\onslide<2->{$f(\vect{x}, \vect{v}, t)$} & 
+			\onslide<2->{Vlasov Equation} \\[0.7cm]
 
-			\multicolumn{2}{l}{$\Downarrow$ \qquad $\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}$} &
-			Gyrokinetic Ordering \pause \\[0.7cm]
+			\multicolumn{2}{>{\onslide<3->}l<{\onslide}}{$\Downarrow$ \qquad $\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}$} &
+			\onslide<3->{Gyrokinetic Ordering} \\[0.7cm]
 
-			$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0$ &
-			$f(\vect{X}, v_{\parallel}, \mu)~;~\frac{\mathrm{d}\mu}{\mathrm{d}t} = 0$ & 
-			Gyrokinetic Formalism \pause \\[0.7cm]
+			\onslide<5->{$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0$} &
+			\onslide<5->{$f(\vect{X}, v_{\parallel}, \mu)~;~\frac{\mathrm{d}\mu}{\mathrm{d}t} = 0$} & 
+			\onslide<5->{Gyrokinetic Formalism} \\[0.7cm]
 
-			\multicolumn{2}{l}{$\Downarrow$ \qquad $f = f_0 + \delta f$} & 
-			$\delta f$ Approx \& Local Limit \pause \\[0.7cm]
+			\multicolumn{2}{>{\onslide<6->}l<{\onslide}}{$\Downarrow$ \qquad $f = f_0 + \delta f$} & 
+			\onslide<6->{$\delta f$ Approx \& Local Limit} \\[0.7cm]
 
-			\multicolumn{2}{l}{$\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S$} &
-			Gyrokinetic Equation \\[0.7cm]
-			\multicolumn{3}{l}{$g \sim (f_0 + \delta f)\qquad S \sim (f + f_0)\qquad \vect{v}_\mathrm{D} \sim \nabla B \qquad \vect{v}_\chi \sim \exb \qquad f(\xcoord, \ycoord, s, v_\parallel, \mu)$} \\
+			\multicolumn{2}{>{\onslide<7->}l<{\onslide}}{$\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S$} &
+			\onslide<7->{Gyrokinetic Equation} \\[0.7cm]
+			\multicolumn{3}{>{\onslide<7->}l<{\onslide}}{$g \sim (f_0 + \delta f)\qquad S \sim (f + f_0)\qquad \vect{v}_\mathrm{D} \sim \nabla B \qquad \vect{v}_\chi \sim \exb \qquad f(\xcoord, \ycoord, s, v_\parallel, \mu)$} \\
 		\end{tabular}
@@
 
-			\only<9>{\textbf{(3) Binormal} \\[0.5cm] $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$}
+			\only<9>{\includegraphics[height = 0.7\paperheight]{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/noise/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_noise_wexb_dens_ene_selection.pdf}}
 
-			\only<10>{\includegraphics[width = 0.8\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}}
+			\only<10>{\textbf{(3) Binormal} \\[0.5cm] $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$}
 
-			\only<11>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
+			\only<11>{\includegraphics[width = 0.8\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}}
+
+			\only<12>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
 		\end{center}
 
-		\only<12>{
-			$\Rightarrow$ Mesoscale pattern size of $\sim 57 - 76\,\rhoth$ 
-			\pause
-			That means profile effects set 
-			\pause
+		\only<13>{
+			\begin{center}
+				$\Rightarrow$ Mesoscale pattern size of: \\[0.3cm]
+				$\boxed{\sim 57 - 76\,\rhoth}$ \bigskip
 				\begin{itemize}
-				\item[(i)] the radial size of the $\exb$ staircase pattern
-				\pause
-				\item[(ii)] the scale of avalanche-like transport events.
+					\item Non-locality is inherent to ITG-driven turbulence
+					\item Avalanches are spatially organized by the E × B staircase pattern
 				\end{itemize}
+			\end{center}
 		}

### 2023-06-21 pictures/Theory/ITG-Turbulence_Presentation.tex
#39: ITG Turbulence picture redone

% \usetikzlibrary{mindmap,backgrounds}
% \usetikzlibrary{decorations.pathmorphing}
% \usetikzlibrary{decorations.markings}
% \usetikzlibrary{arrows.meta,bending}

\begin{tikzpicture}

    % Temperature background
    % \shade [left color=red, right color= white, rounded corners = 0cm] (0,0) rectangle (6,6);
    % \shade [left color=white, right color= blue, rounded corners = 0cm] (6,0) rectangle (12,6);

    \onslide<2->{
    \shade [left color=red!80!white, right color= red!50!white]  (5,0) {[rounded corners=0.5cm] --++ (-5,0)  --++ (0,6)} --++ (5,0)  -- cycle;
    \shade [left color=red!50!white, right color= white] (4.95,0) --++ (1.05,0)  --++ (0,6) --++ (-1.05,0)  -- cycle;

    \shade [left color=blue!50!white, right color= blue!80!white]  (7,0) {[rounded corners=0.5cm] --++ (5,0)  --++ (0,6)} --++ (-5,0)  -- cycle;
    \shade [left color=white, right color= blue!50!white] (6,0) --++ (1.05,0)  --++ (0,6) --++ (-1.05,0)  -- cycle;

    }

    \onslide<7->{
    % Density and Potencial cells
    \draw [thick, gray, opacity=1, fill = gray!70!white] (4.75,4.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
    \draw [thick, gray!70!white, opacity=1, fill = gray] (4.75,3.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
    \draw [thick, gray, opacity=1, fill = gray!70!white] (4.75,2.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
    \draw [thick, gray!70!white, opacity=1, fill = gray] (4.75,1.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
    }

    \onslide<4->{
    % Temperature cells
    \draw [thick, red, opacity=0.7, fill = red]   (7.25,4) ellipse (2.75 and 0.4) node[white, opacity=1]{$T^{+}$};
    \draw [thick, blue, opacity=0.7, fill = blue] (7.25,3) ellipse (2.75 and 0.4) node[white, opacity=1]{$T^{-}$};
    \draw [thick, red, opacity=0.7, fill = red]   (7.25,2) ellipse (2.75 and 0.4) node[white, opacity=1]{$T^{+}$};
    }

    \onslide<3->{
    % Temperature and magnetic field gradient
    \draw[thick, white, <-] (2,5.2) -- (5,5.2) node[midway, above, white]{$\nabla T~;\nabla B$};

    % Magnetic field
    \draw [thick, white] (9,5.2) circle (0.2);
    \filldraw [white] (9,5.2) circle (1.5pt);
    \draw (9.3,4.9) node[white]{$\vect{B}$};
    }

    \onslide<6->{
    % Density gradient
    \draw[thick, white, <-] (2,0.8) -- (5,0.8) node[midway, below, white]{$\nabla n$};
    }

    \onslide<8->{
    % Electric field
    \draw[thick, white, <-] (3,4.35) -- (3,3.65) node[midway, below, left]{$\vect{E}$};
    \draw[thick, white, <-] (6.5,4.35) -- (6.5,3.65) node[midway, below, left]{$\vect{E}$};
    \draw[thick, white, ->] (3,3.35) -- (3,2.65) node[midway, below, left]{$\vect{E}$};
    \draw[thick, white, ->] (6.5,3.35) -- (6.5,2.65) node[midway, below, left]{$\vect{E}$};
    \draw[thick, white, <-] (3,2.35) -- (3,1.65) node[midway, below, left]{$\vect{E}$};
    \draw[thick, white, <-] (6.5,2.35) -- (6.5,1.65) node[midway, below, left]{$\vect{E}$};
    }

    \onslide<5->{
    % Grad B and curvature drift velocity
    \draw[thick, white, ->] (9.5,4.2) -- (9.5,3.5) node[midway, left, white]{$\vect{v}_\mathrm{d}$};
    \draw[thick, white, ->] (9.5,3)   -- (9.5,2.6) node[midway, left, white]{$\vect{v}_\mathrm{d}$};
    \draw[thick, white, ->] (9.5,2.2) -- (9.5,1.5) node[midway, left, white]{$\vect{v}_\mathrm{d}$};
    }

    \onslide<9->{
    % ExB drift velocity
    \draw[thick, white, <-] (10.5,5) -- (11.5,5) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (10.5,4) -- (11.5,4) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (10.5,3) -- (11.5,3) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (10.5,2) -- (11.5,2) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (10.5,1) -- (11.5,1) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (0.5,5)  --  (1.5,5) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (0.5,4)  --  (1.5,4) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (0.5,3)  --  (1.5,3) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, ->] (0.5,2)  --  (1.5,2) node[midway, below, white]{$\vect{v}_{E}$};
    \draw[thick, white, <-] (0.5,1)  --  (1.5,1) node[midway, below, white]{$\vect{v}_{E}$};
    }

\end{tikzpicture}

### 2023-06-21 presentation/presentation.tex
#39: ITG Turbulence picture redone

@@
 
-\title{Size convergence of the E$\times$B \\ Staircase pattern in flux tube \\ simulations of ion temperature \\ gradient driven turbulence}
+\title{Size Convergence of the E$\times$B \\ Staircase Pattern in Flux Tube \\ Simulations of Ion t
+Temperature \\ Gradient-Driven Turbulence}
 \subtitle{~}
@@
 			\begin{itemize}
-				\item <2-> Ion temperature gradient driven turbulence close to marginal stability exhibits Zonal flow pattern formation on mesoscales 
+				\item <2-> Ion temperature gradient-driven turbulence close to marginal stability exhibits Zonal flow pattern formation on mesoscales 
 				\item <3-> Patterns are called $\exb$ staircase structures
@@
 	\begin{frame}
-		\frametitle{Ion Temperature Gradient (ITG) Driven Instability}
+		\frametitle{Ion Temperature Gradient (ITG)-driven Instability}
 		\begin{center}
-			\onslide<2->\input{../pictures/Theory/ITG-Turbulence.tex}
+			\onslide<2->\input{../pictures/Theory/ITG-Turbulence_Presentation.tex}
 		\end{center}
@@
 				\item <3-> Restart Script with \texttt{python} led to further convenience during the task of performing simulations
-				\item <4-> Mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG driven turbulence for Cyclone Base Case parameters
+				\item <4-> Mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG-driven turbulence for Cyclone Base Case parameters
 				\item <5-> Finite heat flux threshold is located at $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$

### 2023-06-21 bachelorthesis/23-iontempgrad.tex
#39: ITG Chapter corrected

@@
 \newpage
-The described mechanism causes perturbation on the outboard low field side of the tokamak this region has a \enquote{bad curvature} and instability happens there.
+The described mechanism causes perturbation on the outboard low field side of the tokamak.
+This region is also referred to as \enquote{bad curvature} because instability happens there.
 On the inboard high field side the opposing $\nabla B$ and $\nabla T$ results in a decay of perturbation which is referred to as \enquote{good curvature}.

### 2023-06-22 presentation/presentation.tex
#39: Minor correction and animations for presentation

@@
 
-\title{Size Convergence of the E$\times$B \\ Staircase Pattern in Flux Tube \\ Simulations of Ion t
+\title{Size Convergence of the E$\times$B \\ Staircase Pattern in Flux Tube \\ Simulations of Ion
 Temperature \\ Gradient-Driven Turbulence}
@@
 
-		Properties:
+		\onslide<3-> Properties:
 		\begin{enumerate}
-			\item[(1)] <3-> They occur on a \textit{radial mesoscale} of order of $\rho_\mathrm{L} < 10\,\rhoth < R$
-			\item[(2)] <4-> The structure are \textit{quasi stationary} in space and vary on time scales much longer than the typical turbulence time scales.
-			\item[(3)] <5-> They have a \textit{typical amplitude} of the order of $10^{-1} \vth/R$.
-			\item[(4)] <6-> Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate.
+			\item[(1)] <4-> They occur on a \textit{radial mesoscale} of order of $\rho_\mathrm{L} < 10\,\rhoth < R$
+			\item[(2)] <5-> The structure are \textit{quasi stationary} in space and vary on time scales much longer than the typical turbulence time scales.
+			\item[(3)] <6-> They have a \textit{typical amplitude} of the order of $10^{-1} \vth/R$.
+			\item[(4)] <7-> Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate.
 		\end{enumerate}
-		\onslide<7->{
+		\onslide<8->{
 		\begin{center}
@@
 			\end{itemize}
-			\onslide<9->{
+			\onslide<4->{
 			\bigskip
@@
 			\onslide<2-> $\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord)$ \\[0.5cm]
-			\onslide<3-> $\kzf = 2\pi \nzf/L_\xcoord$ \\[0.5cm]
-			\onslide<4-> $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$ \\[0.5cm]
+			\onslide<2-> $\kzf = 2\pi \nzf/L_\xcoord$ \\[0.5cm]
+			\onslide<2-> $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$ \\[0.5cm]
 			\begin{itemize}
-				\item <5-> Zonal flow mode that dominates the $\exb$ staircase pattern are called \textbf{basic mode} 
-				\item <6-> The basic mode exhibits the maximum amplitude in the spectrum $\hatwexbamp$
+				\item <3-> Zonal flow mode that dominates the $\exb$ staircase pattern are called \textbf{basic mode} 
+				\item <4-> The basic mode exhibits the maximum amplitude in the spectrum $\hatwexbamp$
 			\end{itemize}

### 2023-06-22 presentation/presentation.tex
#39: Thermal radius in presentation

@@
 				\item <5-> Cyclone Base Case (CBC) parameters $\rlt = 6.0$ 
-				\item <6-> Turbulence level is quantified by the turbulent heat conduction coefficient $\chi$ (normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$)) 
+				\item <6-> Turbulence level is quantified by the turbulent heat conduction coefficient $\chi$ (normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ and $\rhoth = \frac{m \vth}{|q|B}$)) 
 				\item <7-> Quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities

### 2023-06-23 bachelorthesis/04-declaration.tex
#39: Link to Repository in Declaration

@@
 It provides additional plots and paragraphs that were included in the publication.
-The brief communication can be found in Appendix \ref{append:paper}.
+The brief communication can be found in Appendix \ref{append:paper}.\bigskip
+
+Additionally, the author states that every information, except data, regarding this thesis can be found under the GitHub Repository with the link \url{https://github.com/ManeLippert/Bachelorthesis-Shearingrate-Convergence}.

### 2023-06-23 presentation/presentation.tex
#39: Declutter Text on presentation

@@
 			\begin{itemize}
-				\item <2-> Ion temperature gradient-driven turbulence close to marginal stability exhibits Zonal flow pattern formation on mesoscales 
+				\item <2-> Zonal flow pattern formation on mesoscales in Ion temperature gradient-driven turbulence 
 				\item <3-> Patterns are called $\exb$ staircase structures
+				\item <4-> Pattern formation observed in \textbf{local/global flux/gradient-driven} studies
 			\end{itemize}
@@
 		\bigskip
-		\onslide<4-> Observed pattern formation in
 		\begin{center}
 			\begin{itemize}
-				\item <5-> local gradient-driven flux-tube simulations, including collisions and background $\exb$ shear,
-				\item <6-> local flux-driven realizations, including mean electric field shear,
-				\item <7-> global gradient-driven and global flux-driven studies.
+				\item <5-> \textbf{Global studies:} Multiple radial repetitions of staircase structures with a typical pattern size of several ten Larmor radii
+				\item <6-> \textbf{Local studies:} Radial size of $\exb$ staircase structures always converge to the radial box size
 			\end{itemize}
 			\bigskip
-			\onslide<8-> \textbf{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}
+			\onslide<7-> \textbf{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}
 		\end{center}
@@
 		\end{center}
+	\end{frame}
 
-	\end{frame}
 	\begin{frame}
@@
 		\begin{enumerate}
-			\item[(1)] <4-> They occur on a \textit{radial mesoscale} of order of $\rho_\mathrm{L} < 10\,\rhoth < R$
-			\item[(2)] <5-> The structure are \textit{quasi stationary} in space and vary on time scales much longer than the typical turbulence time scales.
-			\item[(3)] <6-> They have a \textit{typical amplitude} of the order of $10^{-1} \vth/R$.
-			\item[(4)] <7-> Turbulent transport occurs in \textit{avalanches} with propagation strongly linked to the local $\exb$ shearing rate.
+			\item[(1)] <4-> \textit{Radial mesoscale} of order of $\rho_\mathrm{L} < 10\,\rhoth < R$
+			\item[(2)] <5-> \textit{Quasi stationary} in space 
+			\item[(3)] <6-> Time scales much longer than the typical turbulence time scales
+			\item[(4)] <7-> \textit{Typical amplitude} of the order of $10^{-1} \vth/R$.
+			\item[(5)] <8-> Turbulent transport strongly linked to the local $\exb$ shearing through \textit{avalanches}
 		\end{enumerate}
-		\onslide<8->{
+		\onslide<9->{
 		\begin{center}
@@
 			\begin{itemize}
-				\item <3-> Zonal flow mode that dominates the $\exb$ staircase pattern are called \textbf{basic mode} 
+				\item <3-> Zonal flow mode that dominates the staircase pattern are called \textbf{basic mode} 
 				\item <4-> The basic mode exhibits the maximum amplitude in the spectrum $\hatwexbamp$
@@
 			\only<12>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
-		\end{center}
 
-		\only<13>{
-			\begin{center}
+			\only<14>{
+				\textbf{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}\bigskip
+
 				$\Rightarrow$ Mesoscale pattern size of: \\[0.3cm]
 				$\boxed{\sim 57 - 76\,\rhoth}$ \bigskip
+
 				\begin{itemize}
 					\item Non-locality is inherent to ITG-driven turbulence
-					\item Avalanches are spatially organized by the E × B staircase pattern
+					\item Avalanches are spatially organized by the $\exb$ staircase pattern
 				\end{itemize}
+			}
 		\end{center}
-		}
 	\end{frame}

### 2023-06-23 bachelorthesis/21-motion.tex
#37, #39: Redo circular motion picture and some grammar

@@
         \begin{tabular}{l l}
-            \textbf{(a)} & Lorentz force $\vect{F}_{\perp, B}$ perpendicular to velocity $\vect{v}$ which causes,\\
-                         & circular motion with Lamor radius $\rho_\mathrm{L}$ and cyclotron frequency $\omega_\mathrm{c}$ \\
-            \textbf{(b)} & Electric force $\vect{F}_{\parallel,E}^\pm$,\\
+            \textbf{(a)} & Lorentz force $\vect{F}_{\perp, B}$ perpendicular to velocity $\vect{v}^\pm$ and \\
+                         & magnetic field $\vect{B}$ which causes, circular motion with different directions for \\
+                         & electron and ions, Lamor radius $\rho_\mathrm{L}$ and cyclotron frequency $\omega_\mathrm{c}$, \\
+            \textbf{(b)} & Electric force $\vect{F}_{\parallel,E}^\pm$ with electric field $\vect{E}$,\\
             \textbf{(c)} & Mirror effect with force $\vect{F}_{\parallel,\nabla_{\!\parallel} B}$ and magnetic moment $\mu$ \\

### 2023-06-23 briefcommunication/briefcommunication.tex
#37: Fixed spacing
#37, #39: Redo circular motion picture and some grammar

@@
 
-\title[Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
-{Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}
+\title[Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient-driven turbulence]
+{Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient-driven turbulence}
 
@@
 	\captionsetup{type=figure}
-	\includegraphics[width=0.85\textwidth]{Briefcommunication/1.pdf}
+	\includegraphics[width=0.\textwidth]{Briefcommunication/1.pdf}
 	\captionof{figure}{
@@
 Again, the $\nzf = 1$ (box scale) zonal flow mode does not grow secularly during the entire turbulent phase.
-Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$. \bigskip
+Also, this test confirms the statistical soundness of the converged pattern size of $\sim 57 - 76\,\rhoth$.\\
+
+\bigskip
 

### 2023-06-23 pictures/Theory/Lorentz-Force.tex
#37, #39: Redo circular motion picture and some grammar

@@
     % Circle
-    \draw (0,0) circle (1.5);
+    \draw[lightgray] (0,0) circle (1.5);
     \filldraw (0,0) circle (1.5pt);
@@
     % Velocity
-    \draw[thick,Notaorange, ->] (x) -- ++(180:1cm) node[above, Notaorange]{$\vect{v}$};
+    \draw[thick,Notaorange, ->] (x) -- ++(180:1cm) node[left, Notaorange]{$\vect{v}^-$};
+    \draw[thick,Notaorange, ->] (0,-1.5) -- ++(180:1cm) node[left, Notaorange]{$\vect{v}^+$};
     % Cyclontron freqency
-    \draw[thick, ->, Notared] (200:1.1*1.5) arc(200:240:1.1*1.5);
-    \draw (240:1.4*1.5) node[Notared] {$\omega_\mathrm{c} = \frac{|q|B}{m}$};
-    % Particle
-    \filldraw [Notablue] (x) circle (2pt);
-    \draw (x) node[above right, Notablue]{Particle};
+    \draw[thick, ->, blue] (90:1.0*1.5) arc(90:150:1.0*1.5);
+    \draw[thick, <-, red] (210:1.0*1.5) arc(210:270:1.0*1.5);
+    \draw (50:1.5*1.5) node[black] {$\omega_\mathrm{c} = \frac{|q|B}{m}$};
+    % Charges
+    \draw [fill=blue, blue] (x) circle (0.2) node[white, font=\boldmath]{$-$};
+    \draw [fill=red, red] (0,-1.5) circle (0.2) node[white, font=\boldmath]{$+$};
     % Magnetic field

### 2023-06-24 bachelorthesis/24-zonalflows.tex
#39: Rearangements and minor corrections

--- 2023-06-20 bachelorthesis/24-zonalflows.tex
@@
 This causes a shear deformation of turbulent structures which is considered to play a significant role in the suppression of turbulence \cite{Biglari1990, Dimits2000}, especially the ion temperature gradient driven turbulence \cite{Nakata2012, Makwana2014, Maeyama2014, Whelan2018, Whelan2019}.
-The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_\mathrm{T,c}$ which leads to a development of ITG-driven turbulence at higher $\rlt$ compared to the linear threshold.
 The shearing process tilts and stretches turbulent eddies, caused by the shear flow, until they break into smaller eddies with reduced radial correlation length [Fig.
@@
 Turbulence will then be reduced because of this process.
-\cite{Biglari1990,Diamond2005,Burnell1997} \bigskip
+\cite{Biglari1990,Diamond2005,Burnell1997} The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_{T,\mathrm{c}}$ which leads to a development of ITG-driven turbulence at higher $\rlt$ compared to the linear threshold.
+\bigskip
 
@@
 The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}.
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ion temperature gradient (ITG)-driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
 \\ \bigskip

### 2023-06-24 bachelorthesis/25-gyrokinetics.tex
#39: Rearangements and minor corrections

@@
 The perpendicular component $k_\perp$ is of the order of the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the scale on which the equilibrium density $n_0$ varies, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$.
-The gradient length compares with the machine size $R$ which leads to the normalized Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak.
+The gradient length compares with the machine size $R$ which leads to the normalized thermal Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak.
 The parallel component $k_\parallel$ on the other hand scales with the machine size $R$.
@@
 \includegraphicsHere{Theory/Gyro-Center-Coordinates.pdf}{
-	Sketch of gyrocenter coordinates where the charged particle performs a circular motion around the gyrocenter. \cite{Krommes2000}
+	Sketch of guidingcenter coordinates where the charged particle performs a circular motion around the gyrocenter. \cite{Krommes2000}
 }{fig:gyrocenterCoords}{0.7}
@@
 The source term $S$ contains the distribution function in the equilibrium $f_0$ and a correction term for the collisions as well as the energy injection term.
-Due to prediodic boundary conditions in the radial direction the radial averaged gradients are fixed in time which is also referred to as \textit{gradient driven} approach.
+Due to periodic boundary conditions in the radial direction the radial averaged gradients are fixed in time which is also referred to as \textit{gradient-driven} approach.
 \cite{Peeters2009}

### 2023-06-24 bachelorthesis/42-boxsizescan.tex
#39: Correction Anna-Maria
#39: Rearangements and minor corrections

@@
 The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} [$\gamma$ is the growth rate of the most unstable linear ion temperature gradient (ITG)-driven Eigenmode], connected by steep flanks where $\wexb$ crosses zero.
 Figure~\ref{fig:wexb-1-2-3-4x1-compariso} or Figure~\ref{fig:wexb-stable-comparison}(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
@@
 		\textbf{(c) binormal:}  & $t_{3\times 1.5}$ & $\in [2000, 3000]$,   & $t_{3\times 2.5}$ & $\in [2000, 3000]$,   \\
-		                        & $t_{3\times 3}$   & $\in [2000, 3000]$    & $t_{3\times 5}$   & $\in [1000, 3000]$.    \\
+		                        & $t_{3\times 3}$   & $\in [2000, 3000]$,   & $t_{3\times 5}$   & $\in [1000, 3000]$.   \\
 	\end{tabular}

### 2023-06-25 bachelorthesis/24-zonalflows.tex
#39: Sheraing rate plot in zonal flow chapter and presentation

@@
 The $\exb$ shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}.
-This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ion temperature gradient (ITG)-driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero.
+This type of structure is characterized by intervals of almost constant shear with alternating sign satisfying the Waltz criterion $|\wexb| \approx \gamma$\cite{Waltz1994, Waltz1998} ($\gamma$ is the growth rate of the most unstable linear ion temperature gradient (ITG)-driven Eigenmode), connected by steep flanks where $\wexb$ crosses zero [Fig.
+\ref{fig:shearingrate}].
 \\ \bigskip
@@
 \cite{Peeters2016,Weikl2017}
+
+\includegraphicsHere{Theory/Shearing-Rate.pdf}{
+	Fully- and partially-developed pattern formation in the radial profile of the $\exb$ shearing rate as temporally averaged radial profiles of the $\exb$ shearing rate in near marginal ion temperature gradient-driven turbulence with adiabatic electrons and the growth rate $\pm\gamma$ of the most unstable linear ITG-driven Eigenmode.
+}{fig:shearingrate}{0.5}

### 2023-06-25 pictures/Theory/ITG-Turbulence.tex
#39: Corrected ITG Plot

@@
 \begin{tikzpicture}
+
     % Temperature background
-    \shade [left color=red, right color= blue, rounded corners = 0.5cm, opacity = 0.7] (0,0) rectangle (12,6);
+    % \shade [left color=red, right color= white, rounded corners = 0cm] (0,0) rectangle (6,6);
+    % \shade [left color=white, right color= blue, rounded corners = 0cm] (6,0) rectangle (12,6);
+
+    \shade [left color=red!90!white, right color= red!50!white]  (5,0) {[rounded corners=0.5cm] --++ (-5,0)  --++ (0,6)} --++ (5,0)  -- cycle;
+    \shade [left color=red!50!white, right color= white] (4.95,0) --++ (1.05,0)  --++ (0,6) --++ (-1.05,0)  -- cycle;
+
+    \shade [left color=blue!50!white, right color= blue!80!white]  (7,0) {[rounded corners=0.5cm] --++ (5,0)  --++ (0,6)} --++ (-5,0)  -- cycle;
+    \shade [left color=white, right color= blue!50!white] (6,0) --++ (1.05,0)  --++ (0,6) --++ (-1.05,0)  -- cycle;
+
     % Magnetic field
-    \draw [thick, white] (8,0.8) circle (0.2);
-    \filldraw [white] (8,0.8) circle (1.5pt);
-    \draw (8.3,1.1) node[white]{$\vect{B}$};
+    \draw [thick, white] (9,5.2) circle (0.2);
+    \filldraw [white] (9,5.2) circle (1.5pt);
+    \draw (9.3,4.9) node[white]{$\vect{B}$};
+
     % Density and Potencial cells
-    \draw [thick, gray, opacity=1, fill = gray] (4.5,4.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
-    \draw [thick, gray, opacity=1, fill = gray] (4.5,3.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
-    \draw [thick, gray, opacity=1, fill = gray] (4.5,2.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
-    \draw [thick, gray, opacity=1, fill = gray] (4.5,1.5) ellipse (2.5 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
+    \draw [thick, gray, opacity=1, fill = gray!70!white] (4.75,4.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
+    \draw [thick, gray!70!white, opacity=1, fill = gray] (4.75,3.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
+    \draw [thick, gray, opacity=1, fill = gray!70!white] (4.75,2.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{-}~;~n^{-}$};
+    \draw [thick, gray!70!white, opacity=1, fill = gray] (4.75,1.5) ellipse (2.75 and 0.4) node[white, opacity=1]{$\phi^{+}~;~n^{+}$};
+
     % Temperature cells
-    \draw [thick, red, opacity=1, fill = red]   (7.5,4) ellipse (2.5 and 0.4) node[white, opacity=1]{$T^{+}$};
-    \draw [thick, blue, opacity=1, fill = blue] (7.5,3) ellipse (2.5 and 0.4) node[white, opacity=1]{$T^{-}$};
-    \draw [thick, red, opacity=1, fill = red]   (7.5,2) ellipse (2.5 and 0.4) node[white, opacity=1]{$T^{+}$};
-    % Temperature and density gradient
-    \draw[thick, white, <-] (3,0.8) -- (6,0.8) node[midway, below, white]{$\nabla n~;\nabla T$};
-    % Outline Density and Potencial cell
-    \draw [thick, lightgray, dashed] (4.5,4.5) ellipse (2.5 and 0.4);
-    \draw [thick, lightgray, dashed] (4.5,3.5) ellipse (2.5 and 0.4);
-    \draw [thick, lightgray, dashed] (4.5,2.5) ellipse (2.5 and 0.4);
-    \draw [thick, lightgray, dashed] (4.5,1.5) ellipse (2.5 and 0.4);
+    \draw [thick, red, opacity=0.7, fill = red]   (7.25,4) ellipse (2.75 and 0.4) node[white, opacity=1]{$T^{+}$};
+    \draw [thick, blue, opacity=0.7, fill = blue] (7.25,3) ellipse (2.75 and 0.4) node[white, opacity=1]{$T^{-}$};
+    \draw [thick, red, opacity=0.7, fill = red]   (7.25,2) ellipse (2.75 and 0.4) node[white, opacity=1]{$T^{+}$};
+
+    % Temperature and magnetic field gradient
+    \draw[thick, white, <-] (2,5.2) -- (5,5.2) node[midway, above, white]{$\nabla T~;\nabla B~;\nabla n$};
+
+    % Density gradient
+    %\draw[thick, white, <-] (2,0.8) -- (5,0.8) node[midway, below, white]{$\nabla n$};
+
+    % Electric field
+    \draw[thick, white, <-] (3,4.35) -- (3,3.65) node[midway, below, left]{$\vect{E}$};
+    \draw[thick, white, <-] (6.5,4.35) -- (6.5,3.65) node[midway, below, left]{$\vect{E}$};
+    \draw[thick, white, ->] (3,3.35) -- (3,2.65) node[midway, below, left]{$\vect{E}$};
+    \draw[thick, white, ->] (6.5,3.35) -- (6.5,2.65) node[midway, below, left]{$\vect{E}$};
+    \draw[thick, white, <-] (3,2.35) -- (3,1.65) node[midway, below, left]{$\vect{E}$};
+    \draw[thick, white, <-] (6.5,2.35) -- (6.5,1.65) node[midway, below, left]{$\vect{E}$};
+
     % Grad B and curvature drift velocity
@@
     \draw[thick, white, ->] (9.5,2.2) -- (9.5,1.5) node[midway, left, white]{$\vect{v}_\mathrm{d}$};
+
     % ExB drift velocity
@@
     \draw[thick, white, <-] (0.5,1)  --  (1.5,1) node[midway, below, white]{$\vect{v}_{E}$};
+
 \end{tikzpicture}

### 2023-06-25 presentation/presentation.tex
#39: Minor corrections and notes for slides
#39: Sheraing rate plot in zonal flow chapter and presentation
#39: Shorter presentation with less theory

@@
 
-	%\begin{frame}{Gliederung}
-	%	\tableofcontents[hideallsubsections]
-	%\end{frame}
-
 	\section*{Motivation}
@@
 	\begin{frame}
-		\frametitle{Magnetic Confinement in Tokamak}
-		\begin{center}
-			\begin{tabular}{>{\onslide<2->}c<{\onslide} >{\onslide<2->}c<{\onslide}}
-				\includegraphics[width=0.6\paperwidth]{Theory/Tokamak-Torus.pdf} & $\beta = \frac{nT}{\mu_0 B^2/2}$
-			\end{tabular}
-		\end{center}
-	\end{frame}
-
-	\begin{frame}
 		\frametitle{Ion Temperature Gradient (ITG)-driven Instability}
@@
 	\begin{frame}
-		\frametitle{Zonal Flows}
+		\frametitle{Zonal Flows \& Shearing Rate}
+
 		\begin{center}
-			\onslide<2->\includegraphics[width = 0.8\paperwidth]{Theory/Zonal-Flow-Generation.pdf}
-			%\begin{itemize}
-			%	\item <3-> Zonal flows are linked to the $\exb$ flows
-			%	\item <4-> Zonal flow is tangential to the flux-surfaces and does not contribute to the turbulent transport across the flux-surfaces
-			%\end{itemize}
+			\only<2>{\includegraphics[width = 0.8\paperwidth]{Theory/Zonal-Flow-Generation.pdf}}
+
+			\only<3>{
+				\begin{tabular}{c c}
+					\includegraphics[height = 0.75\paperheight]{Theory/Shearing-Rate.pdf} &
+					$\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}$\\
+				\end{tabular}
+			}
 		\end{center}
-	\end{frame}
-
-	\begin{frame}
-		\frametitle{Shearing Rate}
-
-		\onslide<2-> Pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. \\\bigskip
-
-		\onslide<3-> Properties:
-		\begin{enumerate}
-			\item[(1)] <4-> \textit{Radial mesoscale} of order of $\rho_\mathrm{L} < 10\,\rhoth < R$
-			\item[(2)] <5-> \textit{Quasi stationary} in space 
-			\item[(3)] <6-> Time scales much longer than the typical turbulence time scales
-			\item[(4)] <7-> \textit{Typical amplitude} of the order of $10^{-1} \vth/R$.
-			\item[(5)] <8-> Turbulent transport strongly linked to the local $\exb$ shearing through \textit{avalanches}
-		\end{enumerate}
-		\onslide<9->{
-		\begin{center}
-			\begin{tabular}{c c}
-				$\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}$ & $\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)$ \\
-			\end{tabular}
-		\end{center}
-		}
-	\end{frame}
-
-	%\begin{frame}
-	%	\frametitle{Vlasov Equation}
-	%	\begin{gather*}
-	%		\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0
-	%	\end{gather*}
-	%	\begin{gather*}
-	%		n = \int \mathrm{d}\vect{v}\,f(\vect{x}, \vect{v}, t) \qquad j = q \int \mathrm{d}\vect{v}\, v f(\vect{x}, \vect{v}, t)
-	%	\end{gather*}
-	%	\begin{gather*}
-	%		\begin{aligned}
-	%			\nabla \cdot \vect{B} &= 0 &\qquad \nabla \times \vect{B} &= \mu_0\left( \sum_\mathrm{species} j + \epsilon_0 \frac{\partial \vect{E}}{\partial t} \right) \\
-	%			\nabla \cdot \vect{E} &= \frac{1}{\epsilon_0} \sum_\mathrm{species} qn &\qquad \nabla \times \vect{E} &= - \frac{\partial \vect{B}}{\partial t}
-	%		\end{aligned}
-	%	\end{gather*}
-	%\end{frame}
-
-	\begin{frame}
-		\frametitle{Gyrokinetics}
-
-		\only<4>{
-			\begin{center}
-				\includegraphics[width=0.85\paperwidth]{Theory/Gyro-Center-Coordinates.pdf}		
-			\end{center}
-		}
-
-		\begin{tabular}{l l l}
-
-			\onslide<2->{$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0$} &
-			\onslide<2->{$f(\vect{x}, \vect{v}, t)$} & 
-			\onslide<2->{Vlasov Equation} \\[0.7cm]
-
-			\multicolumn{2}{>{\onslide<3->}l<{\onslide}}{$\Downarrow$ \qquad $\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}$} &
-			\onslide<3->{Gyrokinetic Ordering} \\[0.7cm]
-
-			\onslide<5->{$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0$} &
-			\onslide<5->{$f(\vect{X}, v_{\parallel}, \mu)~;~\frac{\mathrm{d}\mu}{\mathrm{d}t} = 0$} & 
-			\onslide<5->{Gyrokinetic Formalism} \\[0.7cm]
-
-			\multicolumn{2}{>{\onslide<6->}l<{\onslide}}{$\Downarrow$ \qquad $f = f_0 + \delta f$} & 
-			\onslide<6->{$\delta f$ Approx \& Local Limit} \\[0.7cm]
-
-			\multicolumn{2}{>{\onslide<7->}l<{\onslide}}{$\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S$} &
-			\onslide<7->{Gyrokinetic Equation} \\[0.7cm]
-			\multicolumn{3}{>{\onslide<7->}l<{\onslide}}{$g \sim (f_0 + \delta f)\qquad S \sim (f + f_0)\qquad \vect{v}_\mathrm{D} \sim \nabla B \qquad \vect{v}_\chi \sim \exb \qquad f(\xcoord, \ycoord, s, v_\parallel, \mu)$} \\
-		\end{tabular}
 	\end{frame}
@@
 			\begin{itemize}
+				\item <2-> Simulation setup orientates on Paper from Arthur Peeters in the year 2016
 				\item <2-> Non-linear flux tube version of Gyrokinetic Workshop (GKW) 
@@
 				\item <5-> Cyclone Base Case (CBC) parameters $\rlt = 6.0$ 
-				\item <6-> Turbulence level is quantified by the turbulent heat conduction coefficient $\chi$ (normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ and $\rhoth = \frac{m \vth}{|q|B}$)) 
-				\item <7-> Quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities
-				\item <8-> Standard box size $(L_\xcoord,~L_\ycoord) = (76.3,~89.8)\,\rhoth$
+				\item <6-> Turbulence level is quantified by the turbulent heat conduction coefficient $\chi$
+				%\item <7-> Quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities
+				\item <7-> Standard box size $(L_\xcoord,~L_\ycoord) = (76.3,~89.8)\,\rhoth$
 			\end{itemize}
@@
 				\item <4-> The basic mode exhibits the maximum amplitude in the spectrum $\hatwexbamp$
+				\item <5-> Waltz criterion $|\wexb|\approx\gamma$
 			\end{itemize}
@@
 		\begin{center}
+
+			\only<2>{
 		    \begin{tikzpicture}[node distance=1.5cm]
@@
 		    \end{tikzpicture}
-		\end{center}
-	\end{frame}
+			}
 
-	\begin{frame}
-		\frametitle{Restart Script}
-
-		\newcommand{\distance}{2.5cm}
-
-		\begin{center}
+			\only<3>{
 		    \begin{tikzpicture}[node distance=2cm]
@@
 		    \end{tikzpicture}
-		\end{center}
+			}
 
-	\end{frame}
-
-	\begin{frame}
-		\frametitle{Restart Script}
-
-		\newcommand{\distance}{2.5cm}
-
-		\begin{center}
+			\only<4>{
 		    \begin{tikzpicture}[node distance=1.5cm]
@@
 		    \end{tikzpicture}
+			}
+
 		\end{center}
@@
 
-			\only<9>{\includegraphics[height = 0.7\paperheight]{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/noise/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_noise_wexb_dens_ene_selection.pdf}}
+			\only<9>{\textbf{(3) Binormal} \\[0.5cm] $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$}
 
-			\only<10>{\textbf{(3) Binormal} \\[0.5cm] $\NR \times \NB \in [3\times1.5,~3\times2.5,~3\times3,~3\times5]$}
+			\only<10>{\includegraphics[width = 0.8\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}}
 
-			\only<11>{\includegraphics[width = 0.8\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_comparison_thesis.pdf}}
+			\only<11>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
 
-			\only<12>{\includegraphics[height = 0.5\paperheight]{Comparison/Boxsize/S6_rlt6.0_boxsize3x1-1.5-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison.pdf}}
-
-			\only<14>{
+			\only<12>{
 				\textbf{Does the basic pattern size always converges to the box size, or is there a typical mesoscale size inherent to staircase structures also in a local flux-tube description?}\bigskip
@@
 
+%--------------------------------------------------------------------------
+% Appendix
+%--------------------------------------------------------------------------
+
+	\appendix
+
+	\section*{Theory}
+
+	\begin{frame}
+		\frametitle{Magnetic Confinement in Tokamak}
+		\begin{center}
+			\begin{tabular}{>{\onslide<2->}c<{\onslide} >{\onslide<2->}c<{\onslide}}
+				\includegraphics[width=0.6\paperwidth]{Theory/Tokamak-Torus.pdf} & $\beta = \frac{nT}{\mu_0 B^2/2}$
+			\end{tabular}
+		\end{center}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Shearing Rate}
+
+		\only<10>{
+			\begin{center}
+				\includegraphics[height = 0.75\paperheight]{Theory/Shearing-Rate.pdf}
+			\end{center}
+			\bigskip
+		}
+		\onslide<2-> Pattern formation occurs in the stabilized tokamak plasma the so-called $\exb$ staircase structure. \\\bigskip
+
+		\onslide<3-> Properties:
+		\begin{enumerate}
+			\item[(1)] <4-> \textit{Radial mesoscale} of order of $\rho_\mathrm{L} < 10\,\rhoth < R$
+			\item[(2)] <5-> \textit{Quasi stationary} in space 
+			\item[(3)] <6-> Time scales much longer than the typical turbulence time scales
+			\item[(4)] <7-> \textit{Typical amplitude} of the order of $10^{-1} \vth/R$ (Waltz criterion $|\wexb|\approx\gamma$)
+			\item[(5)] <8-> Turbulent transport strongly linked to the local $\exb$ shearing through \textit{avalanches}
+		\end{enumerate}
+		\onslide<9->{
+		\begin{center}
+			\begin{tabular}{c c}
+				$\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}$ & $\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)$ \\
+			\end{tabular}
+		\end{center}
+		}
+	\end{frame}
+
+	\begin{frame}
+		\frametitle{Gyrokinetics}
+
+		\only<4>{
+			\begin{center}
+				\includegraphics[width=0.85\paperwidth]{Theory/Gyro-Center-Coordinates.pdf}		
+			\end{center}
+		}
+
+		\begin{tabular}{l l l}
+
+			\onslide<2->{$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0$} &
+			\onslide<2->{$f(\vect{x}, \vect{v}, t)$} & 
+			\onslide<2->{Vlasov Equation} \\[0.7cm]
+
+			\multicolumn{2}{>{\onslide<3->}l<{\onslide}}{$\Downarrow$ \qquad $\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}$} &
+			\onslide<3->{Gyrokinetic Ordering} \\[0.7cm]
+
+			\onslide<5->{$\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0$} &
+			\onslide<5->{$f(\vect{X}, v_{\parallel}, \mu)~;~\frac{\mathrm{d}\mu}{\mathrm{d}t} = 0$} & 
+			\onslide<5->{Gyrokinetic Formalism} \\[0.7cm]
+
+			\multicolumn{2}{>{\onslide<6->}l<{\onslide}}{$\Downarrow$ \qquad $f = f_0 + \delta f$} & 
+			\onslide<6->{$\delta f$ Approx \& Local Limit} \\[0.7cm]
+
+			\multicolumn{2}{>{\onslide<7->}l<{\onslide}}{$\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S$} &
+			\onslide<7->{Gyrokinetic Equation} \\[0.7cm]
+			\multicolumn{3}{>{\onslide<7->}l<{\onslide}}{$g \sim (f_0 + \delta f)\qquad S \sim (f + f_0)\qquad \vect{v}_\mathrm{D} \sim \nabla B \qquad \vect{v}_\chi \sim \exb \qquad f(\xcoord, \ycoord, s, v_\parallel, \mu)$} \\
+		\end{tabular}
+	\end{frame}
+
+	\section*{Results}
+
+	\begin{frame}
+		\frametitle{Size Convergence of $\exb$ Staircase Pattern}
+		\begin{center}
+
+			\includegraphics[height = 0.7\paperheight]{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/noise/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_noise_wexb_dens_ene_selection.pdf}
+
+		\end{center}
+	\end{frame}
+
 \end{document}

### 2023-06-25 bachelorthesis/24-zonalflows.tex
#39: Rearange heat flux threshold to zonal flows

@@
 \cite{Biglari1990,Diamond2005,Burnell1997} The nonlinear generation mechanism causes a shift, also known as Dimits shift \cite{Dimits2000}, of the critical temperature gradient $R/L_{T,\mathrm{c}}$ which leads to a development of ITG-driven turbulence at higher $\rlt$ compared to the linear threshold.
-\bigskip
+This results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient also known as \textit{finite heat flux threshold}.
+\cite{Peeters2016,Weikl2017}\bigskip
 
@@
 \cite{Idomura2009,McMillan2009} The fully developed staircase results in almost completely quenched turbulence.
-Its development results in a discontinuous step in the dependency of turbulent fluxes on the temperature gradient also known as \textit{finite heat flux threshold}.
-\cite{Peeters2016,Weikl2017}
 

### 2023-06-29 presentation/presentation.tex
#39: Correction presentation

@@
 		\begin{center}
+			\only<1>{\includegraphics[width=0.6\paperwidth]{Theory/Tokamak-Torus.pdf}}
+
 			\onslide<2->\input{../pictures/Theory/ITG-Turbulence_Presentation.tex}
@@
 			\begin{itemize}
-				\item <2-> Simulation setup orientates on Paper from Arthur Peeters in the year 2016
-				\item <2-> Non-linear flux tube version of Gyrokinetic Workshop (GKW) 
-				\item <3-> Circular concentric flux surfaces 
+				\item <2-> Simulation setup based on Peeters et al. 2016
+				\item <3-> Non-linear flux tube version of Gyrokinetic Workshop (GKW) with field aligned Hamada coordinates
 				\item <4-> Standard resolution with 6th order (S6) 
-				\item <5-> Cyclone Base Case (CBC) parameters $\rlt = 6.0$ 
+				\item <5-> Cyclone Base Case (CBC) parameters with circular concentric flux surfaces and $\rlt = 6.0$ 
 				\item <6-> Turbulence level is quantified by the turbulent heat conduction coefficient $\chi$
@@
 		\begin{enumerate}
-			\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
-			\item Reduce to known the minimum number of grid points to verify result in general.
+			\item Reduce only one number of grid points and look if criteria \textbf{(1), (2)} are satisfied,
+			\item Reduce to the known minimum number of grid points to verify result in general.
 		\end{enumerate}
@@
 				$\Rightarrow$ Mesoscale pattern size of: \\[0.3cm]
-				$\boxed{\sim 57 - 76\,\rhoth}$ \bigskip
+				$\boxed{\sim (57 - 76)\,\rhoth}$ \bigskip
 
@@
 			\begin{itemize}
-				\item <2-> Parallel velocity $\Nvpar$ could be reduced from 64 to 48, which halfed the time until suppression of turbulence
-				\item <3-> Restart Script with \texttt{python} led to further convenience during the task of performing simulations
-				\item <4-> Mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG-driven turbulence for Cyclone Base Case parameters
-				\item <5-> Finite heat flux threshold is located at $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$
+				\item <2-> Restart Script with \texttt{python} led to further convenience during the task of performing simulations
+				\item <3-> Finite heat flux threshold is located at $\rlt|_\mathrm{finite} = 6.3 \pm 0.1$
+				\item <4-> Parallel velocity $\Nvpar$ could be reduced from 64 to 48, which halfed the time until suppression of turbulence
+				\item <5-> Mesoscale pattern size of $\sim (57 - 76)\,\rhoth$ is found to be intrinsic to ITG-driven turbulence for Cyclone Base Case parameters
 			\end{itemize}
@@
 		\begin{minipage}{0.4\paperwidth}
-			\onslide<4-> \includegraphics[width = 0.5\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison_thesis.pdf}
+			\onslide<5-> \includegraphics[width = 0.5\paperwidth]{Comparison/Boxsize/S6_rlt6.0_boxsize1-2-3-4x1-1.5-2-2.5-3-5_Ns16_Nvpar48_Nmu9_wexb_comparison_thesis.pdf}
 		\end{minipage}

### 2023-06-29 bachelorthesis/41-resolution.tex
#39: Reference brief communication and minor corrections

@@
 \begin{enumerate}
-	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
-	\item Reduce to known the minimum number of grid points to verify result in general.
+	\item Reduce only one number of grid points and look if criteria \textbf{(1), (2)} are satisfied
+	\item Reduce to the known minimum number of grid points to verify result in general.
 \end{enumerate}

### 2023-06-30 bachelorthesis/42-boxsizescan.tex
#39: Minor corrections and thermal quantities in presentation

@@
 Each realization exhibits an initial quasi-stationary turbulent phase and a second final \cite{Peeters2016} phase with almost suppressed turbulence [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(a)].
-The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-stable-comparison}. 
-Figure~\ref{fig:wexb-1-2-3-4x1-compariso} or Figure~\ref{fig:wexb-stable-comparison}(a) shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
+The latter state is indicative for the presence of a fully developed staircase pattern as depicted in Fig.~\ref{fig:wexb-1-2-3-4x1-compariso}. 
+Figure~\ref{fig:wexb-1-2-3-4x1-compariso} shows a striking repetition of the staircase structure, with the number of repetitions equal to $\NR$.
 Hence, the basic size of the pattern not only converges with increasing radial box size, but the converged radial size also turns out to at least roughly agree with the standard radial box size of Ref.
@@
 
-Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-stable-comparison} represent the real converged pattern in a statistical sense.
+Due to the lack of a substantial turbulent drive in the final suppressed state no further zonal flow evolution is observed [Fig.~\ref{fig:wexb-eflux-1-2-3-4x1-comparison}(b)] and one might critically ask whether the structures shown in Fig.~\ref{fig:wexb-1-2-3-4x1-compariso} represent the real converged pattern in a statistical sense.
 Note that in the $3 \times 1$ case the initial quasi-stationary turbulent state extends up to a few $\sim 10^4\,R/\vth$.
@@
 As shown in Fig.
-\ref{fig:wexb-1x1-2x2-3x3-comparison} or Fig.~\ref{fig:wexb-stable-comparison}(b) this test also confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
+\ref{fig:wexb-1x1-2x2-3x3-comparison} this test also confirms the convergence of the staircase pattern size to a typical mesoscale that is distinct from the radial box size in the $\NR > 1$ realizations.
 
@@
 By contrast to the radial box size scan the $3\times3$ realization shows a stationary pattern with four repetitions of the fully developed staircase structure, i.e., a somewhat smaller pattern size [Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison}(b), Fig.
-\ref{fig:wexb-1x1-2x2-3x3-comparison} and Fig.
-\ref{fig:wexb-stable-comparison}(b)].
+\ref{fig:eflux-1x1-2x2-3x3-comparison}(b) and Fig.
+\ref{fig:wexb-1x1-2x2-3x3-comparison}].
 Whether this is related to a possible pattern size dependence on the binormal box size or to the competition between patterns with the two sizes $\lambda \in [57,~ 76]\,\rhoth$ as observed in the first test is addressed in the next paragraph.
-The scale of structures developing in the $1.5\times1.5$ and $2.5\times2.5$ realizations (not included in Fig.~\ref{fig:wexb-stable-comparison}, Fig.~\ref{fig:wexb-1x1-2x2-3x3-comparison} and Fig.
-\ref{fig:eflux-1x1-2x2-3x3-comparison} to preserve the clarity of this figure) also lie within the range given above.
+The scale of structures developing in the $1.5\times1.5$ and $2.5\times2.5$ realizations (not included in Fig.
+\ref{fig:eflux-1x1-2x2-3x3-comparison} and Fig.~\ref{fig:wexb-1x1-2x2-3x3-comparison} to preserve the clarity of this figure) also lie within the range given above.
 
@@
 \includegraphicsHere{S6_rlt6.0/boxsize3x3/Ns16/Nvpar48/Nmu9/noise/S6_rlt6.0_boxsize3x3_Ns16_Nvpar48_Nmu9_noise_wexb_dens_ene_selection.pdf}{
-	Time averaged radial proﬁles of (top) the $\exb$ shearing rate $\wexb$ and (bottom) the negative gradient of the perpendicular and parallel energy moment $-\langle \nabla_{\!\!x} E_{\perp, \mathrm{i}}\rangle$ and  $-\langle \nabla_{\!\!x} E_{\parallel, \mathrm{i}}\rangle$, respectively, as well as the electric ﬁeld term $e n_0 \langle \nabla_{\!\!x} \phi\rangle$ of the force balance (on the right axis).
+	Time averaged radial profiles of (top) the $\exb$ shearing rate $\wexb$ and (bottom) the negative gradient of the perpendicular and parallel energy moment $-\langle \nabla_{\!\!x} E_{\perp, \mathrm{i}}\rangle$ and  $-\langle \nabla_{\!\!x} E_{\parallel, \mathrm{i}}\rangle$, respectively, as well as the electric field term $e n_0 \langle \nabla_{\!\!x} \phi\rangle$ of the force balance (on the right axis).
 Shown is a $3\times3$ realization with $\rlt = 6.0$.
@@
 As in the isotropic scan the turbulence subdued and a fully developed staircase pattern forms after $\sim 2000\,R/\vth$ [Fig.~\ref{fig:eflux-3x1.5-2.5-3-5-comparison}(a)].
-The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-3x1.5-2.5-3-5-comparison} or Fig.~\ref{fig:wexb-stable-comparison}(c) and confirms again a size of a typical mesoscale.
-Fig.~\ref{fig:wexb-stable-comparison}(c) also confirms that indeed a competition between patterns with two sizes $\lambda \in [57,~ 76]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
-The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-stable-comparison}(c) in the $3\times 2.5$ realization.
+The convergence of staircase pattern can be seen in Fig.~\ref{fig:wexb-3x1.5-2.5-3-5-comparison} and confirms again a size of a typical mesoscale.
+Fig.~\ref{fig:wexb-3x1.5-2.5-3-5-comparison} also confirms that indeed a competition between patterns with two sizes $\lambda \in [57,~ 76]\,\rhoth$ causing the different results for $3 \times 1$ and $3\times 3$.
+The zonal flow mode number varies between $\nzf = 3,4$ which can be seen in Fig.~\ref{fig:wexb-3x1.5-2.5-3-5-comparison} in the $3\times 2.5$ realization.
 The staircase structure has a pattern between three and four repetitions which get represented in the second repetition with no significant plateau at positive shear.
-Instead the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
+Instead, the pattern returns immediately after reaching the maximum shear ($+ \gamma$) to the minimum shear ($- \gamma$) of the third repetition in a steep flank.
 The Fourier analysis of this case yields no definitely basic mode rather two dominating modes with $\nzf = 3, 4$ with a fraction of the maximum amplitude $\hatwexbamp$ each [Fig.

### 2023-06-30 presentation/presentation.tex
#39: Minor corrections and thermal quantities in presentation

@@
 				$F_{\parallel,\nabla_{\!\parallel} B} = - \underbrace{\frac{mv^2_{\perp}}{2B}}_{\mu} \nabla_{\!\parallel} B$ \\
-				\input{../pictures/Theory/Lorentz-Force.tex} & 
+				\input{../pictures/Theory/Lorentz-Force_Presentation.tex} & 
 				\input{../pictures/Theory/Electric-Force.tex} & 
@@
 				$\Rightarrow$ Mesoscale pattern size of: \\[0.3cm]
-				$\boxed{\sim (57 - 76)\,\rhoth}$ \bigskip
+				$\boxed{\sim 57 - 76\,\rhoth}$ \bigskip
 
@@
 				\item <4-> Parallel velocity $\Nvpar$ could be reduced from 64 to 48, which halfed the time until suppression of turbulence
-				\item <5-> Mesoscale pattern size of $\sim (57 - 76)\,\rhoth$ is found to be intrinsic to ITG-driven turbulence for Cyclone Base Case parameters
+				\item <5-> Mesoscale pattern size of $\sim 57 - 76\,\rhoth$ is found to be intrinsic to ITG-driven turbulence for Cyclone Base Case parameters
 			\end{itemize}
